### 2024-07-16 _tex/index.tex
Built site for gh-pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{agujournal2019}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={La Palma Earthquakes},
  pdfauthor={Steve Purves; Rowan Cockett},
  pdfkeywords={La Palma, Earthquakes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{Earth and Space Science}

\draftfalse

\begin{document}
\title{La Palma Earthquakes}

\authors{Steve Purves\affil{1}, Rowan Cockett\affil{1}}
\affiliation{1}{Curvenote, }
\correspondingauthor{Steve Purves}{steve@curvenote.com}

\begin{abstract}
In September 2021, a significant jump in seismic activity on the island
of La Palma (Canary Islands, Spain) signaled the start of a volcanic
crisis that still continues at the time of writing. Earthquake data is
continually collected and published by the Instituto Geográphico
Nacional (IGN). \ldots{}
\end{abstract}

\section*{Plain Language Summary}
Earthquake data for the island of La Palma from the September 2021
eruption is found \ldots{}

\section{Introduction}\label{introduction}

\textsubscript{Source:
\href{https://jriveraespejo.github.io/papar2_manuscript/index.qmd.html}{Article
Notebook}}

\phantomsection\label{cell-fig-timeline}
\begin{figure}[H]

\centering{

\includegraphics{index_files/figure-pdf/fig-timeline-1.pdf}

}

\caption{\label{fig-timeline}Timeline of recent earthquakes on La Palma}

\end{figure}%

\textsubscript{Source:
\href{https://jriveraespejo.github.io/papar2_manuscript/index.qmd.html}{Article
Notebook}}

\textsubscript{Source:
\href{https://jriveraespejo.github.io/papar2_manuscript/index.qmd.html}{Article
Notebook}}

Based on data up to and including 1971, eruptions on La Palma happen
every 79.8 years on average.

Studies of the magma systems feeding the volcano, such as Marrero et al.
(2019), have proposed that there are two main magma reservoirs feeding
the Cumbre Vieja volcano; one in the mantle (30-40km depth) which
charges and in turn feeds a shallower crustal reservoir (10-20km depth).

Eight eruptions have been recorded since the late 1400s
(Figure~\ref{fig-timeline}).

Data and methods are discussed in Section~\ref{sec-data-methods}.

Let \(x\) denote the number of eruptions in a year. Then, \(x\) can be
modeled by a Poisson distribution

\begin{equation}\phantomsection\label{eq-poisson}{
p(x) = \frac{e^{-\lambda} \lambda^{x}}{x !}
}\end{equation}

where \(\lambda\) is the rate of eruptions per year. Using
Equation~\ref{eq-poisson}, the probability of an eruption in the next
\(t\) years can be calculated.

\begin{longtable}[]{@{}ll@{}}
\caption{Recent historic eruptions on La
Palma}\label{tbl-history}\tabularnewline
\toprule\noalign{}
Name & Year \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Name & Year \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Current & 2021 \\
Teneguía & 1971 \\
Nambroque & 1949 \\
El Charco & 1712 \\
Volcán San Antonio & 1677 \\
Volcán San Martin & 1646 \\
Tajuya near El Paso & 1585 \\
Montaña Quemada & 1492 \\
\end{longtable}

Table~\ref{tbl-history} summarises the eruptions recorded since the
colonization of the islands by Europeans in the late 1400s.

\begin{figure}

\centering{

\includegraphics{images/la-palma-map.png}

}

\caption{\label{fig-map}Map of La Palma}

\end{figure}%

La Palma is one of the west most islands in the Volcanic Archipelago of
the Canary Islands (Figure~\ref{fig-map}).

\begin{figure}[H]

\centering{

\includegraphics{index_files/figure-latex/notebooks-explore-earthquakes-fig-spatial-plot-output-1.png}

}

\caption{\label{fig-spatial-plot}Locations of earthquakes on La Palma
since 2017}

\end{figure}%

\textsubscript{Source:
\href{https://jriveraespejo.github.io/papar2_manuscript/notebooks/explore-earthquakes-preview.html\#cell-fig-spatial-plot}{Explore
Earthquakes}}

Figure~\ref{fig-spatial-plot} shows the location of recent Earthquakes
on La Palma.

\section{Data \& Methods}\label{sec-data-methods}

\section{Conclusion}\label{conclusion}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-marrero2019}
Marrero, J., García, A., Berrocoso, M., Llinares, Á., Rodríguez-Losada,
A., \& Ortiz, R. (2019). Strategies for the development of volcanic
hazard maps in monogenetic volcanic fields: The example of {La} {Palma}
({Canary} {Islands}). \emph{Journal of Applied Volcanology}, \emph{8}.
\url{https://doi.org/10.1186/s13617-019-0085-5}

\end{CSLReferences}

\end{document}

### 2024-07-16 images/tex/SciMod.tex
section:structure, process:write

\documentclass[tikz, border=1mm]{standalone}
\usepackage{tikz} 
\usetikzlibrary{arrows.meta}
\usepackage{pgfplots}

\begin{document}

\begin{tikzpicture}
    % nodes
    \node at (-3,1) {$X_{i}$};
    \node at (-3,0) {$e_{i}$};
    \node at (-1.5,3) {$Z_{k}$};
    \node at (-1.5,2) {$e_{k}$};
    \node at (-1.5,1) {$Y_{r}$};
    \node at (-1.5,0) {$\theta_{i}$};
    \node at (-1.5,-1) {$e_{r_{i}}$};
    \node at (0,2) {$\gamma_{k}$};
    \node at (0,0.1) {$\theta_{ir_{i}}$};
	\node at (1.6,1) {$\delta_{kijr_{ij}}$};
    \node at (3,1) {$O_{kijr_{ij}}$};

	% paths
    % paths
    \draw[{Circle}-{latex}](-3,0.7) to (-1.6,-0.28); % X_i -> theta_i
    \draw[{Circle[open]}-{latex}](-3,-0.3) to (-1.6,-0.32); % e_i -> theta_i
    \draw[{Circle}-{latex}](-1.5,2.7) to (-0.1,1.72); % Z_k -> gamma_k
    \draw[{Circle[open]}-{latex}](-1.5,1.7) to (-0.1,1.68); % e_k -> gamma_k
    \draw[{Circle}-{latex}](-1.5,0.7) to (-0.1,-0.28); % Y_r -> theta_ir
    \draw[{Circle[open]}-{latex}](-1.5,-0.3) to (-0.1,-0.3); % theta_i -> theta_ir
    \draw[{Circle[open]}-{latex}](-1.5,-1.3) to (-0.1,-0.32); % e_ir -> theta_ir
    \draw[{Circle[open]}-{latex}](0,1.7) to (1.4,0.7); % gamma_k -> delta_kijr
    \draw[{Circle[open]}-{latex}](0,-0.3) to (1.4,0.7); % theta_ir -> delta_kijr
    \draw[{Circle[open]}-{latex}{Circle}](1.5,0.7) to (3,0.7); % delta_kijr -> O_kijr
\end{tikzpicture}

\end{document}

### 2024-07-16 index.qmd
section:structure, process:write

---
title: "Causes and effects in Dichotomous Comparative Judgments: a plausible information-theoretical system of mechanism"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    roles: [conceptualization, methodology, data curation, analysis, writing, software, investigation, validation, visualization]
    affiliation:
      - name: University of Antwerp
        department: Education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    roles: [conceptualization, methodology, editing, supervision, project administration, funding, investigation, resources]
    affiliation:
      - name: University of Antwerp
        department: Education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    roles: [conceptualization, methodology, editing, supervision, project administration, funding, investigation, resources]
    affiliation:
      - name: University of Antwerp
        department: Education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    roles: [conceptualization, methodology, editing, supervision, project administration, funding, investigation, resources]
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - comparative judgement
  - directed acycilc graph
  - causal analysis
  - probabilistic statistics
abstract: |
  Dichotomous Comparative Judgment (DCJ, @Pollitt_2012a, @Pollitt_2012b) requires judges to evaluate the relative manifestation of traits between pairs of stimuli, resulting in a dichotomous outcome indicating which stimulus exhibits the trait more strongly. Research has demonstrated DCJ's effectiveness and reliability in various domains [@Pollitt_2012b; @Bartholomew_et_al_2018; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Boonen_et_al_2020]. However, the literature lacks a clear and transparent depiction of the plausible mechanisms underlying DCJ data. Specifically, there is no detail explanation of how the different assessment factors can potentially influence the observed DCJ data. This study aims to fill this gap by applying the framework of causal analysis and Directed Acyclic Graphs (DAG; @Pearl_2009). Using this framework, the study will construct a scientific model to elucidate the causal assumptions and mechanisms inherent the system. This model will enable researchers to draw inferences about causal relationships from DCJ data. Subsequently, the study will translate this model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference. The outcomes of this study will inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

## Introduction {#sec-introduction}

## Theoretical framework {#sec-TheoreticalFramework}

### Research questions and their estimands {#sec-ResearchQuestions}

### A scientific model for the DCJ procedure {#sec-SciMod}

### From the scientific to the Bradley-Terry-Luce model {#sec-Sci2StatMod}

## Discussion {#sec-discussion}

### Limitations and further research {#sec-LFR}

## Conclusion {#sec-conclusion}

{{< pagebreak >}}

## Declarations {.appendix .unnumbered}

**Funding:** The project was founded through the Research Fund of the University of Antwerp (BOF).

**Conflict of interests:** The authors declare no conflict of interest.

**Ethics approval:** The University of Antwerp Research Ethics Committee has confirmed that no ethical approval is required.

**Consent to participate:** Not applicable

**Consent for publication:** All authors have read and agreed to the published version of the manuscript.

**Availability of data and materials:** No data was utilized in this study

**Code availability:** All the code utilized in this research is available in the different notebooks and `CODE LINKS` referenced in the digital document. The digital document is located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).

**Authors' contributions:** *Conceptualization:* S.G., S.dM., and J.M.R.E; *Data curation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Funding acquisition:* S.G. and S.dM; *Investigation:* S.G.; *Methodology:* S.G., S.dM., and J.M.R.E; *Project administration:* S.G. and S.dM.; *Resources:* S.G. and S.dM.; *Software: J.M.R.E.*; *Supervision:* S.G. and S.dM.; *Validation:* J.M.R.E.; *Visualization:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G. and S.dM.

<!-- **Acknowledgements:** NO additional acknowledgements -->

{{< pagebreak >}}

## References {.appendix .unnumbered}

:::{#refs}

:::

### 2024-07-16 _tex/index.tex
Built site for gh-pages

@@
 \documentclass[
-]{agujournal2019}
+  authoryear,
+  preprint,
+  1p]{elsarticle}
 
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\usepackage{url} %this package should fix any errors with URLs in refs.
-\usepackage{lineno}
-\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
-\usepackage{soul}
-\linenumbers
 \makeatletter
@@
 \makeatother
+\journal{Psychometrika}
 
@@
 \fi
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-harv}
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={La Palma Earthquakes},
-  pdfauthor={Steve Purves; Rowan Cockett},
-  pdfkeywords={La Palma, Earthquakes},
+  pdftitle={Causes and effects in Dichotomous Comparative Judgments: a plausible information-theoretical system of mechanism},
+  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
+  pdfkeywords={comparative judgement, directed acycilc graph, causal
+analysis, probabilistic statistics},
   colorlinks=true,
@@
 
-\journalname{Earth and Space Science}
+\setlength{\parindent}{6pt}
+\begin{document}
 
-\draftfalse
+\begin{frontmatter}
+\title{Causes and effects in Dichotomous Comparative Judgments: a
+plausible information-theoretical system of mechanism}
+\author[1]{Jose Manuel Rivera Espejo%
+\corref{cor1}%
+}
+ \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
+\author[1]{Tine van Daal%
+%
+}
+ \ead{tine.vandaal@uantwerpen.be} 
+\author[1]{Sven De Maeyer%
+%
+}
+ \ead{sven.demaeyer@uantwerpen.be} 
+\author[2]{Steven Gillis%
+%
+}
+ \ead{steven.gillis@uantwerpen.be}
 
-\begin{document}
-\title{La Palma Earthquakes}
+\affiliation[1]{organization={University of Antwerp, Education
+sciences},,postcodesep={}}
+\affiliation[2]{organization={University of
+Antwerp, Linguistics},,postcodesep={}}
 
-\authors{Steve Purves\affil{1}, Rowan Cockett\affil{1}}
-\affiliation{1}{Curvenote, }
-\correspondingauthor{Steve Purves}{steve@curvenote.com}
+\cortext[cor1]{Corresponding author}
 
 \begin{abstract}
-In September 2021, a significant jump in seismic activity on the island
-of La Palma (Canary Islands, Spain) signaled the start of a volcanic
-crisis that still continues at the time of writing. Earthquake data is
-continually collected and published by the Instituto Geográphico
-Nacional (IGN). \ldots{}
+Dichotomous Comparative Judgment (DCJ, \citet{Pollitt_2012a},
+\citet{Pollitt_2012b}) requires judges to evaluate the relative
+manifestation of traits between pairs of stimuli, resulting in a
+dichotomous outcome indicating which stimulus exhibits the trait more
+strongly. Research has demonstrated DCJ's effectiveness and reliability
+in various domains
+\citep{Pollitt_2012b, Bartholomew_et_al_2018, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2020, Boonen_et_al_2020}.
+However, the literature lacks a clear and transparent depiction of the
+plausible mechanisms underlying DCJ data. Specifically, there is no
+detail explanation of how the different assessment factors can
+potentially influence the observed DCJ data. This study aims to fill
+this gap by applying the framework of causal analysis and Directed
+Acyclic Graphs (DAG; \citet{Pearl_2009}). Using this framework, the
+study will construct a scientific model to elucidate the causal
+assumptions and mechanisms inherent the system. This model will enable
+researchers to draw inferences about causal relationships from DCJ data.
+Subsequently, the study will translate this model into a probabilistic
+statistical model, aiming to derive statistical estimands for different
+targets of inference. The outcomes of this study will inform the
+planning of DCJ experiments and hold significance for researchers or
+analysts involved in education and assessment procedures who implement
+the DCJ methodology.
 \end{abstract}
 
-\section*{Plain Language Summary}
-Earthquake data for the island of La Palma from the September 2021
-eruption is found \ldots{}
+\begin{keyword}
+    comparative judgement \sep directed acycilc graph \sep causal
+analysis \sep 
+    probabilistic statistics
+\end{keyword}
+\end{frontmatter}
 
-\section{Introduction}\label{introduction}
+\section{Introduction}\label{sec-introduction}
 
-\textsubscript{Source:
-\href{https://jriveraespejo.github.io/papar2_manuscript/index.qmd.html}{Article
-Notebook}}
+\section{Theoretical framework}\label{sec-TheoreticalFramework}
 
-\phantomsection\label{cell-fig-timeline}
-\begin{figure}[H]
+\subsection{Research questions and their
+estimands}\label{sec-ResearchQuestions}
 
-\centering{
+\subsection{A scientific model for the DCJ procedure}\label{sec-SciMod}
 
-\includegraphics{index_files/figure-pdf/fig-timeline-1.pdf}
+\subsection{From the scientific to the Bradley-Terry-Luce
+model}\label{sec-Sci2StatMod}
 
-}
+\section{Discussion}\label{sec-discussion}
 
-\caption{\label{fig-timeline}Timeline of recent earthquakes on La Palma}
+\subsection{Limitations and further research}\label{sec-LFR}
 
-\end{figure}%
+\section{Conclusion}\label{sec-conclusion}
 
-\textsubscript{Source:
-\href{https://jriveraespejo.github.io/papar2_manuscript/index.qmd.html}{Article
-Notebook}}
+\newpage{}
 
-\textsubscript{Source:
-\href{https://jriveraespejo.github.io/papar2_manuscript/index.qmd.html}{Article
-Notebook}}
+\section*{Declarations}\label{declarations}
+\addcontentsline{toc}{section}{Declarations}
 
-Based on data up to and including 1971, eruptions on La Palma happen
-every 79.8 years on average.
+\textbf{Funding:} The project was founded through the Research Fund of
+the University of Antwerp (BOF).
 
-Studies of the magma systems feeding the volcano, such as Marrero et al.
-(2019), have proposed that there are two main magma reservoirs feeding
-the Cumbre Vieja volcano; one in the mantle (30-40km depth) which
-charges and in turn feeds a shallower crustal reservoir (10-20km depth).
+\textbf{Conflict of interests:} The authors declare no conflict of
+interest.
 
-Eight eruptions have been recorded since the late 1400s
-(Figure~\ref{fig-timeline}).
+\textbf{Ethics approval:} The University of Antwerp Research Ethics
+Committee has confirmed that no ethical approval is required.
 
-Data and methods are discussed in Section~\ref{sec-data-methods}.
+\textbf{Consent to participate:} Not applicable
 
-Let \(x\) denote the number of eruptions in a year. Then, \(x\) can be
-modeled by a Poisson distribution
+\textbf{Consent for publication:} All authors have read and agreed to
+the published version of the manuscript.
 
-\begin{equation}\phantomsection\label{eq-poisson}{
-p(x) = \frac{e^{-\lambda} \lambda^{x}}{x !}
-}\end{equation}
+\textbf{Availability of data and materials:} No data was utilized in
+this study
 
-where \(\lambda\) is the rate of eruptions per year. Using
-Equation~\ref{eq-poisson}, the probability of an eruption in the next
-\(t\) years can be calculated.
+\textbf{Code availability:} All the code utilized in this research is
+available in the different notebooks and \texttt{CODE\ LINKS} referenced
+in the digital document. The digital document is located at:
+\url{https://jriveraespejo.github.io/paper2_manuscript/}.
 
-\begin{longtable}[]{@{}ll@{}}
-\caption{Recent historic eruptions on La
-Palma}\label{tbl-history}\tabularnewline
-\toprule\noalign{}
-Name & Year \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-Name & Year \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Current & 2021 \\
-Teneguía & 1971 \\
-Nambroque & 1949 \\
-El Charco & 1712 \\
-Volcán San Antonio & 1677 \\
-Volcán San Martin & 1646 \\
-Tajuya near El Paso & 1585 \\
-Montaña Quemada & 1492 \\
-\end{longtable}
+\textbf{Authors' contributions:} \emph{Conceptualization:} S.G., S.dM.,
+and J.M.R.E; \emph{Data curation:} J.M.R.E.; \emph{Formal Analysis:}
+J.M.R.E.; \emph{Funding acquisition:} S.G. and S.dM;
+\emph{Investigation:} S.G.; \emph{Methodology:} S.G., S.dM., and
+J.M.R.E; \emph{Project administration:} S.G. and S.dM.;
+\emph{Resources:} S.G. and S.dM.; \emph{Software: J.M.R.E.};
+\emph{Supervision:} S.G. and S.dM.; \emph{Validation:} J.M.R.E.;
+\emph{Visualization:} J.M.R.E.; \emph{Writing - original draft:}
+J.M.R.E.; \emph{Writing - review \& editing:} S.G. and S.dM.
 
-Table~\ref{tbl-history} summarises the eruptions recorded since the
-colonization of the islands by Europeans in the late 1400s.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/la-palma-map.png}
-
-}
-
-\caption{\label{fig-map}Map of La Palma}
-
-\end{figure}%
-
-La Palma is one of the west most islands in the Volcanic Archipelago of
-the Canary Islands (Figure~\ref{fig-map}).
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics{index_files/figure-latex/notebooks-explore-earthquakes-fig-spatial-plot-output-1.png}
-
-}
-
-\caption{\label{fig-spatial-plot}Locations of earthquakes on La Palma
-since 2017}
-
-\end{figure}%
-
-\textsubscript{Source:
-\href{https://jriveraespejo.github.io/papar2_manuscript/notebooks/explore-earthquakes-preview.html\#cell-fig-spatial-plot}{Explore
-Earthquakes}}
-
-Figure~\ref{fig-spatial-plot} shows the location of recent Earthquakes
-on La Palma.
-
-\section{Data \& Methods}\label{sec-data-methods}
-
-\section{Conclusion}\label{conclusion}
+\newpage{}
 
@@
 
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\vspace{1em}
-
-\bibitem[\citeproctext]{ref-marrero2019}
-Marrero, J., García, A., Berrocoso, M., Llinares, Á., Rodríguez-Losada,
-A., \& Ortiz, R. (2019). Strategies for the development of volcanic
-hazard maps in monogenetic volcanic fields: The example of {La} {Palma}
-({Canary} {Islands}). \emph{Journal of Applied Volcanology}, \emph{8}.
-\url{https://doi.org/10.1186/s13617-019-0085-5}
-
-\end{CSLReferences}
+\renewcommand{\bibsection}{}
+\bibliography{references.bib}
 

### 2024-07-17 index.qmd
section:introduction; process:write

@@
 ---
-title: "Causes and effects in Dichotomous Comparative Judgments: a plausible information-theoretical system of mechanism"
+title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism"
 author:
@@
     corresponding: true
-    roles: [conceptualization, methodology, data curation, analysis, writing, software, investigation, validation, visualization]
     affiliation:
       - name: University of Antwerp
-        department: Education sciences
+        department: Training and education sciences
         group: Edubron
+    roles: 
+      - conceptualization: lead
+      - methodology: lead
+      - software: only
+      - validation: only
+      - formal analysis: only
+      - investigation: only
+      - resources: no
+      - data curation: only
+      - writing: only
+      - editing: no
+      - visualization: only
+      - supervision: no
+      - project administration: no
+      - funding adquisition: no
   - name: 
@@
     corresponding: false
-    roles: [conceptualization, methodology, editing, supervision, project administration, funding, investigation, resources]
     affiliation:
       - name: University of Antwerp
-        department: Education sciences
+        department: Training and education sciences
         group: Edubron
+    roles: 
+      - conceptualization: support
+      - methodology: support
+      - software: no
+      - validation: no
+      - formal analysis: no
+      - investigation: no
+      - resources: support
+      - data curation: no
+      - writing: no
+      - editing: support
+      - visualization: no
+      - supervision: no
+      - project administration: no
+      - funding adquisition: no
   - name: 
@@
     corresponding: false
-    roles: [conceptualization, methodology, editing, supervision, project administration, funding, investigation, resources]
     affiliation:
       - name: University of Antwerp
-        department: Education sciences
+        department: Training and education sciences
         group: Edubron
+    roles: 
+      - conceptualization: support
+      - methodology: support
+      - software: no
+      - validation: no
+      - formal analysis: no
+      - investigation: no
+      - resources: support
+      - data curation: no
+      - writing: no
+      - editing: support
+      - visualization: no
+      - supervision: yes
+      - project administration: yes
+      - funding adquisition: yes
   - name: 
@@
     corresponding: false
-    roles: [conceptualization, methodology, editing, supervision, project administration, funding, investigation, resources]
     affiliation:
@@
         group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
+    roles: 
+      - conceptualization: support
+      - methodology: no
+      - software: no
+      - validation: no
+      - formal analysis: no
+      - investigation: no
+      - resources: support
+      - data curation: no
+      - writing: no
+      - editing: support
+      - visualization: no
+      - supervision: yes
+      - project administration: yes
+      - funding adquisition: yes
 funding: 
@@
 bibliography: references.bib
+notebook-links: global
 citation: true
@@
 ## Introduction {#sec-introduction}
+
+<!-- 1. What is comparative judgment? -->
+In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *Dichotomous* Comparative Judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b].
+In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli.
+This assessment results in a dichotomous value indicating which stimulus possesses a higher degree of the trait.
+After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced.
+This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959; @Pollitt_2012a], which produces a latent variable corresponding to the trait of interest.
+This latent variable is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.
+
+<!-- 2. in what has been applied on?, and where is the gap? -->
+DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by the works of @Pollitt_2012b, @Bartholomew_et_al_2018, @vanDaal_et_al_2016, @Lesterhuis_2018, and @Bartholomew_et_al_2020.
+However, its application transcends education, as exemplified by the work of @Boonen_et_al_2020.
+The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992].
+Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020].
+Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that give rise to DCJ data.
+Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data.
+
+<!-- 3. What are the factors behind DCJ? -->
+According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and contribute to the reliability of the DCJ method.
+These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used.
+Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles.
+Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome.
+An example of this can be found in @vanDaal_et_al_2016, where the authors assessed the *skills in academic writing* (trait) of Flemish university students, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).
 
@@
 
-**Code availability:** All the code utilized in this research is available in the different notebooks and `CODE LINKS` referenced in the digital document.
-The digital document is located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).
+**Code availability:** All the code utilized in this research is available in the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).
 
-**Authors' contributions:** *Conceptualization:* S.G., S.dM., and J.M.R.E; *Data curation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Funding acquisition:* S.G.
-and S.dM; *Investigation:* S.G.; *Methodology:* S.G., S.dM., and J.M.R.E; *Project administration:* S.G.
-and S.dM.; *Resources:* S.G.
-and S.dM.; *Software: J.M.R.E.*; *Supervision:* S.G.
-and S.dM.; *Validation:* J.M.R.E.; *Visualization:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G.
-and S.dM.
+**Authors' contributions:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
+and S.DM.; *Project administration:* S.G.
+and S.DM.; *Funding acquisition:* S.G.
+and S.DM.
 

### 2024-07-17 _tex/index.tex
Built site for gh-pages

@@
 \hypersetup{
-  pdftitle={Causes and effects in Dichotomous Comparative Judgments: a plausible information-theoretical system of mechanism},
+  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism},
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
@@
 \begin{frontmatter}
-\title{Causes and effects in Dichotomous Comparative Judgments: a
-plausible information-theoretical system of mechanism}
+\title{Causes and effects in Dichotomous Comparative Judgments: an
+information-theoretical system with plausible mechanism}
 \author[1]{Jose Manuel Rivera Espejo%
@@
 
-\affiliation[1]{organization={University of Antwerp, Education
-sciences},,postcodesep={}}
+\affiliation[1]{organization={University of Antwerp, Training and
+education sciences},,postcodesep={}}
 \affiliation[2]{organization={University of
@@
 \section{Introduction}\label{sec-introduction}
+
+In contemporary contexts, Thurstone's law of comparative judgment
+\citeyearpar{Thurstone_1927} primarily refers to the method of
+\emph{Dichotomous} Comparative Judgment
+\citep[DCJ,][]{Pollitt_2012a, Pollitt_2012b}. In DCJ, a judge assesses
+the relative manifestation of a \emph{trait} within a pair of stimuli.
+This assessment results in a dichotomous value indicating which stimulus
+possesses a higher degree of the trait. After different judges perform
+multiple rounds of pairwise comparisons, an outcome vector is produced.
+This vector is modeled using the Bradley-Terry-Luce model
+\citep[BTL,][]{Bradley_et_al_1952, Luce_1959, Pollitt_2012a}, which
+produces a latent variable corresponding to the trait of interest. This
+latent variable is then used to rank the stimuli from lowest to highest
+or to evaluate the influence of certain variables on the stimuli's
+positions in the ranking.
+
+DCJ has proven effective in assessing competencies and traits
+predominantly within the educational realm, as demonstrated by the works
+of \citet{Pollitt_2012b}, \citet{Bartholomew_et_al_2018},
+\citet{vanDaal_et_al_2016}, \citet{Lesterhuis_2018}, and
+\citet{Bartholomew_et_al_2020}. However, its application transcends
+education, as exemplified by the work of \citet{Boonen_et_al_2020}. The
+methodology has also evolved to include multiple, as opposed to pairwise
+comparisons \citep{Luce_1959, Placket_1975}, and to accommodate
+comparisons with ordinal outcomes \citep{Tutz_1986, Agresti_1992}.
+Overall, research suggests that DCJ offers an alternative and efficient
+approach to measurement and evaluation, characterized by its reliability
+and validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
+Nevertheless, despite the method's widespread use, the literature lacks
+a transparent depiction of the DCJ system and the plausible mechanisms
+that give rise to DCJ data. Particularly, there is no detailed
+explanation of how different assessment factors can potentially
+influence the observed DCJ data.
+
+According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
+several assessment factors interact and contribute to the reliability of
+the DCJ method. These factors include the number and characteristics of
+the stimuli, their \emph{proximity} in terms of the assessed trait, the
+number of comparison per stimulus, and the pairing algorithm used.
+Furthermore, since the method relies on judges' assessments, the number
+and characteristics of judges, their \emph{discrimination} abilities,
+and the number of comparisons per judge also play pivotal roles.
+Moreover, when the stimuli represent sub-units of higher-levels units,
+factors such as the number and characteristics of these units, along
+with their \emph{proximity} in terms of the assessed trait, can
+significantly influence the outcome. An example of this can be found in
+\citet{vanDaal_et_al_2016}, where the authors assessed the \emph{skills
+in academic writing} (trait) of Flemish university students, utilizing
+multiple argumentative essays (stimuli, sub-units) originating from
+various students (units).
 
@@
 \textbf{Code availability:} All the code utilized in this research is
-available in the different notebooks and \texttt{CODE\ LINKS} referenced
-in the digital document. The digital document is located at:
+available in the digital document located at:
 \url{https://jriveraespejo.github.io/paper2_manuscript/}.
 
-\textbf{Authors' contributions:} \emph{Conceptualization:} S.G., S.dM.,
-and J.M.R.E; \emph{Data curation:} J.M.R.E.; \emph{Formal Analysis:}
-J.M.R.E.; \emph{Funding acquisition:} S.G. and S.dM;
-\emph{Investigation:} S.G.; \emph{Methodology:} S.G., S.dM., and
-J.M.R.E; \emph{Project administration:} S.G. and S.dM.;
-\emph{Resources:} S.G. and S.dM.; \emph{Software: J.M.R.E.};
-\emph{Supervision:} S.G. and S.dM.; \emph{Validation:} J.M.R.E.;
-\emph{Visualization:} J.M.R.E.; \emph{Writing - original draft:}
-J.M.R.E.; \emph{Writing - review \& editing:} S.G. and S.dM.
+\textbf{Authors' contributions:} \emph{Conceptualization:} S.G., S.DM.,
+T.vD., and J.M.R.E; \emph{Methodology:} S.DM., T.vD., and J.M.R.E;
+\emph{Software:} J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal
+Analysis:} J.M.R.E.; \emph{Investigation:} J.M.R.E; \emph{Resources:}
+S.G., S.DM., and T.vD.; \emph{Data curation:} J.M.R.E.; \emph{Writing -
+original draft:} J.M.R.E.; \emph{Writing - review \& editing:} S.G.,
+S.DM., and T.vD.; \emph{Visualization:} J.M.R.E.; \emph{Supervision:}
+S.G. and S.DM.; \emph{Project administration:} S.G. and S.DM.;
+\emph{Funding acquisition:} S.G. and S.DM.
 

### 2024-07-18 index.qmd
section:introduction; process:write

@@
 abstract: |
-  Dichotomous Comparative Judgment (DCJ, @Pollitt_2012a, @Pollitt_2012b) requires judges to evaluate the relative manifestation of traits between pairs of stimuli, resulting in a dichotomous outcome indicating which stimulus exhibits the trait more strongly.
+  Dichotomous Comparative Judgment [DCJ, @Pollitt_2012a, @Pollitt_2012b] requires judges to evaluate the relative manifestation of traits between pairs of stimuli, resulting in a dichotomous outcome indicating which stimulus exhibits the trait more strongly.
 Research has demonstrated DCJ's effectiveness and reliability in various domains [@Pollitt_2012b; @Bartholomew_et_al_2018; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
-However, the literature lacks a clear and transparent depiction of the plausible mechanisms underlying DCJ data.
-Specifically, there is no detail explanation of how the different assessment factors can potentially influence the observed DCJ data.
-This study aims to fill this gap by applying the framework of causal analysis and Directed Acyclic Graphs (DAG; @Pearl_2009).
+Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that generate the DCJ data.
+Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data.
+This study aims to fill this gap by applying the framework of causal analysis and Directed Acyclic Graphs [DAG; @Pearl_2009].
 Using this framework, the study will construct a scientific model to elucidate the causal assumptions and mechanisms inherent the system.
@@
 After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced.
-This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959; @Pollitt_2012a], which produces a latent variable corresponding to the trait of interest.
-This latent variable is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.
+This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959], which creates a score that corresponds with the trait of interest.
+This score is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.
 
 <!-- 2. in what has been applied on?, and where is the gap? -->
-DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by the works of @Pollitt_2012b, @Bartholomew_et_al_2018, @vanDaal_et_al_2016, @Lesterhuis_2018, and @Bartholomew_et_al_2020.
-However, its application transcends education, as exemplified by the work of @Boonen_et_al_2020.
+DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020.
+However, its application transcends education, as exemplified by @Boonen_et_al_2020.
 The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992].
 Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020].
-Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that give rise to DCJ data.
-Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data.
+Nevertheless, despite the method's widespread use, there is no clear representation in the literature of the plausible mechanisms that generate DCJ data.
+Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.
 
 <!-- 3. What are the factors behind DCJ? -->
-According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and contribute to the reliability of the DCJ method.
+According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the DCJ outcome.
 These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used.
@@
 Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome.
-An example of this can be found in @vanDaal_et_al_2016, where the authors assessed the *skills in academic writing* (trait) of Flemish university students, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).
+For instance, @vanDaal_et_al_2016 assessed university students' skills in academic writing, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).
+
+<!-- 4. Solution to the gap -->
+Although several studies have examined the individual impact of these factors on the method's reliability, including @Bramley_2015, @Pollitt_2012b, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, @vanDaal_et_al_2017, and @Gijsen_et_al_2021, to the best of the authors' knowledge, none have provided such transparent depiction of DCJ system and the plausible mechanisms that generate the DCJ outcome.
+This study aims to fill this gap by utilizing the framework of causal analysis and Directed Acyclic Graphs [DAG; @Pearl_2009; @Pearl_et_al_2016].
+Using this framework, the study will construct a scientific model to elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms of how the DCJ data is generated.
+This model will enable researchers to draw inferences about plausible causal relationships within the DCJ system.
+Furthermore, using a minimal set of assumptions from the framework, the study will translate the scientific model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference.
+Ultimately, the results of this study could inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.
 

### 2024-07-18 _tex/index.tex
Built site for gh-pages

@@
 \begin{abstract}
-Dichotomous Comparative Judgment (DCJ, \citet{Pollitt_2012a},
-\citet{Pollitt_2012b}) requires judges to evaluate the relative
-manifestation of traits between pairs of stimuli, resulting in a
-dichotomous outcome indicating which stimulus exhibits the trait more
-strongly. Research has demonstrated DCJ's effectiveness and reliability
-in various domains
+Dichotomous Comparative Judgment
+\citep[DCJ,][@Pollitt\_2012b]{Pollitt_2012a} requires judges to evaluate
+the relative manifestation of traits between pairs of stimuli, resulting
+in a dichotomous outcome indicating which stimulus exhibits the trait
+more strongly. Research has demonstrated DCJ's effectiveness and
+reliability in various domains
 \citep{Pollitt_2012b, Bartholomew_et_al_2018, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2020, Boonen_et_al_2020}.
-However, the literature lacks a clear and transparent depiction of the
-plausible mechanisms underlying DCJ data. Specifically, there is no
-detail explanation of how the different assessment factors can
-potentially influence the observed DCJ data. This study aims to fill
-this gap by applying the framework of causal analysis and Directed
-Acyclic Graphs (DAG; \citet{Pearl_2009}). Using this framework, the
-study will construct a scientific model to elucidate the causal
-assumptions and mechanisms inherent the system. This model will enable
-researchers to draw inferences about causal relationships from DCJ data.
-Subsequently, the study will translate this model into a probabilistic
-statistical model, aiming to derive statistical estimands for different
-targets of inference. The outcomes of this study will inform the
-planning of DCJ experiments and hold significance for researchers or
-analysts involved in education and assessment procedures who implement
-the DCJ methodology.
+Nevertheless, despite the method's widespread use, the literature lacks
+a transparent depiction of the DCJ system and the plausible mechanisms
+that generate the DCJ data. Particularly, there is no detailed
+explanation of how different assessment factors can potentially
+influence the observed DCJ data. This study aims to fill this gap by
+applying the framework of causal analysis and Directed Acyclic Graphs
+{[}DAG; \citet{Pearl_2009}{]}. Using this framework, the study will
+construct a scientific model to elucidate the causal assumptions and
+mechanisms inherent the system. This model will enable researchers to
+draw inferences about causal relationships from DCJ data. Subsequently,
+the study will translate this model into a probabilistic statistical
+model, aiming to derive statistical estimands for different targets of
+inference. The outcomes of this study will inform the planning of DCJ
+experiments and hold significance for researchers or analysts involved
+in education and assessment procedures who implement the DCJ
+methodology.
 \end{abstract}
@@
 This vector is modeled using the Bradley-Terry-Luce model
-\citep[BTL,][]{Bradley_et_al_1952, Luce_1959, Pollitt_2012a}, which
-produces a latent variable corresponding to the trait of interest. This
-latent variable is then used to rank the stimuli from lowest to highest
-or to evaluate the influence of certain variables on the stimuli's
-positions in the ranking.
+\citep[BTL,][]{Bradley_et_al_1952, Luce_1959}, which creates a score
+that corresponds with the trait of interest. This score is then used to
+rank the stimuli from lowest to highest or to evaluate the influence of
+certain variables on the stimuli's positions in the ranking.
 
 DCJ has proven effective in assessing competencies and traits
-predominantly within the educational realm, as demonstrated by the works
-of \citet{Pollitt_2012b}, \citet{Bartholomew_et_al_2018},
-\citet{vanDaal_et_al_2016}, \citet{Lesterhuis_2018}, and
-\citet{Bartholomew_et_al_2020}. However, its application transcends
-education, as exemplified by the work of \citet{Boonen_et_al_2020}. The
-methodology has also evolved to include multiple, as opposed to pairwise
-comparisons \citep{Luce_1959, Placket_1975}, and to accommodate
-comparisons with ordinal outcomes \citep{Tutz_1986, Agresti_1992}.
-Overall, research suggests that DCJ offers an alternative and efficient
-approach to measurement and evaluation, characterized by its reliability
-and validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
-Nevertheless, despite the method's widespread use, the literature lacks
-a transparent depiction of the DCJ system and the plausible mechanisms
-that give rise to DCJ data. Particularly, there is no detailed
-explanation of how different assessment factors can potentially
-influence the observed DCJ data.
+predominantly within the educational realm, as demonstrated by
+\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
+\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
+\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
+However, its application transcends education, as exemplified by
+\citet{Boonen_et_al_2020}. The methodology has also evolved to include
+multiple, as opposed to pairwise comparisons
+\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
+ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
+suggests that DCJ offers an alternative and efficient approach to
+measurement and evaluation, characterized by its reliability and
+validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
+Nevertheless, despite the method's widespread use, there is no clear
+representation in the literature of the plausible mechanisms that
+generate DCJ data. Particularly, there is no depiction of the complexity
+and the underlying assumptions of the DCJ system, nor how different
+assessment factors can potentially influence the observed DCJ outcome.
 
 According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
-several assessment factors interact and contribute to the reliability of
-the DCJ method. These factors include the number and characteristics of
-the stimuli, their \emph{proximity} in terms of the assessed trait, the
-number of comparison per stimulus, and the pairing algorithm used.
-Furthermore, since the method relies on judges' assessments, the number
-and characteristics of judges, their \emph{discrimination} abilities,
-and the number of comparisons per judge also play pivotal roles.
-Moreover, when the stimuli represent sub-units of higher-levels units,
-factors such as the number and characteristics of these units, along
-with their \emph{proximity} in terms of the assessed trait, can
-significantly influence the outcome. An example of this can be found in
-\citet{vanDaal_et_al_2016}, where the authors assessed the \emph{skills
-in academic writing} (trait) of Flemish university students, utilizing
-multiple argumentative essays (stimuli, sub-units) originating from
-various students (units).
+several assessment factors interact and influence the DCJ outcome. These
+factors include the number and characteristics of the stimuli, their
+\emph{proximity} in terms of the assessed trait, the number of
+comparison per stimulus, and the pairing algorithm used. Furthermore,
+since the method relies on judges' assessments, the number and
+characteristics of judges, their \emph{discrimination} abilities, and
+the number of comparisons per judge also play pivotal roles. Moreover,
+when the stimuli represent sub-units of higher-levels units, factors
+such as the number and characteristics of these units, along with their
+\emph{proximity} in terms of the assessed trait, can significantly
+influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
+university students' skills in academic writing, utilizing multiple
+argumentative essays (stimuli, sub-units) originating from various
+students (units).
+
+Although several studies have examined the individual impact of these
+factors on the method's reliability, including \citet{Bramley_2015},
+\citet{Pollitt_2012b}, \citet{Bramley_et_al_2019},
+\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022},
+\citet{vanDaal_et_al_2017}, and \citet{Gijsen_et_al_2021}, to the best
+of the authors' knowledge, none have provided such transparent depiction
+of DCJ system and the plausible mechanisms that generate the DCJ
+outcome. This study aims to fill this gap by utilizing the framework of
+causal analysis and Directed Acyclic Graphs {[}DAG; \citet{Pearl_2009};
+\citet{Pearl_et_al_2016}{]}. Using this framework, the study will
+construct a scientific model to elucidate the underlying assumptions of
+the DCJ system, providing plausible mechanisms of how the DCJ data is
+generated. This model will enable researchers to draw inferences about
+plausible causal relationships within the DCJ system. Furthermore, using
+a minimal set of assumptions from the framework, the study will
+translate the scientific model into a probabilistic statistical model,
+aiming to derive statistical estimands for different targets of
+inference. Ultimately, the results of this study could inform the
+planning of DCJ experiments and hold significance for researchers or
+analysts involved in education and assessment procedures who implement
+the DCJ methodology.
 

### 2024-07-19 index.qmd
section:introduction; process:write

@@
 
-## Introduction {#sec-introduction}
+# Introduction {#sec-introduction}
 
@@
 Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020].
-Nevertheless, despite the method's widespread use, there is no clear representation in the literature of the plausible mechanisms that generate DCJ data.
+Nevertheless, despite the method's widespread use, the literature does not offer a clear representation of the plausible mechanisms that generate the DCJ data.
 Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.
@@
 
-<!-- 4. Solution to the gap -->
-Although several studies have examined the individual impact of these factors on the method's reliability, including @Bramley_2015, @Pollitt_2012b, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, @vanDaal_et_al_2017, and @Gijsen_et_al_2021, to the best of the authors' knowledge, none have provided such transparent depiction of DCJ system and the plausible mechanisms that generate the DCJ outcome.
-This study aims to fill this gap by utilizing the framework of causal analysis and Directed Acyclic Graphs [DAG; @Pearl_2009; @Pearl_et_al_2016].
-Using this framework, the study will construct a scientific model to elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms of how the DCJ data is generated.
-This model will enable researchers to draw inferences about plausible causal relationships within the DCJ system.
-Furthermore, using a minimal set of assumptions from the framework, the study will translate the scientific model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference.
-Ultimately, the results of this study could inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.
+<!-- 4. Solution to the gap, final objective and relevance -->
+Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the plausible mechanisms generating the DCJ outcome.
+This study aims to fill this gap by representing DCJ within the causal analysis framework [@Pearl_2009; @Pearl_et_al_2016].
+Using Directed Acyclic Graphs [DAG, @Gross_et_al_2018; @Neal_2020], the study will construct a scientific model to elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome could be generated.
+Next, using a minimal set of assumptions embedded in the framework, the study will translate the scientific model into a probabilistic statistical model.
+This model derives statistical estimands to draw inferences about plausible causal relationships within the DCJ system.
+Ultimately, this research aims to extend the law of comparative judgment initially proposed by @Thurstone_1927 and provide a sound probabilistic base for the statistical analysis of DCJ data.
+Consequently, this research holds significance for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.
 
-## Theoretical framework {#sec-TheoreticalFramework}
+# Preliminaries {#sec-prelim}
 
-### Research questions and their estimands {#sec-ResearchQuestions}
+## Causal analysis and DAGs {#sec-prelim-causal}
 
-### A scientific model for the DCJ procedure {#sec-SciMod}
+## The flow of association and causation {#sec-prelim-flow}
 
-### From the scientific to the Bradley-Terry-Luce model {#sec-Sci2StatMod}
+# Theoretical framework {#sec-theory}
 
-## Discussion {#sec-discussion}
+## A scientific model for the DCJ {#sec-theory-scientific}
 
-### Limitations and further research {#sec-LFR}
+## From the scientific to statistical model {#sec-theory-statistics}
 
-## Conclusion {#sec-conclusion}
+## Where does it all fit? {#sec-theory-where}
+
+# Discussion {#sec-discuss}
+
+## Findings {#sec-discuss-finding}
+
+## Limitations and further research {#sec-discuss-limitations}
+
+# Conclusion {#sec-conclusion}
 
@@
 
-## Declarations {.appendix .unnumbered}
+# Declarations {.appendix .unnumbered}
 
@@
 
-## References {.appendix .unnumbered}
+# Appendix {#sec-appendix}
+
+## Why estimate judges appart? {#sec-appA}
+
+<!-- The same idea as interactions: when interactions are significant, the interpretations of the parameters cannot be made in isolation -->
+
+{{< pagebreak >}}
+
+## References {.unnumbered}
 

### 2024-07-19 _tex/index.tex
Built site for gh-pages

@@
 validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
-Nevertheless, despite the method's widespread use, there is no clear
-representation in the literature of the plausible mechanisms that
-generate DCJ data. Particularly, there is no depiction of the complexity
-and the underlying assumptions of the DCJ system, nor how different
-assessment factors can potentially influence the observed DCJ outcome.
+Nevertheless, despite the method's widespread use, the literature does
+not offer a clear representation of the plausible mechanisms that
+generate the DCJ data. Particularly, there is no depiction of the
+complexity and the underlying assumptions of the DCJ system, nor how
+different assessment factors can potentially influence the observed DCJ
+outcome.
 
@@
 Although several studies have examined the individual impact of these
-factors on the method's reliability, including \citet{Bramley_2015},
-\citet{Pollitt_2012b}, \citet{Bramley_et_al_2019},
-\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022},
-\citet{vanDaal_et_al_2017}, and \citet{Gijsen_et_al_2021}, to the best
-of the authors' knowledge, none have provided such transparent depiction
-of DCJ system and the plausible mechanisms that generate the DCJ
-outcome. This study aims to fill this gap by utilizing the framework of
-causal analysis and Directed Acyclic Graphs {[}DAG; \citet{Pearl_2009};
-\citet{Pearl_et_al_2016}{]}. Using this framework, the study will
-construct a scientific model to elucidate the underlying assumptions of
-the DCJ system, providing plausible mechanisms of how the DCJ data is
-generated. This model will enable researchers to draw inferences about
-plausible causal relationships within the DCJ system. Furthermore, using
-a minimal set of assumptions from the framework, the study will
-translate the scientific model into a probabilistic statistical model,
-aiming to derive statistical estimands for different targets of
-inference. Ultimately, the results of this study could inform the
-planning of DCJ experiments and hold significance for researchers or
+factors on the method's reliability
+\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
+none, to the best of the authors' knowledge, have provided a transparent
+depiction of the DCJ system and the plausible mechanisms generating the
+DCJ outcome. This study aims to fill this gap by representing DCJ within
+the causal analysis framework \citep{Pearl_2009, Pearl_et_al_2016}.
+Using Directed Acyclic Graphs
+\citep[DAG,][]{Gross_et_al_2018, Neal_2020}, the study will construct a
+scientific model to elucidate the underlying assumptions of the DCJ
+system, providing plausible mechanisms for how the DCJ outcome could be
+generated. Next, using a minimal set of assumptions embedded in the
+framework, the study will translate the scientific model into a
+probabilistic statistical model. This model derives statistical
+estimands to draw inferences about plausible causal relationships within
+the DCJ system. Ultimately, this research aims to extend the law of
+comparative judgment initially proposed by \citet{Thurstone_1927} and
+provide a sound probabilistic base for the statistical analysis of DCJ
+data. Consequently, this research holds significance for researchers and
 analysts involved in education and assessment procedures who implement
-the DCJ methodology.
+or design DCJ experiments.
 
-\section{Theoretical framework}\label{sec-TheoreticalFramework}
+\section{Preliminaries}\label{sec-prelim}
 
-\subsection{Research questions and their
-estimands}\label{sec-ResearchQuestions}
+\subsection{Causal analysis and DAGs}\label{sec-prelim-causal}
 
-\subsection{A scientific model for the DCJ procedure}\label{sec-SciMod}
+\subsection{The flow of association and
+causation}\label{sec-prelim-flow}
 
-\subsection{From the scientific to the Bradley-Terry-Luce
-model}\label{sec-Sci2StatMod}
+\section{Theoretical framework}\label{sec-theory}
 
-\section{Discussion}\label{sec-discussion}
+\subsection{A scientific model for the DCJ}\label{sec-theory-scientific}
 
-\subsection{Limitations and further research}\label{sec-LFR}
+\subsection{From the scientific to statistical
+model}\label{sec-theory-statistics}
+
+\subsection{Where does it all fit?}\label{sec-theory-where}
+
+\section{Discussion}\label{sec-discuss}
+
+\subsection{Findings}\label{sec-discuss-finding}
+
+\subsection{Limitations and further
+research}\label{sec-discuss-limitations}
 
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
+\section{Appendix}\label{sec-appendix}
+
+\subsection{Why estimate judges appart?}\label{sec-appA}
+
+\newpage{}
+
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
 

### 2024-07-19 index.qmd
section:preliminaries,theory; process:write,images

@@
 
+Causal analysis is the field of experimental design and statistics pertaining to establishing cause and effect from data.
+
+Causal inference
+
+Directed Acyclic Graph (DAG)
+
+the four components of DAGs
+
+Structural Causal Model (SCM)
+
+the four components of DAGs in SCM notation
+
 ## The flow of association and causation {#sec-prelim-flow}
+
+![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=100%}
 
@@
 
+## Where does it all fit? {#sec-theory-where}
+
+![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+
 ## A scientific model for the DCJ {#sec-theory-scientific}
+
+![ DCJ causal diagram, simplified description ](images/figures/SciModel_simp1.png){#fig-SciModel_simp1 fig-align="center" width=80%}
+
+![ DCJ causal diagram, simplified mathematical description ](images/figures/SciModel_simp2.png){#fig-SciModel_simp2 fig-align="center" width=70%}
+
+![ DCJ causal diagram, population mathematical description ](images/figures/SciModel_pop.png){#fig-SciModel_pop fig-align="center" width=70%}
+
+<!-- check the directionality of the arrow from $p_{I}$ to $\sigma_{i(g)}$ and equivalents -->
+
+![ DCJ causal diagram, sample with comparisons mathematical description ](images/figures/SciModel_samp.png){#fig-SciModel_samp fig-align="center" width=80%}
+
+## Probabilitics assumptions of the scientific model {#sec-theory-probability}
+
+<!-- what does it mean in probabilistic terms? describe probabilities and arrive to structural model -->
+
+<!-- commands for d-separation -->
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+
+$$
+\begin{aligned}
+  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
+  \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\
+  \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\
+  \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\
+  \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\
+  & \; e_{k} \dsep e_{i} \\
+  & \; e_{k} \dsep e_{r_{i}} \\
+  & \; e_{i} \dsep e_{r_{i}}
+\end{aligned}
+$$ {#eq-StructuralModel}
 
@@
 
-## Where does it all fit? {#sec-theory-where}
+$$
+\begin{aligned}
+  O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\
+  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
+  \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\
+  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
+  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
+  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
+  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
+  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
+\end{aligned}
+$$ {#eq-StatModel_general}
+
+for identification purposes we can set $\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02$, $\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1$, and $\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1$.
+A special case of this would be to assume that the data comes from the same population, in that case, $\sigma_{k(g)} = \sigma_{k} = 0.02$, $\sigma_{i(g)} = \sigma_{i} = 1$
+
+## Let's talk about Thurstone {#sec-theory-thurstone}
+
+Thurstone's comparative judgment @Thurstone_1927 is based on the formula:
+$$
+X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}}
+$$
+
+where $X_{AB}$ defines the comparative judgment outcome, $S_{A}$ and $S_{B}$ are the modal discriminal processes, $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}$, with $\sigma_{A}$ and $\sigma_{B}$ being the dispersion of discriminal processes $A$ and $B$, respectively, and $\rho$ the correlation between discriminal processes.
+
+The theory identifies five cases:
+
+- **Case 1:** only constant $\rho$ (not $\rho_{ij}$)
+- **Case 2:** $X_{ij}$ becomes $X_{kij}$ with $k=1, \dots, K$ judges (replication, not duplication)
+- **Case 3:** $\rho = 0$, then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$ 
+- **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$
+- **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$
+
+Now using the DAG and statistical notation
+
+$$
+\begin{aligned}
+  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
+  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
+  \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\
+  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
+  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
+  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
+  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
+  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
+\end{aligned}
+$$ {#eq-StatModel_thurstone}
+
+The theory identifies five cases:
+
+- **Case 1:** only constant $\rho \approx \sigma_{i}$
+- **Case 2:** now judges are separated by using $\gamma_{k}$
+- **Case 3:** $\rho \approx \sigma_{e_{i}} = 0$ (no nesting of texts on students), then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$ 
+- **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$
+- **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$
+
+But now can we see other scenarios than just those 5 cases?
+
+- consider different $\rho \approx \sum_{p=1}^{P} \sigma_{p}$, depending on $P$ nesting structures 
+- we can now investigate $\gamma_{k}$
+- we can assume $\sigma_{B} \neq \sigma_{A}$, no need for results on the limit
 
@@
 
-## Why estimate judges appart? {#sec-appA}
+## Why do we need to estimate judges' abilities? {#sec-appA}
 

### 2024-07-19 _tex/index.tex
Built site for gh-pages

@@
 
+Causal analysis is the field of experimental design and statistics
+pertaining to establishing cause and effect from data.
+
+Causal inference
+
+Directed Acyclic Graph (DAG)
+
+the four components of DAGs
+
+Structural Causal Model (SCM)
+
+the four components of DAGs in SCM notation
+
 \subsection{The flow of association and
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/ACflow.png}
+
+}
+
+\caption{\label{fig-ACflow}The flow of association and causation in
+graphs. Extracted from \citet[31]{Neal_2020}}
+
+\end{figure}%
+
 \section{Theoretical framework}\label{sec-theory}
 
+\subsection{Where does it all fit?}\label{sec-theory-where}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
+
+}
+
+\caption{\label{fig-IEflow}Identification-Estimation flowchart.
+Extracted from \citet[32]{Neal_2020}}
+
+\end{figure}%
+
 \subsection{A scientific model for the DCJ}\label{sec-theory-scientific}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/SciModel_simp1.png}
+
+}
+
+\caption{\label{fig-SciModel_simp1}DCJ causal diagram, simplified
+description}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{images/figures/SciModel_simp2.png}
+
+}
+
+\caption{\label{fig-SciModel_simp2}DCJ causal diagram, simplified
+mathematical description}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{images/figures/SciModel_pop.png}
+
+}
+
+\caption{\label{fig-SciModel_pop}DCJ causal diagram, population
+mathematical description}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/SciModel_samp.png}
+
+}
+
+\caption{\label{fig-SciModel_samp}DCJ causal diagram, sample with
+comparisons mathematical description}
+
+\end{figure}%
+
+\subsection{Probabilitics assumptions of the scientific
+model}\label{sec-theory-probability}
+
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+
+\begin{equation}\phantomsection\label{eq-StructuralModel}{
+\begin{aligned}
+  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
+  \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\
+  \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\
+  \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\
+  \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\
+  & \; e_{k} \perp\!\!\!\perp e_{i} \\
+  & \; e_{k} \perp\!\!\!\perp e_{r_{i}} \\
+  & \; e_{i} \perp\!\!\!\perp e_{r_{i}}
+\end{aligned}
+}\end{equation}
 
@@
 
-\subsection{Where does it all fit?}\label{sec-theory-where}
+\begin{equation}\phantomsection\label{eq-StatModel_general}{
+\begin{aligned}
+  O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\
+  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
+  \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\
+  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
+  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
+  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
+  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
+  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
+\end{aligned}
+}\end{equation}
+
+for identification purposes we can set
+\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02\),
+\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1\), and
+\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1\). A special case of this
+would be to assume that the data comes from the same population, in that
+case, \(\sigma_{k(g)} = \sigma_{k} = 0.02\),
+\(\sigma_{i(g)} = \sigma_{i} = 1\)
+
+\subsection{Let's talk about Thurstone}\label{sec-theory-thurstone}
+
+Thurstone's comparative judgment \citet{Thurstone_1927} is based on the
+formula: \[
+X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}}
+\]
+
+where \(X_{AB}\) defines the comparative judgment outcome, \(S_{A}\) and
+\(S_{B}\) are the modal discriminal processes,
+\(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}\),
+with \(\sigma_{A}\) and \(\sigma_{B}\) being the dispersion of
+discriminal processes \(A\) and \(B\), respectively, and \(\rho\) the
+correlation between discriminal processes.
+
+The theory identifies five cases:
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Case 1:} only constant \(\rho\) (not \(\rho_{ij}\))
+\item
+  \textbf{Case 2:} \(X_{ij}\) becomes \(X_{kij}\) with \(k=1, \dots, K\)
+  judges (replication, not duplication)
+\item
+  \textbf{Case 3:} \(\rho = 0\), then
+  \(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}\)
+\item
+  \textbf{Case 4:} \(\sigma_{B}=\sigma_{A}+d\), then
+  \(\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}\)
+\item
+  \textbf{Case 5:} \(\sigma_{B}=\sigma_{A}\), then
+  \(\sigma_{AB} = \sqrt{2}\sigma\)
+\end{itemize}
+
+Now using the DAG and statistical notation
+
+\begin{equation}\phantomsection\label{eq-StatModel_thurstone}{
+\begin{aligned}
+  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
+  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
+  \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\
+  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
+  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
+  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
+  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
+  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
+\end{aligned}
+}\end{equation}
+
+The theory identifies five cases:
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Case 1:} only constant \(\rho \approx \sigma_{i}\)
+\item
+  \textbf{Case 2:} now judges are separated by using \(\gamma_{k}\)
+\item
+  \textbf{Case 3:} \(\rho \approx \sigma_{e_{i}} = 0\) (no nesting of
+  texts on students), then
+  \(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}\)
+\item
+  \textbf{Case 4:} \(\sigma_{B}=\sigma_{A}+d\), then
+  \(\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}\)
+\item
+  \textbf{Case 5:} \(\sigma_{B}=\sigma_{A}\), then
+  \(\sigma_{AB} = \sqrt{2}\sigma\)
+\end{itemize}
+
+But now can we see other scenarios than just those 5 cases?
+
+\begin{itemize}
+\tightlist
+\item
+  consider different \(\rho \approx \sum_{p=1}^{P} \sigma_{p}\),
+  depending on \(P\) nesting structures
+\item
+  we can now investigate \(\gamma_{k}\)
+\item
+  we can assume \(\sigma_{B} \neq \sigma_{A}\), no need for results on
+  the limit
+\end{itemize}
 
@@
 
-\subsection{Why estimate judges appart?}\label{sec-appA}
+\subsection{Why do we need to estimate judges'
+abilities?}\label{sec-appA}
 

### 2024-07-23 index.qmd
section:preliminaries,theory; process:write

@@
 <!-- 1. What is comparative judgment? -->
-In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *Dichotomous* Comparative Judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b].
+In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *dichotomous* comparative judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b].
 In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli.
@@
 Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020].
-Nevertheless, despite the method's widespread use, the literature does not offer a clear representation of the plausible mechanisms that generate the DCJ data.
+Nevertheless, despite the method's widespread use, the literature does not offer a clear representation of the plausible mechanisms that generate DCJ data.
 Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.
@@
 <!-- 3. What are the factors behind DCJ? -->
-According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the DCJ outcome.
+According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the method's outcome.
 These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used.
@@
 Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome.
-For instance, @vanDaal_et_al_2016 assessed university students' skills in academic writing, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).
+For example, @vanDaal_et_al_2016 assessed the academic writing skills of university students (units) using multiple argumentative essays (sub-units).
 
-<!-- 4. Solution to the gap, final objective and relevance -->
-Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the plausible mechanisms generating the DCJ outcome.
-This study aims to fill this gap by representing DCJ within the causal analysis framework [@Pearl_2009; @Pearl_et_al_2016].
-Using Directed Acyclic Graphs [DAG, @Gross_et_al_2018; @Neal_2020], the study will construct a scientific model to elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome could be generated.
-Next, using a minimal set of assumptions embedded in the framework, the study will translate the scientific model into a probabilistic statistical model.
-This model derives statistical estimands to draw inferences about plausible causal relationships within the DCJ system.
+<!-- 4. Solution to the gap -->
+Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome.
+This study aims to fill this gap by representing DCJ within the causal inference framework.
+Specifically, using the *structural approach* to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study aims to construct a scientific model.
+This model will elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome is generated.
+Next, using a minimal set of assumptions, the study will translate the scientific model into a probabilistic statistical model.
+This model will produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.
+
+<!-- 5. final objective and relevance -->
 Ultimately, this research aims to extend the law of comparative judgment initially proposed by @Thurstone_1927 and provide a sound probabilistic base for the statistical analysis of DCJ data.
@@
 
-## Causal analysis and DAGs {#sec-prelim-causal}
+<!-- This section aims to provide the reader with the essential terms, definitions, notation, and assumptions inherent in the structural causal models (SCM) and directed acyclic graphs (DAG) framework.
+-->
 
-Causal analysis is the field of experimental design and statistics pertaining to establishing cause and effect from data.
+## The structural approach to causal inference {#sec-prelim-structural}
 
-Causal inference
+<!-- 1. What is causal inference and counterfactuals? -->
+In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
+Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical representation for analyzing causes and counterfactuals [@Pearl_2009].
+Counterfactuals reflect scenarios *contrary to fact*, where alternative *potential* outcomes have not been observed and cannot be observed [@Neal_2020; @Counterfactual_2024].
+Using counterfactuals, researchers can construct a theory of the world that explains why specific causes have specific effects, and what happens in the absence of those causes.
+Consequently, counterfactuals represent the highest level of abstraction in causal inference [@Pearl_et_al_2018].
 
-Directed Acyclic Graph (DAG)
+<!-- 2. How can we construct counterfactuals and apply causal inference? -->
+Several approaches to constructing counterfactuals and applying causal inference exist, but two are particularly prominent: the potential outcomes approach and the structural approach.
+The potential outcomes approach, also known as the Neyman-Rubin causal model, was developed by @Neyman_et_al_1990 and @Rubin_1974.
+In contrast, the structural approach was pioneered by @Wright_1921, and later extended by Pearl [@Pearl_2009; @Pearl_et_al_2016].
+Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
+The potential outcomes approach relies on counterfactual notation, whereas the structural approach utilizes the do-operator and Structural Causal Models [SCM, @Pearl_2009; @Pearl_et_al_2016].
+Despite these differences, both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].
 
-the four components of DAGs
+<!-- 3. But which approach are we going to use? -->
+However, the structural approach offers a significant advantage over the potential outcomes approach by enabling the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
+These graphical representations provides a transparent depiction of a system's complexity, revealing its underlying assumptions and the plausible mechanisms generating the system's outcome.
+This capability aligns well with the aims of this study.
+Consequently, the structural approach will be the primary method used to address the study's research goal.
 
-Structural Causal Model (SCM)
+<!-- 4. But how do we move from targets of inference to estimates -->
 
-the four components of DAGs in SCM notation
+<!-- Before tackling the main objective of this study, however, it is important to understand how researchers can move from causal estimands to statistical estimates.
+Causal estimands, also known as targets of inference, are the quantities that the researcher wants to estimate.
+Usually, these quantities are articulated through different research questions and they are depicted using the language of the Potential Outcomes framework, Neyman-Rubin causal framework [@Neyman_et_al_1990; @Rubin_1974] mathematical expectation, e.g., $E[Y_{i}(1) - Y_{i}(0)]$ indicating that a researcher is interested in estimating the *average treatment effect* (ATE) -->
+
+<!-- This study uses the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020].
+The flowchart illustrates the process of moving from a target (causal) *estimand* to a corresponding *estimate*, through identification and estimation (@fig-IEflowchart).
+-->
+
+<!-- Initially, it defines the different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method.
+The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data.
+Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets.
+Bayesian statistical methods offer an intuitive approach for the task.
+Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands.
+Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis.
+-->
+
+<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*.
+-->
+
+![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+
+## Graphs, DAGs and SCMs {#sec-prelim-graphs}
+
+<!-- why to use it? -->
+
+<!-- Causal models as a tool for causal inference -->
+
+<!-- DAGs as a graphical representation of an SCM -->
+
+<!-- Terms and definitions -->
+
+<!-- the four components of DAGs -->
+
+<!-- A *scientific (causal) model* describe the causal assumptions and mechanisms underlying a system, and facilitate inferences about causal relationships from data [@Hitchcock_2023]. -->
+
+<!-- To aid in the translation and visualization of the assumption underlying the DCJ procedure, Directed Acyclic Graphs (DAG) and Structural Causal Models (SCM) [@Pearl_2009; @Pearl_et_al_2016; @Cinelli_et_al_2021] are employed.
+DAG are [(...)]{style="color:red;"}.
+While SCM are [(...)]{style="color:red;"}.
+-->
+
+<!-- Directed Acyclic Graph (DAG) -->
+
+<!-- Plays an important role in effectively conveying an assumed -->
+<!-- causal model and determining which variables that should be controlled for in estimation of -->
+<!-- causal effects. -->
+
+<!-- Graph means it is nodes and con- -->
+<!-- nections. Directed means the connections have arrows that indicate directions of causal in- -->
+<!-- fluence. And acyclic means that causes do not eventually flow back on themselves. A DAG -->
+<!-- is a way of describing qualitative causal relationships among variables. -->
+
+<!-- Structural Causal Model (SCM) -->
+<!-- Causal models are mathematical models representing causal relationships within an individual system or population.
+They facilitate inferences about causal relationships from statistical data.
+They can teach us a good deal about the epistemology of causation, and about the relationship between causation and probability.
+They have also been applied to topics of interest to philosophers, such as the logic of counterfactuals, decision theory, and the analysis of actual causation.
+-->
+
+<!-- the four components of DAGs in SCM notation -->
 
@@
 # Theoretical framework {#sec-theory}
-
-## Where does it all fit? {#sec-theory-where}
-
-![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
 
@@
 ## Limitations and further research {#sec-discuss-limitations}
+<!-- and most of the methods depend on unverifiable assumptions, causal inference can still identify causes and estimate their effects, assuming a set of plausibles mechanisms for a system. -->
 
@@
 
-## Why do we need to estimate judges' abilities? {#sec-appA}
+## Additional definitions {#sec-appA}
+
+**Counterfactual:**
+
+## Why do we need to estimate judges' abilities? {#sec-appB}
 

### 2024-07-23 _tex/index.tex
Built site for gh-pages

@@
 \citeyearpar{Thurstone_1927} primarily refers to the method of
-\emph{Dichotomous} Comparative Judgment
+\emph{dichotomous} comparative judgment
 \citep[DCJ,][]{Pollitt_2012a, Pollitt_2012b}. In DCJ, a judge assesses
@@
 not offer a clear representation of the plausible mechanisms that
-generate the DCJ data. Particularly, there is no depiction of the
-complexity and the underlying assumptions of the DCJ system, nor how
-different assessment factors can potentially influence the observed DCJ
-outcome.
+generate DCJ data. Particularly, there is no depiction of the complexity
+and the underlying assumptions of the DCJ system, nor how different
+assessment factors can potentially influence the observed DCJ outcome.
 
 According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
-several assessment factors interact and influence the DCJ outcome. These
-factors include the number and characteristics of the stimuli, their
-\emph{proximity} in terms of the assessed trait, the number of
+several assessment factors interact and influence the method's outcome.
+These factors include the number and characteristics of the stimuli,
+their \emph{proximity} in terms of the assessed trait, the number of
 comparison per stimulus, and the pairing algorithm used. Furthermore,
@@
 \emph{proximity} in terms of the assessed trait, can significantly
-influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
-university students' skills in academic writing, utilizing multiple
-argumentative essays (stimuli, sub-units) originating from various
-students (units).
+influence the outcome. For example, \citet{vanDaal_et_al_2016} assessed
+the academic writing skills of university students (units) using
+multiple argumentative essays (sub-units).
 
@@
 none, to the best of the authors' knowledge, have provided a transparent
-depiction of the DCJ system and the plausible mechanisms generating the
-DCJ outcome. This study aims to fill this gap by representing DCJ within
-the causal analysis framework \citep{Pearl_2009, Pearl_et_al_2016}.
-Using Directed Acyclic Graphs
-\citep[DAG,][]{Gross_et_al_2018, Neal_2020}, the study will construct a
-scientific model to elucidate the underlying assumptions of the DCJ
-system, providing plausible mechanisms for how the DCJ outcome could be
-generated. Next, using a minimal set of assumptions embedded in the
-framework, the study will translate the scientific model into a
-probabilistic statistical model. This model derives statistical
-estimands to draw inferences about plausible causal relationships within
-the DCJ system. Ultimately, this research aims to extend the law of
-comparative judgment initially proposed by \citet{Thurstone_1927} and
-provide a sound probabilistic base for the statistical analysis of DCJ
-data. Consequently, this research holds significance for researchers and
+depiction of the DCJ system and the mechanisms generating the DCJ
+outcome. This study aims to fill this gap by representing DCJ within the
+causal inference framework. Specifically, using the \emph{structural
+approach} to causal inference
+\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study aims to
+construct a scientific model. This model will elucidate the underlying
+assumptions of the DCJ system, providing plausible mechanisms for how
+the DCJ outcome is generated. Next, using a minimal set of assumptions,
+the study will translate the scientific model into a probabilistic
+statistical model. This model will produce statistical estimates to draw
+inferences about plausible causal relationships within the DCJ system.
+
+Ultimately, this research aims to extend the law of comparative judgment
+initially proposed by \citet{Thurstone_1927} and provide a sound
+probabilistic base for the statistical analysis of DCJ data.
+Consequently, this research holds significance for researchers and
 analysts involved in education and assessment procedures who implement
@@
 
-\subsection{Causal analysis and DAGs}\label{sec-prelim-causal}
+\subsection{The structural approach to causal
+inference}\label{sec-prelim-structural}
 
-Causal analysis is the field of experimental design and statistics
-pertaining to establishing cause and effect from data.
+In statistics, \emph{causal inference} refers to the process of
+identifying the causes of a phenomenon and estimating their effects
+using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
+statistical modeling, which focuses solely on summarizing data and
+inferring associations, causal inference provides a coherent
+mathematical representation for analyzing causes and counterfactuals
+\citep{Pearl_2009}. Counterfactuals reflect scenarios \emph{contrary to
+fact}, where alternative \emph{potential} outcomes have not been
+observed and cannot be observed \citep{Neal_2020, Counterfactual_2024}.
+Using counterfactuals, researchers can construct a theory of the world
+that explains why specific causes have specific effects, and what
+happens in the absence of those causes. Consequently, counterfactuals
+represent the highest level of abstraction in causal inference
+\citep{Pearl_et_al_2018}.
 
-Causal inference
+Several approaches to constructing counterfactuals and applying causal
+inference exist, but two are particularly prominent: the potential
+outcomes approach and the structural approach. The potential outcomes
+approach, also known as the Neyman-Rubin causal model, was developed by
+\citet{Neyman_et_al_1990} and \citet{Rubin_1974}. In contrast, the
+structural approach was pioneered by \citet{Wright_1921}, and later
+extended by Pearl \citep{Pearl_2009, Pearl_et_al_2016}. Both approaches
+employ rigorous mathematical notation to characterize causal inference,
+but they do so in different ways \citep{Neal_2020}. The potential
+outcomes approach relies on counterfactual notation, whereas the
+structural approach utilizes the do-operator and Structural Causal
+Models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
+differences, both approaches provide methods for using experimental and
+observational data to estimate causal effects \citep{Pearl_2010}.
 
-Directed Acyclic Graph (DAG)
+However, the structural approach offers a significant advantage over the
+potential outcomes approach by enabling the graphical representation of
+systems through directed acyclic graphs
+\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. These graphical
+representations provides a transparent depiction of a system's
+complexity, revealing its underlying assumptions and the plausible
+mechanisms generating the system's outcome. This capability aligns well
+with the aims of this study. Consequently, the structural approach will
+be the primary method used to address the study's research goal.
 
-the four components of DAGs
+\begin{figure}
 
-Structural Causal Model (SCM)
+\centering{
 
-the four components of DAGs in SCM notation
+\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
+
+}
+
+\caption{\label{fig-IEflow}Identification-Estimation flowchart.
+Extracted from \citet[32]{Neal_2020}}
+
+\end{figure}%
+
+\subsection{Graphs, DAGs and SCMs}\label{sec-prelim-graphs}
 
@@
 \section{Theoretical framework}\label{sec-theory}
-
-\subsection{Where does it all fit?}\label{sec-theory-where}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
-
-}
-
-\caption{\label{fig-IEflow}Identification-Estimation flowchart.
-Extracted from \citet[32]{Neal_2020}}
-
-\end{figure}%
 
@@
 
+\subsection{Additional definitions}\label{sec-appA}
+
+\textbf{Counterfactual:}
+
 \subsection{Why do we need to estimate judges'
-abilities?}\label{sec-appA}
+abilities?}\label{sec-appB}
 

### 2024-07-24 index.qmd
section:background; process:write

@@
 This study aims to fill this gap by representing DCJ within the causal inference framework.
-Specifically, using the *structural approach* to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study aims to construct a scientific model.
+Specifically, using the structural approach to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study aims to construct a scientific model.
 This model will elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome is generated.
@@
 
-# Preliminaries {#sec-prelim}
+# Theoretical background {#sec-background}
 
-<!-- This section aims to provide the reader with the essential terms, definitions, notation, and assumptions inherent in the structural causal models (SCM) and directed acyclic graphs (DAG) framework.
+## The structural approach to causal inference {#sec-background-structural}
+
+<!-- 1. What is causal inference? -->
+In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
+Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical representation for analyzing causes and counterfactuals [@Pearl_2009].
+
+<!-- 2. what are counterfactuals? -->
+According to @Pearl_et_al_2018, counterfactuals occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association, and form the foundation of causal inference.
+Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
+Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in the absence of those causes [@Pearl_et_al_2018].
+Counterfactuals achieves this by translating causal statements to statements about "what would have happened under different circumstances."
+
+<!-- 3. How do we apply causal inference and counterfactuals? -->
+Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016].
+Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
+The potential outcomes approach relies on counterfactual notation, whereas the structural approach utilizes the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016].
+Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].
+
+<!-- 4. But approach are we going to use? -->
+However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
+These graphical representations provide a transparent depiction of a system’s complexity, revealing its underlying assumptions and the plausible causal mechanisms responsible for generating the system’s outcome.
+Ultimately, DAGs play the important role of effectively conveying the assumed causal structure of the system, and also determine wich variables should be controlled for to be able to estimate causal effects.
+Consequently, the structural approach will be the primary method used to address one of the study’s research goals.
+
+## DAGs, SCMs, and the flow of association and causation {#sec-background-dag}
+
+<!-- 1. What are DAGs? and how are they represented? -->
+As previously mentioned, the structural approach employs Directed Acyclic Graphs (DAGs) for the graphical representation of systems.
+A *graph* consists of nodes connected by edges [@Gross_et_al_2018].
+In a *directed graph*, edges extend from a *parent* node to a *child* node, with arrows indicating the direction of causal influence.
+In directed *acyclic* graphs, the direction of causal influences does not loop back on themselves, meaning the graph contains no cycles [@Neal_2020, @McElreath_2020].
+
+<!-- 2. What does a DAG imply? -->
+<!-- and utilizes the do-operator and Structural Causal Models (SCMs) for mathematical notation in causal inference. -->
+<!-- Structural Causal Model (SCM) -->
+<!-- Causal models are mathematical models representing causal relationships within an individual system or population.
+They facilitate inferences about causal relationships from statistical data.
+They can teach us a good deal about the epistemology of causation, and about the relationship between causation and probability.
+They have also been applied to topics of interest to philosophers, such as the logic of counterfactuals, decision theory, and the analysis of actual causation.
 -->
 
-## The structural approach to causal inference {#sec-prelim-structural}
+<!-- 3. But what are those mysterious functions?, and how do they become statistical quantities? -->
+![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=100%}
 
-<!-- 1. What is causal inference and counterfactuals? -->
-In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical representation for analyzing causes and counterfactuals [@Pearl_2009].
-Counterfactuals reflect scenarios *contrary to fact*, where alternative *potential* outcomes have not been observed and cannot be observed [@Neal_2020; @Counterfactual_2024].
-Using counterfactuals, researchers can construct a theory of the world that explains why specific causes have specific effects, and what happens in the absence of those causes.
-Consequently, counterfactuals represent the highest level of abstraction in causal inference [@Pearl_et_al_2018].
+<!-- 4. What is the local markov assumption? and where does it lead? -->
 
-<!-- 2. How can we construct counterfactuals and apply causal inference? -->
-Several approaches to constructing counterfactuals and applying causal inference exist, but two are particularly prominent: the potential outcomes approach and the structural approach.
-The potential outcomes approach, also known as the Neyman-Rubin causal model, was developed by @Neyman_et_al_1990 and @Rubin_1974.
-In contrast, the structural approach was pioneered by @Wright_1921, and later extended by Pearl [@Pearl_2009; @Pearl_et_al_2016].
-Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
-The potential outcomes approach relies on counterfactual notation, whereas the structural approach utilizes the do-operator and Structural Causal Models [SCM, @Pearl_2009; @Pearl_et_al_2016].
-Despite these differences, both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].
+<!-- 5. What is the minimality assumption? and where does it lead? -->
 
-<!-- 3. But which approach are we going to use? -->
-However, the structural approach offers a significant advantage over the potential outcomes approach by enabling the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
-These graphical representations provides a transparent depiction of a system's complexity, revealing its underlying assumptions and the plausible mechanisms generating the system's outcome.
-This capability aligns well with the aims of this study.
-Consequently, the structural approach will be the primary method used to address the study's research goal.
+<!-- 6. What is the causal edges assumption? and where does it lead? -->
+
+<!-- 7. putting it all together -->
+
+<!-- 8. What are the DAG's building blocks?, and what do they imply? (SCMs) -->
+
+## But where does it all fit? {#sec-background-where}
 
@@
 
-## Graphs, DAGs and SCMs {#sec-prelim-graphs}
-
-<!-- why to use it? -->
-
-<!-- Causal models as a tool for causal inference -->
-
-<!-- DAGs as a graphical representation of an SCM -->
-
-<!-- Terms and definitions -->
-
-<!-- the four components of DAGs -->
-
-<!-- A *scientific (causal) model* describe the causal assumptions and mechanisms underlying a system, and facilitate inferences about causal relationships from data [@Hitchcock_2023]. -->
-
-<!-- To aid in the translation and visualization of the assumption underlying the DCJ procedure, Directed Acyclic Graphs (DAG) and Structural Causal Models (SCM) [@Pearl_2009; @Pearl_et_al_2016; @Cinelli_et_al_2021] are employed.
-DAG are [(...)]{style="color:red;"}.
-While SCM are [(...)]{style="color:red;"}.
--->
-
-<!-- Directed Acyclic Graph (DAG) -->
-
-<!-- Plays an important role in effectively conveying an assumed -->
-<!-- causal model and determining which variables that should be controlled for in estimation of -->
-<!-- causal effects. -->
-
-<!-- Graph means it is nodes and con- -->
-<!-- nections. Directed means the connections have arrows that indicate directions of causal in- -->
-<!-- fluence. And acyclic means that causes do not eventually flow back on themselves. A DAG -->
-<!-- is a way of describing qualitative causal relationships among variables. -->
-
-<!-- Structural Causal Model (SCM) -->
-<!-- Causal models are mathematical models representing causal relationships within an individual system or population.
-They facilitate inferences about causal relationships from statistical data.
-They can teach us a good deal about the epistemology of causation, and about the relationship between causation and probability.
-They have also been applied to topics of interest to philosophers, such as the logic of counterfactuals, decision theory, and the analysis of actual causation.
--->
-
-<!-- the four components of DAGs in SCM notation -->
-
-## The flow of association and causation {#sec-prelim-flow}
-
-![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=100%}
-
 # Theoretical framework {#sec-theory}
@@
 ## A scientific model for the DCJ {#sec-theory-scientific}
+
+<!-- A *scientific (causal) model* describe the causal assumptions and mechanisms underlying a system, and facilitate inferences about causal relationships from data [@Hitchcock_2023]. -->
 
@@
 
-**Conflict of interests:** The authors declare no conflict of interest.
+**Financial interests:** The authors have no relevant financial interest to disclose.
+
+**Non-financial interests:** Author XX serve on advisory broad of Company Y but receives no compensation this role.
 
@@
 
-**Availability of data and materials:** No data was utilized in this study
+**Availability of data and materials:** No data was utilized in this study.
 
@@
 
-**Counterfactual:**
-
 ## Why do we need to estimate judges' abilities? {#sec-appB}

### 2024-07-24 _tex/index.tex
Built site for gh-pages

@@
 outcome. This study aims to fill this gap by representing DCJ within the
-causal inference framework. Specifically, using the \emph{structural
-approach} to causal inference
-\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study aims to
-construct a scientific model. This model will elucidate the underlying
-assumptions of the DCJ system, providing plausible mechanisms for how
-the DCJ outcome is generated. Next, using a minimal set of assumptions,
-the study will translate the scientific model into a probabilistic
-statistical model. This model will produce statistical estimates to draw
-inferences about plausible causal relationships within the DCJ system.
+causal inference framework. Specifically, using the structural approach
+to causal inference \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016},
+the study aims to construct a scientific model. This model will
+elucidate the underlying assumptions of the DCJ system, providing
+plausible mechanisms for how the DCJ outcome is generated. Next, using a
+minimal set of assumptions, the study will translate the scientific
+model into a probabilistic statistical model. This model will produce
+statistical estimates to draw inferences about plausible causal
+relationships within the DCJ system.
 
@@
 
-\section{Preliminaries}\label{sec-prelim}
+\section{Theoretical background}\label{sec-background}
 
 \subsection{The structural approach to causal
-inference}\label{sec-prelim-structural}
+inference}\label{sec-background-structural}
 
@@
 mathematical representation for analyzing causes and counterfactuals
-\citep{Pearl_2009}. Counterfactuals reflect scenarios \emph{contrary to
-fact}, where alternative \emph{potential} outcomes have not been
-observed and cannot be observed \citep{Neal_2020, Counterfactual_2024}.
-Using counterfactuals, researchers can construct a theory of the world
-that explains why specific causes have specific effects, and what
-happens in the absence of those causes. Consequently, counterfactuals
-represent the highest level of abstraction in causal inference
-\citep{Pearl_et_al_2018}.
+\citep{Pearl_2009}.
 
-Several approaches to constructing counterfactuals and applying causal
-inference exist, but two are particularly prominent: the potential
-outcomes approach and the structural approach. The potential outcomes
-approach, also known as the Neyman-Rubin causal model, was developed by
-\citet{Neyman_et_al_1990} and \citet{Rubin_1974}. In contrast, the
-structural approach was pioneered by \citet{Wright_1921}, and later
-extended by Pearl \citep{Pearl_2009, Pearl_et_al_2016}. Both approaches
+According to \citet{Pearl_et_al_2018}, counterfactuals occupy the
+highest level of cognitive abstraction in the ladder of causation,
+followed by intervention and association, and form the foundation of
+causal inference. Counterfactuals represent scenarios \emph{contrary to
+fact}, where alternative \emph{potential} outcomes resulting from a
+cause are neither observed nor observable
+\citep{Neal_2020, Counterfactual_2024}. Nevertheless, despite their
+abstract nature, counterfactuals enable the development of a
+\emph{theory of the world} that explains why specific causes have
+specific effects and what occurs in the absence of those causes
+\citep{Pearl_et_al_2018}. Counterfactuals achieves this by translating
+causal statements to statements about ``what would have happened under
+different circumstances.''
+
+Several approaches to causal inference and counterfactuals exist, but
+two are particularly prominent: the potential outcomes approach, also
+known as the Neyman-Rubin causal model
+\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
+\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}. Both approaches
 employ rigorous mathematical notation to characterize causal inference,
@@
 outcomes approach relies on counterfactual notation, whereas the
-structural approach utilizes the do-operator and Structural Causal
-Models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
-differences, both approaches provide methods for using experimental and
-observational data to estimate causal effects \citep{Pearl_2010}.
+structural approach utilizes the do-operator and structural causal
+models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
+differences, both notations can be expressed in terms of the other, and
+both approaches provide methods for using experimental and observational
+data to estimate causal effects \citep{Pearl_2010}.
 
-However, the structural approach offers a significant advantage over the
-potential outcomes approach by enabling the graphical representation of
+However, the structural approach offers a key advantage over the
+potential outcomes approach: it enables the graphical representation of
 systems through directed acyclic graphs
 \citep[DAG,][]{Gross_et_al_2018, Neal_2020}. These graphical
-representations provides a transparent depiction of a system's
+representations provide a transparent depiction of a system's
 complexity, revealing its underlying assumptions and the plausible
-mechanisms generating the system's outcome. This capability aligns well
-with the aims of this study. Consequently, the structural approach will
-be the primary method used to address the study's research goal.
+causal mechanisms responsible for generating the system's outcome.
+Ultimately, DAGs play the important role of effectively conveying the
+assumed causal structure of the system, and also determine wich
+variables should be controlled for to be able to estimate causal
+effects. Consequently, the structural approach will be the primary
+method used to address one of the study's research goals.
+
+\subsection{DAGs, SCMs, and the flow of association and
+causation}\label{sec-background-dag}
+
+As previously mentioned, the structural approach employs Directed
+Acyclic Graphs (DAGs) for the graphical representation of systems. A
+\emph{graph} consists of nodes connected by edges
+\citep{Gross_et_al_2018}. In a \emph{directed graph}, edges extend from
+a \emph{parent} node to a \emph{child} node, with arrows indicating the
+direction of causal influence. In directed \emph{acyclic} graphs, the
+direction of causal influences does not loop back on themselves, meaning
+the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/ACflow.png}
+
+}
+
+\caption{\label{fig-ACflow}The flow of association and causation in
+graphs. Extracted from \citet[31]{Neal_2020}}
+
+\end{figure}%
+
+\subsection{But where does it all fit?}\label{sec-background-where}
 
@@
 Extracted from \citet[32]{Neal_2020}}
-
-\end{figure}%
-
-\subsection{Graphs, DAGs and SCMs}\label{sec-prelim-graphs}
-
-\subsection{The flow of association and
-causation}\label{sec-prelim-flow}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/ACflow.png}
-
-}
-
-\caption{\label{fig-ACflow}The flow of association and causation in
-graphs. Extracted from \citet[31]{Neal_2020}}
 
@@
 
-\textbf{Conflict of interests:} The authors declare no conflict of
-interest.
+\textbf{Financial interests:} The authors have no relevant financial
+interest to disclose.
+
+\textbf{Non-financial interests:} Author XX serve on advisory broad of
+Company Y but receives no compensation this role.
 
@@
 \textbf{Availability of data and materials:} No data was utilized in
-this study
+this study.
 
@@
 
-\textbf{Counterfactual:}
-
 \subsection{Why do we need to estimate judges'

### 2024-07-24 index.qmd
section:background; process:write

@@
 abstract: |
-  Dichotomous Comparative Judgment [DCJ, @Pollitt_2012a, @Pollitt_2012b] requires judges to evaluate the relative manifestation of traits between pairs of stimuli, resulting in a dichotomous outcome indicating which stimulus exhibits the trait more strongly.
-Research has demonstrated DCJ's effectiveness and reliability in various domains [@Pollitt_2012b; @Bartholomew_et_al_2018; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
-Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that generate the DCJ data.
-Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data.
-This study aims to fill this gap by applying the framework of causal analysis and Directed Acyclic Graphs [DAG; @Pearl_2009].
-Using this framework, the study will construct a scientific model to elucidate the causal assumptions and mechanisms inherent the system.
-This model will enable researchers to draw inferences about causal relationships from DCJ data.
-Subsequently, the study will translate this model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference.
-The outcomes of this study will inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.
+  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait.
+DCJ has proven effective and reliable across various fields @Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020].
+However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data are generated.
+This study addresses these issues by integrating DCJ into the framework of causal inference.
+Specifically, utilizing a structural approach to causal inference, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
+It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal connections within the system.
+This research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data.
+The findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.
 key-points:
@@
 <!-- 5. final objective and relevance -->
-Ultimately, this research aims to extend the law of comparative judgment initially proposed by @Thurstone_1927 and provide a sound probabilistic base for the statistical analysis of DCJ data.
-Consequently, this research holds significance for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.
+Ultimately, this research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data.
+Consequently, the findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.
 
@@
 In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical representation for analyzing causes and counterfactuals [@Pearl_2009].
+Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical notation for analyzing causes and counterfactuals [@Pearl_2009].
 
@@
 Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in the absence of those causes [@Pearl_et_al_2018].
-Counterfactuals achieves this by translating causal statements to statements about "what would have happened under different circumstances."
+Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
+They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."
 
@@
 
-<!-- 4. But approach are we going to use? -->
+<!-- 4. But which approach are we going to use? -->
 However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
-These graphical representations provide a transparent depiction of a system’s complexity, revealing its underlying assumptions and the plausible causal mechanisms responsible for generating the system’s outcome.
-Ultimately, DAGs play the important role of effectively conveying the assumed causal structure of the system, and also determine wich variables should be controlled for to be able to estimate causal effects.
-Consequently, the structural approach will be the primary method used to address one of the study’s research goals.
+In this context, DAGs fulfill the role of a heuristic, effectively conveying the assumed causal structure of a system.
+They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].
+
+<!-- A Directed Acyclic Graph (DAG) allows for the transparent depiction of a system’s complexity, revealing its underlying assumptions and the plausible mechanisms generating the system’s outcome, assuming the DAG is true.
+-->
 
@@
 <!-- 1. What are DAGs? and how are they represented? -->
-As previously mentioned, the structural approach employs Directed Acyclic Graphs (DAGs) for the graphical representation of systems.
-A *graph* consists of nodes connected by edges [@Gross_et_al_2018].
-In a *directed graph*, edges extend from a *parent* node to a *child* node, with arrows indicating the direction of causal influence.
-In directed *acyclic* graphs, the direction of causal influences does not loop back on themselves, meaning the graph contains no cycles [@Neal_2020, @McElreath_2020].
+Graph theory is the branch of mathematics focused on the study of graphs.
+Graphs are mathematical structures used to model pairwise relations between objects [@Gross_et_al_2018].
+While graph theory covers a wide array of topics, the field of causal inference, particularly its structural approach, has incorporated some of its concepts to represent causes and counterfactuals formally and transparently.
+A causal graph, or Directed Acyclic Graph (DAG), as its name suggest, is a directed graph without cycles.
+A *graph* is a collection of nodes connected by edges.
+In a *directed graph*, edges extend from a node to another node, with arrows indicating the direction of causal influence.
+In a directed *acyclic* graph, the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles [@Neal_2020, @McElreath_2020].
 
@@
 <!-- 8. What are the DAG's building blocks?, and what do they imply? (SCMs) -->
+<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
 

### 2024-07-24 _tex/index.tex
Built site for gh-pages

@@
 \begin{abstract}
-Dichotomous Comparative Judgment
-\citep[DCJ,][@Pollitt\_2012b]{Pollitt_2012a} requires judges to evaluate
-the relative manifestation of traits between pairs of stimuli, resulting
-in a dichotomous outcome indicating which stimulus exhibits the trait
-more strongly. Research has demonstrated DCJ's effectiveness and
-reliability in various domains
-\citep{Pollitt_2012b, Bartholomew_et_al_2018, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2020, Boonen_et_al_2020}.
-Nevertheless, despite the method's widespread use, the literature lacks
-a transparent depiction of the DCJ system and the plausible mechanisms
-that generate the DCJ data. Particularly, there is no detailed
-explanation of how different assessment factors can potentially
-influence the observed DCJ data. This study aims to fill this gap by
-applying the framework of causal analysis and Directed Acyclic Graphs
-{[}DAG; \citet{Pearl_2009}{]}. Using this framework, the study will
-construct a scientific model to elucidate the causal assumptions and
-mechanisms inherent the system. This model will enable researchers to
-draw inferences about causal relationships from DCJ data. Subsequently,
-the study will translate this model into a probabilistic statistical
-model, aiming to derive statistical estimands for different targets of
-inference. The outcomes of this study will inform the planning of DCJ
-experiments and hold significance for researchers or analysts involved
-in education and assessment procedures who implement the DCJ
-methodology.
+Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
+of stimuli to determine which one exhibits a higher degree of a specific
+trait. DCJ has proven effective and reliable across various fields
+\citet{Pollitt_2012b}; \citet{Jones_2015}; \citet{vanDaal_et_al_2016};
+\citet{Bartholomew_et_al_2018}; \citet{Lesterhuis_2018};
+\citet{Bartholomew_et_al_2020}; \citet{Marshall_et_al_2020};
+\citet{Boonen_et_al_2020}{]}. However, despite the method's widespread
+use, existing literature lacks a clear explanation of the complexities
+and assumptions underpinning the DCJ system, as well as the plausible
+mechanisms through which DCJ data are generated. This study addresses
+these issues by integrating DCJ into the framework of causal inference.
+Specifically, utilizing a structural approach to causal inference, the
+study develops a scientific model to clarify the causal assumptions and
+mechanisms inherent in the DCJ system. It then translates this model
+into a probabilistic statistical framework to estimate statistical
+relationships and infer causal connections within the system. This
+research builds upon Thurstone's law of comparative judgment
+\citeyearpar{Thurstone_1927} and provide a robust probabilistic
+foundation for the statistical analysis of DCJ data. The findings of
+this study offer valuable insights for researchers and analysts involved
+in education and assessment procedures who implement or design DCJ
+experiments.
 \end{abstract}
@@
 
-Ultimately, this research aims to extend the law of comparative judgment
-initially proposed by \citet{Thurstone_1927} and provide a sound
-probabilistic base for the statistical analysis of DCJ data.
-Consequently, this research holds significance for researchers and
+Ultimately, this research builds upon Thurstone's law of comparative
+judgment \citeyearpar{Thurstone_1927} and provide a robust probabilistic
+foundation for the statistical analysis of DCJ data. Consequently, the
+findings of this study offer valuable insights for researchers and
 analysts involved in education and assessment procedures who implement
@@
 inferring associations, causal inference provides a coherent
-mathematical representation for analyzing causes and counterfactuals
+mathematical notation for analyzing causes and counterfactuals
 \citep{Pearl_2009}.
@@
 \emph{theory of the world} that explains why specific causes have
-specific effects and what occurs in the absence of those causes
-\citep{Pearl_et_al_2018}. Counterfactuals achieves this by translating
-causal statements to statements about ``what would have happened under
-different circumstances.''
+specific effects and what occurs in their absence
+\citep{Pearl_et_al_2018}. They achieve this by translating causal
+statements into counterfactual statements, that is, statements about
+``what would have happened in the world under different circumstances.''
 
@@
 systems through directed acyclic graphs
-\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. These graphical
-representations provide a transparent depiction of a system's
-complexity, revealing its underlying assumptions and the plausible
-causal mechanisms responsible for generating the system's outcome.
-Ultimately, DAGs play the important role of effectively conveying the
-assumed causal structure of the system, and also determine wich
-variables should be controlled for to be able to estimate causal
-effects. Consequently, the structural approach will be the primary
-method used to address one of the study's research goals.
+\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. In this context, DAGs
+fulfill the role of a heuristic, effectively conveying the assumed
+causal structure of a system. They do not represent detailed statistical
+models but allow researchers to deduce which statistical models can
+provide valid causal inferences, assuming the causal structure depicted
+in the DAG is accurate \citep{McElreath_2020}.
 
@@
 
-As previously mentioned, the structural approach employs Directed
-Acyclic Graphs (DAGs) for the graphical representation of systems. A
-\emph{graph} consists of nodes connected by edges
-\citep{Gross_et_al_2018}. In a \emph{directed graph}, edges extend from
-a \emph{parent} node to a \emph{child} node, with arrows indicating the
-direction of causal influence. In directed \emph{acyclic} graphs, the
-direction of causal influences does not loop back on themselves, meaning
-the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.
+Graph theory is the branch of mathematics focused on the study of
+graphs. Graphs are mathematical structures used to model pairwise
+relations between objects \citep{Gross_et_al_2018}. While graph theory
+covers a wide array of topics, the field of causal inference,
+particularly its structural approach, has incorporated some of its
+concepts to represent causes and counterfactuals formally and
+transparently. A causal graph, or Directed Acyclic Graph (DAG), as its
+name suggest, is a directed graph without cycles. A \emph{graph} is a
+collection of nodes connected by edges. In a \emph{directed graph},
+edges extend from a node to another node, with arrows indicating the
+direction of causal influence. In a directed \emph{acyclic} graph, the
+direction of causal influences does not loop back on itself, ensuring
+that the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.
 

### 2024-07-25 index.qmd
section:all; process:write,correct

@@
   Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait.
-DCJ has proven effective and reliable across various fields @Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020].
+DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020].
 However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data are generated.
-This study addresses these issues by integrating DCJ into the framework of causal inference.
+This study addresses these issues by representing DCJ within the framework of causal inference.
 Specifically, utilizing a structural approach to causal inference, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
 It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal connections within the system.
-This research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data.
-The findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.
+This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
+Its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.
 key-points:
@@
 Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020].
-Nevertheless, despite the method's widespread use, the literature does not offer a clear representation of the plausible mechanisms that generate DCJ data.
-Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.
+Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data are generated.
+Particularly, there is no depiction of the complexity and the assumptions underpinning the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.
 
@@
 Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome.
-For example, @vanDaal_et_al_2016 assessed the academic writing skills of university students (units) using multiple argumentative essays (sub-units).
+For instance, @vanDaal_et_al_2016 assessed academic writing skills of university students (units) using multiple argumentative essays (sub-units).
 
@@
 Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome.
-This study aims to fill this gap by representing DCJ within the causal inference framework.
-Specifically, using the structural approach to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study aims to construct a scientific model.
-This model will elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome is generated.
-Next, using a minimal set of assumptions, the study will translate the scientific model into a probabilistic statistical model.
-This model will produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.
+This study aims to fill this gap by representing DCJ within the framework of causal inference.
+Specifically, utilizing a structural approach to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
+Next, using a minimal set of assumptions, the study translates the scientific model into a probabilistic statistical model.
+This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.
 
 <!-- 5. final objective and relevance -->
-Ultimately, this research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data.
-Consequently, the findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.
+Ultimately, this research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
+Consequently, its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.
 
@@
 However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
-In this context, DAGs fulfill the role of a heuristic, effectively conveying the assumed causal structure of a system.
+DAGs are heuristics that effectively convey the assumed causal structure of a system.
 They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].
@@
 In a directed *acyclic* graph, the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles [@Neal_2020, @McElreath_2020].
+
+<!-- Graph theory is the branch of mathematics focused on the study of graphs.
+While graph theory covers a wide array of topics, the structural approach to causal inference incorporates some of its concepts to formally represent causes and counterfactuals.
+-->
+
+<!-- Graphs are mathematical structures used to model pairwise relations between objects (Gross, Yellen, and Anderson 2018).
+A graph is a collection of nodes connected by edges.
+In a directed graph, edges extend from a parent node to a child node, with arrows indicating the direction of the causal influence.
+In a directed acyclic graph (DAG), the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles McElreath (2020).
+-->
+
+<!-- To draw a DAG use the exclusion restriction assumption: a variable should not appear in the graph if it is not a part of the causal mechanisms that generate the outcome under analysis (see Neal, 2020).
+-->
 
@@
 <!-- 4. But how do we move from targets of inference to estimates -->
+<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->
 
@@
 
+<!-- From Lawson (2015), another way to see DCJ experiments I'd that you are grouping heterogeneous experimental units (objects) into homogeneous ones (subjects), which are your "blocks" -->
+
 ![ DCJ causal diagram, simplified description ](images/figures/SciModel_simp1.png){#fig-SciModel_simp1 fig-align="center" width=80%}
@@
 ![ DCJ causal diagram, sample with comparisons mathematical description ](images/figures/SciModel_samp.png){#fig-SciModel_samp fig-align="center" width=80%}
+
+<!-- It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
+However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
 
@@
 ## From the scientific to statistical model {#sec-theory-statistics}
+
+<!-- "An appropriate analysis of data cannot be completed without knowledge of what experimental design was used and how the data was collected, and conclusions are not reliable if they are not justified by the proper modeling and analysis of the data" (Lawson, 2015, p.
+7) -->
+
+<!-- "In many cases where observational units are different than the experimental units, the observational units will not be independent" (p. 177) -->
 

### 2024-07-25 _tex/index.tex
Built site for gh-pages

@@
 trait. DCJ has proven effective and reliable across various fields
-\citet{Pollitt_2012b}; \citet{Jones_2015}; \citet{vanDaal_et_al_2016};
-\citet{Bartholomew_et_al_2018}; \citet{Lesterhuis_2018};
-\citet{Bartholomew_et_al_2020}; \citet{Marshall_et_al_2020};
-\citet{Boonen_et_al_2020}{]}. However, despite the method's widespread
-use, existing literature lacks a clear explanation of the complexities
-and assumptions underpinning the DCJ system, as well as the plausible
-mechanisms through which DCJ data are generated. This study addresses
-these issues by integrating DCJ into the framework of causal inference.
-Specifically, utilizing a structural approach to causal inference, the
-study develops a scientific model to clarify the causal assumptions and
-mechanisms inherent in the DCJ system. It then translates this model
-into a probabilistic statistical framework to estimate statistical
-relationships and infer causal connections within the system. This
-research builds upon Thurstone's law of comparative judgment
-\citeyearpar{Thurstone_1927} and provide a robust probabilistic
-foundation for the statistical analysis of DCJ data. The findings of
-this study offer valuable insights for researchers and analysts involved
-in education and assessment procedures who implement or design DCJ
-experiments.
+\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
+However, despite the method's widespread use, existing literature lacks
+a clear explanation of the complexities and assumptions underpinning the
+DCJ system, as well as the plausible mechanisms through which DCJ data
+are generated. This study addresses these issues by representing DCJ
+within the framework of causal inference. Specifically, utilizing a
+structural approach to causal inference, the study develops a scientific
+model to clarify the causal assumptions and mechanisms inherent in the
+DCJ system. It then translates this model into a probabilistic
+statistical framework to estimate statistical relationships and infer
+causal connections within the system. This research provides a robust
+probabilistic foundation for the statistical analysis of DCJ data,
+building upon Thurstone's law of comparative judgment
+\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
+researchers and analysts designing and implementing DCJ experiments.
 \end{abstract}
@@
 validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
-Nevertheless, despite the method's widespread use, the literature does
-not offer a clear representation of the plausible mechanisms that
-generate DCJ data. Particularly, there is no depiction of the complexity
-and the underlying assumptions of the DCJ system, nor how different
-assessment factors can potentially influence the observed DCJ outcome.
+Nevertheless, despite the method's widespread use, existing literature
+lacks a clear representation of the plausible mechanisms through which
+DCJ data are generated. Particularly, there is no depiction of the
+complexity and the assumptions underpinning the DCJ system, nor how
+different assessment factors can potentially influence the observed DCJ
+outcome.
 
@@
 \emph{proximity} in terms of the assessed trait, can significantly
-influence the outcome. For example, \citet{vanDaal_et_al_2016} assessed
-the academic writing skills of university students (units) using
-multiple argumentative essays (sub-units).
+influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
+academic writing skills of university students (units) using multiple
+argumentative essays (sub-units).
 
@@
 outcome. This study aims to fill this gap by representing DCJ within the
-causal inference framework. Specifically, using the structural approach
-to causal inference \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016},
-the study aims to construct a scientific model. This model will
-elucidate the underlying assumptions of the DCJ system, providing
-plausible mechanisms for how the DCJ outcome is generated. Next, using a
-minimal set of assumptions, the study will translate the scientific
-model into a probabilistic statistical model. This model will produce
-statistical estimates to draw inferences about plausible causal
-relationships within the DCJ system.
+framework of causal inference. Specifically, utilizing a structural
+approach to causal inference
+\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study develops a
+scientific model to clarify the causal assumptions and mechanisms
+inherent in the DCJ system. Next, using a minimal set of assumptions,
+the study translates the scientific model into a probabilistic
+statistical model. This model aims to produce statistical estimates to
+draw inferences about plausible causal relationships within the DCJ
+system.
 
-Ultimately, this research builds upon Thurstone's law of comparative
-judgment \citeyearpar{Thurstone_1927} and provide a robust probabilistic
-foundation for the statistical analysis of DCJ data. Consequently, the
-findings of this study offer valuable insights for researchers and
-analysts involved in education and assessment procedures who implement
-or design DCJ experiments.
+Ultimately, this research provides a robust probabilistic foundation for
+the statistical analysis of DCJ data, building upon Thurstone's law of
+comparative judgment \citeyearpar{Thurstone_1927}. Consequently, its
+findings offer valuable insights for researchers and analysts designing
+and implementing DCJ experiments.
 
@@
 systems through directed acyclic graphs
-\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. In this context, DAGs
-fulfill the role of a heuristic, effectively conveying the assumed
-causal structure of a system. They do not represent detailed statistical
-models but allow researchers to deduce which statistical models can
-provide valid causal inferences, assuming the causal structure depicted
-in the DAG is accurate \citep{McElreath_2020}.
+\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs are heuristics that
+effectively convey the assumed causal structure of a system. They do not
+represent detailed statistical models but allow researchers to deduce
+which statistical models can provide valid causal inferences, assuming
+the causal structure depicted in the DAG is accurate
+\citep{McElreath_2020}.
 

### 2024-07-26 index.qmd
section:background; process:write

@@
 DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020].
-However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data are generated.
+However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data could be generated.
 This study addresses these issues by representing DCJ within the framework of causal inference.
-Specifically, utilizing a structural approach to causal inference, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
-It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal connections within the system.
+Specifically, utilizing the structural approach, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
+It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal effects within the system.
 This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
@@
 Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020].
-Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data are generated.
+Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data could be generated.
 Particularly, there is no depiction of the complexity and the assumptions underpinning the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.
@@
 This study aims to fill this gap by representing DCJ within the framework of causal inference.
-Specifically, utilizing a structural approach to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
+Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
 Next, using a minimal set of assumptions, the study translates the scientific model into a probabilistic statistical model.
@@
 <!-- 2. what are counterfactuals? -->
-According to @Pearl_et_al_2018, counterfactuals occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association, and form the foundation of causal inference.
 Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
+According to @Pearl_et_al_2018, counterfactuals form the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association.
+Despite this abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
 They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."
@@
 Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
-The potential outcomes approach relies on counterfactual notation, whereas the structural approach utilizes the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016].
+The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016].
 Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].
@@
 <!-- 4. But which approach are we going to use? -->
-However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
-DAGs are heuristics that effectively convey the assumed causal structure of a system.
+Nevertheless, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
+DAGs are heuristics that can effectively convey the assumed causal structure of a system.
 They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].
@@
 <!-- 1. What are DAGs? and how are they represented? -->
-Graph theory is the branch of mathematics focused on the study of graphs.
-Graphs are mathematical structures used to model pairwise relations between objects [@Gross_et_al_2018].
-While graph theory covers a wide array of topics, the field of causal inference, particularly its structural approach, has incorporated some of its concepts to represent causes and counterfactuals formally and transparently.
-A causal graph, or Directed Acyclic Graph (DAG), as its name suggest, is a directed graph without cycles.
-A *graph* is a collection of nodes connected by edges.
-In a *directed graph*, edges extend from a node to another node, with arrows indicating the direction of causal influence.
-In a directed *acyclic* graph, the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles [@Neal_2020, @McElreath_2020].
-
-<!-- Graph theory is the branch of mathematics focused on the study of graphs.
-While graph theory covers a wide array of topics, the structural approach to causal inference incorporates some of its concepts to formally represent causes and counterfactuals.
+<!-- Graph theory is a branch of mathematics focused on the study of graphs.
+Graphs are mathematical structures that model pairwise relations between objects [@Gross_et_al_2018].
+They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations.
+Graphs have proven useful in various fields, spanning statistics, computer science, operations research, and the natural and social sciences.
+One application in statistics that incorporates concepts from graph theory is causal inference.
+Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of the causal structure of a system [@Neal_2020].
 -->
 
-<!-- Graphs are mathematical structures used to model pairwise relations between objects (Gross, Yellen, and Anderson 2018).
-A graph is a collection of nodes connected by edges.
-In a directed graph, edges extend from a parent node to a child node, with arrows indicating the direction of the causal influence.
-In a directed acyclic graph (DAG), the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles McElreath (2020).
+<!-- A graph is a collection of nodes connected by edges.
+In a directed graph, edges extend from a *parent* node to a *child* node, with arrows indicating the direction of the causal influence.
+Two nodes that are connected by edge are said to be *adjacent* [@Gross_et_al_2018].
+In a directed acyclic graph (DAG), the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles [@Neal_2020; @McElreath_2020].
 -->
@@
 <!-- commands for d-separation -->
-\newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+<!-- \newcommand{\dsep}{\perp\!\!\!\perp} -->
+<!-- \newcommand{\ndsep}{\not\!\perp\!\!\!\perp} -->
 
-$$
-\begin{aligned}
-  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
-  \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\
-  \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\
-  \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\
-  \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\
-  & \; e_{k} \dsep e_{i} \\
-  & \; e_{k} \dsep e_{r_{i}} \\
-  & \; e_{i} \dsep e_{r_{i}}
-\end{aligned}
-$$ {#eq-StructuralModel}
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!--   O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\ -->
+<!--   \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\ -->
+<!--   \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\ -->
+<!--   \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\ -->
+<!--   \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\ -->
+<!--   & \; e_{k} \dsep e_{i} \\ -->
+<!--   & \; e_{k} \dsep e_{r_{i}} \\ -->
+<!--   & \; e_{i} \dsep e_{r_{i}} -->
+<!-- \end{aligned} -->
+<!-- $$ {#eq-StructuralModel} -->
 
@@
 
-$$
-\begin{aligned}
-  O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\
-  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
-  \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\
-  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
-  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
-  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
-  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
-  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
-\end{aligned}
-$$ {#eq-StatModel_general}
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!--   O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\ -->
+<!--   \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\ -->
+<!--   \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\ -->
+<!--   \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\ -->
+<!--   \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\ -->
+<!--   e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\ -->
+<!--   e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\ -->
+<!--   e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)}) -->
+<!-- \end{aligned} -->
+<!-- $$ {#eq-StatModel_general} -->
 
-for identification purposes we can set $\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02$, $\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1$, and $\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1$.
-A special case of this would be to assume that the data comes from the same population, in that case, $\sigma_{k(g)} = \sigma_{k} = 0.02$, $\sigma_{i(g)} = \sigma_{i} = 1$
+<!-- for identification purposes we can set $\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02$, $\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1$, and $\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1$.
+A special case of this would be to assume that the data comes from the same population, in that case, $\sigma_{k(g)} = \sigma_{k} = 0.02$, $\sigma_{i(g)} = \sigma_{i} = 1$ -->
 
@@
 
-Thurstone's comparative judgment @Thurstone_1927 is based on the formula:
-$$
-X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}}
-$$
+<!-- Thurstone's comparative judgment @Thurstone_1927 is based on the formula: -->
+<!-- $$ -->
+<!-- X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}} -->
+<!-- $$ -->
 
-where $X_{AB}$ defines the comparative judgment outcome, $S_{A}$ and $S_{B}$ are the modal discriminal processes, $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}$, with $\sigma_{A}$ and $\sigma_{B}$ being the dispersion of discriminal processes $A$ and $B$, respectively, and $\rho$ the correlation between discriminal processes.
+<!-- where $X_{AB}$ defines the comparative judgment outcome, $S_{A}$ and $S_{B}$ are the modal discriminal processes, $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}$, with $\sigma_{A}$ and $\sigma_{B}$ being the dispersion of discriminal processes $A$ and $B$, respectively, and $\rho$ the correlation between discriminal processes.
+-->
 
-The theory identifies five cases:
+<!-- The theory identifies five cases: -->
 
-- **Case 1:** only constant $\rho$ (not $\rho_{ij}$)
-- **Case 2:** $X_{ij}$ becomes $X_{kij}$ with $k=1, \dots, K$ judges (replication, not duplication)
-- **Case 3:** $\rho = 0$, then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$ 
-- **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$
-- **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$
+<!-- - **Case 1:** only constant $\rho$ (not $\rho_{ij}$) -->
+<!-- - **Case 2:** $X_{ij}$ becomes $X_{kij}$ with $k=1, \dots, K$ judges (replication, not duplication) -->
+<!-- - **Case 3:** $\rho = 0$, then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$  -->
+<!-- - **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$ -->
+<!-- - **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$ -->
 
-Now using the DAG and statistical notation
+<!-- Now using the DAG and statistical notation -->
 
-$$
-\begin{aligned}
-  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
-  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
-  \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\
-  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
-  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
-  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
-  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
-  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
-\end{aligned}
-$$ {#eq-StatModel_thurstone}
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!--   O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\ -->
+<!--   \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\ -->
+<!--   \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\ -->
+<!--   \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\ -->
+<!--   \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\ -->
+<!--   e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\ -->
+<!--   e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\ -->
+<!--   e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)}) -->
+<!-- \end{aligned} -->
+<!-- $$ {#eq-StatModel_thurstone} -->
 
-The theory identifies five cases:
+<!-- The theory identifies five cases: -->
 
-- **Case 1:** only constant $\rho \approx \sigma_{i}$
-- **Case 2:** now judges are separated by using $\gamma_{k}$
-- **Case 3:** $\rho \approx \sigma_{e_{i}} = 0$ (no nesting of texts on students), then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$ 
-- **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$
-- **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$
+<!-- - **Case 1:** only constant $\rho \approx \sigma_{i}$ -->
+<!-- - **Case 2:** now judges are separated by using $\gamma_{k}$ -->
+<!-- - **Case 3:** $\rho \approx \sigma_{e_{i}} = 0$ (no nesting of texts on students), then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$  -->
+<!-- - **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$ -->
+<!-- - **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$ -->
 
-But now can we see other scenarios than just those 5 cases?
+<!-- But now can we see other scenarios than just those 5 cases? -->
 
-- consider different $\rho \approx \sum_{p=1}^{P} \sigma_{p}$, depending on $P$ nesting structures 
-- we can now investigate $\gamma_{k}$
-- we can assume $\sigma_{B} \neq \sigma_{A}$, no need for results on the limit
+<!-- - consider different $\rho \approx \sum_{p=1}^{P} \sigma_{p}$, depending on $P$ nesting structures  -->
+<!-- - we can now investigate $\gamma_{k}$ -->
+<!-- - we can assume $\sigma_{B} \neq \sigma_{A}$, no need for results on the limit -->
 

### 2024-07-26 _tex/index.tex
Built site for gh-pages

@@
 DCJ system, as well as the plausible mechanisms through which DCJ data
-are generated. This study addresses these issues by representing DCJ
-within the framework of causal inference. Specifically, utilizing a
-structural approach to causal inference, the study develops a scientific
-model to clarify the causal assumptions and mechanisms inherent in the
-DCJ system. It then translates this model into a probabilistic
-statistical framework to estimate statistical relationships and infer
-causal connections within the system. This research provides a robust
-probabilistic foundation for the statistical analysis of DCJ data,
-building upon Thurstone's law of comparative judgment
-\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
-researchers and analysts designing and implementing DCJ experiments.
+could be generated. This study addresses these issues by representing
+DCJ within the framework of causal inference. Specifically, utilizing
+the structural approach, the study develops a scientific model to
+clarify the causal assumptions and mechanisms inherent in the DCJ
+system. It then translates this model into a probabilistic statistical
+framework to estimate statistical relationships and infer causal effects
+within the system. This research provides a robust probabilistic
+foundation for the statistical analysis of DCJ data, building upon
+Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
+Its findings offer valuable insights for researchers and analysts
+designing and implementing DCJ experiments.
 \end{abstract}
@@
 lacks a clear representation of the plausible mechanisms through which
-DCJ data are generated. Particularly, there is no depiction of the
+DCJ data could be generated. Particularly, there is no depiction of the
 complexity and the assumptions underpinning the DCJ system, nor how
@@
 outcome. This study aims to fill this gap by representing DCJ within the
-framework of causal inference. Specifically, utilizing a structural
-approach to causal inference
-\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study develops a
-scientific model to clarify the causal assumptions and mechanisms
-inherent in the DCJ system. Next, using a minimal set of assumptions,
-the study translates the scientific model into a probabilistic
-statistical model. This model aims to produce statistical estimates to
-draw inferences about plausible causal relationships within the DCJ
-system.
+framework of causal inference. Specifically, utilizing the structural
+approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
+develops a scientific model to clarify the causal assumptions and
+mechanisms inherent in the DCJ system. Next, using a minimal set of
+assumptions, the study translates the scientific model into a
+probabilistic statistical model. This model aims to produce statistical
+estimates to draw inferences about plausible causal relationships within
+the DCJ system.
 
@@
 
-According to \citet{Pearl_et_al_2018}, counterfactuals occupy the
-highest level of cognitive abstraction in the ladder of causation,
-followed by intervention and association, and form the foundation of
-causal inference. Counterfactuals represent scenarios \emph{contrary to
-fact}, where alternative \emph{potential} outcomes resulting from a
-cause are neither observed nor observable
-\citep{Neal_2020, Counterfactual_2024}. Nevertheless, despite their
-abstract nature, counterfactuals enable the development of a
-\emph{theory of the world} that explains why specific causes have
-specific effects and what occurs in their absence
+Counterfactuals represent scenarios \emph{contrary to fact}, where
+alternative \emph{potential} outcomes resulting from a cause are neither
+observed nor observable \citep{Neal_2020, Counterfactual_2024}.
+According to \citet{Pearl_et_al_2018}, counterfactuals form the
+foundation of causal inference and occupy the highest level of cognitive
+abstraction in the ladder of causation, followed by intervention and
+association. Despite this abstract nature, counterfactuals enable the
+development of a \emph{theory of the world} that explains why specific
+causes have specific effects and what occurs in their absence
 \citep{Pearl_et_al_2018}. They achieve this by translating causal
@@
 outcomes approach relies on counterfactual notation, whereas the
-structural approach utilizes the do-operator and structural causal
-models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
-differences, both notations can be expressed in terms of the other, and
-both approaches provide methods for using experimental and observational
-data to estimate causal effects \citep{Pearl_2010}.
+structural approach employs the do-operator and structural causal models
+\citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these differences,
+both notations can be expressed in terms of the other, and both
+approaches provide methods for using experimental and observational data
+to estimate causal effects \citep{Pearl_2010}.
 
-However, the structural approach offers a key advantage over the
-potential outcomes approach: it enables the graphical representation of
-systems through directed acyclic graphs
+Nevertheless, the structural approach offers an additional key advantage
+over the potential outcomes approach: it enables the graphical
+representation of any system through directed acyclic graphs
 \citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs are heuristics that
-effectively convey the assumed causal structure of a system. They do not
-represent detailed statistical models but allow researchers to deduce
-which statistical models can provide valid causal inferences, assuming
-the causal structure depicted in the DAG is accurate
+can effectively convey the assumed causal structure of a system. They do
+not represent detailed statistical models but allow researchers to
+deduce which statistical models can provide valid causal inferences,
+assuming the causal structure depicted in the DAG is accurate
 \citep{McElreath_2020}.
@@
 causation}\label{sec-background-dag}
-
-Graph theory is the branch of mathematics focused on the study of
-graphs. Graphs are mathematical structures used to model pairwise
-relations between objects \citep{Gross_et_al_2018}. While graph theory
-covers a wide array of topics, the field of causal inference,
-particularly its structural approach, has incorporated some of its
-concepts to represent causes and counterfactuals formally and
-transparently. A causal graph, or Directed Acyclic Graph (DAG), as its
-name suggest, is a directed graph without cycles. A \emph{graph} is a
-collection of nodes connected by edges. In a \emph{directed graph},
-edges extend from a node to another node, with arrows indicating the
-direction of causal influence. In a directed \emph{acyclic} graph, the
-direction of causal influences does not loop back on itself, ensuring
-that the graph contains no cycles \citep[@McElreath\_2020]{Neal_2020}.
 
@@
 
-\newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
-
-\begin{equation}\phantomsection\label{eq-StructuralModel}{
-\begin{aligned}
-  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
-  \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\
-  \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\
-  \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\
-  \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\
-  & \; e_{k} \perp\!\!\!\perp e_{i} \\
-  & \; e_{k} \perp\!\!\!\perp e_{r_{i}} \\
-  & \; e_{i} \perp\!\!\!\perp e_{r_{i}}
-\end{aligned}
-}\end{equation}
-
 \subsection{From the scientific to statistical
@@
 
-\begin{equation}\phantomsection\label{eq-StatModel_general}{
-\begin{aligned}
-  O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\
-  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
-  \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\
-  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
-  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
-  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
-  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
-  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
-\end{aligned}
-}\end{equation}
-
-for identification purposes we can set
-\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02\),
-\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1\), and
-\(\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1\). A special case of this
-would be to assume that the data comes from the same population, in that
-case, \(\sigma_{k(g)} = \sigma_{k} = 0.02\),
-\(\sigma_{i(g)} = \sigma_{i} = 1\)
-
 \subsection{Let's talk about Thurstone}\label{sec-theory-thurstone}
-
-Thurstone's comparative judgment \citet{Thurstone_1927} is based on the
-formula: \[
-X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}}
-\]
-
-where \(X_{AB}\) defines the comparative judgment outcome, \(S_{A}\) and
-\(S_{B}\) are the modal discriminal processes,
-\(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}\),
-with \(\sigma_{A}\) and \(\sigma_{B}\) being the dispersion of
-discriminal processes \(A\) and \(B\), respectively, and \(\rho\) the
-correlation between discriminal processes.
-
-The theory identifies five cases:
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Case 1:} only constant \(\rho\) (not \(\rho_{ij}\))
-\item
-  \textbf{Case 2:} \(X_{ij}\) becomes \(X_{kij}\) with \(k=1, \dots, K\)
-  judges (replication, not duplication)
-\item
-  \textbf{Case 3:} \(\rho = 0\), then
-  \(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}\)
-\item
-  \textbf{Case 4:} \(\sigma_{B}=\sigma_{A}+d\), then
-  \(\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}\)
-\item
-  \textbf{Case 5:} \(\sigma_{B}=\sigma_{A}\), then
-  \(\sigma_{AB} = \sqrt{2}\sigma\)
-\end{itemize}
-
-Now using the DAG and statistical notation
-
-\begin{equation}\phantomsection\label{eq-StatModel_thurstone}{
-\begin{aligned}
-  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
-  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
-  \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\
-  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
-  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
-  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
-  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
-  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
-\end{aligned}
-}\end{equation}
-
-The theory identifies five cases:
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Case 1:} only constant \(\rho \approx \sigma_{i}\)
-\item
-  \textbf{Case 2:} now judges are separated by using \(\gamma_{k}\)
-\item
-  \textbf{Case 3:} \(\rho \approx \sigma_{e_{i}} = 0\) (no nesting of
-  texts on students), then
-  \(\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}\)
-\item
-  \textbf{Case 4:} \(\sigma_{B}=\sigma_{A}+d\), then
-  \(\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}\)
-\item
-  \textbf{Case 5:} \(\sigma_{B}=\sigma_{A}\), then
-  \(\sigma_{AB} = \sqrt{2}\sigma\)
-\end{itemize}
-
-But now can we see other scenarios than just those 5 cases?
-
-\begin{itemize}
-\tightlist
-\item
-  consider different \(\rho \approx \sum_{p=1}^{P} \sigma_{p}\),
-  depending on \(P\) nesting structures
-\item
-  we can now investigate \(\gamma_{k}\)
-\item
-  we can assume \(\sigma_{B} \neq \sigma_{A}\), no need for results on
-  the limit
-\end{itemize}
 

### 2024-07-26 index.qmd
section:background; process:write

@@
 <!-- 1. What are DAGs? and how are they represented? -->
-<!-- Graph theory is a branch of mathematics focused on the study of graphs.
-Graphs are mathematical structures that model pairwise relations between objects [@Gross_et_al_2018].
-They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations.
+Graph theory is a branch of mathematics focused on the study of graphs.
+Graphs are mathematical structures that model pairwise relations between objects.
+They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations[@Gross_et_al_2018].
 Graphs have proven useful in various fields, spanning statistics, computer science, operations research, and the natural and social sciences.
@@
 Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of the causal structure of a system [@Neal_2020].
--->
 
-<!-- A graph is a collection of nodes connected by edges.
+A graph is a collection of nodes connected by edges.
 In a directed graph, edges extend from a *parent* node to a *child* node, with arrows indicating the direction of the causal influence.
@@
 In a directed acyclic graph (DAG), the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles [@Neal_2020; @McElreath_2020].
--->
 

### 2024-07-26 _tex/index.tex
Built site for gh-pages

@@
 
+Graph theory is a branch of mathematics focused on the study of graphs.
+Graphs are mathematical structures that model pairwise relations between
+objects. They can represent physical relations, such as electrical
+circuits and roadways, and less tangible structures, such as ecosystems
+and sociological relations\citep{Gross_et_al_2018}. Graphs have proven
+useful in various fields, spanning statistics, computer science,
+operations research, and the natural and social sciences. One
+application in statistics that incorporates concepts from graph theory
+is causal inference. Specifically, the structural approach to causal
+inference uses directed acyclic graphs (DAG) to provide a formal and
+graphical representation of the causal structure of a system
+\citep{Neal_2020}.
+
+A graph is a collection of nodes connected by edges. In a directed
+graph, edges extend from a \emph{parent} node to a \emph{child} node,
+with arrows indicating the direction of the causal influence. Two nodes
+that are connected by edge are said to be \emph{adjacent}
+\citep{Gross_et_al_2018}. In a directed acyclic graph (DAG), the
+direction of causal influences does not loop back on itself, ensuring
+that the graph contains no cycles \citep{Neal_2020, McElreath_2020}.
+
 \begin{figure}

### 2024-07-29 index.qmd
section:background,dags; process:write

--- 2024-07-26 index.qmd
@@
 
-## DAGs, SCMs, and the flow of association and causation {#sec-background-dag}
+## DAGs, SCMs, and the flow of association and causation in graphs {#sec-background-dag}
 
-<!-- 1. What are DAGs? and how are they represented? -->
-<!-- Graph theory is a branch of mathematics focused on the study of graphs.
-Graphs are mathematical structures that model pairwise relations between objects [@Gross_et_al_2018].
+<!-- 1. Graph theory and the use of graphs -->
+Graph theory is a branch of mathematics focused on the study of graphs.
+Graphs are mathematical structures that model pairwise relations between objects.
 They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations.
-Graphs have proven useful in various fields, spanning statistics, computer science, operations research, and the natural and social sciences.
-One application in statistics that incorporates concepts from graph theory is causal inference.
-Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of the causal structure of a system [@Neal_2020].
--->
+Graphs have proven useful in various fields, including computer science, operations research, and the natural and social sciences [@Gross_et_al_2018].
 
-<!-- A graph is a collection of nodes connected by edges.
-In a directed graph, edges extend from a *parent* node to a *child* node, with arrows indicating the direction of the causal influence.
-Two nodes that are connected by edge are said to be *adjacent* [@Gross_et_al_2018].
-In a directed acyclic graph (DAG), the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles [@Neal_2020; @McElreath_2020].
--->
+<!-- 2. How graphs are use in the structural approach? and what are DAGs? -->
+In statistics, one application that incorporates concepts from graph theory is causal inference.
+Specifically, the structural approach to causal inference uses DAGs to provide a formal and graphical representation of a system's causal structure [@Neal_2020].
+In this context, a graph is a collection of nodes connected by edges, where nodes represent random variables.
+In a directed graph, edges extend from one node to another, with arrows indicating the direction of causal influence.
+In a directed acyclic graph, the direction of causal influences does not form a loop, ensuring that causal influences do not cycle back on themselves [@McElreath_2020].
 
-<!-- To draw a DAG use the exclusion restriction assumption: a variable should not appear in the graph if it is not a part of the causal mechanisms that generate the outcome under analysis (see Neal, 2020).
--->
+<!-- 8. What are the DAG's building blocks?
+<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
+
+<!-- 4. A motivating example -->
+Given the heuristic nature of DAGs, these are easier to understand using a motivating example.
+Consider a system where variables $X$ and $Z$ influence a third variable $Y$.
+In this system, it is assumed that $X$ and $Z$ are dependent on their own processes and are, therefore, independent from each other.
+@fig-dag1a presents the plausible causal structure of this system.
+The DAG shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed.
+The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows.
+Moreover, the exogenous variables $E=\{e_{X},e_{Z},e_{Y}\}$ represent everything else that is chosen not to be modeled explicitly.
+These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature.
+Although this DAG explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag1b [@Cinelli_et_al_2020].
+
+<!-- emphasize that everything can be observed -->
+<!-- change the motivating example to include all building blocks  -->
+
+::: {#fig-dags layout-ncol=2}
+
+![Full DAG](images/figures/dag1a.png){#fig-dag1a fig-align="center" width=90%}
+
+![Simplified DAG](images/figures/dag1b.png){#fig-dag1b fig-align="center" width=35%}
+
+DAGs for a plausible causal structure in a system.
+:::
 
 <!-- 2. What does a DAG imply? -->
-<!-- and utilizes the do-operator and Structural Causal Models (SCMs) for mathematical notation in causal inference. -->
 <!-- Structural Causal Model (SCM) -->
@@
 
+<!-- \begin{aligned} -->
+<!--     Y & := \; f_{Y}( X,Z,e_{Y} ) \\ -->
+<!--     X & := \; f_{X}( e_{X} ) \\ -->
+<!--     Z & := \; f_{Z}( e_{Z} ) \\ -->
+<!--     e_{X} & \dsep e_{Z} \\ -->
+<!--     e_{X} & \dsep e_{Y} \\ -->
+<!--     e_{Z} & \dsep e_{Y} -->
+<!-- \end{aligned}  -->
+
 <!-- 3. But what are those mysterious functions?, and how do they become statistical quantities? -->
@@
 <!-- 4. What is the local markov assumption? and where does it lead? -->
+<!-- 4.1 Bayesian factorization -->
 
@@
 
-<!-- 8. What are the DAG's building blocks?, and what do they imply? (SCMs) -->
-<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
+<!-- 9. dependence and independence in building blocks -->
+
+<!-- 9.1 Blocked path definition-->
+
+<!-- 9.2 d-separation -->
 

### 2024-07-29 _tex/index.tex
Built site for gh-pages

--- 2024-07-26 _tex/index.tex
@@
 
-\subsection{DAGs, SCMs, and the flow of association and
-causation}\label{sec-background-dag}
+\subsection{DAGs, SCMs, and the flow of association and causation in
+graphs}\label{sec-background-dag}
+
+Graph theory is a branch of mathematics focused on the study of graphs.
+Graphs are mathematical structures that model pairwise relations between
+objects. They can represent physical relations, such as electrical
+circuits and roadways, and less tangible structures, such as ecosystems
+and sociological relations. Graphs have proven useful in various fields,
+including computer science, operations research, and the natural and
+social sciences \citep{Gross_et_al_2018}.
+
+In statistics, one application that incorporates concepts from graph
+theory is causal inference. Specifically, the structural approach to
+causal inference uses DAGs to provide a formal and graphical
+representation of a system's causal structure \citep{Neal_2020}. In this
+context, a graph is a collection of nodes connected by edges, where
+nodes represent random variables. In a directed graph, edges extend from
+one node to another, with arrows indicating the direction of causal
+influence. In a directed acyclic graph, the direction of causal
+influences does not form a loop, ensuring that causal influences do not
+cycle back on themselves \citep{McElreath_2020}.
+
+Given the heuristic nature of DAGs, these are easier to understand using
+a motivating example. Consider a system where variables \(X\) and \(Z\)
+influence a third variable \(Y\). In this system, it is assumed that
+\(X\) and \(Z\) are dependent on their own processes and are, therefore,
+independent from each other. Figure~\ref{fig-dag1a} presents the
+plausible causal structure of this system. The DAG shows the endogenous
+variables \(V=\{X,Z,Y\}\) as circled black nodes, indicating these
+variables are observed. The arrows connecting the variables indicate the
+direction of causal influence, while a lack of influence is indicated by
+the absence of arrows. Moreover, the exogenous variables
+\(E=\{e_{X},e_{Z},e_{Y}\}\) represent everything else that is chosen not
+to be modeled explicitly. These exogenous variables, or
+\emph{disturbances}, are usually represented by open circles to indicate
+their unobserved nature. Although this DAG explicitly shows the
+exogenous variables, conventionally these are omitted for brevity,
+resulting in an equivalent graph as shown in Figure~\ref{fig-dag1b}.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.9\textwidth,height=\textheight]{images/figures/dag1a.png}
+
+}
+
+\subcaption{\label{fig-dag1a}Full DAG}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/dag1b.png}
+
+}
+
+\subcaption{\label{fig-dag1b}Simplified DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-dags}DAGs for a plausible causal structure in a
+system.}
+
+\end{figure}%
 

### 2024-07-30 index.qmd
section:background,dag; process:write,images

@@
 This study addresses these issues by representing DCJ within the framework of causal inference.
-Specifically, utilizing the structural approach, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
-It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal effects within the system.
+Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
+The study then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal effects within the system.
 This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
@@
 notebook-links: global
+lightbox: true
 citation: true
@@
 This study aims to fill this gap by representing DCJ within the framework of causal inference.
-Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system.
-Next, using a minimal set of assumptions, the study translates the scientific model into a probabilistic statistical model.
+Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
+The study then translates the scientific model into a probabilistic statistical model.
 This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.
@@
 
-# Theoretical background {#sec-background}
+# Theoretical framework {#sec-framework}
 
-## The structural approach to causal inference {#sec-background-structural}
+## The structural approach to causal inference {#sec-framework-structural}
 
@@
 According to @Pearl_et_al_2018, counterfactuals form the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association.
-Despite this abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
+Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
 They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."
@@
 
-## DAGs, SCMs, and the flow of association and causation in graphs {#sec-background-dag}
+## Directed acyclic graphs (DAG) {#sec-framework-dag}
 
@@
 In statistics, one application that incorporates concepts from graph theory is causal inference.
-Specifically, the structural approach to causal inference uses DAGs to provide a formal and graphical representation of a system's causal structure [@Neal_2020].
-In this context, a graph is a collection of nodes connected by edges, where nodes represent random variables.
-In a directed graph, edges extend from one node to another, with arrows indicating the direction of causal influence.
-In a directed acyclic graph, the direction of causal influences does not form a loop, ensuring that causal influences do not cycle back on themselves [@McElreath_2020].
+Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of a system's causal structure [@Neal_2020].
+In this context, a *graph* is a collection of nodes connected by edges, where nodes represent random variables.
+The term *directed* indicates that the edges of the graph extend from a *parent* node to a *child* node, with arrows showing the direction of causal influence.
+The term *acyclic* indicates that the direction of causal influences does not form a loop, ensuring that the influences do not cycle back on themselves [@McElreath_2020].
 
-<!-- 8. What are the DAG's building blocks?
-<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
+<!-- 3. How do we draw DAGs? -->
+A key advantage of DAGs is that, regardless of the complexity of the causal structure, these structures can be composed using a minimal set of building blocks.
+Moreover, representing any complex structure requires only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+These advantages imply that a researcher seeking to depict and understand the causal assumptions of a complex system only needs to decompose the structure into these basic building blocks and then analyze the causal assumptions associated with each block [@McElreath_2020].
+
+<!-- 4. What are DAG's building blocks?  -->
+@fig-dags presents the five fundamental building blocks of any DAG.
+@fig-dag_bb1 illustrates two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related.
+@fig-dag_bb2 shows two connected nodes, depicting an scenario where a *parent* node $X_{1}$ has a causal influence on a *child* node $X_{2}$.
+This implies that variable $X_{2}$ is a *descendant* of $X_{1}$.
+@fig-dag_bb3 illustrates a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
+This DAG indicates that $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ is a parent node $X_{3}$.
+Furthermore, it shows that $X_{1}$ is an *ancestor* of $X_{3}$, with the relationship between the two variables being mediated entirely by $X_{2}$.
+@fig-dag_bb4 depicts a *fork*, illustrating an scenario where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
+In this scenario $X_{2}$ is a parent node for both $X_{1}$ and $X_{3}$.
+Finally, @fig-dag_bb5 illustrates a *collider* (also known as *inmorality*), where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
+In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.
+Given the heuristic nature of DAGs, the use of fundamental building block to construct a causal structure of a system is easier to understand using a motivating example.
+The motivating example can also serve to point out about other conventions when using DAGs.
+
+::: {#fig-dags layout="[[1,1],[1,1,1]]"}
+
+![Two unconnected nodes](images/figures/dag_bb1.png){#fig-dag_bb1 fig-align="center" width=40%}
+
+![Two connected nodes or descendant](images/figures/dag_bb2.png){#fig-dag_bb2 fig-align="center" width=40%}
+
+![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=100%}
+
+![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=100%}
+
+![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=100%}
+
+DAG's fundamental building blocks.
+:::
 
 <!-- 4. A motivating example -->
-Given the heuristic nature of DAGs, these are easier to understand using a motivating example.
 Consider a system where variables $X$ and $Z$ influence a third variable $Y$.
 In this system, it is assumed that $X$ and $Z$ are dependent on their own processes and are, therefore, independent from each other.
-@fig-dag1a presents the plausible causal structure of this system.
+@fig-dag_example1 presents the plausible causal structure of this system.
 The DAG shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed.
@@
 These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature.
-Although this DAG explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag1b [@Cinelli_et_al_2020].
+Although this DAG explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020].
 
@@
 
-::: {#fig-dags layout-ncol=2}
+::: {#fig-example layout-ncol=2}
 
-![Full DAG](images/figures/dag1a.png){#fig-dag1a fig-align="center" width=90%}
+![Full DAG](images/figures/dag_example1.png){#fig-dag_example1 fig-align="center" width=100%}
 
-![Simplified DAG](images/figures/dag1b.png){#fig-dag1b fig-align="center" width=35%}
+![Simplified DAG](images/figures/dag_example2.png){#fig-dag_example2 fig-align="center" width=40%}
 
@@
 :::
+
+## The flow of association and causation in graphs {#sec-framework-flow}
 
@@
 
+<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
+
 ## But where does it all fit? {#sec-background-where}
@@
 
-# Theoretical framework {#sec-theory}
+# Theory {#sec-theory}
 

### 2024-07-30 _tex/index.tex
Built site for gh-pages

--- 2024-07-26 _tex/index.tex
@@
 the structural approach, the study develops a scientific model to
-clarify the causal assumptions and mechanisms inherent in the DCJ
-system. It then translates this model into a probabilistic statistical
-framework to estimate statistical relationships and infer causal effects
-within the system. This research provides a robust probabilistic
-foundation for the statistical analysis of DCJ data, building upon
-Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
-Its findings offer valuable insights for researchers and analysts
-designing and implementing DCJ experiments.
+clarify plausible causal assumptions and mechanisms inherent in the DCJ
+system. The study then translates this model into a probabilistic
+statistical framework to estimate statistical relationships and infer
+causal effects within the system. This research provides a robust
+probabilistic foundation for the statistical analysis of DCJ data,
+building upon Thurstone's law of comparative judgment
+\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
+researchers and analysts designing and implementing DCJ experiments.
 \end{abstract}
@@
 approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
-develops a scientific model to clarify the causal assumptions and
-mechanisms inherent in the DCJ system. Next, using a minimal set of
-assumptions, the study translates the scientific model into a
-probabilistic statistical model. This model aims to produce statistical
-estimates to draw inferences about plausible causal relationships within
-the DCJ system.
+develops a scientific model to clarify plausible causal assumptions and
+mechanisms inherent in the DCJ system. The study then translates the
+scientific model into a probabilistic statistical model. This model aims
+to produce statistical estimates to draw inferences about plausible
+causal relationships within the DCJ system.
 
@@
 
-\section{Theoretical background}\label{sec-background}
+\section{Theoretical framework}\label{sec-framework}
 
 \subsection{The structural approach to causal
-inference}\label{sec-background-structural}
+inference}\label{sec-framework-structural}
 
@@
 abstraction in the ladder of causation, followed by intervention and
-association. Despite this abstract nature, counterfactuals enable the
-development of a \emph{theory of the world} that explains why specific
-causes have specific effects and what occurs in their absence
-\citep{Pearl_et_al_2018}. They achieve this by translating causal
-statements into counterfactual statements, that is, statements about
-``what would have happened in the world under different circumstances.''
+association. Nevertheless, despite their abstract nature,
+counterfactuals enable the development of a \emph{theory of the world}
+that explains why specific causes have specific effects and what occurs
+in their absence \citep{Pearl_et_al_2018}. They achieve this by
+translating causal statements into counterfactual statements, that is,
+statements about ``what would have happened in the world under different
+circumstances.''
 
@@
 
-\subsection{DAGs, SCMs, and the flow of association and
-causation}\label{sec-background-dag}
+\subsection{Directed acyclic graphs (DAG)}\label{sec-framework-dag}
+
+Graph theory is a branch of mathematics focused on the study of graphs.
+Graphs are mathematical structures that model pairwise relations between
+objects. They can represent physical relations, such as electrical
+circuits and roadways, and less tangible structures, such as ecosystems
+and sociological relations. Graphs have proven useful in various fields,
+including computer science, operations research, and the natural and
+social sciences \citep{Gross_et_al_2018}.
+
+In statistics, one application that incorporates concepts from graph
+theory is causal inference. Specifically, the structural approach to
+causal inference uses directed acyclic graphs (DAG) to provide a formal
+and graphical representation of a system's causal structure
+\citep{Neal_2020}. In this context, a \emph{graph} is a collection of
+nodes connected by edges, where nodes represent random variables. The
+term \emph{directed} indicates that the edges of the graph extend from a
+\emph{parent} node to a \emph{child} node, with arrows showing the
+direction of causal influence. The term \emph{acyclic} indicates that
+the direction of causal influences does not form a loop, ensuring that
+the influences do not cycle back on themselves \citep{McElreath_2020}.
+
+A key advantage of DAGs is that, regardless of the complexity of the
+causal structure, these structures can be composed using a minimal set
+of building blocks. Moreover, representing any complex structure
+requires only five fundamental building blocks
+\citep{Neal_2020, McElreath_2020}. These advantages imply that a
+researcher seeking to depict and understand the causal assumptions of a
+complex system only needs to decompose the structure into these basic
+building blocks and then analyze the causal assumptions associated with
+each block \citep{McElreath_2020}.
+
+Figure~\ref{fig-dags} presents the five fundamental building blocks of
+any DAG. Figure~\ref{fig-dag_bb1} illustrates two unconnected nodes,
+representing an scenario where variables \(X_{1}\) and \(X_{2}\) are not
+causally related. Figure~\ref{fig-dag_bb2} shows two connected nodes,
+depicting an scenario where a \emph{parent} node \(X_{1}\) has a causal
+influence on a \emph{child} node \(X_{2}\). This implies that variable
+\(X_{2}\) is a \emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3}
+illustrates a \emph{chain} (or \emph{pipe}), where \(X_{1}\) influences
+\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). This DAG indicates that
+\(X_{1}\) is a parent node of \(X_{2}\), and \(X_{2}\) is a parent node
+\(X_{3}\). Furthermore, it shows that \(X_{1}\) is an \emph{ancestor} of
+\(X_{3}\), with the relationship between the two variables being
+mediated entirely by \(X_{2}\). Figure~\ref{fig-dag_bb4} depicts a
+\emph{fork}, illustrating an scenario where variables \(X_{1}\) and
+\(X_{3}\) are both influenced by \(X_{2}\). In this scenario \(X_{2}\)
+is a parent node for both \(X_{1}\) and \(X_{3}\). Finally,
+Figure~\ref{fig-dag_bb5} illustrates a \emph{collider} (also known as
+\emph{inmorality}), where variables \(X_{1}\) and \(X_{3}\) are
+concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
+\(X_{3}\) are not causally related to each other but both influence
+\(X_{2}\). Given the heuristic nature of DAGs, the use of fundamental
+building block to construct a causal structure of a system is easier to
+understand using a motivating example. The motivating example can also
+serve to point out about other conventions when using DAGs.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_bb1.png}
+
+}
+
+\subcaption{\label{fig-dag_bb1}Two unconnected nodes}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_bb2.png}
+
+}
+
+\subcaption{\label{fig-dag_bb2}Two connected nodes or descendant}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb3.png}
+
+}
+
+\subcaption{\label{fig-dag_bb3}Chain or pipe}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb4.png}
+
+}
+
+\subcaption{\label{fig-dag_bb4}Fork}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb5.png}
+
+}
+
+\subcaption{\label{fig-dag_bb5}Collider or inmorality}
+
+\end{minipage}%
+
+\caption{\label{fig-dags}DAG's fundamental building blocks.}
+
+\end{figure}%
+
+Consider a system where variables \(X\) and \(Z\) influence a third
+variable \(Y\). In this system, it is assumed that \(X\) and \(Z\) are
+dependent on their own processes and are, therefore, independent from
+each other. Figure~\ref{fig-dag_example1} presents the plausible causal
+structure of this system. The DAG shows the endogenous variables
+\(V=\{X,Z,Y\}\) as circled black nodes, indicating these variables are
+observed. The arrows connecting the variables indicate the direction of
+causal influence, while a lack of influence is indicated by the absence
+of arrows. Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\)
+represent everything else that is chosen not to be modeled explicitly.
+These exogenous variables, or \emph{disturbances}, are usually
+represented by open circles to indicate their unobserved nature.
+Although this DAG explicitly shows the exogenous variables,
+conventionally these are omitted for brevity, resulting in an equivalent
+graph as shown in Figure~\ref{fig-dag_example2}.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_example1.png}
+
+}
+
+\subcaption{\label{fig-dag_example1}Full DAG}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_example2.png}
+
+}
+
+\subcaption{\label{fig-dag_example2}Simplified DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-example}DAGs for a plausible causal structure in a
+system.}
+
+\end{figure}%
+
+\subsection{The flow of association and causation in
+graphs}\label{sec-framework-flow}
 
@@
 
-\section{Theoretical framework}\label{sec-theory}
+\section{Theory}\label{sec-theory}
 

### 2024-07-30 index.qmd
section:background,dag; process:write,images

@@
 <!-- 2. How graphs are use in the structural approach? and what are DAGs? -->
-In statistics, one application that incorporates concepts from graph theory is causal inference.
-Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of a system's causal structure [@Neal_2020].
+In statistics, one application incorporating concepts from graph theory is causal inference.
+Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of the causal structure of a system [@Neal_2020].
 In this context, a *graph* is a collection of nodes connected by edges, where nodes represent random variables.
-The term *directed* indicates that the edges of the graph extend from a *parent* node to a *child* node, with arrows showing the direction of causal influence.
-The term *acyclic* indicates that the direction of causal influences does not form a loop, ensuring that the influences do not cycle back on themselves [@McElreath_2020].
+The term *directed* indicates that the edges of the graph extend from one node to another, with arrows showing the direction of causal influence.
+Moreover, the term *acyclic* indicates the causal influences do not form a loop, meaning the influences do not cycle back on themselves [@McElreath_2020].
 
-<!-- 3. How do we draw DAGs? -->
-A key advantage of DAGs is that, regardless of the complexity of the causal structure, these structures can be composed using a minimal set of building blocks.
-Moreover, representing any complex structure requires only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-These advantages imply that a researcher seeking to depict and understand the causal assumptions of a complex system only needs to decompose the structure into these basic building blocks and then analyze the causal assumptions associated with each block [@McElreath_2020].
-
-<!-- 4. What are DAG's building blocks?  -->
-@fig-dags presents the five fundamental building blocks of any DAG.
-@fig-dag_bb1 illustrates two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related.
-@fig-dag_bb2 shows two connected nodes, depicting an scenario where a *parent* node $X_{1}$ has a causal influence on a *child* node $X_{2}$.
-This implies that variable $X_{2}$ is a *descendant* of $X_{1}$.
-@fig-dag_bb3 illustrates a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
-This DAG indicates that $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ is a parent node $X_{3}$.
-Furthermore, it shows that $X_{1}$ is an *ancestor* of $X_{3}$, with the relationship between the two variables being mediated entirely by $X_{2}$.
-@fig-dag_bb4 depicts a *fork*, illustrating an scenario where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
-In this scenario $X_{2}$ is a parent node for both $X_{1}$ and $X_{3}$.
-Finally, @fig-dag_bb5 illustrates a *collider* (also known as *inmorality*), where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
+<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
+Regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+Each panel of @fig-dags illustrates these building blocks.
+@fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related.
+@fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$.
+Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$.
+@fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
+In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$.
+Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$, and that the relationship between these variables is entirely mediated by $X_{2}$.
+@fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
+In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$.
+Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
 In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.
-Given the heuristic nature of DAGs, the use of fundamental building block to construct a causal structure of a system is easier to understand using a motivating example.
-The motivating example can also serve to point out about other conventions when using DAGs.
 
@@
 <!-- 4. A motivating example -->
+Given the heuristic nature of DAGs, the use of fundamental building block to construct a causal structure of a system is easier to understand using a motivating example.
+The motivating example can also serve to signal about other conventions when using DAGs.
 Consider a system where variables $X$ and $Z$ influence a third variable $Y$.
@@
 <!-- 9.2 d-separation -->
+
+<!-- This further implies that to understand the causal assumptions of a complex system, one can decompose the structure into these basic building blocks and analyze the causal assumptions associated with each block individually [@McElreath_2020].
+-->
 

### 2024-07-30 _tex/index.tex
Built site for gh-pages

@@
 
-In statistics, one application that incorporates concepts from graph
-theory is causal inference. Specifically, the structural approach to
-causal inference uses directed acyclic graphs (DAG) to provide a formal
-and graphical representation of a system's causal structure
+In statistics, one application incorporating concepts from graph theory
+is causal inference. Specifically, the structural approach to causal
+inference uses directed acyclic graphs (DAG) to provide a formal and
+graphical representation of the causal structure of a system
 \citep{Neal_2020}. In this context, a \emph{graph} is a collection of
 nodes connected by edges, where nodes represent random variables. The
-term \emph{directed} indicates that the edges of the graph extend from a
-\emph{parent} node to a \emph{child} node, with arrows showing the
-direction of causal influence. The term \emph{acyclic} indicates that
-the direction of causal influences does not form a loop, ensuring that
-the influences do not cycle back on themselves \citep{McElreath_2020}.
+term \emph{directed} indicates that the edges of the graph extend from
+one node to another, with arrows showing the direction of causal
+influence. Moreover, the term \emph{acyclic} indicates the causal
+influences do not form a loop, meaning the influences do not cycle back
+on themselves \citep{McElreath_2020}.
 
-A key advantage of DAGs is that, regardless of the complexity of the
-causal structure, these structures can be composed using a minimal set
-of building blocks. Moreover, representing any complex structure
-requires only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. These advantages imply that a
-researcher seeking to depict and understand the causal assumptions of a
-complex system only needs to decompose the structure into these basic
-building blocks and then analyze the causal assumptions associated with
-each block \citep{McElreath_2020}.
-
-Figure~\ref{fig-dags} presents the five fundamental building blocks of
-any DAG. Figure~\ref{fig-dag_bb1} illustrates two unconnected nodes,
-representing an scenario where variables \(X_{1}\) and \(X_{2}\) are not
-causally related. Figure~\ref{fig-dag_bb2} shows two connected nodes,
-depicting an scenario where a \emph{parent} node \(X_{1}\) has a causal
-influence on a \emph{child} node \(X_{2}\). This implies that variable
-\(X_{2}\) is a \emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3}
-illustrates a \emph{chain} (or \emph{pipe}), where \(X_{1}\) influences
-\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). This DAG indicates that
-\(X_{1}\) is a parent node of \(X_{2}\), and \(X_{2}\) is a parent node
-\(X_{3}\). Furthermore, it shows that \(X_{1}\) is an \emph{ancestor} of
-\(X_{3}\), with the relationship between the two variables being
-mediated entirely by \(X_{2}\). Figure~\ref{fig-dag_bb4} depicts a
-\emph{fork}, illustrating an scenario where variables \(X_{1}\) and
-\(X_{3}\) are both influenced by \(X_{2}\). In this scenario \(X_{2}\)
-is a parent node for both \(X_{1}\) and \(X_{3}\). Finally,
-Figure~\ref{fig-dag_bb5} illustrates a \emph{collider} (also known as
-\emph{inmorality}), where variables \(X_{1}\) and \(X_{3}\) are
+Regardless of complexity, DAGs can represent various causal structures
+using only five fundamental building blocks
+\citep{Neal_2020, McElreath_2020}. Each panel of Figure~\ref{fig-dags}
+illustrates these building blocks. Figure~\ref{fig-dag_bb1} depicts two
+unconnected nodes, representing an scenario where variables \(X_{1}\)
+and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
+two connected nodes, illustrating a scenario where a parent node
+\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
+Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
+Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
+\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
+this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
+\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
+\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and that the relationship
+between these variables is entirely mediated by \(X_{2}\).
+Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
+\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
+scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
+Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
+as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
 concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
 \(X_{3}\) are not causally related to each other but both influence
-\(X_{2}\). Given the heuristic nature of DAGs, the use of fundamental
-building block to construct a causal structure of a system is easier to
-understand using a motivating example. The motivating example can also
-serve to point out about other conventions when using DAGs.
+\(X_{2}\).
 
@@
 
-Consider a system where variables \(X\) and \(Z\) influence a third
-variable \(Y\). In this system, it is assumed that \(X\) and \(Z\) are
-dependent on their own processes and are, therefore, independent from
-each other. Figure~\ref{fig-dag_example1} presents the plausible causal
-structure of this system. The DAG shows the endogenous variables
-\(V=\{X,Z,Y\}\) as circled black nodes, indicating these variables are
-observed. The arrows connecting the variables indicate the direction of
-causal influence, while a lack of influence is indicated by the absence
-of arrows. Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\)
-represent everything else that is chosen not to be modeled explicitly.
-These exogenous variables, or \emph{disturbances}, are usually
-represented by open circles to indicate their unobserved nature.
-Although this DAG explicitly shows the exogenous variables,
-conventionally these are omitted for brevity, resulting in an equivalent
-graph as shown in Figure~\ref{fig-dag_example2}.
+Given the heuristic nature of DAGs, the use of fundamental building
+block to construct a causal structure of a system is easier to
+understand using a motivating example. The motivating example can also
+serve to signal about other conventions when using DAGs. Consider a
+system where variables \(X\) and \(Z\) influence a third variable \(Y\).
+In this system, it is assumed that \(X\) and \(Z\) are dependent on
+their own processes and are, therefore, independent from each other.
+Figure~\ref{fig-dag_example1} presents the plausible causal structure of
+this system. The DAG shows the endogenous variables \(V=\{X,Z,Y\}\) as
+circled black nodes, indicating these variables are observed. The arrows
+connecting the variables indicate the direction of causal influence,
+while a lack of influence is indicated by the absence of arrows.
+Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\) represent
+everything else that is chosen not to be modeled explicitly. These
+exogenous variables, or \emph{disturbances}, are usually represented by
+open circles to indicate their unobserved nature. Although this DAG
+explicitly shows the exogenous variables, conventionally these are
+omitted for brevity, resulting in an equivalent graph as shown in
+Figure~\ref{fig-dag_example2}.
 

### 2024-07-31 index.qmd
section:background,dag; process:write,figures

@@
 keywords:
+  - causal inference
+  - probability
+  - Thurstone
   - comparative judgement
-  - directed acycilc graph
-  - causal analysis
-  - probabilistic statistics
+  - directed acyclic graph
+  - structural causal models
+  - statistical modeling
 abstract: |
@@
 Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
-The study then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal effects within the system.
+The study then translates this model into a probabilistic statistical model to estimate statistical relationships and infer causal effects within the system.
 This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
@@
 
+This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs.
+It is not a comprehensive description of causal inference methods.
+Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020.
+They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful.
+For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.
+
 ## The structural approach to causal inference {#sec-framework-structural}
@@
 <!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
-Regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-Each panel of @fig-dags illustrates these building blocks.
+One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+@fig-dags illustrates these building blocks.
 @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related.
@@
 In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$.
-Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$, and that the relationship between these variables is entirely mediated by $X_{2}$.
+Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely mediated by $X_{2}$.
 @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
@@
 <!-- 4. A motivating example -->
-Given the heuristic nature of DAGs, the use of fundamental building block to construct a causal structure of a system is easier to understand using a motivating example.
-The motivating example can also serve to signal about other conventions when using DAGs.
-Consider a system where variables $X$ and $Z$ influence a third variable $Y$.
-In this system, it is assumed that $X$ and $Z$ are dependent on their own processes and are, therefore, independent from each other.
-@fig-dag_example1 presents the plausible causal structure of this system.
-The DAG shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed.
+Given the heuristic nature of DAGs, a motivating example can illustrate how these five fundamental building blocks can be used to construct the causal structure of a system.
+This example can also help clarify additional rules and conventions for drawing DAGs.
+
+Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs to be investigated.
+Additionally, the problem indicates that a variable $X$ potentially influences both $Y$ and $T$.
+Such scenarios are not hard to imagine.
+For instance, $T$ could describe different treatments potentially influencing the binary onset of cancer ($Y$) while $X$ could denote the age of a patient.
+Using an example from DCJ, [we need an example]{style="color:red;"}.
+@fig-example presents two graphs describing the plausible causal structure of this system, the *simplified* and the *magnified* DAG.
+
+<!-- 5. DAG for the motivating example -->
+<!-- @fig-dag_example1 shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed.
 The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows.
@@
 These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature.
-Although this DAG explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020].
+Although this DAG, also known as Magnified DAG, explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020].
+Sometimes it is convenient to show magnified graphs to show porblems in conditioning.
+-->
 
 <!-- emphasize that everything can be observed -->
-<!-- change the motivating example to include all building blocks  -->
+<!-- point out that the disturbances are usually stochastic in nature -->
 
@@
 
-![Full DAG](images/figures/dag_example1.png){#fig-dag_example1 fig-align="center" width=100%}
+![Simplied DAG](images/figures/dag_example1.png){#fig-dag_example1 fig-align="center" width=45%}
 
-![Simplified DAG](images/figures/dag_example2.png){#fig-dag_example2 fig-align="center" width=40%}
+![Magnified DAG](images/figures/dag_example2.png){#fig-dag_example2 fig-align="center" width=85%}
 
@@
 
+<!-- (Structural Causal Model (SCM)) A structural causal model is a tuple of the following sets: -->
+<!-- 1. A set ofendogenous variables 푉 2. A set ofexogenous variables 푈 3. A set of functions 푓, one to generate each endogenous variable as a function ofother variables Neal -->
+
 <!-- \begin{aligned} -->
@@
 <!-- \end{aligned}  -->
+
+<!-- these are non-parametrical structural equations -->
+
+<!-- Causal mechanism as a result of markov and minimality assumption -->
+<!-- Modularity assumption also -->
 
@@
 
+<!-- Mention convenient R package daggitty to draw dags and extract the causal assumption inherent in a graph -->
+
+<!-- After detailing how to build graphs and what they imply, Include steps to draw dags using a motivating example -->
+
 ## But where does it all fit? {#sec-background-where}
+
+<!-- The previous chapter gives graphical intuition for causal models, but it doesn’t explain how to identify causal quantities and formalize causal models. Neal -->
 
@@
 ![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+
+<!-- Ardmed with dags and scm, and theit probabilité implication, we can now expand on thé motivating example provided in section xx -->
 
@@
 
+<!-- no fundamental ontological objections to using DAGs to describe causal models and causal relationships. -->
+
 ## Limitations and further research {#sec-discuss-limitations}
@@
 
-## Additional definitions {#sec-appA}
-
-## Why do we need to estimate judges' abilities? {#sec-appB}
+## Why do we need to estimate judges' abilities? {#sec-appA}
 
 <!-- The same idea as interactions: when interactions are significant, the interpretations of the parameters cannot be made in isolation -->
+
+## Latent variables as a mean of imputation {#sec-appB}
+
+<!-- Latent variables and random effects are in essence imputation? (McElreath, p. 504) -->
+<!-- Judge index works as a descendant of the latent variables, or the latent variable is not a reflective, but generative where the index is multiplied by a loading? -->
+<!-- see Morgan et al (2014) -->
+<!-- see Mayer (2019) -->
+
+## Other comparative scenarios  {#sec-appC}
 

### 2024-07-31 _tex/index.tex
Built site for gh-pages

@@
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
-  pdfkeywords={comparative judgement, directed acycilc graph, causal
-analysis, probabilistic statistics},
+  pdfkeywords={causal inference, probability, Thurstone, comparative
+judgement, directed acyclic graph, structural causal models, statistical
+modeling},
   colorlinks=true,
@@
 system. The study then translates this model into a probabilistic
-statistical framework to estimate statistical relationships and infer
-causal effects within the system. This research provides a robust
-probabilistic foundation for the statistical analysis of DCJ data,
-building upon Thurstone's law of comparative judgment
-\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
-researchers and analysts designing and implementing DCJ experiments.
+statistical model to estimate statistical relationships and infer causal
+effects within the system. This research provides a robust probabilistic
+foundation for the statistical analysis of DCJ data, building upon
+Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
+Its findings offer valuable insights for researchers and analysts
+designing and implementing DCJ experiments.
 \end{abstract}
@@
 \begin{keyword}
-    comparative judgement \sep directed acycilc graph \sep causal
-analysis \sep 
-    probabilistic statistics
+    causal inference \sep probability \sep Thurstone \sep comparative
+judgement \sep directed acyclic graph \sep structural causal
+models \sep 
+    statistical modeling
 \end{keyword}
@@
 
+This section introduces fundamental concepts in causal inference, such
+as directed acyclic graphs (DAGs), structural causal models (SCMs), and
+the flow of association and causation in graphs. It is not a
+comprehensive description of causal inference methods. Readers
+interested in deeper exploration should consult introductory papers like
+\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
+\citet{Cinelli_et_al_2020}. They may also find introductory books such
+as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
+\citet{McElreath_2020} useful. For more advanced study, seminal
+intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
+\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
+such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
+\citet{Hernan_et_al_2020} are recommended.
+
 \subsection{The structural approach to causal
@@
 
-Regardless of complexity, DAGs can represent various causal structures
-using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. Each panel of Figure~\ref{fig-dags}
+One key advantage of DAGs is that, regardless of complexity, they can
+represent various causal structures using only five fundamental building
+blocks \citep{Neal_2020, McElreath_2020}. Figure~\ref{fig-dags}
 illustrates these building blocks. Figure~\ref{fig-dag_bb1} depicts two
@@
 \(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
-\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and that the relationship
+\(X_{1}\) is an \emph{ancestor} of \(X_{3}\) and that the relationship
 between these variables is entirely mediated by \(X_{2}\).
@@
 
-Given the heuristic nature of DAGs, the use of fundamental building
-block to construct a causal structure of a system is easier to
-understand using a motivating example. The motivating example can also
-serve to signal about other conventions when using DAGs. Consider a
-system where variables \(X\) and \(Z\) influence a third variable \(Y\).
-In this system, it is assumed that \(X\) and \(Z\) are dependent on
-their own processes and are, therefore, independent from each other.
-Figure~\ref{fig-dag_example1} presents the plausible causal structure of
-this system. The DAG shows the endogenous variables \(V=\{X,Z,Y\}\) as
-circled black nodes, indicating these variables are observed. The arrows
-connecting the variables indicate the direction of causal influence,
-while a lack of influence is indicated by the absence of arrows.
-Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\) represent
-everything else that is chosen not to be modeled explicitly. These
-exogenous variables, or \emph{disturbances}, are usually represented by
-open circles to indicate their unobserved nature. Although this DAG
-explicitly shows the exogenous variables, conventionally these are
-omitted for brevity, resulting in an equivalent graph as shown in
-Figure~\ref{fig-dag_example2}.
+Given the heuristic nature of DAGs, a motivating example can illustrate
+how these five fundamental building blocks can be used to construct the
+causal structure of a system. This example can also help clarify
+additional rules and conventions for drawing DAGs.
+
+Consider a research problem where the causal effect of a variable \(T\)
+on an outcome \(Y\) needs to be investigated. Additionally, the problem
+indicates that a variable \(X\) potentially influences both \(Y\) and
+\(T\). Such scenarios are not hard to imagine. For instance, \(T\) could
+describe different treatments potentially influencing the binary onset
+of cancer (\(Y\)) while \(X\) could denote the age of a patient. Using
+an example from DCJ, {we need an example}. Figure~\ref{fig-example}
+presents two graphs describing the plausible causal structure of this
+system, the \emph{simplified} and the \emph{magnified} DAG.
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_example1.png}
+\includegraphics[width=0.45\textwidth,height=\textheight]{images/figures/dag_example1.png}
 
@@
 
-\subcaption{\label{fig-dag_example1}Full DAG}
+\subcaption{\label{fig-dag_example1}Simplied DAG}
 
@@
 
-\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_example2.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_example2.png}
 
@@
 
-\subcaption{\label{fig-dag_example2}Simplified DAG}
+\subcaption{\label{fig-dag_example2}Magnified DAG}
 
@@
 
-\subsection{Additional definitions}\label{sec-appA}
+\subsection{Why do we need to estimate judges'
+abilities?}\label{sec-appA}
 
-\subsection{Why do we need to estimate judges'
-abilities?}\label{sec-appB}
+\subsection{Latent variables as a mean of imputation}\label{sec-appB}
+
+\subsection{Other comparative scenarios}\label{sec-appC}
 

### 2024-08-01 index.qmd
section:background,dag; process:write

@@
 ---
-title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism"
+title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism"
 author:
@@
 The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992].
-Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020].
+Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020; @Bouwer_et_al_2023].
 Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data could be generated.
@@
 <!-- 4. Solution to the gap -->
-Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome.
+Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021; @Bouwer_et_al_2023], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome.
 This study aims to fill this gap by representing DCJ within the framework of causal inference.
@@
 <!-- 5. final objective and relevance -->
-Ultimately, this research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
+Ultimately, this study provides a robust causal and probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
 Consequently, its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.
@@
 This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs.
-It is not a comprehensive description of causal inference methods.
+The section is not a comprehensive description of causal inference methods.
 Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020.
@@
 Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-According to @Pearl_et_al_2018, counterfactuals form the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association.
+According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association.
 Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
@@
 Graph theory is a branch of mathematics focused on the study of graphs.
-Graphs are mathematical structures that model pairwise relations between objects.
+Graphs are mathematical structures modeling pairwise relations between objects.
 They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations.
@@
 One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-@fig-dags illustrates these building blocks.
+@fig-dags illustrates these blocks.
 @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related.
@@
 In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$.
-Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely mediated by $X_{2}$.
+Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely *mediated* by $X_{2}$.
 @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
@@
 <!-- 4. A motivating example -->
-Given the heuristic nature of DAGs, a motivating example can illustrate how these five fundamental building blocks can be used to construct the causal structure of a system.
-This example can also help clarify additional rules and conventions for drawing DAGs.
+Given the heuristic nature of DAGs, a motivating example can help illustrate the use of the five fundamental building blocks to construct a system's causal structure.
+This example can also help to clarify additional conventions for drawing DAGs.
 
 Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs to be investigated.
-Additionally, the problem indicates that a variable $X$ potentially influences both $Y$ and $T$.
+Additionally, the problem suggests that a variable $X$ potentially influences both $T$ and $Y$.
 Such scenarios are not hard to imagine.
-For instance, $T$ could describe different treatments potentially influencing the binary onset of cancer ($Y$) while $X$ could denote the age of a patient.
-Using an example from DCJ, [we need an example]{style="color:red;"}.
-@fig-example presents two graphs describing the plausible causal structure of this system, the *simplified* and the *magnified* DAG.
+For instance, $T$ might represent different treatments that could affect the recovery from cancer $Y$, while $X$ could denote the patient's age.
+Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status.
+Note that these descriptions do not specify the type of data or the functional forms of the variables, highlighting the heuristic nature of DAGs.
+[(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}
 
 <!-- 5. DAG for the motivating example -->
-<!-- @fig-dag_example1 shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed.
-The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows.
-Moreover, the exogenous variables $E=\{e_{X},e_{Z},e_{Y}\}$ represent everything else that is chosen not to be modeled explicitly.
-These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature.
-Although this DAG, also known as Magnified DAG, explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020].
-Sometimes it is convenient to show magnified graphs to show porblems in conditioning.
--->
+@fig-example presents two graphs illustrating the plausible causal structure of the motivating example.
+@fig-dag_example1 shows the *simplified* DAG.
+In this graph, the variables of interest or *endogenous* variables $V=\{T,X,Y\}$ are depicted as solid black circles, indicating that they are observed variables.
+The arrows in the graph reflect the expected direction of causal influences among the variables.
+@fig-dag_example2, in contrast, depicts the *magnified* version of the DAG.
+This graph includes, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{T},e_{X},e_{Y}\}$.
+These exogenous variables are commonly referred to as *disturbances* or *errors*, because they represent factors not explicitly modeled.
+They are depicted as open circles, indicating their unobserved nature.
+While such exogenous variables are often omitted for simplicity, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts that will be discussed in the next section.
 
-<!-- emphasize that everything can be observed -->
-<!-- point out that the disturbances are usually stochastic in nature -->
+<!-- 6. But where are the building blocks? -->
+<!-- commands for d-separation -->
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+
+Furthermore, a detailed examination of both figures reveals the presence of at least four of the five fundamental building blocks.
+@fig-dag_example1 displays multiple descendants, evident in pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+It also illustrates a fork with $X \rightarrow \{T,Y\}$.
+Similarly, @fig-dag_example2 features multiple two-unconnected nodes, evident in the pairwise relations $\{e_{T} \dsep e_{X}\}$, $\{e_{T} \dsep e_{Y}\}$, and $\{e_{X} \dsep e_{Y}\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
+The symbol $\dsep$ denotes a concept know as *d-separation*, which roughly implies the independence of the variables.
+This concept will be introduced in the next section.
 
@@
 
+<!-- point out that the disturbances are usually stochastic in nature -->
+
 ## But where does it all fit? {#sec-background-where}
@@
 <!-- what does it mean in probabilistic terms? describe probabilities and arrive to structural model -->
-
-<!-- commands for d-separation -->
-<!-- \newcommand{\dsep}{\perp\!\!\!\perp} -->
-<!-- \newcommand{\ndsep}{\not\!\perp\!\!\!\perp} -->
 

### 2024-08-01 _tex/index.tex
Built site for gh-pages

@@
 \hypersetup{
-  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism},
+  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism},
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
@@
 \title{Causes and effects in Dichotomous Comparative Judgments: an
-information-theoretical system with plausible mechanism}
+information-theoretical system of plausible mechanism}
 \author[1]{Jose Manuel Rivera Espejo%
@@
 measurement and evaluation, characterized by its reliability and
-validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
+validity
+\citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020, Bouwer_et_al_2023}.
 Nevertheless, despite the method's widespread use, existing literature
@@
 factors on the method's reliability
-\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
+\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021, Bouwer_et_al_2023},
 none, to the best of the authors' knowledge, have provided a transparent
@@
 
-Ultimately, this research provides a robust probabilistic foundation for
-the statistical analysis of DCJ data, building upon Thurstone's law of
-comparative judgment \citeyearpar{Thurstone_1927}. Consequently, its
-findings offer valuable insights for researchers and analysts designing
-and implementing DCJ experiments.
+Ultimately, this study provides a robust causal and probabilistic
+foundation for the statistical analysis of DCJ data, building upon
+Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
+Consequently, its findings offer valuable insights for researchers and
+analysts designing and implementing DCJ experiments.
 
@@
 as directed acyclic graphs (DAGs), structural causal models (SCMs), and
-the flow of association and causation in graphs. It is not a
+the flow of association and causation in graphs. The section is not a
 comprehensive description of causal inference methods. Readers
@@
 observed nor observable \citep{Neal_2020, Counterfactual_2024}.
-According to \citet{Pearl_et_al_2018}, counterfactuals form the
+According to \citet{Pearl_et_al_2018}, counterfactuals are the
 foundation of causal inference and occupy the highest level of cognitive
@@
 Graph theory is a branch of mathematics focused on the study of graphs.
-Graphs are mathematical structures that model pairwise relations between
+Graphs are mathematical structures modeling pairwise relations between
 objects. They can represent physical relations, such as electrical
@@
 blocks \citep{Neal_2020, McElreath_2020}. Figure~\ref{fig-dags}
-illustrates these building blocks. Figure~\ref{fig-dag_bb1} depicts two
+illustrates these blocks. Figure~\ref{fig-dag_bb1} depicts two
 unconnected nodes, representing an scenario where variables \(X_{1}\)
@@
 \(X_{1}\) is an \emph{ancestor} of \(X_{3}\) and that the relationship
-between these variables is entirely mediated by \(X_{2}\).
+between these variables is entirely \emph{mediated} by \(X_{2}\).
 Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
@@
 
-Given the heuristic nature of DAGs, a motivating example can illustrate
-how these five fundamental building blocks can be used to construct the
-causal structure of a system. This example can also help clarify
-additional rules and conventions for drawing DAGs.
+Given the heuristic nature of DAGs, a motivating example can help
+illustrate the use of the five fundamental building blocks to construct
+a system's causal structure. This example can also help to clarify
+additional conventions for drawing DAGs.
 
@@
 on an outcome \(Y\) needs to be investigated. Additionally, the problem
-indicates that a variable \(X\) potentially influences both \(Y\) and
-\(T\). Such scenarios are not hard to imagine. For instance, \(T\) could
-describe different treatments potentially influencing the binary onset
-of cancer (\(Y\)) while \(X\) could denote the age of a patient. Using
-an example from DCJ, {we need an example}. Figure~\ref{fig-example}
-presents two graphs describing the plausible causal structure of this
-system, the \emph{simplified} and the \emph{magnified} DAG.
+suggests that a variable \(X\) potentially influences both \(T\) and
+\(Y\). Such scenarios are not hard to imagine. For instance, \(T\) might
+represent different treatments that could affect the recovery from
+cancer \(Y\), while \(X\) could denote the patient's age. Similarly, in
+the context of a DCJ study like the one described by
+\citet{Boonen_et_al_2020}, \(T\) could represent the duration of a
+child's cochlear implant use, which might influence the child's overall
+speech quality \(Y\), with \(X\) indicating the child's hearing status.
+Note that these descriptions do not specify the type of data or the
+functional forms of the variables, highlighting the heuristic nature of
+DAGs. {(not a bad example, but I prefer one using writing skills)}
+
+Figure~\ref{fig-example} presents two graphs illustrating the plausible
+causal structure of the motivating example.
+Figure~\ref{fig-dag_example1} shows the \emph{simplified} DAG. In this
+graph, the variables of interest or \emph{endogenous} variables
+\(V=\{T,X,Y\}\) are depicted as solid black circles, indicating that
+they are observed variables. The arrows in the graph reflect the
+expected direction of causal influences among the variables.
+Figure~\ref{fig-dag_example2}, in contrast, depicts the \emph{magnified}
+version of the DAG. This graph includes, in addition to the endogenous
+variables, the \emph{exogenous} variables \(E=\{e_{T},e_{X},e_{Y}\}\).
+These exogenous variables are commonly referred to as
+\emph{disturbances} or \emph{errors}, because they represent factors not
+explicitly modeled. They are depicted as open circles, indicating their
+unobserved nature. While such exogenous variables are often omitted for
+simplicity, including them in the graph can be advantageous in certain
+scenarios, as they can help to highlight potential issues related to
+conditioning and confounding \citep{Cinelli_et_al_2020}, concepts that
+will be discussed in the next section.
+
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+
+Furthermore, a detailed examination of both figures reveals the presence
+of at least four of the five fundamental building blocks.
+Figure~\ref{fig-dag_example1} displays multiple descendants, evident in
+pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
+\(T \rightarrow Y\). It also illustrates a fork with
+\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-dag_example2}
+features multiple two-unconnected nodes, evident in the pairwise
+relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
+\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
+\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
+\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
+symbol \(\perp\!\!\!\perp\) denotes a concept know as
+\emph{d-separation}, which roughly implies the independence of the
+variables. This concept will be introduced in the next section.
 

### 2024-08-01 index.qmd
section:background,dag; process:write;figures

@@
 
-![Two unconnected nodes](images/figures/dag_bb1.png){#fig-dag_bb1 fig-align="center" width=40%}
+![Two unconnected nodes](images/figures/dag_bb1.png){#fig-dag_bb1 fig-align="center" width=30%}
 
-![Two connected nodes or descendant](images/figures/dag_bb2.png){#fig-dag_bb2 fig-align="center" width=40%}
+![Two connected nodes or descendant](images/figures/dag_bb2.png){#fig-dag_bb2 fig-align="center" width=30%}
 
-![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=100%}
+![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=85%}
 
-![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=100%}
+![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=85%}
 
-![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=100%}
+![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=85%}
 

### 2024-08-01 _tex/index.tex
Built site for gh-pages

@@
 
-\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_bb1.png}
+\includegraphics[width=0.3\textwidth,height=\textheight]{images/figures/dag_bb1.png}
 
@@
 
-\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_bb2.png}
+\includegraphics[width=0.3\textwidth,height=\textheight]{images/figures/dag_bb2.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb3.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb3.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb4.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb4.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_bb5.png}
+\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb5.png}
 

### 2024-08-02 index.qmd
section:background,dag; process:write

@@
 This study aims to fill this gap by representing DCJ within the framework of causal inference.
-Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
+Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
 The study then translates the scientific model into a probabilistic statistical model.
@@
 <!-- 3. How do we apply causal inference and counterfactuals? -->
-Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016].
+Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
 Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
@@
 <!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
-One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020; @Morgan_et_al_2014].
 @fig-dags illustrates these blocks.
@@
 
-![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=85%}
+![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=75%}
 
-![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=85%}
+![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=75%}
 
-![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=85%}
+![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=75%}
 
@@
 Given the heuristic nature of DAGs, a motivating example can help illustrate the use of the five fundamental building blocks to construct a system's causal structure.
-This example can also help to clarify additional conventions for drawing DAGs.
+This example can also help to clarify some additional conventions for drawing DAGs.
 
@@
 The arrows in the graph reflect the expected direction of causal influences among the variables.
-@fig-dag_example2, in contrast, depicts the *magnified* version of the DAG.
+In contrast, @fig-dag_example2 depicts the *magnified* version of the DAG.
 This graph includes, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{T},e_{X},e_{Y}\}$.
@@
 They are depicted as open circles, indicating their unobserved nature.
-While such exogenous variables are often omitted for simplicity, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts that will be discussed in the next section.
+These exogenous variables are often omitted for simplicity, resulting in the simplified DAG in @fig-dag_example1.
+However, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+These concepts are introduced in the next section.
 
@@
 The symbol $\dsep$ denotes a concept know as *d-separation*, which roughly implies the independence of the variables.
-This concept will be introduced in the next section.
+This concept is introduced in the next section.
 
@@
 
+<!-- Assuming certain values are decided in the experimental design (proportions, sample size, etc) you can reduce the graph -->
+
 <!-- It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
@@
 <!-- what does it mean in probabilistic terms? describe probabilities and arrive to structural model -->
+
+<!-- Final models does not require the estimation of latent variables, the results can be evaluated without them -->
+
+<!-- Notice that latent variables are kind of like descendants -->
 

### 2024-08-02 _tex/index.tex
Built site for gh-pages

@@
 framework of causal inference. Specifically, utilizing the structural
-approach \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study
-develops a scientific model to clarify plausible causal assumptions and
-mechanisms inherent in the DCJ system. The study then translates the
-scientific model into a probabilistic statistical model. This model aims
-to produce statistical estimates to draw inferences about plausible
-causal relationships within the DCJ system.
+approach, the study develops a scientific model to clarify plausible
+causal assumptions and mechanisms inherent in the DCJ system. The study
+then translates the scientific model into a probabilistic statistical
+model. This model aims to produce statistical estimates to draw
+inferences about plausible causal relationships within the DCJ system.
 
@@
 \citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
-\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}. Both approaches
-employ rigorous mathematical notation to characterize causal inference,
-but they do so in different ways \citep{Neal_2020}. The potential
-outcomes approach relies on counterfactual notation, whereas the
-structural approach employs the do-operator and structural causal models
+\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
+mathematical notation to characterize causal inference, but they do so
+in different ways \citep{Neal_2020}. The potential outcomes approach
+relies on counterfactual notation, whereas the structural approach
+employs the do-operator and structural causal models
 \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these differences,
@@
 represent various causal structures using only five fundamental building
-blocks \citep{Neal_2020, McElreath_2020}. Figure~\ref{fig-dags}
-illustrates these blocks. Figure~\ref{fig-dag_bb1} depicts two
-unconnected nodes, representing an scenario where variables \(X_{1}\)
-and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
-two connected nodes, illustrating a scenario where a parent node
-\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
-Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
-Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
-\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
-this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
-\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
-\(X_{1}\) is an \emph{ancestor} of \(X_{3}\) and that the relationship
-between these variables is entirely \emph{mediated} by \(X_{2}\).
-Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
-\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
-scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
-Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
-as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
-concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
-\(X_{3}\) are not causally related to each other but both influence
-\(X_{2}\).
+blocks \citep{Neal_2020, McElreath_2020, Morgan_et_al_2014}.
+Figure~\ref{fig-dags} illustrates these blocks. Figure~\ref{fig-dag_bb1}
+depicts two unconnected nodes, representing an scenario where variables
+\(X_{1}\) and \(X_{2}\) are not causally related.
+Figure~\ref{fig-dag_bb2} shows two connected nodes, illustrating a
+scenario where a parent node \(X_{1}\) exerts a causal influence on a
+child node \(X_{2}\). Consequently, \(X_{2}\) is considered a
+\emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3} depicts a
+\emph{chain} (or \emph{pipe}), where \(X_{1}\) influences \(X_{2}\), and
+\(X_{2}\) influences \(X_{3}\). In this configuration, \(X_{1}\) is a
+parent node of \(X_{2}\), and \(X_{2}\) a parent node of \(X_{3}\).
+Furthermore, the DAG shows that \(X_{1}\) is an \emph{ancestor} of
+\(X_{3}\) and that the relationship between these variables is entirely
+\emph{mediated} by \(X_{2}\). Figure~\ref{fig-dag_bb4} illustrates a
+\emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are both influenced
+by \(X_{2}\). In this scenario \(X_{2}\) is a parent node of both
+\(X_{1}\) and \(X_{3}\). Finally, Figure~\ref{fig-dag_bb5} depicts a
+\emph{collider}, also known as \emph{inmorality}, where variables
+\(X_{1}\) and \(X_{3}\) are concurrent causes of \(X_{2}\). In this
+configuration, \(X_{1}\) and \(X_{3}\) are not causally related to each
+other but both influence \(X_{2}\).
 
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb3.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb3.png}
 
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb4.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb4.png}
 
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_bb5.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb5.png}
 
@@
 illustrate the use of the five fundamental building blocks to construct
-a system's causal structure. This example can also help to clarify
+a system's causal structure. This example can also help to clarify some
 additional conventions for drawing DAGs.
@@
 they are observed variables. The arrows in the graph reflect the
-expected direction of causal influences among the variables.
-Figure~\ref{fig-dag_example2}, in contrast, depicts the \emph{magnified}
+expected direction of causal influences among the variables. In
+contrast, Figure~\ref{fig-dag_example2} depicts the \emph{magnified}
 version of the DAG. This graph includes, in addition to the endogenous
@@
 explicitly modeled. They are depicted as open circles, indicating their
-unobserved nature. While such exogenous variables are often omitted for
-simplicity, including them in the graph can be advantageous in certain
-scenarios, as they can help to highlight potential issues related to
-conditioning and confounding \citep{Cinelli_et_al_2020}, concepts that
-will be discussed in the next section.
+unobserved nature. These exogenous variables are often omitted for
+simplicity, resulting in the simplified DAG in
+Figure~\ref{fig-dag_example1}. However, including them in the graph can
+be advantageous in certain scenarios, as they can help to highlight
+potential issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}. These concepts are introduced in the next
+section.
 
@@
 \emph{d-separation}, which roughly implies the independence of the
-variables. This concept will be introduced in the next section.
+variables. This concept is introduced in the next section.
 

### 2024-08-02 images/tex/mdag_bb4.tex
section:background,dag; process:write,figures

\documentclass[tikz, border=1mm]{standalone}
\usepackage{tikz} 
\usetikzlibrary{arrows.meta}
\usepackage{pgfplots}

\pgfplotsset{compat=1.18}

\begin{document}

\begin{tikzpicture}

    % dag_bb3
    \node at (-2,0) {$e_{X}$};
    \node at (-1,0) {$X$};
    \node at (-1,0.7) {$e_{Z}$};
    \node at (0,0.7) {$Z$};
    \node at (1,0) {$Y$};
    \node at (2,0) {$e_{Y}$};

    \draw[{Circle[open]}-{latex}](-2,-0.3) to (-1,-0.3); % eX -> X (circle)
    \draw[-{latex}{Circle}](0,0.45) to (-1,-0.3); % X -> Z
    \draw[{Circle[open]}-{latex}](-1,0.47) to (0,0.47); % eZ -> Z (circle)
    \draw[{Circle}-{latex}{Circle}](0,0.5) to (1,-0.3); % Z -> Y
    \draw[{Circle[open]}-{latex}](2,-0.3) to (1,-0.3); % eY -> Y (circle)

\end{tikzpicture}

\end{document}

### 2024-08-02 index.qmd
section:background,dag; process:write,figures

@@
 Nevertheless, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
-DAGs are heuristics that can effectively convey the assumed causal structure of a system.
-They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].
+DAGs serve as heuristics, effectively conveying the assumed causal structure of a system.
+A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015].
+Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].
 
@@
 
-## Directed acyclic graphs (DAG) {#sec-framework-dag}
+## DAGs and SCMs {#sec-framework-dag}
 
@@
 
-<!-- 2. How graphs are use in the structural approach? and what are DAGs? -->
+<!-- 2. What does the structural approach uses?, and what are DAGs? -->
 In statistics, one application incorporating concepts from graph theory is causal inference.
-Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of the causal structure of a system [@Neal_2020].
-In this context, a *graph* is a collection of nodes connected by edges, where nodes represent random variables.
+Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) and its associated structural causal models (SCM) to provide a graphical and formal representation of the causal structure of a system [@Neal_2020].
+In this context, a *graph* denotes a collection of nodes connected by edges, where nodes represent random variables.
 The term *directed* indicates that the edges of the graph extend from one node to another, with arrows showing the direction of causal influence.
@@
 
-<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
-One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020; @Morgan_et_al_2014].
-@fig-dags illustrates these blocks.
-@fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related.
-@fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$.
-Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$.
-@fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
-In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$.
-Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely *mediated* by $X_{2}$.
-@fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
-In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$.
-Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
-In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.
+<!-- 3. The first advantage of DAGs is that you need only five building blocks -->
+<!-- 3.1 how DAG can be depicted? Simplified? -->
+One key advantage of DAGs is that they can represent various causal structures using only five fundamental building blocks, regardless of complexity.
+The left panels of @fig-dags_scms illustrates the *simplified* graphs of these block.
+In these graphs, the variables of interest or *endogenous* variables, $V=\{X,Z,Y\}$, are depicted as solid black circles, indicating that they are observed variables.
+The arrows in the graph reflect the expected direction of causal influences among the variables.
 
-::: {#fig-dags layout="[[1,1],[1,1,1]]"}
+<!-- 3.2 how DAG can be depicted? Magnified? -->
+In contrast, the middle panels of @fig-dags_scms depicts the *magnified* graphs of these building blocks.
+These graph include, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$.
+These exogenous variables, commonly referred to as *disturbances* or *errors*, represent factors not explicitly modeled.
+Moreover, they are depicted as open circles, indicating their unobserved nature.
+These exogenous variables are often omitted for simplicity, resulting in the simplified DAGs shown in the left panels.
+However, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts introduced in the next section.
 
-![Two unconnected nodes](images/figures/dag_bb1.png){#fig-dag_bb1 fig-align="center" width=30%}
+<!-- commands for d-separation -->
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
 
-![Two connected nodes or descendant](images/figures/dag_bb2.png){#fig-dag_bb2 fig-align="center" width=30%}
+::: {#fig-dags_scms layout="[ [1,1,1],[1,1,1],[1,1,1],[1,1,1],[1,1,1] ]"}
 
-![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=75%}
+![Two unconnected nodes](images/figures/sdag_bb1.png){#fig-sdag_bb1 fig-align="center" width=35%}
 
-![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=75%}
+![Two unconnected nodes](images/figures/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" width=80%}
 
-![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=75%}
+::: {#fig-scm_bb1 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
 
-DAG's fundamental building blocks.
+Two unconnected nodes
 :::
 
-<!-- 4. A motivating example -->
+![Two connected nodes or descendant](images/figures/sdag_bb2.png){#fig-sdag_bb2 fig-align="center" width=35%}
+
+![Two connected nodes or descendant](images/figures/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" width=80%}
+
+::: {#fig-scm_bb2 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(X,e_{Y}) \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
+
+Two connected nodes or descendant
+:::
+
+![Chain or pipe](images/figures/sdag_bb3.png){#fig-sdag_bb3 fig-align="center" width=60%}
+
+![Chain or pipe](images/figures/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" width=100%}
+
+::: {#fig-scm_bb3 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Z} \\
+  e_{Z} & \dsep e_{Y}
+\end{aligned}
+$$
+
+Chain or pipe
+:::
+
+![Fork](images/figures/sdag_bb4.png){#fig-sdag_bb4 fig-align="center" width=60%}
+
+![Fork](images/figures/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" width=100%}
+
+::: {#fig-scm_bb4 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(Z,e_{X}) \\
+  Z & := f_{Z}(e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Z} \\
+  e_{Z} & \dsep e_{Y}
+\end{aligned}
+$$
+
+Fork
+:::
+
+![Collider or inmorality](images/figures/sdag_bb5.png){#fig-sdag_bb5 fig-align="center" width=60%}
+
+![Collider or inmorality](images/figures/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" width=100%}
+
+::: {#fig-scm_bb5 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,Y,e_{Z}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Z} \\
+  e_{Z} & \dsep e_{Y}
+\end{aligned}
+$$
+
+Collider or inmorality
+:::
+
+The five fundamental building blocks of DAGs with corresponding SCMs. **Note:** left column shows simplified DAGS, middle column show magnified DAGs, and right column shows the corresponding SCMs.
+:::
+
+<!-- 4. But what does the building blocks mean? -->
+A careful examination of the left and middle panels of @fig-dags_scms reveal what do the building blocks assume.
+Figures [-@fig-sdag_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing an scenario where variables $X$ and $Y$ are not causally related.
+Figures [-@fig-sdag_bb2] and [-@fig-mdag_bb2] show two connected nodes, illustrating a scenario where a parent node $X$ exerts a causal influence on a child node $Y$.
+Consequently, $Y$ is considered a *descendant* of $X$.
+Figures [-@fig-sdag_bb3] and [-@fig-mdag_bb3] depict a *chain* (or *pipe*), where $X$ influences $Z$, and $Z$ influences $Y$.
+In this configuration, $X$ is a parent node of $Z$, and $Z$ a parent node of $Y$.
+Furthermore, the DAG shows that $X$ is an *ancestor* of $Y$ and that the relationship between these variables is entirely *mediated* by $Z$.
+Figures [-@fig-sdag_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
+In this scenario $Z$ is a parent node of both $X$ and $Y$.
+Finally, figures [-@fig-sdag_bb5] and [-@fig-mdag_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$.
+In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$.
+
+<!-- 4. The second advantage of DAGs is that there is no need for functional form -->
+<!-- The second key advantage of DAGs is their ability to represent causal relations in a nonparametric and fully interactive manner.
+This means that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables.
+This generality allows modeling causality without assuming a specific functional form [@Morgan_et_al_2014].
+-->
+
+<!-- 5. A motivating example -->
 Given the heuristic nature of DAGs, a motivating example can help illustrate the use of the five fundamental building blocks to construct a system's causal structure.
@@
 Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status.
-Note that these descriptions do not specify the type of data or the functional forms of the variables, highlighting the heuristic nature of DAGs.
 [(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}
 
-<!-- 5. DAG for the motivating example -->
+<!-- 6. DAG for the motivating example -->
 @fig-example presents two graphs illustrating the plausible causal structure of the motivating example.
-@fig-dag_example1 shows the *simplified* DAG.
-In this graph, the variables of interest or *endogenous* variables $V=\{T,X,Y\}$ are depicted as solid black circles, indicating that they are observed variables.
-The arrows in the graph reflect the expected direction of causal influences among the variables.
-In contrast, @fig-dag_example2 depicts the *magnified* version of the DAG.
-This graph includes, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{T},e_{X},e_{Y}\}$.
-These exogenous variables are commonly referred to as *disturbances* or *errors*, because they represent factors not explicitly modeled.
-They are depicted as open circles, indicating their unobserved nature.
-These exogenous variables are often omitted for simplicity, resulting in the simplified DAG in @fig-dag_example1.
-However, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
-These concepts are introduced in the next section.
-
-<!-- 6. But where are the building blocks? -->
-<!-- commands for d-separation -->
-\newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
-
-Furthermore, a detailed examination of both figures reveals the presence of at least four of the five fundamental building blocks.
-@fig-dag_example1 displays multiple descendants, evident in pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+@fig-sdag_example1 shows the *simplified* graph, while @fig-mdag_example1 depicts the *magnified* graph of the DAG.
+A detailed examination of both figures reveals the presence of at least four of the five fundamental building blocks.
+@fig-sdag_example1 displays multiple descendants, evident in pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
 It also illustrates a fork with $X \rightarrow \{T,Y\}$.
-Similarly, @fig-dag_example2 features multiple two-unconnected nodes, evident in the pairwise relations $\{e_{T} \dsep e_{X}\}$, $\{e_{T} \dsep e_{Y}\}$, and $\{e_{X} \dsep e_{Y}\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
+Similarly, @fig-mdag_example1 features multiple two-unconnected nodes, evident in the pairwise relations $\{e_{T} \dsep e_{X}\}$, $\{e_{T} \dsep e_{Y}\}$, and $\{e_{X} \dsep e_{Y}\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
 The symbol $\dsep$ denotes a concept know as *d-separation*, which roughly implies the independence of the variables.
@@
 
-::: {#fig-example layout-ncol=2}
+::: {#fig-example layout="[ [1,1,1],[1,1,1],[1,1,1] ]"}
 
-![Simplied DAG](images/figures/dag_example1.png){#fig-dag_example1 fig-align="center" width=45%}
+![Simplied DAG](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=35%}
 
-![Magnified DAG](images/figures/dag_example2.png){#fig-dag_example2 fig-align="center" width=85%}
+![Magnified DAG](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=65%}
+
+::: {#fig-scm_example1 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,Y,e_{Z}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Z} \\
+  e_{Z} & \dsep e_{Y}
+\end{aligned}
+$$
+
+Collider or inmorality
+:::
 
@@
 ## The flow of association and causation in graphs {#sec-framework-flow}
+
+<!-- Although there are several benefits  -->
 
@@
 <!-- (Structural Causal Model (SCM)) A structural causal model is a tuple of the following sets: -->
-<!-- 1. A set ofendogenous variables 푉 2. A set ofexogenous variables 푈 3. A set of functions 푓, one to generate each endogenous variable as a function ofother variables Neal -->
+<!-- 1. A set of endogenous variables 푉 2. A set of exogenous variables 푈 3. A set of functions 푓, one to generate each endogenous variable as a function of other variables Neal -->
 

### 2024-08-02 _tex/index.tex
Built site for gh-pages

@@
 representation of any system through directed acyclic graphs
-\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs are heuristics that
-can effectively convey the assumed causal structure of a system. They do
-not represent detailed statistical models but allow researchers to
-deduce which statistical models can provide valid causal inferences,
-assuming the causal structure depicted in the DAG is accurate
-\citep{McElreath_2020}.
+\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
+effectively conveying the assumed causal structure of a system. A
+heuristic is a strategy that simplifies information to make decisions
+more quickly, efficiently, and sometimes more accurately than complex
+methods \citep{Chow_2015}. Consequently, DAGs do not represent detailed
+statistical models but allow researchers to deduce which statistical
+models can provide valid causal inferences, assuming the causal
+structure depicted in the DAG is accurate \citep{McElreath_2020}.
 
-\subsection{Directed acyclic graphs (DAG)}\label{sec-framework-dag}
+\subsection{DAGs and SCMs}\label{sec-framework-dag}
 
@@
 is causal inference. Specifically, the structural approach to causal
-inference uses directed acyclic graphs (DAG) to provide a formal and
-graphical representation of the causal structure of a system
-\citep{Neal_2020}. In this context, a \emph{graph} is a collection of
-nodes connected by edges, where nodes represent random variables. The
-term \emph{directed} indicates that the edges of the graph extend from
-one node to another, with arrows showing the direction of causal
-influence. Moreover, the term \emph{acyclic} indicates the causal
-influences do not form a loop, meaning the influences do not cycle back
-on themselves \citep{McElreath_2020}.
+inference uses directed acyclic graphs (DAG) and its associated
+structural causal models (SCM) to provide a graphical and formal
+representation of the causal structure of a system \citep{Neal_2020}. In
+this context, a \emph{graph} denotes a collection of nodes connected by
+edges, where nodes represent random variables. The term \emph{directed}
+indicates that the edges of the graph extend from one node to another,
+with arrows showing the direction of causal influence. Moreover, the
+term \emph{acyclic} indicates the causal influences do not form a loop,
+meaning the influences do not cycle back on themselves
+\citep{McElreath_2020}.
 
-One key advantage of DAGs is that, regardless of complexity, they can
-represent various causal structures using only five fundamental building
-blocks \citep{Neal_2020, McElreath_2020, Morgan_et_al_2014}.
-Figure~\ref{fig-dags} illustrates these blocks. Figure~\ref{fig-dag_bb1}
-depicts two unconnected nodes, representing an scenario where variables
-\(X_{1}\) and \(X_{2}\) are not causally related.
-Figure~\ref{fig-dag_bb2} shows two connected nodes, illustrating a
-scenario where a parent node \(X_{1}\) exerts a causal influence on a
-child node \(X_{2}\). Consequently, \(X_{2}\) is considered a
-\emph{descendant} of \(X_{1}\). Figure~\ref{fig-dag_bb3} depicts a
-\emph{chain} (or \emph{pipe}), where \(X_{1}\) influences \(X_{2}\), and
-\(X_{2}\) influences \(X_{3}\). In this configuration, \(X_{1}\) is a
-parent node of \(X_{2}\), and \(X_{2}\) a parent node of \(X_{3}\).
-Furthermore, the DAG shows that \(X_{1}\) is an \emph{ancestor} of
-\(X_{3}\) and that the relationship between these variables is entirely
-\emph{mediated} by \(X_{2}\). Figure~\ref{fig-dag_bb4} illustrates a
-\emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are both influenced
-by \(X_{2}\). In this scenario \(X_{2}\) is a parent node of both
-\(X_{1}\) and \(X_{3}\). Finally, Figure~\ref{fig-dag_bb5} depicts a
-\emph{collider}, also known as \emph{inmorality}, where variables
-\(X_{1}\) and \(X_{3}\) are concurrent causes of \(X_{2}\). In this
-configuration, \(X_{1}\) and \(X_{3}\) are not causally related to each
-other but both influence \(X_{2}\).
+One key advantage of DAGs is that they can represent various causal
+structures using only five fundamental building blocks, regardless of
+complexity. The left panels of Figure~\ref{fig-dags_scms} illustrates
+the \emph{simplified} graphs of these block. In these graphs, the
+variables of interest or \emph{endogenous} variables, \(V=\{X,Z,Y\}\),
+are depicted as solid black circles, indicating that they are observed
+variables. The arrows in the graph reflect the expected direction of
+causal influences among the variables.
+
+In contrast, the middle panels of Figure~\ref{fig-dags_scms} depicts the
+\emph{magnified} graphs of these building blocks. These graph include,
+in addition to the endogenous variables, the \emph{exogenous} variables
+\(E=\{e_{X},e_{Z},e_{Y}\}\). These exogenous variables, commonly
+referred to as \emph{disturbances} or \emph{errors}, represent factors
+not explicitly modeled. Moreover, they are depicted as open circles,
+indicating their unobserved nature. These exogenous variables are often
+omitted for simplicity, resulting in the simplified DAGs shown in the
+left panels. However, including them in the graph can be advantageous in
+certain scenarios, as they can help to highlight potential issues
+related to conditioning and confounding \citep{Cinelli_et_al_2020},
+concepts introduced in the next section.
+
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
 
@@
 
-\begin{minipage}{0.50\linewidth}
+\begin{minipage}{0.33\linewidth}
 
@@
 
-\includegraphics[width=0.3\textwidth,height=\textheight]{images/figures/dag_bb1.png}
+\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb1.png}
 
@@
 
-\subcaption{\label{fig-dag_bb1}Two unconnected nodes}
+\subcaption{\label{fig-sdag_bb1}Two unconnected nodes}
 
@@
 %
-\begin{minipage}{0.50\linewidth}
+\begin{minipage}{0.33\linewidth}
 
@@
 
-\includegraphics[width=0.3\textwidth,height=\textheight]{images/figures/dag_bb2.png}
+\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb1.png}
 
@@
 
-\subcaption{\label{fig-dag_bb2}Two connected nodes or descendant}
+\subcaption{\label{fig-mdag_bb1}Two unconnected nodes}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb1}Two unconnected nodes}
 
@@
 
-\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb3.png}
+\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb2.png}
 
@@
 
-\subcaption{\label{fig-dag_bb3}Chain or pipe}
+\subcaption{\label{fig-sdag_bb2}Two connected nodes or descendant}
 
@@
 
-\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb4.png}
+\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb2.png}
 
@@
 
-\subcaption{\label{fig-dag_bb4}Fork}
+\subcaption{\label{fig-mdag_bb2}Two connected nodes or descendant}
 
@@
 
-\includegraphics[width=0.75\textwidth,height=\textheight]{images/figures/dag_bb5.png}
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(X,e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
 
@@
 
-\subcaption{\label{fig-dag_bb5}Collider or inmorality}
+\subcaption{\label{fig-scm_bb2}Two connected nodes or descendant}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb3.png}
+
+}
+
+\subcaption{\label{fig-sdag_bb3}Chain or pipe}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb3.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb3}Chain or pipe}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Z} \\
+  e_{Z} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb3}Chain or pipe}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb4.png}
+
+}
+
+\subcaption{\label{fig-sdag_bb4}Fork}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb4.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb4}Fork}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(Z,e_{X}) \\
+  Z & := f_{Z}(e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Z} \\
+  e_{Z} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb4}Fork}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb5.png}
+
+}
+
+\subcaption{\label{fig-sdag_bb5}Collider or inmorality}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb5.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb5}Collider or inmorality}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,Y,e_{Z}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Z} \\
+  e_{Z} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb5}Collider or inmorality}
 
@@
 
-\caption{\label{fig-dags}DAG's fundamental building blocks.}
+\caption{\label{fig-dags_scms}The five fundamental building blocks of
+DAGs with corresponding SCMs. \textbf{Note:} left column shows
+simplified DAGS, middle column show magnified DAGs, and right column
+shows the corresponding SCMs.}
 
 \end{figure}%
+
+A careful examination of the left and middle panels of
+Figure~\ref{fig-dags_scms} reveal what do the building blocks assume.
+Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1} depict two unconnected
+nodes, representing an scenario where variables \(X\) and \(Y\) are not
+causally related. Figures \ref{fig-sdag_bb2} and \ref{fig-mdag_bb2} show
+two connected nodes, illustrating a scenario where a parent node \(X\)
+exerts a causal influence on a child node \(Y\). Consequently, \(Y\) is
+considered a \emph{descendant} of \(X\). Figures \ref{fig-sdag_bb3} and
+\ref{fig-mdag_bb3} depict a \emph{chain} (or \emph{pipe}), where \(X\)
+influences \(Z\), and \(Z\) influences \(Y\). In this configuration,
+\(X\) is a parent node of \(Z\), and \(Z\) a parent node of \(Y\).
+Furthermore, the DAG shows that \(X\) is an \emph{ancestor} of \(Y\) and
+that the relationship between these variables is entirely
+\emph{mediated} by \(Z\). Figures \ref{fig-sdag_bb4} and
+\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
+\(Y\) are both influenced by \(Z\). In this scenario \(Z\) is a parent
+node of both \(X\) and \(Y\). Finally, figures \ref{fig-sdag_bb5} and
+\ref{fig-mdag_bb5} depict a \emph{collider}, also known as
+\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
+of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
+related to each other but both influence \(Z\).
 
@@
 speech quality \(Y\), with \(X\) indicating the child's hearing status.
-Note that these descriptions do not specify the type of data or the
-functional forms of the variables, highlighting the heuristic nature of
-DAGs. {(not a bad example, but I prefer one using writing skills)}
+{(not a bad example, but I prefer one using writing skills)}
 
@@
 causal structure of the motivating example.
-Figure~\ref{fig-dag_example1} shows the \emph{simplified} DAG. In this
-graph, the variables of interest or \emph{endogenous} variables
-\(V=\{T,X,Y\}\) are depicted as solid black circles, indicating that
-they are observed variables. The arrows in the graph reflect the
-expected direction of causal influences among the variables. In
-contrast, Figure~\ref{fig-dag_example2} depicts the \emph{magnified}
-version of the DAG. This graph includes, in addition to the endogenous
-variables, the \emph{exogenous} variables \(E=\{e_{T},e_{X},e_{Y}\}\).
-These exogenous variables are commonly referred to as
-\emph{disturbances} or \emph{errors}, because they represent factors not
-explicitly modeled. They are depicted as open circles, indicating their
-unobserved nature. These exogenous variables are often omitted for
-simplicity, resulting in the simplified DAG in
-Figure~\ref{fig-dag_example1}. However, including them in the graph can
-be advantageous in certain scenarios, as they can help to highlight
-potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}. These concepts are introduced in the next
-section.
-
-\newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
-
-Furthermore, a detailed examination of both figures reveals the presence
-of at least four of the five fundamental building blocks.
-Figure~\ref{fig-dag_example1} displays multiple descendants, evident in
+Figure~\ref{fig-sdag_example1} shows the \emph{simplified} graph, while
+Figure~\ref{fig-mdag_example1} depicts the \emph{magnified} graph of the
+DAG. A detailed examination of both figures reveals the presence of at
+least four of the five fundamental building blocks.
+Figure~\ref{fig-sdag_example1} displays multiple descendants, evident in
 pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
 \(T \rightarrow Y\). It also illustrates a fork with
-\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-dag_example2}
+\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-mdag_example1}
 features multiple two-unconnected nodes, evident in the pairwise
@@
 
-\includegraphics[width=0.45\textwidth,height=\textheight]{images/figures/dag_example1.png}
+\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_example1.png}
 
@@
 
-\subcaption{\label{fig-dag_example1}Simplied DAG}
+\subcaption{\label{fig-sdag_example1}Simplied DAG}
 
@@
 
-\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_example2.png}
+\includegraphics[width=0.65\textwidth,height=\textheight]{images/figures/mdag_example1.png}
 
@@
 
-\subcaption{\label{fig-dag_example2}Magnified DAG}
+\subcaption{\label{fig-mdag_example1}Magnified DAG}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,Y,e_{Z}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Z} \\
+  e_{Z} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_example1}Collider or inmorality}
 

### 2024-08-05 index.qmd
section:background,dag; process:write

@@
 In statistics, one application incorporating concepts from graph theory is causal inference.
-Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) and its associated structural causal models (SCM) to provide a graphical and formal representation of the causal structure of a system [@Neal_2020].
+Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG), and their associated structural causal models (SCM), to provide a graphical and formal representation of the causal structure of a system [@Neal_2020].
 In this context, a *graph* denotes a collection of nodes connected by edges, where nodes represent random variables.
-The term *directed* indicates that the edges of the graph extend from one node to another, with arrows showing the direction of causal influence.
+The term *directed* indicates the edges of the graph extend from one node to another, with arrows showing the direction of causal influence.
 Moreover, the term *acyclic* indicates the causal influences do not form a loop, meaning the influences do not cycle back on themselves [@McElreath_2020].
 
-<!-- 3. The first advantage of DAGs is that you need only five building blocks -->
-<!-- 3.1 how DAG can be depicted? Simplified? -->
-One key advantage of DAGs is that they can represent various causal structures using only five fundamental building blocks, regardless of complexity.
-The left panels of @fig-dags_scms illustrates the *simplified* graphs of these block.
-In these graphs, the variables of interest or *endogenous* variables, $V=\{X,Z,Y\}$, are depicted as solid black circles, indicating that they are observed variables.
-The arrows in the graph reflect the expected direction of causal influences among the variables.
+<!-- 3. how DAGs can be depicted? -->
+<!-- 3.1 the simplified DAG -->
+Regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+The left panels of @fig-dags_scms illustrate these building blocks using the *standard* representation [@Morgan_et_al_2014].
+In these graphs, the variables of interest, or *endogenous* variables $V=\{X,Z,Y\}$, are depicted as solid black circles to signify that they are observed random variables.
+The arrows in the graphs reflect the expected direction of causal influences among these variables.
 
-<!-- 3.2 how DAG can be depicted? Magnified? -->
-In contrast, the middle panels of @fig-dags_scms depicts the *magnified* graphs of these building blocks.
-These graph include, in addition to the endogenous variables, the *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$.
-These exogenous variables, commonly referred to as *disturbances* or *errors*, represent factors not explicitly modeled.
-Moreover, they are depicted as open circles, indicating their unobserved nature.
-These exogenous variables are often omitted for simplicity, resulting in the simplified DAGs shown in the left panels.
-However, including them in the graph can be advantageous in certain scenarios, as they can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts introduced in the next section.
+<!-- 3.2 the magnified DAG -->
+In contrast, the middle panels of @fig-dags_scms present the building blocks using the *magnified* representation.
+These graphs incorporate *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$, depicted as open circles to signify their unobserved nature.
+Exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
+27,68].
+Thus, they can be viewed as the 'idiosyncratic' causes of each variable.
+Although these variables are often omitted for simplicity, as illustrated in the left panels, including them in the graphs can be beneficial, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the next section.
+
+<!-- 3.3 the SCM -->
+<!-- Lastly, the right panels of @fig-dags_scms present the building blocks using the SCM representation. SCMs are  -->
 
@@
 
-The five fundamental building blocks of DAGs with corresponding SCMs. **Note:** left column shows simplified DAGS, middle column show magnified DAGs, and right column shows the corresponding SCMs.
+The five fundamental building blocks of DAGs with corresponding SCMs.
+**Note:** left panels show the standard representation of DAGs, middle panels show the magnified representation of DAGs, and the right panels show their corresponding SCMs.
 :::
 
-<!-- 4. But what does the building blocks mean? -->
-A careful examination of the left and middle panels of @fig-dags_scms reveal what do the building blocks assume.
-Figures [-@fig-sdag_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing an scenario where variables $X$ and $Y$ are not causally related.
+<!-- 4. What are the advantages of DAGs? -->
+<!-- DAGs can represent causal relations in a nonparametric and fully interactive manner.
+This means that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables.
+This generality allows for modeling causality without assuming a specific functional form [@Morgan_et_al_2014].
+-->
+
+<!-- 5. But what does the building blocks mean? -->
+A careful examination of the left and middle panels of @fig-dags_scms reveals the assumption inherent in these building blocks.
+Figures [-@fig-sdag_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related.
 Figures [-@fig-sdag_bb2] and [-@fig-mdag_bb2] show two connected nodes, illustrating a scenario where a parent node $X$ exerts a causal influence on a child node $Y$.
@@
 Figures [-@fig-sdag_bb3] and [-@fig-mdag_bb3] depict a *chain* (or *pipe*), where $X$ influences $Z$, and $Z$ influences $Y$.
-In this configuration, $X$ is a parent node of $Z$, and $Z$ a parent node of $Y$.
-Furthermore, the DAG shows that $X$ is an *ancestor* of $Y$ and that the relationship between these variables is entirely *mediated* by $Z$.
+In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$.
+Furthermore, the DAG shows that $X$ is an *ancestor* of $Y$ and that $Z$ completely *mediates* the relationship between these variables.
 Figures [-@fig-sdag_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
-In this scenario $Z$ is a parent node of both $X$ and $Y$.
-Finally, figures [-@fig-sdag_bb5] and [-@fig-mdag_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$.
+In this scenario, $Z$ is a parent node of $X$ and $Y$.
+Finally, Figures [-@fig-sdag_bb5] and [-@fig-mdag_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$.
 In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$.
-
-<!-- 4. The second advantage of DAGs is that there is no need for functional form -->
-<!-- The second key advantage of DAGs is their ability to represent causal relations in a nonparametric and fully interactive manner.
-This means that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables.
-This generality allows modeling causality without assuming a specific functional form [@Morgan_et_al_2014].
--->
 

### 2024-08-05 _tex/index.tex
Built site for gh-pages

@@
 is causal inference. Specifically, the structural approach to causal
-inference uses directed acyclic graphs (DAG) and its associated
-structural causal models (SCM) to provide a graphical and formal
+inference uses directed acyclic graphs (DAG), and their associated
+structural causal models (SCM), to provide a graphical and formal
 representation of the causal structure of a system \citep{Neal_2020}. In
@@
 edges, where nodes represent random variables. The term \emph{directed}
-indicates that the edges of the graph extend from one node to another,
-with arrows showing the direction of causal influence. Moreover, the
-term \emph{acyclic} indicates the causal influences do not form a loop,
+indicates the edges of the graph extend from one node to another, with
+arrows showing the direction of causal influence. Moreover, the term
+\emph{acyclic} indicates the causal influences do not form a loop,
 meaning the influences do not cycle back on themselves
@@
 
-One key advantage of DAGs is that they can represent various causal
-structures using only five fundamental building blocks, regardless of
-complexity. The left panels of Figure~\ref{fig-dags_scms} illustrates
-the \emph{simplified} graphs of these block. In these graphs, the
-variables of interest or \emph{endogenous} variables, \(V=\{X,Z,Y\}\),
-are depicted as solid black circles, indicating that they are observed
-variables. The arrows in the graph reflect the expected direction of
-causal influences among the variables.
+Regardless of complexity, DAGs can represent various causal structures
+using only five fundamental building blocks
+\citep{Neal_2020, McElreath_2020}. The left panels of
+Figure~\ref{fig-dags_scms} illustrate these building blocks using the
+\emph{standard} representation \citep{Morgan_et_al_2014}. In these
+graphs, the variables of interest, or \emph{endogenous} variables
+\(V=\{X,Z,Y\}\), are depicted as solid black circles to signify that
+they are observed random variables. The arrows in the graphs reflect the
+expected direction of causal influences among these variables.
 
-In contrast, the middle panels of Figure~\ref{fig-dags_scms} depicts the
-\emph{magnified} graphs of these building blocks. These graph include,
-in addition to the endogenous variables, the \emph{exogenous} variables
-\(E=\{e_{X},e_{Z},e_{Y}\}\). These exogenous variables, commonly
-referred to as \emph{disturbances} or \emph{errors}, represent factors
-not explicitly modeled. Moreover, they are depicted as open circles,
-indicating their unobserved nature. These exogenous variables are often
-omitted for simplicity, resulting in the simplified DAGs shown in the
-left panels. However, including them in the graph can be advantageous in
-certain scenarios, as they can help to highlight potential issues
-related to conditioning and confounding \citep{Cinelli_et_al_2020},
-concepts introduced in the next section.
+In contrast, the middle panels of Figure~\ref{fig-dags_scms} present the
+building blocks using the \emph{magnified} representation. These graphs
+incorporate \emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\),
+depicted as open circles to signify their unobserved nature. Exogenous
+variables represent \emph{errors} or \emph{disturbances} arising from
+omitted factors that the investigator chooses not to model explicitly
+\citep[27,68]{Pearl_2009}. Thus, they can be viewed as the
+`idiosyncratic' causes of each variable. Although these variables are
+often omitted for simplicity, as illustrated in the left panels,
+including them in the graphs can be beneficial, as their presence can
+reveal potential issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}, concepts explored in the next section.
 
@@
 \caption{\label{fig-dags_scms}The five fundamental building blocks of
-DAGs with corresponding SCMs. \textbf{Note:} left column shows
-simplified DAGS, middle column show magnified DAGs, and right column
-shows the corresponding SCMs.}
+DAGs with corresponding SCMs. \textbf{Note:} left panels show the
+standard representation of DAGs, middle panels show the magnified
+representation of DAGs, and the right panels show their corresponding
+SCMs.}
 
@@
 A careful examination of the left and middle panels of
-Figure~\ref{fig-dags_scms} reveal what do the building blocks assume.
-Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1} depict two unconnected
-nodes, representing an scenario where variables \(X\) and \(Y\) are not
-causally related. Figures \ref{fig-sdag_bb2} and \ref{fig-mdag_bb2} show
-two connected nodes, illustrating a scenario where a parent node \(X\)
-exerts a causal influence on a child node \(Y\). Consequently, \(Y\) is
-considered a \emph{descendant} of \(X\). Figures \ref{fig-sdag_bb3} and
-\ref{fig-mdag_bb3} depict a \emph{chain} (or \emph{pipe}), where \(X\)
-influences \(Z\), and \(Z\) influences \(Y\). In this configuration,
-\(X\) is a parent node of \(Z\), and \(Z\) a parent node of \(Y\).
-Furthermore, the DAG shows that \(X\) is an \emph{ancestor} of \(Y\) and
-that the relationship between these variables is entirely
-\emph{mediated} by \(Z\). Figures \ref{fig-sdag_bb4} and
+Figure~\ref{fig-dags_scms} reveals the assumption inherent in these
+building blocks. Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1}
+depict two unconnected nodes, representing a scenario where variables
+\(X\) and \(Y\) are not causally related. Figures \ref{fig-sdag_bb2} and
+\ref{fig-mdag_bb2} show two connected nodes, illustrating a scenario
+where a parent node \(X\) exerts a causal influence on a child node
+\(Y\). Consequently, \(Y\) is considered a \emph{descendant} of \(X\).
+Figures \ref{fig-sdag_bb3} and \ref{fig-mdag_bb3} depict a \emph{chain}
+(or \emph{pipe}), where \(X\) influences \(Z\), and \(Z\) influences
+\(Y\). In this configuration, \(X\) is a parent node of \(Z\), and \(Z\)
+is a parent node of \(Y\). Furthermore, the DAG shows that \(X\) is an
+\emph{ancestor} of \(Y\) and that \(Z\) completely \emph{mediates} the
+relationship between these variables. Figures \ref{fig-sdag_bb4} and
 \ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
-\(Y\) are both influenced by \(Z\). In this scenario \(Z\) is a parent
-node of both \(X\) and \(Y\). Finally, figures \ref{fig-sdag_bb5} and
+\(Y\) are both influenced by \(Z\). In this scenario, \(Z\) is a parent
+node of \(X\) and \(Y\). Finally, Figures \ref{fig-sdag_bb5} and
 \ref{fig-mdag_bb5} depict a \emph{collider}, also known as

### 2024-08-06 index.qmd
section:background,dag; process:write,figures

@@
 In statistics, one application incorporating concepts from graph theory is causal inference.
-Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG), and their associated structural causal models (SCM), to provide a graphical and formal representation of the causal structure of a system [@Neal_2020].
+Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a graphical and formal representation of the causal structure of a system [@Neal_2020].
 In this context, a *graph* denotes a collection of nodes connected by edges, where nodes represent random variables.
@@
 
-<!-- 3. how DAGs can be depicted? -->
-<!-- 3.1 the simplified DAG -->
-Regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-The left panels of @fig-dags_scms illustrate these building blocks using the *standard* representation [@Morgan_et_al_2014].
-In these graphs, the variables of interest, or *endogenous* variables $V=\{X,Z,Y\}$, are depicted as solid black circles to signify that they are observed random variables.
-The arrows in the graphs reflect the expected direction of causal influences among these variables.
+<!-- 3. What are the advantages of DAGs? -->
+DAGs offer two key advantages for modeling causality.
+Firstly, they represent causal relations in a nonparametric and fully interactive manner.
+This generality implies that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014].
+Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+These building blocks can be depicted in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].
 
-<!-- 3.2 the magnified DAG -->
-In contrast, the middle panels of @fig-dags_scms present the building blocks using the *magnified* representation.
-These graphs incorporate *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$, depicted as open circles to signify their unobserved nature.
-Exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
+<!-- 3. how DAGs are depicted? -->
+<!-- 3.1 the magnified DAG -->
+The left panels of @fig-dags_scms illustrate the *magnified* representation.
+These graphs depict the *endogenous* variables $V=\{X,Z,Y\}$ alongside the *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$.
+Endogenous variables are those whose causal mechanisms the investigator chooses to model [@Neal_2020].
+In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
 27,68].
-Thus, they can be viewed as the 'idiosyncratic' causes of each variable.
-Although these variables are often omitted for simplicity, as illustrated in the left panels, including them in the graphs can be beneficial, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the next section.
+The graphs show endogenous variables as solid black circles to signify that they are observed random variables, while endogenous variables are depicted as open circles to signify their unobserved (latent) nature.
+Lastly, the arrows in the graphs reflect the expected direction of causal influences among these variables.
 
-<!-- 3.3 the SCM -->
-<!-- Lastly, the right panels of @fig-dags_scms present the building blocks using the SCM representation. SCMs are  -->
+<!-- 3.2 the simplified DAG -->
+Often, the DAGs omit the exogenous variables for simplicity, resulting in the *standard* representation.
+However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the following section.
+The standard representation is illustrated in the middle panels of @fig-dags_scms.
 
@@
 
+<!-- 3.3 the SCM -->
+Lastly, the right panels of @fig-dags_scms depict the SCM representation of the five fundamental building blocks.
+SCMs are formal mathematical models that represent causal relationships within a system or population [@Hitchcock_2024].
+An SCM is defined by the set of endogenous variables $V$, the set of exogenous variables $E$, and the set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020].
+These functions, referred to as structural equations, specify each endogenous variable as a nonparametric function of other variables.
+Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence.
+The concepts of d-separation and causal (in)dependence are explored in the following section.
+
 ::: {#fig-dags_scms layout="[ [1,1,1],[1,1,1],[1,1,1],[1,1,1],[1,1,1] ]"}
 
+![Two unconnected nodes](images/figures/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" width=80%}
+
 ![Two unconnected nodes](images/figures/sdag_bb1.png){#fig-sdag_bb1 fig-align="center" width=35%}
-
-![Two unconnected nodes](images/figures/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" width=80%}
 
@@
 
+![Two connected nodes or descendant](images/figures/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" width=80%}
+
 ![Two connected nodes or descendant](images/figures/sdag_bb2.png){#fig-sdag_bb2 fig-align="center" width=35%}
-
-![Two connected nodes or descendant](images/figures/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" width=80%}
 
@@
 
-![Chain or pipe](images/figures/sdag_bb3.png){#fig-sdag_bb3 fig-align="center" width=60%}
+![Chain or pipe](images/figures/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" width=100%}
 
-![Chain or pipe](images/figures/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" width=100%}
+![Chain or pipe](images/figures/sdag_bb3.png){#fig-sdag_bb3 fig-align="center" width=55%}
 
@@
 
-![Fork](images/figures/sdag_bb4.png){#fig-sdag_bb4 fig-align="center" width=60%}
+![Fork](images/figures/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" width=100%}
 
-![Fork](images/figures/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" width=100%}
+![Fork](images/figures/sdag_bb4.png){#fig-sdag_bb4 fig-align="center" width=55%}
 
@@
 
-![Collider or inmorality](images/figures/sdag_bb5.png){#fig-sdag_bb5 fig-align="center" width=60%}
+![Collider or inmorality](images/figures/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" width=100%}
 
-![Collider or inmorality](images/figures/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" width=100%}
+![Collider or inmorality](images/figures/sdag_bb5.png){#fig-sdag_bb5 fig-align="center" width=55%}
 
@@
 
-The five fundamental building blocks of DAGs with corresponding SCMs.
-**Note:** left panels show the standard representation of DAGs, middle panels show the magnified representation of DAGs, and the right panels show their corresponding SCMs.
+The five fundamental building blocks of DAGs.
+**Note:** left panels show the magnified representation, middle panels show the standard representation, and the right panels show their corresponding SCM form.
 :::
 
-<!-- 4. What are the advantages of DAGs? -->
-<!-- DAGs can represent causal relations in a nonparametric and fully interactive manner.
-This means that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables.
-This generality allows for modeling causality without assuming a specific functional form [@Morgan_et_al_2014].
--->
-
-<!-- 5. But what does the building blocks mean? -->
-A careful examination of the left and middle panels of @fig-dags_scms reveals the assumption inherent in these building blocks.
-Figures [-@fig-sdag_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related.
-Figures [-@fig-sdag_bb2] and [-@fig-mdag_bb2] show two connected nodes, illustrating a scenario where a parent node $X$ exerts a causal influence on a child node $Y$.
+<!-- 4. But what does the building blocks mean? -->
+A careful examination of @fig-dags_scms highlights the assumptions underlying these building blocks.
+Figures [-@fig-mdag_bb1], [-@fig-sdag_bb1], and SCM [-@fig-scm_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related.
+Figures [-@fig-mdag_bb2], [-@fig-sdag_bb2], and SCM [-@fig-scm_bb2] illustrate two connected nodes, showing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$.
 Consequently, $Y$ is considered a *descendant* of $X$.
-Figures [-@fig-sdag_bb3] and [-@fig-mdag_bb3] depict a *chain* (or *pipe*), where $X$ influences $Z$, and $Z$ influences $Y$.
+Figures [-@fig-mdag_bb3], [-@fig-sdag_bb3], and SCM [-@fig-scm_bb3] depict a *chain* or *pipe*, where $X$ influences $Z$, and $Z$ influences $Y$.
 In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$.
-Furthermore, the DAG shows that $X$ is an *ancestor* of $Y$ and that $Z$ completely *mediates* the relationship between these variables.
-Figures [-@fig-sdag_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
-In this scenario, $Z$ is a parent node of $X$ and $Y$.
-Finally, Figures [-@fig-sdag_bb5] and [-@fig-mdag_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$.
+Furthermore, the DAGs show that $X$ is an *ancestor* of $Y$ and that $Z$ fully *mediates* the relationship between these variables.
+Figures [-@fig-mdag_bb4], [-@fig-sdag_bb4], and SCM [-@fig-scm_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
+Here, $Z$ is a parent node of $X$ and $Y$.
+Finally, Figures [-@fig-mdag_bb5], [-@fig-sdag_bb5], SCM [-@fig-scm_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$.
 In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$.
+Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.
 
 <!-- 5. A motivating example -->
-Given the heuristic nature of DAGs, a motivating example can help illustrate the use of the five fundamental building blocks to construct a system's causal structure.
-This example can also help to clarify some additional conventions for drawing DAGs.
-
-Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs to be investigated.
-Additionally, the problem suggests that a variable $X$ potentially influences both $T$ and $Y$.
+Given the heuristic nature of DAGs, a motivating example can clarify how to use the five fundamental building blocks to construct the causal structure of a system.
+Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation.
+The problem suggests that a variable $X$ potentially influences both $T$ and $Y$.
+Beyond these relationships, the problem does not specify any further variables of interest.
 Such scenarios are not hard to imagine.
@@
 <!-- 6. DAG for the motivating example -->
-@fig-example presents two graphs illustrating the plausible causal structure of the motivating example.
-@fig-sdag_example1 shows the *simplified* graph, while @fig-mdag_example1 depicts the *magnified* graph of the DAG.
-A detailed examination of both figures reveals the presence of at least four of the five fundamental building blocks.
-@fig-sdag_example1 displays multiple descendants, evident in pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
-It also illustrates a fork with $X \rightarrow \{T,Y\}$.
-Similarly, @fig-mdag_example1 features multiple two-unconnected nodes, evident in the pairwise relations $\{e_{T} \dsep e_{X}\}$, $\{e_{T} \dsep e_{Y}\}$, and $\{e_{X} \dsep e_{Y}\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
-The symbol $\dsep$ denotes a concept know as *d-separation*, which roughly implies the independence of the variables.
-This concept is introduced in the next section.
+@fig-example illustrates the plausible causal structure of the motivating example.
+A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks.
+The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
+Finally, the figures illustrate a fork with $X \rightarrow \{T,Y\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
 
@@
 
-![Simplied DAG](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=35%}
+![Magnified representation](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=100%}
 
-![Magnified DAG](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=65%}
+![Standard representation](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%}
 
@@
   X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,Y,e_{Z}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Z} \\
-  e_{Z} & \dsep e_{Y}
+  T & := f_{Z}(X,e_{T}) \\
+  Y & := f_{Y}(T,X,e_{Y}) \\
+  e_{T} & \dsep e_{X} \\
+  e_{T} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Y}
 \end{aligned}
@@
 
-Collider or inmorality
+Structural causal model
 :::
@@
 ## The flow of association and causation in graphs {#sec-framework-flow}
-
-<!-- Although there are several benefits  -->
-
-<!-- 2. What does a DAG imply? -->
-<!-- Structural Causal Model (SCM) -->
-<!-- Causal models are mathematical models representing causal relationships within an individual system or population.
-They facilitate inferences about causal relationships from statistical data.
-They can teach us a good deal about the epistemology of causation, and about the relationship between causation and probability.
-They have also been applied to topics of interest to philosophers, such as the logic of counterfactuals, decision theory, and the analysis of actual causation.
--->
-
-<!-- (Structural Causal Model (SCM)) A structural causal model is a tuple of the following sets: -->
-<!-- 1. A set of endogenous variables 푉 2. A set of exogenous variables 푈 3. A set of functions 푓, one to generate each endogenous variable as a function of other variables Neal -->
 
@@
 <!-- From Lawson (2015), another way to see DCJ experiments I'd that you are grouping heterogeneous experimental units (objects) into homogeneous ones (subjects), which are your "blocks" -->
+
+<!-- one must be able to assume that these structural error terms are -->
+<!-- mutually independent of each other and of all of the other variables in the graph, Morgan -->
 

### 2024-08-06 _tex/index.tex
Built site for gh-pages

@@
 is causal inference. Specifically, the structural approach to causal
-inference uses directed acyclic graphs (DAG), and their associated
-structural causal models (SCM), to provide a graphical and formal
-representation of the causal structure of a system \citep{Neal_2020}. In
-this context, a \emph{graph} denotes a collection of nodes connected by
-edges, where nodes represent random variables. The term \emph{directed}
-indicates the edges of the graph extend from one node to another, with
-arrows showing the direction of causal influence. Moreover, the term
-\emph{acyclic} indicates the causal influences do not form a loop,
-meaning the influences do not cycle back on themselves
+inference uses directed acyclic graphs (DAG) to provide a graphical and
+formal representation of the causal structure of a system
+\citep{Neal_2020}. In this context, a \emph{graph} denotes a collection
+of nodes connected by edges, where nodes represent random variables. The
+term \emph{directed} indicates the edges of the graph extend from one
+node to another, with arrows showing the direction of causal influence.
+Moreover, the term \emph{acyclic} indicates the causal influences do not
+form a loop, meaning the influences do not cycle back on themselves
 \citep{McElreath_2020}.
 
-Regardless of complexity, DAGs can represent various causal structures
-using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. The left panels of
-Figure~\ref{fig-dags_scms} illustrate these building blocks using the
-\emph{standard} representation \citep{Morgan_et_al_2014}. In these
-graphs, the variables of interest, or \emph{endogenous} variables
-\(V=\{X,Z,Y\}\), are depicted as solid black circles to signify that
-they are observed random variables. The arrows in the graphs reflect the
+DAGs offer two key advantages for modeling causality. Firstly, they
+represent causal relations in a nonparametric and fully interactive
+manner. This generality implies that feasible causal analysis strategies
+usually do not require specifying the type of data or the nature of the
+functional dependence among variables \citep{Morgan_et_al_2014}.
+Secondly, regardless of complexity, DAGs can represent various causal
+structures using only five fundamental building blocks
+\citep{Neal_2020, McElreath_2020}. These building blocks can be depicted
+in three ways: the magnified representation, the standard
+representation, and the structural causal model form
+\citep[SCM,][]{Morgan_et_al_2014}.
+
+The left panels of Figure~\ref{fig-dags_scms} illustrate the
+\emph{magnified} representation. These graphs depict the
+\emph{endogenous} variables \(V=\{X,Z,Y\}\) alongside the
+\emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\). Endogenous
+variables are those whose causal mechanisms the investigator chooses to
+model \citep{Neal_2020}. In contrast, exogenous variables represent
+\emph{errors} or \emph{disturbances} arising from omitted factors that
+the investigator chooses not to model explicitly
+\citep[27,68]{Pearl_2009}. The graphs show endogenous variables as solid
+black circles to signify that they are observed random variables, while
+endogenous variables are depicted as open circles to signify their
+unobserved (latent) nature. Lastly, the arrows in the graphs reflect the
 expected direction of causal influences among these variables.
 
-In contrast, the middle panels of Figure~\ref{fig-dags_scms} present the
-building blocks using the \emph{magnified} representation. These graphs
-incorporate \emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\),
-depicted as open circles to signify their unobserved nature. Exogenous
-variables represent \emph{errors} or \emph{disturbances} arising from
-omitted factors that the investigator chooses not to model explicitly
-\citep[27,68]{Pearl_2009}. Thus, they can be viewed as the
-`idiosyncratic' causes of each variable. Although these variables are
-often omitted for simplicity, as illustrated in the left panels,
-including them in the graphs can be beneficial, as their presence can
-reveal potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}, concepts explored in the next section.
+Often, the DAGs omit the exogenous variables for simplicity, resulting
+in the \emph{standard} representation. However, including exogenous
+variables in a graph can be beneficial in some scenarios, as their
+presence can reveal potential issues related to conditioning and
+confounding \citep{Cinelli_et_al_2020}, concepts explored in the
+following section. The standard representation is illustrated in the
+middle panels of Figure~\ref{fig-dags_scms}.
 
@@
 
+Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
+representation of the five fundamental building blocks. SCMs are formal
+mathematical models that represent causal relationships within a system
+or population \citep{Hitchcock_2024}. An SCM is defined by the set of
+endogenous variables \(V\), the set of exogenous variables \(E\), and
+the set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
+\citep{Pearl_2009, Neal_2020}. These functions, referred to as
+structural equations, specify each endogenous variable as a
+nonparametric function of other variables. Moreover, SCMs use the symbol
+`\(:=\)' to indicate the variables' asymmetrical causal dependence and
+the symbol `\(\perp\!\!\!\perp\)' to represent \emph{d-separation},
+which roughly equates to the concept of variable independence. The
+concepts of d-separation and causal (in)dependence are explored in the
+following section.
+
 \begin{figure}
 
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb1.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb1}Two unconnected nodes}
+
+\end{minipage}%
+%
 \begin{minipage}{0.33\linewidth}
@@
 \subcaption{\label{fig-sdag_bb1}Two unconnected nodes}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb1.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb1}Two unconnected nodes}
 
@@
 
-\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb2.png}
+\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb2.png}
 
@@
 
-\subcaption{\label{fig-sdag_bb2}Two connected nodes or descendant}
+\subcaption{\label{fig-mdag_bb2}Two connected nodes or descendant}
 
@@
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb2.png}
+\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb2.png}
 
@@
 
-\subcaption{\label{fig-mdag_bb2}Two connected nodes or descendant}
+\subcaption{\label{fig-sdag_bb2}Two connected nodes or descendant}
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb3.png}
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb3.png}
 
@@
 
-\subcaption{\label{fig-sdag_bb3}Chain or pipe}
+\subcaption{\label{fig-mdag_bb3}Chain or pipe}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb3.png}
+\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb3.png}
 
@@
 
-\subcaption{\label{fig-mdag_bb3}Chain or pipe}
+\subcaption{\label{fig-sdag_bb3}Chain or pipe}
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb4.png}
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb4.png}
 
@@
 
-\subcaption{\label{fig-sdag_bb4}Fork}
+\subcaption{\label{fig-mdag_bb4}Fork}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb4.png}
+\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb4.png}
 
@@
 
-\subcaption{\label{fig-mdag_bb4}Fork}
+\subcaption{\label{fig-sdag_bb4}Fork}
 
@@
 
-\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb5.png}
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb5.png}
 
@@
 
-\subcaption{\label{fig-sdag_bb5}Collider or inmorality}
+\subcaption{\label{fig-mdag_bb5}Collider or inmorality}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb5.png}
+\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb5.png}
 
@@
 
-\subcaption{\label{fig-mdag_bb5}Collider or inmorality}
+\subcaption{\label{fig-sdag_bb5}Collider or inmorality}
 
@@
 \caption{\label{fig-dags_scms}The five fundamental building blocks of
-DAGs with corresponding SCMs. \textbf{Note:} left panels show the
-standard representation of DAGs, middle panels show the magnified
-representation of DAGs, and the right panels show their corresponding
-SCMs.}
+DAGs. \textbf{Note:} left panels show the magnified representation,
+middle panels show the standard representation, and the right panels
+show their corresponding SCM form.}
 
@@
 
-A careful examination of the left and middle panels of
-Figure~\ref{fig-dags_scms} reveals the assumption inherent in these
-building blocks. Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1}
-depict two unconnected nodes, representing a scenario where variables
-\(X\) and \(Y\) are not causally related. Figures \ref{fig-sdag_bb2} and
-\ref{fig-mdag_bb2} show two connected nodes, illustrating a scenario
-where a parent node \(X\) exerts a causal influence on a child node
-\(Y\). Consequently, \(Y\) is considered a \emph{descendant} of \(X\).
-Figures \ref{fig-sdag_bb3} and \ref{fig-mdag_bb3} depict a \emph{chain}
-(or \emph{pipe}), where \(X\) influences \(Z\), and \(Z\) influences
-\(Y\). In this configuration, \(X\) is a parent node of \(Z\), and \(Z\)
-is a parent node of \(Y\). Furthermore, the DAG shows that \(X\) is an
-\emph{ancestor} of \(Y\) and that \(Z\) completely \emph{mediates} the
-relationship between these variables. Figures \ref{fig-sdag_bb4} and
-\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
-\(Y\) are both influenced by \(Z\). In this scenario, \(Z\) is a parent
-node of \(X\) and \(Y\). Finally, Figures \ref{fig-sdag_bb5} and
-\ref{fig-mdag_bb5} depict a \emph{collider}, also known as
-\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
-of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
-related to each other but both influence \(Z\).
+A careful examination of Figure~\ref{fig-dags_scms} highlights the
+assumptions underlying these building blocks. Figures
+\ref{fig-mdag_bb1}, \ref{fig-sdag_bb1}, and SCM \ref{fig-scm_bb1} depict
+two unconnected nodes, representing a scenario where variables \(X\) and
+\(Y\) are not causally related. Figures \ref{fig-mdag_bb2},
+\ref{fig-sdag_bb2}, and SCM \ref{fig-scm_bb2} illustrate two connected
+nodes, showing a scenario where a \emph{parent} node \(X\) exerts a
+causal influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
+considered a \emph{descendant} of \(X\). Figures \ref{fig-mdag_bb3},
+\ref{fig-sdag_bb3}, and SCM \ref{fig-scm_bb3} depict a \emph{chain} or
+\emph{pipe}, where \(X\) influences \(Z\), and \(Z\) influences \(Y\).
+In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
+parent node of \(Y\). Furthermore, the DAGs show that \(X\) is an
+\emph{ancestor} of \(Y\) and that \(Z\) fully \emph{mediates} the
+relationship between these variables. Figures \ref{fig-mdag_bb4},
+\ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4} illustrate a \emph{fork},
+where variables \(X\) and \(Y\) are both influenced by \(Z\). Here,
+\(Z\) is a parent node of \(X\) and \(Y\). Finally, Figures
+\ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM \ref{fig-scm_bb5} depict a
+\emph{collider}, also known as \emph{inmorality}, where variables \(X\)
+and \(Y\) are concurrent causes of \(Z\). In this configuration, \(X\)
+and \(Y\) are not causally related to each other but both influence
+\(Z\). Additionally, in all SCMs, the errors are assumed to be mutually
+independent of each other and from all other variables in the graph, as
+evidenced by the pairwise relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
+\(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).
 
-Given the heuristic nature of DAGs, a motivating example can help
-illustrate the use of the five fundamental building blocks to construct
-a system's causal structure. This example can also help to clarify some
-additional conventions for drawing DAGs.
+Given the heuristic nature of DAGs, a motivating example can clarify how
+to use the five fundamental building blocks to construct the causal
+structure of a system. Consider a research problem where the causal
+effect of a variable \(T\) on an outcome \(Y\) needs investigation. The
+problem suggests that a variable \(X\) potentially influences both \(T\)
+and \(Y\). Beyond these relationships, the problem does not specify any
+further variables of interest. Such scenarios are not hard to imagine.
+For instance, \(T\) might represent different treatments that could
+affect the recovery from cancer \(Y\), while \(X\) could denote the
+patient's age. Similarly, in the context of a DCJ study like the one
+described by \citet{Boonen_et_al_2020}, \(T\) could represent the
+duration of a child's cochlear implant use, which might influence the
+child's overall speech quality \(Y\), with \(X\) indicating the child's
+hearing status. {(not a bad example, but I prefer one using writing
+skills)}
 
-Consider a research problem where the causal effect of a variable \(T\)
-on an outcome \(Y\) needs to be investigated. Additionally, the problem
-suggests that a variable \(X\) potentially influences both \(T\) and
-\(Y\). Such scenarios are not hard to imagine. For instance, \(T\) might
-represent different treatments that could affect the recovery from
-cancer \(Y\), while \(X\) could denote the patient's age. Similarly, in
-the context of a DCJ study like the one described by
-\citet{Boonen_et_al_2020}, \(T\) could represent the duration of a
-child's cochlear implant use, which might influence the child's overall
-speech quality \(Y\), with \(X\) indicating the child's hearing status.
-{(not a bad example, but I prefer one using writing skills)}
-
-Figure~\ref{fig-example} presents two graphs illustrating the plausible
-causal structure of the motivating example.
-Figure~\ref{fig-sdag_example1} shows the \emph{simplified} graph, while
-Figure~\ref{fig-mdag_example1} depicts the \emph{magnified} graph of the
-DAG. A detailed examination of both figures reveals the presence of at
-least four of the five fundamental building blocks.
-Figure~\ref{fig-sdag_example1} displays multiple descendants, evident in
-pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
-\(T \rightarrow Y\). It also illustrates a fork with
-\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-mdag_example1}
-features multiple two-unconnected nodes, evident in the pairwise
-relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
-\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
-\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
-\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
-symbol \(\perp\!\!\!\perp\) denotes a concept know as
-\emph{d-separation}, which roughly implies the independence of the
-variables. This concept is introduced in the next section.
+Figure~\ref{fig-example} illustrates the plausible causal structure of
+the motivating example. A detailed examination of Figures
+\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
+\ref{fig-scm_example1} reveals the presence of at least four of the five
+fundamental building blocks. The figures display multiple descendants,
+as indicated by pairwise relations such as \(X \rightarrow T\),
+\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
+features multiple pairs of unconnected nodes, evident from the relations
+\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
+\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate a fork
+with \(X \rightarrow \{T,Y\}\), as well as colliders such as
+\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).
 
@@
 
-\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_example1.png}
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_example1.png}
 
@@
 
-\subcaption{\label{fig-sdag_example1}Simplied DAG}
+\subcaption{\label{fig-mdag_example1}Magnified representation}
 
@@
 
-\includegraphics[width=0.65\textwidth,height=\textheight]{images/figures/mdag_example1.png}
+\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_example1.png}
 
@@
 
-\subcaption{\label{fig-mdag_example1}Magnified DAG}
+\subcaption{\label{fig-sdag_example1}Standard representation}
 
@@
   X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,Y,e_{Z}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Z} \\
-  e_{Z} & \perp\!\!\!\perp e_{Y}
+  T & := f_{Z}(X,e_{T}) \\
+  Y & := f_{Y}(T,X,e_{Y}) \\
+  e_{T} & \perp\!\!\!\perp e_{X} \\
+  e_{T} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
 \end{aligned}
@@
 
-\subcaption{\label{fig-scm_example1}Collider or inmorality}
+\subcaption{\label{fig-scm_example1}Structural causal model}
 

### 2024-08-06 index.qmd
section:background,dags; process:write

@@
 <!-- 4. But which approach are we going to use? -->
-Nevertheless, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
+However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
 DAGs serve as heuristics, effectively conveying the assumed causal structure of a system.
 A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015].
-Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].
+Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
 
@@
 DAGs offer two key advantages for modeling causality.
-Firstly, they represent causal relations in a nonparametric and fully interactive manner.
-This generality implies that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014].
+Firstly, DAGs represent causal relations in a nonparametric and fully interactive manner.
+This feature implies that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014].
 Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
@@
 <!-- 3.3 the SCM -->
-Lastly, the right panels of @fig-dags_scms depict the SCM representation of the five fundamental building blocks.
+Lastly, the right panels of @fig-dags_scms depict the SCM form of the five fundamental building blocks.
 SCMs are formal mathematical models that represent causal relationships within a system or population [@Hitchcock_2024].
-An SCM is defined by the set of endogenous variables $V$, the set of exogenous variables $E$, and the set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020].
-These functions, referred to as structural equations, specify each endogenous variable as a nonparametric function of other variables.
+An SCM is defined by a set of endogenous variables $V$, a set of exogenous variables $E$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020].
+These functions, referred to as structural equations, specify each endogenous variable as nonparametric functions of other variables.
 Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence.
@@
 In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$.
-Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.
+Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and of all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.
 
 <!-- 5. A motivating example -->
-Given the heuristic nature of DAGs, a motivating example can clarify how to use the five fundamental building blocks to construct the causal structure of a system.
+Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system.
 Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation.
@@
 Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
-Finally, the figures illustrate a fork with $X \rightarrow \{T,Y\}$, as well as colliders such as $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
+Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
 
@@
 ## The flow of association and causation in graphs {#sec-framework-flow}
-
-<!-- \begin{aligned} -->
-<!--     Y & := \; f_{Y}( X,Z,e_{Y} ) \\ -->
-<!--     X & := \; f_{X}( e_{X} ) \\ -->
-<!--     Z & := \; f_{Z}( e_{Z} ) \\ -->
-<!--     e_{X} & \dsep e_{Z} \\ -->
-<!--     e_{X} & \dsep e_{Y} \\ -->
-<!--     e_{Z} & \dsep e_{Y} -->
-<!-- \end{aligned} -->
-
-<!-- these are non-parametrical structural equations -->
 

### 2024-08-06 _tex/index.tex
Built site for gh-pages

@@
 
-Nevertheless, the structural approach offers an additional key advantage
-over the potential outcomes approach: it enables the graphical
-representation of any system through directed acyclic graphs
+However, the structural approach offers an additional key advantage over
+the potential outcomes approach: it enables the graphical representation
+of any system through directed acyclic graphs
 \citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
@@
 models can provide valid causal inferences, assuming the causal
-structure depicted in the DAG is accurate \citep{McElreath_2020}.
+structure depicted in the DAGs are accurate \citep{McElreath_2020}.
 
@@
 
-DAGs offer two key advantages for modeling causality. Firstly, they
+DAGs offer two key advantages for modeling causality. Firstly, DAGs
 represent causal relations in a nonparametric and fully interactive
-manner. This generality implies that feasible causal analysis strategies
+manner. This feature implies that feasible causal analysis strategies
 usually do not require specifying the type of data or the nature of the
@@
 Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
-representation of the five fundamental building blocks. SCMs are formal
+form of the five fundamental building blocks. SCMs are formal
 mathematical models that represent causal relationships within a system
-or population \citep{Hitchcock_2024}. An SCM is defined by the set of
-endogenous variables \(V\), the set of exogenous variables \(E\), and
-the set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
+or population \citep{Hitchcock_2024}. An SCM is defined by a set of
+endogenous variables \(V\), a set of exogenous variables \(E\), and a
+set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
 \citep{Pearl_2009, Neal_2020}. These functions, referred to as
-structural equations, specify each endogenous variable as a
-nonparametric function of other variables. Moreover, SCMs use the symbol
-`\(:=\)' to indicate the variables' asymmetrical causal dependence and
-the symbol `\(\perp\!\!\!\perp\)' to represent \emph{d-separation},
-which roughly equates to the concept of variable independence. The
-concepts of d-separation and causal (in)dependence are explored in the
-following section.
+structural equations, specify each endogenous variable as nonparametric
+functions of other variables. Moreover, SCMs use the symbol `\(:=\)' to
+indicate the variables' asymmetrical causal dependence and the symbol
+`\(\perp\!\!\!\perp\)' to represent \emph{d-separation}, which roughly
+equates to the concept of variable independence. The concepts of
+d-separation and causal (in)dependence are explored in the following
+section.
 
@@
 \(Z\). Additionally, in all SCMs, the errors are assumed to be mutually
-independent of each other and from all other variables in the graph, as
+independent of each other and of all other variables in the graph, as
 evidenced by the pairwise relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
@@
 
-Given the heuristic nature of DAGs, a motivating example can clarify how
-to use the five fundamental building blocks to construct the causal
-structure of a system. Consider a research problem where the causal
-effect of a variable \(T\) on an outcome \(Y\) needs investigation. The
-problem suggests that a variable \(X\) potentially influences both \(T\)
-and \(Y\). Beyond these relationships, the problem does not specify any
-further variables of interest. Such scenarios are not hard to imagine.
-For instance, \(T\) might represent different treatments that could
-affect the recovery from cancer \(Y\), while \(X\) could denote the
-patient's age. Similarly, in the context of a DCJ study like the one
-described by \citet{Boonen_et_al_2020}, \(T\) could represent the
-duration of a child's cochlear implant use, which might influence the
-child's overall speech quality \(Y\), with \(X\) indicating the child's
-hearing status. {(not a bad example, but I prefer one using writing
-skills)}
+Given the heuristic nature of DAGs, a motivating example can help
+clarify how to use the five fundamental building blocks to construct the
+causal structure of a system. Consider a research problem where the
+causal effect of a variable \(T\) on an outcome \(Y\) needs
+investigation. The problem suggests that a variable \(X\) potentially
+influences both \(T\) and \(Y\). Beyond these relationships, the problem
+does not specify any further variables of interest. Such scenarios are
+not hard to imagine. For instance, \(T\) might represent different
+treatments that could affect the recovery from cancer \(Y\), while \(X\)
+could denote the patient's age. Similarly, in the context of a DCJ study
+like the one described by \citet{Boonen_et_al_2020}, \(T\) could
+represent the duration of a child's cochlear implant use, which might
+influence the child's overall speech quality \(Y\), with \(X\)
+indicating the child's hearing status. {(not a bad example, but I prefer
+one using writing skills)}
 
@@
 \(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
-\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate a fork
-with \(X \rightarrow \{T,Y\}\), as well as colliders such as
+\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
+fork \(X \rightarrow \{T,Y\}\), and two colliders with
 \(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).

### 2024-08-06 index.qmd
section:background,flow; process:structure

@@
 Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-These building blocks can be depicted in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].
+This further implies that to understand the causal assumptions of a complex system, one can decompose a structure into these basic building blocks and analyze the causal assumptions associated with each block individually [@McElreath_2020].
+Importantly, these building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].
 
@@
 
-<!-- Causal mechanism as a result of markov and minimality assumption -->
-<!-- Modularity assumption also -->
-
-<!-- 3. But what are those mysterious functions?, and how do they become statistical quantities? -->
+<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
 ![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=100%}
 
-<!-- 4. What is the local markov assumption? and where does it lead? -->
-<!-- 4.1 Bayesian factorization -->
+<!-- 2. What is the local markov assumption? and where does it lead? -->
+<!-- 2.1 Bayesian network factorization -->
 
-<!-- 5. What is the minimality assumption? and where does it lead? -->
+<!-- 3. What is the causal edges assumption? and where does it lead? -->
 
-<!-- 6. What is the causal edges assumption? and where does it lead? -->
+<!-- 4. Combined they are known as the minimality assumption? -->
 
-<!-- 7. putting it all together -->
+<!-- 5. Now, what are those mysterious functions?, and how do they become statistical quantities? -->
 
-<!-- 9. dependence and independence in building blocks -->
+<!-- 6. What is the modularity assumption? and where does it lead? -->
+<!-- 6.1 the Truncated factorization -->
+<!-- 6.2 The law of counterfactuals -->
 
-<!-- 9.1 Blocked path definition-->
+<!-- 7. Causal mechanism -->
+<!-- 7.3 the backdoor criterion (do you need it?) -->
 
-<!-- 9.2 d-separation -->
+<!-- 5. d-separation -->
+<!-- 5.1 Blocked path definition-->
 
-<!-- This further implies that to understand the causal assumptions of a complex system, one can decompose the structure into these basic building blocks and analyze the causal assumptions associated with each block individually [@McElreath_2020].
--->
+<!-- 6. putting it all together -->
+<!-- 6.1. Revising the building block: dependence and independence -->
 
@@
 <!-- point out that the disturbances are usually stochastic in nature -->
+
+<!-- Causal mechanism as a result of markov and minimality assumption -->
+<!-- Modularity assumption also -->
 

### 2024-08-06 _tex/index.tex
Built site for gh-pages

@@
 structures using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. These building blocks can be depicted
-in three ways: the magnified representation, the standard
+\citep{Neal_2020, McElreath_2020}. This further implies that to
+understand the causal assumptions of a complex system, one can decompose
+a structure into these basic building blocks and analyze the causal
+assumptions associated with each block individually
+\citep{McElreath_2020}. Importantly, these building blocks can be
+represented in three ways: the magnified representation, the standard
 representation, and the structural causal model form

### 2024-08-07 index.qmd
section:background,flow; process:write,figures

@@
 However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
-DAGs serve as heuristics, effectively conveying the assumed causal structure of a system.
+DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016].
 A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015].
@@
 Moreover, the term *acyclic* indicates the causal influences do not form a loop, meaning the influences do not cycle back on themselves [@McElreath_2020].
+Several examples of DAGs can be found in @fig-dags_scms.
 
 <!-- 3. What are the advantages of DAGs? -->
-DAGs offer two key advantages for modeling causality.
-Firstly, DAGs represent causal relations in a nonparametric and fully interactive manner.
-This feature implies that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014].
+DAGs offer two key advantages for modeling causal structures.
+Firstly, they represent causal relations in a nonparametric and fully interactive manner.
+This feature allows for feasible causal analysis strategies without needing the specification of the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014].
 Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-This further implies that to understand the causal assumptions of a complex system, one can decompose a structure into these basic building blocks and analyze the causal assumptions associated with each block individually [@McElreath_2020].
-Importantly, these building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].
-
-<!-- 3. how DAGs are depicted? -->
-<!-- 3.1 the magnified DAG -->
-The left panels of @fig-dags_scms illustrate the *magnified* representation.
-These graphs depict the *endogenous* variables $V=\{X,Z,Y\}$ alongside the *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$.
-Endogenous variables are those whose causal mechanisms the investigator chooses to model [@Neal_2020].
-In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
-27,68].
-The graphs show endogenous variables as solid black circles to signify that they are observed random variables, while endogenous variables are depicted as open circles to signify their unobserved (latent) nature.
-Lastly, the arrows in the graphs reflect the expected direction of causal influences among these variables.
-
-<!-- 3.2 the simplified DAG -->
-Often, the DAGs omit the exogenous variables for simplicity, resulting in the *standard* representation.
-However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the following section.
-The standard representation is illustrated in the middle panels of @fig-dags_scms.
+This feature enables the decomposition of complex structures into basic building blocks, facilitating the analysis of these structures by focusing on the causal assumptions associated with each building block individually [@McElreath_2020].
+These building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].
 
@@
 \newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
-
-<!-- 3.3 the SCM -->
-Lastly, the right panels of @fig-dags_scms depict the SCM form of the five fundamental building blocks.
-SCMs are formal mathematical models that represent causal relationships within a system or population [@Hitchcock_2024].
-An SCM is defined by a set of endogenous variables $V$, a set of exogenous variables $E$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020].
-These functions, referred to as structural equations, specify each endogenous variable as nonparametric functions of other variables.
-Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence.
-The concepts of d-separation and causal (in)dependence are explored in the following section.
 
@@
 
+<!-- 3. how DAGs are depicted? -->
+<!-- 3.1 the magnified DAG -->
+The left panels of @fig-dags_scms illustrate the *magnified* representation.
+These graphs depict the *endogenous* variables $V=\{X,Z,Y\}$ alongside the *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$.
+Endogenous variables are those whose causal mechanisms the investigator chooses to model [@Neal_2020].
+In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
+27,68].
+The graphs show endogenous variables as solid black circles to signify that they are observed random variables, while endogenous variables are depicted as open circles to signify their unobserved (latent) nature.
+Lastly, the arrows in the graphs reflect the expected direction of causal influences among these variables.
+
+<!-- 3.2 the simplified DAG -->
+Often, DAGs omit the exogenous variables for simplicity, resulting in the *standard* representation.
+However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the following section.
+The standard representation is illustrated in the middle panels of @fig-dags_scms.
+
+<!-- 3.3 the SCM -->
+Lastly, the right panels of @fig-dags_scms depict the SCM form of the fundamental building blocks.
+SCMs are formal mathematical models defined by a set of endogenous variables $V$, a set of exogenous variables $E$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020].
+These functions, referred to as structural equations, specify each endogenous variable as nonparametric functions of other variables.
+Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence.
+The concepts of d-separation and causal (in)dependence are explored in the following section.
+
 <!-- 4. But what does the building blocks mean? -->
@@
 In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$.
-Furthermore, the DAGs show that $X$ is an *ancestor* of $Y$ and that $Z$ fully *mediates* the relationship between these variables.
+This creates a *directed path* between $X$ and $Y$.
+Consequently, $X$ is an *ancestor* of $Y$, and $Z$ fully *mediates* the relationship between the two.
 Figures [-@fig-mdag_bb4], [-@fig-sdag_bb4], and SCM [-@fig-scm_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
@@
 
-<!-- 5. A motivating example -->
+## The flow of association and causation in graphs {#sec-framework-flow}
+
+<!-- 1. What has been said about DAGs and SCMs -->
+<!-- So far, the explanation of DAGs and SCMs has been fairly concise, sufficient to emphasize their effectiveness in encoding causal structures.
+However, nothing has been said about the assumptions required to use DAGs and SCMs as encoders of causal assumptions.
+This is important because, while encoding causal structures with DAGs can be an intriguing exercise in itself, the primary goal of an investigator is to address a research problem.
+-->
+
+<!-- This study uses the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020].
+The flowchart illustrates the process of moving from a target (causal) *estimand* to a corresponding *estimate*, through identification and estimation (@fig-IEflowchart).
+-->
+
+<!-- Before tackling the main objective of this study, however, it is important to understand how researchers can move from causal estimands to statistical estimates.
+Causal estimands, also known as targets of inference, are the quantities that the researcher wants to estimate.
+Usually, these quantities are articulated through different research questions and they are depicted using the language of the Potential Outcomes framework, Neyman-Rubin causal framework [@Neyman_et_al_1990; @Rubin_1974] mathematical expectation, e.g., $E[Y_{i}(1) - Y_{i}(0)]$ indicating that a researcher is interested in estimating the *average treatment effect* (ATE) -->
+
+<!-- Specifically, the investigator must define tthe different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method.
+The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data.
+Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets.
+Bayesian statistical methods offer an intuitive approach for the task.
+Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands.
+Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis.
+-->
+
+<!-- translate the encoded causal information into a probabilistic statistical model.
+Then, using this model along with data, the investigator must produce an estimate that addresses the research problem of interest.
+This procedure is illustrated in the Identification-Estimation flowchart presented in @fig-IEflow.
+-->
+
+<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*.
+-->
+
+![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+
+<!-- 2. What is the local markov assumption? and where does it lead? -->
+<!-- 2.1 Bayesian network factorization -->
+
+<!-- 3. What is the causal edges assumption? and where does it lead? -->
+
+<!-- 4. Combined they are known as the minimality assumption? -->
+
+<!-- 5. Now, what are those mysterious functions?, and how do they become statistical quantities? -->
+
+<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
+![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=90%}
+
+<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->
+
+<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables -->
+<!-- with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
+-->
+
+<!-- 6. What is the modularity assumption? and where does it lead? -->
+<!-- 6.1 the Truncated factorization -->
+<!-- 6.2 The law of counterfactuals -->
+
+<!-- 7. Causal mechanism -->
+<!-- 7.3 the backdoor criterion (do you need it?) -->
+
+<!-- 5. d-separation -->
+<!-- 5.1 Blocked path definition-->
+
+<!-- 6. putting it all together -->
+<!-- 6.1. Revising the building block: dependence and independence -->
+
+<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
+
+<!-- Mention convenient R package daggitty to draw dags and extract the causal assumption inherent in a graph -->
+
+<!-- After detailing how to build graphs and what they imply, Include steps to draw dags using a motivating example -->
+
+<!-- point out that the disturbances are usually stochastic in nature -->
+
+<!-- Causal mechanism as a result of markov and minimality assumption -->
+<!-- Modularity assumption also -->
+
+<!-- The previous chapter gives graphical intuition for causal models, but it doesn’t explain how to identify causal quantities and formalize causal models. Neal -->
+
+<!-- 4. But how do we move from targets of inference to estimates -->
+<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->
+
+<!-- Ardmed with dags and scm, and theit probabilité implication, we can now expand on thé motivating example provided in section xx -->
+
+## A motivating example {#sec-framework-example}
+
 Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system.
@@
 :::
-
-## The flow of association and causation in graphs {#sec-framework-flow}
-
-<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
-![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=100%}
-
-<!-- 2. What is the local markov assumption? and where does it lead? -->
-<!-- 2.1 Bayesian network factorization -->
-
-<!-- 3. What is the causal edges assumption? and where does it lead? -->
-
-<!-- 4. Combined they are known as the minimality assumption? -->
-
-<!-- 5. Now, what are those mysterious functions?, and how do they become statistical quantities? -->
-
-<!-- 6. What is the modularity assumption? and where does it lead? -->
-<!-- 6.1 the Truncated factorization -->
-<!-- 6.2 The law of counterfactuals -->
-
-<!-- 7. Causal mechanism -->
-<!-- 7.3 the backdoor criterion (do you need it?) -->
-
-<!-- 5. d-separation -->
-<!-- 5.1 Blocked path definition-->
-
-<!-- 6. putting it all together -->
-<!-- 6.1. Revising the building block: dependence and independence -->
-
-<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
-
-<!-- Mention convenient R package daggitty to draw dags and extract the causal assumption inherent in a graph -->
-
-<!-- After detailing how to build graphs and what they imply, Include steps to draw dags using a motivating example -->
-
-<!-- point out that the disturbances are usually stochastic in nature -->
-
-<!-- Causal mechanism as a result of markov and minimality assumption -->
-<!-- Modularity assumption also -->
-
-## But where does it all fit? {#sec-background-where}
-
-<!-- The previous chapter gives graphical intuition for causal models, but it doesn’t explain how to identify causal quantities and formalize causal models. Neal -->
-
-<!-- 4. But how do we move from targets of inference to estimates -->
-<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->
-
-<!-- Before tackling the main objective of this study, however, it is important to understand how researchers can move from causal estimands to statistical estimates.
-Causal estimands, also known as targets of inference, are the quantities that the researcher wants to estimate.
-Usually, these quantities are articulated through different research questions and they are depicted using the language of the Potential Outcomes framework, Neyman-Rubin causal framework [@Neyman_et_al_1990; @Rubin_1974] mathematical expectation, e.g., $E[Y_{i}(1) - Y_{i}(0)]$ indicating that a researcher is interested in estimating the *average treatment effect* (ATE) -->
-
-<!-- This study uses the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020].
-The flowchart illustrates the process of moving from a target (causal) *estimand* to a corresponding *estimate*, through identification and estimation (@fig-IEflowchart).
--->
-
-<!-- Initially, it defines the different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method.
-The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data.
-Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets.
-Bayesian statistical methods offer an intuitive approach for the task.
-Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands.
-Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis.
--->
-
-<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*.
--->
-
-![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
-
-<!-- Ardmed with dags and scm, and theit probabilité implication, we can now expand on thé motivating example provided in section xx -->
 

### 2024-08-07 _tex/index.tex
Built site for gh-pages

--- 2024-08-06 _tex/index.tex
@@
 \citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
-effectively conveying the assumed causal structure of a system. A
-heuristic is a strategy that simplifies information to make decisions
-more quickly, efficiently, and sometimes more accurately than complex
-methods \citep{Chow_2015}. Consequently, DAGs do not represent detailed
-statistical models but allow researchers to deduce which statistical
-models can provide valid causal inferences, assuming the causal
-structure depicted in the DAGs are accurate \citep{McElreath_2020}.
+effectively conveying the assumed causal structure of a system
+\citep{Pearl_et_al_2016}. A heuristic is a strategy that simplifies
+information to make decisions more quickly, efficiently, and sometimes
+more accurately than complex methods \citep{Chow_2015}. Consequently,
+DAGs do not represent detailed statistical models but allow researchers
+to deduce which statistical models can provide valid causal inferences,
+assuming the causal structure depicted in the DAGs are accurate
+\citep{McElreath_2020}.
 
@@
 form a loop, meaning the influences do not cycle back on themselves
-\citep{McElreath_2020}.
+\citep{McElreath_2020}. Several examples of DAGs can be found in
+Figure~\ref{fig-dags_scms}.
 
-DAGs offer two key advantages for modeling causality. Firstly, DAGs
-represent causal relations in a nonparametric and fully interactive
-manner. This feature implies that feasible causal analysis strategies
-usually do not require specifying the type of data or the nature of the
-functional dependence among variables \citep{Morgan_et_al_2014}.
+DAGs offer two key advantages for modeling causal structures. Firstly,
+they represent causal relations in a nonparametric and fully interactive
+manner. This feature allows for feasible causal analysis strategies
+without needing the specification of the type of data or the nature of
+the functional dependence among variables \citep{Morgan_et_al_2014}.
 Secondly, regardless of complexity, DAGs can represent various causal
 structures using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. These building blocks can be depicted
-in three ways: the magnified representation, the standard
-representation, and the structural causal model form
-\citep[SCM,][]{Morgan_et_al_2014}.
-
-The left panels of Figure~\ref{fig-dags_scms} illustrate the
-\emph{magnified} representation. These graphs depict the
-\emph{endogenous} variables \(V=\{X,Z,Y\}\) alongside the
-\emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\). Endogenous
-variables are those whose causal mechanisms the investigator chooses to
-model \citep{Neal_2020}. In contrast, exogenous variables represent
-\emph{errors} or \emph{disturbances} arising from omitted factors that
-the investigator chooses not to model explicitly
-\citep[27,68]{Pearl_2009}. The graphs show endogenous variables as solid
-black circles to signify that they are observed random variables, while
-endogenous variables are depicted as open circles to signify their
-unobserved (latent) nature. Lastly, the arrows in the graphs reflect the
-expected direction of causal influences among these variables.
-
-Often, the DAGs omit the exogenous variables for simplicity, resulting
-in the \emph{standard} representation. However, including exogenous
-variables in a graph can be beneficial in some scenarios, as their
-presence can reveal potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}, concepts explored in the
-following section. The standard representation is illustrated in the
-middle panels of Figure~\ref{fig-dags_scms}.
+\citep{Neal_2020, McElreath_2020}. This feature enables the
+decomposition of complex structures into basic building blocks,
+facilitating the analysis of these structures by focusing on the causal
+assumptions associated with each building block individually
+\citep{McElreath_2020}. These building blocks can be represented in
+three ways: the magnified representation, the standard representation,
+and the structural causal model form \citep[SCM,][]{Morgan_et_al_2014}.
 
@@
 \newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
-
-Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
-form of the five fundamental building blocks. SCMs are formal
-mathematical models that represent causal relationships within a system
-or population \citep{Hitchcock_2024}. An SCM is defined by a set of
-endogenous variables \(V\), a set of exogenous variables \(E\), and a
-set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
-\citep{Pearl_2009, Neal_2020}. These functions, referred to as
-structural equations, specify each endogenous variable as nonparametric
-functions of other variables. Moreover, SCMs use the symbol `\(:=\)' to
-indicate the variables' asymmetrical causal dependence and the symbol
-`\(\perp\!\!\!\perp\)' to represent \emph{d-separation}, which roughly
-equates to the concept of variable independence. The concepts of
-d-separation and causal (in)dependence are explored in the following
-section.
 
@@
 
+The left panels of Figure~\ref{fig-dags_scms} illustrate the
+\emph{magnified} representation. These graphs depict the
+\emph{endogenous} variables \(V=\{X,Z,Y\}\) alongside the
+\emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\). Endogenous
+variables are those whose causal mechanisms the investigator chooses to
+model \citep{Neal_2020}. In contrast, exogenous variables represent
+\emph{errors} or \emph{disturbances} arising from omitted factors that
+the investigator chooses not to model explicitly
+\citep[27,68]{Pearl_2009}. The graphs show endogenous variables as solid
+black circles to signify that they are observed random variables, while
+endogenous variables are depicted as open circles to signify their
+unobserved (latent) nature. Lastly, the arrows in the graphs reflect the
+expected direction of causal influences among these variables.
+
+Often, DAGs omit the exogenous variables for simplicity, resulting in
+the \emph{standard} representation. However, including exogenous
+variables in a graph can be beneficial in some scenarios, as their
+presence can reveal potential issues related to conditioning and
+confounding \citep{Cinelli_et_al_2020}, concepts explored in the
+following section. The standard representation is illustrated in the
+middle panels of Figure~\ref{fig-dags_scms}.
+
+Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
+form of the fundamental building blocks. SCMs are formal mathematical
+models defined by a set of endogenous variables \(V\), a set of
+exogenous variables \(E\), and a set of functions
+\(F=\{f_{X},f_{Z},f_{Y}\}\) \citep{Pearl_2009, Neal_2020}. These
+functions, referred to as structural equations, specify each endogenous
+variable as nonparametric functions of other variables. Moreover, SCMs
+use the symbol `\(:=\)' to indicate the variables' asymmetrical causal
+dependence and the symbol `\(\perp\!\!\!\perp\)' to represent
+\emph{d-separation}, which roughly equates to the concept of variable
+independence. The concepts of d-separation and causal (in)dependence are
+explored in the following section.
+
 A careful examination of Figure~\ref{fig-dags_scms} highlights the
@@
 In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
-parent node of \(Y\). Furthermore, the DAGs show that \(X\) is an
-\emph{ancestor} of \(Y\) and that \(Z\) fully \emph{mediates} the
-relationship between these variables. Figures \ref{fig-mdag_bb4},
-\ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4} illustrate a \emph{fork},
-where variables \(X\) and \(Y\) are both influenced by \(Z\). Here,
-\(Z\) is a parent node of \(X\) and \(Y\). Finally, Figures
-\ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM \ref{fig-scm_bb5} depict a
-\emph{collider}, also known as \emph{inmorality}, where variables \(X\)
-and \(Y\) are concurrent causes of \(Z\). In this configuration, \(X\)
-and \(Y\) are not causally related to each other but both influence
-\(Z\). Additionally, in all SCMs, the errors are assumed to be mutually
-independent of each other and of all other variables in the graph, as
-evidenced by the pairwise relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
+parent node of \(Y\). This creates a \emph{directed path} between \(X\)
+and \(Y\). Consequently, \(X\) is an \emph{ancestor} of \(Y\), and \(Z\)
+fully \emph{mediates} the relationship between the two. Figures
+\ref{fig-mdag_bb4}, \ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4}
+illustrate a \emph{fork}, where variables \(X\) and \(Y\) are both
+influenced by \(Z\). Here, \(Z\) is a parent node of \(X\) and \(Y\).
+Finally, Figures \ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM
+\ref{fig-scm_bb5} depict a \emph{collider}, also known as
+\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
+of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
+related to each other but both influence \(Z\). Additionally, in all
+SCMs, the errors are assumed to be mutually independent of each other
+and of all other variables in the graph, as evidenced by the pairwise
+relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
 \(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).
+
+\subsection{The flow of association and causation in
+graphs}\label{sec-framework-flow}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
+
+}
+
+\caption{\label{fig-IEflow}Identification-Estimation flowchart.
+Extracted from \citet[32]{Neal_2020}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.9\textwidth,height=\textheight]{images/figures/ACflow.png}
+
+}
+
+\caption{\label{fig-ACflow}The flow of association and causation in
+graphs. Extracted from \citet[31]{Neal_2020}}
+
+\end{figure}%
+
+\subsection{A motivating example}\label{sec-framework-example}
 
@@
 system.}
-
-\end{figure}%
-
-\subsection{The flow of association and causation in
-graphs}\label{sec-framework-flow}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/ACflow.png}
-
-}
-
-\caption{\label{fig-ACflow}The flow of association and causation in
-graphs. Extracted from \citet[31]{Neal_2020}}
-
-\end{figure}%
-
-\subsection{But where does it all fit?}\label{sec-background-where}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
-
-}
-
-\caption{\label{fig-IEflow}Identification-Estimation flowchart.
-Extracted from \citet[32]{Neal_2020}}
 

### 2024-08-08 index.qmd
section:background,empirical; process:write,figures

@@
 
+## The empirical research endeavor {#sec-framework-empirical}
+
+<!-- 1. What is the purpose of the research endeavor? -->
+The goal of an investigator is to generate new knowledge or deepen the understanding of specific topics or issues.
+Investigators accomplish this by addressing research problems.
+Research problems fall into two categories: empirical and non-empirical research.
+The first tackles real-world challenges, relying on evidence gathered through observation and experimentation.
+The latter focuses on conceptual inquiries that enhance understanding complex concepts, theories, and phenomena.
+
+<!-- 2. How to approach a research problem?  -->
+<!-- 2.1 the role of a research question -->
+Addressing research problems requires formulating well-defined research questions that provide focus, clarity, and structure to a study.
+In empirical research, these questions are often framed as *targets of inference* or *estimands*, representing the quantities the investigator aims to estimate.
+For instance, an investigator might want to examine how the duration of cochlear implant use affects the average overall speech quality of a child.
+
+<!-- 2.2 what else do you need? -->
+Furthermore, in empirical research, investigators must design appropriate methodologies, collect data, and ensure a systematic and coherent analysis of the results.
+To achieve this, they must accurately identify the factors influencing an outcome and correctly attribute the effects to each factor.
+Consequently, understanding how an investigator transitions from a target of inference to generating statistical estimates is crucial.
+
+<!-- 3. So, how do we move from targets of inference to estimates -->
+@fig-IEflow shows the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], which illustrates the mechanism for transitioning from a estimand to a corresponding *statistical estimate*, through identification and estimation.
+
+<!-- Specifically, the figure an investigator must define the different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method.
+The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data.
+Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets.
+Bayesian statistical methods offer an intuitive approach for the task.
+Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands.
+Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis.
+-->
+
+<!-- translate the encoded causal information into a probabilistic statistical model.
+Then, using this model along with data, the investigator must produce an estimate that addresses the research problem of interest.
+This procedure is illustrated in the Identification-Estimation flowchart presented in @fig-IEflow.
+-->
+
+<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*.
+-->
+
+![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+
 ## The structural approach to causal inference {#sec-framework-structural}
@@
 <!-- 1. What is causal inference? -->
+<!-- Causal inference is essential to the empirical research endeavor, as often, investigators want to make causal claims, rather than merely associational claims.  -->
+
 In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
@@
 
-<!-- A Directed Acyclic Graph (DAG) allows for the transparent depiction of a system’s complexity, revealing its underlying assumptions and the plausible mechanisms generating the system’s outcome, assuming the DAG is true.
--->
-
 ## DAGs and SCMs {#sec-framework-dag}
@@
 Moreover, the term *acyclic* indicates the causal influences do not form a loop, meaning the influences do not cycle back on themselves [@McElreath_2020].
-Several examples of DAGs can be found in @fig-dags_scms.
 
@@
 Often, DAGs omit the exogenous variables for simplicity, resulting in the *standard* representation.
-However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the following section.
+However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in @sec-framework-flow.
 The standard representation is illustrated in the middle panels of @fig-dags_scms.
@@
 Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence.
-The concepts of d-separation and causal (in)dependence are explored in the following section.
+The concepts of d-separation and causal (in)dependence are explored in @sec-framework-flow.
 
@@
 
-<!-- 1. What has been said about DAGs and SCMs -->
-<!-- So far, the explanation of DAGs and SCMs has been fairly concise, sufficient to emphasize their effectiveness in encoding causal structures.
-However, nothing has been said about the assumptions required to use DAGs and SCMs as encoders of causal assumptions.
-This is important because, while encoding causal structures with DAGs can be an intriguing exercise in itself, the primary goal of an investigator is to address a research problem.
+<!-- 1. What has been said about DAGs? -->
+<!-- The previous section gives a concise graphical intuition for DAGs, sufficient to emphasize their effectiveness in encoding causal structures.
+However, so far, nothing has been said about how to identify causal quantities or how to formalize DAGs into probabilistic statistical models.
 -->
 
-<!-- This study uses the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020].
-The flowchart illustrates the process of moving from a target (causal) *estimand* to a corresponding *estimate*, through identification and estimation (@fig-IEflowchart).
--->
+<!-- “identification” in this book, we are referring to the process -->
+<!-- of moving from a causal estimand to an equivalent statistical estimand. -->
+<!-- When we say “estimation,” we are referring to the process of moving from -->
+<!-- a statistical estimand to an estimate. Neal -->
 
-<!-- Before tackling the main objective of this study, however, it is important to understand how researchers can move from causal estimands to statistical estimates.
-Causal estimands, also known as targets of inference, are the quantities that the researcher wants to estimate.
-Usually, these quantities are articulated through different research questions and they are depicted using the language of the Potential Outcomes framework, Neyman-Rubin causal framework [@Neyman_et_al_1990; @Rubin_1974] mathematical expectation, e.g., $E[Y_{i}(1) - Y_{i}(0)]$ indicating that a researcher is interested in estimating the *average treatment effect* (ATE) -->
+<!-- The previous section gives a concise graphical intuition for DAGs, sufficient to emphasize their effectiveness in encoding causal structures.
+However, it doesn’t explain how to identify causal quantities and formalize causal models.
+Neal -->
 
-<!-- Specifically, the investigator must define tthe different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method.
-The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data.
-Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets.
-Bayesian statistical methods offer an intuitive approach for the task.
-Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands.
-Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis.
--->
-
-<!-- translate the encoded causal information into a probabilistic statistical model.
-Then, using this model along with data, the investigator must produce an estimate that addresses the research problem of interest.
-This procedure is illustrated in the Identification-Estimation flowchart presented in @fig-IEflow.
--->
-
-<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*.
--->
-
-![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+<!-- However, the requirements to use DAGs as encoders of causal assumptions and how these can transform into probabilistic statistical models have not yet been discussed. -->
 
@@
 
-<!-- The previous chapter gives graphical intuition for causal models, but it doesn’t explain how to identify causal quantities and formalize causal models. Neal -->
-
 <!-- 4. But how do we move from targets of inference to estimates -->
@@
 
-Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system.
+<!-- Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system.
 Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation.
@@
 Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status.
-[(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}
+[(not a bad example, but I prefer one using writing skills)]{style="color:blue;"} -->
 
 <!-- 6. DAG for the motivating example -->
-@fig-example illustrates the plausible causal structure of the motivating example.
+<!-- @fig-example illustrates the plausible causal structure of the motivating example.
 A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks.
@@
 Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
+-->
 
-::: {#fig-example layout="[ [1,1,1],[1,1,1],[1,1,1] ]"}
+<!-- ::: {#fig-example layout="[ [1,1,1],[1,1,1],[1,1,1] ]"} -->
 
-![Magnified representation](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=100%}
+<!-- ![Magnified representation](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=100%} -->
 
-![Standard representation](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%}
+<!-- ![Standard representation](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%} -->
 
-::: {#fig-scm_example1 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{Z}(X,e_{T}) \\
-  Y & := f_{Y}(T,X,e_{Y}) \\
-  e_{T} & \dsep e_{X} \\
-  e_{T} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Y}
-\end{aligned}
-$$
+<!-- ::: {#fig-scm_example1 fig-align="center"} -->
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!--   X & := f_{X}(e_{X}) \\ -->
+<!--   T & := f_{Z}(X,e_{T}) \\ -->
+<!--   Y & := f_{Y}(T,X,e_{Y}) \\ -->
+<!--   e_{T} & \dsep e_{X} \\ -->
+<!--   e_{T} & \dsep e_{Y} \\ -->
+<!--   e_{X} & \dsep e_{Y} -->
+<!-- \end{aligned} -->
+<!-- $$ -->
 
-Structural causal model
-:::
+<!-- Structural causal model -->
+<!-- ::: -->
 
-DAGs for a plausible causal structure in a system.
-:::
+<!-- DAGs for a plausible causal structure in a system. -->
+<!-- ::: -->
 
@@
 <!-- From Lawson (2015), another way to see DCJ experiments I'd that you are grouping heterogeneous experimental units (objects) into homogeneous ones (subjects), which are your "blocks" -->
-
-<!-- one must be able to assume that these structural error terms are -->
-<!-- mutually independent of each other and of all of the other variables in the graph, Morgan -->
 

### 2024-08-08 _tex/index.tex
Built site for gh-pages

@@
 
+\subsection{The empirical research
+endeavor}\label{sec-framework-empirical}
+
+The goal of an investigator is to generate new knowledge or deepen the
+understanding of specific topics or issues. Investigators accomplish
+this by addressing research problems. Research problems fall into two
+categories: empirical and non-empirical research. The first tackles
+real-world challenges, relying on evidence gathered through observation
+and experimentation. The latter focuses on conceptual inquiries that
+enhance understanding complex concepts, theories, and phenomena.
+
+Addressing research problems requires formulating well-defined research
+questions that provide focus, clarity, and structure to a study. In
+empirical research, these questions are often framed as \emph{targets of
+inference} or \emph{estimands}, representing the quantities the
+investigator aims to estimate. For instance, an investigator might want
+to examine how the duration of cochlear implant use affects the average
+overall speech quality of a child.
+
+Furthermore, in empirical research, investigators must design
+appropriate methodologies, collect data, and ensure a systematic and
+coherent analysis of the results. To achieve this, they must accurately
+identify the factors influencing an outcome and correctly attribute the
+effects to each factor. Consequently, understanding how an investigator
+transitions from a target of inference to generating statistical
+estimates is crucial.
+
+Figure~\ref{fig-IEflow} shows the Identification-Estimation flowchart
+\citep{McElreath_2020, Neal_2020}, which illustrates the mechanism for
+transitioning from a estimand to a corresponding \emph{statistical
+estimate}, through identification and estimation.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
+
+}
+
+\caption{\label{fig-IEflow}Identification-Estimation flowchart.
+Extracted from \citet[32]{Neal_2020}}
+
+\end{figure}%
+
 \subsection{The structural approach to causal
@@
 form a loop, meaning the influences do not cycle back on themselves
-\citep{McElreath_2020}. Several examples of DAGs can be found in
-Figure~\ref{fig-dags_scms}.
+\citep{McElreath_2020}.
 
@@
 presence can reveal potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}, concepts explored in the
-following section. The standard representation is illustrated in the
-middle panels of Figure~\ref{fig-dags_scms}.
+confounding \citep{Cinelli_et_al_2020}, concepts explored in
+Section~\ref{sec-framework-flow}. The standard representation is
+illustrated in the middle panels of Figure~\ref{fig-dags_scms}.
 
@@
 independence. The concepts of d-separation and causal (in)dependence are
-explored in the following section.
+explored in Section~\ref{sec-framework-flow}.
 
@@
 
-\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
-
-}
-
-\caption{\label{fig-IEflow}Identification-Estimation flowchart.
-Extracted from \citet[32]{Neal_2020}}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
 \includegraphics[width=0.9\textwidth,height=\textheight]{images/figures/ACflow.png}
@@
 \subsection{A motivating example}\label{sec-framework-example}
-
-Given the heuristic nature of DAGs, a motivating example can help
-clarify how to use the five fundamental building blocks to construct the
-causal structure of a system. Consider a research problem where the
-causal effect of a variable \(T\) on an outcome \(Y\) needs
-investigation. The problem suggests that a variable \(X\) potentially
-influences both \(T\) and \(Y\). Beyond these relationships, the problem
-does not specify any further variables of interest. Such scenarios are
-not hard to imagine. For instance, \(T\) might represent different
-treatments that could affect the recovery from cancer \(Y\), while \(X\)
-could denote the patient's age. Similarly, in the context of a DCJ study
-like the one described by \citet{Boonen_et_al_2020}, \(T\) could
-represent the duration of a child's cochlear implant use, which might
-influence the child's overall speech quality \(Y\), with \(X\)
-indicating the child's hearing status. {(not a bad example, but I prefer
-one using writing skills)}
-
-Figure~\ref{fig-example} illustrates the plausible causal structure of
-the motivating example. A detailed examination of Figures
-\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
-\ref{fig-scm_example1} reveals the presence of at least four of the five
-fundamental building blocks. The figures display multiple descendants,
-as indicated by pairwise relations such as \(X \rightarrow T\),
-\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
-features multiple pairs of unconnected nodes, evident from the relations
-\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
-\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
-fork \(X \rightarrow \{T,Y\}\), and two colliders with
-\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).
-
-\begin{figure}
-
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_example1.png}
-
-}
-
-\subcaption{\label{fig-mdag_example1}Magnified representation}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_example1.png}
-
-}
-
-\subcaption{\label{fig-sdag_example1}Standard representation}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{Z}(X,e_{T}) \\
-  Y & := f_{Y}(T,X,e_{Y}) \\
-  e_{T} & \perp\!\!\!\perp e_{X} \\
-  e_{T} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_example1}Structural causal model}
-
-\end{minipage}%
-
-\caption{\label{fig-example}DAGs for a plausible causal structure in a
-system.}
-
-\end{figure}%
 

### 2024-08-13 index.qmd
section:background,structural; process:write,figure

@@
 Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome.
-For instance, @vanDaal_et_al_2016 assessed academic writing skills of university students (units) using multiple argumentative essays (sub-units).
+For example, in @vanDaal_et_al_2016, the authors assessed the academic writing skills of university students (units) using multiple argumentative essays (sub-units).
 
@@
 This study aims to fill this gap by representing DCJ within the framework of causal inference.
-Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
+Specifically, utilizing the structural approach to causal inference, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
 The study then translates the scientific model into a probabilistic statistical model.
@@
 They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful.
-For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.
-
-## The empirical research endeavor {#sec-framework-empirical}
-
-<!-- 1. What is the purpose of the research endeavor? -->
-The goal of an investigator is to generate new knowledge or deepen the understanding of specific topics or issues.
-Investigators accomplish this by addressing research problems.
-Research problems fall into two categories: empirical and non-empirical research.
-The first tackles real-world challenges, relying on evidence gathered through observation and experimentation.
-The latter focuses on conceptual inquiries that enhance understanding complex concepts, theories, and phenomena.
-
-<!-- 2. How to approach a research problem?  -->
-<!-- 2.1 the role of a research question -->
-Addressing research problems requires formulating well-defined research questions that provide focus, clarity, and structure to a study.
-In empirical research, these questions are often framed as *targets of inference* or *estimands*, representing the quantities the investigator aims to estimate.
-For instance, an investigator might want to examine how the duration of cochlear implant use affects the average overall speech quality of a child.
-
-<!-- 2.2 what else do you need? -->
-Furthermore, in empirical research, investigators must design appropriate methodologies, collect data, and ensure a systematic and coherent analysis of the results.
-To achieve this, they must accurately identify the factors influencing an outcome and correctly attribute the effects to each factor.
-Consequently, understanding how an investigator transitions from a target of inference to generating statistical estimates is crucial.
-
-<!-- 3. So, how do we move from targets of inference to estimates -->
-@fig-IEflow shows the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], which illustrates the mechanism for transitioning from a estimand to a corresponding *statistical estimate*, through identification and estimation.
-
-<!-- Specifically, the figure an investigator must define the different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method.
-The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data.
-Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets.
-Bayesian statistical methods offer an intuitive approach for the task.
-Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands.
-Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis.
--->
-
-<!-- translate the encoded causal information into a probabilistic statistical model.
-Then, using this model along with data, the investigator must produce an estimate that addresses the research problem of interest.
-This procedure is illustrated in the Identification-Estimation flowchart presented in @fig-IEflow.
--->
-
-<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*.
--->
-
-![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books like @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.
 
@@
 
-<!-- 1. What is causal inference? -->
-<!-- Causal inference is essential to the empirical research endeavor, as often, investigators want to make causal claims, rather than merely associational claims.  -->
+<!-- 1. What does empirical research do first? -->
+Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation.
+In this context, researchers typically frame their research questions as *estimands* or *targets of inference*.
+These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010].
+For instance, a study might seek to answer the question, "To what extent do daily practice hours ($X$) influence students' ability to produce high-quality written texts ($Y$)?".
+To investigate this, the study randomly assigns students to two groups with different levels of daily practice, low $(X=1)$ and medium $(X=2)$.
+The quality of the written texts is evaluated using pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$.
+The research question is then framed as the estimand, "*On average*, which group of students produces higher-quality written texts?" This estimand is mathematically expressed as $E[Y|X=1] - E[Y|X=2]$, where $E[\cdot]$ represents the expected value.
 
-In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical notation for analyzing causes and counterfactuals [@Pearl_2009].
+<!-- 2. What do they do next? -->
+Researchers then proceed to identify the estimands.
+*Identification* is the process of accurately computing an estimand using an estimator.
+An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
+*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010].
+While various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure accurate estimates.
+For example, Fisher's exact test [@Fisher_1922] is an estimator with desirable properties for comparing $2\times2$ tables.
+The test provides accurate estimates for comparing proportions when its assumptions are satisfied.
+@fig-IEflow, known as the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], illustrates the aforementioned process of transitioning from estimands to estimates.
 
-<!-- 2. what are counterfactuals? -->
+![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+
+<!-- 3. But we want these estimands to be causal -->
+However, many studies aim to establish causal relationships rather than merely infer associations.
+In the earlier example, the estimates obtained using Fisher's exact test can be interpreted as causal because the data were collected through a randomized experiment.
+In randomized experiments, the researcher fully controls the treatment assignment mechanism, such as assigning students to different groups.
+This control allows causal estimates to be derived from associational estimates, such as those provided by the Fisher statistic, since randomization effectively eliminates confounding [@Neal_2020].
+*Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious association [@Everitt_et_al_2010].
+
+<!-- 4. causal inference to the rescue -->
+<!-- Nevertheless, in many cases, researchers are not able to perform experiments due to many reasons.
+For instance, .
+In this scenarios, the framework of causal inference can be an alternative to generate causal estimates.
+*Causal inference* is a framework focused on identifying the causes influencing a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
+Unlike classical statistical modeling, which primarily focuses on summarizing data and inferring associations, causal inference provides a coherent mathematical framework for analyzing causes and counterfactuals [@Pearl_2009].
+-->
+
+<!-- 5. How does causal inference defines causal effects? what are counterfactuals? -->
+<!-- Causal inference defines causal effects in terms of counterfactuals.
 Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association.
+Returning to the previous example, if the research aim to determine the $X$ have an impact on an outcome $Y$ -->
+
+<!-- Causal inference approach the *fundamental problem of causal inference* -->
+
+<!-- According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association.
 Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
-They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."
+They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances." -->
 
-<!-- 3. How do we apply causal inference and counterfactuals? -->
-Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
+<!-- 6. How do we apply causal inference and counterfactuals? -->
+<!-- Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
 Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
@@
 Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].
+-->
 
-<!-- 4. But which approach are we going to use? -->
-However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
+<!-- 7. But which approach are we going to use? -->
+<!-- However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
 DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016].
@@
 Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
+-->
 
@@
 <!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
-![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=90%}
+![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=80%}
 

### 2024-08-13 _tex/index.tex
Built site for gh-pages

@@
 \emph{proximity} in terms of the assessed trait, can significantly
-influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
-academic writing skills of university students (units) using multiple
-argumentative essays (sub-units).
+influence the outcome. For example, in \citet{vanDaal_et_al_2016}, the
+authors assessed the academic writing skills of university students
+(units) using multiple argumentative essays (sub-units).
 
@@
 framework of causal inference. Specifically, utilizing the structural
-approach, the study develops a scientific model to clarify plausible
-causal assumptions and mechanisms inherent in the DCJ system. The study
-then translates the scientific model into a probabilistic statistical
-model. This model aims to produce statistical estimates to draw
-inferences about plausible causal relationships within the DCJ system.
+approach to causal inference, the study develops a scientific model to
+clarify plausible causal assumptions and mechanisms inherent in the DCJ
+system. The study then translates the scientific model into a
+probabilistic statistical model. This model aims to produce statistical
+estimates to draw inferences about plausible causal relationships within
+the DCJ system.
 
@@
 \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
-such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
+like \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
 \citet{Hernan_et_al_2020} are recommended.
 
-\subsection{The empirical research
-endeavor}\label{sec-framework-empirical}
+\subsection{The structural approach to causal
+inference}\label{sec-framework-structural}
 
-The goal of an investigator is to generate new knowledge or deepen the
-understanding of specific topics or issues. Investigators accomplish
-this by addressing research problems. Research problems fall into two
-categories: empirical and non-empirical research. The first tackles
-real-world challenges, relying on evidence gathered through observation
-and experimentation. The latter focuses on conceptual inquiries that
-enhance understanding complex concepts, theories, and phenomena.
+Empirical research addresses real-world challenges by relying on
+evidence gathered through observation and experimentation. In this
+context, researchers typically frame their research questions as
+\emph{estimands} or \emph{targets of inference}. These estimands
+represent the specific quantities the study aims to determine
+\citep{Everitt_et_al_2010}. For instance, a study might seek to answer
+the question, ``To what extent do daily practice hours (\(X\)) influence
+students' ability to produce high-quality written texts (\(Y\))?''. To
+investigate this, the study randomly assigns students to two groups with
+different levels of daily practice, low \((X=1)\) and medium \((X=2)\).
+The quality of the written texts is evaluated using pairwise
+comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\). The
+research question is then framed as the estimand, ``\emph{On average},
+which group of students produces higher-quality written texts?'' This
+estimand is mathematically expressed as \(E[Y|X=1] - E[Y|X=2]\), where
+\(E[\cdot]\) represents the expected value.
 
-Addressing research problems requires formulating well-defined research
-questions that provide focus, clarity, and structure to a study. In
-empirical research, these questions are often framed as \emph{targets of
-inference} or \emph{estimands}, representing the quantities the
-investigator aims to estimate. For instance, an investigator might want
-to examine how the duration of cochlear implant use affects the average
-overall speech quality of a child.
-
-Furthermore, in empirical research, investigators must design
-appropriate methodologies, collect data, and ensure a systematic and
-coherent analysis of the results. To achieve this, they must accurately
-identify the factors influencing an outcome and correctly attribute the
-effects to each factor. Consequently, understanding how an investigator
-transitions from a target of inference to generating statistical
-estimates is crucial.
-
-Figure~\ref{fig-IEflow} shows the Identification-Estimation flowchart
-\citep{McElreath_2020, Neal_2020}, which illustrates the mechanism for
-transitioning from a estimand to a corresponding \emph{statistical
-estimate}, through identification and estimation.
+Researchers then proceed to identify the estimands.
+\emph{Identification} is the process of accurately computing an estimand
+using an estimator. An \emph{estimator} is a method or function that
+transforms data into an estimate \citep{Neal_2020}. \emph{Estimates} are
+numerical values that approximate the estimand and are derived through
+\emph{estimation}, which is the process that integrates data with an
+estimator \citep{Everitt_et_al_2010}. While various methods can
+approximate an estimand, researchers prioritize estimators with
+desirable properties that ensure accurate estimates. For example,
+Fisher's exact test \citep{Fisher_1922} is an estimator with desirable
+properties for comparing \(2\times2\) tables. The test provides accurate
+estimates for comparing proportions when its assumptions are satisfied.
+Figure~\ref{fig-IEflow}, known as the Identification-Estimation
+flowchart \citep{McElreath_2020, Neal_2020}, illustrates the
+aforementioned process of transitioning from estimands to estimates.
 
@@
 \caption{\label{fig-IEflow}Identification-Estimation flowchart.
-Extracted from \citet[32]{Neal_2020}}
+Extracted and slightly modified from \citet[32]{Neal_2020}}
 
@@
 
-\subsection{The structural approach to causal
-inference}\label{sec-framework-structural}
-
-In statistics, \emph{causal inference} refers to the process of
-identifying the causes of a phenomenon and estimating their effects
-using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
-statistical modeling, which focuses solely on summarizing data and
-inferring associations, causal inference provides a coherent
-mathematical notation for analyzing causes and counterfactuals
-\citep{Pearl_2009}.
-
-Counterfactuals represent scenarios \emph{contrary to fact}, where
-alternative \emph{potential} outcomes resulting from a cause are neither
-observed nor observable \citep{Neal_2020, Counterfactual_2024}.
-According to \citet{Pearl_et_al_2018}, counterfactuals are the
-foundation of causal inference and occupy the highest level of cognitive
-abstraction in the ladder of causation, followed by intervention and
-association. Nevertheless, despite their abstract nature,
-counterfactuals enable the development of a \emph{theory of the world}
-that explains why specific causes have specific effects and what occurs
-in their absence \citep{Pearl_et_al_2018}. They achieve this by
-translating causal statements into counterfactual statements, that is,
-statements about ``what would have happened in the world under different
-circumstances.''
-
-Several approaches to causal inference and counterfactuals exist, but
-two are particularly prominent: the potential outcomes approach, also
-known as the Neyman-Rubin causal model
-\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
-\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
-mathematical notation to characterize causal inference, but they do so
-in different ways \citep{Neal_2020}. The potential outcomes approach
-relies on counterfactual notation, whereas the structural approach
-employs the do-operator and structural causal models
-\citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these differences,
-both notations can be expressed in terms of the other, and both
-approaches provide methods for using experimental and observational data
-to estimate causal effects \citep{Pearl_2010}.
-
-However, the structural approach offers an additional key advantage over
-the potential outcomes approach: it enables the graphical representation
-of any system through directed acyclic graphs
-\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
-effectively conveying the assumed causal structure of a system
-\citep{Pearl_et_al_2016}. A heuristic is a strategy that simplifies
-information to make decisions more quickly, efficiently, and sometimes
-more accurately than complex methods \citep{Chow_2015}. Consequently,
-DAGs do not represent detailed statistical models but allow researchers
-to deduce which statistical models can provide valid causal inferences,
-assuming the causal structure depicted in the DAGs are accurate
-\citep{McElreath_2020}.
+However, many studies aim to establish causal relationships rather than
+merely infer associations. In the earlier example, the estimates
+obtained using Fisher's exact test can be interpreted as causal because
+the data were collected through a randomized experiment. In randomized
+experiments, the researcher fully controls the treatment assignment
+mechanism, such as assigning students to different groups. This control
+allows causal estimates to be derived from associational estimates, such
+as those provided by the Fisher statistic, since randomization
+effectively eliminates confounding \citep{Neal_2020}. \emph{Confounding}
+occurs when an external variable influences both the outcome and the
+variable of interest, leading to spurious association
+\citep{Everitt_et_al_2010}.
 
@@
 
-\includegraphics[width=0.9\textwidth,height=\textheight]{images/figures/ACflow.png}
+\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/ACflow.png}
 
@@
 \caption{\label{fig-ACflow}The flow of association and causation in
-graphs. Extracted from \citet[31]{Neal_2020}}
+graphs. Extracted and slightly modified from \citet[31]{Neal_2020}}
 

### 2024-08-14 index.qmd
section:background,structural; process:write

@@
 These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010].
-For instance, a study might seek to answer the question, "To what extent do daily practice hours ($X$) influence students' ability to produce high-quality written texts ($Y$)?".
-To investigate this, the study randomly assigns students to two groups with different levels of daily practice, low $(X=1)$ and medium $(X=2)$.
-The quality of the written texts is evaluated using pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$.
-The research question is then framed as the estimand, "*On average*, which group of students produces higher-quality written texts?" This estimand is mathematically expressed as $E[Y|X=1] - E[Y|X=2]$, where $E[\cdot]$ represents the expected value.
+For instance, a study might examine the question, "To what extent do different teaching methods ($X$) influence students' conceptual understanding of algebra ($Y$)?" To investigate this, the study randomly assigns students to two groups, each using different algebra teaching method $(X=\{1,2\})$.
+Students' conceptual understanding of algebra is evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student possesses a higher degree of understanding among those compared.
+The research question is then framed as the estimand, "*On average*, is there a difference in conceptual understanding of algebra between the two groups of student?" This estimand is mathematically expressed as $E[Y| X=1] - E[Y| X=2]$, where $E[\cdot]$ denotes the expected value.
+An example of this approach is seen in @Jones_et_al_2019.
 
@@
 Researchers then proceed to identify the estimands.
-*Identification* is the process of accurately computing an estimand using an estimator.
-An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010].
-While various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure accurate estimates.
-For example, Fisher's exact test [@Fisher_1922] is an estimator with desirable properties for comparing $2\times2$ tables.
-The test provides accurate estimates for comparing proportions when its assumptions are satisfied.
-@fig-IEflow, known as the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], illustrates the aforementioned process of transitioning from estimands to estimates.
-
-![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
+*Identification* refers to the process of accurately computing an estimand using an estimator.
+An *estimator* is a method or function that maps data into an estimate [@Neal_2020].
+*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refer to the process of integrating data with an estimator [@Everitt_et_al_2010].
+Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
+For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, providing accurate estimates for the desired estimand when its assumptions are met.
 
 <!-- 3. But we want these estimands to be causal -->
-However, many studies aim to establish causal relationships rather than merely infer associations.
-In the earlier example, the estimates obtained using Fisher's exact test can be interpreted as causal because the data were collected through a randomized experiment.
-In randomized experiments, the researcher fully controls the treatment assignment mechanism, such as assigning students to different groups.
-This control allows causal estimates to be derived from associational estimates, such as those provided by the Fisher statistic, since randomization effectively eliminates confounding [@Neal_2020].
+However, many studies aim to understand the mechanisms underlying specific data types and also seek to establish causal relationships rather than merely infer associations.
+In the earlier example, the associational estimates obtained using the BTL model can be interpreted as causal, because the data were collected using a randomized experiment.
+Randomized experiments enable the causal interpretation of associational estimates by ensuring several key properties, including common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
+The most crucial property, however, is that randomization effectively eliminates confounding.
 *Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious association [@Everitt_et_al_2010].
+Randomization mitigates this issue by decoupling the intervention assignment mechanism, such as assigning students to different groups, from other variables and outcomes [@Morgan_et_al_2014; @Neal_2020].
 
 <!-- 4. causal inference to the rescue -->
-<!-- Nevertheless, in many cases, researchers are not able to perform experiments due to many reasons.
-For instance, .
-In this scenarios, the framework of causal inference can be an alternative to generate causal estimates.
-*Causal inference* is a framework focused on identifying the causes influencing a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-Unlike classical statistical modeling, which primarily focuses on summarizing data and inferring associations, causal inference provides a coherent mathematical framework for analyzing causes and counterfactuals [@Pearl_2009].
--->
+Although experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], researchers often encounter constraints that limit their ability to conduct experimental studies.
+These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020].
+In these instances, causal inference offers a valuable alternative for generating causal estimates.
+*Causal inference* is a framework designed to identify the causes of phenomena and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
+Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, causal inference provides a rigorous mathematical framework for analyzing causes and counterfactuals [@Pearl_2009].
+Moreover, the framework is beneficial for both observational and experimental studies, as it provides significant theoretical insights that enhance their design [@McElreath_2020].
 
-<!-- 5. How does causal inference defines causal effects? what are counterfactuals? -->
-<!-- Causal inference defines causal effects in terms of counterfactuals.
-Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-Returning to the previous example, if the research aim to determine the $X$ have an impact on an outcome $Y$ -->
-
-<!-- Causal inference approach the *fundamental problem of causal inference* -->
-
-<!-- According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association.
-Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018].
-They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances." -->
+<!-- 5. But how does it work? -->
+Counterfactuals represent the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association [@Pearl_et_al_2018], and a detailed explanation of causes, counterfactuals, and their applications is beyond the scope of this document.
+However, it is important to note that causal inference addresses the question, "What would have happened in the world under different circumstances?" by transforming unobserved *potential outcomes*, known as counterfactuals, into associational estimates.
+*Counterfactuals* represent scenarios *contrary to fact*, where alternative *potential outcomes* resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
+The framework achieves this transformation while ensuring the absence of confounding.
+Ultimately, the *fundamental problem of causality* is reduced to calculating differences between averages of observed outcomes and the transformed counterfactuals, thereby identifying causal effects.
 
 <!-- 6. How do we apply causal inference and counterfactuals? -->
-<!-- Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
+Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
 Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
-The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016].
+The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs do-calculus [@Pearl_2009].
 Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].
--->
 
 <!-- 7. But which approach are we going to use? -->
-<!-- However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020].
+However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical and formal representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020] and and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016].
 DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016].
@@
 Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
--->
+The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, and the application of the sicentific model and data to identify and estimates causal effects.
+
+![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
 
@@
 ## The flow of association and causation in graphs {#sec-framework-flow}
+
+<!-- The main assumption we’ve seen that relates the graph to the distribution -->
+<!-- is the Markov assumption. The Markov assumption tells us if variables are -->
+<!-- d-separated in the graph 𝐺 , then they are independent in the distribution -->
+<!-- 𝑃 (Theorem 3.1): -->
+<!-- Maybe we can detect independencies in the data and then use that -->
+<!-- to infer the causal graph. However, going from independencies in the -->
+<!-- distribution 𝑃 to d-separations in the graph 𝐺 isn’t something that the -->
+<!-- Markov assumption gives us (see Equation 3.20 above). Rather, we need -->
+<!-- the converse of the Markov assumption. This is known as the faithfulness -->
+<!-- assumption. -->
+<!-- Assumption 11.1 (Faithfulness) -->
 

### 2024-08-14 _tex/index.tex
Built site for gh-pages

@@
 represent the specific quantities the study aims to determine
-\citep{Everitt_et_al_2010}. For instance, a study might seek to answer
-the question, ``To what extent do daily practice hours (\(X\)) influence
-students' ability to produce high-quality written texts (\(Y\))?''. To
-investigate this, the study randomly assigns students to two groups with
-different levels of daily practice, low \((X=1)\) and medium \((X=2)\).
-The quality of the written texts is evaluated using pairwise
-comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\). The
-research question is then framed as the estimand, ``\emph{On average},
-which group of students produces higher-quality written texts?'' This
-estimand is mathematically expressed as \(E[Y|X=1] - E[Y|X=2]\), where
-\(E[\cdot]\) represents the expected value.
+\citep{Everitt_et_al_2010}. For instance, a study might examine the
+question, ``To what extent do different teaching methods (\(X\))
+influence students' conceptual understanding of algebra (\(Y\))?'' To
+investigate this, the study randomly assigns students to two groups,
+each using different algebra teaching method \((X=\{1,2\})\). Students'
+conceptual understanding of algebra is evaluated through pairwise
+comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\),
+indicating which student possesses a higher degree of understanding
+among those compared. The research question is then framed as the
+estimand, ``\emph{On average}, is there a difference in conceptual
+understanding of algebra between the two groups of student?'' This
+estimand is mathematically expressed as \(E[Y| X=1] - E[Y| X=2]\), where
+\(E[\cdot]\) denotes the expected value. An example of this approach is
+seen in \citet{Jones_et_al_2019}.
 
 Researchers then proceed to identify the estimands.
-\emph{Identification} is the process of accurately computing an estimand
-using an estimator. An \emph{estimator} is a method or function that
-transforms data into an estimate \citep{Neal_2020}. \emph{Estimates} are
+\emph{Identification} refers to the process of accurately computing an
+estimand using an estimator. An \emph{estimator} is a method or function
+that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
 numerical values that approximate the estimand and are derived through
-\emph{estimation}, which is the process that integrates data with an
-estimator \citep{Everitt_et_al_2010}. While various methods can
+\emph{estimation}, which refer to the process of integrating data with
+an estimator \citep{Everitt_et_al_2010}. Although various methods can
 approximate an estimand, researchers prioritize estimators with
-desirable properties that ensure accurate estimates. For example,
-Fisher's exact test \citep{Fisher_1922} is an estimator with desirable
-properties for comparing \(2\times2\) tables. The test provides accurate
-estimates for comparing proportions when its assumptions are satisfied.
-Figure~\ref{fig-IEflow}, known as the Identification-Estimation
-flowchart \citep{McElreath_2020, Neal_2020}, illustrates the
-aforementioned process of transitioning from estimands to estimates.
+desirable properties that ensure the accuracy of estimates. For
+instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
+estimator known for its effectiveness in modeling pairwise comparisons,
+providing accurate estimates for the desired estimand when its
+assumptions are met.
+
+However, many studies aim to understand the mechanisms underlying
+specific data types and also seek to establish causal relationships
+rather than merely infer associations. In the earlier example, the
+associational estimates obtained using the BTL model can be interpreted
+as causal, because the data were collected using a randomized
+experiment. Randomized experiments enable the causal interpretation of
+associational estimates by ensuring several key properties, including
+common support, no interference, and consistency
+\citep{Morgan_et_al_2014, Neal_2020}. The most crucial property,
+however, is that randomization effectively eliminates confounding.
+\emph{Confounding} occurs when an external variable influences both the
+outcome and the variable of interest, leading to spurious association
+\citep{Everitt_et_al_2010}. Randomization mitigates this issue by
+decoupling the intervention assignment mechanism, such as assigning
+students to different groups, from other variables and outcomes
+\citep{Morgan_et_al_2014, Neal_2020}.
+
+Although experiments are widely regarded as the gold standard in
+evidence-based science \citep{Hariton_et_al_2018, Hansson_2014},
+researchers often encounter constraints that limit their ability to
+conduct experimental studies. These constraints include ethical
+concerns, such as the assignment of individuals to potentially harmful
+interventions, and practical limitations, such as the infeasibility of,
+for example, assigning individuals to genetic modifications or physical
+impairments \citep{Neal_2020}. In these instances, causal inference
+offers a valuable alternative for generating causal estimates.
+\emph{Causal inference} is a framework designed to identify the causes
+of phenomena and estimate their effects using data
+\citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
+modeling, which focuses primarily on summarizing data and inferring
+associations, causal inference provides a rigorous mathematical
+framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
+Moreover, the framework is beneficial for both observational and
+experimental studies, as it provides significant theoretical insights
+that enhance their design \citep{McElreath_2020}.
+
+Counterfactuals represent the highest level of cognitive abstraction in
+the ladder of causation, followed by intervention and association
+\citep{Pearl_et_al_2018}, and a detailed explanation of causes,
+counterfactuals, and their applications is beyond the scope of this
+document. However, it is important to note that causal inference
+addresses the question, ``What would have happened in the world under
+different circumstances?'' by transforming unobserved \emph{potential
+outcomes}, known as counterfactuals, into associational estimates.
+\emph{Counterfactuals} represent scenarios \emph{contrary to fact},
+where alternative \emph{potential outcomes} resulting from a cause are
+neither observed nor observable \citep{Neal_2020, Counterfactual_2024}.
+The framework achieves this transformation while ensuring the absence of
+confounding. Ultimately, the \emph{fundamental problem of causality} is
+reduced to calculating differences between averages of observed outcomes
+and the transformed counterfactuals, thereby identifying causal effects.
+
+Several approaches to causal inference and counterfactuals exist, but
+two are particularly prominent: the potential outcomes approach, also
+known as the Neyman-Rubin causal model
+\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
+\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
+mathematical notation to characterize causal inference, but they do so
+in different ways \citep{Neal_2020}. The potential outcomes approach
+relies on counterfactual notation, whereas the structural approach
+employs do-calculus \citep{Pearl_2009}. Despite these differences, both
+notations can be expressed in terms of the other, and both approaches
+provide methods for using experimental and observational data to
+estimate causal effects \citep{Pearl_2010}.
+
+However, the structural approach offers an additional key advantage over
+the potential outcomes approach: it enables the graphical and formal
+representation of any system through directed acyclic graphs
+\citep[DAG,][]{Gross_et_al_2018, Neal_2020} and and structural causal
+models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. DAGs serve as
+heuristics, effectively conveying the assumed causal structure of a
+system \citep{Pearl_et_al_2016}. A heuristic is a strategy that
+simplifies information to make decisions more quickly, efficiently, and
+sometimes more accurately than complex methods \citep{Chow_2015}.
+Consequently, DAGs do not represent detailed statistical models but
+allow researchers to deduce which statistical models can provide valid
+causal inferences, assuming the causal structure depicted in the DAGs
+are accurate \citep{McElreath_2020}. The Identification-Estimation
+flowchart in Figure~\ref{fig-IEflow} visually represents the process of
+transitioning from estimands to estimates, and the application of the
+sicentific model and data to identify and estimates causal effects.
 
@@
 \end{figure}%
-
-However, many studies aim to establish causal relationships rather than
-merely infer associations. In the earlier example, the estimates
-obtained using Fisher's exact test can be interpreted as causal because
-the data were collected through a randomized experiment. In randomized
-experiments, the researcher fully controls the treatment assignment
-mechanism, such as assigning students to different groups. This control
-allows causal estimates to be derived from associational estimates, such
-as those provided by the Fisher statistic, since randomization
-effectively eliminates confounding \citep{Neal_2020}. \emph{Confounding}
-occurs when an external variable influences both the outcome and the
-variable of interest, leading to spurious association
-\citep{Everitt_et_al_2010}.
 

### 2024-08-16 index.qmd
section:background,structural; process:write

@@
 These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010].
-For instance, a study might examine the question, "To what extent do different teaching methods ($X$) influence students' conceptual understanding of algebra ($Y$)?" To investigate this, the study randomly assigns students to two groups, each using different algebra teaching method $(X=\{1,2\})$.
-Students' conceptual understanding of algebra is evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student possesses a higher degree of understanding among those compared.
-The research question is then framed as the estimand, "*On average*, is there a difference in conceptual understanding of algebra between the two groups of student?" This estimand is mathematically expressed as $E[Y| X=1] - E[Y| X=2]$, where $E[\cdot]$ denotes the expected value.
+For instance, a study might examine the question, “To what extent do different teaching methods ($X$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(X=\{1,2\})$.
+Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding.
+The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed as $E[Y| X=1] - E[Y| X=2]$, where $E[\cdot]$ denotes the expected value.
 An example of this approach is seen in @Jones_et_al_2019.
@@
 Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
-For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, providing accurate estimates for the desired estimand when its assumptions are met.
+For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the target of inference discussed earlier, as long as its assumptions are met.
 
@@
 <!-- 4. causal inference to the rescue -->
-Although experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], researchers often encounter constraints that limit their ability to conduct experimental studies.
+Experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], however, researchers often encounter constraints that limit their ability to conduct experimental studies.
 These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020].
@@
 Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, causal inference provides a rigorous mathematical framework for analyzing causes and counterfactuals [@Pearl_2009].
-Moreover, the framework is beneficial for both observational and experimental studies, as it provides significant theoretical insights that enhance their design [@McElreath_2020].
+Consequently, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].
 
 <!-- 5. But how does it work? -->
-Counterfactuals represent the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association [@Pearl_et_al_2018], and a detailed explanation of causes, counterfactuals, and their applications is beyond the scope of this document.
-However, it is important to note that causal inference addresses the question, "What would have happened in the world under different circumstances?" by transforming unobserved *potential outcomes*, known as counterfactuals, into associational estimates.
-*Counterfactuals* represent scenarios *contrary to fact*, where alternative *potential outcomes* resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-The framework achieves this transformation while ensuring the absence of confounding.
-Ultimately, the *fundamental problem of causality* is reduced to calculating differences between averages of observed outcomes and the transformed counterfactuals, thereby identifying causal effects.
+Although a detailed explanation of causes, and counterfactuals is beyond the scope of this document, it suffices to say that causal inference addresses the *fundamental problem of causality* through the use of counterfactuals [@Neal_2020].
+The framework defines the *individual causal effect* (ICE) as the difference between an individual's observed and unobserved potential outcomes.
+These unobserved potential outcomes, termed *counterfactuals*, represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes for an individual, resulting from a specific cause, are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
+The framework then extends this estimand to groups by introducing the *average causal effect* (ACE).
+The ACE reflects the difference between averages of observed potential outcomes and counterfactuals across groups.
+Finally, similar to an experiment, the framework identifies the ACE from associational estimates by ensuring the absence of confounding.
 
@@
 Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
-Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020].
+Both approaches employ rigorous mathematical notation to characterize the ACE, but they do so in different ways [@Neal_2020].
 The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs do-calculus [@Pearl_2009].
@@
 <!-- 7. But which approach are we going to use? -->
-However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical and formal representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020] and and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016].
-DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016].
-A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015].
-Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
-The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, and the application of the sicentific model and data to identify and estimates causal effects.
+The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of theories through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
+DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model* [@McElreath_2020].
+They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
+The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents this process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.
+
+<!-- A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. -->
 

### 2024-08-16 _tex/index.tex
Built site for gh-pages

@@
 question, ``To what extent do different teaching methods (\(X\))
-influence students' conceptual understanding of algebra (\(Y\))?'' To
-investigate this, the study randomly assigns students to two groups,
-each using different algebra teaching method \((X=\{1,2\})\). Students'
-conceptual understanding of algebra is evaluated through pairwise
-comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\),
-indicating which student possesses a higher degree of understanding
-among those compared. The research question is then framed as the
-estimand, ``\emph{On average}, is there a difference in conceptual
-understanding of algebra between the two groups of student?'' This
-estimand is mathematically expressed as \(E[Y| X=1] - E[Y| X=2]\), where
-\(E[\cdot]\) denotes the expected value. An example of this approach is
-seen in \citet{Jones_et_al_2019}.
+influence students' conceptual understanding of a topic (\(Y\))?'' To
+investigate this, the study could randomly assign students to two
+groups, each using a different teaching method \((X=\{1,2\})\).
+Students' conceptual understanding of the topic could be evaluated
+through pairwise comparisons, resulting in a dichotomous outcome
+\((Y=\{0,1\})\), indicating which student among those compared has a
+higher level of understanding. The research question could be then
+framed as the estimand, ``\emph{On average}, is there a difference in
+conceptual understanding of the topic between the two groups of
+students?'' This estimand could be mathematically expressed as
+\(E[Y| X=1] - E[Y| X=2]\), where \(E[\cdot]\) denotes the expected
+value. An example of this approach is seen in \citet{Jones_et_al_2019}.
 
@@
 estimator known for its effectiveness in modeling pairwise comparisons,
-providing accurate estimates for the desired estimand when its
-assumptions are met.
+yielding accurate estimates for the target of inference discussed
+earlier, as long as its assumptions are met.
 
@@
 
-Although experiments are widely regarded as the gold standard in
-evidence-based science \citep{Hariton_et_al_2018, Hansson_2014},
-researchers often encounter constraints that limit their ability to
-conduct experimental studies. These constraints include ethical
-concerns, such as the assignment of individuals to potentially harmful
-interventions, and practical limitations, such as the infeasibility of,
-for example, assigning individuals to genetic modifications or physical
-impairments \citep{Neal_2020}. In these instances, causal inference
-offers a valuable alternative for generating causal estimates.
-\emph{Causal inference} is a framework designed to identify the causes
-of phenomena and estimate their effects using data
+Experiments are widely regarded as the gold standard in evidence-based
+science \citep{Hariton_et_al_2018, Hansson_2014}, however, researchers
+often encounter constraints that limit their ability to conduct
+experimental studies. These constraints include ethical concerns, such
+as the assignment of individuals to potentially harmful interventions,
+and practical limitations, such as the infeasibility of, for example,
+assigning individuals to genetic modifications or physical impairments
+\citep{Neal_2020}. In these instances, causal inference offers a
+valuable alternative for generating causal estimates. \emph{Causal
+inference} is a framework designed to identify the causes of phenomena
+and estimate their effects using data
 \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
@@
 framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
-Moreover, the framework is beneficial for both observational and
-experimental studies, as it provides significant theoretical insights
-that enhance their design \citep{McElreath_2020}.
+Consequently, the framework offers significant theoretical insights that
+enhance the design of observational and experimental studies
+\citep{McElreath_2020}.
 
-Counterfactuals represent the highest level of cognitive abstraction in
-the ladder of causation, followed by intervention and association
-\citep{Pearl_et_al_2018}, and a detailed explanation of causes,
-counterfactuals, and their applications is beyond the scope of this
-document. However, it is important to note that causal inference
-addresses the question, ``What would have happened in the world under
-different circumstances?'' by transforming unobserved \emph{potential
-outcomes}, known as counterfactuals, into associational estimates.
-\emph{Counterfactuals} represent scenarios \emph{contrary to fact},
-where alternative \emph{potential outcomes} resulting from a cause are
-neither observed nor observable \citep{Neal_2020, Counterfactual_2024}.
-The framework achieves this transformation while ensuring the absence of
-confounding. Ultimately, the \emph{fundamental problem of causality} is
-reduced to calculating differences between averages of observed outcomes
-and the transformed counterfactuals, thereby identifying causal effects.
+Although a detailed explanation of causes, and counterfactuals is beyond
+the scope of this document, it suffices to say that causal inference
+addresses the \emph{fundamental problem of causality} through the use of
+counterfactuals \citep{Neal_2020}. The framework defines the
+\emph{individual causal effect} (ICE) as the difference between an
+individual's observed and unobserved potential outcomes. These
+unobserved potential outcomes, termed \emph{counterfactuals}, represent
+hypothetical scenarios that are \emph{contrary to fact}, where
+alternative outcomes for an individual, resulting from a specific cause,
+are neither observed nor observable
+\citep{Neal_2020, Counterfactual_2024}. The framework then extends this
+estimand to groups by introducing the \emph{average causal effect}
+(ACE). The ACE reflects the difference between averages of observed
+potential outcomes and counterfactuals across groups. Finally, similar
+to an experiment, the framework identifies the ACE from associational
+estimates by ensuring the absence of confounding.
 
@@
 \citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
-mathematical notation to characterize causal inference, but they do so
-in different ways \citep{Neal_2020}. The potential outcomes approach
-relies on counterfactual notation, whereas the structural approach
-employs do-calculus \citep{Pearl_2009}. Despite these differences, both
+mathematical notation to characterize the ACE, but they do so in
+different ways \citep{Neal_2020}. The potential outcomes approach relies
+on counterfactual notation, whereas the structural approach employs
+do-calculus \citep{Pearl_2009}. Despite these differences, both
 notations can be expressed in terms of the other, and both approaches
@@
 
-However, the structural approach offers an additional key advantage over
-the potential outcomes approach: it enables the graphical and formal
-representation of any system through directed acyclic graphs
-\citep[DAG,][]{Gross_et_al_2018, Neal_2020} and and structural causal
-models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. DAGs serve as
-heuristics, effectively conveying the assumed causal structure of a
-system \citep{Pearl_et_al_2016}. A heuristic is a strategy that
-simplifies information to make decisions more quickly, efficiently, and
-sometimes more accurately than complex methods \citep{Chow_2015}.
-Consequently, DAGs do not represent detailed statistical models but
-allow researchers to deduce which statistical models can provide valid
-causal inferences, assuming the causal structure depicted in the DAGs
-are accurate \citep{McElreath_2020}. The Identification-Estimation
-flowchart in Figure~\ref{fig-IEflow} visually represents the process of
-transitioning from estimands to estimates, and the application of the
-sicentific model and data to identify and estimates causal effects.
+The structural approach, however, offers a notable advantage over the
+potential outcomes approach by allowing the graphical and formal
+representation of theories through directed acyclic graphs
+\citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
+DAGs function as heuristics, effectively conveying the presumed causal
+structure of a system, referred to as a \emph{scientific model}
+\citep{McElreath_2020}. They do not represent detailed statistical
+models but allow researchers to deduce which statistical models can
+provide valid causal inferences, assuming the causal structure depicted
+in the DAGs are accurate \citep{McElreath_2020}. The
+Identification-Estimation flowchart in Figure~\ref{fig-IEflow} visually
+represents this process of transitioning from estimands to estimates, as
+well as the application of the scientific model and data to identify and
+estimate causal effects.
 

### 2024-08-16 _tex/index.tex
Built site for gh-pages

@@
 DAGs function as heuristics, effectively conveying the presumed causal
-structure of a system, referred to as a \emph{scientific model}
-\citep{McElreath_2020}. They do not represent detailed statistical
-models but allow researchers to deduce which statistical models can
-provide valid causal inferences, assuming the causal structure depicted
-in the DAGs are accurate \citep{McElreath_2020}. The
-Identification-Estimation flowchart in Figure~\ref{fig-IEflow} visually
-represents this process of transitioning from estimands to estimates, as
-well as the application of the scientific model and data to identify and
-estimate causal effects.
+structure of a system, referred to as a \emph{scientific model}. They do
+not represent detailed statistical models but allow researchers to
+deduce which statistical models can provide valid causal inferences,
+assuming the causal structure depicted in the DAGs are accurate
+\citep{McElreath_2020}. The Identification-Estimation flowchart in
+Figure~\ref{fig-IEflow} visually represents this process of
+transitioning from estimands to estimates, as well as the application of
+the scientific model and data to identify and estimate causal effects.
 

### 2024-08-16 index.qmd
section:background,structural; process:write
section:background,structural,dag; process:write

@@
 
-This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs.
-The section is not a comprehensive description of causal inference methods.
+This section introduces fundamental concepts in causal inference but does not offer a comprehensive description of causal inference methods.
 Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020.
@@
 These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010].
-For instance, a study might examine the question, “To what extent do different teaching methods ($X$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(X=\{1,2\})$.
-Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding.
-The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed as $E[Y| X=1] - E[Y| X=2]$, where $E[\cdot]$ denotes the expected value.
+For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T_{i}=\{1,2\})$.
+Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y_{i}=\{0,1\})$, indicating which student among those compared has a higher level of understanding.
+The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the associational quantity $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, where $E[\cdot]$ denotes the expected value.
 An example of this approach is seen in @Jones_et_al_2019.
@@
 Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
-For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the target of inference discussed earlier, as long as its assumptions are met.
+For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the target of inference discussed earlier, as long as the assumptions of the model are met.
 
@@
 However, many studies aim to understand the mechanisms underlying specific data types and also seek to establish causal relationships rather than merely infer associations.
-In the earlier example, the associational estimates obtained using the BTL model can be interpreted as causal, because the data were collected using a randomized experiment.
+In the earlier example, the differences between groups obtained using the BTL model, referred to as the associational estimates, can be interpreted as causal because the data were collected through a randomized experiment.
 Randomized experiments enable the causal interpretation of associational estimates by ensuring several key properties, including common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
 The most crucial property, however, is that randomization effectively eliminates confounding.
-*Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious association [@Everitt_et_al_2010].
+*Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious associations [@Everitt_et_al_2010].
 Randomization mitigates this issue by decoupling the intervention assignment mechanism, such as assigning students to different groups, from other variables and outcomes [@Morgan_et_al_2014; @Neal_2020].
 
-<!-- 4. causal inference to the rescue -->
+<!-- 4. Experiments are not available to everyone, then causal inference comes to the rescue -->
 Experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], however, researchers often encounter constraints that limit their ability to conduct experimental studies.
@@
 
-<!-- 5. But how does it work? -->
-Although a detailed explanation of causes, and counterfactuals is beyond the scope of this document, it suffices to say that causal inference addresses the *fundamental problem of causality* through the use of counterfactuals [@Neal_2020].
-The framework defines the *individual causal effect* (ICE) as the difference between an individual's observed and unobserved potential outcomes.
-These unobserved potential outcomes, termed *counterfactuals*, represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes for an individual, resulting from a specific cause, are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-The framework then extends this estimand to groups by introducing the *average causal effect* (ACE).
-The ACE reflects the difference between averages of observed potential outcomes and counterfactuals across groups.
-Finally, similar to an experiment, the framework identifies the ACE from associational estimates by ensuring the absence of confounding.
+<!-- 6. But how does it work? -->
+While a detailed explanation of causes and counterfactuals is beyond the scope of this document, it is important to note that causal inference addresses the *fundamental problem of causality* through the use of counterfactuals [@Neal_2020].
+The framework defines *individual causal effects* (ICE) as the difference between an individual's observed and unobserved potential outcomes.
+These unobserved potential outcomes, known as *counterfactuals*, represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes for an individual, resulting from a specific cause, are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
+The framework then extends this estimand to *average causal effects* (ACE), which captures the difference between sample averages of observed potential outcomes and counterfactuals.
+Finally, much like in an experiment, the framework identifies the ACE from associational estimates by conditioning for a *sufficient adjustment set* of variables, ensuring the absence of confounding [@Morgan_et_al_2014].
 
-<!-- 6. How do we apply causal inference and counterfactuals? -->
+<!-- 7. How do we apply causal inference and counterfactuals? -->
 Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
@@
 
-<!-- 7. But which approach are we going to use? -->
+<!-- 8. But which approach are we going to use? -->
 The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of theories through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model* [@McElreath_2020].
+DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model*.
 They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
@@
 
+<!-- 5. continuing with the motivating example -->
+The motivating example in @sec-framework-structural can further illustrate how to use the five fundamental building blocks to construct a system's causal structure.
+In this scenario, the investigator aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y_{i}=\{0,1\})$ between two groups of students $(T_{i}=\{1,2\})$, described by the estimand $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$.
+However, unlike the previous case, an experiment cannot be conducted, and the problem suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$.
+Such scenarios are plausible, especially when the teaching methods depend on software or access to technology, which may be limited in certain countries [(maybe an example with more impact?)]{style="color:blue;"}.
+@fig-example illustrates the plausible causal structure of the motivating example.
+A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks.
+The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
+Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
+
+::: {#fig-example layout="[ [1,1,1],[1,1,1],[1,1,1] ]"}
+
+![Magnified representation](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=100%}
+
+![Standard representation](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%}
+
+::: {#fig-scm_example1 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := f_{Z}(X,e_{T}) \\
+  Y & := f_{Y}(T,X,e_{Y}) \\
+  e_{T} & \dsep e_{X} \\
+  e_{T} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
+
+Structural causal model
+:::
+
+DAGs for a plausible causal structure in a system.
+:::
+
 ## The flow of association and causation in graphs {#sec-framework-flow}
@@
 <!-- Ardmed with dags and scm, and theit probabilité implication, we can now expand on thé motivating example provided in section xx -->
-
-## A motivating example {#sec-framework-example}
-
-<!-- Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system.
-Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation.
-The problem suggests that a variable $X$ potentially influences both $T$ and $Y$.
-Beyond these relationships, the problem does not specify any further variables of interest.
-Such scenarios are not hard to imagine.
-For instance, $T$ might represent different treatments that could affect the recovery from cancer $Y$, while $X$ could denote the patient's age.
-Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status.
-[(not a bad example, but I prefer one using writing skills)]{style="color:blue;"} -->
-
-<!-- 6. DAG for the motivating example -->
-<!-- @fig-example illustrates the plausible causal structure of the motivating example.
-A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks.
-The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
-Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
-Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
--->
-
-<!-- ::: {#fig-example layout="[ [1,1,1],[1,1,1],[1,1,1] ]"} -->
-
-<!-- ![Magnified representation](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=100%} -->
-
-<!-- ![Standard representation](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%} -->
-
-<!-- ::: {#fig-scm_example1 fig-align="center"} -->
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!--   X & := f_{X}(e_{X}) \\ -->
-<!--   T & := f_{Z}(X,e_{T}) \\ -->
-<!--   Y & := f_{Y}(T,X,e_{Y}) \\ -->
-<!--   e_{T} & \dsep e_{X} \\ -->
-<!--   e_{T} & \dsep e_{Y} \\ -->
-<!--   e_{X} & \dsep e_{Y} -->
-<!-- \end{aligned} -->
-<!-- $$ -->
-
-<!-- Structural causal model -->
-<!-- ::: -->
-
-<!-- DAGs for a plausible causal structure in a system. -->
-<!-- ::: -->
 

### 2024-08-16 _tex/index.tex
Built site for gh-pages

@@
 
-This section introduces fundamental concepts in causal inference, such
-as directed acyclic graphs (DAGs), structural causal models (SCMs), and
-the flow of association and causation in graphs. The section is not a
-comprehensive description of causal inference methods. Readers
-interested in deeper exploration should consult introductory papers like
-\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
-\citet{Cinelli_et_al_2020}. They may also find introductory books such
-as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
+This section introduces fundamental concepts in causal inference but
+does not offer a comprehensive description of causal inference methods.
+Readers interested in deeper exploration should consult introductory
+papers like \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019},
+and \citet{Cinelli_et_al_2020}. They may also find introductory books
+such as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
 \citet{McElreath_2020} useful. For more advanced study, seminal
@@
 \citep{Everitt_et_al_2010}. For instance, a study might examine the
-question, ``To what extent do different teaching methods (\(X\))
+question, ``To what extent do different teaching methods (\(T\))
 influence students' conceptual understanding of a topic (\(Y\))?'' To
 investigate this, the study could randomly assign students to two
-groups, each using a different teaching method \((X=\{1,2\})\).
+groups, each using a different teaching method \((T_{i}=\{1,2\})\).
 Students' conceptual understanding of the topic could be evaluated
 through pairwise comparisons, resulting in a dichotomous outcome
-\((Y=\{0,1\})\), indicating which student among those compared has a
+\((Y_{i}=\{0,1\})\), indicating which student among those compared has a
 higher level of understanding. The research question could be then
@@
 conceptual understanding of the topic between the two groups of
-students?'' This estimand could be mathematically expressed as
-\(E[Y| X=1] - E[Y| X=2]\), where \(E[\cdot]\) denotes the expected
-value. An example of this approach is seen in \citet{Jones_et_al_2019}.
+students?'' This estimand could be mathematically expressed by the
+associational quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where
+\(E[\cdot]\) denotes the expected value. An example of this approach is
+seen in \citet{Jones_et_al_2019}.
 
@@
 yielding accurate estimates for the target of inference discussed
-earlier, as long as its assumptions are met.
+earlier, as long as the assumptions of the model are met.
 
@@
 rather than merely infer associations. In the earlier example, the
-associational estimates obtained using the BTL model can be interpreted
-as causal, because the data were collected using a randomized
-experiment. Randomized experiments enable the causal interpretation of
-associational estimates by ensuring several key properties, including
-common support, no interference, and consistency
-\citep{Morgan_et_al_2014, Neal_2020}. The most crucial property,
-however, is that randomization effectively eliminates confounding.
-\emph{Confounding} occurs when an external variable influences both the
-outcome and the variable of interest, leading to spurious association
-\citep{Everitt_et_al_2010}. Randomization mitigates this issue by
-decoupling the intervention assignment mechanism, such as assigning
-students to different groups, from other variables and outcomes
-\citep{Morgan_et_al_2014, Neal_2020}.
+differences between groups obtained using the BTL model, referred to as
+the associational estimates, can be interpreted as causal because the
+data were collected through a randomized experiment. Randomized
+experiments enable the causal interpretation of associational estimates
+by ensuring several key properties, including common support, no
+interference, and consistency \citep{Morgan_et_al_2014, Neal_2020}. The
+most crucial property, however, is that randomization effectively
+eliminates confounding. \emph{Confounding} occurs when an external
+variable influences both the outcome and the variable of interest,
+leading to spurious associations \citep{Everitt_et_al_2010}.
+Randomization mitigates this issue by decoupling the intervention
+assignment mechanism, such as assigning students to different groups,
+from other variables and outcomes \citep{Morgan_et_al_2014, Neal_2020}.
 
@@
 
-Although a detailed explanation of causes, and counterfactuals is beyond
-the scope of this document, it suffices to say that causal inference
+While a detailed explanation of causes and counterfactuals is beyond the
+scope of this document, it is important to note that causal inference
 addresses the \emph{fundamental problem of causality} through the use of
-counterfactuals \citep{Neal_2020}. The framework defines the
-\emph{individual causal effect} (ICE) as the difference between an
+counterfactuals \citep{Neal_2020}. The framework defines
+\emph{individual causal effects} (ICE) as the difference between an
 individual's observed and unobserved potential outcomes. These
-unobserved potential outcomes, termed \emph{counterfactuals}, represent
-hypothetical scenarios that are \emph{contrary to fact}, where
+unobserved potential outcomes, known as \emph{counterfactuals},
+represent hypothetical scenarios that are \emph{contrary to fact}, where
 alternative outcomes for an individual, resulting from a specific cause,
@@
 \citep{Neal_2020, Counterfactual_2024}. The framework then extends this
-estimand to groups by introducing the \emph{average causal effect}
-(ACE). The ACE reflects the difference between averages of observed
-potential outcomes and counterfactuals across groups. Finally, similar
-to an experiment, the framework identifies the ACE from associational
-estimates by ensuring the absence of confounding.
+estimand to \emph{average causal effects} (ACE), which captures the
+difference between sample averages of observed potential outcomes and
+counterfactuals. Finally, much like in an experiment, the framework
+identifies the ACE from associational estimates by conditioning for a
+\emph{sufficient adjustment set} of variables, ensuring the absence of
+confounding \citep{Morgan_et_al_2014}.
 
@@
 
+The motivating example in Section~\ref{sec-framework-structural} can
+further illustrate how to use the five fundamental building blocks to
+construct a system's causal structure. In this scenario, the
+investigator aims to determine whether, \emph{on average}, there is a
+difference in conceptual understanding of a topic \((Y_{i}=\{0,1\})\)
+between two groups of students \((T_{i}=\{1,2\})\), described by the
+estimand \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\). However, unlike the
+previous case, an experiment cannot be conducted, and the problem
+suggests that the country to which a student belongs (\(X\)) may
+influence both \(T\) and \(Y\). Such scenarios are plausible, especially
+when the teaching methods depend on software or access to technology,
+which may be limited in certain countries {(maybe an example with more
+impact?)}. Figure~\ref{fig-example} illustrates the plausible causal
+structure of the motivating example. A detailed examination of Figures
+\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
+\ref{fig-scm_example1} reveals the presence of at least four of the five
+fundamental building blocks. The figures display multiple descendants,
+as indicated by pairwise relations such as \(X \rightarrow T\),
+\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
+features multiple pairs of unconnected nodes, evident from the relations
+\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
+\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
+fork \(X \rightarrow \{T,Y\}\), and two colliders with
+\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_example1.png}
+
+}
+
+\subcaption{\label{fig-mdag_example1}Magnified representation}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_example1.png}
+
+}
+
+\subcaption{\label{fig-sdag_example1}Standard representation}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := f_{Z}(X,e_{T}) \\
+  Y & := f_{Y}(T,X,e_{Y}) \\
+  e_{T} & \perp\!\!\!\perp e_{X} \\
+  e_{T} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_example1}Structural causal model}
+
+\end{minipage}%
+
+\caption{\label{fig-example}DAGs for a plausible causal structure in a
+system.}
+
+\end{figure}%
+
 \subsection{The flow of association and causation in
@@
 \end{figure}%
-
-\subsection{A motivating example}\label{sec-framework-example}
 

### 2024-08-19 index.qmd
section:background,structural; process:write

@@
 These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010].
-For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T_{i}=\{1,2\})$.
-Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y_{i}=\{0,1\})$, indicating which student among those compared has a higher level of understanding.
-The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the associational quantity $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, where $E[\cdot]$ denotes the expected value.
+For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T=\{1,2\})$.
+Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding.
+The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the associational random quantity $E[Y| T=1] - E[Y| T=2]$, where $E[\cdot]$ denotes the expected value.
 An example of this approach is seen in @Jones_et_al_2019.
@@
 An *estimator* is a method or function that maps data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refer to the process of integrating data with an estimator [@Everitt_et_al_2010].
+*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
 Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
-For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the target of inference discussed earlier, as long as the assumptions of the model are met.
+For instance, the Z-test is an estimator known for its effectiveness in comparing two proportions, yielding accurate estimates when its underlying assumptions are met [@Kanji_2006].
+The Z-test is expressed as a signal-to-noise ratio: $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
+The signal is the difference between the group sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y| T=1]$ and $E[Y| T=2]$, respectively.
+The noise $\hat{s}_{p}$ is the unpooled sample variability observed between the two groups.
+
+<!-- the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the comparison of two groups, as long as the assumptions of the model are met.
+Usually these estimates are expressed in terms of the proportions of groups  -->
 
 <!-- 3. But we want these estimands to be causal -->
-However, many studies aim to understand the mechanisms underlying specific data types and also seek to establish causal relationships rather than merely infer associations.
-In the earlier example, the differences between groups obtained using the BTL model, referred to as the associational estimates, can be interpreted as causal because the data were collected through a randomized experiment.
+However, many studies aim to understand the mechanisms underlying specific data and also seek to establish causal relationships rather than merely infer associations.
+In the earlier example, the differences between groups obtained using the Z-test, referred to as the associational estimate, can be interpreted as causal because the data were collected through a randomized experiment.
 Randomized experiments enable the causal interpretation of associational estimates by ensuring several key properties, including common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
@@
 <!-- 4. Experiments are not available to everyone, then causal inference comes to the rescue -->
-Experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], however, researchers often encounter constraints that limit their ability to conduct experimental studies.
+Experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014].
+However, researchers often face constraints that limit their ability to conduct experimental studies.
 These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020].
-In these instances, causal inference offers a valuable alternative for generating causal estimates.
-*Causal inference* is a framework designed to identify the causes of phenomena and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, causal inference provides a rigorous mathematical framework for analyzing causes and counterfactuals [@Pearl_2009].
-Consequently, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].
+In these situations, causal inference provides a valuable alternative for generating causal estimates, particularly when the goal is to understand the mechanisms underlying specific data.
+Moreover, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].
+
+<!-- 5. What is causal inference? -->
+Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, *causal inference* is a framework designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
+This framework employs rigorous mathematical techniques to address the fundamental problem of causality [@Pearl_2009], which is essential for understanding and defining causal effects.
+The *fundamental problem of causality* centers on answering the question, "What would have happened 'in the world' under different circumstances?" a concept known as counterfactuals.
+*Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a specific cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
 
 <!-- 6. But how does it work? --> 
-While a detailed explanation of causes and counterfactuals is beyond the scope of this document, it is important to note that causal inference addresses the *fundamental problem of causality* through the use of counterfactuals [@Neal_2020].
-The framework defines *individual causal effects* (ICE) as the difference between an individual's observed and unobserved potential outcomes.
-These unobserved potential outcomes, known as *counterfactuals*, represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes for an individual, resulting from a specific cause, are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-The framework then extends this estimand to *average causal effects* (ACE), which captures the difference between sample averages of observed potential outcomes and counterfactuals.
-Finally, much like in an experiment, the framework identifies the ACE from associational estimates by conditioning for a *sufficient adjustment set* of variables, ensuring the absence of confounding [@Morgan_et_al_2014].
+Although a comprehensive discussion of causes and counterfactuals exceeds the scope of this document, a brief overview of how the framework addresses the fundamental problem of causality is possible using the previous example.
+The framework starts by defining *individual causal effects* (ICE) as the difference between students' observed and unobserved potential outcomes: $\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)$.
+Notice once a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, and is thus termed a *counterfactual*.
+To address the challenge posed by counterfactuals, the framework extends the ICE to *average causal effects* (ACE).
+The ACE is defined as $\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, and represents the difference between the average of observed potential outcomes and counterfactuals in the sample.
+Finally, akin to experimental studies, the ACE is identified from associational estimates by ensuring the absence of confounding.
+This is accomplished by statistically conditioning on a *sufficient adjustment set* of variables ($X$).
+Consequently, the ACE is expressed as $\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].
 
@@
 <!-- 8. But which approach are we going to use? -->
-The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of theories through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
+The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of the ACE through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
 DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model*.
 They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
-The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents this process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.
+The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.
 

### 2024-08-19 index.qmd
section:background,dag; process:write

@@
 The motivating example in @sec-framework-structural can further illustrate how to use the five fundamental building blocks to construct a system's causal structure.
-In this scenario, the investigator aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y_{i}=\{0,1\})$ between two groups of students $(T_{i}=\{1,2\})$, described by the estimand $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$.
+In this scenario, the investigator aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y=\{0,1\})$ between two groups of students $(T=\{1,2\})$, described by the estimand $E[Y| T=1] - E[Y| T=2]$.
 However, unlike the previous case, an experiment cannot be conducted, and the problem suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$.

### 2024-08-19 _tex/index.tex
Built site for gh-pages

@@
 investigate this, the study could randomly assign students to two
-groups, each using a different teaching method \((T_{i}=\{1,2\})\).
+groups, each using a different teaching method \((T=\{1,2\})\).
 Students' conceptual understanding of the topic could be evaluated
 through pairwise comparisons, resulting in a dichotomous outcome
-\((Y_{i}=\{0,1\})\), indicating which student among those compared has a
+\((Y=\{0,1\})\), indicating which student among those compared has a
 higher level of understanding. The research question could be then
@@
 students?'' This estimand could be mathematically expressed by the
-associational quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where
+associational random quantity \(E[Y| T=1] - E[Y| T=2]\), where
 \(E[\cdot]\) denotes the expected value. An example of this approach is
@@
 numerical values that approximate the estimand and are derived through
-\emph{estimation}, which refer to the process of integrating data with
+\emph{estimation}, which refers to the process of integrating data with
 an estimator \citep{Everitt_et_al_2010}. Although various methods can
@@
 desirable properties that ensure the accuracy of estimates. For
-instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
-estimator known for its effectiveness in modeling pairwise comparisons,
-yielding accurate estimates for the target of inference discussed
-earlier, as long as the assumptions of the model are met.
+instance, the Z-test is an estimator known for its effectiveness in
+comparing two proportions, yielding accurate estimates when its
+underlying assumptions are met \citep{Kanji_2006}. The Z-test is
+expressed as a signal-to-noise ratio:
+\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is the
+difference between the group sample proportions,
+\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
+\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
+\(E[Y| T=1]\) and \(E[Y| T=2]\), respectively. The noise \(\hat{s}_{p}\)
+is the unpooled sample variability observed between the two groups.
 
 However, many studies aim to understand the mechanisms underlying
-specific data types and also seek to establish causal relationships
-rather than merely infer associations. In the earlier example, the
-differences between groups obtained using the BTL model, referred to as
-the associational estimates, can be interpreted as causal because the
-data were collected through a randomized experiment. Randomized
-experiments enable the causal interpretation of associational estimates
-by ensuring several key properties, including common support, no
-interference, and consistency \citep{Morgan_et_al_2014, Neal_2020}. The
-most crucial property, however, is that randomization effectively
-eliminates confounding. \emph{Confounding} occurs when an external
-variable influences both the outcome and the variable of interest,
-leading to spurious associations \citep{Everitt_et_al_2010}.
-Randomization mitigates this issue by decoupling the intervention
-assignment mechanism, such as assigning students to different groups,
-from other variables and outcomes \citep{Morgan_et_al_2014, Neal_2020}.
-
-Experiments are widely regarded as the gold standard in evidence-based
-science \citep{Hariton_et_al_2018, Hansson_2014}, however, researchers
-often encounter constraints that limit their ability to conduct
-experimental studies. These constraints include ethical concerns, such
-as the assignment of individuals to potentially harmful interventions,
-and practical limitations, such as the infeasibility of, for example,
+specific data and also seek to establish causal relationships rather
+than merely infer associations. In the earlier example, the differences
+between groups obtained using the Z-test, referred to as the
+associational estimate, can be interpreted as causal because the data
+were collected through a randomized experiment. Randomized experiments
+enable the causal interpretation of associational estimates by ensuring
+several key properties, including common support, no interference, and
+consistency \citep{Morgan_et_al_2014, Neal_2020}. The most crucial
+property, however, is that randomization effectively eliminates
+confounding. \emph{Confounding} occurs when an external variable
+influences both the outcome and the variable of interest, leading to
+spurious associations \citep{Everitt_et_al_2010}. Randomization
+mitigates this issue by decoupling the intervention assignment
+mechanism, such as assigning students to different groups, from other
+variables and outcomes \citep{Morgan_et_al_2014, Neal_2020}.
+
+Experiments are widely recognized as the gold standard in evidence-based
+science \citep{Hariton_et_al_2018, Hansson_2014}. However, researchers
+often face constraints that limit their ability to conduct experimental
+studies. These constraints include ethical concerns, such as the
+assignment of individuals to potentially harmful interventions, and
+practical limitations, such as the infeasibility of, for example,
 assigning individuals to genetic modifications or physical impairments
-\citep{Neal_2020}. In these instances, causal inference offers a
-valuable alternative for generating causal estimates. \emph{Causal
-inference} is a framework designed to identify the causes of phenomena
-and estimate their effects using data
-\citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
-modeling, which focuses primarily on summarizing data and inferring
-associations, causal inference provides a rigorous mathematical
-framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
-Consequently, the framework offers significant theoretical insights that
+\citep{Neal_2020}. In these situations, causal inference provides a
+valuable alternative for generating causal estimates, particularly when
+the goal is to understand the mechanisms underlying specific data.
+Moreover, the framework offers significant theoretical insights that
 enhance the design of observational and experimental studies
@@
 
-While a detailed explanation of causes and counterfactuals is beyond the
-scope of this document, it is important to note that causal inference
-addresses the \emph{fundamental problem of causality} through the use of
-counterfactuals \citep{Neal_2020}. The framework defines
-\emph{individual causal effects} (ICE) as the difference between an
-individual's observed and unobserved potential outcomes. These
-unobserved potential outcomes, known as \emph{counterfactuals},
-represent hypothetical scenarios that are \emph{contrary to fact}, where
-alternative outcomes for an individual, resulting from a specific cause,
-are neither observed nor observable
-\citep{Neal_2020, Counterfactual_2024}. The framework then extends this
-estimand to \emph{average causal effects} (ACE), which captures the
-difference between sample averages of observed potential outcomes and
-counterfactuals. Finally, much like in an experiment, the framework
-identifies the ACE from associational estimates by conditioning for a
-\emph{sufficient adjustment set} of variables, ensuring the absence of
-confounding \citep{Morgan_et_al_2014}.
+Unlike classical statistical modeling, which focuses primarily on
+summarizing data and inferring associations, \emph{causal inference} is
+a framework designed to identify causes and estimate their effects using
+data \citep{Shaughnessy_et_al_2010, Neal_2020}. This framework employs
+rigorous mathematical techniques to address the fundamental problem of
+causality \citep{Pearl_2009}, which is essential for understanding and
+defining causal effects. The \emph{fundamental problem of causality}
+centers on answering the question, ``What would have happened `in the
+world' under different circumstances?'' a concept known as
+counterfactuals. \emph{Counterfactuals} represent hypothetical scenarios
+that are \emph{contrary to fact}, where alternative outcomes resulting
+from a specific cause are neither observed nor observable
+\citep{Neal_2020, Counterfactual_2024}.
+
+Although a comprehensive discussion of causes and counterfactuals
+exceeds the scope of this document, a brief overview of how the
+framework addresses the fundamental problem of causality is possible
+using the previous example. The framework starts by defining
+\emph{individual causal effects} (ICE) as the difference between
+students' observed and unobserved potential outcomes:
+\(\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)\). Notice once a
+student is assigned to \(T_{i}=1\), the potential outcome under
+\(T_{i}=2\) is no longer observed nor observable, and is thus termed a
+\emph{counterfactual}. To address the challenge posed by
+counterfactuals, the framework extends the ICE to \emph{average causal
+effects} (ACE). The ACE is defined as
+\(\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), and
+represents the difference between the average of observed potential
+outcomes and counterfactuals in the sample. Finally, akin to
+experimental studies, the ACE is identified from associational estimates
+by ensuring the absence of confounding. This is accomplished by
+statistically conditioning on a \emph{sufficient adjustment set} of
+variables (\(X\)). Consequently, the ACE is expressed as
+\(\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]\), where
+\(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
+\citep{Morgan_et_al_2014}.
 
@@
 potential outcomes approach by allowing the graphical and formal
-representation of theories through directed acyclic graphs
+representation of the ACE through directed acyclic graphs
 \citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
@@
 \citep{McElreath_2020}. The Identification-Estimation flowchart in
-Figure~\ref{fig-IEflow} visually represents this process of
-transitioning from estimands to estimates, as well as the application of
-the scientific model and data to identify and estimate causal effects.
+Figure~\ref{fig-IEflow} visually represents the process of transitioning
+from estimands to estimates, as well as the application of the
+scientific model and data to identify and estimate causal effects.
 
@@
 investigator aims to determine whether, \emph{on average}, there is a
-difference in conceptual understanding of a topic \((Y_{i}=\{0,1\})\)
-between two groups of students \((T_{i}=\{1,2\})\), described by the
-estimand \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\). However, unlike the
-previous case, an experiment cannot be conducted, and the problem
-suggests that the country to which a student belongs (\(X\)) may
-influence both \(T\) and \(Y\). Such scenarios are plausible, especially
-when the teaching methods depend on software or access to technology,
-which may be limited in certain countries {(maybe an example with more
-impact?)}. Figure~\ref{fig-example} illustrates the plausible causal
-structure of the motivating example. A detailed examination of Figures
+difference in conceptual understanding of a topic \((Y=\{0,1\})\)
+between two groups of students \((T=\{1,2\})\), described by the
+estimand \(E[Y| T=1] - E[Y| T=2]\). However, unlike the previous case,
+an experiment cannot be conducted, and the problem suggests that the
+country to which a student belongs (\(X\)) may influence both \(T\) and
+\(Y\). Such scenarios are plausible, especially when the teaching
+methods depend on software or access to technology, which may be limited
+in certain countries {(maybe an example with more impact?)}.
+Figure~\ref{fig-example} illustrates the plausible causal structure of
+the motivating example. A detailed examination of Figures
 \ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM

### 2024-08-20 index.qmd
section:background,structural,dag; process:write

@@
 Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding.
-The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the associational random quantity $E[Y| T=1] - E[Y| T=2]$, where $E[\cdot]$ denotes the expected value.
+The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the random quantity $E[Y| T=1] - E[Y| T=2]$, where $E[\cdot]$ denotes the expected value.
 An example of this approach is seen in @Jones_et_al_2019.
@@
 *Identification* refers to the process of accurately computing an estimand using an estimator.
-An *estimator* is a method or function that maps data into an estimate [@Neal_2020].
+An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
 *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
 Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
-For instance, the Z-test is an estimator known for its effectiveness in comparing two proportions, yielding accurate estimates when its underlying assumptions are met [@Kanji_2006].
+For instance, the Z-test is an estimator known for its effectiveness in comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006].
 The Z-test is expressed as a signal-to-noise ratio: $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
@@
 In these situations, causal inference provides a valuable alternative for generating causal estimates, particularly when the goal is to understand the mechanisms underlying specific data.
-Moreover, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].
+Moreover, the framework provides significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].
 
@@
 Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, *causal inference* is a framework designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-This framework employs rigorous mathematical techniques to address the fundamental problem of causality [@Pearl_2009], which is essential for understanding and defining causal effects.
-The *fundamental problem of causality* centers on answering the question, "What would have happened 'in the world' under different circumstances?" a concept known as counterfactuals.
+The framework employs rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009].
+This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" a concept known as counterfactuals, essential for understanding and defining causal effects.
 *Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a specific cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
@@
 <!-- 6. But how does it work? --> 
-Although a comprehensive discussion of causes and counterfactuals exceeds the scope of this document, a brief overview of how the framework addresses the fundamental problem of causality is possible using the previous example.
-The framework starts by defining *individual causal effects* (ICE) as the difference between students' observed and unobserved potential outcomes: $\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)$.
-Notice once a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, and is thus termed a *counterfactual*.
-To address the challenge posed by counterfactuals, the framework extends the ICE to *average causal effects* (ACE).
-The ACE is defined as $\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, and represents the difference between the average of observed potential outcomes and counterfactuals in the sample.
-Finally, akin to experimental studies, the ACE is identified from associational estimates by ensuring the absence of confounding.
+Although a comprehensive discussion of causes and counterfactuals exceeds the scope of this document, a brief overview of how the framework addresses the fundamental problem of causality can be provided.
+The framework begins by defining *individual causal effects* (ICE) as the difference between students' observed and unobserved potential outcomes: $\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)$.
+Notice when a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, and thus termed a *counterfactual*.
+To overcome the challenge posed by counterfactuals, the framework extends the ICE to *average causal effects* (ACE).
+The ACE is defined as $\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, representing the difference between the average of observed potential outcomes and counterfactuals across the sample.
+Finally, similar to experimental studies, the counterfactuals are identified from associational estimates by ensuring the absence of confounding.
 This is accomplished by statistically conditioning on a *sufficient adjustment set* of variables ($X$).
-Consequently, the ACE is expressed as $\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].
+As a result, the ACE is expressed as $\tau = E_{X}\left[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]\right]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].
 
@@
 <!-- 8. But which approach are we going to use? -->
-The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of the ACE through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model*.
+The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of a system through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
+DAGs function as heuristics, effectively conveying the presumed causal structure of the system underlying the ACE, referred to as a *scientific model*.
 They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
@@
 Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-This feature enables the decomposition of complex structures into basic building blocks, facilitating the analysis of these structures by focusing on the causal assumptions associated with each building block individually [@McElreath_2020].
+This feature enables the decomposition of complex structures into basic building blocks, facilitating the analysis of these structures by focusing on the causal assumptions associated with individual building blocks [@McElreath_2020].
 These building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].
@@
 <!-- 5. continuing with the motivating example -->
-The motivating example in @sec-framework-structural can further illustrate how to use the five fundamental building blocks to construct a system's causal structure.
-In this scenario, the investigator aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y=\{0,1\})$ between two groups of students $(T=\{1,2\})$, described by the estimand $E[Y| T=1] - E[Y| T=2]$.
-However, unlike the previous case, an experiment cannot be conducted, and the problem suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$.
+The motivating example in @sec-framework-structural can be used to further illustrate how the five fundamental building blocks help construct a system's causal structure.
+In this scenario, an experiment cannot be conducted but the investigator still aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y=\{0,1\})$ between two groups of students $(T=\{1,2\})$, described by the estimand $E[Y| T=1] - E[Y| T=2]$.
+The problem further suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$.
 Such scenarios are plausible, especially when the teaching methods depend on software or access to technology, which may be limited in certain countries [(maybe an example with more impact?)]{style="color:blue;"}.
-@fig-example illustrates the plausible causal structure of the motivating example.
+@fig-example illustrates the plausible causal structure of this motivating example.
 A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks.
 The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
-Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
+Additionally, the figures feature multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
 Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.

### 2024-08-20 _tex/index.tex
Built site for gh-pages

@@
 students?'' This estimand could be mathematically expressed by the
-associational random quantity \(E[Y| T=1] - E[Y| T=2]\), where
-\(E[\cdot]\) denotes the expected value. An example of this approach is
-seen in \citet{Jones_et_al_2019}.
+random quantity \(E[Y| T=1] - E[Y| T=2]\), where \(E[\cdot]\) denotes
+the expected value. An example of this approach is seen in
+\citet{Jones_et_al_2019}.
 
@@
 estimand using an estimator. An \emph{estimator} is a method or function
-that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
-numerical values that approximate the estimand and are derived through
-\emph{estimation}, which refers to the process of integrating data with
-an estimator \citep{Everitt_et_al_2010}. Although various methods can
-approximate an estimand, researchers prioritize estimators with
-desirable properties that ensure the accuracy of estimates. For
-instance, the Z-test is an estimator known for its effectiveness in
-comparing two proportions, yielding accurate estimates when its
-underlying assumptions are met \citep{Kanji_2006}. The Z-test is
-expressed as a signal-to-noise ratio:
+that transforms data into an estimate \citep{Neal_2020}.
+\emph{Estimates} are numerical values that approximate the estimand and
+are derived through \emph{estimation}, which refers to the process of
+integrating data with an estimator \citep{Everitt_et_al_2010}. Although
+various methods can approximate an estimand, researchers prioritize
+estimators with desirable properties that ensure the accuracy of
+estimates. For instance, the Z-test is an estimator known for its
+effectiveness in comparing groups' proportions, yielding accurate
+estimates when the underlying assumptions of the statistic are met
+\citep{Kanji_2006}. The Z-test is expressed as a signal-to-noise ratio:
 \(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is the
@@
 the goal is to understand the mechanisms underlying specific data.
-Moreover, the framework offers significant theoretical insights that
+Moreover, the framework provides significant theoretical insights that
 enhance the design of observational and experimental studies
@@
 a framework designed to identify causes and estimate their effects using
-data \citep{Shaughnessy_et_al_2010, Neal_2020}. This framework employs
-rigorous mathematical techniques to address the fundamental problem of
-causality \citep{Pearl_2009}, which is essential for understanding and
-defining causal effects. The \emph{fundamental problem of causality}
-centers on answering the question, ``What would have happened `in the
-world' under different circumstances?'' a concept known as
-counterfactuals. \emph{Counterfactuals} represent hypothetical scenarios
-that are \emph{contrary to fact}, where alternative outcomes resulting
-from a specific cause are neither observed nor observable
-\citep{Neal_2020, Counterfactual_2024}.
+data \citep{Shaughnessy_et_al_2010, Neal_2020}. The framework employs
+rigorous mathematical techniques to address the \emph{fundamental
+problem of causality} \citep{Pearl_2009}. This problem revolves around
+the question, ``What would have happened `in the world' under different
+circumstances?'' a concept known as counterfactuals, essential for
+understanding and defining causal effects. \emph{Counterfactuals}
+represent hypothetical scenarios that are \emph{contrary to fact}, where
+alternative outcomes resulting from a specific cause are neither
+observed nor observable \citep{Neal_2020, Counterfactual_2024}.
 
@@
 exceeds the scope of this document, a brief overview of how the
-framework addresses the fundamental problem of causality is possible
-using the previous example. The framework starts by defining
-\emph{individual causal effects} (ICE) as the difference between
-students' observed and unobserved potential outcomes:
-\(\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)\). Notice once a
+framework addresses the fundamental problem of causality can be
+provided. The framework begins by defining \emph{individual causal
+effects} (ICE) as the difference between students' observed and
+unobserved potential outcomes:
+\(\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)\). Notice when a
 student is assigned to \(T_{i}=1\), the potential outcome under
-\(T_{i}=2\) is no longer observed nor observable, and is thus termed a
-\emph{counterfactual}. To address the challenge posed by
+\(T_{i}=2\) is no longer observed nor observable, and thus termed a
+\emph{counterfactual}. To overcome the challenge posed by
 counterfactuals, the framework extends the ICE to \emph{average causal
 effects} (ACE). The ACE is defined as
-\(\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), and
-represents the difference between the average of observed potential
-outcomes and counterfactuals in the sample. Finally, akin to
-experimental studies, the ACE is identified from associational estimates
-by ensuring the absence of confounding. This is accomplished by
-statistically conditioning on a \emph{sufficient adjustment set} of
-variables (\(X\)). Consequently, the ACE is expressed as
-\(\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]\), where
-\(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
+\(\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\),
+representing the difference between the average of observed potential
+outcomes and counterfactuals across the sample. Finally, similar to
+experimental studies, the counterfactuals are identified from
+associational estimates by ensuring the absence of confounding. This is
+accomplished by statistically conditioning on a \emph{sufficient
+adjustment set} of variables (\(X\)). As a result, the ACE is expressed
+as
+\(\tau = E_{X}\left[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]\right]\),
+where \(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
 \citep{Morgan_et_al_2014}.
@@
 potential outcomes approach by allowing the graphical and formal
-representation of the ACE through directed acyclic graphs
+representation of a system through directed acyclic graphs
 \citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
 DAGs function as heuristics, effectively conveying the presumed causal
-structure of a system, referred to as a \emph{scientific model}. They do
-not represent detailed statistical models but allow researchers to
-deduce which statistical models can provide valid causal inferences,
-assuming the causal structure depicted in the DAGs are accurate
-\citep{McElreath_2020}. The Identification-Estimation flowchart in
-Figure~\ref{fig-IEflow} visually represents the process of transitioning
-from estimands to estimates, as well as the application of the
-scientific model and data to identify and estimate causal effects.
+structure of the system underlying the ACE, referred to as a
+\emph{scientific model}. They do not represent detailed statistical
+models but allow researchers to deduce which statistical models can
+provide valid causal inferences, assuming the causal structure depicted
+in the DAGs are accurate \citep{McElreath_2020}. The
+Identification-Estimation flowchart in Figure~\ref{fig-IEflow} visually
+represents the process of transitioning from estimands to estimates, as
+well as the application of the scientific model and data to identify and
+estimate causal effects.
 
@@
 facilitating the analysis of these structures by focusing on the causal
-assumptions associated with each building block individually
+assumptions associated with individual building blocks
 \citep{McElreath_2020}. These building blocks can be represented in
@@
 
-The motivating example in Section~\ref{sec-framework-structural} can
-further illustrate how to use the five fundamental building blocks to
-construct a system's causal structure. In this scenario, the
-investigator aims to determine whether, \emph{on average}, there is a
-difference in conceptual understanding of a topic \((Y=\{0,1\})\)
-between two groups of students \((T=\{1,2\})\), described by the
-estimand \(E[Y| T=1] - E[Y| T=2]\). However, unlike the previous case,
-an experiment cannot be conducted, and the problem suggests that the
-country to which a student belongs (\(X\)) may influence both \(T\) and
-\(Y\). Such scenarios are plausible, especially when the teaching
-methods depend on software or access to technology, which may be limited
-in certain countries {(maybe an example with more impact?)}.
-Figure~\ref{fig-example} illustrates the plausible causal structure of
-the motivating example. A detailed examination of Figures
-\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
-\ref{fig-scm_example1} reveals the presence of at least four of the five
-fundamental building blocks. The figures display multiple descendants,
-as indicated by pairwise relations such as \(X \rightarrow T\),
-\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
-features multiple pairs of unconnected nodes, evident from the relations
-\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
-\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
-fork \(X \rightarrow \{T,Y\}\), and two colliders with
-\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).
+The motivating example in Section~\ref{sec-framework-structural} can be
+used to further illustrate how the five fundamental building blocks help
+construct a system's causal structure. In this scenario, an experiment
+cannot be conducted but the investigator still aims to determine
+whether, \emph{on average}, there is a difference in conceptual
+understanding of a topic \((Y=\{0,1\})\) between two groups of students
+\((T=\{1,2\})\), described by the estimand \(E[Y| T=1] - E[Y| T=2]\).
+The problem further suggests that the country to which a student belongs
+(\(X\)) may influence both \(T\) and \(Y\). Such scenarios are
+plausible, especially when the teaching methods depend on software or
+access to technology, which may be limited in certain countries {(maybe
+an example with more impact?)}. Figure~\ref{fig-example} illustrates the
+plausible causal structure of this motivating example. A detailed
+examination of Figures \ref{fig-mdag_example1}, \ref{fig-sdag_example1},
+and SCM \ref{fig-scm_example1} reveals the presence of at least four of
+the five fundamental building blocks. The figures display multiple
+descendants, as indicated by pairwise relations such as
+\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\).
+Additionally, the figures feature multiple pairs of unconnected nodes,
+evident from the relations \(e_{T} \perp\!\!\!\perp e_{X}\),
+\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
+Finally, the figures illustrate the fork \(X \rightarrow \{T,Y\}\), and
+two colliders with \(\{X,e_{T}\} \rightarrow T\) and
+\(\{X,T,e_{Y}\} \rightarrow Y\).
 

### 2024-10-15 index.qmd
section:introduction; process:write

@@
         group: Edubron
-    roles: 
-      - conceptualization: lead
-      - methodology: lead
-      - software: only
-      - validation: only
-      - formal analysis: only
-      - investigation: only
-      - resources: no
-      - data curation: only
-      - writing: only
-      - editing: no
-      - visualization: only
-      - supervision: no
-      - project administration: no
-      - funding adquisition: no
   - name: 
@@
         group: Edubron
-    roles: 
-      - conceptualization: support
-      - methodology: support
-      - software: no
-      - validation: no
-      - formal analysis: no
-      - investigation: no
-      - resources: support
-      - data curation: no
-      - writing: no
-      - editing: support
-      - visualization: no
-      - supervision: no
-      - project administration: no
-      - funding adquisition: no
   - name: 
@@
         group: Edubron
-    roles: 
-      - conceptualization: support
-      - methodology: support
-      - software: no
-      - validation: no
-      - formal analysis: no
-      - investigation: no
-      - resources: support
-      - data curation: no
-      - writing: no
-      - editing: support
-      - visualization: no
-      - supervision: yes
-      - project administration: yes
-      - funding adquisition: yes
   - name: 
@@
         group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
-    roles: 
-      - conceptualization: support
-      - methodology: no
-      - software: no
-      - validation: no
-      - formal analysis: no
-      - investigation: no
-      - resources: support
-      - data curation: no
-      - writing: no
-      - editing: support
-      - visualization: no
-      - supervision: yes
-      - project administration: yes
-      - funding adquisition: yes
 funding: 
@@
 abstract: |
-  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait.
-DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020].
-However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data could be generated.
-This study addresses these issues by representing DCJ within the framework of causal inference.
-Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
-The study then translates this model into a probabilistic statistical model to estimate statistical relationships and infer causal effects within the system.
-This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
-Its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.
+  (to do)
 key-points:
@@
 
-<!-- 1. What is comparative judgment? -->
-In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *dichotomous* comparative judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b].
-In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli.
-This assessment results in a dichotomous value indicating which stimulus possesses a higher degree of the trait.
-After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced.
-This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959], which creates a score that corresponds with the trait of interest.
-This score is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.
+<!-- recording -->
+<!-- Part 1; 00:00:00 - 00:09:10 -->
+<!-- Part 2; 00:14:30 - 00:21:00 -->
 
-<!-- 2. in what has been applied on?, and where is the gap? -->
-DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020.
-However, its application transcends education, as exemplified by @Boonen_et_al_2020.
-The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992].
-Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020; @Bouwer_et_al_2023].
-Nevertheless, despite the method's widespread use, existing literature lacks a clear representation of the plausible mechanisms through which DCJ data could be generated.
-Particularly, there is no depiction of the complexity and the assumptions underpinning the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.
+<!-- 1. Where are we now with DCJ? -->
 
-<!-- 3. What are the factors behind DCJ? -->
-According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the method's outcome.
-These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used.
-Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles.
-Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome.
-For example, in @vanDaal_et_al_2016, the authors assessed the academic writing skills of university students (units) using multiple argumentative essays (sub-units).
+<!-- Notes: -->
+<!-- - In the past decade there has been a growing number of studies that document and analyze the merits of DCJ as an assessment tool. Define a bit DCJ. -->
+<!-- - It has been a growing field, where people has tested some important aspects of the method.
+In general, is an iterative process.
+People are iterating on particular aspects of the method, such as: -->
+<!-- a. issues within the method as such, critical analyze some aspect of the method -->
+<!-- b. practical use of the method, generate scores with the BTL, and use this scores with some hypotheses, or descriptions. -->
 
-<!-- 4. Solution to the gap -->
-Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021; @Bouwer_et_al_2023], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome.
-This study aims to fill this gap by representing DCJ within the framework of causal inference.
-Specifically, utilizing the structural approach to causal inference, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system.
-The study then translates the scientific model into a probabilistic statistical model.
-This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.
+Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927; @Pollitt_2012a] for assessing competencies and traits.
+These studies have examined the method's strengths from two main perspectives: its capacity to produce reliable and valid trait scores, and its practical applicability.
+In terms of reliability and validity, research indicates that CJ generates precise and consistent scores that accurately represent the traits being measured.
+Key contributions in this area include those by @Bramley_2008, @Pollitt_2012b, @Whitehouse_2012, @Lesterhuis_2018, @vanDaal_et_al_2019, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023.
+Regarding practical applicability, several studies have demonstrated the method's versatility in both educational and non-educational settings, offering an efficient and effective alternative for measurement and evaluation.
+Notable examples include the works of @Jones_2015, @Bartholomew_et_al_2018, @Jones_et_al_2019, @Marshall_et_al_2020, @Bartholomew_et_al_2020, and @Boonen_et_al_2020.
 
-<!-- 5. final objective and relevance -->
-Ultimately, this study provides a robust causal and probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927].
-Consequently, its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.
+<!-- 2. But what are the problems of DCJ as a method? -->
 
-# Theoretical framework {#sec-framework}
+<!-- Notes:  -->
+<!-- - But what it is the problem with this way of using DCJ (just one paragraph) -->
+<!-- a. measurement model is based on case V -->
+<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
+<!--   a.2 It does not consider that stimuli may be nested in individuals -->
+<!-- b. in all literature there is a disconnection between measurement model and the structural part (common practical use) -->
+<!-- c. what about the experimental design?, what are the implications of certain comparison algorithms? -->
 
-This section introduces fundamental concepts in causal inference but does not offer a comprehensive description of causal inference methods.
-Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020.
-They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful.
-For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books like @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.
+<!-- 3. What is the solution? -->
 
-## The structural approach to causal inference {#sec-framework-structural}
-
-<!-- 1. What does empirical research do first? -->
-Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation.
-In this context, researchers typically frame their research questions as *estimands* or *targets of inference*.
-These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010].
-For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T=\{1,2\})$.
-Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding.
-The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the random quantity $E[Y| T=1] - E[Y| T=2]$, where $E[\cdot]$ denotes the expected value.
-An example of this approach is seen in @Jones_et_al_2019.
-
-<!-- 2. What do they do next? -->
-Researchers then proceed to identify the estimands.
-*Identification* refers to the process of accurately computing an estimand using an estimator.
-An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
-Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
-For instance, the Z-test is an estimator known for its effectiveness in comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006].
-The Z-test is expressed as a signal-to-noise ratio: $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
-The signal is the difference between the group sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y| T=1]$ and $E[Y| T=2]$, respectively.
-The noise $\hat{s}_{p}$ is the unpooled sample variability observed between the two groups.
-
-<!-- the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the comparison of two groups, as long as the assumptions of the model are met.
-Usually these estimates are expressed in terms of the proportions of groups  -->
-
-<!-- 3. But we want these estimands to be causal -->
-However, many studies aim to understand the mechanisms underlying specific data and also seek to establish causal relationships rather than merely infer associations.
-In the earlier example, the differences between groups obtained using the Z-test, referred to as the associational estimate, can be interpreted as causal because the data were collected through a randomized experiment.
-Randomized experiments enable the causal interpretation of associational estimates by ensuring several key properties, including common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
-The most crucial property, however, is that randomization effectively eliminates confounding.
-*Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious associations [@Everitt_et_al_2010].
-Randomization mitigates this issue by decoupling the intervention assignment mechanism, such as assigning students to different groups, from other variables and outcomes [@Morgan_et_al_2014; @Neal_2020].
-
-<!-- 4. Experiments are not available to everyone, then causal inference comes to the rescue -->
-Experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014].
-However, researchers often face constraints that limit their ability to conduct experimental studies.
-These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020].
-In these situations, causal inference provides a valuable alternative for generating causal estimates, particularly when the goal is to understand the mechanisms underlying specific data.
-Moreover, the framework provides significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].
-
-<!-- 5. What is causal inference? -->
-Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, *causal inference* is a framework designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-The framework employs rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009].
-This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" a concept known as counterfactuals, essential for understanding and defining causal effects.
-*Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a specific cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-
-<!-- 6. But how does it work? --> 
-Although a comprehensive discussion of causes and counterfactuals exceeds the scope of this document, a brief overview of how the framework addresses the fundamental problem of causality can be provided.
-The framework begins by defining *individual causal effects* (ICE) as the difference between students' observed and unobserved potential outcomes: $\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)$.
-Notice when a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, and thus termed a *counterfactual*.
-To overcome the challenge posed by counterfactuals, the framework extends the ICE to *average causal effects* (ACE).
-The ACE is defined as $\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, representing the difference between the average of observed potential outcomes and counterfactuals across the sample.
-Finally, similar to experimental studies, the counterfactuals are identified from associational estimates by ensuring the absence of confounding.
-This is accomplished by statistically conditioning on a *sufficient adjustment set* of variables ($X$).
-As a result, the ACE is expressed as $\tau = E_{X}\left[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]\right]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].
-
-<!-- 7. How do we apply causal inference and counterfactuals? -->
-Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
-Both approaches employ rigorous mathematical notation to characterize the ACE, but they do so in different ways [@Neal_2020].
-The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs do-calculus [@Pearl_2009].
-Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].
-
-<!-- 8. But which approach are we going to use? -->
-The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of a system through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-DAGs function as heuristics, effectively conveying the presumed causal structure of the system underlying the ACE, referred to as a *scientific model*.
-They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].
-The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.
-
-<!-- A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. -->
-
-![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}
-
-## DAGs and SCMs {#sec-framework-dag}
-
-<!-- 1. Graph theory and the use of graphs -->
-Graph theory is a branch of mathematics focused on the study of graphs.
-Graphs are mathematical structures modeling pairwise relations between objects.
-They can represent physical relations, such as electrical circuits and roadways, and less tangible structures, such as ecosystems and sociological relations.
-Graphs have proven useful in various fields, including computer science, operations research, and the natural and social sciences [@Gross_et_al_2018].
-
-<!-- 2. What does the structural approach uses?, and what are DAGs? -->
-In statistics, one application incorporating concepts from graph theory is causal inference.
-Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a graphical and formal representation of the causal structure of a system [@Neal_2020].
-In this context, a *graph* denotes a collection of nodes connected by edges, where nodes represent random variables.
-The term *directed* indicates the edges of the graph extend from one node to another, with arrows showing the direction of causal influence.
-Moreover, the term *acyclic* indicates the causal influences do not form a loop, meaning the influences do not cycle back on themselves [@McElreath_2020].
-
-<!-- 3. What are the advantages of DAGs? -->
-DAGs offer two key advantages for modeling causal structures.
-Firstly, they represent causal relations in a nonparametric and fully interactive manner.
-This feature allows for feasible causal analysis strategies without needing the specification of the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014].
-Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-This feature enables the decomposition of complex structures into basic building blocks, facilitating the analysis of these structures by focusing on the causal assumptions associated with individual building blocks [@McElreath_2020].
-These building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].
-
-<!-- commands for d-separation -->
-\newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
-
-::: {#fig-dags_scms layout="[ [1,1,1],[1,1,1],[1,1,1],[1,1,1],[1,1,1] ]"}
-
-![Two unconnected nodes](images/figures/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" width=80%}
-
-![Two unconnected nodes](images/figures/sdag_bb1.png){#fig-sdag_bb1 fig-align="center" width=35%}
-
-::: {#fig-scm_bb1 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \dsep e_{Y}
-\end{aligned}
-$$
-
-Two unconnected nodes
-:::
-
-![Two connected nodes or descendant](images/figures/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" width=80%}
-
-![Two connected nodes or descendant](images/figures/sdag_bb2.png){#fig-sdag_bb2 fig-align="center" width=35%}
-
-::: {#fig-scm_bb2 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(X,e_{Y}) \\
-  e_{X} & \dsep e_{Y}
-\end{aligned}
-$$
-
-Two connected nodes or descendant
-:::
-
-![Chain or pipe](images/figures/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" width=100%}
-
-![Chain or pipe](images/figures/sdag_bb3.png){#fig-sdag_bb3 fig-align="center" width=55%}
-
-::: {#fig-scm_bb3 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Z} \\
-  e_{Z} & \dsep e_{Y}
-\end{aligned}
-$$
-
-Chain or pipe
-:::
-
-![Fork](images/figures/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" width=100%}
-
-![Fork](images/figures/sdag_bb4.png){#fig-sdag_bb4 fig-align="center" width=55%}
-
-::: {#fig-scm_bb4 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(Z,e_{X}) \\
-  Z & := f_{Z}(e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Z} \\
-  e_{Z} & \dsep e_{Y}
-\end{aligned}
-$$
-
-Fork
-:::
-
-![Collider or inmorality](images/figures/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" width=100%}
-
-![Collider or inmorality](images/figures/sdag_bb5.png){#fig-sdag_bb5 fig-align="center" width=55%}
-
-::: {#fig-scm_bb5 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,Y,e_{Z}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Z} \\
-  e_{Z} & \dsep e_{Y}
-\end{aligned}
-$$
-
-Collider or inmorality
-:::
-
-The five fundamental building blocks of DAGs.
-**Note:** left panels show the magnified representation, middle panels show the standard representation, and the right panels show their corresponding SCM form.
-:::
-
-<!-- 3. how DAGs are depicted? -->
-<!-- 3.1 the magnified DAG -->
-The left panels of @fig-dags_scms illustrate the *magnified* representation.
-These graphs depict the *endogenous* variables $V=\{X,Z,Y\}$ alongside the *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$.
-Endogenous variables are those whose causal mechanisms the investigator chooses to model [@Neal_2020].
-In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
-27,68].
-The graphs show endogenous variables as solid black circles to signify that they are observed random variables, while endogenous variables are depicted as open circles to signify their unobserved (latent) nature.
-Lastly, the arrows in the graphs reflect the expected direction of causal influences among these variables.
-
-<!-- 3.2 the simplified DAG -->
-Often, DAGs omit the exogenous variables for simplicity, resulting in the *standard* representation.
-However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in @sec-framework-flow.
-The standard representation is illustrated in the middle panels of @fig-dags_scms.
-
-<!-- 3.3 the SCM -->
-Lastly, the right panels of @fig-dags_scms depict the SCM form of the fundamental building blocks.
-SCMs are formal mathematical models defined by a set of endogenous variables $V$, a set of exogenous variables $E$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020].
-These functions, referred to as structural equations, specify each endogenous variable as nonparametric functions of other variables.
-Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence.
-The concepts of d-separation and causal (in)dependence are explored in @sec-framework-flow.
-
-<!-- 4. But what does the building blocks mean? -->
-A careful examination of @fig-dags_scms highlights the assumptions underlying these building blocks.
-Figures [-@fig-mdag_bb1], [-@fig-sdag_bb1], and SCM [-@fig-scm_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related.
-Figures [-@fig-mdag_bb2], [-@fig-sdag_bb2], and SCM [-@fig-scm_bb2] illustrate two connected nodes, showing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$.
-Consequently, $Y$ is considered a *descendant* of $X$.
-Figures [-@fig-mdag_bb3], [-@fig-sdag_bb3], and SCM [-@fig-scm_bb3] depict a *chain* or *pipe*, where $X$ influences $Z$, and $Z$ influences $Y$.
-In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$.
-This creates a *directed path* between $X$ and $Y$.
-Consequently, $X$ is an *ancestor* of $Y$, and $Z$ fully *mediates* the relationship between the two.
-Figures [-@fig-mdag_bb4], [-@fig-sdag_bb4], and SCM [-@fig-scm_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
-Here, $Z$ is a parent node of $X$ and $Y$.
-Finally, Figures [-@fig-mdag_bb5], [-@fig-sdag_bb5], SCM [-@fig-scm_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$.
-In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$.
-Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and of all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.
-
-<!-- 5. continuing with the motivating example -->
-The motivating example in @sec-framework-structural can be used to further illustrate how the five fundamental building blocks help construct a system's causal structure.
-In this scenario, an experiment cannot be conducted but the investigator still aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y=\{0,1\})$ between two groups of students $(T=\{1,2\})$, described by the estimand $E[Y| T=1] - E[Y| T=2]$.
-The problem further suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$.
-Such scenarios are plausible, especially when the teaching methods depend on software or access to technology, which may be limited in certain countries [(maybe an example with more impact?)]{style="color:blue;"}.
-@fig-example illustrates the plausible causal structure of this motivating example.
-A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks.
-The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
-Additionally, the figures feature multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
-Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
-
-::: {#fig-example layout="[ [1,1,1],[1,1,1],[1,1,1] ]"}
-
-![Magnified representation](images/figures/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=100%}
-
-![Standard representation](images/figures/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%}
-
-::: {#fig-scm_example1 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{Z}(X,e_{T}) \\
-  Y & := f_{Y}(T,X,e_{Y}) \\
-  e_{T} & \dsep e_{X} \\
-  e_{T} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Y}
-\end{aligned}
-$$
-
-Structural causal model
-:::
-
-DAGs for a plausible causal structure in a system.
-:::
-
-## The flow of association and causation in graphs {#sec-framework-flow}
-
-<!-- The main assumption we’ve seen that relates the graph to the distribution -->
-<!-- is the Markov assumption. The Markov assumption tells us if variables are -->
-<!-- d-separated in the graph 𝐺 , then they are independent in the distribution -->
-<!-- 𝑃 (Theorem 3.1): -->
-<!-- Maybe we can detect independencies in the data and then use that -->
-<!-- to infer the causal graph. However, going from independencies in the -->
-<!-- distribution 𝑃 to d-separations in the graph 𝐺 isn’t something that the -->
-<!-- Markov assumption gives us (see Equation 3.20 above). Rather, we need -->
-<!-- the converse of the Markov assumption. This is known as the faithfulness -->
-<!-- assumption. -->
-<!-- Assumption 11.1 (Faithfulness) -->
-
-<!-- 1. What has been said about DAGs? -->
-<!-- The previous section gives a concise graphical intuition for DAGs, sufficient to emphasize their effectiveness in encoding causal structures.
-However, so far, nothing has been said about how to identify causal quantities or how to formalize DAGs into probabilistic statistical models.
--->
-
-<!-- “identification” in this book, we are referring to the process -->
-<!-- of moving from a causal estimand to an equivalent statistical estimand. -->
-<!-- When we say “estimation,” we are referring to the process of moving from -->
-<!-- a statistical estimand to an estimate. Neal -->
-
-<!-- The previous section gives a concise graphical intuition for DAGs, sufficient to emphasize their effectiveness in encoding causal structures.
-However, it doesn’t explain how to identify causal quantities and formalize causal models.
-Neal -->
-
-<!-- However, the requirements to use DAGs as encoders of causal assumptions and how these can transform into probabilistic statistical models have not yet been discussed. -->
-
-<!-- 2. What is the local markov assumption? and where does it lead? -->
-<!-- 2.1 Bayesian network factorization -->
-
-<!-- 3. What is the causal edges assumption? and where does it lead? -->
-
-<!-- 4. Combined they are known as the minimality assumption? -->
-
-<!-- 5. Now, what are those mysterious functions?, and how do they become statistical quantities? -->
-
-<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
-![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=80%}
-
-<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->
-
-<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables -->
-<!-- with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
--->
-
-<!-- 6. What is the modularity assumption? and where does it lead? -->
-<!-- 6.1 the Truncated factorization -->
-<!-- 6.2 The law of counterfactuals -->
-
-<!-- 7. Causal mechanism -->
-<!-- 7.3 the backdoor criterion (do you need it?) -->
-
-<!-- 5. d-separation -->
-<!-- 5.1 Blocked path definition-->
-
-<!-- 6. putting it all together -->
-<!-- 6.1. Revising the building block: dependence and independence -->
-
-<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
-
-<!-- Mention convenient R package daggitty to draw dags and extract the causal assumption inherent in a graph -->
-
-<!-- After detailing how to build graphs and what they imply, Include steps to draw dags using a motivating example -->
-
-<!-- point out that the disturbances are usually stochastic in nature -->
-
-<!-- Causal mechanism as a result of markov and minimality assumption -->
-<!-- Modularity assumption also -->
-
-<!-- 4. But how do we move from targets of inference to estimates -->
-<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->
-
-<!-- Ardmed with dags and scm, and theit probabilité implication, we can now expand on thé motivating example provided in section xx -->
+<!-- Notes:  -->
+<!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
+<!-- - This paper tries to systematically integrate all aspects at play in a single scientific model, we will build the model in a stepwise manner using -->
+<!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->
 
 # Theory {#sec-theory}
+<!-- 21:00 recording -->
 
-## A scientific model for the DCJ {#sec-theory-scientific}
+<!-- Notes: -->
+<!-- - But what it is the problem with this way of using DCJ (just one paragraph) -->
+<!-- a. measurement model is based on case V -->
+<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
+<!--   a.2 It does not consider that stimuli may be nested in individuals -->
+<!-- b. in all literature there is a disconnection between measurement model and the structural part (common practical use) -->
+<!-- c. what about the experimental design?, what are the implications of certain comparison algorithms? -->
 
-<!-- A *scientific (causal) model* describe the causal assumptions and mechanisms underlying a system, and facilitate inferences about causal relationships from data [@Hitchcock_2023]. -->
+## A scientific model for the CJ {#sec-theory-scientific}
 
-<!-- From Lawson (2015), another way to see DCJ experiments I'd that you are grouping heterogeneous experimental units (objects) into homogeneous ones (subjects), which are your "blocks" -->
-
-![ DCJ causal diagram, simplified description ](images/figures/SciModel_simp1.png){#fig-SciModel_simp1 fig-align="center" width=80%}
-
-![ DCJ causal diagram, simplified mathematical description ](images/figures/SciModel_simp2.png){#fig-SciModel_simp2 fig-align="center" width=70%}
-
-![ DCJ causal diagram, population mathematical description ](images/figures/SciModel_pop.png){#fig-SciModel_pop fig-align="center" width=70%}
-
-<!-- check the directionality of the arrow from $p_{I}$ to $\sigma_{i(g)}$ and equivalents -->
-
-![ DCJ causal diagram, sample with comparisons mathematical description ](images/figures/SciModel_samp.png){#fig-SciModel_samp fig-align="center" width=80%}
-
-<!-- Assuming certain values are decided in the experimental design (proportions, sample size, etc) you can reduce the graph -->
-
-<!-- It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
-However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
-
-## Probabilitics assumptions of the scientific model {#sec-theory-probability}
-
-<!-- what does it mean in probabilistic terms? describe probabilities and arrive to structural model -->
-
-<!-- Final models does not require the estimation of latent variables, the results can be evaluated without them -->
-
-<!-- Notice that latent variables are kind of like descendants -->
-
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!--   O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\ -->
-<!--   \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\ -->
-<!--   \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\ -->
-<!--   \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\ -->
-<!--   \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\ -->
-<!--   & \; e_{k} \dsep e_{i} \\ -->
-<!--   & \; e_{k} \dsep e_{r_{i}} \\ -->
-<!--   & \; e_{i} \dsep e_{r_{i}} -->
-<!-- \end{aligned} -->
-<!-- $$ {#eq-StructuralModel} -->
-
-## From the scientific to statistical model {#sec-theory-statistics}
-
-<!-- "An appropriate analysis of data cannot be completed without knowledge of what experimental design was used and how the data was collected, and conclusions are not reliable if they are not justified by the proper modeling and analysis of the data" (Lawson, 2015, p.
-7) -->
-
-<!-- "In many cases where observational units are different than the experimental units, the observational units will not be independent" (p. 177) -->
-
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!--   O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\ -->
-<!--   \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\ -->
-<!--   \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\ -->
-<!--   \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\ -->
-<!--   \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\ -->
-<!--   e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\ -->
-<!--   e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\ -->
-<!--   e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)}) -->
-<!-- \end{aligned} -->
-<!-- $$ {#eq-StatModel_general} -->
-
-<!-- for identification purposes we can set $\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02$, $\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1$, and $\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1$.
-A special case of this would be to assume that the data comes from the same population, in that case, $\sigma_{k(g)} = \sigma_{k} = 0.02$, $\sigma_{i(g)} = \sigma_{i} = 1$ -->
-
-## Let's talk about Thurstone {#sec-theory-thurstone}
-
-<!-- Thurstone's comparative judgment @Thurstone_1927 is based on the formula: -->
-<!-- $$ -->
-<!-- X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}} -->
-<!-- $$ -->
-
-<!-- where $X_{AB}$ defines the comparative judgment outcome, $S_{A}$ and $S_{B}$ are the modal discriminal processes, $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}$, with $\sigma_{A}$ and $\sigma_{B}$ being the dispersion of discriminal processes $A$ and $B$, respectively, and $\rho$ the correlation between discriminal processes.
--->
-
-<!-- The theory identifies five cases: -->
-
-<!-- - **Case 1:** only constant $\rho$ (not $\rho_{ij}$) -->
-<!-- - **Case 2:** $X_{ij}$ becomes $X_{kij}$ with $k=1, \dots, K$ judges (replication, not duplication) -->
-<!-- - **Case 3:** $\rho = 0$, then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$  -->
-<!-- - **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$ -->
-<!-- - **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$ -->
-
-<!-- Now using the DAG and statistical notation -->
-
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!--   O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\ -->
-<!--   \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\ -->
-<!--   \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\ -->
-<!--   \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\ -->
-<!--   \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\ -->
-<!--   e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\ -->
-<!--   e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\ -->
-<!--   e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)}) -->
-<!-- \end{aligned} -->
-<!-- $$ {#eq-StatModel_thurstone} -->
-
-<!-- The theory identifies five cases: -->
-
-<!-- - **Case 1:** only constant $\rho \approx \sigma_{i}$ -->
-<!-- - **Case 2:** now judges are separated by using $\gamma_{k}$ -->
-<!-- - **Case 3:** $\rho \approx \sigma_{e_{i}} = 0$ (no nesting of texts on students), then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$  -->
-<!-- - **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$ -->
-<!-- - **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$ -->
-
-<!-- But now can we see other scenarios than just those 5 cases? -->
-
-<!-- - consider different $\rho \approx \sum_{p=1}^{P} \sigma_{p}$, depending on $P$ nesting structures  -->
-<!-- - we can now investigate $\gamma_{k}$ -->
-<!-- - we can assume $\sigma_{B} \neq \sigma_{A}$, no need for results on the limit -->
+## From theory to statistical model {#sec-theory-statmodel}
 
@@
 
-<!-- no fundamental ontological objections to using DAGs to describe causal models and causal relationships. -->
-
 ## Limitations and further research {#sec-discuss-limitations}
-<!-- and most of the methods depend on unverifiable assumptions, causal inference can still identify causes and estimate their effects, assuming a set of plausibles mechanisms for a system. -->
 
@@
 
-<!-- **Acknowledgements:** NO additional acknowledgements -->
+<!-- **Acknowledgements:** -->
 
@@
 # Appendix {#sec-appendix}
-
-## Why do we need to estimate judges' abilities? {#sec-appA}
-
-<!-- The same idea as interactions: when interactions are significant, the interpretations of the parameters cannot be made in isolation -->
-
-## Latent variables as a mean of imputation {#sec-appB}
-
-<!-- Latent variables and random effects are in essence imputation? (McElreath, p. 504) -->
-<!-- Judge index works as a descendant of the latent variables, or the latent variable is not a reflective, but generative where the index is multiplied by a loading? -->
-<!-- see Morgan et al (2014) -->
-<!-- see Mayer (2019) -->
-
-## Other comparative scenarios  {#sec-appC}
 

### 2024-10-15 _tex/index.tex
Built site for gh-pages

--- 2024-08-16 _tex/index.tex
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage[]{natbib}
@@
 \begin{abstract}
-Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
-of stimuli to determine which one exhibits a higher degree of a specific
-trait. DCJ has proven effective and reliable across various fields
-\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
-However, despite the method's widespread use, existing literature lacks
-a clear explanation of the complexities and assumptions underpinning the
-DCJ system, as well as the plausible mechanisms through which DCJ data
-could be generated. This study addresses these issues by representing
-DCJ within the framework of causal inference. Specifically, utilizing
-the structural approach, the study develops a scientific model to
-clarify plausible causal assumptions and mechanisms inherent in the DCJ
-system. The study then translates this model into a probabilistic
-statistical model to estimate statistical relationships and infer causal
-effects within the system. This research provides a robust probabilistic
-foundation for the statistical analysis of DCJ data, building upon
-Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
-Its findings offer valuable insights for researchers and analysts
-designing and implementing DCJ experiments.
+(to do)
 \end{abstract}
@@
 
-In contemporary contexts, Thurstone's law of comparative judgment
-\citeyearpar{Thurstone_1927} primarily refers to the method of
-\emph{dichotomous} comparative judgment
-\citep[DCJ,][]{Pollitt_2012a, Pollitt_2012b}. In DCJ, a judge assesses
-the relative manifestation of a \emph{trait} within a pair of stimuli.
-This assessment results in a dichotomous value indicating which stimulus
-possesses a higher degree of the trait. After different judges perform
-multiple rounds of pairwise comparisons, an outcome vector is produced.
-This vector is modeled using the Bradley-Terry-Luce model
-\citep[BTL,][]{Bradley_et_al_1952, Luce_1959}, which creates a score
-that corresponds with the trait of interest. This score is then used to
-rank the stimuli from lowest to highest or to evaluate the influence of
-certain variables on the stimuli's positions in the ranking.
-
-DCJ has proven effective in assessing competencies and traits
-predominantly within the educational realm, as demonstrated by
-\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
-\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
-\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
-However, its application transcends education, as exemplified by
-\citet{Boonen_et_al_2020}. The methodology has also evolved to include
-multiple, as opposed to pairwise comparisons
-\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
-ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
-suggests that DCJ offers an alternative and efficient approach to
-measurement and evaluation, characterized by its reliability and
-validity
-\citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020, Bouwer_et_al_2023}.
-Nevertheless, despite the method's widespread use, existing literature
-lacks a clear representation of the plausible mechanisms through which
-DCJ data could be generated. Particularly, there is no depiction of the
-complexity and the assumptions underpinning the DCJ system, nor how
-different assessment factors can potentially influence the observed DCJ
-outcome.
-
-According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
-several assessment factors interact and influence the method's outcome.
-These factors include the number and characteristics of the stimuli,
-their \emph{proximity} in terms of the assessed trait, the number of
-comparison per stimulus, and the pairing algorithm used. Furthermore,
-since the method relies on judges' assessments, the number and
-characteristics of judges, their \emph{discrimination} abilities, and
-the number of comparisons per judge also play pivotal roles. Moreover,
-when the stimuli represent sub-units of higher-levels units, factors
-such as the number and characteristics of these units, along with their
-\emph{proximity} in terms of the assessed trait, can significantly
-influence the outcome. For example, in \citet{vanDaal_et_al_2016}, the
-authors assessed the academic writing skills of university students
-(units) using multiple argumentative essays (sub-units).
-
-Although several studies have examined the individual impact of these
-factors on the method's reliability
-\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021, Bouwer_et_al_2023},
-none, to the best of the authors' knowledge, have provided a transparent
-depiction of the DCJ system and the mechanisms generating the DCJ
-outcome. This study aims to fill this gap by representing DCJ within the
-framework of causal inference. Specifically, utilizing the structural
-approach to causal inference, the study develops a scientific model to
-clarify plausible causal assumptions and mechanisms inherent in the DCJ
-system. The study then translates the scientific model into a
-probabilistic statistical model. This model aims to produce statistical
-estimates to draw inferences about plausible causal relationships within
-the DCJ system.
-
-Ultimately, this study provides a robust causal and probabilistic
-foundation for the statistical analysis of DCJ data, building upon
-Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
-Consequently, its findings offer valuable insights for researchers and
-analysts designing and implementing DCJ experiments.
-
-\section{Theoretical framework}\label{sec-framework}
-
-This section introduces fundamental concepts in causal inference but
-does not offer a comprehensive description of causal inference methods.
-Readers interested in deeper exploration should consult introductory
-papers like \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019},
-and \citet{Cinelli_et_al_2020}. They may also find introductory books
-such as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
-\citet{McElreath_2020} useful. For more advanced study, seminal
-intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
-\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
-like \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
-\citet{Hernan_et_al_2020} are recommended.
-
-\subsection{The structural approach to causal
-inference}\label{sec-framework-structural}
-
-Empirical research addresses real-world challenges by relying on
-evidence gathered through observation and experimentation. In this
-context, researchers typically frame their research questions as
-\emph{estimands} or \emph{targets of inference}. These estimands
-represent the specific quantities the study aims to determine
-\citep{Everitt_et_al_2010}. For instance, a study might examine the
-question, ``To what extent do different teaching methods (\(T\))
-influence students' conceptual understanding of a topic (\(Y\))?'' To
-investigate this, the study could randomly assign students to two
-groups, each using a different teaching method \((T_{i}=\{1,2\})\).
-Students' conceptual understanding of the topic could be evaluated
-through pairwise comparisons, resulting in a dichotomous outcome
-\((Y_{i}=\{0,1\})\), indicating which student among those compared has a
-higher level of understanding. The research question could be then
-framed as the estimand, ``\emph{On average}, is there a difference in
-conceptual understanding of the topic between the two groups of
-students?'' This estimand could be mathematically expressed by the
-associational quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where
-\(E[\cdot]\) denotes the expected value. An example of this approach is
-seen in \citet{Jones_et_al_2019}.
-
-Researchers then proceed to identify the estimands.
-\emph{Identification} refers to the process of accurately computing an
-estimand using an estimator. An \emph{estimator} is a method or function
-that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
-numerical values that approximate the estimand and are derived through
-\emph{estimation}, which refer to the process of integrating data with
-an estimator \citep{Everitt_et_al_2010}. Although various methods can
-approximate an estimand, researchers prioritize estimators with
-desirable properties that ensure the accuracy of estimates. For
-instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
-estimator known for its effectiveness in modeling pairwise comparisons,
-yielding accurate estimates for the target of inference discussed
-earlier, as long as the assumptions of the model are met.
-
-However, many studies aim to understand the mechanisms underlying
-specific data types and also seek to establish causal relationships
-rather than merely infer associations. In the earlier example, the
-differences between groups obtained using the BTL model, referred to as
-the associational estimates, can be interpreted as causal because the
-data were collected through a randomized experiment. Randomized
-experiments enable the causal interpretation of associational estimates
-by ensuring several key properties, including common support, no
-interference, and consistency \citep{Morgan_et_al_2014, Neal_2020}. The
-most crucial property, however, is that randomization effectively
-eliminates confounding. \emph{Confounding} occurs when an external
-variable influences both the outcome and the variable of interest,
-leading to spurious associations \citep{Everitt_et_al_2010}.
-Randomization mitigates this issue by decoupling the intervention
-assignment mechanism, such as assigning students to different groups,
-from other variables and outcomes \citep{Morgan_et_al_2014, Neal_2020}.
-
-Experiments are widely regarded as the gold standard in evidence-based
-science \citep{Hariton_et_al_2018, Hansson_2014}, however, researchers
-often encounter constraints that limit their ability to conduct
-experimental studies. These constraints include ethical concerns, such
-as the assignment of individuals to potentially harmful interventions,
-and practical limitations, such as the infeasibility of, for example,
-assigning individuals to genetic modifications or physical impairments
-\citep{Neal_2020}. In these instances, causal inference offers a
-valuable alternative for generating causal estimates. \emph{Causal
-inference} is a framework designed to identify the causes of phenomena
-and estimate their effects using data
-\citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
-modeling, which focuses primarily on summarizing data and inferring
-associations, causal inference provides a rigorous mathematical
-framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
-Consequently, the framework offers significant theoretical insights that
-enhance the design of observational and experimental studies
-\citep{McElreath_2020}.
-
-While a detailed explanation of causes and counterfactuals is beyond the
-scope of this document, it is important to note that causal inference
-addresses the \emph{fundamental problem of causality} through the use of
-counterfactuals \citep{Neal_2020}. The framework defines
-\emph{individual causal effects} (ICE) as the difference between an
-individual's observed and unobserved potential outcomes. These
-unobserved potential outcomes, known as \emph{counterfactuals},
-represent hypothetical scenarios that are \emph{contrary to fact}, where
-alternative outcomes for an individual, resulting from a specific cause,
-are neither observed nor observable
-\citep{Neal_2020, Counterfactual_2024}. The framework then extends this
-estimand to \emph{average causal effects} (ACE), which captures the
-difference between sample averages of observed potential outcomes and
-counterfactuals. Finally, much like in an experiment, the framework
-identifies the ACE from associational estimates by conditioning for a
-\emph{sufficient adjustment set} of variables, ensuring the absence of
-confounding \citep{Morgan_et_al_2014}.
-
-Several approaches to causal inference and counterfactuals exist, but
-two are particularly prominent: the potential outcomes approach, also
-known as the Neyman-Rubin causal model
-\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
-\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
-mathematical notation to characterize the ACE, but they do so in
-different ways \citep{Neal_2020}. The potential outcomes approach relies
-on counterfactual notation, whereas the structural approach employs
-do-calculus \citep{Pearl_2009}. Despite these differences, both
-notations can be expressed in terms of the other, and both approaches
-provide methods for using experimental and observational data to
-estimate causal effects \citep{Pearl_2010}.
-
-The structural approach, however, offers a notable advantage over the
-potential outcomes approach by allowing the graphical and formal
-representation of theories through directed acyclic graphs
-\citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
-DAGs function as heuristics, effectively conveying the presumed causal
-structure of a system, referred to as a \emph{scientific model}. They do
-not represent detailed statistical models but allow researchers to
-deduce which statistical models can provide valid causal inferences,
-assuming the causal structure depicted in the DAGs are accurate
-\citep{McElreath_2020}. The Identification-Estimation flowchart in
-Figure~\ref{fig-IEflow} visually represents this process of
-transitioning from estimands to estimates, as well as the application of
-the scientific model and data to identify and estimate causal effects.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/IEflow.png}
-
-}
-
-\caption{\label{fig-IEflow}Identification-Estimation flowchart.
-Extracted and slightly modified from \citet[32]{Neal_2020}}
-
-\end{figure}%
-
-\subsection{DAGs and SCMs}\label{sec-framework-dag}
-
-Graph theory is a branch of mathematics focused on the study of graphs.
-Graphs are mathematical structures modeling pairwise relations between
-objects. They can represent physical relations, such as electrical
-circuits and roadways, and less tangible structures, such as ecosystems
-and sociological relations. Graphs have proven useful in various fields,
-including computer science, operations research, and the natural and
-social sciences \citep{Gross_et_al_2018}.
-
-In statistics, one application incorporating concepts from graph theory
-is causal inference. Specifically, the structural approach to causal
-inference uses directed acyclic graphs (DAG) to provide a graphical and
-formal representation of the causal structure of a system
-\citep{Neal_2020}. In this context, a \emph{graph} denotes a collection
-of nodes connected by edges, where nodes represent random variables. The
-term \emph{directed} indicates the edges of the graph extend from one
-node to another, with arrows showing the direction of causal influence.
-Moreover, the term \emph{acyclic} indicates the causal influences do not
-form a loop, meaning the influences do not cycle back on themselves
-\citep{McElreath_2020}.
-
-DAGs offer two key advantages for modeling causal structures. Firstly,
-they represent causal relations in a nonparametric and fully interactive
-manner. This feature allows for feasible causal analysis strategies
-without needing the specification of the type of data or the nature of
-the functional dependence among variables \citep{Morgan_et_al_2014}.
-Secondly, regardless of complexity, DAGs can represent various causal
-structures using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. This feature enables the
-decomposition of complex structures into basic building blocks,
-facilitating the analysis of these structures by focusing on the causal
-assumptions associated with each building block individually
-\citep{McElreath_2020}. These building blocks can be represented in
-three ways: the magnified representation, the standard representation,
-and the structural causal model form \citep[SCM,][]{Morgan_et_al_2014}.
-
-\newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
-
-\begin{figure}
-
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb1.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb1}Two unconnected nodes}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb1.png}
-
-}
-
-\subcaption{\label{fig-sdag_bb1}Two unconnected nodes}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb1}Two unconnected nodes}
-
-\end{minipage}%
-\newline
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/mdag_bb2.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb2}Two connected nodes or descendant}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_bb2.png}
-
-}
-
-\subcaption{\label{fig-sdag_bb2}Two connected nodes or descendant}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(X,e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb2}Two connected nodes or descendant}
-
-\end{minipage}%
-\newline
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb3.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb3}Chain or pipe}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb3.png}
-
-}
-
-\subcaption{\label{fig-sdag_bb3}Chain or pipe}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Z} \\
-  e_{Z} & \perp\!\!\!\perp e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb3}Chain or pipe}
-
-\end{minipage}%
-\newline
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb4.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb4}Fork}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb4.png}
-
-}
-
-\subcaption{\label{fig-sdag_bb4}Fork}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(Z,e_{X}) \\
-  Z & := f_{Z}(e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Z} \\
-  e_{Z} & \perp\!\!\!\perp e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb4}Fork}
-
-\end{minipage}%
-\newline
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_bb5.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb5}Collider or inmorality}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb5.png}
-
-}
-
-\subcaption{\label{fig-sdag_bb5}Collider or inmorality}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,Y,e_{Z}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Z} \\
-  e_{Z} & \perp\!\!\!\perp e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb5}Collider or inmorality}
-
-\end{minipage}%
-
-\caption{\label{fig-dags_scms}The five fundamental building blocks of
-DAGs. \textbf{Note:} left panels show the magnified representation,
-middle panels show the standard representation, and the right panels
-show their corresponding SCM form.}
-
-\end{figure}%
-
-The left panels of Figure~\ref{fig-dags_scms} illustrate the
-\emph{magnified} representation. These graphs depict the
-\emph{endogenous} variables \(V=\{X,Z,Y\}\) alongside the
-\emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\). Endogenous
-variables are those whose causal mechanisms the investigator chooses to
-model \citep{Neal_2020}. In contrast, exogenous variables represent
-\emph{errors} or \emph{disturbances} arising from omitted factors that
-the investigator chooses not to model explicitly
-\citep[27,68]{Pearl_2009}. The graphs show endogenous variables as solid
-black circles to signify that they are observed random variables, while
-endogenous variables are depicted as open circles to signify their
-unobserved (latent) nature. Lastly, the arrows in the graphs reflect the
-expected direction of causal influences among these variables.
-
-Often, DAGs omit the exogenous variables for simplicity, resulting in
-the \emph{standard} representation. However, including exogenous
-variables in a graph can be beneficial in some scenarios, as their
-presence can reveal potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}, concepts explored in
-Section~\ref{sec-framework-flow}. The standard representation is
-illustrated in the middle panels of Figure~\ref{fig-dags_scms}.
-
-Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
-form of the fundamental building blocks. SCMs are formal mathematical
-models defined by a set of endogenous variables \(V\), a set of
-exogenous variables \(E\), and a set of functions
-\(F=\{f_{X},f_{Z},f_{Y}\}\) \citep{Pearl_2009, Neal_2020}. These
-functions, referred to as structural equations, specify each endogenous
-variable as nonparametric functions of other variables. Moreover, SCMs
-use the symbol `\(:=\)' to indicate the variables' asymmetrical causal
-dependence and the symbol `\(\perp\!\!\!\perp\)' to represent
-\emph{d-separation}, which roughly equates to the concept of variable
-independence. The concepts of d-separation and causal (in)dependence are
-explored in Section~\ref{sec-framework-flow}.
-
-A careful examination of Figure~\ref{fig-dags_scms} highlights the
-assumptions underlying these building blocks. Figures
-\ref{fig-mdag_bb1}, \ref{fig-sdag_bb1}, and SCM \ref{fig-scm_bb1} depict
-two unconnected nodes, representing a scenario where variables \(X\) and
-\(Y\) are not causally related. Figures \ref{fig-mdag_bb2},
-\ref{fig-sdag_bb2}, and SCM \ref{fig-scm_bb2} illustrate two connected
-nodes, showing a scenario where a \emph{parent} node \(X\) exerts a
-causal influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
-considered a \emph{descendant} of \(X\). Figures \ref{fig-mdag_bb3},
-\ref{fig-sdag_bb3}, and SCM \ref{fig-scm_bb3} depict a \emph{chain} or
-\emph{pipe}, where \(X\) influences \(Z\), and \(Z\) influences \(Y\).
-In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
-parent node of \(Y\). This creates a \emph{directed path} between \(X\)
-and \(Y\). Consequently, \(X\) is an \emph{ancestor} of \(Y\), and \(Z\)
-fully \emph{mediates} the relationship between the two. Figures
-\ref{fig-mdag_bb4}, \ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4}
-illustrate a \emph{fork}, where variables \(X\) and \(Y\) are both
-influenced by \(Z\). Here, \(Z\) is a parent node of \(X\) and \(Y\).
-Finally, Figures \ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM
-\ref{fig-scm_bb5} depict a \emph{collider}, also known as
-\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
-of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
-related to each other but both influence \(Z\). Additionally, in all
-SCMs, the errors are assumed to be mutually independent of each other
-and of all other variables in the graph, as evidenced by the pairwise
-relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
-\(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).
-
-The motivating example in Section~\ref{sec-framework-structural} can
-further illustrate how to use the five fundamental building blocks to
-construct a system's causal structure. In this scenario, the
-investigator aims to determine whether, \emph{on average}, there is a
-difference in conceptual understanding of a topic \((Y_{i}=\{0,1\})\)
-between two groups of students \((T_{i}=\{1,2\})\), described by the
-estimand \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\). However, unlike the
-previous case, an experiment cannot be conducted, and the problem
-suggests that the country to which a student belongs (\(X\)) may
-influence both \(T\) and \(Y\). Such scenarios are plausible, especially
-when the teaching methods depend on software or access to technology,
-which may be limited in certain countries {(maybe an example with more
-impact?)}. Figure~\ref{fig-example} illustrates the plausible causal
-structure of the motivating example. A detailed examination of Figures
-\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
-\ref{fig-scm_example1} reveals the presence of at least four of the five
-fundamental building blocks. The figures display multiple descendants,
-as indicated by pairwise relations such as \(X \rightarrow T\),
-\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
-features multiple pairs of unconnected nodes, evident from the relations
-\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
-\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
-fork \(X \rightarrow \{T,Y\}\), and two colliders with
-\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).
-
-\begin{figure}
-
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_example1.png}
-
-}
-
-\subcaption{\label{fig-mdag_example1}Magnified representation}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_example1.png}
-
-}
-
-\subcaption{\label{fig-sdag_example1}Standard representation}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{Z}(X,e_{T}) \\
-  Y & := f_{Y}(T,X,e_{Y}) \\
-  e_{T} & \perp\!\!\!\perp e_{X} \\
-  e_{T} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_example1}Structural causal model}
-
-\end{minipage}%
-
-\caption{\label{fig-example}DAGs for a plausible causal structure in a
-system.}
-
-\end{figure}%
-
-\subsection{The flow of association and causation in
-graphs}\label{sec-framework-flow}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/ACflow.png}
-
-}
-
-\caption{\label{fig-ACflow}The flow of association and causation in
-graphs. Extracted and slightly modified from \citet[31]{Neal_2020}}
-
-\end{figure}%
+Over the past decade, numerous studies have documented the effectiveness
+of the \emph{comparative judgment} (CJ) method
+\citep{Thurstone_1927, Pollitt_2012a} for assessing competencies and
+traits. These studies have examined the method's strengths from two main
+perspectives: its capacity to produce reliable and valid trait scores,
+and its practical applicability. In terms of reliability and validity,
+research indicates that CJ generates precise and consistent scores that
+accurately represent the traits being measured. Key contributions in
+this area include those by \citet{Bramley_2008}, \citet{Pollitt_2012b},
+\citet{Whitehouse_2012}, \citet{Lesterhuis_2018},
+\citet{vanDaal_et_al_2019}, \citet{Bramley_et_al_2019},
+\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022}, and
+\citet{Bouwer_et_al_2023}. Regarding practical applicability, several
+studies have demonstrated the method's versatility in both educational
+and non-educational settings, offering an efficient and effective
+alternative for measurement and evaluation. Notable examples include the
+works of \citet{Jones_2015}, \citet{Bartholomew_et_al_2018},
+\citet{Jones_et_al_2019}, \citet{Marshall_et_al_2020},
+\citet{Bartholomew_et_al_2020}, and \citet{Boonen_et_al_2020}.
 
@@
 
-\subsection{A scientific model for the DCJ}\label{sec-theory-scientific}
+\subsection{A scientific model for the CJ}\label{sec-theory-scientific}
 
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/SciModel_simp1.png}
-
-}
-
-\caption{\label{fig-SciModel_simp1}DCJ causal diagram, simplified
-description}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{images/figures/SciModel_simp2.png}
-
-}
-
-\caption{\label{fig-SciModel_simp2}DCJ causal diagram, simplified
-mathematical description}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{images/figures/SciModel_pop.png}
-
-}
-
-\caption{\label{fig-SciModel_pop}DCJ causal diagram, population
-mathematical description}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/SciModel_samp.png}
-
-}
-
-\caption{\label{fig-SciModel_samp}DCJ causal diagram, sample with
-comparisons mathematical description}
-
-\end{figure}%
-
-\subsection{Probabilitics assumptions of the scientific
-model}\label{sec-theory-probability}
-
-\subsection{From the scientific to statistical
-model}\label{sec-theory-statistics}
-
-\subsection{Let's talk about Thurstone}\label{sec-theory-thurstone}
+\subsection{From theory to statistical
+model}\label{sec-theory-statmodel}
 
@@
 
-\subsection{Why do we need to estimate judges'
-abilities?}\label{sec-appA}
-
-\subsection{Latent variables as a mean of imputation}\label{sec-appB}
-
-\subsection{Other comparative scenarios}\label{sec-appC}
-
 \newpage{}

### 2024-10-16 index.qmd
section:introduction; process:write

@@
 Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927; @Pollitt_2012a] for assessing competencies and traits.
-These studies have examined the method's strengths from two main perspectives: its capacity to produce reliable and valid trait scores, and its practical applicability.
-In terms of reliability and validity, research indicates that CJ generates precise and consistent scores that accurately represent the traits being measured.
-Key contributions in this area include those by @Bramley_2008, @Pollitt_2012b, @Whitehouse_2012, @Lesterhuis_2018, @vanDaal_et_al_2019, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023.
-Regarding practical applicability, several studies have demonstrated the method's versatility in both educational and non-educational settings, offering an efficient and effective alternative for measurement and evaluation.
-Notable examples include the works of @Jones_2015, @Bartholomew_et_al_2018, @Jones_et_al_2019, @Marshall_et_al_2020, @Bartholomew_et_al_2020, and @Boonen_et_al_2020.
+These studies have evaluated CJ from two main perspectives: its ability to produce reliable and valid trait scores, and its practical applicability.
+In terms of reliability and validity, research has shown that CJ can generate precise and consistent scores that accurately represent the traits being measured.
+Notable contributions in this research area include studies by @Pollitt_2012b, @Whitehouse_2012, @vanDaal_et_al_2016, @Lesterhuis_2018, @vanDaal_et_al_2019, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023.
+Regarding practical applicability, several studies have highlighted the method's versatility in both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation.
+Key examples in this research area include the works of @Jones_2015, @Bartholomew_et_al_2018, @Jones_et_al_2019, @Marshall_et_al_2020, @Bartholomew_et_al_2020, and @Boonen_et_al_2020.
 
@@
 
+Nevertheless, despite the growing number of CJ studies, the research approaches employed in the literature have been unsystematic and non-integrated, leading to the oversight of several critical issues related to the method.
+These issues include concerns about the measurement model responsible for generating the CJ scores, the structural component used for further data analysis and hypothesis testing, and challenges related to the design of CJ experiments.
+For instance, a notable concern regarding the measurement model is the prevalent reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
+Although Case 5 was originally articulated to produce a "rough measurement" or "rather coarse scaling" of traits [@Thurstone_1927, p.
+268-269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003
+
+<!-- a practice that may not be justifiable. -->
+
+<!-- One issue related to the structural component, is the disconnection existing in the literature between the measurement model and the structural component used for hypothesis testing,  -->
+
+<!-- and issues related to the method's experimental design. These issues can affect the method's reliability, validity, and practical applicability. -->
+
 <!-- 3. What is the solution? -->
@@
 
+## Let's talk about Thurstone co. {#sec-theory-thurstone}
+
 <!-- Notes: -->
@@
 <!-- c. what about the experimental design?, what are the implications of certain comparison algorithms? -->
+
+<!-- Use the SEM description of measurement model and structural model. -->
+
+<!-- The issues fall into three categories: those related to the measurement model, the structural model, and the experimental design. -->
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]. -->
+
+<!-- Measurement model -->
+
+<!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:  -->
+
+<!-- (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects.
+This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects.
+-->
+<!-- This is the case of confounding, and @Andrich_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->
+
+<!-- possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @Arlett_2003, @Guthrie_2003 and @Gijsen_et_al_2021 -->
+
+<!-- (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
+There is an amount of correlation that it is not accounted for.
+-->
+<!-- This is the case of clustering or measurement error, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.
+-->
+
+<!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->
+
+<!-- (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
+-->
+
+<!-- This ties up with the next subject -->
+
+<!-- Structural model -->
+
+<!-- Regarding this, it is apparent that in all literature there is a disconnection between measurement model and the structural part, where hypothesis and summaries are performed. -->
+
+<!-- The current literature estimates the CJ scores to later used them into other analysis, such as investigating bias and misfits among judges, or for the comparison of groups of individuals.
+This ignores measurement error.
+-->
+
+<!-- Experimental design -->
+
+<!-- In this regard, although the literature has tested multiple features of the experimental design of the method, such as the comparison algorithm and the number of comparisons, it is not clear where this procedures fit in the whole process of comparison, and what are their implications for the outcome of the method.
+-->
+
+<!-- present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required. Adhoc rules of thumb. -->
 

### 2024-10-16 _tex/index.tex
Built site for gh-pages

@@
 \citep{Thurstone_1927, Pollitt_2012a} for assessing competencies and
-traits. These studies have examined the method's strengths from two main
-perspectives: its capacity to produce reliable and valid trait scores,
-and its practical applicability. In terms of reliability and validity,
-research indicates that CJ generates precise and consistent scores that
-accurately represent the traits being measured. Key contributions in
-this area include those by \citet{Bramley_2008}, \citet{Pollitt_2012b},
-\citet{Whitehouse_2012}, \citet{Lesterhuis_2018},
-\citet{vanDaal_et_al_2019}, \citet{Bramley_et_al_2019},
-\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022}, and
-\citet{Bouwer_et_al_2023}. Regarding practical applicability, several
-studies have demonstrated the method's versatility in both educational
-and non-educational settings, offering an efficient and effective
-alternative for measurement and evaluation. Notable examples include the
-works of \citet{Jones_2015}, \citet{Bartholomew_et_al_2018},
+traits. These studies have evaluated CJ from two main perspectives: its
+ability to produce reliable and valid trait scores, and its practical
+applicability. In terms of reliability and validity, research has shown
+that CJ can generate precise and consistent scores that accurately
+represent the traits being measured. Notable contributions in this
+research area include studies by \citet{Pollitt_2012b},
+\citet{Whitehouse_2012}, \citet{vanDaal_et_al_2016},
+\citet{Lesterhuis_2018}, \citet{vanDaal_et_al_2019},
+\citet{Bramley_et_al_2019}, \citet{Verhavert_et_al_2019},
+\citet{Crompvoets_et_al_2022}, and \citet{Bouwer_et_al_2023}. Regarding
+practical applicability, several studies have highlighted the method's
+versatility in both educational and non-educational contexts, presenting
+it as an efficient and effective alternative for measurement and
+evaluation. Key examples in this research area include the works of
+\citet{Jones_2015}, \citet{Bartholomew_et_al_2018},
 \citet{Jones_et_al_2019}, \citet{Marshall_et_al_2020},
@@
 
+Nevertheless, despite the growing number of CJ studies, the research
+approaches employed in the literature have been unsystematic and
+non-integrated, leading to the oversight of several critical issues
+related to the method. These issues include concerns about the
+measurement model responsible for generating the CJ scores, the
+structural component used for further data analysis and hypothesis
+testing, and challenges related to the design of CJ experiments. For
+instance, a notable concern regarding the measurement model is the
+prevalent reliance on the assumptions of Case 5 from Thurstone's law of
+comparative judgment \citeyearpar{Thurstone_1927}. Although Case 5 was
+originally articulated to produce a ``rough measurement'' or ``rather
+coarse scaling'' of traits \citep[p.~268-269]{Thurstone_1927}, its
+assumptions have become predominant in the literature due to its
+implementation through the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959}. This leaves issues such as the
+presence of judge' biases hinted by \citet{Bramley_2008} and
+\citet{Kelly_et_al_2022}, and evidenced by \citet{Pollitt_et_al_2003},
+\citet{Arlett_2003}, and \citet{Guthrie_2003}
+
 \section{Theory}\label{sec-theory}
+
+\subsection{Let's talk about Thurstone co.}\label{sec-theory-thurstone}
 

### 2024-10-17 index.qmd
section:introduction; process:write

@@
 
-Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927; @Pollitt_2012a] for assessing competencies and traits.
+Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927] for assessing competencies and traits.
 These studies have evaluated CJ from two main perspectives: its ability to produce reliable and valid trait scores, and its practical applicability.
 In terms of reliability and validity, research has shown that CJ can generate precise and consistent scores that accurately represent the traits being measured.
-Notable contributions in this research area include studies by @Pollitt_2012b, @Whitehouse_2012, @vanDaal_et_al_2016, @Lesterhuis_2018, @vanDaal_et_al_2019, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023.
+Notable contributions in this research area include studies by Pollit [-@Pollitt_2012a; -@Pollitt_2012b], @Whitehouse_2012, @vanDaal_et_al_2016, @Lesterhuis_2018, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023.
 Regarding practical applicability, several studies have highlighted the method's versatility in both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation.
@@
 
-Nevertheless, despite the growing number of CJ studies, the research approaches employed in the literature have been unsystematic and non-integrated, leading to the oversight of several critical issues related to the method.
-These issues include concerns about the measurement model responsible for generating the CJ scores, the structural component used for further data analysis and hypothesis testing, and challenges related to the design of CJ experiments.
-For instance, a notable concern regarding the measurement model is the prevalent reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
-Although Case 5 was originally articulated to produce a "rough measurement" or "rather coarse scaling" of traits [@Thurstone_1927, p.
-268-269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003
+Despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning CJ.
+These issues can be categorized into concerns about the method's structural model, measurement model, and experimental design.
 
-<!-- a practice that may not be justifiable. -->
+A key issue in the first category is the apparent disconnect between the method's structural and measurement models.
+A common practice in CJ literature involves performing data analysis and hypothesis testing on scores previously estimated using the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+These studies use the scores or their transformations to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017], detect 'bias' in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], or test various hypothesis [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+However, since CJ scores are parameter estimates with inherent uncertainty, the statistical literature suggests that separating the analysis from this uncertainty may artificially inflate the precision and power of the results.
+This, in turn, could increase the risk of committing a type I error, where a null hypothesis is wrongly rejected [@McElreath_2020].
 
-<!-- One issue related to the structural component, is the disconnection existing in the literature between the measurement model and the structural component used for hypothesis testing,  -->
+<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
+As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
+-->
+
+<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
+
+<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
+269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
 
@@
 
-<!-- possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @Arlett_2003, @Guthrie_2003 and @Gijsen_et_al_2021 -->
+<!-- possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @Arlett_2003, @Guthrie_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
 

### 2024-10-17 _tex/index.tex
Built site for gh-pages

--- 2024-10-15 _tex/index.tex
@@
 Over the past decade, numerous studies have documented the effectiveness
-of the \emph{comparative judgment} (CJ) method
-\citep{Thurstone_1927, Pollitt_2012a} for assessing competencies and
-traits. These studies have examined the method's strengths from two main
-perspectives: its capacity to produce reliable and valid trait scores,
-and its practical applicability. In terms of reliability and validity,
-research indicates that CJ generates precise and consistent scores that
-accurately represent the traits being measured. Key contributions in
-this area include those by \citet{Bramley_2008}, \citet{Pollitt_2012b},
-\citet{Whitehouse_2012}, \citet{Lesterhuis_2018},
-\citet{vanDaal_et_al_2019}, \citet{Bramley_et_al_2019},
+of the \emph{comparative judgment} (CJ) method \citep{Thurstone_1927}
+for assessing competencies and traits. These studies have evaluated CJ
+from two main perspectives: its ability to produce reliable and valid
+trait scores, and its practical applicability. In terms of reliability
+and validity, research has shown that CJ can generate precise and
+consistent scores that accurately represent the traits being measured.
+Notable contributions in this research area include studies by Pollit
+\citetext{\citeyear{Pollitt_2012a}; \citeyear{Pollitt_2012b}},
+\citet{Whitehouse_2012}, \citet{vanDaal_et_al_2016},
+\citet{Lesterhuis_2018}, \citet{Bramley_et_al_2019},
 \citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022}, and
 \citet{Bouwer_et_al_2023}. Regarding practical applicability, several
-studies have demonstrated the method's versatility in both educational
-and non-educational settings, offering an efficient and effective
-alternative for measurement and evaluation. Notable examples include the
-works of \citet{Jones_2015}, \citet{Bartholomew_et_al_2018},
-\citet{Jones_et_al_2019}, \citet{Marshall_et_al_2020},
-\citet{Bartholomew_et_al_2020}, and \citet{Boonen_et_al_2020}.
+studies have highlighted the method's versatility in both educational
+and non-educational contexts, presenting it as an efficient and
+effective alternative for measurement and evaluation. Key examples in
+this research area include the works of \citet{Jones_2015},
+\citet{Bartholomew_et_al_2018}, \citet{Jones_et_al_2019},
+\citet{Marshall_et_al_2020}, \citet{Bartholomew_et_al_2020}, and
+\citet{Boonen_et_al_2020}.
+
+Despite the growing number of CJ studies, the unsystematic and
+fragmented research approaches employed in the literature have
+overlooked several critical issues concerning CJ. These issues can be
+categorized into concerns about the method's structural model,
+measurement model, and experimental design.
+
+A key issue in the first category is the apparent disconnect between the
+method's structural and measurement models. A common practice in CJ
+literature involves performing data analysis and hypothesis testing on
+scores previously estimated using the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959}. These studies use the scores or
+their transformations to identify `misfit' judges and stimuli
+\citep{Pollitt_2012b, vanDaal_et_al_2017}, detect `bias' in judges'
+ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, or test various
+hypothesis
+\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+However, since CJ scores are parameter estimates with inherent
+uncertainty, the statistical literature suggests that separating the
+analysis from this uncertainty may artificially inflate the precision
+and power of the results. This, in turn, could increase the risk of
+committing a type I error, where a null hypothesis is wrongly rejected
+\citep{McElreath_2020}.
 
 \section{Theory}\label{sec-theory}
+
+\subsection{Let's talk about Thurstone co.}\label{sec-theory-thurstone}
 

### 2024-10-18 index.qmd
section:introduction; process: write

@@
 These studies have evaluated CJ from two main perspectives: its ability to produce reliable and valid trait scores, and its practical applicability.
-In terms of reliability and validity, research has shown that CJ can generate precise and consistent scores that accurately represent the traits being measured.
-Notable contributions in this research area include studies by Pollit [-@Pollitt_2012a; -@Pollitt_2012b], @Whitehouse_2012, @vanDaal_et_al_2016, @Lesterhuis_2018, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023.
-Regarding practical applicability, several studies have highlighted the method's versatility in both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation.
-Key examples in this research area include the works of @Jones_2015, @Bartholomew_et_al_2018, @Jones_et_al_2019, @Marshall_et_al_2020, @Bartholomew_et_al_2020, and @Boonen_et_al_2020.
+Research on reliability and validity has shown that CJ can generate precise and consistent scores that accurately represent the traits being measured [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @Bouwer_et_al_2023].
+Regarding practical applicability, several studies have highlighted CJ's versatility across both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 
-Despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning CJ.
-These issues can be categorized into concerns about the method's structural model, measurement model, and experimental design.
+Despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning the method.
+These issues fall into three main categories: concerns about the structural model, the measurement model, and the experimental design of CJ.
+In the following sections, each issue will be discussed in detail, followed by the introduction of an approach that addresses all three concerns simultaneously.
 
 A key issue in the first category is the apparent disconnect between the method's structural and measurement models.
-A common practice in CJ literature involves performing data analysis and hypothesis testing on scores previously estimated using the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-These studies use the scores or their transformations to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017], detect 'bias' in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], or test various hypothesis [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, since CJ scores are parameter estimates with inherent uncertainty, the statistical literature suggests that separating the analysis from this uncertainty may artificially inflate the precision and power of the results.
+In CJ literature, it is common to perform data analysis and hypothesis testing on scores estimated using the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+Several studies use the scores generated by the BTL model or their transformations to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017], detect 'bias' in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], or test various hypotheses about the underlying trait being measured [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+However, since these scores are parameter estimates with inherent uncertainty, the statistical literature suggests that separating the analysis from this uncertainty may artificially inflate the precision and power of the results.
 This, in turn, could increase the risk of committing a type I error, where a null hypothesis is wrongly rejected [@McElreath_2020].
+To address this issue properly, the approach should follow a strategy similar to that used in Structural Equation Modeling (SEM), where data analysis and hypothesis testing occur at the structural model level, while the BTL model functions as the measurement model.
 
@@
 
-<!-- The issues fall into three categories: those related to the measurement model, the structural model, and the experimental design. -->
-
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]. -->
-
 <!-- Measurement model -->
@@
 <!-- possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @Arlett_2003, @Guthrie_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]. -->
 

### 2024-10-18 _tex/index.tex
Built site for gh-pages

@@
 from two main perspectives: its ability to produce reliable and valid
-trait scores, and its practical applicability. In terms of reliability
-and validity, research has shown that CJ can generate precise and
-consistent scores that accurately represent the traits being measured.
-Notable contributions in this research area include studies by Pollit
-\citetext{\citeyear{Pollitt_2012a}; \citeyear{Pollitt_2012b}},
-\citet{Whitehouse_2012}, \citet{vanDaal_et_al_2016},
-\citet{Lesterhuis_2018}, \citet{Bramley_et_al_2019},
-\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022}, and
-\citet{Bouwer_et_al_2023}. Regarding practical applicability, several
-studies have highlighted the method's versatility in both educational
-and non-educational contexts, presenting it as an efficient and
-effective alternative for measurement and evaluation. Key examples in
-this research area include the works of \citet{Jones_2015},
-\citet{Bartholomew_et_al_2018}, \citet{Jones_et_al_2019},
-\citet{Marshall_et_al_2020}, \citet{Bartholomew_et_al_2020}, and
-\citet{Boonen_et_al_2020}.
+trait scores, and its practical applicability. Research on reliability
+and validity has shown that CJ can generate precise and consistent
+scores that accurately represent the traits being measured
+\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, Bouwer_et_al_2023}.
+Regarding practical applicability, several studies have highlighted CJ's
+versatility across both educational and non-educational contexts,
+presenting it as an efficient and effective alternative for measurement
+and evaluation
+\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
 
@@
 fragmented research approaches employed in the literature have
-overlooked several critical issues concerning CJ. These issues can be
-categorized into concerns about the method's structural model,
-measurement model, and experimental design.
+overlooked several critical issues concerning the method. These issues
+fall into three main categories: concerns about the structural model,
+the measurement model, and the experimental design of CJ. In the
+following sections, each issue will be discussed in detail, followed by
+the introduction of an approach that addresses all three concerns
+simultaneously.
 
 A key issue in the first category is the apparent disconnect between the
-method's structural and measurement models. A common practice in CJ
-literature involves performing data analysis and hypothesis testing on
-scores previously estimated using the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959}. These studies use the scores or
-their transformations to identify `misfit' judges and stimuli
-\citep{Pollitt_2012b, vanDaal_et_al_2017}, detect `bias' in judges'
-ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, or test various
-hypothesis
+method's structural and measurement models. In CJ literature, it is
+common to perform data analysis and hypothesis testing on scores
+estimated using the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959}. Several studies use the scores
+generated by the BTL model or their transformations to identify `misfit'
+judges and stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017}, detect
+`bias' in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, or
+test various hypotheses about the underlying trait being measured
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-However, since CJ scores are parameter estimates with inherent
+However, since these scores are parameter estimates with inherent
 uncertainty, the statistical literature suggests that separating the
@@
 committing a type I error, where a null hypothesis is wrongly rejected
-\citep{McElreath_2020}.
+\citep{McElreath_2020}. To address this issue properly, the approach
+should follow a strategy similar to that used in Structural Equation
+Modeling (SEM), where data analysis and hypothesis testing occur at the
+structural model level, while the BTL model functions as the measurement
+model.
 

### 2024-10-18 index.qmd
section:introduction; process:write

@@
 These studies have evaluated CJ from two main perspectives: its ability to produce reliable and valid trait scores, and its practical applicability.
-Research on reliability and validity has shown that CJ can generate precise and consistent scores that accurately represent the traits being measured [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @Bouwer_et_al_2023].
+Research on reliability and validity has shown that CJ can generate precise and consistent scores [@Pollitt_2012a; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @Bouwer_et_al_2023] that accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018].
 Regarding practical applicability, several studies have highlighted CJ's versatility across both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
@@
 
-Despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning the method.
+Nevertheless, despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning the method.
 These issues fall into three main categories: concerns about the structural model, the measurement model, and the experimental design of CJ.
@@
 
-A key issue in the first category is the apparent disconnect between the method's structural and measurement models.
-In CJ literature, it is common to perform data analysis and hypothesis testing on scores estimated using the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-Several studies use the scores generated by the BTL model or their transformations to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017], detect 'bias' in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], or test various hypotheses about the underlying trait being measured [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, since these scores are parameter estimates with inherent uncertainty, the statistical literature suggests that separating the analysis from this uncertainty may artificially inflate the precision and power of the results.
-This, in turn, could increase the risk of committing a type I error, where a null hypothesis is wrongly rejected [@McElreath_2020].
-To address this issue properly, the approach should follow a strategy similar to that used in Structural Equation Modeling (SEM), where data analysis and hypothesis testing occur at the structural model level, while the BTL model functions as the measurement model.
-
-<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
-As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
--->
-
-<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
-
-<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
-269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
-
-<!-- and issues related to the method's experimental design. These issues can affect the method's reliability, validity, and practical applicability. -->
-
-<!-- 3. What is the solution? -->
-
-<!-- Notes:  -->
-<!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
-<!-- - This paper tries to systematically integrate all aspects at play in a single scientific model, we will build the model in a stepwise manner using -->
-<!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->
-
-# Theory {#sec-theory}
-<!-- 21:00 recording -->
-
-## Let's talk about Thurstone co. {#sec-theory-thurstone}
+# Let's talk about Thurstone co. {#sec-theory-thurstone}
 
@@
 <!-- Use the SEM description of measurement model and structural model. -->
+
+## The disconect between structural and measurement model
+
+<!-- Structural model -->
+
+<!-- Regarding this, it is apparent that in all literature there is a disconnection between measurement model and the structural part, where hypothesis and summaries are performed. -->
+
+<!-- The current literature estimates the CJ scores to later used them into other analysis, such as investigating bias and misfits among judges, or for the comparison of groups of individuals.
+This ignores measurement error.
+-->
+
+A key issue in the first category is the apparent disconnect between the method's structural and measurement models.
+In CJ literature, it is common to perform data analysis and hypothesis testing on scores estimated using the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+Several studies use the scores generated by the BTL model or their transformations to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017], detect 'bias' in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], or test various hypotheses about the underlying trait being measured [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+In the motivating example, ...
+
+However, since these scores are parameter estimates with inherent uncertainty, the statistical literature suggests that separating the analysis from this uncertainty may artificially inflate the precision and power of the results.
+This, in turn, could increase the risk of committing a type I error, where a null hypothesis is wrongly rejected [@McElreath_2020].
+In the motivating example
+
+To address this issue properly, the approach should follow a strategy similar to that used in Structural Equation Modeling (SEM), where data analysis and hypothesis testing occur at the structural model level, while the BTL model functions as the measurement model.
 
@@
 
-<!-- Structural model -->
+<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
+As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
+-->
 
-<!-- Regarding this, it is apparent that in all literature there is a disconnection between measurement model and the structural part, where hypothesis and summaries are performed. -->
+<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
 
-<!-- The current literature estimates the CJ scores to later used them into other analysis, such as investigating bias and misfits among judges, or for the comparison of groups of individuals.
-This ignores measurement error.
--->
+<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
+269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
 
@@
 <!-- present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required. Adhoc rules of thumb. -->
+
+<!-- and issues related to the method's experimental design. These issues can affect the method's reliability, validity, and practical applicability. -->
+
+<!-- 3. What is the solution? -->
+
+<!-- Notes:  -->
+<!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
+<!-- - This paper tries to systematically integrate all aspects at play in a single scientific model, we will build the model in a stepwise manner using -->
+<!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->
+
+# Theory {#sec-theory}
+<!-- 21:00 recording -->
 

### 2024-10-18 _tex/index.tex
Built site for gh-pages

@@
 and validity has shown that CJ can generate precise and consistent
-scores that accurately represent the traits being measured
-\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, Bouwer_et_al_2023}.
-Regarding practical applicability, several studies have highlighted CJ's
+scores
+\citep{Pollitt_2012a, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, Bouwer_et_al_2023}
+that accurately represent the traits being measured
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018}. Regarding
+practical applicability, several studies have highlighted CJ's
 versatility across both educational and non-educational contexts,
@@
 
-Despite the growing number of CJ studies, the unsystematic and
-fragmented research approaches employed in the literature have
+Nevertheless, despite the growing number of CJ studies, the unsystematic
+and fragmented research approaches employed in the literature have
 overlooked several critical issues concerning the method. These issues
@@
 simultaneously.
+
+\section{Let's talk about Thurstone co.}\label{sec-theory-thurstone}
+
+\subsection{The disconect between structural and measurement
+model}\label{the-disconect-between-structural-and-measurement-model}
 
@@
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+In the motivating example, \ldots{}
+
 However, since these scores are parameter estimates with inherent
@@
 committing a type I error, where a null hypothesis is wrongly rejected
-\citep{McElreath_2020}. To address this issue properly, the approach
-should follow a strategy similar to that used in Structural Equation
-Modeling (SEM), where data analysis and hypothesis testing occur at the
-structural model level, while the BTL model functions as the measurement
-model.
+\citep{McElreath_2020}. In the motivating example
+
+To address this issue properly, the approach should follow a strategy
+similar to that used in Structural Equation Modeling (SEM), where data
+analysis and hypothesis testing occur at the structural model level,
+while the BTL model functions as the measurement model.
 
 \section{Theory}\label{sec-theory}
-
-\subsection{Let's talk about Thurstone co.}\label{sec-theory-thurstone}
 

### 2024-10-22 index.qmd
section:issues; process:write

--- 2024-10-17 index.qmd
@@
 ---
-title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism"
+title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
 author:
@@
 
-<!-- recording -->
+<!-- recording 1 -->
 <!-- Part 1; 00:00:00 - 00:09:10 -->
-<!-- Part 2; 00:14:30 - 00:21:00 -->
+
+<!-- recording 2 -->
+<!-- Part 2; 00:00:00 - 00:17:30 -->
 
 <!-- 1. Where are we now with DCJ? -->
-
-<!-- Notes: -->
-<!-- - In the past decade there has been a growing number of studies that document and analyze the merits of DCJ as an assessment tool. Define a bit DCJ. -->
-<!-- - It has been a growing field, where people has tested some important aspects of the method.
-In general, is an iterative process.
-People are iterating on particular aspects of the method, such as: -->
-<!-- a. issues within the method as such, critical analyze some aspect of the method -->
-<!-- b. practical use of the method, generate scores with the BTL, and use this scores with some hypotheses, or descriptions. -->
-
 Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927] for assessing competencies and traits.
-These studies have evaluated CJ from two main perspectives: its ability to produce reliable and valid trait scores, and its practical applicability.
-In terms of reliability and validity, research has shown that CJ can generate precise and consistent scores that accurately represent the traits being measured.
-Notable contributions in this research area include studies by Pollit [-@Pollitt_2012a; -@Pollitt_2012b], @Whitehouse_2012, @vanDaal_et_al_2016, @Lesterhuis_2018, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023.
-Regarding practical applicability, several studies have highlighted the method's versatility in both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation.
-Key examples in this research area include the works of @Jones_2015, @Bartholomew_et_al_2018, @Jones_et_al_2019, @Marshall_et_al_2020, @Bartholomew_et_al_2020, and @Boonen_et_al_2020.
+These studies have evaluated CJ from three main perspectives: its ability to produce reliable and valid trait scores, its practical applicability, and its time efficiency.
+Research on reliability and validity shows that CJ can generate precise and consistent scores [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @Bouwer_et_al_2023] that accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bouwer_et_al_2023].
+Research on practical applicability highlights CJ's versatility across both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation [@Pollitt_2004; @Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Lastly, research on time efficiency suggests that CJ can offer at least equal, if not significant, time savings when evaluating stimuli compared to traditional marking methods [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018].
 
 <!-- 2. But what are the problems of DCJ as a method? -->
+Nevertheless, despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning the method.
+These issues fall into three main categories: the disconnect between CJ's structural and measurement model, the over-reliance on Thurstone's Case 5 [-@Thurstone_1927] in its measurement model, and the role of the comparison algorithms and the number of comparisons on the method's reliability and validity.
+In the following sections, each issue will be discussed in detail, followed by the introduction of a theoretical model and its translation into a statistical model that addresses all three concerns simultaneously.
 
-<!-- Notes:  -->
-<!-- - But what it is the problem with this way of using DCJ (just one paragraph) -->
+# Overlooked issues in CJ literature {#sec-theory-issues}
+
+## The disconnect between structural and measurement model
+
+<!-- 1. The BTL model is CJ's measurement model -->
+In the CJ literature, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ because it specifies how latent variables are estimated from observed ones [@Hoyle_et_al_2023; @Kline_et_al_2023].
+In a CJ study, multiple judges conduct several rounds of pairwise comparisons to evaluate the relative manifestation of a trait between a pair of stimuli.
+This evaluation results in a dichotomous outcome that indicates which stimulus is perceived to exhibit a higher degree of the trait.
+The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].
+
+<!-- 2. But there is no structural model, only post estimation analyses -->
+Moreover, in the CJ literature, it is common practice for BTL-generated scores, or their transformations, to undergo additional data analysis and hypothesis testing.
+Researchers have utilized these scores in independent analyses to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+
+<!-- 3. But what is the problem with this? -->
+<!-- However, the statistical literature cautions when using estimated scores for further independent data analysis and hypothesis testing.
+Since BTL-generated scores are parameter estimates that carry inherent uncertainty, separating the subsequent analyses from this uncertainty may lead to an artificial inflation of the precision and power of the analyses and tests.
+Consequently, this practice could increase the risk of committing a type I error, wherein a null hypothesis is incorrectly rejected [@McElreath_2020].
+An appropriate approach to address this issue, consistent with the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023], is that these analyses should be conducted at the structural level rather than as independent analyses.
+Consequently, an approach that integrates the structural and measurement model should be followed.
+-->
+
+<!-- However, there is often a lack of integration between the structural model, which captures the relationships among the variables, and the measurement model, which generates the CJ scores.
+This disconnect can lead to inaccurate inferences about the traits being measured.
+-->
+
+## Case 5 and the measurement model
+
+<!-- Measurement model -->
+
 <!-- a. measurement model is based on case V -->
@@
 <!--   a.2 It does not consider that stimuli may be nested in individuals -->
-<!-- b. in all literature there is a disconnection between measurement model and the structural part (common practical use) -->
-<!-- c. what about the experimental design?, what are the implications of certain comparison algorithms? -->
-
-Despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning CJ.
-These issues can be categorized into concerns about the method's structural model, measurement model, and experimental design.
-
-A key issue in the first category is the apparent disconnect between the method's structural and measurement models.
-A common practice in CJ literature involves performing data analysis and hypothesis testing on scores previously estimated using the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-These studies use the scores or their transformations to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017], detect 'bias' in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], or test various hypothesis [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, since CJ scores are parameter estimates with inherent uncertainty, the statistical literature suggests that separating the analysis from this uncertainty may artificially inflate the precision and power of the results.
-This, in turn, could increase the risk of committing a type I error, where a null hypothesis is wrongly rejected [@McElreath_2020].
-
-<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
-As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
--->
-
-<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
-
-<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
-269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
-
-<!-- and issues related to the method's experimental design. These issues can affect the method's reliability, validity, and practical applicability. -->
-
-<!-- 3. What is the solution? -->
-
-<!-- Notes:  -->
-<!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
-<!-- - This paper tries to systematically integrate all aspects at play in a single scientific model, we will build the model in a stepwise manner using -->
-<!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->
-
-# Theory {#sec-theory}
-<!-- 21:00 recording -->
-
-## Let's talk about Thurstone co. {#sec-theory-thurstone}
-
-<!-- Notes: -->
-<!-- - But what it is the problem with this way of using DCJ (just one paragraph) -->
-<!-- a. measurement model is based on case V -->
-<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
-<!--   a.2 It does not consider that stimuli may be nested in individuals -->
-<!-- b. in all literature there is a disconnection between measurement model and the structural part (common practical use) -->
-<!-- c. what about the experimental design?, what are the implications of certain comparison algorithms? -->
-
-<!-- Use the SEM description of measurement model and structural model. -->
-
-<!-- The issues fall into three categories: those related to the measurement model, the structural model, and the experimental design. -->
-
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]. -->
-
-<!-- Measurement model -->
 
@@
 -->
-<!-- This is the case of confounding, and @Andrich_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->
+<!-- This is the case of confounding, and @Andrich_1978; @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
+-->
 
 <!-- possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @Arlett_2003, @Guthrie_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
+
+<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind.
+This is studied by analysing the residuals that remain after the model is fitted to the data.
+Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large.
+Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability.
+indicates that one judge – Judge 5 – is significantly biased in some way.
+(Consider the problem of using adhoc methods) -->
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]. -->
 
@@
 
-<!-- Structural model -->
+<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
+As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
+-->
 
-<!-- Regarding this, it is apparent that in all literature there is a disconnection between measurement model and the structural part, where hypothesis and summaries are performed. -->
+<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
 
-<!-- The current literature estimates the CJ scores to later used them into other analysis, such as investigating bias and misfits among judges, or for the comparison of groups of individuals.
-This ignores measurement error.
--->
+<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
+269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
+
+<!-- All studies use SSR as a measure of reliability -->
+
+## The efficiency and reliability of comparison algorithms
 
@@
 
-<!-- present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required. Adhoc rules of thumb. -->
+<!-- @Bramley_et_al_2019 and @Bramley_2015 says adaptivity inflates reliability.
+present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required.
+Adhoc rules of thumb.
+-->
 
-## A scientific model for the CJ {#sec-theory-scientific}
+<!-- @Gray_et_al_2024 entropy-driven active learning pair-selection algorithm to compare stimuli, vs random and no repeating pairs -->
+
+<!-- @DeVrindt_et_al_2024 cold start problem to optimize time efficiency. -->
+
+<!-- @Mikhailiuk_et_al_2021 proposed a fully Bayesian active sampling strategy for pairwise comparisons via Approximate Message Passing and Information Gain Maximization.
+ASAP computes the full posterior distribution, which is crucial to achieving accurate EIG estimates, and thus the accuracy of active sampling.
+We recommend ASAP, as it offered the highest accuracy of inferred scores compared to existing methods in experiments with real and synthetic data.
+-->
+
+<!-- @Verhavert_et_al_2022 The current article proposes a new adaptive selection algorithm using a previously calibrated reference set.
+Using a reference set should eliminate the reliability inflation.
+It is proven that this adaptive selection algorithm is more efficient without reducing the accuracy of the results and without increasing the standard deviation of the assessment results.
+Akin to BIB designs and more specifically to BTIB designs  -->
+
+<!-- and issues related to the method's experimental design. These issues can affect the method's reliability, validity, and practical applicability. -->
+
+<!-- @Coertjens_et_al_2017 assess reliability and time investment.
+We conclude that pairwise comparison requires a similar time investment from evaluators as the criterion list method.
+We want to note that the outcomes in this study depend heavily on the design.
+First, the time taken for an assessment is strongly determined by the length of the criteria list used and the length of the texts (Breland, 1983).
+Not adaptive -->
+
+<!-- Although the process of pairwise comparisons is quite easy and fast for raters to apply (@Laming_2004; @Pollitt_2004), even without training, the increase in time investment that is needed for CJ decreases its feasibility in educational practice.
+-->
+
+<!-- Messick_1989 talks about efficiency and feasibility for application  -->
+
+<!-- @Bramley_2008 proposes ranking instead of comparisons -->
+
+<!-- talk about the number of comparisons @Verhavert_et_al_2019, @Crompvoets_et_al_2022, @Pollitt_2004 -->
+
+<!-- @Pollitt_2012b concludes that speed is not related to the quality of assessments in pairwise comparisons -->
+
+## What is the solution
+
+<!-- 3. What is the solution? -->
+
+<!-- Notes:  -->
+<!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
+<!-- - This paper tries to systematically integrate all aspects at play in a single scientific model, we will build the model in a stepwise manner using -->
+<!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->
+
+# Theory {#sec-theory}
+<!-- recording 1 -->
+<!-- Part 2; 00:14:30 - 00:21:00 -->
+
+## A theoretical model for the CJ {#sec-theory-scientific}
 
 ## From theory to statistical model {#sec-theory-statmodel}
+
+<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
 
@@
 ## Findings {#sec-discuss-finding}
+
+<!-- Pollitt (2012a) argues that assessors in pairwise comparisons need less training, given intuitive comparison and decision making.
+However, there is a distinct lack of research on the role of training.
+-->
 

### 2024-10-22 _tex/index.tex
Built site for gh-pages

--- 2024-10-17 _tex/index.tex
@@
 \hypersetup{
-  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism},
+  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
@@
 \begin{frontmatter}
-\title{Causes and effects in Dichotomous Comparative Judgments: an
-information-theoretical system of plausible mechanism}
+\title{Let's talk about Thurstone \& Co.: An information-theoretical
+model for comparative judgments, and its statistical translation}
 \author[1]{Jose Manuel Rivera Espejo%
@@
 for assessing competencies and traits. These studies have evaluated CJ
-from two main perspectives: its ability to produce reliable and valid
-trait scores, and its practical applicability. In terms of reliability
-and validity, research has shown that CJ can generate precise and
-consistent scores that accurately represent the traits being measured.
-Notable contributions in this research area include studies by Pollit
-\citetext{\citeyear{Pollitt_2012a}; \citeyear{Pollitt_2012b}},
-\citet{Whitehouse_2012}, \citet{vanDaal_et_al_2016},
-\citet{Lesterhuis_2018}, \citet{Bramley_et_al_2019},
-\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022}, and
-\citet{Bouwer_et_al_2023}. Regarding practical applicability, several
-studies have highlighted the method's versatility in both educational
-and non-educational contexts, presenting it as an efficient and
-effective alternative for measurement and evaluation. Key examples in
-this research area include the works of \citet{Jones_2015},
-\citet{Bartholomew_et_al_2018}, \citet{Jones_et_al_2019},
-\citet{Marshall_et_al_2020}, \citet{Bartholomew_et_al_2020}, and
-\citet{Boonen_et_al_2020}.
+from three main perspectives: its ability to produce reliable and valid
+trait scores, its practical applicability, and its time efficiency.
+Research on reliability and validity shows that CJ can generate precise
+and consistent scores
+\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018, Verhavert_et_al_2019, Crompvoets_et_al_2022, Bouwer_et_al_2023}
+that accurately represent the traits being measured
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bouwer_et_al_2023}.
+Research on practical applicability highlights CJ's versatility across
+both educational and non-educational contexts, presenting it as an
+efficient and effective alternative for measurement and evaluation
+\citep{Pollitt_2004, Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
+Lastly, research on time efficiency suggests that CJ can offer at least
+equal, if not significant, time savings when evaluating stimuli compared
+to traditional marking methods
+\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018}.
 
-Despite the growing number of CJ studies, the unsystematic and
-fragmented research approaches employed in the literature have
-overlooked several critical issues concerning CJ. These issues can be
-categorized into concerns about the method's structural model,
-measurement model, and experimental design.
+Nevertheless, despite the growing number of CJ studies, the unsystematic
+and fragmented research approaches employed in the literature have
+overlooked several critical issues concerning the method. These issues
+fall into three main categories: the disconnect between CJ's structural
+and measurement model, the over-reliance on Thurstone's Case 5
+\citeyearpar{Thurstone_1927} in its measurement model, and the role of
+the comparison algorithms and the number of comparisons on the method's
+reliability and validity. In the following sections, each issue will be
+discussed in detail, followed by the introduction of a theoretical model
+and its translation into a statistical model that addresses all three
+concerns simultaneously.
 
-A key issue in the first category is the apparent disconnect between the
-method's structural and measurement models. A common practice in CJ
-literature involves performing data analysis and hypothesis testing on
-scores previously estimated using the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959}. These studies use the scores or
-their transformations to identify `misfit' judges and stimuli
-\citep{Pollitt_2012b, vanDaal_et_al_2017}, detect `bias' in judges'
-ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, or test various
-hypothesis
+\section{Overlooked issues in CJ literature}\label{sec-theory-issues}
+
+\subsection{The disconnect between structural and measurement
+model}\label{the-disconnect-between-structural-and-measurement-model}
+
+In the CJ literature, the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
+for CJ because it specifies how latent variables are estimated from
+observed ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In a CJ study,
+multiple judges conduct several rounds of pairwise comparisons to
+evaluate the relative manifestation of a trait between a pair of
+stimuli. This evaluation results in a dichotomous outcome that indicates
+which stimulus is perceived to exhibit a higher degree of the trait. The
+BTL model then uses these observed outcomes to estimate scores that
+represent the latent trait of interest
+\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.
+
+Moreover, in the CJ literature, it is common practice for BTL-generated
+scores, or their transformations, to undergo additional data analysis
+and hypothesis testing. Researchers have utilized these scores in
+independent analyses to identify `misfit' judges and stimuli
+\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
+biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
+calculate correlations with other scoring methods
+\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
+related to the trait of interest
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-However, since CJ scores are parameter estimates with inherent
-uncertainty, the statistical literature suggests that separating the
-analysis from this uncertainty may artificially inflate the precision
-and power of the results. This, in turn, could increase the risk of
-committing a type I error, where a null hypothesis is wrongly rejected
-\citep{McElreath_2020}.
+
+\subsection{Case 5 and the measurement
+model}\label{case-5-and-the-measurement-model}
+
+\subsection{The efficiency and reliability of comparison
+algorithms}\label{the-efficiency-and-reliability-of-comparison-algorithms}
+
+\subsection{What is the solution}\label{what-is-the-solution}
 
@@
 
-\subsection{Let's talk about Thurstone co.}\label{sec-theory-thurstone}
-
-\subsection{A scientific model for the CJ}\label{sec-theory-scientific}
+\subsection{A theoretical model for the CJ}\label{sec-theory-scientific}
 

### 2024-10-23 _tex/index.tex
Built site for gh-pages

@@
 
-Nevertheless, despite the growing number of CJ studies, the unsystematic
-and fragmented research approaches employed in the literature have
-overlooked several critical issues concerning the method. These issues
-fall into three main categories: the disconnect between CJ's structural
-and measurement model, the over-reliance on Thurstone's Case 5
-\citeyearpar{Thurstone_1927} in its measurement model, and the role of
-the comparison algorithms and the number of comparisons on the method's
-reliability and validity. In the following sections, each issue will be
-discussed in detail, followed by the introduction of a theoretical model
-and its translation into a statistical model that addresses all three
-concerns simultaneously.
+Nevertheless, despite the growing number of studies on CJ, unsystematic
+and fragmented research approaches in the literature have left several
+critical issues unaddressed. These issues can be grouped into three main
+categories: the disconnect between CJ's structural and measurement
+model, the over-reliance on the assumptions of Thurstone's Case 5
+\citeyearpar{Thurstone_1927} in its measurement model, and the unclear
+role and impact of comparison algorithms and the number of comparisons
+on the method's reliability and validity. The following sections will
+discuss each of these issues in detail,, followed by the introduction of
+a theoretical model and its statistical translation, which aim to
+address all three concerns simultaneously.
 
-\section{Overlooked issues in CJ literature}\label{sec-theory-issues}
+\section{The three critical issues in CJ
+literature}\label{sec-theory-issues}
 
 \subsection{The disconnect between structural and measurement
-model}\label{the-disconnect-between-structural-and-measurement-model}
+models}\label{the-disconnect-between-structural-and-measurement-models}
 
-In the CJ literature, the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
-for CJ because it specifies how latent variables are estimated from
-observed ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In a CJ study,
-multiple judges conduct several rounds of pairwise comparisons to
-evaluate the relative manifestation of a trait between a pair of
-stimuli. This evaluation results in a dichotomous outcome that indicates
-which stimulus is perceived to exhibit a higher degree of the trait. The
-BTL model then uses these observed outcomes to estimate scores that
-represent the latent trait of interest
+In a typical CJ study, the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model.
+This model specifies how latent variables are estimated from observed
+ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In the context of CJ,
+multiple judges engage in several rounds of pairwise comparisons to
+assess the relative manifestation of a trait between two stimuli. Each
+comparison generates a dichotomous outcome, indicating which stimulus is
+perceived to exhibit a higher degree of the trait. The BTL model then
+uses these observed outcomes to estimate scores that represent the
+latent trait of interest
 \citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.
 
-Moreover, in the CJ literature, it is common practice for BTL-generated
-scores, or their transformations, to undergo additional data analysis
-and hypothesis testing. Researchers have utilized these scores in
-independent analyses to identify `misfit' judges and stimuli
+Moreover, researchers often use these BTL-generated scores or their
+transformations separately, applying additional analyses or hypothesis
+testing. The CJ literature shows that these scores have been employed to
+identify `misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
@@
 
-\subsection{Case 5 and the measurement
-model}\label{case-5-and-the-measurement-model}
+However, the statistical literature warns against using estimated scores
+for separate analyses and hypothesis testing. BTL-generated scores are
+parameter estimates that inherently carry uncertainty. Neglecting this
+uncertainty can inflate the precision and statistical power in
+independent analyses and tests. This oversight, in turn, may increase
+the risk of committing a type I error, where the null hypothesis is
+incorrectly rejected \citep{McElreath_2020}. To mitigate this risk,
+principles from Structural Equation Modeling (SEM)
+\citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting such
+analyses within the structural model of CJ rather than separately.
+Therefore, an integrated approach combining both structural and
+measurement models is advisable.
 
-\subsection{The efficiency and reliability of comparison
-algorithms}\label{the-efficiency-and-reliability-of-comparison-algorithms}
+\subsection{The assumptions of Case 5 and the measurement
+model}\label{the-assumptions-of-case-5-and-the-measurement-model}
 
-\subsection{What is the solution}\label{what-is-the-solution}
+\subsection{The role and impact of comparison
+algorithms}\label{the-role-and-impact-of-comparison-algorithms}
 
@@
 
-\subsection{A theoretical model for the CJ}\label{sec-theory-scientific}
+\subsection{A theoretical model for CJ}\label{sec-theory-theoretical}
 
-\subsection{From theory to statistical
-model}\label{sec-theory-statmodel}
+\subsection{From theory to statistics}\label{sec-theory-statistics}
 

### 2024-10-23 index.qmd
section:issues; process:write

@@
 ---
-title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
+title: "Let's talk about Thurstone & Co. "
+subtitle: "An information-theoretical model for comparative judgments, and its statistical translation"
 author:
@@
 <!-- 2. But what are the problems of DCJ as a method? -->
-Nevertheless, despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning the method.
-These issues fall into three main categories: the disconnect between CJ's structural and measurement model, the over-reliance on Thurstone's Case 5 [-@Thurstone_1927] in its measurement model, and the role of the comparison algorithms and the number of comparisons on the method's reliability and validity.
-In the following sections, each issue will be discussed in detail, followed by the introduction of a theoretical model and its translation into a statistical model that addresses all three concerns simultaneously.
+Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
+These issues can be grouped into three main categories: the disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in its measurement model, and the unclear role and impact of comparison algorithms and the number of comparisons on the method's reliability and validity.
+The following sections will discuss each of these issues in detail,, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.
 
-# Overlooked issues in CJ literature {#sec-theory-issues}
+# The three critical issues in CJ literature {#sec-theory-issues}
 
-## The disconnect between structural and measurement model
+## The disconnect between structural and measurement models
 
 <!-- 1. The BTL model is CJ's measurement model -->
-In the CJ literature, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ because it specifies how latent variables are estimated from observed ones [@Hoyle_et_al_2023; @Kline_et_al_2023].
-In a CJ study, multiple judges conduct several rounds of pairwise comparisons to evaluate the relative manifestation of a trait between a pair of stimuli.
-This evaluation results in a dichotomous outcome that indicates which stimulus is perceived to exhibit a higher degree of the trait.
+In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model.
+This model specifies how latent variables are estimated from observed ones [@Hoyle_et_al_2023; @Kline_et_al_2023].
+In the context of CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli.
+Each comparison generates a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher degree of the trait.
 The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].
@@
 <!-- 2. But there is no structural model, only post estimation analyses -->
-Moreover, in the CJ literature, it is common practice for BTL-generated scores, or their transformations, to undergo additional data analysis and hypothesis testing.
-Researchers have utilized these scores in independent analyses to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Moreover, researchers often use these BTL-generated scores or their transformations separately, applying additional analyses or hypothesis testing.
+The CJ literature shows that these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
-<!-- 3. But what is the problem with this? -->
-<!-- However, the statistical literature cautions when using estimated scores for further independent data analysis and hypothesis testing.
-Since BTL-generated scores are parameter estimates that carry inherent uncertainty, separating the subsequent analyses from this uncertainty may lead to an artificial inflation of the precision and power of the analyses and tests.
-Consequently, this practice could increase the risk of committing a type I error, wherein a null hypothesis is incorrectly rejected [@McElreath_2020].
-An appropriate approach to address this issue, consistent with the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023], is that these analyses should be conducted at the structural level rather than as independent analyses.
-Consequently, an approach that integrates the structural and measurement model should be followed.
--->
+<!-- 3. But what is the problem with this? and what is the solution? -->
+However, the statistical literature warns against using estimated scores for separate analyses and hypothesis testing.
+BTL-generated scores are parameter estimates that inherently carry uncertainty.
+Neglecting this uncertainty can inflate the precision and statistical power in independent analyses and tests.
+This oversight, in turn, may increase the risk of committing a type I error, where the null hypothesis is incorrectly rejected [@McElreath_2020].
+To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] recommend conducting such analyses within the structural model of CJ rather than separately.
+Therefore, an integrated approach combining both structural and measurement models is advisable.
 
-<!-- However, there is often a lack of integration between the structural model, which captures the relationships among the variables, and the measurement model, which generates the CJ scores.
-This disconnect can lead to inaccurate inferences about the traits being measured.
--->
-
-## Case 5 and the measurement model
+## The assumptions of Case 5 and the measurement model
 
@@
 
-## The efficiency and reliability of comparison algorithms
+## The role and impact of comparison algorithms
 
@@
 
-## What is the solution
+# Theory {#sec-theory}
+<!-- recording 1 -->
+<!-- Part 2; 00:14:30 - 00:21:00 -->
 
@@
 
-# Theory {#sec-theory}
-<!-- recording 1 -->
-<!-- Part 2; 00:14:30 - 00:21:00 -->
+## A theoretical model for CJ {#sec-theory-theoretical}
 
-## A theoretical model for the CJ {#sec-theory-scientific}
-
-## From theory to statistical model {#sec-theory-statmodel}
+## From theory to statistics {#sec-theory-statistics}
 

### 2024-10-23 _tex/index.tex
Built site for gh-pages

@@
 \hypersetup{
-  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
+  pdftitle={Let's talk about Thurstone \& Co.~},
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
@@
 \begin{frontmatter}
-\title{Let's talk about Thurstone \& Co.: An information-theoretical
-model for comparative judgments, and its statistical translation}
+\title{Let's talk about Thurstone \& Co.~ \\\large{An
+information-theoretical model for comparative judgments, and its
+statistical translation} }
 \author[1]{Jose Manuel Rivera Espejo%

### 2024-10-23 index.qmd
section:issues; process:write

@@
 ---
-title: "Let's talk about Thurstone & Co. "
-subtitle: "An information-theoretical model for comparative judgments, and its statistical translation"
+title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
 author:

### 2024-10-24 index.qmd
section:issues; process:write

@@
 Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
-These issues can be grouped into three main categories: the disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in its measurement model, and the unclear role and impact of comparison algorithms and the number of comparisons on the method's reliability and validity.
-The following sections will discuss each of these issues in detail,, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.
+This research primarily focuses on three: the apparent disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role and impact of comparison algorithms and the number of comparisons on the method's reliability and validity.
+The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.
 
@@
 In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model.
-This model specifies how latent variables are estimated from observed ones [@Hoyle_et_al_2023; @Kline_et_al_2023].
-In the context of CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli.
+This model specifies how latent variables are estimated from observed variables [@Everitt_et_al_2010].
+In CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli.
 Each comparison generates a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher degree of the trait.
@@
 <!-- 2. But there is no structural model, only post estimation analyses -->
-Moreover, researchers often use these BTL-generated scores or their transformations separately, applying additional analyses or hypothesis testing.
-The CJ literature shows that these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing.
+The CJ literature highlights how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
 <!-- 3. But what is the problem with this? and what is the solution? -->
-However, the statistical literature warns against using estimated scores for separate analyses and hypothesis testing.
-BTL-generated scores are parameter estimates that inherently carry uncertainty.
-Neglecting this uncertainty can inflate the precision and statistical power in independent analyses and tests.
-This oversight, in turn, may increase the risk of committing a type I error, where the null hypothesis is incorrectly rejected [@McElreath_2020].
-To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] recommend conducting such analyses within the structural model of CJ rather than separately.
-Therefore, an integrated approach combining both structural and measurement models is advisable.
+However, the statistical literature cautions against using estimated scores for separate analyses and hypothesis testing.
+A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty.
+Ignoring this uncertainty when conducting separate analyses and test can inflate their precision and statistical power.
+This, in turn, increases the risk of committing a type I error, which is when a null hypothesis is incorrectly rejected [@McElreath_2020].
+To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] recommend conducting these analyses within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately.
+Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.
 

### 2024-10-24 _tex/index.tex
Built site for gh-pages

--- 2024-10-23 _tex/index.tex
@@
 and fragmented research approaches in the literature have left several
-critical issues unaddressed. These issues can be grouped into three main
-categories: the disconnect between CJ's structural and measurement
-model, the over-reliance on the assumptions of Thurstone's Case 5
-\citeyearpar{Thurstone_1927} in its measurement model, and the unclear
+critical issues unaddressed. This research primarily focuses on three:
+the apparent disconnect between CJ's structural and measurement model,
+the over-reliance on the assumptions of Thurstone's Case 5
+\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
 role and impact of comparison algorithms and the number of comparisons
 on the method's reliability and validity. The following sections will
-discuss each of these issues in detail,, followed by the introduction of
+discuss each of these issues in detail, followed by the introduction of
 a theoretical model and its statistical translation, which aim to
@@
 This model specifies how latent variables are estimated from observed
-ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In the context of CJ,
-multiple judges engage in several rounds of pairwise comparisons to
-assess the relative manifestation of a trait between two stimuli. Each
-comparison generates a dichotomous outcome, indicating which stimulus is
-perceived to exhibit a higher degree of the trait. The BTL model then
-uses these observed outcomes to estimate scores that represent the
-latent trait of interest
+variables \citep{Everitt_et_al_2010}. In CJ, multiple judges engage in
+several rounds of pairwise comparisons to assess the relative
+manifestation of a trait between two stimuli. Each comparison generates
+a dichotomous outcome, indicating which stimulus is perceived to exhibit
+a higher degree of the trait. The BTL model then uses these observed
+outcomes to estimate scores that represent the latent trait of interest
 \citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.
@@
 Moreover, researchers often use these BTL-generated scores or their
-transformations separately, applying additional analyses or hypothesis
-testing. The CJ literature shows that these scores have been employed to
-identify `misfit' judges and stimuli
+transformations separately, conducting further analyses or hypothesis
+testing. The CJ literature highlights how these scores have been
+employed to identify `misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
@@
 
-However, the statistical literature warns against using estimated scores
-for separate analyses and hypothesis testing. BTL-generated scores are
-parameter estimates that inherently carry uncertainty. Neglecting this
-uncertainty can inflate the precision and statistical power in
-independent analyses and tests. This oversight, in turn, may increase
-the risk of committing a type I error, where the null hypothesis is
-incorrectly rejected \citep{McElreath_2020}. To mitigate this risk,
-principles from Structural Equation Modeling (SEM)
-\citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting such
-analyses within the structural model of CJ rather than separately.
-Therefore, an integrated approach combining both structural and
-measurement models is advisable.
+However, the statistical literature cautions against using estimated
+scores for separate analyses and hypothesis testing. A key consideration
+is that BTL-generated scores are parameter estimates that inherently
+carry uncertainty. Ignoring this uncertainty when conducting separate
+analyses and test can inflate their precision and statistical power.
+This, in turn, increases the risk of committing a type I error, which is
+when a null hypothesis is incorrectly rejected \citep{McElreath_2020}.
+To mitigate this risk, principles from Structural Equation Modeling
+(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting
+these analyses within a structural model that accounts for both the
+scores and their uncertainties, rather than treating them separately.
+Thus, an integrated approach combining CJ's structural and measurement
+models can offer significant advantages.
 

### 2024-10-24 index.qmd
section:issues; process:write

@@
 Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
-This research primarily focuses on three: the apparent disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role and impact of comparison algorithms and the number of comparisons on the method's reliability and validity.
+This research primarily focuses on three: the apparent disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role and impact of comparison algorithms on the method's reliability and validity.
 The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.
 
-# The three critical issues in CJ literature {#sec-theory-issues}
+# Three critical issues in CJ literature {#sec-theory-issues}
 
@@
 <!-- 1. The BTL model is CJ's measurement model -->
-In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model.
-This model specifies how latent variables are estimated from observed variables [@Everitt_et_al_2010].
+In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model [@Andrich_1978; @Bramley_2008].
+This model specifies how latent variables are estimated from manifest variables [@Everitt_et_al_2010].
 In CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli.
@@
 Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing.
-The CJ literature highlights how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+The CJ literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
 <!-- 3. But what is the problem with this? and what is the solution? -->
-However, the statistical literature cautions against using estimated scores for separate analyses and hypothesis testing.
+However, the statistical literature cautions against using estimated scores for separate analyses and tests.
 A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty when conducting separate analyses and test can inflate their precision and statistical power.
-This, in turn, increases the risk of committing a type I error, which is when a null hypothesis is incorrectly rejected [@McElreath_2020].
-To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] recommend conducting these analyses within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately.
+Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power.
+This, in turn, increases the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
+To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately.
 Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.
@@
 
+<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
+As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
+-->
+
+<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
+
+<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
+269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
+
 <!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:  -->
@@
 -->
+
 <!-- This is the case of confounding, and @Andrich_1978; @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
@@
 <!-- This ties up with the next subject -->
-
-<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
-As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
--->
-
-<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
-
-<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
-269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
 

### 2024-10-24 _tex/index.tex
Built site for gh-pages

@@
 \citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
-role and impact of comparison algorithms and the number of comparisons
-on the method's reliability and validity. The following sections will
-discuss each of these issues in detail, followed by the introduction of
-a theoretical model and its statistical translation, which aim to
-address all three concerns simultaneously.
+role and impact of comparison algorithms on the method's reliability and
+validity. The following sections will discuss each of these issues in
+detail, followed by the introduction of a theoretical model and its
+statistical translation, which aim to address all three concerns
+simultaneously.
 
-\section{The three critical issues in CJ
+\section{Three critical issues in CJ
 literature}\label{sec-theory-issues}
@@
 In a typical CJ study, the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model.
-This model specifies how latent variables are estimated from observed
-variables \citep{Everitt_et_al_2010}. In CJ, multiple judges engage in
-several rounds of pairwise comparisons to assess the relative
-manifestation of a trait between two stimuli. Each comparison generates
-a dichotomous outcome, indicating which stimulus is perceived to exhibit
-a higher degree of the trait. The BTL model then uses these observed
-outcomes to estimate scores that represent the latent trait of interest
+\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
+\citep{Andrich_1978, Bramley_2008}. This model specifies how latent
+variables are estimated from manifest variables
+\citep{Everitt_et_al_2010}. In CJ, multiple judges engage in several
+rounds of pairwise comparisons to assess the relative manifestation of a
+trait between two stimuli. Each comparison generates a dichotomous
+outcome, indicating which stimulus is perceived to exhibit a higher
+degree of the trait. The BTL model then uses these observed outcomes to
+estimate scores that represent the latent trait of interest
 \citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.
@@
 transformations separately, conducting further analyses or hypothesis
-testing. The CJ literature highlights how these scores have been
-employed to identify `misfit' judges and stimuli
+testing. The CJ literature shows how these scores have been employed to
+identify `misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
@@
 However, the statistical literature cautions against using estimated
-scores for separate analyses and hypothesis testing. A key consideration
-is that BTL-generated scores are parameter estimates that inherently
-carry uncertainty. Ignoring this uncertainty when conducting separate
-analyses and test can inflate their precision and statistical power.
-This, in turn, increases the risk of committing a type I error, which is
-when a null hypothesis is incorrectly rejected \citep{McElreath_2020}.
-To mitigate this risk, principles from Structural Equation Modeling
-(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting
-these analyses within a structural model that accounts for both the
-scores and their uncertainties, rather than treating them separately.
-Thus, an integrated approach combining CJ's structural and measurement
-models can offer significant advantages.
+scores for separate analyses and tests. A key consideration is that
+BTL-generated scores are parameter estimates that inherently carry
+uncertainty. Ignoring this uncertainty when conducting separate analyses
+and tests can inflate their precision and statistical power. This, in
+turn, increases the risk of committing a type I error
+\citep{McElreath_2020}, which is when a null hypothesis is incorrectly
+rejected \citep{Everitt_et_al_2010}. To mitigate this risk, principles
+from Structural Equation Modeling (SEM)
+\citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response Theory
+(IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
+conducting these analyses and tests within a structural model that
+accounts for both the scores and their uncertainties, rather than
+treating them separately. Thus, an integrated approach combining CJ's
+structural and measurement models can offer significant advantages.
 

### 2024-10-25 index.qmd
section:issues,measurement; process:write

@@
 Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing.
-The CJ literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+The literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
@@
 A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power.
-This, in turn, increases the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
+Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
 To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately.
@@
 
-<!-- Measurement model -->
+1. the BTL model is the statistical articulation of Case 5
 
-<!-- a. measurement model is based on case V -->
-<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
-<!--   a.2 It does not consider that stimuli may be nested in individuals -->
+  From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008
 
-<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927].
-As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model.
--->
+2. But this model hold the greater number of assumptions
 
-<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->
+  Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
+269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions.
 
-<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
-269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->
+  @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:
 
-<!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:  -->
-
-<!-- (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects.
+  (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects.
 This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects.
--->
 
@@
 <!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind.
-This is studied by analysing the residuals that remain after the model is fitted to the data.
+This is studied by analyzing the residuals that remain after the model is fitted to the data.
 Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large.
@@
 
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]. -->
+  (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
+Meaning there is an amount of correlation that it is not accounted for.
 
-<!-- (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
-There is an amount of correlation that it is not accounted for.
--->
 <!-- This is the case of clustering or measurement error, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.
@@
 
-<!-- (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
--->
+  (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
+
+3. What problems does this bring?
+
+<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
+<!--   a.2 It does not consider that stimuli may be nested in individuals -->
+
+  This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003
+
+  [Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}.
+
+<!-- All studies use SSR as a measure of reliability -->
 
 <!-- This ties up with the next subject -->
-
-<!-- All studies use SSR as a measure of reliability -->
 

### 2024-10-25 _tex/index.tex
Built site for gh-pages

@@
 transformations separately, conducting further analyses or hypothesis
-testing. The CJ literature shows how these scores have been employed to
+testing. The literature shows how these scores have been employed to
 identify `misfit' judges and stimuli
@@
 uncertainty. Ignoring this uncertainty when conducting separate analyses
-and tests can inflate their precision and statistical power. This, in
-turn, increases the risk of committing a type I error
-\citep{McElreath_2020}, which is when a null hypothesis is incorrectly
-rejected \citep{Everitt_et_al_2010}. To mitigate this risk, principles
-from Structural Equation Modeling (SEM)
+and tests can inflate their precision and statistical power, increasing
+the risk of committing a type I error \citep{McElreath_2020}, which is
+when a null hypothesis is incorrectly rejected
+\citep{Everitt_et_al_2010}. To mitigate this risk, principles from
+Structural Equation Modeling (SEM)
 \citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response Theory
@@
 model}\label{the-assumptions-of-case-5-and-the-measurement-model}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  the BTL model is the statistical articulation of Case 5
+\end{enumerate}
+
+From early on in the literature, it has been clear that the BTL model
+represents a statistical articulation of Thurstone's Case 5
+\citeyearpar{Thurstone_1927}. Talk about \citet{Pollitt_et_al_2003} and
+\citet{Bramley_2008}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  But this model hold the greater number of assumptions
+\end{enumerate}
+
+Although Case 5 was originally articulated to produce a ``rather coarse
+scaling'' of traits \citep[p.~269]{Thurstone_1927}, its assumptions have
+become predominant in the literature, but Thurstone's Case 5 is the case
+that holds the greater amount of simplifying assumptions.
+
+\citet{Thurstone_1927} justify the use of case V, on multiple
+assumptions, but the most important for our purpose are three:
+
+\begin{enumerate}
+\def\labelenumi{(\alph{enumi})}
+\tightlist
+\item
+  related to Case 3, it assumes the correlation between stimuli is zero,
+  this translates into the cancellation of judges effects by mean of
+  opposing and equally weigthed `mood' and `simultaneous contrasts'
+  effects. This is demonstrated using the additivive nature of the logit
+  scale, which helps to cancel `bias' judges effects.
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{(\alph{enumi})}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  related to Case 3, it assumes the correlation between stimuli is zero,
+  this also translates into the idea that stimuli are the main focus of
+  estimation and analysis, but what happens when the focus of analysis
+  is the individuals that generated those stimuli. Meaning there is an
+  amount of correlation that it is not accounted for.
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{(\alph{enumi})}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  related to case 5, discriminal dispersions of the stimuli are equal,
+  but it is not hard to imagine that certain individuals can produce
+  good quality texts with more precision than other individuals.
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  What problems does this bring?
+\end{enumerate}
+
+This leaves issues such as the presence of judge' biases hinted by
+\citet{Bramley_2008} and \citet{Kelly_et_al_2022}, and evidenced by
+\citet{Pollitt_et_al_2003}, \citet{Arlett_2003}, and
+\citet{Guthrie_2003}
+
+{Reliability is a necessary but not sufficient condition for validity.
+Reliability can exist without validity but validity cannot exist without
+reliability \citep{Perron_et_al_2015}}.
 

### 2024-10-31 index.qmd
section:introduction,issues,structural; process:write

@@
 
-<!-- 1. Where are we now with DCJ? -->
-Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927] for assessing competencies and traits.
-These studies have evaluated CJ from three main perspectives: its ability to produce reliable and valid trait scores, its practical applicability, and its time efficiency.
-Research on reliability and validity shows that CJ can generate precise and consistent scores [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @Bouwer_et_al_2023] that accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bouwer_et_al_2023].
-Research on practical applicability highlights CJ's versatility across both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation [@Pollitt_2004; @Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
-Lastly, research on time efficiency suggests that CJ can offer at least equal, if not significant, time savings when evaluating stimuli compared to traditional marking methods [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018].
+<!-- []{style="color:red;"}. -->
 
-<!-- 2. But what are the problems of DCJ as a method? -->
+<!-- 1. What does CJ do? -->
+[In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
+The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher level of the trait.
+After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest.
+CJ has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}
+
+<!-- Specifically, CJ has been used to evaluate the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023] and to assess the quality of speech [@Boonen_et_al_2020].
+Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_2015], engineering design skills [@Bartholomew_et_al_2018], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowdlege [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020].
+-->
+
+<!-- 2. Where are we now with DCJ? -->
+Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade.
+These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability.
+Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
+On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
+Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+
+<!-- rubrics [@Coertjens_et_al_2017; @Goossens_et_al_2018]  -->
+<!-- holistic benchmark ratings [@Bouwer_et_al_2023] -->
+
+<!-- Adaptive comparison algorithms dynamically present stimuli to judges, based on the results of previous comparisons.
+They aim to enhance the informativeness of comparisons while optimizing judges' time [@Bramley_2015].
+-->
+
+<!-- 3. But what are the problems of DCJ as a method? -->
 Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
-This research primarily focuses on three: the apparent disconnect between CJ's structural and measurement model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role and impact of comparison algorithms on the method's reliability and validity.
+This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity.
 The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.
@@
 <!-- 1. The BTL model is CJ's measurement model -->
-In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model [@Andrich_1978; @Bramley_2008].
-This model specifies how latent variables are estimated from manifest variables [@Everitt_et_al_2010].
-In CJ, multiple judges engage in several rounds of pairwise comparisons to assess the relative manifestation of a trait between two stimuli.
-Each comparison generates a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher degree of the trait.
-The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].
+In a typical CJ study, the BTL model serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008].
+A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
+For example, when evaluating text quality, the BTL model uses the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
 <!-- 2. But there is no structural model, only post estimation analyses -->
-Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing.
-The literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing.
+The literature indicates that these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the latent trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
-<!-- 3. But what is the problem with this? and what is the solution? -->
-However, the statistical literature cautions against using estimated scores for separate analyses and tests.
-A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
-To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately.
+<!-- 3. But what is the problem with this? -->
+However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests.
+A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
+Ignoring this uncertainty when conducting additional analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
+
+<!-- 4. So, what is the solution?  -->
+To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately.
 Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.
@@
 
-1. the BTL model is the statistical articulation of Case 5
+<!-- 1. the BTL model is the statistical articulation of Case 5 -->
 
-  From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008
-
-2. But this model hold the greater number of assumptions
-
-  Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
-269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions.
-
-  @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:
-
-  (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects.
-This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects.
-
-<!-- This is the case of confounding, and @Andrich_1978; @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
+<!--   From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927].
+Talk about @Pollitt_et_al_2003 and @Bramley_2008.
 -->
 
-<!-- possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @Arlett_2003, @Guthrie_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
+<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->
+
+<!--   Table with assumptions -->
+
+<!-- 2. But this model hold the greater number of assumptions -->
+
+<!--   Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
+269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions.
+-->
+
+<!--   @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three: -->
+
+<!--   (a) related to Case 2 and case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects.
+This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.
+-->
+
+<!--   Particularly on how CJ is used now, use the text example -->
+
+<!--   What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
+This, in the best scenario, cause less precise estimates, and in the worst scenario, biased estimates.
+possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
 
@@
 
-  (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
+<!--   Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
+-->
+
+<!--   use the example of writing!!! -->
+
+<!--   (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
 Meaning there is an amount of correlation that it is not accounted for.
+-->
+
+<!--   Here we loose precision, because we are not controlling for clustering -->
 
@@
 
-  (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
+<!--   (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
+Example of exaggerated comparison of individuals, university students vs secondary school students.
+-->
 
-3. What problems does this bring?
+<!--   here we face misspecification in the model -->
 
-<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
-<!--   a.2 It does not consider that stimuli may be nested in individuals -->
+<!-- If your scores are not reliable they cannot be valid.
+People have been employing a model with assumptions that they do not know if those assumptions apply for their cases.
+And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid.
+-->
 
-  This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003
-
-  [Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}.
-
-<!-- All studies use SSR as a measure of reliability -->
-
-<!-- This ties up with the next subject -->
+<!--   [Reliability is a necessary but not sufficient condition for validity.
+Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}.
+-->
 
@@
 Adhoc rules of thumb.
+-->
+
+<!-- @Bramley_2015 -->
+<!-- The implication is that the SSR statistic is worthless as an indicator of the quality / consistency / reliability of a scale constructed from an ACJ study that has involved a significant number of adaptive rounds.
+With random or fixed rounds, the SSR values were all below 0.25, which would generally be taken as an indicator of failure to create a meaningful scale.
+To put it another way, a low value of SSR almost certainly indicates low reliability, but a high value of SSR does not necessarily indicate high reliability.
+-->
+<!-- This is not to say that the ACJ studies listed in Table 1 (or indeed other studies not listed) were themselves worthless, just that not much should be made of claims about very high reliability in ACJ studies on the basis of high values of the SSR statistic.
+In fact, many of those studies also showed moderate to high correlations with variables external to the analysis, which were rightly interpreted as evidence of concurrent validity.
+-->
+<!-- The conclusion is therefore that the SSR statistic is at best misleading and at worst worthless as an indicator of scale reliability whenever a CJ study has involved a significant amount of adaptivity.
 -->
@@
 
+<!-- All studies use SSR as a measure of reliability -->
+
 # Theory {#sec-theory}
@@
 However, there is a distinct lack of research on the role of training.
+-->
+
+<!-- Due to efficiency of scoring, most of research only colects only one text.
+This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
+However, this cannot be known ad-hoc.
 -->
@@
 
-**Authors' contributions:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
+**AI-assisted technologies in the writing process:** The authors used ChatGPT, an AI language model, during the preparation of this work.
+They occasionally employed the tool to refine phrasing and optimize wording, ensuring appropriate language use and enhancing the manuscript's clarity and coherence.
+The authors take full responsibility for the final content of the publication.
+
+**CRediT authorship contribution statement:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
 and S.DM.; *Project administration:* S.G.

### 2024-10-31 _tex/index.tex
Built site for gh-pages

--- 2024-10-24 _tex/index.tex
@@
 
-Over the past decade, numerous studies have documented the effectiveness
-of the \emph{comparative judgment} (CJ) method \citep{Thurstone_1927}
-for assessing competencies and traits. These studies have evaluated CJ
-from three main perspectives: its ability to produce reliable and valid
-trait scores, its practical applicability, and its time efficiency.
-Research on reliability and validity shows that CJ can generate precise
-and consistent scores
-\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018, Verhavert_et_al_2019, Crompvoets_et_al_2022, Bouwer_et_al_2023}
-that accurately represent the traits being measured
-\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bouwer_et_al_2023}.
-Research on practical applicability highlights CJ's versatility across
-both educational and non-educational contexts, presenting it as an
-efficient and effective alternative for measurement and evaluation
-\citep{Pollitt_2004, Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
-Lastly, research on time efficiency suggests that CJ can offer at least
-equal, if not significant, time savings when evaluating stimuli compared
-to traditional marking methods
-\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018}.
+{In \emph{comparative judgment} (CJ) studies, judges assess the presence
+of a trait or competence by conducting pairwise comparisons of stimuli
+\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
+produces a dichotomous outcome, indicating which stimulus is perceived
+to possess a higher level of the trait. After conducting multiple rounds
+of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL)
+model \citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
+estimate scores that reflect the underlying trait of interest. CJ has
+been successfully employed in assessing the quality of written texts
+\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.}
+
+Numerous studies have documented the effectiveness of CJ in assessing
+various traits and competencies over the past decade. These studies have
+emphasized three aspects of the method's effectiveness: its reliability,
+validity, and practical applicability. Research on reliability indicates
+that CJ requires a relatively small number of pairwise comparisons
+\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
+scores that are as precise and consistent as those generated by other
+assessment methods
+\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
+Furthermore, evidence suggests that the reliability and time efficiency
+of CJ are comparable, if not superior, to those of other assessment
+methods when employing adaptive comparison algorithms
+\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
+the other hand, research on validity suggests that scores generated by
+CJ can accurately represent the traits being measured
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
+Finally, research on practical applicability highlights the method's
+versatility across both educational and non-educational contexts
+\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
 
@@
 critical issues unaddressed. This research primarily focuses on three:
-the apparent disconnect between CJ's structural and measurement model,
+the apparent disconnect between CJ's measurement and structural model,
 the over-reliance on the assumptions of Thurstone's Case 5
 \citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
-role and impact of comparison algorithms on the method's reliability and
-validity. The following sections will discuss each of these issues in
-detail, followed by the introduction of a theoretical model and its
-statistical translation, which aim to address all three concerns
-simultaneously.
+role of comparison algorithms on the method's reliability and validity.
+The following sections will discuss each of these issues in detail,
+followed by the introduction of a theoretical model and its statistical
+translation, which aim to address all three concerns simultaneously.
 
@@
 
-In a typical CJ study, the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
-\citep{Andrich_1978, Bramley_2008}. This model specifies how latent
-variables are estimated from manifest variables
-\citep{Everitt_et_al_2010}. In CJ, multiple judges engage in several
-rounds of pairwise comparisons to assess the relative manifestation of a
-trait between two stimuli. Each comparison generates a dichotomous
-outcome, indicating which stimulus is perceived to exhibit a higher
-degree of the trait. The BTL model then uses these observed outcomes to
-estimate scores that represent the latent trait of interest
-\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.
+In a typical CJ study, the BTL model serves as the measurement model for
+CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
+manifest variables contribute to the estimation of latent variables
+\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
+the BTL model uses the dichotomous outcomes resulting from the pairwise
+comparisons (the manifest variables) to estimate scores that reflect the
+underlying quality level of the texts (the latent variable)
+\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
 
-Moreover, researchers often use these BTL-generated scores or their
-transformations separately, conducting further analyses or hypothesis
-testing. The CJ literature shows how these scores have been employed to
-identify `misfit' judges and stimuli
+Researchers then typically use the estimated BTL scores, or their
+transformations, to conduct additional analyses or hypothesis testing.
+The literature indicates that these scores have been used to identify
+`misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
@@
 \citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
-related to the trait of interest
+related to the latent trait of interest
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 However, the statistical literature cautions against using estimated
-scores for separate analyses and tests. A key consideration is that
-BTL-generated scores are parameter estimates that inherently carry
-uncertainty. Ignoring this uncertainty when conducting separate analyses
-and tests can inflate their precision and statistical power. This, in
-turn, increases the risk of committing a type I error
-\citep{McElreath_2020}, which is when a null hypothesis is incorrectly
-rejected \citep{Everitt_et_al_2010}. To mitigate this risk, principles
-from Structural Equation Modeling (SEM)
-\citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response Theory
-(IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
-conducting these analyses and tests within a structural model that
-accounts for both the scores and their uncertainties, rather than
+scores to conduct additional analyses or tests. A key consideration is
+that BTL scores are parameter estimates that inherently carry
+uncertainty. Ignoring this uncertainty when conducting additional
+analyses and tests can inflate their precision and statistical power,
+increasing the risk of committing a type I error \citep{McElreath_2020},
+which is when a null hypothesis is incorrectly rejected
+\citep{Everitt_et_al_2010}.
+
+To mitigate these risks, principles from Structural Equation Modeling
+(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response
+Theory (IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017}
+recommend conducting these analyses and tests within a structural model
+that accounts for both the scores and their uncertainties, rather than
 treating them separately. Thus, an integrated approach combining CJ's
@@
 
-\textbf{Authors' contributions:} \emph{Conceptualization:} S.G., S.DM.,
-T.vD., and J.M.R.E; \emph{Methodology:} S.DM., T.vD., and J.M.R.E;
-\emph{Software:} J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal
-Analysis:} J.M.R.E.; \emph{Investigation:} J.M.R.E; \emph{Resources:}
-S.G., S.DM., and T.vD.; \emph{Data curation:} J.M.R.E.; \emph{Writing -
-original draft:} J.M.R.E.; \emph{Writing - review \& editing:} S.G.,
-S.DM., and T.vD.; \emph{Visualization:} J.M.R.E.; \emph{Supervision:}
-S.G. and S.DM.; \emph{Project administration:} S.G. and S.DM.;
-\emph{Funding acquisition:} S.G. and S.DM.
+\textbf{AI-assisted technologies in the writing process:} The authors
+used ChatGPT, an AI language model, during the preparation of this work.
+They occasionally employed the tool to refine phrasing and optimize
+wording, ensuring appropriate language use and enhancing the
+manuscript's clarity and coherence. The authors take full responsibility
+for the final content of the publication.
+
+\textbf{CRediT authorship contribution statement:}
+\emph{Conceptualization:} S.G., S.DM., T.vD., and J.M.R.E;
+\emph{Methodology:} S.DM., T.vD., and J.M.R.E; \emph{Software:}
+J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal Analysis:} J.M.R.E.;
+\emph{Investigation:} J.M.R.E; \emph{Resources:} S.G., S.DM., and T.vD.;
+\emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
+J.M.R.E.; \emph{Writing - review \& editing:} S.G., S.DM., and T.vD.;
+\emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;
+\emph{Project administration:} S.G. and S.DM.; \emph{Funding
+acquisition:} S.G. and S.DM.
 

### 2024-11-01 index.qmd
section:introduction,issues,structural; process:write

@@
 [In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
-The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher level of the trait.
+The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level.
 After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest.
-CJ has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}
+This method has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}
 
@@
 Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
-Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023], while research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 <!-- 3. But what are the problems of DCJ as a method? -->
-Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
+Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
 This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity.
@@
 Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing.
-The literature indicates that these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the latent trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+The literature indicates that the scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
@@
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty when conducting additional analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
+Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
 
 <!-- 4. So, what is the solution?  -->
-To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately.
-Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.
+To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model.
+A structural model specify how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
+This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements.
+Therefore, an integrated approach that combines CJ’s structural and measurement models can offer significant advantages.
 

### 2024-11-01 _tex/index.tex
Built site for gh-pages

@@
 produces a dichotomous outcome, indicating which stimulus is perceived
-to possess a higher level of the trait. After conducting multiple rounds
-of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL)
-model \citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
-estimate scores that reflect the underlying trait of interest. CJ has
-been successfully employed in assessing the quality of written texts
+to possess a higher trait level. After conducting multiple rounds of
+pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
+estimate scores that reflect the underlying trait of interest. This
+method has been successfully employed in assessing the quality of
+written texts
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.}
@@
 methods when employing adaptive comparison algorithms
-\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
-the other hand, research on validity suggests that scores generated by
-CJ can accurately represent the traits being measured
-\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
-Finally, research on practical applicability highlights the method's
+\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
+Additionally, research on validity suggests that scores generated by CJ
+can accurately represent the traits under measurement
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023},
+while research on practical applicability highlights the method's
 versatility across both educational and non-educational contexts
@@
 
-Nevertheless, despite the growing number of studies on CJ, unsystematic
-and fragmented research approaches in the literature have left several
+Nevertheless, despite the growing number of CJ studies, unsystematic and
+fragmented research approaches in the literature have left several
 critical issues unaddressed. This research primarily focuses on three:
@@
 transformations, to conduct additional analyses or hypothesis testing.
-The literature indicates that these scores have been used to identify
+The literature indicates that the scores have been used to identify
 `misfit' judges and stimuli
@@
 biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
-calculate correlations with other scoring methods
+calculate correlations with other assessment methods
 \citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
-related to the latent trait of interest
+related to the underlying trait of interest
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 that BTL scores are parameter estimates that inherently carry
-uncertainty. Ignoring this uncertainty when conducting additional
-analyses and tests can inflate their precision and statistical power,
-increasing the risk of committing a type I error \citep{McElreath_2020},
-which is when a null hypothesis is incorrectly rejected
+uncertainty. Ignoring this uncertainty when conducting separate analyses
+and tests can inflate their precision and statistical power, increasing
+the risk of committing a type I error \citep{McElreath_2020}, which is
+when a null hypothesis is incorrectly rejected
 \citep{Everitt_et_al_2010}.
@@
 Theory (IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017}
-recommend conducting these analyses and tests within a structural model
-that accounts for both the scores and their uncertainties, rather than
-treating them separately. Thus, an integrated approach combining CJ's
+recommend conducting these analyses and tests within a structural model.
+A structural model specify how different manifest or latent variables
+influence the latent variable of interest \citep{Everitt_et_al_2010}.
+This approach allows analyses that can account for both the scores and
+their uncertainties simultaneously, rather than treating them as
+separate elements. Therefore, an integrated approach that combines CJ's
 structural and measurement models can offer significant advantages.

### 2024-11-05 index.qmd
section:issues; process:code

@@
 keywords:
-  - causal inference
-  - probability
-  - Thurstone
-  - comparative judgement
-  - directed acyclic graph
-  - structural causal models
-  - statistical modeling
+  - Probability
+  - Directed Acyclic Graphs
+  - Bayesian methods
+  - Thurstonian model
+  - Comparative judgement
+  - Structural Causal Models
+  - Statistical modeling
 abstract: |
@@
 
+```{r}
+#| results: hide
+#| label: code-package
+
+# # load packages
+# libraries = c('RColorBrewer','stringr','dplyr','tidyverse',
+#               'rechape2','knitr','kableExtra',
+#               'rstan','StanHeaders','runjags','rethinking')
+# sapply(libraries, require, character.only=T)
+
+```
+
+```{r}
+#| results: hide
+#| label: code-usd
+
+# # load functions
+# main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript'
+# source( file.path( main_dir, 'code', 'user-defined-functions.R') )
+
+```
+
 # Introduction {#sec-introduction}
@@
 
-<!-- []{style="color:red;"}. -->
-
 <!-- 1. What does CJ do? -->
-[In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
+In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
 The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level.
 After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest.
-This method has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}
+This method has been successfully employed in assessing the quality of written texts, where quality describes the underlying trait of interest and the texts serve as the stimuli [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023], while research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
+Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 <!-- 3. But what are the problems of DCJ as a method? -->
-Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
+Nevertheless, despite the growing number of CJ studies, the literature's unsystematic and fragmented research approaches have left several critical issues unaddressed.
 This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity.
-The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.
+The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.
 
@@
 Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing.
-The literature indicates that the scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+The scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
@@
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
+Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020].
+A type I error results when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
 
@@
 
+<!-- []{style="color:red;"}. -->
+
+<!-- 1. What are the five cases of Thurstone? -->
+
 <!-- 1. the BTL model is the statistical articulation of Case 5 -->
 
-<!--   From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927].
-Talk about @Pollitt_et_al_2003 and @Bramley_2008.
--->
+From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008.
 
-<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->
+  What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution
 
-<!--   Table with assumptions -->
+  Table with assumptions
 
@@
 
+## Appendix A: Ignoring uncertainty {#sec-appendix-A}
+
+```{r}
+#| label: code-AppA-simulation
+
+# # simulate units and sub-units
+# USd = sim_units_trait( Un=10,
+#                        Us=c(1,1),
+#                        Ub=c(0,-0.2,0.2),
+#                        Useed = 45789,
+#                        Sn=6,
+#                        Ss=c(0.8,0.8),
+#                        Sb=c(0,0,0),
+#                        Sseed = 9478 )
+# 
+# # simulate judges bias
+# Jd = sim_judges_bias( Jn=10,
+#                       Js=c(0.02,0.02),
+#                       Jseed = 79985 )
+# 
+# 
+# # simulate full comparison data
+# d = cm_full( USd=USd$USd,
+#              Jd=Jd$Jd )
+# 
+# 
+# # set data in list
+# dL = list_data( d )
+# str(dL)
+
+```
+
+## Appendix B: The five cases of Thurstone {#sec-appendix-B}
+
+<!-- Thurstone's comparative judgment (CJ) theory [@Thurstone_1927] is composed of two key elements: the discriminal process and the law of comparative judgment.   -->
+
+<!-- In a CJ study evaluating text quality quality of written texts -->
+
 {{< pagebreak >}}
 
-## References {.unnumbered}
+# References {.unnumbered}
 

### 2024-11-05 code/user-defined-functions.R

@@ -1,0 +1,468 @@
+require(docstring)
+
+rm(list=ls())
+
+
+#' Simulating predictor variables
+#'
+#' @param n sample size, default=10
+#' @param sup_c support for continuous variable, default=c(-3,3)
+#' @param sup_d support for discrete variable, default=1:2
+#' @param prop_d sample proportion for sup_d, default=c(0.5,0.5) 
+#'
+#' @return data.frame of generated variables
+#' @export
+#'
+#' @examples
+sim_var = function( n = 30,
+                    sup_c = c(-3,3),
+                    sup_d = 1:2,
+                    prop_d = c(0.5,0.5) ){
+  
+  # # test
+  # n = 30
+  # sup_c = c(-3,3)
+  # sup_d = 1:2
+  # prop_d = c(0.4,0.6)
+  
+  
+  # continuous variable
+  delta = ( sup_c[2]-sup_c[1] ) / n # effective minimum difference
+  start = sup_c[1] + delta/2
+  end = sup_c[2] - delta/2 
+  Xc = seq( start, end, length.out=n )
+  
+  idx = sample( 1:n, size=n, replace=F ) # break correlation
+  Xc = Xc[idx]
+  
+  
+  # dichotomous variable
+  n_d = round( n*prop_d ) # samples based on proportion
+  Xd = c()
+  for( i in sup_d ){
+    Xd = c( Xd, rep( sup_d[i], each=n_d[i] ) )
+  }
+  if( sum(n_d) < n ){
+    extra = sample(sup_d, size=n-sum(n_d), replace=T)
+    Xd = c(Xd, extra)
+  }
+  
+  idx = sample( 1:n, size=n, replace=F ) # break correlation
+  Xd = Xd[idx]
+  
+  
+  # join data
+  X = data.frame( Xc, Xd )
+  
+  
+  # return object
+  return(X)
+  
+}
+
+
+
+#' Simulate judges' bias
+#'
+#' @param Jn sample size, default=10
+#' @param Jb vector with betas, size=1+length(Jsup_d), default=c(0,0,0), no judges differences
+#' @param Js vector with standard error, size=length(Jsup_d), default=c(1,1)
+#' @param Jsup_c support for continuous variable, default=c(-3,3)
+#' @param Jsup_d support for discrete variable, default=1:2
[…]

### 2024-11-05 _tex/index.tex
Built site for gh-pages

@@
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
-  pdfkeywords={causal inference, probability, Thurstone, comparative
-judgement, directed acyclic graph, structural causal models, statistical
-modeling},
+  pdfkeywords={Probability, Directed Acyclic Graphs, Bayesian
+methods, Thurstonian model, Comparative judgement, Structural Causal
+Models, Statistical modeling},
   colorlinks=true,
@@
 \begin{keyword}
-    causal inference \sep probability \sep Thurstone \sep comparative
-judgement \sep directed acyclic graph \sep structural causal
-models \sep 
-    statistical modeling
+    Probability \sep Directed Acyclic Graphs \sep Bayesian
+methods \sep Thurstonian model \sep Comparative
+judgement \sep Structural Causal Models \sep 
+    Statistical modeling
 \end{keyword}
@@
 
-{In \emph{comparative judgment} (CJ) studies, judges assess the presence
+In \emph{comparative judgment} (CJ) studies, judges assess the presence
 of a trait or competence by conducting pairwise comparisons of stimuli
@@
 method has been successfully employed in assessing the quality of
-written texts
-\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.}
+written texts, where quality describes the underlying trait of interest
+and the texts serve as the stimuli
+\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
 
@@
 can accurately represent the traits under measurement
-\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023},
-while research on practical applicability highlights the method's
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
+Finally, research on practical applicability highlights the method's
 versatility across both educational and non-educational contexts
@@
 
-Nevertheless, despite the growing number of CJ studies, unsystematic and
-fragmented research approaches in the literature have left several
+Nevertheless, despite the growing number of CJ studies, the literature's
+unsystematic and fragmented research approaches have left several
 critical issues unaddressed. This research primarily focuses on three:
@@
 followed by the introduction of a theoretical model and its statistical
-translation, which aim to address all three concerns simultaneously.
+translation, which aims to address all three concerns simultaneously.
 
@@
 transformations, to conduct additional analyses or hypothesis testing.
-The literature indicates that the scores have been used to identify
-`misfit' judges and stimuli
+The scores have been used to identify `misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
@@
 and tests can inflate their precision and statistical power, increasing
-the risk of committing a type I error \citep{McElreath_2020}, which is
-when a null hypothesis is incorrectly rejected
+the risk of committing a type I error \citep{McElreath_2020}. A type I
+error results when a null hypothesis is incorrectly rejected
 \citep{Everitt_et_al_2010}.
@@
 model}\label{the-assumptions-of-case-5-and-the-measurement-model}
+
+From early on in the literature, it has been clear that the BTL model
+represents a statistical articulation of Thurstone's Case 5
+\citeyearpar{Thurstone_1927}. Talk about \citet{Pollitt_et_al_2003} and
+\citet{Bramley_2008}.
+
+What case 5 implies, Assumptions. Not a normal distribution but a
+logistic distribution
+
+Table with assumptions
 
@@
 
+\subsection{Appendix A: Ignoring uncertainty}\label{sec-appendix-A}
+
+\subsection{Appendix B: The five cases of
+Thurstone}\label{sec-appendix-B}
+
 \newpage{}
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 

### 2024-11-07 index.qmd
section:introduction,issues,structural; process:write

@@
 
-```{r}
-#| results: hide
-#| label: code-package
-
-# # load packages
-# libraries = c('RColorBrewer','stringr','dplyr','tidyverse',
-#               'rechape2','knitr','kableExtra',
-#               'rstan','StanHeaders','runjags','rethinking')
-# sapply(libraries, require, character.only=T)
-
-```
-
-```{r}
-#| results: hide
-#| label: code-usd
-
-# # load functions
-# main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript'
-# source( file.path( main_dir, 'code', 'user-defined-functions.R') )
-
-```
-
 # Introduction {#sec-introduction}
@@
 <!-- 1. What does CJ do? -->
-In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
+In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence in different stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
 The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level.
-After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest.
+After multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the resulting outcomes and estimate scores that reflect the underlying trait of interest.
 This method has been successfully employed in assessing the quality of written texts, where quality describes the underlying trait of interest and the texts serve as the stimuli [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
@@
 <!-- 2. But there is no structural model, only post estimation analyses -->
-Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing.
-The scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
+For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
 <!-- 3. But what is the problem with this? -->
-However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests.
+However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests.
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020].
-A type I error results when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].
+Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests.
+Notably, the direction and magnitude of this bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, p.
+25; @Hoyle_et_al_2023, p.
+137].
+Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
+
+<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->
 
 <!-- 4. So, what is the solution?  -->
-To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model.
-A structural model specify how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
+To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Kline_et_al_2023; @Hoyle_et_al_2023, p.
+138] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model.
+A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
 This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements.
-Therefore, an integrated approach that combines CJ’s structural and measurement models can offer significant advantages.
+Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.
 
@@
 
-From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008.
+<!-- From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927].
+Talk about @Pollitt_et_al_2003 and @Bramley_2008.
+-->
 
-  What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution
+<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->
 
-  Table with assumptions
+<!--   Table with assumptions -->
 
@@
 
-## Appendix A: Ignoring uncertainty {#sec-appendix-A}
-
-```{r}
-#| label: code-AppA-simulation
-
-# # simulate units and sub-units
-# USd = sim_units_trait( Un=10,
-#                        Us=c(1,1),
-#                        Ub=c(0,-0.2,0.2),
-#                        Useed = 45789,
-#                        Sn=6,
-#                        Ss=c(0.8,0.8),
-#                        Sb=c(0,0,0),
-#                        Sseed = 9478 )
-# 
-# # simulate judges bias
-# Jd = sim_judges_bias( Jn=10,
-#                       Js=c(0.02,0.02),
-#                       Jseed = 79985 )
-# 
-# 
-# # simulate full comparison data
-# d = cm_full( USd=USd$USd,
-#              Jd=Jd$Jd )
-# 
-# 
-# # set data in list
-# dL = list_data( d )
-# str(dL)
-
-```
-
-## Appendix B: The five cases of Thurstone {#sec-appendix-B}
-
-<!-- Thurstone's comparative judgment (CJ) theory [@Thurstone_1927] is composed of two key elements: the discriminal process and the law of comparative judgment.   -->
-
-<!-- In a CJ study evaluating text quality quality of written texts -->
-
 {{< pagebreak >}}

### 2024-11-07 _tex/index.tex
Built site for gh-pages

@@
 In \emph{comparative judgment} (CJ) studies, judges assess the presence
-of a trait or competence by conducting pairwise comparisons of stimuli
-\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
-produces a dichotomous outcome, indicating which stimulus is perceived
-to possess a higher trait level. After conducting multiple rounds of
+of a trait or competence in different stimuli by performing pairwise
+comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The
+comparison produces a dichotomous outcome, indicating which stimulus is
+perceived to possess a higher trait level. After multiple rounds of
 pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
-estimate scores that reflect the underlying trait of interest. This
+\citep{Bradley_et_al_1952, Luce_1959} to process the resulting outcomes
+and estimate scores that reflect the underlying trait of interest. This
 method has been successfully employed in assessing the quality of
@@
 Researchers then typically use the estimated BTL scores, or their
-transformations, to conduct additional analyses or hypothesis testing.
-The scores have been used to identify `misfit' judges and stimuli
-\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
-biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
-calculate correlations with other assessment methods
-\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
-related to the underlying trait of interest
+transformations, to conduct additional analyses or hypothesis tests. For
+example, these scores have been used to identify `misfit' judges and
+stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018},
+detect biases in judges' ratings
+\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
+other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
+or test hypotheses related to the underlying trait of interest
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
 
-However, the statistical literature cautions against using estimated
-scores to conduct additional analyses or tests. A key consideration is
-that BTL scores are parameter estimates that inherently carry
-uncertainty. Ignoring this uncertainty when conducting separate analyses
-and tests can inflate their precision and statistical power, increasing
-the risk of committing a type I error \citep{McElreath_2020}. A type I
-error results when a null hypothesis is incorrectly rejected
-\citep{Everitt_et_al_2010}.
+However, the statistical literature advises caution when using estimated
+scores to conduct additional analyses or hypotheses tests. A key
+consideration is that BTL scores are parameter estimates that inherently
+carry uncertainty. Ignoring this uncertainty can introduce bias into the
+analysis and reduce the precision of hypothesis tests. Notably, the
+direction and magnitude of this bias are often unpredictable; results
+may be attenuated, exaggerated, or remain unaffected, depending on the
+uncertainty in the scores and the actual effects being tested
+\citetext{\citealp[p.~25]{Kline_et_al_2023}; \citealp[p.~137]{Hoyle_et_al_2023}}.
+Furthermore, reduced precision in hypothesis tests weakens their
+statistical power, ultimately increasing the likelihood of committing
+type-I or type-II errors \citep{McElreath_2020}.
 
 To mitigate these risks, principles from Structural Equation Modeling
-(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response
-Theory (IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017}
-recommend conducting these analyses and tests within a structural model.
-A structural model specify how different manifest or latent variables
+(SEM) \citep[p.~138]{Kline_et_al_2023, Hoyle_et_al_2023} and Item
+Response Theory (IRT)
+\citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
+conducting these analyses and tests within a structural model. A
+structural model specifies how different manifest or latent variables
 influence the latent variable of interest \citep{Everitt_et_al_2010}.
@@
 separate elements. Therefore, an integrated approach that combines CJ's
-structural and measurement models can offer significant advantages.
+measurement and structural models can offer significant advantages.
 
@@
 model}\label{the-assumptions-of-case-5-and-the-measurement-model}
-
-From early on in the literature, it has been clear that the BTL model
-represents a statistical articulation of Thurstone's Case 5
-\citeyearpar{Thurstone_1927}. Talk about \citet{Pollitt_et_al_2003} and
-\citet{Bramley_2008}.
-
-What case 5 implies, Assumptions. Not a normal distribution but a
-logistic distribution
-
-Table with assumptions
 
@@
 
-\subsection{Appendix A: Ignoring uncertainty}\label{sec-appendix-A}
-
-\subsection{Appendix B: The five cases of
-Thurstone}\label{sec-appendix-B}
-
 \newpage{}

### 2024-11-07 code/user-defined-functions.R

@@ -1,468 +1,0 @@
-require(docstring)
-
-rm(list=ls())
-
-
-#' Simulating predictor variables
-#'
-#' @param n sample size, default=10
-#' @param sup_c support for continuous variable, default=c(-3,3)
-#' @param sup_d support for discrete variable, default=1:2
-#' @param prop_d sample proportion for sup_d, default=c(0.5,0.5) 
-#'
-#' @return data.frame of generated variables
-#' @export
-#'
-#' @examples
-sim_var = function( n = 30,
-                    sup_c = c(-3,3),
-                    sup_d = 1:2,
-                    prop_d = c(0.5,0.5) ){
-  
-  # # test
-  # n = 30
-  # sup_c = c(-3,3)
-  # sup_d = 1:2
-  # prop_d = c(0.4,0.6)
-  
-  
-  # continuous variable
-  delta = ( sup_c[2]-sup_c[1] ) / n # effective minimum difference
-  start = sup_c[1] + delta/2
-  end = sup_c[2] - delta/2 
-  Xc = seq( start, end, length.out=n )
-  
-  idx = sample( 1:n, size=n, replace=F ) # break correlation
-  Xc = Xc[idx]
-  
-  
-  # dichotomous variable
-  n_d = round( n*prop_d ) # samples based on proportion
-  Xd = c()
-  for( i in sup_d ){
-    Xd = c( Xd, rep( sup_d[i], each=n_d[i] ) )
-  }
-  if( sum(n_d) < n ){
-    extra = sample(sup_d, size=n-sum(n_d), replace=T)
-    Xd = c(Xd, extra)
-  }
-  
-  idx = sample( 1:n, size=n, replace=F ) # break correlation
-  Xd = Xd[idx]
-  
-  
-  # join data
-  X = data.frame( Xc, Xd )
-  
-  
-  # return object
-  return(X)
-  
-}
-
-
-
-#' Simulate judges' bias
-#'
-#' @param Jn sample size, default=10
-#' @param Jb vector with betas, size=1+length(Jsup_d), default=c(0,0,0), no judges differences
-#' @param Js vector with standard error, size=length(Jsup_d), default=c(1,1)
-#' @param Jsup_c support for continuous variable, default=c(-3,3)
-#' @param Jsup_d support for discrete variable, default=1:2
[…]

### 2024-11-08 index.qmd
section:introduction,issues,structural; process:write

@@
 
-<!-- recording 1 -->
-<!-- Part 1; 00:00:00 - 00:09:10 -->
-
-<!-- recording 2 -->
-<!-- Part 2; 00:00:00 - 00:17:30 -->
+<!-- recording: Sven 24.10.04; time: 00:00:00 - 00:09:10 -->
+<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
+<!-- recording: Sven and Tine 24.10.18; time: 00:04:05 - 00:08:30 -->
+<!-- recording: Sven and Tine 24.10.25; time: 00:00:00 - 00:17:00 -->
 
 <!-- 1. What does CJ do? -->
-In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence in different stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
-The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level.
-After multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the resulting outcomes and estimate scores that reflect the underlying trait of interest.
-This method has been successfully employed in assessing the quality of written texts, where quality describes the underlying trait of interest and the texts serve as the stimuli [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+In *comparative judgment* (CJ) studies, judges assess a specific trait or competence across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
+Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to manifest a higher trait level.
+For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
-<!-- Specifically, CJ has been used to evaluate the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023] and to assess the quality of speech [@Boonen_et_al_2020].
-Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_2015], engineering design skills [@Bartholomew_et_al_2018], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowdlege [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020].
+<!-- and spoken language [@Boonen_et_al_2020].
+Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_2015], engineering design skills [@Bartholomew_et_al_2018], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020].
 -->
@@
 <!-- 2. Where are we now with DCJ? -->
-Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade.
+Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade.
 These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability.
@@
 Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
+On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
 Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
@@
 <!-- 3. But what are the problems of DCJ as a method? -->
-Nevertheless, despite the growing number of CJ studies, the literature's unsystematic and fragmented research approaches have left several critical issues unaddressed.
+Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
 This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity.
@@
 
+<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
+<!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
+<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:00 -->
+
 <!-- 1. The BTL model is CJ's measurement model -->
-In a typical CJ study, the BTL model serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008].
+In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008].
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
-For example, when evaluating text quality, the BTL model uses the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests.
-Notably, the direction and magnitude of this bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, p.
-25; @Hoyle_et_al_2023, p.
+Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp.
+25; @Hoyle_et_al_2023, pp.
 137].
@@
 <!-- 4. So, what is the solution?  -->
-To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Kline_et_al_2023; @Hoyle_et_al_2023, p.
-138] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model.
+To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+138] and Item Response Theory (IRT) [@Fox_2010, chap.
+6; @vanderLinden_et_al_2017_I, chap.
+24] recommend conducting these analyses and tests within a structural model.
 A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
@@
 ## The assumptions of Case 5 and the measurement model
+
+<!-- recording: Sven and Tine 24.10.25; time: 00:31:00 - 00:52:20 -->
 
@@
 
-<!--   [Reliability is a necessary but not sufficient condition for validity.
-Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}.
--->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015] -->
 
@@
 # Theory {#sec-theory}
-<!-- recording 1 -->
-<!-- Part 2; 00:14:30 - 00:21:00 -->
 
@@
 
+<!-- general development -->
+<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
+<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
+<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
+
+<!-- comparison mechanism as missing mechanism -->
+<!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
+<!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->
+
 ## From theory to statistics {#sec-theory-statistics}
+
+<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
+
+<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independece assumption -->
+
+<!-- [@deAyala_2009, p.
+291-29] think about indeterminancies in MIRT.
+to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
 

### 2024-11-08 _tex/index.tex
Built site for gh-pages

@@
 
-In \emph{comparative judgment} (CJ) studies, judges assess the presence
-of a trait or competence in different stimuli by performing pairwise
-comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The
+In \emph{comparative judgment} (CJ) studies, judges assess a specific
+trait or competence across various stimuli by performing pairwise
+comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. Each
 comparison produces a dichotomous outcome, indicating which stimulus is
-perceived to possess a higher trait level. After multiple rounds of
-pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959} to process the resulting outcomes
-and estimate scores that reflect the underlying trait of interest. This
-method has been successfully employed in assessing the quality of
-written texts, where quality describes the underlying trait of interest
-and the texts serve as the stimuli
+perceived to manifest a higher trait level. For example, when assessing
+text quality, judges compare pairs of written texts (the stimuli) to
+determine the relative quality each text exhibit (the trait)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
@@
 Numerous studies have documented the effectiveness of CJ in assessing
-various traits and competencies over the past decade. These studies have
+traits and competencies over the past decade. These studies have
 emphasized three aspects of the method's effectiveness: its reliability,
@@
 methods when employing adaptive comparison algorithms
-\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
-Additionally, research on validity suggests that scores generated by CJ
-can accurately represent the traits under measurement
+\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
+the other hand, research on validity suggests that scores generated by
+CJ can accurately represent the traits under measurement
 \citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
@@
 
-Nevertheless, despite the growing number of CJ studies, the literature's
-unsystematic and fragmented research approaches have left several
+Nevertheless, despite the growing number of CJ studies, unsystematic and
+fragmented research approaches in the literature have left several
 critical issues unaddressed. This research primarily focuses on three:
@@
 
-In a typical CJ study, the BTL model serves as the measurement model for
-CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
-manifest variables contribute to the estimation of latent variables
+In a typical CJ study, the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
+for CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies
+how manifest variables contribute to the estimation of latent variables
 \citep{Everitt_et_al_2010}. For example, when evaluating text quality,
-the BTL model uses the dichotomous outcomes resulting from the pairwise
-comparisons (the manifest variables) to estimate scores that reflect the
-underlying quality level of the texts (the latent variable)
+researchers use the BTL model to process the dichotomous outcomes
+resulting from the pairwise comparisons (the manifest variables) to
+estimate scores that reflect the underlying quality level of the texts
+(the latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
@@
 analysis and reduce the precision of hypothesis tests. Notably, the
-direction and magnitude of this bias are often unpredictable; results
-may be attenuated, exaggerated, or remain unaffected, depending on the
-uncertainty in the scores and the actual effects being tested
-\citetext{\citealp[p.~25]{Kline_et_al_2023}; \citealp[p.~137]{Hoyle_et_al_2023}}.
+direction and magnitude of the bias are often unpredictable; results may
+be attenuated, exaggerated, or remain unaffected, depending on the
+amount of uncertainty present in the scores and the actual effects being
+tested
+\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
 Furthermore, reduced precision in hypothesis tests weakens their
@@
 To mitigate these risks, principles from Structural Equation Modeling
-(SEM) \citep[p.~138]{Kline_et_al_2023, Hoyle_et_al_2023} and Item
-Response Theory (IRT)
-\citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
-conducting these analyses and tests within a structural model. A
-structural model specifies how different manifest or latent variables
+(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
+\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
+recommend conducting these analyses and tests within a structural model.
+A structural model specifies how different manifest or latent variables
 influence the latent variable of interest \citep{Everitt_et_al_2010}.

### 2024-11-12 index.qmd
section:issues,caseV; process:write

@@
 
+```{r}
+#| results: hide
+#| label: code-package
+
+# main directory
+main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript'
+
+# load packages
+libraries = c('tidyverse','tibble','gt')
+sapply(libraries, require, character.only=T)
+
+```
+
 # Introduction {#sec-introduction}
@@
 Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
-This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity.
+This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case V [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity.
 The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.
@@
 <!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
-<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:00 -->
+<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:10 -->
 
@@
 Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
-
 <!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->
@@
 A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
-This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements.
+This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
 Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.
 
-## The assumptions of Case 5 and the measurement model
+## The assumptions of Case V and the measurement model
 
-<!-- recording: Sven and Tine 24.10.25; time: 00:31:00 - 00:52:20 -->
-
-<!-- []{style="color:red;"}. -->
-
-<!-- 1. What are the five cases of Thurstone? -->
+<!-- recording: Sven and Tine 24.10.25; time: 00:31:10 - 00:52:20 -->
 
 <!-- 1. the BTL model is the statistical articulation of Case 5 -->
-
-<!-- From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927].
-Talk about @Pollitt_et_al_2003 and @Bramley_2008.
--->
-
-<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->
-
-<!--   Table with assumptions -->
+In the CJ literature, researchers use the BTL model as the statistical representation of Thurstone's theory, particularly Thurstone's Case V.
+The BTL model closely aligns with the assumptions of Case V, with one notable distinction: the mathematical function linking dichotomous outcomes to trait scores, referred to as the *discriminal process distribution*.
+While Case V uses a cumulative normal distribution, the BTL model employs the more tractable logistic distribution [@Bramley_2008, pp.
+254] (see @tbl-thurstone_cases).
+Importantly, this modification does not have a significant impact on the estimation or interpretation of the model compared with Case V, mainly because the logistic and normal distributions are related to each other by a factor of $1.7$ [@vanderLinden_et_al_2017_I, pp.
+16].
 
 <!-- 2. But this model hold the greater number of assumptions -->
+However, Case V incorporates the largest number of simplifying assumptions [@Bramley_2008, pp.
+253], prioritizing ease of use from the statistical point of view over precision in trait measurement.
+This focus on simplicity reflects Case V’s original purpose: to provide a "rather coarse scaling" of traits [@Thurstone_1927, pp.
+269].
+@tbl-thurstone_cases presents each of Thurstone's cases along with their underlying assumptions.
+These assumptions fall into four categories: assumptions concerning the *discriminal process distribution* and *discriminal dispersion* of stimuli, assumptions about correlations among stimuli, and assumptions regarding who makes the comparisons, either a single judge or multiple judges.
+For a comprehensive explanation of this theoretical progression, see @Thurstone_1927 and @Bramley_2008 [pp.
+248-253].
 
-<!--   Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
-269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions.
--->
+```{r}
+# #| results: hide
+# #| label: tbl-thurstone_cases
+# #| tbl-cap: "Thurstone's cases and BTL model assumptions."
+# #| classes: plain
+# 
+# assump = tibble( 
+#   Assumption = c('Discriminal process distribution', 'Discriminal dispersion', 
+#                  'Correlation between stimuli', 'Who compares?'),
+#   `Case I` = c('Cummulative Normal','Different','Constant','Single judge'),
+#   `Case II` = c('Cummulative Normal','Different','Constant','Multiple judges'),
+#   `Case III` = c('Cummulative Normal','Different','Zero','Multiple judges'),
+#   `Case IV` = c('Cummulative Normal','Similar','Zero','Multiple judges'),
+#   `Case V` = c('Cummulative Normal','Equal','Zero','Multiple judges'),
+#   `BTL model` = c('Logistic','Equal','Zero','Multiple judges') ) 
+# 
+# assump = assump %>% 
+#   mutate( Assumption = str_replace_all( Assumption, " ", "<br>" ),
+#           `Case I` = str_replace_all( `Case I`, " ", "<br>" ),
+#           `Case II` = str_replace_all( `Case II`, " ", "<br>" ),
+#           `Case III` = str_replace_all( `Case III`, " ", "<br>" ),
+#           `Case IV` = str_replace_all( `Case IV`, " ", "<br>" ),
+#           `Case V` = str_replace_all( `Case V`, " ", "<br>" ),
+#           `BTL model` = str_replace_all( `BTL model`, " ", "<br>" ) )
+# 
+# gt_assump = gt(assump) |>
+#   tab_spanner( label=md("**Thurstone's**"), columns=2:6 ) |>
+#   cols_align( align="center", columns=1:7 ) |>
+#   tab_style( style = cell_text(weight="bold"), locations=cells_column_labels() ) |>
+#   tab_options( table.font.size=pct(90) ) |>
+#   fmt_markdown( columns=everything() ) 
+# 
+# gt_assump
 
-<!--   @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three: -->
+```
+
+@Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:
 
@@
 
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015] -->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->
 
 ## The role and impact of comparison algorithms
+
+<!-- []{style="color:red;"}. -->
 

### 2024-11-14 images/tables/thurstone_cases.tex
section:issues; process:write

\documentclass[border=3mm, varwidth=500mm]{standalone}

\usepackage{booktabs} 
\usepackage{color, colortbl}
\usepackage[usestackEOL]{stackengine}

\definecolor{gray.95}{gray}{0.95}

\begin{document}

\centering
\begin{table}[h!]
    \begin{center}
    \begin{tabular}{ccccccc}
        & \multicolumn{5}{c}{ \textbf{Thurstone's} } & \textbf{BTL} \\
        \cline{2-6}
        \textbf{Assumption} & \textbf{Case I} & \textbf{Case II} & \textbf{Case III} & \textbf{Case IV} & \textbf{Case V} & \textbf{model} \\
        \hline

        \rowcolor{gray.95} \Centerstack{ Discriminal process \\ (distribution) } & Normal & Normal & Normal & Normal & Normal & Logistic \\

        \Centerstack{ Discriminal dispersion \\ (between stimuli) } & Different & Different & Different & Similar & Equal & Equal \\

        \rowcolor{gray.95} \Centerstack{ Correlation \\ (between stimuli) } & Constant & Constant & Zero & Zero & Zero & Zero \\

        \Centerstack{ How many \\ judges compare? } & Single & Multiple & Multiple & Multiple & Multiple & Multiple \\

        %\rowcolor{gray.95} \Centerstack{ Bias effects \\ in judges? } & Possible & Possible & \Centerstack{ Cancels \\ out } & \Centerstack{ Cancels \\ out } & \Centerstack{ Cancels \\ out } & \Centerstack{ Cancels \\ out } \\

    \end{tabular}
    \end{center}
\end{table}

\end{document}

### 2024-11-14 index.qmd
section:issues; process:write

--- 2024-11-08 index.qmd
@@
 <!-- 1. What does CJ do? -->
-In *comparative judgment* (CJ) studies, judges assess a specific trait or competence across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
-Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to manifest a higher trait level.
+In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
+Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level.
 For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
@@
 <!-- 3. But what are the problems of DCJ as a method? -->
-Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed.
-This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity.
+Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed.
+This research primarily focuses on three: the over-reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role and influence of comparison algorithms on the method's reliability and validity.
 The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.
@@
 
-## The disconnect between structural and measurement models
+## The Case V and the statistical analysis of CJ data
 
-<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
-<!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
-<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:00 -->
+<!-- recording: Sven and Tine 24.10.25; time: 00:31:10 - 00:52:20 -->
+<!-- recording: Sven and Tine 24.11.12; time: 00:00:00 - 00:00:00 -->
 
-<!-- 1. The BTL model is CJ's measurement model -->
-In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008].
-A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
-For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+<!-- 1. What is the general from of the CJ theory? -->
+In its most general form, Thurstone's theory [-@Thurstone_1927] posits that when a judge compares two stimuli, the resulting dichotomous outcome depends on two main factors: the discriminal process and the law of comparative judgment.
+The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus.
+Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution.
+In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus.
+@fig-discriminal_process shows a visual depiction of the discriminal process of two stimuli (objects).
 
-<!-- 2. But there is no structural model, only post estimation analyses -->
-Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
-For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+![Discriminal processes of two stimuli (objects). Extracted from @Bramley_2008 [pp. 249] ](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
 
-<!-- 3. But what is the problem with this? -->
-However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests.
-A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests.
-Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp.
-25; @Hoyle_et_al_2023, pp.
-137].
-Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
+However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential.
+This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait.
+Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*.
+@fig-comparative_judgment shows a visual depiction of the discriminal difference of two stimuli (objects).
 
-<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->
+![Discriminal difference between two stimuli (objects). Extracted from @Bramley_2008 [pp. 251] ](/images/figures/comparative_judgment.png){#fig-comparative_judgment width=65%}
 
-<!-- 4. So, what is the solution?  -->
-To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
-138] and Item Response Theory (IRT) [@Fox_2010, chap.
-6; @vanderLinden_et_al_2017_I, chap.
-24] recommend conducting these analyses and tests within a structural model.
-A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
-This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements.
-Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.
+<!-- 2. But several simplifying assumption are made to ease its use -->
+Thurstone’s general form of the theory primarily applies to pairwise comparisons of stimuli made by a single judge [@Thurstone_1927, pp.
+267].
+Therefore, to enhance the theory's practical applicability, he developed five distinct cases, each incorporating a series of increasingly simplifying assumptions.
+@fig-thurstone_cases outlines each case along with their key assumptions.
+For a detailed explanation of this theoretical progression, see @Thurstone_1927 and @Bramley_2008 [pp.
+248-253].
 
-## The assumptions of Case 5 and the measurement model
+![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#fig-thurstone_cases width=85%}
 
-<!-- recording: Sven and Tine 24.10.25; time: 00:31:00 - 00:52:20 -->
+<!-- 3. And yet Case V is is the most commonly used -->
+Surprisingly, Case V is the most commonly used case in the CJ literature, despite incorporating the largest number of simplifying assumptions [@Bramley_2008, pp.
+253].
+This widespread adoption is primarily due to the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959], which offers a simpler statistical representation of the case.
+The BTL model closely aligns with Case V’s assumptions, with one key difference: the distribution of the discriminal process.
+While Case V assumes a Normal distribution, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp.
+254] (see @fig-thurstone_cases).
+However, this adjustment does not significantly affect the model’s estimation or interpretation, as the logistic and normal distributions are related to each other by a factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
+16].
 
-<!-- []{style="color:red;"}. -->
+<!-- the mathematical function linking dichotomous outcomes to trait scores -->
 
-<!-- 1. What are the five cases of Thurstone? -->
+<!-- 4. But two assumptions of Case V might not hold -->
+However, because Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927, pp.
+269], prioritizing statistical ease of use over precision in trait measurement, two of its assumptions may be causing more problems than expected: the assumption of zero correlation between the stimuli, and the assumption of equal dispersion between stimuli.
 
-<!-- 1. the BTL model is the statistical articulation of Case 5 -->
+<!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are two: -->
 
-<!-- From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927].
-Talk about @Pollitt_et_al_2003 and @Bramley_2008.
--->
-
-<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->
-
-<!--   Table with assumptions -->
-
-<!-- 2. But this model hold the greater number of assumptions -->
-
-<!--   Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p.
-269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions.
--->
-
-<!--   @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three: -->
-
-<!--   (a) related to Case 2 and case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects.
+<!--   (a) it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects.
 This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.
@@
 
-<!--   (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
+<!--   (b) it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
 Meaning there is an amount of correlation that it is not accounted for.
@@
 
-<!--   (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
+<!-- (c) discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
 Example of exaggerated comparison of individuals, university students vs secondary school students.
@@
 <!--   here we face misspecification in the model -->
+
+<!-- "It is probably true that this variability of the discriminal dispersions on the psychological continuum is of relatively less serious importance in dealing with strictly homogenous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogenous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927].
+-->
 
@@
 
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015] -->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->
+
+## The disconnect between structural and measurement models
+
+<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
+<!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
+<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:10 -->
+
+<!-- 1. The BTL model is CJ's measurement model -->
+As outlined in the previous section, the BTL model typically functions as the measurement model in a CJ study [@Andrich_1978; @Bramley_2008].
+A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
+For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+
+<!-- 2. But there is no structural model, only post estimation analyses -->
+Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
+For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+
+<!-- 3. But what is the problem with this? -->
+However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests.
+A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
+Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests.
+Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp.
+25; @Hoyle_et_al_2023, pp.
+137].
+Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
+<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->
+
+<!-- 4. So, what is the solution?  -->
+To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+138] and Item Response Theory (IRT) [@Fox_2010, chap.
+6; @vanderLinden_et_al_2017_I, chap.
+24] recommend conducting these analyses and tests within a structural model.
+A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
+This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
+Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.
 
 ## The role and impact of comparison algorithms
+
+<!-- []{style="color:red;"}. -->
 
@@
 <!-- All studies use SSR as a measure of reliability -->
+
+<!-- It is assumed that the single observer compares each pair of stimuli a sufficient number of times so that a proportion may be determined for eah pair [@Thurstone_1927]. -->
 

### 2024-11-14 _tex/index.tex
Built site for gh-pages

@@
 In \emph{comparative judgment} (CJ) studies, judges assess a specific
-trait or competence across various stimuli by performing pairwise
+trait or attribute across various stimuli by performing pairwise
 comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. Each
 comparison produces a dichotomous outcome, indicating which stimulus is
-perceived to manifest a higher trait level. For example, when assessing
+perceived to exhibit a higher trait level. For example, when assessing
 text quality, judges compare pairs of written texts (the stimuli) to
@@
 
-Nevertheless, despite the growing number of CJ studies, unsystematic and
-fragmented research approaches in the literature have left several
-critical issues unaddressed. This research primarily focuses on three:
-the apparent disconnect between CJ's measurement and structural model,
-the over-reliance on the assumptions of Thurstone's Case 5
-\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
-role of comparison algorithms on the method's reliability and validity.
-The following sections will discuss each of these issues in detail,
-followed by the introduction of a theoretical model and its statistical
-translation, which aims to address all three concerns simultaneously.
+Nevertheless, despite the increasing number of CJ studies, unsystematic
+and fragmented research approaches have left several critical issues
+unaddressed. This research primarily focuses on three: the over-reliance
+on Thurstone's Case V assumptions in the statistical analysis of CJ
+data, the apparent disconnect between CJ's trait measurement and
+hypothesis testing, and the unclear role and influence of comparison
+algorithms on the method's reliability and validity. The following
+sections will discuss each of these issues in detail, followed by the
+introduction of a theoretical model and its statistical translation,
+which aims to address all three concerns simultaneously.
 
@@
 
+\subsection{The Case V and the statistical analysis of CJ
+data}\label{the-case-v-and-the-statistical-analysis-of-cj-data}
+
+In its most general form, Thurstone's theory
+\citeyearpar{Thurstone_1927} posits that when a judge compares two
+stimuli, the resulting dichotomous outcome depends on two main factors:
+the discriminal process and the law of comparative judgment. The
+\emph{discriminal process} refers to the psychological effect each
+stimulus has on the judges, or more simply stated, the judges'
+perception of the trait level of each stimulus. Thurstone assumes that
+the discriminal process for each stimulus follows a Normal distribution.
+In this distribution, the mode (mean), known as the \emph{modal
+discriminal process}, represents the position of the stimulus on the
+trait continuum, while the dispersion, known as the \emph{discriminal
+dispersion}, reflects variability in the perceived trait level of the
+stimulus. Figure~\ref{fig-discriminal_process} shows a visual depiction
+of the discriminal process of two stimuli (objects).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}
+
+}
+
+\caption{\label{fig-discriminal_process}Discriminal processes of two
+stimuli (objects). Extracted from \citet[pp.~249]{Bramley_2008}}
+
+\end{figure}%
+
+However, since the discriminal mode and dispersion of a single stimulus
+are not directly observable without comparing it to other stimuli, the
+\emph{law of comparative judgment} becomes essential. This law asserts
+that when assessing a specific trait by comparing two stimuli, the
+stimulus positioned further along the continuum is perceived as having a
+higher level of that trait. Thus, the observed dichotomous outcome is
+determined by the distribution of the difference between the discriminal
+processes of the two stimuli, called the \emph{discriminal difference}.
+Figure~\ref{fig-comparative_judgment} shows a visual depiction of the
+discriminal difference of two stimuli (objects).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{./images/figures/comparative_judgment.png}
+
+}
+
+\caption{\label{fig-comparative_judgment}Discriminal difference between
+two stimuli (objects). Extracted from \citet[pp.~251]{Bramley_2008}}
+
+\end{figure}%
+
+Thurstone's general form of the theory primarily applies to pairwise
+comparisons of stimuli made by a single judge
+\citep[pp.~267]{Thurstone_1927}. Therefore, to enhance the theory's
+practical applicability, he developed five distinct cases, each
+incorporating a series of increasingly simplifying assumptions.
+Figure~\ref{fig-thurstone_cases} outlines each case along with their key
+assumptions. For a detailed explanation of this theoretical progression,
+see \citet{Thurstone_1927} and \citet[pp.~248-253]{Bramley_2008}.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}
+
+}
+
+\caption{\label{fig-thurstone_cases}Thurstones cases and asumptions}
+
+\end{figure}%
+
+Surprisingly, Case V is the most commonly used case in the CJ
+literature, despite incorporating the largest number of simplifying
+assumptions \citep[pp.~253]{Bramley_2008}. This widespread adoption is
+primarily due to the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959}, which offers a simpler
+statistical representation of the case. The BTL model closely aligns
+with Case V's assumptions, with one key difference: the distribution of
+the discriminal process. While Case V assumes a Normal distribution, the
+BTL model uses the more mathematically tractable logistic distribution
+\citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
+However, this adjustment does not significantly affect the model's
+estimation or interpretation, as the logistic and normal distributions
+are related to each other by a factor of approximately \(1.7\)
+\citep[pp.~16]{vanderLinden_et_al_2017_I}.
+
+However, because Case V was originally developed to provide a ``rather
+coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927}, prioritizing
+statistical ease of use over precision in trait measurement, two of its
+assumptions may be causing more problems than expected: the assumption
+of zero correlation between the stimuli, and the assumption of equal
+dispersion between stimuli.
+
 \subsection{The disconnect between structural and measurement
@@
 
-In a typical CJ study, the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
-for CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies
-how manifest variables contribute to the estimation of latent variables
+As outlined in the previous section, the BTL model typically functions
+as the measurement model in a CJ study
+\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
+manifest variables contribute to the estimation of latent variables
 \citep{Everitt_et_al_2010}. For example, when evaluating text quality,
@@
 influence the latent variable of interest \citep{Everitt_et_al_2010}.
-This approach allows analyses that can account for both the scores and
-their uncertainties simultaneously, rather than treating them as
+This approach allows analyses that can account for both the BTL scores
+and their uncertainties simultaneously, rather than treating them as
 separate elements. Therefore, an integrated approach that combines CJ's
 measurement and structural models can offer significant advantages.
-
-\subsection{The assumptions of Case 5 and the measurement
-model}\label{the-assumptions-of-case-5-and-the-measurement-model}
 

### 2024-11-15 index.qmd
section:issues; process:write

@@
 
-<!-- 1. What does CJ do? -->
-In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a].
+<!-- 1. Overview of CJ method in assessing traits -->
+In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b].
 Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level.
@@
 
-<!-- 2. Where are we now with DCJ? -->
+<!-- 2. What is the effectiveness of CJ -->
 Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade.
@@
 
-<!-- 3. But what are the problems of DCJ as a method? -->
+<!-- 3. But what critical issues remain CJ Research -->
 Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed.
-This research primarily focuses on three: the over-reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role and influence of comparison algorithms on the method's reliability and validity.
+This research primarily focuses on three: the over-reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role of comparison algorithms on the method's reliability and validity.
 The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.
@@
 # Three critical issues in CJ literature {#sec-theory-issues}
+
+<!-- 1. Thurstone's Theory of CJ: discriminal process -->
+In its most general form, Thurstone's theory [-@Thurstone_1927b] posits that the dichotomous outcome resulting from comparing two stimuli is determined by two factors: the discriminal process of each stimulus and their discriminal difference.
+The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus.
+Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution.
+In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus.
+@fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).
+
+![Example distributions of discriminal processes for two stimuli (objects). Extracted from @Bramley_2008 [pp. 249]. ](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
+
+<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
+However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential.
+This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait.
+Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*.
+@fig-comparative_judgment shows an example distribution of the discriminal difference for two stimuli (objects).
+
+![Distribution of the discriminal difference between two stimuli (objects). Extracted from @Bramley_2008 [pp. 251]. ](/images/figures/comparative_judgment.png){#fig-comparative_judgment width=65%}
+
+<!-- 3. Thurstone's Cases and their simplifying assumptions  -->
+Importantly, the theory's general form primarily addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
+267].
+Consequently, to enhance its practical applicability, Thurstone introduced five distinct cases, each defined by progressively simplifying assumptions.
+@fig-thurstone_cases summarizes these cases, focusing on key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersion across stimuli, the correlation between stimuli, and which judges perform the comparisons.
+For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
+248-253].
+
+![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#fig-thurstone_cases width=85%}
 
@@
 
-<!-- 1. What is the general from of the CJ theory? -->
-In its most general form, Thurstone's theory [-@Thurstone_1927] posits that when a judge compares two stimuli, the resulting dichotomous outcome depends on two main factors: the discriminal process and the law of comparative judgment.
-The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus.
-Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution.
-In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus.
-@fig-discriminal_process shows a visual depiction of the discriminal process of two stimuli (objects).
-
-![Discriminal processes of two stimuli (objects). Extracted from @Bramley_2008 [pp. 249] ](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
-
-However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential.
-This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait.
-Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*.
-@fig-comparative_judgment shows a visual depiction of the discriminal difference of two stimuli (objects).
-
-![Discriminal difference between two stimuli (objects). Extracted from @Bramley_2008 [pp. 251] ](/images/figures/comparative_judgment.png){#fig-comparative_judgment width=65%}
-
-<!-- 2. But several simplifying assumption are made to ease its use -->
-Thurstone’s general form of the theory primarily applies to pairwise comparisons of stimuli made by a single judge [@Thurstone_1927, pp.
-267].
-Therefore, to enhance the theory's practical applicability, he developed five distinct cases, each incorporating a series of increasingly simplifying assumptions.
-@fig-thurstone_cases outlines each case along with their key assumptions.
-For a detailed explanation of this theoretical progression, see @Thurstone_1927 and @Bramley_2008 [pp.
-248-253].
-
-![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#fig-thurstone_cases width=85%}
-
-<!-- 3. And yet Case V is is the most commonly used -->
-Surprisingly, Case V is the most commonly used case in the CJ literature, despite incorporating the largest number of simplifying assumptions [@Bramley_2008, pp.
-253].
-This widespread adoption is primarily due to the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959], which offers a simpler statistical representation of the case.
-The BTL model closely aligns with Case V’s assumptions, with one key difference: the distribution of the discriminal process.
-While Case V assumes a Normal distribution, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp.
+<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
+Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp.
+253], Case V remains the most widely used case in the CJ literature.
+This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp.
 254] (see @fig-thurstone_cases).
-However, this adjustment does not significantly affect the model’s estimation or interpretation, as the logistic and normal distributions are related to each other by a factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
-16].
+This substitution has little effect on the model's estimation or interpretation, as the Normal and logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
+16] (refer to @fig-logistic_vs_normal).
 
@@
 
-<!-- 4. But two assumptions of Case V might not hold -->
-However, because Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927, pp.
-269], prioritizing statistical ease of use over precision in trait measurement, two of its assumptions may be causing more problems than expected: the assumption of zero correlation between the stimuli, and the assumption of equal dispersion between stimuli.
+::: {#fig-logistic_vs_normal layout-ncol=2}
 
-<!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are two: -->
+![Probability density](/images/figures/density.png){#fig-density width=95%}
 
-<!--   (a) it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects.
+![Cummulative probability](/images/figures/cumulative.png){#fig-cumulative width=95%}
+
+Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255].
+:::
+
+<!-- 2. But what are limitations of Case V (and the BTL model) -->
+However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+269], prioritizing statistical simplicity over precision in trait measurement.
+As a result, its assumptions may not be suitable for applications beyond the psychophysical contexts, which typically involve simple, homogeneous stimuli.
+Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+270], further acknowledging that some assumptions could become problematic with less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp.
+374].
+Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretical or in practice: the assumption of zero correlation and equal dispersion between stimuli.
+
+<!-- 3. The problems of assuming zero correlation between stimuli -->
+[Regarding the assumption of *zero correlation between stimuli* ... ]{style="color:red;"}
+
+<!-- this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects.
 This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.
 -->
-
-<!--   Particularly on how CJ is used now, use the text example -->
 
@@
 
-<!--   Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
--->
-
-<!--   use the example of writing!!! -->
-
-<!--   (b) it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
+<!-- this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
 Meaning there is an amount of correlation that it is not accounted for.
--->
+example of Boonen -->
 
@@
 
-<!-- (c) discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
+<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
+268].
+-->
+
+<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
+-->
+
+<!--   use the example of writing!!! -->
+
+<!-- 4. The problems of assuming equal dispersion between stimuli -->
+
+<!-- it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
 Example of exaggerated comparison of individuals, university students vs secondary school students.
@@
 
-<!-- "It is probably true that this variability of the discriminal dispersions on the psychological continuum is of relatively less serious importance in dealing with strictly homogenous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogenous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927].
+<!-- "It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927a, pp.
+374].
 -->
@@
 
-## The disconnect between structural and measurement models
+## The disconnect between trait measurement and hypothesis testing
 
@@
 
-<!-- 1. The BTL model is CJ's measurement model -->
-As outlined in the previous section, the BTL model typically functions as the measurement model in a CJ study [@Andrich_1978; @Bramley_2008].
+<!-- 1. The BTL model as a measurement model for estimating latent traits -->
+Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait under assessment [@Andrich_1978; @Bramley_2008].
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
@@
 
-<!-- 2. But there is no structural model, only post estimation analyses -->
+<!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
 Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
@@
 
-<!-- 3. But what is the problem with this? -->
+<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
 However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests.
@@
 
-<!-- 4. So, what is the solution?  -->
+<!-- 4. So, how do we mitigate these risks?  -->
 To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
@@
 
-<!-- It is assumed that the single observer compares each pair of stimuli a sufficient number of times so that a proportion may be determined for eah pair [@Thurstone_1927]. -->
+<!-- It is assumed that the single observer compares each pair of stimuli a sufficient number of times so that a proportion may be determined for each pair [@Thurstone_1927b, pp. 267]. -->
 

### 2024-11-15 _tex/index.tex
Built site for gh-pages

@@
 trait or attribute across various stimuli by performing pairwise
-comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. Each
-comparison produces a dichotomous outcome, indicating which stimulus is
-perceived to exhibit a higher trait level. For example, when assessing
-text quality, judges compare pairs of written texts (the stimuli) to
+comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
+produces a dichotomous outcome, indicating which stimulus is perceived
+to exhibit a higher trait level. For example, when assessing text
+quality, judges compare pairs of written texts (the stimuli) to
 determine the relative quality each text exhibit (the trait)
@@
 data, the apparent disconnect between CJ's trait measurement and
-hypothesis testing, and the unclear role and influence of comparison
-algorithms on the method's reliability and validity. The following
-sections will discuss each of these issues in detail, followed by the
-introduction of a theoretical model and its statistical translation,
-which aims to address all three concerns simultaneously.
+hypothesis testing, and the unclear role of comparison algorithms on the
+method's reliability and validity. The following sections will discuss
+each of these issues in detail, followed by the introduction of a
+theoretical model and its statistical translation, which aims to address
+all three concerns simultaneously.
 
@@
 
-\subsection{The Case V and the statistical analysis of CJ
-data}\label{the-case-v-and-the-statistical-analysis-of-cj-data}
-
 In its most general form, Thurstone's theory
-\citeyearpar{Thurstone_1927} posits that when a judge compares two
-stimuli, the resulting dichotomous outcome depends on two main factors:
-the discriminal process and the law of comparative judgment. The
-\emph{discriminal process} refers to the psychological effect each
+\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
+resulting from comparing two stimuli is determined by two factors: the
+discriminal process of each stimulus and their discriminal difference.
+The \emph{discriminal process} refers to the psychological effect each
 stimulus has on the judges, or more simply stated, the judges'
@@
 dispersion}, reflects variability in the perceived trait level of the
-stimulus. Figure~\ref{fig-discriminal_process} shows a visual depiction
-of the discriminal process of two stimuli (objects).
+stimulus. Figure~\ref{fig-discriminal_process} shows example
+distributions of discriminal process for two stimuli (objects).
 
@@
 
-\caption{\label{fig-discriminal_process}Discriminal processes of two
-stimuli (objects). Extracted from \citet[pp.~249]{Bramley_2008}}
+\caption{\label{fig-discriminal_process}Example distributions of
+discriminal processes for two stimuli (objects). Extracted from
+\citet[pp.~249]{Bramley_2008}.}
 
@@
 processes of the two stimuli, called the \emph{discriminal difference}.
-Figure~\ref{fig-comparative_judgment} shows a visual depiction of the
-discriminal difference of two stimuli (objects).
+Figure~\ref{fig-comparative_judgment} shows an example distribution of
+the discriminal difference for two stimuli (objects).
 
@@
 
-\caption{\label{fig-comparative_judgment}Discriminal difference between
-two stimuli (objects). Extracted from \citet[pp.~251]{Bramley_2008}}
+\caption{\label{fig-comparative_judgment}Distribution of the discriminal
+difference between two stimuli (objects). Extracted from
+\citet[pp.~251]{Bramley_2008}.}
 
@@
 
-Thurstone's general form of the theory primarily applies to pairwise
+Importantly, the theory's general form primarily addresses pairwise
 comparisons of stimuli made by a single judge
-\citep[pp.~267]{Thurstone_1927}. Therefore, to enhance the theory's
-practical applicability, he developed five distinct cases, each
-incorporating a series of increasingly simplifying assumptions.
-Figure~\ref{fig-thurstone_cases} outlines each case along with their key
-assumptions. For a detailed explanation of this theoretical progression,
-see \citet{Thurstone_1927} and \citet[pp.~248-253]{Bramley_2008}.
+\citep[pp.~267]{Thurstone_1927b}. Consequently, to enhance its practical
+applicability, Thurstone introduced five distinct cases, each defined by
+progressively simplifying assumptions. Figure~\ref{fig-thurstone_cases}
+summarizes these cases, focusing on key assumptions such as the
+distribution of discriminal processes, the similarity of discriminal
+dispersion across stimuli, the correlation between stimuli, and which
+judges perform the comparisons. For a comprehensive discussion of this
+progression, refer to \citet{Thurstone_1927b} and
+\citet[pp.~248-253]{Bramley_2008}.
 
@@
 
-Surprisingly, Case V is the most commonly used case in the CJ
-literature, despite incorporating the largest number of simplifying
-assumptions \citep[pp.~253]{Bramley_2008}. This widespread adoption is
-primarily due to the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959}, which offers a simpler
-statistical representation of the case. The BTL model closely aligns
-with Case V's assumptions, with one key difference: the distribution of
-the discriminal process. While Case V assumes a Normal distribution, the
-BTL model uses the more mathematically tractable logistic distribution
+\subsection{The Case V and the statistical analysis of CJ
+data}\label{the-case-v-and-the-statistical-analysis-of-cj-data}
+
+Surprisingly, despite its reliance on the largest number of simplifying
+assumptions \citep[pp.~253]{Bramley_2008}, Case V remains the most
+widely used case in the CJ literature. This popularity is largely due to
+its simplified statistical representation in the Bradley-Terry-Luce
+(BTL) model \citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors
+Case V's assumptions, with one key difference: while Case V assumes a
+Normal distribution for the discriminal processes, the BTL model uses
+the more mathematically tractable logistic distribution
 \citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
-However, this adjustment does not significantly affect the model's
-estimation or interpretation, as the logistic and normal distributions
-are related to each other by a factor of approximately \(1.7\)
-\citep[pp.~16]{vanderLinden_et_al_2017_I}.
+This substitution has little effect on the model's estimation or
+interpretation, as the Normal and logistic distributions differ by a
+scaling factor of approximately \(1.7\)
+\citep[pp.~16]{vanderLinden_et_al_2017_I} (refer to
+Figure~\ref{fig-logistic_vs_normal}).
 
-However, because Case V was originally developed to provide a ``rather
-coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927}, prioritizing
-statistical ease of use over precision in trait measurement, two of its
-assumptions may be causing more problems than expected: the assumption
-of zero correlation between the stimuli, and the assumption of equal
-dispersion between stimuli.
+\begin{figure}
 
-\subsection{The disconnect between structural and measurement
-models}\label{the-disconnect-between-structural-and-measurement-models}
+\begin{minipage}{0.50\linewidth}
 
-As outlined in the previous section, the BTL model typically functions
-as the measurement model in a CJ study
-\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
-manifest variables contribute to the estimation of latent variables
-\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
-researchers use the BTL model to process the dichotomous outcomes
-resulting from the pairwise comparisons (the manifest variables) to
-estimate scores that reflect the underlying quality level of the texts
-(the latent variable)
+\centering{
+
+\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}
+
+}
+
+\subcaption{\label{fig-density}Probability density}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/cumulative.png}
+
+}
+
+\subcaption{\label{fig-cumulative}Cummulative probability}
+
+\end{minipage}%
+
+\caption{\label{fig-logistic_vs_normal}Probability density and
+cumulative probability of the logistic and Normal distributions.
+Extracted from \citet[pp.~254-255]{Bramley_2008}.}
+
+\end{figure}%
+
+However, Case V (and its BTL model counterpart) was originally developed
+to provide a ``rather coarse scaling'' of traits
+\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
+over precision in trait measurement. As a result, its assumptions may
+not be suitable for applications beyond the psychophysical contexts,
+which typically involve simple, homogeneous stimuli. Thurstone himself
+cautioned that its use ``should not be made without (an) experimental
+test'' \citep[pp.~270]{Thurstone_1927b}, further acknowledging that some
+assumptions could become problematic with less homogeneous stimuli, such
+as handwriting or English compositions \citep[pp.~374]{Thurstone_1927a}.
+Consequently, given that current CJ applications often involve more
+complex, less homogeneous stimuli, two assumptions of Case V may not
+always hold theoretical or in practice: the assumption of zero
+correlation and equal dispersion between stimuli.
+
+{Regarding the assumption of \emph{zero correlation between stimuli}
+\ldots{}}
+
+\subsection{The disconnect between trait measurement and hypothesis
+testing}\label{the-disconnect-between-trait-measurement-and-hypothesis-testing}
+
+Building on the previous section, it is evident that the BTL model
+typically functions as the measurement model for the trait under
+assessment \citep{Andrich_1978, Bramley_2008}. A measurement model
+specifies how manifest variables contribute to the estimation of latent
+variables \citep{Everitt_et_al_2010}. For example, when evaluating text
+quality, researchers use the BTL model to process the dichotomous
+outcomes resulting from the pairwise comparisons (the manifest
+variables) to estimate scores that reflect the underlying quality level
+of the texts (the latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

### 2024-11-15 index.qmd
section:issues; process:write

@@
 This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp.
+The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Bramley_2008, pp.
 254] (see @fig-thurstone_cases).
-This substitution has little effect on the model's estimation or interpretation, as the Normal and logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
+This substitution has little effect on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
 16] (refer to @fig-logistic_vs_normal).
@@
 374].
-Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretical or in practice: the assumption of zero correlation and equal dispersion between stimuli.
+Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretically or in practice: the assumption of zero correlation and equal dispersion between stimuli.
 

### 2024-11-15 _tex/index.tex
Built site for gh-pages

@@
 Normal distribution for the discriminal processes, the BTL model uses
-the more mathematically tractable logistic distribution
+the more mathematically tractable Logistic distribution
 \citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
 This substitution has little effect on the model's estimation or
-interpretation, as the Normal and logistic distributions differ by a
+interpretation, as the Normal and Logistic distributions differ by a
 scaling factor of approximately \(1.7\)
@@
 complex, less homogeneous stimuli, two assumptions of Case V may not
-always hold theoretical or in practice: the assumption of zero
+always hold theoretically or in practice: the assumption of zero
 correlation and equal dispersion between stimuli.

### 2024-11-16 index.qmd
section:issues; process:write

@@
 Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution.
-In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus.
+In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the stimulus' perceived trait level.
 @fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).
@@
 <!-- 2. Thurstone's Theory of CJ: discriminal difference -->
-However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential.
+However, since the discriminal mode and dispersion of a single stimulus are not directly observable except through comparison, the *law of comparative judgment* becomes essential.
 This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait.
-Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*.
+Thus, the observed dichotomous outcome is determined by the distribution of the difference between the stimuli's discriminal processes, called the *discriminal difference*.
 @fig-comparative_judgment shows an example distribution of the discriminal difference for two stimuli (objects).
@@
 Consequently, to enhance its practical applicability, Thurstone introduced five distinct cases, each defined by progressively simplifying assumptions.
-@fig-thurstone_cases summarizes these cases, focusing on key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersion across stimuli, the correlation between stimuli, and which judges perform the comparisons.
+@tbl-thurstone_cases summarizes these cases, focusing on key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersions across stimuli, the correlation between stimuli, and which judges perform the comparisons.
 For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
@@
 
-![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#fig-thurstone_cases width=85%}
+![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}
 
@@
 This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Bramley_2008, pp.
-254] (see @fig-thurstone_cases).
+The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Bramley_2008, pp.
+254] (see @tbl-thurstone_cases).
 This substitution has little effect on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
@@
 269], prioritizing statistical simplicity over precision in trait measurement.
-As a result, its assumptions may not be suitable for applications beyond the psychophysical contexts, which typically involve simple, homogeneous stimuli.
+As a result, its assumptions may not suit applications beyond the psychophysical contexts for which it was created.
 Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], further acknowledging that some assumptions could become problematic with less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp.
+270], acknowledging that some assumptions could prove problematic with more complex, less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp.
 374].
-Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretically or in practice: the assumption of zero correlation and equal dispersion between stimuli.
+Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
 
 <!-- 3. The problems of assuming zero correlation between stimuli -->
-[Regarding the assumption of *zero correlation between stimuli* ... ]{style="color:red;"}
+[The *zero correlation between stimuli* represents two additional assumtions about the method:   ]{style="color:red;"}
 
@@
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait under assessment [@Andrich_1978; @Bramley_2008].
+Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait of interest [@Andrich_1978; @Bramley_2008].
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
-For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying texts' quality level (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 

### 2024-11-16 _tex/index.tex
Built site for gh-pages

@@
 trait continuum, while the dispersion, known as the \emph{discriminal
-dispersion}, reflects variability in the perceived trait level of the
-stimulus. Figure~\ref{fig-discriminal_process} shows example
-distributions of discriminal process for two stimuli (objects).
+dispersion}, reflects variability in the stimulus' perceived trait
+level. Figure~\ref{fig-discriminal_process} shows example distributions
+of discriminal process for two stimuli (objects).
 
@@
 However, since the discriminal mode and dispersion of a single stimulus
-are not directly observable without comparing it to other stimuli, the
-\emph{law of comparative judgment} becomes essential. This law asserts
-that when assessing a specific trait by comparing two stimuli, the
-stimulus positioned further along the continuum is perceived as having a
-higher level of that trait. Thus, the observed dichotomous outcome is
-determined by the distribution of the difference between the discriminal
-processes of the two stimuli, called the \emph{discriminal difference}.
+are not directly observable except through comparison, the \emph{law of
+comparative judgment} becomes essential. This law asserts that when
+assessing a specific trait by comparing two stimuli, the stimulus
+positioned further along the continuum is perceived as having a higher
+level of that trait. Thus, the observed dichotomous outcome is
+determined by the distribution of the difference between the stimuli's
+discriminal processes, called the \emph{discriminal difference}.
 Figure~\ref{fig-comparative_judgment} shows an example distribution of
@@
 applicability, Thurstone introduced five distinct cases, each defined by
-progressively simplifying assumptions. Figure~\ref{fig-thurstone_cases}
+progressively simplifying assumptions. Table~\ref{tbl-thurstone_cases}
 summarizes these cases, focusing on key assumptions such as the
 distribution of discriminal processes, the similarity of discriminal
-dispersion across stimuli, the correlation between stimuli, and which
+dispersions across stimuli, the correlation between stimuli, and which
 judges perform the comparisons. For a comprehensive discussion of this
@@
 
-\begin{figure}
+\begin{table}
+
+\caption{\label{tbl-thurstone_cases}Thurstones cases and asumptions}
 
@@
 
-\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}
 
@@
 
-\caption{\label{fig-thurstone_cases}Thurstones cases and asumptions}
-
-\end{figure}%
+\end{table}%
 
@@
 Case V's assumptions, with one key difference: while Case V assumes a
-Normal distribution for the discriminal processes, the BTL model uses
-the more mathematically tractable Logistic distribution
-\citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
+Normal distribution for the stimuli's discriminal processes, the BTL
+model uses the more mathematically tractable Logistic distribution
+\citep[pp.~254]{Bramley_2008} (see Table~\ref{tbl-thurstone_cases}).
 This substitution has little effect on the model's estimation or
@@
 over precision in trait measurement. As a result, its assumptions may
-not be suitable for applications beyond the psychophysical contexts,
-which typically involve simple, homogeneous stimuli. Thurstone himself
-cautioned that its use ``should not be made without (an) experimental
-test'' \citep[pp.~270]{Thurstone_1927b}, further acknowledging that some
-assumptions could become problematic with less homogeneous stimuli, such
-as handwriting or English compositions \citep[pp.~374]{Thurstone_1927a}.
-Consequently, given that current CJ applications often involve more
-complex, less homogeneous stimuli, two assumptions of Case V may not
-always hold theoretically or in practice: the assumption of zero
-correlation and equal dispersion between stimuli.
+not suit applications beyond the psychophysical contexts for which it
+was created. Thurstone himself cautioned that its use ``should not be
+made without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b},
+acknowledging that some assumptions could prove problematic with more
+complex, less homogeneous stimuli, such as handwriting or English
+compositions \citep[pp.~374]{Thurstone_1927a}. Consequently, given that
+current CJ applications often deal with complex, less homogeneous
+stimuli, two key assumptions of Case V may not consistently hold in
+theory or practice: the zero correlation and equal dispersion between
+stimuli.
 
-{Regarding the assumption of \emph{zero correlation between stimuli}
-\ldots{}}
+{The \emph{zero correlation between stimuli} represents two additional
+assumtions about the method:}
 
@@
 Building on the previous section, it is evident that the BTL model
-typically functions as the measurement model for the trait under
-assessment \citep{Andrich_1978, Bramley_2008}. A measurement model
-specifies how manifest variables contribute to the estimation of latent
-variables \citep{Everitt_et_al_2010}. For example, when evaluating text
-quality, researchers use the BTL model to process the dichotomous
-outcomes resulting from the pairwise comparisons (the manifest
-variables) to estimate scores that reflect the underlying quality level
-of the texts (the latent variable)
+typically functions as the measurement model for the trait of interest
+\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
+manifest variables contribute to the estimation of latent variables
+\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
+researchers use the BTL model to process the dichotomous outcomes
+resulting from the pairwise comparisons (the manifest variables) to
+estimate scores that reflect the underlying texts' quality level (the
+latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

### 2024-11-16 index.qmd
section:issues; process:write

@@
 <!-- and spoken language [@Boonen_et_al_2020].
-Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_2015], engineering design skills [@Bartholomew_et_al_2018], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020].
+Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_et_al_2015], engineering design skills [@Bartholomew_et_al_2018], build real-time web-based portfolios of performance [@@Kimbell_2012], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020].
 -->
@@
 On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
-Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 <!-- 3. The problems of assuming zero correlation between stimuli -->
-[The *zero correlation between stimuli* represents two additional assumtions about the method:   ]{style="color:red;"}
+[The *zero correlation between stimuli* represents two additional assumptions about the method:   ]{style="color:red;"}
 
@@
 (Consider the problem of using adhoc methods) -->
+
+<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_2020; @vanDaal_et_al_2017] -->
 

### 2024-11-16 _tex/index.tex
Built site for gh-pages

@@
 versatility across both educational and non-educational contexts
-\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
+\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
 
@@
 {The \emph{zero correlation between stimuli} represents two additional
-assumtions about the method:}
+assumptions about the method:}
 

### 2024-11-16 index.qmd
section:issues; process:write

@@
 Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp.
-253], Case V remains the most widely used case in the CJ literature.
+253; @Kelly_et_al_2022, pp.
+677], Case V remains the most widely used case in the CJ literature.
 This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Bramley_2008, pp.
+The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
 254] (see @tbl-thurstone_cases).
@@
 However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
-269], prioritizing statistical simplicity over precision in trait measurement.
+269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
+677].
 As a result, its assumptions may not suit applications beyond the psychophysical contexts for which it was created.
@@
 Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
+
+<!-- Adding assumptions simplifies the formula and improves its tractability, but one cannot place as much confidence in the outputs of the formula if those assumptions are not met [@Kelly_et_al_2022, pp.
+677].
+-->
 
@@
 <!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_2020; @vanDaal_et_al_2017] -->
+
+<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
+Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
+[@Kelly_et_al_2022, pp.
+683] -->
+
+<!-- Such differences may not be detected through analyses of bias and misfit.
+McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
+Where students prioritised factual recall, teachers prioritised scientific explanation.
+Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
+Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
+[@Kelly_et_al_2022, pp.
+683] -->
 
@@
 <!-- @Bramley_et_al_2019 and @Bramley_2015 says adaptivity inflates reliability.
-present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required.
+present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required or how many judges should be involved [@Kelly_et_al_2022, pp.
+682].
 Adhoc rules of thumb.
@@
 
+<!-- "If Thurstone’s Law is to be used in support of comparative judgment for -->
+<!-- assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes." [@Kelly_et_al_2022, pp. 678] -->
+
 <!-- Pollitt (2012a) argues that assessors in pairwise comparisons need less training, given intuitive comparison and decision making.
@@
 
-<!-- Due to efficiency of scoring, most of research only colects only one text.
+<!-- Due to efficiency of scoring, most of research only collects only one text.
 This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
@@
 
+<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
+If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
+...
+It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
+681] -->
+
+<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
+
 ## Limitations and further research {#sec-discuss-limitations}
+
+<!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need?
+Who should judge?
+How many judgments should they make?
+And crucially, how can we be confident that they are basing their judgments on acceptable criteria?
+If judges are not using construct-relevant criteria, then the argument that comparative judgments are necessarily valid cannot be upheld.
+We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used.
+However, work to derive some general principles for good practice would be valuable.[@Kelly_et_al_2022, pp.
+683] -->
 

### 2024-11-16 _tex/index.tex
Built site for gh-pages

@@
 Surprisingly, despite its reliance on the largest number of simplifying
-assumptions \citep[pp.~253]{Bramley_2008}, Case V remains the most
-widely used case in the CJ literature. This popularity is largely due to
-its simplified statistical representation in the Bradley-Terry-Luce
-(BTL) model \citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors
-Case V's assumptions, with one key difference: while Case V assumes a
-Normal distribution for the stimuli's discriminal processes, the BTL
-model uses the more mathematically tractable Logistic distribution
-\citep[pp.~254]{Bramley_2008} (see Table~\ref{tbl-thurstone_cases}).
-This substitution has little effect on the model's estimation or
-interpretation, as the Normal and Logistic distributions differ by a
-scaling factor of approximately \(1.7\)
+assumptions
+\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
+Case V remains the most widely used case in the CJ literature. This
+popularity is largely due to its simplified statistical representation
+in the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors Case V's
+assumptions, with one key difference: while Case V assumes a Normal
+distribution for the stimuli's discriminal processes, the BTL model uses
+the more mathematically tractable Logistic distribution
+\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
+Table~\ref{tbl-thurstone_cases}). This substitution has little effect on
+the model's estimation or interpretation, as the Normal and Logistic
+distributions differ by a scaling factor of approximately \(1.7\)
 \citep[pp.~16]{vanderLinden_et_al_2017_I} (refer to
@@
 \citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
-over precision in trait measurement. As a result, its assumptions may
-not suit applications beyond the psychophysical contexts for which it
-was created. Thurstone himself cautioned that its use ``should not be
-made without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b},
-acknowledging that some assumptions could prove problematic with more
-complex, less homogeneous stimuli, such as handwriting or English
-compositions \citep[pp.~374]{Thurstone_1927a}. Consequently, given that
-current CJ applications often deal with complex, less homogeneous
-stimuli, two key assumptions of Case V may not consistently hold in
-theory or practice: the zero correlation and equal dispersion between
-stimuli.
+over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
+As a result, its assumptions may not suit applications beyond the
+psychophysical contexts for which it was created. Thurstone himself
+cautioned that its use ``should not be made without (an) experimental
+test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging that some
+assumptions could prove problematic with more complex, less homogeneous
+stimuli, such as handwriting or English compositions
+\citep[pp.~374]{Thurstone_1927a}. Consequently, given that current CJ
+applications often deal with complex, less homogeneous stimuli, two key
+assumptions of Case V may not consistently hold in theory or practice:
+the zero correlation and equal dispersion between stimuli.
 

### 2024-11-18 index.qmd
section:issues; process:write

@@
 Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level.
-For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
+On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
 Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
@@
 Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution.
-In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the stimulus' perceived trait level.
+In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects the variability in the stimulus' perceived trait level.
 @fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).
@@
 677].
-As a result, its assumptions may not suit applications beyond the psychophysical contexts for which it was created.
+As a result, its assumptions may not suit applications beyond the psycho-physical contexts for which it was created.
 Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
@@
 
-<!-- 3. The problems of assuming zero correlation between stimuli -->
-[The *zero correlation between stimuli* represents two additional assumptions about the method:   ]{style="color:red;"}
+<!-- 3. What does the zero correlation between stimuli imply? -->
+The assumption of *zero correlation between stimuli* is best illustrated with an example.
+For instance, when evaluating text quality, this assumption suggests that a judge’s perception of quality in one text does not influence their perception of the same trait in the comparison text.
+Thurstone attributes this independence to the cancellation of potential judges' biases, driven by opposing and equally weighted factors; commonly known as 'mood' and 'simultaneous contrast' effects, that operate during pairwise comparisons [@Thurstone_1927b, pp.
+268].
+This cancellation of bias has been mathematically demonstrated by @Andrich_1978, assuming additive biases and a logit scale derived from the BTL model.
 
-<!-- this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects.
-This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.
--->
+<!-- 4. But what are the problems of assuming zero correlation? -->
+However, given that evidence has shown judges' quality assessments are influenced by multiple, complex aspects of the texts [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], it is plausible that the accuracy of these judgments also varies based on factors such as age, culture, education, expertise [@Kelly_et_al_2022, pp.
+683], and ultimately, based on the individual judge themselves [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020].
+
+[This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder ]{style="color:red;"}
+
+Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].
+
+<!-- Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women? [@Kelly_et_al_2022, pp. 683] -->
+
+<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020] -->
 
@@
 
-<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_2020; @vanDaal_et_al_2017] -->
-
-<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
-Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
-[@Kelly_et_al_2022, pp.
-683] -->
-
 <!-- Such differences may not be detected through analyses of bias and misfit.
@@
 Meaning there is an amount of correlation that it is not accounted for.
-example of Boonen -->
+Use example of Boonen!!
+-->
 
@@
 -->
-
-<!--   use the example of writing!!! -->
 
@@
 
+<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
+Thurstone clearly did not expect this to hold for any but the most simple stimuli.
+The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naïve to expect this assumption to hold here [@Bramley_2008].
+-->
+
 ## The disconnect between trait measurement and hypothesis testing
@@
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait of interest [@Andrich_1978; @Bramley_2008].
+Building on the previous section, it is evident that the BTL model commonly functions as the measurement model for the trait of interest in CJ experiments [@Andrich_1978; @Bramley_2008].
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
-For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying texts' quality level (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 
-<!-- []{style="color:red;"}. -->
+<!-- []{style="color:red;"} -->
 

### 2024-11-18 _tex/index.tex
Built site for gh-pages

@@
 determine the relative quality each text exhibit (the trait)
-\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
+\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
 
@@
 CJ can accurately represent the traits under measurement
-\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
 Finally, research on practical applicability highlights the method's
@@
 trait continuum, while the dispersion, known as the \emph{discriminal
-dispersion}, reflects variability in the stimulus' perceived trait
+dispersion}, reflects the variability in the stimulus' perceived trait
 level. Figure~\ref{fig-discriminal_process} shows example distributions
@@
 As a result, its assumptions may not suit applications beyond the
-psychophysical contexts for which it was created. Thurstone himself
+psycho-physical contexts for which it was created. Thurstone himself
 cautioned that its use ``should not be made without (an) experimental
@@
 
-{The \emph{zero correlation between stimuli} represents two additional
-assumptions about the method:}
+The assumption of \emph{zero correlation between stimuli} is best
+illustrated with an example. For instance, when evaluating text quality,
+this assumption suggests that a judge's perception of quality in one
+text does not influence their perception of the same trait in the
+comparison text. Thurstone attributes this independence to the
+cancellation of potential judges' biases, driven by opposing and equally
+weighted factors; commonly known as `mood' and `simultaneous contrast'
+effects, that operate during pairwise comparisons
+\citep[pp.~268]{Thurstone_1927b}. This cancellation of bias has been
+mathematically demonstrated by \citet{Andrich_1978}, assuming additive
+biases and a logit scale derived from the BTL model.
+
+However, given that evidence has shown judges' quality assessments are
+influenced by multiple, complex aspects of the texts
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}, it is
+plausible that the accuracy of these judgments also varies based on
+factors such as age, culture, education, expertise
+\citep[pp.~683]{Kelly_et_al_2022}, and ultimately, based on the
+individual judge themselves
+\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020}.
+
+{This means that the cancellation of judges' biases may not hold in
+theory or practice, and that judges' biases may depend on the stimuli,
+by means of an interaction (but present) or because the judges works as
+a confounder}
+
+Such differences may not be detected through analyses of bias and misfit
+\citep[pp.~683]{Kelly_et_al_2022}.
 
@@
 Building on the previous section, it is evident that the BTL model
-typically functions as the measurement model for the trait of interest
-\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
-manifest variables contribute to the estimation of latent variables
-\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
-researchers use the BTL model to process the dichotomous outcomes
-resulting from the pairwise comparisons (the manifest variables) to
-estimate scores that reflect the underlying texts' quality level (the
-latent variable)
-\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
+commonly functions as the measurement model for the trait of interest in
+CJ experiments \citep{Andrich_1978, Bramley_2008}. A measurement model
+specifies how manifest variables contribute to the estimation of latent
+variables \citep{Everitt_et_al_2010}. For example, when evaluating text
+quality, researchers use the BTL model to process the dichotomous
+outcomes resulting from the pairwise comparisons (the manifest
+variables) to estimate scores that reflect the underlying quality level
+of texts (the latent variable)
+\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
 

### 2024-11-18 index.qmd
section:issues; process:write

@@
 <!-- 2. But what are limitations of Case V (and the BTL model) -->
-However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+However, Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
@@
 As a result, its assumptions may not suit applications beyond the psycho-physical contexts for which it was created.
-Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions could prove problematic with more complex, less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp.
+Thurstone himself cautioned that its use "should not be made without experimental test" [@Thurstone_1927b, pp.
+270], acknowledging that some assumptions could prove problematic with complex traits or less homogeneous stimuli [@Thurstone_1927a, pp.
 374].
-Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
+Consequently, given that current CJ applications often deal these types of traits and stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
 
@@
 The assumption of *zero correlation between stimuli* is best illustrated with an example.
-For instance, when evaluating text quality, this assumption suggests that a judge’s perception of quality in one text does not influence their perception of the same trait in the comparison text.
-Thurstone attributes this independence to the cancellation of potential judges' biases, driven by opposing and equally weighted factors; commonly known as 'mood' and 'simultaneous contrast' effects, that operate during pairwise comparisons [@Thurstone_1927b, pp.
+For instance, when evaluating text quality, the assumption suggests that a judge’s perception of quality in one text does not influence the perception of the same trait in the comparison text.
+Thurstone attributes this independence to the cancellation of potential judges' biases, driven by opposing and equally weighted factors that operate during pairwise comparisons, called 'mood' and 'simultaneous contrast' effects [@Thurstone_1927b, pp.
 268].
-This cancellation of bias has been mathematically demonstrated by @Andrich_1978, assuming additive biases and a logit scale derived from the BTL model.
+This cancellation of bias has been mathematically demonstrated by @Andrich_1978, assuming discriminal processes with additive biases and a logit scale derived from the BTL model.
 
 <!-- 4. But what are the problems of assuming zero correlation? -->
-However, given that evidence has shown judges' quality assessments are influenced by multiple, complex aspects of the texts [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], it is plausible that the accuracy of these judgments also varies based on factors such as age, culture, education, expertise [@Kelly_et_al_2022, pp.
-683], and ultimately, based on the individual judge themselves [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020].
+However, two types of scenarios make it plausible that potential judges' biases may not cancel: those involving complex traits and less homogeneous stimuli, and those where judges differ in what they value in their assessments.
+In the first scenario, Thurstone noted that for complex traits and non-homogeneous stimuli, such as handwriting or English compositions, CJ data might not align with the assumptions of Case V [@Thurstone_1927a, pp.
+374].
+This insight likely extends to other similarly complex traits and stimuli.
+In the second scenario, evidence indicates that judges' assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+Moreover, factors like age, culture, education, expertise [@Kelly_et_al_2022, pp.
+683], and even individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020] can influence judgment accuracy.
+These scenarios can ultimately result in non-additive biases that resist cancellation, driven by the characteristics or location of stimuli within the trait continuum.
 
-[This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder ]{style="color:red;"}
+<!-- seen as a strength of the method [@vanDaal_et_al_2016], and commonly referred to as "the wisdom of crowds" [@Surowiecki_2004] -->
 
-Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].
+<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->
 
@@
 
-<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020] -->
-
-<!-- What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
-This, in the best scenario, cause less precise estimates, and in the worst scenario, biased estimates.
-possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
-
-<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind.
-This is studied by analyzing the residuals that remain after the model is fitted to the data.
-Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large.
-Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability.
-indicates that one judge – Judge 5 – is significantly biased in some way.
-(Consider the problem of using adhoc methods) -->
-
-<!-- Such differences may not be detected through analyses of bias and misfit.
-McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
-Where students prioritised factual recall, teachers prioritised scientific explanation.
-Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
-Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
-[@Kelly_et_al_2022, pp.
-683] -->
-
-<!-- this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
+<!-- 5. But there is another problem associated with the zero correlation that no one is talking -->
+[This also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
 Meaning there is an amount of correlation that it is not accounted for.
 Use example of Boonen!!
--->
+]{style="color:red;"}
 
@@
 
-<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
--->
+Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].
+
+<!-- Such differences may not be detected through analyses of bias and misfit.
+McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
+Where students prioritised factual recall, teachers prioritised scientific explanation.
+Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
+Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
+[@Kelly_et_al_2022, pp.
+683] -->
+<!-- What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
+This, in the best scenario, cause less precise estimates, and in the worst scenario, biased estimates.
+possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
+
+<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind.
+This is studied by analyzing the residuals that remain after the model is fitted to the data.
+Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large.
+Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability.
+indicates that one judge – Judge 5 – is significantly biased in some way.
+(Consider the problem of using adhoc methods) -->
 
@@
 The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naïve to expect this assumption to hold here [@Bramley_2008].
+-->
+
+<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
 -->

### 2024-11-18 _tex/index.tex
Built site for gh-pages

@@
 
-However, Case V (and its BTL model counterpart) was originally developed
-to provide a ``rather coarse scaling'' of traits
-\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
-over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
-As a result, its assumptions may not suit applications beyond the
-psycho-physical contexts for which it was created. Thurstone himself
-cautioned that its use ``should not be made without (an) experimental
-test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging that some
-assumptions could prove problematic with more complex, less homogeneous
-stimuli, such as handwriting or English compositions
-\citep[pp.~374]{Thurstone_1927a}. Consequently, given that current CJ
-applications often deal with complex, less homogeneous stimuli, two key
-assumptions of Case V may not consistently hold in theory or practice:
-the zero correlation and equal dispersion between stimuli.
+However, Case V was originally developed to provide a ``rather coarse
+scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
+statistical simplicity over precision in trait measurement
+\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
+suit applications beyond the psycho-physical contexts for which it was
+created. Thurstone himself cautioned that its use ``should not be made
+without experimental test'' \citep[pp.~270]{Thurstone_1927b},
+acknowledging that some assumptions could prove problematic with complex
+traits or less homogeneous stimuli \citep[pp.~374]{Thurstone_1927a}.
+Consequently, given that current CJ applications often deal these types
+of traits and stimuli, two key assumptions of Case V may not
+consistently hold in theory or practice: the zero correlation and equal
+dispersion between stimuli.
 
@@
 illustrated with an example. For instance, when evaluating text quality,
-this assumption suggests that a judge's perception of quality in one
-text does not influence their perception of the same trait in the
-comparison text. Thurstone attributes this independence to the
-cancellation of potential judges' biases, driven by opposing and equally
-weighted factors; commonly known as `mood' and `simultaneous contrast'
-effects, that operate during pairwise comparisons
-\citep[pp.~268]{Thurstone_1927b}. This cancellation of bias has been
-mathematically demonstrated by \citet{Andrich_1978}, assuming additive
+the assumption suggests that a judge's perception of quality in one text
+does not influence the perception of the same trait in the comparison
+text. Thurstone attributes this independence to the cancellation of
+potential judges' biases, driven by opposing and equally weighted
+factors that operate during pairwise comparisons, called `mood' and
+`simultaneous contrast' effects \citep[pp.~268]{Thurstone_1927b}. This
+cancellation of bias has been mathematically demonstrated by
+\citet{Andrich_1978}, assuming discriminal processes with additive
 biases and a logit scale derived from the BTL model.
 
-However, given that evidence has shown judges' quality assessments are
-influenced by multiple, complex aspects of the texts
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}, it is
-plausible that the accuracy of these judgments also varies based on
-factors such as age, culture, education, expertise
-\citep[pp.~683]{Kelly_et_al_2022}, and ultimately, based on the
-individual judge themselves
-\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020}.
+However, two types of scenarios make it plausible that potential judges'
+biases may not cancel: those involving complex traits and less
+homogeneous stimuli, and those where judges differ in what they value in
+their assessments. In the first scenario, Thurstone noted that for
+complex traits and non-homogeneous stimuli, such as handwriting or
+English compositions, CJ data might not align with the assumptions of
+Case V \citep[pp.~374]{Thurstone_1927a}. This insight likely extends to
+other similarly complex traits and stimuli. In the second scenario,
+evidence indicates that judges' assessments are influenced by multiple,
+intricate aspects of the stimuli
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
+Moreover, factors like age, culture, education, expertise
+\citep[pp.~683]{Kelly_et_al_2022}, and even individual differences among
+judges \citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020} can
+influence judgment accuracy. These scenarios can ultimately result in
+non-additive biases that resist cancellation, driven by the
+characteristics or location of stimuli within the trait continuum.
 
-{This means that the cancellation of judges' biases may not hold in
-theory or practice, and that judges' biases may depend on the stimuli,
-by means of an interaction (but present) or because the judges works as
-a confounder}
+{This also translates into the idea that stimuli are the main focus of
+estimation and analysis, but what happens when the focus of analysis is
+the individuals that generated those stimuli. Meaning there is an amount
+of correlation that it is not accounted for. Use example of Boonen!!}
 

### 2024-11-21 index.qmd
section:issues; process:write

--- 2024-11-18 index.qmd
@@
 Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
-Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Meanwhile, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023], while research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 
-<!-- 3. But what critical issues remain CJ Research -->
+<!-- 3. But what critical issues remain in CJ Research?, what is the organization of the study? -->
 Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed.
-This research primarily focuses on three: the over-reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role of comparison algorithms on the method's reliability and validity.
-The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.
+The present study primarily focuses on three: the over-reliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role of the diverse assessment design features on CJ's reliability and validity.
+The following sections begin with a brief overview of Thurstone's theory and a detailed discussion of these issues.
+Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address all three concerns simultaneously.
 
-# Three critical issues in CJ literature {#sec-theory-issues}
+# Thurstone's theory {#sec-thurstone_theory}
+<!-- recording: Sven and Tine 24.11.18; time: 00:04:30 - 00:14:50 -->
+<!-- recording: Sven and Tine 24.11.18; time: 00:19:04 - 00:21:37 -->
 
 <!-- 1. Thurstone's Theory of CJ: discriminal process -->
-In its most general form, Thurstone's theory [-@Thurstone_1927b] posits that the dichotomous outcome resulting from comparing two stimuli is determined by two factors: the discriminal process of each stimulus and their discriminal difference.
-The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus.
-Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution.
-In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects the variability in the stimulus' perceived trait level.
-@fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).
-
-![Example distributions of discriminal processes for two stimuli (objects). Extracted from @Bramley_2008 [pp. 249]. ](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
+In its most general form, Thurstone's theory [-@Thurstone_1927b] suggests that two factors determine the dichotomous outcome of pairwise comparisons: the discriminal process of each stimulus and their discriminal difference.
+The *discriminal process* refers to the psychological effect each stimulus exerts on the judges, or more simply, the underlying perception of the stimulus' trait level.
+According to the theory, the discriminal process for each stimulus follows a Normal distribution.
+The mode (mean) of this distribution, referred to as the *modal discriminal process*, represents the stimulus' position on the trait continuum.
+Meanwhile, the dispersion of the distribution, referred to as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
 
 <!-- 2. Thurstone's Theory of CJ: discriminal difference -->
-However, since the discriminal mode and dispersion of a single stimulus are not directly observable except through comparison, the *law of comparative judgment* becomes essential.
-This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait.
-Thus, the observed dichotomous outcome is determined by the distribution of the difference between the stimuli's discriminal processes, called the *discriminal difference*.
-@fig-comparative_judgment shows an example distribution of the discriminal difference for two stimuli (objects).
+However, since the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential.
+This law states that in pairwise comparisons, the stimulus positioned further along the trait continuum is perceived as having a higher level of that trait.
+Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.
+This suggests that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.
 
-![Distribution of the discriminal difference between two stimuli (objects). Extracted from @Bramley_2008 [pp. 251]. ](/images/figures/comparative_judgment.png){#fig-comparative_judgment width=65%}
+<!-- 3. Example-->
+These concepts can be illustrated through an example.
+For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum.
+Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area.
+Consequently, the dichotomous outcome of this comparison would favor text A.
 
-<!-- 3. Thurstone's Cases and their simplifying assumptions  -->
-Importantly, the theory's general form primarily addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
+::: {#fig-thurstone_theory layout-ncol=2}
+
+![Discriminal processes](/images/figures/discriminal_process.png){#fig-discriminal_process width=95%}
+
+![Discriminal difference](/images/figures/comparative_judgment.png){#fig-discriminal_difference width=100%}
+
+Example distribution of discriminal processes and their discriminal difference for two written texts (stimuli or objects). Extracted from @Bramley_2008 [pp. 249-251].
+:::
+
+<!-- 4. Thurstone's Cases and their simplifying assumptions  -->
+Importantly, the general form of Thurstone's theory primarily addressed pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
 267].
-Consequently, to enhance its practical applicability, Thurstone introduced five distinct cases, each defined by progressively simplifying assumptions.
-@tbl-thurstone_cases summarizes these cases, focusing on key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersions across stimuli, the correlation between stimuli, and which judges perform the comparisons.
+Thus, for practical application, Thurstone introduced five distinct cases derived from this general form, each defined by progressively simplifying assumptions.
+@tbl-thurstone_cases summarizes these cases, highlighting key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersions across stimuli, the correlation between stimuli, and the number of judges performing the comparisons.
 For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
@@
 
-## The Case V and the statistical analysis of CJ data
+# Three critical issues in CJ literature {#sec-theory-issues}
+
+## The Case V and the statistical analysis of CJ data {#sec-theory-issue1}
 
@@
 <!-- recording: Sven and Tine 24.11.12; time: 00:00:00 - 00:00:00 -->
+<!-- recording: Sven and Tine 24.11.18; time: 00:00:25 - 00:04:30 -->
+<!-- recording: Sven and Tine 24.11.18; time: 00:14:50 - 00:19:04 -->
+<!-- recording: Sven and Tine 24.11.18; time: 00:21:37 - 00:47:20 -->
 
 <!-- 1. The Case V and its simpler statistical representation: the BTL model -->
-Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp.
+Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp.
 253; @Kelly_et_al_2022, pp.
@@
 This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
+The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
 254] (see @tbl-thurstone_cases).
-This substitution has little effect on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
-16] (refer to @fig-logistic_vs_normal).
+This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
+16] (see @fig-logistic_vs_normal).
 
@@
 
-![Probability density](/images/figures/density.png){#fig-density width=95%}
+![Probability density](/images/figures/density.png){#fig-density width=85%}
 
-![Cummulative probability](/images/figures/cumulative.png){#fig-cumulative width=95%}
+![Cummulative probability](/images/figures/cumulative.png){#fig-cumulative width=85%}
 
@@
 
-<!-- 2. But what are limitations of Case V (and the BTL model) -->
-However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+<!-- 2. But what are limitations of Case V (and the BTL model)? -->
+However, Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
 677].
-As a result, its assumptions may not suit applications beyond the psycho-physical contexts for which it was created.
+As a result, its assumptions may not be suitable for applications beyond the psycho-physical contexts for which it was created.
 Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions could prove problematic with more complex, less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp.
+270], acknowledging that some assumptions could prove problematic in the presence of complex traits or heterogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp.
 374].
-Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
+Consequently, given that modern CJ applications frequently involve these types of traits and stimuli, two main assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
 
@@
 
-<!-- 3. What does the zero correlation between stimuli imply? -->
-The assumption of *zero correlation between stimuli* is best illustrated with an example.
-For instance, when evaluating text quality, this assumption suggests that a judge’s perception of quality in one text does not influence their perception of the same trait in the comparison text.
-Thurstone attributes this independence to the cancellation of potential judges' biases, driven by opposing and equally weighted factors; commonly known as 'mood' and 'simultaneous contrast' effects, that operate during pairwise comparisons [@Thurstone_1927b, pp.
+<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
+The assumption of *zero correlation between stimuli* is more easily understood through an example.
+For instance, when using pairwise comparisons to evaluate text quality, the assumption suggests that a judge's perception of the trait in one text does not influence his perception of the same trait on another.
+Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
 268].
-This cancellation of bias has been mathematically demonstrated by @Andrich_1978, assuming additive biases and a logit scale derived from the BTL model.
+This cancellation was mathematically demonstrated by @Andrich_1978, assuming discriminal processes with additive biases.
+However, it is easy to imagine two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.
 
-<!-- 4. But what are the problems of assuming zero correlation? -->
-However, given that evidence has shown judges' quality assessments are influenced by multiple, complex aspects of the texts [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], it is plausible that the accuracy of these judgments also varies based on factors such as age, culture, education, expertise [@Kelly_et_al_2022, pp.
-683], and ultimately, based on the individual judge themselves [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020].
+<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
+268].
+-->
 
-[This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder ]{style="color:red;"}
+<!-- 4. The first scenario where the assumption does not hold -->
+In the first scenario, evidence suggests that when judges are confronted with multi-dimensional, complex traits and heterogeneous stimuli, such as when evaluating written text quality, their assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+In such cases, it is not inconceivable that these various aspects of the stimuli unevenly influence the judges’ perception of one text over another, leading to biases that resist cancellation, as the effects of the factors involved in the comparison are neither equally weighted nor opposing.
+For example, this may occur when a judge, while assessing the argumentative quality of a text, places disproportionate emphasis on grammatical errors, ultimately favoring texts with fewer errors but weaker arguments.
+Although the literature does not provide direct evidence for this specific argument, studies such as @Pollitt_et_al_2003 do offer evidence of judges' biases, supporting the idea that the factors involved in pairwise comparisons may not always cancel out.
 
-Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].
+<!-- research shows that, in the presence of these traits and stimuli, the accuracy of the judges' trait perception is influenced by the rank-order distance of the stimuli [@vanDaal_et_al_2017; @Gijsen_et_al_2021], a coarse measure of the relative positioning of the stimuli within the trait continuum.
+-->
 
-<!-- Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women? [@Kelly_et_al_2022, pp. 683] -->
+<!-- 5. The second scenario where the assumption does not hold -->
+<!-- In the second scenario, the literature has not even considered the issues related with the existence of an additional hierarchical structure relevant to the stimuli, such as in cases where data includes multiple samples of stimuli within different individuals.
+In this scenario, there is an amount of correlation that it is not accounted for -->
 
-<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020] -->
+<!-- 6. But what are the problems when the zero correlation assumption does not hold? -->
+<!-- The best the literature has done is to estimate scores and then use those scores to run a multilevel model as in @Boonen_et_al_2020, which in itself brings its own problems.
+Refer to @sec-theory-issue2 -->
 
-<!-- What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
-This, in the best scenario, cause less precise estimates, and in the worst scenario, biased estimates.
-possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
+<!-- Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].  -->
 
@@
 (Consider the problem of using adhoc methods) -->
+
+<!-- Cancellation of judges: What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
+This, in the best scenario, cause less precise estimates by ignoring clusters, and in the worst scenario, biased estimates if the judges do register biases.
+-->
+
+<!-- Multilevel structure: Here we loose precision, because we are not controlling for clustering, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.
+-->
+
+<!-- (This a is a validity contention) -->
+<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
+Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp.
+683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement.
+As @Kelly_et_al_2022 noted [pp.
+683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->
 
@@
 
-<!-- this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli.
-Meaning there is an amount of correlation that it is not accounted for.
-Use example of Boonen!!
--->
-
-<!--   Here we loose precision, because we are not controlling for clustering -->
-
-<!-- This is the case of clustering or measurement error, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.
+<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
 -->
@@
 <!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->
-
-<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
-268].
--->
-
-<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
--->
 
@@
 
-## The disconnect between trait measurement and hypothesis testing
+## The disconnect between trait measurement and hypothesis testing {#sec-theory-issue2}
 
@@
 <!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
-For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses and tests, or to make decisions regarding the exclusion of certain data in these analyses and tests.
+The literature shows that these scores have been employed to calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023] or to test hypotheses related to the underlying traits of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Additionally, the BTL scores have been used to detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], as well as to identify "misfit" judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], with considerations for their possible exclusion.
 
 <!-- 3. But what are the challenges in using BTL scores for further analysis? -->
-However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests.
+However, the statistical literature advises caution when using estimated scores to conduct additional analyses and tests while warning against eliminating data through ad-hoc univariate procedures.
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests.
-Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp.
+Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests.
+Notably, the direction and magnitude of such biases are often unpredictable.
+Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp.
 25; @Hoyle_et_al_2023, pp.
 137].
-Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
+Moreover, eliminating data using ad-hoc univariate procedures using these scores compounds this issue by excluding possible valuable information from the analyses [@McElreath_2020], further introducing bias.
+Finally, the reduced precision in hypothesis tests weakens their statistical power, increasing the likelihood of type-I or type-II errors [@McElreath_2020].
+
 <!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->
@@
 
-## The role and impact of comparison algorithms
+## Assessment design features and their role on reliability and validity {#sec-theory-issue3}
+
+<!-- recording: Sven and Tine 24.11.18; time: 00:47:20 - 00:56:41 -->
 
@@
 
-# Theory {#sec-theory}
+# An updated theoretical and statistical model for CJ {#sec-theory}
 
@@
 
-## A theoretical model for CJ {#sec-theory-theoretical}
+## The theoretical model {#sec-theory-theoretical}
 
@@
 <!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->
+
+<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->
 
@@
 
-**Non-financial interests:** Author XX serve on advisory broad of Company Y but receives no compensation this role.
+**Non-financial interests:** The authors have no relevant non-financial interest to disclose.
 
@@
 
-**CRediT authorship contribution statement:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
+**CRediT authorship contribution statement:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review and editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
 and S.DM.; *Project administration:* S.G.

### 2024-11-21 _tex/index.tex
Built site for gh-pages

--- 2024-11-18 _tex/index.tex
@@
 methods when employing adaptive comparison algorithms
-\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
-the other hand, research on validity suggests that scores generated by
-CJ can accurately represent the traits under measurement
-\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
-Finally, research on practical applicability highlights the method's
+\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
+Meanwhile, research on validity suggests that scores generated by CJ can
+accurately represent the traits under measurement
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023},
+while research on practical applicability highlights the method's
 versatility across both educational and non-educational contexts
@@
 and fragmented research approaches have left several critical issues
-unaddressed. This research primarily focuses on three: the over-reliance
-on Thurstone's Case V assumptions in the statistical analysis of CJ
-data, the apparent disconnect between CJ's trait measurement and
-hypothesis testing, and the unclear role of comparison algorithms on the
-method's reliability and validity. The following sections will discuss
-each of these issues in detail, followed by the introduction of a
-theoretical model and its statistical translation, which aims to address
-all three concerns simultaneously.
+unaddressed. The present study primarily focuses on three: the
+over-reliance on the assumptions of Thurstone's Case V in the
+statistical analysis of CJ data, the apparent disconnect between CJ's
+trait measurement and hypothesis testing, and the unclear role of the
+diverse assessment design features on CJ's reliability and validity. The
+following sections begin with a brief overview of Thurstone's theory and
+a detailed discussion of these issues. Subsequently, the study
+introduces a theoretical model for CJ that builds upon Thurstone's
+theory, alongside its statistical translation, designed to address all
+three concerns simultaneously.
 
-\section{Three critical issues in CJ
-literature}\label{sec-theory-issues}
+\section{Thurstone's theory}\label{sec-thurstone_theory}
 
 In its most general form, Thurstone's theory
-\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
-resulting from comparing two stimuli is determined by two factors: the
-discriminal process of each stimulus and their discriminal difference.
-The \emph{discriminal process} refers to the psychological effect each
-stimulus has on the judges, or more simply stated, the judges'
-perception of the trait level of each stimulus. Thurstone assumes that
-the discriminal process for each stimulus follows a Normal distribution.
-In this distribution, the mode (mean), known as the \emph{modal
-discriminal process}, represents the position of the stimulus on the
-trait continuum, while the dispersion, known as the \emph{discriminal
-dispersion}, reflects the variability in the stimulus' perceived trait
-level. Figure~\ref{fig-discriminal_process} shows example distributions
-of discriminal process for two stimuli (objects).
+\citeyearpar{Thurstone_1927b} suggests that two factors determine the
+dichotomous outcome of pairwise comparisons: the discriminal process of
+each stimulus and their discriminal difference. The \emph{discriminal
+process} refers to the psychological effect each stimulus exerts on the
+judges, or more simply, the underlying perception of the stimulus' trait
+level. According to the theory, the discriminal process for each
+stimulus follows a Normal distribution. The mode (mean) of this
+distribution, referred to as the \emph{modal discriminal process},
+represents the stimulus' position on the trait continuum. Meanwhile, the
+dispersion of the distribution, referred to as the \emph{discriminal
+dispersion}, reflects the variability in the perceived trait level of
+the stimulus.
+
+However, since the discriminal process of a single stimulus is not
+directly observable, the \emph{law of comparative judgment} becomes
+essential. This law states that in pairwise comparisons, the stimulus
+positioned further along the trait continuum is perceived as having a
+higher level of that trait. Thus, the theory assumes the observed
+dichotomous outcome is determined by the distribution of the difference
+between the underlying discriminal processes of the stimuli, referred to
+as the \emph{discriminal difference}. This suggests that the outcome
+depends on the relative distance between stimuli, rather than their
+absolute positions on the trait continuum.
+
+These concepts can be illustrated through an example. For instance, in
+the context of evaluating text quality,
+Figure~\ref{fig-discriminal_process} could depict the underlying
+discriminal process distributions for two written texts, highlighting
+differences in their discriminal dispersions and modal discriminal
+processes along the quality trait continuum. Furthermore,
+Figure~\ref{fig-discriminal_difference} could display the discriminal
+difference distribution for these texts, showing that text A is
+perceived to exhibit significantly higher quality than text B, as
+indicated by the shaded gray area. Consequently, the dichotomous outcome
+of this comparison would favor text A.
 
@@
 
+\begin{minipage}{0.50\linewidth}
+
 \centering{
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}
+\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}
 
@@
 
-\caption{\label{fig-discriminal_process}Example distributions of
-discriminal processes for two stimuli (objects). Extracted from
-\citet[pp.~249]{Bramley_2008}.}
+\subcaption{\label{fig-discriminal_process}Discriminal processes}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/figures/comparative_judgment.png}
+
+}
+
+\subcaption{\label{fig-discriminal_difference}Discriminal difference}
+
+\end{minipage}%
+
+\caption{\label{fig-thurstone_theory}Example distribution of discriminal
+processes and their discriminal difference for two written texts
+(stimuli or objects). Extracted from \citet[pp.~249-251]{Bramley_2008}.}
 
@@
 
-However, since the discriminal mode and dispersion of a single stimulus
-are not directly observable except through comparison, the \emph{law of
-comparative judgment} becomes essential. This law asserts that when
-assessing a specific trait by comparing two stimuli, the stimulus
-positioned further along the continuum is perceived as having a higher
-level of that trait. Thus, the observed dichotomous outcome is
-determined by the distribution of the difference between the stimuli's
-discriminal processes, called the \emph{discriminal difference}.
-Figure~\ref{fig-comparative_judgment} shows an example distribution of
-the discriminal difference for two stimuli (objects).
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{./images/figures/comparative_judgment.png}
-
-}
-
-\caption{\label{fig-comparative_judgment}Distribution of the discriminal
-difference between two stimuli (objects). Extracted from
-\citet[pp.~251]{Bramley_2008}.}
-
-\end{figure}%
-
-Importantly, the theory's general form primarily addresses pairwise
-comparisons of stimuli made by a single judge
-\citep[pp.~267]{Thurstone_1927b}. Consequently, to enhance its practical
-applicability, Thurstone introduced five distinct cases, each defined by
-progressively simplifying assumptions. Table~\ref{tbl-thurstone_cases}
-summarizes these cases, focusing on key assumptions such as the
-distribution of discriminal processes, the similarity of discriminal
-dispersions across stimuli, the correlation between stimuli, and which
-judges perform the comparisons. For a comprehensive discussion of this
-progression, refer to \citet{Thurstone_1927b} and
-\citet[pp.~248-253]{Bramley_2008}.
+Importantly, the general form of Thurstone's theory primarily addressed
+pairwise comparisons of stimuli made by a single judge
+\citep[pp.~267]{Thurstone_1927b}. Thus, for practical application,
+Thurstone introduced five distinct cases derived from this general form,
+each defined by progressively simplifying assumptions.
+Table~\ref{tbl-thurstone_cases} summarizes these cases, highlighting key
+assumptions such as the distribution of discriminal processes, the
+similarity of discriminal dispersions across stimuli, the correlation
+between stimuli, and the number of judges performing the comparisons.
+For a comprehensive discussion of this progression, refer to
+\citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.
 
@@
 
+\section{Three critical issues in CJ
+literature}\label{sec-theory-issues}
+
 \subsection{The Case V and the statistical analysis of CJ
-data}\label{the-case-v-and-the-statistical-analysis-of-cj-data}
+data}\label{sec-theory-issue1}
 
-Surprisingly, despite its reliance on the largest number of simplifying
-assumptions
+Despite its reliance on the largest number of simplifying assumptions
 \citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
@@
 in the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors Case V's
-assumptions, with one key difference: while Case V assumes a Normal
-distribution for the stimuli's discriminal processes, the BTL model uses
-the more mathematically tractable Logistic distribution
+\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
+assumptions of Case V, with one key difference: while Case V assumes a
+Normal distribution for the discriminal processes of stimuli, the BTL
+model uses the more mathematically tractable Logistic distribution
 \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
-Table~\ref{tbl-thurstone_cases}). This substitution has little effect on
+Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
 the model's estimation or interpretation, as the Normal and Logistic
 distributions differ by a scaling factor of approximately \(1.7\)
-\citep[pp.~16]{vanderLinden_et_al_2017_I} (refer to
+\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
 Figure~\ref{fig-logistic_vs_normal}).
@@
 
-\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}
+\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}
 
@@
 
-\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/cumulative.png}
+\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/figures/cumulative.png}
 
@@
 
-However, Case V (and its BTL model counterpart) was originally developed
-to provide a ``rather coarse scaling'' of traits
-\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
-over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
-As a result, its assumptions may not suit applications beyond the
-psycho-physical contexts for which it was created. Thurstone himself
-cautioned that its use ``should not be made without (an) experimental
-test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging that some
-assumptions could prove problematic with more complex, less homogeneous
-stimuli, such as handwriting or English compositions
-\citep[pp.~374]{Thurstone_1927a}. Consequently, given that current CJ
-applications often deal with complex, less homogeneous stimuli, two key
-assumptions of Case V may not consistently hold in theory or practice:
-the zero correlation and equal dispersion between stimuli.
+However, Case V was originally developed to provide a ``rather coarse
+scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
+statistical simplicity over precision in trait measurement
+\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
+be suitable for applications beyond the psycho-physical contexts for
+which it was created. Thurstone himself cautioned that its use ``should
+not be made without (an) experimental test''
+\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
+could prove problematic in the presence of complex traits or
+heterogeneous stimuli, such as handwriting or English compositions
+\citep[pp.~374]{Thurstone_1927a}. Consequently, given that modern CJ
+applications frequently involve these types of traits and stimuli, two
+main assumptions of Case V may not consistently hold in theory or
+practice: the zero correlation and equal dispersion between stimuli.
 
-The assumption of \emph{zero correlation between stimuli} is best
-illustrated with an example. For instance, when evaluating text quality,
-this assumption suggests that a judge's perception of quality in one
-text does not influence their perception of the same trait in the
-comparison text. Thurstone attributes this independence to the
-cancellation of potential judges' biases, driven by opposing and equally
-weighted factors; commonly known as `mood' and `simultaneous contrast'
-effects, that operate during pairwise comparisons
-\citep[pp.~268]{Thurstone_1927b}. This cancellation of bias has been
-mathematically demonstrated by \citet{Andrich_1978}, assuming additive
-biases and a logit scale derived from the BTL model.
+The assumption of \emph{zero correlation between stimuli} is more easily
+understood through an example. For instance, when using pairwise
+comparisons to evaluate text quality, the assumption suggests that a
+judge's perception of the trait in one text does not influence his
+perception of the same trait on another. Thurstone attributed this
+independence to the cancellation of potential judges' biases, driven by
+two opposing and equally weighted effects occurring during the pairwise
+comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
+mathematically demonstrated by \citet{Andrich_1978}, assuming
+discriminal processes with additive biases. However, it is easy to
+imagine two scenarios where the zero correlation assumption almost
+certainly does not hold: when the pairwise comparison involves
+multidimensional, complex traits with heterogeneous stimuli, and when an
+additional hierarchical structure is relevant to the stimuli.
 
-However, given that evidence has shown judges' quality assessments are
-influenced by multiple, complex aspects of the texts
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}, it is
-plausible that the accuracy of these judgments also varies based on
-factors such as age, culture, education, expertise
-\citep[pp.~683]{Kelly_et_al_2022}, and ultimately, based on the
-individual judge themselves
-\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020}.
-
-{This means that the cancellation of judges' biases may not hold in
-theory or practice, and that judges' biases may depend on the stimuli,
-by means of an interaction (but present) or because the judges works as
-a confounder}
-
-Such differences may not be detected through analyses of bias and misfit
-\citep[pp.~683]{Kelly_et_al_2022}.
+In the first scenario, evidence suggests that when judges are confronted
+with multi-dimensional, complex traits and heterogeneous stimuli, such
+as when evaluating written text quality, their assessments are
+influenced by multiple, intricate aspects of the stimuli
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
+such cases, it is not inconceivable that these various aspects of the
+stimuli unevenly influence the judges' perception of one text over
+another, leading to biases that resist cancellation, as the effects of
+the factors involved in the comparison are neither equally weighted nor
+opposing. For example, this may occur when a judge, while assessing the
+argumentative quality of a text, places disproportionate emphasis on
+grammatical errors, ultimately favoring texts with fewer errors but
+weaker arguments. Although the literature does not provide direct
+evidence for this specific argument, studies such as
+\citet{Pollitt_et_al_2003} do offer evidence of judges' biases,
+supporting the idea that the factors involved in pairwise comparisons
+may not always cancel out.
 
 \subsection{The disconnect between trait measurement and hypothesis
-testing}\label{the-disconnect-between-trait-measurement-and-hypothesis-testing}
+testing}\label{sec-theory-issue2}
 
@@
 Researchers then typically use the estimated BTL scores, or their
-transformations, to conduct additional analyses or hypothesis tests. For
-example, these scores have been used to identify `misfit' judges and
-stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018},
-detect biases in judges' ratings
-\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
-other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
-or test hypotheses related to the underlying trait of interest
+transformations, to conduct additional analyses and tests, or to make
+decisions regarding the exclusion of certain data in these analyses and
+tests. The literature shows that these scores have been employed to
+calculate correlations with other assessment methods
+\citep{Goossens_et_al_2018, Bouwer_et_al_2023} or to test hypotheses
+related to the underlying traits of interest
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+Additionally, the BTL scores have been used to detect biases in judges'
+ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, as well as to
+identify ``misfit'' judges and stimuli
+\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, with
+considerations for their possible exclusion.
 
 However, the statistical literature advises caution when using estimated
-scores to conduct additional analyses or hypotheses tests. A key
+scores to conduct additional analyses and tests while warning against
+eliminating data through ad-hoc univariate procedures. A key
 consideration is that BTL scores are parameter estimates that inherently
-carry uncertainty. Ignoring this uncertainty can introduce bias into the
-analysis and reduce the precision of hypothesis tests. Notably, the
-direction and magnitude of the bias are often unpredictable; results may
-be attenuated, exaggerated, or remain unaffected, depending on the
-amount of uncertainty present in the scores and the actual effects being
-tested
+carry uncertainty. Ignoring this uncertainty can bias the analysis and
+reduce the precision of hypothesis tests. Notably, the direction and
+magnitude of such biases are often unpredictable. Results may be
+attenuated, exaggerated, or remain unaffected depending on the degree of
+uncertainty in the scores and the actual effects being tested
 \citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
-Furthermore, reduced precision in hypothesis tests weakens their
-statistical power, ultimately increasing the likelihood of committing
-type-I or type-II errors \citep{McElreath_2020}.
+Moreover, eliminating data using ad-hoc univariate procedures using
+these scores compounds this issue by excluding possible valuable
+information from the analyses \citep{McElreath_2020}, further
+introducing bias. Finally, the reduced precision in hypothesis tests
+weakens their statistical power, increasing the likelihood of type-I or
+type-II errors \citep{McElreath_2020}.
 
@@
 
-\subsection{The role and impact of comparison
-algorithms}\label{the-role-and-impact-of-comparison-algorithms}
+\subsection{Assessment design features and their role on reliability and
+validity}\label{sec-theory-issue3}
 
-\section{Theory}\label{sec-theory}
+\section{An updated theoretical and statistical model for
+CJ}\label{sec-theory}
 
-\subsection{A theoretical model for CJ}\label{sec-theory-theoretical}
+\subsection{The theoretical model}\label{sec-theory-theoretical}
 
@@
 
-\textbf{Non-financial interests:} Author XX serve on advisory broad of
-Company Y but receives no compensation this role.
+\textbf{Non-financial interests:} The authors have no relevant
+non-financial interest to disclose.
 
@@
 \emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
-J.M.R.E.; \emph{Writing - review \& editing:} S.G., S.DM., and T.vD.;
+J.M.R.E.; \emph{Writing - review and editing:} S.G., S.DM., and T.vD.;
 \emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;

### 2024-11-21 index.qmd
section:issues; process:write

@@
 Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.
-This suggests that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.
+This indicates that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.
 
 <!-- 3. Example-->
-These concepts can be illustrated through an example.
+These concepts are more easily understood through an example.
 For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum.
@@
 The assumption of *zero correlation between stimuli* is more easily understood through an example.
-For instance, when using pairwise comparisons to evaluate text quality, the assumption suggests that a judge's perception of the trait in one text does not influence his perception of the same trait on another.
+For instance, when using pairwise comparisons to evaluate text quality, the assumption suggests that the judge's perception of the trait in one text does not influence his perception of the same trait on another.
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
@@
 In the first scenario, evidence suggests that when judges are confronted with multi-dimensional, complex traits and heterogeneous stimuli, such as when evaluating written text quality, their assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-In such cases, it is not inconceivable that these various aspects of the stimuli unevenly influence the judges’ perception of one text over another, leading to biases that resist cancellation, as the effects of the factors involved in the comparison are neither equally weighted nor opposing.
+In such cases, it is not inconceivable that various aspects of the stimuli unevenly influence the judges’ perception of one text over another, leading to biases that resist cancellation, as the effects of the factors involved in the comparison are neither equally weighted nor opposing.
 For example, this may occur when a judge, while assessing the argumentative quality of a text, places disproportionate emphasis on grammatical errors, ultimately favoring texts with fewer errors but weaker arguments.
-Although the literature does not provide direct evidence for this specific argument, studies such as @Pollitt_et_al_2003 do offer evidence of judges' biases, supporting the idea that the factors involved in pairwise comparisons may not always cancel out.
+Although the literature does not provide direct evidence for this specific argument, studies such as @Pollitt_et_al_2003 do offer evidence of judges' biases, supporting the idea that the factors involved in the pairwise comparisons may not always cancel out.
 
@@
 <!-- 5. The second scenario where the assumption does not hold -->
-<!-- In the second scenario, the literature has not even considered the issues related with the existence of an additional hierarchical structure relevant to the stimuli, such as in cases where data includes multiple samples of stimuli within different individuals.
-In this scenario, there is an amount of correlation that it is not accounted for -->
+In the second scenario, although the CJ literature has acknowledged the existence of additional hierarchical structures relevant to the stimuli, the statistical handling of this extra source of correlation has been inadequate.
+For example, in cases where data included multiple samples of stimuli within different individuals, researchers have often relied on using (averaged) estimated BTL scores to conduct further analyses and tests at the hierarchical level of the individuals [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+However, this approach has the notable limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).
 
-<!-- 6. But what are the problems when the zero correlation assumption does not hold? -->
-<!-- The best the literature has done is to estimate scores and then use those scores to run a multilevel model as in @Boonen_et_al_2020, which in itself brings its own problems.
-Refer to @sec-theory-issue2 -->
-
-<!-- Such differences may not be detected through analyses of bias and misfit [@Kelly_et_al_2022, pp. 683].  -->
-
-<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind.
-This is studied by analyzing the residuals that remain after the model is fitted to the data.
-Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large.
-Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability.
-indicates that one judge – Judge 5 – is significantly biased in some way.
-(Consider the problem of using adhoc methods) -->
-
-<!-- Cancellation of judges: What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
-This, in the best scenario, cause less precise estimates by ignoring clusters, and in the worst scenario, biased estimates if the judges do register biases.
--->
-
-<!-- Multilevel structure: Here we loose precision, because we are not controlling for clustering, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.
--->
-
-<!-- (This a is a validity contention) -->
+<!-- (Is this a validity contention?) -->
 <!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
@@
 
-<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
+<!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
+In contrast, the assumption of *equal dispersion between stimuli* suggests that the perceived trait of all texts varies around their modal discriminal process in the same way across comparisons.
+While Thurstone noted that the assumption of equal dispersion may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp.
+374], no study explicitly proposes that this assumption may also be violated due to the presence of an additional hierarchical structure relevant to the texts.
+One such scenario could occur when comparing university and secondary school students.
+In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students might exhibit a broader range of writing abilities, leading to greater variability in their text quality.
+Despite the contrived nature of the example, it helps to effectively illustrate how assuming equal dispersions in the texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
+
+<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
+Thurstone clearly did not expect this to hold for any but the most simple stimuli.
+The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
 -->
 
-<!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->
+<!-- X. But what are the problems when the zero correlation assumption does not hold? -->
 
-<!-- 4. The problems of assuming equal dispersion between stimuli -->
-
-<!-- it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.
-Example of exaggerated comparison of individuals, university students vs secondary school students.
+<!-- Cancellation of judges: What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
+This, in the best scenario, cause less precise estimates by ignoring clusters, and in the worst scenario, biased estimates if the judges do register biases.
 -->
 
-<!--   here we face misspecification in the model -->
+<!-- Multilevel structure: Here we loose precision, because we are not controlling for clustering, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.
+-->
 
-<!-- "It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927a, pp.
-374].
--->
+<!-- Different dispersion: here we face misspecification in the model? -->
+
+<!-- This observation highlights how variability in the quality of stimuli should be accounted for when comparing heterogeneous groups or evaluating complex attributes like writing ability. -->
 
@@
 
-<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
-Thurstone clearly did not expect this to hold for any but the most simple stimuli.
-The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naïve to expect this assumption to hold here [@Bramley_2008].
+<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
 -->
+
+<!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->
 
@@
 <!-- 3. But what are the challenges in using BTL scores for further analysis? -->
-However, the statistical literature advises caution when using estimated scores to conduct additional analyses and tests while warning against eliminating data through ad-hoc univariate procedures.
+However, the statistical literature advises caution when using estimated scores for additional analyses and tests, as well as when eliminating data through ad hoc univariate procedures.
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
@@
 137].
-Moreover, eliminating data using ad-hoc univariate procedures using these scores compounds this issue by excluding possible valuable information from the analyses [@McElreath_2020], further introducing bias.
-Finally, the reduced precision in hypothesis tests weakens their statistical power, increasing the likelihood of type-I or type-II errors [@McElreath_2020].
+Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020].
+Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of type-I or type-II errors [@McElreath_2020].
 
@@
 
-## Assessment design features and their role on reliability and validity {#sec-theory-issue3}
+## The assessment design features and their role on reliability and validity {#sec-theory-issue3}
 

### 2024-11-21 _tex/index.tex
Built site for gh-pages

@@
 between the underlying discriminal processes of the stimuli, referred to
-as the \emph{discriminal difference}. This suggests that the outcome
+as the \emph{discriminal difference}. This indicates that the outcome
 depends on the relative distance between stimuli, rather than their
@@
 
-These concepts can be illustrated through an example. For instance, in
-the context of evaluating text quality,
+These concepts are more easily understood through an example. For
+instance, in the context of evaluating text quality,
 Figure~\ref{fig-discriminal_process} could depict the underlying
@@
 understood through an example. For instance, when using pairwise
-comparisons to evaluate text quality, the assumption suggests that a
+comparisons to evaluate text quality, the assumption suggests that the
 judge's perception of the trait in one text does not influence his
@@
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
-such cases, it is not inconceivable that these various aspects of the
-stimuli unevenly influence the judges' perception of one text over
-another, leading to biases that resist cancellation, as the effects of
-the factors involved in the comparison are neither equally weighted nor
+such cases, it is not inconceivable that various aspects of the stimuli
+unevenly influence the judges' perception of one text over another,
+leading to biases that resist cancellation, as the effects of the
+factors involved in the comparison are neither equally weighted nor
 opposing. For example, this may occur when a judge, while assessing the
@@
 \citet{Pollitt_et_al_2003} do offer evidence of judges' biases,
-supporting the idea that the factors involved in pairwise comparisons
-may not always cancel out.
+supporting the idea that the factors involved in the pairwise
+comparisons may not always cancel out.
+
+In the second scenario, although the CJ literature has acknowledged the
+existence of additional hierarchical structures relevant to the stimuli,
+the statistical handling of this extra source of correlation has been
+inadequate. For example, in cases where data included multiple samples
+of stimuli within different individuals, researchers have often relied
+on using (averaged) estimated BTL scores to conduct further analyses and
+tests at the hierarchical level of the individuals
+\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+However, this approach has the notable limitation of ignoring the
+uncertainty associated with the scores (refer to section
+Section~\ref{sec-theory-issue2} for a detailed discussion of this
+issue).
+
+In contrast, the assumption of \emph{equal dispersion between stimuli}
+suggests that the perceived trait of all texts varies around their modal
+discriminal process in the same way across comparisons. While Thurstone
+noted that the assumption of equal dispersion may be violated when
+``dealing with less conspicuous attributes or with less homogeneous
+stimuli'' \citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes
+that this assumption may also be violated due to the presence of an
+additional hierarchical structure relevant to the texts. One such
+scenario could occur when comparing university and secondary school
+students. In this case, university students may consistently (or more
+precisely) produce higher-quality texts, while secondary school students
+might exhibit a broader range of writing abilities, leading to greater
+variability in their text quality. Despite the contrived nature of the
+example, it helps to effectively illustrate how assuming equal
+dispersions in the texts can overlook meaningful differences in the
+reliability of text quality across groups or individuals.
 
@@
 However, the statistical literature advises caution when using estimated
-scores to conduct additional analyses and tests while warning against
-eliminating data through ad-hoc univariate procedures. A key
-consideration is that BTL scores are parameter estimates that inherently
-carry uncertainty. Ignoring this uncertainty can bias the analysis and
-reduce the precision of hypothesis tests. Notably, the direction and
-magnitude of such biases are often unpredictable. Results may be
-attenuated, exaggerated, or remain unaffected depending on the degree of
-uncertainty in the scores and the actual effects being tested
+scores for additional analyses and tests, as well as when eliminating
+data through ad hoc univariate procedures. A key consideration is that
+BTL scores are parameter estimates that inherently carry uncertainty.
+Ignoring this uncertainty can bias the analysis and reduce the precision
+of hypothesis tests. Notably, the direction and magnitude of such biases
+are often unpredictable. Results may be attenuated, exaggerated, or
+remain unaffected depending on the degree of uncertainty in the scores
+and the actual effects being tested
 \citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
-Moreover, eliminating data using ad-hoc univariate procedures using
-these scores compounds this issue by excluding possible valuable
-information from the analyses \citep{McElreath_2020}, further
-introducing bias. Finally, the reduced precision in hypothesis tests
-weakens their statistical power, increasing the likelihood of type-I or
-type-II errors \citep{McElreath_2020}.
+Moreover, excluding data using ad hoc univariate procedures can compound
+these issues by discarding potentially valuable information, further
+exacerbating the bias \citep{Zimmerman_1994, McElreath_2020}. Finally,
+the reduced precision in hypothesis tests diminishes their statistical
+power, increasing the likelihood of type-I or type-II errors
+\citep{McElreath_2020}.
 
@@
 
-\subsection{Assessment design features and their role on reliability and
-validity}\label{sec-theory-issue3}
+\subsection{The assessment design features and their role on reliability
+and validity}\label{sec-theory-issue3}
 

### 2024-11-22 index.qmd
section:issues, aseV; process:write

@@
 Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area.
-Consequently, the dichotomous outcome of this comparison would favor text A.
+Consequently, the dichotomous outcome of this comparison would probably favor text A.
 
@@
 254] (see @tbl-thurstone_cases).
-This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
+This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
 16] (see @fig-logistic_vs_normal).
@@
 <!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
-The assumption of *zero correlation between stimuli* is more easily understood through an example.
-For instance, when using pairwise comparisons to evaluate text quality, the assumption suggests that the judge's perception of the trait in one text does not influence his perception of the same trait on another.
+The assumption of *zero correlation between stimuli* can be better understood through an example.
+For instance, when using pairwise comparisons to evaluate text quality, the assumption implies that a judge's perception of a trait in one text does not influence his perception of the same trait in another text.
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
 268].
-This cancellation was mathematically demonstrated by @Andrich_1978, assuming discriminal processes with additive biases.
-However, it is easy to imagine two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.
+This cancellation was mathematically demonstrated by @Andrich_1978, using the BTL model under the assumption of additive biases in the discriminal processes.
+However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.
 
@@
 <!-- 4. The first scenario where the assumption does not hold -->
-In the first scenario, evidence suggests that when judges are confronted with multi-dimensional, complex traits and heterogeneous stimuli, such as when evaluating written text quality, their assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-In such cases, it is not inconceivable that various aspects of the stimuli unevenly influence the judges’ perception of one text over another, leading to biases that resist cancellation, as the effects of the factors involved in the comparison are neither equally weighted nor opposing.
-For example, this may occur when a judge, while assessing the argumentative quality of a text, places disproportionate emphasis on grammatical errors, ultimately favoring texts with fewer errors but weaker arguments.
-Although the literature does not provide direct evidence for this specific argument, studies such as @Pollitt_et_al_2003 do offer evidence of judges' biases, supporting the idea that the factors involved in the pairwise comparisons may not always cancel out.
+In the first scenario, the intricate aspects of multidimensional, complex traits and heterogeneous stimuli may introduce dependencies between stimuli.
+Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+In this context, it is not inconceivable that these aspects, being neither equally weighted nor opposing, may unevenly influence judges' perceptions, resulting in biases that resist cancellation.
+For example, this might occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments.
+While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of judges' biases, supporting the idea that the different factors influencing pairwise comparisons may not always cancel out.
 
@@
 <!-- 5. The second scenario where the assumption does not hold -->
-In the second scenario, although the CJ literature has acknowledged the existence of additional hierarchical structures relevant to the stimuli, the statistical handling of this extra source of correlation has been inadequate.
-For example, in cases where data included multiple samples of stimuli within different individuals, researchers have often relied on using (averaged) estimated BTL scores to conduct further analyses and tests at the hierarchical level of the individuals [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, this approach has the notable limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).
+In the second scenario, the shared context or inherent connections created by the additional hierarchical structure may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
+Nevertheless, despite recognizing such hierarchical structures in CJ data, the statistical handling of this extra source of dependency in the CJ literature has been inadequate.
+For instance, in cases where the CJ data included multiple samples of stimuli from the same individuals, researchers have often relied on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+This approach, however, has the significant limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).
 
@@
 <!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
-In contrast, the assumption of *equal dispersion between stimuli* suggests that the perceived trait of all texts varies around their modal discriminal process in the same way across comparisons.
-While Thurstone noted that the assumption of equal dispersion may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp.
-374], no study explicitly proposes that this assumption may also be violated due to the presence of an additional hierarchical structure relevant to the texts.
-One such scenario could occur when comparing university and secondary school students.
-In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students might exhibit a broader range of writing abilities, leading to greater variability in their text quality.
-Despite the contrived nature of the example, it helps to effectively illustrate how assuming equal dispersions in the texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
+In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimuli is the same across stimuli.
+While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp.
+374], no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
+One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
+In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
+Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
 
@@
 
-<!-- X. But what are the problems when the zero correlation assumption does not hold? -->
+<!-- 7. But what are the challenges when these assumptions are violated? -->
+["however" paragraph]{style="color:red;"}
 
@@
 
+<!-- 8. So, how do we mitigate these risks? -->
+["to mitigate" paragraph]{style="color:red;"}
+
 <!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
@@
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is evident that the BTL model commonly functions as the measurement model for the trait of interest in CJ experiments [@Andrich_1978; @Bramley_2008].
+Building on the previous section, it is evident that the BTL model commonly functions as the trait's measurement model in CJ experiments [@Andrich_1978; @Bramley_2008].
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
@@
 Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020].
-Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of type-I or type-II errors [@McElreath_2020].
+Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
 
@@
 
-## The assessment design features and their role on reliability and validity {#sec-theory-issue3}
+## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3}
 

### 2024-11-22 _tex/index.tex
Built site for gh-pages

--- 2024-11-21 _tex/index.tex
@@
 indicated by the shaded gray area. Consequently, the dichotomous outcome
-of this comparison would favor text A.
+of this comparison would probably favor text A.
 
@@
 the model's estimation or interpretation, as the Normal and Logistic
-distributions differ by a scaling factor of approximately \(1.7\)
+distributions share similar statistical properties, differing only by a
+scaling factor of approximately \(1.7\)
 \citep[pp.~16]{vanderLinden_et_al_2017_I} (see
@@
 
-The assumption of \emph{zero correlation between stimuli} is more easily
+The assumption of \emph{zero correlation between stimuli} can be better
 understood through an example. For instance, when using pairwise
-comparisons to evaluate text quality, the assumption suggests that a
-judge's perception of the trait in one text does not influence his
-perception of the same trait on another. Thurstone attributed this
+comparisons to evaluate text quality, the assumption implies that a
+judge's perception of a trait in one text does not influence his
+perception of the same trait in another text. Thurstone attributed this
 independence to the cancellation of potential judges' biases, driven by
@@
 comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
-mathematically demonstrated by \citet{Andrich_1978}, assuming
-discriminal processes with additive biases. However, it is easy to
-imagine two scenarios where the zero correlation assumption almost
-certainly does not hold: when the pairwise comparison involves
-multidimensional, complex traits with heterogeneous stimuli, and when an
-additional hierarchical structure is relevant to the stimuli.
+mathematically demonstrated by \citet{Andrich_1978}, using the BTL model
+under the assumption of additive biases in the discriminal processes.
+However, it is easy to imagine at least two scenarios where the zero
+correlation assumption almost certainly does not hold: when the pairwise
+comparison involves multidimensional, complex traits with heterogeneous
+stimuli, and when an additional hierarchical structure is relevant to
+the stimuli.
 
-In the first scenario, evidence suggests that when judges are confronted
-with multi-dimensional, complex traits and heterogeneous stimuli, such
-as when evaluating written text quality, their assessments are
-influenced by multiple, intricate aspects of the stimuli
+In the first scenario, the intricate aspects of multidimensional,
+complex traits and heterogeneous stimuli may introduce dependencies
+between stimuli. Research on text quality suggests that when judges
+evaluate these traits, they often rely on various intricate aspects of
+the stimuli to form their judgments
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
-such cases, it is not inconceivable that these various aspects of the
-stimuli unevenly influence the judges' perception of one text over
-another, leading to biases that resist cancellation, as the effects of
-the factors involved in the comparison are neither equally weighted nor
-opposing. For example, this may occur when a judge, while assessing the
-argumentative quality of a text, places disproportionate emphasis on
-grammatical errors, ultimately favoring texts with fewer errors but
-weaker arguments. Although the literature does not provide direct
-evidence for this specific argument, studies such as
-\citet{Pollitt_et_al_2003} do offer evidence of judges' biases,
-supporting the idea that the factors involved in pairwise comparisons
-may not always cancel out.
+this context, it is not inconceivable that these aspects, being neither
+equally weighted nor opposing, may unevenly influence judges'
+perceptions, resulting in biases that resist cancellation. For example,
+this might occur when a judge assessing the argumentative quality of a
+text places disproportionate emphasis on grammatical accuracy,
+ultimately favoring texts with fewer errors but weaker arguments. While
+direct evidence for this specific scenario is lacking, studies such as
+\citet{Pollitt_et_al_2003} demonstrate the presence of judges' biases,
+supporting the idea that the different factors influencing pairwise
+comparisons may not always cancel out.
+
+In the second scenario, the shared context or inherent connections
+created by the additional hierarchical structure may introduce
+dependencies between stimuli, a statistical phenomenon commonly known as
+clustering \citep{Everitt_et_al_2010}. Nevertheless, despite recognizing
+such hierarchical structures in CJ data, the statistical handling of
+this extra source of dependency in the CJ literature has been
+inadequate. For instance, in cases where the CJ data included multiple
+samples of stimuli from the same individuals, researchers have often
+relied on (averaged) estimated BTL scores to conduct subsequent analyses
+and tests at the individual hierarchical level
+\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+This approach, however, has the significant limitation of ignoring the
+uncertainty associated with the scores (refer to section
+Section~\ref{sec-theory-issue2} for a detailed discussion of this
+issue).
+
+In contrast, the assumption of \emph{equal dispersion between stimuli}
+suggests that the variability in the perceived trait level of the
+stimuli is the same across stimuli. While Thurstone acknowledged that
+this assumption may be violated when ``dealing with less conspicuous
+attributes or with less homogeneous stimuli''
+\citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes that this
+assumption could also be violated due to the presence of an additional
+hierarchical (grouping) structure relevant to the texts. One such
+scenario might arise, for example, when comparing texts produced by
+university and secondary school students. In this case, university
+students may consistently (or more precisely) produce higher-quality
+texts, while secondary school students, who exhibit a broader range of
+writing abilities, would show greater variability in the quality of
+their texts. Although this example is somewhat contrived, it effectively
+illustrates how assuming equal dispersions across texts can overlook
+meaningful differences in the reliability of text quality across groups
+or individuals.
+
+{``however'' paragraph}
+
+{``to mitigate'' paragraph}
 
@@
 Building on the previous section, it is evident that the BTL model
-commonly functions as the measurement model for the trait of interest in
-CJ experiments \citep{Andrich_1978, Bramley_2008}. A measurement model
-specifies how manifest variables contribute to the estimation of latent
-variables \citep{Everitt_et_al_2010}. For example, when evaluating text
-quality, researchers use the BTL model to process the dichotomous
-outcomes resulting from the pairwise comparisons (the manifest
-variables) to estimate scores that reflect the underlying quality level
-of texts (the latent variable)
+commonly functions as the trait's measurement model in CJ experiments
+\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
+manifest variables contribute to the estimation of latent variables
+\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
+researchers use the BTL model to process the dichotomous outcomes
+resulting from the pairwise comparisons (the manifest variables) to
+estimate scores that reflect the underlying quality level of texts (the
+latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
@@
 However, the statistical literature advises caution when using estimated
-scores to conduct additional analyses and tests while warning against
-eliminating data through ad-hoc univariate procedures. A key
-consideration is that BTL scores are parameter estimates that inherently
-carry uncertainty. Ignoring this uncertainty can bias the analysis and
-reduce the precision of hypothesis tests. Notably, the direction and
-magnitude of such biases are often unpredictable. Results may be
-attenuated, exaggerated, or remain unaffected depending on the degree of
-uncertainty in the scores and the actual effects being tested
+scores for additional analyses and tests, as well as when eliminating
+data through ad hoc univariate procedures. A key consideration is that
+BTL scores are parameter estimates that inherently carry uncertainty.
+Ignoring this uncertainty can bias the analysis and reduce the precision
+of hypothesis tests. Notably, the direction and magnitude of such biases
+are often unpredictable. Results may be attenuated, exaggerated, or
+remain unaffected depending on the degree of uncertainty in the scores
+and the actual effects being tested
 \citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
-Moreover, eliminating data using ad-hoc univariate procedures using
-these scores compounds this issue by excluding possible valuable
-information from the analyses \citep{McElreath_2020}, further
-introducing bias. Finally, the reduced precision in hypothesis tests
-weakens their statistical power, increasing the likelihood of type-I or
-type-II errors \citep{McElreath_2020}.
+Moreover, excluding data using ad hoc univariate procedures can compound
+these issues by discarding potentially valuable information, further
+exacerbating the bias \citep{Zimmerman_1994, McElreath_2020}. Finally,
+the reduced precision in hypothesis tests diminishes their statistical
+power, increasing the likelihood of committing type-I or type-II errors
+\citep{McElreath_2020}.
 
@@
 
-\subsection{Assessment design features and their role on reliability and
-validity}\label{sec-theory-issue3}
+\subsection{The diverse assessment design features and their role on
+reliability and validity}\label{sec-theory-issue3}
 

### 2024-11-22 index.qmd
section:issues,CaseV; process:write

@@
 <!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
-In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimuli is the same across stimuli.
+In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimulus is the same across stimuli.
 While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp.

### 2024-11-22 _tex/index.tex
Built site for gh-pages

@@
 suggests that the variability in the perceived trait level of the
-stimuli is the same across stimuli. While Thurstone acknowledged that
+stimulus is the same across stimuli. While Thurstone acknowledged that
 this assumption may be violated when ``dealing with less conspicuous

### 2024-11-25 index.qmd
section:issues,CaseV; process:write

--- 2024-11-22 index.qmd
@@
 <!-- 2. Thurstone's Theory of CJ: discriminal difference -->
-However, since the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential.
+Nevertheless, the theory posits that because the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential.
 This law states that in pairwise comparisons, the stimulus positioned further along the trait continuum is perceived as having a higher level of that trait.
+This emphasize that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.
 Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.
-This indicates that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.
 
@@
 These concepts are more easily understood through an example.
-For instance, in the context of evaluating text quality, @fig-discriminal_process could depict the underlying discriminal process distributions for two written texts, highlighting differences in their discriminal dispersions and modal discriminal processes along the quality trait continuum.
-Furthermore, @fig-discriminal_difference could display the discriminal difference distribution for these texts, showing that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area.
-Consequently, the dichotomous outcome of this comparison would probably favor text A.
+For instance, in the context of evaluating text quality, @fig-discriminal_process depicts the underlying discriminal process distributions for two written texts.
+The figure highlights differences in the texts' positions along the quality trait continuum, represented by their modal discriminal processes, as well as differences in the variability of their perceived trait levels, represented by their discriminal dispersions.
+Furthermore, @fig-discriminal_difference displays the discriminal difference distribution for these texts.
+This figure shows that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area.
+Consequently, the dichotomous outcome of this comparison would likely favor text A.
 
@@
 267].
-Thus, for practical application, Thurstone introduced five distinct cases derived from this general form, each defined by progressively simplifying assumptions.
-@tbl-thurstone_cases summarizes these cases, highlighting key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersions across stimuli, the correlation between stimuli, and the number of judges performing the comparisons.
-For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
+Thus, to facilitate its practical application, Thurstone developed five distinct cases derived from this general form.
+Each case progressively introduces additional simplifying assumptions related to different features of the theory, such as the distribution of discriminal processes, the uniformity of discriminal dispersions across stimuli, the correlation between stimuli, and the number of judges performing the comparisons.
+A summary of these cases is provided in @tbl-thurstone_cases.
+For a detailed examination of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
 248-253].
@@
 This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
+The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of the stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
 254] (see @tbl-thurstone_cases).
@@
 268].
-This cancellation was mathematically demonstrated by @Andrich_1978, using the BTL model under the assumption of additive biases in the discriminal processes.
+This cancellation was mathematically demonstrated by @Andrich_1978, using the BTL model under the assumption of discriminal processes with additive biases.
 However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.
@@
 <!-- 4. The first scenario where the assumption does not hold -->
-In the first scenario, the intricate aspects of multidimensional, complex traits and heterogeneous stimuli may introduce dependencies between stimuli.
-Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between heterogeneous stimuli due to certain judges' biases that resist cancellation.
+Research on text quality indicates that when judges evaluate such traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
 In this context, it is not inconceivable that these aspects, being neither equally weighted nor opposing, may unevenly influence judges' perceptions, resulting in biases that resist cancellation.
-For example, this might occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments.
-While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of judges' biases, supporting the idea that the different factors influencing pairwise comparisons may not always cancel out.
+For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments.
+While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of judges' biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
 
@@
 For instance, in cases where the CJ data included multiple samples of stimuli from the same individuals, researchers have often relied on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-This approach, however, has the significant limitation of ignoring the uncertainty associated with the scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).
+This approach, however, has the significant limitation of ignoring the uncertainty associated with the BTL scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).
 
@@
 <!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
-In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimuli is the same across stimuli.
+In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimulus is the same across all stimuli.
 While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp.
@@
 <!-- 7. But what are the challenges when these assumptions are violated? -->
-["however" paragraph]{style="color:red;"}
+["However" paragraph]{style="color:red;"}
+<!-- But the psychometric literature warns against ignoring additional traits that might confound the measurement of the trait of interest.
+Particularly, @vanderLinden_et_al_2017_I, [pp.
+346] suggests that when the assumption of unidimensionality, i.e, when the model expects to measure only one trait, is violated and more than one dimension is necessary to describe the position of the stimuli in the trait space, The association between the responses to the items given one proficiency parameter does then not vanish, i.e.
+violating the usual conditional independence assumption -->
+
+<!-- @Hoyle_et_al_2023 [pp.
+340-341] indicates that the implication of forcing multidimensional data into a unidimensional model is that the standard measurement error for the trait become too small, due to the multidimensionality.
+Moreover,  -->
+
+<!-- Suppose the assumption of unidimensionality is violated and more than one dimension is necessary to describe the test takers’ position in the latent space.
+The association between the responses to the items given one proficiency parameter does then not vanish, i.e.
+violating the usual conditional independence assumption [@vanderLinden_et_al_2017_I, pp.
+346] -->
+
+<!-- Furthermore, statistical literature also caution about ignoring clustering (grouping) structures present in the data, as these might  -->
 

### 2024-11-25 _tex/index.tex
Built site for gh-pages

--- 2024-11-22 _tex/index.tex
@@
 
-However, since the discriminal process of a single stimulus is not
-directly observable, the \emph{law of comparative judgment} becomes
-essential. This law states that in pairwise comparisons, the stimulus
-positioned further along the trait continuum is perceived as having a
-higher level of that trait. Thus, the theory assumes the observed
-dichotomous outcome is determined by the distribution of the difference
-between the underlying discriminal processes of the stimuli, referred to
-as the \emph{discriminal difference}. This indicates that the outcome
-depends on the relative distance between stimuli, rather than their
-absolute positions on the trait continuum.
+Nevertheless, the theory posits that because the discriminal process of
+a single stimulus is not directly observable, the \emph{law of
+comparative judgment} becomes essential. This law states that in
+pairwise comparisons, the stimulus positioned further along the trait
+continuum is perceived as having a higher level of that trait. This
+emphasize that the outcome depends on the relative distance between
+stimuli, rather than their absolute positions on the trait continuum.
+Thus, the theory assumes the observed dichotomous outcome is determined
+by the distribution of the difference between the underlying discriminal
+processes of the stimuli, referred to as the \emph{discriminal
+difference}.
 
@@
 instance, in the context of evaluating text quality,
-Figure~\ref{fig-discriminal_process} could depict the underlying
-discriminal process distributions for two written texts, highlighting
-differences in their discriminal dispersions and modal discriminal
-processes along the quality trait continuum. Furthermore,
-Figure~\ref{fig-discriminal_difference} could display the discriminal
-difference distribution for these texts, showing that text A is
-perceived to exhibit significantly higher quality than text B, as
+Figure~\ref{fig-discriminal_process} depicts the underlying discriminal
+process distributions for two written texts. The figure highlights
+differences in the texts' positions along the quality trait continuum,
+represented by their modal discriminal processes, as well as differences
+in the variability of their perceived trait levels, represented by their
+discriminal dispersions. Furthermore,
+Figure~\ref{fig-discriminal_difference} displays the discriminal
+difference distribution for these texts. This figure shows that text A
+is perceived to exhibit significantly higher quality than text B, as
 indicated by the shaded gray area. Consequently, the dichotomous outcome
-of this comparison would probably favor text A.
+of this comparison would likely favor text A.
 
@@
 pairwise comparisons of stimuli made by a single judge
-\citep[pp.~267]{Thurstone_1927b}. Thus, for practical application,
-Thurstone introduced five distinct cases derived from this general form,
-each defined by progressively simplifying assumptions.
-Table~\ref{tbl-thurstone_cases} summarizes these cases, highlighting key
-assumptions such as the distribution of discriminal processes, the
-similarity of discriminal dispersions across stimuli, the correlation
-between stimuli, and the number of judges performing the comparisons.
-For a comprehensive discussion of this progression, refer to
-\citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.
+\citep[pp.~267]{Thurstone_1927b}. Thus, to facilitate its practical
+application, Thurstone developed five distinct cases derived from this
+general form. Each case progressively introduces additional simplifying
+assumptions related to different features of the theory, such as the
+distribution of discriminal processes, the uniformity of discriminal
+dispersions across stimuli, the correlation between stimuli, and the
+number of judges performing the comparisons. A summary of these cases is
+provided in Table~\ref{tbl-thurstone_cases}. For a detailed examination
+of this progression, refer to \citet{Thurstone_1927b} and
+\citet[pp.~248-253]{Bramley_2008}.
 
@@
 assumptions of Case V, with one key difference: while Case V assumes a
-Normal distribution for the discriminal processes of stimuli, the BTL
-model uses the more mathematically tractable Logistic distribution
+Normal distribution for the discriminal processes of the stimuli, the
+BTL model uses the more mathematically tractable Logistic distribution
 \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
@@
 mathematically demonstrated by \citet{Andrich_1978}, using the BTL model
-under the assumption of additive biases in the discriminal processes.
+under the assumption of discriminal processes with additive biases.
 However, it is easy to imagine at least two scenarios where the zero
@@
 In the first scenario, the intricate aspects of multidimensional,
-complex traits and heterogeneous stimuli may introduce dependencies
-between stimuli. Research on text quality suggests that when judges
-evaluate these traits, they often rely on various intricate aspects of
-the stimuli to form their judgments
+complex traits may introduce dependencies between heterogeneous stimuli
+due to certain judges' biases that resist cancellation. Research on text
+quality indicates that when judges evaluate such traits, they often rely
+on various intricate aspects of the stimuli to form their judgments
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
@@
 perceptions, resulting in biases that resist cancellation. For example,
-this might occur when a judge assessing the argumentative quality of a
+this could occur when a judge assessing the argumentative quality of a
 text places disproportionate emphasis on grammatical accuracy,
@@
 \citet{Pollitt_et_al_2003} demonstrate the presence of judges' biases,
-supporting the idea that the different factors influencing pairwise
-comparisons may not always cancel out.
+supporting the idea that the factors influencing pairwise comparisons
+may not always cancel out.
 
@@
 This approach, however, has the significant limitation of ignoring the
-uncertainty associated with the scores (refer to section
+uncertainty associated with the BTL scores (refer to section
 Section~\ref{sec-theory-issue2} for a detailed discussion of this
@@
 suggests that the variability in the perceived trait level of the
-stimuli is the same across stimuli. While Thurstone acknowledged that
-this assumption may be violated when ``dealing with less conspicuous
-attributes or with less homogeneous stimuli''
+stimulus is the same across all stimuli. While Thurstone acknowledged
+that this assumption may be violated when ``dealing with less
+conspicuous attributes or with less homogeneous stimuli''
 \citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes that this
@@
 
-{``however'' paragraph}
+{``However'' paragraph}
 

### 2024-11-25 index.qmd
section:issues,CaseV; process:write

@@
 <!-- 7. But what are the challenges when these assumptions are violated? -->
-["However" paragraph]{style="color:red;"}
-<!-- But the psychometric literature warns against ignoring additional traits that might confound the measurement of the trait of interest.
-Particularly, @vanderLinden_et_al_2017_I, [pp.
-346] suggests that when the assumption of unidimensionality, i.e, when the model expects to measure only one trait, is violated and more than one dimension is necessary to describe the position of the stimuli in the trait space, The association between the responses to the items given one proficiency parameter does then not vanish, i.e.
-violating the usual conditional independence assumption -->
+But the psychometric and statistical literature strongly advises against ignoring additional traits that could confound the measurement of the trait of interest, overlooking clustering (grouping) structures, or mis-specifying the trait’s measurement model.
+First, when additional traits are ignored in favor of a unidimensional trait structure, such as when judges' biases are disregarded, the association between stimuli does not vanish [@vanderLinden_et_al_2017_II, pp.
+346].
+This mismatch can result in a potential overestimation of precision (reliability) in the measurement process [@Hoyle_et_al_2023, pp.
+340-341], or worse, introduce bias into the trait's measurement [@Ackerman_1989].
+Second, when relevant hierarchical (or grouping) structures are present in the data, ignoring these structures can further exacerbate the inflation of measurement reliability (citation needed).
+Lastly, mis-specifying the measurement model by assuming, for example, equal discriminal dispersions across stimuli can overlook meaningful differences in the reliability of the trait.
+Taken together, these mis-specifications and oversights can lead to misleading conclusions about the trait measurement, ultimately casting doubts about the validity of the trait, as validity cannot exist without reliability [@Perron_et_al_2015, pp.
+2].
 
-<!-- @Hoyle_et_al_2023 [pp.
-340-341] indicates that the implication of forcing multidimensional data into a unidimensional model is that the standard measurement error for the trait become too small, due to the multidimensionality.
-Moreover,  -->
+<!-- forcing a unidimensional trait structure on multidimensional data can inflate the perceived measurement precision (reliability) of the trait [@Hoyle_et_al_2023, pp. 340-341] -->
 
-<!-- Suppose the assumption of unidimensionality is violated and more than one dimension is necessary to describe the test takers’ position in the latent space.
-The association between the responses to the items given one proficiency parameter does then not vanish, i.e.
-violating the usual conditional independence assumption [@vanderLinden_et_al_2017_I, pp.
-346] -->
-
-<!-- Furthermore, statistical literature also caution about ignoring clustering (grouping) structures present in the data, as these might  -->
-
-<!-- Cancellation of judges: What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder.
-This, in the best scenario, cause less precise estimates by ignoring clusters, and in the worst scenario, biased estimates if the judges do register biases.
--->
-
-<!-- Multilevel structure: Here we loose precision, because we are not controlling for clustering, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.
--->
-
-<!-- Different dispersion: here we face misspecification in the model? -->
-
-<!-- This observation highlights how variability in the quality of stimuli should be accounted for when comparing heterogeneous groups or evaluating complex attributes like writing ability. -->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->
 
@@
 -->
-
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->
 

### 2024-11-25 _tex/index.tex
Built site for gh-pages

@@
 
-{``However'' paragraph}
+But the psychometric and statistical literature strongly advises against
+ignoring additional traits that could confound the measurement of the
+trait of interest, overlooking clustering (grouping) structures, or
+mis-specifying the trait's measurement model. First, when additional
+traits are ignored in favor of a unidimensional trait structure, such as
+when judges' biases are disregarded, the association between stimuli
+does not vanish \citep[pp.~346]{vanderLinden_et_al_2017_II}. This
+mismatch can result in a potential overestimation of precision
+(reliability) in the measurement process
+\citep[pp.~340-341]{Hoyle_et_al_2023}, or worse, introduce bias into the
+trait's measurement \citep{Ackerman_1989}. Second, when relevant
+hierarchical (or grouping) structures are present in the data, ignoring
+these structures can further exacerbate the inflation of measurement
+reliability (citation needed). Lastly, mis-specifying the measurement
+model by assuming, for example, equal discriminal dispersions across
+stimuli can overlook meaningful differences in the reliability of the
+trait. Taken together, these mis-specifications and oversights can lead
+to misleading conclusions about the trait measurement, ultimately
+casting doubts about the validity of the trait, as validity cannot exist
+without reliability \citep[pp.~2]{Perron_et_al_2015}.
 

### 2024-11-30 index.qmd
section:issues; process:write,figures

@@
 
+<!-- abstract -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:58:20 - 00:59:20 -->
+
 # Introduction {#sec-introduction}
@@
 Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed.
-The present study primarily focuses on three: the over-reliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role of the diverse assessment design features on CJ's reliability and validity.
+The present study primarily focuses on two: the over-reliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data, and the apparent disconnect between CJ's trait measurement and hypothesis testing.
 The following sections begin with a brief overview of Thurstone's theory and a detailed discussion of these issues.
-Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address all three concerns simultaneously.
+Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address the two concerns simultaneously.
 
@@
 <!-- recording: Sven and Tine 24.11.18; time: 00:19:04 - 00:21:37 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:00:35 - 00:06:55 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:30:35 - 00:35:10 -->
 
 <!-- 1. Thurstone's Theory of CJ: discriminal process -->
-In its most general form, Thurstone's theory [-@Thurstone_1927b] suggests that two factors determine the dichotomous outcome of pairwise comparisons: the discriminal process of each stimulus and their discriminal difference.
-The *discriminal process* refers to the psychological effect each stimulus exerts on the judges, or more simply, the underlying perception of the stimulus' trait level.
-According to the theory, the discriminal process for each stimulus follows a Normal distribution.
-The mode (mean) of this distribution, referred to as the *modal discriminal process*, represents the stimulus' position on the trait continuum.
-Meanwhile, the dispersion of the distribution, referred to as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
+In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
+267].
+The theory proposes that two key factors determine the dichotomous outcome of these designs: the discriminal process of each stimulus and their discriminal difference.
+The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level.
+According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (or mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
+
+For instance, @fig-discriminal_process illustrates the discriminal process distributions along a quality trait continuum for two written texts.
+The figure shows that these processes follow a Normal distribution.
+Moreover, it depicts differences in the texts' positions along the quality trait continuum, where text B is positioned two units further along the continuum than text A, as indicated by their modal discriminal processes ($S_{B}=2$ and $S_{A}=0$).
+Finally, it highlights differences in the texts' discriminal dispersions ($\sigma_{B}=1$ and $\sigma_{A}=0.5$), showing that text B exhibits a greater variability in its perceived quality than text A, as reflected by its wider distribution.
+
+![Example distributions of the discriminal processes for two written texts](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
 
 <!-- 2. Thurstone's Theory of CJ: discriminal difference -->
-Nevertheless, the theory posits that because the discriminal process of a single stimulus is not directly observable, the *law of comparative judgment* becomes essential.
-This law states that in pairwise comparisons, the stimulus positioned further along the trait continuum is perceived as having a higher level of that trait.
-This emphasize that the outcome depends on the relative distance between stimuli, rather than their absolute positions on the trait continuum.
-Thus, the theory assumes the observed dichotomous outcome is determined by the distribution of the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.
+However, because the discriminal process of a single stimulus is not directly observable, the theory introduces the *law of comparative judgment*.
+This law posits that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait.
+This principle highlights that the outcome of a pairwise comparison likely depends on the relative distance between stimuli rather than their absolute positions on the trait continuum.
 
-<!-- 3. Example-->
-These concepts are more easily understood through an example.
-For instance, in the context of evaluating text quality, @fig-discriminal_process depicts the underlying discriminal process distributions for two written texts.
-The figure highlights differences in the texts' positions along the quality trait continuum, represented by their modal discriminal processes, as well as differences in the variability of their perceived trait levels, represented by their discriminal dispersions.
-Furthermore, @fig-discriminal_difference displays the discriminal difference distribution for these texts.
-This figure shows that text A is perceived to exhibit significantly higher quality than text B, as indicated by the shaded gray area.
-Consequently, the dichotomous outcome of this comparison would likely favor text A.
+Thus, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*.
+Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution.
+The mode (or mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$.
+Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$.
+Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between their discriminal processes.
+This correlation quantifies how much the judge's perception of the trait in one stimulus influences his perception of the same trait in the other.
 
-::: {#fig-thurstone_theory layout-ncol=2}
+@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process, assuming a correlation of $\rho = 0.6$.
+The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$.
+As a result, the dichotomous outcome of this comparison almost certainly favors text B over text A.
 
-![Discriminal processes](/images/figures/discriminal_process.png){#fig-discriminal_process width=95%}
+![Example distribution of the discriminal difference for the two texts shown in @fig-discriminal_process, assuming a correlation of 0.6](/images/figures/discriminal_difference.png){#fig-discriminal_difference width=70%}
 
-![Discriminal difference](/images/figures/comparative_judgment.png){#fig-discriminal_difference width=100%}
+Additionally, @fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the same two texts.
+The figure shows that as the correlation increases, indicating a stronger influence of the judge's perception of quality in one text on his perception of the other, the distribution of the discriminal difference becomes narrower.
+This narrowing affects the area under the curve that determines the comparison outcome and, ultimately, the conclusions drawn from this outcome.
 
-Example distribution of discriminal processes and their discriminal difference for two written texts (stimuli or objects). Extracted from @Bramley_2008 [pp. 249-251].
-:::
+For instance, when two texts clearly differ in quality, as depicted in @fig-discriminal_process, higher correlations increase the likelihood that the discriminal difference more distinctly favors text B over text A, as indicated by the shaded gray area under the curve in @fig-correlation.
+Conversely, it is easy to infer that when the two texts have similar or identical quality levels, higher correlations reduce the likelihood that the discriminal difference distinctly favors one text over the other, as the distribution narrows around zero (not depicted in the figure).
 
-<!-- 4. Thurstone's Cases and their simplifying assumptions  -->
-Importantly, the general form of Thurstone's theory primarily addressed pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
-267].
-Thus, to facilitate its practical application, Thurstone developed five distinct cases derived from this general form.
-Each case progressively introduces additional simplifying assumptions related to different features of the theory, such as the distribution of discriminal processes, the uniformity of discriminal dispersions across stimuli, the correlation between stimuli, and the number of judges performing the comparisons.
-A summary of these cases is provided in @tbl-thurstone_cases.
-For a detailed examination of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
-248-253].
-
-![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}
+![The effect of correlation on the distribution of the discriminal difference of the same two written text](/images/figures/correlation.png){#fig-correlation width=70%}
 
@@
 <!-- recording: Sven and Tine 24.11.18; time: 00:21:37 - 00:47:20 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:06:55 - 00:24:50 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:35:10 - 00:49:30 -->
 
-<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
+<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
+The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli.
+For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions.
+Case I represents the general form of the theory.
+Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge.
+Case III introduces the assumption of zero correlation between stimuli.
+Case IV builds on this by assuming that the stimuli have similar dispersions.
+Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions.
+@tbl-thurstone_cases summarizes these cases and their assumptions.
+For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
+248-253].
+
+<!-- , suggesting all stimuli have the same variability in their perceived trait levels -->
+
+![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}
+
+<!-- 2. The Case V and its simpler statistical representation: the BTL model -->
 Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp.
@@
 677], Case V remains the most widely used case in the CJ literature.
-This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
+This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
 The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of the stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
@@
 
-::: {#fig-logistic_vs_normal layout-ncol=2}
+::: {#fig-logistic_vs_normal layout-nrow=2}
 
-![Probability density](/images/figures/density.png){#fig-density width=85%}
+![Probability density](/images/figures/density.png){#fig-density width=70%}
 
-![Cummulative probability](/images/figures/cumulative.png){#fig-cumulative width=85%}
+![Cummulative probability](/images/figures/cummulative.png){#fig-cumulative width=70%}
 
@@
 
-<!-- 2. But what are limitations of Case V (and the BTL model)? -->
-However, Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+<!-- 3. But what are limitations of Case V (and the BTL model)? -->
+However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
@@
 374].
-Consequently, given that modern CJ applications frequently involve these types of traits and stimuli, two main assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
+Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
 
@@
 
-<!-- 3. What does the zero correlation between stimuli imply? and does it hold? -->
-The assumption of *zero correlation between stimuli* can be better understood through an example.
-For instance, when using pairwise comparisons to evaluate text quality, the assumption implies that a judge's perception of a trait in one text does not influence his perception of the same trait in another text.
+<!-- 4. What does the zero correlation between stimuli imply? and does it hold? -->
+As outlined in the previous section, the assumption of *zero correlation between stimuli* suggests that during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @fig-correlation, $\rho=0$).
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
 268].
-This cancellation was mathematically demonstrated by @Andrich_1978, using the BTL model under the assumption of discriminal processes with additive biases.
-However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli, and when an additional hierarchical structure is relevant to the stimuli.
+@Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
+However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.
 
@@
 
-<!-- 4. The first scenario where the assumption does not hold -->
+<!-- 5. The first scenario where the assumption does not hold -->
 In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between heterogeneous stimuli due to certain judges' biases that resist cancellation.
-Research on text quality indicates that when judges evaluate such traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-In this context, it is not inconceivable that these aspects, being neither equally weighted nor opposing, may unevenly influence judges' perceptions, resulting in biases that resist cancellation.
+Research on text quality indicates that when judges evaluate these traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+These aspects, which are likely neither equally weighted nor opposing, may influence judges' perceptions unevenly.
 For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments.
-While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of judges' biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
+Ignoring these relevant additional "traits" can result in biases that resist cancellation, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
+346] and ultimately violating the assumption of zero correlation.
+While direct evidence for the specific mechanisms causing biases is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
+
+<!-- If the test taker’s position on one latent trait is fixed, the assumption of local stochastic independence requires the association between the responses to the items to vanish.
+Suppose the assumption is violated and more than one dimension is necessary to describe the test takers’ position in the latent space.
+The association between the responses to the items given one proficiency parameter does then not vanish.
+[@vanderLinden_et_al_2017_II, pp.
+346].
+-->
 
@@
 
-<!-- 5. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections created by the additional hierarchical structure may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
-Nevertheless, despite recognizing such hierarchical structures in CJ data, the statistical handling of this extra source of dependency in the CJ literature has been inadequate.
-For instance, in cases where the CJ data included multiple samples of stimuli from the same individuals, researchers have often relied on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-This approach, however, has the significant limitation of ignoring the uncertainty associated with the BTL scores (refer to section @sec-theory-issue2 for a detailed discussion of this issue).
+<!-- 6. The second scenario where the assumption does not hold -->
+In the second scenario, the shared context or inherent connections created by additional hierarchical structures may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
+Although the CJ literature acknowledges the presence of such hierarchical structures in CJ data, the statistical handling of this extra source of dependency has been inadequate.
+For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+This approach, however, has the significant limitation of ignoring the uncertainty associated with the BTL scores, which generates additional statistical and measurement issues, as discussed in section @sec-theory-issue2.
 
-<!-- (Is this a validity contention?) -->
-<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
-Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp.
-683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement.
-As @Kelly_et_al_2022 noted [pp.
-683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->
+<!-- 7. But what are the challenges when these assumptions are violated? -->
+In any case, the psychometric and statistical literature strongly advises against ignoring relevant additional traits or overlooking clustering (grouping) structures that may introduce dependencies between stimuli.
+First, ignoring additional aspects pertinent to the trait of interest can cause a dimensional mismatch, which can lead to an overestimation of the precision (reliability) of the trait's measurement [@Hoyle_et_al_2023, pp.
+340-341] or, worse, introduce bias into the measurement process [@Ackerman_1989].
+Additionally, overlooking relevant hierarchical (or grouping) structures can exacerbate the inflation of reliability by reducing the accuracy of the model's parameters [@Hoyle_et_al_2023, pp.
+482].
+Taken together, these oversights undermine the reliability of the measurement process and, since validity cannot exist without reliability [@Perron_et_al_2015, pp.
+2], ultimately compromise the validity of the results derived from it.
 
-<!-- Such differences may not be detected through analyses of bias and misfit.
-McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
-Where students prioritised factual recall, teachers prioritised scientific explanation.
-Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
-Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
-[@Kelly_et_al_2022, pp.
-683] -->
+<!-- forcing a unidimensional trait structure on multidimensional data can inflate the perceived measurement precision (reliability) of the trait [@Hoyle_et_al_2023, pp. 340-341] -->
 
-<!-- 6. What does the equal dispersion between stimuli imply? and does it hold? -->
-In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimulus is the same across all stimuli.
+<!-- Reliability is a necessary but not sufficient condition for validity.
+Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp.
+2].
+If your scores are not reliable they cannot be valid.
+People have been employing a model with assumptions that they do not know if those assumptions apply for their cases.
+And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid.
+-->
+
+<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem. -->
+
+<!-- 8. So, how do we mitigate these risks? -->
+<!-- Luckily, @Andrich_1978 and @Wainer_et_al_1978 already proposed a solution to the first scenario where the assumption of zero correlation is violated.
+Both authors demonstrated that by parametrizing judges' biases in the measurement model of the trait of interest, their effects are experimentally eliminated from the trait's measurement process.
+Furthermore, a wealth of statistical literature offers a solution to the second scenario: incorporate the hierarchical structure into the measurement model of the trait of interest, similar to a multilevel SEM model [@Hoyle_et_al_2023, chap.
+26].
+-->
+
+<!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
+
+<!-- 9. What does the equal dispersion between stimuli imply? and does it hold? -->
+<!-- In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimulus is the same across all stimuli.
 While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp.
-374], no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
-One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
-In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
-Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
+374], ...
+-->
+
+<!-- It is a ridiculous assumption, because multidimensional, complex traits -->
+<!-- The dispersion is not the same for every text, a weak writer will produce consistently weak quality texts, in all the dimensions, this implies that every judge will agree on its lower quality, thereby the dispersion of the texts is going to be narrow.
+In contrast, when the texts move further along the quality trait continuum, implying that some areas of the texts are improved while some other remains, then there is a higher potential that there are discrepancies between the judges, as they focus in different aspects of the stimuli.
+So it seem that the extreme of the discriminal process distribution of the stimuli have higher reliability than the center?
+-->
+
+<!-- Some text will provoke more variability in the judgments, and that is because of the text. There is something in the text that cause doubts in the judges. -->
+
+<!-- Example of creativity from Tine.
+Misfiting text might be texts with larger dispersion, because there is a lot of confusion about them.
+When assessing creativity, for instance, student can do an assignment in really creative but an unexpected way, but because the judges did not expected such a way, they do not what to do with it, and that creates more dispersion in the judgment.
+Then the discriminal difference is all over the place.
+-->
 
@@
 
-<!-- 7. But what are the challenges when these assumptions are violated? -->
-But the psychometric and statistical literature strongly advises against ignoring additional traits that could confound the measurement of the trait of interest, overlooking clustering (grouping) structures, or mis-specifying the trait’s measurement model.
-First, when additional traits are ignored in favor of a unidimensional trait structure, such as when judges' biases are disregarded, the association between stimuli does not vanish [@vanderLinden_et_al_2017_II, pp.
-346].
-This mismatch can result in a potential overestimation of precision (reliability) in the measurement process [@Hoyle_et_al_2023, pp.
-340-341], or worse, introduce bias into the trait's measurement [@Ackerman_1989].
-Second, when relevant hierarchical (or grouping) structures are present in the data, ignoring these structures can further exacerbate the inflation of measurement reliability (citation needed).
-Lastly, mis-specifying the measurement model by assuming, for example, equal discriminal dispersions across stimuli can overlook meaningful differences in the reliability of the trait.
+<!-- 10. But what are the challenges when these assumptions are violated? -->
+<!-- But the statistical literature warns against mis-specifying the trait’s measurement model.
+When mis-specifying the measurement model by assuming, for example, equal discriminal dispersions across stimuli, the model can overlook meaningful differences in the reliability of the trait.
 Taken together, these mis-specifications and oversights can lead to misleading conclusions about the trait measurement, ultimately casting doubts about the validity of the trait, as validity cannot exist without reliability [@Perron_et_al_2015, pp.
 2].
-
-<!-- forcing a unidimensional trait structure on multidimensional data can inflate the perceived measurement precision (reliability) of the trait [@Hoyle_et_al_2023, pp. 340-341] -->
+-->
 
@@
 
-<!-- 8. So, how do we mitigate these risks? -->
-["to mitigate" paragraph]{style="color:red;"}
-
-<!-- Moreover, a solution was already envisioned, @Andrich_1978 and @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.
--->
+<!-- 11. So, how do we mitigate these risks? -->
+<!-- ["to mitigate" paragraph]{style="color:red;"} -->
 
@@
 <!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
-<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:10 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
 
@@
 <!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses and tests, or to make decisions regarding the exclusion of certain data in these analyses and tests.
-The literature shows that these scores have been employed to calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023] or to test hypotheses related to the underlying traits of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-Additionally, the BTL scores have been used to detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], as well as to identify "misfit" judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], with considerations for their possible exclusion.
+Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
+For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
 <!-- 3. But what are the challenges in using BTL scores for further analysis? -->
-However, the statistical literature advises caution when using estimated scores for additional analyses and tests, as well as when eliminating data through ad hoc univariate procedures.
+However, the statistical literature advises caution when using estimated scores for additional analyses and tests.
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
@@
 137].
-Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020].
 Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
@@
 
-## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3}
+<!-- ## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3} -->
+<!-- Do not include yet!, until you see if you need it when you write about the next section -->
+<!-- recording: Sven and Tine 24.11.18; time: 00:47:20 - 00:56:41 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:53:30 - 01:00:00 -->
 
-<!-- recording: Sven and Tine 24.11.18; time: 00:47:20 - 00:56:41 -->
+# An updated theoretical and statistical model for CJ {#sec-theory}
 
 <!-- []{style="color:red;"} -->
-
-<!-- Experimental design -->
-
-<!-- In this regard, although the literature has tested multiple features of the experimental design of the method, such as the comparison algorithm and the number of comparisons, it is not clear where this procedures fit in the whole process of comparison, and what are their implications for the outcome of the method.
--->
-
-<!-- @Bramley_et_al_2019 and @Bramley_2015 says adaptivity inflates reliability.
-present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required or how many judges should be involved [@Kelly_et_al_2022, pp.
-682].
-Adhoc rules of thumb.
--->
-
-<!-- @Bramley_2015 -->
-<!-- The implication is that the SSR statistic is worthless as an indicator of the quality / consistency / reliability of a scale constructed from an ACJ study that has involved a significant number of adaptive rounds.
-With random or fixed rounds, the SSR values were all below 0.25, which would generally be taken as an indicator of failure to create a meaningful scale.
-To put it another way, a low value of SSR almost certainly indicates low reliability, but a high value of SSR does not necessarily indicate high reliability.
--->
-<!-- This is not to say that the ACJ studies listed in Table 1 (or indeed other studies not listed) were themselves worthless, just that not much should be made of claims about very high reliability in ACJ studies on the basis of high values of the SSR statistic.
-In fact, many of those studies also showed moderate to high correlations with variables external to the analysis, which were rightly interpreted as evidence of concurrent validity.
--->
-<!-- The conclusion is therefore that the SSR statistic is at best misleading and at worst worthless as an indicator of scale reliability whenever a CJ study has involved a significant amount of adaptivity.
--->
-
-<!-- @Gray_et_al_2024 entropy-driven active learning pair-selection algorithm to compare stimuli, vs random and no repeating pairs -->
-
-<!-- @DeVrindt_et_al_2024 cold start problem to optimize time efficiency. -->
-
-<!-- @Mikhailiuk_et_al_2021 proposed a fully Bayesian active sampling strategy for pairwise comparisons via Approximate Message Passing and Information Gain Maximization.
-ASAP computes the full posterior distribution, which is crucial to achieving accurate EIG estimates, and thus the accuracy of active sampling.
-We recommend ASAP, as it offered the highest accuracy of inferred scores compared to existing methods in experiments with real and synthetic data.
--->
-
-<!-- @Verhavert_et_al_2022 The current article proposes a new adaptive selection algorithm using a previously calibrated reference set.
-Using a reference set should eliminate the reliability inflation.
-It is proven that this adaptive selection algorithm is more efficient without reducing the accuracy of the results and without increasing the standard deviation of the assessment results.
-Akin to BIB designs and more specifically to BTIB designs  -->
-
-<!-- and issues related to the method's experimental design. These issues can affect the method's reliability, validity, and practical applicability. -->
-
-<!-- @Coertjens_et_al_2017 assess reliability and time investment.
-We conclude that pairwise comparison requires a similar time investment from evaluators as the criterion list method.
-We want to note that the outcomes in this study depend heavily on the design.
-First, the time taken for an assessment is strongly determined by the length of the criteria list used and the length of the texts (Breland, 1983).
-Not adaptive -->
-
-<!-- Although the process of pairwise comparisons is quite easy and fast for raters to apply (@Laming_2004; @Pollitt_2004), even without training, the increase in time investment that is needed for CJ decreases its feasibility in educational practice.
--->
-
-<!-- Messick_1989 talks about efficiency and feasibility for application  -->
-
-<!-- @Bramley_2008 proposes ranking instead of comparisons -->
-
-<!-- talk about the number of comparisons @Verhavert_et_al_2019, @Crompvoets_et_al_2022, @Pollitt_2004 -->
-
-<!-- @Pollitt_2012b concludes that speed is not related to the quality of assessments in pairwise comparisons -->
-
-<!-- All studies use SSR as a measure of reliability -->
-
-<!-- It is assumed that the single observer compares each pair of stimuli a sufficient number of times so that a proportion may be determined for each pair [@Thurstone_1927b, pp. 267]. -->
-
-# An updated theoretical and statistical model for CJ {#sec-theory}
 
@@
 <!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
+<!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->
 
@@
 
+<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
+One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
+In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
+Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
+-->
+
 ## From theory to statistics {#sec-theory-statistics}
@@
 <!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
 
@@
 
+<!-- , as well as when eliminating data through ad hoc univariate procedures -->
+<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
+
 # Discussion {#sec-discuss}
@@
 
-<!-- "If Thurstone’s Law is to be used in support of comparative judgment for -->
-<!-- assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes." [@Kelly_et_al_2022, pp. 678] -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:56:10 -->
+
+<!-- "If Thurstone’s Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes." [@Kelly_et_al_2022, pp.
+678] -->
 
@@
 -->
+
+<!-- (Is this a validity contention?) -->
+<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
+Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp.
+683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement.
+As @Kelly_et_al_2022 noted [pp.
+683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->
+
+<!-- Such differences may not be detected through analyses of bias and misfit.
+McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
+Where students prioritized factual recall, teachers prioritized scientific explanation.
+Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
+Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
+[@Kelly_et_al_2022, pp.
+683] -->
 
@@
 683] -->
+
+<!-- @Bramley_2008 proposes ranking instead of comparisons -->
 
@@
 
-# Appendix {#sec-appendix}
-
-{{< pagebreak >}}
-
 # References {.unnumbered}

### 2024-11-30 _tex/index.tex
Built site for gh-pages

--- 2024-11-25 _tex/index.tex
@@
 and fragmented research approaches have left several critical issues
-unaddressed. The present study primarily focuses on three: the
+unaddressed. The present study primarily focuses on two: the
 over-reliance on the assumptions of Thurstone's Case V in the
-statistical analysis of CJ data, the apparent disconnect between CJ's
-trait measurement and hypothesis testing, and the unclear role of the
-diverse assessment design features on CJ's reliability and validity. The
-following sections begin with a brief overview of Thurstone's theory and
-a detailed discussion of these issues. Subsequently, the study
-introduces a theoretical model for CJ that builds upon Thurstone's
-theory, alongside its statistical translation, designed to address all
-three concerns simultaneously.
+statistical analysis of CJ data, and the apparent disconnect between
+CJ's trait measurement and hypothesis testing. The following sections
+begin with a brief overview of Thurstone's theory and a detailed
+discussion of these issues. Subsequently, the study introduces a
+theoretical model for CJ that builds upon Thurstone's theory, alongside
+its statistical translation, designed to address the two concerns
+simultaneously.
 
@@
 
-In its most general form, Thurstone's theory
-\citeyearpar{Thurstone_1927b} suggests that two factors determine the
-dichotomous outcome of pairwise comparisons: the discriminal process of
-each stimulus and their discriminal difference. The \emph{discriminal
-process} refers to the psychological effect each stimulus exerts on the
-judges, or more simply, the underlying perception of the stimulus' trait
-level. According to the theory, the discriminal process for each
-stimulus follows a Normal distribution. The mode (mean) of this
-distribution, referred to as the \emph{modal discriminal process},
-represents the stimulus' position on the trait continuum. Meanwhile, the
-dispersion of the distribution, referred to as the \emph{discriminal
-dispersion}, reflects the variability in the perceived trait level of
-the stimulus.
+In its most general form, Thurstone's theory deals with pairwise
+comparisons of stimuli made by a single judge
+\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
+factors determine the dichotomous outcome of these designs: the
+discriminal process of each stimulus and their discriminal difference.
+The \emph{discriminal process} represents the psychological impact each
+stimulus has on judges or, more simply, their underlying perception of
+the stimulus' trait level. According to the theory, the discriminal
+process for each stimulus follows a Normal distribution, where its mode
+(or mean), referred to as the \emph{modal discriminal process},
+indicates the stimulus' position on the trait continuum, and its
+dispersion, known as the \emph{discriminal dispersion}, reflects the
+variability in the perceived trait level of the stimulus.
 
-Nevertheless, the theory posits that because the discriminal process of
-a single stimulus is not directly observable, the \emph{law of
-comparative judgment} becomes essential. This law states that in
-pairwise comparisons, the stimulus positioned further along the trait
-continuum is perceived as having a higher level of that trait. This
-emphasize that the outcome depends on the relative distance between
-stimuli, rather than their absolute positions on the trait continuum.
-Thus, the theory assumes the observed dichotomous outcome is determined
-by the distribution of the difference between the underlying discriminal
-processes of the stimuli, referred to as the \emph{discriminal
-difference}.
-
-These concepts are more easily understood through an example. For
-instance, in the context of evaluating text quality,
-Figure~\ref{fig-discriminal_process} depicts the underlying discriminal
-process distributions for two written texts. The figure highlights
-differences in the texts' positions along the quality trait continuum,
-represented by their modal discriminal processes, as well as differences
-in the variability of their perceived trait levels, represented by their
-discriminal dispersions. Furthermore,
-Figure~\ref{fig-discriminal_difference} displays the discriminal
-difference distribution for these texts. This figure shows that text A
-is perceived to exhibit significantly higher quality than text B, as
-indicated by the shaded gray area. Consequently, the dichotomous outcome
-of this comparison would likely favor text A.
+For instance, Figure~\ref{fig-discriminal_process} illustrates the
+discriminal process distributions along a quality trait continuum for
+two written texts. The figure shows that these processes follow a Normal
+distribution. Moreover, it depicts differences in the texts' positions
+along the quality trait continuum, where text B is positioned two units
+further along the continuum than text A, as indicated by their modal
+discriminal processes (\(S_{B}=2\) and \(S_{A}=0\)). Finally, it
+highlights differences in the texts' discriminal dispersions
+(\(\sigma_{B}=1\) and \(\sigma_{A}=0.5\)), showing that text B exhibits
+a greater variability in its perceived quality than text A, as reflected
+by its wider distribution.
 
@@
 
-\begin{minipage}{0.50\linewidth}
+\centering{
+
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}
+
+}
+
+\caption{\label{fig-discriminal_process}Example distributions of the
+discriminal processes for two written texts}
+
+\end{figure}%
+
+However, because the discriminal process of a single stimulus is not
+directly observable, the theory introduces the \emph{law of comparative
+judgment}. This law posits that in pairwise comparisons, a judge
+perceives the stimulus positioned further along the trait continuum as
+having a higher level of that trait. This principle highlights that the
+outcome of a pairwise comparison likely depends on the relative distance
+between stimuli rather than their absolute positions on the trait
+continuum.
+
+Thus, the theory assumes that the observed dichotomous outcome arises
+from the distribution of the difference between the underlying
+discriminal processes of the stimuli, known as the \emph{discriminal
+difference}. Since the individual discriminal processes follow a Normal
+distribution, their difference also follows a Normal distribution. The
+mode (or mean) of this distribution, representing the (average) relative
+separation, is given by the difference between the modal discriminal
+processes of the stimuli \(S_{BA} = S_{B} - S_{A}\). Meanwhile, the
+dispersion of the distribution, reflecting the variability in the
+relative separation, is calculated as
+\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
+Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
+discriminal dispersions, while \(\rho\) measures the correlation between
+their discriminal processes. This correlation quantifies how much the
+judge's perception of the trait in one stimulus influences his
+perception of the same trait in the other.
+
+Figure~\ref{fig-discriminal_difference} shows the distribution of the
+discriminal difference for the texts depicted in
+Figure~\ref{fig-discriminal_process}, assuming a correlation of
+\(\rho = 0.6\). The figure reveals that, under these conditions, the
+judge perceives text B as having significantly higher quality than text
+A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
+a result, the dichotomous outcome of this comparison almost certainly
+favors text B over text A.
+
+\begin{figure}
 
@@
 
-\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_difference.png}
 
@@
 
-\subcaption{\label{fig-discriminal_process}Discriminal processes}
+\caption{\label{fig-discriminal_difference}Example distribution of the
+discriminal difference for the two texts shown in
+Figure~\ref{fig-discriminal_process}, assuming a correlation of 0.6}
 
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
+\end{figure}%
+
+Additionally, Figure~\ref{fig-correlation} illustrates how varying
+correlations influence the distribution of the discriminal difference
+for the same two texts. The figure shows that as the correlation
+increases, indicating a stronger influence of the judge's perception of
+quality in one text on his perception of the other, the distribution of
+the discriminal difference becomes narrower. This narrowing affects the
+area under the curve that determines the comparison outcome and,
+ultimately, the conclusions drawn from this outcome.
+
+For instance, when two texts clearly differ in quality, as depicted in
+Figure~\ref{fig-discriminal_process}, higher correlations increase the
+likelihood that the discriminal difference more distinctly favors text B
+over text A, as indicated by the shaded gray area under the curve in
+Figure~\ref{fig-correlation}. Conversely, it is easy to infer that when
+the two texts have similar or identical quality levels, higher
+correlations reduce the likelihood that the discriminal difference
+distinctly favors one text over the other, as the distribution narrows
+around zero (not depicted in the figure).
+
+\begin{figure}
 
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/figures/comparative_judgment.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/correlation.png}
 
@@
 
-\subcaption{\label{fig-discriminal_difference}Discriminal difference}
-
-\end{minipage}%
-
-\caption{\label{fig-thurstone_theory}Example distribution of discriminal
-processes and their discriminal difference for two written texts
-(stimuli or objects). Extracted from \citet[pp.~249-251]{Bramley_2008}.}
+\caption{\label{fig-correlation}The effect of correlation on the
+distribution of the discriminal difference of the same two written text}
 
@@
 
-Importantly, the general form of Thurstone's theory primarily addressed
-pairwise comparisons of stimuli made by a single judge
-\citep[pp.~267]{Thurstone_1927b}. Thus, to facilitate its practical
-application, Thurstone developed five distinct cases derived from this
-general form. Each case progressively introduces additional simplifying
-assumptions related to different features of the theory, such as the
-distribution of discriminal processes, the uniformity of discriminal
-dispersions across stimuli, the correlation between stimuli, and the
-number of judges performing the comparisons. A summary of these cases is
-provided in Table~\ref{tbl-thurstone_cases}. For a detailed examination
-of this progression, refer to \citet{Thurstone_1927b} and
-\citet[pp.~248-253]{Bramley_2008}.
+\section{Three critical issues in CJ
+literature}\label{sec-theory-issues}
+
+\subsection{The Case V and the statistical analysis of CJ
+data}\label{sec-theory-issue1}
+
+The previous section outlines the general form of Thurstone's theory,
+which applies to a CJ design where a single judge evaluates multiple
+stimuli. For the practical application of the theory, Thurstone
+developed four additional cases derived from this general form, where
+each successive case incorporates additional simplifying assumptions.
+Case I represents the general form of the theory. Case II extends this
+by allowing multiple judges to make comparisons rather than restricting
+the comparisons to a single judge. Case III introduces the assumption of
+zero correlation between stimuli. Case IV builds on this by assuming
+that the stimuli have similar dispersions. Finally, Case V replaces this
+assumption with the condition that the stimuli have equal discriminal
+dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
+their assumptions. For a detailed discussion of this progression, refer
+to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.
 
@@
 
-\section{Three critical issues in CJ
-literature}\label{sec-theory-issues}
-
-\subsection{The Case V and the statistical analysis of CJ
-data}\label{sec-theory-issue1}
-
 Despite its reliance on the largest number of simplifying assumptions
@@
 Case V remains the most widely used case in the CJ literature. This
-popularity is largely due to its simplified statistical representation
+popularity stems mainly from its simplified statistical representation
 in the Bradley-Terry-Luce (BTL) model
@@
 
-\begin{minipage}{0.50\linewidth}
+\begin{minipage}{\linewidth}
 
@@
 
-\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}
 
@@
 \end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
+\newline
+\begin{minipage}{\linewidth}
 
@@
 
-\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/figures/cumulative.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/cummulative.png}
 
@@
 
-However, Case V was originally developed to provide a ``rather coarse
-scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
-statistical simplicity over precision in trait measurement
+However, Thurstone originally developed Case V to provide a ``rather
+coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
+prioritizing statistical simplicity over precision in trait measurement
 \citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
@@
 \citep[pp.~374]{Thurstone_1927a}. Consequently, given that modern CJ
-applications frequently involve these types of traits and stimuli, two
-main assumptions of Case V may not consistently hold in theory or
-practice: the zero correlation and equal dispersion between stimuli.
+applications frequently involve these traits and stimuli, two main
+assumptions of Case V may not consistently hold in theory or practice:
+the zero correlation and equal dispersion between stimuli.
 
-The assumption of \emph{zero correlation between stimuli} can be better
-understood through an example. For instance, when using pairwise
-comparisons to evaluate text quality, the assumption implies that a
-judge's perception of a trait in one text does not influence his
-perception of the same trait in another text. Thurstone attributed this
+As outlined in the previous section, the assumption of \emph{zero
+correlation between stimuli} suggests that during a pairwise comparison,
+a judge's perception of quality in one text does not influence his
+perception of the same trait in another text (see
+Figure~\ref{fig-correlation}, \(\rho=0\)). Thurstone attributed this
 independence to the cancellation of potential judges' biases, driven by
 two opposing and equally weighted effects occurring during the pairwise
-comparisons \citep[pp.~268]{Thurstone_1927b}. This cancellation was
-mathematically demonstrated by \citet{Andrich_1978}, using the BTL model
-under the assumption of discriminal processes with additive biases.
-However, it is easy to imagine at least two scenarios where the zero
-correlation assumption almost certainly does not hold: when the pairwise
-comparison involves multidimensional, complex traits with heterogeneous
-stimuli, and when an additional hierarchical structure is relevant to
-the stimuli.
+comparisons \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978}
+mathematically demonstrated this cancellation using the BTL model under
+the assumption of discriminal processes with additive biases. However,
+it is easy to imagine at least two scenarios where the zero correlation
+assumption almost certainly does not hold: when the pairwise comparison
+involves multidimensional, complex traits with heterogeneous stimuli and
+when an additional hierarchical structure is relevant to the stimuli.
 
@@
 due to certain judges' biases that resist cancellation. Research on text
-quality indicates that when judges evaluate such traits, they often rely
-on various intricate aspects of the stimuli to form their judgments
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. In
-this context, it is not inconceivable that these aspects, being neither
-equally weighted nor opposing, may unevenly influence judges'
-perceptions, resulting in biases that resist cancellation. For example,
-this could occur when a judge assessing the argumentative quality of a
-text places disproportionate emphasis on grammatical accuracy,
-ultimately favoring texts with fewer errors but weaker arguments. While
-direct evidence for this specific scenario is lacking, studies such as
-\citet{Pollitt_et_al_2003} demonstrate the presence of judges' biases,
+quality indicates that when judges evaluate these traits, they often
+rely on various intricate aspects of the stimuli to form their judgments
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
+aspects, which are likely neither equally weighted nor opposing, may
+influence judges' perceptions unevenly. For example, this could occur
+when a judge assessing the argumentative quality of a text places
+disproportionate emphasis on grammatical accuracy, ultimately favoring
+texts with fewer errors but weaker arguments. Ignoring these relevant
+additional ``traits'' can result in biases that resist cancellation,
+introducing dependencies between stimuli
+\citep[pp.~346]{vanderLinden_et_al_2017_II} and ultimately violating the
+assumption of zero correlation. While direct evidence for the specific
+mechanisms causing biases is lacking, studies such as
+\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
 supporting the idea that the factors influencing pairwise comparisons
@@
 In the second scenario, the shared context or inherent connections
-created by the additional hierarchical structure may introduce
-dependencies between stimuli, a statistical phenomenon commonly known as
-clustering \citep{Everitt_et_al_2010}. Nevertheless, despite recognizing
-such hierarchical structures in CJ data, the statistical handling of
-this extra source of dependency in the CJ literature has been
-inadequate. For instance, in cases where the CJ data included multiple
-samples of stimuli from the same individuals, researchers have often
-relied on (averaged) estimated BTL scores to conduct subsequent analyses
-and tests at the individual hierarchical level
+created by additional hierarchical structures may introduce dependencies
+between stimuli, a statistical phenomenon commonly known as clustering
+\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
+presence of such hierarchical structures in CJ data, the statistical
+handling of this extra source of dependency has been inadequate. For
+example, when CJ data includes multiple samples of stimuli from the same
+individuals, researchers often rely on (average) estimated BTL scores to
+conduct subsequent analyses and tests at the individual hierarchical
+level
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
 This approach, however, has the significant limitation of ignoring the
-uncertainty associated with the BTL scores (refer to section
-Section~\ref{sec-theory-issue2} for a detailed discussion of this
-issue).
+uncertainty associated with the BTL scores, which generates additional
+statistical and measurement issues, as discussed in section
+Section~\ref{sec-theory-issue2}.
 
-In contrast, the assumption of \emph{equal dispersion between stimuli}
-suggests that the variability in the perceived trait level of the
-stimulus is the same across all stimuli. While Thurstone acknowledged
-that this assumption may be violated when ``dealing with less
-conspicuous attributes or with less homogeneous stimuli''
-\citep[pp.~374]{Thurstone_1927a}, no study explicitly proposes that this
-assumption could also be violated due to the presence of an additional
-hierarchical (grouping) structure relevant to the texts. One such
-scenario might arise, for example, when comparing texts produced by
-university and secondary school students. In this case, university
-students may consistently (or more precisely) produce higher-quality
-texts, while secondary school students, who exhibit a broader range of
-writing abilities, would show greater variability in the quality of
-their texts. Although this example is somewhat contrived, it effectively
-illustrates how assuming equal dispersions across texts can overlook
-meaningful differences in the reliability of text quality across groups
-or individuals.
-
-{``However'' paragraph}
-
-{``to mitigate'' paragraph}
+In any case, the psychometric and statistical literature strongly
+advises against ignoring relevant additional traits or overlooking
+clustering (grouping) structures that may introduce dependencies between
+stimuli. First, ignoring additional aspects pertinent to the trait of
+interest can cause a dimensional mismatch, which can lead to an
+overestimation of the precision (reliability) of the trait's measurement
+\citep[pp.~340-341]{Hoyle_et_al_2023} or, worse, introduce bias into the
+measurement process \citep{Ackerman_1989}. Additionally, overlooking
+relevant hierarchical (or grouping) structures can exacerbate the
+inflation of reliability by reducing the accuracy of the model's
+parameters \citep[pp.~482]{Hoyle_et_al_2023}. Taken together, these
+oversights undermine the reliability of the measurement process and,
+since validity cannot exist without reliability
+\citep[pp.~2]{Perron_et_al_2015}, ultimately compromise the validity of
+the results derived from it.
 
@@
 Researchers then typically use the estimated BTL scores, or their
-transformations, to conduct additional analyses and tests, or to make
-decisions regarding the exclusion of certain data in these analyses and
-tests. The literature shows that these scores have been employed to
-calculate correlations with other assessment methods
-\citep{Goossens_et_al_2018, Bouwer_et_al_2023} or to test hypotheses
-related to the underlying traits of interest
+transformations, to conduct additional analyses or hypothesis tests. For
+example, these scores have been used to identify `misfit' judges and
+stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018},
+detect biases in judges' ratings
+\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
+other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
+or test hypotheses related to the underlying trait of interest
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-Additionally, the BTL scores have been used to detect biases in judges'
-ratings \citep{Pollitt_et_al_2003, Pollitt_2012b}, as well as to
-identify ``misfit'' judges and stimuli
-\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, with
-considerations for their possible exclusion.
 
 However, the statistical literature advises caution when using estimated
-scores for additional analyses and tests, as well as when eliminating
-data through ad hoc univariate procedures. A key consideration is that
+scores for additional analyses and tests. A key consideration is that
 BTL scores are parameter estimates that inherently carry uncertainty.
@@
 \citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
-Moreover, excluding data using ad hoc univariate procedures can compound
-these issues by discarding potentially valuable information, further
-exacerbating the bias \citep{Zimmerman_1994, McElreath_2020}. Finally,
-the reduced precision in hypothesis tests diminishes their statistical
-power, increasing the likelihood of committing type-I or type-II errors
-\citep{McElreath_2020}.
+Finally, the reduced precision in hypothesis tests diminishes their
+statistical power, increasing the likelihood of committing type-I or
+type-II errors \citep{McElreath_2020}.
 
@@
 measurement and structural models can offer significant advantages.
-
-\subsection{The diverse assessment design features and their role on
-reliability and validity}\label{sec-theory-issue3}
 
@@
 
-\section{Appendix}\label{sec-appendix}
-
-\newpage{}
-
 \section*{References}\label{references}

### 2024-11-30 code/figures.R

@@ -1,0 +1,192 @@
+# working area ####
+rm(list=ls())
+
+main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript/images/figures/'
+
+
+
+# functions ####
+colorArea <- function(from, to, density, ..., col="lightgray", dens=NULL){
+  y_seq = seq(from, to, length.out=500)
+  d = c(0, density(y_seq, ...), 0)
+  polygon(c(from, y_seq, to), d, col=col, density=dens, border=NA)
+}
+
+
+
+# figure 1a ####
+
+# distributional parameters
+dpA = c(0, 0.5) # distribution text A (mean, sd)
+dpB = c(2, 1) # distribution text B (mean, sd)
+rAB = 0.6 # correlation between discriminal processes
+
+# figure
+png( filename=file.path(main_dir, 'discriminal_process.png'), 
+     height=10, width=15, units='cm', res=400  )
+
+# initial figure
+par( mar=c(2,0,0.5,0) )
+plot( NULL, xlim=c(-2,6), ylim=c(0,0.9), 
+      xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
+axis( side=1, labels=F, lwd.ticks=0 )
+axis( side=1, at=c(dpA[1], dpB[1]), tick=F, 
+      label=c( expression( S[A] ), expression( S[B] ) ) )
+
+# figure parameters
+pplotA = c( dnorm( dpA[1], mean=dpA[1], sd=dpA[2] ),
+            dnorm( dpA[1], mean=dpA[1], sd=dpA[2] ) + 0.05,
+            dnorm( dpA[1] + dpA[2], mean=dpA[1], sd=dpA[2]),
+            dnorm( dpA[1]+0.25, mean=dpA[1], sd=dpA[2] )-0.25 )
+
+pplotB = c( dnorm( dpB[1], mean=dpB[1], sd=dpB[2] ),
+            dnorm( dpB[1], mean=dpB[1], sd=dpB[2] ) + 0.05,
+            dnorm( dpB[1] + dpB[2], mean=dpB[1], sd=dpB[2]),
+            dnorm( dpB[1]+0.45, mean=dpB[1], sd=dpB[2] )-0.15 )
+
+
[…]

### 2024-11-30 index.qmd
section:issues; process:write,figures

@@
 The figure shows that these processes follow a Normal distribution.
-Moreover, it depicts differences in the texts' positions along the quality trait continuum, where text B is positioned two units further along the continuum than text A, as indicated by their modal discriminal processes ($S_{B}=2$ and $S_{A}=0$).
-Finally, it highlights differences in the texts' discriminal dispersions ($\sigma_{B}=1$ and $\sigma_{A}=0.5$), showing that text B exhibits a greater variability in its perceived quality than text A, as reflected by its wider distribution.
+Moreover, it depicts differences in the texts' positions along the quality trait continuum, where text B is positioned further along the continuum than text A, as indicated by their modal discriminal processes ($S_{B}$ and $S_{A}$).
+Finally, it highlights differences in the texts' discriminal dispersions ($\sigma_{B}$ and $\sigma_{A}$), showing that text B exhibits a greater variability in its perceived quality than text A, as reflected by its wider distribution.
 
@@
 
-Thus, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*.
+Indeed, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*.
 Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution.
@@
 For instance, when two texts clearly differ in quality, as depicted in @fig-discriminal_process, higher correlations increase the likelihood that the discriminal difference more distinctly favors text B over text A, as indicated by the shaded gray area under the curve in @fig-correlation.
-Conversely, it is easy to infer that when the two texts have similar or identical quality levels, higher correlations reduce the likelihood that the discriminal difference distinctly favors one text over the other, as the distribution narrows around zero (not depicted in the figure).
+Conversely, when the two texts have similar or identical quality levels, higher correlations reduce the likelihood that the discriminal difference distinctly favors one text over the other, as the distribution narrows around zero (not depicted in the figure).
 

### 2024-11-30 _tex/index.tex
Built site for gh-pages

@@
 distribution. Moreover, it depicts differences in the texts' positions
-along the quality trait continuum, where text B is positioned two units
-further along the continuum than text A, as indicated by their modal
-discriminal processes (\(S_{B}=2\) and \(S_{A}=0\)). Finally, it
-highlights differences in the texts' discriminal dispersions
-(\(\sigma_{B}=1\) and \(\sigma_{A}=0.5\)), showing that text B exhibits
-a greater variability in its perceived quality than text A, as reflected
-by its wider distribution.
+along the quality trait continuum, where text B is positioned further
+along the continuum than text A, as indicated by their modal discriminal
+processes (\(S_{B}\) and \(S_{A}\)). Finally, it highlights differences
+in the texts' discriminal dispersions (\(\sigma_{B}\) and
+\(\sigma_{A}\)), showing that text B exhibits a greater variability in
+its perceived quality than text A, as reflected by its wider
+distribution.
 
@@
 
-Thus, the theory assumes that the observed dichotomous outcome arises
+Indeed, the theory assumes that the observed dichotomous outcome arises
 from the distribution of the difference between the underlying
@@
 over text A, as indicated by the shaded gray area under the curve in
-Figure~\ref{fig-correlation}. Conversely, it is easy to infer that when
-the two texts have similar or identical quality levels, higher
-correlations reduce the likelihood that the discriminal difference
-distinctly favors one text over the other, as the distribution narrows
-around zero (not depicted in the figure).
+Figure~\ref{fig-correlation}. Conversely, when the two texts have
+similar or identical quality levels, higher correlations reduce the
+likelihood that the discriminal difference distinctly favors one text
+over the other, as the distribution narrows around zero (not depicted in
+the figure).
 

### 2024-12-02 index.qmd
section:thurstone; process:write

@@
 267].
-The theory proposes that two key factors determine the dichotomous outcome of these designs: the discriminal process of each stimulus and their discriminal difference.
+The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
 The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level.
-According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (or mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
+According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (or the mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
 
+<!-- 2. Example of discriminal processes -->
 For instance, @fig-discriminal_process illustrates the discriminal process distributions along a quality trait continuum for two written texts.
@@
 
-<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
+<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
 However, because the discriminal process of a single stimulus is not directly observable, the theory introduces the *law of comparative judgment*.
@@
 Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution.
-The mode (or mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$.
+The mode (or the mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$.
 Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$.
-Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between their discriminal processes.
-This correlation quantifies how much the judge's perception of the trait in one stimulus influences his perception of the same trait in the other.
+Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between the discriminal processes of the stimuli.
+This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in the other.
 
+<!-- 4. Example of discriminal difference -->
 @fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process, assuming a correlation of $\rho = 0.6$.
@@
 
-Additionally, @fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the same two texts.
-The figure shows that as the correlation increases, indicating a stronger influence of the judge's perception of quality in one text on his perception of the other, the distribution of the discriminal difference becomes narrower.
-This narrowing affects the area under the curve that determines the comparison outcome and, ultimately, the conclusions drawn from this outcome.
+<!-- 5. The effects of correlation on the discriminal difference -->
+Notably, the correlation between the discriminal processes, $\rho$, plays a pivotal role in determining comparison outcomes by shaping the distribution of the discriminal difference between the stimuli.
+Specifically, as the correlation increases, reflecting a stronger dependence of the judge's perception of quality in one stimulus on his perception of the other, the distribution of the discriminal difference narrows.
+This narrowing ultimately impacts the area under the curve that determines the comparison outcome and, consequently, the conclusions drawn from the comparison.
 
-For instance, when two texts clearly differ in quality, as depicted in @fig-discriminal_process, higher correlations increase the likelihood that the discriminal difference more distinctly favors text B over text A, as indicated by the shaded gray area under the curve in @fig-correlation.
-Conversely, when the two texts have similar or identical quality levels, higher correlations reduce the likelihood that the discriminal difference distinctly favors one text over the other, as the distribution narrows around zero (not depicted in the figure).
+<!-- 6. example of the effects of correlation on the discriminal difference -->
+@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the two texts depicted in @fig-discriminal_process.
+Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A.
+This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation).
+Conversely, it is not hard to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance that the discriminal difference distinctly favors one text over the other.
+This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero (not illustrated).
 

### 2024-12-02 _tex/index.tex
Built site for gh-pages

@@
 \citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
-factors determine the dichotomous outcome of these designs: the
+factors determine the dichotomous outcome of these comparisons: the
 discriminal process of each stimulus and their discriminal difference.
@@
 process for each stimulus follows a Normal distribution, where its mode
-(or mean), referred to as the \emph{modal discriminal process},
+(or the mean), referred to as the \emph{modal discriminal process},
 indicates the stimulus' position on the trait continuum, and its
@@
 distribution, their difference also follows a Normal distribution. The
-mode (or mean) of this distribution, representing the (average) relative
-separation, is given by the difference between the modal discriminal
-processes of the stimuli \(S_{BA} = S_{B} - S_{A}\). Meanwhile, the
-dispersion of the distribution, reflecting the variability in the
-relative separation, is calculated as
+mode (or the mean) of this distribution, representing the (average)
+relative separation, is given by the difference between the modal
+discriminal processes of the stimuli \(S_{BA} = S_{B} - S_{A}\).
+Meanwhile, the dispersion of the distribution, reflecting the
+variability in the relative separation, is calculated as
 \(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
@@
 discriminal dispersions, while \(\rho\) measures the correlation between
-their discriminal processes. This correlation quantifies how much the
-judge's perception of the trait in one stimulus influences his
-perception of the same trait in the other.
+the discriminal processes of the stimuli. This correlation quantifies
+the dependence of the judge's perception of the trait in one stimulus on
+his perception of the same trait in the other.
 
@@
 
-Additionally, Figure~\ref{fig-correlation} illustrates how varying
-correlations influence the distribution of the discriminal difference
-for the same two texts. The figure shows that as the correlation
-increases, indicating a stronger influence of the judge's perception of
-quality in one text on his perception of the other, the distribution of
-the discriminal difference becomes narrower. This narrowing affects the
-area under the curve that determines the comparison outcome and,
-ultimately, the conclusions drawn from this outcome.
+Notably, the correlation between the discriminal processes, \(\rho\),
+plays a pivotal role in determining comparison outcomes by shaping the
+distribution of the discriminal difference between the stimuli.
+Specifically, as the correlation increases, reflecting a stronger
+dependence of the judge's perception of quality in one stimulus on his
+perception of the other, the distribution of the discriminal difference
+narrows. This narrowing ultimately impacts the area under the curve that
+determines the comparison outcome and, consequently, the conclusions
+drawn from the comparison.
 
-For instance, when two texts clearly differ in quality, as depicted in
-Figure~\ref{fig-discriminal_process}, higher correlations increase the
-likelihood that the discriminal difference more distinctly favors text B
-over text A, as indicated by the shaded gray area under the curve in
-Figure~\ref{fig-correlation}. Conversely, when the two texts have
-similar or identical quality levels, higher correlations reduce the
-likelihood that the discriminal difference distinctly favors one text
-over the other, as the distribution narrows around zero (not depicted in
-the figure).
+Figure~\ref{fig-correlation} illustrates how varying correlations
+influence the distribution of the discriminal difference for the two
+texts depicted in Figure~\ref{fig-discriminal_process}. Since the texts
+differ in quality, higher correlations increase the likelihood that the
+discriminal difference distinctly favors text B over text A. This is
+evident from the larger proportion of the area under the curve,
+\(P(B > A)\), that lies above zero (shaded gray area in
+Figure~\ref{fig-correlation}). Conversely, it is not hard to infer that
+if the texts had similar or identical quality levels, higher
+correlations would likely reduce the chance that the discriminal
+difference distinctly favors one text over the other. This probability
+reduction occurs because the distribution of the discriminal difference
+would become more narrowly centered around zero (not illustrated).
 

### 2024-12-02 index.qmd
section:issues,CaseV; process:write

@@
 Indeed, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*.
-Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution.
+Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution [@Andrich_1978].
 The mode (or the mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$.
@@
 The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$.
-As a result, the dichotomous outcome of this comparison almost certainly favors text B over text A.
+As a result, the dichotomous outcome of this comparison would mostly favor text B over text A.
 
@@
 Specifically, as the correlation increases, reflecting a stronger dependence of the judge's perception of quality in one stimulus on his perception of the other, the distribution of the discriminal difference narrows.
-This narrowing ultimately impacts the area under the curve that determines the comparison outcome and, consequently, the conclusions drawn from the comparison.
+This narrowing ultimately impacts the area under the curve that determines the comparison outcome and, consequently, the conclusions drawn from this outcome.
 
@@
 Case I represents the general form of the theory.
-Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge.
+Case II extends this by allowing multiple judges to make comparisons rather than restricting them to a single judge.
 Case III introduces the assumption of zero correlation between stimuli.
@@
 <!-- 4. What does the zero correlation between stimuli imply? and does it hold? -->
-As outlined in the previous section, the assumption of *zero correlation between stimuli* suggests that during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @fig-correlation, $\rho=0$).
+As outlined in @sec-thurstone_theory, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text.
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
@@
 <!-- 5. The first scenario where the assumption does not hold -->
-In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between heterogeneous stimuli due to certain judges' biases that resist cancellation.
-Research on text quality indicates that when judges evaluate these traits, they often rely on various intricate aspects of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These aspects, which are likely neither equally weighted nor opposing, may influence judges' perceptions unevenly.
+In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation.
+Research on text quality suggests that when judges evaluating these traits, they often rely on various intricate characteristics of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, consequently, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
+346].
 For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments.
-Ignoring these relevant additional "traits" can result in biases that resist cancellation, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
-346] and ultimately violating the assumption of zero correlation.
-While direct evidence for the specific mechanisms causing biases is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
+While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
 
@@
 <!-- 6. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections created by additional hierarchical structures may introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
+In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
 Although the CJ literature acknowledges the presence of such hierarchical structures in CJ data, the statistical handling of this extra source of dependency has been inadequate.
@@
 
-<!-- 7. But what are the challenges when these assumptions are violated? -->
-In any case, the psychometric and statistical literature strongly advises against ignoring relevant additional traits or overlooking clustering (grouping) structures that may introduce dependencies between stimuli.
-First, ignoring additional aspects pertinent to the trait of interest can cause a dimensional mismatch, which can lead to an overestimation of the precision (reliability) of the trait's measurement [@Hoyle_et_al_2023, pp.
-340-341] or, worse, introduce bias into the measurement process [@Ackerman_1989].
-Additionally, overlooking relevant hierarchical (or grouping) structures can exacerbate the inflation of reliability by reducing the accuracy of the model's parameters [@Hoyle_et_al_2023, pp.
+<!-- 7. But what are the challenges when these assumptions are violated? and how to solve them? -->
+In any case, the psychometric and statistical literature underscores the importance of incorporating additional relevant traits and clustering (grouping) structures into the modeling of CJ data to address potential dependencies between stimuli.
+Oversights in these areas can lead to misleading conclusions about the trait's reliability and ultimately compromise its validity [@Perron_et_al_2015, pp.
+2].
+Neglecting additional traits relevant to the stimuli or the trait of interest, such as judges' biases, introduces a dimensional mismatch in the modeling process of the CJ data.
+This mismatch often results in an overestimation of the trait's reliability [@Hoyle_et_al_2023, pp.
+341] or, worse, introduces biases into the estimated trait [@Ackerman_1989].
+Therefore, these traits must be incorporated into the modeling of CJ data, as demonstrated by @Andrich_1978 and @Wainer_et_al_1978.
+Similarly, ignoring hierarchical (grouping) structures reduces the accuracy of model parameter estimates and further exacerbates the overestimation of reliability [@Hoyle_et_al_2023, pp.
 482].
-Taken together, these oversights undermine the reliability of the measurement process and, since validity cannot exist without reliability [@Perron_et_al_2015, pp.
-2], ultimately compromise the validity of the results derived from it.
+To address this, any additional hierarchical structure relevant to the stimuli should also be included in the model, as in Multilevel Structural Equation models (MSEM) [@Hoyle_et_al_2023, chap.
+26].
 
-<!-- forcing a unidimensional trait structure on multidimensional data can inflate the perceived measurement precision (reliability) of the trait [@Hoyle_et_al_2023, pp. 340-341] -->
+<!-- This further implies that the shape of the discriminal difference distribution that determines the comparison outcome and, consequently, the conclusions drawn from this outcome (see @sec-thurstone_theory, @fig-correlation).
+-->
+
+<!-- To show the practical consequences, in Figure 18.1, we present the test information for the two models; clearly, the information is positively biased when the multidimensional data are forced into a unidimensional IRT model.
+Moreover, in this same figure, we show the test response curves (how expected scores change as a function of trait level) for the two models; again, they, clearly, are not overlapping.
+The estimated implication is that if one applied a unidimensional IRT model, psychometric information is inflated due to the multidimensionality and examinee standard errors, which are the inverse square root of information, are too small.
+[@Hoyle_et_al_2023, pp.
+340-341] -->
+
+<!-- Accounting for measurement error in defining latent variables improves the accuracy of the model’s estimated structural parameters between the latent constructs, which in turn can provide a stronger test of a proposed theoretical model.
+[@Hoyle_et_al_2023, pp.
+482] -->
 
@@
 <!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem. -->
-
-<!-- 8. So, how do we mitigate these risks? -->
-<!-- Luckily, @Andrich_1978 and @Wainer_et_al_1978 already proposed a solution to the first scenario where the assumption of zero correlation is violated.
-Both authors demonstrated that by parametrizing judges' biases in the measurement model of the trait of interest, their effects are experimentally eliminated from the trait's measurement process.
-Furthermore, a wealth of statistical literature offers a solution to the second scenario: incorporate the hierarchical structure into the measurement model of the trait of interest, similar to a multilevel SEM model [@Hoyle_et_al_2023, chap.
-26].
--->
 
@@
 
+<!-- The correlation $\rho$, considered above is not the kind of correlation considered by Bock (1958) in studies of preference and choice of objects.
+The Bock correlation arises from a specific interaction between a subject and an item because the subject has a particular "preference", such as a taste preference, for the object.
+This preference for the object is then reflected in its comparisons with all other objects.
+However, it is less likely that subjects would have such particular preferences for some items in a test and, in any case, the assumption is open to empirical checking [@Andrich_1978, pp.
+455].
+-->
+
 ## From theory to statistics {#sec-theory-statistics}

### 2024-12-02 code/figures.R

@@ -18,7 +18,7 @@ colorArea <- function(from, to, density, ..., col="lightgray", dens=NULL){
 
 # distributional parameters
 dpA = c(0, 0.5) # distribution text A (mean, sd)
-dpB = c(2, 1) # distribution text B (mean, sd)
+dpB = c(1, 1) # distribution text B (mean, sd)
 rAB = 0.6 # correlation between discriminal processes
 
 # figure
@@ -27,7 +27,7 @@ png( filename=file.path(main_dir, 'discriminal_process.png'),
 
 # initial figure
 par( mar=c(2,0,0.5,0) )
-plot( NULL, xlim=c(-2,6), ylim=c(0,0.9), 
+plot( NULL, xlim=c(-2,5), ylim=c(0,0.9), 
       xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=c(dpA[1], dpB[1]), tick=F, 
@@ -75,7 +75,7 @@ png( filename=file.path(main_dir, 'discriminal_difference.png'),
 
 # initial figure
 par( mar=c(2,0,0.5,0) )
-plot( NULL, xlim=c(-2,6), ylim=c(0,0.5), 
+plot( NULL, xlim=c(-2,5), ylim=c(0,0.5), 
       xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=c(0, dd[1]), tick=F, 
@@ -101,8 +101,8 @@ text( x=dd[1]+2, y=pplotD[4], cex=0.8,
 text( x=dd[1]+2.45, y=pplotD[4]-0.04, cex=0.8,
       expression( sigma[BA] == sqrt( sigma[B]^2 + sigma[A]^2 - rho*sigma[B]*sigma[A]) ) )
 lines( x=c(0,0), y=c(pplotD[5],-0.1), lty=2, lwd=0.7 )
-text( x=-0.7, y=0.05, cex=0.8, expression( P(B<A) ) ) 
-text( x=0.8, y=0.05, cex=0.8, expression( P(B>A) ) ) 
+# text( x=-0.5, y=0.05, cex=0.8, expression( P(B<A) ) ) 
+text( x=0.5, y=0.05, cex=0.8, expression( P(B>A) ) ) 
 
[…]

### 2024-12-02 _tex/index.tex
Built site for gh-pages

@@
 difference}. Since the individual discriminal processes follow a Normal
-distribution, their difference also follows a Normal distribution. The
-mode (or the mean) of this distribution, representing the (average)
-relative separation, is given by the difference between the modal
-discriminal processes of the stimuli \(S_{BA} = S_{B} - S_{A}\).
-Meanwhile, the dispersion of the distribution, reflecting the
-variability in the relative separation, is calculated as
+distribution, their difference also follows a Normal distribution
+\citep{Andrich_1978}. The mode (or the mean) of this distribution,
+representing the (average) relative separation, is given by the
+difference between the modal discriminal processes of the stimuli
+\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
+distribution, reflecting the variability in the relative separation, is
+calculated as
 \(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
@@
 A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
-a result, the dichotomous outcome of this comparison almost certainly
-favors text B over text A.
+a result, the dichotomous outcome of this comparison would mostly favor
+text B over text A.
 
@@
 determines the comparison outcome and, consequently, the conclusions
-drawn from the comparison.
+drawn from this outcome.
 
@@
 by allowing multiple judges to make comparisons rather than restricting
-the comparisons to a single judge. Case III introduces the assumption of
-zero correlation between stimuli. Case IV builds on this by assuming
-that the stimuli have similar dispersions. Finally, Case V replaces this
+them to a single judge. Case III introduces the assumption of zero
+correlation between stimuli. Case IV builds on this by assuming that the
+stimuli have similar dispersions. Finally, Case V replaces this
 assumption with the condition that the stimuli have equal discriminal
@@
 
-As outlined in the previous section, the assumption of \emph{zero
-correlation between stimuli} suggests that during a pairwise comparison,
-a judge's perception of quality in one text does not influence his
-perception of the same trait in another text (see
-Figure~\ref{fig-correlation}, \(\rho=0\)). Thurstone attributed this
-independence to the cancellation of potential judges' biases, driven by
-two opposing and equally weighted effects occurring during the pairwise
-comparisons \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978}
-mathematically demonstrated this cancellation using the BTL model under
-the assumption of discriminal processes with additive biases. However,
-it is easy to imagine at least two scenarios where the zero correlation
-assumption almost certainly does not hold: when the pairwise comparison
-involves multidimensional, complex traits with heterogeneous stimuli and
-when an additional hierarchical structure is relevant to the stimuli.
+As outlined in Section~\ref{sec-thurstone_theory}, the assumption of
+\emph{zero correlation between stimuli} implies that, during a pairwise
+comparison, a judge's perception of quality in one text does not
+influence his perception of the same trait in another text. Thurstone
+attributed this independence to the cancellation of potential judges'
+biases, driven by two opposing and equally weighted effects occurring
+during the pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
+\citet{Andrich_1978} mathematically demonstrated this cancellation using
+the BTL model under the assumption of discriminal processes with
+additive biases. However, it is easy to imagine at least two scenarios
+where the zero correlation assumption almost certainly does not hold:
+when the pairwise comparison involves multidimensional, complex traits
+with heterogeneous stimuli and when an additional hierarchical structure
+is relevant to the stimuli.
 
 In the first scenario, the intricate aspects of multidimensional,
-complex traits may introduce dependencies between heterogeneous stimuli
-due to certain judges' biases that resist cancellation. Research on text
-quality indicates that when judges evaluate these traits, they often
-rely on various intricate aspects of the stimuli to form their judgments
+complex traits may introduce dependencies between the stimuli due to
+certain judges' biases that resist cancellation. Research on text
+quality suggests that when judges evaluating these traits, they often
+rely on various intricate characteristics of the stimuli
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
-aspects, which are likely neither equally weighted nor opposing, may
-influence judges' perceptions unevenly. For example, this could occur
-when a judge assessing the argumentative quality of a text places
+additional relevant characteristics, which are unlikely to be equally
+weighted or opposing, can unevenly influence judges' perceptions,
+creating biases in their judgments and, consequently, introducing
+dependencies between stimuli
+\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
+occur when a judge assessing the argumentative quality of a text places
 disproportionate emphasis on grammatical accuracy, ultimately favoring
-texts with fewer errors but weaker arguments. Ignoring these relevant
-additional ``traits'' can result in biases that resist cancellation,
-introducing dependencies between stimuli
-\citep[pp.~346]{vanderLinden_et_al_2017_II} and ultimately violating the
-assumption of zero correlation. While direct evidence for the specific
-mechanisms causing biases is lacking, studies such as
+texts with fewer errors but weaker arguments. While direct evidence for
+this specific scenario is lacking, studies such as
 \citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
@@
 In the second scenario, the shared context or inherent connections
-created by additional hierarchical structures may introduce dependencies
-between stimuli, a statistical phenomenon commonly known as clustering
-\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
-presence of such hierarchical structures in CJ data, the statistical
-handling of this extra source of dependency has been inadequate. For
-example, when CJ data includes multiple samples of stimuli from the same
-individuals, researchers often rely on (average) estimated BTL scores to
-conduct subsequent analyses and tests at the individual hierarchical
-level
+created by additional hierarchical structures may further introduce
+dependencies between stimuli, a statistical phenomenon commonly known as
+clustering \citep{Everitt_et_al_2010}. Although the CJ literature
+acknowledges the presence of such hierarchical structures in CJ data,
+the statistical handling of this extra source of dependency has been
+inadequate. For example, when CJ data includes multiple samples of
+stimuli from the same individuals, researchers often rely on (average)
+estimated BTL scores to conduct subsequent analyses and tests at the
+individual hierarchical level
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 
-In any case, the psychometric and statistical literature strongly
-advises against ignoring relevant additional traits or overlooking
-clustering (grouping) structures that may introduce dependencies between
-stimuli. First, ignoring additional aspects pertinent to the trait of
-interest can cause a dimensional mismatch, which can lead to an
-overestimation of the precision (reliability) of the trait's measurement
-\citep[pp.~340-341]{Hoyle_et_al_2023} or, worse, introduce bias into the
-measurement process \citep{Ackerman_1989}. Additionally, overlooking
-relevant hierarchical (or grouping) structures can exacerbate the
-inflation of reliability by reducing the accuracy of the model's
-parameters \citep[pp.~482]{Hoyle_et_al_2023}. Taken together, these
-oversights undermine the reliability of the measurement process and,
-since validity cannot exist without reliability
-\citep[pp.~2]{Perron_et_al_2015}, ultimately compromise the validity of
-the results derived from it.
+In any case, the psychometric and statistical literature underscores the
+importance of incorporating additional relevant traits and clustering
+(grouping) structures into the modeling of CJ data to address potential
+dependencies between stimuli. Oversights in these areas can lead to
+misleading conclusions about the trait's reliability and ultimately
+compromise its validity \citep[pp.~2]{Perron_et_al_2015}. Neglecting
+additional traits relevant to the stimuli or the trait of interest, such
+as judges' biases, introduces a dimensional mismatch in the modeling
+process of the CJ data. This mismatch often results in an overestimation
+of the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
+introduces biases into the estimated trait \citep{Ackerman_1989}.
+Therefore, these traits must be incorporated into the modeling of CJ
+data, as demonstrated by \citet{Andrich_1978} and
+\citet{Wainer_et_al_1978}. Similarly, ignoring hierarchical (grouping)
+structures reduces the accuracy of model parameter estimates and further
+exacerbates the overestimation of reliability
+\citep[pp.~482]{Hoyle_et_al_2023}. To address this, any additional
+hierarchical structure relevant to the stimuli should also be included
+in the model, as in Multilevel Structural Equation models (MSEM)
+\citep[chap.~26]{Hoyle_et_al_2023}.
 

### 2024-12-03 index.qmd
section:issues,CaseV; process:write

@@
 The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level.
-According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (or the mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
+According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
 
@@
 Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution [@Andrich_1978].
-The mode (or the mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$.
+The mode (mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$.
 Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$.
 Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between the discriminal processes of the stimuli.
-This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in the other.
+This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in another.
 
@@
 Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A.
-This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation).
+This is evident from the larger proportion of area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation).
 Conversely, it is not hard to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance that the discriminal difference distinctly favors one text over the other.
@@
 Case I represents the general form of the theory.
-Case II extends this by allowing multiple judges to make comparisons rather than restricting them to a single judge.
+Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge.
 Case III introduces the assumption of zero correlation between stimuli.
@@
 In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation.
-Research on text quality suggests that when judges evaluating these traits, they often rely on various intricate characteristics of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
 These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, consequently, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
@@
 
-<!-- 7. But what are the challenges when these assumptions are violated? and how to solve them? -->
-In any case, the psychometric and statistical literature underscores the importance of incorporating additional relevant traits and clustering (grouping) structures into the modeling of CJ data to address potential dependencies between stimuli.
-Oversights in these areas can lead to misleading conclusions about the trait's reliability and ultimately compromise its validity [@Perron_et_al_2015, pp.
-2].
-Neglecting additional traits relevant to the stimuli or the trait of interest, such as judges' biases, introduces a dimensional mismatch in the modeling process of the CJ data.
-This mismatch often results in an overestimation of the trait's reliability [@Hoyle_et_al_2023, pp.
-341] or, worse, introduces biases into the estimated trait [@Ackerman_1989].
-Therefore, these traits must be incorporated into the modeling of CJ data, as demonstrated by @Andrich_1978 and @Wainer_et_al_1978.
-Similarly, ignoring hierarchical (grouping) structures reduces the accuracy of model parameter estimates and further exacerbates the overestimation of reliability [@Hoyle_et_al_2023, pp.
+<!-- 7. But what are the challenges when this assumption is violated? -->
+In any case, the psychometric and statistical literature emphasizes the importance of addressing factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity.
+For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the model.
+This mismatch can potentially inflate the reliability of the trait [@Hoyle_et_al_2023, pp.
+341] or, worse, introduce bias into the trait's estimates [@Ackerman_1989].
+Similarly, failing to account for hierarchical (grouping) structures reduces the precision of model parameter estimates, further amplifying the overestimation of reliability [@Hoyle_et_al_2023, pp.
 482].
-To address this, any additional hierarchical structure relevant to the stimuli should also be included in the model, as in Multilevel Structural Equation models (MSEM) [@Hoyle_et_al_2023, chap.
-26].
+These issues collectively undermine the trait's reliability, and because "validity cannot exist without reliability" [@Perron_et_al_2015, pp.
+2], they ultimately compromise the validity of results derived from the the data modeling.
 
@@
 
-<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem. -->
+<!-- 8. How to solve the violation of the assumption? -->
+Fortunately, the same literature provides solutions for addressing the factors that create dependencies between stimuli.
+@Andrich_1978 and @Wainer_et_al_1978 propose integrating judges' biases into the CJ model.
+This integration would produce a model similar to a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap.
+15], depending on the theoretical assumptions a researcher is willing to make regarding the judges' biases.
+The literature also suggests incorporating relevant hierarchical structures into the model to account for dependencies in the stimuli, resulting in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
+26].
+
+<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem for latent variables?. -->
 

### 2024-12-03 _tex/index.tex
Built site for gh-pages

@@
 process for each stimulus follows a Normal distribution, where its mode
-(or the mean), referred to as the \emph{modal discriminal process},
-indicates the stimulus' position on the trait continuum, and its
-dispersion, known as the \emph{discriminal dispersion}, reflects the
-variability in the perceived trait level of the stimulus.
+(mean), referred to as the \emph{modal discriminal process}, indicates
+the stimulus' position on the trait continuum, and its dispersion, known
+as the \emph{discriminal dispersion}, reflects the variability in the
+perceived trait level of the stimulus.
 
@@
 distribution, their difference also follows a Normal distribution
-\citep{Andrich_1978}. The mode (or the mean) of this distribution,
-representing the (average) relative separation, is given by the
-difference between the modal discriminal processes of the stimuli
+\citep{Andrich_1978}. The mode (mean) of this distribution, representing
+the (average) relative separation, is given by the difference between
+the modal discriminal processes of the stimuli
 \(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
@@
 the dependence of the judge's perception of the trait in one stimulus on
-his perception of the same trait in the other.
+his perception of the same trait in another.
 
@@
 discriminal difference distinctly favors text B over text A. This is
-evident from the larger proportion of the area under the curve,
+evident from the larger proportion of area under the curve,
 \(P(B > A)\), that lies above zero (shaded gray area in
@@
 by allowing multiple judges to make comparisons rather than restricting
-them to a single judge. Case III introduces the assumption of zero
-correlation between stimuli. Case IV builds on this by assuming that the
-stimuli have similar dispersions. Finally, Case V replaces this
+the comparisons to a single judge. Case III introduces the assumption of
+zero correlation between stimuli. Case IV builds on this by assuming
+that the stimuli have similar dispersions. Finally, Case V replaces this
 assumption with the condition that the stimuli have equal discriminal
@@
 certain judges' biases that resist cancellation. Research on text
-quality suggests that when judges evaluating these traits, they often
-rely on various intricate characteristics of the stimuli
+quality suggests that when judges evaluate these traits, they often rely
+on various intricate characteristics of the stimuli to form their
+judgments
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
@@
 
-In any case, the psychometric and statistical literature underscores the
-importance of incorporating additional relevant traits and clustering
-(grouping) structures into the modeling of CJ data to address potential
-dependencies between stimuli. Oversights in these areas can lead to
-misleading conclusions about the trait's reliability and ultimately
-compromise its validity \citep[pp.~2]{Perron_et_al_2015}. Neglecting
-additional traits relevant to the stimuli or the trait of interest, such
-as judges' biases, introduces a dimensional mismatch in the modeling
-process of the CJ data. This mismatch often results in an overestimation
-of the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
-introduces biases into the estimated trait \citep{Ackerman_1989}.
-Therefore, these traits must be incorporated into the modeling of CJ
-data, as demonstrated by \citet{Andrich_1978} and
-\citet{Wainer_et_al_1978}. Similarly, ignoring hierarchical (grouping)
-structures reduces the accuracy of model parameter estimates and further
-exacerbates the overestimation of reliability
-\citep[pp.~482]{Hoyle_et_al_2023}. To address this, any additional
-hierarchical structure relevant to the stimuli should also be included
-in the model, as in Multilevel Structural Equation models (MSEM)
-\citep[chap.~26]{Hoyle_et_al_2023}.
+In any case, the psychometric and statistical literature emphasizes the
+importance of addressing factors that create dependencies between
+stimuli, as failing to do so can lead to inaccurate conclusions about a
+trait's reliability and, by extension, its validity. For instance,
+neglecting additional traits relevant to the stimuli, such as judges'
+biases, often leads to a dimensional mismatch in the model. This
+mismatch can potentially inflate the reliability of the trait
+\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias into the
+trait's estimates \citep{Ackerman_1989}. Similarly, failing to account
+for hierarchical (grouping) structures reduces the precision of model
+parameter estimates, further amplifying the overestimation of
+reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
+undermine the trait's reliability, and because ``validity cannot exist
+without reliability'' \citep[pp.~2]{Perron_et_al_2015}, they ultimately
+compromise the validity of results derived from the the data modeling.
+
+Fortunately, the same literature provides solutions for addressing the
+factors that create dependencies between stimuli. \citet{Andrich_1978}
+and \citet{Wainer_et_al_1978} propose integrating judges' biases into
+the CJ model. This integration would produce a model similar to a
+multidimensional or two-parameter logistic IRT model
+\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
+assumptions a researcher is willing to make regarding the judges'
+biases. The literature also suggests incorporating relevant hierarchical
+structures into the model to account for dependencies in the stimuli,
+resulting in a model resembling a Multilevel Structural Equation Model
+(MSEM) \citep[chap.~26]{Hoyle_et_al_2023}.
 

### 2024-12-03 index.qmd
section:issues,CaseV; process:write

@@
 374].
-Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
+Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
 
@@
 Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, consequently, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
+These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
 346].
@@
 In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
-Although the CJ literature acknowledges the presence of such hierarchical structures in CJ data, the statistical handling of this extra source of dependency has been inadequate.
+Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between the stimuli has been inadequate.
 For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 These issues collectively undermine the trait's reliability, and because "validity cannot exist without reliability" [@Perron_et_al_2015, pp.
-2], they ultimately compromise the validity of results derived from the the data modeling.
+2], they ultimately compromise the validity of results derived from the data modeling.
 
@@
 <!-- 8. How to solve the violation of the assumption? -->
-Fortunately, the same literature provides solutions for addressing the factors that create dependencies between stimuli.
-@Andrich_1978 and @Wainer_et_al_1978 propose integrating judges' biases into the CJ model.
-This integration would produce a model similar to a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap.
-15], depending on the theoretical assumptions a researcher is willing to make regarding the judges' biases.
-The literature also suggests incorporating relevant hierarchical structures into the model to account for dependencies in the stimuli, resulting in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
-26].
+Fortunately, the same literature offers solutions for addressing the factors that create these dependencies between stimuli.
+@Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
+Furthermore, the literature advocates for incorporating relevant hierarchical structures into the data model to account for these dependencies.
+Taken together, these additions can result in a model that combines a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
+26] with a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap.
+15], depending on the theoretical assumptions a researcher makes about the judges' biases.
 

### 2024-12-03 _tex/index.tex
Built site for gh-pages

@@
 applications frequently involve these traits and stimuli, two main
-assumptions of Case V may not consistently hold in theory or practice:
-the zero correlation and equal dispersion between stimuli.
+assumptions of Case V and, by extension, of the BTL model may not
+consistently hold in theory or practice: the zero correlation and equal
+dispersion between stimuli.
 
@@
 weighted or opposing, can unevenly influence judges' perceptions,
-creating biases in their judgments and, consequently, introducing
+creating biases in their judgments and, ultimately, introducing
 dependencies between stimuli
@@
 clustering \citep{Everitt_et_al_2010}. Although the CJ literature
-acknowledges the presence of such hierarchical structures in CJ data,
-the statistical handling of this extra source of dependency has been
-inadequate. For example, when CJ data includes multiple samples of
-stimuli from the same individuals, researchers often rely on (average)
-estimated BTL scores to conduct subsequent analyses and tests at the
-individual hierarchical level
+acknowledges the presence of such hierarchical structures, the
+statistical handling of this extra source of dependency between the
+stimuli has been inadequate. For example, when CJ data includes multiple
+samples of stimuli from the same individuals, researchers often rely on
+(average) estimated BTL scores to conduct subsequent analyses and tests
+at the individual hierarchical level
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 without reliability'' \citep[pp.~2]{Perron_et_al_2015}, they ultimately
-compromise the validity of results derived from the the data modeling.
+compromise the validity of results derived from the data modeling.
 
-Fortunately, the same literature provides solutions for addressing the
-factors that create dependencies between stimuli. \citet{Andrich_1978}
-and \citet{Wainer_et_al_1978} propose integrating judges' biases into
-the CJ model. This integration would produce a model similar to a
-multidimensional or two-parameter logistic IRT model
-\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
-assumptions a researcher is willing to make regarding the judges'
-biases. The literature also suggests incorporating relevant hierarchical
-structures into the model to account for dependencies in the stimuli,
-resulting in a model resembling a Multilevel Structural Equation Model
-(MSEM) \citep[chap.~26]{Hoyle_et_al_2023}.
+Fortunately, the same literature offers solutions for addressing the
+factors that create these dependencies between stimuli.
+\citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend integrating
+judges' biases into the BTL model. Furthermore, the literature advocates
+for incorporating relevant hierarchical structures into the data model
+to account for these dependencies. Taken together, these additions can
+result in a model that combines a Multilevel Structural Equation Model
+(MSEM) \citep[chap.~26]{Hoyle_et_al_2023} with a multidimensional or
+two-parameter logistic IRT model \citep[chap.~15]{Hoyle_et_al_2023},
+depending on the theoretical assumptions a researcher makes about the
+judges' biases.
 

### 2024-12-03 index.qmd
section:issues,CaseV; process:write

@@
 Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions could prove problematic in the presence of complex traits or heterogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp.
-374].
-Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.
+270], acknowledging that some assumptions could prove problematic in the presence of complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+376].
+Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the equal dispersion and zero correlation between stimuli.
 
@@
 
-<!-- 4. What does the zero correlation between stimuli imply? and does it hold? -->
-As outlined in @sec-thurstone_theory, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text.
+<!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
+On the one hand, the assumption of *equal dispersion between stimuli* asserts that the variability in perceived trait levels remains consistent across all stimuli.
+However, Thurstone noted that this assumption may fail when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+376], as these can introduce greater discrepancies into judgments due to their unique characteristics.
+One example of this is the evaluation of creativity.
+In this context, for instance, it is plausible that a student completes an assignment in a highly creative yet unexpected way.
+Judges, unprepared for such originality, may experience greater discrepancies in their judgments of the assignment.
+As a result, the discriminal difference associated with this assignment is expected to show a broader dispersion, reflecting the larger variation in its discriminal process compared to others.
+Although direct evidence for this specific scenario is lacking, [studies such as @vanDaal_2020 [chap.
+8] suggest that misfitting texts tend to exhibit larger dispersion, likely due to increased confusion among judges regarding the texts.]{style="color:red;"}
+
+<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
+[@Thurstone_1927a, pp.
+376] -->
+
+<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
+Thurstone clearly did not expect this to hold for any but the most simple stimuli.
+The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
+-->
+
+<!-- "weak" writers consistently produce texts of low quality in all aspects that determine the quality of a text, such as grammatical accuracy, organization, and language use.
+In such cases, judges typically agree on their inferior quality, leading to a clear disfavoring of these texts compared to others.
+This agreement ultimately imply narrow dispersions for the discriminal differences related to these texts and, by extension, narrow discriminal processes for them.
+However, as texts improve along the trait continuum, with certain aspects showing improvement and others reaming unaltered, these variations introduce uncertainty into the judgments.
+Now, since judges often rely on various intricate characteristics of the texts to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], these changes lead to greater discrepancies in their evaluations.
+This divergence ultimately imply broader discriminal differences and, by extension, more varied discriminal processes for the texts.
+-->
+
+<!-- So it seem that the extreme of the discriminal process distribution of the stimuli have higher reliability than the center? -->
+
+<!-- Misfiting assignments might be assignments with larger dispersion, because there is a lot of confusion about them.
+When assessing creativity, for instance, a student might complete an assignment in a highly creative but unexpected way.
+Judges, unprepared for such originality, may experience uncertainty, leading to greater variability in their judgments about the assignment.
+Then the discriminal difference is all over the place.
+-->
+
+<!-- 5. But what are the challenges when these assumptions are violated? -->
+[what are the challenges when the equal dispersion assumption is violated?]{style="color:red;"}
+
+<!-- But the statistical literature warns against mis-specifying the trait’s measurement model.
+When mis-specifying the measurement model by assuming, for example, equal discriminal dispersions across stimuli, the model can overlook meaningful differences in the reliability of the trait.
+Taken together, these mis-specifications and oversights can lead to misleading conclusions about the trait measurement, ultimately casting doubts about the validity of the trait, as validity cannot exist without reliability [@Perron_et_al_2015, pp.
+2].
+-->
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->
+
+<!-- If your scores are not reliable they cannot be valid.
+People have been employing a model with assumptions that they do not know if those assumptions apply for their cases.
+And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid.
+-->
+
+<!-- 6. So, how do we mitigate these risks? -->
+[how do we mitigate these risks?]{style="color:red;"}
+
+<!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
+
+<!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->
+
+<!-- 7. What does the zero correlation between stimuli imply? and does it hold? -->
+On the other hand hand, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @sec-thurstone_theory).
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
@@
 
-<!-- 5. The first scenario where the assumption does not hold -->
+<!-- 8. The first scenario where the assumption does not hold -->
 In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation.
@@
 
-<!-- 6. The second scenario where the assumption does not hold -->
+<!-- 9. The second scenario where the assumption does not hold -->
 In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
-Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between the stimuli has been inadequate.
+Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between stimuli has been inadequate.
 For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 
-<!-- 7. But what are the challenges when this assumption is violated? -->
-In any case, the psychometric and statistical literature emphasizes the importance of addressing factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity.
-For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the model.
+<!-- 10. But what are the challenges when this assumption is violated? -->
+In any case, the psychometric and statistical literature emphasizes the importance of addressing the factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity.
+For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the statistical model used for analysis.
 This mismatch can potentially inflate the reliability of the trait [@Hoyle_et_al_2023, pp.
@@
 482].
-These issues collectively undermine the trait's reliability, and because "validity cannot exist without reliability" [@Perron_et_al_2015, pp.
-2], they ultimately compromise the validity of results derived from the data modeling.
+These issues collectively undermine the trait's reliability and ultimately compromise the validity of the trait's estimates [@Perron_et_al_2015, pp.
+2].
 
@@
 
-<!-- 8. How to solve the violation of the assumption? -->
-Fortunately, the same literature offers solutions for addressing the factors that create these dependencies between stimuli.
+<!-- 11. How to solve the violation of the assumption? -->
+Fortunately, the same literature offers solutions for addressing these issues.
 @Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
-Furthermore, the literature advocates for incorporating relevant hierarchical structures into the data model to account for these dependencies.
-Taken together, these additions can result in a model that combines a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
-26] with a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap.
-15], depending on the theoretical assumptions a researcher makes about the judges' biases.
+Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
+Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
+26] combined with a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap.
+15], depending on the theoretical and statistical treatment of judges' biases.
 
@@
 <!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
-
-<!-- 9. What does the equal dispersion between stimuli imply? and does it hold? -->
-<!-- In contrast, the assumption of *equal dispersion between stimuli* suggests that the variability in the perceived trait level of the stimulus is the same across all stimuli.
-While Thurstone acknowledged that this assumption may be violated when "dealing with less conspicuous attributes or with less homogeneous stimuli" [@Thurstone_1927a, pp.
-374], ...
--->
-
-<!-- It is a ridiculous assumption, because multidimensional, complex traits -->
-<!-- The dispersion is not the same for every text, a weak writer will produce consistently weak quality texts, in all the dimensions, this implies that every judge will agree on its lower quality, thereby the dispersion of the texts is going to be narrow.
-In contrast, when the texts move further along the quality trait continuum, implying that some areas of the texts are improved while some other remains, then there is a higher potential that there are discrepancies between the judges, as they focus in different aspects of the stimuli.
-So it seem that the extreme of the discriminal process distribution of the stimuli have higher reliability than the center?
--->
-
-<!-- Some text will provoke more variability in the judgments, and that is because of the text. There is something in the text that cause doubts in the judges. -->
-
-<!-- Example of creativity from Tine.
-Misfiting text might be texts with larger dispersion, because there is a lot of confusion about them.
-When assessing creativity, for instance, student can do an assignment in really creative but an unexpected way, but because the judges did not expected such a way, they do not what to do with it, and that creates more dispersion in the judgment.
-Then the discriminal difference is all over the place.
--->
-
-<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
-Thurstone clearly did not expect this to hold for any but the most simple stimuli.
-The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
--->
-
-<!-- 10. But what are the challenges when these assumptions are violated? -->
-<!-- But the statistical literature warns against mis-specifying the trait’s measurement model.
-When mis-specifying the measurement model by assuming, for example, equal discriminal dispersions across stimuli, the model can overlook meaningful differences in the reliability of the trait.
-Taken together, these mis-specifications and oversights can lead to misleading conclusions about the trait measurement, ultimately casting doubts about the validity of the trait, as validity cannot exist without reliability [@Perron_et_al_2015, pp.
-2].
--->
-
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->
-
-<!-- If your scores are not reliable they cannot be valid.
-People have been employing a model with assumptions that they do not know if those assumptions apply for their cases.
-And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid.
--->
-
-<!-- 11. So, how do we mitigate these risks? -->
-<!-- ["to mitigate" paragraph]{style="color:red;"} -->
-
-<!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->
 

### 2024-12-03 _tex/index.tex
Built site for gh-pages

@@
 could prove problematic in the presence of complex traits or
-heterogeneous stimuli, such as handwriting or English compositions
-\citep[pp.~374]{Thurstone_1927a}. Consequently, given that modern CJ
-applications frequently involve these traits and stimuli, two main
-assumptions of Case V and, by extension, of the BTL model may not
-consistently hold in theory or practice: the zero correlation and equal
-dispersion between stimuli.
+heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}. Consequently,
+given that modern CJ applications frequently involve these traits and
+stimuli, two main assumptions of Case V and, by extension, of the BTL
+model may not consistently hold in theory or practice: the equal
+dispersion and zero correlation between stimuli.
 
-As outlined in Section~\ref{sec-thurstone_theory}, the assumption of
-\emph{zero correlation between stimuli} implies that, during a pairwise
-comparison, a judge's perception of quality in one text does not
-influence his perception of the same trait in another text. Thurstone
-attributed this independence to the cancellation of potential judges'
-biases, driven by two opposing and equally weighted effects occurring
-during the pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
-\citet{Andrich_1978} mathematically demonstrated this cancellation using
-the BTL model under the assumption of discriminal processes with
-additive biases. However, it is easy to imagine at least two scenarios
-where the zero correlation assumption almost certainly does not hold:
-when the pairwise comparison involves multidimensional, complex traits
-with heterogeneous stimuli and when an additional hierarchical structure
-is relevant to the stimuli.
+On the one hand, the assumption of \emph{equal dispersion between
+stimuli} asserts that the variability in perceived trait levels remains
+consistent across all stimuli. However, Thurstone noted that this
+assumption may fail when researchers assess complex traits or
+heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these can
+introduce greater discrepancies into judgments due to their unique
+characteristics. One example of this is the evaluation of creativity. In
+this context, for instance, it is plausible that a student completes an
+assignment in a highly creative yet unexpected way. Judges, unprepared
+for such originality, may experience greater discrepancies in their
+judgments of the assignment. As a result, the discriminal difference
+associated with this assignment is expected to show a broader
+dispersion, reflecting the larger variation in its discriminal process
+compared to others. Although direct evidence for this specific scenario
+is lacking, {studies such as \citet[chap.~8]{vanDaal_2020} suggest that
+misfitting texts tend to exhibit larger dispersion, likely due to
+increased confusion among judges regarding the texts.}
+
+{what are the challenges when the equal dispersion assumption is
+violated?}
+
+{how do we mitigate these risks?}
+
+On the other hand hand, the assumption of \emph{zero correlation between
+stimuli} implies that, during a pairwise comparison, a judge's
+perception of quality in one text does not influence his perception of
+the same trait in another text (see Section~\ref{sec-thurstone_theory}).
+Thurstone attributed this independence to the cancellation of potential
+judges' biases, driven by two opposing and equally weighted effects
+occurring during the pairwise comparisons
+\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} mathematically
+demonstrated this cancellation using the BTL model under the assumption
+of discriminal processes with additive biases. However, it is easy to
+imagine at least two scenarios where the zero correlation assumption
+almost certainly does not hold: when the pairwise comparison involves
+multidimensional, complex traits with heterogeneous stimuli and when an
+additional hierarchical structure is relevant to the stimuli.
 
@@
 acknowledges the presence of such hierarchical structures, the
-statistical handling of this extra source of dependency between the
-stimuli has been inadequate. For example, when CJ data includes multiple
-samples of stimuli from the same individuals, researchers often rely on
+statistical handling of this extra source of dependency between stimuli
+has been inadequate. For example, when CJ data includes multiple samples
+of stimuli from the same individuals, researchers often rely on
 (average) estimated BTL scores to conduct subsequent analyses and tests
@@
 In any case, the psychometric and statistical literature emphasizes the
-importance of addressing factors that create dependencies between
+importance of addressing the factors that create dependencies between
 stimuli, as failing to do so can lead to inaccurate conclusions about a
@@
 neglecting additional traits relevant to the stimuli, such as judges'
-biases, often leads to a dimensional mismatch in the model. This
-mismatch can potentially inflate the reliability of the trait
-\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias into the
-trait's estimates \citep{Ackerman_1989}. Similarly, failing to account
-for hierarchical (grouping) structures reduces the precision of model
-parameter estimates, further amplifying the overestimation of
+biases, often leads to a dimensional mismatch in the statistical model
+used for analysis. This mismatch can potentially inflate the reliability
+of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias
+into the trait's estimates \citep{Ackerman_1989}. Similarly, failing to
+account for hierarchical (grouping) structures reduces the precision of
+model parameter estimates, further amplifying the overestimation of
 reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
-undermine the trait's reliability, and because ``validity cannot exist
-without reliability'' \citep[pp.~2]{Perron_et_al_2015}, they ultimately
-compromise the validity of results derived from the data modeling.
+undermine the trait's reliability and ultimately compromise the validity
+of the trait's estimates \citep[pp.~2]{Perron_et_al_2015}.
 
-Fortunately, the same literature offers solutions for addressing the
-factors that create these dependencies between stimuli.
-\citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend integrating
-judges' biases into the BTL model. Furthermore, the literature advocates
-for incorporating relevant hierarchical structures into the data model
-to account for these dependencies. Taken together, these additions can
-result in a model that combines a Multilevel Structural Equation Model
-(MSEM) \citep[chap.~26]{Hoyle_et_al_2023} with a multidimensional or
-two-parameter logistic IRT model \citep[chap.~15]{Hoyle_et_al_2023},
-depending on the theoretical assumptions a researcher makes about the
-judges' biases.
+Fortunately, the same literature offers solutions for addressing these
+issues. \citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend
+integrating judges' biases into the BTL model. Moreover, the literature
+advocates for the incorporation of relevant hierarchical structures into
+the statistical model to account for these dependencies. Together, these
+additions can result in a model resembling a Multilevel Structural
+Equation Model (MSEM) \citep[chap.~26]{Hoyle_et_al_2023} combined with a
+multidimensional or two-parameter logistic IRT model
+\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical and
+statistical treatment of judges' biases.
 

### 2024-12-04 index.qmd
section:issues,CaseV; process:write

@@
 Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$.
-Here, $\sigma_{B}$ and $\sigma_{A}$ are the previously defined discriminal dispersions, while $\rho$ measures the correlation between the discriminal processes of the stimuli.
+Here, $\sigma_{B}$ and $\sigma_{A}$ denote the discriminal dispersions of the stimuli, while $\rho$ represents the correlation between their discriminal processes.
 This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in another.
@@
 <!-- 5. The effects of correlation on the discriminal difference -->
-Notably, the correlation between the discriminal processes, $\rho$, plays a pivotal role in determining comparison outcomes by shaping the distribution of the discriminal difference between the stimuli.
+Notably, the correlation between the discriminal processes, $\rho$, plays a pivotal role in determining the comparison outcome by shaping the distribution of the discriminal difference between the stimuli.
 Specifically, as the correlation increases, reflecting a stronger dependence of the judge's perception of quality in one stimulus on his perception of the other, the distribution of the discriminal difference narrows.
@@
 <!-- 6. example of the effects of correlation on the discriminal difference -->
-@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the two texts depicted in @fig-discriminal_process.
+@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process.
 Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A.
-This is evident from the larger proportion of area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation).
-Conversely, it is not hard to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance that the discriminal difference distinctly favors one text over the other.
-This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero (not illustrated).
+This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation) compared to curves with lower correlations.
+Moreover, although the figure does not illustrate this scenario, it is reasonable to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance of the discriminal difference distinctly favoring one text over the other.
+This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero.
 
@@
 <!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
-On the one hand, the assumption of *equal dispersion between stimuli* asserts that the variability in perceived trait levels remains consistent across all stimuli.
-However, Thurstone noted that this assumption may fail when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376], as these can introduce greater discrepancies into judgments due to their unique characteristics.
-One example of this is the evaluation of creativity.
-In this context, for instance, it is plausible that a student completes an assignment in a highly creative yet unexpected way.
-Judges, unprepared for such originality, may experience greater discrepancies in their judgments of the assignment.
-As a result, the discriminal difference associated with this assignment is expected to show a broader dispersion, reflecting the larger variation in its discriminal process compared to others.
-Although direct evidence for this specific scenario is lacking, [studies such as @vanDaal_2020 [chap.
-8] suggest that misfitting texts tend to exhibit larger dispersion, likely due to increased confusion among judges regarding the texts.]{style="color:red;"}
+On the one hand, the assumption of *equal dispersion between stimuli* suggests that variability in perceived trait levels is consistent across all stimuli.
+However, Thurstone observed that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+376], as these traits and stimuli can introduce greater judgment discrepancies due to their unique characteristics.
+In fact, research on text quality may already provide evidence of the violation of this assumption, particularly through the use of misfit statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
+Misfit statistics quantify the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp.
+12; @Goossens_et_al_2018, pp.
+20].
+In this context, texts identified as "misfits" display greater judgment discrepancies than other texts.
+This implies that the discriminal difference associated with these texts shows a broader dispersion, suggesting that their discriminal processes also exhibit greater variation than other texts.
+This reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations due to the unique or extreme characteristics of the stimuli.
 
@@
 The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
+-->
+
+<!-- For instance, in evaluating creativity, a student might complete an assignment in a highly creative but unexpected manner.
+Judges, unprepared for such originality, may produce more divergent evaluations of the assignment.
+These increased discrepancies in judgments suggest that the discriminal difference associated with the assignment has a broader dispersion, indicating greater variation in the discriminal process for this assignment compared to others.
 -->
@@
 -->
+
+<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
+[@Pollitt_2012a, pp.
+165] -->
+
+<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
+
+<!-- a misfit statistic.
+Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
+It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
+12] -->
 
@@
 Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
-For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 

### 2024-12-04 _tex/index.tex
Built site for gh-pages

@@
 \(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
-Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
-discriminal dispersions, while \(\rho\) measures the correlation between
-the discriminal processes of the stimuli. This correlation quantifies
-the dependence of the judge's perception of the trait in one stimulus on
-his perception of the same trait in another.
+Here, \(\sigma_{B}\) and \(\sigma_{A}\) denote the discriminal
+dispersions of the stimuli, while \(\rho\) represents the correlation
+between their discriminal processes. This correlation quantifies the
+dependence of the judge's perception of the trait in one stimulus on his
+perception of the same trait in another.
 
@@
 Notably, the correlation between the discriminal processes, \(\rho\),
-plays a pivotal role in determining comparison outcomes by shaping the
-distribution of the discriminal difference between the stimuli.
+plays a pivotal role in determining the comparison outcome by shaping
+the distribution of the discriminal difference between the stimuli.
 Specifically, as the correlation increases, reflecting a stronger
@@
 Figure~\ref{fig-correlation} illustrates how varying correlations
-influence the distribution of the discriminal difference for the two
-texts depicted in Figure~\ref{fig-discriminal_process}. Since the texts
-differ in quality, higher correlations increase the likelihood that the
+influence the distribution of the discriminal difference for the texts
+depicted in Figure~\ref{fig-discriminal_process}. Since the texts differ
+in quality, higher correlations increase the likelihood that the
 discriminal difference distinctly favors text B over text A. This is
-evident from the larger proportion of area under the curve,
+evident from the larger proportion of the area under the curve,
 \(P(B > A)\), that lies above zero (shaded gray area in
-Figure~\ref{fig-correlation}). Conversely, it is not hard to infer that
-if the texts had similar or identical quality levels, higher
-correlations would likely reduce the chance that the discriminal
-difference distinctly favors one text over the other. This probability
-reduction occurs because the distribution of the discriminal difference
-would become more narrowly centered around zero (not illustrated).
+Figure~\ref{fig-correlation}) compared to curves with lower
+correlations. Moreover, although the figure does not illustrate this
+scenario, it is reasonable to infer that if the texts had similar or
+identical quality levels, higher correlations would likely reduce the
+chance of the discriminal difference distinctly favoring one text over
+the other. This probability reduction occurs because the distribution of
+the discriminal difference would become more narrowly centered around
+zero.
 
@@
 On the one hand, the assumption of \emph{equal dispersion between
-stimuli} asserts that the variability in perceived trait levels remains
-consistent across all stimuli. However, Thurstone noted that this
-assumption may fail when researchers assess complex traits or
-heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these can
-introduce greater discrepancies into judgments due to their unique
-characteristics. One example of this is the evaluation of creativity. In
-this context, for instance, it is plausible that a student completes an
-assignment in a highly creative yet unexpected way. Judges, unprepared
-for such originality, may experience greater discrepancies in their
-judgments of the assignment. As a result, the discriminal difference
-associated with this assignment is expected to show a broader
-dispersion, reflecting the larger variation in its discriminal process
-compared to others. Although direct evidence for this specific scenario
-is lacking, {studies such as \citet[chap.~8]{vanDaal_2020} suggest that
-misfitting texts tend to exhibit larger dispersion, likely due to
-increased confusion among judges regarding the texts.}
+stimuli} suggests that variability in perceived trait levels is
+consistent across all stimuli. However, Thurstone observed that this
+assumption may not hold when researchers assess complex traits or
+heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
+and stimuli can introduce greater judgment discrepancies due to their
+unique characteristics. In fact, research on text quality may already
+provide evidence of the violation of this assumption, particularly
+through the use of misfit statistics
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
+Misfit statistics quantify the judgment discrepancies associated with a
+given stimulus
+\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
+In this context, texts identified as ``misfits'' display greater
+judgment discrepancies than other texts. This implies that the
+discriminal difference associated with these texts shows a broader
+dispersion, suggesting that their discriminal processes also exhibit
+greater variation than other texts. This reasoning also applies to
+``misfit'' judges, whose evaluations reflect substantial deviations due
+to the unique or extreme characteristics of the stimuli.
 
@@
 example, these scores have been used to identify `misfit' judges and
-stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018},
+stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
 detect biases in judges' ratings

### 2024-12-04 index.qmd
section:issues,CaseV; process:write

@@
 The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level.
-According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.
+According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the perceived trait variability of the stimulus.
 
@@
 
+<!-- Model misspecification happens when the set of probability distributions considered by the statistician does not include the distribution that generated the observed data. -->
+
 ::: {#fig-logistic_vs_normal layout-nrow=2}
@@
 <!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
-On the one hand, the assumption of *equal dispersion between stimuli* suggests that variability in perceived trait levels is consistent across all stimuli.
+On the one hand, the assumption of *equal dispersion between stimuli* suggests that the perceived trait variability is consistent across all stimuli.
 However, Thurstone observed that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376], as these traits and stimuli can introduce greater judgment discrepancies due to their unique characteristics.
-In fact, research on text quality may already provide evidence of the violation of this assumption, particularly through the use of misfit statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
+376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics.
+Research on text quality may already provide evidence of this assumption's violation, particularly when considering the role of misfit statistics in the analysis of CJ data [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
 Misfit statistics quantify the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp.
@@
 20].
-In this context, texts identified as "misfits" display greater judgment discrepancies than other texts.
-This implies that the discriminal difference associated with these texts shows a broader dispersion, suggesting that their discriminal processes also exhibit greater variation than other texts.
-This reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations due to the unique or extreme characteristics of the stimuli.
-
-<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
-[@Thurstone_1927a, pp.
-376] -->
+For instance, texts identified as "misfits" usually exhibit more judgment discrepancies than others.
+This, in turn, implies that the discriminal difference associated with these texts shows a broader dispersion, ultimately suggesting that their discriminal processes also exhibit more variation than other texts.
+Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations due to the unique characteristics of the stimuli or the judges themselves.
 
@@
 -->
+
+<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
+[@Thurstone_1927a, pp.
+376] -->
 
@@
 
-<!-- Misfiting assignments might be assignments with larger dispersion, because there is a lot of confusion about them.
-When assessing creativity, for instance, a student might complete an assignment in a highly creative but unexpected way.
-Judges, unprepared for such originality, may experience uncertainty, leading to greater variability in their judgments about the assignment.
-Then the discriminal difference is all over the place.
--->
-
 <!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
@@
 <!-- 5. But what are the challenges when these assumptions are violated? -->
-[what are the challenges when the equal dispersion assumption is violated?]{style="color:red;"}
-
-<!-- But the statistical literature warns against mis-specifying the trait’s measurement model.
-When mis-specifying the measurement model by assuming, for example, equal discriminal dispersions across stimuli, the model can overlook meaningful differences in the reliability of the trait.
-Taken together, these mis-specifications and oversights can lead to misleading conclusions about the trait measurement, ultimately casting doubts about the validity of the trait, as validity cannot exist without reliability [@Perron_et_al_2015, pp.
+Moreover, assuming equal dispersion between stimuli despite its violation causes Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli.
+Moreover, if researchers recognize that misfit statistics help to identify such violations, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020].
+The direction and magnitude of these biases remain unpredictable because they depend on which stimuli researchers exclude from the analysis.
+Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
 2].
--->
 
@@
 
-<!-- If your scores are not reliable they cannot be valid.
-People have been employing a model with assumptions that they do not know if those assumptions apply for their cases.
-And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid.
--->
+<!-- People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. -->
 
@@
 
+<!-- Although no CJ literature offer solutions to mitigate the aforementioned issues, we can take lesson from the Structural Equation Model (SEM) literature. More specifically, from @Depaoli_2021. -->
+
 <!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
-
-<!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->
 
@@
 
-<!-- Reliability is a necessary but not sufficient condition for validity.
-Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp.
-2].
-If your scores are not reliable they cannot be valid.
-People have been employing a model with assumptions that they do not know if those assumptions apply for their cases.
-And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid.
--->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. -->
 
@@
 -->
+
+<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
 

### 2024-12-04 _tex/index.tex
Built site for gh-pages

@@
 the stimulus' position on the trait continuum, and its dispersion, known
-as the \emph{discriminal dispersion}, reflects the variability in the
-perceived trait level of the stimulus.
+as the \emph{discriminal dispersion}, reflects the perceived trait
+variability of the stimulus.
 
@@
 On the one hand, the assumption of \emph{equal dispersion between
-stimuli} suggests that variability in perceived trait levels is
-consistent across all stimuli. However, Thurstone observed that this
-assumption may not hold when researchers assess complex traits or
-heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
-and stimuli can introduce greater judgment discrepancies due to their
-unique characteristics. In fact, research on text quality may already
-provide evidence of the violation of this assumption, particularly
-through the use of misfit statistics
+stimuli} suggests that the perceived trait variability is consistent
+across all stimuli. However, Thurstone observed that this assumption may
+not hold when researchers assess complex traits or heterogeneous stimuli
+\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
+introduce judgment discrepancies due to their unique characteristics.
+Research on text quality may already provide evidence of this
+assumption's violation, particularly when considering the role of misfit
+statistics in the analysis of CJ data
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
@@
 \citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
-In this context, texts identified as ``misfits'' display greater
-judgment discrepancies than other texts. This implies that the
+For instance, texts identified as ``misfits'' usually exhibit more
+judgment discrepancies than others. This, in turn, implies that the
 discriminal difference associated with these texts shows a broader
-dispersion, suggesting that their discriminal processes also exhibit
-greater variation than other texts. This reasoning also applies to
-``misfit'' judges, whose evaluations reflect substantial deviations due
-to the unique or extreme characteristics of the stimuli.
+dispersion, ultimately suggesting that their discriminal processes also
+exhibit more variation than other texts. Notably, this reasoning also
+applies to ``misfit'' judges, whose evaluations reflect substantial
+deviations due to the unique characteristics of the stimuli or the
+judges themselves.
 
-{what are the challenges when the equal dispersion assumption is
-violated?}
+Moreover, assuming equal dispersion between stimuli despite its
+violation causes Case V (and the BTL model) to overlook critical
+differences in the reliability of the trait across stimuli. Moreover, if
+researchers recognize that misfit statistics help to identify such
+violations, the usual practice in the CJ literature of excluding stimuli
+based on these statistics
+\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
+discarding valuable information and introducing bias into the trait's
+estimates \citep{Zimmerman_1994, McElreath_2020}. The direction and
+magnitude of these biases remain unpredictable because they depend on
+which stimuli researchers exclude from the analysis. Together, these
+oversights undermine the reliability of the trait and ultimately
+compromise its validity \citep[pp.~2]{Perron_et_al_2015}.
 

### 2024-12-05 index.qmd
section:issues,CaseV; process:write

@@
 The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
-The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level.
-According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the perceived trait variability of the stimulus.
+The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait.
+According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability of the stimulus' perceived trait.
 
@@
 This law posits that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait.
-This principle highlights that the outcome of a pairwise comparison likely depends on the relative distance between stimuli rather than their absolute positions on the trait continuum.
+This emphasizes that the outcome of a pairwise comparison likely depends on the relative distance between stimuli rather than their absolute positions on the trait continuum.
 
@@
 The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$.
-As a result, the dichotomous outcome of this comparison would mostly favor text B over text A.
+As a result, the dichotomous outcome of this comparison would likely favor text B over text A.
 
@@
 Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A.
-This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation) compared to curves with lower correlations.
+This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero compared to curves with lower correlations.
 Moreover, although the figure does not illustrate this scenario, it is reasonable to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance of the discriminal difference distinctly favoring one text over the other.
@@
 <!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
-On the one hand, the assumption of *equal dispersion between stimuli* suggests that the perceived trait variability is consistent across all stimuli.
-However, Thurstone observed that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics.
-Research on text quality may already provide evidence of this assumption's violation, particularly when considering the role of misfit statistics in the analysis of CJ data [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
-Misfit statistics quantify the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp.
+On the one hand, the assumption of *equal dispersion between stimuli* suggests that the variability of the perceived trait remains consistent across all stimuli.
+However, Thurstone contended that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+376], as these traits and stimuli can introduce judgment discrepancies due to their unique features.
+Indeed, evidence of such violation may already exist in the CJ literature as misfit statistics.
+These statistics measure the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp.
 12; @Goossens_et_al_2018, pp.
 20].
-For instance, texts identified as "misfits" usually exhibit more judgment discrepancies than others.
-This, in turn, implies that the discriminal difference associated with these texts shows a broader dispersion, ultimately suggesting that their discriminal processes also exhibit more variation than other texts.
-Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations due to the unique characteristics of the stimuli or the judges themselves.
+For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than comparisons involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
+This finding implies that the discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts.
+Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
 
@@
 <!-- 5. But what are the challenges when these assumptions are violated? -->
-Moreover, assuming equal dispersion between stimuli despite its violation causes Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli.
-Moreover, if researchers recognize that misfit statistics help to identify such violations, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020].
-The direction and magnitude of these biases remain unpredictable because they depend on which stimuli researchers exclude from the analysis.
-Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
+However, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli.
+Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the common practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+12].
+The direction and magnitude of these biases are unpredictable, as they depend on the particular stimuli researchers exclude from the analysis.
+Taken together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
 2].
@@
 Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
-26] combined with a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap.
+26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap.
 15], depending on the theoretical and statistical treatment of judges' biases.
@@
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is evident that the BTL model commonly functions as the trait's measurement model in CJ experiments [@Andrich_1978; @Bramley_2008].
+Building on the previous section, it is evident that, besides its inconveniences, the BTL model commonly functions as the trait's measurement model for CJ data [@Andrich_1978; @Bramley_2008].
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
@@
 This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
-Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.
 

### 2024-12-05 _tex/index.tex
Built site for gh-pages

--- 2024-12-04 _tex/index.tex
@@
 stimulus has on judges or, more simply, their underlying perception of
-the stimulus' trait level. According to the theory, the discriminal
-process for each stimulus follows a Normal distribution, where its mode
-(mean), referred to as the \emph{modal discriminal process}, indicates
-the stimulus' position on the trait continuum, and its dispersion, known
-as the \emph{discriminal dispersion}, reflects the variability in the
-perceived trait level of the stimulus.
+the stimulus' trait. According to the theory, the discriminal process
+for each stimulus follows a Normal distribution, where its mode (mean),
+referred to as the \emph{modal discriminal process}, indicates the
+stimulus' position on the trait continuum, and its dispersion, known as
+the \emph{discriminal dispersion}, reflects the variability of the
+stimulus' perceived trait.
 
@@
 perceives the stimulus positioned further along the trait continuum as
-having a higher level of that trait. This principle highlights that the
-outcome of a pairwise comparison likely depends on the relative distance
-between stimuli rather than their absolute positions on the trait
-continuum.
+having a higher level of that trait. This emphasizes that the outcome of
+a pairwise comparison likely depends on the relative distance between
+stimuli rather than their absolute positions on the trait continuum.
 
@@
 A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
-a result, the dichotomous outcome of this comparison would mostly favor
+a result, the dichotomous outcome of this comparison would likely favor
 text B over text A.
@@
 evident from the larger proportion of the area under the curve,
-\(P(B > A)\), that lies above zero (shaded gray area in
-Figure~\ref{fig-correlation}) compared to curves with lower
+\(P(B > A)\), that lies above zero compared to curves with lower
 correlations. Moreover, although the figure does not illustrate this
@@
 On the one hand, the assumption of \emph{equal dispersion between
-stimuli} suggests that variability in perceived trait levels is
-consistent across all stimuli. However, Thurstone observed that this
+stimuli} suggests that the variability of the perceived trait remains
+consistent across all stimuli. However, Thurstone contended that this
 assumption may not hold when researchers assess complex traits or
 heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
-and stimuli can introduce greater judgment discrepancies due to their
-unique characteristics. In fact, research on text quality may already
-provide evidence of the violation of this assumption, particularly
-through the use of misfit statistics
+and stimuli can introduce judgment discrepancies due to their unique
+features. Indeed, evidence of such violation may already exist in the CJ
+literature as misfit statistics. These statistics measure the judgment
+discrepancies associated with a given stimulus
+\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
+For instance, labeling texts as ``misfits'' indicates that comparisons
+involving these texts result in more judgment discrepancies than
+comparisons involving other texts
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
-Misfit statistics quantify the judgment discrepancies associated with a
-given stimulus
-\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
-In this context, texts identified as ``misfits'' display greater
-judgment discrepancies than other texts. This implies that the
-discriminal difference associated with these texts shows a broader
-dispersion, suggesting that their discriminal processes also exhibit
-greater variation than other texts. This reasoning also applies to
-``misfit'' judges, whose evaluations reflect substantial deviations due
-to the unique or extreme characteristics of the stimuli.
+This finding implies that the discriminal differences associated with
+``misfits'' texts usually display a broader dispersion, suggesting that
+their discriminal processes also exhibit more variation than other
+texts. Notably, this reasoning also applies to ``misfit'' judges, whose
+evaluations reflect substantial deviations from the shared consensus due
+to the unique characteristics of the stimuli or the judges themselves.
 
-{what are the challenges when the equal dispersion assumption is
-violated?}
+However, assuming equal dispersions between stimuli despite its
+violation can lead Case V (and the BTL model) to overlook critical
+differences in the reliability of the trait across stimuli. Furthermore,
+if researchers acknowledge that misfit statistics help identify these
+critical differences, the common practice in the CJ literature of
+excluding stimuli based on these statistics
+\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
+discarding valuable information and introducing bias into the trait's
+estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
+direction and magnitude of these biases are unpredictable, as they
+depend on the particular stimuli researchers exclude from the analysis.
+Taken together, these oversights undermine the reliability of the trait
+and ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.
 
@@
 Equation Model (MSEM) \citep[chap.~26]{Hoyle_et_al_2023} combined with a
-multidimensional or two-parameter logistic IRT model
-\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical and
-statistical treatment of judges' biases.
+multidimensional or two-parameter logistic Item Response Theory (IRT)
+model \citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
+and statistical treatment of judges' biases.
 
@@
 
-Building on the previous section, it is evident that the BTL model
-commonly functions as the trait's measurement model in CJ experiments
-\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
-manifest variables contribute to the estimation of latent variables
-\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
-researchers use the BTL model to process the dichotomous outcomes
-resulting from the pairwise comparisons (the manifest variables) to
-estimate scores that reflect the underlying quality level of texts (the
-latent variable)
+Building on the previous section, it is evident that, besides its
+inconveniences, the BTL model commonly functions as the trait's
+measurement model for CJ data \citep{Andrich_1978, Bramley_2008}. A
+measurement model specifies how manifest variables contribute to the
+estimation of latent variables \citep{Everitt_et_al_2010}. For example,
+when evaluating text quality, researchers use the BTL model to process
+the dichotomous outcomes resulting from the pairwise comparisons (the
+manifest variables) to estimate scores that reflect the underlying
+quality level of texts (the latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
@@
 and their uncertainties simultaneously, rather than treating them as
-separate elements. Therefore, an integrated approach that combines CJ's
-measurement and structural models can offer significant advantages.
+separate elements.
 

### 2024-12-05 index.qmd
section:issues,CaseV; process:write

@@
 <!-- 3. But what are limitations of Case V (and the BTL model)? -->
-However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+Nevertheless, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
@@
 <!-- 5. But what are the challenges when these assumptions are violated? -->
-However, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli.
+Nonetheless, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp.
+370].
 Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the common practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
-The direction and magnitude of these biases are unpredictable, as they depend on the particular stimuli researchers exclude from the analysis.
+The direction and magnitude of these biases are unpredictable, as they depend on the specific stimuli researchers exclude from the analysis.
 Taken together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
@@
 <!-- 6. So, how do we mitigate these risks? -->
-[how do we mitigate these risks?]{style="color:red;"}
-
-<!-- Although no CJ literature offer solutions to mitigate the aforementioned issues, we can take lesson from the Structural Equation Model (SEM) literature. More specifically, from @Depaoli_2021. -->
+Fortunately, the statistical literature offers a solution to address the abovementioned issues.
+This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
+12].
 
@@
 <!-- 7. What does the zero correlation between stimuli imply? and does it hold? -->
-On the other hand hand, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @sec-thurstone_theory).
+On the other hand, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @sec-thurstone_theory).
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
@@
 346].
-For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments.
+For example, this could occur when a judge assessing the argumentative quality of a text places more weight on grammatical accuracy than other judges, ultimately favoring texts with fewer errors but weaker arguments.
 While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
@@
 482].
-These issues collectively undermine the trait's reliability and ultimately compromise the validity of the trait's estimates [@Perron_et_al_2015, pp.
+These issues collectively also undermine the trait's reliability and ultimately compromise the validity of the trait's estimates [@Perron_et_al_2015, pp.
 2].
@@
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is evident that, besides its inconveniences, the BTL model commonly functions as the trait's measurement model for CJ data [@Andrich_1978; @Bramley_2008].
+Building on the previous section, it is evident that, besides its inconveniences, the BTL model commonly functions as the trait's measurement model in CJ assessments.
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
@@
 <!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
+Researchers then typically use these estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
 For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
-138] and Item Response Theory (IRT) [@Fox_2010, chap.
+138] and IRT [@Fox_2010, chap.
 6; @vanderLinden_et_al_2017_I, chap.

### 2024-12-05 _tex/index.tex
Built site for gh-pages

@@
 
-However, Thurstone originally developed Case V to provide a ``rather
-coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
+Nevertheless, Thurstone originally developed Case V to provide a
+``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
 prioritizing statistical simplicity over precision in trait measurement
@@
 
-However, assuming equal dispersions between stimuli despite its
+Nonetheless, assuming equal dispersions between stimuli despite its
 violation can lead Case V (and the BTL model) to overlook critical
-differences in the reliability of the trait across stimuli. Furthermore,
-if researchers acknowledge that misfit statistics help identify these
-critical differences, the common practice in the CJ literature of
-excluding stimuli based on these statistics
+differences in the reliability of the trait across stimuli, resulting in
+inaccurate conclusions about the trait's estimates
+\citep[pp.~370]{McElreath_2020}. Furthermore, if researchers acknowledge
+that misfit statistics help identify these critical differences, the
+common practice in the CJ literature of excluding stimuli based on these
+statistics
 \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
@@
 direction and magnitude of these biases are unpredictable, as they
-depend on the particular stimuli researchers exclude from the analysis.
+depend on the specific stimuli researchers exclude from the analysis.
 Taken together, these oversights undermine the reliability of the trait
@@
 
-{how do we mitigate these risks?}
+Fortunately, the statistical literature offers a solution to address the
+abovementioned issues. This solution involves using models that extend
+traditional approaches to account for the different variability in the
+stimuli, referred to as over-dispersed models
+\citep[chap.~12]{McElreath_2020}.
 
-On the other hand hand, the assumption of \emph{zero correlation between
+On the other hand, the assumption of \emph{zero correlation between
 stimuli} implies that, during a pairwise comparison, a judge's
@@
 occur when a judge assessing the argumentative quality of a text places
-disproportionate emphasis on grammatical accuracy, ultimately favoring
-texts with fewer errors but weaker arguments. While direct evidence for
-this specific scenario is lacking, studies such as
+more weight on grammatical accuracy than other judges, ultimately
+favoring texts with fewer errors but weaker arguments. While direct
+evidence for this specific scenario is lacking, studies such as
 \citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
@@
 reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
-undermine the trait's reliability and ultimately compromise the validity
-of the trait's estimates \citep[pp.~2]{Perron_et_al_2015}.
+also undermine the trait's reliability and ultimately compromise the
+validity of the trait's estimates \citep[pp.~2]{Perron_et_al_2015}.
 
@@
 inconveniences, the BTL model commonly functions as the trait's
-measurement model for CJ data \citep{Andrich_1978, Bramley_2008}. A
-measurement model specifies how manifest variables contribute to the
-estimation of latent variables \citep{Everitt_et_al_2010}. For example,
-when evaluating text quality, researchers use the BTL model to process
-the dichotomous outcomes resulting from the pairwise comparisons (the
-manifest variables) to estimate scores that reflect the underlying
-quality level of texts (the latent variable)
+measurement model in CJ assessments. A measurement model specifies how
+manifest variables contribute to the estimation of latent variables
+\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
+researchers use the BTL model to process the dichotomous outcomes
+resulting from the pairwise comparisons (the manifest variables) to
+estimate scores that reflect the underlying quality level of texts (the
+latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
 
-Researchers then typically use the estimated BTL scores, or their
+Researchers then typically use these estimated BTL scores, or their
 transformations, to conduct additional analyses or hypothesis tests. For
@@
 To mitigate these risks, principles from Structural Equation Modeling
-(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
+(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and IRT
 \citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}

### 2024-12-06 index.qmd
section:issues,CaseV; process:write

@@
 Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed.
-The present study primarily focuses on two: the over-reliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data, and the apparent disconnect between CJ's trait measurement and hypothesis testing.
-The following sections begin with a brief overview of Thurstone's theory and a detailed discussion of these issues.
+The present study primarily focuses on two: the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data, and the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+The following sections begin with a brief overview of Thurstone's theory and a detailed examination of these issues.
 Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address the two concerns simultaneously.
@@
 <!-- 1. Thurstone's Theory of CJ: discriminal process -->
-In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
+In its most general form, Thurstone’s theory addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
 267].
-The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
-The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait.
-According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability of the stimulus' perceived trait.
+The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
+The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their perception of the stimulus trait.
+According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), called the *modal discriminal process*, indicates the stimulus position on the trait continuum, while its dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.
 
 <!-- 2. Example of discriminal processes -->
-For instance, @fig-discriminal_process illustrates the discriminal process distributions along a quality trait continuum for two written texts.
-The figure shows that these processes follow a Normal distribution.
-Moreover, it depicts differences in the texts' positions along the quality trait continuum, where text B is positioned further along the continuum than text A, as indicated by their modal discriminal processes ($S_{B}$ and $S_{A}$).
-Finally, it highlights differences in the texts' discriminal dispersions ($\sigma_{B}$ and $\sigma_{A}$), showing that text B exhibits a greater variability in its perceived quality than text A, as reflected by its wider distribution.
+@fig-discriminal_process illustrates the discriminal processes along a quality trait continuum for two written texts.
+The figure shows that these processes individually follow a Normal distribution.
+Furthermore, it shows that the modal discriminal process for Text B is positioned further along the continuum than Text A $(S_{B} > S_{A})$.
+Additionally, the figure highlights the broader distribution of Text B compared to Text A as a result of its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
 
-![Example distributions of the discriminal processes for two written texts](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
+![Discriminal processes for two written texts](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
 
 <!-- 3. Thurstone's Theory of CJ: discriminal difference -->
-However, because the discriminal process of a single stimulus is not directly observable, the theory introduces the *law of comparative judgment*.
+However, since the discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
 This law posits that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait.
-This emphasizes that the outcome of a pairwise comparison likely depends on the relative distance between stimuli rather than their absolute positions on the trait continuum.
+This emphasizes that the relative distance between stimuli, rather than their absolute positions on the continuum, determines the outcome of the pairwise comparison.
+Indeed, the theory assumes that the observed dichotomous outcome arises from the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.
+Since the individual discriminal processes follow a Normal distribution, the discriminal difference also follows a Normal distribution [@Andrich_1978], with its mode (mean) representing the relative separation between the stimuli, while its dispersion reflects the variability in that relative separation.
 
-Indeed, the theory assumes that the observed dichotomous outcome arises from the distribution of the difference between the underlying discriminal processes of the stimuli, known as the *discriminal difference*.
-Since the individual discriminal processes follow a Normal distribution, their difference also follows a Normal distribution [@Andrich_1978].
-The mode (mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA} = S_{B} - S_{A}$.
+<!-- The mode (mean) of this distribution, representing the (average) relative separation, is given by the difference between the modal discriminal processes of the stimuli $S_{BA}=S_{B}-S_{A}$.
 Meanwhile, the dispersion of the distribution, reflecting the variability in the relative separation, is calculated as $\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}$.
@@
 This correlation quantifies the dependence of the judge's perception of the trait in one stimulus on his perception of the same trait in another.
+-->
 
 <!-- 4. Example of discriminal difference -->
-@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process, assuming a correlation of $\rho = 0.6$.
-The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$.
-As a result, the dichotomous outcome of this comparison would likely favor text B over text A.
+@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process.
+The figure illustrates that the judge perceives Text B as having significantly higher quality than Text A, as indicated by their discriminal difference $(S_{B} - S_{A} > 0)$ and the shaded gray area under the curve $P(B > A)$.
+Therefore, the dichotomous outcome of this comparison would likely favor Text B over Text A.
 
-![Example distribution of the discriminal difference for the two texts shown in @fig-discriminal_process, assuming a correlation of 0.6](/images/figures/discriminal_difference.png){#fig-discriminal_difference width=70%}
+![Discriminal difference for two written texts](/images/figures/discriminal_difference.png){#fig-discriminal_difference width=70%}
 
-<!-- 5. The effects of correlation on the discriminal difference -->
-Notably, the correlation between the discriminal processes, $\rho$, plays a pivotal role in determining the comparison outcome by shaping the distribution of the discriminal difference between the stimuli.
-Specifically, as the correlation increases, reflecting a stronger dependence of the judge's perception of quality in one stimulus on his perception of the other, the distribution of the discriminal difference narrows.
-This narrowing ultimately impacts the area under the curve that determines the comparison outcome and, consequently, the conclusions drawn from this outcome.
+# The two critical issues in CJ literature {#sec-theory-issues}
 
-<!-- 6. example of the effects of correlation on the discriminal difference -->
-@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process.
-Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A.
-This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero compared to curves with lower correlations.
-Moreover, although the figure does not illustrate this scenario, it is reasonable to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance of the discriminal difference distinctly favoring one text over the other.
-This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero.
-
-![The effect of correlation on the distribution of the discriminal difference of the same two written text](/images/figures/correlation.png){#fig-correlation width=70%}
-
-# Three critical issues in CJ literature {#sec-theory-issues}
+This section examines the two critical issues in the CJ literature that serve as the primary focus of this study.
+The first is the overreliance on Thurstone’s Case V assumptions in the statistical analysis of CJ data.
+The second is the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
 
@@
 677].
-As a result, its assumptions may not be suitable for applications beyond the psycho-physical contexts for which it was created.
-Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions could prove problematic in the presence of complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+Moreover, he warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+270], acknowledging that some assumptions might be problematic when complex traits or heterogeneous stimuli are involved [@Thurstone_1927a, pp.
 376].
-Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the equal dispersion and zero correlation between stimuli.
+Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.
 
@@
 
-<!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
-On the one hand, the assumption of *equal dispersion between stimuli* suggests that the variability of the perceived trait remains consistent across all stimuli.
+### The assumption of equal dispersions between stimuli
+
+<!-- 1. What does the equal dispersion between stimuli imply? -->
+The discriminal dispersions of stimuli are crucial in determining the dichotomous outcomes of pairwise comparisons.
+Discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome.
+For instance, @fig-dispersion illustrates how more uncertainty in the trait perception of a text over another broadens the distribution of their discriminal difference.
+This broadening affects the area under the curve, $P(B>A)$, determining the comparison outcome.
+Additionally, the figure shows that when there are no discrepancies in the texts' discriminal dispersions, meaning the *texts have equal dispersions* $(\sigma_{B}-\sigma_{A}=0)$, the likelihood that the discriminal difference distinctly favors one text over another increases.
+
+![The effect of different dispersions between stimuli on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
+
+<!-- 2. But does it hold?  -->
 However, Thurstone contended that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
@@
 
-<!-- 5. But what are the challenges when these assumptions are violated? -->
-Nonetheless, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp.
+<!-- 3. But what are the challenges when these assumptions are violated? -->
+<!-- Therefore, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to inflate the reliability of the observed dichotomous outcome  -->
+
+<!-- to overlook critical differences in the reliability of the trait across stimuli,  -->
+
+<!-- resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp.
 370].
@@
 2].
+-->
 
@@
 
-<!-- 6. So, how do we mitigate these risks? -->
-Fortunately, the statistical literature offers a solution to address the abovementioned issues.
-This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
-12].
+### The assumption of zero correlation between stimuli
 
-<!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
+<!-- 1. What does the zero correlation between stimuli imply? -->
+Similar to the discriminal dispersions, the correlation between the discriminal processes is also crucial in determining the dichotomous outcomes of pairwise comparisons.
+This correlation shapes the distribution of the discriminal difference, directly influencing the comparison outcome.
+For instance, @fig-correlation illustrates how the stronger dependence of the trait perception in one text over another narrows the distribution of their discriminal difference.
+This narrowing impacts the area under the curve, $P(B>A)$, determining the comparison outcome.
+Additionally, the figure shows that when the trait perception in one text is independent of another, meaning the *texts have zero correlation*, the likelihood that the discriminal difference distinctly favors one text over another is lower compared to when the texts are highly correlated.
 
-<!-- 7. What does the zero correlation between stimuli imply? and does it hold? -->
-On the other hand, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @sec-thurstone_theory).
+<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
+268].
+-->
+
+![The effect of correlation on the distribution of the discriminal difference](/images/figures/correlation.png){#fig-correlation width=70%}
+
+<!-- 2. But does it hold?  -->
+Notably, Thurstone's Case V and the BTL model assume independent trait perceptions across stimuli.
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
@@
 
-<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
-268].
--->
-
-<!-- 8. The first scenario where the assumption does not hold -->
+<!-- 3. The first scenario where the assumption does not hold -->
 In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation.
@@
 
-<!-- 9. The second scenario where the assumption does not hold -->
+<!-- 4. The second scenario where the assumption does not hold -->
 In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
@@
 
-<!-- 10. But what are the challenges when this assumption is violated? -->
-In any case, the psychometric and statistical literature emphasizes the importance of addressing the factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity.
+<!-- 5. But what are the challenges when this assumption is violated? -->
+<!-- In any case, the psychometric and statistical literature emphasizes the importance of addressing the factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity.
 For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the statistical model used for analysis.
@@
 2].
-
-<!-- This further implies that the shape of the discriminal difference distribution that determines the comparison outcome and, consequently, the conclusions drawn from this outcome (see @sec-thurstone_theory, @fig-correlation).
 -->
@@
 <!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. -->
-
-<!-- 11. How to solve the violation of the assumption? -->
-Fortunately, the same literature offers solutions for addressing these issues.
-@Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
-Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
-Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
-26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap.
-15], depending on the theoretical and statistical treatment of judges' biases.
-
-<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem for latent variables?. -->
-
-<!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
 
@@
 
-<!-- 4. So, how do we mitigate these risks? -->
-To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
-138] and IRT [@Fox_2010, chap.
-6; @vanderLinden_et_al_2017_I, chap.
-24] recommend conducting these analyses and tests within a structural model.
-A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
-This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
-
 <!-- ## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3} -->
@@
 <!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
-<!-- - This paper tries to systematically integrate all aspects at play in a single scientific model, we will build the model in a stepwise manner using -->
+<!-- - This paper tries to systematically integrate all aspects at play in a single theoretical model, we will build the model in a stepwise manner using -->
 <!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->
@@
 <!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->
+
+<!-- 6. So, how do we mitigate these risks? -->
+<!-- Fortunately, the statistical literature offers a solution to address the abovementioned issues.
+This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
+12].
+-->
+
+<!-- 11. How to solve the violation of the assumption? -->
+<!-- Fortunately, the same literature offers solutions for addressing these issues.
+@Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
+Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
+Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
+26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap.
+15], depending on the theoretical and statistical treatment of judges' biases.
+-->
+
+<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem for latent variables?. -->
+
+<!-- 4. So, how do we mitigate these risks? -->
+<!-- To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+138] and IRT [@Fox_2010, chap.
+6; @vanderLinden_et_al_2017_I, chap.
+24] recommend conducting these analyses and tests within a structural model.
+A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
+This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
+-->
+
+<!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
 

### 2024-12-06 _tex/index.tex
Built site for gh-pages

@@
 unaddressed. The present study primarily focuses on two: the
-over-reliance on the assumptions of Thurstone's Case V in the
-statistical analysis of CJ data, and the apparent disconnect between
-CJ's trait measurement and hypothesis testing. The following sections
+overreliance on the assumptions of Thurstone's Case V in the statistical
+analysis of CJ data, and the apparent disconnect between CJ's approach
+to trait measurement and hypothesis testing. The following sections
 begin with a brief overview of Thurstone's theory and a detailed
-discussion of these issues. Subsequently, the study introduces a
+examination of these issues. Subsequently, the study introduces a
 theoretical model for CJ that builds upon Thurstone's theory, alongside
@@
 
-In its most general form, Thurstone's theory deals with pairwise
+In its most general form, Thurstone's theory addresses pairwise
 comparisons of stimuli made by a single judge
-\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
-factors determine the dichotomous outcome of these comparisons: the
-discriminal process of each stimulus and their discriminal difference.
-The \emph{discriminal process} represents the psychological impact each
-stimulus has on judges or, more simply, their underlying perception of
-the stimulus' trait. According to the theory, the discriminal process
-for each stimulus follows a Normal distribution, where its mode (mean),
-referred to as the \emph{modal discriminal process}, indicates the
-stimulus' position on the trait continuum, and its dispersion, known as
-the \emph{discriminal dispersion}, reflects the variability of the
-stimulus' perceived trait.
+\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
+determine the dichotomous outcome of these comparisons: the discriminal
+process of each stimulus and their discriminal difference. The
+\emph{discriminal process} represents the psychological impact each
+stimulus has on judges or, more simply, their perception of the stimulus
+trait. According to the theory, the discriminal process for each
+stimulus follows a Normal distribution, where its mode (mean), called
+the \emph{modal discriminal process}, indicates the stimulus position on
+the trait continuum, while its dispersion, known as the
+\emph{discriminal dispersion}, reflects variability in the perceived
+trait of the stimulus.
 
-For instance, Figure~\ref{fig-discriminal_process} illustrates the
-discriminal process distributions along a quality trait continuum for
-two written texts. The figure shows that these processes follow a Normal
-distribution. Moreover, it depicts differences in the texts' positions
-along the quality trait continuum, where text B is positioned further
-along the continuum than text A, as indicated by their modal discriminal
-processes (\(S_{B}\) and \(S_{A}\)). Finally, it highlights differences
-in the texts' discriminal dispersions (\(\sigma_{B}\) and
-\(\sigma_{A}\)), showing that text B exhibits a greater variability in
-its perceived quality than text A, as reflected by its wider
-distribution.
+Figure~\ref{fig-discriminal_process} illustrates the discriminal
+processes along a quality trait continuum for two written texts. The
+figure shows that these processes individually follow a Normal
+distribution. Furthermore, it shows that the modal discriminal process
+for Text B is positioned further along the continuum than Text A
+\((S_{B} > S_{A})\). Additionally, the figure highlights the broader
+distribution of Text B compared to Text A as a result of its larger
+discriminal dispersion \((\sigma_{B} > \sigma_{A})\).
 
@@
 
-\caption{\label{fig-discriminal_process}Example distributions of the
-discriminal processes for two written texts}
+\caption{\label{fig-discriminal_process}Discriminal processes for two
+written texts}
 
@@
 
-However, because the discriminal process of a single stimulus is not
-directly observable, the theory introduces the \emph{law of comparative
+However, since the discriminal processes of the stimuli are not directly
+observable, the theory introduces the \emph{law of comparative
 judgment}. This law posits that in pairwise comparisons, a judge
 perceives the stimulus positioned further along the trait continuum as
-having a higher level of that trait. This emphasizes that the outcome of
-a pairwise comparison likely depends on the relative distance between
-stimuli rather than their absolute positions on the trait continuum.
-
-Indeed, the theory assumes that the observed dichotomous outcome arises
-from the distribution of the difference between the underlying
-discriminal processes of the stimuli, known as the \emph{discriminal
-difference}. Since the individual discriminal processes follow a Normal
-distribution, their difference also follows a Normal distribution
-\citep{Andrich_1978}. The mode (mean) of this distribution, representing
-the (average) relative separation, is given by the difference between
-the modal discriminal processes of the stimuli
-\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
-distribution, reflecting the variability in the relative separation, is
-calculated as
-\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
-Here, \(\sigma_{B}\) and \(\sigma_{A}\) denote the discriminal
-dispersions of the stimuli, while \(\rho\) represents the correlation
-between their discriminal processes. This correlation quantifies the
-dependence of the judge's perception of the trait in one stimulus on his
-perception of the same trait in another.
+having a higher level of that trait. This emphasizes that the relative
+distance between stimuli, rather than their absolute positions on the
+continuum, determines the outcome of the pairwise comparison. Indeed,
+the theory assumes that the observed dichotomous outcome arises from the
+difference between the underlying discriminal processes of the stimuli,
+referred to as the \emph{discriminal difference}. Since the individual
+discriminal processes follow a Normal distribution, the discriminal
+difference also follows a Normal distribution \citep{Andrich_1978}, with
+its mode (mean) representing the relative separation between the
+stimuli, while its dispersion reflects the variability in that relative
+separation.
 
@@
 discriminal difference for the texts depicted in
-Figure~\ref{fig-discriminal_process}, assuming a correlation of
-\(\rho = 0.6\). The figure reveals that, under these conditions, the
-judge perceives text B as having significantly higher quality than text
-A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
-a result, the dichotomous outcome of this comparison would likely favor
-text B over text A.
+Figure~\ref{fig-discriminal_process}. The figure illustrates that the
+judge perceives Text B as having significantly higher quality than Text
+A, as indicated by their discriminal difference \((S_{B} - S_{A} > 0)\)
+and the shaded gray area under the curve \(P(B > A)\). Therefore, the
+dichotomous outcome of this comparison would likely favor Text B over
+Text A.
 
@@
 
-\caption{\label{fig-discriminal_difference}Example distribution of the
-discriminal difference for the two texts shown in
-Figure~\ref{fig-discriminal_process}, assuming a correlation of 0.6}
+\caption{\label{fig-discriminal_difference}Discriminal difference for
+two written texts}
 
@@
 
-Notably, the correlation between the discriminal processes, \(\rho\),
-plays a pivotal role in determining the comparison outcome by shaping
-the distribution of the discriminal difference between the stimuli.
-Specifically, as the correlation increases, reflecting a stronger
-dependence of the judge's perception of quality in one stimulus on his
-perception of the other, the distribution of the discriminal difference
-narrows. This narrowing ultimately impacts the area under the curve that
-determines the comparison outcome and, consequently, the conclusions
-drawn from this outcome.
+\section{The two critical issues in CJ
+literature}\label{sec-theory-issues}
 
-Figure~\ref{fig-correlation} illustrates how varying correlations
-influence the distribution of the discriminal difference for the texts
-depicted in Figure~\ref{fig-discriminal_process}. Since the texts differ
-in quality, higher correlations increase the likelihood that the
-discriminal difference distinctly favors text B over text A. This is
-evident from the larger proportion of the area under the curve,
-\(P(B > A)\), that lies above zero compared to curves with lower
-correlations. Moreover, although the figure does not illustrate this
-scenario, it is reasonable to infer that if the texts had similar or
-identical quality levels, higher correlations would likely reduce the
-chance of the discriminal difference distinctly favoring one text over
-the other. This probability reduction occurs because the distribution of
-the discriminal difference would become more narrowly centered around
-zero.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/correlation.png}
-
-}
-
-\caption{\label{fig-correlation}The effect of correlation on the
-distribution of the discriminal difference of the same two written text}
-
-\end{figure}%
-
-\section{Three critical issues in CJ
-literature}\label{sec-theory-issues}
+This section examines the two critical issues in the CJ literature that
+serve as the primary focus of this study. The first is the overreliance
+on Thurstone's Case V assumptions in the statistical analysis of CJ
+data. The second is the apparent disconnect between CJ's approach to
+trait measurement and hypothesis testing.
 
@@
 prioritizing statistical simplicity over precision in trait measurement
-\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
-be suitable for applications beyond the psycho-physical contexts for
-which it was created. Thurstone himself cautioned that its use ``should
-not be made without (an) experimental test''
-\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
-could prove problematic in the presence of complex traits or
-heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}. Consequently,
-given that modern CJ applications frequently involve these traits and
-stimuli, two main assumptions of Case V and, by extension, of the BTL
-model may not consistently hold in theory or practice: the equal
-dispersion and zero correlation between stimuli.
+\citep[pp.~677]{Kelly_et_al_2022}. Moreover, he warned against its
+untested application, stating that its use ``should not be made without
+(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
+that some assumptions might be problematic when complex traits or
+heterogeneous stimuli are involved \citep[pp.~376]{Thurstone_1927a}.
+Consequently, given that modern CJ applications frequently involve such
+traits and stimuli, two main assumptions of Case V and, by extension, of
+the BTL model may not consistently hold in theory or practice: the
+assumption of equal dispersion and zero correlation between stimuli.
 
-On the one hand, the assumption of \emph{equal dispersion between
-stimuli} suggests that the variability of the perceived trait remains
-consistent across all stimuli. However, Thurstone contended that this
-assumption may not hold when researchers assess complex traits or
-heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
-and stimuli can introduce judgment discrepancies due to their unique
-features. Indeed, evidence of such violation may already exist in the CJ
-literature as misfit statistics. These statistics measure the judgment
-discrepancies associated with a given stimulus
+\subsubsection{The assumption of equal dispersions between
+stimuli}\label{the-assumption-of-equal-dispersions-between-stimuli}
+
+The discriminal dispersions of stimuli are crucial in determining the
+dichotomous outcomes of pairwise comparisons. Discrepancies in these
+dispersions shape the distribution of the discriminal difference,
+directly influencing the comparison outcome. For instance,
+Figure~\ref{fig-dispersion} illustrates how more uncertainty in the
+trait perception of a text over another broadens the distribution of
+their discriminal difference. This broadening affects the area under the
+curve, \(P(B>A)\), determining the comparison outcome. Additionally, the
+figure shows that when there are no discrepancies in the texts'
+discriminal dispersions, meaning the \emph{texts have equal dispersions}
+\((\sigma_{B}-\sigma_{A}=0)\), the likelihood that the discriminal
+difference distinctly favors one text over another increases.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/dispersion.png}
+
+}
+
+\caption{\label{fig-dispersion}The effect of different dispersions
+between stimuli on the distribution of the discriminal difference}
+
+\end{figure}%
+
+However, Thurstone contended that this assumption may not hold when
+researchers assess complex traits or heterogeneous stimuli
+\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
+introduce judgment discrepancies due to their unique features. Indeed,
+evidence of such violation may already exist in the CJ literature as
+misfit statistics. These statistics measure the judgment discrepancies
+associated with a given stimulus
 \citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
@@
 
-Nonetheless, assuming equal dispersions between stimuli despite its
-violation can lead Case V (and the BTL model) to overlook critical
-differences in the reliability of the trait across stimuli, resulting in
-inaccurate conclusions about the trait's estimates
-\citep[pp.~370]{McElreath_2020}. Furthermore, if researchers acknowledge
-that misfit statistics help identify these critical differences, the
-common practice in the CJ literature of excluding stimuli based on these
-statistics
-\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
-discarding valuable information and introducing bias into the trait's
-estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
-direction and magnitude of these biases are unpredictable, as they
-depend on the specific stimuli researchers exclude from the analysis.
-Taken together, these oversights undermine the reliability of the trait
-and ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.
+\subsubsection{The assumption of zero correlation between
+stimuli}\label{the-assumption-of-zero-correlation-between-stimuli}
 
-Fortunately, the statistical literature offers a solution to address the
-abovementioned issues. This solution involves using models that extend
-traditional approaches to account for the different variability in the
-stimuli, referred to as over-dispersed models
-\citep[chap.~12]{McElreath_2020}.
+Similar to the discriminal dispersions, the correlation between the
+discriminal processes is also crucial in determining the dichotomous
+outcomes of pairwise comparisons. This correlation shapes the
+distribution of the discriminal difference, directly influencing the
+comparison outcome. For instance, Figure~\ref{fig-correlation}
+illustrates how the stronger dependence of the trait perception in one
+text over another narrows the distribution of their discriminal
+difference. This narrowing impacts the area under the curve, \(P(B>A)\),
+determining the comparison outcome. Additionally, the figure shows that
+when the trait perception in one text is independent of another, meaning
+the \emph{texts have zero correlation}, the likelihood that the
+discriminal difference distinctly favors one text over another is lower
+compared to when the texts are highly correlated.
 
-On the other hand, the assumption of \emph{zero correlation between
-stimuli} implies that, during a pairwise comparison, a judge's
-perception of quality in one text does not influence his perception of
-the same trait in another text (see Section~\ref{sec-thurstone_theory}).
-Thurstone attributed this independence to the cancellation of potential
-judges' biases, driven by two opposing and equally weighted effects
-occurring during the pairwise comparisons
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/correlation.png}
+
+}
+
+\caption{\label{fig-correlation}The effect of correlation on the
+distribution of the discriminal difference}
+
+\end{figure}%
+
+Notably, Thurstone's Case V and the BTL model assume independent trait
+perceptions across stimuli. Thurstone attributed this independence to
+the cancellation of potential judges' biases, driven by two opposing and
+equally weighted effects occurring during the pairwise comparisons
 \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} mathematically
@@
 
-In any case, the psychometric and statistical literature emphasizes the
-importance of addressing the factors that create dependencies between
-stimuli, as failing to do so can lead to inaccurate conclusions about a
-trait's reliability and, by extension, its validity. For instance,
-neglecting additional traits relevant to the stimuli, such as judges'
-biases, often leads to a dimensional mismatch in the statistical model
-used for analysis. This mismatch can potentially inflate the reliability
-of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias
-into the trait's estimates \citep{Ackerman_1989}. Similarly, failing to
-account for hierarchical (grouping) structures reduces the precision of
-model parameter estimates, further amplifying the overestimation of
-reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
-also undermine the trait's reliability and ultimately compromise the
-validity of the trait's estimates \citep[pp.~2]{Perron_et_al_2015}.
-
-Fortunately, the same literature offers solutions for addressing these
-issues. \citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend
-integrating judges' biases into the BTL model. Moreover, the literature
-advocates for the incorporation of relevant hierarchical structures into
-the statistical model to account for these dependencies. Together, these
-additions can result in a model resembling a Multilevel Structural
-Equation Model (MSEM) \citep[chap.~26]{Hoyle_et_al_2023} combined with a
-multidimensional or two-parameter logistic Item Response Theory (IRT)
-model \citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical
-and statistical treatment of judges' biases.
-
 \subsection{The disconnect between trait measurement and hypothesis
@@
 type-II errors \citep{McElreath_2020}.
-
-To mitigate these risks, principles from Structural Equation Modeling
-(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and IRT
-\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
-recommend conducting these analyses and tests within a structural model.
-A structural model specifies how different manifest or latent variables
-influence the latent variable of interest \citep{Everitt_et_al_2010}.
-This approach allows analyses that can account for both the BTL scores
-and their uncertainties simultaneously, rather than treating them as
-separate elements.
 

### 2024-12-06 code/figures.R

@@ -14,12 +14,12 @@ colorArea <- function(from, to, density, ..., col="lightgray", dens=NULL){
 
 
 
-# figure 1a ####
+# figure 1 ####
 
 # distributional parameters
 dpA = c(0, 0.5) # distribution text A (mean, sd)
 dpB = c(1, 1) # distribution text B (mean, sd)
-rAB = 0.6 # correlation between discriminal processes
+rAB = 0 # correlation between discriminal processes
 
 # figure
 png( filename=file.path(main_dir, 'discriminal_process.png'), 
@@ -63,11 +63,10 @@ dev.off()
 
 
 
-# figure 1b ####
+# figure 2 ####
 
 # distributional parameters
-dd = c( dpB[1]-dpA[1], sqrt( dpB[2]^2 + dpA[2]^2 - 2*rAB*dpB[2]*dpA[2] )) 
-dd_alt = c( dpB[1]-dpA[1], sqrt( dpB[2]^2 + dpA[2]^2 )) 
+dd = c( dpB[1]-dpA[1], sqrt( dpB[2]^2 + dpA[2]^2 - 2*rAB*dpB[2]*dpA[2] ) )
 
 # figure
 png( filename=file.path(main_dir, 'discriminal_difference.png'), 
@@ -75,17 +74,17 @@ png( filename=file.path(main_dir, 'discriminal_difference.png'),
 
 # initial figure
 par( mar=c(2,0,0.5,0) )
-plot( NULL, xlim=c(-2,5), ylim=c(0,0.5), 
+plot( NULL, xlim=c(-2,5), ylim=c(0,0.4), 
       xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=c(0, dd[1]), tick=F, 
-      label=c(0, expression( S[BA] ) ) )
+      label=c(0, expression( S[B] - S[A] ) ) )
 
 # figure parameters
 pplotD = c( dnorm( dd[1], mean=dd[1], sd=dd[2] ),
             dnorm( dd[1], mean=dd[1], sd=dd[2] ) + 0.05,
             dnorm( dd[1] + dd[2], mean=dd[1], sd=dd[2]),
-            dnorm( dd[1]+0.4, mean=dd[1], sd=dd[2] )-0.16,
[…]

### 2024-12-07 index.qmd
section:issues,CaseV; process:write

@@
 However, since the discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
-This law posits that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait.
-This emphasizes that the relative distance between stimuli, rather than their absolute positions on the continuum, determines the outcome of the pairwise comparison.
+This law states that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait.
+This law emphasizes that the relative distance between stimuli, rather than their absolute positions on the continuum, determines the outcome of the pairwise comparison.
 Indeed, the theory assumes that the observed dichotomous outcome arises from the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.
-Since the individual discriminal processes follow a Normal distribution, the discriminal difference also follows a Normal distribution [@Andrich_1978], with its mode (mean) representing the relative separation between the stimuli, while its dispersion reflects the variability in that relative separation.
+Since the individual discriminal processes follow a Normal distribution, the discriminal difference also follows a Normal distribution [@Andrich_1978].
+In this distribution, the mode (mean) represents the relative separation between the stimuli and the dispersion captures the variability of that relative separation.
 
@@
 <!-- 4. Example of discriminal difference -->
-@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process.
-The figure illustrates that the judge perceives Text B as having significantly higher quality than Text A, as indicated by their discriminal difference $(S_{B} - S_{A} > 0)$ and the shaded gray area under the curve $P(B > A)$.
-Therefore, the dichotomous outcome of this comparison would likely favor Text B over Text A.
+@fig-discriminal_difference illustrates the distribution of the discriminal difference for the texts presented in @fig-discriminal_process.
+The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
+This conclusion is evident from the positive difference in the modal discriminal processes of the texts $(S_{B} - S_{A} > 0)$ and the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$ (shaded gray area).
+As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
 
@@
 Case III introduces the assumption of zero correlation between stimuli.
-Case IV builds on this by assuming that the stimuli have similar dispersions.
+Case IV builds on this by assuming the stimuli have similar dispersions.
 Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions.
@@
 This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the discriminal processes of the stimuli, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
+The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
 254] (see @tbl-thurstone_cases).
@@
 677].
-Moreover, he warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
 270], acknowledging that some assumptions might be problematic when complex traits or heterogeneous stimuli are involved [@Thurstone_1927a, pp.
@@
 
-### The assumption of equal dispersions between stimuli
+### The assumption of equal dispersions between stimuli {#sec-theory-issue1a}
 
@@
 Discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome.
-For instance, @fig-dispersion illustrates how more uncertainty in the trait perception of a text over another broadens the distribution of their discriminal difference.
-This broadening affects the area under the curve, $P(B>A)$, determining the comparison outcome.
-Additionally, the figure shows that when there are no discrepancies in the texts' discriminal dispersions, meaning the *texts have equal dispersions* $(\sigma_{B}-\sigma_{A}=0)$, the likelihood that the discriminal difference distinctly favors one text over another increases.
+@fig-dispersion illustrates how more uncertainty in the trait perception of one text relative to another broadens the distribution of their discriminal difference.
+This broadening affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B > A)$.
+Additionally, the figure shows that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor one text over the other than when their dispersions differ (shaded gray area).
 
@@
 <!-- 2. But does it hold?  -->
-However, Thurstone contended that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376], as these traits and stimuli can introduce judgment discrepancies due to their unique features.
+However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
 Indeed, evidence of such violation may already exist in the CJ literature as misfit statistics.
@@
 20].
-For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than comparisons involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
-This finding implies that the discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts.
+For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than others [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
+This finding implies that discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts.
 Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
+Moreover, the presence of these "misfit" judges and their deviations can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue1b.
+
+<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
+[@Thurstone_1927a, pp.
+376] -->
 
@@
 
-<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
-[@Thurstone_1927a, pp.
-376] -->
+<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
+[@Pollitt_2012a, pp.
+165] -->
+
+<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
+
+<!-- a misfit statistic.
+Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
+It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
+12] -->
 
@@
 
-<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
-[@Pollitt_2012a, pp.
-165] -->
-
-<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
-
-<!-- a misfit statistic.
-Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
-It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
-12] -->
-
 <!-- 3. But what are the challenges when these assumptions are violated? -->
-<!-- Therefore, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to inflate the reliability of the observed dichotomous outcome  -->
-
-<!-- to overlook critical differences in the reliability of the trait across stimuli,  -->
-
-<!-- resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp.
+Therefore, assuming equal dispersions between stimuli, despite its violation, can cause Case V (and the BTL model) to inflate the reliability of the comparison outcome.
+Moreover, ignoring the difference in dispersions can lead to the neglect of critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp.
 370].
-Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the common practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
 The direction and magnitude of these biases are unpredictable, as they depend on the specific stimuli researchers exclude from the analysis.
-Taken together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
+Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
 2].
--->
 
@@
 
-### The assumption of zero correlation between stimuli
+### The assumption of zero correlation between stimuli {#sec-theory-issue1b}
 
 <!-- 1. What does the zero correlation between stimuli imply? -->
-Similar to the discriminal dispersions, the correlation between the discriminal processes is also crucial in determining the dichotomous outcomes of pairwise comparisons.
-This correlation shapes the distribution of the discriminal difference, directly influencing the comparison outcome.
-For instance, @fig-correlation illustrates how the stronger dependence of the trait perception in one text over another narrows the distribution of their discriminal difference.
-This narrowing impacts the area under the curve, $P(B>A)$, determining the comparison outcome.
-Additionally, the figure shows that when the trait perception in one text is independent of another, meaning the *texts have zero correlation*, the likelihood that the discriminal difference distinctly favors one text over another is lower compared to when the texts are highly correlated.
+Similar to the discriminal dispersions, the correlation between discriminal processes plays a crucial role in determining the dichotomous outcomes of pairwise comparisons.
+This correlation also shapes the distribution of the discriminal difference, directly influencing the comparison outcome.
+@fig-correlation illustrates how the dependence in trait perception between two texts narrows the distribution of their discriminal difference.
+This narrowing affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B>A)$.
+Moreover, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor one text over the other than when the texts are highly correlated (shaded gray area).
 
@@
 346].
-For example, this could occur when a judge assessing the argumentative quality of a text places more weight on grammatical accuracy than other judges, ultimately favoring texts with fewer errors but weaker arguments.
+For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, ultimately favoring texts with fewer errors but weaker arguments.
 While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
@@
 For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-This approach, however, has the significant limitation of ignoring the uncertainty associated with the BTL scores, which generates additional statistical and measurement issues, as discussed in section @sec-theory-issue2.
+This approach, however, can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue2.
 
 <!-- 5. But what are the challenges when this assumption is violated? -->
-<!-- In any case, the psychometric and statistical literature emphasizes the importance of addressing the factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity.
-For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the statistical model used for analysis.
-This mismatch can potentially inflate the reliability of the trait [@Hoyle_et_al_2023, pp.
-341] or, worse, introduce bias into the trait's estimates [@Ackerman_1989].
-Similarly, failing to account for hierarchical (grouping) structures reduces the precision of model parameter estimates, further amplifying the overestimation of reliability [@Hoyle_et_al_2023, pp.
+In any case, the psychometric and statistical literature emphasizes the need to address factors that create dependencies between stimuli, as failing to do so can affect the reliability of the comparison outcomes and lead to inaccurate conclusions about the trait's estimates.
+For instance, researchers who overlook additional relevant traits, such as judges' biases, can cause dimensional mismatches in the statistical model used for analysis.
+This mismatch can artificially inflate the reliability of the trait [@Hoyle_et_al_2023, pp.
+341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989].
+Furthermore, as discussed in @sec-theory-issue1a, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+12].
+Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of reliability [@Hoyle_et_al_2023, pp.
 482].
-These issues collectively also undermine the trait's reliability and ultimately compromise the validity of the trait's estimates [@Perron_et_al_2015, pp.
+Together, these issues undermine the reliability of the trait and compromise the validity of its estimates [@Perron_et_al_2015, pp.
 2].
--->
 
@@
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is evident that, besides its inconveniences, the BTL model commonly functions as the trait's measurement model in CJ assessments.
+Building on the previous section, it is clear that, despite its limitations, the BTL model is commonly used as the measurement model in CJ assessments.
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
-For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 

### 2024-12-07 code/figures.R

@@ -81,27 +81,23 @@ axis( side=1, at=c(0, dd[1]), tick=F,
       label=c(0, expression( S[B] - S[A] ) ) )
 
 # figure parameters
-pplotD = c( dnorm( dd[1], mean=dd[1], sd=dd[2] ),
+pplotD = c( dnorm( 0, mean=dd[1], sd=dd[2] ),
+            dnorm( dd[1], mean=dd[1], sd=dd[2] ),
+            dnorm( dd[1]+dd[2], mean=dd[1], sd=dd[2]),
             dnorm( dd[1], mean=dd[1], sd=dd[2] ) + 0.05,
-            dnorm( dd[1] + dd[2], mean=dd[1], sd=dd[2]),
-            dnorm( dd[1]+0.4, mean=dd[1], sd=dd[2] )-0.14,
-            dnorm( 0, mean=dd[1], sd=dd[2] ) )
+            dnorm( dd[1]+0.4, mean=dd[1], sd=dd[2] )-0.14 )
 
 # discriminal difference
-colorArea( from=0, to=7, density=dnorm, mean=dd[1], sd=dd[2],col=rgb(0,0,0,0.05) )
+colorArea( from=0, to=7, density=dnorm, mean=dd[1], sd=dd[2], col=rgb(0,0,0,0.05) )
 curve( dnorm(x, mean=dd[1], sd=dd[2]), lwd=1.5, add=T ) 
-# text( x=dd[1], y=pplotD[2], cex=0.8, 'Text B' ) 
-lines( x=c(dd[1], dd[1]), y=c(pplotD[1], -0.1), lty=2, lwd=0.7 )
+lines( x=c(0,0), y=c(pplotD[1],-0.1), lty=2, lwd=0.7 )
+text( x=0.45, y=0.03, cex=0.8, expression( P(B>A) ) ) 
+lines( x=c(dd[1], dd[1]), y=c(pplotD[2], -0.1), lty=2, lwd=0.7 )
 lines( x=c(dd[1], dd[1]+dd[2]), y=rep( pplotD[3], 2), lty=2, lwd=0.7 )
-text( x=dd[1]+0.5, y=pplotD[4], cex=0.8, expression( sigma[AB] ) ) 
-# text( x=dd[1]+1.65, y=0.32, cex=0.8, 'where:' )
-# text( x=dd[1]+2, y=pplotD[4], cex=0.8,
[…]

### 2024-12-07 _tex/index.tex
Built site for gh-pages

@@
 observable, the theory introduces the \emph{law of comparative
-judgment}. This law posits that in pairwise comparisons, a judge
+judgment}. This law states that in pairwise comparisons, a judge
 perceives the stimulus positioned further along the trait continuum as
-having a higher level of that trait. This emphasizes that the relative
-distance between stimuli, rather than their absolute positions on the
-continuum, determines the outcome of the pairwise comparison. Indeed,
-the theory assumes that the observed dichotomous outcome arises from the
-difference between the underlying discriminal processes of the stimuli,
-referred to as the \emph{discriminal difference}. Since the individual
-discriminal processes follow a Normal distribution, the discriminal
-difference also follows a Normal distribution \citep{Andrich_1978}, with
-its mode (mean) representing the relative separation between the
-stimuli, while its dispersion reflects the variability in that relative
-separation.
+having a higher level of that trait. This law emphasizes that the
+relative distance between stimuli, rather than their absolute positions
+on the continuum, determines the outcome of the pairwise comparison.
+Indeed, the theory assumes that the observed dichotomous outcome arises
+from the difference between the underlying discriminal processes of the
+stimuli, referred to as the \emph{discriminal difference}. Since the
+individual discriminal processes follow a Normal distribution, the
+discriminal difference also follows a Normal distribution
+\citep{Andrich_1978}. In this distribution, the mode (mean) represents
+the relative separation between the stimuli and the dispersion captures
+the variability of that relative separation.
 
-Figure~\ref{fig-discriminal_difference} shows the distribution of the
-discriminal difference for the texts depicted in
-Figure~\ref{fig-discriminal_process}. The figure illustrates that the
+Figure~\ref{fig-discriminal_difference} illustrates the distribution of
+the discriminal difference for the texts presented in
+Figure~\ref{fig-discriminal_process}. The figure indicates that the
 judge perceives Text B as having significantly higher quality than Text
-A, as indicated by their discriminal difference \((S_{B} - S_{A} > 0)\)
-and the shaded gray area under the curve \(P(B > A)\). Therefore, the
-dichotomous outcome of this comparison would likely favor Text B over
-Text A.
+A. This conclusion is evident from the positive difference in the modal
+discriminal processes of the texts \((S_{B} - S_{A} > 0)\) and the area
+under the curve where the discriminal difference distinctly favors Text
+B over Text A, denoted as \(P(B > A)\) (shaded gray area). As a result,
+the dichotomous outcome of this comparison is more likely to favor Text
+B over Text A.
 
@@
 the comparisons to a single judge. Case III introduces the assumption of
-zero correlation between stimuli. Case IV builds on this by assuming
-that the stimuli have similar dispersions. Finally, Case V replaces this
+zero correlation between stimuli. Case IV builds on this by assuming the
+stimuli have similar dispersions. Finally, Case V replaces this
 assumption with the condition that the stimuli have equal discriminal
@@
 assumptions of Case V, with one key difference: while Case V assumes a
-Normal distribution for the discriminal processes of the stimuli, the
-BTL model uses the more mathematically tractable Logistic distribution
+Normal distribution for the stimuli's discriminal processes, the BTL
+model uses the more mathematically tractable Logistic distribution
 \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
@@
 prioritizing statistical simplicity over precision in trait measurement
-\citep[pp.~677]{Kelly_et_al_2022}. Moreover, he warned against its
+\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
 untested application, stating that its use ``should not be made without
@@
 \subsubsection{The assumption of equal dispersions between
-stimuli}\label{the-assumption-of-equal-dispersions-between-stimuli}
+stimuli}\label{sec-theory-issue1a}
 
@@
 dispersions shape the distribution of the discriminal difference,
-directly influencing the comparison outcome. For instance,
-Figure~\ref{fig-dispersion} illustrates how more uncertainty in the
-trait perception of a text over another broadens the distribution of
-their discriminal difference. This broadening affects the area under the
-curve, \(P(B>A)\), determining the comparison outcome. Additionally, the
-figure shows that when there are no discrepancies in the texts'
-discriminal dispersions, meaning the \emph{texts have equal dispersions}
-\((\sigma_{B}-\sigma_{A}=0)\), the likelihood that the discriminal
-difference distinctly favors one text over another increases.
+directly influencing the comparison outcome. Figure~\ref{fig-dispersion}
+illustrates how more uncertainty in the trait perception of one text
+relative to another broadens the distribution of their discriminal
+difference. This broadening affects the area under the curve where the
+discriminal difference distinctly favors one text over the other,
+\(P(B > A)\). Additionally, the figure shows that when the discriminal
+dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
+discriminal difference is more likely to favor one text over the other
+than when their dispersions differ (shaded gray area).
 
@@
 
-However, Thurstone contended that this assumption may not hold when
-researchers assess complex traits or heterogeneous stimuli
-\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
-introduce judgment discrepancies due to their unique features. Indeed,
-evidence of such violation may already exist in the CJ literature as
-misfit statistics. These statistics measure the judgment discrepancies
-associated with a given stimulus
+However, Thurstone contended that the assumption of equal dispersions
+may not hold when researchers assess complex traits or heterogeneous
+stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
+can introduce judgment discrepancies due to their unique features
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
+Indeed, evidence of such violation may already exist in the CJ
+literature as misfit statistics. These statistics measure the judgment
+discrepancies associated with a given stimulus
 \citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
 For instance, labeling texts as ``misfits'' indicates that comparisons
-involving these texts result in more judgment discrepancies than
-comparisons involving other texts
+involving these texts result in more judgment discrepancies than others
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
-This finding implies that the discriminal differences associated with
+This finding implies that discriminal differences associated with
 ``misfits'' texts usually display a broader dispersion, suggesting that
@@
 to the unique characteristics of the stimuli or the judges themselves.
+Moreover, the presence of these ``misfit'' judges and their deviations
+can introduce additional statistical and measurement issues, as
+discussed in Section~\ref{sec-theory-issue1b}.
+
+Therefore, assuming equal dispersions between stimuli, despite its
+violation, can cause Case V (and the BTL model) to inflate the
+reliability of the comparison outcome. Moreover, ignoring the difference
+in dispersions can lead to the neglect of critical differences in the
+reliability of the trait across stimuli, resulting in inaccurate
+conclusions about the trait's estimates \citep[pp.~370]{McElreath_2020}.
+Furthermore, if researchers acknowledge that misfit statistics help
+identify these critical differences, the usual practice in the CJ
+literature of excluding stimuli based on these statistics
+\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
+discarding valuable information and introducing bias into the trait's
+estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
+direction and magnitude of these biases are unpredictable, as they
+depend on the specific stimuli researchers exclude from the analysis.
+Together, these oversights undermine the reliability of the trait and
+ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.
 
 \subsubsection{The assumption of zero correlation between
-stimuli}\label{the-assumption-of-zero-correlation-between-stimuli}
+stimuli}\label{sec-theory-issue1b}
 
-Similar to the discriminal dispersions, the correlation between the
-discriminal processes is also crucial in determining the dichotomous
-outcomes of pairwise comparisons. This correlation shapes the
-distribution of the discriminal difference, directly influencing the
-comparison outcome. For instance, Figure~\ref{fig-correlation}
-illustrates how the stronger dependence of the trait perception in one
-text over another narrows the distribution of their discriminal
-difference. This narrowing impacts the area under the curve, \(P(B>A)\),
-determining the comparison outcome. Additionally, the figure shows that
-when the trait perception in one text is independent of another, meaning
-the \emph{texts have zero correlation}, the likelihood that the
-discriminal difference distinctly favors one text over another is lower
-compared to when the texts are highly correlated.
+Similar to the discriminal dispersions, the correlation between
+discriminal processes plays a crucial role in determining the
+dichotomous outcomes of pairwise comparisons. This correlation also
+shapes the distribution of the discriminal difference, directly
+influencing the comparison outcome. Figure~\ref{fig-correlation}
+illustrates how the dependence in trait perception between two texts
+narrows the distribution of their discriminal difference. This narrowing
+affects the area under the curve where the discriminal difference
+distinctly favors one text over the other, \(P(B>A)\). Moreover, the
+figure shows that when two texts are independent or uncorrelated
+\((\rho=0)\), their discriminal difference is less likely to favor one
+text over the other than when the texts are highly correlated (shaded
+gray area).
 
@@
 occur when a judge assessing the argumentative quality of a text places
-more weight on grammatical accuracy than other judges, ultimately
+more weight on its grammatical accuracy than other judges, ultimately
 favoring texts with fewer errors but weaker arguments. While direct
@@
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-This approach, however, has the significant limitation of ignoring the
-uncertainty associated with the BTL scores, which generates additional
-statistical and measurement issues, as discussed in section
-Section~\ref{sec-theory-issue2}.
+This approach, however, can introduce additional statistical and
+measurement issues, as discussed in Section~\ref{sec-theory-issue2}.
+
+In any case, the psychometric and statistical literature emphasizes the
+need to address factors that create dependencies between stimuli, as
+failing to do so can affect the reliability of the comparison outcomes
+and lead to inaccurate conclusions about the trait's estimates. For
+instance, researchers who overlook additional relevant traits, such as
+judges' biases, can cause dimensional mismatches in the statistical
+model used for analysis. This mismatch can artificially inflate the
+reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
+worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
+Furthermore, as discussed in Section~\ref{sec-theory-issue1a},
+researchers who exclude judges based on misfit statistics can risk
+discarding valuable information, further biasing the trait's estimates
+\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly, researchers
+who fail to account for hierarchical (grouping) structures can reduce
+the precision of model parameter estimates, which may amplify the
+overestimation of reliability \citep[pp.~482]{Hoyle_et_al_2023}.
+Together, these issues undermine the reliability of the trait and
+compromise the validity of its estimates
+\citep[pp.~2]{Perron_et_al_2015}.
 
@@
 
-Building on the previous section, it is evident that, besides its
-inconveniences, the BTL model commonly functions as the trait's
-measurement model in CJ assessments. A measurement model specifies how
-manifest variables contribute to the estimation of latent variables
+Building on the previous section, it is clear that, despite its
+limitations, the BTL model is commonly used as the measurement model in
+CJ assessments. A measurement model specifies how manifest variables
+contribute to the estimation of latent variables
 \citep{Everitt_et_al_2010}. For example, when evaluating text quality,
@@
 resulting from the pairwise comparisons (the manifest variables) to
-estimate scores that reflect the underlying quality level of texts (the
-latent variable)
+estimate scores that reflect the underlying quality level of the texts
+(the latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

### 2024-12-07 index.qmd
section:issues,CaseV; process:write

@@
 
-![The effect of different dispersions between stimuli on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
+![Discrepancies in the stimuli dispersions and tehir effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
 
@@
 
-![The effect of correlation on the distribution of the discriminal difference](/images/figures/correlation.png){#fig-correlation width=70%}
+![The correlation between stimuli and its effect on the distribution of the discriminal difference](/images/figures/correlation.png){#fig-correlation width=70%}
 

### 2024-12-07 _tex/index.tex
Built site for gh-pages

@@
 
-\caption{\label{fig-dispersion}The effect of different dispersions
-between stimuli on the distribution of the discriminal difference}
+\caption{\label{fig-dispersion}Discrepancies in the stimuli dispersions
+and tehir effect on the distribution of the discriminal difference}
 
@@
 
-\caption{\label{fig-correlation}The effect of correlation on the
-distribution of the discriminal difference}
+\caption{\label{fig-correlation}The correlation between stimuli and its
+effect on the distribution of the discriminal difference}
 

### 2024-12-07 index.qmd
section:issues,CaseV; process:write

@@
 
-![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}
+![Thurstones cases and their asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}
 
@@
 
-![Discrepancies in the stimuli dispersions and tehir effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
+![The Discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
 

### 2024-12-07 _tex/index.tex
Built site for gh-pages

@@
 
-\caption{\label{tbl-thurstone_cases}Thurstones cases and asumptions}
+\caption{\label{tbl-thurstone_cases}Thurstones cases and their
+asumptions}
 
@@
 
-\caption{\label{fig-dispersion}Discrepancies in the stimuli dispersions
-and tehir effect on the distribution of the discriminal difference}
+\caption{\label{fig-dispersion}The Discrepancy in the dispersions of
+stimuli and their effect on the distribution of the discriminal
+difference}
 

### 2024-12-09 index.qmd
section:issues; process:write

@@
 @fig-discriminal_process illustrates the discriminal processes along a quality trait continuum for two written texts.
-The figure shows that these processes individually follow a Normal distribution.
-Furthermore, it shows that the modal discriminal process for Text B is positioned further along the continuum than Text A $(S_{B} > S_{A})$.
-Additionally, the figure highlights the broader distribution of Text B compared to Text A as a result of its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
+The figure shows that each discriminal process follows a Normal distribution.
+It also indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$, signifying a higher quality for Text B.
+Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
+
+<!-- 1. Thurstone's Theory of CJ: discriminal process -->
+<!-- In its most general form, Thurstone’s theory addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
+267].
+The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
+The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their perception of the stimulus trait.
+For instance, according to the theory, the discriminal process for a text along a quality trait continuum follows a Normal distribution, as depicted in @fig-discriminal_process.
+In this distribution, the mode or mean, termed the *modal discriminal process*, represents the stimulus's position on the trait continuum.
+@fig-discriminal_process show the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$.
+Additionally, the dispersion of this distribution, called the *discriminal dispersion*, captures variability in the perceived trait of the stimulus.
+@fig-discriminal_process highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
+-->
 
@@
 The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
-This conclusion is evident from the positive difference in the modal discriminal processes of the texts $(S_{B} - S_{A} > 0)$ and the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$ (shaded gray area).
+This conclusion is evident from the positive difference in their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$ (shaded gray area).
 As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
@@
 <!-- 1. Thurstone's Cases and their simplifying assumptions  -->
-The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli.
-For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions.
+Since the general form of Thurstone's theory, outlined in @sec-thurstone_theory, applies to a CJ design where a single judge evaluates multiple stimuli, Thurstone developed four additional cases for the theory's practical application, each progressively incorporating additional simplifying assumptions.
 Case I represents the general form of the theory.
@@
 677].
-He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions might be problematic when complex traits or heterogeneous stimuli are involved [@Thurstone_1927a, pp.
+Thurstone explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+270], acknowledging that some assumptions could be problematic when researchers asesss complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376].
@@
 The discriminal dispersions of stimuli are crucial in determining the dichotomous outcomes of pairwise comparisons.
-Discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome.
-@fig-dispersion illustrates how more uncertainty in the trait perception of one text relative to another broadens the distribution of their discriminal difference.
+Holding all other modeling factors constant, discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome.
+@fig-dispersion illustrates how more uncertainty in the trait perception of one text relative to another, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
 This broadening affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B > A)$.
@@
 
-![The Discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
+![The discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
 
 <!-- 2. But does it hold?  -->
-However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+However, Thurstone contended that this assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
@@
 For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than others [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
-This finding implies that discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts.
+This finding implies that discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that the discriminal processes of these texts also exhibit more variation than other texts.
 Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
-Moreover, the presence of these "misfit" judges and their deviations can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue1b.
+Moreover, the presence of these "misfit" judges and their deviations can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue1b.
 
@@
 <!-- 3. But what are the challenges when these assumptions are violated? -->
-Therefore, assuming equal dispersions between stimuli, despite its violation, can cause Case V (and the BTL model) to inflate the reliability of the comparison outcome.
-Moreover, ignoring the difference in dispersions can lead to the neglect of critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp.
+Therefore, assuming equal dispersions between stimuli, despite its violation, can cause Case V (and the BTL model) to inflate the reliability of the outcome, resulting in inaccurate conclusions about the comparison.
+Moreover, ignoring the disparity in dispersions can lead to the neglect of critical differences in the reliability of the trait across these stimuli, leading to erroneous conclusions about the trait's estimates [@McElreath_2020, pp.
 370].
-Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Additionally, if researchers acknowledge that misfit statistics help identify these critical differences in dispersion, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
-The direction and magnitude of these biases are unpredictable, as they depend on the specific stimuli researchers exclude from the analysis.
-Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
-2].
+The direction and magnitude of these biases are often unpredictable, as they depend on the specific stimuli researchers exclude from the analysis.
 
@@
 Similar to the discriminal dispersions, the correlation between discriminal processes plays a crucial role in determining the dichotomous outcomes of pairwise comparisons.
-This correlation also shapes the distribution of the discriminal difference, directly influencing the comparison outcome.
+Holding all other modeling factors constant, this correlation shapes the distribution of the discriminal difference, directly influencing the comparison outcome.
 @fig-correlation illustrates how the dependence in trait perception between two texts narrows the distribution of their discriminal difference.
@@
 For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-This approach, however, can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue2.
+However, this approach can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue2.
 
@@
 12].
-Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of reliability [@Hoyle_et_al_2023, pp.
+Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp.
 482].
-Together, these issues undermine the reliability of the trait and compromise the validity of its estimates [@Perron_et_al_2015, pp.
-2].
 
@@
 
+In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, along with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
+2].
+Consequently, adopting a more systematic and integrated approach to examining what happens when judges compare two stimuli could offer several statistical and measurement benefits, including addressing these issues.
+
 <!-- ## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3} -->
@@
 
-<!-- []{style="color:red;"} -->
+This section presents a theoretical model for CJ that extends Thurstone's theory.
+The model systematically incorporates all factors involved when judges make pairwise comparisons.
+Additionally, the section develops the statistical translation of the theoretical model based on assumptions informed by the CJ theory.
 
-<!-- 3. What is the solution? -->
-
-<!-- Notes:  -->
-<!-- - There is some gain on considering a more integrated overarching systematic way of looking on what happen in DCJ when people compare two stimuli -->
-<!-- - This paper tries to systematically integrate all aspects at play in a single theoretical model, we will build the model in a stepwise manner using -->
-<!-- - we will build a scientific model (more theoretical model) that integrates different aspects of the method that are at play when people use DCJ. -->
+<!-- in a stepwise manner -->
 
 ## The theoretical model {#sec-theory-theoretical}
+
+<!-- []{style="color:red;"} -->
 

### 2024-12-09 _tex/index.tex
Built site for gh-pages

@@
 processes along a quality trait continuum for two written texts. The
-figure shows that these processes individually follow a Normal
-distribution. Furthermore, it shows that the modal discriminal process
-for Text B is positioned further along the continuum than Text A
-\((S_{B} > S_{A})\). Additionally, the figure highlights the broader
-distribution of Text B compared to Text A as a result of its larger
+figure shows that each discriminal process follows a Normal
+distribution. It also indicates that the modal discriminal process for
+Text B is positioned further along the continuum than that of Text A
+\((S_{B} > S_{A})\), signifying a higher quality for Text B.
+Additionally, the figure highlights that Text B has a broader
+distribution compared to Text A, which arises from its larger
 discriminal dispersion \((\sigma_{B} > \sigma_{A})\).
@@
 judge perceives Text B as having significantly higher quality than Text
-A. This conclusion is evident from the positive difference in the modal
-discriminal processes of the texts \((S_{B} - S_{A} > 0)\) and the area
-under the curve where the discriminal difference distinctly favors Text
-B over Text A, denoted as \(P(B > A)\) (shaded gray area). As a result,
-the dichotomous outcome of this comparison is more likely to favor Text
-B over Text A.
+A. This conclusion is evident from the positive difference in their
+modal discriminal processes \((S_{B} - S_{A} > 0)\) and the area under
+the curve where the discriminal difference distinctly favors Text B over
+Text A, denoted as \(P(B > A)\) (shaded gray area). As a result, the
+dichotomous outcome of this comparison is more likely to favor Text B
+over Text A.
 
@@
 
-The previous section outlines the general form of Thurstone's theory,
-which applies to a CJ design where a single judge evaluates multiple
-stimuli. For the practical application of the theory, Thurstone
-developed four additional cases derived from this general form, where
-each successive case incorporates additional simplifying assumptions.
-Case I represents the general form of the theory. Case II extends this
-by allowing multiple judges to make comparisons rather than restricting
-the comparisons to a single judge. Case III introduces the assumption of
+Since the general form of Thurstone's theory, outlined in
+Section~\ref{sec-thurstone_theory}, applies to a CJ design where a
+single judge evaluates multiple stimuli, Thurstone developed four
+additional cases for the theory's practical application, each
+progressively incorporating additional simplifying assumptions. Case I
+represents the general form of the theory. Case II extends this by
+allowing multiple judges to make comparisons rather than restricting the
+comparisons to a single judge. Case III introduces the assumption of
 zero correlation between stimuli. Case IV builds on this by assuming the
@@
 prioritizing statistical simplicity over precision in trait measurement
-\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
-untested application, stating that its use ``should not be made without
-(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
-that some assumptions might be problematic when complex traits or
-heterogeneous stimuli are involved \citep[pp.~376]{Thurstone_1927a}.
-Consequently, given that modern CJ applications frequently involve such
-traits and stimuli, two main assumptions of Case V and, by extension, of
-the BTL model may not consistently hold in theory or practice: the
-assumption of equal dispersion and zero correlation between stimuli.
+\citep[pp.~677]{Kelly_et_al_2022}. Thurstone explicitly warned against
+its untested application, stating that its use ``should not be made
+without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b},
+acknowledging that some assumptions could be problematic when
+researchers asesss complex traits or heterogeneous stimuli
+\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
+applications frequently involve such traits and stimuli, two main
+assumptions of Case V and, by extension, of the BTL model may not
+consistently hold in theory or practice: the assumption of equal
+dispersion and zero correlation between stimuli.
 
@@
 The discriminal dispersions of stimuli are crucial in determining the
-dichotomous outcomes of pairwise comparisons. Discrepancies in these
-dispersions shape the distribution of the discriminal difference,
-directly influencing the comparison outcome. Figure~\ref{fig-dispersion}
-illustrates how more uncertainty in the trait perception of one text
-relative to another broadens the distribution of their discriminal
-difference. This broadening affects the area under the curve where the
-discriminal difference distinctly favors one text over the other,
-\(P(B > A)\). Additionally, the figure shows that when the discriminal
-dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
-discriminal difference is more likely to favor one text over the other
-than when their dispersions differ (shaded gray area).
+dichotomous outcomes of pairwise comparisons. Holding all other modeling
+factors constant, discrepancies in these dispersions shape the
+distribution of the discriminal difference, directly influencing the
+comparison outcome. Figure~\ref{fig-dispersion} illustrates how more
+uncertainty in the trait perception of one text relative to another,
+\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
+discriminal difference. This broadening affects the area under the curve
+where the discriminal difference distinctly favors one text over the
+other, \(P(B > A)\). Additionally, the figure shows that when the
+discriminal dispersions of the texts are equal
+\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
+to favor one text over the other than when their dispersions differ
+(shaded gray area).
 
@@
 
-\caption{\label{fig-dispersion}The Discrepancy in the dispersions of
+\caption{\label{fig-dispersion}The discrepancy in the dispersions of
 stimuli and their effect on the distribution of the discriminal
@@
 
-However, Thurstone contended that the assumption of equal dispersions
+However, Thurstone contended that this assumption of equal dispersions
 may not hold when researchers assess complex traits or heterogeneous
@@
 ``misfits'' texts usually display a broader dispersion, suggesting that
-their discriminal processes also exhibit more variation than other
-texts. Notably, this reasoning also applies to ``misfit'' judges, whose
-evaluations reflect substantial deviations from the shared consensus due
-to the unique characteristics of the stimuli or the judges themselves.
-Moreover, the presence of these ``misfit'' judges and their deviations
-can introduce additional statistical and measurement issues, as
-discussed in Section~\ref{sec-theory-issue1b}.
+the discriminal processes of these texts also exhibit more variation
+than other texts. Notably, this reasoning also applies to ``misfit''
+judges, whose evaluations reflect substantial deviations from the shared
+consensus due to the unique characteristics of the stimuli or the judges
+themselves. Moreover, the presence of these ``misfit'' judges and their
+deviations can introduce additional statistical and measurement issues,
+which we discuss in Section~\ref{sec-theory-issue1b}.
 
@@
 violation, can cause Case V (and the BTL model) to inflate the
-reliability of the comparison outcome. Moreover, ignoring the difference
-in dispersions can lead to the neglect of critical differences in the
-reliability of the trait across stimuli, resulting in inaccurate
-conclusions about the trait's estimates \citep[pp.~370]{McElreath_2020}.
-Furthermore, if researchers acknowledge that misfit statistics help
-identify these critical differences, the usual practice in the CJ
-literature of excluding stimuli based on these statistics
+reliability of the outcome, resulting in inaccurate conclusions about
+the comparison. Moreover, ignoring the disparity in dispersions can lead
+to the neglect of critical differences in the reliability of the trait
+across these stimuli, leading to erroneous conclusions about the trait's
+estimates \citep[pp.~370]{McElreath_2020}. Additionally, if researchers
+acknowledge that misfit statistics help identify these critical
+differences in dispersion, the usual practice in the CJ literature of
+excluding stimuli based on these statistics
 \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
@@
 estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
-direction and magnitude of these biases are unpredictable, as they
+direction and magnitude of these biases are often unpredictable, as they
 depend on the specific stimuli researchers exclude from the analysis.
-Together, these oversights undermine the reliability of the trait and
-ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.
 
@@
 discriminal processes plays a crucial role in determining the
-dichotomous outcomes of pairwise comparisons. This correlation also
-shapes the distribution of the discriminal difference, directly
-influencing the comparison outcome. Figure~\ref{fig-correlation}
-illustrates how the dependence in trait perception between two texts
-narrows the distribution of their discriminal difference. This narrowing
-affects the area under the curve where the discriminal difference
-distinctly favors one text over the other, \(P(B>A)\). Moreover, the
-figure shows that when two texts are independent or uncorrelated
-\((\rho=0)\), their discriminal difference is less likely to favor one
-text over the other than when the texts are highly correlated (shaded
-gray area).
+dichotomous outcomes of pairwise comparisons. Holding all other modeling
+factors constant, this correlation shapes the distribution of the
+discriminal difference, directly influencing the comparison outcome.
+Figure~\ref{fig-correlation} illustrates how the dependence in trait
+perception between two texts narrows the distribution of their
+discriminal difference. This narrowing affects the area under the curve
+where the discriminal difference distinctly favors one text over the
+other, \(P(B>A)\). Moreover, the figure shows that when two texts are
+independent or uncorrelated \((\rho=0)\), their discriminal difference
+is less likely to favor one text over the other than when the texts are
+highly correlated (shaded gray area).
 
@@
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-This approach, however, can introduce additional statistical and
-measurement issues, as discussed in Section~\ref{sec-theory-issue2}.
+However, this approach can introduce additional statistical and
+measurement issues, which we discuss in Section~\ref{sec-theory-issue2}.
 
@@
 the precision of model parameter estimates, which may amplify the
-overestimation of reliability \citep[pp.~482]{Hoyle_et_al_2023}.
-Together, these issues undermine the reliability of the trait and
-compromise the validity of its estimates
-\citep[pp.~2]{Perron_et_al_2015}.
+overestimation of the trait's reliability
+\citep[pp.~482]{Hoyle_et_al_2023}.
 
@@
 
+In aggregate, researchers' inadequate handling of violations to the
+assumptions of equal dispersion and zero correlation between stimuli,
+along with the apparent disconnect between CJ's approach to trait
+measurement and hypothesis testing, can undermine the reliability of the
+trait and ultimately compromise its validity
+\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
+systematic and integrated approach to examining what happens when judges
+compare two stimuli could offer several statistical and measurement
+benefits, including addressing these issues.
+
 \section{An updated theoretical and statistical model for
 CJ}\label{sec-theory}
+
+This section presents a theoretical model for CJ that extends
+Thurstone's theory. The model systematically incorporates all factors
+involved when judges make pairwise comparisons. Additionally, the
+section develops the statistical translation of the theoretical model
+based on assumptions informed by the CJ theory.
 

### 2024-12-09 code/figures.R

@@ -163,14 +163,16 @@ plot( NULL, xlim=c(-2,5), ylim=c(0,0.6),
       xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=0, tick=F, label='0' )
+axis( side=1, at=c(0, dd[1]), tick=F, 
+      label=c(0, expression( S[B] - S[A] ) ) )
 
 # figure parameters
 l_ty = c( 1, rep(2,2), rep(3,2) )
 l_wd = c( 1.5, seq(1,0.5,length.out=4) )
-text = paste0( ' ~ sigma[B] - sigma[A] == ', dpB_d - dpA )
+text = paste0( ' ~ sigma[B] - sigma[A] == ', dpB_d - dpA[2] )
 pplotD = c( dnorm( 0, mean=dd[[1]], sd=dd[[2]][1] ),
             dnorm( dd[[1]], mean=dd[[1]], sd=dd[[2]][1] ),
-            dnorm( dd[[1]] + 2*dd[[2]][1], mean=dd[[1]], sd=dd[[2]][1] ),
+            dnorm( dd[[1]]+2*dd[[2]][1], mean=dd[[1]], sd=dd[[2]][1] ),
             dnorm( dd[[1]]+0.4, mean=dd[[1]], sd=dd[[2]][1] )-0.14 )
 
 
@@ -182,9 +184,17 @@ for(i in 1:length(dd[[2]])){
 }
 lines( x=c(0,0), y=c(pplotD[1],-0.1), lty=2, lwd=0.7 )
 text( x=0.45, y=0.03, cex=0.8, expression( P(B>A) ) ) 
+lines( x=c(dd[[1]], dd[[1]]), y=c(pplotD[2], -0.1), lty=2, lwd=0.7 )
+lines( x=c(dd[[1]], dd[[1]]+2*dd[[2]][1]), y=rep(pplotD[3], 2), lty=2, lwd=0.7 )
+text( x=dd[[1]]+0.65, y=0.05, cex=0.8, expression( sigma[BA] ) ) 
+text( x=dd[[1]]+1.7, y=0.26, cex=0.8, 'where:' )
+text( x=dd[[1]]+2.45, y=0.22, cex=0.8,
+      expression( sigma[BA] == sqrt( sigma[B]^2 + sigma[A]^2 - 2*rho*sigma[B]*sigma[A]) ) )
[…]

### 2024-12-09 code/figures.R
section:issues; process:write

@@ -83,7 +83,7 @@ axis( side=1, at=c(0, dd[1]), tick=F,
 # figure parameters
 pplotD = c( dnorm( 0, mean=dd[1], sd=dd[2] ),
             dnorm( dd[1], mean=dd[1], sd=dd[2] ),
-            dnorm( dd[1]+dd[2], mean=dd[1], sd=dd[2]),
+            dnorm( dd[1]+2*dd[2], mean=dd[1], sd=dd[2]),
             dnorm( dd[1], mean=dd[1], sd=dd[2] ) + 0.05,
             dnorm( dd[1]+0.4, mean=dd[1], sd=dd[2] )-0.14 )
 
@@ -93,10 +93,10 @@ curve( dnorm(x, mean=dd[1], sd=dd[2]), lwd=1.5, add=T )
 lines( x=c(0,0), y=c(pplotD[1],-0.1), lty=2, lwd=0.7 )
 text( x=0.45, y=0.03, cex=0.8, expression( P(B>A) ) ) 
 lines( x=c(dd[1], dd[1]), y=c(pplotD[2], -0.1), lty=2, lwd=0.7 )
-lines( x=c(dd[1], dd[1]+dd[2]), y=rep( pplotD[3], 2), lty=2, lwd=0.7 )
-text( x=dd[1]+0.6, y=pplotD[3]-0.02, cex=0.8, expression( sigma[AB] ) ) 
+lines( x=c(dd[1], dd[1]+2*dd[2]), y=rep( pplotD[3], 2), lty=2, lwd=0.7 )
+text( x=dd[1]+1, y=pplotD[3]-0.02, cex=0.8, expression( sigma[BA] ) ) 
 text( x=dd[1]+1.7, y=0.25, cex=0.8, 'where:' )
-text( x=dd[[1]]+2.4, y=0.22, cex=0.8,
+text( x=dd[1]+2.4, y=0.22, cex=0.8,
       expression( sigma[BA] == sqrt( sigma[B]^2 + sigma[A]^2 - rho*sigma[B]*sigma[A]) ) )
 
 dev.off()

### 2024-12-13 index.qmd
section:issues; process:write

--- 2024-12-07 index.qmd
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:30:35 - 00:35:10 -->
+<!-- recording: Sven and Tine 24.12.09; time: 00:00:00 - 00:03:20 -->
 
 <!-- 1. Thurstone's Theory of CJ: discriminal process -->
-In its most general form, Thurstone’s theory addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp.
+In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp.
 267].
 The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
-The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their perception of the stimulus trait.
-According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), called the *modal discriminal process*, indicates the stimulus position on the trait continuum, while its dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.
+The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait.
+The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp.
+266].
+The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.
+
+<!-- The psychological continuum or scale is so constructed or defined that the frequencies of the respective discriminal processes for any given stimulus form a normal distribution on the psychological scale.
+This involves no assumption of a normal distribution or of anything else.
+The psychological scale is at best an artificial construct.
+If it has any physical reality, we certainly have not the remotest idea of what it may be like.
+...
+We define the scale in terms of the frequencies of the discriminal processes for any stimulus.
+This artificial construct, the psychological scale, is so spaced off that the frequencies of the discriminal processes for any given stimulus form a normal distribution on the scale.[@Thurstone_1927b, pp.
+266] -->
 
 <!-- 2. Example of discriminal processes -->
-@fig-discriminal_process illustrates the discriminal processes along a quality trait continuum for two written texts.
-The figure shows that these processes individually follow a Normal distribution.
-Furthermore, it shows that the modal discriminal process for Text B is positioned further along the continuum than Text A $(S_{B} > S_{A})$.
-Additionally, the figure highlights the broader distribution of Text B compared to Text A as a result of its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
+@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts.
+The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$, suggesting that Text B exhibits higher quality.
+Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
 
-![Discriminal processes for two written texts](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}
+![Discriminal processes for two written texts](/images/png/discriminal_process.png){#fig-discriminal_process width=70%}
 
 <!-- 3. Thurstone's Theory of CJ: discriminal difference -->
-However, since the discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
-This law states that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait.
-This law emphasizes that the relative distance between stimuli, rather than their absolute positions on the continuum, determines the outcome of the pairwise comparison.
-Indeed, the theory assumes that the observed dichotomous outcome arises from the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*.
-Since the individual discriminal processes follow a Normal distribution, the discriminal difference also follows a Normal distribution [@Andrich_1978].
-In this distribution, the mode (mean) represents the relative separation between the stimuli and the dispersion captures the variability of that relative separation.
+However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces *the law of comparative judgment*.
+This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp.
+251].
+This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons.
+Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as *the discriminal difference*, determines the observed dichotomous outcome.
+Moreover, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
+In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.
 
@@
 <!-- 4. Example of discriminal difference -->
-@fig-discriminal_difference illustrates the distribution of the discriminal difference for the texts presented in @fig-discriminal_process.
+@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process.
 The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
-This conclusion is evident from the positive difference in the modal discriminal processes of the texts $(S_{B} - S_{A} > 0)$ and the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$ (shaded gray area).
+This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
 As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
 
-![Discriminal difference for two written texts](/images/figures/discriminal_difference.png){#fig-discriminal_difference width=70%}
+![Discriminal difference for two written texts](/images/png/discriminal_difference.png){#fig-discriminal_difference width=70%}
 
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:35:10 - 00:49:30 -->
+<!-- recording: Sven and Tine 24.12.09; time: 00:03:20 - 00:08:10 -->
 
-<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
-The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli.
-For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions.
-Case I represents the general form of the theory.
-Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge.
-Case III introduces the assumption of zero correlation between stimuli.
-Case IV builds on this by assuming the stimuli have similar dispersions.
-Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions.
-@tbl-thurstone_cases summarizes these cases and their assumptions.
-For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
-248-253].
+<!-- 1. What is the problem with the general form -->
+Thurstone observed that the general form of the theory, outlined in @sec-thurstone_theory, created a trait scaling problem.
+The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp.
+267].
+To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form.
+Each case progressively incorporated additional simplifying assumptions into the model.
 
-<!-- , suggesting all stimuli have the same variability in their perceived trait levels -->
+<!-- 2. Thurstone's Cases and their simplifying assumptions -->
+In Case I, Thurstone assumed that pairs of stimuli maintained a constant correlation across all comparisons.
+In Case II, he allowed multiple judges to make comparisons instead of restricting evaluations to a single judge.
+In Case III, he introduced the assumption of zero correlation between stimuli.
+In Case IV, he assumed stimuli exhibited similar dispersions.
+Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions.
+@tbl-thurstone_cases summarizes the assumptions of the general form and the five cases.
+For an in-depth discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
+248–253].
 
-![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}
+<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->
 
-<!-- 2. The Case V and its simpler statistical representation: the BTL model -->
-Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp.
+![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}
+
+<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
+Despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp.
 253; @Kelly_et_al_2022, pp.
@@
 This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
-16] (see @fig-logistic_vs_normal).
+16].
 
@@
 
-::: {#fig-logistic_vs_normal layout-nrow=2}
+<!-- ::: {#fig-logistic_vs_normal layout-nrow=2} -->
 
-![Probability density](/images/figures/density.png){#fig-density width=70%}
+<!-- ![Probability density](/images/png/density.png){#fig-density width=70%} -->
 
-![Cummulative probability](/images/figures/cummulative.png){#fig-cumulative width=70%}
+<!-- ![Cummulative probability](/images/png/cummulative.png){#fig-cumulative width=70%} -->
 
-Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255].
-:::
+<!-- Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255]. -->
+<!-- ::: -->
 
-<!-- 3. But what are limitations of Case V (and the BTL model)? -->
-Nevertheless, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+<!-- 4. But what are limitations of Case V (and the BTL model)? -->
+However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
@@
 He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions might be problematic when complex traits or heterogeneous stimuli are involved [@Thurstone_1927a, pp.
+270], acknowledging that some assumptions could prove problematic when researchers asesss complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376].
@@
 
+<!-- recording: Sven and Tine 24.12.09; time: 00:08:10 - 00:37:00 -->
+
 <!-- 1. What does the equal dispersion between stimuli imply? -->
-The discriminal dispersions of stimuli are crucial in determining the dichotomous outcomes of pairwise comparisons.
-Discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome.
-@fig-dispersion illustrates how more uncertainty in the trait perception of one text relative to another broadens the distribution of their discriminal difference.
-This broadening affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B > A)$.
-Additionally, the figure shows that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor one text over the other than when their dispersions differ (shaded gray area).
+According to the theory, the discriminal dispersions of stimuli play a critical role in determining the outcome of pairwise comparisons.
+Specifically, discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome.
+@fig-dispersion illustrates this idea, assuming a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
+The figure also considers that the discriminal dispersion for Text A remains constant and that the texts have no correlation $(\rho=0)$.
 
-![Discrepancies in the stimuli dispersions and tehir effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}
+@fig-dispersion reveals that more uncertainty in the trait perception of Text B compared to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
+This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome.
+Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor Text B over Text A (shaded gray area), compared to situations where their dispersions differ.
 
-<!-- 2. But does it hold?  -->
+![The discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/png/dispersion.png){#fig-dispersion width=70%}
+
+<!-- 2. But this is not the scenario we observe -->
+In experimental practice, however, this process occurs in reverse.
+Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their discriminal processes [@Thurstone_1927a, pp.
+373].
+For example, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
+This scenario is illustrated with @fig-dispersion when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
+The estimation accuracy of this discriminal difference, in turn, ensures reliable discriminal process estimates for the texts (citation needed?).
+It then intuitively follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.
+
+<!-- In experimental practice the procedure is the reverse of this hypothesis because the frequencies are known first experimentally, and from these frequencies we construct the psychological continuum.
+[@Thurstone_1927a, pp.
+373] -->
+
+<!-- 3. But does it hold?  -->
 However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-Indeed, evidence of such violation may already exist in the CJ literature as misfit statistics.
-These statistics measure the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp.
+Indeed, evidence of this violation may already exist in the CJ literature as misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp.
 12; @Goossens_et_al_2018, pp.
 20].
-For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than others [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
-This finding implies that discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts.
-Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
-Moreover, the presence of these "misfit" judges and their deviations can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue1b.
+For example, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
+These discrepancies, in turn, suggest that the discriminal differences for "misfit" texts have broader distributions, indicating that their discriminal processes may also exhibit more variation than that of other texts.
+A similar reasoning applies to "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
+Moreover, these "misfit" judges and their deviations can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue1b.
 
 <!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
+In measurements of the type known as judgment scales the discriminal dispersion on the psychological continuum becomes one of the unknowns to be determined as well as the scale value of the specimen.
+Every specimen in such a series presents two unknown values to be determined, namely, the scale value of its modal discriminal process on the psychological continuum and its discriminal dispersion.
 [@Thurstone_1927a, pp.
@@
 
-<!-- 3. But what are the challenges when these assumptions are violated? -->
-Therefore, assuming equal dispersions between stimuli, despite its violation, can cause Case V (and the BTL model) to inflate the reliability of the comparison outcome.
-Moreover, ignoring the difference in dispersions can lead to the neglect of critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp.
-370].
-Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+<!-- 4. So, what happens when the assumption does not hold? -->
+Then, incorrectly assuming equal dispersions between stimuli can lead the BTL model to introduce various statistical and measurement issues.
+For example, the model could overestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli.
+This overestimation may result in spurious inferences about these differences [@McElreath_2020, pp.
+370] and, by extension, about the stimuli's discriminal processes.
+@fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
+Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the common practice in CJ literature of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] may inadvertently discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
-The direction and magnitude of these biases are unpredictable, as they depend on the specific stimuli researchers exclude from the analysis.
-Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp.
-2].
+These biases are often unpredictable, as they depend on the specific stimuli excluded from the analysis.
 
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->
-
-<!-- People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. -->
+<!-- But ignoring over-dispersion can also lead to all of the same problems as ignoring any predictor variable.
+Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
+[@McElreath_2020, pp.
+370] -->
 
@@
 <!-- 1. What does the zero correlation between stimuli imply? -->
-Similar to the discriminal dispersions, the correlation between discriminal processes plays a crucial role in determining the dichotomous outcomes of pairwise comparisons.
-This correlation also shapes the distribution of the discriminal difference, directly influencing the comparison outcome.
-@fig-correlation illustrates how the dependence in trait perception between two texts narrows the distribution of their discriminal difference.
-This narrowing affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B>A)$.
-Moreover, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor one text over the other than when the texts are highly correlated (shaded gray area).
+Denoted by $\rho$, the correlation measures the dependence of a judge's perception of the trait in one stimulus on his perception of the same trait in another.
+Like the discriminal dispersions, this correlation shapes the distribution of the discriminal difference and directly influences the outcomes of pairwise comparisons.
+@fig-correlation illustrates this concept, assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
+The figure also considers that the discriminal dispersions for both texts remain constant.
+
+@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
+This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
+Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area), compared to scenarios when the texts are highly correlated.
 
@@
 
-![The correlation between stimuli and its effect on the distribution of the discriminal difference](/images/figures/correlation.png){#fig-correlation width=70%}
+![The correlation between stimuli and its effect on the distribution of the discriminal difference](/images/png/correlation.png){#fig-correlation width=70%}
+
+<!-- 2. But this is not the scenario we observe -->
+However, in experimental practice, researchers typically follow this process in reverse.
+For example, when they observe a large sample of outcomes favoring Text B over Text A and correctly assume zero correlation between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
+This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$.
+The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts (citation needed?).
+
+<!-- It then follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the assumption of zero correlation. -->
 
 <!-- 2. But does it hold?  -->
-Notably, Thurstone's Case V and the BTL model assume independent trait perceptions across stimuli.
+Notably, Thurstone's Case V and the BTL model assume independent discriminal processes across comparisons.
 Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
@@
 For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-This approach, however, can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue2.
+However, this approach can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue2.
 
 <!-- 5. But what are the challenges when this assumption is violated? -->
-In any case, the psychometric and statistical literature emphasizes the need to address factors that create dependencies between stimuli, as failing to do so can affect the reliability of the comparison outcomes and lead to inaccurate conclusions about the trait's estimates.
-For instance, researchers who overlook additional relevant traits, such as judges' biases, can cause dimensional mismatches in the statistical model used for analysis.
-This mismatch can artificially inflate the reliability of the trait [@Hoyle_et_al_2023, pp.
+[In any case, similar to @sec-theory-issue1a, incorrectly assuming zero correlation between stimuli can lead the BTL model to introduce various statistical and measurement issues.
+For instance, the model could over- or underestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli.
+This over- underestimation may result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes [@Hoyle_et_al_2023, pp.
+341].
+@fig-correlation also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
+This missaligment can be due to the overlook of additional relevant traits, such as judges' biases, which cause dimensional mismatches in the BTL model, artificially inflating the reliability of the trait [@Hoyle_et_al_2023, pp.
 341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989].
-Furthermore, as discussed in @sec-theory-issue1a, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Furthermore, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
-Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of reliability [@Hoyle_et_al_2023, pp.
-482].
-Together, these issues undermine the reliability of the trait and compromise the validity of its estimates [@Perron_et_al_2015, pp.
-2].
+Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp.
+482].]{style="color:red;"}
 
@@
 482] -->
-
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. -->
 
@@
 
+[In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, along with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can undermine the reliability of the trait estimates and ultimately compromise its validity [@Perron_et_al_2015, pp.
+2].
+Consequently, adopting a more systematic and integrated approach to examining what happens when judges compare two stimuli could offer several statistical and measurement benefits, including addressing these issues.]{style="color:red;"}
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. -->
+
+<!-- People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. -->
+
 <!-- ## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3} -->
@@
 <!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->
+
+The (latent) discriminal difference of the stimuli directly determines the (manifest) outcome of the pairwise comparisons
+
+![Theoretical model A1$](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=29%}
+
+The (latent) "perceived" discriminal processes for the stimuli directly determines their discriminal difference
+
+![Theoretical model A2$](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=40%}
+
+The (latent) "true" discriminal processes for the stimuli and the judges' biases directly determines their (latent) "perceived" discriminal processes
+
+![Theoretical model A3$](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=60%}
+
+without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in
+
+<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->
+
+<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
+One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
+In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
+Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
+-->
+
+<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
+
+<!-- Notably, the theory assumes the judge compares each pair of stimuli a "sufficient number of times" to accurately calculate the relative separation probability [@Thurstone_1927b, pp. 267]. -->
+
+<!-- It is assumed that the single observer compares each pair of stimuli sufficient number of times so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
+## From theory to statistics {#sec-theory-statistics}
+
+<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
+
+<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
+
+<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->
+
+<!-- [@deAyala_2009, p.
+291-29] think about indeterminancies in MIRT.
+to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
+
+<!-- , as well as when eliminating data through ad hoc univariate procedures -->
+<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
 
@@
 
-<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->
-
-<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
-One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
-In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
-Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
--->
-
 <!-- The correlation $\rho$, considered above is not the kind of correlation considered by Bock (1958) in studies of preference and choice of objects.
@@
 -->
-
-<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
-
-## From theory to statistics {#sec-theory-statistics}
-
-<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
-<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
-<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
-
-<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independece assumption -->
-
-<!-- [@deAyala_2009, p.
-291-29] think about indeterminancies in MIRT.
-to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
-
-<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
-
-<!-- , as well as when eliminating data through ad hoc univariate procedures -->
-<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
 

### 2024-12-13 _tex/index.tex
Built site for gh-pages

@@
 In its most general form, Thurstone's theory addresses pairwise
-comparisons of stimuli made by a single judge
+comparisons where a single judge evaluates multiple stimuli
 \citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
@@
 process of each stimulus and their discriminal difference. The
-\emph{discriminal process} represents the psychological impact each
-stimulus has on judges or, more simply, their perception of the stimulus
-trait. According to the theory, the discriminal process for each
-stimulus follows a Normal distribution, where its mode (mean), called
-the \emph{modal discriminal process}, indicates the stimulus position on
-the trait continuum, while its dispersion, known as the
+\emph{discriminal process} captures the psychological impact each
+stimulus exerts on the judge or, more simply, his perception of the
+stimulus trait. The theory assumes that the discriminal process for any
+given stimulus forms a Normal distribution along the trait continuum
+\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
+known as the \emph{modal discriminal process}, indicates the stimulus
+position on this continuum, while its dispersion, referred to as the
 \emph{discriminal dispersion}, reflects variability in the perceived
@@
 
-Figure~\ref{fig-discriminal_process} illustrates the discriminal
-processes along a quality trait continuum for two written texts. The
-figure shows that each discriminal process follows a Normal
-distribution. It also indicates that the modal discriminal process for
-Text B is positioned further along the continuum than that of Text A
-\((S_{B} > S_{A})\), signifying a higher quality for Text B.
+Figure~\ref{fig-discriminal_process} illustrates hypothetical
+discriminal processes along a quality trait continuum for two written
+texts. The figure indicates that the modal discriminal process for Text
+B is positioned further along the continuum than that of Text A
+\((S_{B} > S_{A})\), suggesting that Text B exhibits higher quality.
 Additionally, the figure highlights that Text B has a broader
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_process.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_process.png}
 
@@
 
-However, since the discriminal processes of the stimuli are not directly
-observable, the theory introduces the \emph{law of comparative
-judgment}. This law states that in pairwise comparisons, a judge
-perceives the stimulus positioned further along the trait continuum as
-having a higher level of that trait. This law emphasizes that the
-relative distance between stimuli, rather than their absolute positions
-on the continuum, determines the outcome of the pairwise comparison.
-Indeed, the theory assumes that the observed dichotomous outcome arises
-from the difference between the underlying discriminal processes of the
-stimuli, referred to as the \emph{discriminal difference}. Since the
-individual discriminal processes follow a Normal distribution, the
-discriminal difference also follows a Normal distribution
-\citep{Andrich_1978}. In this distribution, the mode (mean) represents
-the relative separation between the stimuli and the dispersion captures
-the variability of that relative separation.
+However, since the individual discriminal processes of the stimuli are
+not directly observable, the theory introduces \emph{the law of
+comparative judgment}. This law posits that in pairwise comparisons, a
+judge perceives the stimulus with a discriminal process positioned
+further along the trait continuum as possessing more of the trait
+\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
+between stimuli, rather than their absolute positions on the continuum,
+likely defines the outcome of pairwise comparisons. Indeed, the theory
+assumes that the difference between the underlying discriminal processes
+of the stimuli, referred to as \emph{the discriminal difference},
+determines the observed dichotomous outcome. Moreover, the theory
+assumes that because the individual discriminal processes form a Normal
+distribution on the continuum, the discriminal difference will also
+conform to a Normal distribution \citep{Andrich_1978}. In this
+distribution, the mode (mean) represents the relative separation between
+the stimuli, and its dispersion indicates the variability of that
+separation.
 
 Figure~\ref{fig-discriminal_difference} illustrates the distribution of
-the discriminal difference for the texts presented in
+the discriminal difference for the hypothetical texts depicted in
 Figure~\ref{fig-discriminal_process}. The figure indicates that the
 judge perceives Text B as having significantly higher quality than Text
-A. This conclusion is evident from the positive difference in their
-modal discriminal processes \((S_{B} - S_{A} > 0)\) and the area under
-the curve where the discriminal difference distinctly favors Text B over
-Text A, denoted as \(P(B > A)\) (shaded gray area). As a result, the
-dichotomous outcome of this comparison is more likely to favor Text B
-over Text A.
+A. This conclusion rests on two key observations: the positive
+difference between their modal discriminal processes
+\((S_{B} - S_{A} > 0)\) and the probability area where the discriminal
+difference distinctly favors Text B over Text A, represented by the
+shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
+outcome of this comparison is more likely to favor Text B over Text A.
 
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/discriminal_difference.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_difference.png}
 
@@
 
-Since the general form of Thurstone's theory, outlined in
-Section~\ref{sec-thurstone_theory}, applies to a CJ design where a
-single judge evaluates multiple stimuli, Thurstone developed four
-additional cases for the theory's practical application, each
-progressively incorporating additional simplifying assumptions. Case I
-represents the general form of the theory. Case II extends this by
-allowing multiple judges to make comparisons rather than restricting the
-comparisons to a single judge. Case III introduces the assumption of
-zero correlation between stimuli. Case IV builds on this by assuming the
-stimuli have similar dispersions. Finally, Case V replaces this
-assumption with the condition that the stimuli have equal discriminal
-dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
-their assumptions. For a detailed discussion of this progression, refer
-to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.
+Thurstone observed that the general form of the theory, outlined in
+Section~\ref{sec-thurstone_theory}, created a trait scaling problem. The
+model required estimating more ``unknown'' parameters than the available
+pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To address this
+issue and facilitate the practical application of the theory, he
+developed five cases derived from this general form. Each case
+progressively incorporated additional simplifying assumptions into the
+model.
+
+In Case I, Thurstone assumed that pairs of stimuli maintained a constant
+correlation across all comparisons. In Case II, he allowed multiple
+judges to make comparisons instead of restricting evaluations to a
+single judge. In Case III, he introduced the assumption of zero
+correlation between stimuli. In Case IV, he assumed stimuli exhibited
+similar dispersions. Finally, in Case V, he replaced this assumption
+with the condition that stimuli had equal discriminal dispersions.
+Table~\ref{tbl-thurstone_cases} summarizes the assumptions of the
+general form and the five cases. For an in-depth discussion of these
+cases and their progression, refer to \citet{Thurstone_1927b} and
+\citet[pp.~248--253]{Bramley_2008}.
 
@@
 
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/thurstone_cases.png}
 
@@
 
-Despite its reliance on the largest number of simplifying assumptions
+Despite relying on the most extensive set of simplifying assumptions
 \citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
@@
 scaling factor of approximately \(1.7\)
-\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
-Figure~\ref{fig-logistic_vs_normal}).
+\citep[pp.~16]{vanderLinden_et_al_2017_I}.
 
-\begin{figure}
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}
-
-}
-
-\subcaption{\label{fig-density}Probability density}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/cummulative.png}
-
-}
-
-\subcaption{\label{fig-cumulative}Cummulative probability}
-
-\end{minipage}%
-
-\caption{\label{fig-logistic_vs_normal}Probability density and
-cumulative probability of the logistic and Normal distributions.
-Extracted from \citet[pp.~254-255]{Bramley_2008}.}
-
-\end{figure}%
-
-Nevertheless, Thurstone originally developed Case V to provide a
-``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
+However, Thurstone originally developed Case V to provide a ``rather
+coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
 prioritizing statistical simplicity over precision in trait measurement
-\citep[pp.~677]{Kelly_et_al_2022}. Thurstone explicitly warned against
-its untested application, stating that its use ``should not be made
-without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b},
-acknowledging that some assumptions could be problematic when
-researchers asesss complex traits or heterogeneous stimuli
+\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
+untested application, stating that its use ``should not be made without
+(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
+that some assumptions could prove problematic when researchers asesss
+complex traits or heterogeneous stimuli
 \citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
@@
 
-The discriminal dispersions of stimuli are crucial in determining the
-dichotomous outcomes of pairwise comparisons. Holding all other modeling
-factors constant, discrepancies in these dispersions shape the
-distribution of the discriminal difference, directly influencing the
-comparison outcome. Figure~\ref{fig-dispersion} illustrates how more
-uncertainty in the trait perception of one text relative to another,
-\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
-discriminal difference. This broadening affects the area under the curve
-where the discriminal difference distinctly favors one text over the
-other, \(P(B > A)\). Additionally, the figure shows that when the
-discriminal dispersions of the texts are equal
+According to the theory, the discriminal dispersions of stimuli play a
+critical role in determining the outcome of pairwise comparisons.
+Specifically, discrepancies in these dispersions shape the distribution
+of the discriminal difference, directly influencing the comparison
+outcome. Figure~\ref{fig-dispersion} illustrates this idea, assuming a
+researcher can observe the discriminal processes for the texts shown in
+Figure~\ref{fig-discriminal_process}. The figure also considers that the
+discriminal dispersion for Text A remains constant and that the texts
+have no correlation \((\rho=0)\).
+
+Figure~\ref{fig-dispersion} reveals that more uncertainty in the trait
+perception of Text B compared to Text A, \((\sigma_{B}-\sigma_{A})\),
+broadens the distribution of their discriminal difference. This
+broadening affects the probability area where the discriminal difference
+distinctly favors Text B over Text A, expressed as \(P(B > A)\),
+ultimately influencing the comparison outcome. Additionally, the figure
+reveals that when the discriminal dispersions of the texts are equal
 \((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
-to favor one text over the other than when their dispersions differ
-(shaded gray area).
+to favor Text B over Text A (shaded gray area), compared to situations
+where their dispersions differ.
 
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/dispersion.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/dispersion.png}
 
@@
 
-However, Thurstone contended that this assumption of equal dispersions
+In experimental practice, however, this process occurs in reverse.
+Researchers first observe the comparison outcome and then use the BTL
+model to infer the discriminal difference between the stimuli and their
+discriminal processes \citep[pp.~373]{Thurstone_1927a}. For example,
+when researchers observe a large sample of outcomes favoring Text B over
+Text A and correctly assume equal dispersions between the texts, the BTL
+model estimates a discriminal difference distribution that accurately
+represents the ``true'' discriminal difference of the texts. This
+scenario is illustrated with Figure~\ref{fig-dispersion} when the
+discriminal difference distribution of the model aligns with the
+``true'' distribution, represented by the thick continuous line
+corresponding to \(\sigma_{B}-\sigma_{A}=0\). The estimation accuracy of
+this discriminal difference, in turn, ensures reliable discriminal
+process estimates for the texts (citation needed?). It then intuitively
+follows that the outcome's ability to represent the ``true'' differences
+between stimuli largely depends on the validity of the model's
+assumptions, particularly the assumption of equal dispersions.
+
+However, Thurstone contended that the assumption of equal dispersions
 may not hold when researchers assess complex traits or heterogeneous
@@
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
-Indeed, evidence of such violation may already exist in the CJ
-literature as misfit statistics. These statistics measure the judgment
-discrepancies associated with a given stimulus
+Indeed, evidence of this violation may already exist in the CJ
+literature as misfit statistics, which measure judgment discrepancies
+associated with specific stimuli
 \citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
-For instance, labeling texts as ``misfits'' indicates that comparisons
-involving these texts result in more judgment discrepancies than others
+For example, labeling texts as ``misfits'' indicates that comparisons
+involving these texts result in more judgment discrepancies than those
+involving other texts
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
-This finding implies that discriminal differences associated with
-``misfits'' texts usually display a broader dispersion, suggesting that
-the discriminal processes of these texts also exhibit more variation
-than other texts. Notably, this reasoning also applies to ``misfit''
-judges, whose evaluations reflect substantial deviations from the shared
-consensus due to the unique characteristics of the stimuli or the judges
-themselves. Moreover, the presence of these ``misfit'' judges and their
-deviations can introduce additional statistical and measurement issues,
-which we discuss in Section~\ref{sec-theory-issue1b}.
+These discrepancies, in turn, suggest that the discriminal differences
+for ``misfit'' texts have broader distributions, indicating that their
+discriminal processes may also exhibit more variation than that of other
+texts. A similar reasoning applies to ``misfit'' judges, whose
+evaluations deviate substantially from the shared consensus due to the
+unique characteristics of the stimuli or the judges themselves.
+Moreover, these ``misfit'' judges and their deviations can introduce
+additional statistical and measurement issues, which we discuss in
+Section~\ref{sec-theory-issue1b}.
 
-Therefore, assuming equal dispersions between stimuli, despite its
-violation, can cause Case V (and the BTL model) to inflate the
-reliability of the outcome, resulting in inaccurate conclusions about
-the comparison. Moreover, ignoring the disparity in dispersions can lead
-to the neglect of critical differences in the reliability of the trait
-across these stimuli, leading to erroneous conclusions about the trait's
-estimates \citep[pp.~370]{McElreath_2020}. Additionally, if researchers
-acknowledge that misfit statistics help identify these critical
-differences in dispersion, the usual practice in the CJ literature of
-excluding stimuli based on these statistics
-\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
-discarding valuable information and introducing bias into the trait's
-estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
-direction and magnitude of these biases are often unpredictable, as they
-depend on the specific stimuli researchers exclude from the analysis.
+Then, incorrectly assuming equal dispersions between stimuli can lead
+the BTL model to introduce various statistical and measurement issues.
+For example, the model could overestimate the accuracy of the outcome in
+reflecting the ``true'' discriminal differences between stimuli. This
+overestimation may result in spurious inferences about these differences
+\citep[pp.~370]{McElreath_2020} and, by extension, about the stimuli's
+discriminal processes. Figure~\ref{fig-dispersion} also illustrates this
+scenario when the model's discriminal difference distribution aligns
+with the thick continuous line for \(\sigma_{B}-\sigma_{A}=0\), while
+the ``true'' discriminal difference follows any discontinuous line where
+\(\sigma_{B}-\sigma_{A} \neq 0\). Moreover, if researchers recognize
+that misfit statistics highlight critical differences in dispersions,
+the common practice in CJ literature of excluding stimuli based on these
+statistics
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
+may inadvertently discard valuable information, introducing bias into
+trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. These
+biases are often unpredictable, as they depend on the specific stimuli
+excluded from the analysis.
 
@@
 
-Similar to the discriminal dispersions, the correlation between
-discriminal processes plays a crucial role in determining the
-dichotomous outcomes of pairwise comparisons. Holding all other modeling
-factors constant, this correlation shapes the distribution of the
-discriminal difference, directly influencing the comparison outcome.
-Figure~\ref{fig-correlation} illustrates how the dependence in trait
-perception between two texts narrows the distribution of their
-discriminal difference. This narrowing affects the area under the curve
-where the discriminal difference distinctly favors one text over the
-other, \(P(B>A)\). Moreover, the figure shows that when two texts are
-independent or uncorrelated \((\rho=0)\), their discriminal difference
-is less likely to favor one text over the other than when the texts are
-highly correlated (shaded gray area).
+Denoted by \(\rho\), the correlation measures the dependence of a
+judge's perception of the trait in one stimulus on his perception of the
+same trait in another. Like the discriminal dispersions, this
+correlation shapes the distribution of the discriminal difference and
+directly influences the outcomes of pairwise comparisons.
+Figure~\ref{fig-correlation} illustrates this concept, assuming the
+researcher can observe the discriminal processes for the texts shown in
+Figure~\ref{fig-discriminal_process}. The figure also considers that the
+discriminal dispersions for both texts remain constant.
+
+Figure~\ref{fig-correlation} reveals that as the correlation between the
+texts increases, the distribution of their discriminal difference
+becomes narrower. This narrowing affects the area under the curve where
+the discriminal difference distinctly favors Text B over Text A, denoted
+as \(P(B > A)\), thus influencing the comparison outcome. Furthermore,
+the figure shows that when two texts are independent or uncorrelated
+\((\rho=0)\), their discriminal difference is less likely to favor Text
+B over Text A (shaded gray area), compared to scenarios when the texts
+are highly correlated.
 
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/figures/correlation.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/correlation.png}
 
@@
 
-Notably, Thurstone's Case V and the BTL model assume independent trait
-perceptions across stimuli. Thurstone attributed this independence to
-the cancellation of potential judges' biases, driven by two opposing and
-equally weighted effects occurring during the pairwise comparisons
-\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} mathematically
-demonstrated this cancellation using the BTL model under the assumption
-of discriminal processes with additive biases. However, it is easy to
-imagine at least two scenarios where the zero correlation assumption
-almost certainly does not hold: when the pairwise comparison involves
-multidimensional, complex traits with heterogeneous stimuli and when an
-additional hierarchical structure is relevant to the stimuli.
+However, in experimental practice, researchers typically follow this
+process in reverse. For example, when they observe a large sample of
+outcomes favoring Text B over Text A and correctly assume zero
+correlation between the texts, the BTL model estimates a discriminal
+difference distribution that accurately represents the ``true''
+discriminal difference of the texts. This scenario is illustrated with
+Figure~\ref{fig-correlation} when the discriminal difference
+distribution of the model aligns with the ``true'' distribution,
+represented by the thick continuous line corresponding to \(\rho=0\).
+The accuracy of this discriminal difference estimation, in turn, ensures
+reliable estimates for the discriminal process of the texts (citation
+needed?).
+
+Notably, Thurstone's Case V and the BTL model assume independent
+discriminal processes across comparisons. Thurstone attributed this
+independence to the cancellation of potential judges' biases, driven by
+two opposing and equally weighted effects occurring during the pairwise
+comparisons \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978}
+mathematically demonstrated this cancellation using the BTL model under
+the assumption of discriminal processes with additive biases. However,
+it is easy to imagine at least two scenarios where the zero correlation
+assumption almost certainly does not hold: when the pairwise comparison
+involves multidimensional, complex traits with heterogeneous stimuli and
+when an additional hierarchical structure is relevant to the stimuli.
 
@@
 
-In any case, the psychometric and statistical literature emphasizes the
-need to address factors that create dependencies between stimuli, as
-failing to do so can affect the reliability of the comparison outcomes
-and lead to inaccurate conclusions about the trait's estimates. For
-instance, researchers who overlook additional relevant traits, such as
-judges' biases, can cause dimensional mismatches in the statistical
-model used for analysis. This mismatch can artificially inflate the
+{In any case, similar to Section~\ref{sec-theory-issue1a}, incorrectly
+assuming zero correlation between stimuli can lead the BTL model to
+introduce various statistical and measurement issues. For instance, the
+model could over- or underestimate the accuracy of the outcome in
+reflecting the ``true'' discriminal differences between stimuli. This
+over- underestimation may result in spurious inferences about these
+differences and, by extension, about the stimuli's discriminal processes
+\citep[pp.~341]{Hoyle_et_al_2023}. Figure~\ref{fig-correlation} also
+illustrates this scenario when the model's discriminal difference
+distribution aligns with the thick continuous line for \(\rho=0\), while
+the ``true'' discriminal difference follows any discontinuous line where
+\(\rho \neq 0\). This missaligment can be due to the overlook of
+additional relevant traits, such as judges' biases, which cause
+dimensional mismatches in the BTL model, artificially inflating the
 reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
 worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
-Furthermore, as discussed in Section~\ref{sec-theory-issue1a},
-researchers who exclude judges based on misfit statistics can risk
-discarding valuable information, further biasing the trait's estimates
-\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly, researchers
-who fail to account for hierarchical (grouping) structures can reduce
-the precision of model parameter estimates, which may amplify the
-overestimation of the trait's reliability
-\citep[pp.~482]{Hoyle_et_al_2023}.
+Furthermore, researchers who exclude judges based on misfit statistics
+can risk discarding valuable information, further biasing the trait's
+estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly,
+researchers who fail to account for hierarchical (grouping) structures
+can reduce the precision of model parameter estimates, which may amplify
+the overestimation of the trait's reliability
+\citep[pp.~482]{Hoyle_et_al_2023}.}
 
@@
 
-In aggregate, researchers' inadequate handling of violations to the
+{In aggregate, researchers' inadequate handling of violations to the
 assumptions of equal dispersion and zero correlation between stimuli,
@@
 measurement and hypothesis testing, can undermine the reliability of the
-trait and ultimately compromise its validity
+trait estimates and ultimately compromise its validity
 \citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
@@
 compare two stimuli could offer several statistical and measurement
-benefits, including addressing these issues.
+benefits, including addressing these issues.}
 
@@
 \subsection{The theoretical model}\label{sec-theory-theoretical}
+
+The (latent) discriminal difference of the stimuli directly determines
+the (manifest) outcome of the pairwise comparisons
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.29\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}
+
+}
+
+\caption{\label{fig-CJ_TM_A1}Theoretical model A1\$}
+
+\end{figure}%
+
+The (latent) ``perceived'' discriminal processes for the stimuli
+directly determines their discriminal difference
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}
+
+}
+
+\caption{\label{fig-CJ_TM_A2}Theoretical model A2\$}
+
+\end{figure}%
+
+The (latent) ``true'' discriminal processes for the stimuli and the
+judges' biases directly determines their (latent) ``perceived''
+discriminal processes
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}
+
+}
+
+\caption{\label{fig-CJ_TM_A3}Theoretical model A3\$}
+
+\end{figure}%
+
+without loosing generality, the (latent) ``perceived'' and ``true''
+discriminal processes for the stimuli can be depicted in a vector for
+each judge, as in
 

### 2024-12-13 code/figures.R

@@ -105,7 +105,7 @@ dev.off()
 
 
 
-# figure 3a ####
+# figure XX ####
 png( filename=file.path(main_dir, 'density.png'), 
      height=10, width=15, units='cm', res=400  )
 
@@ -125,7 +125,7 @@ dev.off()
 
 
 
-# figure 3b ####
+# figure XX ####
 png( filename=file.path(main_dir, 'cummulative.png'), 
      height=10, width=15, units='cm', res=400  )
 
@@ -146,7 +146,7 @@ dev.off()
 
 
 
-# figure 4 ####
+# figure 3 ####
 # distributional parameters
 dpB_d = round( seq(0.5, 1.5, by=0.25), 2 ) # correlation between discriminal processes
 dd = list()
@@ -197,7 +197,7 @@ dev.off()
 
 
 
-# figure 5 ####
+# figure 4 ####
 # distributional parameters
 rAB = round( seq(-0.5, 0.5, by=0.25), 2 ) # correlation between discriminal processes
 dd = list()

### 2024-12-17 paper2.qmd
section:issues; process:write,figures

--- 2024-12-13 index.qmd
@@
 keywords:
-  - Probability
+  - Causal analysis
   - Directed Acyclic Graphs
-  - Bayesian methods
+  - Bayesian statistical methods
   - Thurstonian model
   - Comparative judgement
-  - Structural Causal Models
+  - Probability
   - Statistical modeling
 abstract: |
-  (to do)
+  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches.
+Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data.
+Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation.
+The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data.
+The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences.
+This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
 key-points:
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:58:20 - 00:59:20 -->
+<!-- recording: Sven and Tine 24.12.17; time: 00:11:50 - 00:26:10, (Here we did it!) -->
 
@@
 
-![Discriminal processes for two written texts](/images/png/discriminal_process.png){#fig-discriminal_process width=70%}
+::: {#fig-thurstone_theory layout-ncol=2}
+
+![Discriminal processes](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}
+
+![Discriminal difference](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}
+
+Hypothetical discriminal processes and discriminant difference along a quality trait continuum for two written texts.
+:::
 
@@
 <!-- 4. Example of discriminal difference -->
-@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process.
+@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in panel @fig-discriminal_process.
 The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
@@
 As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
-
-![Discriminal difference for two written texts](/images/png/discriminal_difference.png){#fig-discriminal_difference width=70%}
 
@@
 <!-- recording: Sven and Tine 24.12.09; time: 00:03:20 - 00:08:10 -->
+<!-- recording: Sven and Tine 24.12.17; time: 00:00:00 - 00:06:10 -->
 
@@
 267].
-To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form.
-Each case progressively incorporated additional simplifying assumptions into the model.
+To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form, each case progressively incorporating additional simplifying assumptions into the model.
 
@@
 
-<!-- Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255]. -->
+<!-- Probability density and cumulative probability of the logistic and Normal distributions. -->
 <!-- ::: -->
@@
 He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions could prove problematic when researchers asesss complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+270], acknowledging that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376].
@@
 <!-- recording: Sven and Tine 24.12.09; time: 00:08:10 - 00:37:00 -->
+<!-- recording: Sven and Tine 24.12.17; time: 00:06:10 - 00:08:50 -->
 
 <!-- 1. What does the equal dispersion between stimuli imply? -->
-According to the theory, the discriminal dispersions of stimuli play a critical role in determining the outcome of pairwise comparisons.
-Specifically, discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome.
-@fig-dispersion illustrates this idea, assuming a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
-The figure also considers that the discriminal dispersion for Text A remains constant and that the texts have no correlation $(\rho=0)$.
-
-@fig-dispersion reveals that more uncertainty in the trait perception of Text B compared to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
+According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons.
+@fig-dispersion illustrates this idea using a thought experiment in which a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
+Furthermore, the figure assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
+The figure reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
 This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome.
@@
 
-![The discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/png/dispersion.png){#fig-dispersion width=70%}
-
 <!-- 2. But this is not the scenario we observe -->
 In experimental practice, however, this process occurs in reverse.
-Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their discriminal processes [@Thurstone_1927a, pp.
+Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp.
 373].
-For example, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
-This scenario is illustrated with @fig-dispersion when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
-The estimation accuracy of this discriminal difference, in turn, ensures reliable discriminal process estimates for the texts (citation needed?).
-It then intuitively follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.
+For instance, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
+This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
+The accuracy of this estimated discriminal difference ensures reliable estimates of the texts’ discriminal processes [(citation needed?){style="color:red;"}.
+Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.
 
@@
 <!-- 4. So, what happens when the assumption does not hold? -->
-Then, incorrectly assuming equal dispersions between stimuli can lead the BTL model to introduce various statistical and measurement issues.
-For example, the model could overestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli.
-This overestimation may result in spurious inferences about these differences [@McElreath_2020, pp.
-370] and, by extension, about the stimuli's discriminal processes.
-@fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the common practice in CJ literature of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] may inadvertently discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Therefore, incorrectly assuming equal dispersions between stimuli is not without harm, as it can cause the BTL model to introduce significant statistical and measurement issues.
+For instance, the model may overestimate how accurately the outcome reflects the "true" discriminal differences between stimuli.
+This overestimation can lead researchers to draw spurious conclusions about these differences [@McElreath_2020, pp.
+370] and, by extension, of the stimuli underlying discriminal processes.
+@fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows a discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
+Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the usual CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information.
+This exclusion can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
-These biases are often unpredictable, as they depend on the specific stimuli excluded from the analysis.
+The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.
 
@@
 
+::: {#fig-caseV_issues layout-ncol=2}
+
+![Discrepancies in the dispersions of stimuli](/images/png/dispersion.png){#fig-dispersion width=100%}
+
+![Correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}
+
+The effect of dispersion discrepancies and stimulus correlation on the distribution of the discriminal difference.
+:::
+
 ### The assumption of zero correlation between stimuli {#sec-theory-issue1b}
+
+<!-- recording: Sven and Tine 24.12.17; time: 00:08:50 - 00:10:30 -->
 
@@
 Like the discriminal dispersions, this correlation shapes the distribution of the discriminal difference and directly influences the outcomes of pairwise comparisons.
-@fig-correlation illustrates this concept, assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
+@fig-correlation illustrates this concept, again assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
 The figure also considers that the discriminal dispersions for both texts remain constant.
@@
 
-![The correlation between stimuli and its effect on the distribution of the discriminal difference](/images/png/correlation.png){#fig-correlation width=70%}
-
 <!-- 2. But this is not the scenario we observe -->
@@
 This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$.
-The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts (citation needed?).
+The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts [(citation needed?){style="color:red;"}.
 
@@
 <!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->
+<!-- recording: Sven and Tine 24.12.17; time: 00:26:10 - 00:54:00 -->
 
@@
 We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used.
-However, work to derive some general principles for good practice would be valuable.[@Kelly_et_al_2022, pp.
+However, work to derive some general principles for good practice would be valuable.
+[@Kelly_et_al_2022, pp.
 683] -->

### 2024-12-17 _tex/paper2.tex
Built site for gh-pages

--- 2024-12-13 _tex/index.tex
@@
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
-  pdfkeywords={Probability, Directed Acyclic Graphs, Bayesian
-methods, Thurstonian model, Comparative judgement, Structural Causal
-Models, Statistical modeling},
+  pdfkeywords={Causal analysis, Directed Acyclic Graphs, Bayesian
+statistical methods, Thurstonian model, Comparative
+judgement, Probability, Statistical modeling},
   colorlinks=true,
@@
 \begin{abstract}
-(to do)
+This study revisits Thurstone's law of comparative judgments (CJ) by
+addressing two key limitations in traditional approaches. Firstly, it
+addresses the overreliance on the assumptions of Thurstone's Case V in
+the statistical analysis of CJ data. Secondly, it addresses the apparent
+disconnect between CJ's approach to trait measurement and hypothesis
+testing. We put forward a systematic approach based on causal analysis
+and Bayesian statistical methods, which results in a model that
+facilitates a more comprehensive understanding of the factors
+influencing CJ experiments while offering a robust statistical
+translation. The new model accommodates unequal dispersions and
+correlations between stimuli, enhancing the reliability and validity of
+CJ's trait estimation, thereby ensuring the accurate measurement and
+interpretation of comparative data. The paper highlights the relevance
+of this updated framework for modern empirical research, particularly in
+education and social sciences. This contribution advances current
+research methodologies, providing a robust foundation for future
+applications in diverse fields.
 \end{abstract}
@@
 \begin{keyword}
-    Probability \sep Directed Acyclic Graphs \sep Bayesian
-methods \sep Thurstonian model \sep Comparative
-judgement \sep Structural Causal Models \sep 
+    Causal analysis \sep Directed Acyclic Graphs \sep Bayesian
+statistical methods \sep Thurstonian model \sep Comparative
+judgement \sep Probability \sep 
     Statistical modeling
@@
 
+\setstretch{2}
 \section{Introduction}\label{sec-introduction}
@@
 
+\begin{minipage}{0.50\linewidth}
+
 \centering{
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_process.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_process.png}
 
@@
 
-\caption{\label{fig-discriminal_process}Discriminal processes for two
-written texts}
+\subcaption{\label{fig-discriminal_process}Discriminal processes}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_difference.png}
+
+}
+
+\subcaption{\label{fig-discriminal_difference}Discriminal difference}
+
+\end{minipage}%
+
+\caption{\label{fig-thurstone_theory}Hypothetical discriminal processes
+and discriminant difference along a quality trait continuum for two
+written texts.}
 
@@
 Figure~\ref{fig-discriminal_difference} illustrates the distribution of
-the discriminal difference for the hypothetical texts depicted in
+the discriminal difference for the hypothetical texts depicted in panel
 Figure~\ref{fig-discriminal_process}. The figure indicates that the
@@
 outcome of this comparison is more likely to favor Text B over Text A.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_difference.png}
-
-}
-
-\caption{\label{fig-discriminal_difference}Discriminal difference for
-two written texts}
-
-\end{figure}%
 
@@
 issue and facilitate the practical application of the theory, he
-developed five cases derived from this general form. Each case
-progressively incorporated additional simplifying assumptions into the
+developed five cases derived from this general form, each case
+progressively incorporating additional simplifying assumptions into the
 model.
@@
 (an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
-that some assumptions could prove problematic when researchers asesss
+that some assumptions could prove problematic when researchers assess
 complex traits or heterogeneous stimuli
@@
 
-According to the theory, the discriminal dispersions of stimuli play a
-critical role in determining the outcome of pairwise comparisons.
-Specifically, discrepancies in these dispersions shape the distribution
-of the discriminal difference, directly influencing the comparison
-outcome. Figure~\ref{fig-dispersion} illustrates this idea, assuming a
-researcher can observe the discriminal processes for the texts shown in
-Figure~\ref{fig-discriminal_process}. The figure also considers that the
-discriminal dispersion for Text A remains constant and that the texts
-have no correlation \((\rho=0)\).
-
-Figure~\ref{fig-dispersion} reveals that more uncertainty in the trait
-perception of Text B compared to Text A, \((\sigma_{B}-\sigma_{A})\),
-broadens the distribution of their discriminal difference. This
-broadening affects the probability area where the discriminal difference
-distinctly favors Text B over Text A, expressed as \(P(B > A)\),
-ultimately influencing the comparison outcome. Additionally, the figure
-reveals that when the discriminal dispersions of the texts are equal
-\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
-to favor Text B over Text A (shaded gray area), compared to situations
-where their dispersions differ.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/dispersion.png}
-
-}
-
-\caption{\label{fig-dispersion}The discrepancy in the dispersions of
-stimuli and their effect on the distribution of the discriminal
-difference}
-
-\end{figure}%
+According to the theory, discrepancies in the discriminal dispersions of
+stimuli shape the distribution of the discriminal difference, exerting a
+direct influence on the outcome of pairwise comparisons.
+Figure~\ref{fig-dispersion} illustrates this idea using a thought
+experiment in which a researcher can observe the discriminal processes
+for the texts shown in Figure~\ref{fig-discriminal_process}.
+Furthermore, the figure assumes that the discriminal dispersion for Text
+A remains constant and that the texts are uncorrelated \((\rho=0)\). The
+figure reveals that an increase in the uncertainty associated with the
+perception of Text B in comparison to Text A,
+\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
+discriminal difference. This broadening affects the probability area
+where the discriminal difference distinctly favors Text B over Text A,
+expressed as \(P(B > A)\), ultimately influencing the comparison
+outcome. Additionally, the figure reveals that when the discriminal
+dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
+discriminal difference is more likely to favor Text B over Text A
+(shaded gray area), compared to situations where their dispersions
+differ.
 
@@
 model to infer the discriminal difference between the stimuli and their
-discriminal processes \citep[pp.~373]{Thurstone_1927a}. For example,
-when researchers observe a large sample of outcomes favoring Text B over
-Text A and correctly assume equal dispersions between the texts, the BTL
-model estimates a discriminal difference distribution that accurately
-represents the ``true'' discriminal difference of the texts. This
-scenario is illustrated with Figure~\ref{fig-dispersion} when the
-discriminal difference distribution of the model aligns with the
+respective discriminal processes \citep[pp.~373]{Thurstone_1927a}. For
+instance, when researchers observe a large sample of outcomes favoring
+Text B over Text A and correctly assume equal dispersions between the
+texts, the BTL model estimates a discriminal difference distribution
+that accurately represents the ``true'' discriminal difference of the
+texts. This scenario is illustrated in Figure~\ref{fig-dispersion},
+where the model's discriminal difference distribution aligns with the
 ``true'' distribution, represented by the thick continuous line
-corresponding to \(\sigma_{B}-\sigma_{A}=0\). The estimation accuracy of
-this discriminal difference, in turn, ensures reliable discriminal
-process estimates for the texts (citation needed?). It then intuitively
-follows that the outcome's ability to represent the ``true'' differences
-between stimuli largely depends on the validity of the model's
-assumptions, particularly the assumption of equal dispersions.
+corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
+estimated discriminal difference ensures reliable estimates of the
+texts' discriminal processes {[}(citation
+needed?)\{style=``color:red;''\}. Therefore, it is reasonable to assume
+that the outcome's ability to reflect the ``true'' differences between
+stimuli largely depends on the validity of the model's assumptions,
+particularly the assumption of equal dispersions.
 
@@
 
-Then, incorrectly assuming equal dispersions between stimuli can lead
-the BTL model to introduce various statistical and measurement issues.
-For example, the model could overestimate the accuracy of the outcome in
-reflecting the ``true'' discriminal differences between stimuli. This
-overestimation may result in spurious inferences about these differences
-\citep[pp.~370]{McElreath_2020} and, by extension, about the stimuli's
-discriminal processes. Figure~\ref{fig-dispersion} also illustrates this
-scenario when the model's discriminal difference distribution aligns
-with the thick continuous line for \(\sigma_{B}-\sigma_{A}=0\), while
-the ``true'' discriminal difference follows any discontinuous line where
-\(\sigma_{B}-\sigma_{A} \neq 0\). Moreover, if researchers recognize
-that misfit statistics highlight critical differences in dispersions,
-the common practice in CJ literature of excluding stimuli based on these
-statistics
+Therefore, incorrectly assuming equal dispersions between stimuli is not
+without harm, as it can cause the BTL model to introduce significant
+statistical and measurement issues. For instance, the model may
+overestimate how accurately the outcome reflects the ``true''
+discriminal differences between stimuli. This overestimation can lead
+researchers to draw spurious conclusions about these differences
+\citep[pp.~370]{McElreath_2020} and, by extension, of the stimuli
+underlying discriminal processes. Figure~\ref{fig-dispersion}
+illustrates this issue when the model's discriminal difference
+distribution aligns with the thick continuous line for
+\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
+follows a discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
+Moreover, if researchers recognize that misfit statistics highlight
+critical differences in dispersions, the usual CJ practice of excluding
+stimuli based on these statistics
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
-may inadvertently discard valuable information, introducing bias into
-trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. These
-biases are often unpredictable, as they depend on the specific stimuli
-excluded from the analysis.
+can unintentionally discard valuable information. This exclusion can
+introduce bias into trait estimates
+\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
+magnitude of these biases are often unpredictable, as they depend on
+which stimuli are excluded from the analysis.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/dispersion.png}
+
+}
+
+\subcaption{\label{fig-dispersion}Discrepancies in the dispersions of
+stimuli}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/correlation.png}
+
+}
+
+\subcaption{\label{fig-correlation}Correlation between stimuli}
+
+\end{minipage}%
+
+\caption{\label{fig-caseV_issues}The effect of dispersion discrepancies
+and stimulus correlation on the distribution of the discriminal
+difference.}
+
+\end{figure}%
 
@@
 directly influences the outcomes of pairwise comparisons.
-Figure~\ref{fig-correlation} illustrates this concept, assuming the
-researcher can observe the discriminal processes for the texts shown in
-Figure~\ref{fig-discriminal_process}. The figure also considers that the
-discriminal dispersions for both texts remain constant.
+Figure~\ref{fig-correlation} illustrates this concept, again assuming
+the researcher can observe the discriminal processes for the texts shown
+in Figure~\ref{fig-discriminal_process}. The figure also considers that
+the discriminal dispersions for both texts remain constant.
 
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/correlation.png}
-
-}
-
-\caption{\label{fig-correlation}The correlation between stimuli and its
-effect on the distribution of the discriminal difference}
-
-\end{figure}%
-
 However, in experimental practice, researchers typically follow this
@@
 The accuracy of this discriminal difference estimation, in turn, ensures
-reliable estimates for the discriminal process of the texts (citation
-needed?).
+reliable estimates for the discriminal process of the texts {[}(citation
+needed?)\{style=``color:red;''\}.
 

### 2024-12-18 paper2.qmd
section:issues,model; process:write,figures

@@
 bibliography: references.bib
-notebook-links: global
-lightbox: true
-citation: true
-  # type: article-journal
-  # container-title: "Psychometrika"
-  # doi: ""
-  # # url: https://example.com/summarizing-output
-execute: 
-  cache: true
-  # eval: true
-  echo: false
-  # output: true
-  # include: true
-  warning: false
-  error: false
-  message: false
 ---
@@
 @fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts.
-The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$, suggesting that Text B exhibits higher quality.
+The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
 Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
@@
 <!-- 4. Example of discriminal difference -->
-@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in panel @fig-discriminal_process.
+@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process.
 The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
-This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
+This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
 As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
@@
 This section examines the two critical issues in the CJ literature that serve as the primary focus of this study.
-The first is the overreliance on Thurstone’s Case V assumptions in the statistical analysis of CJ data.
+The first is the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
 The second is the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
@@
 <!-- 1. What is the problem with the general form -->
-Thurstone observed that the general form of the theory, outlined in @sec-thurstone_theory, created a trait scaling problem.
+Thurstone noted that the general form of the theory, as delineated in @sec-thurstone_theory, gave rise to a trait scaling problem.
 The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp.
 267].
-To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form, each case progressively incorporating additional simplifying assumptions into the model.
+To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each case progressively incorporated additional simplifying assumptions into the model.
 
 <!-- 2. Thurstone's Cases and their simplifying assumptions -->
-In Case I, Thurstone assumed that pairs of stimuli maintained a constant correlation across all comparisons.
-In Case II, he allowed multiple judges to make comparisons instead of restricting evaluations to a single judge.
-In Case III, he introduced the assumption of zero correlation between stimuli.
-In Case IV, he assumed stimuli exhibited similar dispersions.
+In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons.
+In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge.
+In Case III, he posited that there was no correlation between stimuli.
+In Case IV, he assumed that the stimuli exhibited similar dispersions.
 Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions.
 @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases.
-For an in-depth discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
+For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
 248–253].
@@
 This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
+The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
 254] (see @tbl-thurstone_cases).
-This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
+This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
 16].
@@
 He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], acknowledging that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+270].
+Furthermore, he acknowledged that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376].
-Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.
+Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice, namely the assumption of equal dispersion and zero correlation between stimuli.
 
@@
 According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons.
-@fig-dispersion illustrates this idea using a thought experiment in which a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
+@fig-dispersion presents a thought experiment to illustrate this idea.
+In this experiment, a researcher can observe the discriminal processes for the texts depicted in @fig-discriminal_process.
 Furthermore, the figure assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
@@
 373].
-For instance, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
+Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.
+For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
 This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
-The accuracy of this estimated discriminal difference ensures reliable estimates of the texts’ discriminal processes [(citation needed?){style="color:red;"}.
-Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.
+The accuracy of these discriminal difference ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.
 
@@
 <!-- 3. But does it hold?  -->
-However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+However, Thurstone argued that the assumption of equal dispersions may not be applicable when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-Indeed, evidence of this violation may already exist in the CJ literature as misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp.
+Indeed, evidence of this violation may already be present in the CJ literature in the form of misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp.
 12; @Goossens_et_al_2018, pp.
@@
 These discrepancies, in turn, suggest that the discriminal differences for "misfit" texts have broader distributions, indicating that their discriminal processes may also exhibit more variation than that of other texts.
-A similar reasoning applies to "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
-Moreover, these "misfit" judges and their deviations can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue1b.
+A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
+These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.
 
@@
 <!-- 4. So, what happens when the assumption does not hold? -->
-Therefore, incorrectly assuming equal dispersions between stimuli is not without harm, as it can cause the BTL model to introduce significant statistical and measurement issues.
-For instance, the model may overestimate how accurately the outcome reflects the "true" discriminal differences between stimuli.
-This overestimation can lead researchers to draw spurious conclusions about these differences [@McElreath_2020, pp.
-370] and, by extension, of the stimuli underlying discriminal processes.
-@fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows a discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the usual CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information.
-This exclusion can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Nevertheless, model misspecification, in the form of an erroneous assumption of equal dispersions between stimuli, can give rise to significant statistical and measurement issues.
+For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli.
+This overestimation can result in researchers drawing spurious conclusions about these differences [@McElreath_2020, pp.
+370] and, by extension, about the underlying discriminal processes of stimuli.
+@fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
+
+Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information.
+Such exclusions can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
@@
 <!-- 1. What does the zero correlation between stimuli imply? -->
-Denoted by $\rho$, the correlation measures the dependence of a judge's perception of the trait in one stimulus on his perception of the same trait in another.
-Like the discriminal dispersions, this correlation shapes the distribution of the discriminal difference and directly influences the outcomes of pairwise comparisons.
-@fig-correlation illustrates this concept, again assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process.
-The figure also considers that the discriminal dispersions for both texts remain constant.
-
-@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
+The correlation, represented by the symbol $\rho$, measures how much a judge's perception of a specific trait in one stimulus depends on their perception of the same trait in another.
+As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
+@fig-correlation presents a similar thought experiment as in @sec-theory-issue1a to illustrate this idea.
+The illustration now assumes that the discriminal dispersions for both texts remain constant.
+The figure shows that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
 This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
-Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area), compared to scenarios when the texts are highly correlated.
+Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area) compared to scenarios where the texts are highly correlated.
 
@@
 <!-- 2. But this is not the scenario we observe -->
-However, in experimental practice, researchers typically follow this process in reverse.
-For example, when they observe a large sample of outcomes favoring Text B over Text A and correctly assume zero correlation between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
+However, in practice, researchers approach this process in reverse.
+They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli.
+Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts.
 This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$.
-The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts [(citation needed?){style="color:red;"}.
+Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.
 
@@
 <!-- 2. But does it hold?  -->
-Notably, Thurstone's Case V and the BTL model assume independent discriminal processes across comparisons.
-Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp.
+Notably, Thurstone attributed the independence of stimuli to the cancellation of potential judges' biases.
+He argued that this cancellation resulted from two opposing and equally weighted effects occurring during pairwise comparisons [@Thurstone_1927b, pp.
 268].
-@Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
-However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.
+@Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
+However, it is easy to imagine at least two scenarios in which the zero correlation assumption is almost certainly invalid: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.
 
@@
 Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
+These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can exert an uneven influence on judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
 346].
-For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, ultimately favoring texts with fewer errors but weaker arguments.
-While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.
+For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments.
+While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
 
@@
 In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
-Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between stimuli has been inadequate.
-For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, this approach can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue2.
+Despite the CJ literature acknowledging the existence of such hierarchical structures, the statistical handling of this additional source of dependence between stimuli has been inadequate.
+For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
 
 <!-- 5. But what are the challenges when this assumption is violated? -->
-[In any case, similar to @sec-theory-issue1a, incorrectly assuming zero correlation between stimuli can lead the BTL model to introduce various statistical and measurement issues.
-For instance, the model could over- or underestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli.
-This over- underestimation may result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes [@Hoyle_et_al_2023, pp.
-341].
-@fig-correlation also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
-This missaligment can be due to the overlook of additional relevant traits, such as judges' biases, which cause dimensional mismatches in the BTL model, artificially inflating the reliability of the trait [@Hoyle_et_al_2023, pp.
-341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989].
-Furthermore, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+In any case, similar to @sec-theory-issue1a, model misspecification due to an erroneous assumption of zero correlation between stimuli can lead to significant statistical and measurement issues.
+For instance, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli.
+Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes.
+This scenario is illustrated in @fig-correlation, where the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
+
+This misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures.
+Neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp.
+341] or, worse, introducing bias into the trait's estimates [@Ackerman_1989].
+Excluding judges based on misfit statistics risks discarding valuable information, which may further bias the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
-Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp.
-482].]{style="color:red;"}
+Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially amplifying the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp.
+482].
 
@@
 
-[In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, along with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can undermine the reliability of the trait estimates and ultimately compromise its validity [@Perron_et_al_2015, pp.
+In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can potentially compromise the reliability of the trait estimates and, by extension, their validity [@Perron_et_al_2015, pp.
 2].
-Consequently, adopting a more systematic and integrated approach to examining what happens when judges compare two stimuli could offer several statistical and measurement benefits, including addressing these issues.]{style="color:red;"}
+Consequently, adopting a more systematic and integrated approach to examine the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address these issues.
 
@@
 
-This section presents a theoretical model for CJ that extends Thurstone's theory.
-The model systematically incorporates all factors involved when judges make pairwise comparisons.
-Additionally, the section develops the statistical translation of the theoretical model based on assumptions informed by the CJ theory.
-
-<!-- in a stepwise manner -->
+This section introduces a theoretical model for Comparative Judgment (CJ) that builds upon Thurstone's theory.
+The model systematically accounts for the factors influencing judges during CJ experiments.
+These theoretical and practical insights are then translated into a statistical model.
 
@@
 
-![Theoretical model A1$](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=29%}
-
 The (latent) "perceived" discriminal processes for the stimuli directly determines their discriminal difference
-
-![Theoretical model A2$](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=40%}
 
@@
 
-![Theoretical model A3$](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=60%}
+::: {.panel-tabset group="Figure"}
+
+#### Figure 3
+
+![](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=20%}
+
+#### Figure 4
+
+![](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=30%}
+
+#### Figure 5
+
+![](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=42%}
+
+#### Figure 6
+
+![](/images/png/CJ_TM_A4.png){#fig-CJ_TM_A4 width=42%}
+
+:::
 
 without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in
+
+::: panel-tabset
+
+#### Figure 7
+
+![](/images/png/CJ_TM_B1.png){#fig-CJ_TM_B1 width=42%}
+
+#### Figure 8
+
+![](/images/png/CJ_TM_B2.png){#fig-CJ_TM_B2 width=62%}
+
+#### Figure 9
+
+![](/images/png/CJ_TM_B3.png){#fig-CJ_TM_B3 width=72%}
+
+#### Figure 10
+
+![](/images/png/CJ_TM_B4.png){#fig-CJ_TM_B4 width=72%}
+
+#### Figure 11
+
+![](/images/png/CJ_TM_B5.png){#fig-CJ_TM_B5 width=72%}
+
+#### Figure 12
+
+![](/images/png/CJ_TM_B6.png){#fig-CJ_TM_B6 width=72%}
+
+:::
+
+Considering the sampling process
+
+::: panel-tabset
+
+#### Figure 13
+
+![](/images/png/CJ_TM_B7.png){#fig-CJ_TM_B7 width=90%}
+
+#### Figure 14
+
+![](/images/png/CJ_TM_B8.png){#fig-CJ_TM_B8 width=100%}
+
+#### Figure 15
+
+![](/images/png/CJ_TM_B9.png){#fig-CJ_TM_B9 width=100%}
+
+:::
 
@@
 
-**AI-assisted technologies in the writing process:** The authors used ChatGPT, an AI language model, during the preparation of this work.
-They occasionally employed the tool to refine phrasing and optimize wording, ensuring appropriate language use and enhancing the manuscript's clarity and coherence.
+**AI-assisted technologies in the writing process:** The authors utilized a range of AI-based language tools throughout the preparation of this work.
+They occasionally employed the tools to refine phrasing and optimize wording, ensuring appropriate language use and enhancing the manuscript's clarity and coherence.
 The authors take full responsibility for the final content of the publication.

### 2024-12-18 code/figures.R

@@ -1,7 +1,7 @@
 # working area ####
 rm(list=ls())
 
-main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript/images/figures/'
+main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript/images/png/'
 
 
 
@@ -31,7 +31,7 @@ plot( NULL, xlim=c(-2,5), ylim=c(0,0.9),
       xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=c(dpA[1], dpB[1]), tick=F, 
-      label=c( expression( S[A] ), expression( S[B] ) ) )
+      label=c( expression( T[A] ), expression( T[B] ) ) )
 
 # figure parameters
 pplotA = c( dnorm( dpA[1], mean=dpA[1], sd=dpA[2] ),
@@ -78,7 +78,7 @@ plot( NULL, xlim=c(-2,5), ylim=c(0,0.4),
       xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=c(0, dd[1]), tick=F, 
-      label=c(0, expression( S[B] - S[A] ) ) )
+      label=c(0, expression( T[B] - T[A] ) ) )
 
 # figure parameters
 pplotD = c( dnorm( 0, mean=dd[1], sd=dd[2] ),
@@ -164,7 +164,7 @@ plot( NULL, xlim=c(-2,5), ylim=c(0,0.6),
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=0, tick=F, label='0' )
 axis( side=1, at=c(0, dd[1]), tick=F, 
-      label=c(0, expression( S[B] - S[A] ) ) )
+      label=c(0, expression( T[B] - T[A] ) ) )
 
 # figure parameters
 l_ty = c( 1, rep(2,2), rep(3,2) )
@@ -215,7 +215,7 @@ plot( NULL, xlim=c(-2,5), ylim=c(0,0.5),
 axis( side=1, labels=F, lwd.ticks=0 )
 axis( side=1, at=0, tick=F, label='0' )
 axis( side=1, at=c(0, dd[1]), tick=F, 
[…]

### 2024-12-18 _tex/paper2.tex
Built site for gh-pages

@@
   authoryear,
-  preprint,
+  review,
   1p]{elsarticle}
@@
 \usepackage{amsmath,amssymb}
-\usepackage{setspace}
 \usepackage{iftex}
@@
 
-\setstretch{2}
 \section{Introduction}\label{sec-introduction}
@@
 B is positioned further along the continuum than that of Text A
-\((S_{B} > S_{A})\), suggesting that Text B exhibits higher quality.
+\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
 Additionally, the figure highlights that Text B has a broader
@@
 Figure~\ref{fig-discriminal_difference} illustrates the distribution of
-the discriminal difference for the hypothetical texts depicted in panel
+the discriminal difference for the hypothetical texts depicted in
 Figure~\ref{fig-discriminal_process}. The figure indicates that the
@@
 difference between their modal discriminal processes
-\((S_{B} - S_{A} > 0)\) and the probability area where the discriminal
+\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
 difference distinctly favors Text B over Text A, represented by the
@@
 
-Thurstone observed that the general form of the theory, outlined in
-Section~\ref{sec-thurstone_theory}, created a trait scaling problem. The
-model required estimating more ``unknown'' parameters than the available
-pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To address this
-issue and facilitate the practical application of the theory, he
-developed five cases derived from this general form, each case
-progressively incorporating additional simplifying assumptions into the
-model.
+Thurstone noted that the general form of the theory, as delineated in
+Section~\ref{sec-thurstone_theory}, gave rise to a trait scaling
+problem. The model required estimating more ``unknown'' parameters than
+the available pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To
+address this issue and facilitate the practical implementation of the
+theory, he developed five cases derived from this general form, each
+case progressively incorporated additional simplifying assumptions into
+the model.
 
-In Case I, Thurstone assumed that pairs of stimuli maintained a constant
-correlation across all comparisons. In Case II, he allowed multiple
-judges to make comparisons instead of restricting evaluations to a
-single judge. In Case III, he introduced the assumption of zero
-correlation between stimuli. In Case IV, he assumed stimuli exhibited
-similar dispersions. Finally, in Case V, he replaced this assumption
-with the condition that stimuli had equal discriminal dispersions.
-Table~\ref{tbl-thurstone_cases} summarizes the assumptions of the
-general form and the five cases. For an in-depth discussion of these
-cases and their progression, refer to \citet{Thurstone_1927b} and
+In Case I, Thurstone postulated that pairs of stimuli would maintain a
+constant correlation across all comparisons. In Case II, he allowed
+multiple judges to undertake comparisons instead of confining
+evaluations to a single judge. In Case III, he posited that there was no
+correlation between stimuli. In Case IV, he assumed that the stimuli
+exhibited similar dispersions. Finally, in Case V, he replaced this
+assumption with the condition that stimuli had equal discriminal
+dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
+of the general form and the five cases. For a detailed discussion of
+these cases and their progression, refer to \citet{Thurstone_1927b} and
 \citet[pp.~248--253]{Bramley_2008}.
@@
 \citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
-assumptions of Case V, with one key difference: while Case V assumes a
-Normal distribution for the stimuli's discriminal processes, the BTL
-model uses the more mathematically tractable Logistic distribution
-\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
+assumptions of Case V, with one notable distinction: whereas Case V
+assumes a Normal distribution for the stimuli's discriminal processes,
+the BTL model uses the more mathematically tractable Logistic
+distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
 Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
 the model's estimation or interpretation, as the Normal and Logistic
-distributions share similar statistical properties, differing only by a
-scaling factor of approximately \(1.7\)
+distributions exhibit analogous statistical properties, differing only
+by a scaling factor of approximately \(1.7\)
 \citep[pp.~16]{vanderLinden_et_al_2017_I}.
@@
 untested application, stating that its use ``should not be made without
-(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
-that some assumptions could prove problematic when researchers assess
-complex traits or heterogeneous stimuli
+(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}. Furthermore,
+he acknowledged that some assumptions could prove problematic when
+researchers assess complex traits or heterogeneous stimuli
 \citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
@@
 assumptions of Case V and, by extension, of the BTL model may not
-consistently hold in theory or practice: the assumption of equal
+consistently hold in theory or practice, namely the assumption of equal
 dispersion and zero correlation between stimuli.
@@
 direct influence on the outcome of pairwise comparisons.
-Figure~\ref{fig-dispersion} illustrates this idea using a thought
-experiment in which a researcher can observe the discriminal processes
-for the texts shown in Figure~\ref{fig-discriminal_process}.
-Furthermore, the figure assumes that the discriminal dispersion for Text
-A remains constant and that the texts are uncorrelated \((\rho=0)\). The
-figure reveals that an increase in the uncertainty associated with the
-perception of Text B in comparison to Text A,
-\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
-discriminal difference. This broadening affects the probability area
-where the discriminal difference distinctly favors Text B over Text A,
-expressed as \(P(B > A)\), ultimately influencing the comparison
-outcome. Additionally, the figure reveals that when the discriminal
-dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
-discriminal difference is more likely to favor Text B over Text A
-(shaded gray area), compared to situations where their dispersions
-differ.
+Figure~\ref{fig-dispersion} presents a thought experiment to illustrate
+this idea. In this experiment, a researcher can observe the discriminal
+processes for the texts depicted in
+Figure~\ref{fig-discriminal_process}. Furthermore, the figure assumes
+that the discriminal dispersion for Text A remains constant and that the
+texts are uncorrelated \((\rho=0)\). The figure reveals that an increase
+in the uncertainty associated with the perception of Text B in
+comparison to Text A, \((\sigma_{B}-\sigma_{A})\), broadens the
+distribution of their discriminal difference. This broadening affects
+the probability area where the discriminal difference distinctly favors
+Text B over Text A, expressed as \(P(B > A)\), ultimately influencing
+the comparison outcome. Additionally, the figure reveals that when the
+discriminal dispersions of the texts are equal
+\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
+to favor Text B over Text A (shaded gray area), compared to situations
+where their dispersions differ.
 
@@
 model to infer the discriminal difference between the stimuli and their
-respective discriminal processes \citep[pp.~373]{Thurstone_1927a}. For
-instance, when researchers observe a large sample of outcomes favoring
-Text B over Text A and correctly assume equal dispersions between the
-texts, the BTL model estimates a discriminal difference distribution
-that accurately represents the ``true'' discriminal difference of the
-texts. This scenario is illustrated in Figure~\ref{fig-dispersion},
-where the model's discriminal difference distribution aligns with the
-``true'' distribution, represented by the thick continuous line
-corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
-estimated discriminal difference ensures reliable estimates of the
-texts' discriminal processes {[}(citation
-needed?)\{style=``color:red;''\}. Therefore, it is reasonable to assume
-that the outcome's ability to reflect the ``true'' differences between
-stimuli largely depends on the validity of the model's assumptions,
-particularly the assumption of equal dispersions.
+respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
+Therefore, it is reasonable to assume that the outcome's ability to
+reflect the ``true'' differences between stimuli largely depends on the
+validity of the model's assumptions, particularly the assumption of
+equal dispersions. For instance, when researchers observe a sample of
+outcomes favoring Text B over Text A and correctly assume equal
+dispersions between the texts, the BTL model estimates a discriminal
+difference distribution that accurately represents the ``true''
+discriminal difference of the texts. This scenario is illustrated in
+Figure~\ref{fig-dispersion}, where the model's discriminal difference
+distribution aligns with the ``true'' distribution, represented by the
+thick continuous line corresponding to \(\sigma_{B}-\sigma_{A}=0\). The
+accuracy of these discriminal difference ensures reliable estimates for
+the texts' discriminal processes {(citation needed?)}.
 
-However, Thurstone contended that the assumption of equal dispersions
-may not hold when researchers assess complex traits or heterogeneous
-stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
-can introduce judgment discrepancies due to their unique features
+However, Thurstone argued that the assumption of equal dispersions may
+not be applicable when researchers assess complex traits or
+heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
+and stimuli can introduce judgment discrepancies due to their unique
+features
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
-Indeed, evidence of this violation may already exist in the CJ
-literature as misfit statistics, which measure judgment discrepancies
-associated with specific stimuli
+Indeed, evidence of this violation may already be present in the CJ
+literature in the form of misfit statistics, which measure judgment
+discrepancies associated with specific stimuli
 \citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
@@
 discriminal processes may also exhibit more variation than that of other
-texts. A similar reasoning applies to ``misfit'' judges, whose
-evaluations deviate substantially from the shared consensus due to the
-unique characteristics of the stimuli or the judges themselves.
-Moreover, these ``misfit'' judges and their deviations can introduce
-additional statistical and measurement issues, which we discuss in
-Section~\ref{sec-theory-issue1b}.
+texts. A similar line of reasoning applies to the concept of ``misfit''
+judges, whose evaluations deviate substantially from the shared
+consensus due to the unique characteristics of the stimuli or the judges
+themselves. These ``misfit'' judges and their associated deviations can
+give rise to additional statistical and measurement issues, which we
+discuss in more detail in Section~\ref{sec-theory-issue1b}.
 
-Therefore, incorrectly assuming equal dispersions between stimuli is not
-without harm, as it can cause the BTL model to introduce significant
-statistical and measurement issues. For instance, the model may
-overestimate how accurately the outcome reflects the ``true''
-discriminal differences between stimuli. This overestimation can lead
-researchers to draw spurious conclusions about these differences
-\citep[pp.~370]{McElreath_2020} and, by extension, of the stimuli
-underlying discriminal processes. Figure~\ref{fig-dispersion}
+Nevertheless, model misspecification, in the form of an erroneous
+assumption of equal dispersions between stimuli, can give rise to
+significant statistical and measurement issues. For instance, the model
+may overestimate the degree to which the outcome accurately reflects the
+``true'' discriminal differences between stimuli. This overestimation
+can result in researchers drawing spurious conclusions about these
+differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
+underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
 illustrates this issue when the model's discriminal difference
@@
 \(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
-follows a discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
-Moreover, if researchers recognize that misfit statistics highlight
-critical differences in dispersions, the usual CJ practice of excluding
-stimuli based on these statistics
+follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
+
+Additionally, if researchers recognize that misfit statistics highlight
+these critical differences in dispersions, the conventional CJ practice
+of excluding stimuli based on these statistics
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
-can unintentionally discard valuable information. This exclusion can
+can unintentionally discard valuable information. Such exclusions can
 introduce bias into trait estimates
@@
 
-Denoted by \(\rho\), the correlation measures the dependence of a
-judge's perception of the trait in one stimulus on his perception of the
-same trait in another. Like the discriminal dispersions, this
-correlation shapes the distribution of the discriminal difference and
-directly influences the outcomes of pairwise comparisons.
-Figure~\ref{fig-correlation} illustrates this concept, again assuming
-the researcher can observe the discriminal processes for the texts shown
-in Figure~\ref{fig-discriminal_process}. The figure also considers that
-the discriminal dispersions for both texts remain constant.
-
-Figure~\ref{fig-correlation} reveals that as the correlation between the
+The correlation, represented by the symbol \(\rho\), measures how much a
+judge's perception of a specific trait in one stimulus depends on their
+perception of the same trait in another. As with the discriminal
+dispersions, this correlation shapes the distribution of the discriminal
+difference, directly impacting the outcomes of pairwise comparisons.
+Figure~\ref{fig-correlation} presents a similar thought experiment as in
+Section~\ref{sec-theory-issue1a} to illustrate this idea. The
+illustration now assumes that the discriminal dispersions for both texts
+remain constant. The figure shows that as the correlation between the
 texts increases, the distribution of their discriminal difference
@@
 \((\rho=0)\), their discriminal difference is less likely to favor Text
-B over Text A (shaded gray area), compared to scenarios when the texts
+B over Text A (shaded gray area) compared to scenarios where the texts
 are highly correlated.
 
-However, in experimental practice, researchers typically follow this
-process in reverse. For example, when they observe a large sample of
-outcomes favoring Text B over Text A and correctly assume zero
-correlation between the texts, the BTL model estimates a discriminal
-difference distribution that accurately represents the ``true''
-discriminal difference of the texts. This scenario is illustrated with
+However, in practice, researchers approach this process in reverse. They
+begin by observing the sample of outcomes favoring Text B over Text A
+and then use the BTL model to estimate the discriminal difference and
+the discriminal processes of the stimuli. Given that the BTL model
+assumes independent discriminal processes across comparisons, if this
+assumption holds, then the model estimates a discriminal difference
+distribution that accurately reflects the ``true'' discriminal
+difference of the texts. This scenario is illustrated with
 Figure~\ref{fig-correlation} when the discriminal difference
@@
 represented by the thick continuous line corresponding to \(\rho=0\).
-The accuracy of this discriminal difference estimation, in turn, ensures
-reliable estimates for the discriminal process of the texts {[}(citation
-needed?)\{style=``color:red;''\}.
+Once more, the estimation accuracy of the discriminal difference ensures
+reliable estimates for the discriminal processes of the texts {(citation
+needed?)}.
 
-Notably, Thurstone's Case V and the BTL model assume independent
-discriminal processes across comparisons. Thurstone attributed this
-independence to the cancellation of potential judges' biases, driven by
-two opposing and equally weighted effects occurring during the pairwise
-comparisons \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978}
-mathematically demonstrated this cancellation using the BTL model under
-the assumption of discriminal processes with additive biases. However,
-it is easy to imagine at least two scenarios where the zero correlation
-assumption almost certainly does not hold: when the pairwise comparison
-involves multidimensional, complex traits with heterogeneous stimuli and
-when an additional hierarchical structure is relevant to the stimuli.
+Notably, Thurstone attributed the independence of stimuli to the
+cancellation of potential judges' biases. He argued that this
+cancellation resulted from two opposing and equally weighted effects
+occurring during pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
+\citet{Andrich_1978} provided a mathematical demonstration of this
+cancellation using the BTL model under the assumption of discriminal
+processes with additive biases. However, it is easy to imagine at least
+two scenarios in which the zero correlation assumption is almost
+certainly invalid: when the pairwise comparison involves
+multidimensional, complex traits with heterogeneous stimuli and when an
+additional hierarchical structure is relevant to the stimuli.
 
@@
 additional relevant characteristics, which are unlikely to be equally
-weighted or opposing, can unevenly influence judges' perceptions,
-creating biases in their judgments and, ultimately, introducing
-dependencies between stimuli
+weighted or opposing, can exert an uneven influence on judges'
+perceptions, creating biases in their judgments and, ultimately,
+introducing dependencies between stimuli
 \citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
 occur when a judge assessing the argumentative quality of a text places
-more weight on its grammatical accuracy than other judges, ultimately
+more weight on its grammatical accuracy than other judges, thereby
 favoring texts with fewer errors but weaker arguments. While direct
-evidence for this specific scenario is lacking, studies such as
+evidence for this particular scenario is lacking, studies such as
 \citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
-supporting the idea that the factors influencing pairwise comparisons
+supporting the notion that the factors influencing pairwise comparisons
 may not always cancel out.
@@
 dependencies between stimuli, a statistical phenomenon commonly known as
-clustering \citep{Everitt_et_al_2010}. Although the CJ literature
-acknowledges the presence of such hierarchical structures, the
-statistical handling of this extra source of dependency between stimuli
-has been inadequate. For example, when CJ data includes multiple samples
-of stimuli from the same individuals, researchers often rely on
-(average) estimated BTL scores to conduct subsequent analyses and tests
-at the individual hierarchical level
+clustering \citep{Everitt_et_al_2010}. Despite the CJ literature
+acknowledging the existence of such hierarchical structures, the
+statistical handling of this additional source of dependence between
+stimuli has been inadequate. For instance, when CJ data incorporates
+multiple samples of stimuli from the same individuals, researchers
+frequently rely on (average) estimated BTL scores to conduct subsequent
+analyses and tests at the individual hierarchical level
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
 However, this approach can introduce additional statistical and
-measurement issues, which we discuss in Section~\ref{sec-theory-issue2}.
+measurement issues, which we discuss in greater detail in
+Section~\ref{sec-theory-issue2}.
 
-{In any case, similar to Section~\ref{sec-theory-issue1a}, incorrectly
-assuming zero correlation between stimuli can lead the BTL model to
-introduce various statistical and measurement issues. For instance, the
-model could over- or underestimate the accuracy of the outcome in
-reflecting the ``true'' discriminal differences between stimuli. This
-over- underestimation may result in spurious inferences about these
-differences and, by extension, about the stimuli's discriminal processes
-\citep[pp.~341]{Hoyle_et_al_2023}. Figure~\ref{fig-correlation} also
-illustrates this scenario when the model's discriminal difference
+In any case, similar to Section~\ref{sec-theory-issue1a}, model
+misspecification due to an erroneous assumption of zero correlation
+between stimuli can lead to significant statistical and measurement
+issues. For instance, the model may over- or underestimate how
+accurately the outcome reflects the ``true'' discriminal differences
+between stimuli. Such inaccuracies can result in spurious inferences
+about these differences and, by extension, about the stimuli's
+discriminal processes. This scenario is illustrated in
+Figure~\ref{fig-correlation}, where the model's discriminal difference
 distribution aligns with the thick continuous line for \(\rho=0\), while
 the ``true'' discriminal difference follows any discontinuous line where
-\(\rho \neq 0\). This missaligment can be due to the overlook of
-additional relevant traits, such as judges' biases, which cause
-dimensional mismatches in the BTL model, artificially inflating the
-reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
-worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
-Furthermore, researchers who exclude judges based on misfit statistics
-can risk discarding valuable information, further biasing the trait's
-estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly,
-researchers who fail to account for hierarchical (grouping) structures
-can reduce the precision of model parameter estimates, which may amplify
-the overestimation of the trait's reliability
-\citep[pp.~482]{Hoyle_et_al_2023}.}
+\(\rho \neq 0\).
+
+This misspecification may arise from neglecting additional relevant
+traits, excluding judges based on misfit statistics, or ignoring
+hierarchical (grouping) structures. Neglecting relevant traits, such as
+judges' biases, can cause dimensional mismatches in the BTL model,
+artificially inflating the trait's reliability
+\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introducing bias into the
+trait's estimates \citep{Ackerman_1989}. Excluding judges based on
+misfit statistics risks discarding valuable information, which may
+further bias the trait's estimates
+\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
+hierarchical structures may reduce the precision of model parameter
+estimates, potentially amplifying the overestimation of the trait's
+reliability \citep[pp.~482]{Hoyle_et_al_2023}.
 
@@
 
-{In aggregate, researchers' inadequate handling of violations to the
+In aggregate, researchers' inadequate handling of violations to the
 assumptions of equal dispersion and zero correlation between stimuli,
-along with the apparent disconnect between CJ's approach to trait
-measurement and hypothesis testing, can undermine the reliability of the
-trait estimates and ultimately compromise its validity
+coupled with the apparent disconnect between CJ's approach to trait
+measurement and hypothesis testing, can potentially compromise the
+reliability of the trait estimates and, by extension, their validity
 \citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
-systematic and integrated approach to examining what happens when judges
-compare two stimuli could offer several statistical and measurement
-benefits, including addressing these issues.}
+systematic and integrated approach to examine the factors influencing CJ
+experiments could offer several statistical and measurement benefits,
+including the ability to address these issues.
 
@@
 
-This section presents a theoretical model for CJ that extends
-Thurstone's theory. The model systematically incorporates all factors
-involved when judges make pairwise comparisons. Additionally, the
-section develops the statistical translation of the theoretical model
-based on assumptions informed by the CJ theory.
+This section introduces a theoretical model for Comparative Judgment
+(CJ) that builds upon Thurstone's theory. The model systematically
+accounts for the factors influencing judges during CJ experiments. These
+theoretical and practical insights are then translated into a
+statistical model.
 
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.29\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}
-
-}
-
-\caption{\label{fig-CJ_TM_A1}Theoretical model A1\$}
-
-\end{figure}%
-
 The (latent) ``perceived'' discriminal processes for the stimuli
 directly determines their discriminal difference
+
+The (latent) ``true'' discriminal processes for the stimuli and the
+judges' biases directly determines their (latent) ``perceived''
+discriminal processes
+
+\paragraph{Figure 3}
 
@@
 
-\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}
+\includegraphics[width=0.2\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}
 
@@
 
-\caption{\label{fig-CJ_TM_A2}Theoretical model A2\$}
+\caption{\label{fig-CJ_TM_A1}}
 
@@
 
-The (latent) ``true'' discriminal processes for the stimuli and the
-judges' biases directly determines their (latent) ``perceived''
-discriminal processes
+\paragraph{Figure 4}
 
@@
 
-\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}
+\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}
 
@@
 
-\caption{\label{fig-CJ_TM_A3}Theoretical model A3\$}
+\caption{\label{fig-CJ_TM_A2}}
+
+\end{figure}%
+
+\paragraph{Figure 5}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}
+
+}
+
+\caption{\label{fig-CJ_TM_A3}}
+
+\end{figure}%
+
+\paragraph{Figure 6}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A4.png}
+
+}
+
+\caption{\label{fig-CJ_TM_A4}}
 
@@
 each judge, as in
+
+\paragraph{Figure 7}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B1.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B1}}
+
+\end{figure}%
+
+\paragraph{Figure 8}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.62\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B2.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B2}}
+
+\end{figure}%
+
+\paragraph{Figure 9}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B3.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B3}}
+
+\end{figure}%
+
+\paragraph{Figure 10}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B4.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B4}}
+
+\end{figure}%
+
+\paragraph{Figure 11}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B5.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B5}}
+
+\end{figure}%
+
+\paragraph{Figure 12}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B6.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B6}}
+
+\end{figure}%
+
+Considering the sampling process
+
+\paragraph{Figure 13}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B7.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B7}}
+
+\end{figure}%
+
+\paragraph{Figure 14}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B8.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B8}}
+
+\end{figure}%
+
+\paragraph{Figure 15}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B9.png}
+
+}
+
+\caption{\label{fig-CJ_TM_B9}}
+
+\end{figure}%
 
@@
 \textbf{AI-assisted technologies in the writing process:} The authors
-used ChatGPT, an AI language model, during the preparation of this work.
-They occasionally employed the tool to refine phrasing and optimize
-wording, ensuring appropriate language use and enhancing the
-manuscript's clarity and coherence. The authors take full responsibility
-for the final content of the publication.
+utilized a range of AI-based language tools throughout the preparation
+of this work. They occasionally employed the tools to refine phrasing
+and optimize wording, ensuring appropriate language use and enhancing
+the manuscript's clarity and coherence. The authors take full
+responsibility for the final content of the publication.
 

### 2024-12-20 paper2.qmd
section:model; process:write

@@
 The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait.
-The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp.
+The theory assumes that the DP for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp.
 266].
@@
 @fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts.
-The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
+The figure indicates that the mDP for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
 Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
@@
 373].
-Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.
+Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
+150], particularly the assumption of equal dispersions.
 For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
@@
 373] -->
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
 
@@
 <!-- 2. But this is not the scenario we observe -->
-However, in practice, researchers approach this process in reverse.
+Off course, in experimental practice, researchers approach this process in reverse.
 They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli.
@@
 
-This section introduces a theoretical model for Comparative Judgment (CJ) that builds upon Thurstone's theory.
-The model systematically accounts for the factors influencing judges during CJ experiments.
-These theoretical and practical insights are then translated into a statistical model.
+This section presents a theoretical model for CJ that extends Thurstone's theory.
+It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014; @Hernan_et_al_2020] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory.
+The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and the number of comparisons.
+In addition, the study uses Bayesian statistical methods [@McElreath_2020] to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
 
@@
 
-#### Figure 3
+#### Figure
+
+#### 3
 
@@
 
-#### Figure 4
+#### 4
 
@@
 
-#### Figure 5
+#### 5
 
@@
 
-#### Figure 6
+#### 6
 
@@
 
-#### Figure 7
+#### Figure
+
+#### 7
 
@@
 
-#### Figure 8
+#### 8
 
@@
 
-#### Figure 9
+#### 9
 
@@
 
-#### Figure 10
+#### 10
 
@@
 
-#### Figure 11
+#### 11
 
@@
 
-#### Figure 12
+#### 12
 
@@
 
-Considering the sampling process
+Considering the sampling and comparison mechanisms
 
@@
 
-#### Figure 13
+#### Figure
+
+#### 13
 
@@
 
-#### Figure 14
+#### 14
 
@@
 
-#### Figure 15
+#### 15
 
@@
 267] -->
+
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
 

### 2024-12-20 _tex/paper2.tex
Built site for gh-pages

@@
 stimulus exerts on the judge or, more simply, his perception of the
-stimulus trait. The theory assumes that the discriminal process for any
-given stimulus forms a Normal distribution along the trait continuum
+stimulus trait. The theory assumes that the DP for any given stimulus
+forms a Normal distribution along the trait continuum
 \citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
@@
 discriminal processes along a quality trait continuum for two written
-texts. The figure indicates that the modal discriminal process for Text
-B is positioned further along the continuum than that of Text A
-\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
-Additionally, the figure highlights that Text B has a broader
-distribution compared to Text A, which arises from its larger
-discriminal dispersion \((\sigma_{B} > \sigma_{A})\).
+texts. The figure indicates that the mDP for Text B is positioned
+further along the continuum than that of Text A \((T_{B} > T_{A})\),
+suggesting that Text B exhibits higher quality. Additionally, the figure
+highlights that Text B has a broader distribution compared to Text A,
+which arises from its larger discriminal dispersion
+\((\sigma_{B} > \sigma_{A})\).
 
@@
 reflect the ``true'' differences between stimuli largely depends on the
-validity of the model's assumptions, particularly the assumption of
-equal dispersions. For instance, when researchers observe a sample of
-outcomes favoring Text B over Text A and correctly assume equal
-dispersions between the texts, the BTL model estimates a discriminal
-difference distribution that accurately represents the ``true''
-discriminal difference of the texts. This scenario is illustrated in
-Figure~\ref{fig-dispersion}, where the model's discriminal difference
-distribution aligns with the ``true'' distribution, represented by the
-thick continuous line corresponding to \(\sigma_{B}-\sigma_{A}=0\). The
-accuracy of these discriminal difference ensures reliable estimates for
-the texts' discriminal processes {(citation needed?)}.
+validity of the model's assumptions \citep[pp.~150]{Kohler_et_al_2019},
+particularly the assumption of equal dispersions. For instance, when
+researchers observe a sample of outcomes favoring Text B over Text A and
+correctly assume equal dispersions between the texts, the BTL model
+estimates a discriminal difference distribution that accurately
+represents the ``true'' discriminal difference of the texts. This
+scenario is illustrated in Figure~\ref{fig-dispersion}, where the
+model's discriminal difference distribution aligns with the ``true''
+distribution, represented by the thick continuous line corresponding to
+\(\sigma_{B}-\sigma_{A}=0\). The accuracy of these discriminal
+difference ensures reliable estimates for the texts' discriminal
+processes {(citation needed?)}.
 
@@
 
-However, in practice, researchers approach this process in reverse. They
-begin by observing the sample of outcomes favoring Text B over Text A
-and then use the BTL model to estimate the discriminal difference and
-the discriminal processes of the stimuli. Given that the BTL model
-assumes independent discriminal processes across comparisons, if this
-assumption holds, then the model estimates a discriminal difference
-distribution that accurately reflects the ``true'' discriminal
-difference of the texts. This scenario is illustrated with
+Off course, in experimental practice, researchers approach this process
+in reverse. They begin by observing the sample of outcomes favoring Text
+B over Text A and then use the BTL model to estimate the discriminal
+difference and the discriminal processes of the stimuli. Given that the
+BTL model assumes independent discriminal processes across comparisons,
+if this assumption holds, then the model estimates a discriminal
+difference distribution that accurately reflects the ``true''
+discriminal difference of the texts. This scenario is illustrated with
 Figure~\ref{fig-correlation} when the discriminal difference
@@
 
-This section introduces a theoretical model for Comparative Judgment
-(CJ) that builds upon Thurstone's theory. The model systematically
-accounts for the factors influencing judges during CJ experiments. These
-theoretical and practical insights are then translated into a
-statistical model.
+This section presents a theoretical model for CJ that extends
+Thurstone's theory. It uses causal analysis
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014, Hernan_et_al_2020}
+and, in particular, directed acyclic graphs (DAGs)
+\citep{Gross_et_al_2018, Neal_2020}, to articulate the core theoretical
+principles of CJ theory. The model also incorporates several practical
+factors that influence judges in CJ experiments, such as the selection
+of judges, stimuli, and the number of comparisons. In addition, the
+study uses Bayesian statistical methods \citep{McElreath_2020} to
+translate these theoretical and practical elements into a statistical
+model that facilitates the analysis of pairwise comparison data.
 
@@
 
-\paragraph{Figure 3}
+\paragraph{Figure}
+
+\paragraph{3}
 
@@
 
-\paragraph{Figure 4}
+\paragraph{4}
 
@@
 
-\paragraph{Figure 5}
+\paragraph{5}
 
@@
 
-\paragraph{Figure 6}
+\paragraph{6}
 
@@
 
-\paragraph{Figure 7}
+\paragraph{Figure}
+
+\paragraph{7}
 
@@
 
-\paragraph{Figure 8}
+\paragraph{8}
 
@@
 
-\paragraph{Figure 9}
+\paragraph{9}
 
@@
 
-\paragraph{Figure 10}
+\paragraph{10}
 
@@
 
-\paragraph{Figure 11}
+\paragraph{11}
 
@@
 
-\paragraph{Figure 12}
+\paragraph{12}
 
@@
 
-Considering the sampling process
+Considering the sampling and comparison mechanisms
 
-\paragraph{Figure 13}
+\paragraph{Figure}
+
+\paragraph{13}
 
@@
 
-\paragraph{Figure 14}
+\paragraph{14}
 
@@
 
-\paragraph{Figure 15}
+\paragraph{15}
 

### 2025-01-06 paper2.qmd
section:model; process:write,figures

@@
 The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait.
-The theory assumes that the DP for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp.
+The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp.
 266].
@@
 @fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts.
-The figure indicates that the mDP for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
+The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
 Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
@@
 <!-- 3. Thurstone's Theory of CJ: discriminal difference -->
-However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces *the law of comparative judgment*.
+However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
 This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp.
@@
 This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons.
-Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as *the discriminal difference*, determines the observed dichotomous outcome.
+Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome.
 Moreover, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
@@
 The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
-This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
+This conclusion is supported by two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
 As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
@@
 <!-- 1. What is the problem with the general form -->
-Thurstone noted that the general form of the theory, as delineated in @sec-thurstone_theory, gave rise to a trait scaling problem.
+Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a problem of trait scaling.
 The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp.
@@
 <!-- 3. The Case V and its simpler statistical representation: the BTL model -->
-Despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp.
+Notably, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp.
 253; @Kelly_et_al_2022, pp.
@@
 
+::: {#fig-caseV_issues layout-ncol=2}
+
+![Discrepancies in the dispersions of stimuli](/images/png/dispersion.png){#fig-dispersion width=100%}
+
+![Correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}
+
+The effect of dispersion discrepancies and stimulus correlation on the distribution of the discriminal difference.
+:::
+
 <!-- 2. But this is not the scenario we observe -->
@@
 373].
-Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
+Therefore, the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
 150], particularly the assumption of equal dispersions.
@@
 <!-- 4. So, what happens when the assumption does not hold? -->
-Nevertheless, model misspecification, in the form of an erroneous assumption of equal dispersions between stimuli, can give rise to significant statistical and measurement issues.
+Thus, model misspecification, in the form of an erroneous assumption of equal dispersions between stimuli, can give rise to significant statistical and measurement issues.
 For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli.
@@
 370] and, by extension, about the underlying discriminal processes of stimuli.
-@fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-
+@fig-dispersion also illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
 Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information.
@@
 
-::: {#fig-caseV_issues layout-ncol=2}
-
-![Discrepancies in the dispersions of stimuli](/images/png/dispersion.png){#fig-dispersion width=100%}
-
-![Correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}
-
-The effect of dispersion discrepancies and stimulus correlation on the distribution of the discriminal difference.
-:::
-
 ### The assumption of zero correlation between stimuli {#sec-theory-issue1b}
@@
 The illustration now assumes that the discriminal dispersions for both texts remain constant.
-The figure shows that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
+The figure reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
 This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
@@
 Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts.
-This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$.
+This scenario is also illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$.
 Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.
@@
 Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes.
-This scenario is illustrated in @fig-correlation, where the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
+This scenario is also illustrated by @fig-correlation, when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
 
-This misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures.
+The misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures.
 Neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp.
@@
 This section presents a theoretical model for CJ that extends Thurstone's theory.
-It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014; @Hernan_et_al_2020] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory.
-The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and the number of comparisons.
+It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory.
+The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and comparisons.
 In addition, the study uses Bayesian statistical methods [@McElreath_2020] to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
@@
 
-::: {.panel-tabset group="Figure"}
+<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->
 
-#### Figure
+<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->
 
-#### 3
+<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->
 
-![](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=20%}
-
-#### 4
-
-![](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=30%}
-
-#### 5
-
-![](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=42%}
-
-#### 6
-
-![](/images/png/CJ_TM_A4.png){#fig-CJ_TM_A4 width=42%}
-
-:::
+![](/images/png/CJ_TM_04.png){#fig-CJ_TM_04 width=45%}
 
@@
 
-::: panel-tabset
+<!-- ![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%} -->
 
-#### Figure
+<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->
 
-#### 7
+<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->
 
-![](/images/png/CJ_TM_B1.png){#fig-CJ_TM_B1 width=42%}
+<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->
 
-#### 8
+<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->
 
-![](/images/png/CJ_TM_B2.png){#fig-CJ_TM_B2 width=62%}
+![](/images/png/CJ_TM_10.png){#fig-CJ_TM_10 width=68%}
 
-#### 9
+Considering the sampling mechanism
 
-![](/images/png/CJ_TM_B3.png){#fig-CJ_TM_B3 width=72%}
+<!-- ![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%} -->
 
-#### 10
+![](/images/png/CJ_TM_12.png){#fig-CJ_TM_12 width=80%}
 
-![](/images/png/CJ_TM_B4.png){#fig-CJ_TM_B4 width=72%}
+Considering comparison mechanisms
 
-#### 11
+<!-- ![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%} -->
 
-![](/images/png/CJ_TM_B5.png){#fig-CJ_TM_B5 width=72%}
+![](/images/png/CJ_TM_14.png){#fig-CJ_TM_14 width=90%}
 
-#### 12
-
-![](/images/png/CJ_TM_B6.png){#fig-CJ_TM_B6 width=72%}
-
-:::
-
-Considering the sampling and comparison mechanisms
-
-::: panel-tabset
-
-#### Figure
-
-#### 13
-
-![](/images/png/CJ_TM_B7.png){#fig-CJ_TM_B7 width=90%}
-
-#### 14
-
-![](/images/png/CJ_TM_B8.png){#fig-CJ_TM_B8 width=100%}
-
-#### 15
-
-![](/images/png/CJ_TM_B9.png){#fig-CJ_TM_B9 width=100%}
-
-:::
+<!-- ![](/images/png/CJ_TM_15.png){#fig-CJ_TM_15 width=90%} -->
 

### 2025-01-06 _tex/paper2.tex
Built site for gh-pages

@@
 stimulus exerts on the judge or, more simply, his perception of the
-stimulus trait. The theory assumes that the DP for any given stimulus
-forms a Normal distribution along the trait continuum
+stimulus trait. The theory assumes that the discriminal process for any
+given stimulus forms a Normal distribution along the trait continuum
 \citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
@@
 discriminal processes along a quality trait continuum for two written
-texts. The figure indicates that the mDP for Text B is positioned
-further along the continuum than that of Text A \((T_{B} > T_{A})\),
-suggesting that Text B exhibits higher quality. Additionally, the figure
-highlights that Text B has a broader distribution compared to Text A,
-which arises from its larger discriminal dispersion
-\((\sigma_{B} > \sigma_{A})\).
+texts. The figure indicates that the modal discriminal process for Text
+B is positioned further along the continuum than that of Text A
+\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
+Additionally, the figure highlights that Text B has a broader
+distribution compared to Text A, which arises from its larger
+discriminal dispersion \((\sigma_{B} > \sigma_{A})\).
 
@@
 However, since the individual discriminal processes of the stimuli are
-not directly observable, the theory introduces \emph{the law of
+not directly observable, the theory introduces the \emph{law of
 comparative judgment}. This law posits that in pairwise comparisons, a
@@
 assumes that the difference between the underlying discriminal processes
-of the stimuli, referred to as \emph{the discriminal difference},
+of the stimuli, referred to as the \emph{discriminal difference},
 determines the observed dichotomous outcome. Moreover, the theory
@@
 judge perceives Text B as having significantly higher quality than Text
-A. This conclusion rests on two key observations: the positive
+A. This conclusion is supported by two key observations: the positive
 difference between their modal discriminal processes
@@
 
-Thurstone noted that the general form of the theory, as delineated in
-Section~\ref{sec-thurstone_theory}, gave rise to a trait scaling
-problem. The model required estimating more ``unknown'' parameters than
-the available pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To
-address this issue and facilitate the practical implementation of the
-theory, he developed five cases derived from this general form, each
-case progressively incorporated additional simplifying assumptions into
-the model.
+Thurstone noted from the outset that the general form of the theory, as
+outlined in Section~\ref{sec-thurstone_theory}, gave rise to a problem
+of trait scaling. The model required estimating more ``unknown''
+parameters than the available pairwise comparisons
+\citep[pp.~267]{Thurstone_1927b}. To address this issue and facilitate
+the practical implementation of the theory, he developed five cases
+derived from this general form, each case progressively incorporated
+additional simplifying assumptions into the model.
 
@@
 
-Despite relying on the most extensive set of simplifying assumptions
+Notably, despite relying on the most extensive set of simplifying
+assumptions
 \citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
@@
 
-In experimental practice, however, this process occurs in reverse.
-Researchers first observe the comparison outcome and then use the BTL
-model to infer the discriminal difference between the stimuli and their
-respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
-Therefore, it is reasonable to assume that the outcome's ability to
-reflect the ``true'' differences between stimuli largely depends on the
-validity of the model's assumptions \citep[pp.~150]{Kohler_et_al_2019},
-particularly the assumption of equal dispersions. For instance, when
-researchers observe a sample of outcomes favoring Text B over Text A and
-correctly assume equal dispersions between the texts, the BTL model
-estimates a discriminal difference distribution that accurately
-represents the ``true'' discriminal difference of the texts. This
-scenario is illustrated in Figure~\ref{fig-dispersion}, where the
-model's discriminal difference distribution aligns with the ``true''
-distribution, represented by the thick continuous line corresponding to
-\(\sigma_{B}-\sigma_{A}=0\). The accuracy of these discriminal
-difference ensures reliable estimates for the texts' discriminal
-processes {(citation needed?)}.
-
-However, Thurstone argued that the assumption of equal dispersions may
-not be applicable when researchers assess complex traits or
-heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
-and stimuli can introduce judgment discrepancies due to their unique
-features
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
-Indeed, evidence of this violation may already be present in the CJ
-literature in the form of misfit statistics, which measure judgment
-discrepancies associated with specific stimuli
-\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
-For example, labeling texts as ``misfits'' indicates that comparisons
-involving these texts result in more judgment discrepancies than those
-involving other texts
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
-These discrepancies, in turn, suggest that the discriminal differences
-for ``misfit'' texts have broader distributions, indicating that their
-discriminal processes may also exhibit more variation than that of other
-texts. A similar line of reasoning applies to the concept of ``misfit''
-judges, whose evaluations deviate substantially from the shared
-consensus due to the unique characteristics of the stimuli or the judges
-themselves. These ``misfit'' judges and their associated deviations can
-give rise to additional statistical and measurement issues, which we
-discuss in more detail in Section~\ref{sec-theory-issue1b}.
-
-Nevertheless, model misspecification, in the form of an erroneous
-assumption of equal dispersions between stimuli, can give rise to
-significant statistical and measurement issues. For instance, the model
-may overestimate the degree to which the outcome accurately reflects the
-``true'' discriminal differences between stimuli. This overestimation
-can result in researchers drawing spurious conclusions about these
-differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
-underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
-illustrates this issue when the model's discriminal difference
-distribution aligns with the thick continuous line for
-\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
-follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
-
-Additionally, if researchers recognize that misfit statistics highlight
-these critical differences in dispersions, the conventional CJ practice
-of excluding stimuli based on these statistics
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
-can unintentionally discard valuable information. Such exclusions can
-introduce bias into trait estimates
-\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
-magnitude of these biases are often unpredictable, as they depend on
-which stimuli are excluded from the analysis.
-
 \begin{figure}
@@
 
+In experimental practice, however, this process occurs in reverse.
+Researchers first observe the comparison outcome and then use the BTL
+model to infer the discriminal difference between the stimuli and their
+respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
+Therefore, the outcome's ability to reflect the ``true'' differences
+between stimuli largely depends on the validity of the model's
+assumptions \citep[pp.~150]{Kohler_et_al_2019}, particularly the
+assumption of equal dispersions. For instance, when researchers observe
+a sample of outcomes favoring Text B over Text A and correctly assume
+equal dispersions between the texts, the BTL model estimates a
+discriminal difference distribution that accurately represents the
+``true'' discriminal difference of the texts. This scenario is
+illustrated in Figure~\ref{fig-dispersion}, where the model's
+discriminal difference distribution aligns with the ``true''
+distribution, represented by the thick continuous line corresponding to
+\(\sigma_{B}-\sigma_{A}=0\). The accuracy of these discriminal
+difference ensures reliable estimates for the texts' discriminal
+processes {(citation needed?)}.
+
+However, Thurstone argued that the assumption of equal dispersions may
+not be applicable when researchers assess complex traits or
+heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
+and stimuli can introduce judgment discrepancies due to their unique
+features
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
+Indeed, evidence of this violation may already be present in the CJ
+literature in the form of misfit statistics, which measure judgment
+discrepancies associated with specific stimuli
+\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
+For example, labeling texts as ``misfits'' indicates that comparisons
+involving these texts result in more judgment discrepancies than those
+involving other texts
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
+These discrepancies, in turn, suggest that the discriminal differences
+for ``misfit'' texts have broader distributions, indicating that their
+discriminal processes may also exhibit more variation than that of other
+texts. A similar line of reasoning applies to the concept of ``misfit''
+judges, whose evaluations deviate substantially from the shared
+consensus due to the unique characteristics of the stimuli or the judges
+themselves. These ``misfit'' judges and their associated deviations can
+give rise to additional statistical and measurement issues, which we
+discuss in more detail in Section~\ref{sec-theory-issue1b}.
+
+Thus, model misspecification, in the form of an erroneous assumption of
+equal dispersions between stimuli, can give rise to significant
+statistical and measurement issues. For instance, the model may
+overestimate the degree to which the outcome accurately reflects the
+``true'' discriminal differences between stimuli. This overestimation
+can result in researchers drawing spurious conclusions about these
+differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
+underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
+also illustrates this issue when the model's discriminal difference
+distribution aligns with the thick continuous line for
+\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
+follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
+Additionally, if researchers recognize that misfit statistics highlight
+these critical differences in dispersions, the conventional CJ practice
+of excluding stimuli based on these statistics
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
+can unintentionally discard valuable information. Such exclusions can
+introduce bias into trait estimates
+\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
+magnitude of these biases are often unpredictable, as they depend on
+which stimuli are excluded from the analysis.
+
 \subsubsection{The assumption of zero correlation between
@@
 illustration now assumes that the discriminal dispersions for both texts
-remain constant. The figure shows that as the correlation between the
+remain constant. The figure reveals that as the correlation between the
 texts increases, the distribution of their discriminal difference
@@
 difference distribution that accurately reflects the ``true''
-discriminal difference of the texts. This scenario is illustrated with
-Figure~\ref{fig-correlation} when the discriminal difference
+discriminal difference of the texts. This scenario is also illustrated
+in Figure~\ref{fig-correlation} when the discriminal difference
 distribution of the model aligns with the ``true'' distribution,
@@
 about these differences and, by extension, about the stimuli's
-discriminal processes. This scenario is illustrated in
-Figure~\ref{fig-correlation}, where the model's discriminal difference
+discriminal processes. This scenario is also illustrated by
+Figure~\ref{fig-correlation}, when the model's discriminal difference
 distribution aligns with the thick continuous line for \(\rho=0\), while
@@
 
-This misspecification may arise from neglecting additional relevant
+The misspecification may arise from neglecting additional relevant
 traits, excluding judges based on misfit statistics, or ignoring
@@
 Thurstone's theory. It uses causal analysis
-\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014, Hernan_et_al_2020}
-and, in particular, directed acyclic graphs (DAGs)
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014} and, in
+particular, directed acyclic graphs (DAGs)
 \citep{Gross_et_al_2018, Neal_2020}, to articulate the core theoretical
@@
 factors that influence judges in CJ experiments, such as the selection
-of judges, stimuli, and the number of comparisons. In addition, the
-study uses Bayesian statistical methods \citep{McElreath_2020} to
-translate these theoretical and practical elements into a statistical
-model that facilitates the analysis of pairwise comparison data.
+of judges, stimuli, and comparisons. In addition, the study uses
+Bayesian statistical methods \citep{McElreath_2020} to translate these
+theoretical and practical elements into a statistical model that
+facilitates the analysis of pairwise comparison data.
 
@@
 
-\paragraph{Figure}
-
-\paragraph{3}
-
 \begin{figure}
@@
 
-\includegraphics[width=0.2\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A1.png}
+\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
 
@@
 
-\caption{\label{fig-CJ_TM_A1}}
-
-\end{figure}%
-
-\paragraph{4}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A2.png}
-
-}
-
-\caption{\label{fig-CJ_TM_A2}}
-
-\end{figure}%
-
-\paragraph{5}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A3.png}
-
-}
-
-\caption{\label{fig-CJ_TM_A3}}
-
-\end{figure}%
-
-\paragraph{6}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_A4.png}
-
-}
-
-\caption{\label{fig-CJ_TM_A4}}
+\caption{\label{fig-CJ_TM_04}}
 
@@
 
-\paragraph{Figure}
+\begin{figure}
 
-\paragraph{7}
+\centering{
+
+\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_10.png}
+
+}
+
+\caption{\label{fig-CJ_TM_10}}
+
+\end{figure}%
+
+Considering the sampling mechanism
 
@@
 
-\includegraphics[width=0.42\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B1.png}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}
 
@@
 
-\caption{\label{fig-CJ_TM_B1}}
+\caption{\label{fig-CJ_TM_12}}
 
@@
 
-\paragraph{8}
+Considering comparison mechanisms
 
@@
 
-\includegraphics[width=0.62\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B2.png}
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}
 
@@
 
-\caption{\label{fig-CJ_TM_B2}}
-
-\end{figure}%
-
-\paragraph{9}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B3.png}
-
-}
-
-\caption{\label{fig-CJ_TM_B3}}
-
-\end{figure}%
-
-\paragraph{10}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B4.png}
-
-}
-
-\caption{\label{fig-CJ_TM_B4}}
-
-\end{figure}%
-
-\paragraph{11}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B5.png}
-
-}
-
-\caption{\label{fig-CJ_TM_B5}}
-
-\end{figure}%
-
-\paragraph{12}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B6.png}
-
-}
-
-\caption{\label{fig-CJ_TM_B6}}
-
-\end{figure}%
-
-Considering the sampling and comparison mechanisms
-
-\paragraph{Figure}
-
-\paragraph{13}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B7.png}
-
-}
-
-\caption{\label{fig-CJ_TM_B7}}
-
-\end{figure}%
-
-\paragraph{14}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B8.png}
-
-}
-
-\caption{\label{fig-CJ_TM_B8}}
-
-\end{figure}%
-
-\paragraph{15}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_B9.png}
-
-}
-
-\caption{\label{fig-CJ_TM_B9}}
+\caption{\label{fig-CJ_TM_14}}
 

### 2025-01-09 images/text/mdag_bb2.tex
section:appendix; process:write

--- 2024-08-02 images/tex/mdag_bb4.tex
@@
 
-    % dag_bb3
+    % dag_bb2
     \node at (-2,0) {$e_{X}$};
     \node at (-1,0) {$X$};
-    \node at (-1,0.7) {$e_{Z}$};
-    \node at (0,0.7) {$Z$};
-    \node at (1,0) {$Y$};
-    \node at (2,0) {$e_{Y}$};
+    \node at (0,0) {$Y$};
+    \node at (1,0) {$e_{Y}$};
 
-    \draw[{Circle[open]}-{latex}](-2,-0.3) to (-1,-0.3); % eX -> X (circle)
-    \draw[-{latex}{Circle}](0,0.45) to (-1,-0.3); % X -> Z
-    \draw[{Circle[open]}-{latex}](-1,0.47) to (0,0.47); % eZ -> Z (circle)
-    \draw[{Circle}-{latex}{Circle}](0,0.5) to (1,-0.3); % Z -> Y
-    \draw[{Circle[open]}-{latex}](2,-0.3) to (1,-0.3); % eY -> Y (circle)
+    \draw[{Circle[open]}-{latex}](-2,-0.3) to (-1.05,-0.3); % eX -> X (circle)
+    \draw[{Circle[open]}-{latex}](1,-0.3) to (0.05,-0.3); % eY -> Y (circle)
+    \draw[{Circle}-{latex}{Circle}](-1.05,-0.3) to (0.05,-0.3); % X -> Y (circle)
 

### 2025-01-09 paper2.qmd
section:appendix; process:write

@@
 keywords:
-  - Causal analysis
-  - Directed Acyclic Graphs
-  - Bayesian statistical methods
-  - Thurstonian model
-  - Comparative judgement
-  - Probability
-  - Statistical modeling
+  - causal inference
+  - directed acyclic graphs
+  - structural causal models
+  - bayesian statistical methods
+  - thurstonian model
+  - comparative judgement
+  - probability
+  - statistical modeling
 abstract: |
@@
 <!-- 1. Overview of CJ method in assessing traits -->
-In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b].
-Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level.
-For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across different stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b].
+Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to have a higher trait level.
+For example, when assessing writing quality, judges compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade.
-These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability.
-Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
-Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-Meanwhile, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023], while research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability.
+Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
+Meanwhile, research on the validity of CJ scores indicates their capacity to represent the traits under measurement accurately [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
+Moreover, research on CJ's practical applicability highlights its versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 <!-- 3. But what critical issues remain in CJ Research?, what is the organization of the study? -->
-Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed.
-The present study primarily focuses on two: the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data, and the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-The following sections begin with a brief overview of Thurstone's theory and a detailed examination of these issues.
+Nevertheless, despite the increasing number of CJ studies, the prevalence of unsystematic and fragmented research approaches has left several critical issues unaddressed.
+The present study primarily focuses on two issues: the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data and the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+The following sections begin with a brief overview of Thurstone's theory followed by a detailed examination of these issues.
 Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address the two concerns simultaneously.
@@
 <!-- 1. Thurstone's Theory of CJ: discriminal process -->
-In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp.
+In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b, pp.
 267].
@@
 <!-- 3. Thurstone's Theory of CJ: discriminal difference -->
-However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
+However, given that the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
 This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp.
@@
 Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome.
-Moreover, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
+Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
 In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.
@@
 
-This section examines the two critical issues in the CJ literature that serve as the primary focus of this study.
-The first is the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second is the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+This section examines the two critical issues in the CJ literature that serve as the primary focus of the present study.
+The first is related to the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+The second concern with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
 
@@
 
+<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
+Thurstone clearly did not expect this to hold for any but the most simple stimuli.
+The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
+-->
+
 <!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
@@
 376] -->
-
-<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
-Thurstone clearly did not expect this to hold for any but the most simple stimuli.
-The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
--->
-
-<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
-[@Pollitt_2012a, pp.
-165] -->
-
-<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
-
-<!-- a misfit statistic.
-Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
-It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
-12] -->
 
@@
 
+<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
+[@Pollitt_2012a, pp.
+165] -->
+
+<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
+
+<!-- a misfit statistic.
+Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
+It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
+12] -->
+
 <!-- 4. So, what happens when the assumption does not hold? -->
@@
 The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.
+
+<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
 
@@
 
+<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
+One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
+In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
+Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
+-->
+
 <!-- 5. But what are the challenges when this assumption is violated? -->
@@
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is clear that, despite its limitations, the BTL model is commonly used as the measurement model in CJ assessments.
+Building on the previous section, it is clear that, despite its limitations, the BTL model is commonly used as a measurement model in CJ assessments.
 A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
-For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, when evaluating writing quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying level of writing quality (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 2].
-Consequently, adopting a more systematic and integrated approach to examine the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address these issues.
+Consequently, adopting a more systematic and integrated approach to handling these assumptions and examining the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address these issues.
 
@@
 
-# An updated theoretical and statistical model for CJ {#sec-theory}
+# Updating CJ's theoretical and statistical model {#sec-theory}
 
 This section presents a theoretical model for CJ that extends Thurstone's theory.
-It uses causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014] and, in particular, directed acyclic graphs (DAGs) [@Gross_et_al_2018; @Neal_2020], to articulate the core theoretical principles of CJ theory.
-The model also incorporates several practical factors that influence judges in CJ experiments, such as the selection of judges, stimuli, and comparisons.
-In addition, the study uses Bayesian statistical methods [@McElreath_2020] to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
+It uses causal inference and, in particular, the structural approach where structural causal models (SCMs) and directed acyclic graphs (DAGs) articulate the core theoretical principles of CJ theory.
+The model also incorporates several assessment design features that influence CJ experiments, such as the selection of judges, stimuli, and comparisons.
+In addition, the study uses Bayesian statistical methods to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
 
@@
 
+### The population model {#sec-theory-theoretical_P}
 <!-- []{style="color:red;"} -->
@@
 <!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->
+
+structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]
+
+causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]
+
+<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
+
+<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+4].
+-->
+
+<!-- The main reason to focus on identification as a first step is that if a parameter cannot be identified using “infinite” and error free data, we certainly cannot learn anything about it using finite data.1 So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+5] -->
+
+Assuming population data or more commonly known as census data, we ...
 
@@
 
+<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
+
+### The assessment design features {#sec-theory-theoretical_AD}
+
 Considering the sampling mechanism
@@
 ![](/images/png/CJ_TM_12.png){#fig-CJ_TM_12 width=80%}
+
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
 
@@
 
-<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->
-
-<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
-One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
-In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
-Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
--->
-
-<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
-
-<!-- Notably, the theory assumes the judge compares each pair of stimuli a "sufficient number of times" to accurately calculate the relative separation probability [@Thurstone_1927b, pp. 267]. -->
-
-<!-- It is assumed that the single observer compares each pair of stimuli sufficient number of times so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
 267] -->
-
-<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
 
@@
 
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are [@Kohler_et_al_2019, pp.
+150] -->
+
 <!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
-
-<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->
-
-<!-- [@deAyala_2009, p.
-291-29] think about indeterminancies in MIRT.
-to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
-
-<!-- , as well as when eliminating data through ad hoc univariate procedures -->
-<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
 
@@
 <!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
+
+<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->
+
+<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->
+
+<!-- [@deAyala_2009, p.
+291-29] think about indeterminancies in MIRT.
+to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
 
@@
 
-# Declarations {.appendix .unnumbered}
+# Declarations {.unnumbered .appendix appendix-style=plain}
 
@@
 
-# References {.unnumbered}
+# Appendix {#sec-appendix .appendix appendix-style=plain}
+
+This section introduces fundamental statistical and causal inference concepts necessary for understanding the core theoretical principles described in @sec-theory.
+It does not, however, offer a comprehensive overview of statistical and causal inference methods.
+Readers seeking more in-depth understanding may wish to explore introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020.
+They may also find it helpful to consult introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020.
+For more advanced study, readers may refer to seminal intermediate papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020.
+
+## Empirical research and randomized experiments {#sec-appendix-A .appendix appendix-style=plain}
+
+<!-- 1. What does empirical research do first? -->
+Empirical research uses evidence from observation and experimentation to address real-world challenges.
+In this context, researchers typically formulate their research questions as *estimands* or *targets of inference*, i.e., the specific quantities they seek to determine [@Everitt_et_al_2010].
+For instance, researchers might be interested in answering the following question: “To what extent do different teaching methods $(T)$ influence students' ability to produce high-quality written texts $(Y)$?" To investigate this, researchers could randomly assign students to two groups, each exposed to a different teaching method $(T_{i} = \{1,2\})$.
+Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability.
+In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random quantity $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, where $E[\cdot]$ denotes the expected value.
+
+<!-- 2. What do they do next? -->
+Researchers would then proceed to identify the estimands.
+*Identification* refers to the process of accurately computing an estimand using an estimator.
+An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
+*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
+The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.
+
+![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}
+
+<!-- 3. But, what do we need to produce a good estimate? -->
+While numerous methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
+For instance, the Z-test is an estimator known for its effectiveness in comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006].
+If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
+The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i}| T_{i}=1]$ and $E[Y_{i}| T_{i}=2]$, respectively.
+The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.
+
+<!-- the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the comparison of two groups, as long as the assumptions of the model are met.
+Usually these estimates are expressed in terms of the proportions of groups  -->
+
+<!-- 4. But we want these estimands to be causal -->
+However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply identify associations.
+In the example, researchers can interpret the associational estimate represented by the Z-statistic as causal.
+This interpretation relies on the data meeting the assumptions of the Z-test and the data being collected through a randomized experiment.
+
+Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014].
+This recognition stems from their ability to enable researchers to interpret associational estimates as causal.
+They achieve this by ensuring data, and by extension an estimator, satisfies several key properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
+The most critical property, however, is the elimination of confounding.
+*Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010].
+Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from other variables $X$ and the outcome $Y$ [@Morgan_et_al_2014; @Neal_2020].
+
+<!-- 5. Experiments are not available to everyone, then causal inference comes to the rescue -->
+Nevertheless, researchers often face constraints that limit their ability to conduct randomized experiments.
+These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020].
+In these cases, causal inference offers a valuable alternative for generating causal estimates and understanding the mechanisms underlying specific data.
+In addition, the framework can provide significant theoretical insights that can enhance the design of experimental and observational studies [@McElreath_2020].
+
+## Empirical research under causal inference {#sec-appendix-B .appendix appendix-style=plain}
+
+<!-- 1. What is causal inference? -->
+Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, the *causal inference* framework is designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
+The framework uses rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" This question introduces the concept of counterfactuals, which are instrumental in understanding and defining causal effects.
+
+<!-- 2. But how does it work? -->
+*Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
+Although a comprehensive discussion of causes and counterfactuals is beyond the scope of this document, it is possible to provide a brief overview of how the framework addresses the fundamental problem of causality.
+Using the example outlined in @sec-appendix-B, the framework begins by defining the *individual causal effect* (ICE) as the difference between the students' potential outcomes: $\tau_{i} = Y^{1}_{i} - Y^{2}_{i}$.
+Here, $Y^{1}_{i}$ represents the potential outcome observed under $T_{i}=1$, while $Y^{2}_{i}$ represents the potential outcome observed under $T_{i}=2$.
+Note that when a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, making it a counterfactual.
+To address the challenge posed by these counterfactuals, the framework extends the ICE to the *average causal effect* (ACE).
+Researchers define the ACE as $\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]$, which represents the difference between the average of observed potential outcomes and counterfactuals.
+
+Even when data originates from an observational study, researchers can identify the ACE from associational estimates through causal inference.
+They achieve this by performing statistical conditioning on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+This *sufficient* set (potentially an empty set) should block all (and not open any) non-causal paths between $T$ to $Y$, thereby ensuring that the estimator for the ACE satisfies several key properties, including eliminating confounding.
+If such a set exists, researchers can express the ACE from associational random quantities as $\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].
+Of course, the validity of the claims about the effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set.
+However, as @Kohler_et_al_2019 [pp.
+150] noted, "statements about the real world from observed (not observational) data require assumptions (and) this is true no matter what kind of data we have".
+For example, if researchers cannot conduct the randomized experiments described in @sec-appendix-A and must rely on observational data, they can still estimate $\tau$ if a variable $X$, such as the socio-economic status of the school, blocks all (and not open any) non-causal paths from the teaching method $T$ to the outcome $Y$.
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
+
+<!-- 4. How do we apply causal inference and counterfactuals? -->
+Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
+Both approaches use rigorous mathematical notation to characterize the ACE, *not only for dichotomous but also for continuous variables* $T$, though they do so in different ways [@Neal_2020].
+The potential outcomes approach relies on counterfactual notation, as seen in the above examples, whereas the structural approach uses do-calculus [@Pearl_2009].
+Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to identify and estimate causal effects [@Pearl_2010].
+
+## SCMs and DAGs {#sec-appendix-C .appendix appendix-style=plain}
+
+<!-- commands for d-separation -->
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+
+<!-- 1. What is the benefit of causal inference? -->
+The structural approach to causal inference, however, offers a notable advantage over the potential outcomes approach.
+It enables researchers to formally and graphically represent a system using SCMs and DAGs [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
+SCMs and DAGs function as conceptual models describing the presumed causal structure underlying the ACE.
+In essence, these models act as *theoretical models* that guide researchers in determining which statistical models can yield valid causal inferences, assuming the depicted causal structure of the models is correct [@McElreath_2020].
+Notably, every SCM has a associated DAG [@Cinelli_et_al_2020].
+@fig-IEflow provides a visual representation of the role of theoretical models in the inference process.
+
+<!-- A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. -->
+
+<!-- 2. What are the advantages of SCMs and DAGs? -->
+SCMs and DAGs offer three key advantages for modeling causal structures.
+First, they enable researchers to represent causal relations in a non-parametric and fully interactive manner.
+This feature allows for feasible ACE identification strategies without specifying the data type or the nature of the functional dependence among variables [@Morgan_et_al_2014].
+Second, regardless of complexity, they can represent a wide range of causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+This feature allows researchers to decompose complex structures, facilitating their analysis by focusing on the causal assumptions associated with each building block [@McElreath_2020].
+Finally, they help to formalize (causal) theoretical models into (probabilistic) statistical models for identification purposes [@McElreath_2020] [@McElreath_2020].
+
+<!-- 3. the SCMs -->
+Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display the five fundamental building blocks of SCMs and DAGs.
+The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X,Z,Y\}$, a set of *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020].
+Endogenous variables are those whose causal mechanisms a researchers chooses to model [@Neal_2020].
+In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
+27,68].
+Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
+These functions use the symbol '$:=$' to denote the asymmetrical causal dependence of the variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+
+<!-- 4. The DAGs -->
+The right panels of the figures display the complementary DAGs.
+A DAG consists of nodes connected by edges, where nodes represent random variables.
+The term *directed* means the edges extend from one node to another, with arrows indicating the direction of causal influence.
+The term *acyclic* signifies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
+DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
+Finally, the arrows in the graphs show the expected direction of causal influences among these variables.
+Often, DAGs omit the exogenous variables for simplicity (the *standard* representation).
+However, including exogenous variables in a graph (the *magnified* representation) can be beneficial, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+
+::: {#fig-dags_scms1 layout-ncol=2}
+
+::: {#fig-scm_bb1 fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](images/png/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" fig-pos="H" width=50%}
+
+<!-- ![Two unconnected nodes](images/png/sdag_bb1.png){#fig-sdag_bb1 fig-align="center" fig-pos="H" width=35%} -->
+
+Two unconnected nodes
+:::
+
+::: {#fig-dags_scms2 layout-ncol=2}
+
+::: {#fig-scm_bb2 fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(X,e_{Y}) \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](images/png/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" fig-pos="H" width=50%}
+
+<!-- ![Two connected nodes or descendant](images/png/sdag_bb2.png){#fig-sdag_bb2 fig-align="center" fig-pos="H" width=35%} -->
+
+Two connected nodes or descendant
+:::
+
+::: {#fig-dags_scms3 layout-ncol=2}
+
+::: {#fig-scm_bb3 fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Z} \\
+  e_{Z} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](images/png/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" fig-pos="H" width=60%}
+
+<!-- ![Chain or pipe](images/png/sdag_bb3.png){#fig-sdag_bb3 fig-align="center" fig-pos="H" width=55%} -->
+
+Chain or mediator
+:::
+
+::: {#fig-dags_scms4 layout-ncol=2}
+
+::: {#fig-scm_bb4 fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  X & := f_{X}(Z,e_{X}) \\
+  Z & := f_{Z}(e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Z} \\
+  e_{Z} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](images/png/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" fig-pos="H" width=60%}
+
+<!-- ![Fork](images/png/sdag_bb4.png){#fig-sdag_bb4 fig-align="center" fig-pos="H" width=55%} -->
+
+Fork or confounder
+:::
+
+::: {#fig-dags_scms5 layout-ncol=2}
+
+::: {#fig-scm_bb5 fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,Y,e_{Z}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Z} \\
+  e_{Z} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](images/png/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" fig-pos="H" width=60%}
+
+<!-- ![Collider or inmorality](images/png/sdag_bb5.png){#fig-sdag_bb5 fig-align="center" fig-pos="H" width=55%} -->
+
+Collider or inmorality
+:::
+
+<!-- 5. Implications of the basic building block of the SCMS and DAGs -->
+A careful examination of the figures highlights the assumptions underlying these building blocks.
+Figures [-@fig-scm_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related.
+Figures [-@fig-scm_bb2] and [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$.
+Consequently, $Y$ is considered a *descendant* of $X$.
+Figures [-@fig-scm_bb3] and [-@fig-mdag_bb3] depict a *chain*, where $X$ influences $Z$, and $Z$ influences $Y$.
+In this configuration, $X$ is a parent node of $Z$, which is a parent node of $Y$.
+This structure creates a *directed path* between $X$ and $Y$.
+Consequently, $X$ is an *ancestor* of $Y$, and $Z$ fully *mediates* the relationship between the two.
+Figures [-@fig-scm_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
+Here, $Z$ is a parent node that *confounds* the relationship between $X$ and $Y$.
+Finally, Figures [-@fig-scm_bb5] and [-@fig-mdag_bb5] show a *collider*, where variables $X$ and $Y$ are concurrent causes of $Z$.
+In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$ (an *inmorality*).
+Notably, in all SCMs, the errors are assumed to be independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.
+
+<!-- 6. continuing with the motivating example -->
+Researchers can use these building blocks to depict the scenario described in @sec-appendix-B.
+In this scenario, they rely on observational data because it is not feasible to run a randomized experiment.
+Furthermore, the scenario indicates that researchers have measured a variable $X$ (the socio-economic status of the school) that is assumed to block all non-causal paths from the teaching method $T$ to the outcome $Y$.
+Figures [-@fig-scm_example1] and [-@fig-mdag_example1] illustrate the plausible causal structure for this example.
+The figures highlight the presence of at least four of the five fundamental building blocks.
+The figures display multiple descendants, as shown by pairwise relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+They also feature multiple pairs of unconnected nodes, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
+Finally, the figures illustrate one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.
+
+::: {#fig-example layout-ncol=2}
+
+::: {#fig-scm_example1 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := f_{Z}(X,e_{T}) \\
+  Y & := f_{Y}(T,X,e_{Y}) \\
+  e_{T} & \dsep e_{X} \\
+  e_{T} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](images/png/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=60%}
+
+<!-- ![Standard representation](images/png/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%} -->
+
+Plausible causal structure of the example.
+:::
+
+## The flow of association and causation in DAGs {#sec-appendix-D .appendix appendix-style=plain}
+
+<!-- 1. But what does the building blocks mean? -->
+<!-- [The advantages of SCMs and DAGs to identify causal quantities are realized after researchers examine the causal and probabilistic assumptions underlying these building blocks.]{style="color:red;"} -->
+
+<!-- The main assumption we’ve seen that relates the graph to the distribution -->
+<!-- is the Markov assumption. The Markov assumption tells us if variables are -->
+<!-- d-separated in the graph 𝐺 , then they are independent in the distribution -->
+<!-- 𝑃 (Theorem 3.1): -->
+<!-- Maybe we can detect independencies in the data and then use that -->
+<!-- to infer the causal graph. However, going from independencies in the -->
+<!-- distribution 𝑃 to d-separations in the graph 𝐺 isn’t something that the -->
+<!-- Markov assumption gives us (see Equation 3.20 above). Rather, we need -->
+<!-- the converse of the Markov assumption. This is known as the faithfulness -->
+<!-- assumption. -->
+<!-- Assumption 11.1 (Faithfulness) -->
+
+<!-- 2. What is the local markov assumption? and where does it lead? -->
+<!-- 2.1 Bayesian network factorization -->
+
+<!-- 3. What is the causal edges assumption? and where does it lead? -->
+
+<!-- 4. Combined they are known as the minimality assumption? -->
+
+<!-- 5. Now, what are those mysterious functions?, and how do they become statistical quantities? -->
+
+<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->
+
+<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables -->
+<!-- with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
+-->
+
+<!-- 6. What is the modularity assumption? and where does it lead? -->
+<!-- 6.1 the Truncated factorization -->
+<!-- 6.2 The law of counterfactuals -->
+
+<!-- 7. Causal mechanism -->
+<!-- 7.3 the backdoor criterion (do you need it?) -->
+
+<!-- 5. d-separation -->
+<!-- 5.1 Blocked path definition-->
+
+<!-- 6. putting it all together -->
+<!-- 6.1. Revising the building block: dependence and independence -->
+
+<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
+
+<!-- Mention convenient R package daggitty to draw dags and extract the causal assumption inherent in a graph -->
+
+<!-- After detailing how to build graphs and what they imply, Include steps to draw dags using a motivating example -->
+
+<!-- point out that the disturbances are usually stochastic in nature -->
+
+<!-- Causal mechanism as a result of markov and minimality assumption -->
+<!-- Modularity assumption also -->
+
+<!-- 4. But how do we move from targets of inference to estimates -->
+<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->
+
+![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%}
+
+{{< pagebreak >}}
+
+# References {.unnumbered .appendix appendix-style=plain}
+
+<!-- {.unnumbered} -->
 

### 2025-01-09 _tex/paper2.tex
Built site for gh-pages

@@
   pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
-  pdfkeywords={Causal analysis, Directed Acyclic Graphs, Bayesian
-statistical methods, Thurstonian model, Comparative
-judgement, Probability, Statistical modeling},
+  pdfkeywords={causal inference, directed acyclic graphs, structural
+causal models, bayesian statistical methods, thurstonian
+model, comparative judgement, probability, statistical modeling},
   colorlinks=true,
@@
 \begin{keyword}
-    Causal analysis \sep Directed Acyclic Graphs \sep Bayesian
-statistical methods \sep Thurstonian model \sep Comparative
-judgement \sep Probability \sep 
-    Statistical modeling
+    causal inference \sep directed acyclic graphs \sep structural causal
+models \sep bayesian statistical methods \sep thurstonian
+model \sep comparative judgement \sep probability \sep 
+    statistical modeling
 \end{keyword}
@@
 In \emph{comparative judgment} (CJ) studies, judges assess a specific
-trait or attribute across various stimuli by performing pairwise
+trait or attribute across different stimuli by performing pairwise
 comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
 produces a dichotomous outcome, indicating which stimulus is perceived
-to exhibit a higher trait level. For example, when assessing text
+to have a higher trait level. For example, when assessing writing
 quality, judges compare pairs of written texts (the stimuli) to
-determine the relative quality each text exhibit (the trait)
+determine the relative writing quality each text exhibit (the trait)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
@@
 traits and competencies over the past decade. These studies have
-emphasized three aspects of the method's effectiveness: its reliability,
-validity, and practical applicability. Research on reliability indicates
-that CJ requires a relatively small number of pairwise comparisons
-\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
-scores that are as precise and consistent as those generated by other
-assessment methods
-\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
-Furthermore, evidence suggests that the reliability and time efficiency
+highlighted three aspects of the method's effectiveness: its
+reliability, validity, and practical applicability. Research on
+reliability suggests that CJ requires a relatively modest number of
+pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
+to generate trait scores that are as precise and consistent as those
+generated by other assessment methods
+\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
+addition, the evidence suggests that the reliability and time efficiency
 of CJ are comparable, if not superior, to those of other assessment
@@
 \citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
-Meanwhile, research on validity suggests that scores generated by CJ can
-accurately represent the traits under measurement
-\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023},
-while research on practical applicability highlights the method's
+Meanwhile, research on the validity of CJ scores indicates their
+capacity to represent the traits under measurement accurately
+\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
+Moreover, research on CJ's practical applicability highlights its
 versatility across both educational and non-educational contexts
@@
 
-Nevertheless, despite the increasing number of CJ studies, unsystematic
-and fragmented research approaches have left several critical issues
-unaddressed. The present study primarily focuses on two: the
-overreliance on the assumptions of Thurstone's Case V in the statistical
-analysis of CJ data, and the apparent disconnect between CJ's approach
-to trait measurement and hypothesis testing. The following sections
-begin with a brief overview of Thurstone's theory and a detailed
-examination of these issues. Subsequently, the study introduces a
-theoretical model for CJ that builds upon Thurstone's theory, alongside
-its statistical translation, designed to address the two concerns
-simultaneously.
+Nevertheless, despite the increasing number of CJ studies, the
+prevalence of unsystematic and fragmented research approaches has left
+several critical issues unaddressed. The present study primarily focuses
+on two issues: the overreliance on Thurstone's Case V assumptions in the
+statistical analysis of CJ data and the apparent disconnect between CJ's
+approach to trait measurement and hypothesis testing. The following
+sections begin with a brief overview of Thurstone's theory followed by a
+detailed examination of these issues. Subsequently, the study introduces
+a theoretical model for CJ that builds upon Thurstone's theory,
+alongside its statistical translation, designed to address the two
+concerns simultaneously.
 
@@
 In its most general form, Thurstone's theory addresses pairwise
-comparisons where a single judge evaluates multiple stimuli
+comparisons wherein a single judge evaluates multiple stimuli
 \citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
@@
 
-However, since the individual discriminal processes of the stimuli are
-not directly observable, the theory introduces the \emph{law of
+However, given that the individual discriminal processes of the stimuli
+are not directly observable, the theory introduces the \emph{law of
 comparative judgment}. This law posits that in pairwise comparisons, a
@@
 of the stimuli, referred to as the \emph{discriminal difference},
-determines the observed dichotomous outcome. Moreover, the theory
+determines the observed dichotomous outcome. Furthermore, the theory
 assumes that because the individual discriminal processes form a Normal
@@
 This section examines the two critical issues in the CJ literature that
-serve as the primary focus of this study. The first is the overreliance
-on Thurstone's Case V assumptions in the statistical analysis of CJ
-data. The second is the apparent disconnect between CJ's approach to
-trait measurement and hypothesis testing.
+serve as the primary focus of the present study. The first is related to
+the overreliance on Thurstone's Case V assumptions in the statistical
+analysis of CJ data. The second concern with the apparent disconnect
+between CJ's approach to trait measurement and hypothesis testing.
 
@@
 Building on the previous section, it is clear that, despite its
-limitations, the BTL model is commonly used as the measurement model in
-CJ assessments. A measurement model specifies how manifest variables
+limitations, the BTL model is commonly used as a measurement model in CJ
+assessments. A measurement model specifies how manifest variables
 contribute to the estimation of latent variables
-\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
-researchers use the BTL model to process the dichotomous outcomes
-resulting from the pairwise comparisons (the manifest variables) to
-estimate scores that reflect the underlying quality level of the texts
-(the latent variable)
+\citep{Everitt_et_al_2010}. For example, when evaluating writing
+quality, researchers use the BTL model to process the dichotomous
+outcomes resulting from the pairwise comparisons (the manifest
+variables) to estimate scores that reflect the underlying level of
+writing quality (the latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
@@
 \citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
-systematic and integrated approach to examine the factors influencing CJ
-experiments could offer several statistical and measurement benefits,
-including the ability to address these issues.
-
-\section{An updated theoretical and statistical model for
-CJ}\label{sec-theory}
+systematic and integrated approach to handling these assumptions and
+examining the factors influencing CJ experiments could offer several
+statistical and measurement benefits, including the ability to address
+these issues.
+
+\section{Updating CJ's theoretical and statistical
+model}\label{sec-theory}
 
 This section presents a theoretical model for CJ that extends
-Thurstone's theory. It uses causal analysis
-\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014} and, in
-particular, directed acyclic graphs (DAGs)
-\citep{Gross_et_al_2018, Neal_2020}, to articulate the core theoretical
-principles of CJ theory. The model also incorporates several practical
-factors that influence judges in CJ experiments, such as the selection
-of judges, stimuli, and comparisons. In addition, the study uses
-Bayesian statistical methods \citep{McElreath_2020} to translate these
-theoretical and practical elements into a statistical model that
-facilitates the analysis of pairwise comparison data.
+Thurstone's theory. It uses causal inference and, in particular, the
+structural approach where structural causal models (SCMs) and directed
+acyclic graphs (DAGs) articulate the core theoretical principles of CJ
+theory. The model also incorporates several assessment design features
+that influence CJ experiments, such as the selection of judges, stimuli,
+and comparisons. In addition, the study uses Bayesian statistical
+methods to translate these theoretical and practical elements into a
+statistical model that facilitates the analysis of pairwise comparison
+data.
 
 \subsection{The theoretical model}\label{sec-theory-theoretical}
+
+\subsubsection{The population model}\label{sec-theory-theoretical_P}
+
+structural causal models (SCMs) and directed acyclic graphs (DAGs)
+\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}
+
+causal analysis \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}
+
+Assuming population data or more commonly known as census data, we
+\ldots{}
 
@@
 \end{figure}%
+
+\subsubsection{The assessment design
+features}\label{sec-theory-theoretical_AD}
 
@@
 
+\section{Appendix}\label{sec-appendix}
+
+This section introduces fundamental statistical and causal inference
+concepts necessary for understanding the core theoretical principles
+described in Section~\ref{sec-theory}. It does not, however, offer a
+comprehensive overview of statistical and causal inference methods.
+Readers seeking more in-depth understanding may wish to explore
+introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
+\citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}. They may also find
+it helpful to consult introductory books like \citet{Pearl_et_al_2018},
+\citet{Neal_2020}, and \citet{McElreath_2020}. For more advanced study,
+readers may refer to seminal intermediate papers such as
+\citet{Neyman_et_al_1923}, \citet{Rubin_1974},
+\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
+such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
+\citet{Hernan_et_al_2020}.
+
+\subsection{Empirical research and randomized
+experiments}\label{sec-appendix-A}
+
+Empirical research uses evidence from observation and experimentation to
+address real-world challenges. In this context, researchers typically
+formulate their research questions as \emph{estimands} or \emph{targets
+of inference}, i.e., the specific quantities they seek to determine
+\citep{Everitt_et_al_2010}. For instance, researchers might be
+interested in answering the following question: ``To what extent do
+different teaching methods \((T)\) influence students' ability to
+produce high-quality written texts \((Y)\)?'' To investigate this,
+researchers could randomly assign students to two groups, each exposed
+to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
+perform pairwise comparisons, generating a dichotomous outcome
+\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
+ability. In this scenario, the research question can be rephrased as the
+estimand, ``\emph{On average}, is there a difference in the ability to
+produce high-quality written texts between the two groups of students?''
+and this estimand can be mathematically represented by the random
+quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where \(E[\cdot]\)
+denotes the expected value.
+
+Researchers would then proceed to identify the estimands.
+\emph{Identification} refers to the process of accurately computing an
+estimand using an estimator. An \emph{estimator} is a method or function
+that transforms data into an estimate \citep{Neal_2020}.
+\emph{Estimates} are numerical values that approximate the estimand and
+are derived through \emph{estimation}, which refers to the process of
+integrating data with an estimator \citep{Everitt_et_al_2010}. The
+Identification-Estimation flowchart \citep{McElreath_2020, Neal_2020} in
+Figure~\ref{fig-IEflow} provides a visual representation of the process
+of transitioning from estimands to estimates.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{images/png/IEflow.png}
+
+}
+
+\caption{\label{fig-IEflow}Identification-Estimation flowchart.
+Extracted and slightly modified from \citet[pp.~32]{Neal_2020}}
+
+\end{figure}%
+
+While numerous methods can approximate an estimand, researchers
+prioritize estimators with desirable properties that ensure the accuracy
+of estimates. For instance, the Z-test is an estimator known for its
+effectiveness in comparing groups' proportions, yielding accurate
+estimates when the underlying assumptions of the statistic are met
+\citep{Kanji_2006}. If this is the case, the Z-test is expressed as a
+signal-to-noise statistic
+\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
+as the difference between the groups' sample proportions,
+\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
+\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
+\(E[Y_{i}| T_{i}=1]\) and \(E[Y_{i}| T_{i}=2]\), respectively. The
+noise, represented by \(\hat{s}_{p}\), is defined as the unpooled sample
+variability observed between the two groups.
+
+However, researchers often seek to uncover the mechanisms underlying
+specific data and establish causal relationships rather than simply
+identify associations. In the example, researchers can interpret the
+associational estimate represented by the Z-statistic as causal. This
+interpretation relies on the data meeting the assumptions of the Z-test
+and the data being collected through a randomized experiment.
+
+Randomized experiments are widely recognized as the gold standard in
+evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
+recognition stems from their ability to enable researchers to interpret
+associational estimates as causal. They achieve this by ensuring data,
+and by extension an estimator, satisfies several key properties, such as
+common support, no interference, and consistency
+\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
+however, is the elimination of confounding. \emph{Confounding} occurs
+when an external variable \(X\) simultaneously influences the outcome
+\(Y\) and the variable of interest \(T\), resulting in spurious
+associations \citep{Everitt_et_al_2010}. Randomization addresses this
+issue by decoupling the association between the intervention allocation
+\(T\) from other variables \(X\) and the outcome \(Y\)
+\citep{Morgan_et_al_2014, Neal_2020}.
+
+Nevertheless, researchers often face constraints that limit their
+ability to conduct randomized experiments. These constraints include
+ethical concerns, such as the assignment of individuals to potentially
+harmful interventions, and practical limitations, such as the
+infeasibility of, for example, assigning individuals to genetic
+modifications or physical impairments \citep{Neal_2020}. In these cases,
+causal inference offers a valuable alternative for generating causal
+estimates and understanding the mechanisms underlying specific data. In
+addition, the framework can provide significant theoretical insights
+that can enhance the design of experimental and observational studies
+\citep{McElreath_2020}.
+
+\subsection{Empirical research under causal
+inference}\label{sec-appendix-B}
+
+Unlike classical statistical modeling, which focuses primarily on
+summarizing data and inferring associations, the \emph{causal inference}
+framework is designed to identify causes and estimate their effects
+using data \citep{Shaughnessy_et_al_2010, Neal_2020}. The framework uses
+rigorous mathematical techniques to address the \emph{fundamental
+problem of causality}
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This problem
+revolves around the question, ``What would have happened `in the world'
+under different circumstances?'' This question introduces the concept of
+counterfactuals, which are instrumental in understanding and defining
+causal effects.
+
+\emph{Counterfactuals} represent hypothetical scenarios that are
+\emph{contrary to fact}, where alternative outcomes resulting from a
+given cause are neither observed nor observable
+\citep{Neal_2020, Counterfactual_2024}. Although a comprehensive
+discussion of causes and counterfactuals is beyond the scope of this
+document, it is possible to provide a brief overview of how the
+framework addresses the fundamental problem of causality. Using the
+example outlined in Section~\ref{sec-appendix-B}, the framework begins
+by defining the \emph{individual causal effect} (ICE) as the difference
+between the students' potential outcomes:
+\(\tau_{i} = Y^{1}_{i} - Y^{2}_{i}\). Here, \(Y^{1}_{i}\) represents the
+potential outcome observed under \(T_{i}=1\), while \(Y^{2}_{i}\)
+represents the potential outcome observed under \(T_{i}=2\). Note that
+when a student is assigned to \(T_{i}=1\), the potential outcome under
+\(T_{i}=2\) is no longer observed nor observable, making it a
+counterfactual. To address the challenge posed by these counterfactuals,
+the framework extends the ICE to the \emph{average causal effect} (ACE).
+Researchers define the ACE as
+\(\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]\), which represents
+the difference between the average of observed potential outcomes and
+counterfactuals.
+
+Even when data originates from an observational study, researchers can
+identify the ACE from associational estimates through causal inference.
+They achieve this by performing statistical conditioning on a
+\emph{sufficient adjustment set} of variables \(X\)
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
+\emph{sufficient} set (potentially an empty set) should block all (and
+not open any) non-causal paths between \(T\) to \(Y\), thereby ensuring
+that the estimator for the ACE satisfies several key properties,
+including eliminating confounding. If such a set exists, researchers can
+express the ACE from associational random quantities as
+\(\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
+where \(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
+\citep{Morgan_et_al_2014}. Of course, the validity of the claims about
+the effects of \(T\) on \(Y\) now hinges on the assumption that \(X\)
+serves as a sufficient adjustment set. However, as
+\citet[pp.~150]{Kohler_et_al_2019} noted, ``statements about the real
+world from observed (not observational) data require assumptions (and)
+this is true no matter what kind of data we have''. For example, if
+researchers cannot conduct the randomized experiments described in
+Section~\ref{sec-appendix-A} and must rely on observational data, they
+can still estimate \(\tau\) if a variable \(X\), such as the
+socio-economic status of the school, blocks all (and not open any)
+non-causal paths from the teaching method \(T\) to the outcome \(Y\).
+
+Several approaches to causal inference and counterfactuals exist, but
+two are particularly prominent: the potential outcomes approach, also
+known as the Neyman-Rubin causal model
+\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
+\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches use rigorous
+mathematical notation to characterize the ACE, \emph{not only for
+dichotomous but also for continuous variables} \(T\), though they do so
+in different ways \citep{Neal_2020}. The potential outcomes approach
+relies on counterfactual notation, as seen in the above examples,
+whereas the structural approach uses do-calculus \citep{Pearl_2009}.
+Despite these differences, both notations can be expressed in terms of
+the other, and both approaches provide methods for using experimental
+and observational data to identify and estimate causal effects
+\citep{Pearl_2010}.
+
+\subsection{SCMs and DAGs}\label{sec-appendix-C}
+
+\newcommand{\dsep}{\perp\!\!\!\perp}
+\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+
+The structural approach to causal inference, however, offers a notable
+advantage over the potential outcomes approach. It enables researchers
+to formally and graphically represent a system using SCMs and DAGs
+\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. SCMs
+and DAGs function as conceptual models describing the presumed causal
+structure underlying the ACE. In essence, these models act as
+\emph{theoretical models} that guide researchers in determining which
+statistical models can yield valid causal inferences, assuming the
+depicted causal structure of the models is correct
+\citep{McElreath_2020}. Notably, every SCM has a associated DAG
+\citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a visual
+representation of the role of theoretical models in the inference
+process.
+
+SCMs and DAGs offer three key advantages for modeling causal structures.
+First, they enable researchers to represent causal relations in a
+non-parametric and fully interactive manner. This feature allows for
+feasible ACE identification strategies without specifying the data type
+or the nature of the functional dependence among variables
+\citep{Morgan_et_al_2014}. Second, regardless of complexity, they can
+represent a wide range of causal structures using only five fundamental
+building blocks \citep{Neal_2020, McElreath_2020}. This feature allows
+researchers to decompose complex structures, facilitating their analysis
+by focusing on the causal assumptions associated with each building
+block \citep{McElreath_2020}. Finally, they help to formalize (causal)
+theoretical models into (probabilistic) statistical models for
+identification purposes \citep{McElreath_2020} \citep{McElreath_2020}.
+
+Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
+\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
+display the five fundamental building blocks of SCMs and DAGs. The left
+panels of the figures show the formal mathematical models, represented
+by the SCMs, defined in terms of a set of \emph{endogenous} variables
+\(V=\{X,Z,Y\}\), a set of \emph{exogenous} variables
+\(E=\{e_{X},e_{Z},e_{Y}\}\), and a set of functions
+\(F=\{f_{X},f_{Z},f_{Y}\}\)
+\citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
+are those whose causal mechanisms a researchers chooses to model
+\citep{Neal_2020}. In contrast, exogenous variables represent
+\emph{errors} or \emph{disturbances} arising from omitted factors that
+the investigator chooses not to model explicitly
+\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
+\emph{structural equations}, express the endogenous variables as
+non-parametric functions of other variables. These functions use the
+symbol `\(:=\)' to denote the asymmetrical causal dependence of the
+variables and the symbol `\(\perp\!\!\!\perp\)' to represent
+\emph{d-separation}, a concept akin to (conditional) independence.
+
+The right panels of the figures display the complementary DAGs. A DAG
+consists of nodes connected by edges, where nodes represent random
+variables. The term \emph{directed} means the edges extend from one node
+to another, with arrows indicating the direction of causal influence.
+The term \emph{acyclic} signifies that the causal influences do not form
+loops, ensuring the influences do not cycle back on themselves
+\citep{McElreath_2020}. DAGs represent observed variables as solid black
+circles, while they use open circles for unobserved (latent) variables
+\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs show the
+expected direction of causal influences among these variables. Often,
+DAGs omit the exogenous variables for simplicity (the \emph{standard}
+representation). However, including exogenous variables in a graph (the
+\emph{magnified} representation) can be beneficial, as their presence
+can reveal potential issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}.
+
+\begin{figure}[H]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb1}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb1}DAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\end{minipage}%
+
+\caption{\label{fig-dags_scms1}Two unconnected nodes}
+
+\end{figure}%
+
+\begin{figure}[H]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Y & := f_{Y}(X,e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb2}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb2}DAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\end{minipage}%
+
+\caption{\label{fig-dags_scms2}Two connected nodes or descendant}
+
+\end{figure}%
+
+\begin{figure}[H]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Z} \\
+  e_{Z} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb3}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb3}DAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\end{minipage}%
+
+\caption{\label{fig-dags_scms3}Chain or mediator}
+
+\end{figure}%
+
+\begin{figure}[H]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(Z,e_{X}) \\
+  Z & := f_{Z}(e_{Z}) \\
+  Y & := f_{Y}(Z,e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Z} \\
+  e_{Z} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb4}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb4}DAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\end{minipage}%
+
+\caption{\label{fig-dags_scms4}Fork or confounder}
+
+\end{figure}%
+
+\begin{figure}[H]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  Z & := f_{Z}(X,Y,e_{Z}) \\
+  Y & := f_{Y}(e_{Y}) \\
+  e_{X} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Z} \\
+  e_{Z} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_bb5}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}
+
+}
+
+\subcaption{\label{fig-mdag_bb5}DAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\end{minipage}%
+
+\caption{\label{fig-dags_scms5}Collider or inmorality}
+
+\end{figure}%
+
+A careful examination of the figures highlights the assumptions
+underlying these building blocks. Figures \ref{fig-scm_bb1} and
+\ref{fig-mdag_bb1} depict two unconnected nodes, representing a scenario
+where variables \(X\) and \(Y\) are not causally related. Figures
+\ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two connected nodes,
+representing a scenario where a \emph{parent} node \(X\) exerts a causal
+influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
+considered a \emph{descendant} of \(X\). Figures \ref{fig-scm_bb3} and
+\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X\) influences \(Z\),
+and \(Z\) influences \(Y\). In this configuration, \(X\) is a parent
+node of \(Z\), which is a parent node of \(Y\). This structure creates a
+\emph{directed path} between \(X\) and \(Y\). Consequently, \(X\) is an
+\emph{ancestor} of \(Y\), and \(Z\) fully \emph{mediates} the
+relationship between the two. Figures \ref{fig-scm_bb4} and
+\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
+\(Y\) are both influenced by \(Z\). Here, \(Z\) is a parent node that
+\emph{confounds} the relationship between \(X\) and \(Y\). Finally,
+Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a \emph{collider},
+where variables \(X\) and \(Y\) are concurrent causes of \(Z\). In this
+configuration, \(X\) and \(Y\) are not causally related to each other
+but both influence \(Z\) (an \emph{inmorality}). Notably, in all SCMs,
+the errors are assumed to be independent of each other and from all
+other variables in the graph, as evidenced by the pairwise relations
+\(e_{X} \perp\!\!\!\perp e_{Y}\), \(e_{X} \perp\!\!\!\perp e_{Z}\), and
+\(e_{Z} \perp\!\!\!\perp e_{Y}\).
+
+Researchers can use these building blocks to depict the scenario
+described in Section~\ref{sec-appendix-B}. In this scenario, they rely
+on observational data because it is not feasible to run a randomized
+experiment. Furthermore, the scenario indicates that researchers have
+measured a variable \(X\) (the socio-economic status of the school) that
+is assumed to block all non-causal paths from the teaching method \(T\)
+to the outcome \(Y\). Figures \ref{fig-scm_example1} and
+\ref{fig-mdag_example1} illustrate the plausible causal structure for
+this example. The figures highlight the presence of at least four of the
+five fundamental building blocks. The figures display multiple
+descendants, as shown by pairwise relationships such as
+\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
+also feature multiple pairs of unconnected nodes, evident from the
+relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
+\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
+Finally, the figures illustrate one fork, \(X \rightarrow \{T, Y\}\),
+and two colliders: \(\{X, e_{T}\} \rightarrow T\) and
+\(\{X, T, e_{Y}\} \rightarrow Y\).
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := f_{Z}(X,e_{T}) \\
+  Y & := f_{Y}(T,X,e_{Y}) \\
+  e_{T} & \perp\!\!\!\perp e_{X} \\
+  e_{T} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_example1}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}
+
+}
+
+\subcaption{\label{fig-mdag_example1}DAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{0.50\linewidth}
+
+\end{minipage}%
+
+\caption{\label{fig-example}Plausible causal structure of the example.}
+
+\end{figure}%
+
+\subsection{The flow of association and causation in
+DAGs}\label{sec-appendix-D}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{images/png/ACflow.png}
+
+}
+
+\caption{\label{fig-ACflow}The flow of association and causation in
+graphs. Extracted and slightly modified from \citet[pp.~31]{Neal_2020}}
+
+\end{figure}%
+
+\newpage{}
+
 \section*{References}\label{references}

### 2025-01-09 paper2.qmd
section:appendix; process:write

@@
 <!-- 2. What are the advantages of SCMs and DAGs? -->
-SCMs and DAGs offer three key advantages for modeling causal structures.
+SCMs and DAGs offer two key advantages for modeling causal structures.
 First, they enable researchers to represent causal relations in a non-parametric and fully interactive manner.
@@
 This feature allows researchers to decompose complex structures, facilitating their analysis by focusing on the causal assumptions associated with each building block [@McElreath_2020].
-Finally, they help to formalize (causal) theoretical models into (probabilistic) statistical models for identification purposes [@McElreath_2020] [@McElreath_2020].
 
@@
 
-Plausible causal structure of the example.
+Plausible causal structure for the example.
 :::
@@
 <!-- 1. But what does the building blocks mean? -->
+<!-- Ultimately, SCMs and DAGS help to formalize (causal) theoretical models into (probabilistic) statistical models for identification purposes [@McElreath_2020]. -->
+
 <!-- [The advantages of SCMs and DAGs to identify causal quantities are realized after researchers examine the causal and probabilistic assumptions underlying these building blocks.]{style="color:red;"} -->

### 2025-01-09 _tex/paper2.tex
Built site for gh-pages

@@
 
-SCMs and DAGs offer three key advantages for modeling causal structures.
+SCMs and DAGs offer two key advantages for modeling causal structures.
 First, they enable researchers to represent causal relations in a
@@
 by focusing on the causal assumptions associated with each building
-block \citep{McElreath_2020}. Finally, they help to formalize (causal)
-theoretical models into (probabilistic) statistical models for
-identification purposes \citep{McElreath_2020} \citep{McElreath_2020}.
+block \citep{McElreath_2020}.
 
@@
 
-\caption{\label{fig-example}Plausible causal structure of the example.}
+\caption{\label{fig-example}Plausible causal structure for the example.}
 

### 2025-01-10 paper2.qmd
section:appendix,model; process:write

@@
 
-# Updating CJ's theoretical and statistical model {#sec-theory}
-
-This section presents a theoretical model for CJ that extends Thurstone's theory.
-It uses causal inference and, in particular, the structural approach where structural causal models (SCMs) and directed acyclic graphs (DAGs) articulate the core theoretical principles of CJ theory.
-The model also incorporates several assessment design features that influence CJ experiments, such as the selection of judges, stimuli, and comparisons.
-In addition, the study uses Bayesian statistical methods to translate these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
+# Updating the theoretical and statistical model {#sec-theory}
+
+This section uses the structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] to articulate a theoretical model that captures the core principles of Thurstone's theory.
+The model also incorporates various assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+Finally, the section employs Bayesian statistical methods to transform these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
+See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section.
 
 ## The theoretical model {#sec-theory-theoretical}
+
+The theoretical model uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of the CJ system.
+First, *the population model* is created to represent a conceptual population of CJ experiments.
+The model then integrates various assessment design features relevant to CJ experiments, leading to the development of *the sample-comparison model*.
 
@@
 
-structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]
-
-causal analysis [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]
-
 <!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
@@
 
-### The assessment design features {#sec-theory-theoretical_AD}
+### The sample-comparison model {#sec-theory-theoretical_SC}
 
@@
 This section introduces fundamental statistical and causal inference concepts necessary for understanding the core theoretical principles described in @sec-theory.
-It does not, however, offer a comprehensive overview of statistical and causal inference methods.
+It does not, however, offer a comprehensive overview of causal inference methods.
 Readers seeking more in-depth understanding may wish to explore introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020.
@@
 <!-- 4. But we want these estimands to be causal -->
-However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply identify associations.
+However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply estimate associations.
 In the example, researchers can interpret the associational estimate represented by the Z-statistic as causal.
-This interpretation relies on the data meeting the assumptions of the Z-test and the data being collected through a randomized experiment.
+This interpretation relies on the data meeting the assumptions of the Z-test and being collected through a randomized experiment.
 
@@
 *Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010].
-Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from other variables $X$ and the outcome $Y$ [@Morgan_et_al_2014; @Neal_2020].
+Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from any other variable $X$ [@Morgan_et_al_2014; @Neal_2020].
 
@@
 
-<!-- 2. But how does it work? -->
-*Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-Although a comprehensive discussion of causes and counterfactuals is beyond the scope of this document, it is possible to provide a brief overview of how the framework addresses the fundamental problem of causality.
-Using the example outlined in @sec-appendix-B, the framework begins by defining the *individual causal effect* (ICE) as the difference between the students' potential outcomes: $\tau_{i} = Y^{1}_{i} - Y^{2}_{i}$.
-Here, $Y^{1}_{i}$ represents the potential outcome observed under $T_{i}=1$, while $Y^{2}_{i}$ represents the potential outcome observed under $T_{i}=2$.
-Note that when a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, making it a counterfactual.
-To address the challenge posed by these counterfactuals, the framework extends the ICE to the *average causal effect* (ACE).
-Researchers define the ACE as $\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]$, which represents the difference between the average of observed potential outcomes and counterfactuals.
-
-Even when data originates from an observational study, researchers can identify the ACE from associational estimates through causal inference.
+<!-- 2. What are counterfactuals? -->
+*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
+The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals.
+For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes: $\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)$.
+Here, $do(T_{i}=t)$ represents the intervention operator, $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
+Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable.
+To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts.
+Notably, the approach extends the ACEs for categorical and continuous variables $T$, ensuring broad applicability across different causal scenarios.
+
+<!-- [^1]: The potential outcomes approach [@Neyman_et_al_1923; @Rubin_1974] defines the ICE as $\tau_{i} = Y^{1}_{i} - Y^{2}_{i}$, while it defines the ACE as $\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]$.
+-->
+
+<!-- 3. But how does it work? -->
+Even when data originates from an observational study, researchers can identify the ACE from associational estimates using the structural approach.
 They achieve this by performing statistical conditioning on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-This *sufficient* set (potentially an empty set) should block all (and not open any) non-causal paths between $T$ to $Y$, thereby ensuring that the estimator for the ACE satisfies several key properties, including eliminating confounding.
-If such a set exists, researchers can express the ACE from associational random quantities as $\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].
-Of course, the validity of the claims about the effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set.
+This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones, ensuring the ACE estimator satisfies several key properties, including confounding elimination.
+If such a set exists, then $T$ and $Y$ are *d-separated* by $X$ [@Pearl_2009], meaning researchers can estimate the ACE from associational random quantities [@Morgan_et_al_2014].
+Naturally, the validity of claims about the effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set.
 However, as @Kohler_et_al_2019 [pp.
-150] noted, "statements about the real world from observed (not observational) data require assumptions (and) this is true no matter what kind of data we have".
-For example, if researchers cannot conduct the randomized experiments described in @sec-appendix-A and must rely on observational data, they can still estimate $\tau$ if a variable $X$, such as the socio-economic status of the school, blocks all (and not open any) non-causal paths from the teaching method $T$ to the outcome $Y$.
+150] observed, drawing conclusions about the real world from observed data inevitably requires assumptions.
+This holds true for both observational and experimental data.
+
+For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and instead rely on observational data, they can still estimate the ACE provided a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ to the outcome $Y$  without opening any new ones.
+Under these conditions, the ACE can be estimated from associational quantities as $\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
 
@@
 
-<!-- 4. How do we apply causal inference and counterfactuals? -->
-Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016].
-Both approaches use rigorous mathematical notation to characterize the ACE, *not only for dichotomous but also for continuous variables* $T$, though they do so in different ways [@Neal_2020].
-The potential outcomes approach relies on counterfactual notation, as seen in the above examples, whereas the structural approach uses do-calculus [@Pearl_2009].
-Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to identify and estimate causal effects [@Pearl_2010].
-
 ## SCMs and DAGs {#sec-appendix-C .appendix appendix-style=plain}
@@
 
-<!-- 1. What is the benefit of causal inference? -->
-The structural approach to causal inference, however, offers a notable advantage over the potential outcomes approach.
-It enables researchers to formally and graphically represent a system using SCMs and DAGs [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-SCMs and DAGs function as conceptual models describing the presumed causal structure underlying the ACE.
-In essence, these models act as *theoretical models* that guide researchers in determining which statistical models can yield valid causal inferences, assuming the depicted causal structure of the models is correct [@McElreath_2020].
-Notably, every SCM has a associated DAG [@Cinelli_et_al_2020].
+<!-- 1. What does the structural approach uses? -->
+The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
+In essence, SCMs and DAGs act as *conceptual models* that guide researchers in determining which statistical models can yield valid causal inferences, assuming the depicted causal structure of the models is correct [@McElreath_2020].
+Notably, every SCM has an associated DAG [@Cinelli_et_al_2020].
 @fig-IEflow provides a visual representation of the role of theoretical models in the inference process.
@@
 SCMs and DAGs offer two key advantages for modeling causal structures.
-First, they enable researchers to represent causal relations in a non-parametric and fully interactive manner.
+First, they enable the representation of causal relations in a non-parametric and fully interactive manner.
 This feature allows for feasible ACE identification strategies without specifying the data type or the nature of the functional dependence among variables [@Morgan_et_al_2014].
@@
 <!-- 3. the SCMs -->
-Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display the five fundamental building blocks of SCMs and DAGs.
+Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display these five fundamental building blocks.
 The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X,Z,Y\}$, a set of *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020].
-Endogenous variables are those whose causal mechanisms a researchers chooses to model [@Neal_2020].
+Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
 In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
@@
 Finally, the arrows in the graphs show the expected direction of causal influences among these variables.
-Often, DAGs omit the exogenous variables for simplicity (the *standard* representation).
-However, including exogenous variables in a graph (the *magnified* representation) can be beneficial, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+While DAGs often omit exogenous variables for simplicity (the *standard* representation), including these variables is advantageous (the *magnified* representation shown in the figures), as their inclusion helps to identify potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 <!-- 6. continuing with the motivating example -->
-Researchers can use these building blocks to depict the scenario described in @sec-appendix-B.
-In this scenario, they rely on observational data because it is not feasible to run a randomized experiment.
-Furthermore, the scenario indicates that researchers have measured a variable $X$ (the socio-economic status of the school) that is assumed to block all non-causal paths from the teaching method $T$ to the outcome $Y$.
-Figures [-@fig-scm_example1] and [-@fig-mdag_example1] illustrate the plausible causal structure for this example.
-The figures highlight the presence of at least four of the five fundamental building blocks.
-The figures display multiple descendants, as shown by pairwise relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
-They also feature multiple pairs of unconnected nodes, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
-Finally, the figures illustrate one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.
+Researchers can use these building blocks to represent the scenario outlined in @sec-appendix-B.
+Figures [-@fig-scm_example1] and [-@fig-mdag_example1] depict the plausible causal structure for this example, assuming that the variable $X$ (socio-economic status of the school) acts as a confounder in the relationship between the teaching method $T$ and the outcome $Y$.
+The figures show several descendant, such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$, and also highlight multiple pairs of unconnected nodes, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
+Additional, the figures depict one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.
 
@@
 
-Plausible causal structure for the example.
+Plausible causal structure the scenario outlined in @sec-appendix-B.
 :::
@@
 
-<!-- 1. But what does the building blocks mean? -->
-<!-- Ultimately, SCMs and DAGS help to formalize (causal) theoretical models into (probabilistic) statistical models for identification purposes [@McElreath_2020]. -->
+<!-- 1. The general probabilistic description of data -->
+
+<!-- Maximum a posteriori (MAP) estimates are the set of parameters that maximizes the probability of $\theta$ given the data.
+By Bayes theorem the MAP estimate can be obtained by maximizing the following conditional distribution: -->
+
+<!-- $$ -->
+<!-- P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})} -->
+<!-- $$ -->
+<!-- where $P(\theta)$ is the prior distribution for the parameters $\theta$ and $P(X_{1}, X_{2}, \dots, X_{n})$ is the probability of the data averaged over all parameters.
+Since the denominator is independent of $\theta$, the Bayesian estimator is obtained by maximizing $P(X_{1}, X_{2}, \dots, X_{n} | \theta)$ with respect to $\theta$.
+This indicates that when researchers model association, they generally model the joint probability of the data given the parameters $\theta$.
+-->
+
+<!-- This join probability can be further re-expressed using the chain rule of probability as $P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )$.
+However, from this description, it is easy to see how quickly the parametrization becomes intractable as the number of observed variables $n$ grows [@Neal_2020].
+To deal with this complexity, researchers typically simplify the modeling process by assuming local (in)dependencies between variables.
+-->
+
+<!-- 2. But what does the building blocks mean? -->
+<!-- The five building blocks help to express joint distributions very efficiently, by formalizing local (in)dependencies into (probabilistic) statistical models [@Neal_2020].
+Notably, this formalization rest on two key assumptions: the local Markov and causal edges assumptions.
+The local Markov assumption states that given its parents in the DAG, a node is independent of all its non-descendants.
+This lead to the Bayesian Network factorization.
+However, this assumption only give us information about independencias in the distribution that the DAG implies, we need also to know about the dependencies.
+The causal edges assumption states that in a DAG, every parent is a direct cause of all its children.
+In this sense this assumption is indicating that parents and their children are dependent, and more precisely, they are causally dependent [@Neal_2020].
+-->
+
+<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->
+
+<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables -->
+<!-- with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
+-->
 
@@
 
-<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->
-
-<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables -->
-<!-- with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
--->
-
 <!-- 6. What is the modularity assumption? and where does it lead? -->
@@
 
+<!-- For instance, in the example outlined in @sec-appendix-C, the collected observational data can be expressed as joint probability $P(X,T,Y)$.
+Using the chain rule, this probability can be further factorized in $P(X,T,Y) = P(X) P(T|X) P(Y|T,X)$.
+Although this description of this data does not seem hard   -->
+
+<!-- These building blocks can then guide researchers to identify feasible ACE identification strategies.
+Researchers can use these building blocks to depict the scenario described in @sec-appendix-B.
+Assuming the variable $X$ (the socio-economic status of the school) blocks all non-causal paths from the teaching method $T$ to the outcome $Y$.
+-->
+
+<!-- Conditioning on $T=t$ just means that we are restricting our focus to the subset of the population to those who received treatment $t$.
+In contrast, an intervention would be to take the whole population and give everyone treatment $t$ [@Neal_2020, pp.
+32].
+-->
+
 {{< pagebreak >}}

### 2025-01-10 _tex/paper2.tex
Built site for gh-pages

@@
 
-\section{Updating CJ's theoretical and statistical
+\section{Updating the theoretical and statistical
 model}\label{sec-theory}
 
-This section presents a theoretical model for CJ that extends
-Thurstone's theory. It uses causal inference and, in particular, the
-structural approach where structural causal models (SCMs) and directed
-acyclic graphs (DAGs) articulate the core theoretical principles of CJ
-theory. The model also incorporates several assessment design features
-that influence CJ experiments, such as the selection of judges, stimuli,
-and comparisons. In addition, the study uses Bayesian statistical
-methods to translate these theoretical and practical elements into a
-statistical model that facilitates the analysis of pairwise comparison
-data.
+This section uses the structural approach to causal inference
+\citep{Pearl_2009, Pearl_et_al_2016} to articulate a theoretical model
+that captures the core principles of Thurstone's theory. The model also
+incorporates various assessment design features relevant to CJ
+experiments, such as the selection of judges, stimuli, and comparisons.
+Finally, the section employs Bayesian statistical methods to transform
+these theoretical and practical elements into a statistical model that
+facilitates the analysis of pairwise comparison data. See
+Section~\ref{sec-appendix} for an overview of the statistical and causal
+inference concepts required for the development of this section.
 
@@
 
+The theoretical model uses structural causal models (SCMs) and directed
+acyclic graphs (DAGs)
+\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
+formally and graphically represent the assumed causal structure of the
+CJ system. First, \emph{the population model} is created to represent a
+conceptual population of CJ experiments. The model then integrates
+various assessment design features relevant to CJ experiments, leading
+to the development of \emph{the sample-comparison model}.
+
 \subsubsection{The population model}\label{sec-theory-theoretical_P}
-
-structural causal models (SCMs) and directed acyclic graphs (DAGs)
-\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}
-
-causal analysis \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}
 
@@
 
-\subsubsection{The assessment design
-features}\label{sec-theory-theoretical_AD}
+\subsubsection{The sample-comparison
+model}\label{sec-theory-theoretical_SC}
 
@@
 described in Section~\ref{sec-theory}. It does not, however, offer a
-comprehensive overview of statistical and causal inference methods.
-Readers seeking more in-depth understanding may wish to explore
-introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
-\citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}. They may also find
-it helpful to consult introductory books like \citet{Pearl_et_al_2018},
-\citet{Neal_2020}, and \citet{McElreath_2020}. For more advanced study,
-readers may refer to seminal intermediate papers such as
-\citet{Neyman_et_al_1923}, \citet{Rubin_1974},
-\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
-such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
-\citet{Hernan_et_al_2020}.
+comprehensive overview of causal inference methods. Readers seeking more
+in-depth understanding may wish to explore introductory papers such as
+\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
+\citet{Cinelli_et_al_2020}. They may also find it helpful to consult
+introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
+\citet{McElreath_2020}. For more advanced study, readers may refer to
+seminal intermediate papers such as \citet{Neyman_et_al_1923},
+\citet{Rubin_1974}, \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009},
+as well as books such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
+and \citet{Hernan_et_al_2020}.
 
@@
 specific data and establish causal relationships rather than simply
-identify associations. In the example, researchers can interpret the
+estimate associations. In the example, researchers can interpret the
 associational estimate represented by the Z-statistic as causal. This
 interpretation relies on the data meeting the assumptions of the Z-test
-and the data being collected through a randomized experiment.
+and being collected through a randomized experiment.
 
@@
 issue by decoupling the association between the intervention allocation
-\(T\) from other variables \(X\) and the outcome \(Y\)
+\(T\) from any other variable \(X\)
 \citep{Morgan_et_al_2014, Neal_2020}.
@@
 
-\emph{Counterfactuals} represent hypothetical scenarios that are
+\emph{Counterfactuals} are hypothetical scenarios that are
 \emph{contrary to fact}, where alternative outcomes resulting from a
 given cause are neither observed nor observable
-\citep{Neal_2020, Counterfactual_2024}. Although a comprehensive
-discussion of causes and counterfactuals is beyond the scope of this
-document, it is possible to provide a brief overview of how the
-framework addresses the fundamental problem of causality. Using the
-example outlined in Section~\ref{sec-appendix-B}, the framework begins
-by defining the \emph{individual causal effect} (ICE) as the difference
-between the students' potential outcomes:
-\(\tau_{i} = Y^{1}_{i} - Y^{2}_{i}\). Here, \(Y^{1}_{i}\) represents the
-potential outcome observed under \(T_{i}=1\), while \(Y^{2}_{i}\)
-represents the potential outcome observed under \(T_{i}=2\). Note that
-when a student is assigned to \(T_{i}=1\), the potential outcome under
-\(T_{i}=2\) is no longer observed nor observable, making it a
-counterfactual. To address the challenge posed by these counterfactuals,
-the framework extends the ICE to the \emph{average causal effect} (ACE).
-Researchers define the ACE as
-\(\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]\), which represents
-the difference between the average of observed potential outcomes and
-counterfactuals.
+\citep{Neal_2020, Counterfactual_2024}. The structural approach to
+causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
+framework for defining counterfactuals. For instance, in the scenario
+described in Section~\ref{sec-appendix-A}, the approach begins by
+defining the \emph{individual causal effect} (ICE) as the difference
+between each student's potential outcomes:
+\(\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)\). Here,
+\(do(T_{i}=t)\) represents the intervention operator,
+\(Y_{i}|do(T_{i}=1)\) represents the potential outcome under
+intervention \(T_{i}=1\), and \(Y_{i}|do(T_{i}=1)\) represents the
+potential outcome under intervention \(T_{i}=2\). Note that if a student
+is assigned to intervention \(T_{i}=1\), the potential outcome under
+\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
+observable. To address the challenge of unobserved counterfactuals, the
+approach extends the ICE to the \emph{average causal effect} (ACE):
+\(\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]\),
+representing the average difference between observed potential outcomes
+and their counterfactual counterparts. Notably, the approach extends the
+ACEs for categorical and continuous variables \(T\), ensuring broad
+applicability across different causal scenarios.
 
 Even when data originates from an observational study, researchers can
-identify the ACE from associational estimates through causal inference.
-They achieve this by performing statistical conditioning on a
+identify the ACE from associational estimates using the structural
+approach. They achieve this by performing statistical conditioning on a
 \emph{sufficient adjustment set} of variables \(X\)
 \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
-\emph{sufficient} set (potentially an empty set) should block all (and
-not open any) non-causal paths between \(T\) to \(Y\), thereby ensuring
-that the estimator for the ACE satisfies several key properties,
-including eliminating confounding. If such a set exists, researchers can
-express the ACE from associational random quantities as
+\emph{sufficient} set (potentially empty) must block all non-causal
+paths between \(T\) to \(Y\) without opening new ones, ensuring the ACE
+estimator satisfies several key properties, including confounding
+elimination. If such a set exists, then \(T\) and \(Y\) are
+\emph{d-separated} by \(X\) \citep{Pearl_2009}, meaning researchers can
+estimate the ACE from associational random quantities
+\citep{Morgan_et_al_2014}. Naturally, the validity of claims about the
+effects of \(T\) on \(Y\) now hinges on the assumption that \(X\) serves
+as a sufficient adjustment set. However, as
+\citet[pp.~150]{Kohler_et_al_2019} observed, drawing conclusions about
+the real world from observed data inevitably requires assumptions. This
+holds true for both observational and experimental data.
+
+For instance, if researchers are unable to conduct the randomized
+experiments described in Section~\ref{sec-appendix-A} and instead rely
+on observational data, they can still estimate the ACE provided a
+variable \(X\), such as the socio-economic status of the school, blocks
+all non-causal paths between the teaching method \(T\) to the outcome
+\(Y\) without opening any new ones. Under these conditions, the ACE can
+be estimated from associational quantities as
 \(\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
-where \(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
-\citep{Morgan_et_al_2014}. Of course, the validity of the claims about
-the effects of \(T\) on \(Y\) now hinges on the assumption that \(X\)
-serves as a sufficient adjustment set. However, as
-\citet[pp.~150]{Kohler_et_al_2019} noted, ``statements about the real
-world from observed (not observational) data require assumptions (and)
-this is true no matter what kind of data we have''. For example, if
-researchers cannot conduct the randomized experiments described in
-Section~\ref{sec-appendix-A} and must rely on observational data, they
-can still estimate \(\tau\) if a variable \(X\), such as the
-socio-economic status of the school, blocks all (and not open any)
-non-causal paths from the teaching method \(T\) to the outcome \(Y\).
-
-Several approaches to causal inference and counterfactuals exist, but
-two are particularly prominent: the potential outcomes approach, also
-known as the Neyman-Rubin causal model
-\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
-\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches use rigorous
-mathematical notation to characterize the ACE, \emph{not only for
-dichotomous but also for continuous variables} \(T\), though they do so
-in different ways \citep{Neal_2020}. The potential outcomes approach
-relies on counterfactual notation, as seen in the above examples,
-whereas the structural approach uses do-calculus \citep{Pearl_2009}.
-Despite these differences, both notations can be expressed in terms of
-the other, and both approaches provide methods for using experimental
-and observational data to identify and estimate causal effects
-\citep{Pearl_2010}.
+where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
+\citep{Morgan_et_al_2014}.
 
@@
 
-The structural approach to causal inference, however, offers a notable
-advantage over the potential outcomes approach. It enables researchers
-to formally and graphically represent a system using SCMs and DAGs
-\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. SCMs
-and DAGs function as conceptual models describing the presumed causal
-structure underlying the ACE. In essence, these models act as
-\emph{theoretical models} that guide researchers in determining which
-statistical models can yield valid causal inferences, assuming the
-depicted causal structure of the models is correct
-\citep{McElreath_2020}. Notably, every SCM has a associated DAG
-\citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a visual
-representation of the role of theoretical models in the inference
+The structural approach to causal inference uses SCMs and DAGs to
+formally and graphically represent the presumed causal structure
+underlying the ACE
+\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. In
+essence, SCMs and DAGs act as \emph{conceptual models} that guide
+researchers in determining which statistical models can yield valid
+causal inferences, assuming the depicted causal structure of the models
+is correct \citep{McElreath_2020}. Notably, every SCM has an associated
+DAG \citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a
+visual representation of the role of theoretical models in the inference
 process.
@@
 SCMs and DAGs offer two key advantages for modeling causal structures.
-First, they enable researchers to represent causal relations in a
+First, they enable the representation of causal relations in a
 non-parametric and fully interactive manner. This feature allows for
@@
 \ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
-display the five fundamental building blocks of SCMs and DAGs. The left
-panels of the figures show the formal mathematical models, represented
-by the SCMs, defined in terms of a set of \emph{endogenous} variables
+display these five fundamental building blocks. The left panels of the
+figures show the formal mathematical models, represented by the SCMs,
+defined in terms of a set of \emph{endogenous} variables
 \(V=\{X,Z,Y\}\), a set of \emph{exogenous} variables
@@
 \citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
-are those whose causal mechanisms a researchers chooses to model
+are those whose causal mechanisms a researcher chooses to model
 \citep{Neal_2020}. In contrast, exogenous variables represent
@@
 \citep{Morgan_et_al_2014}. Finally, the arrows in the graphs show the
-expected direction of causal influences among these variables. Often,
-DAGs omit the exogenous variables for simplicity (the \emph{standard}
-representation). However, including exogenous variables in a graph (the
-\emph{magnified} representation) can be beneficial, as their presence
-can reveal potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}.
+expected direction of causal influences among these variables. While
+DAGs often omit exogenous variables for simplicity (the \emph{standard}
+representation), including these variables is advantageous (the
+\emph{magnified} representation shown in the figures), as their
+inclusion helps to identify potential issues related to conditioning and
+confounding \citep{Cinelli_et_al_2020}.
 
@@
 
-Researchers can use these building blocks to depict the scenario
-described in Section~\ref{sec-appendix-B}. In this scenario, they rely
-on observational data because it is not feasible to run a randomized
-experiment. Furthermore, the scenario indicates that researchers have
-measured a variable \(X\) (the socio-economic status of the school) that
-is assumed to block all non-causal paths from the teaching method \(T\)
-to the outcome \(Y\). Figures \ref{fig-scm_example1} and
-\ref{fig-mdag_example1} illustrate the plausible causal structure for
-this example. The figures highlight the presence of at least four of the
-five fundamental building blocks. The figures display multiple
-descendants, as shown by pairwise relationships such as
-\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
-also feature multiple pairs of unconnected nodes, evident from the
-relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
+Researchers can use these building blocks to represent the scenario
+outlined in Section~\ref{sec-appendix-B}. Figures \ref{fig-scm_example1}
+and \ref{fig-mdag_example1} depict the plausible causal structure for
+this example, assuming that the variable \(X\) (socio-economic status of
+the school) acts as a confounder in the relationship between the
+teaching method \(T\) and the outcome \(Y\). The figures show several
+descendant, such as \(X \rightarrow T\), \(X \rightarrow Y\), and
+\(T \rightarrow Y\), and also highlight multiple pairs of unconnected
+nodes, evident from the relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
 \(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
-Finally, the figures illustrate one fork, \(X \rightarrow \{T, Y\}\),
-and two colliders: \(\{X, e_{T}\} \rightarrow T\) and
+Additional, the figures depict one fork, \(X \rightarrow \{T, Y\}\), and
+two colliders: \(\{X, e_{T}\} \rightarrow T\) and
 \(\{X, T, e_{Y}\} \rightarrow Y\).
@@
 
-\caption{\label{fig-example}Plausible causal structure for the example.}
+\caption{\label{fig-example}Plausible causal structure the scenario
+outlined in Section~\ref{sec-appendix-B}.}
 

### 2025-01-11 paper2.qmd
section:appendix; process:write,figures

@@
 The model also incorporates various assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-Finally, the section employs Bayesian statistical methods to transform these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
+Finally, the section employs Bayesian inference methods to transform these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
 See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section.
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
+
+<!-- 2. What are the benefits of Bayesian inference?-->
+Bayesian inference procedures offer three key advantages.
+First, they are well-suited to handling complex and overparameterized models, enabling researchers to estimate models where the number of parameters exceeds the number of observations for estimation [@Baker_1998; @Kim_et_al_1999].
+Second, they allow researchers to incorporate prior information, which helps constrain parameters within specified bounds.
+This capability addresses challenges such as non-convergence or improper parameter estimation that often arise in complex models when analyzed with frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
+Finally, Bayesian methods are particularly effective at drawing inferences from small sample sizes, where relying on the asymptotic properties of frequentist approaches may not be justified [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
 
@@
 
-## Empirical research under causal inference {#sec-appendix-B .appendix appendix-style=plain}
+## Identification under causal inference {#sec-appendix-B .appendix appendix-style=plain}
 
@@
 The framework uses rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" This question introduces the concept of counterfactuals, which are instrumental in understanding and defining causal effects.
+This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" This question introduces the concept of counterfactuals, which are instrumental in defining and identifying causal effects.
 
@@
 To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts.
-Notably, the approach extends the ACEs for categorical and continuous variables $T$, ensuring broad applicability across different causal scenarios.
 
@@
 For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and instead rely on observational data, they can still estimate the ACE provided a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ to the outcome $Y$  without opening any new ones.
-Under these conditions, the ACE can be estimated from associational quantities as $\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
+Under these conditions, the ACE can be estimated from associational quantities as $\tau =$ $E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)] =$ $E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
+Notably, the approach extends the ACE for a continuous variable $T$ as $\tau =$ $E[Y_{i}|do(T_{i}=t)] =$ $d E_{X}\left[ E[Y_{i}| T_{i}=t, X]\right]/ dt$, ensuring broad applicability across different causal scenarios [@Neal_2020, pp.
+45].
 
@@
 Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display these five fundamental building blocks.
-The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X,Z,Y\}$, a set of *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020].
+The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \dsep e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{3} & := f_{X3}(e_{X3}) \\
+  e_{X1} & \dsep e_{X3}
 \end{aligned}
@@
 
-<!-- ![Two unconnected nodes](images/png/sdag_bb1.png){#fig-sdag_bb1 fig-align="center" fig-pos="H" width=35%} -->
-
 Two unconnected nodes
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(X,e_{Y}) \\
-  e_{X} & \dsep e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{3} & := f_{X3}(X_{1},e_{X3}) \\
+  e_{X1} & \dsep e_{X3}
 \end{aligned}
@@
 
-<!-- ![Two connected nodes or descendant](images/png/sdag_bb2.png){#fig-sdag_bb2 fig-align="center" fig-pos="H" width=35%} -->
-
 Two connected nodes or descendant
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Z} \\
-  e_{Z} & \dsep e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{2} & := f_{X2}(X_{1},e_{X2}) \\
+  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
+  e_{X1} & \dsep e_{X2} \\
+  e_{X1} & \dsep e_{X3} \\
+  e_{X2} & \dsep e_{X3}
 \end{aligned}
@@
 
-<!-- ![Chain or pipe](images/png/sdag_bb3.png){#fig-sdag_bb3 fig-align="center" fig-pos="H" width=55%} -->
-
 Chain or mediator
@@
 \begin{aligned}
-  X & := f_{X}(Z,e_{X}) \\
-  Z & := f_{Z}(e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Z} \\
-  e_{Z} & \dsep e_{Y}
+  X_{1} & := f_{X1}(X_{2},e_{X1}) \\
+  X_{2} & := f_{X2}(e_{X2}) \\
+  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
+  e_{X1} & \dsep e_{X2} \\
+  e_{X1} & \dsep e_{X3} \\
+  e_{X2} & \dsep e_{X3}
 \end{aligned}
@@
 
-<!-- ![Fork](images/png/sdag_bb4.png){#fig-sdag_bb4 fig-align="center" fig-pos="H" width=55%} -->
-
 Fork or confounder
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,Y,e_{Z}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Z} \\
-  e_{Z} & \dsep e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{2} & := f_{X2}(X_{1},X_{3},e_{X2}) \\
+  X_{3} & := f_{X3}(e_{X3}) \\
+  e_{X1} & \dsep e_{X2} \\
+  e_{X1} & \dsep e_{X3} \\
+  e_{X2} & \dsep e_{X3}
 \end{aligned}
@@
 ![DAG](images/png/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" fig-pos="H" width=60%}
-
-<!-- ![Collider or inmorality](images/png/sdag_bb5.png){#fig-sdag_bb5 fig-align="center" fig-pos="H" width=55%} -->
 
@@
 A careful examination of the figures highlights the assumptions underlying these building blocks.
-Figures [-@fig-scm_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related.
-Figures [-@fig-scm_bb2] and [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$.
-Consequently, $Y$ is considered a *descendant* of $X$.
-Figures [-@fig-scm_bb3] and [-@fig-mdag_bb3] depict a *chain*, where $X$ influences $Z$, and $Z$ influences $Y$.
-In this configuration, $X$ is a parent node of $Z$, which is a parent node of $Y$.
-This structure creates a *directed path* between $X$ and $Y$.
-Consequently, $X$ is an *ancestor* of $Y$, and $Z$ fully *mediates* the relationship between the two.
-Figures [-@fig-scm_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$.
-Here, $Z$ is a parent node that *confounds* the relationship between $X$ and $Y$.
-Finally, Figures [-@fig-scm_bb5] and [-@fig-mdag_bb5] show a *collider*, where variables $X$ and $Y$ are concurrent causes of $Z$.
-In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$ (an *inmorality*).
-Notably, in all SCMs, the errors are assumed to be independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.
+Figures [-@fig-scm_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are not causally related.
+Figures [-@fig-scm_bb2] and [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X_{1}$ exerts a causal influence on a *child* node $X_{3}$.
+In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$.
+Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because an edge directly connects them.
+Figures [-@fig-scm_bb3] and [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
+In this configuration, $X_{1}$ is a parent node of $X_{2}$, which is a parent node of $X_{3}$.
+This structure creates a *directed path* between $X_{1}$ and $X_{3}$.
+Consequently, $X_{1}$ is an *ancestor* of $X_{3}$, and $X_{2}$ fully *mediates* the relationship between the two.
+Figures [-@fig-scm_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
+Here, $X_{2}$ is a parent node that *confounds* the relationship between $X_{1}$ and $X_{3}$.
+Finally, Figures [-@fig-scm_bb5] and [-@fig-mdag_bb5] show a *collider*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
+In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$ (an *inmorality*).
+Notably, in all SCMs, the errors are assumed to be independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X1} \dsep e_{X2}$, $e_{X1} \dsep e_{X3}$, and $e_{X2} \dsep e_{X3}$.
 
@@
 
-<!-- ![Standard representation](images/png/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%} -->
-
 Plausible causal structure the scenario outlined in @sec-appendix-B.
@@
 
-## The flow of association and causation in DAGs {#sec-appendix-D .appendix appendix-style=plain}
-
-<!-- 1. The general probabilistic description of data -->
-
-<!-- Maximum a posteriori (MAP) estimates are the set of parameters that maximizes the probability of $\theta$ given the data.
-By Bayes theorem the MAP estimate can be obtained by maximizing the following conditional distribution: -->
-
-<!-- $$ -->
-<!-- P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})} -->
-<!-- $$ -->
-<!-- where $P(\theta)$ is the prior distribution for the parameters $\theta$ and $P(X_{1}, X_{2}, \dots, X_{n})$ is the probability of the data averaged over all parameters.
-Since the denominator is independent of $\theta$, the Bayesian estimator is obtained by maximizing $P(X_{1}, X_{2}, \dots, X_{n} | \theta)$ with respect to $\theta$.
-This indicates that when researchers model association, they generally model the joint probability of the data given the parameters $\theta$.
--->
-
-<!-- This join probability can be further re-expressed using the chain rule of probability as $P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )$.
-However, from this description, it is easy to see how quickly the parametrization becomes intractable as the number of observed variables $n$ grows [@Neal_2020].
-To deal with this complexity, researchers typically simplify the modeling process by assuming local (in)dependencies between variables.
--->
-
-<!-- 2. But what does the building blocks mean? -->
-<!-- The five building blocks help to express joint distributions very efficiently, by formalizing local (in)dependencies into (probabilistic) statistical models [@Neal_2020].
-Notably, this formalization rest on two key assumptions: the local Markov and causal edges assumptions.
-The local Markov assumption states that given its parents in the DAG, a node is independent of all its non-descendants.
-This lead to the Bayesian Network factorization.
-However, this assumption only give us information about independencias in the distribution that the DAG implies, we need also to know about the dependencies.
-The causal edges assumption states that in a DAG, every parent is a direct cause of all its children.
-In this sense this assumption is indicating that parents and their children are dependent, and more precisely, they are causally dependent [@Neal_2020].
--->
-
-<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->
+## The probabilistic implications of DAGs {#sec-appendix-D .appendix appendix-style=plain}
+
+<!-- 1. How do we use Bayesian inference? -->
+After completing the identification process outlined in @sec-appendix-B, researchers estimate the ACE using Bayesian inference methods.
+Bayesian inference calculates the probability of a set of parameters $\theta$, enabling researchers to use their distributions to represent the ACE.
+The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{n} | \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010].
+Here, $X_{n}$ represents a random variable, and for simplicity, $\theta$ denotes a parameter space of dimension one.
+After observing empirical data, researchers update these priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:
+
+$$
+P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})}
+$$ {#eq-bayes_rule}
+
+Researchers can further simplify the posterior updating process in @eq-bayes_rule into two steps: integrating new empirical data, which is defined by the likelihood, with the parameter update, which is determined by the priors, as demonstrated in equation @eq-prop_rule.
+This simplification is possible because the denominator on the right-hand side of @eq-bayes_rule acts as a normalizing constant, independent of $\theta$.
+
+$$
+P(\theta| X_{1}, X_{2}, \dots, X_{n}) \propto P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)
+$$ {#eq-prop_rule}
+
+<!-- check: -->
+<!-- https://en.wikipedia.org/wiki/Maximum_a_posteriori_estimation -->
+<!-- https://en.wikipedia.org/wiki/Bayesian_inference -->
+
+<!-- 2. But what is the problem? -->
+Temporarily setting aside the definition of prior distributions, it is important to note that the posterior updating process depends on the assumptions underlying the likelihood of the data.
+However, as the number of random variables, $n$, increases, this joint distribution becomes quickly intractable [@Neal_2020].
+This is evident from @eq-like_chain, where the likelihood distribution is re-expressed in terms of chained conditional distributions.
+
+$$
+P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )
+$$ {#eq-like_chain}
+
+<!-- 3. How do DAGs can help? -->
+To address the complexity of the likelihood, researchers simplify the distribution by assuming that variables exhibit specific local (in)dependencies.
+These assumptions improve model tractability and streamline the estimation process.
+Researchers formalize these local (in)dependencies using directed acyclic graphs (DAGs).
+DAGs represent the probabilistic implications of these assumptions by relying on three key principles: the local Markov assumption, the minimality assumption, and the causal edges assumption.
+The *local Markov assumption* defines the statistical independencies implied by a DAG.
+It states that a variable is independent of all its non-descendants, given its parents [@Neal_2020].
+The *minimality assumption* identifies the statistical dependencies implied by a DAG.
+It asserts that every pair of adjacent variables are dependent [@Neal_2020].
+Finally, the *causal edges assumption* establishes causal relationships between parents and their children in a DAG by stating that every parent is a direct cause of their children [@Neal_2020].
+Figure @fig-ACflow illustrates the flow of association and causation in graphs.
+
+![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%}
+
+<!-- 4. What are the implications of this? -->
+This lead to the *Bayesian Network factorization* of @eq-like_chain.
+
+$$
+P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | pa(X_{i}), \theta )
+$$ {#eq-net_factor}
 
@@
 -->
-
-<!-- [The advantages of SCMs and DAGs to identify causal quantities are realized after researchers examine the causal and probabilistic assumptions underlying these building blocks.]{style="color:red;"} -->
-
-<!-- The main assumption we’ve seen that relates the graph to the distribution -->
-<!-- is the Markov assumption. The Markov assumption tells us if variables are -->
-<!-- d-separated in the graph 𝐺 , then they are independent in the distribution -->
-<!-- 𝑃 (Theorem 3.1): -->
-<!-- Maybe we can detect independencies in the data and then use that -->
-<!-- to infer the causal graph. However, going from independencies in the -->
-<!-- distribution 𝑃 to d-separations in the graph 𝐺 isn’t something that the -->
-<!-- Markov assumption gives us (see Equation 3.20 above). Rather, we need -->
-<!-- the converse of the Markov assumption. This is known as the faithfulness -->
-<!-- assumption. -->
-<!-- Assumption 11.1 (Faithfulness) -->
-
-<!-- 2. What is the local markov assumption? and where does it lead? -->
-<!-- 2.1 Bayesian network factorization -->
-
-<!-- 3. What is the causal edges assumption? and where does it lead? -->
-
-<!-- 4. Combined they are known as the minimality assumption? -->
 
@@
 
-<!-- Causal mechanism as a result of markov and minimality assumption -->
-<!-- Modularity assumption also -->
-
-<!-- 4. But how do we move from targets of inference to estimates -->
-<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->
-
-![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%}
-
 <!-- For instance, in the example outlined in @sec-appendix-C, the collected observational data can be expressed as joint probability $P(X,T,Y)$.

### 2025-01-11 _tex/paper2.tex
Built site for gh-pages

@@
 
-\section{Introduction}\label{sec-introduction}
+Introduction
 
@@
 
-\section{Thurstone's theory}\label{sec-thurstone_theory}
+Thurstone's theory
 
@@
 
-\section{The two critical issues in CJ
-literature}\label{sec-theory-issues}
+The two critical issues in CJ literature
 
@@
 
-\subsection{The Case V and the statistical analysis of CJ
+\section{The Case V and the statistical analysis of CJ
 data}\label{sec-theory-issue1}
@@
 
-\subsubsection{The assumption of equal dispersions between
+\subsection{The assumption of equal dispersions between
 stimuli}\label{sec-theory-issue1a}
@@
 
-\subsubsection{The assumption of zero correlation between
+\subsection{The assumption of zero correlation between
 stimuli}\label{sec-theory-issue1b}
@@
 
-\subsection{The disconnect between trait measurement and hypothesis
+\section{The disconnect between trait measurement and hypothesis
 testing}\label{sec-theory-issue2}
@@
 
-\section{Updating the theoretical and statistical
-model}\label{sec-theory}
+Updating the theoretical and statistical model
 
@@
 experiments, such as the selection of judges, stimuli, and comparisons.
-Finally, the section employs Bayesian statistical methods to transform
+Finally, the section employs Bayesian inference methods to transform
 these theoretical and practical elements into a statistical model that
@@
 
-\subsection{The theoretical model}\label{sec-theory-theoretical}
+\section{The theoretical model}\label{sec-theory-theoretical}
 
@@
 
-\subsubsection{The population model}\label{sec-theory-theoretical_P}
+\subsection{The population model}\label{sec-theory-theoretical_P}
 
@@
 
-\subsubsection{The sample-comparison
+\subsection{The sample-comparison
 model}\label{sec-theory-theoretical_SC}
@@
 
-\subsection{From theory to statistics}\label{sec-theory-statistics}
-
-\section{Discussion}\label{sec-discuss}
-
-\subsection{Findings}\label{sec-discuss-finding}
-
-\subsection{Limitations and further
+\section{From theory to statistics}\label{sec-theory-statistics}
+
+Bayesian inference procedures offer three key advantages. First, they
+are well-suited to handling complex and overparameterized models,
+enabling researchers to estimate models where the number of parameters
+exceeds the number of observations for estimation
+\citep{Baker_1998, Kim_et_al_1999}. Second, they allow researchers to
+incorporate prior information, which helps constrain parameters within
+specified bounds. This capability addresses challenges such as
+non-convergence or improper parameter estimation that often arise in
+complex models when analyzed with frequentist methods
+\citep{Martin_et_al_1975, Seaman_et_al_2011}. Finally, Bayesian methods
+are particularly effective at drawing inferences from small sample
+sizes, where relying on the asymptotic properties of frequentist
+approaches may not be justified
+\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
+
+Discussion
+
+\section{Findings}\label{sec-discuss-finding}
+
+\section{Limitations and further
 research}\label{sec-discuss-limitations}
 
-\section{Conclusion}\label{sec-conclusion}
+Conclusion
 
@@
 
-\section*{Declarations}\label{declarations}
-\addcontentsline{toc}{section}{Declarations}
+Declarations
 
@@
 
-\section{Appendix}\label{sec-appendix}
+Appendix
 
@@
 
-\subsection{Empirical research and randomized
+\section{Empirical research and randomized
 experiments}\label{sec-appendix-A}
@@
 
-\subsection{Empirical research under causal
-inference}\label{sec-appendix-B}
+\section{Identification under causal inference}\label{sec-appendix-B}
 
@@
 under different circumstances?'' This question introduces the concept of
-counterfactuals, which are instrumental in understanding and defining
+counterfactuals, which are instrumental in defining and identifying
 causal effects.
@@
 representing the average difference between observed potential outcomes
-and their counterfactual counterparts. Notably, the approach extends the
-ACEs for categorical and continuous variables \(T\), ensuring broad
-applicability across different causal scenarios.
+and their counterfactual counterparts.
 
@@
 \(Y\) without opening any new ones. Under these conditions, the ACE can
-be estimated from associational quantities as
-\(\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
+be estimated from associational quantities as \(\tau =\)
+\(E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)] =\)
+\(E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
 where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
-\citep{Morgan_et_al_2014}.
-
-\subsection{SCMs and DAGs}\label{sec-appendix-C}
+\citep{Morgan_et_al_2014}. Notably, the approach extends the ACE for a
+continuous variable \(T\) as \(\tau =\) \(E[Y_{i}|do(T_{i}=t)] =\)
+\(d E_{X}\left[ E[Y_{i}| T_{i}=t, X]\right]/ dt\), ensuring broad
+applicability across different causal scenarios
+\citep[pp.~45]{Neal_2020}.
+
+\section{SCMs and DAGs}\label{sec-appendix-C}
 
@@
 defined in terms of a set of \emph{endogenous} variables
-\(V=\{X,Z,Y\}\), a set of \emph{exogenous} variables
-\(E=\{e_{X},e_{Z},e_{Y}\}\), and a set of functions
-\(F=\{f_{X},f_{Z},f_{Y}\}\)
+\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
+\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
+\(F=\{f_{X1},f_{X2},f_{X3}\}\)
 \citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{3} & := f_{X3}(e_{X3}) \\
+  e_{X1} & \perp\!\!\!\perp e_{X3}
 \end{aligned}
@@
 \end{minipage}%
-\newline
-\begin{minipage}{0.50\linewidth}
-
-\end{minipage}%
 
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Y & := f_{Y}(X,e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{3} & := f_{X3}(X_{1},e_{X3}) \\
+  e_{X1} & \perp\!\!\!\perp e_{X3}
 \end{aligned}
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Z} \\
-  e_{Z} & \perp\!\!\!\perp e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{2} & := f_{X2}(X_{1},e_{X2}) \\
+  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
+  e_{X1} & \perp\!\!\!\perp e_{X2} \\
+  e_{X1} & \perp\!\!\!\perp e_{X3} \\
+  e_{X2} & \perp\!\!\!\perp e_{X3}
 \end{aligned}
@@
 \begin{aligned}
-  X & := f_{X}(Z,e_{X}) \\
-  Z & := f_{Z}(e_{Z}) \\
-  Y & := f_{Y}(Z,e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Z} \\
-  e_{Z} & \perp\!\!\!\perp e_{Y}
+  X_{1} & := f_{X1}(X_{2},e_{X1}) \\
+  X_{2} & := f_{X2}(e_{X2}) \\
+  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
+  e_{X1} & \perp\!\!\!\perp e_{X2} \\
+  e_{X1} & \perp\!\!\!\perp e_{X3} \\
+  e_{X2} & \perp\!\!\!\perp e_{X3}
 \end{aligned}
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  Z & := f_{Z}(X,Y,e_{Z}) \\
-  Y & := f_{Y}(e_{Y}) \\
-  e_{X} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Z} \\
-  e_{Z} & \perp\!\!\!\perp e_{Y}
+  X_{1} & := f_{X1}(e_{X1}) \\
+  X_{2} & := f_{X2}(X_{1},X_{3},e_{X2}) \\
+  X_{3} & := f_{X3}(e_{X3}) \\
+  e_{X1} & \perp\!\!\!\perp e_{X2} \\
+  e_{X1} & \perp\!\!\!\perp e_{X3} \\
+  e_{X2} & \perp\!\!\!\perp e_{X3}
 \end{aligned}
@@
 \subcaption{\label{fig-mdag_bb5}DAG}
-
-\end{minipage}%
-\newline
-\begin{minipage}{0.50\linewidth}
 
@@
 \ref{fig-mdag_bb1} depict two unconnected nodes, representing a scenario
-where variables \(X\) and \(Y\) are not causally related. Figures
-\ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two connected nodes,
-representing a scenario where a \emph{parent} node \(X\) exerts a causal
-influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
-considered a \emph{descendant} of \(X\). Figures \ref{fig-scm_bb3} and
-\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X\) influences \(Z\),
-and \(Z\) influences \(Y\). In this configuration, \(X\) is a parent
-node of \(Z\), which is a parent node of \(Y\). This structure creates a
-\emph{directed path} between \(X\) and \(Y\). Consequently, \(X\) is an
-\emph{ancestor} of \(Y\), and \(Z\) fully \emph{mediates} the
-relationship between the two. Figures \ref{fig-scm_bb4} and
-\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
-\(Y\) are both influenced by \(Z\). Here, \(Z\) is a parent node that
-\emph{confounds} the relationship between \(X\) and \(Y\). Finally,
-Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a \emph{collider},
-where variables \(X\) and \(Y\) are concurrent causes of \(Z\). In this
-configuration, \(X\) and \(Y\) are not causally related to each other
-but both influence \(Z\) (an \emph{inmorality}). Notably, in all SCMs,
-the errors are assumed to be independent of each other and from all
-other variables in the graph, as evidenced by the pairwise relations
-\(e_{X} \perp\!\!\!\perp e_{Y}\), \(e_{X} \perp\!\!\!\perp e_{Z}\), and
-\(e_{Z} \perp\!\!\!\perp e_{Y}\).
+where variables \(X_{1}\) and \(X_{3}\) are not causally related.
+Figures \ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two
+connected nodes, representing a scenario where a \emph{parent} node
+\(X_{1}\) exerts a causal influence on a \emph{child} node \(X_{3}\). In
+this setup, \(X_{3}\) is considered a \emph{descendant} of \(X_{1}\).
+Additionally, \(X_{1}\) and \(X_{3}\) are described as \emph{adjacent}
+because an edge directly connects them. Figures \ref{fig-scm_bb3} and
+\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X_{1}\) influences
+\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In this configuration,
+\(X_{1}\) is a parent node of \(X_{2}\), which is a parent node of
+\(X_{3}\). This structure creates a \emph{directed path} between
+\(X_{1}\) and \(X_{3}\). Consequently, \(X_{1}\) is an \emph{ancestor}
+of \(X_{3}\), and \(X_{2}\) fully \emph{mediates} the relationship
+between the two. Figures \ref{fig-scm_bb4} and \ref{fig-mdag_bb4}
+illustrate a \emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are
+both influenced by \(X_{2}\). Here, \(X_{2}\) is a parent node that
+\emph{confounds} the relationship between \(X_{1}\) and \(X_{3}\).
+Finally, Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a
+\emph{collider}, where variables \(X_{1}\) and \(X_{3}\) are concurrent
+causes of \(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are
+not causally related to each other but both influence \(X_{2}\) (an
+\emph{inmorality}). Notably, in all SCMs, the errors are assumed to be
+independent of each other and from all other variables in the graph, as
+evidenced by the pairwise relations \(e_{X1} \perp\!\!\!\perp e_{X2}\),
+\(e_{X1} \perp\!\!\!\perp e_{X3}\), and
+\(e_{X2} \perp\!\!\!\perp e_{X3}\).
 
@@
 
-\subsection{The flow of association and causation in
-DAGs}\label{sec-appendix-D}
+\section{The probabilistic implications of DAGs}\label{sec-appendix-D}
+
+After completing the identification process outlined in
+Section~\ref{sec-appendix-B}, researchers estimate the ACE using
+Bayesian inference methods. Bayesian inference calculates the
+probability of a set of parameters \(\theta\), enabling researchers to
+use their distributions to represent the ACE. The approach begins by
+defining two probability distributions: the likelihood of the data,
+\(P(X_{1}, X_{2}, \dots, X_{n} | \theta)\), and the prior distribution,
+\(P(\theta)\) \citep{Everitt_et_al_2010}. Here, \(X_{n}\) represents a
+random variable, and for simplicity, \(\theta\) denotes a parameter
+space of dimension one. After observing empirical data, researchers
+update these priors to posterior distributions using Bayes' rule
+\citep{Jeffreys_1998}:
+
+\begin{equation}\phantomsection\label{eq-bayes_rule}{
+P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})}
+}\end{equation}
+
+Researchers can further simplify the posterior updating process in
+Equation~\ref{eq-bayes_rule} into two steps: integrating new empirical
+data, which is defined by the likelihood, with the parameter update,
+which is determined by the priors, as demonstrated in equation
+Equation~\ref{eq-prop_rule}. This simplification is possible because the
+denominator on the right-hand side of Equation~\ref{eq-bayes_rule} acts
+as a normalizing constant, independent of \(\theta\).
+
+\begin{equation}\phantomsection\label{eq-prop_rule}{
+P(\theta| X_{1}, X_{2}, \dots, X_{n}) \propto P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)
+}\end{equation}
+
+Temporarily setting aside the definition of prior distributions, it is
+important to note that the posterior updating process depends on the
+assumptions underlying the likelihood of the data. However, as the
+number of random variables, \(n\), increases, this joint distribution
+becomes quickly intractable \citep{Neal_2020}. This is evident from
+Equation~\ref{eq-like_chain}, where the likelihood distribution is
+re-expressed in terms of chained conditional distributions.
+
+\begin{equation}\phantomsection\label{eq-like_chain}{
+P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )
+}\end{equation}
+
+To address the complexity of the likelihood, researchers simplify the
+distribution by assuming that variables exhibit specific local
+(in)dependencies. These assumptions improve model tractability and
+streamline the estimation process. Researchers formalize these local
+(in)dependencies using directed acyclic graphs (DAGs). DAGs represent
+the probabilistic implications of these assumptions by relying on three
+key principles: the local Markov assumption, the minimality assumption,
+and the causal edges assumption. The \emph{local Markov assumption}
+defines the statistical independencies implied by a DAG. It states that
+a variable is independent of all its non-descendants, given its parents
+\citep{Neal_2020}. The \emph{minimality assumption} identifies the
+statistical dependencies implied by a DAG. It asserts that every pair of
+adjacent variables are dependent \citep{Neal_2020}. Finally, the
+\emph{causal edges assumption} establishes causal relationships between
+parents and their children in a DAG by stating that every parent is a
+direct cause of their children \citep{Neal_2020}. Figure
+Figure~\ref{fig-ACflow} illustrates the flow of association and
+causation in graphs.
 
@@
 
+This lead to the \emph{Bayesian Network factorization} of
+Equation~\ref{eq-like_chain}.
+
+\begin{equation}\phantomsection\label{eq-net_factor}{
+P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | pa(X_{i}), \theta )
+}\end{equation}
+
 \newpage{}
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
+References
 

### 2025-01-14 paper2.qmd
section:appendix; process:write

@@
 
+In this section we focus on the identification of several parameters of interest from the theoretical model, together with various assessment design features relevant to CJ experiments.
+The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+5]
+
 <!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
@@
 
-<!-- The main reason to focus on identification as a first step is that if a parameter cannot be identified using “infinite” and error free data, we certainly cannot learn anything about it using finite data.1 So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
-5] -->
-
 Assuming population data or more commonly known as census data, we ...
@@
 
-<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->
-
-<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->
-
-<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->
+![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%}
+
+![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%}
+
+![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%}
 
@@
 
-<!-- ![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%} -->
-
-<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->
-
-<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->
+![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%}
+
+![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%}
+
+![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%}
+
+![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%}
+
+![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%}
 
@@
 
-<!-- ![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%} -->
+![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%}
 
@@
 
-<!-- ![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%} -->
+![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%}
 
@@
 
-<!-- ![](/images/png/CJ_TM_15.png){#fig-CJ_TM_15 width=90%} -->
+![](/images/png/CJ_TM_15.png){#fig-CJ_TM_15 width=90%}
 
@@
 Researchers would then proceed to identify the estimands.
-*Identification* refers to the process of accurately computing an estimand using an estimator.
+*Identification* is concerned with determining whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+4].
 An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
@@
 The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.
+
+<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+4].
+-->
 
@@
 
-<!-- 1. What does the structural approach uses? -->
+<!-- 1. The benefits of SCMs and DAGs -->
 The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-In essence, SCMs and DAGs act as *conceptual models* that guide researchers in determining which statistical models can yield valid causal inferences, assuming the depicted causal structure of the models is correct [@McElreath_2020].
-Notably, every SCM has an associated DAG [@Cinelli_et_al_2020].
-@fig-IEflow provides a visual representation of the role of theoretical models in the inference process.
+Assuming the depicted causal structure is correct, SCMs and DAGs serve as *conceptual (theoretical) models* that help researchers determine which statistical models can identify an estimand, thus enabling valid causal inference [@McElreath_2020].
+They achieve this by representing various causal structures using only five fundamental building blocks, regardless of complexity [@Neal_2020; @McElreath_2020].
+This feature allows researchers to decompose complex structures into manageable components [@McElreath_2020].
+Furthermore, SCMs and DAGs can depict causal relationships in a non-parametric and fully interactive way.
+This flexibility enables feasible ACE identification strategies without requiring detailed specifications of the functional relating the variables, their parameters, or their data types [@Pearl_et_al_2016].
+@fig-IEflow shows how theoretical models guide the inference process.
 
@@
 
-<!-- 2. What are the advantages of SCMs and DAGs? -->
-SCMs and DAGs offer two key advantages for modeling causal structures.
-First, they enable the representation of causal relations in a non-parametric and fully interactive manner.
-This feature allows for feasible ACE identification strategies without specifying the data type or the nature of the functional dependence among variables [@Morgan_et_al_2014].
-Second, regardless of complexity, they can represent a wide range of causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-This feature allows researchers to decompose complex structures, facilitating their analysis by focusing on the causal assumptions associated with each building block [@McElreath_2020].
-
-<!-- 3. the SCMs -->
+<!-- 2. the SCMs -->
 Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display these five fundamental building blocks.
-The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020].
+The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 
-<!-- 4. The DAGs -->
-The right panels of the figures display the complementary DAGs.
-A DAG consists of nodes connected by edges, where nodes represent random variables.
-The term *directed* means the edges extend from one node to another, with arrows indicating the direction of causal influence.
-The term *acyclic* signifies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
+<!-- 3. The DAGs -->
+Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
+The right panels of the figures display these DAGs.
+A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables.
+The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence.
+The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
 DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-Finally, the arrows in the graphs show the expected direction of causal influences among these variables.
+Finally, the arrows in the graphs indicate the expected direction of causal influences between these variables.
 While DAGs often omit exogenous variables for simplicity (the *standard* representation), including these variables is advantageous (the *magnified* representation shown in the figures), as their inclusion helps to identify potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
@@
 
-<!-- 5. Implications of the basic building block of the SCMS and DAGs -->
-A careful examination of the figures highlights the assumptions underlying these building blocks.
+<!-- 4. Theoretical implications of the basic building block of the SCMS and DAGs -->
+A careful examination of the figures highlights the theoretical assumptions underlying these building blocks.
 Figures [-@fig-scm_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are not causally related.
@@
 
-<!-- 6. continuing with the motivating example -->
+<!-- 5. Continuing with the motivating example -->
 Researchers can use these building blocks to represent the scenario outlined in @sec-appendix-B.
@@
 
-## The probabilistic implications of DAGs {#sec-appendix-D .appendix appendix-style=plain}
+## The probabilistic implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}
+
+<!-- 1. How the SCMS and DAGs encode probabilistic information? -->
+In addition to their graphical capabilities, SCMs and DAGs can encode probabilistic information about causal structures.
+They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, and the causal edges assumption.
+The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020].
+Meanwhile, the *minimality assumption* encodes probabilistic dependencies among variables by stating that every pair of adjacent nodes exhibits a dependency [@Neal_2020].
+Finally, the *causal edges assumption* encodes causal relationships between variables by declaring that each parent node acts as a direct cause of its children [@Neal_2020].
+@fig-ACflow illustrates how these assumptions influence the statistical and causal interpretations of graphs.
+
+![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%}
+
+<!-- 2. What are the general implications of this? -->
+The encoding of probabilistic information in SCMs and DAGs facilitates the efficient representation of the joint distribution of variables [@Pearl_et_al_2016, pp.
+29].
+A notable implication of the assumptions underlying this encoding is that any conceptual model described by an SCM and DAG can express the joint distribution of variables as a product of conditional probability distributions (CPDs) in the form $P(child \: | \: parents)$.
+This property is formally known as the *Bayesian Network factorization* (@eq-net_factor) [@Pearl_et_al_2016, pp.
+29; @Neal_2020, pp.
+21].
+In this factorization, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.
+
+$$
+P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \: | \: pa(X_{i}) )
+$$ {#eq-net_factor}
+
+<!-- 3. What is the net result of this? -->
+Ultimately, this encoding enables researchers with conceptual (theoretical) knowledge in the form of an SCM and DAG to predict patterns of (in)dependencies in the data.
+As highlighted by @Pearl_et_al_2016 [pp.
+35], these predictions depend solely on the structure of these conceptual models without requiring the quantitative details of the equations or the distributions of the errors.
+Moreover, once researchers observe empirical data, the patterns of (in)dependencies in this data can provide significant insights into the validity of the proposed conceptual model.
+
+<!-- The net result is that a researcher who has scientific knowledge in the form of a structural equation model is able to predict patterns of independencies in the data, based solely on the structure of the model’s graph, without relying on any quantitative information carried by the equations or by the distributions of the errors [@Pearl_et_al_2016, pp.
+35].
+-->
+
+The concept of predicting patterns of (in)dependencies in the data is more clearly illustrated through the five fundamental building blocks outlined in @sec-appendix-C. 
+<!-- For instance,the joint probability for the variables in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] can be expressed in the following form: .
+Furthermore, the (in)dependencies shared by data sets generated by SCMs and DAGs with the same structure are the following: -->
+
+<!-- SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2] -->
+
+<!-- SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] -->
+<!-- Chain: Ux -> X -> Y -> Z <- Uz [@Pearl_et_al_2016, pp. 37] -->
+<!-- 1. Z and Y are dependent -->
+<!-- For some z, y, P(Z = z|Y = y) ≠ P(Z = z) -->
+<!-- 2. Y and X are dependent -->
+<!-- For some y, x, P(Y = y|X = x) ≠ P(Y = y) -->
+<!-- 3. Z and X are likely dependent -->
+<!-- For some z, x, P(Z = z|X = x) ≠ P(Z = z) -->
+<!-- 4. Z and X are independent, conditional on Y -->
+<!-- For all x, y, z, P(Z = z|X = x, Y = y) = P(Z = z|Y = y)' -->
+
+<!-- SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4] -->
+<!-- Fork: Uy -> Y <- X -> Z <- Uz [@Pearl_et_al_2016, pp. 40] -->
+<!-- 1. X and Y are dependent. -->
+<!-- For some x, y, P(X = x|Y = y) ≠ P(X = x) -->
+<!-- 2. X and Z are dependent. -->
+<!-- For some x, z, P(X = x|Z = z) ≠ P(X = x) -->
+<!-- 3. Z and Y are likely dependent. -->
+<!-- For some z, y, P(Z = z|Y = y) ≠ P(Z = z) -->
+<!-- 4. Y and Z are independent, conditional on X. -->
+<!-- For all 4x, y, z, P(Y = y|Z = z, X = x) = P(Y = y|X = x) -->
+
+<!-- SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] -->
+<!-- Collider: Ux -> X -> Z <- Y <- Uy [@Pearl_et_al_2016, pp. 31] -->
+<!-- 1. X and Z are dependent. -->
+<!-- For some x, z, P(X = x|Z = z) ≠ P(X = x) -->
+<!-- 2. Y and Z are dependent. -->
+<!-- For some y, z, P(Y = y|Z = z) ≠ P(Y = y) -->
+<!-- 3. X and Y are independent. -->
+<!-- For all x, y, P(X = x|Y = y) = P(X = x) -->
+<!-- 4. X and Y are dependent conditional on Z. -->
+<!-- For some x, y, z, P(X = x|Y = y, Z = z) ≠ P(X = x|Z = z) -->
 
 <!-- 1. How do we use Bayesian inference? -->
-After completing the identification process outlined in @sec-appendix-B, researchers estimate the ACE using Bayesian inference methods.
+<!-- Researchers can estimate the ACE using Bayesian inference methods.
 Bayesian inference calculates the probability of a set of parameters $\theta$, enabling researchers to use their distributions to represent the ACE.
-The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{n} | \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010].
-Here, $X_{n}$ represents a random variable, and for simplicity, $\theta$ denotes a parameter space of dimension one.
-After observing empirical data, researchers update these priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:
-
-$$
-P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})}
-$$ {#eq-bayes_rule}
-
-Researchers can further simplify the posterior updating process in @eq-bayes_rule into two steps: integrating new empirical data, which is defined by the likelihood, with the parameter update, which is determined by the priors, as demonstrated in equation @eq-prop_rule.
+The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} | \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010].
+Here, $X_{P}$ represents a random variable, and for simplicity, $\theta$ denotes a parameter space of dimension one.
+After observing empirical data, researchers update these priors to posterior distributions using Bayes' rule [@Jeffreys_1998]: -->
+
+<!-- $$ -->
+<!-- P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})} -->
+<!-- $$ {#eq-bayes_rule} -->
+
+<!-- Researchers can further simplify the posterior updating process in @eq-bayes_rule into two steps: integrating new empirical data, which is defined by the likelihood, with the parameter update, which is determined by the priors, as demonstrated in equation @eq-prop_rule.
 This simplification is possible because the denominator on the right-hand side of @eq-bayes_rule acts as a normalizing constant, independent of $\theta$.
-
-$$
-P(\theta| X_{1}, X_{2}, \dots, X_{n}) \propto P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)
-$$ {#eq-prop_rule}
+-->
+
+<!-- $$ -->
+<!-- P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\: | \: \theta) \cdot P(\theta) -->
+<!-- $$ {#eq-prop_rule} -->
 
@@
 <!-- 2. But what is the problem? -->
-Temporarily setting aside the definition of prior distributions, it is important to note that the posterior updating process depends on the assumptions underlying the likelihood of the data.
-However, as the number of random variables, $n$, increases, this joint distribution becomes quickly intractable [@Neal_2020].
+<!-- Temporarily setting aside the definition of prior distributions, it is important to note that the posterior updating process depends on the assumptions underlying the likelihood of the data.
+However, as the number of random variables, $p$, increases, this joint distribution becomes quickly intractable [@Neal_2020].
 This is evident from @eq-like_chain, where the likelihood distribution is re-expressed in terms of chained conditional distributions.
-
-$$
-P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )
-$$ {#eq-like_chain}
+-->
+
+<!-- $$ -->
+<!-- P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: X_{i-1}, \dots, X_{1}, \theta ) -->
+<!-- $$ {#eq-like_chain} -->
 
 <!-- 3. How do DAGs can help? -->
-To address the complexity of the likelihood, researchers simplify the distribution by assuming that variables exhibit specific local (in)dependencies.
+<!-- To address the complexity of the likelihood, researchers simplify the distribution by assuming that variables exhibit specific local (in)dependencies.
 These assumptions improve model tractability and streamline the estimation process.
 Researchers formalize these local (in)dependencies using directed acyclic graphs (DAGs).
-DAGs represent the probabilistic implications of these assumptions by relying on three key principles: the local Markov assumption, the minimality assumption, and the causal edges assumption.
-The *local Markov assumption* defines the statistical independencies implied by a DAG.
-It states that a variable is independent of all its non-descendants, given its parents [@Neal_2020].
-The *minimality assumption* identifies the statistical dependencies implied by a DAG.
-It asserts that every pair of adjacent variables are dependent [@Neal_2020].
-Finally, the *causal edges assumption* establishes causal relationships between parents and their children in a DAG by stating that every parent is a direct cause of their children [@Neal_2020].
-Figure @fig-ACflow illustrates the flow of association and causation in graphs.
-
-![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%}
-
-<!-- 4. What are the implications of this? -->
-This lead to the *Bayesian Network factorization* of @eq-like_chain.
-
-$$
-P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | pa(X_{i}), \theta )
-$$ {#eq-net_factor}
-
-<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables -->
-<!-- with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
 -->

### 2025-01-14 _tex/paper2.tex
Built site for gh-pages

@@
 
+In this section we focus on the identification of several parameters of
+interest from the theoretical model, together with various assessment
+design features relevant to CJ experiments. The main reason to focus on
+identification is that if these parameters cannot be identified using
+``infinite'' and error free data, we certainly cannot learn anything
+about it using finite data. So, for the time being, we assume large
+samples and error-free measurements \citep[pp.~5]{Schuessler_et_al_2023}
+
 Assuming population data or more commonly known as census data, we
@@
 
+\includegraphics[width=0.23\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_01.png}
+
+}
+
+\caption{\label{fig-CJ_TM_01}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.32\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_02.png}
+
+}
+
+\caption{\label{fig-CJ_TM_02}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
+
+}
+
+\caption{\label{fig-CJ_TM_03}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
@@
 
+\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_05.png}
+
+}
+
+\caption{\label{fig-CJ_TM_05}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.59\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_06.png}
+
+}
+
+\caption{\label{fig-CJ_TM_06}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_07.png}
+
+}
+
+\caption{\label{fig-CJ_TM_07}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_08.png}
+
+}
+
+\caption{\label{fig-CJ_TM_08}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
+
+}
+
+\caption{\label{fig-CJ_TM_09}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_10.png}
@@
 
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}
+
+}
+
+\caption{\label{fig-CJ_TM_11}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}
@@
 
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}
+
+}
+
+\caption{\label{fig-CJ_TM_13}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}
@@
 \caption{\label{fig-CJ_TM_14}}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}
+
+}
+
+\caption{\label{fig-CJ_TM_15}}
 
@@
 Researchers would then proceed to identify the estimands.
-\emph{Identification} refers to the process of accurately computing an
-estimand using an estimator. An \emph{estimator} is a method or function
-that transforms data into an estimate \citep{Neal_2020}.
-\emph{Estimates} are numerical values that approximate the estimand and
-are derived through \emph{estimation}, which refers to the process of
-integrating data with an estimator \citep{Everitt_et_al_2010}. The
-Identification-Estimation flowchart \citep{McElreath_2020, Neal_2020} in
-Figure~\ref{fig-IEflow} provides a visual representation of the process
-of transitioning from estimands to estimates.
+\emph{Identification} is concerned with determining whether an estimator
+can accurately compute an estimand based on its assumptions, regardless
+of random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
+\emph{estimator} is a method or function that transforms data into an
+estimate \citep{Neal_2020}. \emph{Estimates} are numerical values that
+approximate the estimand and are derived through \emph{estimation},
+which refers to the process of integrating data with an estimator
+\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
+\citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow} provides a
+visual representation of the process of transitioning from estimands to
+estimates.
 
@@
 underlying the ACE
-\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. In
-essence, SCMs and DAGs act as \emph{conceptual models} that guide
-researchers in determining which statistical models can yield valid
-causal inferences, assuming the depicted causal structure of the models
-is correct \citep{McElreath_2020}. Notably, every SCM has an associated
-DAG \citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a
-visual representation of the role of theoretical models in the inference
-process.
-
-SCMs and DAGs offer two key advantages for modeling causal structures.
-First, they enable the representation of causal relations in a
-non-parametric and fully interactive manner. This feature allows for
-feasible ACE identification strategies without specifying the data type
-or the nature of the functional dependence among variables
-\citep{Morgan_et_al_2014}. Second, regardless of complexity, they can
-represent a wide range of causal structures using only five fundamental
-building blocks \citep{Neal_2020, McElreath_2020}. This feature allows
-researchers to decompose complex structures, facilitating their analysis
-by focusing on the causal assumptions associated with each building
-block \citep{McElreath_2020}.
+\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
+Assuming the depicted causal structure is correct, SCMs and DAGs serve
+as \emph{conceptual (theoretical) models} that help researchers
+determine which statistical models can identify an estimand, thus
+enabling valid causal inference \citep{McElreath_2020}. They achieve
+this by representing various causal structures using only five
+fundamental building blocks, regardless of complexity
+\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
+decompose complex structures into manageable components
+\citep{McElreath_2020}. Furthermore, SCMs and DAGs can depict causal
+relationships in a non-parametric and fully interactive way. This
+flexibility enables feasible ACE identification strategies without
+requiring detailed specifications of the functional relating the
+variables, their parameters, or their data types
+\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows how theoretical
+models guide the inference process.
 
@@
 \(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
-\(F=\{f_{X1},f_{X2},f_{X3}\}\)
-\citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
-are those whose causal mechanisms a researcher chooses to model
-\citep{Neal_2020}. In contrast, exogenous variables represent
-\emph{errors} or \emph{disturbances} arising from omitted factors that
-the investigator chooses not to model explicitly
+\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
+Endogenous variables are those whose causal mechanisms a researcher
+chooses to model \citep{Neal_2020}. In contrast, exogenous variables
+represent \emph{errors} or \emph{disturbances} arising from omitted
+factors that the investigator chooses not to model explicitly
 \citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
@@
 
-The right panels of the figures display the complementary DAGs. A DAG
-consists of nodes connected by edges, where nodes represent random
-variables. The term \emph{directed} means the edges extend from one node
-to another, with arrows indicating the direction of causal influence.
-The term \emph{acyclic} signifies that the causal influences do not form
-loops, ensuring the influences do not cycle back on themselves
+Notably, every SCM has an associated DAG
+\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
+figures display these DAGs. A DAG is a graph consisting of nodes
+connected by edges, where the nodes represent random variables. The term
+\emph{directed} means that the edges extend from one node to another,
+with arrows indicating the direction of causal influence. The term
+\emph{acyclic} implies that the causal influences do not form loops,
+ensuring the influences do not cycle back on themselves
 \citep{McElreath_2020}. DAGs represent observed variables as solid black
 circles, while they use open circles for unobserved (latent) variables
-\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs show the
-expected direction of causal influences among these variables. While
-DAGs often omit exogenous variables for simplicity (the \emph{standard}
-representation), including these variables is advantageous (the
-\emph{magnified} representation shown in the figures), as their
-inclusion helps to identify potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}.
+\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs indicate
+the expected direction of causal influences between these variables.
+While DAGs often omit exogenous variables for simplicity (the
+\emph{standard} representation), including these variables is
+advantageous (the \emph{magnified} representation shown in the figures),
+as their inclusion helps to identify potential issues related to
+conditioning and confounding \citep{Cinelli_et_al_2020}.
 
@@
 
-A careful examination of the figures highlights the assumptions
-underlying these building blocks. Figures \ref{fig-scm_bb1} and
-\ref{fig-mdag_bb1} depict two unconnected nodes, representing a scenario
-where variables \(X_{1}\) and \(X_{3}\) are not causally related.
-Figures \ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two
+A careful examination of the figures highlights the theoretical
+assumptions underlying these building blocks. Figures \ref{fig-scm_bb1}
+and \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
+scenario where variables \(X_{1}\) and \(X_{3}\) are not causally
+related. Figures \ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two
 connected nodes, representing a scenario where a \emph{parent} node
@@
 
-\section{The probabilistic implications of DAGs}\label{sec-appendix-D}
-
-After completing the identification process outlined in
-Section~\ref{sec-appendix-B}, researchers estimate the ACE using
-Bayesian inference methods. Bayesian inference calculates the
-probability of a set of parameters \(\theta\), enabling researchers to
-use their distributions to represent the ACE. The approach begins by
-defining two probability distributions: the likelihood of the data,
-\(P(X_{1}, X_{2}, \dots, X_{n} | \theta)\), and the prior distribution,
-\(P(\theta)\) \citep{Everitt_et_al_2010}. Here, \(X_{n}\) represents a
-random variable, and for simplicity, \(\theta\) denotes a parameter
-space of dimension one. After observing empirical data, researchers
-update these priors to posterior distributions using Bayes' rule
-\citep{Jeffreys_1998}:
-
-\begin{equation}\phantomsection\label{eq-bayes_rule}{
-P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})}
-}\end{equation}
-
-Researchers can further simplify the posterior updating process in
-Equation~\ref{eq-bayes_rule} into two steps: integrating new empirical
-data, which is defined by the likelihood, with the parameter update,
-which is determined by the priors, as demonstrated in equation
-Equation~\ref{eq-prop_rule}. This simplification is possible because the
-denominator on the right-hand side of Equation~\ref{eq-bayes_rule} acts
-as a normalizing constant, independent of \(\theta\).
-
-\begin{equation}\phantomsection\label{eq-prop_rule}{
-P(\theta| X_{1}, X_{2}, \dots, X_{n}) \propto P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)
-}\end{equation}
-
-Temporarily setting aside the definition of prior distributions, it is
-important to note that the posterior updating process depends on the
-assumptions underlying the likelihood of the data. However, as the
-number of random variables, \(n\), increases, this joint distribution
-becomes quickly intractable \citep{Neal_2020}. This is evident from
-Equation~\ref{eq-like_chain}, where the likelihood distribution is
-re-expressed in terms of chained conditional distributions.
-
-\begin{equation}\phantomsection\label{eq-like_chain}{
-P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )
-}\end{equation}
-
-To address the complexity of the likelihood, researchers simplify the
-distribution by assuming that variables exhibit specific local
-(in)dependencies. These assumptions improve model tractability and
-streamline the estimation process. Researchers formalize these local
-(in)dependencies using directed acyclic graphs (DAGs). DAGs represent
-the probabilistic implications of these assumptions by relying on three
-key principles: the local Markov assumption, the minimality assumption,
-and the causal edges assumption. The \emph{local Markov assumption}
-defines the statistical independencies implied by a DAG. It states that
-a variable is independent of all its non-descendants, given its parents
-\citep{Neal_2020}. The \emph{minimality assumption} identifies the
-statistical dependencies implied by a DAG. It asserts that every pair of
-adjacent variables are dependent \citep{Neal_2020}. Finally, the
-\emph{causal edges assumption} establishes causal relationships between
-parents and their children in a DAG by stating that every parent is a
-direct cause of their children \citep{Neal_2020}. Figure
-Figure~\ref{fig-ACflow} illustrates the flow of association and
-causation in graphs.
+\section{The probabilistic implications of SCMs and
+DAGs}\label{sec-appendix-D}
+
+In addition to their graphical capabilities, SCMs and DAGs can encode
+probabilistic information about causal structures. They achieve this
+encoding by relying on three fundamental assumptions: the local Markov,
+the minimality, and the causal edges assumption. The \emph{local Markov
+assumption} encodes probabilistic independencies between variables by
+declaring that nodes in a graph are independent of all its
+non-descendants, given its parents \citep{Neal_2020}. Meanwhile, the
+\emph{minimality assumption} encodes probabilistic dependencies among
+variables by stating that every pair of adjacent nodes exhibits a
+dependency \citep{Neal_2020}. Finally, the \emph{causal edges
+assumption} encodes causal relationships between variables by declaring
+that each parent node acts as a direct cause of its children
+\citep{Neal_2020}. Figure~\ref{fig-ACflow} illustrates how these
+assumptions influence the statistical and causal interpretations of
+graphs.
 
@@
 
-This lead to the \emph{Bayesian Network factorization} of
-Equation~\ref{eq-like_chain}.
+The encoding of probabilistic information in SCMs and DAGs facilitates
+the efficient representation of the joint distribution of variables
+\citep[pp.~29]{Pearl_et_al_2016}. A notable implication of the
+assumptions underlying this encoding is that any conceptual model
+described by an SCM and DAG can express the joint distribution of
+variables as a product of conditional probability distributions (CPDs)
+in the form \(P(child \: | \: parents)\). This property is formally
+known as the \emph{Bayesian Network factorization}
+(Equation~\ref{eq-net_factor})
+\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
+In this factorization, \(pa(X_{i})\) denotes the set of variables that
+are the parents of \(X_{i}\).
 
 \begin{equation}\phantomsection\label{eq-net_factor}{
-P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | pa(X_{i}), \theta )
+P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \: | \: pa(X_{i}) )
 }\end{equation}
+
+Ultimately, this encoding enables researchers with conceptual
+(theoretical) knowledge in the form of an SCM and DAG to predict
+patterns of (in)dependencies in the data. As highlighted by
+\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
+structure of these conceptual models without requiring the quantitative
+details of the equations or the distributions of the errors. Moreover,
+once researchers observe empirical data, the patterns of
+(in)dependencies in this data can provide significant insights into the
+validity of the proposed conceptual model.
+
+The concept of predicting patterns of (in)dependencies in the data is
+more clearly illustrated through the five fundamental building blocks
+outlined in Section~\ref{sec-appendix-C}.
 

### 2025-01-16 images/text/mdag_example1a.tex
section:appendix; process:write,figues,equations

\documentclass[tikz, border=1mm]{standalone}
\usepackage{tikz} 
\usetikzlibrary{arrows.meta}
\usepackage{pgfplots}

\pgfplotsset{compat=1.18}

\begin{document}

\begin{tikzpicture}

    % dag_bb3
    \node at (-2,0) {$e_{T}$};
    \node at (-1,0) {$T$};
    \node[gray] at (-1,0.7) {$e_{X}$};
    \node[rectangle,draw,gray] at (0,0.8) {$X$};
    \node at (1,0) {$Y$};
    \node at (2,0) {$e_{Y}$};

    \draw[{Circle[open]}-{latex}{Circle}](-2,-0.27) to (-0.9,-0.27); % eT -> T (circle)
    \draw[-{latex},gray](0,0.45) to (-0.9,-0.2); % X -> T
    \draw[{Circle[open]}-{latex},gray](-1,0.47) to (0,0.47); % eX -> X (circle)
    \draw[-{latex}](-0.9,-0.27) to (0.9,-0.27); % T -> Y
    \draw[{Circle}-{latex},gray](0,0.5) to (0.9,-0.2); % X -> Y
    \draw[{Circle[open]}-{latex}{Circle}](2,-0.27) to (0.9,-0.27); % eY -> Y (circle)

\end{tikzpicture}

\end{document}

### 2025-01-16 paper2.qmd
section:appendix; process:write,figues,equations

@@
 ### The population model {#sec-theory-theoretical_P}
-<!-- []{style="color:red;"} -->
 
@@
 
-This section introduces fundamental statistical and causal inference concepts necessary for understanding the core theoretical principles described in @sec-theory.
+This section introduces fundamental statistical and causal inference concepts necessary for understanding the core theoretical principles described in this document.
 It does not, however, offer a comprehensive overview of causal inference methods.
@@
 Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability.
-In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random quantity $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, where $E[\cdot]$ denotes the expected value.
+In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random quantity $E[Y_{i} \: | \: T_{i}=1] - E[Y_{i} \: | \: T_{i}=2]$, where $E[\cdot]$ denotes the expected value.
 
@@
 An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
+*Estimates* are numerical values that approximate the estimand.
+They are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
 The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.
@@
 <!-- 3. But, what do we need to produce a good estimate? -->
-While numerous methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates.
-For instance, the Z-test is an estimator known for its effectiveness in comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006].
+While numerous methods can approximate an estimand, researchers prioritize estimators with desirable identification properties that ensure the accuracy of estimates.
+For instance, the Z-test is an estimator known for its effectiveness when comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006].
 If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
-The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i}| T_{i}=1]$ and $E[Y_{i}| T_{i}=2]$, respectively.
+The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \: | \: T_{i}=1]$ and $E[Y_{i} \: | \: T_{i}=2]$, respectively.
 The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.
@@
 The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals.
-For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes: $\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)$.
-Here, $do(T_{i}=t)$ represents the intervention operator, $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
+For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes: $\tau_{i} = Y_{i} \: | \: do(T_{i}=1) - Y_{i} \: | \: do(T_{i}=2)$.
+Here, $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \: | \: do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \: | \: do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
 Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable.
-To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts.
+To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i} \: | \: do(T_{i}=1)]- E[Y_{i} \: | \: do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts.
 
@@
 
-For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and instead rely on observational data, they can still estimate the ACE provided a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ to the outcome $Y$  without opening any new ones.
-Under these conditions, the ACE can be estimated from associational quantities as $\tau =$ $E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)] =$ $E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
-Notably, the approach extends the ACE for a continuous variable $T$ as $\tau =$ $E[Y_{i}|do(T_{i}=t)] =$ $d E_{X}\left[ E[Y_{i}| T_{i}=t, X]\right]/ dt$, ensuring broad applicability across different causal scenarios [@Neal_2020, pp.
-45].
-
 <!-- Statements about the real world from observed data require assumptions.
@@
 
+For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and must rely on observational data, they can still estimate the ACE, provided that a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ and the outcome $Y$ without opening any new ones.
+Under these conditions, the ACE can be estimated from the *conditional average causal effect* (CACE), defined by $E[Y_{i} \: | \: T_{i}=t, X]$.
+Using the CACE, the ACE can be expressed as $\tau =$ $E[Y_{i} \: | \: do(T_{i}=1)] - E[Y_{i} \: | \: do(T_{i}=2)] =$ $E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
+Notably, the approach extends the ACE for a continuous variable $T$ as $\tau =$ $E[Y_{i} \: | \: do(T_{i}=t)] =$ $d E_{X}\left[ E[Y_{i} \: | \: T_{i}=t, X]\right]/ dt$, ensuring broad applicability across different causal scenarios [@Neal_2020, pp.
+45].
+
 ## SCMs and DAGs {#sec-appendix-C .appendix appendix-style=plain}
@@
 \newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
+<!-- \newcommand{\ndsep}{\not\!\perp\!\!\!\perp} -->
 
@@
 The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-Assuming the depicted causal structure is correct, SCMs and DAGs serve as *conceptual (theoretical) models* that help researchers determine which statistical models can identify an estimand, thus enabling valid causal inference [@McElreath_2020].
-They achieve this by representing various causal structures using only five fundamental building blocks, regardless of complexity [@Neal_2020; @McElreath_2020].
-This feature allows researchers to decompose complex structures into manageable components [@McElreath_2020].
-Furthermore, SCMs and DAGs can depict causal relationships in a non-parametric and fully interactive way.
-This flexibility enables feasible ACE identification strategies without requiring detailed specifications of the functional relating the variables, their parameters, or their data types [@Pearl_et_al_2016].
-@fig-IEflow shows how theoretical models guide the inference process.
+Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that help researchers determine which statistical models can identify an estimand, assuming the depicted causal structure is correct, thus enabling valid causal inference [@McElreath_2020].
+They achieve this by representing various causal structures, regardless of complexity, using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
+Furthermore, SCMs and DAGs depict causal relationships in a non-parametric and fully interactive way.
+This flexibility enables feasible ACE identification strategies without requiring detailed specifications of functional relationships relating the variables, their parameters, or their data types [@Pearl_et_al_2016].
+@fig-IEflow shows the role of theoretical models in the inference process.
 
@@
 Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display these five fundamental building blocks.
-The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020].
+The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $X=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-Finally, the arrows in the graphs indicate the expected direction of causal influences between these variables.
-While DAGs often omit exogenous variables for simplicity (the *standard* representation), including these variables is advantageous (the *magnified* representation shown in the figures), as their inclusion helps to identify potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+Although DAGs typically omit exogenous variables for simplicity (the *standard representation*), including these variables, as demonstrated in the *magnified representation* in the figures, can be advantageous.
+Their inclusion helps to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 
-![DAG](images/png/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" fig-pos="H" width=50%}
+![DAG](images/png/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" fig-pos="H" width=54%}
 
@@
 
-![DAG](images/png/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" fig-pos="H" width=50%}
+![DAG](images/png/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" fig-pos="H" width=54%}
 
@@
 
-![DAG](images/png/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" fig-pos="H" width=60%}
+![DAG](images/png/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" fig-pos="H" width=65%}
 
@@
 
-![DAG](images/png/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" fig-pos="H" width=60%}
+![DAG](images/png/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" fig-pos="H" width=65%}
 
@@
 
-![DAG](images/png/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" fig-pos="H" width=60%}
+![DAG](images/png/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" fig-pos="H" width=65%}
 
@@
 A careful examination of the figures highlights the theoretical assumptions underlying these building blocks.
-Figures [-@fig-scm_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are not causally related.
-Figures [-@fig-scm_bb2] and [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X_{1}$ exerts a causal influence on a *child* node $X_{3}$.
+SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are independent or not causally related.
+SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X_{1}$ exerts a causal influence on a *child* node $X_{3}$.
 In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$.
 Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because an edge directly connects them.
-Figures [-@fig-scm_bb3] and [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
+SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
 In this configuration, $X_{1}$ is a parent node of $X_{2}$, which is a parent node of $X_{3}$.
@@
 Consequently, $X_{1}$ is an *ancestor* of $X_{3}$, and $X_{2}$ fully *mediates* the relationship between the two.
-Figures [-@fig-scm_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
+SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4] illustrate a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
 Here, $X_{2}$ is a parent node that *confounds* the relationship between $X_{1}$ and $X_{3}$.
-Finally, Figures [-@fig-scm_bb5] and [-@fig-mdag_bb5] show a *collider*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
+Finally, SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] show a *collider*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
 In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$ (an *inmorality*).
-Notably, in all SCMs, the errors are assumed to be independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X1} \dsep e_{X2}$, $e_{X1} \dsep e_{X3}$, and $e_{X2} \dsep e_{X3}$.
+Notably, all building blocks assume the errors are independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X1} \dsep e_{X2}$, $e_{X1} \dsep e_{X3}$, and $e_{X2} \dsep e_{X3}$.
 
 <!-- 5. Continuing with the motivating example -->
-Researchers can use these building blocks to represent the scenario outlined in @sec-appendix-B.
-Figures [-@fig-scm_example1] and [-@fig-mdag_example1] depict the plausible causal structure for this example, assuming that the variable $X$ (socio-economic status of the school) acts as a confounder in the relationship between the teaching method $T$ and the outcome $Y$.
-The figures show several descendant, such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$, and also highlight multiple pairs of unconnected nodes, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
-Additional, the figures depict one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.
+Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B.
+SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example.
+In this context, the variable $X$ (socio-economic status of the school) acts as a confounder in the relationship between the teaching method $T$ and the outcome $Y$.
+The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
+Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.
 
@@
 
-![DAG](images/png/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=60%}
+![DAG](images/png/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=65%}
 
@@
 
-## The probabilistic implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}
+## The probabilistic and causal implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}
 
 <!-- 1. How the SCMS and DAGs encode probabilistic information? -->
-In addition to their graphical capabilities, SCMs and DAGs can encode probabilistic information about causal structures.
-They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, and the causal edges assumption.
-The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020].
-Meanwhile, the *minimality assumption* encodes probabilistic dependencies among variables by stating that every pair of adjacent nodes exhibits a dependency [@Neal_2020].
-Finally, the *causal edges assumption* encodes causal relationships between variables by declaring that each parent node acts as a direct cause of its children [@Neal_2020].
+Beyond their graphical capabilities, SCMs and DAGs can encode the probabilistic information embedded within a causal structure.
+They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges and the modularity assumptions.
+The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020, pp.
+20].
+Meanwhile, the *minimality assumption* encodes probabilistic dependencies among variables by stating that every pair of adjacent nodes exhibits a dependency [@Neal_2020, pp.
+21].
+Finally, the *causal edges assumption* encodes causal relationships between variables by declaring that each parent node acts as a direct cause of its children [@Neal_2020, pp.
+22].
 @fig-ACflow illustrates how these assumptions influence the statistical and causal interpretations of graphs.
@@
 29].
-A notable implication of the assumptions underlying this encoding is that any conceptual model described by an SCM and DAG can express the joint distribution of variables as a product of conditional probability distributions (CPDs) in the form $P(child \: | \: parents)$.
-This property is formally known as the *Bayesian Network factorization* (@eq-net_factor) [@Pearl_et_al_2016, pp.
+A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can express the joint distribution of variables as a product of conditional probability distributions (CPDs) in the form $P(child \: | \: parents)$.
+This property is formally known as the *Bayesian Network factorization* (BNF) [@Pearl_et_al_2016, pp.
 29; @Neal_2020, pp.
 21].
-In this factorization, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.
+In this factorization, depicted in @eq-net_factor, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.
 
@@
 <!-- 3. What is the net result of this? -->
-Ultimately, this encoding enables researchers with conceptual (theoretical) knowledge in the form of an SCM and DAG to predict patterns of (in)dependencies in the data.
+This encoding enables researchers with conceptual (theoretical) knowledge in the form of an SCM and DAG to predict patterns of (in)dependencies in the data.
 As highlighted by @Pearl_et_al_2016 [pp.
@@
 Moreover, once researchers observe empirical data, the patterns of (in)dependencies in this data can provide significant insights into the validity of the proposed conceptual model.
+Moreover, it allows to express
 
@@
 
-The concept of predicting patterns of (in)dependencies in the data is more clearly illustrated through the five fundamental building blocks outlined in @sec-appendix-C. 
-<!-- For instance,the joint probability for the variables in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] can be expressed in the following form: .
-Furthermore, the (in)dependencies shared by data sets generated by SCMs and DAGs with the same structure are the following: -->
-
-<!-- SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2] -->
-
-<!-- SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] -->
-<!-- Chain: Ux -> X -> Y -> Z <- Uz [@Pearl_et_al_2016, pp. 37] -->
-<!-- 1. Z and Y are dependent -->
-<!-- For some z, y, P(Z = z|Y = y) ≠ P(Z = z) -->
-<!-- 2. Y and X are dependent -->
-<!-- For some y, x, P(Y = y|X = x) ≠ P(Y = y) -->
-<!-- 3. Z and X are likely dependent -->
-<!-- For some z, x, P(Z = z|X = x) ≠ P(Z = z) -->
-<!-- 4. Z and X are independent, conditional on Y -->
-<!-- For all x, y, z, P(Z = z|X = x, Y = y) = P(Z = z|Y = y)' -->
-
-<!-- SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4] -->
-<!-- Fork: Uy -> Y <- X -> Z <- Uz [@Pearl_et_al_2016, pp. 40] -->
-<!-- 1. X and Y are dependent. -->
-<!-- For some x, y, P(X = x|Y = y) ≠ P(X = x) -->
-<!-- 2. X and Z are dependent. -->
-<!-- For some x, z, P(X = x|Z = z) ≠ P(X = x) -->
-<!-- 3. Z and Y are likely dependent. -->
-<!-- For some z, y, P(Z = z|Y = y) ≠ P(Z = z) -->
-<!-- 4. Y and Z are independent, conditional on X. -->
-<!-- For all 4x, y, z, P(Y = y|Z = z, X = x) = P(Y = y|X = x) -->
-
-<!-- SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] -->
-<!-- Collider: Ux -> X -> Z <- Y <- Uy [@Pearl_et_al_2016, pp. 31] -->
-<!-- 1. X and Z are dependent. -->
-<!-- For some x, z, P(X = x|Z = z) ≠ P(X = x) -->
-<!-- 2. Y and Z are dependent. -->
-<!-- For some y, z, P(Y = y|Z = z) ≠ P(Y = y) -->
-<!-- 3. X and Y are independent. -->
-<!-- For all x, y, P(X = x|Y = y) = P(X = x) -->
-<!-- 4. X and Y are dependent conditional on Z. -->
-<!-- For some x, y, z, P(X = x|Y = y, Z = z) ≠ P(X = x|Z = z) -->
+<!-- 4. can you illustrate this concept? -->
+The five fundamental building blocks described in @sec-appendix-C illustrate how to predict (in)dependencies in data.
+For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$).
+This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, which defines probabilistic independence [@Neal_2020, pp.
+24].
+Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$, as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \: | \: X_{1}) P(X_{1})$.
+Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in the figures of @sec-appendix-C and the CPDs.
+
+<!-- ($X_{1} \ndsep X_{3}$) -->
+
+Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of (in)dependencies from chains, forks, and colliders.
+By leveraging these patterns, researchers can identify the ACE from associational estimates through statistical conditioning, thereby disentangling causation from mere association.
+For instance, researchers applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] can represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}) P(X_{3} \: | \: X_{2})$.
+From this expression, researchers can infer additional (in)dependencies assumed in any data generated by this structure.
+Specifically, the structure implies that $X_{1}$ and $X_{3}$ are unconditionally dependent, but conditionally independent when controlling for $X_{2}$ ($X_{1} \dsep X_{3} \: | \: X_{2}$) [@Pearl_et_al_2016, pp.
+37; @Neal_2020, pp.
+25].
+In the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1} \: | \: X_{2}) P(X_{2}) P(X_{3} \: | \: X_{2})$.
+This expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent, but conditionally independent when controlling for $X_{2}$ ($X_{1} \dsep X_{3}  \: | \: X_{2}$) [@Pearl_et_al_2016, pp.
+40].
+Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}, X_{3}) P(X_{3})$.
+This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$), but conditionally dependent when controlling for $X_{2}$ [@Pearl_et_al_2016, pp.
+41; @Neal_2020, pp.
+26].
+
+<!-- ($X_{1} \ndsep X_{3}$) -->
+<!-- ($X_{1} \ndsep X_{3}$) -->
+<!-- ($X_{1} \ndsep X_{3} \: | \: X_{2}$)  -->
+
+<!-- 5. But, how do we identify causal effects?  -->
+Ultimately, after encoding probabilistic information, researchers can use SCMs and DAGs to identify causal effects.
+This process relies on one additional assumption: the *modularity assumption*.
+This assumption asserts that intervening on a node only alters its causal mechanism, implying that interventions are local [@Neal_2020, pp.
+34].
+From this assumption, concepts such as manipulated DAGs and the Truncated factorization emerge.
+*Manipulated graphs* are DAGs with removed edges, signifying intervention.
+Similarly, the *Truncated factorization* represents these manipulated DAGs within SCMs (@eq-trunc_factor).
+In this expression, $S$ denotes the subset of manipulated variables $X_{p}$.
+
+$$
+\begin{aligned}
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = \prod_{p \not\in S} P(X_{i} \: | \: pa(X_{i}) ) \\
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = 1 \quad \quad (\text{otherwise})
+\end{aligned}
+$$ {#eq-trunc_factor}
+
+<!-- 6. Continuing with the motivating example -->
+Researchers can apply these new insights to revisit the scenario in @sec-appendix-B.
+From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can infer that the effect of $T$ on $Y$ is confounded by $X$.
+This is evident because the effect decomposes into a direct causal effect ($T \rightarrow Y$, black path) and a non-causal effect arising from the confounder $X$ ($T \leftarrow X \rightarrow Y$, gray path).
+Assuming the causal structure is correct, researchers can identify the CACE of $T$ on $Y$ by conditioning on $X$ and then marginalize this effect over $X$ to obtain the ACE.
+Conditioning on $X$ means restricting the focus to schools with a specific socio-economic status $X$ [@Neal_2020, pp.
+32].
+Notably, researchers often depict conditioning as a square around a variable.
+Furthermore, the conditioning process leads to the manipulated DAG [-@fig-mdag_example1b], equivalent to the conditioned DAG [-@fig-mdag_example1a].
+
+Using @eq-net_factor and @eq-trunc_factor, researchers can show that the joint probability of the observed variables in these structures can be expressed as $P(Y, T, X) = P(Y \: | \: T, X) P(T \: | \: X) P(X)$.
+Upon intervention, this expression becomes $P(Y, X \: | \: do(T=t)) = P(Y \: | \: T,X) P(X)$, and by marginalizing over $X$, they obtain $P(Y \: | \: do(T=t)) = \sum_{X} P(Y \: | \: T,X) P(X)$ [@Neal_2020, pp.
+36].
+From these distributions, the ACE can be expressed as $\tau = E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]$ (see @sec-appendix-B).
+
+::: {#fig-example layout-ncol=3}
+
+::: {#fig-scm_example1 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := t \\
+  Y & := f_{Y}(t,X,e_{Y}) \\
+  e_{T} & \dsep e_{X} \\
+  e_{T} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example1a fig-align="center" width=100%}
+
+![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example1b fig-align="center" width=100%}
+
+Plausible causal structure the scenario outlined in @sec-appendix-B.
+:::
+
+## The estimation process {#sec-appendix-E .appendix appendix-style=plain}
 
 <!-- 1. How do we use Bayesian inference? -->
-<!-- Researchers can estimate the ACE using Bayesian inference methods.
-Bayesian inference calculates the probability of a set of parameters $\theta$, enabling researchers to use their distributions to represent the ACE.
-The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} | \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010].
-Here, $X_{P}$ represents a random variable, and for simplicity, $\theta$ denotes a parameter space of dimension one.
-After observing empirical data, researchers update these priors to posterior distributions using Bayes' rule [@Jeffreys_1998]: -->
-
-<!-- $$ -->
-<!-- P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})} -->
-<!-- $$ {#eq-bayes_rule} -->
-
-<!-- Researchers can further simplify the posterior updating process in @eq-bayes_rule into two steps: integrating new empirical data, which is defined by the likelihood, with the parameter update, which is determined by the priors, as demonstrated in equation @eq-prop_rule.
-This simplification is possible because the denominator on the right-hand side of @eq-bayes_rule acts as a normalizing constant, independent of $\theta$.
--->
-
-<!-- $$ -->
-<!-- P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\: | \: \theta) \cdot P(\theta) -->
-<!-- $$ {#eq-prop_rule} -->
+Researchers can use Bayesian inference methods to estimate the ACE.
+The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
+After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:
+
+$$
+P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
+$$ {#eq-bayes_rule}
+
+Given that the denominator on the right-hand side of @eq-bayes_rule serves as a normalizing constant independent of the parameter $\theta$, researchers can simplify the posterior updating process into the three steps depicted by @eq-prop_rule.
+First, they integrate new empirical data through the likelihood.
+Second, they update the parameters' priors to a posterior distribution according to @eq-prop_rule.
+Ultimately, they normalize these results to obtain a valid probability distribution.
+
+$$
+P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\: | \: \theta) \cdot P(\theta)
+$$ {#eq-prop_rule}
 
@@
 <!-- 2. But what is the problem? -->
-<!-- Temporarily setting aside the definition of prior distributions, it is important to note that the posterior updating process depends on the assumptions underlying the likelihood of the data.
-However, as the number of random variables, $p$, increases, this joint distribution becomes quickly intractable [@Neal_2020].
-This is evident from @eq-like_chain, where the likelihood distribution is re-expressed in terms of chained conditional distributions.
--->
-
-<!-- $$ -->
-<!-- P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: X_{i-1}, \dots, X_{1}, \theta ) -->
-<!-- $$ {#eq-like_chain} -->
+
+Temporarily setting aside the definition of prior distributions $P(\theta)$, note that the posterior updating process depends heavily on the assumptions underlying the likelihood of the data.
+However, as the number of random variables, $P$, increases, this joint distribution becomes quickly intractable [@Neal_2020].
+This intractability is evident from @eq-like_chain, where the likelihood distribution is expressed by multiple chained CPDs.
+
+$$
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: X_{i-1}, \dots, X_{1}, \theta )
+$$ {#eq-like_chain}
 
 <!-- 3. How do DAGs can help? -->
-<!-- To address the complexity of the likelihood, researchers simplify the distribution by assuming that variables exhibit specific local (in)dependencies.
-These assumptions improve model tractability and streamline the estimation process.
-Researchers formalize these local (in)dependencies using directed acyclic graphs (DAGs).
--->
-
-<!-- 5. Now, what are those mysterious functions?, and how do they become statistical quantities? -->
-
-<!-- 6. What is the modularity assumption? and where does it lead? -->
-<!-- 6.1 the Truncated factorization -->
-<!-- 6.2 The law of counterfactuals -->
-
-<!-- 7. Causal mechanism -->
-<!-- 7.3 the backdoor criterion (do you need it?) -->
-
-<!-- 5. d-separation -->
-<!-- 5.1 Blocked path definition-->
-
-<!-- 6. putting it all together -->
-<!-- 6.1. Revising the building block: dependence and independence -->
-
-<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
-
-<!-- Mention convenient R package daggitty to draw dags and extract the causal assumption inherent in a graph -->
-
-<!-- After detailing how to build graphs and what they imply, Include steps to draw dags using a motivating example -->
-
-<!-- point out that the disturbances are usually stochastic in nature -->
-
-<!-- For instance, in the example outlined in @sec-appendix-C, the collected observational data can be expressed as joint probability $P(X,T,Y)$.
-Using the chain rule, this probability can be further factorized in $P(X,T,Y) = P(X) P(T|X) P(Y|T,X)$.
-Although this description of this data does not seem hard   -->
-
-<!-- These building blocks can then guide researchers to identify feasible ACE identification strategies.
-Researchers can use these building blocks to depict the scenario described in @sec-appendix-B.
-Assuming the variable $X$ (the socio-economic status of the school) blocks all non-causal paths from the teaching method $T$ to the outcome $Y$.
--->
-
-<!-- Conditioning on $T=t$ just means that we are restricting our focus to the subset of the population to those who received treatment $t$.
-In contrast, an intervention would be to take the whole population and give everyone treatment $t$ [@Neal_2020, pp.
-32].
--->
+Nevertheless, researchers can manage the complexity of the likelihood by assuming that variables exhibit specific local (in)dependencies.
+These assumptions enhance model tractability and facilitate the estimation process.
+As described in @sec-appendix-D, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF).
+From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution from its priors using @eq-bayes_rule.
+
+$$
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: pa(X_{i}), \theta )
+$$ {#eq-like_BNF}
 

### 2025-01-16 _tex/paper2.tex
Built site for gh-pages

@@
 
-Introduction
+\section{Introduction}\label{sec-introduction}
 
@@
 
-Thurstone's theory
+\section{Thurstone's theory}\label{sec-thurstone_theory}
 
@@
 
-The two critical issues in CJ literature
+\section{The two critical issues in CJ
+literature}\label{sec-theory-issues}
 
@@
 
-\section{The Case V and the statistical analysis of CJ
+\subsection{The Case V and the statistical analysis of CJ
 data}\label{sec-theory-issue1}
@@
 
-\subsection{The assumption of equal dispersions between
+\subsubsection{The assumption of equal dispersions between
 stimuli}\label{sec-theory-issue1a}
@@
 
-\subsection{The assumption of zero correlation between
+\subsubsection{The assumption of zero correlation between
 stimuli}\label{sec-theory-issue1b}
@@
 
-\section{The disconnect between trait measurement and hypothesis
+\subsection{The disconnect between trait measurement and hypothesis
 testing}\label{sec-theory-issue2}
@@
 
-Updating the theoretical and statistical model
+\section{Updating the theoretical and statistical
+model}\label{sec-theory}
 
@@
 
-\section{The theoretical model}\label{sec-theory-theoretical}
+\subsection{The theoretical model}\label{sec-theory-theoretical}
 
@@
 
-\subsection{The population model}\label{sec-theory-theoretical_P}
+\subsubsection{The population model}\label{sec-theory-theoretical_P}
 
@@
 
-\subsection{The sample-comparison
+\subsubsection{The sample-comparison
 model}\label{sec-theory-theoretical_SC}
@@
 
-\section{From theory to statistics}\label{sec-theory-statistics}
+\subsection{From theory to statistics}\label{sec-theory-statistics}
 
@@
 
-Discussion
-
-\section{Findings}\label{sec-discuss-finding}
-
-\section{Limitations and further
+\section{Discussion}\label{sec-discuss}
+
+\subsection{Findings}\label{sec-discuss-finding}
+
+\subsection{Limitations and further
 research}\label{sec-discuss-limitations}
 
-Conclusion
+\section{Conclusion}\label{sec-conclusion}
 
@@
 
-Declarations
+\section*{Declarations}\label{declarations}
+\addcontentsline{toc}{section}{Declarations}
 
@@
 
-Appendix
+\section{Appendix}\label{sec-appendix}
 
@@
 concepts necessary for understanding the core theoretical principles
-described in Section~\ref{sec-theory}. It does not, however, offer a
-comprehensive overview of causal inference methods. Readers seeking more
-in-depth understanding may wish to explore introductory papers such as
+described in this document. It does not, however, offer a comprehensive
+overview of causal inference methods. Readers seeking more in-depth
+understanding may wish to explore introductory papers such as
 \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
@@
 
-\section{Empirical research and randomized
+\subsection{Empirical research and randomized
 experiments}\label{sec-appendix-A}
@@
 and this estimand can be mathematically represented by the random
-quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where \(E[\cdot]\)
-denotes the expected value.
+quantity \(E[Y_{i} \: | \: T_{i}=1] - E[Y_{i} \: | \: T_{i}=2]\), where
+\(E[\cdot]\) denotes the expected value.
 
@@
 estimate \citep{Neal_2020}. \emph{Estimates} are numerical values that
-approximate the estimand and are derived through \emph{estimation},
+approximate the estimand. They are derived through \emph{estimation},
 which refers to the process of integrating data with an estimator
@@
 While numerous methods can approximate an estimand, researchers
-prioritize estimators with desirable properties that ensure the accuracy
-of estimates. For instance, the Z-test is an estimator known for its
-effectiveness in comparing groups' proportions, yielding accurate
-estimates when the underlying assumptions of the statistic are met
-\citep{Kanji_2006}. If this is the case, the Z-test is expressed as a
-signal-to-noise statistic
+prioritize estimators with desirable identification properties that
+ensure the accuracy of estimates. For instance, the Z-test is an
+estimator known for its effectiveness when comparing groups'
+proportions, yielding accurate estimates when the underlying assumptions
+of the statistic are met \citep{Kanji_2006}. If this is the case, the
+Z-test is expressed as a signal-to-noise statistic
 \(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
@@
 \(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
-\(E[Y_{i}| T_{i}=1]\) and \(E[Y_{i}| T_{i}=2]\), respectively. The
-noise, represented by \(\hat{s}_{p}\), is defined as the unpooled sample
-variability observed between the two groups.
+\(E[Y_{i} \: | \: T_{i}=1]\) and \(E[Y_{i} \: | \: T_{i}=2]\),
+respectively. The noise, represented by \(\hat{s}_{p}\), is defined as
+the unpooled sample variability observed between the two groups.
 
@@
 
-\section{Identification under causal inference}\label{sec-appendix-B}
+\subsection{Identification under causal inference}\label{sec-appendix-B}
 
@@
 between each student's potential outcomes:
-\(\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)\). Here,
-\(do(T_{i}=t)\) represents the intervention operator,
-\(Y_{i}|do(T_{i}=1)\) represents the potential outcome under
-intervention \(T_{i}=1\), and \(Y_{i}|do(T_{i}=1)\) represents the
-potential outcome under intervention \(T_{i}=2\). Note that if a student
-is assigned to intervention \(T_{i}=1\), the potential outcome under
-\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
-observable. To address the challenge of unobserved counterfactuals, the
-approach extends the ICE to the \emph{average causal effect} (ACE):
-\(\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]\),
+\(\tau_{i} = Y_{i} \: | \: do(T_{i}=1) - Y_{i} \: | \: do(T_{i}=2)\).
+Here, \(do(T_{i}=t)\) represents the intervention operator,
+\(Y_{i} \: | \: do(T_{i}=1)\) represents the potential outcome under
+intervention \(T_{i}=1\), and \(Y_{i} \: | \: do(T_{i}=1)\) represents
+the potential outcome under intervention \(T_{i}=2\). Note that if a
+student is assigned to intervention \(T_{i}=1\), the potential outcome
+under \(T_{i}=2\) becomes a counterfactual, as it is no longer observed
+nor observable. To address the challenge of unobserved counterfactuals,
+the approach extends the ICE to the \emph{average causal effect} (ACE):
+\(\tau = E[\tau_{i}] = E[Y_{i} \: | \: do(T_{i}=1)]- E[Y_{i} \: | \: do(T_{i}=2)]\),
 representing the average difference between observed potential outcomes
@@
 For instance, if researchers are unable to conduct the randomized
-experiments described in Section~\ref{sec-appendix-A} and instead rely
-on observational data, they can still estimate the ACE provided a
+experiments described in Section~\ref{sec-appendix-A} and must rely on
+observational data, they can still estimate the ACE, provided that a
 variable \(X\), such as the socio-economic status of the school, blocks
-all non-causal paths between the teaching method \(T\) to the outcome
+all non-causal paths between the teaching method \(T\) and the outcome
 \(Y\) without opening any new ones. Under these conditions, the ACE can
-be estimated from associational quantities as \(\tau =\)
-\(E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)] =\)
-\(E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
+be estimated from the \emph{conditional average causal effect} (CACE),
+defined by \(E[Y_{i} \: | \: T_{i}=t, X]\). Using the CACE, the ACE can
+be expressed as \(\tau =\)
+\(E[Y_{i} \: | \: do(T_{i}=1)] - E[Y_{i} \: | \: do(T_{i}=2)] =\)
+\(E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]\),
 where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
 \citep{Morgan_et_al_2014}. Notably, the approach extends the ACE for a
-continuous variable \(T\) as \(\tau =\) \(E[Y_{i}|do(T_{i}=t)] =\)
-\(d E_{X}\left[ E[Y_{i}| T_{i}=t, X]\right]/ dt\), ensuring broad
+continuous variable \(T\) as \(\tau =\)
+\(E[Y_{i} \: | \: do(T_{i}=t)] =\)
+\(d E_{X}\left[ E[Y_{i} \: | \: T_{i}=t, X]\right]/ dt\), ensuring broad
 applicability across different causal scenarios
@@
 
-\section{SCMs and DAGs}\label{sec-appendix-C}
+\subsection{SCMs and DAGs}\label{sec-appendix-C}
 
 \newcommand{\dsep}{\perp\!\!\!\perp}
-\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}
 
@@
 \citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
-Assuming the depicted causal structure is correct, SCMs and DAGs serve
-as \emph{conceptual (theoretical) models} that help researchers
-determine which statistical models can identify an estimand, thus
-enabling valid causal inference \citep{McElreath_2020}. They achieve
-this by representing various causal structures using only five
-fundamental building blocks, regardless of complexity
+Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
+models} that help researchers determine which statistical models can
+identify an estimand, assuming the depicted causal structure is correct,
+thus enabling valid causal inference \citep{McElreath_2020}. They
+achieve this by representing various causal structures, regardless of
+complexity, using only five fundamental building blocks
 \citep{Neal_2020, McElreath_2020}. This feature allows researchers to
-decompose complex structures into manageable components
-\citep{McElreath_2020}. Furthermore, SCMs and DAGs can depict causal
-relationships in a non-parametric and fully interactive way. This
+decompose complex structures into manageable components, facilitating
+their analysis \citep{McElreath_2020}. Furthermore, SCMs and DAGs depict
+causal relationships in a non-parametric and fully interactive way. This
 flexibility enables feasible ACE identification strategies without
-requiring detailed specifications of the functional relating the
-variables, their parameters, or their data types
-\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows how theoretical
-models guide the inference process.
+requiring detailed specifications of functional relationships relating
+the variables, their parameters, or their data types
+\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows the role of
+theoretical models in the inference process.
 
@@
 defined in terms of a set of \emph{endogenous} variables
-\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
+\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
 \(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
@@
 circles, while they use open circles for unobserved (latent) variables
-\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs indicate
-the expected direction of causal influences between these variables.
-While DAGs often omit exogenous variables for simplicity (the
-\emph{standard} representation), including these variables is
-advantageous (the \emph{magnified} representation shown in the figures),
-as their inclusion helps to identify potential issues related to
-conditioning and confounding \citep{Cinelli_et_al_2020}.
+\citep{Morgan_et_al_2014}. Although DAGs typically omit exogenous
+variables for simplicity (the \emph{standard representation}), including
+these variables, as demonstrated in the \emph{magnified representation}
+in the figures, can be advantageous. Their inclusion helps to highlight
+potential issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}.
 
@@
 
-\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}
+\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}
 
@@
 
-\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}
+\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}
 
@@
 
-\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}
+\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}
 
@@
 
-\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}
+\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}
 
@@
 
-\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}
+\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}
 
@@
 A careful examination of the figures highlights the theoretical
-assumptions underlying these building blocks. Figures \ref{fig-scm_bb1}
-and \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
-scenario where variables \(X_{1}\) and \(X_{3}\) are not causally
-related. Figures \ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two
-connected nodes, representing a scenario where a \emph{parent} node
-\(X_{1}\) exerts a causal influence on a \emph{child} node \(X_{3}\). In
-this setup, \(X_{3}\) is considered a \emph{descendant} of \(X_{1}\).
-Additionally, \(X_{1}\) and \(X_{3}\) are described as \emph{adjacent}
-because an edge directly connects them. Figures \ref{fig-scm_bb3} and
-\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X_{1}\) influences
-\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In this configuration,
-\(X_{1}\) is a parent node of \(X_{2}\), which is a parent node of
-\(X_{3}\). This structure creates a \emph{directed path} between
-\(X_{1}\) and \(X_{3}\). Consequently, \(X_{1}\) is an \emph{ancestor}
-of \(X_{3}\), and \(X_{2}\) fully \emph{mediates} the relationship
-between the two. Figures \ref{fig-scm_bb4} and \ref{fig-mdag_bb4}
-illustrate a \emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are
-both influenced by \(X_{2}\). Here, \(X_{2}\) is a parent node that
-\emph{confounds} the relationship between \(X_{1}\) and \(X_{3}\).
-Finally, Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a
-\emph{collider}, where variables \(X_{1}\) and \(X_{3}\) are concurrent
-causes of \(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are
-not causally related to each other but both influence \(X_{2}\) (an
-\emph{inmorality}). Notably, in all SCMs, the errors are assumed to be
-independent of each other and from all other variables in the graph, as
-evidenced by the pairwise relations \(e_{X1} \perp\!\!\!\perp e_{X2}\),
-\(e_{X1} \perp\!\!\!\perp e_{X3}\), and
-\(e_{X2} \perp\!\!\!\perp e_{X3}\).
-
-Researchers can use these building blocks to represent the scenario
-outlined in Section~\ref{sec-appendix-B}. Figures \ref{fig-scm_example1}
-and \ref{fig-mdag_example1} depict the plausible causal structure for
-this example, assuming that the variable \(X\) (socio-economic status of
-the school) acts as a confounder in the relationship between the
-teaching method \(T\) and the outcome \(Y\). The figures show several
-descendant, such as \(X \rightarrow T\), \(X \rightarrow Y\), and
-\(T \rightarrow Y\), and also highlight multiple pairs of unconnected
-nodes, evident from the relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
-\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
-Additional, the figures depict one fork, \(X \rightarrow \{T, Y\}\), and
-two colliders: \(\{X, e_{T}\} \rightarrow T\) and
-\(\{X, T, e_{Y}\} \rightarrow Y\).
+assumptions underlying these building blocks. SCM \ref{fig-scm_bb1} and
+DAG \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
+scenario where variables \(X_{1}\) and \(X_{3}\) are independent or not
+causally related. SCM \ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}
+illustrate two connected nodes, representing a scenario where a
+\emph{parent} node \(X_{1}\) exerts a causal influence on a \emph{child}
+node \(X_{3}\). In this setup, \(X_{3}\) is considered a
+\emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
+are described as \emph{adjacent} because an edge directly connects them.
+SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a \emph{chain},
+where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences
+\(X_{3}\). In this configuration, \(X_{1}\) is a parent node of
+\(X_{2}\), which is a parent node of \(X_{3}\). This structure creates a
+\emph{directed path} between \(X_{1}\) and \(X_{3}\). Consequently,
+\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and \(X_{2}\) fully
+\emph{mediates} the relationship between the two. SCM \ref{fig-scm_bb4}
+and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables
+\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). Here,
+\(X_{2}\) is a parent node that \emph{confounds} the relationship
+between \(X_{1}\) and \(X_{3}\). Finally, SCM \ref{fig-scm_bb5} and DAG
+\ref{fig-mdag_bb5} show a \emph{collider}, where variables \(X_{1}\) and
+\(X_{3}\) are concurrent causes of \(X_{2}\). In this configuration,
+\(X_{1}\) and \(X_{3}\) are not causally related to each other but both
+influence \(X_{2}\) (an \emph{inmorality}). Notably, all building blocks
+assume the errors are independent of each other and from all other
+variables in the graph, as evidenced by the pairwise relations
+\(e_{X1} \perp\!\!\!\perp e_{X2}\), \(e_{X1} \perp\!\!\!\perp e_{X3}\),
+and \(e_{X2} \perp\!\!\!\perp e_{X3}\).
+
+Researchers can then use these building blocks to represent the scenario
+outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
+DAG \ref{fig-mdag_example1} depict the plausible causal structure for
+this example. In this context, the variable \(X\) (socio-economic status
+of the school) acts as a confounder in the relationship between the
+teaching method \(T\) and the outcome \(Y\). The figures display
+multiple descendant relationships such as \(X \rightarrow T\),
+\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
+unconnected node pairs, evident from the relationships
+\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
+\(e_{X} \perp\!\!\!\perp e_{Y}\). Additional, the figures show one fork,
+\(X \rightarrow \{T, Y\}\), and two colliders:
+\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).
 
@@
 
-\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}
+\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}
 
@@
 
-\section{The probabilistic implications of SCMs and
+\subsection{The probabilistic and causal implications of SCMs and
 DAGs}\label{sec-appendix-D}
 
-In addition to their graphical capabilities, SCMs and DAGs can encode
-probabilistic information about causal structures. They achieve this
-encoding by relying on three fundamental assumptions: the local Markov,
-the minimality, and the causal edges assumption. The \emph{local Markov
-assumption} encodes probabilistic independencies between variables by
-declaring that nodes in a graph are independent of all its
-non-descendants, given its parents \citep{Neal_2020}. Meanwhile, the
-\emph{minimality assumption} encodes probabilistic dependencies among
-variables by stating that every pair of adjacent nodes exhibits a
-dependency \citep{Neal_2020}. Finally, the \emph{causal edges
-assumption} encodes causal relationships between variables by declaring
-that each parent node acts as a direct cause of its children
-\citep{Neal_2020}. Figure~\ref{fig-ACflow} illustrates how these
-assumptions influence the statistical and causal interpretations of
-graphs.
+Beyond their graphical capabilities, SCMs and DAGs can encode the
+probabilistic information embedded within a causal structure. They
+achieve this encoding by relying on three fundamental assumptions: the
+local Markov, the minimality, the causal edges and the modularity
+assumptions. The \emph{local Markov assumption} encodes probabilistic
+independencies between variables by declaring that nodes in a graph are
+independent of all its non-descendants, given its parents
+\citep[pp.~20]{Neal_2020}. Meanwhile, the \emph{minimality assumption}
+encodes probabilistic dependencies among variables by stating that every
+pair of adjacent nodes exhibits a dependency \citep[pp.~21]{Neal_2020}.
+Finally, the \emph{causal edges assumption} encodes causal relationships
+between variables by declaring that each parent node acts as a direct
+cause of its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
+illustrates how these assumptions influence the statistical and causal
+interpretations of graphs.
 
@@
 \citep[pp.~29]{Pearl_et_al_2016}. A notable implication of the
-assumptions underlying this encoding is that any conceptual model
-described by an SCM and DAG can express the joint distribution of
+assumptions underlying the probabilistic encoding is that any conceptual
+model described by an SCM and DAG can express the joint distribution of
 variables as a product of conditional probability distributions (CPDs)
 in the form \(P(child \: | \: parents)\). This property is formally
-known as the \emph{Bayesian Network factorization}
-(Equation~\ref{eq-net_factor})
+known as the \emph{Bayesian Network factorization} (BNF)
 \citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
-In this factorization, \(pa(X_{i})\) denotes the set of variables that
-are the parents of \(X_{i}\).
+In this factorization, depicted in Equation~\ref{eq-net_factor},
+\(pa(X_{i})\) denotes the set of variables that are the parents of
+\(X_{i}\).
 
@@
 
-Ultimately, this encoding enables researchers with conceptual
-(theoretical) knowledge in the form of an SCM and DAG to predict
-patterns of (in)dependencies in the data. As highlighted by
+This encoding enables researchers with conceptual (theoretical)
+knowledge in the form of an SCM and DAG to predict patterns of
+(in)dependencies in the data. As highlighted by
 \citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
@@
 (in)dependencies in this data can provide significant insights into the
-validity of the proposed conceptual model.
-
-The concept of predicting patterns of (in)dependencies in the data is
-more clearly illustrated through the five fundamental building blocks
-outlined in Section~\ref{sec-appendix-C}.
+validity of the proposed conceptual model. Moreover, it allows to
+express
+
+The five fundamental building blocks described in
+Section~\ref{sec-appendix-C} illustrate how to predict (in)dependencies
+in data. For instance, applying the BNF to the causal structure shown in
+the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} supports the
+theoretical assumption that the observed variables \(X_{1}\) and
+\(X_{3}\) are unconditionally independent
+(\(X_{1} \perp\!\!\!\perp X_{3}\)). This conclusion follows from
+expressing their joint probability distribution as
+\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\), which defines probabilistic
+independence \citep[pp.~24]{Neal_2020}. Conversely, when \(X_{3}\) is
+unconditionally dependent on \(X_{1}\), as depicted in the SCM
+\ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}, the BNF express their
+joint probability distribution as
+\(P(X_{1}, X_{3}) = P(X_{3} \: | \: X_{1}) P(X_{1})\). Notably, these
+descriptions demonstrate the clear correspondence between the structural
+equations illustrated in the figures of Section~\ref{sec-appendix-C} and
+the CPDs.
+
+Beyond the insights gained from two-node structures, researchers can
+uncover more nuanced patterns of (in)dependencies from chains, forks,
+and colliders. By leveraging these patterns, researchers can identify
+the ACE from associational estimates through statistical conditioning,
+thereby disentangling causation from mere association. For instance,
+researchers applying the BNF to the chain structure depicted in the SCM
+\ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} can represent the joint
+distribution for the observed variables as
+\(P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}) P(X_{3} \: | \: X_{2})\).
+From this expression, researchers can infer additional (in)dependencies
+assumed in any data generated by this structure. Specifically, the
+structure implies that \(X_{1}\) and \(X_{3}\) are unconditionally
+dependent, but conditionally independent when controlling for \(X_{2}\)
+(\(X_{1} \perp\!\!\!\perp X_{3} \: | \: X_{2}\))
+\citetext{\citealp[pp.~37]{Pearl_et_al_2016}; \citealp[pp.~25]{Neal_2020}}.
+In the fork structure shown in the SCM \ref{fig-scm_bb4} and DAG
+\ref{fig-mdag_bb4}, researchers can express the joint distribution of
+the observed variables as
+\(P(X_{1},X_{2},X_{3}) = P(X_{1} \: | \: X_{2}) P(X_{2}) P(X_{3} \: | \: X_{2})\).
+This expression allows researchers to further infer that \(X_{1}\) and
+\(X_{3}\) are unconditionally dependent, but conditionally independent
+when controlling for \(X_{2}\)
+(\(X_{1} \perp\!\!\!\perp X_{3}  \: | \: X_{2}\))
+\citep[pp.~40]{Pearl_et_al_2016}. Finally, researchers analyzing the
+collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
+\ref{fig-mdag_bb5} can express the joint distribution of the observed
+variables as
+\(P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}, X_{3}) P(X_{3})\).
+This representation allows researchers to infer that \(X_{1}\) and
+\(X_{3}\) are unconditionally independent
+(\(X_{1} \perp\!\!\!\perp X_{3}\)), but conditionally dependent when
+controlling for \(X_{2}\)
+\citetext{\citealp[pp.~41]{Pearl_et_al_2016}; \citealp[pp.~26]{Neal_2020}}.
+
+Ultimately, after encoding probabilistic information, researchers can
+use SCMs and DAGs to identify causal effects. This process relies on one
+additional assumption: the \emph{modularity assumption}. This assumption
+asserts that intervening on a node only alters its causal mechanism,
+implying that interventions are local \citep[pp.~34]{Neal_2020}. From
+this assumption, concepts such as manipulated DAGs and the Truncated
+factorization emerge. \emph{Manipulated graphs} are DAGs with removed
+edges, signifying intervention. Similarly, the \emph{Truncated
+factorization} represents these manipulated DAGs within SCMs
+(Equation~\ref{eq-trunc_factor}). In this expression, \(S\) denotes the
+subset of manipulated variables \(X_{p}\).
+
+\begin{equation}\phantomsection\label{eq-trunc_factor}{
+\begin{aligned}
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = \prod_{p \not\in S} P(X_{i} \: | \: pa(X_{i}) ) \\
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = 1 \quad \quad (\text{otherwise})
+\end{aligned}
+}\end{equation}
+
+Researchers can apply these new insights to revisit the scenario in
+Section~\ref{sec-appendix-B}. From the SCM \ref{fig-scm_example1} and
+DAG \ref{fig-mdag_example1a}, they can infer that the effect of \(T\) on
+\(Y\) is confounded by \(X\). This is evident because the effect
+decomposes into a direct causal effect (\(T \rightarrow Y\), black path)
+and a non-causal effect arising from the confounder \(X\)
+(\(T \leftarrow X \rightarrow Y\), gray path). Assuming the causal
+structure is correct, researchers can identify the CACE of \(T\) on
+\(Y\) by conditioning on \(X\) and then marginalize this effect over
+\(X\) to obtain the ACE. Conditioning on \(X\) means restricting the
+focus to schools with a specific socio-economic status \(X\)
+\citep[pp.~32]{Neal_2020}. Notably, researchers often depict
+conditioning as a square around a variable. Furthermore, the
+conditioning process leads to the manipulated DAG
+\ref{fig-mdag_example1b}, equivalent to the conditioned DAG
+\ref{fig-mdag_example1a}.
+
+Using Equation~\ref{eq-net_factor} and Equation~\ref{eq-trunc_factor},
+researchers can show that the joint probability of the observed
+variables in these structures can be expressed as
+\(P(Y, T, X) = P(Y \: | \: T, X) P(T \: | \: X) P(X)\). Upon
+intervention, this expression becomes
+\(P(Y, X \: | \: do(T=t)) = P(Y \: | \: T,X) P(X)\), and by
+marginalizing over \(X\), they obtain
+\(P(Y \: | \: do(T=t)) = \sum_{X} P(Y \: | \: T,X) P(X)\)
+\citep[pp.~36]{Neal_2020}. From these distributions, the ACE can be
+expressed as
+\(\tau = E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]\)
+(see Section~\ref{sec-appendix-B}).
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := t \\
+  Y & := f_{Y}(t,X,e_{Y}) \\
+  e_{T} & \perp\!\!\!\perp e_{X} \\
+  e_{T} & \perp\!\!\!\perp e_{Y} \\
+  e_{X} & \perp\!\!\!\perp e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_example1}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1a.png}
+
+}
+
+\subcaption{\label{fig-mdag_example1a}Conditioned DAG}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1b.png}
+
+}
+
+\subcaption{\label{fig-mdag_example1b}Manipulated DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-example}Plausible causal structure the scenario
+outlined in Section~\ref{sec-appendix-B}.}
+
+\end{figure}%
+
+\subsection{The estimation process}\label{sec-appendix-E}
+
+Researchers can use Bayesian inference methods to estimate the ACE. The
+approach begins by defining two probability distributions: the
+likelihood of the data,
+\(P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta)\), and the prior
+distribution, \(P(\theta)\) \citep{Everitt_et_al_2010}, where \(X_{P}\)
+represents a random variable, and \(\theta\) represents a
+one-dimensional parameter space for simplicity. After observing
+empirical data, researchers can update the priors to posterior
+distributions using Bayes' rule \citep{Jeffreys_1998}:
+
+\begin{equation}\phantomsection\label{eq-bayes_rule}{
+P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
+}\end{equation}
+
+Given that the denominator on the right-hand side of
+Equation~\ref{eq-bayes_rule} serves as a normalizing constant
+independent of the parameter \(\theta\), researchers can simplify the
+posterior updating process into the three steps depicted by
+Equation~\ref{eq-prop_rule}. First, they integrate new empirical data
+through the likelihood. Second, they update the parameters' priors to a
+posterior distribution according to Equation~\ref{eq-prop_rule}.
+Ultimately, they normalize these results to obtain a valid probability
+distribution.
+
+\begin{equation}\phantomsection\label{eq-prop_rule}{
+P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\: | \: \theta) \cdot P(\theta)
+}\end{equation}
+
+Temporarily setting aside the definition of prior distributions
+\(P(\theta)\), note that the posterior updating process depends heavily
+on the assumptions underlying the likelihood of the data. However, as
+the number of random variables, \(P\), increases, this joint
+distribution becomes quickly intractable \citep{Neal_2020}. This
+intractability is evident from Equation~\ref{eq-like_chain}, where the
+likelihood distribution is expressed by multiple chained CPDs.
+
+\begin{equation}\phantomsection\label{eq-like_chain}{
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: X_{i-1}, \dots, X_{1}, \theta )
+}\end{equation}
+
+Nevertheless, researchers can manage the complexity of the likelihood by
+assuming that variables exhibit specific local (in)dependencies. These
+assumptions enhance model tractability and facilitate the estimation
+process. As described in Section~\ref{sec-appendix-D}, researchers can
+use SCMs and DAGs to formalize these local (in)dependencies, which leads
+to the BNF of the likelihood (Equation~\ref{eq-like_BNF}). From this
+point, any probabilistic programming language can model this simpler
+structure and generate the parameter's posterior distribution from its
+priors using Equation~\ref{eq-bayes_rule}.
+
+\begin{equation}\phantomsection\label{eq-like_BNF}{
+P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: pa(X_{i}), \theta )
+}\end{equation}
 
@@
 
-References
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 

### 2025-01-17 paper2.qmd
section:appendix; process:write

@@
 
+<!-- commands for d-separation -->
+\newcommand{\dsep}{\:\bot\:}
+\newcommand{\ndsep}{\:\not\bot\:}
+\newcommand{\cond}{\:|\:}
+
 ## Empirical research and randomized experiments {#sec-appendix-A .appendix appendix-style=plain}
@@
 Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability.
-In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random quantity $E[Y_{i} \: | \: T_{i}=1] - E[Y_{i} \: | \: T_{i}=2]$, where $E[\cdot]$ denotes the expected value.
+In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random associational quantity in @eq-group_diff, where $E[\cdot]$ denotes the expected value.
+
+$$
+E[Y_{i} \cond T_{i}=1] - E[Y_{i} \cond T_{i}=2]
+$$ {#eq-group_diff}
 
@@
 If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
-The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \: | \: T_{i}=1]$ and $E[Y_{i} \: | \: T_{i}=2]$, respectively.
+The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \cond T_{i}=1]$ and $E[Y_{i} \cond T_{i}=2]$, respectively.
 The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.
@@
 Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014].
-This recognition stems from their ability to enable researchers to interpret associational estimates as causal.
+This recognition stems from their ability to enable researchers interpret associational estimates as causal.
 They achieve this by ensuring data, and by extension an estimator, satisfies several key properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
@@
 The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals.
-For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes: $\tau_{i} = Y_{i} \: | \: do(T_{i}=1) - Y_{i} \: | \: do(T_{i}=2)$.
-Here, $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \: | \: do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \: | \: do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
+For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes, as in @eq-ICE.
+
+$$
+\tau_{i} = Y_{i} \cond do(T_{i}=1) - Y_{i} \cond do(T_{i}=2)
+$$ {#eq-ICE}
+
+where $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \cond do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \cond do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
+Here, an *intervention* involves assigning a constant value to the treatment variable for each student's potential outcomes.
+In contrast, conditioning involves restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
+32] (see @eq-group_diff).
+<!-- Conditioning on T=t just means that we are restricting our focus to the subset of the population to those who received treatment t.
+In contrast, an intervention would be to take the whole population and give everyone treatment t [@Neal_2020, pp.
+32].
+-->
+
 Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable.
-To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i} \: | \: do(T_{i}=1)]- E[Y_{i} \: | \: do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts.
+To address this challenge, the structural approach extends the ICE to the *average causal effect* (ACE, @eq-ACE), representing the average difference between observed potential outcomes and their counterfactual counterparts.
+
+$$
+\tau = E[\tau_{i}] = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)]
+$$ {#eq-ACE}
 
@@
 <!-- 3. But how does it work? -->
-Even when data originates from an observational study, researchers can identify the ACE from associational estimates using the structural approach.
-They achieve this by performing statistical conditioning on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones, ensuring the ACE estimator satisfies several key properties, including confounding elimination.
-If such a set exists, then $T$ and $Y$ are *d-separated* by $X$ [@Pearl_2009], meaning researchers can estimate the ACE from associational random quantities [@Morgan_et_al_2014].
-Naturally, the validity of claims about the effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set.
+Even though counterfactuals are unobserved and unobservable, researchers can still identify the ACE from associational estimates by leveraging the structural approach.
+The approach identifies the ACE by statistically conditioning data on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones.
+If such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
+
+Conditioning on a sufficient adjustment set enables researchers to estimate the ACE, even when the data comes from an observational study.
+This process is feasible because such conditioning ensures that the ACE estimator satisfies several critical properties, including confounding elimination [@Morgan_et_al_2014].
+Naturally, the validity of claims about the causal effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set.
 However, as @Kohler_et_al_2019 [pp.
-150] observed, drawing conclusions about the real world from observed data inevitably requires assumptions.
-This holds true for both observational and experimental data.
+150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions.
+This requirement holds true for both observational and experimental data.
 
@@
 
-For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and must rely on observational data, they can still estimate the ACE, provided that a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ and the outcome $Y$ without opening any new ones.
-Under these conditions, the ACE can be estimated from the *conditional average causal effect* (CACE), defined by $E[Y_{i} \: | \: T_{i}=t, X]$.
-Using the CACE, the ACE can be expressed as $\tau =$ $E[Y_{i} \: | \: do(T_{i}=1)] - E[Y_{i} \: | \: do(T_{i}=2)] =$ $E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
-Notably, the approach extends the ACE for a continuous variable $T$ as $\tau =$ $E[Y_{i} \: | \: do(T_{i}=t)] =$ $d E_{X}\left[ E[Y_{i} \: | \: T_{i}=t, X]\right]/ dt$, ensuring broad applicability across different causal scenarios [@Neal_2020, pp.
-45].
+For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as a variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
+Under these circumstances, researchers firt identify the *conditional average causal effect* (CACE, @eq-CACE)
+
+$$
+E[Y_{i} \cond T_{i}=t, X]
+$$ {#eq-CACE}
+
+From the CACE, they identify the ACE from associational quantities as in @eq-mACE. Here, $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
+
+$$
+\begin{aligned}
+  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
+\end{aligned}
+$$ {#eq-mACE}
+
+Notably, the approach extends the ACE identification for a continuous variable $T$ as in @eq-mACE_cont, ensuring broad applicability across different causal scenarios [@Neal_2020, pp. 45]
+
+$$
+\begin{aligned}
+  \tau &= E[Y_{i} \cond do(T_{i}=t)] \\
+  & = d E_{X}\left[ E[Y_{i} \cond T_{i}=t, X]\right]/ dt
+  \end{aligned}
+$$ {#eq-mACE_cont}
 
 ## SCMs and DAGs {#sec-appendix-C .appendix appendix-style=plain}
-
-<!-- commands for d-separation -->
-\newcommand{\dsep}{\perp\!\!\!\perp}
-<!-- \newcommand{\ndsep}{\not\!\perp\!\!\!\perp} -->
 
@@
 The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that help researchers determine which statistical models can identify an estimand, assuming the depicted causal structure is correct, thus enabling valid causal inference [@McElreath_2020].
-They achieve this by representing various causal structures, regardless of complexity, using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference by guiding researchers in determining which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020].
+
+SCMs and DAGs support causal inference through two key advantages.
+First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
 This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
-Furthermore, SCMs and DAGs depict causal relationships in a non-parametric and fully interactive way.
-This flexibility enables feasible ACE identification strategies without requiring detailed specifications of functional relationships relating the variables, their parameters, or their data types [@Pearl_et_al_2016].
+Second, they depict causal relationships in a non-parametric and fully interactive way.
+This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp.
+35].
 @fig-IEflow shows the role of theoretical models in the inference process.
@@
 <!-- 4. Theoretical implications of the basic building block of the SCMS and DAGs -->
-A careful examination of the figures highlights the theoretical assumptions underlying these building blocks.
+A careful examination of the figures highlights the theoretical assumptions underlying the observed variables in these building blocks.
 SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are independent or not causally related.
@@
 In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$.
-Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because an edge directly connects them.
+Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because there is a *direct path* connecting.
 SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
@@
 
-## The probabilistic and causal implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}
+## The probabilistic implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}
 
@@
 Beyond their graphical capabilities, SCMs and DAGs can encode the probabilistic information embedded within a causal structure.
-They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges and the modularity assumptions.
+They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges assumptions.
 The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020, pp.
@@
 <!-- 2. What are the general implications of this? -->
-The encoding of probabilistic information in SCMs and DAGs facilitates the efficient representation of the joint distribution of variables [@Pearl_et_al_2016, pp.
+A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can represent the joint distribution of variables more efficiently [@Pearl_et_al_2016, pp.
 29].
-A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can express the joint distribution of variables as a product of conditional probability distributions (CPDs) in the form $P(child \: | \: parents)$.
-This property is formally known as the *Bayesian Network factorization* (BNF) [@Pearl_et_al_2016, pp.
+This expression takes the form of a product of conditional probability distributions (CPDs) $P(child \cond parents)$.
+This property is formally known as the *Bayesian Network factorization* (BNF, @eq-net_factor) [@Pearl_et_al_2016, pp.
 29; @Neal_2020, pp.
 21].
-In this factorization, depicted in @eq-net_factor, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.
-
-$$
-P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \: | \: pa(X_{i}) )
+In this expression, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.
+
+$$
+P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \cond pa(X_{i}) )
 $$ {#eq-net_factor}
@@
 Moreover, once researchers observe empirical data, the patterns of (in)dependencies in this data can provide significant insights into the validity of the proposed conceptual model.
-Moreover, it allows to express
 
@@
 <!-- 4. can you illustrate this concept? -->
-The five fundamental building blocks described in @sec-appendix-C illustrate how to predict (in)dependencies in data.
+The five fundamental building blocks described in @sec-appendix-C clearly illustrate which (in)dependencies can SMCs and DAGs predict.
 For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$).
@@
 24].
-Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$, as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \: | \: X_{1}) P(X_{1})$.
+Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$.
 Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in the figures of @sec-appendix-C and the CPDs.
 
-<!-- ($X_{1} \ndsep X_{3}$) -->
-
 Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of (in)dependencies from chains, forks, and colliders.
-By leveraging these patterns, researchers can identify the ACE from associational estimates through statistical conditioning, thereby disentangling causation from mere association.
-For instance, researchers applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] can represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}) P(X_{3} \: | \: X_{2})$.
+For instance, researchers applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] can represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}) P(X_{3} \cond X_{2})$.
 From this expression, researchers can infer additional (in)dependencies assumed in any data generated by this structure.
-Specifically, the structure implies that $X_{1}$ and $X_{3}$ are unconditionally dependent, but conditionally independent when controlling for $X_{2}$ ($X_{1} \dsep X_{3} \: | \: X_{2}$) [@Pearl_et_al_2016, pp.
-37; @Neal_2020, pp.
-25].
-In the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1} \: | \: X_{2}) P(X_{2}) P(X_{3} \: | \: X_{2})$.
-This expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent, but conditionally independent when controlling for $X_{2}$ ($X_{1} \dsep X_{3}  \: | \: X_{2}$) [@Pearl_et_al_2016, pp.
-40].
-Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}, X_{3}) P(X_{3})$.
-This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$), but conditionally dependent when controlling for $X_{2}$ [@Pearl_et_al_2016, pp.
-41; @Neal_2020, pp.
-26].
-
-<!-- ($X_{1} \ndsep X_{3}$) -->
-<!-- ($X_{1} \ndsep X_{3}$) -->
-<!-- ($X_{1} \ndsep X_{3} \: | \: X_{2}$)  -->
-
-<!-- 5. But, how do we identify causal effects?  -->
-Ultimately, after encoding probabilistic information, researchers can use SCMs and DAGs to identify causal effects.
-This process relies on one additional assumption: the *modularity assumption*.
-This assumption asserts that intervening on a node only alters its causal mechanism, implying that interventions are local [@Neal_2020, pp.
+Specifically, the structure implies that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
+In the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \cond X_{2}) P(X_{2}) P(X_{3} \cond X_{2})$.
+This expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
+Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}, X_{3}) P(X_{3})$.
+This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent $(X_{1} \dsep X_{3})$, but conditionally dependent when controlling for $X_{2}$ $(X_{1} \ndsep X_{3} \cond X_{2})$.
+The authors @Pearl_et_al_2016 [pp.
+37, 40, 41] and @Neal_2020 [pp.
+25–26] provide the mathematical proofs for these conclusions.
+
+<!-- 5. Continuing with the motivating example -->
+<!-- Researchers can apply these new insights to revisit the scenario in @sec-appendix-B.
+From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can express the joint probability for the observed variables as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$.
+From this expression, researchers can say that.
+-->
+
+## From probability to causality {#sec-appendix-E .appendix appendix-style=plain}
+
+<!-- 1. But, how do we identify causal effects?  -->
+<!-- Researchers can use SCMs and DAGs to represent interventions of the type $P(X_{a} \cond do(X_{b}))$.
+These representations rely on the *modularity assumption*, which asserts that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
 34].
-From this assumption, concepts such as manipulated DAGs and the Truncated factorization emerge.
-*Manipulated graphs* are DAGs with removed edges, signifying intervention.
-Similarly, the *Truncated factorization* represents these manipulated DAGs within SCMs (@eq-trunc_factor).
-In this expression, $S$ denotes the subset of manipulated variables $X_{p}$.
+-->
+
+<!-- By leveraging these patterns, researchers can identify the ACE from associational estimates through statistical conditioning, thereby disentangling causation from mere association.  -->
+
+<!-- Intervention relies on the concepts of Truncated factorization and manipulated graphs.
+The *Truncated factorization* (TF) simulates physical interventions by removing specific functions from the conceptual model and substituting them with constants, while keeping the remaining structure unchanged [Pearl_2010].
+The probabilistic implications for this factorization are expressed in @eq-trunc_factor.
+Here, $S$ denotes the subset of variables $X_{p}$ directly impacted by the intervention.
+Complementary to the TF, *manipulated graphs* are DAGs with certain edges removed to represent the effects of interventions [@Neal_2020, pp.
+34].
+-->
+
+$$
+P(X_{1}, X_{2}, \dots, X_{P} \cond do(S)) =
+\begin{cases}
+  \prod P(X_{i} \cond pa(X_{i}) ) & \text{if} \: p \not\in S \\
+  1 \quad & \text{otherwise}
+\end{cases}
+$$ {#eq-trunc_factor}
+
+<!-- The Truncated Factorization enables researchers to define the *backdoor adjustment*.
+This adjustment specifies that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+-->
+
+$$
+P(X_{a} \cond do(X_{b}=x)) = \sum_{Xp} P(X_{a} \cond X_{b}=x, X_{p}) P(X_{p})
+$$ {#eq-backdoor}
+
+<!-- Ultimately, backdoor adjustment enables researchers to express the ACE as: -->
 
@@
 \begin{aligned}
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = \prod_{p \not\in S} P(X_{i} \: | \: pa(X_{i}) ) \\
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = 1 \quad \quad (\text{otherwise})
+\tau &= E[\tau_{i}] \\
+  & = E[X_{a} \cond do(X_{b}=1)]- E[X_{a} \cond do(X_{b}=2)] \\
+  & = \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=1, X_{p}) P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=2, X_{p}) P(X_{p})
 \end{aligned}
-$$ {#eq-trunc_factor}
+$$ {#eq-backdoor_adjustment}
 
 <!-- 6. Continuing with the motivating example -->
-Researchers can apply these new insights to revisit the scenario in @sec-appendix-B.
+<!-- Researchers can apply these new insights to revisit the scenario in @sec-appendix-B.
 From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can infer that the effect of $T$ on $Y$ is confounded by $X$.
-This is evident because the effect decomposes into a direct causal effect ($T \rightarrow Y$, black path) and a non-causal effect arising from the confounder $X$ ($T \leftarrow X \rightarrow Y$, gray path).
-Assuming the causal structure is correct, researchers can identify the CACE of $T$ on $Y$ by conditioning on $X$ and then marginalize this effect over $X$ to obtain the ACE.
+-->
+
+<!-- This is evident because the effect decomposes into a *direct causal effect* ($T \rightarrow Y$, black path) and a non-causal effect arising from the confounder $X$ ($T \leftarrow X \rightarrow Y$, gray path).
+-->
+
+<!-- Using @eq-net_factor and @eq-trunc_factor, researchers can show that the joint probability of the observed variables in these structures can be expressed as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$.
+Assuming intervention, this expression becomes $P(Y, X \cond do(T=t)) = P(Y \cond T,X) P(X)$, and by marginalizing over $X$, they obtain $P(Y \cond do(T=t)) = \sum_{X} P(Y \cond T,X) P(X)$ [@Neal_2020, pp.
+36].
+-->
+
+<!-- Assuming the causal structure is correct, researchers can identify the CACE of $T$ on $Y$ by conditioning on $X$ and then marginalize this effect over $X$ to obtain the ACE.
 Conditioning on $X$ means restricting the focus to schools with a specific socio-economic status $X$ [@Neal_2020, pp.
@@
 Furthermore, the conditioning process leads to the manipulated DAG [-@fig-mdag_example1b], equivalent to the conditioned DAG [-@fig-mdag_example1a].
-
-Using @eq-net_factor and @eq-trunc_factor, researchers can show that the joint probability of the observed variables in these structures can be expressed as $P(Y, T, X) = P(Y \: | \: T, X) P(T \: | \: X) P(X)$.
-Upon intervention, this expression becomes $P(Y, X \: | \: do(T=t)) = P(Y \: | \: T,X) P(X)$, and by marginalizing over $X$, they obtain $P(Y \: | \: do(T=t)) = \sum_{X} P(Y \: | \: T,X) P(X)$ [@Neal_2020, pp.
-36].
-From these distributions, the ACE can be expressed as $\tau = E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]$ (see @sec-appendix-B).
+-->
+
+<!--  From these distributions, the ACE can be expressed as $\tau = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]$ (see @sec-appendix-B). -->
 
@@
 
-## The estimation process {#sec-appendix-E .appendix appendix-style=plain}
+## The estimation process {#sec-appendix-F .appendix appendix-style=plain}
 
@@
 Researchers can use Bayesian inference methods to estimate the ACE.
-The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
+The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \cond \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
 After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:
@@
 $$
-P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
+P(\theta \cond X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \cond \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
 $$ {#eq-bayes_rule}
@@
 $$
-P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\: | \: \theta) \cdot P(\theta)
+P(\theta \cond X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\cond \theta) \cdot P(\theta)
 $$ {#eq-prop_rule}
@@
 $$
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: X_{i-1}, \dots, X_{1}, \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \cond \theta) = P(X_{1} \cond \theta) \prod^{P}_{p=2} P(X_{i} \cond X_{i-1}, \dots, X_{1}, \theta )
 $$ {#eq-like_chain}
@@
 $$
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: pa(X_{i}), \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \cond \theta) = P(X_{1} \cond \theta) \prod^{P}_{p=2} P(X_{i} \cond pa(X_{i}), \theta )
 $$ {#eq-like_BNF}

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

@@
 
+\newcommand{\dsep}{\:\bot\:}
+\newcommand{\ndsep}{\:\not\bot\:}
+\newcommand{\cond}{\:|\:}
+
 \subsection{Empirical research and randomized
@@
 and this estimand can be mathematically represented by the random
-quantity \(E[Y_{i} \: | \: T_{i}=1] - E[Y_{i} \: | \: T_{i}=2]\), where
+associational quantity in Equation~\ref{eq-group_diff}, where
 \(E[\cdot]\) denotes the expected value.
+
+\begin{equation}\phantomsection\label{eq-group_diff}{
+E[Y_{i} \:|\:T_{i}=1] - E[Y_{i} \:|\:T_{i}=2]
+}\end{equation}
 
@@
 \(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
-\(E[Y_{i} \: | \: T_{i}=1]\) and \(E[Y_{i} \: | \: T_{i}=2]\),
-respectively. The noise, represented by \(\hat{s}_{p}\), is defined as
-the unpooled sample variability observed between the two groups.
+\(E[Y_{i} \:|\:T_{i}=1]\) and \(E[Y_{i} \:|\:T_{i}=2]\), respectively.
+The noise, represented by \(\hat{s}_{p}\), is defined as the unpooled
+sample variability observed between the two groups.
 
@@
 evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
-recognition stems from their ability to enable researchers to interpret
+recognition stems from their ability to enable researchers interpret
 associational estimates as causal. They achieve this by ensuring data,
@@
 defining the \emph{individual causal effect} (ICE) as the difference
-between each student's potential outcomes:
-\(\tau_{i} = Y_{i} \: | \: do(T_{i}=1) - Y_{i} \: | \: do(T_{i}=2)\).
-Here, \(do(T_{i}=t)\) represents the intervention operator,
-\(Y_{i} \: | \: do(T_{i}=1)\) represents the potential outcome under
-intervention \(T_{i}=1\), and \(Y_{i} \: | \: do(T_{i}=1)\) represents
-the potential outcome under intervention \(T_{i}=2\). Note that if a
-student is assigned to intervention \(T_{i}=1\), the potential outcome
-under \(T_{i}=2\) becomes a counterfactual, as it is no longer observed
-nor observable. To address the challenge of unobserved counterfactuals,
-the approach extends the ICE to the \emph{average causal effect} (ACE):
-\(\tau = E[\tau_{i}] = E[Y_{i} \: | \: do(T_{i}=1)]- E[Y_{i} \: | \: do(T_{i}=2)]\),
-representing the average difference between observed potential outcomes
-and their counterfactual counterparts.
-
-Even when data originates from an observational study, researchers can
-identify the ACE from associational estimates using the structural
-approach. They achieve this by performing statistical conditioning on a
-\emph{sufficient adjustment set} of variables \(X\)
-\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
-\emph{sufficient} set (potentially empty) must block all non-causal
-paths between \(T\) to \(Y\) without opening new ones, ensuring the ACE
-estimator satisfies several key properties, including confounding
-elimination. If such a set exists, then \(T\) and \(Y\) are
-\emph{d-separated} by \(X\) \citep{Pearl_2009}, meaning researchers can
-estimate the ACE from associational random quantities
-\citep{Morgan_et_al_2014}. Naturally, the validity of claims about the
-effects of \(T\) on \(Y\) now hinges on the assumption that \(X\) serves
-as a sufficient adjustment set. However, as
-\citet[pp.~150]{Kohler_et_al_2019} observed, drawing conclusions about
-the real world from observed data inevitably requires assumptions. This
-holds true for both observational and experimental data.
-
-For instance, if researchers are unable to conduct the randomized
-experiments described in Section~\ref{sec-appendix-A} and must rely on
-observational data, they can still estimate the ACE, provided that a
-variable \(X\), such as the socio-economic status of the school, blocks
-all non-causal paths between the teaching method \(T\) and the outcome
-\(Y\) without opening any new ones. Under these conditions, the ACE can
-be estimated from the \emph{conditional average causal effect} (CACE),
-defined by \(E[Y_{i} \: | \: T_{i}=t, X]\). Using the CACE, the ACE can
-be expressed as \(\tau =\)
-\(E[Y_{i} \: | \: do(T_{i}=1)] - E[Y_{i} \: | \: do(T_{i}=2)] =\)
-\(E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]\),
-where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
-\citep{Morgan_et_al_2014}. Notably, the approach extends the ACE for a
-continuous variable \(T\) as \(\tau =\)
-\(E[Y_{i} \: | \: do(T_{i}=t)] =\)
-\(d E_{X}\left[ E[Y_{i} \: | \: T_{i}=t, X]\right]/ dt\), ensuring broad
+between each student's potential outcomes, as in Equation~\ref{eq-ICE}.
+
+\begin{equation}\phantomsection\label{eq-ICE}{
+\tau_{i} = Y_{i} \:|\:do(T_{i}=1) - Y_{i} \:|\:do(T_{i}=2)
+}\end{equation}
+
+where \(do(T_{i}=t)\) represents the intervention operator,
+\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
+intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
+potential outcome under intervention \(T_{i}=2\). Here, an
+\emph{intervention} involves assigning a constant value to the treatment
+variable for each student's potential outcomes. In contrast,
+conditioning involves restricting the focus to the subset of the
+population defined by the conditioning variable
+\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).\\
+
+Note that if a student is assigned to intervention \(T_{i}=1\), the
+potential outcome under \(T_{i}=2\) becomes a counterfactual, as it is
+no longer observed nor observable. To address this challenge, the
+structural approach extends the ICE to the \emph{average causal effect}
+(ACE, Equation~\ref{eq-ACE}), representing the average difference
+between observed potential outcomes and their counterfactual
+counterparts.
+
+\begin{equation}\phantomsection\label{eq-ACE}{
+\tau = E[\tau_{i}] = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
+}\end{equation}
+
+Even though counterfactuals are unobserved and unobservable, researchers
+can still identify the ACE from associational estimates by leveraging
+the structural approach. The approach identifies the ACE by
+statistically conditioning data on a \emph{sufficient adjustment set} of
+variables \(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
+This \emph{sufficient} set (potentially empty) must block all non-causal
+paths between \(T\) to \(Y\) without opening new ones. If such a set
+exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
+(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
+\emph{backdoor criterion} \citep[pp 37]{Neal_2020}.
+
+Conditioning on a sufficient adjustment set enables researchers to
+estimate the ACE, even when the data comes from an observational study.
+This process is feasible because such conditioning ensures that the ACE
+estimator satisfies several critical properties, including confounding
+elimination \citep{Morgan_et_al_2014}. Naturally, the validity of claims
+about the causal effects of \(T\) on \(Y\) now hinges on the assumption
+that \(X\) serves as a sufficient adjustment set. However, as
+\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
+real world from observed data inevitably requires assumptions. This
+requirement holds true for both observational and experimental data.
+
+For instance, If researchers cannot conduct the randomized experiments
+described in Section~\ref{sec-appendix-A} and must instead rely on
+observational data, they can still identify the ACE as long as a
+variable \(X\), such as the socio-economic status of the school,
+satisfies the backdoor criterion. Under these circumstances, researchers
+firt identify the \emph{conditional average causal effect} (CACE,
+Equation~\ref{eq-CACE})
+
+\begin{equation}\phantomsection\label{eq-CACE}{
+E[Y_{i} \:|\:T_{i}=t, X]
+}\end{equation}
+
+From the CACE, they identify the ACE from associational quantities as in
+Equation~\ref{eq-mACE}. Here, \(E_{X}[\cdot]\) represents the marginal
+expected value over \(X\) \citep{Morgan_et_al_2014}.
+
+\begin{equation}\phantomsection\label{eq-mACE}{
+\begin{aligned}
+  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
+\end{aligned}
+}\end{equation}
+
+Notably, the approach extends the ACE identification for a continuous
+variable \(T\) as in Equation~\ref{eq-mACE_cont}, ensuring broad
 applicability across different causal scenarios
-\citep[pp.~45]{Neal_2020}.
+\citep[pp.~45]{Neal_2020}
+
+\begin{equation}\phantomsection\label{eq-mACE_cont}{
+\begin{aligned}
+  \tau &= E[Y_{i} \:|\:do(T_{i}=t)] \\
+  & = d E_{X}\left[ E[Y_{i} \:|\:T_{i}=t, X]\right]/ dt
+  \end{aligned}
+}\end{equation}
 
 \subsection{SCMs and DAGs}\label{sec-appendix-C}
-
-\newcommand{\dsep}{\perp\!\!\!\perp}
 
@@
 Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
-models} that help researchers determine which statistical models can
-identify an estimand, assuming the depicted causal structure is correct,
-thus enabling valid causal inference \citep{McElreath_2020}. They
-achieve this by representing various causal structures, regardless of
-complexity, using only five fundamental building blocks
+models} that facilitate valid causal inference by guiding researchers in
+determining which statistical models can identify an estimand (ACE,
+CACE, or other), assuming the depicted causal structure is correct
+\citep{McElreath_2020}.
+
+SCMs and DAGs support causal inference through two key advantages.
+First, regardless of complexity, they can represent various causal
+structures using only five fundamental building blocks
 \citep{Neal_2020, McElreath_2020}. This feature allows researchers to
 decompose complex structures into manageable components, facilitating
-their analysis \citep{McElreath_2020}. Furthermore, SCMs and DAGs depict
-causal relationships in a non-parametric and fully interactive way. This
+their analysis \citep{McElreath_2020}. Second, they depict causal
+relationships in a non-parametric and fully interactive way. This
 flexibility enables feasible ACE identification strategies without
-requiring detailed specifications of functional relationships relating
-the variables, their parameters, or their data types
-\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows the role of
-theoretical models in the inference process.
+defining the variables' data types, the functional form between them, or
+their parameters \citep[pp.~35]{Pearl_et_al_2016}.
+Figure~\ref{fig-IEflow} shows the role of theoretical models in the
+inference process.
 
@@
 symbol `\(:=\)' to denote the asymmetrical causal dependence of the
-variables and the symbol `\(\perp\!\!\!\perp\)' to represent
+variables and the symbol `\(\:\bot\:\)' to represent
 \emph{d-separation}, a concept akin to (conditional) independence.
@@
   X_{3} & := f_{X3}(e_{X3}) \\
-  e_{X1} & \perp\!\!\!\perp e_{X3}
+  e_{X1} & \:\bot\:e_{X3}
 \end{aligned}
@@
   X_{3} & := f_{X3}(X_{1},e_{X3}) \\
-  e_{X1} & \perp\!\!\!\perp e_{X3}
+  e_{X1} & \:\bot\:e_{X3}
 \end{aligned}
@@
   X_{3} & := f_{X3}(X_{2},e_{X3}) \\
-  e_{X1} & \perp\!\!\!\perp e_{X2} \\
-  e_{X1} & \perp\!\!\!\perp e_{X3} \\
-  e_{X2} & \perp\!\!\!\perp e_{X3}
+  e_{X1} & \:\bot\:e_{X2} \\
+  e_{X1} & \:\bot\:e_{X3} \\
+  e_{X2} & \:\bot\:e_{X3}
 \end{aligned}
@@
   X_{3} & := f_{X3}(e_{X3}) \\
-  e_{X1} & \perp\!\!\!\perp e_{X2} \\
-  e_{X1} & \perp\!\!\!\perp e_{X3} \\
-  e_{X2} & \perp\!\!\!\perp e_{X3}
+  e_{X1} & \:\bot\:e_{X2} \\
+  e_{X1} & \:\bot\:e_{X3} \\
+  e_{X2} & \:\bot\:e_{X3}
 \end{aligned}
@@
 A careful examination of the figures highlights the theoretical
-assumptions underlying these building blocks. SCM \ref{fig-scm_bb1} and
-DAG \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
-scenario where variables \(X_{1}\) and \(X_{3}\) are independent or not
-causally related. SCM \ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}
-illustrate two connected nodes, representing a scenario where a
-\emph{parent} node \(X_{1}\) exerts a causal influence on a \emph{child}
-node \(X_{3}\). In this setup, \(X_{3}\) is considered a
-\emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
-are described as \emph{adjacent} because an edge directly connects them.
-SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a \emph{chain},
-where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences
-\(X_{3}\). In this configuration, \(X_{1}\) is a parent node of
-\(X_{2}\), which is a parent node of \(X_{3}\). This structure creates a
-\emph{directed path} between \(X_{1}\) and \(X_{3}\). Consequently,
-\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and \(X_{2}\) fully
-\emph{mediates} the relationship between the two. SCM \ref{fig-scm_bb4}
-and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables
-\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). Here,
-\(X_{2}\) is a parent node that \emph{confounds} the relationship
-between \(X_{1}\) and \(X_{3}\). Finally, SCM \ref{fig-scm_bb5} and DAG
-\ref{fig-mdag_bb5} show a \emph{collider}, where variables \(X_{1}\) and
-\(X_{3}\) are concurrent causes of \(X_{2}\). In this configuration,
-\(X_{1}\) and \(X_{3}\) are not causally related to each other but both
-influence \(X_{2}\) (an \emph{inmorality}). Notably, all building blocks
-assume the errors are independent of each other and from all other
-variables in the graph, as evidenced by the pairwise relations
-\(e_{X1} \perp\!\!\!\perp e_{X2}\), \(e_{X1} \perp\!\!\!\perp e_{X3}\),
-and \(e_{X2} \perp\!\!\!\perp e_{X3}\).
+assumptions underlying the observed variables in these building blocks.
+SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
+nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
+are independent or not causally related. SCM \ref{fig-scm_bb2} and DAG
+\ref{fig-mdag_bb2} illustrate two connected nodes, representing a
+scenario where a \emph{parent} node \(X_{1}\) exerts a causal influence
+on a \emph{child} node \(X_{3}\). In this setup, \(X_{3}\) is considered
+a \emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
+are described as \emph{adjacent} because there is a \emph{direct path}
+connecting. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a
+\emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
+influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
+of \(X_{2}\), which is a parent node of \(X_{3}\). This structure
+creates a \emph{directed path} between \(X_{1}\) and \(X_{3}\).
+Consequently, \(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and
+\(X_{2}\) fully \emph{mediates} the relationship between the two. SCM
+\ref{fig-scm_bb4} and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork},
+where variables \(X_{1}\) and \(X_{3}\) are both influenced by
+\(X_{2}\). Here, \(X_{2}\) is a parent node that \emph{confounds} the
+relationship between \(X_{1}\) and \(X_{3}\). Finally, SCM
+\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} show a \emph{collider},
+where variables \(X_{1}\) and \(X_{3}\) are concurrent causes of
+\(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are not
+causally related to each other but both influence \(X_{2}\) (an
+\emph{inmorality}). Notably, all building blocks assume the errors are
+independent of each other and from all other variables in the graph, as
+evidenced by the pairwise relations \(e_{X1} \:\bot\:e_{X2}\),
+\(e_{X1} \:\bot\:e_{X3}\), and \(e_{X2} \:\bot\:e_{X3}\).
 
@@
 unconnected node pairs, evident from the relationships
-\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
-\(e_{X} \perp\!\!\!\perp e_{Y}\). Additional, the figures show one fork,
+\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
+\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
 \(X \rightarrow \{T, Y\}\), and two colliders:
@@
   Y & := f_{Y}(T,X,e_{Y}) \\
-  e_{T} & \perp\!\!\!\perp e_{X} \\
-  e_{T} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
+  e_{T} & \:\bot\:e_{X} \\
+  e_{T} & \:\bot\:e_{Y} \\
+  e_{X} & \:\bot\:e_{Y}
 \end{aligned}
@@
 
-\subsection{The probabilistic and causal implications of SCMs and
+\subsection{The probabilistic implications of SCMs and
 DAGs}\label{sec-appendix-D}
@@
 achieve this encoding by relying on three fundamental assumptions: the
-local Markov, the minimality, the causal edges and the modularity
-assumptions. The \emph{local Markov assumption} encodes probabilistic
-independencies between variables by declaring that nodes in a graph are
-independent of all its non-descendants, given its parents
-\citep[pp.~20]{Neal_2020}. Meanwhile, the \emph{minimality assumption}
-encodes probabilistic dependencies among variables by stating that every
-pair of adjacent nodes exhibits a dependency \citep[pp.~21]{Neal_2020}.
-Finally, the \emph{causal edges assumption} encodes causal relationships
-between variables by declaring that each parent node acts as a direct
-cause of its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
+local Markov, the minimality, the causal edges assumptions. The
+\emph{local Markov assumption} encodes probabilistic independencies
+between variables by declaring that nodes in a graph are independent of
+all its non-descendants, given its parents \citep[pp.~20]{Neal_2020}.
+Meanwhile, the \emph{minimality assumption} encodes probabilistic
+dependencies among variables by stating that every pair of adjacent
+nodes exhibits a dependency \citep[pp.~21]{Neal_2020}. Finally, the
+\emph{causal edges assumption} encodes causal relationships between
+variables by declaring that each parent node acts as a direct cause of
+its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
 illustrates how these assumptions influence the statistical and causal
@@
 
-The encoding of probabilistic information in SCMs and DAGs facilitates
-the efficient representation of the joint distribution of variables
-\citep[pp.~29]{Pearl_et_al_2016}. A notable implication of the
-assumptions underlying the probabilistic encoding is that any conceptual
-model described by an SCM and DAG can express the joint distribution of
-variables as a product of conditional probability distributions (CPDs)
-in the form \(P(child \: | \: parents)\). This property is formally
-known as the \emph{Bayesian Network factorization} (BNF)
+A notable implication of the assumptions underlying the probabilistic
+encoding is that any conceptual model described by an SCM and DAG can
+represent the joint distribution of variables more efficiently
+\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
+product of conditional probability distributions (CPDs)
+\(P(child \:|\:parents)\). This property is formally known as the
+\emph{Bayesian Network factorization} (BNF,
+Equation~\ref{eq-net_factor})
 \citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
-In this factorization, depicted in Equation~\ref{eq-net_factor},
-\(pa(X_{i})\) denotes the set of variables that are the parents of
-\(X_{i}\).
+In this expression, \(pa(X_{i})\) denotes the set of variables that are
+the parents of \(X_{i}\).
 
 \begin{equation}\phantomsection\label{eq-net_factor}{
-P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \: | \: pa(X_{i}) )
+P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \:|\:pa(X_{i}) )
 }\end{equation}
@@
 (in)dependencies in this data can provide significant insights into the
-validity of the proposed conceptual model. Moreover, it allows to
-express
+validity of the proposed conceptual model.
 
 The five fundamental building blocks described in
-Section~\ref{sec-appendix-C} illustrate how to predict (in)dependencies
-in data. For instance, applying the BNF to the causal structure shown in
-the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} supports the
-theoretical assumption that the observed variables \(X_{1}\) and
-\(X_{3}\) are unconditionally independent
-(\(X_{1} \perp\!\!\!\perp X_{3}\)). This conclusion follows from
-expressing their joint probability distribution as
+Section~\ref{sec-appendix-C} clearly illustrate which (in)dependencies
+can SMCs and DAGs predict. For instance, applying the BNF to the causal
+structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
+supports the theoretical assumption that the observed variables
+\(X_{1}\) and \(X_{3}\) are unconditionally independent
+(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
+joint probability distribution as
 \(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\), which defines probabilistic
 independence \citep[pp.~24]{Neal_2020}. Conversely, when \(X_{3}\) is
-unconditionally dependent on \(X_{1}\), as depicted in the SCM
-\ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}, the BNF express their
-joint probability distribution as
-\(P(X_{1}, X_{3}) = P(X_{3} \: | \: X_{1}) P(X_{1})\). Notably, these
+unconditionally dependent on \(X_{1}\) (\(X_{1} \:\not\bot\:X_{3}\)), as
+depicted in the SCM \ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}, the
+BNF express their joint probability distribution as
+\(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\). Notably, these
 descriptions demonstrate the clear correspondence between the structural
@@
 uncover more nuanced patterns of (in)dependencies from chains, forks,
-and colliders. By leveraging these patterns, researchers can identify
-the ACE from associational estimates through statistical conditioning,
-thereby disentangling causation from mere association. For instance,
-researchers applying the BNF to the chain structure depicted in the SCM
-\ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} can represent the joint
-distribution for the observed variables as
-\(P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}) P(X_{3} \: | \: X_{2})\).
-From this expression, researchers can infer additional (in)dependencies
-assumed in any data generated by this structure. Specifically, the
-structure implies that \(X_{1}\) and \(X_{3}\) are unconditionally
-dependent, but conditionally independent when controlling for \(X_{2}\)
-(\(X_{1} \perp\!\!\!\perp X_{3} \: | \: X_{2}\))
-\citetext{\citealp[pp.~37]{Pearl_et_al_2016}; \citealp[pp.~25]{Neal_2020}}.
-In the fork structure shown in the SCM \ref{fig-scm_bb4} and DAG
+and colliders. For instance, researchers applying the BNF to the chain
+structure depicted in the SCM \ref{fig-scm_bb3} and DAG
+\ref{fig-mdag_bb3} can represent the joint distribution for the observed
+variables as \(P(X_{1},X_{2},X_{3}) =\)
+\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). From this
+expression, researchers can infer additional (in)dependencies assumed in
+any data generated by this structure. Specifically, the structure
+implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
+\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
+controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). In the
+fork structure shown in the SCM \ref{fig-scm_bb4} and DAG
 \ref{fig-mdag_bb4}, researchers can express the joint distribution of
-the observed variables as
-\(P(X_{1},X_{2},X_{3}) = P(X_{1} \: | \: X_{2}) P(X_{2}) P(X_{3} \: | \: X_{2})\).
-This expression allows researchers to further infer that \(X_{1}\) and
-\(X_{3}\) are unconditionally dependent, but conditionally independent
-when controlling for \(X_{2}\)
-(\(X_{1} \perp\!\!\!\perp X_{3}  \: | \: X_{2}\))
-\citep[pp.~40]{Pearl_et_al_2016}. Finally, researchers analyzing the
+the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
+\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
+allows researchers to further infer that \(X_{1}\) and \(X_{3}\) are
+unconditionally dependent \((X_{1} \:\not\bot\:X_{3})\), but
+conditionally independent when controlling for \(X_{2}\)
+\((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally, researchers analyzing the
 collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
 \ref{fig-mdag_bb5} can express the joint distribution of the observed
-variables as
-\(P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}, X_{3}) P(X_{3})\).
-This representation allows researchers to infer that \(X_{1}\) and
-\(X_{3}\) are unconditionally independent
-(\(X_{1} \perp\!\!\!\perp X_{3}\)), but conditionally dependent when
-controlling for \(X_{2}\)
-\citetext{\citealp[pp.~41]{Pearl_et_al_2016}; \citealp[pp.~26]{Neal_2020}}.
-
-Ultimately, after encoding probabilistic information, researchers can
-use SCMs and DAGs to identify causal effects. This process relies on one
-additional assumption: the \emph{modularity assumption}. This assumption
-asserts that intervening on a node only alters its causal mechanism,
-implying that interventions are local \citep[pp.~34]{Neal_2020}. From
-this assumption, concepts such as manipulated DAGs and the Truncated
-factorization emerge. \emph{Manipulated graphs} are DAGs with removed
-edges, signifying intervention. Similarly, the \emph{Truncated
-factorization} represents these manipulated DAGs within SCMs
-(Equation~\ref{eq-trunc_factor}). In this expression, \(S\) denotes the
-subset of manipulated variables \(X_{p}\).
+variables as \(P(X_{1},X_{2},X_{3}) =\)
+\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
+allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
+unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
+dependent when controlling for \(X_{2}\)
+\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
+41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
+mathematical proofs for these conclusions.
+
+\subsection{From probability to causality}\label{sec-appendix-E}
 
 \begin{equation}\phantomsection\label{eq-trunc_factor}{
+P(X_{1}, X_{2}, \dots, X_{P} \:|\:do(S)) =
+\begin{cases}
+  \prod P(X_{i} \:|\:pa(X_{i}) ) & \text{if} \: p \not\in S \\
+  1 \quad & \text{otherwise}
+\end{cases}
+}\end{equation}
+
+\begin{equation}\phantomsection\label{eq-backdoor}{
+P(X_{a} \:|\:do(X_{b}=x)) = \sum_{Xp} P(X_{a} \:|\:X_{b}=x, X_{p}) P(X_{p})
+}\end{equation}
+
+\begin{equation}\phantomsection\label{eq-backdoor_adjustment}{
 \begin{aligned}
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = \prod_{p \not\in S} P(X_{i} \: | \: pa(X_{i}) ) \\
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = 1 \quad \quad (\text{otherwise})
+\tau &= E[\tau_{i}] \\
+  & = E[X_{a} \:|\:do(X_{b}=1)]- E[X_{a} \:|\:do(X_{b}=2)] \\
+  & = \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=1, X_{p}) P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=2, X_{p}) P(X_{p})
 \end{aligned}
 }\end{equation}
-
-Researchers can apply these new insights to revisit the scenario in
-Section~\ref{sec-appendix-B}. From the SCM \ref{fig-scm_example1} and
-DAG \ref{fig-mdag_example1a}, they can infer that the effect of \(T\) on
-\(Y\) is confounded by \(X\). This is evident because the effect
-decomposes into a direct causal effect (\(T \rightarrow Y\), black path)
-and a non-causal effect arising from the confounder \(X\)
-(\(T \leftarrow X \rightarrow Y\), gray path). Assuming the causal
-structure is correct, researchers can identify the CACE of \(T\) on
-\(Y\) by conditioning on \(X\) and then marginalize this effect over
-\(X\) to obtain the ACE. Conditioning on \(X\) means restricting the
-focus to schools with a specific socio-economic status \(X\)
-\citep[pp.~32]{Neal_2020}. Notably, researchers often depict
-conditioning as a square around a variable. Furthermore, the
-conditioning process leads to the manipulated DAG
-\ref{fig-mdag_example1b}, equivalent to the conditioned DAG
-\ref{fig-mdag_example1a}.
-
-Using Equation~\ref{eq-net_factor} and Equation~\ref{eq-trunc_factor},
-researchers can show that the joint probability of the observed
-variables in these structures can be expressed as
-\(P(Y, T, X) = P(Y \: | \: T, X) P(T \: | \: X) P(X)\). Upon
-intervention, this expression becomes
-\(P(Y, X \: | \: do(T=t)) = P(Y \: | \: T,X) P(X)\), and by
-marginalizing over \(X\), they obtain
-\(P(Y \: | \: do(T=t)) = \sum_{X} P(Y \: | \: T,X) P(X)\)
-\citep[pp.~36]{Neal_2020}. From these distributions, the ACE can be
-expressed as
-\(\tau = E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]\)
-(see Section~\ref{sec-appendix-B}).
 
@@
   Y & := f_{Y}(t,X,e_{Y}) \\
-  e_{T} & \perp\!\!\!\perp e_{X} \\
-  e_{T} & \perp\!\!\!\perp e_{Y} \\
-  e_{X} & \perp\!\!\!\perp e_{Y}
+  e_{T} & \:\bot\:e_{X} \\
+  e_{T} & \:\bot\:e_{Y} \\
+  e_{X} & \:\bot\:e_{Y}
 \end{aligned}
@@
 
-\subsection{The estimation process}\label{sec-appendix-E}
+\subsection{The estimation process}\label{sec-appendix-F}
 
@@
 approach begins by defining two probability distributions: the
-likelihood of the data,
-\(P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta)\), and the prior
-distribution, \(P(\theta)\) \citep{Everitt_et_al_2010}, where \(X_{P}\)
-represents a random variable, and \(\theta\) represents a
-one-dimensional parameter space for simplicity. After observing
+likelihood of the data, \(P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta)\),
+and the prior distribution, \(P(\theta)\) \citep{Everitt_et_al_2010},
+where \(X_{P}\) represents a random variable, and \(\theta\) represents
+a one-dimensional parameter space for simplicity. After observing
 empirical data, researchers can update the priors to posterior
@@
 \begin{equation}\phantomsection\label{eq-bayes_rule}{
-P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
+P(\theta \:|\:X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-prop_rule}{
-P(\theta \: | \: X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\: | \: \theta) \cdot P(\theta)
+P(\theta \:|\:X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\:|\:\theta) \cdot P(\theta)
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-like_chain}{
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: X_{i-1}, \dots, X_{1}, \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta) = P(X_{1} \:|\:\theta) \prod^{P}_{p=2} P(X_{i} \:|\:X_{i-1}, \dots, X_{1}, \theta )
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-like_BNF}{
-P(X_{1}, X_{2}, \dots, X_{P} \: | \: \theta) = P(X_{1} \: | \: \theta) \prod^{P}_{p=2} P(X_{i} \: | \: pa(X_{i}), \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta) = P(X_{1} \:|\:\theta) \prod^{P}_{p=2} P(X_{i} \:|\:pa(X_{i}), \theta )
 }\end{equation}

### 2025-01-17 paper2.qmd
section:appendix; process:write

@@
 Here, an *intervention* involves assigning a constant value to the treatment variable for each student's potential outcomes.
-In contrast, conditioning involves restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
+In contrast, *conditioning* involves restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
 32] (see @eq-group_diff).
@@
   X & := f_{X}(e_{X}) \\
-  T & := f_{Z}(X,e_{T}) \\
+  T & := f_{T}(X,e_{T}) \\
   Y & := f_{Y}(T,X,e_{Y}) \\
@@
 Beyond their graphical capabilities, SCMs and DAGs can encode the probabilistic information embedded within a causal structure.
-They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges assumptions.
+They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges assumption.
 The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020, pp.
@@
 29].
-This expression takes the form of a product of conditional probability distributions (CPDs) $P(child \cond parents)$.
+This expression takes the form of a product of conditional probability distributions (CPDs) of the type $P(child \cond parents)$.
 This property is formally known as the *Bayesian Network factorization* (BNF, @eq-net_factor) [@Pearl_et_al_2016, pp.
@@
 The five fundamental building blocks described in @sec-appendix-C clearly illustrate which (in)dependencies can SMCs and DAGs predict.
-For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$).
-This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, which defines probabilistic independence [@Neal_2020, pp.
+For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent or d-separated ($X_{1} \dsep X_{3}$).
+This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$ [@Neal_2020, pp.
 24].
 Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$.
-Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in the figures of @sec-appendix-C and the CPDs.
+Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendix-C and the CPDs.
 
 Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of (in)dependencies from chains, forks, and colliders.
-For instance, researchers applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] can represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}) P(X_{3} \cond X_{2})$.
-From this expression, researchers can infer additional (in)dependencies assumed in any data generated by this structure.
-Specifically, the structure implies that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
-In the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \cond X_{2}) P(X_{2}) P(X_{3} \cond X_{2})$.
+These (in)dependencies apply to any data generated by these structures.
+For instance, applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] allow researchers to represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}) P(X_{3} \cond X_{2})$.
+This expression implies that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
+Moreover, in the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \cond X_{2}) P(X_{2}) P(X_{3} \cond X_{2})$.
 This expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
@@
 <!-- 5. Continuing with the motivating example -->
-<!-- Researchers can apply these new insights to revisit the scenario in @sec-appendix-B.
-From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can express the joint probability for the observed variables as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$.
-From this expression, researchers can say that.
--->
+Researchers can revisit the scenario in @sec-appendix-B using these insights.
+By applying the BNF to the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$.
+Assuming the *direct effect* of $T$ on $Y$ ($T \rightarrow Y$, black path) is absent, researchers can infer from the joint probability that $Y$ is not unconditionally independent of $T$ $(Y \dsep T)$.
+This dependency arises because a non-causal path exists between $T$ and $Y$ through $X$ ($T \leftarrow X \rightarrow Y$, gray path), indicating that $X$ is a confounder.
+However, researchers can observe that conditioning on $X$, depicted as a square in DAG [-@fig-mdag_example1a], results in $Y \dsep T \cond X$, eliminating the confounding effect and allowing the identification of the direct effect $T \rightarrow Y$ if it exists.
+
+::: {#fig-example layout-ncol=2}
+
+::: {#fig-scm_example1 fig-align="center"}
+$$
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := f_{T}(X,e_{T}) \\
+  Y & := f_{Y}(t,X,e_{Y}) \\
+  e_{T} & \dsep e_{X} \\
+  e_{T} & \dsep e_{Y} \\
+  e_{X} & \dsep e_{Y}
+\end{aligned}
+$$
+
+SCM
+:::
+
+![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example1a fig-align="center" width=65%}
+
+Plausible causal structure the scenario outlined in @sec-appendix-B.
+:::
 
@@
 <!-- 1. But, how do we identify causal effects?  -->
-<!-- Researchers can use SCMs and DAGs to represent interventions of the type $P(X_{a} \cond do(X_{b}))$.
-These representations rely on the *modularity assumption*, which asserts that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
+In addition to the assumptions required to encode probabilistic information, SCMs and DAGs rely on the modularity assumption to represent interventions $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B).
+Defining these interventions enables researchers to identify the ACE from associational quantities through statistical conditioning, effectively disentangling causation from mere association.
+
+The modularity assumption underpins the concepts of Truncated Factorization and manipulated graphs, which are essential for representing interventions within SCMs and DAGs.
+The *modularity assumption* posits that intervening on a node modifies only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
 34].
--->
-
-<!-- By leveraging these patterns, researchers can identify the ACE from associational estimates through statistical conditioning, thereby disentangling causation from mere association.  -->
-
-<!-- Intervention relies on the concepts of Truncated factorization and manipulated graphs.
-The *Truncated factorization* (TF) simulates physical interventions by removing specific functions from the conceptual model and substituting them with constants, while keeping the remaining structure unchanged [Pearl_2010].
-The probabilistic implications for this factorization are expressed in @eq-trunc_factor.
-Here, $S$ denotes the subset of variables $X_{p}$ directly impacted by the intervention.
-Complementary to the TF, *manipulated graphs* are DAGs with certain edges removed to represent the effects of interventions [@Neal_2020, pp.
+Building on this principle, the *Truncated Factorization* (TF) simulates physical interventions by removing specific functions from the conceptual model and substituting them with constants, preserving the rest of the structure unchanged [Pearl_2010].
+The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ denotes the subset of variables $X_{p}$ directly influenced by the intervention.
+Complementing the TF, *manipulated graphs* are DAGs where specific edges have been removed to reflect the effects of interventions [@Neal_2020, pp.
 34].
--->
 
@@
 \begin{cases}
-  \prod P(X_{i} \cond pa(X_{i}) ) & \text{if} \: p \not\in S \\
+  \prod P(X_{p} \cond pa(X_{p}) ) & \text{if} \: p \not\in S \\
   1 \quad & \text{otherwise}
@@
 
-<!-- The Truncated Factorization enables researchers to define the *backdoor adjustment*.
-This adjustment specifies that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
--->
+Using the TF, researchers can define the *backdoor adjustment*, a method for identifying causal effects.
+This adjustment specifies that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014] (see @sec-appendix-B).
 
@@
 
-<!-- Ultimately, backdoor adjustment enables researchers to express the ACE as: -->
+Ultimately, backdoor adjustment enables researchers to express the ACE as:
 
@@
 
-::: {#fig-example layout-ncol=3}
+::: {#fig-example layout-ncol=2}
 
@@
 
-![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example1a fig-align="center" width=100%}
-
-![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example1b fig-align="center" width=100%}
+![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example1b fig-align="center" width=65%}
 

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

@@
 variable for each student's potential outcomes. In contrast,
-conditioning involves restricting the focus to the subset of the
+\emph{conditioning} involves restricting the focus to the subset of the
 population defined by the conditioning variable
@@
 achieve this encoding by relying on three fundamental assumptions: the
-local Markov, the minimality, the causal edges assumptions. The
+local Markov, the minimality, the causal edges assumption. The
 \emph{local Markov assumption} encodes probabilistic independencies
@@
 \citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
-product of conditional probability distributions (CPDs)
+product of conditional probability distributions (CPDs) of the type
 \(P(child \:|\:parents)\). This property is formally known as the
@@
 supports the theoretical assumption that the observed variables
-\(X_{1}\) and \(X_{3}\) are unconditionally independent
+\(X_{1}\) and \(X_{3}\) are unconditionally independent or d-separated
 (\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
 joint probability distribution as
-\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\), which defines probabilistic
-independence \citep[pp.~24]{Neal_2020}. Conversely, when \(X_{3}\) is
-unconditionally dependent on \(X_{1}\) (\(X_{1} \:\not\bot\:X_{3}\)), as
-depicted in the SCM \ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}, the
-BNF express their joint probability distribution as
-\(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\). Notably, these
-descriptions demonstrate the clear correspondence between the structural
-equations illustrated in the figures of Section~\ref{sec-appendix-C} and
+\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
+Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
+(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
+and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
+distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
+Notably, these descriptions demonstrate the clear correspondence between
+the structural equations illustrated in Section~\ref{sec-appendix-C} and
 the CPDs.
@@
 uncover more nuanced patterns of (in)dependencies from chains, forks,
-and colliders. For instance, researchers applying the BNF to the chain
-structure depicted in the SCM \ref{fig-scm_bb3} and DAG
-\ref{fig-mdag_bb3} can represent the joint distribution for the observed
+and colliders. These (in)dependencies apply to any data generated by
+these structures. For instance, applying the BNF to the chain structure
+depicted in the SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} allow
+researchers to represent the joint distribution for the observed
 variables as \(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). From this
-expression, researchers can infer additional (in)dependencies assumed in
-any data generated by this structure. Specifically, the structure
+\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
 implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
 \((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
-controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). In the
-fork structure shown in the SCM \ref{fig-scm_bb4} and DAG
-\ref{fig-mdag_bb4}, researchers can express the joint distribution of
-the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
+controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\).
+Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
+DAG \ref{fig-mdag_bb4}, researchers can express the joint distribution
+of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
 \(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
@@
 
+Researchers can revisit the scenario in Section~\ref{sec-appendix-B}
+using these insights. By applying the BNF to the SCM
+\ref{fig-scm_example1} and DAG \ref{fig-mdag_example1a} structures, they
+can represent the joint probability of the observed variables as
+\(P(Y, T, X) = P(Y \:|\:T, X) P(T \:|\:X) P(X)\). Assuming the
+\emph{direct effect} of \(T\) on \(Y\) (\(T \rightarrow Y\), black path)
+is absent, researchers can infer from the joint probability that \(Y\)
+is not unconditionally independent of \(T\) \((Y \:\bot\:T)\). This
+dependency arises because a non-causal path exists between \(T\) and
+\(Y\) through \(X\) (\(T \leftarrow X \rightarrow Y\), gray path),
+indicating that \(X\) is a confounder. However, researchers can observe
+that conditioning on \(X\), depicted as a square in DAG
+\ref{fig-mdag_example1a}, results in \(Y \:\bot\:T \:|\:X\), eliminating
+the confounding effect and allowing the identification of the direct
+effect \(T \rightarrow Y\) if it exists.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  X & := f_{X}(e_{X}) \\
+  T & := f_{T}(X,e_{T}) \\
+  Y & := f_{Y}(t,X,e_{Y}) \\
+  e_{T} & \:\bot\:e_{X} \\
+  e_{T} & \:\bot\:e_{Y} \\
+  e_{X} & \:\bot\:e_{Y}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-scm_example1}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1a.png}
+
+}
+
+\subcaption{\label{fig-mdag_example1a}Conditioned DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-example}Plausible causal structure the scenario
+outlined in Section~\ref{sec-appendix-B}.}
+
+\end{figure}%
+
 \subsection{From probability to causality}\label{sec-appendix-E}
+
+In addition to the assumptions required to encode probabilistic
+information, SCMs and DAGs rely on the modularity assumption to
+represent interventions \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
+Section~\ref{sec-appendix-B}). Defining these interventions enables
+researchers to identify the ACE from associational quantities through
+statistical conditioning, effectively disentangling causation from mere
+association.
+
+The modularity assumption underpins the concepts of Truncated
+Factorization and manipulated graphs, which are essential for
+representing interventions within SCMs and DAGs. The \emph{modularity
+assumption} posits that intervening on a node modifies only the causal
+mechanism of that node, leaving others unchanged
+\citep[pp.~34]{Neal_2020}. Building on this principle, the
+\emph{Truncated Factorization} (TF) simulates physical interventions by
+removing specific functions from the conceptual model and substituting
+them with constants, preserving the rest of the structure unchanged
+{[}Pearl\_2010{]}. The probabilistic implications of this factorization
+are formalized in Equation~\ref{eq-trunc_factor}, where \(S\) denotes
+the subset of variables \(X_{p}\) directly influenced by the
+intervention. Complementing the TF, \emph{manipulated graphs} are DAGs
+where specific edges have been removed to reflect the effects of
+interventions \citep[pp.~34]{Neal_2020}.
 
@@
 \begin{cases}
-  \prod P(X_{i} \:|\:pa(X_{i}) ) & \text{if} \: p \not\in S \\
+  \prod P(X_{p} \:|\:pa(X_{p}) ) & \text{if} \: p \not\in S \\
   1 \quad & \text{otherwise}
@@
 
+Using the TF, researchers can define the \emph{backdoor adjustment}, a
+method for identifying causal effects. This adjustment specifies that if
+a variable \(X_{p} \in S\) serves as a \emph{sufficient adjustment set}
+for the effect of \(X_{a}\) on \(X_{b}\), then the ACE can be identified
+using Equation~\ref{eq-backdoor} (see Section~\ref{sec-appendix-B}).
+
 \begin{equation}\phantomsection\label{eq-backdoor}{
@@
 }\end{equation}
+
+Ultimately, backdoor adjustment enables researchers to express the ACE
+as:
 
@@
 
-\begin{minipage}{0.33\linewidth}
+\begin{minipage}{0.50\linewidth}
 
@@
 %
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1a.png}
-
-}
-
-\subcaption{\label{fig-mdag_example1a}Conditioned DAG}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1b.png}
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1b.png}
 

### 2025-01-17 paper2.qmd
section:appendix; process:write

@@
 
+### The general idea {#sec-appendix-B1 .appendix appendix-style=plain}
+
 <!-- 1. What is causal inference? -->
@@
 For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as a variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
-Under these circumstances, researchers firt identify the *conditional average causal effect* (CACE, @eq-CACE)
+Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)
 
@@
 
-From the CACE, they identify the ACE from associational quantities as in @eq-mACE. Here, $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
+From the CACE, they identify the ACE from associational quantities as in @eq-mACE1.
+This identification process is commonly known as the *backdoor adjustment*.
+Here, $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
 
@@
 \end{aligned}
-$$ {#eq-mACE}
+$$ {#eq-mACE1}
 
@@
 
-## SCMs and DAGs {#sec-appendix-C .appendix appendix-style=plain}
+### Diving into the specifics {#sec-appendix-B2 .appendix appendix-style=plain}
+
+#### SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}
 
@@
 <!-- 5. Continuing with the motivating example -->
-Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B.
+Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B1.
 SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example.
@@
 
-Plausible causal structure the scenario outlined in @sec-appendix-B.
-:::
-
-## The probabilistic implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}
+Plausible causal structure the scenario outlined in @sec-appendix-B1.
+:::
+
+#### The probabilistic implications of SCMs and DAGs {#sec-appendix-B22 .appendix appendix-style=plain}
 
@@
 <!-- 4. can you illustrate this concept? -->
-The five fundamental building blocks described in @sec-appendix-C clearly illustrate which (in)dependencies can SMCs and DAGs predict.
+The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which (in)dependencies can SMCs and DAGs predict.
 For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent or d-separated ($X_{1} \dsep X_{3}$).
@@
 Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$.
-Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendix-C and the CPDs.
+Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendix-B21 and the CPDs.
 
@@
 <!-- 5. Continuing with the motivating example -->
-Researchers can revisit the scenario in @sec-appendix-B using these insights.
-By applying the BNF to the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$.
-Assuming the *direct effect* of $T$ on $Y$ ($T \rightarrow Y$, black path) is absent, researchers can infer from the joint probability that $Y$ is not unconditionally independent of $T$ $(Y \dsep T)$.
-This dependency arises because a non-causal path exists between $T$ and $Y$ through $X$ ($T \leftarrow X \rightarrow Y$, gray path), indicating that $X$ is a confounder.
-However, researchers can observe that conditioning on $X$, depicted as a square in DAG [-@fig-mdag_example1a], results in $Y \dsep T \cond X$, eliminating the confounding effect and allowing the identification of the direct effect $T \rightarrow Y$ if it exists.
+Researchers can revisit the scenario in @sec-appendix-B1 using these insights.
+By applying the BNF to the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
+From this expression, researchers can infer that the outcome $Y$ is not unconditionally independent of the teaching method $T$ $(Y \dsep T)$.
+This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path).
+
+These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on $X$ and depicted by a square in DAG [-@fig-mdag_example1a], effectively blocks all confounding non-causal path introduced by the fork structure.
+Nevertheless, since $Y$ remains not d-separated from $T$ even after conditioning $(Y \ndsep T \cond X)$, this lack of d-separation can only be attributed to the direct causal effect $T \rightarrow Y$.
 
@@
 
-::: {#fig-scm_example1 fig-align="center"}
+::: {#fig-scm_example1a fig-align="center"}
 $$
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{T}(X,e_{T}) \\
-  Y & := f_{Y}(t,X,e_{Y}) \\
+  X & := x \\
+  T & := f_{T}(x,e_{T}) \\
+  Y & := f_{Y}(T,x,e_{Y}) \\
   e_{T} & \dsep e_{X} \\
@@
 
-Plausible causal structure the scenario outlined in @sec-appendix-B.
-:::
-
-## From probability to causality {#sec-appendix-E .appendix appendix-style=plain}
-
-<!-- 1. But, how do we identify causal effects?  -->
-In addition to the assumptions required to encode probabilistic information, SCMs and DAGs rely on the modularity assumption to represent interventions $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B).
-Defining these interventions enables researchers to identify the ACE from associational quantities through statistical conditioning, effectively disentangling causation from mere association.
-
-The modularity assumption underpins the concepts of Truncated Factorization and manipulated graphs, which are essential for representing interventions within SCMs and DAGs.
-The *modularity assumption* posits that intervening on a node modifies only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
+Plausible causal structure the scenario outlined in @sec-appendix-B1.
+:::
+
+#### From probability to causality {#sec-appendix-B23 .appendix appendix-style=plain}
+
+<!-- 1. Transforming probabilistic association into causation?  -->
+The structural approach to causal inference translate the probabilistic insights from @sec-appendix-B23 into actionable strategies for identifying the ACE from associational quantities.
+The approach achieve this by relying on the modularity assumption and defining interventions of the form $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B1).
+The *modularity assumption* posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
 34].
-Building on this principle, the *Truncated Factorization* (TF) simulates physical interventions by removing specific functions from the conceptual model and substituting them with constants, preserving the rest of the structure unchanged [Pearl_2010].
-The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ denotes the subset of variables $X_{p}$ directly influenced by the intervention.
-Complementing the TF, *manipulated graphs* are DAGs where specific edges have been removed to reflect the effects of interventions [@Neal_2020, pp.
+
+<!-- 2. additional necessary concepts  -->
+The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions within SCMs and DAGs.
+*Manipulated graphs* simulate physical interventions by removing specific edges from a DAG, while preserving the remaining structure unchanged [@Neal_2020, pp.
 34].
+In parallel, *Truncated Factorization* (TF) achieves a similar simulation by removing specific functions from the conceptual model and replacing them with constants, while keeping the rest of the structure unchanged [Pearl_2010].
+The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ represents the subset of variables $X_{p}$ directly influenced by the intervention.
 
@@
 
-Using the TF, researchers can define the *backdoor adjustment*, a method for identifying causal effects.
-This adjustment specifies that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014] (see @sec-appendix-B).
+<!-- 3. How to use these concepts to finally define the ACE -->
+Using the TF, researchers can define the *backdoor adjustment* to identify the ACE.
+This adjustment states that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+The sufficient adjustment set (potentially empty) must block all non-causal paths between $X_{a}$ and $X_{b}$ without introducing new paths.
+If such a set exists, then $X_{a}$ and $X_{b}$ are *d-separated* by $X_{p}$ ($X_{a} \dsep X_{b} \cond X_{p}$) [@Pearl_2009], and $X_{p}$ satisfies the *backdoor criterion* [@Neal_2020, pp.
+37].
 
@@
 
-<!-- 6. Continuing with the motivating example -->
-<!-- Researchers can apply these new insights to revisit the scenario in @sec-appendix-B.
-From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can infer that the effect of $T$ on $Y$ is confounded by $X$.
--->
-
-<!-- This is evident because the effect decomposes into a *direct causal effect* ($T \rightarrow Y$, black path) and a non-causal effect arising from the confounder $X$ ($T \leftarrow X \rightarrow Y$, gray path).
--->
-
-<!-- Using @eq-net_factor and @eq-trunc_factor, researchers can show that the joint probability of the observed variables in these structures can be expressed as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$.
-Assuming intervention, this expression becomes $P(Y, X \cond do(T=t)) = P(Y \cond T,X) P(X)$, and by marginalizing over $X$, they obtain $P(Y \cond do(T=t)) = \sum_{X} P(Y \cond T,X) P(X)$ [@Neal_2020, pp.
-36].
--->
-
-<!-- Assuming the causal structure is correct, researchers can identify the CACE of $T$ on $Y$ by conditioning on $X$ and then marginalize this effect over $X$ to obtain the ACE.
-Conditioning on $X$ means restricting the focus to schools with a specific socio-economic status $X$ [@Neal_2020, pp.
-32].
-Notably, researchers often depict conditioning as a square around a variable.
-Furthermore, the conditioning process leads to the manipulated DAG [-@fig-mdag_example1b], equivalent to the conditioned DAG [-@fig-mdag_example1a].
--->
-
-<!--  From these distributions, the ACE can be expressed as $\tau = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]$ (see @sec-appendix-B). -->
+<!-- 4. revisiting the motivating example -->
+With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] is correct.
+This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure.
+In this context, researchers can apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$.
+They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE, as expressed in @eq-mACE2 (see @sec-appendix-B).
+Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example1a] is equivalent to the manipulated DAG [-@fig-mdag_example1b], because $X$ satisfies the backdoor criterion.
+
+$$
+\begin{aligned}
+  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
+\end{aligned}
+$$ {#eq-mACE2}
 
@@
 
-::: {#fig-scm_example1 fig-align="center"}
+::: {#fig-scm_example1b fig-align="center"}
 $$
@@
 
-Plausible causal structure the scenario outlined in @sec-appendix-B.
-:::
-
-## The estimation process {#sec-appendix-F .appendix appendix-style=plain}
+Plausible causal structure the scenario outlined in @sec-appendix-B22.
+:::
+
+## The estimation process {#sec-appendix-C .appendix appendix-style=plain}
 
 <!-- 1. How do we use Bayesian inference? -->
-Researchers can use Bayesian inference methods to estimate the ACE.
+Ultimately, researchers can use Bayesian inference methods to estimate the ACE.
 The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \cond \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
@@
 These assumptions enhance model tractability and facilitate the estimation process.
-As described in @sec-appendix-D, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF).
+As described in @sec-appendix-B22, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF).
 From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution from its priors using @eq-bayes_rule.

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

@@
 
+\subsubsection{The general idea}\label{sec-appendix-B1}
+
 Unlike classical statistical modeling, which focuses primarily on
@@
 satisfies the backdoor criterion. Under these circumstances, researchers
-firt identify the \emph{conditional average causal effect} (CACE,
+first identify the \emph{conditional average causal effect} (CACE,
 Equation~\ref{eq-CACE})
@@
 From the CACE, they identify the ACE from associational quantities as in
-Equation~\ref{eq-mACE}. Here, \(E_{X}[\cdot]\) represents the marginal
-expected value over \(X\) \citep{Morgan_et_al_2014}.
-
-\begin{equation}\phantomsection\label{eq-mACE}{
+Equation~\ref{eq-mACE1}. This identification process is commonly known
+as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\) represents the
+marginal expected value over \(X\) \citep{Morgan_et_al_2014}.
+
+\begin{equation}\phantomsection\label{eq-mACE1}{
 \begin{aligned}
@@
 
-\subsection{SCMs and DAGs}\label{sec-appendix-C}
+\subsubsection{Diving into the specifics}\label{sec-appendix-B2}
+
+\paragraph{SCMs and DAGs}\label{sec-appendix-B21}
 
@@
 Researchers can then use these building blocks to represent the scenario
-outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
-DAG \ref{fig-mdag_example1} depict the plausible causal structure for
-this example. In this context, the variable \(X\) (socio-economic status
-of the school) acts as a confounder in the relationship between the
-teaching method \(T\) and the outcome \(Y\). The figures display
+outlined in Section~\ref{sec-appendix-B1}. SCM \ref{fig-scm_example1}
+and DAG \ref{fig-mdag_example1} depict the plausible causal structure
+for this example. In this context, the variable \(X\) (socio-economic
+status of the school) acts as a confounder in the relationship between
+the teaching method \(T\) and the outcome \(Y\). The figures display
 multiple descendant relationships such as \(X \rightarrow T\),
@@
 \caption{\label{fig-example}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B}.}
-
-\end{figure}%
-
-\subsection{The probabilistic implications of SCMs and
-DAGs}\label{sec-appendix-D}
+outlined in Section~\ref{sec-appendix-B1}.}
+
+\end{figure}%
+
+\paragraph{The probabilistic implications of SCMs and
+DAGs}\label{sec-appendix-B22}
 
@@
 The five fundamental building blocks described in
-Section~\ref{sec-appendix-C} clearly illustrate which (in)dependencies
+Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
 can SMCs and DAGs predict. For instance, applying the BNF to the causal
@@
 Notably, these descriptions demonstrate the clear correspondence between
-the structural equations illustrated in Section~\ref{sec-appendix-C} and
-the CPDs.
+the structural equations illustrated in Section~\ref{sec-appendix-B21}
+and the CPDs.
 
@@
 
-Researchers can revisit the scenario in Section~\ref{sec-appendix-B}
+Researchers can revisit the scenario in Section~\ref{sec-appendix-B1}
 using these insights. By applying the BNF to the SCM
-\ref{fig-scm_example1} and DAG \ref{fig-mdag_example1a} structures, they
-can represent the joint probability of the observed variables as
-\(P(Y, T, X) = P(Y \:|\:T, X) P(T \:|\:X) P(X)\). Assuming the
-\emph{direct effect} of \(T\) on \(Y\) (\(T \rightarrow Y\), black path)
-is absent, researchers can infer from the joint probability that \(Y\)
-is not unconditionally independent of \(T\) \((Y \:\bot\:T)\). This
-dependency arises because a non-causal path exists between \(T\) and
-\(Y\) through \(X\) (\(T \leftarrow X \rightarrow Y\), gray path),
-indicating that \(X\) is a confounder. However, researchers can observe
-that conditioning on \(X\), depicted as a square in DAG
-\ref{fig-mdag_example1a}, results in \(Y \:\bot\:T \:|\:X\), eliminating
-the confounding effect and allowing the identification of the direct
-effect \(T \rightarrow Y\) if it exists.
+\ref{fig-scm_example1a} and DAG \ref{fig-mdag_example1a} structures,
+they can represent the joint probability of the observed variables as
+\(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this
+expression, researchers can infer that the outcome \(Y\) is not
+unconditionally independent of the teaching method \(T\)
+\((Y \:\bot\:T)\). This dependency arises from two key structures: a
+direct causal path from the teaching method \(T\) to the outcome \(Y\),
+represented by the two-connected-nodes structure \(T \rightarrow Y\)
+(black path), and a confounding non-causal path from the teaching method
+\(T\) to the outcome \(Y\) through the socio-economic status of the
+school \(X\), represented by the fork structure
+\(T \leftarrow X \rightarrow Y\) (gray path).
+
+These probabilistic insights also indicate that narrowing the focus to a
+subset of the population defined by the socio-economic status of the
+school \(X\), achieved by conditioning on \(X\) and depicted by a square
+in DAG \ref{fig-mdag_example1a}, effectively blocks all confounding
+non-causal path introduced by the fork structure. Nevertheless, since
+\(Y\) remains not d-separated from \(T\) even after conditioning
+\((Y \:\not\bot\:T \:|\:X)\), this lack of d-separation can only be
+attributed to the direct causal effect \(T \rightarrow Y\).
 
@@
 \begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{T}(X,e_{T}) \\
-  Y & := f_{Y}(t,X,e_{Y}) \\
+  X & := x \\
+  T & := f_{T}(x,e_{T}) \\
+  Y & := f_{Y}(T,x,e_{Y}) \\
   e_{T} & \:\bot\:e_{X} \\
@@
 
-\subcaption{\label{fig-scm_example1}SCM}
+\subcaption{\label{fig-scm_example1a}SCM}
 
@@
 \caption{\label{fig-example}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B}.}
-
-\end{figure}%
-
-\subsection{From probability to causality}\label{sec-appendix-E}
-
-In addition to the assumptions required to encode probabilistic
-information, SCMs and DAGs rely on the modularity assumption to
-represent interventions \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
-Section~\ref{sec-appendix-B}). Defining these interventions enables
-researchers to identify the ACE from associational quantities through
-statistical conditioning, effectively disentangling causation from mere
-association.
-
-The modularity assumption underpins the concepts of Truncated
-Factorization and manipulated graphs, which are essential for
-representing interventions within SCMs and DAGs. The \emph{modularity
-assumption} posits that intervening on a node modifies only the causal
-mechanism of that node, leaving others unchanged
-\citep[pp.~34]{Neal_2020}. Building on this principle, the
-\emph{Truncated Factorization} (TF) simulates physical interventions by
-removing specific functions from the conceptual model and substituting
-them with constants, preserving the rest of the structure unchanged
-{[}Pearl\_2010{]}. The probabilistic implications of this factorization
-are formalized in Equation~\ref{eq-trunc_factor}, where \(S\) denotes
-the subset of variables \(X_{p}\) directly influenced by the
-intervention. Complementing the TF, \emph{manipulated graphs} are DAGs
-where specific edges have been removed to reflect the effects of
-interventions \citep[pp.~34]{Neal_2020}.
+outlined in Section~\ref{sec-appendix-B1}.}
+
+\end{figure}%
+
+\paragraph{From probability to causality}\label{sec-appendix-B23}
+
+The structural approach to causal inference translate the probabilistic
+insights from Section~\ref{sec-appendix-B23} into actionable strategies
+for identifying the ACE from associational quantities. The approach
+achieve this by relying on the modularity assumption and defining
+interventions of the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
+Section~\ref{sec-appendix-B1}). The \emph{modularity assumption} posits
+that intervening on a node alters only the causal mechanism of that
+node, leaving others unchanged \citep[pp.~34]{Neal_2020}.
+
+The modularity assumption underpins the concepts of manipulated graphs
+and Truncated Factorization, which are essential for representing
+interventions within SCMs and DAGs. \emph{Manipulated graphs} simulate
+physical interventions by removing specific edges from a DAG, while
+preserving the remaining structure unchanged \citep[pp.~34]{Neal_2020}.
+In parallel, \emph{Truncated Factorization} (TF) achieves a similar
+simulation by removing specific functions from the conceptual model and
+replacing them with constants, while keeping the rest of the structure
+unchanged {[}Pearl\_2010{]}. The probabilistic implications of this
+factorization are formalized in Equation~\ref{eq-trunc_factor}, where
+\(S\) represents the subset of variables \(X_{p}\) directly influenced
+by the intervention.
 
@@
 
-Using the TF, researchers can define the \emph{backdoor adjustment}, a
-method for identifying causal effects. This adjustment specifies that if
-a variable \(X_{p} \in S\) serves as a \emph{sufficient adjustment set}
-for the effect of \(X_{a}\) on \(X_{b}\), then the ACE can be identified
-using Equation~\ref{eq-backdoor} (see Section~\ref{sec-appendix-B}).
+Using the TF, researchers can define the \emph{backdoor adjustment} to
+identify the ACE. This adjustment states that if a variable
+\(X_{p} \in S\) serves as a \emph{sufficient adjustment set} for the
+effect of \(X_{a}\) on \(X_{b}\), then the ACE can be identified using
+Equation~\ref{eq-backdoor}. The sufficient adjustment set (potentially
+empty) must block all non-causal paths between \(X_{a}\) and \(X_{b}\)
+without introducing new paths. If such a set exists, then \(X_{a}\) and
+\(X_{b}\) are \emph{d-separated} by \(X_{p}\)
+(\(X_{a} \:\bot\:X_{b} \:|\:X_{p}\)) \citep{Pearl_2009}, and \(X_{p}\)
+satisfies the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020}.
 
@@
   & = \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=1, X_{p}) P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=2, X_{p}) P(X_{p})
+\end{aligned}
+}\end{equation}
+
+With these new insights, researchers revisiting the scenario in
+Section~\ref{sec-appendix-B22} can infer that the socio-economic status
+of the school, \(X\), satisfies the backdoor criterion, assuming the
+causal structure depicted by the SCM \ref{fig-scm_example1a} and DAG
+\ref{fig-mdag_example1a} is correct. This means that \(X\) serves as a
+sufficient adjustment set, as it effectively blocks all confounding
+non-causal paths introduced by the fork structure. In this context,
+researchers can apply the \emph{backdoor adjustment} to identify the ACE
+of \(T\) on \(Y\). They achieve this by first identifying the CACE of
+\(T\) on \(Y\) by conditioning on \(X\), and then marginalizing this
+effect over \(X\) to obtain the ACE, as expressed in
+Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
+the purpose of identification, the conditioned DAG
+\ref{fig-mdag_example1a} is equivalent to the manipulated DAG
+\ref{fig-mdag_example1b}, because \(X\) satisfies the backdoor
+criterion.
+
+\begin{equation}\phantomsection\label{eq-mACE2}{
+\begin{aligned}
+  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
 \end{aligned}
@@
 
-\subcaption{\label{fig-scm_example1}SCM}
+\subcaption{\label{fig-scm_example1b}SCM}
 
@@
 \caption{\label{fig-example}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B}.}
-
-\end{figure}%
-
-\subsection{The estimation process}\label{sec-appendix-F}
-
-Researchers can use Bayesian inference methods to estimate the ACE. The
-approach begins by defining two probability distributions: the
-likelihood of the data, \(P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta)\),
-and the prior distribution, \(P(\theta)\) \citep{Everitt_et_al_2010},
-where \(X_{P}\) represents a random variable, and \(\theta\) represents
-a one-dimensional parameter space for simplicity. After observing
+outlined in Section~\ref{sec-appendix-B22}.}
+
+\end{figure}%
+
+\subsection{The estimation process}\label{sec-appendix-C}
+
+Ultimately, researchers can use Bayesian inference methods to estimate
+the ACE. The approach begins by defining two probability distributions:
+the likelihood of the data,
+\(P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta)\), and the prior
+distribution, \(P(\theta)\) \citep{Everitt_et_al_2010}, where \(X_{P}\)
+represents a random variable, and \(\theta\) represents a
+one-dimensional parameter space for simplicity. After observing
 empirical data, researchers can update the priors to posterior
@@
 assumptions enhance model tractability and facilitate the estimation
-process. As described in Section~\ref{sec-appendix-D}, researchers can
+process. As described in Section~\ref{sec-appendix-B22}, researchers can
 use SCMs and DAGs to formalize these local (in)dependencies, which leads

### 2025-01-17 paper2.qmd
section:appendix; process:write

@@
 
-::: {#fig-example layout-ncol=2}
+::: {#fig-example1 layout-ncol=2}
 
@@
 Researchers can revisit the scenario in @sec-appendix-B1 using these insights.
-By applying the BNF to the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
+By applying the BNF to the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
 From this expression, researchers can infer that the outcome $Y$ is not unconditionally independent of the teaching method $T$ $(Y \dsep T)$.
@@
 
-These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on $X$ and depicted by a square in DAG [-@fig-mdag_example1a], effectively blocks all confounding non-causal path introduced by the fork structure.
+These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on $X$ and depicted by a square in DAG [-@fig-mdag_example2], effectively blocks all confounding non-causal path introduced by the fork structure.
 Nevertheless, since $Y$ remains not d-separated from $T$ even after conditioning $(Y \ndsep T \cond X)$, this lack of d-separation can only be attributed to the direct causal effect $T \rightarrow Y$.
 
-::: {#fig-example layout-ncol=2}
-
-::: {#fig-scm_example1a fig-align="center"}
+::: {#fig-example2 layout-ncol=2}
+
+::: {#fig-scm_example2 fig-align="center"}
 $$
@@
 
-![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example1a fig-align="center" width=65%}
+![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example2 fig-align="center" width=65%}
 
@@
 <!-- 4. revisiting the motivating example -->
-With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] is correct.
+With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct.
 This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure.
@@
 They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE, as expressed in @eq-mACE2 (see @sec-appendix-B).
-Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example1a] is equivalent to the manipulated DAG [-@fig-mdag_example1b], because $X$ satisfies the backdoor criterion.
+Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.
 
@@
 
-::: {#fig-example layout-ncol=2}
-
-::: {#fig-scm_example1b fig-align="center"}
+::: {#fig-example3 layout-ncol=2}
+
+::: {#fig-scm_example3 fig-align="center"}
 $$
@@
 
-![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example1b fig-align="center" width=65%}
+![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example3 fig-align="center" width=65%}
 

### 2025-01-17 _tex/paper2.tex
Built site for gh-pages

@@
 
-\caption{\label{fig-example}Plausible causal structure the scenario
+\caption{\label{fig-example1}Plausible causal structure the scenario
 outlined in Section~\ref{sec-appendix-B1}.}
@@
 using these insights. By applying the BNF to the SCM
-\ref{fig-scm_example1a} and DAG \ref{fig-mdag_example1a} structures,
-they can represent the joint probability of the observed variables as
+\ref{fig-scm_example2} and DAG \ref{fig-mdag_example2} structures, they
+can represent the joint probability of the observed variables as
 \(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this
@@
 school \(X\), achieved by conditioning on \(X\) and depicted by a square
-in DAG \ref{fig-mdag_example1a}, effectively blocks all confounding
+in DAG \ref{fig-mdag_example2}, effectively blocks all confounding
 non-causal path introduced by the fork structure. Nevertheless, since
@@
 
-\subcaption{\label{fig-scm_example1a}SCM}
+\subcaption{\label{fig-scm_example2}SCM}
 
@@
 
-\subcaption{\label{fig-mdag_example1a}Conditioned DAG}
+\subcaption{\label{fig-mdag_example2}Conditioned DAG}
 
@@
 
-\caption{\label{fig-example}Plausible causal structure the scenario
+\caption{\label{fig-example2}Plausible causal structure the scenario
 outlined in Section~\ref{sec-appendix-B1}.}
@@
 of the school, \(X\), satisfies the backdoor criterion, assuming the
-causal structure depicted by the SCM \ref{fig-scm_example1a} and DAG
-\ref{fig-mdag_example1a} is correct. This means that \(X\) serves as a
+causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
+\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
 sufficient adjustment set, as it effectively blocks all confounding
@@
 the purpose of identification, the conditioned DAG
-\ref{fig-mdag_example1a} is equivalent to the manipulated DAG
-\ref{fig-mdag_example1b}, because \(X\) satisfies the backdoor
-criterion.
+\ref{fig-mdag_example2} is equivalent to the manipulated DAG
+\ref{fig-mdag_example3}, because \(X\) satisfies the backdoor criterion.
 
@@
 
-\subcaption{\label{fig-scm_example1b}SCM}
+\subcaption{\label{fig-scm_example3}SCM}
 
@@
 
-\subcaption{\label{fig-mdag_example1b}Manipulated DAG}
+\subcaption{\label{fig-mdag_example3}Manipulated DAG}
 
@@
 
-\caption{\label{fig-example}Plausible causal structure the scenario
+\caption{\label{fig-example3}Plausible causal structure the scenario
 outlined in Section~\ref{sec-appendix-B22}.}

### 2025-01-20 paper2.qmd
section:appendix; process:write

@@
 Researchers would then proceed to identify the estimands.
-*Identification* is concerned with determining whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+*Identification* determines whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
 4].
 An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand.
-They are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
+*Estimates* are numerical values that approximate the estimand, and they are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
 The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.
@@
 
-### The general idea {#sec-appendix-B1 .appendix appendix-style=plain}
+<!-- ### The general idea {#sec-appendix-B1 .appendix appendix-style=plain} -->
 
@@
 Here, an *intervention* involves assigning a constant value to the treatment variable for each student's potential outcomes.
-In contrast, *conditioning* involves restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
-32] (see @eq-group_diff).
-<!-- Conditioning on T=t just means that we are restricting our focus to the subset of the population to those who received treatment t.
-In contrast, an intervention would be to take the whole population and give everyone treatment t [@Neal_2020, pp.
-32].
--->
-
 Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable.
@@
 $$
-\tau = E[\tau_{i}] = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)]
+\begin{aligned}
+\tau & = E[\tau_{i}] \\
+  & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)]
+\end{aligned}
 $$ {#eq-ACE}
@@
 If such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
+Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
+32] (see @eq-group_diff).
+
+<!-- Conditioning on T=t just means that we are restricting our focus to the subset of the population to those who received treatment t.
+In contrast, an intervention would be to take the whole population and give everyone treatment t [@Neal_2020, pp.
+32].
+-->
 
@@
 $$
-E[Y_{i} \cond T_{i}=t, X]
+CACE_{t} = E[Y_{i} \cond T_{i}=t, X]
 $$ {#eq-CACE}
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
+  \tau & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)] \\
+  & = E_{X}[CACE_{1} - CACE_{2}] \\
   & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
@@
 
-#### SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}
-
 <!-- 1. The benefits of SCMs and DAGs -->
@@
 Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference by guiding researchers in determining which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020].
+@fig-IEflow shows the role of theoretical models in the inference process.
 
@@
 35].
-@fig-IEflow shows the role of theoretical models in the inference process.
+@sec-appendix-B21 and @sec-appendix-B22 elaborate on the first advantage, while @sec-appendix-B22 and @sec-appendix-B23 do so for the second.
 
@@
 
+#### SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}
+
 <!-- 2. the SCMs -->
-Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display these five fundamental building blocks.
+Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display the five fundamental building blocks for SCMs and DAGs.
 The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $X=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020].
@@
 DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-Although DAGs typically omit exogenous variables for simplicity (the *standard representation*), including these variables, as demonstrated in the *magnified representation* in the figures, can be advantageous.
-Their inclusion helps to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+Although the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* depicted in the figures offers some advantages: including exogenous variables can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$.
-Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because there is a *direct path* connecting.
+Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because there is a *direct path* connecting them.
 SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
@@
 <!-- 5. Continuing with the motivating example -->
-Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B1.
+Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B.
 SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example.
@@
 
-Plausible causal structure the scenario outlined in @sec-appendix-B1.
+Plausible causal structure the scenario outlined in @sec-appendix-B.
 :::
@@
 35], these predictions depend solely on the structure of these conceptual models without requiring the quantitative details of the equations or the distributions of the errors.
-Moreover, once researchers observe empirical data, the patterns of (in)dependencies in this data can provide significant insights into the validity of the proposed conceptual model.
+Moreover, once researchers observe empirical data, the patterns of (in)dependencies in the data can provide significant insights into the validity of the proposed conceptual model.
 
@@
 <!-- 4. can you illustrate this concept? -->
-The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which (in)dependencies can SMCs and DAGs predict.
-For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent or d-separated ($X_{1} \dsep X_{3}$).
+The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which variables' (in)dependencies can SMCs and DAGs predict.
+For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$).
 This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$ [@Neal_2020, pp.
@@
 
-Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of (in)dependencies from chains, forks, and colliders.
+Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of variables' (in)dependencies from chains, forks, and colliders.
 These (in)dependencies apply to any data generated by these structures.
@@
 <!-- 5. Continuing with the motivating example -->
-Researchers can revisit the scenario in @sec-appendix-B1 using these insights.
-By applying the BNF to the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
+Then, researchers can revisit the scenario in @sec-appendix-B using these insights.
+In this context, applying the BNF to the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] structures, can help researchers to represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
 From this expression, researchers can infer that the outcome $Y$ is not unconditionally independent of the teaching method $T$ $(Y \dsep T)$.
@@
 
-These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on $X$ and depicted by a square in DAG [-@fig-mdag_example2], effectively blocks all confounding non-causal path introduced by the fork structure.
-Nevertheless, since $Y$ remains not d-separated from $T$ even after conditioning $(Y \ndsep T \cond X)$, this lack of d-separation can only be attributed to the direct causal effect $T \rightarrow Y$.
+These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on the variable and depicted by a square in DAG [-@fig-mdag_example2], effectively blocks all confounding non-causal path introduced by the fork structure.
+Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.
 
@@
 The structural approach to causal inference translate the probabilistic insights from @sec-appendix-B23 into actionable strategies for identifying the ACE from associational quantities.
-The approach achieve this by relying on the modularity assumption and defining interventions of the form $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B1).
+The approach achieve this by relying on the modularity assumption and by re-defining interventions in the form $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B).
 The *modularity assumption* posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
@@
 34].
-In parallel, *Truncated Factorization* (TF) achieves a similar simulation by removing specific functions from the conceptual model and replacing them with constants, while keeping the rest of the structure unchanged [Pearl_2010].
+In parallel, *Truncated Factorization* (TF) achieves a similar simulation by removing specific functions from the conceptual model and replacing them with constants, while keeping the rest of the structure unchanged [@Pearl_2010].
 The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ represents the subset of variables $X_{p}$ directly influenced by the intervention.
@@
 
-Ultimately, backdoor adjustment enables researchers to express the ACE as:
+Ultimately, the backdoor adjustment enables researchers to express the ACE as:
 
@@
 \begin{aligned}
-\tau &= E[\tau_{i}] \\
-  & = E[X_{a} \cond do(X_{b}=1)]- E[X_{a} \cond do(X_{b}=2)] \\
-  & = \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=1, X_{p}) P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=2, X_{p}) P(X_{p})
+\tau & = E[X_{a} \cond do(X_{b}=1)]- E[X_{a} \cond do(X_{b}=2)] \\
+  & = E_{Xp}\left[ E[X_{a} \cond do(X_{b}=1), X_{p}]- E[X_{a} \cond do(X_{b}=2), X_{p}] \right] \\
+  & = \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=2, X_{p}) \cdot P(X_{p})
 \end{aligned}
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
+  \tau & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right] \\
+  & = \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=2, X) \cdot P(X)
 \end{aligned}
@@
 The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \cond \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
-After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:
+After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule in @eq-bayes_rule [@Jeffreys_1998]:
 
@@
 
-Given that the denominator on the right-hand side of @eq-bayes_rule serves as a normalizing constant independent of the parameter $\theta$, researchers can simplify the posterior updating process into the three steps depicted by @eq-prop_rule.
+Given that the denominator on the right-hand side of @eq-bayes_rule serves as a normalizing constant independent of the parameter $\theta$, researchers can simplify the posterior updating process into three steps.
 First, they integrate new empirical data through the likelihood.
@@
 Temporarily setting aside the definition of prior distributions $P(\theta)$, note that the posterior updating process depends heavily on the assumptions underlying the likelihood of the data.
-However, as the number of random variables, $P$, increases, this joint distribution becomes quickly intractable [@Neal_2020].
+However, as the number of random variables, $P$, increases, this joint distribution quickly becomes intractable [@Neal_2020].
 This intractability is evident from @eq-like_chain, where the likelihood distribution is expressed by multiple chained CPDs.
@@
 As described in @sec-appendix-B22, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF).
-From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution from its priors using @eq-bayes_rule.
+From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution using @eq-bayes_rule.
 

### 2025-01-20 _tex/paper2.tex
Built site for gh-pages

@@
 Researchers would then proceed to identify the estimands.
-\emph{Identification} is concerned with determining whether an estimator
-can accurately compute an estimand based on its assumptions, regardless
-of random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
-\emph{estimator} is a method or function that transforms data into an
-estimate \citep{Neal_2020}. \emph{Estimates} are numerical values that
-approximate the estimand. They are derived through \emph{estimation},
-which refers to the process of integrating data with an estimator
-\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
-\citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow} provides a
-visual representation of the process of transitioning from estimands to
-estimates.
+\emph{Identification} determines whether an estimator can accurately
+compute an estimand based on its assumptions, regardless of random
+variability \citep[pp.~4]{Schuessler_et_al_2023}. An \emph{estimator} is
+a method or function that transforms data into an estimate
+\citep{Neal_2020}. \emph{Estimates} are numerical values that
+approximate the estimand, and they are derived through
+\emph{estimation}, which refers to the process of integrating data with
+an estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
+flowchart \citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow}
+provides a visual representation of the process of transitioning from
+estimands to estimates.
 
@@
 
-\subsubsection{The general idea}\label{sec-appendix-B1}
-
 Unlike classical statistical modeling, which focuses primarily on
@@
 \emph{intervention} involves assigning a constant value to the treatment
-variable for each student's potential outcomes. In contrast,
-\emph{conditioning} involves restricting the focus to the subset of the
-population defined by the conditioning variable
-\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).\\
-
-Note that if a student is assigned to intervention \(T_{i}=1\), the
-potential outcome under \(T_{i}=2\) becomes a counterfactual, as it is
-no longer observed nor observable. To address this challenge, the
-structural approach extends the ICE to the \emph{average causal effect}
-(ACE, Equation~\ref{eq-ACE}), representing the average difference
-between observed potential outcomes and their counterfactual
-counterparts.
+variable for each student's potential outcomes. Note that if a student
+is assigned to intervention \(T_{i}=1\), the potential outcome under
+\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
+observable. To address this challenge, the structural approach extends
+the ICE to the \emph{average causal effect} (ACE,
+Equation~\ref{eq-ACE}), representing the average difference between
+observed potential outcomes and their counterfactual counterparts.
 
 \begin{equation}\phantomsection\label{eq-ACE}{
-\tau = E[\tau_{i}] = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
+\begin{aligned}
+\tau & = E[\tau_{i}] \\
+  & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
+\end{aligned}
 }\end{equation}
@@
 (\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
-\emph{backdoor criterion} \citep[pp 37]{Neal_2020}.
+\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
+\emph{conditioning} describes the process of restricting the focus to
+the subset of the population defined by the conditioning variable
+\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).
 
@@
 \begin{equation}\phantomsection\label{eq-CACE}{
-E[Y_{i} \:|\:T_{i}=t, X]
+CACE_{t} = E[Y_{i} \:|\:T_{i}=t, X]
 }\end{equation}
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
+  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
+  & = E_{X}[CACE_{1} - CACE_{2}] \\
   & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
@@
 \subsubsection{Diving into the specifics}\label{sec-appendix-B2}
-
-\paragraph{SCMs and DAGs}\label{sec-appendix-B21}
 
@@
 CACE, or other), assuming the depicted causal structure is correct
-\citep{McElreath_2020}.
+\citep{McElreath_2020}. Figure~\ref{fig-IEflow} shows the role of
+theoretical models in the inference process.
 
@@
 their parameters \citep[pp.~35]{Pearl_et_al_2016}.
-Figure~\ref{fig-IEflow} shows the role of theoretical models in the
-inference process.
+Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
+elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
+and Section~\ref{sec-appendix-B23} do so for the second.
+
+\paragraph{SCMs and DAGs}\label{sec-appendix-B21}
 
@@
 \ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
-display these five fundamental building blocks. The left panels of the
-figures show the formal mathematical models, represented by the SCMs,
-defined in terms of a set of \emph{endogenous} variables
+display the five fundamental building blocks for SCMs and DAGs. The left
+panels of the figures show the formal mathematical models, represented
+by the SCMs, defined in terms of a set of \emph{endogenous} variables
 \(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
@@
 circles, while they use open circles for unobserved (latent) variables
-\citep{Morgan_et_al_2014}. Although DAGs typically omit exogenous
-variables for simplicity (the \emph{standard representation}), including
-these variables, as demonstrated in the \emph{magnified representation}
-in the figures, can be advantageous. Their inclusion helps to highlight
+\citep{Morgan_et_al_2014}. Although the \emph{standard representation}
+of DAGs typically omits exogenous variables for simplicity, the
+\emph{magnified representation} depicted in the figures offers some
+advantages: including exogenous variables can help to highlight
 potential issues related to conditioning and confounding
@@
 are described as \emph{adjacent} because there is a \emph{direct path}
-connecting. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a
-\emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
+connecting them. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict
+a \emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
 influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
@@
 Researchers can then use these building blocks to represent the scenario
-outlined in Section~\ref{sec-appendix-B1}. SCM \ref{fig-scm_example1}
-and DAG \ref{fig-mdag_example1} depict the plausible causal structure
-for this example. In this context, the variable \(X\) (socio-economic
-status of the school) acts as a confounder in the relationship between
-the teaching method \(T\) and the outcome \(Y\). The figures display
+outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
+DAG \ref{fig-mdag_example1} depict the plausible causal structure for
+this example. In this context, the variable \(X\) (socio-economic status
+of the school) acts as a confounder in the relationship between the
+teaching method \(T\) and the outcome \(Y\). The figures display
 multiple descendant relationships such as \(X \rightarrow T\),
@@
 \caption{\label{fig-example1}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B1}.}
+outlined in Section~\ref{sec-appendix-B}.}
 
@@
 once researchers observe empirical data, the patterns of
-(in)dependencies in this data can provide significant insights into the
+(in)dependencies in the data can provide significant insights into the
 validity of the proposed conceptual model.
@@
 The five fundamental building blocks described in
-Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
-can SMCs and DAGs predict. For instance, applying the BNF to the causal
-structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
-supports the theoretical assumption that the observed variables
-\(X_{1}\) and \(X_{3}\) are unconditionally independent or d-separated
+Section~\ref{sec-appendix-B21} clearly illustrate which variables'
+(in)dependencies can SMCs and DAGs predict. For instance, applying the
+BNF to the causal structure shown in the SCM \ref{fig-scm_bb1} and DAG
+\ref{fig-mdag_bb1} supports the theoretical assumption that the observed
+variables \(X_{1}\) and \(X_{3}\) are unconditionally independent
 (\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
@@
 Beyond the insights gained from two-node structures, researchers can
-uncover more nuanced patterns of (in)dependencies from chains, forks,
-and colliders. These (in)dependencies apply to any data generated by
-these structures. For instance, applying the BNF to the chain structure
-depicted in the SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} allow
-researchers to represent the joint distribution for the observed
-variables as \(P(X_{1},X_{2},X_{3}) =\)
+uncover more nuanced patterns of variables' (in)dependencies from
+chains, forks, and colliders. These (in)dependencies apply to any data
+generated by these structures. For instance, applying the BNF to the
+chain structure depicted in the SCM \ref{fig-scm_bb3} and DAG
+\ref{fig-mdag_bb3} allow researchers to represent the joint distribution
+for the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
 \(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
@@
 
-Researchers can revisit the scenario in Section~\ref{sec-appendix-B1}
-using these insights. By applying the BNF to the SCM
-\ref{fig-scm_example2} and DAG \ref{fig-mdag_example2} structures, they
-can represent the joint probability of the observed variables as
-\(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this
-expression, researchers can infer that the outcome \(Y\) is not
-unconditionally independent of the teaching method \(T\)
-\((Y \:\bot\:T)\). This dependency arises from two key structures: a
-direct causal path from the teaching method \(T\) to the outcome \(Y\),
-represented by the two-connected-nodes structure \(T \rightarrow Y\)
-(black path), and a confounding non-causal path from the teaching method
-\(T\) to the outcome \(Y\) through the socio-economic status of the
-school \(X\), represented by the fork structure
-\(T \leftarrow X \rightarrow Y\) (gray path).
+Then, researchers can revisit the scenario in
+Section~\ref{sec-appendix-B} using these insights. In this context,
+applying the BNF to the SCM \ref{fig-scm_example2} and DAG
+\ref{fig-mdag_example2} structures, can help researchers to represent
+the joint probability of the observed variables as \(P(Y, T, X) =\)
+\(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this expression, researchers
+can infer that the outcome \(Y\) is not unconditionally independent of
+the teaching method \(T\) \((Y \:\bot\:T)\). This dependency arises from
+two key structures: a direct causal path from the teaching method \(T\)
+to the outcome \(Y\), represented by the two-connected-nodes structure
+\(T \rightarrow Y\) (black path), and a confounding non-causal path from
+the teaching method \(T\) to the outcome \(Y\) through the
+socio-economic status of the school \(X\), represented by the fork
+structure \(T \leftarrow X \rightarrow Y\) (gray path).
 
@@
 subset of the population defined by the socio-economic status of the
-school \(X\), achieved by conditioning on \(X\) and depicted by a square
-in DAG \ref{fig-mdag_example2}, effectively blocks all confounding
-non-causal path introduced by the fork structure. Nevertheless, since
-\(Y\) remains not d-separated from \(T\) even after conditioning
-\((Y \:\not\bot\:T \:|\:X)\), this lack of d-separation can only be
+school \(X\), achieved by conditioning on the variable and depicted by a
+square in DAG \ref{fig-mdag_example2}, effectively blocks all
+confounding non-causal path introduced by the fork structure.
+Nevertheless, since \(Y\) remains dependent on \(T\) even after
+conditioning \((Y \:\not\bot\:T \:|\:X)\), this dependency can only be
 attributed to the direct causal effect \(T \rightarrow Y\).
@@
 \caption{\label{fig-example2}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B1}.}
+outlined in Section~\ref{sec-appendix-B}.}
 
@@
 for identifying the ACE from associational quantities. The approach
-achieve this by relying on the modularity assumption and defining
-interventions of the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
-Section~\ref{sec-appendix-B1}). The \emph{modularity assumption} posits
+achieve this by relying on the modularity assumption and by re-defining
+interventions in the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
+Section~\ref{sec-appendix-B}). The \emph{modularity assumption} posits
 that intervening on a node alters only the causal mechanism of that
@@
 replacing them with constants, while keeping the rest of the structure
-unchanged {[}Pearl\_2010{]}. The probabilistic implications of this
+unchanged \citep{Pearl_2010}. The probabilistic implications of this
 factorization are formalized in Equation~\ref{eq-trunc_factor}, where
@@
 
-Ultimately, backdoor adjustment enables researchers to express the ACE
-as:
+Ultimately, the backdoor adjustment enables researchers to express the
+ACE as:
 
@@
 \begin{aligned}
-\tau &= E[\tau_{i}] \\
-  & = E[X_{a} \:|\:do(X_{b}=1)]- E[X_{a} \:|\:do(X_{b}=2)] \\
-  & = \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=1, X_{p}) P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=2, X_{p}) P(X_{p})
+\tau & = E[X_{a} \:|\:do(X_{b}=1)]- E[X_{a} \:|\:do(X_{b}=2)] \\
+  & = E_{Xp}\left[ E[X_{a} \:|\:do(X_{b}=1), X_{p}]- E[X_{a} \:|\:do(X_{b}=2), X_{p}] \right] \\
+  & = \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=2, X_{p}) \cdot P(X_{p})
 \end{aligned}
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \:|\:do(T_{i}=1)] - E[Y_{i} \:|\:do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
+  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right] \\
+  & = \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=2, X) \cdot P(X)
 \end{aligned}
@@
 empirical data, researchers can update the priors to posterior
-distributions using Bayes' rule \citep{Jeffreys_1998}:
+distributions using Bayes' rule in Equation~\ref{eq-bayes_rule}:
 
@@
 independent of the parameter \(\theta\), researchers can simplify the
-posterior updating process into the three steps depicted by
-Equation~\ref{eq-prop_rule}. First, they integrate new empirical data
-through the likelihood. Second, they update the parameters' priors to a
-posterior distribution according to Equation~\ref{eq-prop_rule}.
-Ultimately, they normalize these results to obtain a valid probability
-distribution.
+posterior updating process into three steps. First, they integrate new
+empirical data through the likelihood. Second, they update the
+parameters' priors to a posterior distribution according to
+Equation~\ref{eq-prop_rule}. Ultimately, they normalize these results to
+obtain a valid probability distribution.
 
@@
 the number of random variables, \(P\), increases, this joint
-distribution becomes quickly intractable \citep{Neal_2020}. This
+distribution quickly becomes intractable \citep{Neal_2020}. This
 intractability is evident from Equation~\ref{eq-like_chain}, where the
@@
 point, any probabilistic programming language can model this simpler
-structure and generate the parameter's posterior distribution from its
-priors using Equation~\ref{eq-bayes_rule}.
+structure and generate the parameter's posterior distribution using
+Equation~\ref{eq-bayes_rule}.
 

### 2025-01-21 paper2.qmd
section:appendix; process:write

@@
 An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand, and they are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010].
+*Estimates* are numerical values that approximate the estimand, and they are derived through *estimation*, referring to the process of integrating data with an estimator [@Everitt_et_al_2010].
 The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.
@@
 <!-- 3. But, what do we need to produce a good estimate? -->
-While numerous methods can approximate an estimand, researchers prioritize estimators with desirable identification properties that ensure the accuracy of estimates.
+While numerous estimators can approximate an estimand, researchers prioritize estimators with desirable identification properties that ensure the accuracy of estimates.
 For instance, the Z-test is an estimator known for its effectiveness when comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006].
@@
 However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply estimate associations.
-In the example, researchers can interpret the associational estimate represented by the Z-statistic as causal.
-This interpretation relies on the data meeting the assumptions of the Z-test and being collected through a randomized experiment.
+In the example, researchers can interpret the associational estimand in @eq-group_diff and the associational estimate represented by the Z-statistic as causal.
+This interpretation is valid as long as the data satisfy the assumptions of the Z-test and are collected through a randomized experiment.
 
@@
 Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable.
-To address this challenge, the structural approach extends the ICE to the *average causal effect* (ACE, @eq-ACE), representing the average difference between observed potential outcomes and their counterfactual counterparts.
+To address this challenge, the structural approach extends the ICE to the *average causal effect* (ACE, @eq-ACE), representing the average difference between the students' observed potential outcomes and their counterfactual counterparts.
 
@@
 <!-- 3. But how does it work? -->
-Even though counterfactuals are unobserved and unobservable, researchers can still identify the ACE from associational estimates by leveraging the structural approach.
+Even though counterfactuals are unobservable, researchers can still identify the ACE from associational estimates by leveraging the structural approach.
 The approach identifies the ACE by statistically conditioning data on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
 This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones.
-If such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
+When such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
 Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
@@
 
-For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as a variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
+For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
 Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)
@@
 
-From the CACE, they identify the ACE from associational quantities as in @eq-mACE1.
+From the CACE, researchers can identify the ACE from associational quantities as in @eq-mACE1.
 This identification process is commonly known as the *backdoor adjustment*.
@@
 The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference by guiding researchers in determining which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020].
+Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference.
+They achieve this by helping researchers to determine which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020].
 @fig-IEflow shows the role of theoretical models in the inference process.
@@
 35].
+
 @sec-appendix-B21 and @sec-appendix-B22 elaborate on the first advantage, while @sec-appendix-B22 and @sec-appendix-B23 do so for the second.
@@
 
-#### SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}
+#### The five fundamental block for SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}
 
@@
 DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-Although the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* depicted in the figures offers some advantages: including exogenous variables can help to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+Although the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* depicted in the figures offers one key advantage: including exogenous variables can help researchers highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 <!-- 4. Theoretical implications of the basic building block of the SCMS and DAGs -->
-A careful examination of the figures highlights the theoretical assumptions underlying the observed variables in these building blocks.
+A careful examination of these building blocks highlights the theoretical assumptions underlying their observed variables.
 SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are independent or not causally related.
@@
 SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example.
-In this context, the variable $X$ (socio-economic status of the school) acts as a confounder in the relationship between the teaching method $T$ and the outcome $Y$.
-The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+In this context, the variable $X$ (socio-economic status of the school) is thought to be a confounder in the relationship between the teaching method $T$ and the outcome $Y$.
+In this scenario, the figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
 They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
@@
 
-#### The probabilistic implications of SCMs and DAGs {#sec-appendix-B22 .appendix appendix-style=plain}
+#### The probabilistic implications of these blocks {#sec-appendix-B22 .appendix appendix-style=plain}
 
@@
 $$
-P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \cond pa(X_{i}) )
+\begin{aligned}
+P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \cond X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
+& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \cond pa(X_{i}) ) & (\small{\text{by BNF}})
+\end{aligned}
 $$ {#eq-net_factor}
@@
 <!-- 4. can you illustrate this concept? -->
-The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which variables' (in)dependencies can SMCs and DAGs predict.
-For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$).
-This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$ [@Neal_2020, pp.
+The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which (in)dependencies can SMCs and DAGs predict.
+For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] enables researchers to express the joint probability distribution of the observed variables as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, supporting the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$) [@Neal_2020, pp.
 24].
@@
 
-Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of variables' (in)dependencies from chains, forks, and colliders.
-These (in)dependencies apply to any data generated by these structures.
+Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of(in)dependencies from chains, forks, and colliders.
+These (in)dependencies apply to any data set generated by a causal model with those structures, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, pp.
+36].
 For instance, applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] allow researchers to represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}) P(X_{3} \cond X_{2})$.
@@
 Moreover, in the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \cond X_{2}) P(X_{2}) P(X_{3} \cond X_{2})$.
-This expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
+Similar to the chain structure, this expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
 Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}, X_{3}) P(X_{3})$.
@@
 
+<!-- These (in)dependencies will be true of every data set generated by a causal model with that graphical structure, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, pp.
+36].
+-->
+
 <!-- 5. Continuing with the motivating example -->
-Then, researchers can revisit the scenario in @sec-appendix-B using these insights.
-In this context, applying the BNF to the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] structures, can help researchers to represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
-From this expression, researchers can infer that the outcome $Y$ is not unconditionally independent of the teaching method $T$ $(Y \dsep T)$.
-This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path).
-
-These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on the variable and depicted by a square in DAG [-@fig-mdag_example2], effectively blocks all confounding non-causal path introduced by the fork structure.
-Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.
+Using these additional probabilistic insights, researchers can revisit the scenario in @sec-appendix-B.
+In this context, applying the BNF to the SCM [-@fig-scm_example2] structure, enables the representation of the joint probability distribution of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
+From this expression, researchers can infer that the outcome $Y$ is unconditionally dependent on the teaching method $T$ $(Y \ndsep T)$.
+This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path in DAG [-@fig-mdag_example2]), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path in DAG [-@fig-mdag_example2]).
 
@@
 <!-- 1. Transforming probabilistic association into causation?  -->
-The structural approach to causal inference translate the probabilistic insights from @sec-appendix-B23 into actionable strategies for identifying the ACE from associational quantities.
-The approach achieve this by relying on the modularity assumption and by re-defining interventions in the form $P(Y_{i} \cond do(T_{i}=t))$ (see @sec-appendix-B).
-The *modularity assumption* posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
+The structural approach to causal inference translate the probabilistic insights from @sec-appendix-B23 into actionable strategies seeking to identify the ACE from associational quantities.
+It achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
 34].
@@
 <!-- 2. additional necessary concepts  -->
-The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions within SCMs and DAGs.
+The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions $P(Y_{i} \cond do(T_{i}=t))$ within SCMs and DAGs.
 *Manipulated graphs* simulate physical interventions by removing specific edges from a DAG, while preserving the remaining structure unchanged [@Neal_2020, pp.
@@
 In parallel, *Truncated Factorization* (TF) achieves a similar simulation by removing specific functions from the conceptual model and replacing them with constants, while keeping the rest of the structure unchanged [@Pearl_2010].
-The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ represents the subset of variables $X_{p}$ directly influenced by the intervention.
+The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ represents the subset of variables $X_{p}$ directly influenced by the intervention, while an example illustrating these concepts follows below.
 
@@
 This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure.
-In this context, researchers can apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$.
-They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE, as expressed in @eq-mACE2 (see @sec-appendix-B).
+Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.
+
+Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$.
+They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE.
+This process is expressed in @eq-mACE2 (see @sec-appendix-B).
 Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.
@@
 
-## The estimation process {#sec-appendix-C .appendix appendix-style=plain}
+#### The estimation process {#sec-appendix-C .appendix appendix-style=plain}
 

### 2025-01-21 _tex/paper2.tex
Built site for gh-pages

@@
 approximate the estimand, and they are derived through
-\emph{estimation}, which refers to the process of integrating data with
-an estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
+\emph{estimation}, referring to the process of integrating data with an
+estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
 flowchart \citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow}
@@
 
-While numerous methods can approximate an estimand, researchers
+While numerous estimators can approximate an estimand, researchers
 prioritize estimators with desirable identification properties that
@@
 estimate associations. In the example, researchers can interpret the
+associational estimand in Equation~\ref{eq-group_diff} and the
 associational estimate represented by the Z-statistic as causal. This
-interpretation relies on the data meeting the assumptions of the Z-test
-and being collected through a randomized experiment.
+interpretation is valid as long as the data satisfy the assumptions of
+the Z-test and are collected through a randomized experiment.
 
@@
 the ICE to the \emph{average causal effect} (ACE,
-Equation~\ref{eq-ACE}), representing the average difference between
-observed potential outcomes and their counterfactual counterparts.
+Equation~\ref{eq-ACE}), representing the average difference between the
+students' observed potential outcomes and their counterfactual
+counterparts.
 
@@
 
-Even though counterfactuals are unobserved and unobservable, researchers
-can still identify the ACE from associational estimates by leveraging
-the structural approach. The approach identifies the ACE by
-statistically conditioning data on a \emph{sufficient adjustment set} of
-variables \(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
-This \emph{sufficient} set (potentially empty) must block all non-causal
-paths between \(T\) to \(Y\) without opening new ones. If such a set
+Even though counterfactuals are unobservable, researchers can still
+identify the ACE from associational estimates by leveraging the
+structural approach. The approach identifies the ACE by statistically
+conditioning data on a \emph{sufficient adjustment set} of variables
+\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
+\emph{sufficient} set (potentially empty) must block all non-causal
+paths between \(T\) to \(Y\) without opening new ones. When such a set
 exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
@@
 described in Section~\ref{sec-appendix-A} and must instead rely on
-observational data, they can still identify the ACE as long as a
-variable \(X\), such as the socio-economic status of the school,
-satisfies the backdoor criterion. Under these circumstances, researchers
-first identify the \emph{conditional average causal effect} (CACE,
-Equation~\ref{eq-CACE})
+observational data, they can still identify the ACE as long as an
+observed variable \(X\), such as the socio-economic status of the
+school, satisfies the backdoor criterion. Under these circumstances,
+researchers first identify the \emph{conditional average causal effect}
+(CACE, Equation~\ref{eq-CACE})
 
@@
 
-From the CACE, they identify the ACE from associational quantities as in
-Equation~\ref{eq-mACE1}. This identification process is commonly known
-as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\) represents the
-marginal expected value over \(X\) \citep{Morgan_et_al_2014}.
+From the CACE, researchers can identify the ACE from associational
+quantities as in Equation~\ref{eq-mACE1}. This identification process is
+commonly known as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\)
+represents the marginal expected value over \(X\)
+\citep{Morgan_et_al_2014}.
 
@@
 Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
-models} that facilitate valid causal inference by guiding researchers in
-determining which statistical models can identify an estimand (ACE,
-CACE, or other), assuming the depicted causal structure is correct
-\citep{McElreath_2020}. Figure~\ref{fig-IEflow} shows the role of
-theoretical models in the inference process.
+models} that facilitate valid causal inference. They achieve this by
+helping researchers to determine which statistical models can identify
+an estimand (ACE, CACE, or other), assuming the depicted causal
+structure is correct \citep{McElreath_2020}. Figure~\ref{fig-IEflow}
+shows the role of theoretical models in the inference process.
 
@@
 their parameters \citep[pp.~35]{Pearl_et_al_2016}.
+
 Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
@@
 
-\paragraph{SCMs and DAGs}\label{sec-appendix-B21}
+\paragraph{The five fundamental block for SCMs and
+DAGs}\label{sec-appendix-B21}
 
@@
 of DAGs typically omits exogenous variables for simplicity, the
-\emph{magnified representation} depicted in the figures offers some
-advantages: including exogenous variables can help to highlight
+\emph{magnified representation} depicted in the figures offers one key
+advantage: including exogenous variables can help researchers highlight
 potential issues related to conditioning and confounding
@@
 
-A careful examination of the figures highlights the theoretical
-assumptions underlying the observed variables in these building blocks.
-SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
+A careful examination of these building blocks highlights the
+theoretical assumptions underlying their observed variables. SCM
+\ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
 nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
@@
 this example. In this context, the variable \(X\) (socio-economic status
-of the school) acts as a confounder in the relationship between the
-teaching method \(T\) and the outcome \(Y\). The figures display
-multiple descendant relationships such as \(X \rightarrow T\),
-\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
-unconnected node pairs, evident from the relationships
+of the school) is thought to be a confounder in the relationship between
+the teaching method \(T\) and the outcome \(Y\). In this scenario, the
+figures display multiple descendant relationships such as
+\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
+also highlight unconnected node pairs, evident from the relationships
 \(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
@@
 
-\paragraph{The probabilistic implications of SCMs and
-DAGs}\label{sec-appendix-B22}
+\paragraph{The probabilistic implications of these
+blocks}\label{sec-appendix-B22}
 
@@
 \begin{equation}\phantomsection\label{eq-net_factor}{
-P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \:|\:pa(X_{i}) )
+\begin{aligned}
+P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
+& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:pa(X_{i}) ) & (\small{\text{by BNF}})
+\end{aligned}
 }\end{equation}
@@
 The five fundamental building blocks described in
-Section~\ref{sec-appendix-B21} clearly illustrate which variables'
-(in)dependencies can SMCs and DAGs predict. For instance, applying the
-BNF to the causal structure shown in the SCM \ref{fig-scm_bb1} and DAG
-\ref{fig-mdag_bb1} supports the theoretical assumption that the observed
-variables \(X_{1}\) and \(X_{3}\) are unconditionally independent
-(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
-joint probability distribution as
-\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
-Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
+Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
+can SMCs and DAGs predict. For instance, applying the BNF to the causal
+structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
+enables researchers to express the joint probability distribution of the
+observed variables as \(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\),
+supporting the theoretical assumption that the observed variables
+\(X_{1}\) and \(X_{3}\) are unconditionally independent
+(\(X_{1} \:\bot\:X_{3}\)) \citep[pp.~24]{Neal_2020}. Conversely, when
+\(X_{3}\) is unconditionally dependent on \(X_{1}\)
 (\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
@@
 Beyond the insights gained from two-node structures, researchers can
-uncover more nuanced patterns of variables' (in)dependencies from
-chains, forks, and colliders. These (in)dependencies apply to any data
-generated by these structures. For instance, applying the BNF to the
-chain structure depicted in the SCM \ref{fig-scm_bb3} and DAG
-\ref{fig-mdag_bb3} allow researchers to represent the joint distribution
-for the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
+uncover more nuanced patterns of(in)dependencies from chains, forks, and
+colliders. These (in)dependencies apply to any data set generated by a
+causal model with those structures, regardless of the specific functions
+attached to the SCM \citep[pp.~36]{Pearl_et_al_2016}. For instance,
+applying the BNF to the chain structure depicted in the SCM
+\ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} allow researchers to
+represent the joint distribution for the observed variables as
+\(P(X_{1},X_{2},X_{3}) =\)
 \(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
@@
 of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
-allows researchers to further infer that \(X_{1}\) and \(X_{3}\) are
-unconditionally dependent \((X_{1} \:\not\bot\:X_{3})\), but
-conditionally independent when controlling for \(X_{2}\)
-\((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally, researchers analyzing the
-collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
-\ref{fig-mdag_bb5} can express the joint distribution of the observed
-variables as \(P(X_{1},X_{2},X_{3}) =\)
+\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). Similar to the
+chain structure, this expression allows researchers to further infer
+that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
+\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
+controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally,
+researchers analyzing the collider structure illustrated in the SCM
+\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} can express the joint
+distribution of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
 \(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
@@
 
-Then, researchers can revisit the scenario in
-Section~\ref{sec-appendix-B} using these insights. In this context,
-applying the BNF to the SCM \ref{fig-scm_example2} and DAG
-\ref{fig-mdag_example2} structures, can help researchers to represent
-the joint probability of the observed variables as \(P(Y, T, X) =\)
-\(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this expression, researchers
-can infer that the outcome \(Y\) is not unconditionally independent of
-the teaching method \(T\) \((Y \:\bot\:T)\). This dependency arises from
-two key structures: a direct causal path from the teaching method \(T\)
-to the outcome \(Y\), represented by the two-connected-nodes structure
-\(T \rightarrow Y\) (black path), and a confounding non-causal path from
-the teaching method \(T\) to the outcome \(Y\) through the
-socio-economic status of the school \(X\), represented by the fork
-structure \(T \leftarrow X \rightarrow Y\) (gray path).
-
-These probabilistic insights also indicate that narrowing the focus to a
-subset of the population defined by the socio-economic status of the
-school \(X\), achieved by conditioning on the variable and depicted by a
-square in DAG \ref{fig-mdag_example2}, effectively blocks all
-confounding non-causal path introduced by the fork structure.
-Nevertheless, since \(Y\) remains dependent on \(T\) even after
-conditioning \((Y \:\not\bot\:T \:|\:X)\), this dependency can only be
-attributed to the direct causal effect \(T \rightarrow Y\).
+Using these additional probabilistic insights, researchers can revisit
+the scenario in Section~\ref{sec-appendix-B}. In this context, applying
+the BNF to the SCM \ref{fig-scm_example2} structure, enables the
+representation of the joint probability distribution of the observed
+variables as \(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From
+this expression, researchers can infer that the outcome \(Y\) is
+unconditionally dependent on the teaching method \(T\)
+\((Y \:\not\bot\:T)\). This dependency arises from two key structures: a
+direct causal path from the teaching method \(T\) to the outcome \(Y\),
+represented by the two-connected-nodes structure \(T \rightarrow Y\)
+(black path in DAG \ref{fig-mdag_example2}), and a confounding
+non-causal path from the teaching method \(T\) to the outcome \(Y\)
+through the socio-economic status of the school \(X\), represented by
+the fork structure \(T \leftarrow X \rightarrow Y\) (gray path in DAG
+\ref{fig-mdag_example2}).
 
@@
 insights from Section~\ref{sec-appendix-B23} into actionable strategies
-for identifying the ACE from associational quantities. The approach
-achieve this by relying on the modularity assumption and by re-defining
-interventions in the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
-Section~\ref{sec-appendix-B}). The \emph{modularity assumption} posits
-that intervening on a node alters only the causal mechanism of that
-node, leaving others unchanged \citep[pp.~34]{Neal_2020}.
+seeking to identify the ACE from associational quantities. It achieves
+this by relying on the \emph{modularity assumption}, which posits that
+intervening on a node alters only the causal mechanism of that node,
+leaving others unchanged \citep[pp.~34]{Neal_2020}.
 
@@
 and Truncated Factorization, which are essential for representing
-interventions within SCMs and DAGs. \emph{Manipulated graphs} simulate
-physical interventions by removing specific edges from a DAG, while
-preserving the remaining structure unchanged \citep[pp.~34]{Neal_2020}.
-In parallel, \emph{Truncated Factorization} (TF) achieves a similar
-simulation by removing specific functions from the conceptual model and
-replacing them with constants, while keeping the rest of the structure
-unchanged \citep{Pearl_2010}. The probabilistic implications of this
-factorization are formalized in Equation~\ref{eq-trunc_factor}, where
-\(S\) represents the subset of variables \(X_{p}\) directly influenced
-by the intervention.
+interventions \(P(Y_{i} \:|\:do(T_{i}=t))\) within SCMs and DAGs.
+\emph{Manipulated graphs} simulate physical interventions by removing
+specific edges from a DAG, while preserving the remaining structure
+unchanged \citep[pp.~34]{Neal_2020}. In parallel, \emph{Truncated
+Factorization} (TF) achieves a similar simulation by removing specific
+functions from the conceptual model and replacing them with constants,
+while keeping the rest of the structure unchanged \citep{Pearl_2010}.
+The probabilistic implications of this factorization are formalized in
+Equation~\ref{eq-trunc_factor}, where \(S\) represents the subset of
+variables \(X_{p}\) directly influenced by the intervention, while an
+example illustrating these concepts follows below.
 
@@
 sufficient adjustment set, as it effectively blocks all confounding
-non-causal paths introduced by the fork structure. In this context,
-researchers can apply the \emph{backdoor adjustment} to identify the ACE
-of \(T\) on \(Y\). They achieve this by first identifying the CACE of
-\(T\) on \(Y\) by conditioning on \(X\), and then marginalizing this
-effect over \(X\) to obtain the ACE, as expressed in
+non-causal paths introduced by the fork structure. Nevertheless, since
+\(Y\) remains dependent on \(T\) even after conditioning
+\((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
+the direct causal effect \(T \rightarrow Y\).
+
+Researchers can then apply the \emph{backdoor adjustment} to identify
+the ACE of \(T\) on \(Y\). They achieve this by first identifying the
+CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
+this effect over \(X\) to obtain the ACE. This process is expressed in
 Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
@@
 
-\subsection{The estimation process}\label{sec-appendix-C}
+\paragraph{The estimation process}\label{sec-appendix-C}
 

### 2025-01-23 paper2.qmd
section:theory; process:write

@@
 
+<!-- commands for d-separation -->
+\newcommand{\dsep}{\:\bot\:}
+\newcommand{\ndsep}{\:\not\bot\:}
+\newcommand{\cond}{\:|\:}
+
 <!-- abstract -->
@@
 In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-Meanwhile, research on the validity of CJ scores indicates their capacity to represent the traits under measurement accurately [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
+Meanwhile, research on the validity of CJ scores indicates their capacity to represent accurately the traits under measurement  [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
 Moreover, research on CJ's practical applicability highlights its versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
@@
 <!-- 3. But what critical issues remain in CJ Research?, what is the organization of the study? -->
-Nevertheless, despite the increasing number of CJ studies, the prevalence of unsystematic and fragmented research approaches has left several critical issues unaddressed.
-The present study primarily focuses on two issues: the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data and the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-The following sections begin with a brief overview of Thurstone's theory followed by a detailed examination of these issues.
-Subsequently, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, alongside its statistical translation, designed to address the two concerns simultaneously.
+Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unresolved.
+This study examines two such problems that can compromise the reliability and validity of CJ's trait estimates.
+The first issue concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+
+The study has five main sections.
+It begins with a brief overview of Thurstone's theory in @sec-thurstone_theory, followed by a detailed examination of these issues in @sec-theory-issues.
+In @sec-theory, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, along with its statistical translation, designed to address both issues simultaneously.
+Finally, @sec-discussion discusses the concepts developed in the study, their implications, and avenues for future exploration, while @sec-conclusion presents the conclusions.
 
@@
 <!-- 2. Example of discriminal processes -->
-@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts.
+@fig-discriminal_process illustrates the hypothetical discriminal processes along a quality trait continuum for two written texts.
 The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
@@
 <!-- 3. Thurstone's Theory of CJ: discriminal difference -->
-However, given that the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
+However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
 This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp.
@@
 
-This section examines the two critical issues in the CJ literature that serve as the primary focus of the present study.
-The first is related to the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second concern with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+This section examines the two critical issues in CJ's theory that can compromise the reliability and validity of CJ's trait estimates.
+In @sec-theory-issue1, the study examines the first issue: the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+In @sec-theory-issue2, the study explores the second issue: the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
 
@@
 According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons.
-@fig-dispersion presents a thought experiment to illustrate this idea.
-In this experiment, a researcher can observe the discriminal processes for the texts depicted in @fig-discriminal_process.
-Furthermore, the figure assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
-The figure reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
+A thought experiment can illustrate this idea.
+In this experiment, the researcher can observe the discriminal processes for the texts depicted in @fig-discriminal_process.
+Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
+In this scenario, @fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
 This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome.
@@
 <!-- 2. But this is not the scenario we observe -->
-In experimental practice, however, this process occurs in reverse.
+In experimental practice, however, the thought experiment occurs in reverse.
 Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp.
@@
 For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
-This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
-The accuracy of these discriminal difference ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.
+This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
+The accuracy of the discriminal difference then ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.
 
@@
 @fig-dispersion also illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information.
-Such exclusions can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
@@
 As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
-@fig-correlation presents a similar thought experiment as in @sec-theory-issue1a to illustrate this idea.
-The illustration now assumes that the discriminal dispersions for both texts remain constant.
-The figure reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
+A similar thought experiment, as the one in @sec-theory-issue1a, can illustrate this idea.
+The experiment now assumes that the discriminal dispersions for both texts remain constant.
+@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
 This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
@@
 <!-- 2. But does it hold?  -->
-Notably, Thurstone attributed the independence of stimuli to the cancellation of potential judges' biases.
+Notably, Thurstone attributed the lack of correlation (independence) between stimuli to the cancellation of potential judges' biases.
 He argued that this cancellation resulted from two opposing and equally weighted effects occurring during pairwise comparisons [@Thurstone_1927b, pp.
@@
 <!-- 4. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010].
-Despite the CJ literature acknowledging the existence of such hierarchical structures, the statistical handling of this additional source of dependence between stimuli has been inadequate.
+In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, creating a statistical phenomenon known as clustering [@Everitt_et_al_2010].
+Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical handling of this additional source of dependence between stimuli has been inadequate.
 For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
+This approach, however, can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
 
@@
 
-The misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures.
+The model misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures.
 Neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp.
@@
 12].
-Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially amplifying the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp.
+Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially overestimating the trait's reliability [@Hoyle_et_al_2023, pp.
 482].
@@
 2].
-Consequently, adopting a more systematic and integrated approach to handling these assumptions and examining the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address these issues.
+Consequently, adopting a more systematic and integrated approach to handling these assumptions and examining the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address the aforementioned issues.
 
@@
 
-This section uses the structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] to articulate a theoretical model that captures the core principles of Thurstone's theory.
-The model also incorporates various assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-Finally, the section employs Bayesian inference methods to transform these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data.
-See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section.
+The objective of this section is twofold.
+First, it aims to refine CJ's theoretical model using the *structural approach* to causal inference.
+This approach, as outlined in @sec-theory-theoretical, defines a model that integrates the core principles of Thurstone's theory and key assessment design features relevant to CJ experiments.
+Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data.
+
+<!-- See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section. -->
 
@@
 
-The theoretical model uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of the CJ system.
-First, *the population model* is created to represent a conceptual population of CJ experiments.
-The model then integrates various assessment design features relevant to CJ experiments, leading to the development of *the sample-comparison model*.
+<!-- 1. The structural approach to causal inference -->
+The *structural approach* to causal inference provides a formal framework to identify causes and estimate their effects using data.
+The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system.
+Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference.
+They achieve this by helping researchers to determine which statistical models can identify an estimand, assuming the depicted causal structure is correct [@McElreath_2020].
+*Identification* determines whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+4].
+
+<!-- 2. The SCMs -->
+SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020].
+Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
+In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
+27,68].
+Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
+These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+
+<!-- 3. The DAGs -->
+Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
+A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables.
+The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence.
+The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
+DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
+While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+
+<!-- 4. The benefits of SCMs and DAGs -->
+SCMs and DAGs support causal inference through two key advantages.
+First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
+Second, they depict causal relationships in a non-parametric and fully interactive way.
+This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp.
+35].
+
+<!-- 5. what are we doing with them? -->
+Therefore, this section leverages SCMs and DAGs to represent the conceptual population of CJ experiments, referred to as *the population model* in @sec-theory-theoretical_P, and later integrates various assessment design features relevant to CJ experiments.
+This process leads to the development of *the sample-comparison model* in @sec-theory-theoretical_SC.
 
@@
 <!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->
-
-In this section we focus on the identification of several parameters of interest from the theoretical model, together with various assessment design features relevant to CJ experiments.
-The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
-So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
-5]
 
@@
 
-# Discussion {#sec-discuss}
-
-## Findings {#sec-discuss-finding}
+# Discussion {#sec-discussion}
+
+## Findings {#sec-discussion-finding}
 
@@
 
-## Limitations and further research {#sec-discuss-limitations}
+## Limitations and further research {#sec-discussion-limitations}
 
@@
 
-<!-- commands for d-separation -->
-\newcommand{\dsep}{\:\bot\:}
-\newcommand{\ndsep}{\:\not\bot\:}
-\newcommand{\cond}{\:|\:}
-
 ## Empirical research and randomized experiments {#sec-appendix-A .appendix appendix-style=plain}
@@
 
+<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+5].
+-->
+
 <!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
@@
 4].
+-->
+
+<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+5].
 -->

### 2025-01-23 _tex/paper2.tex
Built site for gh-pages

@@
 
+\newcommand{\dsep}{\:\bot\:}
+\newcommand{\ndsep}{\:\not\bot\:}
+\newcommand{\cond}{\:|\:}
+
 \section{Introduction}\label{sec-introduction}
@@
 Meanwhile, research on the validity of CJ scores indicates their
-capacity to represent the traits under measurement accurately
+capacity to represent accurately the traits under measurement
 \citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
@@
 
-Nevertheless, despite the increasing number of CJ studies, the
-prevalence of unsystematic and fragmented research approaches has left
-several critical issues unaddressed. The present study primarily focuses
-on two issues: the overreliance on Thurstone's Case V assumptions in the
-statistical analysis of CJ data and the apparent disconnect between CJ's
-approach to trait measurement and hypothesis testing. The following
-sections begin with a brief overview of Thurstone's theory followed by a
-detailed examination of these issues. Subsequently, the study introduces
-a theoretical model for CJ that builds upon Thurstone's theory,
-alongside its statistical translation, designed to address the two
-concerns simultaneously.
+Nevertheless, despite the growing number of CJ studies, unsystematic and
+fragmented research approaches have left several critical issues
+unresolved. This study examines two such problems that can compromise
+the reliability and validity of CJ's trait estimates. The first issue
+concerns the overreliance on Thurstone's Case V assumptions in the
+statistical analysis of CJ data. The second involves the apparent
+disconnect between CJ's approach to trait measurement and hypothesis
+testing.
+
+The study has five main sections. It begins with a brief overview of
+Thurstone's theory in Section~\ref{sec-thurstone_theory}, followed by a
+detailed examination of these issues in Section~\ref{sec-theory-issues}.
+In Section~\ref{sec-theory}, the study introduces a theoretical model
+for CJ that builds upon Thurstone's theory, along with its statistical
+translation, designed to address both issues simultaneously. Finally,
+Section~\ref{sec-discussion} discusses the concepts developed in the
+study, their implications, and avenues for future exploration, while
+Section~\ref{sec-conclusion} presents the conclusions.
 
@@
 
-Figure~\ref{fig-discriminal_process} illustrates hypothetical
+Figure~\ref{fig-discriminal_process} illustrates the hypothetical
 discriminal processes along a quality trait continuum for two written
@@
 
-However, given that the individual discriminal processes of the stimuli
-are not directly observable, the theory introduces the \emph{law of
+However, since the individual discriminal processes of the stimuli are
+not directly observable, the theory introduces the \emph{law of
 comparative judgment}. This law posits that in pairwise comparisons, a
@@
 
-This section examines the two critical issues in the CJ literature that
-serve as the primary focus of the present study. The first is related to
-the overreliance on Thurstone's Case V assumptions in the statistical
-analysis of CJ data. The second concern with the apparent disconnect
-between CJ's approach to trait measurement and hypothesis testing.
+This section examines the two critical issues in CJ's theory that can
+compromise the reliability and validity of CJ's trait estimates. In
+Section~\ref{sec-theory-issue1}, the study examines the first issue: the
+overreliance on Thurstone's Case V assumptions in the statistical
+analysis of CJ data. In Section~\ref{sec-theory-issue2}, the study
+explores the second issue: the apparent disconnect between CJ's approach
+to trait measurement and hypothesis testing.
 
@@
 stimuli shape the distribution of the discriminal difference, exerting a
-direct influence on the outcome of pairwise comparisons.
-Figure~\ref{fig-dispersion} presents a thought experiment to illustrate
-this idea. In this experiment, a researcher can observe the discriminal
-processes for the texts depicted in
-Figure~\ref{fig-discriminal_process}. Furthermore, the figure assumes
-that the discriminal dispersion for Text A remains constant and that the
-texts are uncorrelated \((\rho=0)\). The figure reveals that an increase
-in the uncertainty associated with the perception of Text B in
-comparison to Text A, \((\sigma_{B}-\sigma_{A})\), broadens the
-distribution of their discriminal difference. This broadening affects
-the probability area where the discriminal difference distinctly favors
-Text B over Text A, expressed as \(P(B > A)\), ultimately influencing
-the comparison outcome. Additionally, the figure reveals that when the
-discriminal dispersions of the texts are equal
-\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
-to favor Text B over Text A (shaded gray area), compared to situations
-where their dispersions differ.
+direct influence on the outcome of pairwise comparisons. A thought
+experiment can illustrate this idea. In this experiment, the researcher
+can observe the discriminal processes for the texts depicted in
+Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
+assumes that the discriminal dispersion for Text A remains constant and
+that the texts are uncorrelated \((\rho=0)\). In this scenario,
+Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
+associated with the perception of Text B in comparison to Text A,
+\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
+discriminal difference. This broadening affects the probability area
+where the discriminal difference distinctly favors Text B over Text A,
+expressed as \(P(B > A)\), ultimately influencing the comparison
+outcome. Additionally, the figure reveals that when the discriminal
+dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
+discriminal difference is more likely to favor Text B over Text A
+(shaded gray area), compared to situations where their dispersions
+differ.
 
@@
 
-In experimental practice, however, this process occurs in reverse.
-Researchers first observe the comparison outcome and then use the BTL
-model to infer the discriminal difference between the stimuli and their
-respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
-Therefore, the outcome's ability to reflect the ``true'' differences
-between stimuli largely depends on the validity of the model's
-assumptions \citep[pp.~150]{Kohler_et_al_2019}, particularly the
-assumption of equal dispersions. For instance, when researchers observe
-a sample of outcomes favoring Text B over Text A and correctly assume
-equal dispersions between the texts, the BTL model estimates a
-discriminal difference distribution that accurately represents the
-``true'' discriminal difference of the texts. This scenario is
-illustrated in Figure~\ref{fig-dispersion}, where the model's
+In experimental practice, however, the thought experiment occurs in
+reverse. Researchers first observe the comparison outcome and then use
+the BTL model to infer the discriminal difference between the stimuli
+and their respective discriminal processes
+\citep[pp.~373]{Thurstone_1927a}. Therefore, the outcome's ability to
+reflect the ``true'' differences between stimuli largely depends on the
+validity of the model's assumptions \citep[pp.~150]{Kohler_et_al_2019},
+particularly the assumption of equal dispersions. For instance, when
+researchers observe a sample of outcomes favoring Text B over Text A and
+correctly assume equal dispersions between the texts, the BTL model
+estimates a discriminal difference distribution that accurately
+represents the ``true'' discriminal difference of the texts. This
+scenario is illustrated in Figure~\ref{fig-dispersion}, when the model's
 discriminal difference distribution aligns with the ``true''
 distribution, represented by the thick continuous line corresponding to
-\(\sigma_{B}-\sigma_{A}=0\). The accuracy of these discriminal
-difference ensures reliable estimates for the texts' discriminal
-processes {(citation needed?)}.
+\(\sigma_{B}-\sigma_{A}=0\). The accuracy of the discriminal difference
+then ensures reliable estimates for the texts' discriminal processes
+{(citation needed?)}.
 
@@
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
-can unintentionally discard valuable information. Such exclusions can
-introduce bias into trait estimates
-\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
-magnitude of these biases are often unpredictable, as they depend on
-which stimuli are excluded from the analysis.
+can unintentionally discard valuable information, introducing bias into
+trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
+direction and magnitude of these biases are often unpredictable, as they
+depend on which stimuli are excluded from the analysis.
 
@@
 dispersions, this correlation shapes the distribution of the discriminal
-difference, directly impacting the outcomes of pairwise comparisons.
-Figure~\ref{fig-correlation} presents a similar thought experiment as in
-Section~\ref{sec-theory-issue1a} to illustrate this idea. The
-illustration now assumes that the discriminal dispersions for both texts
-remain constant. The figure reveals that as the correlation between the
-texts increases, the distribution of their discriminal difference
-becomes narrower. This narrowing affects the area under the curve where
-the discriminal difference distinctly favors Text B over Text A, denoted
-as \(P(B > A)\), thus influencing the comparison outcome. Furthermore,
-the figure shows that when two texts are independent or uncorrelated
-\((\rho=0)\), their discriminal difference is less likely to favor Text
-B over Text A (shaded gray area) compared to scenarios where the texts
-are highly correlated.
+difference, directly impacting the outcomes of pairwise comparisons. A
+similar thought experiment, as the one in
+Section~\ref{sec-theory-issue1a}, can illustrate this idea. The
+experiment now assumes that the discriminal dispersions for both texts
+remain constant. Figure~\ref{fig-correlation} reveals that as the
+correlation between the texts increases, the distribution of their
+discriminal difference becomes narrower. This narrowing affects the area
+under the curve where the discriminal difference distinctly favors Text
+B over Text A, denoted as \(P(B > A)\), thus influencing the comparison
+outcome. Furthermore, the figure shows that when two texts are
+independent or uncorrelated \((\rho=0)\), their discriminal difference
+is less likely to favor Text B over Text A (shaded gray area) compared
+to scenarios where the texts are highly correlated.
 
@@
 
-Notably, Thurstone attributed the independence of stimuli to the
-cancellation of potential judges' biases. He argued that this
-cancellation resulted from two opposing and equally weighted effects
-occurring during pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
-\citet{Andrich_1978} provided a mathematical demonstration of this
-cancellation using the BTL model under the assumption of discriminal
-processes with additive biases. However, it is easy to imagine at least
-two scenarios in which the zero correlation assumption is almost
-certainly invalid: when the pairwise comparison involves
-multidimensional, complex traits with heterogeneous stimuli and when an
-additional hierarchical structure is relevant to the stimuli.
+Notably, Thurstone attributed the lack of correlation (independence)
+between stimuli to the cancellation of potential judges' biases. He
+argued that this cancellation resulted from two opposing and equally
+weighted effects occurring during pairwise comparisons
+\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
+mathematical demonstration of this cancellation using the BTL model
+under the assumption of discriminal processes with additive biases.
+However, it is easy to imagine at least two scenarios in which the zero
+correlation assumption is almost certainly invalid: when the pairwise
+comparison involves multidimensional, complex traits with heterogeneous
+stimuli and when an additional hierarchical structure is relevant to the
+stimuli.
 
@@
 created by additional hierarchical structures may further introduce
-dependencies between stimuli, a statistical phenomenon commonly known as
-clustering \citep{Everitt_et_al_2010}. Despite the CJ literature
-acknowledging the existence of such hierarchical structures, the
+dependencies between stimuli, creating a statistical phenomenon known as
+clustering \citep{Everitt_et_al_2010}. Although the CJ literature
+acknowledges the existence of such hierarchical structures, the
 statistical handling of this additional source of dependence between
@@
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-However, this approach can introduce additional statistical and
+This approach, however, can introduce additional statistical and
 measurement issues, which we discuss in greater detail in
@@
 
-The misspecification may arise from neglecting additional relevant
+The model misspecification may arise from neglecting additional relevant
 traits, excluding judges based on misfit statistics, or ignoring
@@
 hierarchical structures may reduce the precision of model parameter
-estimates, potentially amplifying the overestimation of the trait's
-reliability \citep[pp.~482]{Hoyle_et_al_2023}.
+estimates, potentially overestimating the trait's reliability
+\citep[pp.~482]{Hoyle_et_al_2023}.
 
@@
 statistical and measurement benefits, including the ability to address
-these issues.
+the aforementioned issues.
 
@@
 
-This section uses the structural approach to causal inference
-\citep{Pearl_2009, Pearl_et_al_2016} to articulate a theoretical model
-that captures the core principles of Thurstone's theory. The model also
-incorporates various assessment design features relevant to CJ
-experiments, such as the selection of judges, stimuli, and comparisons.
-Finally, the section employs Bayesian inference methods to transform
-these theoretical and practical elements into a statistical model that
-facilitates the analysis of pairwise comparison data. See
-Section~\ref{sec-appendix} for an overview of the statistical and causal
-inference concepts required for the development of this section.
+The objective of this section is twofold. First, it aims to refine CJ's
+theoretical model using the \emph{structural approach} to causal
+inference. This approach, as outlined in
+Section~\ref{sec-theory-theoretical}, defines a model that integrates
+the core principles of Thurstone's theory and key assessment design
+features relevant to CJ experiments. Second,
+Section~\ref{sec-theory-statistics} uses \emph{Bayesian inference
+methods} to convert these theoretical and practical elements into a
+statistical model that supports the analysis of paired comparison data.
 
@@
 
-The theoretical model uses structural causal models (SCMs) and directed
-acyclic graphs (DAGs)
+The \emph{structural approach} to causal inference provides a formal
+framework to identify causes and estimate their effects using data. The
+approach uses structural causal models (SCMs) and directed acyclic
+graphs (DAGs)
 \citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
-formally and graphically represent the assumed causal structure of the
-CJ system. First, \emph{the population model} is created to represent a
-conceptual population of CJ experiments. The model then integrates
-various assessment design features relevant to CJ experiments, leading
-to the development of \emph{the sample-comparison model}.
+formally and graphically represent the assumed causal structure of a
+system. Essentially, SCMs and DAGs serve as \emph{conceptual
+(theoretical) models} that facilitate valid causal inference. They
+achieve this by helping researchers to determine which statistical
+models can identify an estimand, assuming the depicted causal structure
+is correct \citep{McElreath_2020}. \emph{Identification} determines
+whether an estimator can accurately compute an estimand based on its
+assumptions, regardless of random variability
+\citep[pp.~4]{Schuessler_et_al_2023}.
+
+SCMs are formal mathematical models defined by a set of
+\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
+set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
+variables are those whose causal mechanisms a researcher chooses to
+model \citep{Neal_2020}. In contrast, exogenous variables represent
+\emph{errors} or \emph{disturbances} arising from omitted factors that
+the investigator chooses not to model explicitly
+\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
+\emph{structural equations}, express the endogenous variables as
+non-parametric functions of other variables. These functions use the
+symbol `\(:=\)' to denote the asymmetrical causal dependence between
+variables and the symbol `\(\:\bot\:\)' to represent
+\emph{d-separation}, a concept akin to (conditional) independence.
+
+Notably, every SCM has an associated DAG
+\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
+consisting of nodes connected by edges, where the nodes represent random
+variables. The term \emph{directed} means that the edges extend from one
+node to another, with arrows indicating the direction of causal
+influence. The term \emph{acyclic} implies that the causal influences do
+not form loops, ensuring the influences do not cycle back on themselves
+\citep{McElreath_2020}. DAGs represent observed variables as solid black
+circles, while they use open circles for unobserved (latent) variables
+\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
+DAGs typically omits exogenous variables for simplicity, the
+\emph{magnified representation} includes them, offering offering the
+advantage of highlighting potential issues related to conditioning and
+confounding \citep{Cinelli_et_al_2020}.
+
+SCMs and DAGs support causal inference through two key advantages.
+First, regardless of complexity, they can represent various causal
+structures using only five fundamental building blocks
+\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
+decompose complex structures into manageable components, facilitating
+their analysis \citep{McElreath_2020}. Second, they depict causal
+relationships in a non-parametric and fully interactive way. This
+flexibility enables feasible ACE identification strategies without
+defining the variables' data types, the functional form between them, or
+their parameters \citep[pp.~35]{Pearl_et_al_2016}.
+
+Therefore, this section leverages SCMs and DAGs to represent the
+conceptual population of CJ experiments, referred to as \emph{the
+population model} in Section~\ref{sec-theory-theoretical_P}, and later
+integrates various assessment design features relevant to CJ
+experiments. This process leads to the development of \emph{the
+sample-comparison model} in Section~\ref{sec-theory-theoretical_SC}.
 
 \subsubsection{The population model}\label{sec-theory-theoretical_P}
-
-In this section we focus on the identification of several parameters of
-interest from the theoretical model, together with various assessment
-design features relevant to CJ experiments. The main reason to focus on
-identification is that if these parameters cannot be identified using
-``infinite'' and error free data, we certainly cannot learn anything
-about it using finite data. So, for the time being, we assume large
-samples and error-free measurements \citep[pp.~5]{Schuessler_et_al_2023}
 
@@
 
-\section{Discussion}\label{sec-discuss}
-
-\subsection{Findings}\label{sec-discuss-finding}
+\section{Discussion}\label{sec-discussion}
+
+\subsection{Findings}\label{sec-discussion-finding}
 
 \subsection{Limitations and further
-research}\label{sec-discuss-limitations}
+research}\label{sec-discussion-limitations}
 
@@
 and \citet{Hernan_et_al_2020}.
-
-\newcommand{\dsep}{\:\bot\:}
-\newcommand{\ndsep}{\:\not\bot\:}
-\newcommand{\cond}{\:|\:}
 

### 2025-01-24 paper2.qmd
section:theory; process:write

@@
 
-<!-- 3. But what critical issues remain in CJ Research?, what is the organization of the study? -->
+<!-- 3. But what critical issues remain in CJ Research? -->
 Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unresolved.
-This study examines two such problems that can compromise the reliability and validity of CJ's trait estimates.
-The first issue concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-
-The study has five main sections.
-It begins with a brief overview of Thurstone's theory in @sec-thurstone_theory, followed by a detailed examination of these issues in @sec-theory-issues.
-In @sec-theory, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, along with its statistical translation, designed to address both issues simultaneously.
-Finally, @sec-discussion discusses the concepts developed in the study, their implications, and avenues for future exploration, while @sec-conclusion presents the conclusions.
+This study addresses two such issues that can compromise the reliability and validity of CJ's trait estimates.
+The first concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+The second focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+
+<!-- 4. what is the organization of the study? -->
+The study is organized into six main sections.
+It begins with an overview of Thurstone's theory in @sec-thurstone_theory, followed by a detailed examination of these issues in @sec-theory-issues.
+In @sec-theoretical, the study develops a theoretical model for CJ that simultaneously addresses these issues.
+The model integrates Thurstone's core principles with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+@sec-statistical translates these theoretical and practical elements into a statistical model that supports the analysis of pairwise comparison data.
+Finally, @sec-discussion discusses the concepts developed in the study, their implications, and avenues for future exploration, while @sec-conclusion summarizes the findings.
 
@@
 
-This section examines the two critical issues in CJ's theory that can compromise the reliability and validity of CJ's trait estimates.
-In @sec-theory-issue1, the study examines the first issue: the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-In @sec-theory-issue2, the study explores the second issue: the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+This section explores the two critical issues in CJ's theory that have the potential to compromise the reliability and validity of CJ's trait estimates.
+@sec-theory-issue1 inspects the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+@sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
 
@@
 
-# Updating the theoretical and statistical model {#sec-theory}
-
-The objective of this section is twofold.
-First, it aims to refine CJ's theoretical model using the *structural approach* to causal inference.
-This approach, as outlined in @sec-theory-theoretical, defines a model that integrates the core principles of Thurstone's theory and key assessment design features relevant to CJ experiments.
-Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data.
+# Updating the theoretical model {#sec-theoretical}
+
+<!-- This study examines two such problems that can compromise the reliability and validity of CJ's trait estimates. -->
+
+<!-- This section aims to refine CJ's theoretical model using the *structural approach* to causal inference with the purpose of addressing the two.  -->
+
+<!-- This approach, as outlined in @sec-theory-theoretical, defines a model that integrates the core principles of Thurstone's theory, as well as other key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and the number of comparisons.
+-->
 
@@
 
-## The theoretical model {#sec-theory-theoretical}
-
 <!-- 1. The structural approach to causal inference -->
-The *structural approach* to causal inference provides a formal framework to identify causes and estimate their effects using data.
+<!-- The *structural approach* to causal inference provides a formal framework to identify causes and estimate their effects using data.
 The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system.
@@
 4].
+-->
 
 <!-- 2. The SCMs -->
-SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020].
+<!-- SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+-->
 
 <!-- 3. The DAGs -->
-Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
+<!-- Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
 A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables.
@@
 While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+-->
 
 <!-- 4. The benefits of SCMs and DAGs -->
-SCMs and DAGs support causal inference through two key advantages.
+<!-- SCMs and DAGs support causal inference through two key advantages.
 First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
@@
 35].
+-->
 
 <!-- 5. what are we doing with them? -->
-Therefore, this section leverages SCMs and DAGs to represent the conceptual population of CJ experiments, referred to as *the population model* in @sec-theory-theoretical_P, and later integrates various assessment design features relevant to CJ experiments.
+<!-- Therefore, this section leverages SCMs and DAGs to represent the conceptual population of CJ experiments, referred to as *the population model* in @sec-theory-theoretical_P, and later integrates various assessment design features relevant to CJ experiments.
 This process leads to the development of *the sample-comparison model* in @sec-theory-theoretical_SC.
-
-### The population model {#sec-theory-theoretical_P}
+-->
+
+## The population model {#sec-theory-theoretical_P}
 
@@
 
-<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
-Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
-4].
--->
-
 Assuming population data or more commonly known as census data, we ...
@@
 
-![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%}
-
-![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%}
-
-![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%}
+<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->
+
+<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->
+
+<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->
 
@@
 
-![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%}
-
-![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%}
-
-![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%}
-
-![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%}
+<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->
+
+<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->
+
+<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->
+
+<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->
 
@@
 
-### The sample-comparison model {#sec-theory-theoretical_SC}
+## The sample-comparison model {#sec-theory-theoretical_SC}
 
@@
 
-![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%}
+<!-- ![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%} -->
 
@@
 
-![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%}
+<!-- ![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%} -->
 
@@
 
-## From theory to statistics {#sec-theory-statistics}
+# From theory to statistics {#sec-statistical}
+
+<!-- Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data.
+-->
 
@@
 <!-- 2. What are the benefits of Bayesian inference?-->
-Bayesian inference procedures offer three key advantages.
+<!-- Bayesian inference procedures offer three key advantages.
 First, they are well-suited to handling complex and overparameterized models, enabling researchers to estimate models where the number of parameters exceeds the number of observations for estimation [@Baker_1998; @Kim_et_al_1999].
@@
 Finally, Bayesian methods are particularly effective at drawing inferences from small sample sizes, where relying on the asymptotic properties of frequentist approaches may not be justified [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
+-->
 
@@
 <!-- 2. What do they do next? -->
-Researchers would then proceed to identify the estimands.
-*Identification* determines whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+Ideally, researchers then proceed to identify the estimands.
+*Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
 4].
-An *estimator* is a method or function that transforms data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand, and they are derived through *estimation*, referring to the process of integrating data with an estimator [@Everitt_et_al_2010].
-The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.
+Identification is crucial because if an estimator cannot reliably compute an estimand using "infinite" and error-free data, it is impossible to gain meaningful insights about that estimand using finite data [@Schuessler_et_al_2023, pp.
+5].
+An *estimator* refers to a method or function that transforms data into an estimate [@Neal_2020].
+*Estimates* are numerical values that approximate the estimand derived through the process of *estimation*, which integrates data with an estimator [@Everitt_et_al_2010].
+The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], shown in @fig-IEflow, visually represents the transition from estimands to estimates.
 
@@
 
-<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
-So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
-5].
--->
-
 ![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}
@@
 <!-- 3. But, what do we need to produce a good estimate? -->
-While numerous estimators can approximate an estimand, researchers prioritize estimators with desirable identification properties that ensure the accuracy of estimates.
-For instance, the Z-test is an estimator known for its effectiveness when comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006].
-If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
+Consequently, although many estimators can approximate an estimand, researchers prioritize those with desirable identification properties to ensure accurate and reliable estimates.
+For instance, the Z-test is a widely used estimator for comparing group proportions, providing accurate estimates when its underlying assumptions are satisfied [@Kanji_2006].
+Moreover, researchers can interpret the Z-test estimates as causal, provided they collect the data through a randomized experiment.
+
+<!-- If both assumptions are met, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
 The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \cond T_{i}=1]$ and $E[Y_{i} \cond T_{i}=2]$, respectively.
 The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.
+-->
 
@@
 
-<!-- 4. But we want these estimands to be causal -->
-However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply estimate associations.
-In the example, researchers can interpret the associational estimand in @eq-group_diff and the associational estimate represented by the Z-statistic as causal.
-This interpretation is valid as long as the data satisfy the assumptions of the Z-test and are collected through a randomized experiment.
-
+<!-- However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply estimate associations. -->
+
+<!-- 4. Why does causality stem from randomize experiments? -->
 Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014].
 This recognition stems from their ability to enable researchers interpret associational estimates as causal.
-They achieve this by ensuring data, and by extension an estimator, satisfies several key properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
+They achieve this by ensuring data, and by extension an estimator, satisfies several key identification properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
 The most critical property, however, is the elimination of confounding.
@@
 The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference.
-They achieve this by helping researchers to determine which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020].
+Essentially, these tools serve as *conceptual (theoretical) models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
+4].
+Specifically, they guide researchers in determining which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020], thus enabling valid causal inference.
 @fig-IEflow shows the role of theoretical models in the inference process.
+
+<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+4].
+-->
 

### 2025-01-24 _tex/paper2.tex
Built site for gh-pages

@@
 fragmented research approaches have left several critical issues
-unresolved. This study examines two such problems that can compromise
-the reliability and validity of CJ's trait estimates. The first issue
-concerns the overreliance on Thurstone's Case V assumptions in the
-statistical analysis of CJ data. The second involves the apparent
-disconnect between CJ's approach to trait measurement and hypothesis
-testing.
-
-The study has five main sections. It begins with a brief overview of
-Thurstone's theory in Section~\ref{sec-thurstone_theory}, followed by a
-detailed examination of these issues in Section~\ref{sec-theory-issues}.
-In Section~\ref{sec-theory}, the study introduces a theoretical model
-for CJ that builds upon Thurstone's theory, along with its statistical
-translation, designed to address both issues simultaneously. Finally,
-Section~\ref{sec-discussion} discusses the concepts developed in the
-study, their implications, and avenues for future exploration, while
-Section~\ref{sec-conclusion} presents the conclusions.
+unresolved. This study addresses two such issues that can compromise the
+reliability and validity of CJ's trait estimates. The first concerns the
+overreliance on Thurstone's Case V assumptions in the statistical
+analysis of CJ data. The second focuses on the apparent disconnect
+between CJ's approach to trait measurement and hypothesis testing.
+
+The study is organized into six main sections. It begins with an
+overview of Thurstone's theory in Section~\ref{sec-thurstone_theory},
+followed by a detailed examination of these issues in
+Section~\ref{sec-theory-issues}. In Section~\ref{sec-theoretical}, the
+study develops a theoretical model for CJ that simultaneously addresses
+these issues. The model integrates Thurstone's core principles with key
+assessment design features relevant to CJ experiments, such as the
+selection of judges, stimuli, and comparisons.
+Section~\ref{sec-statistical} translates these theoretical and practical
+elements into a statistical model that supports the analysis of pairwise
+comparison data. Finally, Section~\ref{sec-discussion} discusses the
+concepts developed in the study, their implications, and avenues for
+future exploration, while Section~\ref{sec-conclusion} summarizes the
+findings.
 
@@
 
-This section examines the two critical issues in CJ's theory that can
-compromise the reliability and validity of CJ's trait estimates. In
-Section~\ref{sec-theory-issue1}, the study examines the first issue: the
-overreliance on Thurstone's Case V assumptions in the statistical
-analysis of CJ data. In Section~\ref{sec-theory-issue2}, the study
-explores the second issue: the apparent disconnect between CJ's approach
-to trait measurement and hypothesis testing.
+This section explores the two critical issues in CJ's theory that have
+the potential to compromise the reliability and validity of CJ's trait
+estimates. Section~\ref{sec-theory-issue1} inspects the overreliance on
+Thurstone's Case V assumptions in the statistical analysis of CJ data.
+Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
+between CJ's approach to trait measurement and hypothesis testing.
 
@@
 
-\section{Updating the theoretical and statistical
-model}\label{sec-theory}
-
-The objective of this section is twofold. First, it aims to refine CJ's
-theoretical model using the \emph{structural approach} to causal
-inference. This approach, as outlined in
-Section~\ref{sec-theory-theoretical}, defines a model that integrates
-the core principles of Thurstone's theory and key assessment design
-features relevant to CJ experiments. Second,
-Section~\ref{sec-theory-statistics} uses \emph{Bayesian inference
-methods} to convert these theoretical and practical elements into a
-statistical model that supports the analysis of paired comparison data.
-
-\subsection{The theoretical model}\label{sec-theory-theoretical}
-
-The \emph{structural approach} to causal inference provides a formal
-framework to identify causes and estimate their effects using data. The
-approach uses structural causal models (SCMs) and directed acyclic
-graphs (DAGs)
-\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
-formally and graphically represent the assumed causal structure of a
-system. Essentially, SCMs and DAGs serve as \emph{conceptual
-(theoretical) models} that facilitate valid causal inference. They
-achieve this by helping researchers to determine which statistical
-models can identify an estimand, assuming the depicted causal structure
-is correct \citep{McElreath_2020}. \emph{Identification} determines
-whether an estimator can accurately compute an estimand based on its
-assumptions, regardless of random variability
-\citep[pp.~4]{Schuessler_et_al_2023}.
-
-SCMs are formal mathematical models defined by a set of
-\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
-set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
-variables are those whose causal mechanisms a researcher chooses to
-model \citep{Neal_2020}. In contrast, exogenous variables represent
-\emph{errors} or \emph{disturbances} arising from omitted factors that
-the investigator chooses not to model explicitly
-\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
-\emph{structural equations}, express the endogenous variables as
-non-parametric functions of other variables. These functions use the
-symbol `\(:=\)' to denote the asymmetrical causal dependence between
-variables and the symbol `\(\:\bot\:\)' to represent
-\emph{d-separation}, a concept akin to (conditional) independence.
-
-Notably, every SCM has an associated DAG
-\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
-consisting of nodes connected by edges, where the nodes represent random
-variables. The term \emph{directed} means that the edges extend from one
-node to another, with arrows indicating the direction of causal
-influence. The term \emph{acyclic} implies that the causal influences do
-not form loops, ensuring the influences do not cycle back on themselves
-\citep{McElreath_2020}. DAGs represent observed variables as solid black
-circles, while they use open circles for unobserved (latent) variables
-\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
-DAGs typically omits exogenous variables for simplicity, the
-\emph{magnified representation} includes them, offering offering the
-advantage of highlighting potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}.
-
-SCMs and DAGs support causal inference through two key advantages.
-First, regardless of complexity, they can represent various causal
-structures using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
-decompose complex structures into manageable components, facilitating
-their analysis \citep{McElreath_2020}. Second, they depict causal
-relationships in a non-parametric and fully interactive way. This
-flexibility enables feasible ACE identification strategies without
-defining the variables' data types, the functional form between them, or
-their parameters \citep[pp.~35]{Pearl_et_al_2016}.
-
-Therefore, this section leverages SCMs and DAGs to represent the
-conceptual population of CJ experiments, referred to as \emph{the
-population model} in Section~\ref{sec-theory-theoretical_P}, and later
-integrates various assessment design features relevant to CJ
-experiments. This process leads to the development of \emph{the
-sample-comparison model} in Section~\ref{sec-theory-theoretical_SC}.
-
-\subsubsection{The population model}\label{sec-theory-theoretical_P}
+\section{Updating the theoretical model}\label{sec-theoretical}
+
+\subsection{The population model}\label{sec-theory-theoretical_P}
 
@@
 
-\includegraphics[width=0.23\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_01.png}
-
-}
-
-\caption{\label{fig-CJ_TM_01}}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.32\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_02.png}
-
-}
-
-\caption{\label{fig-CJ_TM_02}}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
-
-}
-
-\caption{\label{fig-CJ_TM_03}}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
 \includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
@@
 
-\includegraphics[width=0.59\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_06.png}
-
-}
-
-\caption{\label{fig-CJ_TM_06}}
+\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_10.png}
+
+}
+
+\caption{\label{fig-CJ_TM_10}}
 
@@
 
+\subsection{The sample-comparison
+model}\label{sec-theory-theoretical_SC}
+
+Considering the sampling mechanism
+
 \begin{figure}
@@
 
-\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_07.png}
-
-}
-
-\caption{\label{fig-CJ_TM_07}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}
+
+}
+
+\caption{\label{fig-CJ_TM_12}}
 
@@
 
+Considering comparison mechanisms
+
 \begin{figure}
@@
 
-\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_08.png}
-
-}
-
-\caption{\label{fig-CJ_TM_08}}
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}
+
+}
+
+\caption{\label{fig-CJ_TM_14}}
 
@@
 
-\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
-
-}
-
-\caption{\label{fig-CJ_TM_09}}
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}
+
+}
+
+\caption{\label{fig-CJ_TM_15}}
 
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_10.png}
-
-}
-
-\caption{\label{fig-CJ_TM_10}}
-
-\end{figure}%
-
-\subsubsection{The sample-comparison
-model}\label{sec-theory-theoretical_SC}
-
-Considering the sampling mechanism
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}
-
-}
-
-\caption{\label{fig-CJ_TM_11}}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}
-
-}
-
-\caption{\label{fig-CJ_TM_12}}
-
-\end{figure}%
-
-Considering comparison mechanisms
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}
-
-}
-
-\caption{\label{fig-CJ_TM_13}}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}
-
-}
-
-\caption{\label{fig-CJ_TM_14}}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}
-
-}
-
-\caption{\label{fig-CJ_TM_15}}
-
-\end{figure}%
-
-\subsection{From theory to statistics}\label{sec-theory-statistics}
-
-Bayesian inference procedures offer three key advantages. First, they
-are well-suited to handling complex and overparameterized models,
-enabling researchers to estimate models where the number of parameters
-exceeds the number of observations for estimation
-\citep{Baker_1998, Kim_et_al_1999}. Second, they allow researchers to
-incorporate prior information, which helps constrain parameters within
-specified bounds. This capability addresses challenges such as
-non-convergence or improper parameter estimation that often arise in
-complex models when analyzed with frequentist methods
-\citep{Martin_et_al_1975, Seaman_et_al_2011}. Finally, Bayesian methods
-are particularly effective at drawing inferences from small sample
-sizes, where relying on the asymptotic properties of frequentist
-approaches may not be justified
-\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
+\section{From theory to statistics}\label{sec-statistical}
 
@@
 
-Researchers would then proceed to identify the estimands.
+Ideally, researchers then proceed to identify the estimands.
 \emph{Identification} determines whether an estimator can accurately
-compute an estimand based on its assumptions, regardless of random
-variability \citep[pp.~4]{Schuessler_et_al_2023}. An \emph{estimator} is
-a method or function that transforms data into an estimate
+compute the estimand based solely on its assumptions, regardless of
+random variability \citep[pp.~4]{Schuessler_et_al_2023}. Identification
+is crucial because if an estimator cannot reliably compute an estimand
+using ``infinite'' and error-free data, it is impossible to gain
+meaningful insights about that estimand using finite data
+\citep[pp.~5]{Schuessler_et_al_2023}. An \emph{estimator} refers to a
+method or function that transforms data into an estimate
 \citep{Neal_2020}. \emph{Estimates} are numerical values that
-approximate the estimand, and they are derived through
-\emph{estimation}, referring to the process of integrating data with an
-estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
-flowchart \citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow}
-provides a visual representation of the process of transitioning from
-estimands to estimates.
+approximate the estimand derived through the process of
+\emph{estimation}, which integrates data with an estimator
+\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
+\citep{McElreath_2020, Neal_2020}, shown in Figure~\ref{fig-IEflow},
+visually represents the transition from estimands to estimates.
 
@@
 
-While numerous estimators can approximate an estimand, researchers
-prioritize estimators with desirable identification properties that
-ensure the accuracy of estimates. For instance, the Z-test is an
-estimator known for its effectiveness when comparing groups'
-proportions, yielding accurate estimates when the underlying assumptions
-of the statistic are met \citep{Kanji_2006}. If this is the case, the
-Z-test is expressed as a signal-to-noise statistic
-\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
-as the difference between the groups' sample proportions,
-\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
-\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
-\(E[Y_{i} \:|\:T_{i}=1]\) and \(E[Y_{i} \:|\:T_{i}=2]\), respectively.
-The noise, represented by \(\hat{s}_{p}\), is defined as the unpooled
-sample variability observed between the two groups.
-
-However, researchers often seek to uncover the mechanisms underlying
-specific data and establish causal relationships rather than simply
-estimate associations. In the example, researchers can interpret the
-associational estimand in Equation~\ref{eq-group_diff} and the
-associational estimate represented by the Z-statistic as causal. This
-interpretation is valid as long as the data satisfy the assumptions of
-the Z-test and are collected through a randomized experiment.
+Consequently, although many estimators can approximate an estimand,
+researchers prioritize those with desirable identification properties to
+ensure accurate and reliable estimates. For instance, the Z-test is a
+widely used estimator for comparing group proportions, providing
+accurate estimates when its underlying assumptions are satisfied
+\citep{Kanji_2006}. Moreover, researchers can interpret the Z-test
+estimates as causal, provided they collect the data through a randomized
+experiment.
 
@@
 associational estimates as causal. They achieve this by ensuring data,
-and by extension an estimator, satisfies several key properties, such as
-common support, no interference, and consistency
+and by extension an estimator, satisfies several key identification
+properties, such as common support, no interference, and consistency
 \citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
@@
 \citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
-Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
-models} that facilitate valid causal inference. They achieve this by
-helping researchers to determine which statistical models can identify
-an estimand (ACE, CACE, or other), assuming the depicted causal
-structure is correct \citep{McElreath_2020}. Figure~\ref{fig-IEflow}
-shows the role of theoretical models in the inference process.
+Essentially, these tools serve as \emph{conceptual (theoretical) models}
+on which identification analysis rests
+\citep[pp.~4]{Schuessler_et_al_2023}. Specifically, they guide
+researchers in determining which statistical models can identify an
+estimand (ACE, CACE, or other), assuming the depicted causal structure
+is correct \citep{McElreath_2020}, thus enabling valid causal inference.
+Figure~\ref{fig-IEflow} shows the role of theoretical models in the
+inference process.
 

### 2025-01-27 paper2.qmd
section:appendix; process:write

@@
 <!-- 3. But what critical issues remain in CJ Research? -->
-Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unresolved.
-This study addresses two such issues that can compromise the reliability and validity of CJ's trait estimates.
-The first concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+Nevertheless, despite the increasing number of studies on CJ, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
+This study addresses two prominent issues in CJ literature that can undermine the reliability and validity of CJ's trait estimates.
+The first problem concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
 
@@
 The study is organized into six main sections.
-It begins with an overview of Thurstone's theory in @sec-thurstone_theory, followed by a detailed examination of these issues in @sec-theory-issues.
-In @sec-theoretical, the study develops a theoretical model for CJ that simultaneously addresses these issues.
-The model integrates Thurstone's core principles with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-@sec-statistical translates these theoretical and practical elements into a statistical model that supports the analysis of pairwise comparison data.
-Finally, @sec-discussion discusses the concepts developed in the study, their implications, and avenues for future exploration, while @sec-conclusion summarizes the findings.
+An overview of Thurstone's theory is presented in @sec-thurstone_theory, followed by a detailed examination of the identified issues in @sec-theory-issues.
+In @sec-theoretical, the study proposes a theoretical model for CJ that addresses these challenges.
+The model incorporates Thurstone's core principles alongside key assessment design features relevant to CJ experiments, such as judges, stimuli, and comparison selections.
+In @sec-statistical, these practical and theoretical considerations are translated into a statistical model for the analysis of pairwise comparison data.
+Finally, @sec-discussion explores the implications of the study's findings and the avenues for future exploration, and @sec-conclusion provides a summary of the key insights.
 
@@
 
-# The two critical issues in CJ literature {#sec-theory-issues}
-
-This section explores the two critical issues in CJ's theory that have the potential to compromise the reliability and validity of CJ's trait estimates.
-@sec-theory-issue1 inspects the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+# Two prominent issues in CJ literature {#sec-theory-issues}
+
+This section discusses the two prominent issues in CJ literature that can undermine the reliability and validity of CJ's trait estimates.
+@sec-theory-issue1 examines the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
 @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
@@
 <!-- 1. What is the problem with the general form -->
-Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a problem of trait scaling.
-The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp.
+Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a trait scaling problem.
+Specifically, the model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp.
 267].
-To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each case progressively incorporated additional simplifying assumptions into the model.
+To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions into the model.
 
@@
 
-In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can potentially compromise the reliability of the trait estimates and, by extension, their validity [@Perron_et_al_2015, pp.
+<!-- 4. segway for next section -->
+In aggregate, researchers' overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates.
+This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp.
 2].
-Consequently, adopting a more systematic and integrated approach to handling these assumptions and examining the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address the aforementioned issues.
+Consequently, addressing these issues and examining the factors influencing CJ experiments by adopting a more systematic and integrated approach could offer several statistical and measurement benefits.
 
@@
 
-## The population model {#sec-theory-theoretical_P}
+## The conceptual population model {#sec-theory-theoretical_P}
 
@@
 <!-- 2. What do they do next? -->
-Ideally, researchers then proceed to identify the estimands.
+Researchers then proceed to identify the estimands.
 *Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
 4].
-Identification is crucial because if an estimator cannot reliably compute an estimand using "infinite" and error-free data, it is impossible to gain meaningful insights about that estimand using finite data [@Schuessler_et_al_2023, pp.
-5].
 An *estimator* refers to a method or function that transforms data into an estimate [@Neal_2020].
@@
 
+<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+4].
+-->
+
+![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}
+
+<!-- 3. the importance of identification -->
+Identification is essential because it is a necessary condition for *consistency*.
+An estimator is *consistent* if it converges to the "true" value of an estimand as the sample size approaches infinity [@Everitt_et_al_2010].
+Without identification, an estimator cannot achieve consistency, rendering it incapable of reliably computing the estimand, even with "infinite" and error-free data.
+Consequently, deriving meaningful insights about the estimand from finite data becomes impossible [@Schuessler_et_al_2023, pp.
+5].
+Therefore, to ensure accurate and reliable estimates, researchers prioritize estimators with desirable identification properties.
+For instance, the Z-test is a widely used estimator for comparing group proportions, yielding accurate estimates when its underlying assumptions are satisfied [@Kanji_2006].
+Furthermore, researchers can interpret estimates from the Z-test as causal, provided the data is collected through a randomized experiment.
+
 <!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
@@
 -->
-
-<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
-Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
-4].
--->
-
-![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}
-
-<!-- 3. But, what do we need to produce a good estimate? -->
-Consequently, although many estimators can approximate an estimand, researchers prioritize those with desirable identification properties to ensure accurate and reliable estimates.
-For instance, the Z-test is a widely used estimator for comparing group proportions, providing accurate estimates when its underlying assumptions are satisfied [@Kanji_2006].
-Moreover, researchers can interpret the Z-test estimates as causal, provided they collect the data through a randomized experiment.
 
@@
 Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
-32] (see @eq-group_diff).
+32] (see @eq-CACE).
 
@@
 
-For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
+For instance, if researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
 Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)
@@
 35].
-
 @sec-appendix-B21 and @sec-appendix-B22 elaborate on the first advantage, while @sec-appendix-B22 and @sec-appendix-B23 do so for the second.
@@
 <!-- 1. Transforming probabilistic association into causation?  -->
-The structural approach to causal inference translate the probabilistic insights from @sec-appendix-B23 into actionable strategies seeking to identify the ACE from associational quantities.
-It achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
+The structural approach to causal inference translates probabilistic insights into actionable strategies seeking to identify the ACE from associational quantities.
+The approach achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
 34].
@@
 Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.
-
-Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$.
-They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE.
-This process is expressed in @eq-mACE2 (see @sec-appendix-B).
 Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.
-
-$$
-\begin{aligned}
-  \tau & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right] \\
-  & = \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=2, X) \cdot P(X)
-\end{aligned}
-$$ {#eq-mACE2}
 
@@
 
+Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$.
+They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE.
+This process is expressed in @eq-mACE2 (see @sec-appendix-B).
+
+$$
+\begin{aligned}
+  \tau & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right] \\
+  & = \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=2, X) \cdot P(X)
+\end{aligned}
+$$ {#eq-mACE2}
+
 #### The estimation process {#sec-appendix-C .appendix appendix-style=plain}
@@
 <!-- 3. How do DAGs can help? -->
-Nevertheless, researchers can manage the complexity of the likelihood by assuming that variables exhibit specific local (in)dependencies.
-These assumptions enhance model tractability and facilitate the estimation process.
-As described in @sec-appendix-B22, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF).
-From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution using @eq-bayes_rule.
+Nevertheless, researchers can manage the complexity of the likelihood by assuming specific local (in)dependencies among variables.
+SCMs and DAGs provide a formal framework to represent these assumptions, as detailed in @sec-appendix-B22.
+These assumptions improve model tractability and simplify the estimation process by enabling the derivation of the BNF of the likelihood (@eq-like_BNF).
+With this simplified structure, any probabilistic programming language can model the system and compute the parameter's posterior distribution using @eq-bayes_rule.
 

### 2025-01-27 _tex/paper2.tex
Built site for gh-pages

@@
 
-Nevertheless, despite the growing number of CJ studies, unsystematic and
-fragmented research approaches have left several critical issues
-unresolved. This study addresses two such issues that can compromise the
-reliability and validity of CJ's trait estimates. The first concerns the
-overreliance on Thurstone's Case V assumptions in the statistical
-analysis of CJ data. The second focuses on the apparent disconnect
-between CJ's approach to trait measurement and hypothesis testing.
-
-The study is organized into six main sections. It begins with an
-overview of Thurstone's theory in Section~\ref{sec-thurstone_theory},
-followed by a detailed examination of these issues in
+Nevertheless, despite the increasing number of studies on CJ, research
+in this domain remains unsystematic and fragmented, leaving several
+critical issues unresolved. This study addresses two prominent issues in
+CJ literature that can undermine the reliability and validity of CJ's
+trait estimates. The first problem concerns the overreliance on
+Thurstone's Case V assumptions in the statistical analysis of CJ data.
+The second involves the apparent disconnect between CJ's approach to
+trait measurement and hypothesis testing.
+
+The study is organized into six main sections. An overview of
+Thurstone's theory is presented in Section~\ref{sec-thurstone_theory},
+followed by a detailed examination of the identified issues in
 Section~\ref{sec-theory-issues}. In Section~\ref{sec-theoretical}, the
-study develops a theoretical model for CJ that simultaneously addresses
-these issues. The model integrates Thurstone's core principles with key
-assessment design features relevant to CJ experiments, such as the
-selection of judges, stimuli, and comparisons.
-Section~\ref{sec-statistical} translates these theoretical and practical
-elements into a statistical model that supports the analysis of pairwise
-comparison data. Finally, Section~\ref{sec-discussion} discusses the
-concepts developed in the study, their implications, and avenues for
-future exploration, while Section~\ref{sec-conclusion} summarizes the
-findings.
+study proposes a theoretical model for CJ that addresses these
+challenges. The model incorporates Thurstone's core principles alongside
+key assessment design features relevant to CJ experiments, such as
+judges, stimuli, and comparison selections. In
+Section~\ref{sec-statistical}, these practical and theoretical
+considerations are translated into a statistical model for the analysis
+of pairwise comparison data. Finally, Section~\ref{sec-discussion}
+explores the implications of the study's findings and the avenues for
+future exploration, and Section~\ref{sec-conclusion} provides a summary
+of the key insights.
 
@@
 
-\section{The two critical issues in CJ
-literature}\label{sec-theory-issues}
-
-This section explores the two critical issues in CJ's theory that have
-the potential to compromise the reliability and validity of CJ's trait
-estimates. Section~\ref{sec-theory-issue1} inspects the overreliance on
-Thurstone's Case V assumptions in the statistical analysis of CJ data.
+\section{Two prominent issues in CJ literature}\label{sec-theory-issues}
+
+This section discusses the two prominent issues in CJ literature that
+can undermine the reliability and validity of CJ's trait estimates.
+Section~\ref{sec-theory-issue1} examines the overreliance on Thurstone's
+Case V assumptions in the statistical analysis of CJ data.
 Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
@@
 Thurstone noted from the outset that the general form of the theory, as
-outlined in Section~\ref{sec-thurstone_theory}, gave rise to a problem
-of trait scaling. The model required estimating more ``unknown''
-parameters than the available pairwise comparisons
+outlined in Section~\ref{sec-thurstone_theory}, gave rise to a trait
+scaling problem. Specifically, the model required estimating more
+``unknown'' parameters than the available pairwise comparisons
 \citep[pp.~267]{Thurstone_1927b}. To address this issue and facilitate
 the practical implementation of the theory, he developed five cases
-derived from this general form, each case progressively incorporated
+derived from this general form, each progressively incorporating
 additional simplifying assumptions into the model.
@@
 
-In aggregate, researchers' inadequate handling of violations to the
-assumptions of equal dispersion and zero correlation between stimuli,
-coupled with the apparent disconnect between CJ's approach to trait
-measurement and hypothesis testing, can potentially compromise the
-reliability of the trait estimates and, by extension, their validity
-\citep[pp.~2]{Perron_et_al_2015}. Consequently, adopting a more
-systematic and integrated approach to handling these assumptions and
-examining the factors influencing CJ experiments could offer several
-statistical and measurement benefits, including the ability to address
-the aforementioned issues.
+In aggregate, researchers' overreliance on Thurstone's Case V
+assumptions in the statistical analysis of CJ data can compromise the
+reliability of CJ's trait estimates. This overreliance, coupled with the
+apparent disconnect between CJ's approach to trait measurement and
+hypothesis testing, may also undermine the validity of these estimates
+\citep[pp.~2]{Perron_et_al_2015}. Consequently, addressing these issues
+and examining the factors influencing CJ experiments by adopting a more
+systematic and integrated approach could offer several statistical and
+measurement benefits.
 
@@
 
-\subsection{The population model}\label{sec-theory-theoretical_P}
+\subsection{The conceptual population
+model}\label{sec-theory-theoretical_P}
 
@@
 
-Ideally, researchers then proceed to identify the estimands.
+Researchers then proceed to identify the estimands.
 \emph{Identification} determines whether an estimator can accurately
 compute the estimand based solely on its assumptions, regardless of
-random variability \citep[pp.~4]{Schuessler_et_al_2023}. Identification
-is crucial because if an estimator cannot reliably compute an estimand
-using ``infinite'' and error-free data, it is impossible to gain
-meaningful insights about that estimand using finite data
-\citep[pp.~5]{Schuessler_et_al_2023}. An \emph{estimator} refers to a
-method or function that transforms data into an estimate
-\citep{Neal_2020}. \emph{Estimates} are numerical values that
-approximate the estimand derived through the process of
+random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
+\emph{estimator} refers to a method or function that transforms data
+into an estimate \citep{Neal_2020}. \emph{Estimates} are numerical
+values that approximate the estimand derived through the process of
 \emph{estimation}, which integrates data with an estimator
@@
 
-Consequently, although many estimators can approximate an estimand,
-researchers prioritize those with desirable identification properties to
-ensure accurate and reliable estimates. For instance, the Z-test is a
-widely used estimator for comparing group proportions, providing
-accurate estimates when its underlying assumptions are satisfied
-\citep{Kanji_2006}. Moreover, researchers can interpret the Z-test
-estimates as causal, provided they collect the data through a randomized
-experiment.
+Identification is essential because it is a necessary condition for
+\emph{consistency}. An estimator is \emph{consistent} if it converges to
+the ``true'' value of an estimand as the sample size approaches infinity
+\citep{Everitt_et_al_2010}. Without identification, an estimator cannot
+achieve consistency, rendering it incapable of reliably computing the
+estimand, even with ``infinite'' and error-free data. Consequently,
+deriving meaningful insights about the estimand from finite data becomes
+impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore, to ensure
+accurate and reliable estimates, researchers prioritize estimators with
+desirable identification properties. For instance, the Z-test is a
+widely used estimator for comparing group proportions, yielding accurate
+estimates when its underlying assumptions are satisfied
+\citep{Kanji_2006}. Furthermore, researchers can interpret estimates
+from the Z-test as causal, provided the data is collected through a
+randomized experiment.
 
@@
 the subset of the population defined by the conditioning variable
-\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).
+\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).
 
@@
 
-For instance, If researchers cannot conduct the randomized experiments
+For instance, if researchers cannot conduct the randomized experiments
 described in Section~\ref{sec-appendix-A} and must instead rely on
@@
 their parameters \citep[pp.~35]{Pearl_et_al_2016}.
-
 Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
@@
 
-The structural approach to causal inference translate the probabilistic
-insights from Section~\ref{sec-appendix-B23} into actionable strategies
-seeking to identify the ACE from associational quantities. It achieves
-this by relying on the \emph{modularity assumption}, which posits that
-intervening on a node alters only the causal mechanism of that node,
-leaving others unchanged \citep[pp.~34]{Neal_2020}.
+The structural approach to causal inference translates probabilistic
+insights into actionable strategies seeking to identify the ACE from
+associational quantities. The approach achieves this by relying on the
+\emph{modularity assumption}, which posits that intervening on a node
+alters only the causal mechanism of that node, leaving others unchanged
+\citep[pp.~34]{Neal_2020}.
 
@@
 \((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
-the direct causal effect \(T \rightarrow Y\).
-
-Researchers can then apply the \emph{backdoor adjustment} to identify
-the ACE of \(T\) on \(Y\). They achieve this by first identifying the
-CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
-this effect over \(X\) to obtain the ACE. This process is expressed in
-Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
-the purpose of identification, the conditioned DAG
-\ref{fig-mdag_example2} is equivalent to the manipulated DAG
-\ref{fig-mdag_example3}, because \(X\) satisfies the backdoor criterion.
-
-\begin{equation}\phantomsection\label{eq-mACE2}{
-\begin{aligned}
-  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right] \\
-  & = \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=2, X) \cdot P(X)
-\end{aligned}
-}\end{equation}
+the direct causal effect \(T \rightarrow Y\). Notably, for the purpose
+of identification, the conditioned DAG \ref{fig-mdag_example2} is
+equivalent to the manipulated DAG \ref{fig-mdag_example3}, because \(X\)
+satisfies the backdoor criterion.
 
@@
 
+Researchers can then apply the \emph{backdoor adjustment} to identify
+the ACE of \(T\) on \(Y\). They achieve this by first identifying the
+CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
+this effect over \(X\) to obtain the ACE. This process is expressed in
+Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}).
+
+\begin{equation}\phantomsection\label{eq-mACE2}{
+\begin{aligned}
+  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right] \\
+  & = \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=2, X) \cdot P(X)
+\end{aligned}
+}\end{equation}
+
 \paragraph{The estimation process}\label{sec-appendix-C}
@@
 Nevertheless, researchers can manage the complexity of the likelihood by
-assuming that variables exhibit specific local (in)dependencies. These
-assumptions enhance model tractability and facilitate the estimation
-process. As described in Section~\ref{sec-appendix-B22}, researchers can
-use SCMs and DAGs to formalize these local (in)dependencies, which leads
-to the BNF of the likelihood (Equation~\ref{eq-like_BNF}). From this
-point, any probabilistic programming language can model this simpler
-structure and generate the parameter's posterior distribution using
-Equation~\ref{eq-bayes_rule}.
+assuming specific local (in)dependencies among variables. SCMs and DAGs
+provide a formal framework to represent these assumptions, as detailed
+in Section~\ref{sec-appendix-B22}. These assumptions improve model
+tractability and simplify the estimation process by enabling the
+derivation of the BNF of the likelihood (Equation~\ref{eq-like_BNF}).
+With this simplified structure, any probabilistic programming language
+can model the system and compute the parameter's posterior distribution
+using Equation~\ref{eq-bayes_rule}.
 

### 2025-01-28 paper2.qmd
section:theoretical; process:write

@@
 <!-- recording: Sven and Tine 24.10.25; time: 00:00:00 - 00:17:00 -->
+<!-- recording: Sven and Tine 24.10.25; time: 00:00:00 - 00:17:00 -->
+<!-- recording: Sven and Tine 25.01.22; time: 00:06:30 - 00:08:25 -->
 
@@
 <!-- 3. But what critical issues remain in CJ Research? -->
-Nevertheless, despite the increasing number of studies on CJ, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
-This study addresses two prominent issues in CJ literature that can undermine the reliability and validity of CJ's trait estimates.
-The first problem concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
+This study discusses two prominent issues that can compromise the reliability and validity of CJ's trait estimates.
+The first issue concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
 The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+The study then addresses these challenges by adopting a more systematic and integrated approach grounded in causal inference and Bayesian inference methods.
 
 <!-- 4. what is the organization of the study? -->
-The study is organized into six main sections.
-An overview of Thurstone's theory is presented in @sec-thurstone_theory, followed by a detailed examination of the identified issues in @sec-theory-issues.
-In @sec-theoretical, the study proposes a theoretical model for CJ that addresses these challenges.
-The model incorporates Thurstone's core principles alongside key assessment design features relevant to CJ experiments, such as judges, stimuli, and comparison selections.
-In @sec-statistical, these practical and theoretical considerations are translated into a statistical model for the analysis of pairwise comparison data.
-Finally, @sec-discussion explores the implications of the study's findings and the avenues for future exploration, and @sec-conclusion provides a summary of the key insights.
+As a result, the study divides its content into six main sections.
+@sec-thurstone_theory provides an overview of Thurstone's theory.
+@sec-theory-issues discusses the identified issues in detail.
+@sec-theoretical updates the Thurstonian model to address these challenges.
+The model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+@sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of pairwise comparison data.
+@sec-discussion discusses the implications of the findings and explores avenues for future research.
+Finally, @sec-conclusion summarizes the study's key insights.
 
@@
 
-This section discusses the two prominent issues in CJ literature that can undermine the reliability and validity of CJ's trait estimates.
+This section discusses the two prominent issues that can undermine the reliability and validity of CJ's trait estimates.
 @sec-theory-issue1 examines the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
@@
 According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons.
-A thought experiment can illustrate this idea.
-In this experiment, the researcher can observe the discriminal processes for the texts depicted in @fig-discriminal_process.
+A thought experiment can illustrate this concept.
+In this experiment, the researcher observes the discriminal processes for the texts depicted in @fig-discriminal_process.
 Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
@@
 This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome.
-Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor Text B over Text A (shaded gray area), compared to situations where their dispersions differ.
+Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ.
+Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area)
 
@@
 <!-- 4. So, what happens when the assumption does not hold? -->
-Thus, model misspecification, in the form of an erroneous assumption of equal dispersions between stimuli, can give rise to significant statistical and measurement issues.
+Thus, erroneously assuming equal dispersions between stimuli, can give rise to significant statistical and measurement issues.
 For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli.
@@
 As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
-A similar thought experiment, as the one in @sec-theory-issue1a, can illustrate this idea.
+A similar thought experiment, as the one in @sec-theory-issue1a, can illustrate this concept.
 The experiment now assumes that the discriminal dispersions for both texts remain constant.
@@
 This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
-Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area) compared to scenarios where the texts are highly correlated.
+Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are highly correlated.
+Consequently, the discriminal difference is less likely to favor Text B over Text A (shaded gray area).
 
@@
 <!-- 2. But this is not the scenario we observe -->
-Off course, in experimental practice, researchers approach this process in reverse.
+Again, in experimental practice, researchers approach this process in reverse.
 They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli.
@@
 This scenario is also illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$.
-Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.
+Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.
 
@@
 <!-- 2. But does it hold?  -->
-Notably, Thurstone attributed the lack of correlation (independence) between stimuli to the cancellation of potential judges' biases.
+Notably, Thurstone attributed the lack of correlation between stimuli to the cancellation of potential judges' biases.
 He argued that this cancellation resulted from two opposing and equally weighted effects occurring during pairwise comparisons [@Thurstone_1927b, pp.
@@
 <!-- 5. But what are the challenges when this assumption is violated? -->
-In any case, similar to @sec-theory-issue1a, model misspecification due to an erroneous assumption of zero correlation between stimuli can lead to significant statistical and measurement issues.
-For instance, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli.
+In any case, similar to @sec-theory-issue1a, assuming zero correlation between stimuli by neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures can cause significant statistical and measurement issues.
+In general, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli.
 Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes.
-This scenario is also illustrated by @fig-correlation, when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
-
-The model misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures.
-Neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp.
+This scenario is illustrated by @fig-correlation, when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
+
+In particular, neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp.
 341] or, worse, introducing bias into the trait's estimates [@Ackerman_1989].
@@
 <!-- 4. segway for next section -->
-In aggregate, researchers' overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates.
+In aggregate, the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates.
 This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp.
 2].
-Consequently, addressing these issues and examining the factors influencing CJ experiments by adopting a more systematic and integrated approach could offer several statistical and measurement benefits.
+Consequently, addressing these issues by adopting a more systematic and integrated approach could offer several statistical and measurement benefits.
 
@@
 
-# Updating the theoretical model {#sec-theoretical}
-
-<!-- This study examines two such problems that can compromise the reliability and validity of CJ's trait estimates. -->
-
-<!-- This section aims to refine CJ's theoretical model using the *structural approach* to causal inference with the purpose of addressing the two.  -->
-
-<!-- This approach, as outlined in @sec-theory-theoretical, defines a model that integrates the core principles of Thurstone's theory, as well as other key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and the number of comparisons.
--->
-
-<!-- See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section. -->
-
-<!-- 1. The structural approach to causal inference -->
-<!-- The *structural approach* to causal inference provides a formal framework to identify causes and estimate their effects using data.
-The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system.
-Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference.
-They achieve this by helping researchers to determine which statistical models can identify an estimand, assuming the depicted causal structure is correct [@McElreath_2020].
-*Identification* determines whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+# Extending Thurstone's model {#sec-theoretical}
+
+<!-- recording: Sven and Tine 25.01.22; time: 00:05:40 - 00:06:30 -->
+<!-- recording: Sven and Tine 25.01.22; time: 00:08:25 - 00:31:56 -->
+
+<!-- 1. What the section will do? -->
+This section addresses the challenges outlined in @sec-theory-issues by leveraging the structural approach to causal inference to extend Thurstone's model.
+The extension combines the core theoretical principles from @sec-thurstone_theory with several key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+Specifically, @sec-theory-theoretical_P incorporates the theoretical principles into a *conceptual-population model*.
+Conversely, @sec-theory-theoretical_SC integrates the assessment design features into a *sample-comparison model*.
+Combining these elements helps to elucidate how all actors and processes interact during CJ experiments.
+
+<!-- Generalizability is the degree to which you can apply the results of your study to a broader context.  -->
+<!-- https://www.scribbr.com/research-bias/generalizability/ -->
+
+<!-- A theoretical model is a framework that researchers create to structure a study process and plan how to approach a specific research inquiry.
+It can allow you to define the purpose of your research and develop an informed perspective.
+Creating a theoretical model typically involves analyzing a subject's literature, including previous research studies and journal articles.
+Researchers often include a written description of a theoretical model in a beginning section of a dissertation or similar document so a reader can gain more context.
+-->
+<!-- url: https://www.indeed.com/career-advice/career-development/theoretical-model -->
+
+<!-- 2. The structural approach to causal inference -->
+The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data.
+This approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments.
+Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
 4].
--->
-
-<!-- 2. The SCMs -->
-<!-- SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020].
-Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
-In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
-27,68].
-Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
-These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
--->
-
-<!-- 3. The DAGs -->
-<!-- Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
-A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables.
-The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence.
-The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
-DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
--->
-
-<!-- 4. The benefits of SCMs and DAGs -->
-<!-- SCMs and DAGs support causal inference through two key advantages.
+Identification analysis helps researchers determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+4].
+Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010].
+*Estimators* denote the methods or functions that transform data into an estimate [@Neal_2020], while *estimates* are the numerical values approximating the estimand.
+These estimates are obtained through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010].
+
+<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+4].
+-->
+
+<!-- 4. example of an estimand and an estimator in CJ -->
+For instance, researchers might aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each exposed to a different teaching method.
+Afterward, judges would compare pairs of written texts, generating a dichotomous outcome that reflects the relative quality of each text.
+In this scenario, researchers can rephrase the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
+Researchers would then use estimates from the BTL model, or its transformations, to approximate this estimand.
+
+<!-- 5. But the estimator does not satisfy minimum identification properties -->
+However, @sec-theory-issues provides compelling evidence that the BTL model suffers from several statistical and measurement issues, including identification problems.
+These issues ultimately raise concerns about the consistency of the model's estimates.
+An estimator is *consistent* if it converges to the "true" value of an estimand as the data size approaches infinity [@Everitt_et_al_2010].
+Since identification is a prerequisite for consistency, the absence of proper identification means that researchers cannot achieve consistency, even with "infinite" and error-free data.
+As a result, deriving meaningful insights from finite data also becomes impossible [@Schuessler_et_al_2023, pp.
+5].
+
+<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+5].
+-->
+
+<!-- 5. So, how do SCMs and DAGs support identification? -->
+Luckily, SCMs and DAGs support identification analysis through two key advantages.
 First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
 This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
-Second, they depict causal relationships in a non-parametric and fully interactive way.
-This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp.
+Second, they depict causal relationships in an interactive, non-parametric manner.
+This flexibility enables feasible identification strategies without defining the variables' data types, the functional forms, or their parameters [@Pearl_et_al_2016, pp.
 35].
--->
-
-<!-- 5. what are we doing with them? -->
-<!-- Therefore, this section leverages SCMs and DAGs to represent the conceptual population of CJ experiments, referred to as *the population model* in @sec-theory-theoretical_P, and later integrates various assessment design features relevant to CJ experiments.
-This process leads to the development of *the sample-comparison model* in @sec-theory-theoretical_SC.
--->
-
-## The conceptual population model {#sec-theory-theoretical_P}
+
+<!-- See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section. -->
+
+## The conceptual-population model {#sec-theory-theoretical_P}
 
@@
 
+<!-- 1. The SCMs -->
+Before integrating the theoretical principles from @sec-thurstone_theory, it is important to define SCMs and DAGs.
+SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020].
+Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
+In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
+27,68].
+Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
+These function use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+
+<!-- 2. The DAGs -->
+Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
+A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables.
+The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence.
+The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
+DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
+While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+
+<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
+
+<!-- 3. The begining -->
+@fig-CJ1 displays CJ's outcome, discriminal difference, discriminal processes and judges' biases assumed by the Thurstone's theory. Notice that in the figure, the structural
+
+<!-- Assuming population data or more commonly known as census data, we ... -->
+
+<!-- The (latent) discriminal difference of the stimuli directly determines the (manifest) outcome of the pairwise comparisons -->
+
+<!-- The (latent) "perceived" discriminal processes for the stimuli directly determines their discriminal difference -->
+
+<!-- The (latent) "true" discriminal processes for the stimuli and the judges' biases directly determines their (latent) "perceived" discriminal processes -->
+
+::: {#fig-CJ1 layout-nrow=2}
+
+::: {#fig-CJ1_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{kjabi_{1}i_{2}} & := f_{O}(D_{kjabi_{1}i_{2}}) \\
+  D_{kjabi_{1}i_{2}} & := f_{D}(PT_{kjai_{1}}, PT_{kjbi_{2}}) \\
+  PT_{kjai_{1}} & := f_{PT}(T_{ai_{1}}, T_{bi_{2}}, B_{kj})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_04.png){#fig-CJ1_DAG fig-align="center" fig-pos="H" width=45%}
+
+CJ's outcome, discriminal difference, discriminal processes and judges' biases
+:::
+
+<!-- To declare a matrix \( P \) in \( \mathbb{N}^2 \) that represents the paired combinations of elements in \( I \), follow these steps: -->
+
+<!-- ### Definitions: -->
+<!-- 1. \( I \) is a subset of the natural numbers, \( I \subseteq \mathbb{N} \). -->
+<!-- 2. The paired combination of elements in \( I \) can be written as \( (i, j) \) for \( i, j \in I \). -->
+
+<!-- ### Constructing the Matrix \( P \): -->
+<!-- 1. Each row of \( P \) represents one pairing \( (i, j) \). -->
+<!-- 2. If \( I = \{a_1, a_2, \ldots, a_n\} \), then \( P \) contains \( n^2 \) rows for all pairs \( (i, j) \), including repetitions where \( i = j \). -->
+<!-- 3. Specifically, the matrix \( P \) is structured as: -->
+<!--    \[ -->
+<!--    P = \begin{bmatrix} -->
+<!--    i_1 & j_1 \\ -->
+<!--    i_2 & j_2 \\ -->
+<!--    \vdots & \vdots \\ -->
+<!--    i_{n^2} & j_{n^2} -->
+<!--    \end{bmatrix}, -->
+<!--    \] -->
+<!--    where \( i_k, j_k \in I \) for all \( k \), and each pair \( (i_k, j_k) \) corresponds to a combination of elements in \( I \). -->
+
+<!-- ### Algorithm to Generate \( P \): -->
+<!-- Given \( I = \{a_1, a_2, \ldots, a_n\} \): -->
+<!-- 1. Loop over all \( i \in I \) (outer loop). -->
+<!-- 2. For each \( i \), loop over all \( j \in I \) (inner loop). -->
+<!-- 3. Append the pair \( (i, j) \) as a new row in \( P \). -->
+
+<!-- ### Example: -->
+<!-- If \( I = \{1, 2, 3\} \), then the matrix \( P \) is: -->
+<!-- \[ -->
+<!-- P = \begin{bmatrix} -->
+<!-- 1 & 1 \\ -->
+<!-- 1 & 2 \\ -->
+<!-- 1 & 3 \\ -->
+<!-- 2 & 1 \\ -->
+<!-- 2 & 2 \\ -->
+<!-- 2 & 3 \\ -->
+<!-- 3 & 1 \\ -->
+<!-- 3 & 2 \\ -->
+<!-- 3 & 3 -->
+<!-- \end{bmatrix}. -->
+<!-- \] -->
+
+<!-- ### Mathematical Notation: -->
+<!-- 1. The matrix \( P \) can be defined as: -->
+<!--    \[ -->
+<!--    P = \{(i, j) \mid i \in I, j \in I\}. -->
+<!--    \] -->
+<!-- 2. In terms of dimensions, \( P \) has \( |I|^2 \) rows and 2 columns. -->
+
+<!-- This formalization ensures that \( P \) represents all paired combinations of elements from \( I \) in \( \mathbb{N}^2 \). -->
+
+<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->
+
+<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->
+
+<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->
+
+without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in
+
+![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%}
+
+<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->
+
+<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->
+
+<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->
+
+<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->
+
+![](/images/png/CJ_TM_10.png){#fig-CJ_TM_10 width=68%}
+
+<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
+
+## The sample-comparison model {#sec-theory-theoretical_SC}
+
 <!-- comparison mechanism as missing mechanism -->
@@
 
-<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
-
-Assuming population data or more commonly known as census data, we ...
-
-The (latent) discriminal difference of the stimuli directly determines the (manifest) outcome of the pairwise comparisons
-
-The (latent) "perceived" discriminal processes for the stimuli directly determines their discriminal difference
-
-The (latent) "true" discriminal processes for the stimuli and the judges' biases directly determines their (latent) "perceived" discriminal processes
-
-<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->
-
-<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->
-
-<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->
-
-![](/images/png/CJ_TM_04.png){#fig-CJ_TM_04 width=45%}
-
-without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in
-
-![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%}
-
-<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->
-
-<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->
-
-![](/images/png/CJ_TM_10.png){#fig-CJ_TM_10 width=68%}
-
-<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
-
-## The sample-comparison model {#sec-theory-theoretical_SC}
-
 Considering the sampling mechanism
@@
 
-# From theory to statistics {#sec-statistical}
+# Extending the BTL model {#sec-statistical}
 
@@
 
+<!-- recording: Sven and Tine 25.01.22; time: 00:00:00 - 00:05:40 -->
+
 This section introduces fundamental statistical and causal inference concepts necessary for understanding the core theoretical principles described in this document.
@@
 In this context, researchers typically formulate their research questions as *estimands* or *targets of inference*, i.e., the specific quantities they seek to determine [@Everitt_et_al_2010].
-For instance, researchers might be interested in answering the following question: “To what extent do different teaching methods $(T)$ influence students' ability to produce high-quality written texts $(Y)$?" To investigate this, researchers could randomly assign students to two groups, each exposed to a different teaching method $(T_{i} = \{1,2\})$.
+For instance, researchers might be interested in answering the following question: "To what extent do different teaching methods $(T)$ influence students' ability to produce high-quality written texts $(Y)$?" To investigate this, researchers could randomly assign students to two groups, each exposed to a different teaching method $(T_{i} = \{1,2\})$.
 Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability.
-In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random associational quantity in @eq-group_diff, where $E[\cdot]$ denotes the expected value.
+In this scenario, the research question can be rephrased as the estimand, "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?" and this estimand can be mathematically represented by the random associational quantity in @eq-group_diff, where $E[\cdot]$ denotes the expected value.
 
@@
 <!-- 3. the importance of identification -->
-Identification is essential because it is a necessary condition for *consistency*.
-An estimator is *consistent* if it converges to the "true" value of an estimand as the sample size approaches infinity [@Everitt_et_al_2010].
-Without identification, an estimator cannot achieve consistency, rendering it incapable of reliably computing the estimand, even with "infinite" and error-free data.
-Consequently, deriving meaningful insights about the estimand from finite data becomes impossible [@Schuessler_et_al_2023, pp.
+Identification is a necessary condition to ensure *consistent* estimators.
+An estimator achieves *consistency* when it converges to the "true" value of an estimand as the data size approaches infinity [@Everitt_et_al_2010].
+Without identification, researchers cannot achieve consistency, even with "infinite" and error-free data.
+As a result, deriving meaningful insights about an estimand from finite data becomes impossible [@Schuessler_et_al_2023, pp.
 5].
@@
 4].
-Specifically, they guide researchers in determining which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020], thus enabling valid causal inference.
+Thus, using these tools, researchers can determine which statistical models can identify (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020], thus enabling valid causal inference.
 @fig-IEflow shows the role of theoretical models in the inference process.
@@
 
-SCMs and DAGs support causal inference through two key advantages.
+SCMs and DAGs support identification analysis through two key advantages.
 First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].

### 2025-01-28 _tex/paper2.tex
Built site for gh-pages

@@
 
-Nevertheless, despite the increasing number of studies on CJ, research
-in this domain remains unsystematic and fragmented, leaving several
-critical issues unresolved. This study addresses two prominent issues in
-CJ literature that can undermine the reliability and validity of CJ's
-trait estimates. The first problem concerns the overreliance on
-Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second involves the apparent disconnect between CJ's approach to
-trait measurement and hypothesis testing.
-
-The study is organized into six main sections. An overview of
-Thurstone's theory is presented in Section~\ref{sec-thurstone_theory},
-followed by a detailed examination of the identified issues in
-Section~\ref{sec-theory-issues}. In Section~\ref{sec-theoretical}, the
-study proposes a theoretical model for CJ that addresses these
-challenges. The model incorporates Thurstone's core principles alongside
-key assessment design features relevant to CJ experiments, such as
-judges, stimuli, and comparison selections. In
-Section~\ref{sec-statistical}, these practical and theoretical
-considerations are translated into a statistical model for the analysis
-of pairwise comparison data. Finally, Section~\ref{sec-discussion}
-explores the implications of the study's findings and the avenues for
-future exploration, and Section~\ref{sec-conclusion} provides a summary
-of the key insights.
+Nevertheless, despite the increasing number of CJ studies, research in
+this domain remains unsystematic and fragmented, leaving several
+critical issues unresolved. This study discusses two prominent issues
+that can compromise the reliability and validity of CJ's trait
+estimates. The first issue concerns the overreliance on Thurstone's Case
+V assumptions in the statistical analysis of CJ data. The second
+involves the apparent disconnect between CJ's approach to trait
+measurement and hypothesis testing. The study then addresses these
+challenges by adopting a more systematic and integrated approach
+grounded in causal inference and Bayesian inference methods.
+
+As a result, the study divides its content into six main sections.
+Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
+theory. Section~\ref{sec-theory-issues} discusses the identified issues
+in detail. Section~\ref{sec-theoretical} updates the Thurstonian model
+to address these challenges. The model integrates the theoretical core
+principles alongside key assessment design features relevant to CJ
+experiments, such as the selection of judges, stimuli, and comparisons.
+Section~\ref{sec-statistical} translates these theoretical and practical
+elements into a probabilistic statistical model for the analysis of
+pairwise comparison data. Section~\ref{sec-discussion} discusses the
+implications of the findings and explores avenues for future research.
+Finally, Section~\ref{sec-conclusion} summarizes the study's key
+insights.
 
@@
 
-This section discusses the two prominent issues in CJ literature that
-can undermine the reliability and validity of CJ's trait estimates.
+This section discusses the two prominent issues that can undermine the
+reliability and validity of CJ's trait estimates.
 Section~\ref{sec-theory-issue1} examines the overreliance on Thurstone's
@@
 direct influence on the outcome of pairwise comparisons. A thought
-experiment can illustrate this idea. In this experiment, the researcher
-can observe the discriminal processes for the texts depicted in
+experiment can illustrate this concept. In this experiment, the
+researcher observes the discriminal processes for the texts depicted in
 Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
@@
 dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
-discriminal difference is more likely to favor Text B over Text A
-(shaded gray area), compared to situations where their dispersions
-differ.
+discriminal difference is more narrow compared to situations where their
+dispersions differ. Consequently, the discriminal difference is more
+likely to favor Text B over Text A (shaded gray area)
 
@@
 
-Thus, model misspecification, in the form of an erroneous assumption of
-equal dispersions between stimuli, can give rise to significant
-statistical and measurement issues. For instance, the model may
-overestimate the degree to which the outcome accurately reflects the
-``true'' discriminal differences between stimuli. This overestimation
-can result in researchers drawing spurious conclusions about these
-differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
-underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
-also illustrates this issue when the model's discriminal difference
-distribution aligns with the thick continuous line for
-\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
-follows any discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
-Additionally, if researchers recognize that misfit statistics highlight
-these critical differences in dispersions, the conventional CJ practice
-of excluding stimuli based on these statistics
+Thus, erroneously assuming equal dispersions between stimuli, can give
+rise to significant statistical and measurement issues. For instance,
+the model may overestimate the degree to which the outcome accurately
+reflects the ``true'' discriminal differences between stimuli. This
+overestimation can result in researchers drawing spurious conclusions
+about these differences \citep[pp.~370]{McElreath_2020} and, by
+extension, about the underlying discriminal processes of stimuli.
+Figure~\ref{fig-dispersion} also illustrates this issue when the model's
+discriminal difference distribution aligns with the thick continuous
+line for \(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal
+difference follows any discontinuous line where
+\(\sigma_{B}-\sigma_{A} \neq 0\). Additionally, if researchers recognize
+that misfit statistics highlight these critical differences in
+dispersions, the conventional CJ practice of excluding stimuli based on
+these statistics
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
@@
 similar thought experiment, as the one in
-Section~\ref{sec-theory-issue1a}, can illustrate this idea. The
+Section~\ref{sec-theory-issue1a}, can illustrate this concept. The
 experiment now assumes that the discriminal dispersions for both texts
@@
 independent or uncorrelated \((\rho=0)\), their discriminal difference
-is less likely to favor Text B over Text A (shaded gray area) compared
-to scenarios where the texts are highly correlated.
-
-Off course, in experimental practice, researchers approach this process
-in reverse. They begin by observing the sample of outcomes favoring Text
-B over Text A and then use the BTL model to estimate the discriminal
+is less narrow compared to scenarios where the texts are highly
+correlated. Consequently, the discriminal difference is less likely to
+favor Text B over Text A (shaded gray area).
+
+Again, in experimental practice, researchers approach this process in
+reverse. They begin by observing the sample of outcomes favoring Text B
+over Text A and then use the BTL model to estimate the discriminal
 difference and the discriminal processes of the stimuli. Given that the
@@
 represented by the thick continuous line corresponding to \(\rho=0\).
-Once more, the estimation accuracy of the discriminal difference ensures
-reliable estimates for the discriminal processes of the texts {(citation
+Once more, the accuracy of the discriminal difference ensures reliable
+estimates for the discriminal processes of the texts {(citation
 needed?)}.
 
-Notably, Thurstone attributed the lack of correlation (independence)
-between stimuli to the cancellation of potential judges' biases. He
-argued that this cancellation resulted from two opposing and equally
-weighted effects occurring during pairwise comparisons
-\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
-mathematical demonstration of this cancellation using the BTL model
-under the assumption of discriminal processes with additive biases.
-However, it is easy to imagine at least two scenarios in which the zero
-correlation assumption is almost certainly invalid: when the pairwise
-comparison involves multidimensional, complex traits with heterogeneous
-stimuli and when an additional hierarchical structure is relevant to the
-stimuli.
+Notably, Thurstone attributed the lack of correlation between stimuli to
+the cancellation of potential judges' biases. He argued that this
+cancellation resulted from two opposing and equally weighted effects
+occurring during pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
+\citet{Andrich_1978} provided a mathematical demonstration of this
+cancellation using the BTL model under the assumption of discriminal
+processes with additive biases. However, it is easy to imagine at least
+two scenarios in which the zero correlation assumption is almost
+certainly invalid: when the pairwise comparison involves
+multidimensional, complex traits with heterogeneous stimuli and when an
+additional hierarchical structure is relevant to the stimuli.
 
@@
 
-In any case, similar to Section~\ref{sec-theory-issue1a}, model
-misspecification due to an erroneous assumption of zero correlation
-between stimuli can lead to significant statistical and measurement
-issues. For instance, the model may over- or underestimate how
-accurately the outcome reflects the ``true'' discriminal differences
-between stimuli. Such inaccuracies can result in spurious inferences
-about these differences and, by extension, about the stimuli's
-discriminal processes. This scenario is also illustrated by
-Figure~\ref{fig-correlation}, when the model's discriminal difference
-distribution aligns with the thick continuous line for \(\rho=0\), while
-the ``true'' discriminal difference follows any discontinuous line where
-\(\rho \neq 0\).
-
-The model misspecification may arise from neglecting additional relevant
-traits, excluding judges based on misfit statistics, or ignoring
-hierarchical (grouping) structures. Neglecting relevant traits, such as
-judges' biases, can cause dimensional mismatches in the BTL model,
-artificially inflating the trait's reliability
-\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introducing bias into the
-trait's estimates \citep{Ackerman_1989}. Excluding judges based on
-misfit statistics risks discarding valuable information, which may
-further bias the trait's estimates
+In any case, similar to Section~\ref{sec-theory-issue1a}, assuming zero
+correlation between stimuli by neglecting additional relevant traits,
+excluding judges based on misfit statistics, or ignoring hierarchical
+(grouping) structures can cause significant statistical and measurement
+issues. In general, the model may over- or underestimate how accurately
+the outcome reflects the ``true'' discriminal differences between
+stimuli. Such inaccuracies can result in spurious inferences about these
+differences and, by extension, about the stimuli's discriminal
+processes. This scenario is illustrated by Figure~\ref{fig-correlation},
+when the model's discriminal difference distribution aligns with the
+thick continuous line for \(\rho=0\), while the ``true'' discriminal
+difference follows any discontinuous line where \(\rho \neq 0\).
+
+In particular, neglecting relevant traits, such as judges' biases, can
+cause dimensional mismatches in the BTL model, artificially inflating
+the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
+introducing bias into the trait's estimates \citep{Ackerman_1989}.
+Excluding judges based on misfit statistics risks discarding valuable
+information, which may further bias the trait's estimates
 \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
@@
 
-In aggregate, researchers' overreliance on Thurstone's Case V
-assumptions in the statistical analysis of CJ data can compromise the
-reliability of CJ's trait estimates. This overreliance, coupled with the
-apparent disconnect between CJ's approach to trait measurement and
-hypothesis testing, may also undermine the validity of these estimates
+In aggregate, the overreliance on Thurstone's Case V assumptions in the
+statistical analysis of CJ data can compromise the reliability of CJ's
+trait estimates. This overreliance, coupled with the apparent disconnect
+between CJ's approach to trait measurement and hypothesis testing, may
+also undermine the validity of these estimates
 \citep[pp.~2]{Perron_et_al_2015}. Consequently, addressing these issues
-and examining the factors influencing CJ experiments by adopting a more
-systematic and integrated approach could offer several statistical and
-measurement benefits.
-
-\section{Updating the theoretical model}\label{sec-theoretical}
-
-\subsection{The conceptual population
+by adopting a more systematic and integrated approach could offer
+several statistical and measurement benefits.
+
+\section{Extending Thurstone's model}\label{sec-theoretical}
+
+This section addresses the challenges outlined in
+Section~\ref{sec-theory-issues} by leveraging the structural approach to
+causal inference to extend Thurstone's model. The extension combines the
+core theoretical principles from Section~\ref{sec-thurstone_theory} with
+several key assessment design features relevant to CJ experiments, such
+as the selection of judges, stimuli, and comparisons. Specifically,
+Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
+principles into a \emph{conceptual-population model}. Conversely,
+Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
+features into a \emph{sample-comparison model}. Combining these elements
+helps to elucidate how all actors and processes interact during CJ
+experiments.
+
+The \emph{structural approach} to causal inference provides a formal
+framework for identifying causes and estimating their effects using
+data. This approach uses structural causal models (SCMs) and directed
+acyclic graphs (DAGs)
+\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
+formally and graphically represent the assumed causal structure of a
+system, such as the one found in CJ experiments. Essentially, SCMs and
+DAGs function as \emph{conceptual models} on which identification
+analysis rests \citep[pp.~4]{Schuessler_et_al_2023}. Identification
+analysis helps researchers determine whether an estimator can accurately
+compute an estimand based solely on its (causal) assumptions, regardless
+of random variability \citep[pp.~4]{Schuessler_et_al_2023}. Here,
+\emph{estimands} represent the specific quantities researchers aim to
+determine \citep{Everitt_et_al_2010}. \emph{Estimators} denote the
+methods or functions that transform data into an estimate
+\citep{Neal_2020}, while \emph{estimates} are the numerical values
+approximating the estimand. These estimates are obtained through
+\emph{estimation}, which is the process that integrates data with an
+estimator \citep{Everitt_et_al_2010}.
+
+For instance, researchers might aim to answer the question: ``To what
+extent do different teaching methods influence students' ability to
+produce high-quality written texts?'' To investigate this, they could
+design a CJ experiment by randomly assigning students to two groups,
+each exposed to a different teaching method. Afterward, judges would
+compare pairs of written texts, generating a dichotomous outcome that
+reflects the relative quality of each text. In this scenario,
+researchers can rephrase the research question as the estimand:
+``\emph{On average}, is there a difference in the ability to produce
+high-quality written texts between the two groups of students?''.
+Researchers would then use estimates from the BTL model, or its
+transformations, to approximate this estimand.
+
+However, Section~\ref{sec-theory-issues} provides compelling evidence
+that the BTL model suffers from several statistical and measurement
+issues, including identification problems. These issues ultimately raise
+concerns about the consistency of the model's estimates. An estimator is
+\emph{consistent} if it converges to the ``true'' value of an estimand
+as the data size approaches infinity \citep{Everitt_et_al_2010}. Since
+identification is a prerequisite for consistency, the absence of proper
+identification means that researchers cannot achieve consistency, even
+with ``infinite'' and error-free data. As a result, deriving meaningful
+insights from finite data also becomes impossible
+\citep[pp.~5]{Schuessler_et_al_2023}.
+
+Luckily, SCMs and DAGs support identification analysis through two key
+advantages. First, regardless of complexity, they can represent various
+causal structures using only five fundamental building blocks
+\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
+decompose complex structures into manageable components, facilitating
+their analysis \citep{McElreath_2020}. Second, they depict causal
+relationships in an interactive, non-parametric manner. This flexibility
+enables feasible identification strategies without defining the
+variables' data types, the functional forms, or their parameters
+\citep[pp.~35]{Pearl_et_al_2016}.
+
+\subsection{The conceptual-population
 model}\label{sec-theory-theoretical_P}
 
-Assuming population data or more commonly known as census data, we
-\ldots{}
-
-The (latent) discriminal difference of the stimuli directly determines
-the (manifest) outcome of the pairwise comparisons
-
-The (latent) ``perceived'' discriminal processes for the stimuli
-directly determines their discriminal difference
-
-The (latent) ``true'' discriminal processes for the stimuli and the
-judges' biases directly determines their (latent) ``perceived''
-discriminal processes
-
-\begin{figure}
+Before integrating the theoretical principles from
+Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
+DAGs. SCMs are formal mathematical models defined by a set of
+\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
+set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
+variables are those whose causal mechanisms a researcher chooses to
+model \citep{Neal_2020}. In contrast, exogenous variables represent
+\emph{errors} or \emph{disturbances} arising from omitted factors that
+the investigator chooses not to model explicitly
+\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
+\emph{structural equations}, express the endogenous variables as
+non-parametric functions of other variables. These function use the
+symbol `\(:=\)' to denote the asymmetrical causal dependence between
+variables and the symbol `\(\:\bot\:\)' to represent
+\emph{d-separation}, a concept akin to (conditional) independence.
+
+Notably, every SCM has an associated DAG
+\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
+consisting of nodes connected by edges, where the nodes represent random
+variables. The term \emph{directed} means that the edges extend from one
+node to another, with arrows indicating the direction of causal
+influence. The term \emph{acyclic} implies that the causal influences do
+not form loops, ensuring the influences do not cycle back on themselves
+\citep{McElreath_2020}. DAGs represent observed variables as solid black
+circles, while they use open circles for unobserved (latent) variables
+\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
+DAGs typically omits exogenous variables for simplicity, the
+\emph{magnified representation} includes them, offering offering the
+advantage of highlighting potential issues related to conditioning and
+confounding \citep{Cinelli_et_al_2020}.
+
+Figure~\ref{fig-CJ1} displays CJ's outcome, discriminal difference,
+discriminal processes and judges' biases assumed by the Thurstone's
+theory. Notice that in the figure, the structural
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{kjabi_{1}i_{2}} & := f_{O}(D_{kjabi_{1}i_{2}}) \\
+  D_{kjabi_{1}i_{2}} & := f_{D}(PT_{kjai_{1}}, PT_{kjbi_{2}}) \\
+  PT_{kjai_{1}} & := f_{PT}(T_{ai_{1}}, T_{bi_{2}}, B_{kj})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-CJ1_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
 
@@
 
-\caption{\label{fig-CJ_TM_04}}
+\subcaption{\label{fig-CJ1_DAG}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-CJ1}CJ's outcome, discriminal difference,
+discriminal processes and judges' biases}
 
@@
 
-\section{From theory to statistics}\label{sec-statistical}
+\section{Extending the BTL model}\label{sec-statistical}
 
@@
 
-Identification is essential because it is a necessary condition for
-\emph{consistency}. An estimator is \emph{consistent} if it converges to
-the ``true'' value of an estimand as the sample size approaches infinity
-\citep{Everitt_et_al_2010}. Without identification, an estimator cannot
-achieve consistency, rendering it incapable of reliably computing the
-estimand, even with ``infinite'' and error-free data. Consequently,
-deriving meaningful insights about the estimand from finite data becomes
-impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore, to ensure
-accurate and reliable estimates, researchers prioritize estimators with
-desirable identification properties. For instance, the Z-test is a
-widely used estimator for comparing group proportions, yielding accurate
-estimates when its underlying assumptions are satisfied
-\citep{Kanji_2006}. Furthermore, researchers can interpret estimates
-from the Z-test as causal, provided the data is collected through a
-randomized experiment.
+Identification is a necessary condition to ensure \emph{consistent}
+estimators. An estimator achieves \emph{consistency} when it converges
+to the ``true'' value of an estimand as the data size approaches
+infinity \citep{Everitt_et_al_2010}. Without identification, researchers
+cannot achieve consistency, even with ``infinite'' and error-free data.
+As a result, deriving meaningful insights about an estimand from finite
+data becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore,
+to ensure accurate and reliable estimates, researchers prioritize
+estimators with desirable identification properties. For instance, the
+Z-test is a widely used estimator for comparing group proportions,
+yielding accurate estimates when its underlying assumptions are
+satisfied \citep{Kanji_2006}. Furthermore, researchers can interpret
+estimates from the Z-test as causal, provided the data is collected
+through a randomized experiment.
 
@@
 on which identification analysis rests
-\citep[pp.~4]{Schuessler_et_al_2023}. Specifically, they guide
-researchers in determining which statistical models can identify an
-estimand (ACE, CACE, or other), assuming the depicted causal structure
-is correct \citep{McElreath_2020}, thus enabling valid causal inference.
+\citep[pp.~4]{Schuessler_et_al_2023}. Thus, using these tools,
+researchers can determine which statistical models can identify (ACE,
+CACE, or other), assuming the depicted causal structure is correct
+\citep{McElreath_2020}, thus enabling valid causal inference.
 Figure~\ref{fig-IEflow} shows the role of theoretical models in the
@@
 
-SCMs and DAGs support causal inference through two key advantages.
-First, regardless of complexity, they can represent various causal
-structures using only five fundamental building blocks
+SCMs and DAGs support identification analysis through two key
+advantages. First, regardless of complexity, they can represent various
+causal structures using only five fundamental building blocks
 \citep{Neal_2020, McElreath_2020}. This feature allows researchers to

### 2025-01-29 paper2.qmd
section:theoretical; process:write

@@
 Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
-This study discusses two prominent issues that can compromise the reliability and validity of CJ's trait estimates.
-The first issue concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-The study then addresses these challenges by adopting a more systematic and integrated approach grounded in causal inference and Bayesian inference methods.
+This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates.
+The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
+The study then addresses these issues by extending Thurstone's model through a systematic and integrated approach based on causal and Bayesian inference methods.
 
@@
 @sec-theory-issues discusses the identified issues in detail.
-@sec-theoretical updates the Thurstonian model to address these challenges.
+@sec-theoretical extends Thurstone's model to address these challenges.
 The model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
@@
 
-This section discusses the two prominent issues that can undermine the reliability and validity of CJ's trait estimates.
-@sec-theory-issue1 examines the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+This section identifies and discusses the two prominent issues that can undermine the reliability and validity of CJ's trait estimates.
+@sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
 @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
@@
 
-<!-- 1. What the section will do? -->
-This section addresses the challenges outlined in @sec-theory-issues by leveraging the structural approach to causal inference to extend Thurstone's model.
-The extension combines the core theoretical principles from @sec-thurstone_theory with several key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-Specifically, @sec-theory-theoretical_P incorporates the theoretical principles into a *conceptual-population model*.
-Conversely, @sec-theory-theoretical_SC integrates the assessment design features into a *sample-comparison model*.
-Combining these elements helps to elucidate how all actors and processes interact during CJ experiments.
+<!-- 1. The structural approach to causal inference -->
+The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data.
+The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments.
+
+<!-- 2. How does SCMs and DAGs are used? -->
+Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
+4].
+*Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+4].
+Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010].
+*Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].
+
+<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+4].
+-->
+
+<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
+For instance, researchers might aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each receiving a different teaching method.
+Judges would then compare pairs of students' written texts, producing a dichotomous outcome that reflects the relative quality of each text.
+In this scenario, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
+Researchers would then use estimates from the BTL model, or its transformations, to approximate this estimand.
+
+<!-- 4. But why identification is so important? -->
+However, @sec-theory-issues presents strong evidence that the BTL model has several statistical and measurement limitations.
+These limitations hinder the model's ability to identify distinct estimands relevant to CJ inquiries, including the one presented in the previous paragraph.
+Identification is crucial because it is a necessary condition for ensuring consistent estimators.
+Without identification, researchers cannot achieve consistency even with "infinite" and error-free data.
+In other words, the estimates derived from the estimator fail to converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010].
+As a result, deriving meaningful insights from finite data also becomes unattainable [@Schuessler_et_al_2023, pp.
+5].
+
+<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+5].
+-->
+
+<!-- 5. So, how do SCMs and DAGs support identification? -->
+Luckily, SCMs and DAGs support identification analysis through two key advantages.
+First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
+Second, they depict causal relationships in an interactive, non-parametric way.
+This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp.
+35].
+
+<!-- 5. What the section will do? -->
+As a result, this section addresses the identified limitations by extending Thurstone's model through a systematic and integrated approach based on the structural approach to causal inference.
+Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments.
+These features include the selection of judges, stimuli, and comparisons.
+Combining these components clarifies the interactions among all actors and processes involved in CJ experiments.
+Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into a *conceptual-population model*.
+Meanwhile, @sec-theory-theoretical_SC integrates the assessment design features into a *sample-comparison model*.
 
@@
 <!-- url: https://www.indeed.com/career-advice/career-development/theoretical-model -->
-
-<!-- 2. The structural approach to causal inference -->
-The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data.
-This approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments.
-Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
-4].
-Identification analysis helps researchers determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
-4].
-Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010].
-*Estimators* denote the methods or functions that transform data into an estimate [@Neal_2020], while *estimates* are the numerical values approximating the estimand.
-These estimates are obtained through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010].
-
-<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
-Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
-4].
--->
-
-<!-- 4. example of an estimand and an estimator in CJ -->
-For instance, researchers might aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each exposed to a different teaching method.
-Afterward, judges would compare pairs of written texts, generating a dichotomous outcome that reflects the relative quality of each text.
-In this scenario, researchers can rephrase the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-Researchers would then use estimates from the BTL model, or its transformations, to approximate this estimand.
-
-<!-- 5. But the estimator does not satisfy minimum identification properties -->
-However, @sec-theory-issues provides compelling evidence that the BTL model suffers from several statistical and measurement issues, including identification problems.
-These issues ultimately raise concerns about the consistency of the model's estimates.
-An estimator is *consistent* if it converges to the "true" value of an estimand as the data size approaches infinity [@Everitt_et_al_2010].
-Since identification is a prerequisite for consistency, the absence of proper identification means that researchers cannot achieve consistency, even with "infinite" and error-free data.
-As a result, deriving meaningful insights from finite data also becomes impossible [@Schuessler_et_al_2023, pp.
-5].
-
-<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
-So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
-5].
--->
-
-<!-- 5. So, how do SCMs and DAGs support identification? -->
-Luckily, SCMs and DAGs support identification analysis through two key advantages.
-First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
-Second, they depict causal relationships in an interactive, non-parametric manner.
-This flexibility enables feasible identification strategies without defining the variables' data types, the functional forms, or their parameters [@Pearl_et_al_2016, pp.
-35].
-
-<!-- See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section. -->
 
@@
 Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
-These function use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+
+For instance, @fig-CJ1_scm display the SCM relating the CJ outcome $(O_{kjabi_{1}i_{2}})$, discriminal difference $(D_{kjabi_{1}i_{2}})$, perceived discriminal processes for each text within each student $(PT_{kjai_{1}}, PT_{kjbi_{2}})$, the "true" discriminal processes for each text within each student $(T_{kjai_{1}}, T_{kjbi_{2}})$, and the judges' biases $(B_{kj})$.
+Notice that the structural equations use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
 
@@
 <!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
-
-<!-- 3. The begining -->
-@fig-CJ1 displays CJ's outcome, discriminal difference, discriminal processes and judges' biases assumed by the Thurstone's theory. Notice that in the figure, the structural
 
@@
 
+<!-- These estimates are obtained through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010]. -->
+
 <!-- 2. What are the benefits of Bayesian inference?-->

### 2025-01-29 _tex/paper2.tex
Built site for gh-pages

@@
 this domain remains unsystematic and fragmented, leaving several
-critical issues unresolved. This study discusses two prominent issues
-that can compromise the reliability and validity of CJ's trait
-estimates. The first issue concerns the overreliance on Thurstone's Case
-V assumptions in the statistical analysis of CJ data. The second
-involves the apparent disconnect between CJ's approach to trait
-measurement and hypothesis testing. The study then addresses these
-challenges by adopting a more systematic and integrated approach
-grounded in causal inference and Bayesian inference methods.
+critical issues unresolved. This study identifies and discusses two
+prominent issues that can undermine the reliability and validity of CJ's
+trait estimates. The first issue arises from the excessive reliance on
+Thurstone's Case V assumptions in the statistical analysis of CJ data.
+The second stems from the apparent disconnect between CJ's approach to
+trait measurement and hypothesis testing. The study then addresses these
+issues by extending Thurstone's model through a systematic and
+integrated approach based on causal and Bayesian inference methods.
 
@@
 theory. Section~\ref{sec-theory-issues} discusses the identified issues
-in detail. Section~\ref{sec-theoretical} updates the Thurstonian model
-to address these challenges. The model integrates the theoretical core
+in detail. Section~\ref{sec-theoretical} extends Thurstone's model to
+address these challenges. The model integrates the theoretical core
 principles alongside key assessment design features relevant to CJ
@@
 
-This section discusses the two prominent issues that can undermine the
-reliability and validity of CJ's trait estimates.
-Section~\ref{sec-theory-issue1} examines the overreliance on Thurstone's
-Case V assumptions in the statistical analysis of CJ data.
+This section identifies and discusses the two prominent issues that can
+undermine the reliability and validity of CJ's trait estimates.
+Section~\ref{sec-theory-issue1} examines the the excessive reliance on
+Thurstone's Case V assumptions in the statistical analysis of CJ data.
 Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
@@
 
-This section addresses the challenges outlined in
-Section~\ref{sec-theory-issues} by leveraging the structural approach to
-causal inference to extend Thurstone's model. The extension combines the
-core theoretical principles from Section~\ref{sec-thurstone_theory} with
-several key assessment design features relevant to CJ experiments, such
-as the selection of judges, stimuli, and comparisons. Specifically,
-Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
-principles into a \emph{conceptual-population model}. Conversely,
-Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
-features into a \emph{sample-comparison model}. Combining these elements
-helps to elucidate how all actors and processes interact during CJ
-experiments.
-
 The \emph{structural approach} to causal inference provides a formal
 framework for identifying causes and estimating their effects using
-data. This approach uses structural causal models (SCMs) and directed
+data. The approach uses structural causal models (SCMs) and directed
 acyclic graphs (DAGs)
@@
 formally and graphically represent the assumed causal structure of a
-system, such as the one found in CJ experiments. Essentially, SCMs and
-DAGs function as \emph{conceptual models} on which identification
-analysis rests \citep[pp.~4]{Schuessler_et_al_2023}. Identification
-analysis helps researchers determine whether an estimator can accurately
-compute an estimand based solely on its (causal) assumptions, regardless
-of random variability \citep[pp.~4]{Schuessler_et_al_2023}. Here,
-\emph{estimands} represent the specific quantities researchers aim to
-determine \citep{Everitt_et_al_2010}. \emph{Estimators} denote the
-methods or functions that transform data into an estimate
-\citep{Neal_2020}, while \emph{estimates} are the numerical values
-approximating the estimand. These estimates are obtained through
-\emph{estimation}, which is the process that integrates data with an
-estimator \citep{Everitt_et_al_2010}.
+system, such as the one found in CJ experiments.
+
+Essentially, SCMs and DAGs function as \emph{conceptual models} on which
+identification analysis rests \citep[pp.~4]{Schuessler_et_al_2023}.
+\emph{Identification analysis} helps researchers to determine whether an
+estimator can accurately compute an estimand based solely on its
+(causal) assumptions, regardless of random variability
+\citep[pp.~4]{Schuessler_et_al_2023}. Here, \emph{estimands} represent
+the specific quantities researchers aim to determine
+\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
+functions that transform data into an estimate, while \emph{estimates}
+are the numerical values approximating the estimand
+\citep{Neal_2020, Everitt_et_al_2010}.
 
@@
 design a CJ experiment by randomly assigning students to two groups,
-each exposed to a different teaching method. Afterward, judges would
-compare pairs of written texts, generating a dichotomous outcome that
+each receiving a different teaching method. Judges would then compare
+pairs of students' written texts, producing a dichotomous outcome that
 reflects the relative quality of each text. In this scenario,
-researchers can rephrase the research question as the estimand:
+researchers can reformulate the research question as the estimand:
 ``\emph{On average}, is there a difference in the ability to produce
@@
 
-However, Section~\ref{sec-theory-issues} provides compelling evidence
-that the BTL model suffers from several statistical and measurement
-issues, including identification problems. These issues ultimately raise
-concerns about the consistency of the model's estimates. An estimator is
-\emph{consistent} if it converges to the ``true'' value of an estimand
-as the data size approaches infinity \citep{Everitt_et_al_2010}. Since
-identification is a prerequisite for consistency, the absence of proper
-identification means that researchers cannot achieve consistency, even
-with ``infinite'' and error-free data. As a result, deriving meaningful
-insights from finite data also becomes impossible
+However, Section~\ref{sec-theory-issues} presents strong evidence that
+the BTL model has several statistical and measurement limitations. These
+limitations hinder the model's ability to identify distinct estimands
+relevant to CJ inquiries, including the one presented in the previous
+paragraph. Identification is crucial because it is a necessary condition
+for ensuring consistent estimators. Without identification, researchers
+cannot achieve consistency even with ``infinite'' and error-free data.
+In other words, the estimates derived from the estimator fail to
+converge to the ``true'' value of the estimand as the data size
+approaches infinity \citep{Everitt_et_al_2010}. As a result, deriving
+meaningful insights from finite data also becomes unattainable
 \citep[pp.~5]{Schuessler_et_al_2023}.
@@
 their analysis \citep{McElreath_2020}. Second, they depict causal
-relationships in an interactive, non-parametric manner. This flexibility
-enables feasible identification strategies without defining the
-variables' data types, the functional forms, or their parameters
+relationships in an interactive, non-parametric way. This flexibility
+enables feasible identification strategies without requiring
+specification of the types of variables, the functional forms relating
+them, or the parameters of those functional forms
 \citep[pp.~35]{Pearl_et_al_2016}.
+
+As a result, this section addresses the identified limitations by
+extending Thurstone's model through a systematic and integrated approach
+based on the structural approach to causal inference. Specifically, it
+leverages the capabilities of this approach to formalize the combination
+of the core theoretical principles outlined in
+Section~\ref{sec-thurstone_theory} with key assessment design features
+relevant to CJ experiments. These features include the selection of
+judges, stimuli, and comparisons. Combining these components clarifies
+the interactions among all actors and processes involved in CJ
+experiments. Accordingly, Section~\ref{sec-theory-theoretical_P}
+incorporates the theoretical principles into a
+\emph{conceptual-population model}. Meanwhile,
+Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
+features into a \emph{sample-comparison model}.
 
@@
 \emph{structural equations}, express the endogenous variables as
-non-parametric functions of other variables. These function use the
-symbol `\(:=\)' to denote the asymmetrical causal dependence between
-variables and the symbol `\(\:\bot\:\)' to represent
-\emph{d-separation}, a concept akin to (conditional) independence.
+non-parametric functions of other variables.
+
+For instance, Figure~\ref{fig-CJ1_scm} display the SCM relating the CJ
+outcome \((O_{kjabi_{1}i_{2}})\), discriminal difference
+\((D_{kjabi_{1}i_{2}})\), perceived discriminal processes for each text
+within each student \((PT_{kjai_{1}}, PT_{kjbi_{2}})\), the ``true''
+discriminal processes for each text within each student
+\((T_{kjai_{1}}, T_{kjbi_{2}})\), and the judges' biases \((B_{kj})\).
+Notice that the structural equations use the symbol `\(:=\)' to denote
+the asymmetrical causal dependence between variables and the symbol
+`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
+(conditional) independence.
 
@@
 confounding \citep{Cinelli_et_al_2020}.
-
-Figure~\ref{fig-CJ1} displays CJ's outcome, discriminal difference,
-discriminal processes and judges' biases assumed by the Thurstone's
-theory. Notice that in the figure, the structural
 

### 2025-01-30 paper2.qmd
section:theoretical; process:write

@@
 
-::: {#fig-caseV_issues layout-ncol=2}
+::: {#fig-casev_issues layout-ncol=2}
 
@@
 Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
-
-For instance, @fig-CJ1_scm display the SCM relating the CJ outcome $(O_{kjabi_{1}i_{2}})$, discriminal difference $(D_{kjabi_{1}i_{2}})$, perceived discriminal processes for each text within each student $(PT_{kjai_{1}}, PT_{kjbi_{2}})$, the "true" discriminal processes for each text within each student $(T_{kjai_{1}}, T_{kjbi_{2}})$, and the judges' biases $(B_{kj})$.
-Notice that the structural equations use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+
+For instance, following the scenario outline in @sec-theoretical, @fig-cj04_scm displays the SCM that relates the outcome of the CJ experiment $(O_{abhijk})$, to several endogenous variables $V$.
+These variables include the discriminal difference $(D_{abhijk})$, perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$, formally expressed as $V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$.
+The structural equations $\{f_{O},f_{D},f_{PT}\}$ define the dependencies among these variables.
+Notably, in this scenario, the judges produce a dichotomous outcome $(O_{abhijk}) = \{0,1\}$.
+Moreover, the indices $\{a,b\}$ represent the compared texts, $\{h,i\}$ label the students who wrote these texts, $j$ denotes the judge, and $k$ represents the comparison number.
+Regarding the $k$ index, the index accounts for cases in which judges perform multiple comparisons of the same texts written by the same students, as the one discussed in @sec-theoretical.
 
@@
 
-::: {#fig-CJ1 layout-nrow=2}
-
-::: {#fig-CJ1_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{kjabi_{1}i_{2}} & := f_{O}(D_{kjabi_{1}i_{2}}) \\
-  D_{kjabi_{1}i_{2}} & := f_{D}(PT_{kjai_{1}}, PT_{kjbi_{2}}) \\
-  PT_{kjai_{1}} & := f_{PT}(T_{ai_{1}}, T_{bi_{2}}, B_{kj})
+<!-- ![](/images/png/CJ_TM_01.png){#fig-cj01_dag fig-align="center" fig-pos="H" width=23%} -->
+
+<!-- ![](/images/png/CJ_TM_02.png){#fig-cj02_dag fig-align="center" fig-pos="H" width=32%} -->
+
+<!-- ![](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=45%} -->
+
+::: {#fig-cj04 layout-nrow=2}
+
+::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{abhijk} & := f_{O}(D_{abhijk}) \\
+  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
+  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
+  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
 \end{aligned}
@@
 
-![DAG](/images/png/CJ_TM_04.png){#fig-CJ1_DAG fig-align="center" fig-pos="H" width=45%}
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=45%}
 
@@
 :::
+
+Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in
 
@@
 
-<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->
-
-<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->
-
-<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->
-
-without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in
-
-![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%}
-
-<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->
-
-<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->
-
-![](/images/png/CJ_TM_10.png){#fig-CJ_TM_10 width=68%}
+::: {#fig-cj05 layout-nrow=2}
+
+::: {#fig-cj05_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_05.png){#fig-cj05_dag fig-align="center" fig-pos="H" width=44%}
+
+CJ's outcome, discriminal difference, discriminal processes and judges' biases
+:::
+
+<!-- ![](/images/png/CJ_TM_06.png){#fig-cj06_dag fig-align="center" fig-pos="H" width=59%} -->
+
+<!-- ![](/images/png/CJ_TM_07.png){#fig-cj07_dag fig-align="center" fig-pos="H" width=68%} -->
+
+<!-- ![](/images/png/CJ_TM_08.png){#fig-cj08_dag fig-align="center" fig-pos="H" width=68%} -->
+
+<!-- ![](/images/png/CJ_TM_09.png){#fig-cj09_dag fig-align="center" fig-pos="H" width=68%} -->
+
+::: {#fig-cj10 layout-nrow=2}
+
+::: {#fig-cj10_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_10.png){#fig-cj10_dag fig-align="center" fig-pos="H" width=68%}
+
+CJ's outcome, discriminal difference, discriminal processes and judges' biases
+:::
 
@@
 
-<!-- ![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%} -->
-
-![](/images/png/CJ_TM_12.png){#fig-CJ_TM_12 width=80%}
+<!-- ![](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=80%} -->
+
+::: {#fig-cj12 layout-nrow=2}
+
+::: {#fig-cj12_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_12.png){#fig-cj12_dag fig-align="center" fig-pos="H" width=80%}
+
+CJ's outcome, discriminal difference, discriminal processes and judges' biases
+:::
 
@@
 
-<!-- ![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%} -->
-
-![](/images/png/CJ_TM_14.png){#fig-CJ_TM_14 width=90%}
-
-![](/images/png/CJ_TM_15.png){#fig-CJ_TM_15 width=90%}
+<!-- ![](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=90%} -->
+
+::: {#fig-cj14 layout-nrow=2}
+
+::: {#fig-cj14_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
+  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_14.png){#fig-cj14_dag fig-align="center" fig-pos="H" width=90%}
+
+CJ's outcome, discriminal difference, discriminal processes and judges' biases
+:::
+
+::: {#fig-cj15 layout-nrow=2}
+
+::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
+  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_15.png){#fig-cj15_dag fig-align="center" fig-pos="H" width=90%}
+
+CJ's outcome, discriminal difference, discriminal processes and judges' biases
+:::
 

### 2025-01-30 _tex/paper2.tex
Built site for gh-pages

@@
 
-\caption{\label{fig-caseV_issues}The effect of dispersion discrepancies
+\caption{\label{fig-casev_issues}The effect of dispersion discrepancies
 and stimulus correlation on the distribution of the discriminal
@@
 \emph{structural equations}, express the endogenous variables as
-non-parametric functions of other variables.
-
-For instance, Figure~\ref{fig-CJ1_scm} display the SCM relating the CJ
-outcome \((O_{kjabi_{1}i_{2}})\), discriminal difference
-\((D_{kjabi_{1}i_{2}})\), perceived discriminal processes for each text
-within each student \((PT_{kjai_{1}}, PT_{kjbi_{2}})\), the ``true''
-discriminal processes for each text within each student
-\((T_{kjai_{1}}, T_{kjbi_{2}})\), and the judges' biases \((B_{kj})\).
-Notice that the structural equations use the symbol `\(:=\)' to denote
-the asymmetrical causal dependence between variables and the symbol
-`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
-(conditional) independence.
+non-parametric functions of other variables. These functions use the
+symbol `\(:=\)' to denote the asymmetrical causal dependence between
+variables and the symbol `\(\:\bot\:\)' to represent
+\emph{d-separation}, a concept akin to (conditional) independence.
+
+For instance, following the scenario outline in
+Section~\ref{sec-theoretical}, Figure~\ref{fig-cj04_scm} displays the
+SCM that relates the outcome of the CJ experiment \((O_{abhijk})\), to
+several endogenous variables \(V\). These variables include the
+discriminal difference \((D_{abhijk})\), perceived discriminal processes
+\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
+\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\), formally
+expressed as
+\(V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
+The structural equations \(\{f_{O},f_{D},f_{PT}\}\) define the
+dependencies among these variables. Notably, in this scenario, the
+judges produce a dichotomous outcome \((O_{abhijk}) = \{0,1\}\).
+Moreover, the indices \(\{a,b\}\) represent the compared texts,
+\(\{h,i\}\) label the students who wrote these texts, \(j\) denotes the
+judge, and \(k\) represents the comparison number. Regarding the \(k\)
+index, the index accounts for cases in which judges perform multiple
+comparisons of the same texts written by the same students, as the one
+discussed in Section~\ref{sec-theoretical}.
 
@@
 \begin{aligned}
-  O_{kjabi_{1}i_{2}} & := f_{O}(D_{kjabi_{1}i_{2}}) \\
-  D_{kjabi_{1}i_{2}} & := f_{D}(PT_{kjai_{1}}, PT_{kjbi_{2}}) \\
-  PT_{kjai_{1}} & := f_{PT}(T_{ai_{1}}, T_{bi_{2}}, B_{kj})
+  O_{abhijk} & := f_{O}(D_{abhijk}) \\
+  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
+  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
+  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
 \end{aligned}
@@
 
-\subcaption{\label{fig-CJ1_scm}SCM}
+\subcaption{\label{fig-cj04_scm}SCM}
 
@@
 
-\subcaption{\label{fig-CJ1_DAG}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-CJ1}CJ's outcome, discriminal difference,
+\subcaption{\label{fig-cj04_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
 discriminal processes and judges' biases}
@@
 
-without loosing generality, the (latent) ``perceived'' and ``true''
-discriminal processes for the stimuli can be depicted in a vector for
-each judge, as in
-
-\begin{figure}
+Because it quickly becomes intractable, and without loosing generality,
+the (latent) ``perceived'' and ``true'' discriminal processes for the
+stimuli can be depicted in a vector for each judge, as in
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj05_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
 
@@
 
-\caption{\label{fig-CJ_TM_05}}
+\subcaption{\label{fig-cj05_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj05}CJ's outcome, discriminal difference,
+discriminal processes and judges' biases}
 
@@
 
-\begin{figure}
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj10_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
 
@@
 
-\caption{\label{fig-CJ_TM_10}}
+\subcaption{\label{fig-cj10_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj10}CJ's outcome, discriminal difference,
+discriminal processes and judges' biases}
 
@@
 
-\begin{figure}
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj12_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
 
@@
 
-\caption{\label{fig-CJ_TM_12}}
+\subcaption{\label{fig-cj12_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj12}CJ's outcome, discriminal difference,
+discriminal processes and judges' biases}
 
@@
 
-\begin{figure}
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
+  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj14_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
 
@@
 
-\caption{\label{fig-CJ_TM_14}}
+\subcaption{\label{fig-cj14_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj14}CJ's outcome, discriminal difference,
+discriminal processes and judges' biases}
 
@@
 
-\begin{figure}
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
+  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O_{PRQ} & := f_{O}(D_{PRQ}) \\
+  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
+  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
+  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj15_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
 
@@
 
-\caption{\label{fig-CJ_TM_15}}
+\subcaption{\label{fig-cj15_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj15}CJ's outcome, discriminal difference,
+discriminal processes and judges' biases}
 

### 2025-01-31 paper2.qmd
section:theoretical; process:write,scms,dags

@@
 The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-The study then addresses these issues by extending Thurstone's model through a systematic and integrated approach based on causal and Bayesian inference methods.
+The study then addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on causal and Bayesian inference methods.
 
@@
 @sec-theory-issues discusses the identified issues in detail.
-@sec-theoretical extends Thurstone's model to address these challenges.
-The model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+@sec-theoretical extends Thurstone's general form to address these challenges.
+The new model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
 @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of pairwise comparison data.
@@
 Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
-In this scenario, @fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
+@fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
 This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome.
@@
 <!-- 4. segway for next section -->
-In aggregate, the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates.
+In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates.
 This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp.
 2].
-Consequently, addressing these issues by adopting a more systematic and integrated approach could offer several statistical and measurement benefits.
+Thus, addressing these issues through a more systematic and integrated approach could provide significant statistical and measurement benefits.
+For this reason, the study employs the structural approach to causal inference, as it fulfills these requirements.
 
@@
 <!-- recording: Sven and Tine 25.01.22; time: 00:08:25 - 00:31:56 -->
+<!-- recording: Sven and Tine 25.01.30 p1; time: 00:00:00 - 00:23:25 -->
 
@@
 
+<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
+
 <!-- 3. Example of a CJ estimand and estimator, and its limitations -->
-For instance, researchers might aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each receiving a different teaching method.
+For instance, a motivating example that illustrates these concepts throughout the document could involve researchers aiming to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each receiving a different teaching method.
 Judges would then compare pairs of students' written texts, producing a dichotomous outcome that reflects the relative quality of each text.
-In this scenario, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-Researchers would then use estimates from the BTL model, or its transformations, to approximate this estimand.
+Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
+Following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
 
 <!-- 4. But why identification is so important? -->
-However, @sec-theory-issues presents strong evidence that the BTL model has several statistical and measurement limitations.
-These limitations hinder the model's ability to identify distinct estimands relevant to CJ inquiries, including the one presented in the previous paragraph.
+However, @sec-theory-issues presents compelling evidence that the BTL model has several statistical and measurement limitations.
+These limitations hinder the model's ability to identify distinct estimands relevant to CJ inquiries, including the one from the motivating example.
 Identification is crucial because it is a necessary condition for ensuring consistent estimators.
-Without identification, researchers cannot achieve consistency even with "infinite" and error-free data.
-In other words, the estimates derived from the estimator fail to converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010].
+Without identification, researchers cannot achieve *consistency* even with "infinite" and error-free data.
+In other words, the estimates derived from the estimator would fail to converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010].
 As a result, deriving meaningful insights from finite data also becomes unattainable [@Schuessler_et_al_2023, pp.
@@
 <!-- 5. What the section will do? -->
-As a result, this section addresses the identified limitations by extending Thurstone's model through a systematic and integrated approach based on the structural approach to causal inference.
-Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments.
-These features include the selection of judges, stimuli, and comparisons.
-Combining these components clarifies the interactions among all actors and processes involved in CJ experiments.
-Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into a *conceptual-population model*.
-Meanwhile, @sec-theory-theoretical_SC integrates the assessment design features into a *sample-comparison model*.
+As a result, this section addresses the identified limitations by extending Thurstone's general form (see @tbl-thurstone_cases) through a systematic and integrated approach based on the structural approach to causal inference.
+Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+The extended model clarifies the interactions among all actors and processes involved in CJ experiments.
+Furthermore, it allows researchers to shift the current comparative data analysis paradigm from accepting the model's assumptions as given to testing these assumptions for specific comparative data.
+For details on these assumptions, refer to @sec-theory-issues.
+Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what the study calls the *conceptual-population model*.
+Meanwhile, @sec-theory-theoretical_SC integrates the assessment design features into what the study terms the *sample-comparison model*.
 
@@
 <!-- recording: Sven and Tine 24.12.17; time: 00:26:10 - 00:54:00 -->
+<!-- recording: Sven and Tine 25.01.30 p1; time: 00:23:25 - 00:48:00 -->
 
 <!-- 1. The SCMs -->
-Before integrating the theoretical principles from @sec-thurstone_theory, it is important to define SCMs and DAGs.
-SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020].
+Before integrating the theoretical principles from @sec-thurstone_theory, the study defines SCMs and DAGs.
+SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 
-For instance, following the scenario outline in @sec-theoretical, @fig-cj04_scm displays the SCM that relates the outcome of the CJ experiment $(O_{abhijk})$, to several endogenous variables $V$.
-These variables include the discriminal difference $(D_{abhijk})$, perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$, formally expressed as $V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$.
-The structural equations $\{f_{O},f_{D},f_{PT}\}$ define the dependencies among these variables.
-Notably, in this scenario, the judges produce a dichotomous outcome $(O_{abhijk}) = \{0,1\}$.
-Moreover, the indices $\{a,b\}$ represent the compared texts, $\{h,i\}$ label the students who wrote these texts, $j$ denotes the judge, and $k$ represents the comparison number.
-Regarding the $k$ index, the index accounts for cases in which judges perform multiple comparisons of the same texts written by the same students, as the one discussed in @sec-theoretical.
-
-<!-- 2. The DAGs -->
+<!-- 2. the preliminary SCM for CJ -->
+@fig-cj04 illustrates the preliminary integration of the theoretical principles underlying the motivating example.
+Specifically, [SCM -@fig-cj04_scm] displays the relationship between the outcome of the CJ experiment, $O_{abhijk}$, to several other variables.
+Here, the indices $\{a,b\}$ represent the compared texts, $\{h,i\}$ label the students who wrote these texts, $j$ denotes the judge, and $k$ accounts for cases where the judges perform multiple comparisons of the same texts written by the same students, i.e., the comparison number for a *repeated measures design* [@Lawson_2015, pp.
+366-376].
+The other variables related to the outcome include the discriminal difference $(D_{abhijk})$, the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$.
+Together with the outcome, these variables form the preliminary set of endogenous variables $V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$.
+Finally, the preliminary set of structural equations $F = \{f_{O},f_{D},f_{PT}\}$ defines the dependencies among these variables.
+
+<!-- ![](/images/png/CJ_TM_01.png){#fig-cj01_dag fig-align="center" fig-pos="H" width=23%} -->
+
+<!-- ![](/images/png/CJ_TM_02.png){#fig-cj02_dag fig-align="center" fig-pos="H" width=32%} -->
+
+<!-- ![](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=45%} -->
+
+::: {#fig-cj04 layout-nrow=2}
+
+::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{abhijk} & := f_{O}(D_{abhijk}) \\
+  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
+  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
+  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=45%}
+
+CJ's outcome, discriminal difference, discriminal processes and judges' biases
+:::
+
+<!-- 3. The DAGs -->
 Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
@@
 DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
-
-<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
+While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+
+<!-- 4. the preliminary DAG for CJ -->
+[DAG -@fig-cj04_dag] displays the preliminary DAG associated with [SCM -@fig-cj04_scm].
+The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory.
+It also distinguishes between observable endogenous variables, such as the outcome $O_{abhijk}$, and variables considered latent according to the theory and practice of CJ, such as the discriminal difference $(D_{abhijk})$, the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$.
+
+<!-- Notice,  -->
+<!-- the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$ -->
 
@@
 
-<!-- The (latent) discriminal difference of the stimuli directly determines the (manifest) outcome of the pairwise comparisons -->
-
-<!-- The (latent) "perceived" discriminal processes for the stimuli directly determines their discriminal difference -->
-
-<!-- The (latent) "true" discriminal processes for the stimuli and the judges' biases directly determines their (latent) "perceived" discriminal processes -->
-
-<!-- ![](/images/png/CJ_TM_01.png){#fig-cj01_dag fig-align="center" fig-pos="H" width=23%} -->
-
-<!-- ![](/images/png/CJ_TM_02.png){#fig-cj02_dag fig-align="center" fig-pos="H" width=32%} -->
-
-<!-- ![](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=45%} -->
-
-::: {#fig-cj04 layout-nrow=2}
-
-::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{abhijk} & := f_{O}(D_{abhijk}) \\
-  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
-  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
-  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=45%}
-
-CJ's outcome, discriminal difference, discriminal processes and judges' biases
-:::
-
-Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in
+<!-- Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in   -->
 
@@
 In this context, the variable $X$ (socio-economic status of the school) is thought to be a confounder in the relationship between the teaching method $T$ and the outcome $Y$.
-In this scenario, the figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
+The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
 They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.

### 2025-01-31 _tex/paper2.tex
Built site for gh-pages

@@
 trait measurement and hypothesis testing. The study then addresses these
-issues by extending Thurstone's model through a systematic and
+issues by extending Thurstone's general form through a systematic and
 integrated approach based on causal and Bayesian inference methods.
@@
 theory. Section~\ref{sec-theory-issues} discusses the identified issues
-in detail. Section~\ref{sec-theoretical} extends Thurstone's model to
-address these challenges. The model integrates the theoretical core
-principles alongside key assessment design features relevant to CJ
-experiments, such as the selection of judges, stimuli, and comparisons.
-Section~\ref{sec-statistical} translates these theoretical and practical
-elements into a probabilistic statistical model for the analysis of
-pairwise comparison data. Section~\ref{sec-discussion} discusses the
-implications of the findings and explores avenues for future research.
-Finally, Section~\ref{sec-conclusion} summarizes the study's key
-insights.
+in detail. Section~\ref{sec-theoretical} extends Thurstone's general
+form to address these challenges. The new model integrates the
+theoretical core principles alongside key assessment design features
+relevant to CJ experiments, such as the selection of judges, stimuli,
+and comparisons. Section~\ref{sec-statistical} translates these
+theoretical and practical elements into a probabilistic statistical
+model for the analysis of pairwise comparison data.
+Section~\ref{sec-discussion} discusses the implications of the findings
+and explores avenues for future research. Finally,
+Section~\ref{sec-conclusion} summarizes the study's key insights.
 
@@
 assumes that the discriminal dispersion for Text A remains constant and
-that the texts are uncorrelated \((\rho=0)\). In this scenario,
+that the texts are uncorrelated \((\rho=0)\).
 Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
@@
 
-In aggregate, the overreliance on Thurstone's Case V assumptions in the
-statistical analysis of CJ data can compromise the reliability of CJ's
-trait estimates. This overreliance, coupled with the apparent disconnect
-between CJ's approach to trait measurement and hypothesis testing, may
-also undermine the validity of these estimates
-\citep[pp.~2]{Perron_et_al_2015}. Consequently, addressing these issues
-by adopting a more systematic and integrated approach could offer
-several statistical and measurement benefits.
+In aggregate, the excessive reliance on Thurstone's Case V assumptions
+in the statistical analysis of CJ data can compromise the reliability of
+CJ's trait estimates. This overreliance, coupled with the apparent
+disconnect between CJ's approach to trait measurement and hypothesis
+testing, may also undermine the validity of these estimates
+\citep[pp.~2]{Perron_et_al_2015}. Thus, addressing these issues through
+a more systematic and integrated approach could provide significant
+statistical and measurement benefits. For this reason, the study employs
+the structural approach to causal inference, as it fulfills these
+requirements.
 
@@
 
-For instance, researchers might aim to answer the question: ``To what
-extent do different teaching methods influence students' ability to
-produce high-quality written texts?'' To investigate this, they could
-design a CJ experiment by randomly assigning students to two groups,
-each receiving a different teaching method. Judges would then compare
-pairs of students' written texts, producing a dichotomous outcome that
-reflects the relative quality of each text. In this scenario,
-researchers can reformulate the research question as the estimand:
-``\emph{On average}, is there a difference in the ability to produce
-high-quality written texts between the two groups of students?''.
-Researchers would then use estimates from the BTL model, or its
-transformations, to approximate this estimand.
-
-However, Section~\ref{sec-theory-issues} presents strong evidence that
-the BTL model has several statistical and measurement limitations. These
-limitations hinder the model's ability to identify distinct estimands
-relevant to CJ inquiries, including the one presented in the previous
-paragraph. Identification is crucial because it is a necessary condition
-for ensuring consistent estimators. Without identification, researchers
-cannot achieve consistency even with ``infinite'' and error-free data.
-In other words, the estimates derived from the estimator fail to
-converge to the ``true'' value of the estimand as the data size
-approaches infinity \citep{Everitt_et_al_2010}. As a result, deriving
-meaningful insights from finite data also becomes unattainable
-\citep[pp.~5]{Schuessler_et_al_2023}.
+For instance, a motivating example that illustrates these concepts
+throughout the document could involve researchers aiming to answer the
+question: ``To what extent do different teaching methods influence
+students' ability to produce high-quality written texts?'' To
+investigate this, they could design a CJ experiment by randomly
+assigning students to two groups, each receiving a different teaching
+method. Judges would then compare pairs of students' written texts,
+producing a dichotomous outcome that reflects the relative quality of
+each text. Based on this setup, researchers can reformulate the research
+question as the estimand: ``\emph{On average}, is there a difference in
+the ability to produce high-quality written texts between the two groups
+of students?''. Following current CJ practices, researchers would then
+rely on estimates from the BTL model, or its transformations, to
+approximate this estimand.
+
+However, Section~\ref{sec-theory-issues} presents compelling evidence
+that the BTL model has several statistical and measurement limitations.
+These limitations hinder the model's ability to identify distinct
+estimands relevant to CJ inquiries, including the one from the
+motivating example. Identification is crucial because it is a necessary
+condition for ensuring consistent estimators. Without identification,
+researchers cannot achieve \emph{consistency} even with ``infinite'' and
+error-free data. In other words, the estimates derived from the
+estimator would fail to converge to the ``true'' value of the estimand
+as the data size approaches infinity \citep{Everitt_et_al_2010}. As a
+result, deriving meaningful insights from finite data also becomes
+unattainable \citep[pp.~5]{Schuessler_et_al_2023}.
 
@@
 As a result, this section addresses the identified limitations by
-extending Thurstone's model through a systematic and integrated approach
-based on the structural approach to causal inference. Specifically, it
-leverages the capabilities of this approach to formalize the combination
-of the core theoretical principles outlined in
-Section~\ref{sec-thurstone_theory} with key assessment design features
-relevant to CJ experiments. These features include the selection of
-judges, stimuli, and comparisons. Combining these components clarifies
-the interactions among all actors and processes involved in CJ
-experiments. Accordingly, Section~\ref{sec-theory-theoretical_P}
-incorporates the theoretical principles into a
-\emph{conceptual-population model}. Meanwhile,
-Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
-features into a \emph{sample-comparison model}.
+extending Thurstone's general form (see Table~\ref{tbl-thurstone_cases})
+through a systematic and integrated approach based on the structural
+approach to causal inference. Specifically, it leverages the
+capabilities of this approach to formalize the combination of the core
+theoretical principles outlined in Section~\ref{sec-thurstone_theory}
+with key assessment design features relevant to CJ experiments, such as
+the selection of judges, stimuli, and comparisons. The extended model
+clarifies the interactions among all actors and processes involved in CJ
+experiments. Furthermore, it allows researchers to shift the current
+comparative data analysis paradigm from accepting the model's
+assumptions as given to testing these assumptions for specific
+comparative data. For details on these assumptions, refer to
+Section~\ref{sec-theory-issues}. Accordingly,
+Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
+principles into what the study calls the \emph{conceptual-population
+model}. Meanwhile, Section~\ref{sec-theory-theoretical_SC} integrates
+the assessment design features into what the study terms the
+\emph{sample-comparison model}.
 
@@
 Before integrating the theoretical principles from
-Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
-DAGs. SCMs are formal mathematical models defined by a set of
-\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
-set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
-variables are those whose causal mechanisms a researcher chooses to
-model \citep{Neal_2020}. In contrast, exogenous variables represent
-\emph{errors} or \emph{disturbances} arising from omitted factors that
-the investigator chooses not to model explicitly
-\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
-\emph{structural equations}, express the endogenous variables as
-non-parametric functions of other variables. These functions use the
-symbol `\(:=\)' to denote the asymmetrical causal dependence between
-variables and the symbol `\(\:\bot\:\)' to represent
-\emph{d-separation}, a concept akin to (conditional) independence.
-
-For instance, following the scenario outline in
-Section~\ref{sec-theoretical}, Figure~\ref{fig-cj04_scm} displays the
-SCM that relates the outcome of the CJ experiment \((O_{abhijk})\), to
-several endogenous variables \(V\). These variables include the
-discriminal difference \((D_{abhijk})\), perceived discriminal processes
+Section~\ref{sec-thurstone_theory}, the study defines SCMs and DAGs.
+SCMs are formal mathematical models characterized by a set of
+\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
+\(E\), and a set of functions \(F\)
+\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
+whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
+In contrast, exogenous variables represent \emph{errors} or
+\emph{disturbances} arising from omitted factors that the investigator
+chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
+the functions, referred to as \emph{structural equations}, express the
+endogenous variables as non-parametric functions of other variables.
+These functions use the symbol `\(:=\)' to denote the asymmetrical
+causal dependence between variables and the symbol `\(\:\bot\:\)' to
+represent \emph{d-separation}, a concept akin to (conditional)
+independence.
+
+Figure~\ref{fig-cj04} illustrates the preliminary integration of the
+theoretical principles underlying the motivating example. Specifically,
+SCM~\ref{fig-cj04_scm} displays the relationship between the outcome of
+the CJ experiment, \(O_{abhijk}\), to several other variables. Here, the
+indices \(\{a,b\}\) represent the compared texts, \(\{h,i\}\) label the
+students who wrote these texts, \(j\) denotes the judge, and \(k\)
+accounts for cases where the judges perform multiple comparisons of the
+same texts written by the same students, i.e., the comparison number for
+a \emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}. The
+other variables related to the outcome include the discriminal
+difference \((D_{abhijk})\), the perceived discriminal processes
 \((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
-\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\), formally
-expressed as
-\(V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
-The structural equations \(\{f_{O},f_{D},f_{PT}\}\) define the
-dependencies among these variables. Notably, in this scenario, the
-judges produce a dichotomous outcome \((O_{abhijk}) = \{0,1\}\).
-Moreover, the indices \(\{a,b\}\) represent the compared texts,
-\(\{h,i\}\) label the students who wrote these texts, \(j\) denotes the
-judge, and \(k\) represents the comparison number. Regarding the \(k\)
-index, the index accounts for cases in which judges perform multiple
-comparisons of the same texts written by the same students, as the one
-discussed in Section~\ref{sec-theoretical}.
+\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\). Together
+with the outcome, these variables form the preliminary set of endogenous
+variables
+\(V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
+Finally, the preliminary set of structural equations
+\(F = \{f_{O},f_{D},f_{PT}\}\) defines the dependencies among these
+variables.
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{abhijk} & := f_{O}(D_{abhijk}) \\
+  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
+  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
+  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj04_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+
+}
+
+\subcaption{\label{fig-cj04_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
+discriminal processes and judges' biases}
+
+\end{figure}%
 
@@
 DAGs typically omits exogenous variables for simplicity, the
-\emph{magnified representation} includes them, offering offering the
-advantage of highlighting potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}.
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O_{abhijk} & := f_{O}(D_{abhijk}) \\
-  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
-  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
-  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj04_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
-
-}
-
-\subcaption{\label{fig-cj04_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
-
-\end{figure}%
-
-Because it quickly becomes intractable, and without loosing generality,
-the (latent) ``perceived'' and ``true'' discriminal processes for the
-stimuli can be depicted in a vector for each judge, as in
+\emph{magnified representation} includes them, offering the advantage of
+highlighting potential issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}.
+
+DAG~\ref{fig-cj04_dag} displays the preliminary DAG associated with
+SCM~\ref{fig-cj04_scm}. The graph shows the direction of causal
+influences among variables, reflecting the assumptions of Thurstone's
+theory. It also distinguishes between observable endogenous variables,
+such as the outcome \(O_{abhijk}\), and variables considered latent
+according to the theory and practice of CJ, such as the discriminal
+difference \((D_{abhijk})\), the perceived discriminal processes
+\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
+\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\).
 
@@
 of the school) is thought to be a confounder in the relationship between
-the teaching method \(T\) and the outcome \(Y\). In this scenario, the
-figures display multiple descendant relationships such as
-\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
-also highlight unconnected node pairs, evident from the relationships
+the teaching method \(T\) and the outcome \(Y\). The figures display
+multiple descendant relationships such as \(X \rightarrow T\),
+\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
+unconnected node pairs, evident from the relationships
 \(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and

### 2025-02-01 paper2.qmd
section:theoretical; process:write

@@
 <!-- 3. Example of a CJ estimand and estimator, and its limitations -->
-For instance, a motivating example that illustrates these concepts throughout the document could involve researchers aiming to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each receiving a different teaching method.
-Judges would then compare pairs of students' written texts, producing a dichotomous outcome that reflects the relative quality of each text.
-Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
+A motivating example helps to clarify these concepts.
+The example will appear throughout the document to illustrate various aspects of the proposed extension.
+In this scenario, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+Judges would then compare pairs of students' written texts (stimuli), producing a dichotomous outcome that reflects the relative quality of each text (trait).
+Based on this setup, researchers could reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
 Following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
@@
 However, @sec-theory-issues presents compelling evidence that the BTL model has several statistical and measurement limitations.
-These limitations hinder the model's ability to identify distinct estimands relevant to CJ inquiries, including the one from the motivating example.
+These limitations hinder the model's ability to identify various estimands relevant to CJ inquiries, including the one from the motivating example.
 Identification is crucial because it is a necessary condition for ensuring consistent estimators.
-Without identification, researchers cannot achieve *consistency* even with "infinite" and error-free data.
-In other words, the estimates derived from the estimator would fail to converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010].
+Without identification, *consistency* --the property of an estimator whose estimates converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]-- cannot be achieved, even with "infinite" and error-free data.
 As a result, deriving meaningful insights from finite data also becomes unattainable [@Schuessler_et_al_2023, pp.
@@
 Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-The extended model clarifies the interactions among all actors and processes involved in CJ experiments.
-Furthermore, it allows researchers to shift the current comparative data analysis paradigm from accepting the model's assumptions as given to testing these assumptions for specific comparative data.
-For details on these assumptions, refer to @sec-theory-issues.
+
+<!-- 6. What are the advantages? -->
+In addition to addressing the identified limitations, the extended model offers two key advantages.
+First, it clarifies the interactions among all actors and processes involved in CJ experiments.
+Second, it shifts the current data analysis paradigm from passively accepting CJ's model assumptions to actively testing them with the specific data under analysis.
+For more details on these assumptions, refer to @sec-theory-issues.
+
+<!-- 7. How the section is divided? -->
 Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what the study calls the *conceptual-population model*.
-Meanwhile, @sec-theory-theoretical_SC integrates the assessment design features into what the study terms the *sample-comparison model*.
+This model assumes that researchers have access to a conceptual population of comparison data, meaning data representing all repeated comparisons made by every available judge for each pair of stimuli produced by each pair of individuals in the population, thus its name.
+Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what the study calls the *sample-comparison model*.
+This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
 
@@
 <!-- 1. The SCMs -->
-Before integrating the theoretical principles from @sec-thurstone_theory, the study defines SCMs and DAGs.
-SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
+Before incorporating the theoretical principles into the *conceptual-population model*, the study defines SCMs, which are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 27,68].
-Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
+Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other endogenous and exogenous variables.
 These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
@@
 <!-- 2. the preliminary SCM for CJ -->
-@fig-cj04 illustrates the preliminary integration of the theoretical principles underlying the motivating example.
-Specifically, [SCM -@fig-cj04_scm] displays the relationship between the outcome of the CJ experiment, $O_{abhijk}$, to several other variables.
-Here, the indices $\{a,b\}$ represent the compared texts, $\{h,i\}$ label the students who wrote these texts, $j$ denotes the judge, and $k$ accounts for cases where the judges perform multiple comparisons of the same texts written by the same students, i.e., the comparison number for a *repeated measures design* [@Lawson_2015, pp.
+@fig-cj04 illustrates the preliminary integration of theoretical principles into the conceptual-population model.
+Specifically, @fig-cj04_scm shows the SCM that underlies the motivating example.
+This SCM maps the relationship between the outcome of the CJ experiment $(O_{abhijk})$ to several other variables.
+In this model, the indices $\{a,b\}$ represent the compared texts (stimuli), with $a \neq b$.
+The indices $\{h,i\}$ label the students who wrote the texts (individuals), where $h = i$ is possible.
+The index $j$ denotes the judge, and $k$ denotes the comparison number, which accounts for experimental cases where judges perform multiple comparisons of the same texts written by the same students, i.e., a *repeated measures design* [@Lawson_2015, pp.
 366-376].
-The other variables related to the outcome include the discriminal difference $(D_{abhijk})$, the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$.
-Together with the outcome, these variables form the preliminary set of endogenous variables $V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$.
-Finally, the preliminary set of structural equations $F = \{f_{O},f_{D},f_{PT}\}$ defines the dependencies among these variables.
+Finally, the SCM presents the preliminary set of structural equations $F = \{f_{O},f_{D},f_{PT}\}$, which define the non-parametric dependencies among these variables.
+
+<!-- 3. What are the innovations of this? -->
+Note that the [SCM -@fig-cj04_scm] incorporates not only the variables described by Thurstone's theory, such as the discriminal difference $(D_{abhijk})$ and the texts' *true* discriminal processes $(T_{ahjk}, T_{bijk})$ (refer to @sec-thurstone_theory), but also innovates by incorporating additional variables central to CJ experiments, such as the texts' *perceived* discriminal processes $(PT_{ahjk}, PT_{bijk})$ and the judges' perceptions $(B_{kj})$.
+Together with the outcome, these variables constitute the preliminary set of endogenous variables $V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$.
+
+By incorporating the texts' perceived discriminal processes into the conceptual-population model, this study considers a crucial aspect of CJ experiments that previous literature has overlooked.
+It recognizes that the texts' true discriminal processes become an observable outcome only through the judges' perceptions.
+In other words, researchers observe the CJ outcome only when both variables create a *pure interaction effect*, meaning that neither variable has a direct effect on the outcome, but their interaction produces an effect [@Attia_et_al_2022].
+
+Conversely, based on the evidence presented in @sec-theory-issue1b, the study assumes that other relevant stimulus characteristics influence the judges' perceptions, leading to biases that resist cancellation and ultimately introduce dependencies between the stimuli.
+To address this, the study incorporates the judges' perceptions into the conceptual-population model, following the recommendations of @Andrich_1978 and @Wainer_et_al_1978.
 
@@
 
-<!-- 3. The DAGs -->
+<!-- 4. The DAGs -->
 Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
@@
 While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
-
-<!-- 4. the preliminary DAG for CJ -->
-[DAG -@fig-cj04_dag] displays the preliminary DAG associated with [SCM -@fig-cj04_scm].
-The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory.
-It also distinguishes between observable endogenous variables, such as the outcome $O_{abhijk}$, and variables considered latent according to the theory and practice of CJ, such as the discriminal difference $(D_{abhijk})$, the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$.
-
-<!-- Notice,  -->
-<!-- the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$ -->
-
-<!-- Assuming population data or more commonly known as census data, we ... -->
+Thus, this study uses the magnified representation.
+
+<!-- 5. the preliminary DAG for CJ -->
+@fig-cj04_dag displays the DAG associated with [SCM -@fig-cj04_scm].
+The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and CJ experiments.
+It also distinguishes between observable endogenous variables, such as the outcome $O_{abhijk}$, and variables considered latent according to the theory and practice of CJ, such as the discriminal difference, the perceived discriminal processes, the true discriminal processes, and the judges' perceptions.
+
+<!-- 6. The problem of continuing this way -->
+It is easy to intuit that describing the CJ system by assigning an index and a variable to each discriminal process could quickly become intractable and cumbersome due to the large size of the DAG and the repetition of its components.
+Therefore, to maintain generality, the study groups pairs of discriminal processes with different indices into matrices that contain these pairs.
+First, the study defines the vectors $U_{A} \in \mathbb{N}^{A}$, $U_{I} \in \mathbb{N}^{I}$ containing the indices of the stimuli and individuals, where:
+
+$$
+U_{A} = [1, \dots, a, \dots, b, \dots, A] ;\quad U_{I} = [1, \dots, a, \dots, b, \dots, I]
+$$
+
+Then, the study defines the matrices of stimuli and individual comparisons, $P \in \mathbb{N}^{m \times 2}$ and $R \in \mathbb{N}^{n \times 2}$, respectively.
+Here, each row of $P$ represents a pairing $\{a,b\}$ of stimuli indices and each row of $R$ represents a pairing $\{h,i\}$ of individual indices, i.e.,
+
+$$
+P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\}
+$$
+
+Here, $m$ and $n$ represents the number of unique pairwise comparisons between stimuli, and unique pairwise comparison per individuals, which corresponds to the number of ways to choose two elements from $U_{A}$ and $U_{I}$, typically given by:
+
+$$
+m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!}
+$$
 
 <!-- Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in   -->
-
-<!-- To declare a matrix \( P \) in \( \mathbb{N}^2 \) that represents the paired combinations of elements in \( I \), follow these steps: -->
-
-<!-- ### Definitions: -->
-<!-- 1. \( I \) is a subset of the natural numbers, \( I \subseteq \mathbb{N} \). -->
-<!-- 2. The paired combination of elements in \( I \) can be written as \( (i, j) \) for \( i, j \in I \). -->
-
-<!-- ### Constructing the Matrix \( P \): -->
-<!-- 1. Each row of \( P \) represents one pairing \( (i, j) \). -->
-<!-- 2. If \( I = \{a_1, a_2, \ldots, a_n\} \), then \( P \) contains \( n^2 \) rows for all pairs \( (i, j) \), including repetitions where \( i = j \). -->
-<!-- 3. Specifically, the matrix \( P \) is structured as: -->
-<!--    \[ -->
-<!--    P = \begin{bmatrix} -->
-<!--    i_1 & j_1 \\ -->
-<!--    i_2 & j_2 \\ -->
-<!--    \vdots & \vdots \\ -->
-<!--    i_{n^2} & j_{n^2} -->
-<!--    \end{bmatrix}, -->
-<!--    \] -->
-<!--    where \( i_k, j_k \in I \) for all \( k \), and each pair \( (i_k, j_k) \) corresponds to a combination of elements in \( I \). -->
-
-<!-- ### Algorithm to Generate \( P \): -->
-<!-- Given \( I = \{a_1, a_2, \ldots, a_n\} \): -->
-<!-- 1. Loop over all \( i \in I \) (outer loop). -->
-<!-- 2. For each \( i \), loop over all \( j \in I \) (inner loop). -->
-<!-- 3. Append the pair \( (i, j) \) as a new row in \( P \). -->
-
-<!-- ### Example: -->
-<!-- If \( I = \{1, 2, 3\} \), then the matrix \( P \) is: -->
-<!-- \[ -->
-<!-- P = \begin{bmatrix} -->
-<!-- 1 & 1 \\ -->
-<!-- 1 & 2 \\ -->
-<!-- 1 & 3 \\ -->
-<!-- 2 & 1 \\ -->
-<!-- 2 & 2 \\ -->
-<!-- 2 & 3 \\ -->
-<!-- 3 & 1 \\ -->
-<!-- 3 & 2 \\ -->
-<!-- 3 & 3 -->
-<!-- \end{bmatrix}. -->
-<!-- \] -->
-
-<!-- ### Mathematical Notation: -->
-<!-- 1. The matrix \( P \) can be defined as: -->
-<!--    \[ -->
-<!--    P = \{(i, j) \mid i \in I, j \in I\}. -->
-<!--    \] -->
-<!-- 2. In terms of dimensions, \( P \) has \( |I|^2 \) rows and 2 columns. -->
-
-<!-- This formalization ensures that \( P \) represents all paired combinations of elements from \( I \) in \( \mathbb{N}^2 \). -->
 
@@
 Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display the five fundamental building blocks for SCMs and DAGs.
-The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $X=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020].
+The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].

### 2025-02-01 _tex/paper2.tex
Built site for gh-pages

@@
 
-For instance, a motivating example that illustrates these concepts
-throughout the document could involve researchers aiming to answer the
-question: ``To what extent do different teaching methods influence
-students' ability to produce high-quality written texts?'' To
-investigate this, they could design a CJ experiment by randomly
-assigning students to two groups, each receiving a different teaching
-method. Judges would then compare pairs of students' written texts,
-producing a dichotomous outcome that reflects the relative quality of
-each text. Based on this setup, researchers can reformulate the research
+A motivating example helps to clarify these concepts. The example will
+appear throughout the document to illustrate various aspects of the
+proposed extension. In this scenario, researchers aim to determine: ``To
+what extent do different teaching methods influence students' ability to
+produce high-quality written texts?'' To investigate this, they could
+design a CJ experiment by randomly assigning students (individuals) to
+two groups, each receiving a different teaching method. Judges would
+then compare pairs of students' written texts (stimuli), producing a
+dichotomous outcome that reflects the relative quality of each text
+(trait). Based on this setup, researchers could reformulate the research
 question as the estimand: ``\emph{On average}, is there a difference in
@@
 that the BTL model has several statistical and measurement limitations.
-These limitations hinder the model's ability to identify distinct
+These limitations hinder the model's ability to identify various
 estimands relevant to CJ inquiries, including the one from the
@@
 condition for ensuring consistent estimators. Without identification,
-researchers cannot achieve \emph{consistency} even with ``infinite'' and
-error-free data. In other words, the estimates derived from the
-estimator would fail to converge to the ``true'' value of the estimand
-as the data size approaches infinity \citep{Everitt_et_al_2010}. As a
-result, deriving meaningful insights from finite data also becomes
-unattainable \citep[pp.~5]{Schuessler_et_al_2023}.
+\emph{consistency} --the property of an estimator whose estimates
+converge to the ``true'' value of the estimand as the data size
+approaches infinity \citep{Everitt_et_al_2010}-- cannot be achieved,
+even with ``infinite'' and error-free data. As a result, deriving
+meaningful insights from finite data also becomes unattainable
+\citep[pp.~5]{Schuessler_et_al_2023}.
 
@@
 with key assessment design features relevant to CJ experiments, such as
-the selection of judges, stimuli, and comparisons. The extended model
-clarifies the interactions among all actors and processes involved in CJ
-experiments. Furthermore, it allows researchers to shift the current
-comparative data analysis paradigm from accepting the model's
-assumptions as given to testing these assumptions for specific
-comparative data. For details on these assumptions, refer to
-Section~\ref{sec-theory-issues}. Accordingly,
-Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
-principles into what the study calls the \emph{conceptual-population
-model}. Meanwhile, Section~\ref{sec-theory-theoretical_SC} integrates
-the assessment design features into what the study terms the
-\emph{sample-comparison model}.
+the selection of judges, stimuli, and comparisons.
+
+In addition to addressing the identified limitations, the extended model
+offers two key advantages. First, it clarifies the interactions among
+all actors and processes involved in CJ experiments. Second, it shifts
+the current data analysis paradigm from passively accepting CJ's model
+assumptions to actively testing them with the specific data under
+analysis. For more details on these assumptions, refer to
+Section~\ref{sec-theory-issues}.
+
+Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
+theoretical principles into what the study calls the
+\emph{conceptual-population model}. This model assumes that researchers
+have access to a conceptual population of comparison data, meaning data
+representing all repeated comparisons made by every available judge for
+each pair of stimuli produced by each pair of individuals in the
+population, thus its name. Conversely,
+Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
+features into what the study calls the \emph{sample-comparison model}.
+This model assumes a more realistic scenario where researchers only have
+access to a sample of judges, individuals, stimuli, and comparisons from
+the conceptual population.
 
@@
 
-Before integrating the theoretical principles from
-Section~\ref{sec-thurstone_theory}, the study defines SCMs and DAGs.
-SCMs are formal mathematical models characterized by a set of
-\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
-\(E\), and a set of functions \(F\)
-\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
-whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
-In contrast, exogenous variables represent \emph{errors} or
-\emph{disturbances} arising from omitted factors that the investigator
-chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
-the functions, referred to as \emph{structural equations}, express the
-endogenous variables as non-parametric functions of other variables.
+Before incorporating the theoretical principles into the
+\emph{conceptual-population model}, the study defines SCMs, which are
+formal mathematical models characterized by a set of \emph{endogenous}
+variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
+functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
+variables are those whose causal mechanisms a researcher chooses to
+model \citep{Neal_2020}. In contrast, exogenous variables represent
+\emph{errors} or \emph{disturbances} arising from omitted factors that
+the investigator chooses not to model explicitly
+\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
+\emph{structural equations}, express the endogenous variables as
+non-parametric functions of other endogenous and exogenous variables.
 These functions use the symbol `\(:=\)' to denote the asymmetrical
@@
 
-Figure~\ref{fig-cj04} illustrates the preliminary integration of the
-theoretical principles underlying the motivating example. Specifically,
-SCM~\ref{fig-cj04_scm} displays the relationship between the outcome of
-the CJ experiment, \(O_{abhijk}\), to several other variables. Here, the
-indices \(\{a,b\}\) represent the compared texts, \(\{h,i\}\) label the
-students who wrote these texts, \(j\) denotes the judge, and \(k\)
-accounts for cases where the judges perform multiple comparisons of the
-same texts written by the same students, i.e., the comparison number for
-a \emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}. The
-other variables related to the outcome include the discriminal
-difference \((D_{abhijk})\), the perceived discriminal processes
-\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
-\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\). Together
-with the outcome, these variables form the preliminary set of endogenous
-variables
+Figure~\ref{fig-cj04} illustrates the preliminary integration of
+theoretical principles into the conceptual-population model.
+Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
+motivating example. This SCM maps the relationship between the outcome
+of the CJ experiment \((O_{abhijk})\) to several other variables. In
+this model, the indices \(\{a,b\}\) represent the compared texts
+(stimuli), with \(a \neq b\). The indices \(\{h,i\}\) label the students
+who wrote the texts (individuals), where \(h = i\) is possible. The
+index \(j\) denotes the judge, and \(k\) denotes the comparison number,
+which accounts for experimental cases where judges perform multiple
+comparisons of the same texts written by the same students, i.e., a
+\emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}.
+Finally, the SCM presents the preliminary set of structural equations
+\(F = \{f_{O},f_{D},f_{PT}\}\), which define the non-parametric
+dependencies among these variables.
+
+Note that the SCM~\ref{fig-cj04_scm} incorporates not only the variables
+described by Thurstone's theory, such as the discriminal difference
+\((D_{abhijk})\) and the texts' \emph{true} discriminal processes
+\((T_{ahjk}, T_{bijk})\) (refer to Section~\ref{sec-thurstone_theory}),
+but also innovates by incorporating additional variables central to CJ
+experiments, such as the texts' \emph{perceived} discriminal processes
+\((PT_{ahjk}, PT_{bijk})\) and the judges' perceptions \((B_{kj})\).
+Together with the outcome, these variables constitute the preliminary
+set of endogenous variables
 \(V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
-Finally, the preliminary set of structural equations
-\(F = \{f_{O},f_{D},f_{PT}\}\) defines the dependencies among these
-variables.
+
+By incorporating the texts' perceived discriminal processes into the
+conceptual-population model, this study considers a crucial aspect of CJ
+experiments that previous literature has overlooked. It recognizes that
+the texts' true discriminal processes become an observable outcome only
+through the judges' perceptions. In other words, researchers observe the
+CJ outcome only when both variables create a \emph{pure interaction
+effect}, meaning that neither variable has a direct effect on the
+outcome, but their interaction produces an effect
+\citep{Attia_et_al_2022}.
+
+Conversely, based on the evidence presented in
+Section~\ref{sec-theory-issue1b}, the study assumes that other relevant
+stimulus characteristics influence the judges' perceptions, leading to
+biases that resist cancellation and ultimately introduce dependencies
+between the stimuli. To address this, the study incorporates the judges'
+perceptions into the conceptual-population model, following the
+recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978}.
 
@@
 highlighting potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}.
-
-DAG~\ref{fig-cj04_dag} displays the preliminary DAG associated with
+\citep{Cinelli_et_al_2020}. Thus, this study uses the magnified
+representation.
+
+Figure~\ref{fig-cj04_dag} displays the DAG associated with
 SCM~\ref{fig-cj04_scm}. The graph shows the direction of causal
 influences among variables, reflecting the assumptions of Thurstone's
-theory. It also distinguishes between observable endogenous variables,
-such as the outcome \(O_{abhijk}\), and variables considered latent
-according to the theory and practice of CJ, such as the discriminal
-difference \((D_{abhijk})\), the perceived discriminal processes
-\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
-\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\).
+theory and CJ experiments. It also distinguishes between observable
+endogenous variables, such as the outcome \(O_{abhijk}\), and variables
+considered latent according to the theory and practice of CJ, such as
+the discriminal difference, the perceived discriminal processes, the
+true discriminal processes, and the judges' perceptions.
+
+It is easy to intuit that describing the CJ system by assigning an index
+and a variable to each discriminal process could quickly become
+intractable and cumbersome due to the large size of the DAG and the
+repetition of its components. Therefore, to maintain generality, the
+study groups pairs of discriminal processes with different indices into
+matrices that contain these pairs. First, the study defines the vectors
+\(U_{A} \in \mathbb{N}^{A}\), \(U_{I} \in \mathbb{N}^{I}\) containing
+the indices of the stimuli and individuals, where:
+
+\[
+U_{A} = [1, \dots, a, \dots, b, \dots, A] ;\quad U_{I} = [1, \dots, a, \dots, b, \dots, I]
+\]
+
+Then, the study defines the matrices of stimuli and individual
+comparisons, \(P \in \mathbb{N}^{m \times 2}\) and
+\(R \in \mathbb{N}^{n \times 2}\), respectively. Here, each row of \(P\)
+represents a pairing \(\{a,b\}\) of stimuli indices and each row of
+\(R\) represents a pairing \(\{h,i\}\) of individual indices, i.e.,
+
+\[
+P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\}
+\]
+
+Here, \(m\) and \(n\) represents the number of unique pairwise
+comparisons between stimuli, and unique pairwise comparison per
+individuals, which corresponds to the number of ways to choose two
+elements from \(U_{A}\) and \(U_{I}\), typically given by:
+
+\[
+m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!}
+\]
 
@@
 by the SCMs, defined in terms of a set of \emph{endogenous} variables
-\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
+\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
 \(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions

### 2025-02-09 paper2.qmd
section:theoretical; process:write,figures,code

@@
 Thus, addressing these issues through a more systematic and integrated approach could provide significant statistical and measurement benefits.
-For this reason, the study employs the structural approach to causal inference, as it fulfills these requirements.
+For this reason, the study employs the structural approach to causal inference.
 
@@
 
-# Extending Thurstone's model {#sec-theoretical}
+# Extending Thurstone's general form {#sec-theoretical}
 
@@
 A motivating example helps to clarify these concepts.
-The example will appear throughout the document to illustrate various aspects of the proposed extension.
-In this scenario, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
-Judges would then compare pairs of students' written texts (stimuli), producing a dichotomous outcome that reflects the relative quality of each text (trait).
+This example will appear throughout the document to illustrate various aspects of the proposed extension.
+In this scenario, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+Judges then compare pairs of students’ written texts (stimuli), producing a dichotomous outcome that reflects the relative quality of each text (trait).
 Based on this setup, researchers could reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
@@
 <!-- 5. What the section will do? -->
-As a result, this section addresses the identified limitations by extending Thurstone's general form (see @tbl-thurstone_cases) through a systematic and integrated approach based on the structural approach to causal inference.
+Thus, this section addresses the identified limitations by extending Thurstone’s general form (see @tbl-thurstone_cases) through a systematic and integrated approach based on the structural approach to causal inference.
 Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
@@
 <!-- 6. What are the advantages? -->
-In addition to addressing the identified limitations, the extended model offers two key advantages.
+In addition to overcoming the identified limitations, the extended model offers two key advantages.
 First, it clarifies the interactions among all actors and processes involved in CJ experiments.
-Second, it shifts the current data analysis paradigm from passively accepting CJ's model assumptions to actively testing them with the specific data under analysis.
+Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing if those assumptions fit the specific data under analysis.
 For more details on these assumptions, refer to @sec-theory-issues.
@@
 <!-- recording: Sven and Tine 25.01.30 p1; time: 00:23:25 - 00:48:00 -->
+<!-- recording: Sven and Tine 25.02.07; time: 00:00:00 - 00:00:00 -->
 
 <!-- 1. The SCMs -->
-Before incorporating the theoretical principles into the *conceptual-population model*, the study defines SCMs, which are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
+Before incorporating the theoretical principles into the *conceptual-population model*, the study defines SCMs.
+SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 
-<!-- ![](/images/png/CJ_TM_01.png){#fig-cj01_dag fig-align="center" fig-pos="H" width=23%} -->
-
-<!-- ![](/images/png/CJ_TM_02.png){#fig-cj02_dag fig-align="center" fig-pos="H" width=32%} -->
-
-<!-- ![](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=45%} -->
-
-::: {#fig-cj04 layout-nrow=2}
-
-::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{abhijk} & := f_{O}(D_{abhijk}) \\
-  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
-  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
-  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
+::: {#fig-cj03 layout-nrow=2}
+
+::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
+  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
 \end{aligned}
@@
 
-![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=45%}
+![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=35%}
 
@@
 
-::: {#fig-cj05 layout-nrow=2}
-
-::: {#fig-cj05_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
+::: {#fig-cj04 layout-nrow=2}
+
+::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK})
 \end{aligned}
@@
 
-![DAG](/images/png/CJ_TM_05.png){#fig-cj05_dag fig-align="center" fig-pos="H" width=44%}
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
 
@@
 
-<!-- ![](/images/png/CJ_TM_06.png){#fig-cj06_dag fig-align="center" fig-pos="H" width=59%} -->
-
-<!-- ![](/images/png/CJ_TM_07.png){#fig-cj07_dag fig-align="center" fig-pos="H" width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_08.png){#fig-cj08_dag fig-align="center" fig-pos="H" width=68%} -->
-
-<!-- ![](/images/png/CJ_TM_09.png){#fig-cj09_dag fig-align="center" fig-pos="H" width=68%} -->
-
-::: {#fig-cj10 layout-nrow=2}
-
-::: {#fig-cj10_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+::: {#fig-cj09 layout-nrow=2}
+
+::: {#fig-cj09_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) 
@@
 
-![DAG](/images/png/CJ_TM_10.png){#fig-cj10_dag fig-align="center" fig-pos="H" width=68%}
+![DAG](/images/png/CJ_TM_09.png){#fig-cj09_dag fig-align="center" fig-pos="H" width=70%}
 
@@
 
-<!-- ![](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=80%} -->
-
-::: {#fig-cj12 layout-nrow=2}
-
-::: {#fig-cj12_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+::: {#fig-cj11 layout-nrow=2}
+
+::: {#fig-cj11_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) 
@@
 
-![DAG](/images/png/CJ_TM_12.png){#fig-cj12_dag fig-align="center" fig-pos="H" width=80%}
+![DAG](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=84%}
 
@@
 
-<!-- ![](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=90%} -->
-
-::: {#fig-cj14 layout-nrow=2}
-
-::: {#fig-cj14_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
-  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+::: {#fig-cj13 layout-nrow=2}
+
+::: {#fig-cj13_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) 
@@
 
-![DAG](/images/png/CJ_TM_14.png){#fig-cj14_dag fig-align="center" fig-pos="H" width=90%}
-
-CJ's outcome, discriminal difference, discriminal processes and judges' biases
-:::
-
-::: {#fig-cj15 layout-nrow=2}
-
-::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
-  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) 
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_15.png){#fig-cj15_dag fig-align="center" fig-pos="H" width=90%}
+![DAG](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=100%}
 
@@
 
-**Funding:** The project was founded through the Research Fund of the University of Antwerp (BOF).
-
-**Financial interests:** The authors have no relevant financial interest to disclose.
-
-**Non-financial interests:** The authors have no relevant non-financial interest to disclose.
-
-**Ethics approval:** The University of Antwerp Research Ethics Committee has confirmed that no ethical approval is required.
+**Funding:** The Research Fund (BOF) of the University of Antwerp funded this project.
+
+**Financial interests:** The authors declare no relevant financial interests.
+
+**Non-financial interests:** The authors declare no relevant non-financial interests.
+
+**Ethics approval:** The University of Antwerp Research Ethics Committee confirmed that this study does not require ethical approval.
 
@@
 
-**Consent for publication:** All authors have read and agreed to the published version of the manuscript.
-
-**Availability of data and materials:** No data was utilized in this study.
-
-**Code availability:** All the code utilized in this research is available in the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).
-
-**AI-assisted technologies in the writing process:** The authors utilized a range of AI-based language tools throughout the preparation of this work.
-They occasionally employed the tools to refine phrasing and optimize wording, ensuring appropriate language use and enhancing the manuscript's clarity and coherence.
-The authors take full responsibility for the final content of the publication.
+**Consent for publication:** All authors have read and approved the final version of the manuscript for publication.
+
+**Data availability:** This study did not use any data.
+
+**Materials and code availability:** The `CODE LINK` section at the top of the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/) provides access to all materials and code.
+
+**AI-assisted technologies in the writing process:** The authors used various AI-based language tools to refine phrasing, optimize wording, and enhance clarity and coherence throughout the manuscript.
+They take full responsibility for the final content of the publication.
 
@@
 # Appendix {#sec-appendix .appendix appendix-style=plain}
+
+## Statistics and Causal inference {#sec-appendixB .appendix appendix-style=plain}
 
@@
 
-## Empirical research and randomized experiments {#sec-appendix-A .appendix appendix-style=plain}
+### Empirical research and randomized experiments {#sec-appendixB1 .appendix appendix-style=plain}
 
@@
 
-## Identification under causal inference {#sec-appendix-B .appendix appendix-style=plain}
-
-<!-- ### The general idea {#sec-appendix-B1 .appendix appendix-style=plain} -->
+### Identification under causal inference {#sec-appendixB2 .appendix appendix-style=plain}
 
@@
 The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals.
-For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes, as in @eq-ICE.
+For instance, in the scenario described in @sec-appendixB1, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes, as in @eq-ICE.
 
@@
 
-For instance, if researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
+For instance, if researchers cannot conduct the randomized experiments described in @sec-appendixB1 and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
 Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)
@@
 
-### Diving into the specifics {#sec-appendix-B2 .appendix appendix-style=plain}
+### Diving into the specifics {#sec-appendixB3 .appendix appendix-style=plain}
 
@@
 35].
-@sec-appendix-B21 and @sec-appendix-B22 elaborate on the first advantage, while @sec-appendix-B22 and @sec-appendix-B23 do so for the second.
+
+Thus, @sec-appendixB31 and @sec-appendixB32 elaborate on the first advantage, while @sec-appendixB32 and @sec-appendixB33 do so for the second.
+Finally, @sec-appendixB34 explains how researchers use SCMs and DAGs alongside Bayesian inference methods in the estimation process.
 
@@
 
-#### The five fundamental block for SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}
+#### The five fundamental block for SCMs and DAGs {#sec-appendixB31 .appendix appendix-style=plain}
 
@@
 <!-- 5. Continuing with the motivating example -->
-Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B.
+Researchers can then use these building blocks to represent the scenario outlined in @sec-appendixB2.
 SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example.
@@
 
-Plausible causal structure the scenario outlined in @sec-appendix-B.
-:::
-
-#### The probabilistic implications of these blocks {#sec-appendix-B22 .appendix appendix-style=plain}
+Plausible causal structure the scenario outlined in @sec-appendixB2.
+:::
+
+#### The probabilistic implications of these blocks {#sec-appendixB32 .appendix appendix-style=plain}
 
@@
 <!-- 4. can you illustrate this concept? -->
-The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which (in)dependencies can SMCs and DAGs predict.
+The five fundamental building blocks described in @sec-appendixB31 clearly illustrate which (in)dependencies can SMCs and DAGs predict.
 For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] enables researchers to express the joint probability distribution of the observed variables as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, supporting the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$) [@Neal_2020, pp.
@@
 Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$.
-Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendix-B21 and the CPDs.
+Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendixB31 and the CPDs.
 
@@
 <!-- 5. Continuing with the motivating example -->
-Using these additional probabilistic insights, researchers can revisit the scenario in @sec-appendix-B.
+Using these additional probabilistic insights, researchers can revisit the scenario in @sec-appendixB2.
 In this context, applying the BNF to the SCM [-@fig-scm_example2] structure, enables the representation of the joint probability distribution of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
@@
 
-Plausible causal structure the scenario outlined in @sec-appendix-B.
-:::
-
-#### From probability to causality {#sec-appendix-B23 .appendix appendix-style=plain}
+Plausible causal structure the scenario outlined in @sec-appendixB2.
+:::
+
+#### From probability to causality {#sec-appendixB33 .appendix appendix-style=plain}
 
@@
 <!-- 4. revisiting the motivating example -->
-With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct.
+With these new insights, researchers revisiting the scenario in @sec-appendixB32 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct.
 This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure.
@@
 
-Plausible causal structure the scenario outlined in @sec-appendix-B22.
+Plausible causal structure the scenario outlined in @sec-appendixB32.
 :::
@@
 They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE.
-This process is expressed in @eq-mACE2 (see @sec-appendix-B).
+This process is expressed in @eq-mACE2 (see @sec-appendixB2).
 
@@
 
-#### The estimation process {#sec-appendix-C .appendix appendix-style=plain}
+#### The estimation process {#sec-appendixB34 .appendix appendix-style=plain}
 
@@
 Nevertheless, researchers can manage the complexity of the likelihood by assuming specific local (in)dependencies among variables.
-SCMs and DAGs provide a formal framework to represent these assumptions, as detailed in @sec-appendix-B22.
+SCMs and DAGs provide a formal framework to represent these assumptions, as detailed in @sec-appendixB32.
 These assumptions improve model tractability and simplify the estimation process by enabling the derivation of the BNF of the likelihood (@eq-like_BNF).

### 2025-02-09 _tex/paper2.tex
Built site for gh-pages

@@
 statistical and measurement benefits. For this reason, the study employs
-the structural approach to causal inference, as it fulfills these
-requirements.
-
-\section{Extending Thurstone's model}\label{sec-theoretical}
+the structural approach to causal inference.
+
+\section{Extending Thurstone's general form}\label{sec-theoretical}
 
@@
 
-A motivating example helps to clarify these concepts. The example will
+A motivating example helps to clarify these concepts. This example will
 appear throughout the document to illustrate various aspects of the
@@
 what extent do different teaching methods influence students' ability to
-produce high-quality written texts?'' To investigate this, they could
-design a CJ experiment by randomly assigning students (individuals) to
-two groups, each receiving a different teaching method. Judges would
-then compare pairs of students' written texts (stimuli), producing a
-dichotomous outcome that reflects the relative quality of each text
-(trait). Based on this setup, researchers could reformulate the research
-question as the estimand: ``\emph{On average}, is there a difference in
-the ability to produce high-quality written texts between the two groups
-of students?''. Following current CJ practices, researchers would then
-rely on estimates from the BTL model, or its transformations, to
-approximate this estimand.
+produce high-quality written texts?'' To investigate this, they design a
+CJ experiment by randomly assigning students (individuals) to two
+groups, each receiving a different teaching method. Judges then compare
+pairs of students' written texts (stimuli), producing a dichotomous
+outcome that reflects the relative quality of each text (trait). Based
+on this setup, researchers could reformulate the research question as
+the estimand: ``\emph{On average}, is there a difference in the ability
+to produce high-quality written texts between the two groups of
+students?''. Following current CJ practices, researchers would then rely
+on estimates from the BTL model, or its transformations, to approximate
+this estimand.
 
@@
 
-As a result, this section addresses the identified limitations by
-extending Thurstone's general form (see Table~\ref{tbl-thurstone_cases})
-through a systematic and integrated approach based on the structural
-approach to causal inference. Specifically, it leverages the
-capabilities of this approach to formalize the combination of the core
-theoretical principles outlined in Section~\ref{sec-thurstone_theory}
-with key assessment design features relevant to CJ experiments, such as
-the selection of judges, stimuli, and comparisons.
-
-In addition to addressing the identified limitations, the extended model
+Thus, this section addresses the identified limitations by extending
+Thurstone's general form (see Table~\ref{tbl-thurstone_cases}) through a
+systematic and integrated approach based on the structural approach to
+causal inference. Specifically, it leverages the capabilities of this
+approach to formalize the combination of the core theoretical principles
+outlined in Section~\ref{sec-thurstone_theory} with key assessment
+design features relevant to CJ experiments, such as the selection of
+judges, stimuli, and comparisons.
+
+In addition to overcoming the identified limitations, the extended model
 offers two key advantages. First, it clarifies the interactions among
 all actors and processes involved in CJ experiments. Second, it shifts
-the current data analysis paradigm from passively accepting CJ's model
-assumptions to actively testing them with the specific data under
-analysis. For more details on these assumptions, refer to
-Section~\ref{sec-theory-issues}.
+the current comparative data analysis paradigm from passively accepting
+the model assumptions to actively testing if those assumptions fit the
+specific data under analysis. For more details on these assumptions,
+refer to Section~\ref{sec-theory-issues}.
 
@@
 Before incorporating the theoretical principles into the
-\emph{conceptual-population model}, the study defines SCMs, which are
+\emph{conceptual-population model}, the study defines SCMs. SCMs are
 formal mathematical models characterized by a set of \emph{endogenous}
@@
 \begin{aligned}
-  O_{abhijk} & := f_{O}(D_{abhijk}) \\
-  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
-  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
-  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
+  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
+  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
 \end{aligned}
@@
 
-\subcaption{\label{fig-cj04_scm}SCM}
+\subcaption{\label{fig-cj03_scm}SCM}
 
@@
 
-\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
-
-}
-
-\subcaption{\label{fig-cj04_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
+\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
+
+}
+
+\subcaption{\label{fig-cj03_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj03}CJ's outcome, discriminal difference,
 discriminal processes and judges' biases}
@@
 \begin{aligned}
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK})
 \end{aligned}
@@
 
-\subcaption{\label{fig-cj05_scm}SCM}
+\subcaption{\label{fig-cj04_scm}SCM}
 
@@
 
-\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_05.png}
-
-}
-
-\subcaption{\label{fig-cj05_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj05}CJ's outcome, discriminal difference,
+\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+
+}
+
+\subcaption{\label{fig-cj04_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
 discriminal processes and judges' biases}
@@
 \begin{aligned}
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) 
@@
 
-\subcaption{\label{fig-cj10_scm}SCM}
+\subcaption{\label{fig-cj09_scm}SCM}
 
@@
 
-\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_10.png}
-
-}
-
-\subcaption{\label{fig-cj10_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj10}CJ's outcome, discriminal difference,
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
+
+}
+
+\subcaption{\label{fig-cj09_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj09}CJ's outcome, discriminal difference,
 discriminal processes and judges' biases}
@@
 \begin{aligned}
-  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) 
@@
 
-\subcaption{\label{fig-cj12_scm}SCM}
+\subcaption{\label{fig-cj11_scm}SCM}
 
@@
 
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}
-
-}
-
-\subcaption{\label{fig-cj12_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj12}CJ's outcome, discriminal difference,
+\includegraphics[width=0.84\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}
+
+}
+
+\subcaption{\label{fig-cj11_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj11}CJ's outcome, discriminal difference,
 discriminal processes and judges' biases}
@@
 \begin{aligned}
-  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
-  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
+  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) 
@@
 
-\subcaption{\label{fig-cj14_scm}SCM}
+\subcaption{\label{fig-cj13_scm}SCM}
 
@@
 
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}
-
-}
-
-\subcaption{\label{fig-cj14_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj14}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
-
-\end{figure}%
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
-  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
-  O_{PRQ} & := f_{O}(D_{PRQ}) \\
-  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
-  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
-  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) 
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj15_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}
-
-}
-
-\subcaption{\label{fig-cj15_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj15}CJ's outcome, discriminal difference,
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}
+
+}
+
+\subcaption{\label{fig-cj13_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj13}CJ's outcome, discriminal difference,
 discriminal processes and judges' biases}
@@
 
-\textbf{Funding:} The project was founded through the Research Fund of
-the University of Antwerp (BOF).
-
-\textbf{Financial interests:} The authors have no relevant financial
-interest to disclose.
-
-\textbf{Non-financial interests:} The authors have no relevant
-non-financial interest to disclose.
+\textbf{Funding:} The Research Fund (BOF) of the University of Antwerp
+funded this project.
+
+\textbf{Financial interests:} The authors declare no relevant financial
+interests.
+
+\textbf{Non-financial interests:} The authors declare no relevant
+non-financial interests.
 
 \textbf{Ethics approval:} The University of Antwerp Research Ethics
-Committee has confirmed that no ethical approval is required.
+Committee confirmed that this study does not require ethical approval.
 
@@
 
-\textbf{Consent for publication:} All authors have read and agreed to
-the published version of the manuscript.
-
-\textbf{Availability of data and materials:} No data was utilized in
-this study.
-
-\textbf{Code availability:} All the code utilized in this research is
-available in the digital document located at:
-\url{https://jriveraespejo.github.io/paper2_manuscript/}.
+\textbf{Consent for publication:} All authors have read and approved the
+final version of the manuscript for publication.
+
+\textbf{Data availability:} This study did not use any data.
+
+\textbf{Materials and code availability:} The \texttt{CODE\ LINK}
+section at the top of the digital document located at:
+\url{https://jriveraespejo.github.io/paper2_manuscript/} provides access
+to all materials and code.
 
 \textbf{AI-assisted technologies in the writing process:} The authors
-utilized a range of AI-based language tools throughout the preparation
-of this work. They occasionally employed the tools to refine phrasing
-and optimize wording, ensuring appropriate language use and enhancing
-the manuscript's clarity and coherence. The authors take full
-responsibility for the final content of the publication.
+used various AI-based language tools to refine phrasing, optimize
+wording, and enhance clarity and coherence throughout the manuscript.
+They take full responsibility for the final content of the publication.
 
@@
 \section{Appendix}\label{sec-appendix}
+
+\subsection{Statistics and Causal inference}\label{sec-appendixB}
 
@@
 
-\subsection{Empirical research and randomized
-experiments}\label{sec-appendix-A}
+\subsubsection{Empirical research and randomized
+experiments}\label{sec-appendixB1}
 
@@
 
-\subsection{Identification under causal inference}\label{sec-appendix-B}
+\subsubsection{Identification under causal
+inference}\label{sec-appendixB2}
 
@@
 framework for defining counterfactuals. For instance, in the scenario
-described in Section~\ref{sec-appendix-A}, the approach begins by
+described in Section~\ref{sec-appendixB1}, the approach begins by
 defining the \emph{individual causal effect} (ICE) as the difference
@@
 For instance, if researchers cannot conduct the randomized experiments
-described in Section~\ref{sec-appendix-A} and must instead rely on
+described in Section~\ref{sec-appendixB1} and must instead rely on
 observational data, they can still identify the ACE as long as an
@@
 
-\subsubsection{Diving into the specifics}\label{sec-appendix-B2}
+\subsubsection{Diving into the specifics}\label{sec-appendixB3}
 
@@
 their parameters \citep[pp.~35]{Pearl_et_al_2016}.
-Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
-elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
-and Section~\ref{sec-appendix-B23} do so for the second.
+
+Thus, Section~\ref{sec-appendixB31} and Section~\ref{sec-appendixB32}
+elaborate on the first advantage, while Section~\ref{sec-appendixB32}
+and Section~\ref{sec-appendixB33} do so for the second. Finally,
+Section~\ref{sec-appendixB34} explains how researchers use SCMs and DAGs
+alongside Bayesian inference methods in the estimation process.
 
 \paragraph{The five fundamental block for SCMs and
-DAGs}\label{sec-appendix-B21}
+DAGs}\label{sec-appendixB31}
 
@@
 Researchers can then use these building blocks to represent the scenario
-outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
+outlined in Section~\ref{sec-appendixB2}. SCM \ref{fig-scm_example1} and
 DAG \ref{fig-mdag_example1} depict the plausible causal structure for
@@
 \caption{\label{fig-example1}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B}.}
+outlined in Section~\ref{sec-appendixB2}.}
 
@@
 \paragraph{The probabilistic implications of these
-blocks}\label{sec-appendix-B22}
+blocks}\label{sec-appendixB32}
 
@@
 The five fundamental building blocks described in
-Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
+Section~\ref{sec-appendixB31} clearly illustrate which (in)dependencies
 can SMCs and DAGs predict. For instance, applying the BNF to the causal
@@
 Notably, these descriptions demonstrate the clear correspondence between
-the structural equations illustrated in Section~\ref{sec-appendix-B21}
+the structural equations illustrated in Section~\ref{sec-appendixB31}
 and the CPDs.
@@
 Using these additional probabilistic insights, researchers can revisit
-the scenario in Section~\ref{sec-appendix-B}. In this context, applying
+the scenario in Section~\ref{sec-appendixB2}. In this context, applying
 the BNF to the SCM \ref{fig-scm_example2} structure, enables the
@@
 \caption{\label{fig-example2}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B}.}
+outlined in Section~\ref{sec-appendixB2}.}
 
@@
 
-\paragraph{From probability to causality}\label{sec-appendix-B23}
+\paragraph{From probability to causality}\label{sec-appendixB33}
 
@@
 With these new insights, researchers revisiting the scenario in
-Section~\ref{sec-appendix-B22} can infer that the socio-economic status
+Section~\ref{sec-appendixB32} can infer that the socio-economic status
 of the school, \(X\), satisfies the backdoor criterion, assuming the
@@
 \caption{\label{fig-example3}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendix-B22}.}
+outlined in Section~\ref{sec-appendixB32}.}
 
@@
 this effect over \(X\) to obtain the ACE. This process is expressed in
-Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}).
+Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendixB2}).
 
@@
 
-\paragraph{The estimation process}\label{sec-appendix-C}
+\paragraph{The estimation process}\label{sec-appendixB34}
 
@@
 provide a formal framework to represent these assumptions, as detailed
-in Section~\ref{sec-appendix-B22}. These assumptions improve model
+in Section~\ref{sec-appendixB32}. These assumptions improve model
 tractability and simplify the estimation process by enabling the

### 2025-02-09 code/CJmod_add.R

@@ -1,0 +1,155 @@
+mcmc_code = "
+data{
+
+    // data dimensions
+    int ns;     // number of experimental runs
+    int nsI;    // max. number of individuals
+    int nsA;    // max. number of stimuli
+    int nsJ;    // max. number of judges
+    int nsK;    // max. number of repeated comparisons
+    
+    // categories discrete variables
+    int sup_dXI;  // max. number of categories in XI
+    int sup_dXA;  // max. number of categories in XA
+    int sup_dZJ;  // max. number of categories in ZJ
+    int sup_dZK;  // max. number of categories in ZK
+    
+    // comparison data
+    array[ns] int<lower=1, upper=nsI> Rs1;   // individual (left)
+    array[ns] int<lower=1, upper=nsA> Ps1;   // stimuli (left)
+    array[ns] int<lower=1, upper=nsI> Rs2;   // individual (right)
+    array[ns] int<lower=1, upper=nsA> Ps2;   // stimuli (right)
+    array[ns] int<lower=1, upper=nsJ> Us;    // judges
+    array[ns] int<lower=1, upper=nsK> Vs;    // repeated comparisons
+    array[ns] int<lower=0, upper=1> ORPUV;   // outcome (dichotomous)
+    
+    // individual-stimuli data
+    array[nsI*nsA] int<lower=1, upper=nsI> Is;      // individuals
+    array[nsI*nsA] int<lower=1, upper=nsA> As;      // stimuli
+    array[nsI*nsA] real XIc;                        // individual predictor (continuous)
+    array[nsI*nsA] int<lower=1, upper=sup_dXI> XId; // individual predictor (discrete)
+    array[nsI*nsA] real XAc;                        // stimuli predictor (continuous)
+    array[nsI*nsA] int<lower=1, upper=sup_dXA> XAd; // stimuli predictor (discrete)
+    
+    // juges-repeated comparisons data
+    array[nsJ*nsK] int<lower=1, upper=nsJ> Js;      // judges
+    array[nsJ*nsK] int<lower=1, upper=nsK> Ks;      // repeated comparisons
+    array[nsJ*nsK] real ZJc;                        // judges predictor (continuous)
[…]

### 2025-02-10 paper2.qmd
section:theoretical; process:write,code

@@
 Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can exert an uneven influence on judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp.
+These additional relevant characteristics are unlikely to be equally weighted or opposing.
+As a result, they may exert an uneven influence on judges' perceptions, creating biases in their judgments.
+Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases may ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
 346].
@@
 <!-- 4. segway for next section -->
-In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates.
-This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp.
-2].
-Thus, addressing these issues through a more systematic and integrated approach could provide significant statistical and measurement benefits.
-For this reason, the study employs the structural approach to causal inference.
+In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of the trait estimates.
+This overreliance may also undermine their validity [@Perron_et_al_2015, pp.
+2], especially when combined with the disconnect between CJ's approach to trait measurement and hypothesis testing.
+However, the structural approach to causal inference can address these issues by providing a systematic and integrated framework that enhances statistical accuracy while strengthening measurement reliability and validity.
 
@@
 <!-- recording: Sven and Tine 25.01.30 p1; time: 00:00:00 - 00:23:25 -->
+<!-- recording: Sven and Tine 25.02.07; time: 00:02:20 - 00:06:20 -->
 
@@
 A motivating example helps to clarify these concepts.
-This example will appear throughout the document to illustrate various aspects of the proposed extension.
-In this scenario, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
-Judges then compare pairs of students’ written texts (stimuli), producing a dichotomous outcome that reflects the relative quality of each text (trait).
+This example will appear throughout the document to illustrate various aspects of the proposed development.
+In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researchers could design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+Judges would then compare pairs of students’ written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait).
 Based on this setup, researchers could reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-Following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
+Finally, following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
 
@@
 However, @sec-theory-issues presents compelling evidence that the BTL model has several statistical and measurement limitations.
-These limitations hinder the model's ability to identify various estimands relevant to CJ inquiries, including the one from the motivating example.
+These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the motivating example.
 Identification is crucial because it is a necessary condition for ensuring consistent estimators.
-Without identification, *consistency* --the property of an estimator whose estimates converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]-- cannot be achieved, even with "infinite" and error-free data.
-As a result, deriving meaningful insights from finite data also becomes unattainable [@Schuessler_et_al_2023, pp.
+*Consistency* refers to the property of an estimator whose estimates converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010].
+Without identification, consistency cannot be achieved, even with "infinite" and error-free data.
+Consequently, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023, pp.
 5].
@@
 
-<!-- 5. What the section will do? -->
-Thus, this section addresses the identified limitations by extending Thurstone’s general form (see @tbl-thurstone_cases) through a systematic and integrated approach based on the structural approach to causal inference.
+<!-- 6. What the section will do? -->
+Thus, this section addresses the identified issues in @sec-theory-issues by extending Thurstone’s general form using the structural approach to causal inference.
 Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-
-<!-- 6. What are the advantages? -->
-In addition to overcoming the identified limitations, the extended model offers two key advantages.
+In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
 First, it clarifies the interactions among all actors and processes involved in CJ experiments.
-Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing if those assumptions fit the specific data under analysis.
-For more details on these assumptions, refer to @sec-theory-issues.
+Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.
 
 <!-- 7. How the section is divided? -->
-Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what the study calls the *conceptual-population model*.
-This model assumes that researchers have access to a conceptual population of comparison data, meaning data representing all repeated comparisons made by every available judge for each pair of stimuli produced by each pair of individuals in the population, thus its name.
-Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what the study calls the *sample-comparison model*.
+Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we call the *conceptual-population model*.
+This model assumes that researchers have access to a conceptual population of comparison data, that is, data representing all repeated comparisons made by every available judge for each pair of stimuli produced by each pair of individuals in the population, hence its name.
+Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we call the *sample-comparison model*.
 This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
@@
 <!-- recording: Sven and Tine 25.01.30 p1; time: 00:23:25 - 00:48:00 -->
-<!-- recording: Sven and Tine 25.02.07; time: 00:00:00 - 00:00:00 -->
+<!-- recording: Sven and Tine 25.02.07; time: 00:00:00 - 00:02:20 -->
 
 <!-- 1. The SCMs -->
-Before incorporating the theoretical principles into the *conceptual-population model*, the study defines SCMs.
+Before incorporating the theoretical principles into the *conceptual-population model*, it is essential to define SCMs.
 SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
@@
 Specifically, @fig-cj04_scm shows the SCM that underlies the motivating example.
-This SCM maps the relationship between the outcome of the CJ experiment $(O_{abhijk})$ to several other variables.
-In this model, the indices $\{a,b\}$ represent the compared texts (stimuli), with $a \neq b$.
+This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{abhijk})$ to several other variables.
+In this SCM, the subscripts $\{a,b\}$ represent the compared texts (stimuli), with $a \neq b$.
 The indices $\{h,i\}$ label the students who wrote the texts (individuals), where $h = i$ is possible.
-The index $j$ denotes the judge, and $k$ denotes the comparison number, which accounts for experimental cases where judges perform multiple comparisons of the same texts written by the same students, i.e., a *repeated measures design* [@Lawson_2015, pp.
+Moreover, the index $j$ denotes the judge, and $k$ denotes the judgment number, which accounts for experimental cases where judges perform multiple judgments of the same texts written by the same students, i.e., a *repeated measures design* [@Lawson_2015, pp.
 366-376].
-Finally, the SCM presents the preliminary set of structural equations $F = \{f_{O},f_{D},f_{PT}\}$, which define the non-parametric dependencies among these variables.
+
+The other variables in [SCM -@fig-cj04_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory).
+Additionally, based on the evidence presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, we incorporate the judges' biases $(B_{kj})$ into the model.
+Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}$.
+Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
 
 <!-- 3. What are the innovations of this? -->
-Note that the [SCM -@fig-cj04_scm] incorporates not only the variables described by Thurstone's theory, such as the discriminal difference $(D_{abhijk})$ and the texts' *true* discriminal processes $(T_{ahjk}, T_{bijk})$ (refer to @sec-thurstone_theory), but also innovates by incorporating additional variables central to CJ experiments, such as the texts' *perceived* discriminal processes $(PT_{ahjk}, PT_{bijk})$ and the judges' perceptions $(B_{kj})$.
-Together with the outcome, these variables constitute the preliminary set of endogenous variables $V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$.
-
-By incorporating the texts' perceived discriminal processes into the conceptual-population model, this study considers a crucial aspect of CJ experiments that previous literature has overlooked.
-It recognizes that the texts' true discriminal processes become an observable outcome only through the judges' perceptions.
-In other words, researchers observe the CJ outcome only when both variables create a *pure interaction effect*, meaning that neither variable has a direct effect on the outcome, but their interaction produces an effect [@Attia_et_al_2022].
-
-Conversely, based on the evidence presented in @sec-theory-issue1b, the study assumes that other relevant stimulus characteristics influence the judges' perceptions, leading to biases that resist cancellation and ultimately introduce dependencies between the stimuli.
-To address this, the study incorporates the judges' perceptions into the conceptual-population model, following the recommendations of @Andrich_1978 and @Wainer_et_al_1978.
+By incorporating judges' biases into the conceptual-population model, we assume that additional relevant characteristics of the stimuli are unlikely to be equally weighted or opposing.
+Thus, these characteristics may exert an uneven influence on judges' perceptions, creating biases in their judgments.
+Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases may ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
+346], hence the need for their integration.
+The full statistical elaboration of this point is presented in @sec-statistical.
 
@@
 <!-- 5. the preliminary DAG for CJ -->
-@fig-cj04_dag displays the DAG associated with [SCM -@fig-cj04_scm].
+@fig-cj04_dag displays the DAG associated with [SCM -@fig-cj03_scm].
 The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and CJ experiments.
-It also distinguishes between observable endogenous variables, such as the outcome $O_{abhijk}$, and variables considered latent according to the theory and practice of CJ, such as the discriminal difference, the perceived discriminal processes, the true discriminal processes, and the judges' perceptions.
+It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent according to the theory and practice of CJ, such as the the text's discriminal processes, their discriminal difference, and the judges' biases.
 
 <!-- 6. The problem of continuing this way -->
-It is easy to intuit that describing the CJ system by assigning an index and a variable to each discriminal process could quickly become intractable and cumbersome due to the large size of the DAG and the repetition of its components.
-Therefore, to maintain generality, the study groups pairs of discriminal processes with different indices into matrices that contain these pairs.
-First, the study defines the vectors $U_{A} \in \mathbb{N}^{A}$, $U_{I} \in \mathbb{N}^{I}$ containing the indices of the stimuli and individuals, where:
-
-$$
-U_{A} = [1, \dots, a, \dots, b, \dots, A] ;\quad U_{I} = [1, \dots, a, \dots, b, \dots, I]
-$$
-
-Then, the study defines the matrices of stimuli and individual comparisons, $P \in \mathbb{N}^{m \times 2}$ and $R \in \mathbb{N}^{n \times 2}$, respectively.
-Here, each row of $P$ represents a pairing $\{a,b\}$ of stimuli indices and each row of $R$ represents a pairing $\{h,i\}$ of individual indices, i.e.,
-
-$$
-P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\}
-$$
-
-Here, $m$ and $n$ represents the number of unique pairwise comparisons between stimuli, and unique pairwise comparison per individuals, which corresponds to the number of ways to choose two elements from $U_{A}$ and $U_{I}$, typically given by:
-
-$$
-m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!}
-$$
+It is easy to see that representing the CJ system using indices for a single pair of individuals and stimuli, along with the indices for the judgment number and the judge responsible for the comparison, can quickly become intractable and cumbersome.
+Thus, to maintain generality and simplify the notation, certain indices are grouped into vectors or matrices.
+First, we define the matrices $IA$ and $JK$ as:
+
+$$
+IA = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+\vdots & \vdots \\
+h & a \\
+\vdots & \vdots \\
+i & b \\
+\vdots & \vdots \\
+n_{I} & n_{A}-1 \\
+n_{I} & n_{A} \\
+\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
+JK = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+\vdots & \vdots \\
+j & k \\
+\vdots & \vdots \\
+n_{J} & n_{K}-1 \\
+n_{I} & n_{A} \\
+\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
+$$
+
+Each row of $IA$ represents a unique pairing of an individual $\{h,i\}$ with a stimulus $\{a,b\}$.
+Here, $n_{I}$ represents the number of individuals and $n_{A}$ the number of stimulus within individuals in the population.
+Similarly, each row of $JK$ represents a unique pairing of a judge $j$ with a judgment $k$.
+Here, $n_{J}$ represents the number of judges and $n_{K}$ the number of judgments performed by each judge in the population.
+
+<!-- Then, the study defines the matrices of stimuli and individual comparisons, $P \in \mathbb{N}^{m \times 2}$ and $R \in \mathbb{N}^{n \times 2}$, respectively.
+Here, each row of $P$ represents a pairing $\{a,b\}$ of stimuli indices and each row of $R$ represents a pairing $\{h,i\}$ of individual indices, i.e., -->
+
+<!-- $$ -->
+<!-- P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\} -->
+<!-- $$ -->
+
+<!-- Here, $m$ and $n$ represents the number of unique pairwise comparisons between stimuli, and unique pairwise comparison per individuals, which corresponds to the number of ways to choose two elements from $U_{A}$ and $U_{I}$, typically given by: -->
+
+<!-- $$ -->
+<!-- m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!} -->
+<!-- $$ -->
 
@@
 
-# Extending the BTL model {#sec-statistical}
+# Abandoning the BTL model {#sec-statistical}
 
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
+
+<!-- In other words, researchers observe the CJ outcome only when texts' discriminal processes and the judges biases create a *pure interaction effect*, meaning that neither of them has a direct effect on the outcome, but their interaction produces an effect [@Attia_et_al_2022].
+-->
 

### 2025-02-10 _tex/paper2.tex
Built site for gh-pages

@@
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
-additional relevant characteristics, which are unlikely to be equally
-weighted or opposing, can exert an uneven influence on judges'
-perceptions, creating biases in their judgments and, ultimately,
-introducing dependencies between stimuli
+additional relevant characteristics are unlikely to be equally weighted
+or opposing. As a result, they may exert an uneven influence on judges'
+perceptions, creating biases in their judgments. Furthermore, since the
+discriminal difference of the stimuli becomes an observable outcome only
+through the judges' perceptions, these biases may ultimately introduce
+dependencies between the stimuli
 \citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
@@
 In aggregate, the excessive reliance on Thurstone's Case V assumptions
-in the statistical analysis of CJ data can compromise the reliability of
-CJ's trait estimates. This overreliance, coupled with the apparent
-disconnect between CJ's approach to trait measurement and hypothesis
-testing, may also undermine the validity of these estimates
-\citep[pp.~2]{Perron_et_al_2015}. Thus, addressing these issues through
-a more systematic and integrated approach could provide significant
-statistical and measurement benefits. For this reason, the study employs
-the structural approach to causal inference.
+in the statistical analysis of comparative data can compromise the
+reliability of the trait estimates. This overreliance may also undermine
+their validity \citep[pp.~2]{Perron_et_al_2015}, especially when
+combined with the disconnect between CJ's approach to trait measurement
+and hypothesis testing. However, the structural approach to causal
+inference can address these issues by providing a systematic and
+integrated framework that enhances statistical accuracy while
+strengthening measurement reliability and validity.
 
@@
 appear throughout the document to illustrate various aspects of the
-proposed extension. In this scenario, researchers aim to determine: ``To
-what extent do different teaching methods influence students' ability to
-produce high-quality written texts?'' To investigate this, they design a
-CJ experiment by randomly assigning students (individuals) to two
-groups, each receiving a different teaching method. Judges then compare
-pairs of students' written texts (stimuli), producing a dichotomous
-outcome that reflects the relative quality of each text (trait). Based
-on this setup, researchers could reformulate the research question as
-the estimand: ``\emph{On average}, is there a difference in the ability
-to produce high-quality written texts between the two groups of
-students?''. Following current CJ practices, researchers would then rely
-on estimates from the BTL model, or its transformations, to approximate
-this estimand.
+proposed development. In this example, researchers aim to determine:
+``To what extent do different teaching methods influence students'
+ability to produce high-quality written texts?'' To investigate this,
+researchers could design a CJ experiment by randomly assigning students
+(individuals) to two groups, each receiving a different teaching method.
+Judges would then compare pairs of students' written texts (stimuli), to
+produce a dichotomous outcome that reflects the relative quality of each
+text (trait). Based on this setup, researchers could reformulate the
+research question as the estimand: ``\emph{On average}, is there a
+difference in the ability to produce high-quality written texts between
+the two groups of students?''. Finally, following current CJ practices,
+researchers would then rely on estimates from the BTL model, or its
+transformations, to approximate this estimand.
 
@@
 These limitations hinder the model's ability to identify various
-estimands relevant to CJ inquiries, including the one from the
+estimands relevant to CJ inquiries, including the one described in the
 motivating example. Identification is crucial because it is a necessary
-condition for ensuring consistent estimators. Without identification,
-\emph{consistency} --the property of an estimator whose estimates
-converge to the ``true'' value of the estimand as the data size
-approaches infinity \citep{Everitt_et_al_2010}-- cannot be achieved,
-even with ``infinite'' and error-free data. As a result, deriving
-meaningful insights from finite data also becomes unattainable
+condition for ensuring consistent estimators. \emph{Consistency} refers
+to the property of an estimator whose estimates converge to the ``true''
+value of the estimand as the data size approaches infinity
+\citep{Everitt_et_al_2010}. Without identification, consistency cannot
+be achieved, even with ``infinite'' and error-free data. Consequently,
+deriving meaningful insights from finite data becomes impossible
 \citep[pp.~5]{Schuessler_et_al_2023}.
@@
 
-Thus, this section addresses the identified limitations by extending
-Thurstone's general form (see Table~\ref{tbl-thurstone_cases}) through a
-systematic and integrated approach based on the structural approach to
-causal inference. Specifically, it leverages the capabilities of this
-approach to formalize the combination of the core theoretical principles
-outlined in Section~\ref{sec-thurstone_theory} with key assessment
-design features relevant to CJ experiments, such as the selection of
-judges, stimuli, and comparisons.
-
-In addition to overcoming the identified limitations, the extended model
-offers two key advantages. First, it clarifies the interactions among
-all actors and processes involved in CJ experiments. Second, it shifts
-the current comparative data analysis paradigm from passively accepting
-the model assumptions to actively testing if those assumptions fit the
-specific data under analysis. For more details on these assumptions,
-refer to Section~\ref{sec-theory-issues}.
+Thus, this section addresses the identified issues in
+Section~\ref{sec-theory-issues} by extending Thurstone's general form
+using the structural approach to causal inference. Specifically, it
+leverages the capabilities of this approach to formalize the combination
+of the core theoretical principles outlined in
+Section~\ref{sec-thurstone_theory} with key assessment design features
+relevant to CJ experiments, such as the selection of judges, stimuli,
+and comparisons. In addition to enhancing statistical accuracy and
+strengthening measurement reliability and validity, the approach offers
+two key advantages. First, it clarifies the interactions among all
+actors and processes involved in CJ experiments. Second, it shifts the
+current comparative data analysis paradigm from passively accepting the
+model assumptions to actively testing whether those assumptions fit the
+data under analysis.
 
 Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
-theoretical principles into what the study calls the
-\emph{conceptual-population model}. This model assumes that researchers
-have access to a conceptual population of comparison data, meaning data
-representing all repeated comparisons made by every available judge for
-each pair of stimuli produced by each pair of individuals in the
-population, thus its name. Conversely,
-Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
-features into what the study calls the \emph{sample-comparison model}.
-This model assumes a more realistic scenario where researchers only have
-access to a sample of judges, individuals, stimuli, and comparisons from
-the conceptual population.
+theoretical principles into what we call the \emph{conceptual-population
+model}. This model assumes that researchers have access to a conceptual
+population of comparison data, that is, data representing all repeated
+comparisons made by every available judge for each pair of stimuli
+produced by each pair of individuals in the population, hence its name.
+Conversely, Section~\ref{sec-theory-theoretical_SC} integrates the
+assessment design features into what we call the \emph{sample-comparison
+model}. This model assumes a more realistic scenario where researchers
+only have access to a sample of judges, individuals, stimuli, and
+comparisons from the conceptual population.
 
@@
 Before incorporating the theoretical principles into the
-\emph{conceptual-population model}, the study defines SCMs. SCMs are
-formal mathematical models characterized by a set of \emph{endogenous}
-variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
-functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
-variables are those whose causal mechanisms a researcher chooses to
-model \citep{Neal_2020}. In contrast, exogenous variables represent
-\emph{errors} or \emph{disturbances} arising from omitted factors that
-the investigator chooses not to model explicitly
-\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
-\emph{structural equations}, express the endogenous variables as
-non-parametric functions of other endogenous and exogenous variables.
-These functions use the symbol `\(:=\)' to denote the asymmetrical
-causal dependence between variables and the symbol `\(\:\bot\:\)' to
-represent \emph{d-separation}, a concept akin to (conditional)
-independence.
+\emph{conceptual-population model}, it is essential to define SCMs. SCMs
+are formal mathematical models characterized by a set of
+\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
+\(E\), and a set of functions \(F\)
+\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
+whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
+In contrast, exogenous variables represent \emph{errors} or
+\emph{disturbances} arising from omitted factors that the investigator
+chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
+the functions, referred to as \emph{structural equations}, express the
+endogenous variables as non-parametric functions of other endogenous and
+exogenous variables. These functions use the symbol `\(:=\)' to denote
+the asymmetrical causal dependence between variables and the symbol
+`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
+(conditional) independence.
 
@@
 Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
-motivating example. This SCM maps the relationship between the outcome
-of the CJ experiment \((O_{abhijk})\) to several other variables. In
-this model, the indices \(\{a,b\}\) represent the compared texts
-(stimuli), with \(a \neq b\). The indices \(\{h,i\}\) label the students
-who wrote the texts (individuals), where \(h = i\) is possible. The
-index \(j\) denotes the judge, and \(k\) denotes the comparison number,
-which accounts for experimental cases where judges perform multiple
-comparisons of the same texts written by the same students, i.e., a
-\emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}.
-Finally, the SCM presents the preliminary set of structural equations
-\(F = \{f_{O},f_{D},f_{PT}\}\), which define the non-parametric
-dependencies among these variables.
-
-Note that the SCM~\ref{fig-cj04_scm} incorporates not only the variables
-described by Thurstone's theory, such as the discriminal difference
-\((D_{abhijk})\) and the texts' \emph{true} discriminal processes
-\((T_{ahjk}, T_{bijk})\) (refer to Section~\ref{sec-thurstone_theory}),
-but also innovates by incorporating additional variables central to CJ
-experiments, such as the texts' \emph{perceived} discriminal processes
-\((PT_{ahjk}, PT_{bijk})\) and the judges' perceptions \((B_{kj})\).
-Together with the outcome, these variables constitute the preliminary
-set of endogenous variables
-\(V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
-
-By incorporating the texts' perceived discriminal processes into the
-conceptual-population model, this study considers a crucial aspect of CJ
-experiments that previous literature has overlooked. It recognizes that
-the texts' true discriminal processes become an observable outcome only
-through the judges' perceptions. In other words, researchers observe the
-CJ outcome only when both variables create a \emph{pure interaction
-effect}, meaning that neither variable has a direct effect on the
-outcome, but their interaction produces an effect
-\citep{Attia_et_al_2022}.
-
-Conversely, based on the evidence presented in
-Section~\ref{sec-theory-issue1b}, the study assumes that other relevant
-stimulus characteristics influence the judges' perceptions, leading to
-biases that resist cancellation and ultimately introduce dependencies
-between the stimuli. To address this, the study incorporates the judges'
-perceptions into the conceptual-population model, following the
-recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978}.
+motivating example. This SCM maps the relationship between the
+conceptual-population outcome \((O^{cp}_{abhijk})\) to several other
+variables. In this SCM, the subscripts \(\{a,b\}\) represent the
+compared texts (stimuli), with \(a \neq b\). The indices \(\{h,i\}\)
+label the students who wrote the texts (individuals), where \(h = i\) is
+possible. Moreover, the index \(j\) denotes the judge, and \(k\) denotes
+the judgment number, which accounts for experimental cases where judges
+perform multiple judgments of the same texts written by the same
+students, i.e., a \emph{repeated measures design}
+\citep[pp.~366-376]{Lawson_2015}.
+
+The other variables in SCM~\ref{fig-cj04_scm} include the texts'
+discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
+difference \((D_{abhijk})\), in line with Thurstone's theory (see
+Section~\ref{sec-thurstone_theory}). Additionally, based on the evidence
+presented in Section~\ref{sec-theory-issue1b} and the recommendations of
+\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, we incorporate the
+judges' biases \((B_{kj})\) into the model. Together with the outcome,
+these variables form the preliminary set of endogenous variables,
+\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
+Finally, the SCM shows the preliminary set of structural equations,
+\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
+among these variables.
+
+By incorporating judges' biases into the conceptual-population model, we
+assume that additional relevant characteristics of the stimuli are
+unlikely to be equally weighted or opposing. Thus, these characteristics
+may exert an uneven influence on judges' perceptions, creating biases in
+their judgments. Furthermore, since the discriminal difference of the
+stimuli becomes an observable outcome only through the judges'
+perceptions, these biases may ultimately introduce dependencies between
+the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}, hence the need
+for their integration. The full statistical elaboration of this point is
+presented in Section~\ref{sec-statistical}.
 
@@
 Figure~\ref{fig-cj04_dag} displays the DAG associated with
-SCM~\ref{fig-cj04_scm}. The graph shows the direction of causal
+SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
 influences among variables, reflecting the assumptions of Thurstone's
 theory and CJ experiments. It also distinguishes between observable
-endogenous variables, such as the outcome \(O_{abhijk}\), and variables
-considered latent according to the theory and practice of CJ, such as
-the discriminal difference, the perceived discriminal processes, the
-true discriminal processes, and the judges' perceptions.
-
-It is easy to intuit that describing the CJ system by assigning an index
-and a variable to each discriminal process could quickly become
-intractable and cumbersome due to the large size of the DAG and the
-repetition of its components. Therefore, to maintain generality, the
-study groups pairs of discriminal processes with different indices into
-matrices that contain these pairs. First, the study defines the vectors
-\(U_{A} \in \mathbb{N}^{A}\), \(U_{I} \in \mathbb{N}^{I}\) containing
-the indices of the stimuli and individuals, where:
+endogenous variables, such as the outcome, and variables considered
+latent according to the theory and practice of CJ, such as the the
+text's discriminal processes, their discriminal difference, and the
+judges' biases.
+
+It is easy to see that representing the CJ system using indices for a
+single pair of individuals and stimuli, along with the indices for the
+judgment number and the judge responsible for the comparison, can
+quickly become intractable and cumbersome. Thus, to maintain generality
+and simplify the notation, certain indices are grouped into vectors or
+matrices. First, we define the matrices \(IA\) and \(JK\) as:
 
 \[
-U_{A} = [1, \dots, a, \dots, b, \dots, A] ;\quad U_{I} = [1, \dots, a, \dots, b, \dots, I]
+IA = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+\vdots & \vdots \\
+h & a \\
+\vdots & \vdots \\
+i & b \\
+\vdots & \vdots \\
+n_{I} & n_{A}-1 \\
+n_{I} & n_{A} \\
+\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
+JK = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+\vdots & \vdots \\
+j & k \\
+\vdots & \vdots \\
+n_{J} & n_{K}-1 \\
+n_{I} & n_{A} \\
+\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
 \]
 
-Then, the study defines the matrices of stimuli and individual
-comparisons, \(P \in \mathbb{N}^{m \times 2}\) and
-\(R \in \mathbb{N}^{n \times 2}\), respectively. Here, each row of \(P\)
-represents a pairing \(\{a,b\}\) of stimuli indices and each row of
-\(R\) represents a pairing \(\{h,i\}\) of individual indices, i.e.,
-
-\[
-P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\}
-\]
-
-Here, \(m\) and \(n\) represents the number of unique pairwise
-comparisons between stimuli, and unique pairwise comparison per
-individuals, which corresponds to the number of ways to choose two
-elements from \(U_{A}\) and \(U_{I}\), typically given by:
-
-\[
-m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!}
-\]
+Each row of \(IA\) represents a unique pairing of an individual
+\(\{h,i\}\) with a stimulus \(\{a,b\}\). Here, \(n_{I}\) represents the
+number of individuals and \(n_{A}\) the number of stimulus within
+individuals in the population. Similarly, each row of \(JK\) represents
+a unique pairing of a judge \(j\) with a judgment \(k\). Here, \(n_{J}\)
+represents the number of judges and \(n_{K}\) the number of judgments
+performed by each judge in the population.
 
@@
 
-\section{Extending the BTL model}\label{sec-statistical}
+\section{Abandoning the BTL model}\label{sec-statistical}
 

### 2025-02-10 code/CJmod_add.R

@@ -6,7 +6,7 @@ data{
     int nsI;    // max. number of individuals
     int nsA;    // max. number of stimuli
     int nsJ;    // max. number of judges
-    int nsK;    // max. number of repeated comparisons
+    int nsK;    // max. number of judgments
     
     // categories discrete variables
     int sup_dXI;  // max. number of categories in XI
@@ -20,24 +20,24 @@ data{
     array[ns] int<lower=1, upper=nsI> Rs2;   // individual (right)
     array[ns] int<lower=1, upper=nsA> Ps2;   // stimuli (right)
     array[ns] int<lower=1, upper=nsJ> Us;    // judges
-    array[ns] int<lower=1, upper=nsK> Vs;    // repeated comparisons
+    array[ns] int<lower=1, upper=nsK> Vs;    // judgments
     array[ns] int<lower=0, upper=1> ORPUV;   // outcome (dichotomous)
     
     // individual-stimuli data
-    array[nsI*nsA] int<lower=1, upper=nsI> Is;      // individuals
-    array[nsI*nsA] int<lower=1, upper=nsA> As;      // stimuli
+    array[nsI*nsA] int<lower=1, upper=nsI> IA1s;    // individuals
+    array[nsI*nsA] int<lower=1, upper=nsA> IA2s;    // stimuli
     array[nsI*nsA] real XIc;                        // individual predictor (continuous)
     array[nsI*nsA] int<lower=1, upper=sup_dXI> XId; // individual predictor (discrete)
     array[nsI*nsA] real XAc;                        // stimuli predictor (continuous)
     array[nsI*nsA] int<lower=1, upper=sup_dXA> XAd; // stimuli predictor (discrete)
     
     // juges-repeated comparisons data
-    array[nsJ*nsK] int<lower=1, upper=nsJ> Js;      // judges
-    array[nsJ*nsK] int<lower=1, upper=nsK> Ks;      // repeated comparisons
+    array[nsJ*nsK] int<lower=1, upper=nsJ> JK1s;    // judges
[…]

### 2025-02-11 paper2.qmd
section:theoretical; process:write,figures

@@
 <!-- 2. the preliminary SCM for CJ -->
-@fig-cj04 illustrates the preliminary integration of theoretical principles into the conceptual-population model.
-Specifically, @fig-cj04_scm shows the SCM that underlies the motivating example.
-This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{abhijk})$ to several other variables.
-In this SCM, the subscripts $\{a,b\}$ represent the compared texts (stimuli), with $a \neq b$.
-The indices $\{h,i\}$ label the students who wrote the texts (individuals), where $h = i$ is possible.
-Moreover, the index $j$ denotes the judge, and $k$ denotes the judgment number, which accounts for experimental cases where judges perform multiple judgments of the same texts written by the same students, i.e., a *repeated measures design* [@Lawson_2015, pp.
+[SCM -@fig-cj03_scm] illustrates the preliminary integration of theoretical principles into the conceptual-population model.
+This SCM maps the relationship between the conceptual-population outcome, $(O^{cp}_{abhijk})$, and several related variables.
+The subscripts $a$ and $b$ represent the compared texts (stimuli), while the indices $h$ and $i$ label the students who authored them (individuals).
+Moreover, the index $j$ refers to the judge, and $k$ specifies the judgment number, accounting for experimental conditions where judges compare the same texts written by the same students multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
 366-376].
-
-The other variables in [SCM -@fig-cj04_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory).
-Additionally, based on the evidence presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, we incorporate the judges' biases $(B_{kj})$ into the model.
+It is important to note that the indexing system allows for comparisons between different texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
+However, it excludes cases where judges compare a student’s text to itself ($h = i$, $a = b$), as such a comparison lacks practical meaning.
+
+The other variables in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory).
+Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
 Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}$.
@@
 
-<!-- 3. What are the innovations of this? -->
-By incorporating judges' biases into the conceptual-population model, we assume that additional relevant characteristics of the stimuli are unlikely to be equally weighted or opposing.
-Thus, these characteristics may exert an uneven influence on judges' perceptions, creating biases in their judgments.
-Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases may ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
-346], hence the need for their integration.
-The full statistical elaboration of this point is presented in @sec-statistical.
-
 ::: {#fig-cj03 layout-nrow=2}
@@
 \begin{aligned}
-  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
-  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
+  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
+  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
 \end{aligned}
@@
 Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
-A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables.
-The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence.
-The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
-DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
-Thus, this study uses the magnified representation.
+A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables.
+The term *directed* indicates that edges or arrows extend from one node to another showing the direction of causal influence.
+The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
+DAGs depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
+The *standard representation* of DAGs typically omits exogenous variables for simplicity, while the *magnified representation* includes them to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+This section adopts the magnified representation to enhance clarity.
 
 <!-- 5. the preliminary DAG for CJ -->
-@fig-cj04_dag displays the DAG associated with [SCM -@fig-cj03_scm].
-The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and CJ experiments.
-It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent according to the theory and practice of CJ, such as the the text's discriminal processes, their discriminal difference, and the judges' biases.
+@fig-cj03_dag displays the DAG associated with [SCM -@fig-cj03_scm].
+The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and CJ's experiments.
+It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent according to the theory and practice of CJ, such as the text's discriminal processes, their discriminal difference, and the judges' biases.
 
 <!-- 6. The problem of continuing this way -->
-It is easy to see that representing the CJ system using indices for a single pair of individuals and stimuli, along with the indices for the judgment number and the judge responsible for the comparison, can quickly become intractable and cumbersome.
-Thus, to maintain generality and simplify the notation, certain indices are grouped into vectors or matrices.
-First, we define the matrices $IA$ and $JK$ as:
-
-$$
-IA = \begin{bmatrix}
+Although SCMs and DAGs do not necessitate specifying the data structure of the modeled system, defining these structures can sometimes enhance their clarity and simplify their notation.
+Thus, we organize certain indices into vectors or matrices.
+First, we define the matrices $IA$ and $JK$ as follows:
+
+$$
+IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
 1 & 1 \\
 1 & 2 \\
+\vdots & \vdots \\
+1 & n_{A}-1 \\
+1 & n_{A} \\
 \vdots & \vdots \\
@@
 n_{I} & n_{A} \\
-\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
-JK = \begin{bmatrix}
+\end{bmatrix} \in \mathbb{N} \; ; \quad
+JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
 1 & 1 \\
 1 & 2 \\
+\vdots & \vdots \\
+1 & n_{K}-1 \\
+1 & n_{K} \\
 \vdots & \vdots \\
@@
 n_{J} & n_{K}-1 \\
-n_{I} & n_{A} \\
-\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
-$$
-
-Each row of $IA$ represents a unique pairing of an individual $\{h,i\}$ with a stimulus $\{a,b\}$.
-Here, $n_{I}$ represents the number of individuals and $n_{A}$ the number of stimulus within individuals in the population.
-Similarly, each row of $JK$ represents a unique pairing of a judge $j$ with a judgment $k$.
-Here, $n_{J}$ represents the number of judges and $n_{K}$ the number of judgments performed by each judge in the population.
-
-<!-- Then, the study defines the matrices of stimuli and individual comparisons, $P \in \mathbb{N}^{m \times 2}$ and $R \in \mathbb{N}^{n \times 2}$, respectively.
-Here, each row of $P$ represents a pairing $\{a,b\}$ of stimuli indices and each row of $R$ represents a pairing $\{h,i\}$ of individual indices, i.e., -->
-
-<!-- $$ -->
-<!-- P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\} -->
-<!-- $$ -->
-
-<!-- Here, $m$ and $n$ represents the number of unique pairwise comparisons between stimuli, and unique pairwise comparison per individuals, which corresponds to the number of ways to choose two elements from $U_{A}$ and $U_{I}$, typically given by: -->
-
-<!-- $$ -->
-<!-- m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!} -->
-<!-- $$ -->
-
-<!-- Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in   -->
+n_{J} & n_{K} \\
+\end{bmatrix} \in \mathbb{N}
+$$ {#eq-mat12}
+
+where each row of $IA$ corresponds to a unique pairing of individuals $h, i$ with stimuli $a, b$.
+Here, $n_{I}$ represents the number of individuals, while $n_{A}$ denotes the number of stimuli available per individual.
+Thus, the matrix $IA$ is composed of natural numbers $\mathbb{N}$, with $n_{I} \cdot n_{A}$ rows and $2$ columns.
+Similarly, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment $k$.
+In this case, $n_{J}$ indicates the total number of judges, and $n_{K}$ represents the number of judgments each judge performs.
+Therefore, the matrix $JK$ is also composed of natural numbers $\mathbb{N},$ with $n_{J} \cdot n_{K}$ rows and $2$ columns.
+
+Importantly, since the discriminal difference of the stimuli becomes an observable outcome only through judges' perceptions, it is necessary to define the matrix $RPUV$, where each row of the matrix uniquely pairs the corresponding rows from the matrices $IA$ and $JK$ as follows:
+
+$$
+RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
+1 & 1 & 1 & 2 & 1 & 1 \\
+1 & 1 & 1 & 2 & 1 & 2 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
+1 & 1 & 1 & 2 & 1 & n_{K} \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+h & a & i & b & j & k \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 2 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}-1 \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K} \\
+\end{bmatrix} \in \mathbb{N}
+$$ {#eq-mat3}
+
+Here, the $RPUV$ matrix is also composed of natural numbers $\mathbb{N}$, and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
+The dimension of the matrix results from the conceptual-population model's assumption that researchers have access to data representing all repeated comparisons made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+Here, the term ${n_{I} \cdot n_{a} \choose 2}$ represents the Binomial coefficient, which quantifies the number of individual-stimuli comparison pairs, and it is defined as:
+
+$$
+{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
+$$
+
+Using the motivating example outlined in @sec-theoretical, we can visualize the shape of these structures.
+For illustrative purposes, assume the conceptual population consists of $n_{I} = 5$ individuals, each with $n_{A} = 2$ available stimuli.
+Additionally, assume that each pair of stimuli is evaluated by $n_{J} = 3$ judges, with each judge making $n_{K} = 3$ comparative judgments.
+Given this setup, the matrices in ([-@eq-mat12]) and ([-@eq-mat3]) are represented as follows:
+
+$$
+IA_{10 \times 2} = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+2 & 1 \\
+2 & 2 \\
+3 & 1 \\
+3 & 2 \\
+4 & 1 \\
+4 & 2 \\
+5 & 1 \\
+5 & 2 \\
+\end{bmatrix}\; ; \quad
+JK_{9 \times 2} = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+1 & 3 \\
+2 & 1 \\
+2 & 2 \\
+2 & 3 \\
+3 & 1 \\
+3 & 2 \\
+3 & 3 \\
+\end{bmatrix} \; ; \quad
+RPUV_{405  \times 6} = \begin{bmatrix}
+1 & 1 & 1 & 2 & 1 & 1 \\
+1 & 1 & 1 & 2 & 1 & 2 \\
+1 & 1 & 1 & 2 & 1 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 5 & 1 & 1 & 1 \\
+1 & 1 & 5 & 1 & 1 & 2 \\
+1 & 1 & 5 & 1 & 1 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 5 & 1 & 3 & 1 \\
+1 & 1 & 5 & 1 & 3 & 2 \\
+1 & 1 & 5 & 1 & 3 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+4 & 2 & 5 & 2 & 3 & 1 \\
+4 & 2 & 5 & 2 & 3 & 2 \\
+4 & 2 & 5 & 2 & 3 & 3 \\
+\end{bmatrix}
+$$
+
+With these matrices we can express an equivalent version of [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in a vectorized form.
 
@@
 
-## Statistics and Causal inference {#sec-appendixB .appendix appendix-style=plain}
+## Statistical and Causal inference {#sec-appendixB .appendix appendix-style=plain}
 

### 2025-02-11 _tex/paper2.tex
Built site for gh-pages

@@
 
-Figure~\ref{fig-cj04} illustrates the preliminary integration of
-theoretical principles into the conceptual-population model.
-Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
-motivating example. This SCM maps the relationship between the
-conceptual-population outcome \((O^{cp}_{abhijk})\) to several other
-variables. In this SCM, the subscripts \(\{a,b\}\) represent the
-compared texts (stimuli), with \(a \neq b\). The indices \(\{h,i\}\)
-label the students who wrote the texts (individuals), where \(h = i\) is
-possible. Moreover, the index \(j\) denotes the judge, and \(k\) denotes
-the judgment number, which accounts for experimental cases where judges
-perform multiple judgments of the same texts written by the same
-students, i.e., a \emph{repeated measures design}
-\citep[pp.~366-376]{Lawson_2015}.
-
-The other variables in SCM~\ref{fig-cj04_scm} include the texts'
+SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
+theoretical principles into the conceptual-population model. This SCM
+maps the relationship between the conceptual-population outcome,
+\((O^{cp}_{abhijk})\), and several related variables. The subscripts
+\(a\) and \(b\) represent the compared texts (stimuli), while the
+indices \(h\) and \(i\) label the students who authored them
+(individuals). Moreover, the index \(j\) refers to the judge, and \(k\)
+specifies the judgment number, accounting for experimental conditions
+where judges compare the same texts written by the same students
+multiple times, i.e., a \emph{repeated measures design}
+\citep[pp.~366-376]{Lawson_2015}. It is important to note that the
+indexing system allows for comparisons between different texts written
+by the same student (\(h = i\), \(a \neq b\)) and between texts written
+by distinct students (\(h \neq i\), where \(a = b\) is possible).
+However, it excludes cases where judges compare a student's text to
+itself (\(h = i\), \(a = b\)), as such a comparison lacks practical
+meaning.
+
+The other variables in SCM~\ref{fig-cj03_scm} include the texts'
 discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
 difference \((D_{abhijk})\), in line with Thurstone's theory (see
-Section~\ref{sec-thurstone_theory}). Additionally, based on the evidence
-presented in Section~\ref{sec-theory-issue1b} and the recommendations of
-\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, we incorporate the
-judges' biases \((B_{kj})\) into the model. Together with the outcome,
-these variables form the preliminary set of endogenous variables,
+Section~\ref{sec-thurstone_theory}). Additionally, based on the
+arguments presented in Section~\ref{sec-theory-issue1b} and the
+recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
+the SCM incorporates the judges' biases \((B_{kj})\). Together with the
+outcome, these variables form the preliminary set of endogenous
+variables,
 \(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
@@
 
-By incorporating judges' biases into the conceptual-population model, we
-assume that additional relevant characteristics of the stimuli are
-unlikely to be equally weighted or opposing. Thus, these characteristics
-may exert an uneven influence on judges' perceptions, creating biases in
-their judgments. Furthermore, since the discriminal difference of the
-stimuli becomes an observable outcome only through the judges'
-perceptions, these biases may ultimately introduce dependencies between
-the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}, hence the need
-for their integration. The full statistical elaboration of this point is
-presented in Section~\ref{sec-statistical}.
-
 \begin{figure}[H]
@@
 \citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
-consisting of nodes connected by edges, where the nodes represent random
-variables. The term \emph{directed} means that the edges extend from one
-node to another, with arrows indicating the direction of causal
-influence. The term \emph{acyclic} implies that the causal influences do
+consisting of nodes connected by edges, where nodes represent random
+variables. The term \emph{directed} indicates that edges or arrows
+extend from one node to another showing the direction of causal
+influence. The term \emph{acyclic} means that the causal influences do
 not form loops, ensuring the influences do not cycle back on themselves
-\citep{McElreath_2020}. DAGs represent observed variables as solid black
-circles, while they use open circles for unobserved (latent) variables
-\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
-DAGs typically omits exogenous variables for simplicity, the
-\emph{magnified representation} includes them, offering the advantage of
-highlighting potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}. Thus, this study uses the magnified
-representation.
-
-Figure~\ref{fig-cj04_dag} displays the DAG associated with
+\citep{McElreath_2020}. DAGs depict observed variables as solid black
+circles and unobserved (latent) variables as open circles
+\citep{Morgan_et_al_2014}. The \emph{standard representation} of DAGs
+typically omits exogenous variables for simplicity, while the
+\emph{magnified representation} includes them to highlight potential
+issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}. This section adopts the magnified
+representation to enhance clarity.
+
+Figure~\ref{fig-cj03_dag} displays the DAG associated with
 SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
 influences among variables, reflecting the assumptions of Thurstone's
-theory and CJ experiments. It also distinguishes between observable
+theory and CJ's experiments. It also distinguishes between observable
 endogenous variables, such as the outcome, and variables considered
-latent according to the theory and practice of CJ, such as the the
-text's discriminal processes, their discriminal difference, and the
-judges' biases.
-
-It is easy to see that representing the CJ system using indices for a
-single pair of individuals and stimuli, along with the indices for the
-judgment number and the judge responsible for the comparison, can
-quickly become intractable and cumbersome. Thus, to maintain generality
-and simplify the notation, certain indices are grouped into vectors or
-matrices. First, we define the matrices \(IA\) and \(JK\) as:
-
-\[
-IA = \begin{bmatrix}
+latent according to the theory and practice of CJ, such as the text's
+discriminal processes, their discriminal difference, and the judges'
+biases.
+
+Although SCMs and DAGs do not necessitate specifying the data structure
+of the modeled system, defining these structures can sometimes enhance
+their clarity and simplify their notation. Thus, we organize certain
+indices into vectors or matrices. First, we define the matrices \(IA\)
+and \(JK\) as follows:
+
+\begin{equation}\phantomsection\label{eq-mat12}{
+IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
 1 & 1 \\
 1 & 2 \\
+\vdots & \vdots \\
+1 & n_{A}-1 \\
+1 & n_{A} \\
 \vdots & \vdots \\
@@
 n_{J} & n_{K}-1 \\
-n_{I} & n_{A} \\
-\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
+n_{J} & n_{K} \\
+\end{bmatrix} \in \mathbb{N}
+}\end{equation}
+
+where each row of \(IA\) corresponds to a unique pairing of individuals
+\(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents the number of
+individuals, while \(n_{A}\) denotes the number of stimuli available per
+individual. Thus, the matrix \(IA\) is composed of natural numbers
+\(\mathbb{N}\), with \(n_{I} \cdot n_{A}\) rows and \(2\) columns.
+Similarly, each row of \(JK\) specifies a unique pairing of a judge
+\(j\) with a judgment \(k\). In this case, \(n_{J}\) indicates the total
+number of judges, and \(n_{K}\) represents the number of judgments each
+judge performs. Therefore, the matrix \(JK\) is also composed of natural
+numbers \(\mathbb{N},\) with \(n_{J} \cdot n_{K}\) rows and \(2\)
+columns.
+
+Importantly, since the discriminal difference of the stimuli becomes an
+observable outcome only through judges' perceptions, it is necessary to
+define the matrix \(RPUV\), where each row of the matrix uniquely pairs
+the corresponding rows from the matrices \(IA\) and \(JK\) as follows:
+
+\begin{equation}\phantomsection\label{eq-mat3}{
+RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
+1 & 1 & 1 & 2 & 1 & 1 \\
+1 & 1 & 1 & 2 & 1 & 2 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
+1 & 1 & 1 & 2 & 1 & n_{K} \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+h & a & i & b & j & k \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 2 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}-1 \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K} \\
+\end{bmatrix} \in \mathbb{N}
+}\end{equation}
+
+Here, the \(RPUV\) matrix is also composed of natural numbers
+\(\mathbb{N}\), and has a dimension of
+\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
+columns. The dimension of the matrix results from the
+conceptual-population model's assumption that researchers have access to
+data representing all repeated comparisons made by every available judge
+for each pair of stimuli produced by each pair of individuals in the
+population. Here, the term \({n_{I} \cdot n_{a} \choose 2}\) represents
+the Binomial coefficient, which quantifies the number of
+individual-stimuli comparison pairs, and it is defined as:
+
+\[
+{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
 \]
 
-Each row of \(IA\) represents a unique pairing of an individual
-\(\{h,i\}\) with a stimulus \(\{a,b\}\). Here, \(n_{I}\) represents the
-number of individuals and \(n_{A}\) the number of stimulus within
-individuals in the population. Similarly, each row of \(JK\) represents
-a unique pairing of a judge \(j\) with a judgment \(k\). Here, \(n_{J}\)
-represents the number of judges and \(n_{K}\) the number of judgments
-performed by each judge in the population.
+Using the motivating example outlined in Section~\ref{sec-theoretical},
+we can visualize the shape of these structures. For illustrative
+purposes, assume the conceptual population consists of \(n_{I} = 5\)
+individuals, each with \(n_{A} = 2\) available stimuli. Additionally,
+assume that each pair of stimuli is evaluated by \(n_{J} = 3\) judges,
+with each judge making \(n_{K} = 3\) comparative judgments. Given this
+setup, the matrices in (\ref{eq-mat12}) and (\ref{eq-mat3}) are
+represented as follows:
+
+\[
+IA_{10 \times 2} = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+2 & 1 \\
+2 & 2 \\
+3 & 1 \\
+3 & 2 \\
+4 & 1 \\
+4 & 2 \\
+5 & 1 \\
+5 & 2 \\
+\end{bmatrix}\; ; \quad
+JK_{9 \times 2} = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+1 & 3 \\
+2 & 1 \\
+2 & 2 \\
+2 & 3 \\
+3 & 1 \\
+3 & 2 \\
+3 & 3 \\
+\end{bmatrix} \; ; \quad
+RPUV_{405  \times 6} = \begin{bmatrix}
+1 & 1 & 1 & 2 & 1 & 1 \\
+1 & 1 & 1 & 2 & 1 & 2 \\
+1 & 1 & 1 & 2 & 1 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 5 & 1 & 1 & 1 \\
+1 & 1 & 5 & 1 & 1 & 2 \\
+1 & 1 & 5 & 1 & 1 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 5 & 1 & 3 & 1 \\
+1 & 1 & 5 & 1 & 3 & 2 \\
+1 & 1 & 5 & 1 & 3 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+4 & 2 & 5 & 2 & 3 & 1 \\
+4 & 2 & 5 & 2 & 3 & 2 \\
+4 & 2 & 5 & 2 & 3 & 3 \\
+\end{bmatrix}
+\]
+
+With these matrices we can express an equivalent version of
+SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in a vectorized form.
 
@@
 
-\subsection{Statistics and Causal inference}\label{sec-appendixB}
+\subsection{Statistical and Causal inference}\label{sec-appendixB}
 

### 2025-02-12 paper2.qmd
section:theoretical; process:write,figures

@@
 Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical handling of this additional source of dependence between stimuli has been inadequate.
-For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 This approach, however, can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
@@
 Researchers then typically use these estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
-For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
@@
 Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we call the *conceptual-population model*.
-This model assumes that researchers have access to a conceptual population of comparison data, that is, data representing all repeated comparisons made by every available judge for each pair of stimuli produced by each pair of individuals in the population, hence its name.
+This model assumes that researchers have access to comparison data from a conceptual population, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population, hence its name.
 Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we call the *sample-comparison model*.
@@
 <!-- 2. the preliminary SCM for CJ -->
-[SCM -@fig-cj03_scm] illustrates the preliminary integration of theoretical principles into the conceptual-population model.
-This SCM maps the relationship between the conceptual-population outcome, $(O^{cp}_{abhijk})$, and several related variables.
+[SCM -@fig-cj03_scm] illustrates the preliminary integration of theoretical principles into the conceptual-population model of the motivating example.
+This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{abhijk})$ and several related variables.
 The subscripts $a$ and $b$ represent the compared texts (stimuli), while the indices $h$ and $i$ label the students who authored them (individuals).
@@
 366-376].
-It is important to note that the indexing system allows for comparisons between different texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
-However, it excludes cases where judges compare a student’s text to itself ($h = i$, $a = b$), as such a comparison lacks practical meaning.
-
-The other variables in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory).
-Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
+Notably, the indexing system allow comparisons between different texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
+However, it excludes cases where judges compare a student’s text to itself ($h = i$, $a = b$), as such a comparison lacks practical meaning within the CJ framework.
+
+The other variables present in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory).
+Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the conceptual-population model incorporates the judges' biases $(B_{kj})$.
 Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}$.
@@
 
+<!-- @Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
+Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
+Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
+26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap.
+15], depending on the theoretical and statistical treatment of judges' biases.
+-->
+
 ::: {#fig-cj03 layout-nrow=2}
@@
 
-CJ's outcome, discriminal difference, discriminal processes and judges' biases
+Conceptual-population model, preliminary integration of theoretical principles.
 :::
@@
 @fig-cj03_dag displays the DAG associated with [SCM -@fig-cj03_scm].
-The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and CJ's experiments.
-It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent according to the theory and practice of CJ, such as the text's discriminal processes, their discriminal difference, and the judges' biases.
+This graph illustrates the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and the practical aspects of CJ's experiments.
+It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent within the CJ framework, such as the text's discriminal processes, their discriminal difference, and the judges' biases.
 
 <!-- 6. The problem of continuing this way -->
-Although SCMs and DAGs do not necessitate specifying the data structure of the modeled system, defining these structures can sometimes enhance their clarity and simplify their notation.
-Thus, we organize certain indices into vectors or matrices.
+Although specifying a system's data structure is not necessary requirement for using SCMs and DAGs, defining one can improve clarity and simplify their notation.
+Thus, we define specific data structures to facilitate the description of the CJ system.
 First, we define the matrices $IA$ and $JK$ as follows:
@@
 Here, $n_{I}$ represents the number of individuals, while $n_{A}$ denotes the number of stimuli available per individual.
-Thus, the matrix $IA$ is composed of natural numbers $\mathbb{N}$, with $n_{I} \cdot n_{A}$ rows and $2$ columns.
+Thus, the matrix $IA$ is composed of natural numbers $\mathbb{N}$, with $(n_{I} \cdot n_{A})$ rows and $2$ columns.
 Similarly, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment $k$.
 In this case, $n_{J}$ indicates the total number of judges, and $n_{K}$ represents the number of judgments each judge performs.
-Therefore, the matrix $JK$ is also composed of natural numbers $\mathbb{N},$ with $n_{J} \cdot n_{K}$ rows and $2$ columns.
+Therefore, the matrix $JK$ is also composed of natural numbers $\mathbb{N},$ with $(n_{J} \cdot n_{K})$ rows and $2$ columns.
 
@@
 
-Here, the $RPUV$ matrix is also composed of natural numbers $\mathbb{N}$, and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
-The dimension of the matrix results from the conceptual-population model's assumption that researchers have access to data representing all repeated comparisons made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
-Here, the term ${n_{I} \cdot n_{a} \choose 2}$ represents the Binomial coefficient, which quantifies the number of individual-stimuli comparison pairs, and it is defined as:
+The $RPUV$ matrix is also composed of natural numbers $\mathbb{N}$, and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
+The dimension of the matrix results from the conceptual-population model's assumption that researchers have access to data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs in the conceptual population.
+The coefficient is defined as:
 
@@
 
-Using the motivating example outlined in @sec-theoretical, we can visualize the shape of these structures.
-For illustrative purposes, assume the conceptual population consists of $n_{I} = 5$ individuals, each with $n_{A} = 2$ available stimuli.
-Additionally, assume that each pair of stimuli is evaluated by $n_{J} = 3$ judges, with each judge making $n_{K} = 3$ comparative judgments.
-Given this setup, the matrices in ([-@eq-mat12]) and ([-@eq-mat3]) are represented as follows:
+Readers can better visualize the structure of these matrices by considering relatively small values for the parameter corresponding to number of individuals $(n_{I} = 5)$, the number of available stimuli per individual $(n_{A} = 2)$, the total number of judges $(n_{J} = 3)$, and the number of judgments each judge performs $(n_{K} = 3)$.
+With these values, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) are represented as follows:
 
@@
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-4 & 2 & 5 & 2 & 3 & 1 \\
-4 & 2 & 5 & 2 & 3 & 2 \\
-4 & 2 & 5 & 2 & 3 & 3 \\
+5 & 1 & 5 & 2 & 3 & 1 \\
+5 & 1 & 5 & 2 & 3 & 2 \\
+5 & 1 & 5 & 2 & 3 & 3 \\
 \end{bmatrix}
@@
 
-With these matrices we can express an equivalent version of [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in a vectorized form.
+Thus, with the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can now re-express the [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
+In this form, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are no longer scalar values but are instead vectors.
+These vectors encompass all the values from the conceptual population.
+Specifically, $O^{cp}_{RPUV}$ and $D_{RPUV}$ are vectors with a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ values each, while $T_{IA}$ and $B_{JK}$ have dimensions of $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$ values, respectively.
 
@@
 
-CJ's outcome, discriminal difference, discriminal processes and judges' biases
-:::
+Conceptual-population model, vectorized form of [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag].
+:::
+
+Finally, building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+138] and Item Response Theory (IRT) [@Fox_2010, chap.
+6; @vanderLinden_et_al_2017_I, chap.
+24], the conceptual-population model includes two (hierarchical) structural components.
+Each component defines how different manifest or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
+This addition enables analyses that simultaneously account for the traits and their associated uncertainties, linking the trait measurement process with hypothesis testing, thereby reducing bias and enhancing the measurement precision.
+
+<!-- principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+138] and IRT [@Fox_2010, chap.
+6; @vanderLinden_et_al_2017_I, chap.
+24] recommend conducting these analyses and tests within a structural model.
+A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
+This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
+-->
+
+@fig-cj09 displays the two (hierarchical) structural components in the motivating example.
+Specifically, the [DAG -@fig-cj09_dag] shows that the first component describes how the students' overall writing quality trait $T_{I}$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written quality trait $T_{IA}$.
+The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' quality that neither the students' traits nor other relevant text-related variables can explain.
+In this description, $T_{I}$ is a vectors of dimension $n_{I}$, $e_{IA}$ is a vector of dimension $(n_{I} \cdot n_{A})$, while $X_{IA}$ is a matrix with $(n_{I} \cdot n_{A})$ rows and $p_{IA}$ columns.
+
+At a higher level in the hierarchy, the [DAG -@fig-cj09_dag] also describes how a set of *relevant* student-related variables $X_{I}$ causally influence the students' overall writing quality trait $T_{I}$.
+Additionally, the DAG accounts for idiosyncratic errors $e_{I}$ that capture variations in the students' writing quality that the relevant student-related variables cannot explain.
+Here, $e_{I}$ is a vector of dimension $n_{I}$, and $X_{I}$ is a matrix with $n_{I}$ rows and $p_{I}$ columns.
+
+Similarly, the [DAG -@fig-cj09_dag] shows that the second component describes how the judges' overall bias $B_{J}$ and a set of *relevant* judgment-related variables $Z_{JK}$ causally influence the biases associated with each text $B_{JK}$.
+The DAG also highlights the idiosyncratic errors $e_{JK}$ that capture variations in the texts' biases that neither the judges' bias nor other relevant judgment-related variables can explain.
+In this description, $B_{J}$ is a vectors of dimension $n_{J}$, $e_{JK}$ is a vector of dimension $(n_{J} \cdot n_{K})$, while $Z_{JK}$ is a matrix with $(n_{J} \cdot n_{K})$ rows and $p_{JK}$ columns.
+
+Furthermore, at a higher level, the [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $Z_{J}$ causally influence the judges' overall bias $B_{J}$.
+The DAG also accounts for idiosyncratic errors $e_{J}$, which capture the variations in the judges' biases that the relevant judge-related variables cannot explain.
+Here, $e_{J}$ is a vector of dimension $n_{J}$, and $Z_{J}$ is a matrix with $n_{J}$ rows and $p_{J}$ columns.
 
@@
 
-CJ's outcome, discriminal difference, discriminal processes and judges' biases
-:::
-
-<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
+Conceptual-population model, full integration of theoretical principles.
+:::
+
+A cautious reader will notice that these descriptions rely on several sets of relevant variables.
+In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refer to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
+37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+
+Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
+This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
+Here, *conditioning* means restricting the statistical analysis to a subset of the population defined by the values of a given variable [@Neal_2020, pp.
+32].
+
+Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
+Nevertheless, as @Kohler_et_al_2019 [pp.
+150] emphasized, drawing conclusions about the real world from observed data inevitably requires assumptions, and this requirement holds true for both observational and experimental data.
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
 
@@
 
-<!-- Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data.
--->
-
 <!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
@@
 
-<!-- In other words, researchers observe the CJ outcome only when texts' discriminal processes and the judges biases create a *pure interaction effect*, meaning that neither of them has a direct effect on the outcome, but their interaction produces an effect [@Attia_et_al_2022].
+<!-- Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data.
 -->
@@
 <!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
+
+<!-- In other words, researchers observe the CJ outcome only when texts' discriminal processes and the judges biases create a *pure interaction effect*, meaning that neither of them has a direct effect on the outcome, but their interaction produces an effect [@Attia_et_al_2022].
+-->
 

### 2025-02-12 _tex/paper2.tex
Built site for gh-pages

@@
 multiple samples of stimuli from the same individuals, researchers
-frequently rely on (average) estimated BTL scores to conduct subsequent
+frequently rely on (averaged) estimated BTL scores to conduct subsequent
 analyses and tests at the individual hierarchical level
@@
 or test hypotheses related to the underlying trait of interest
-\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
 
@@
 theoretical principles into what we call the \emph{conceptual-population
-model}. This model assumes that researchers have access to a conceptual
-population of comparison data, that is, data representing all repeated
-comparisons made by every available judge for each pair of stimuli
-produced by each pair of individuals in the population, hence its name.
-Conversely, Section~\ref{sec-theory-theoretical_SC} integrates the
-assessment design features into what we call the \emph{sample-comparison
-model}. This model assumes a more realistic scenario where researchers
-only have access to a sample of judges, individuals, stimuli, and
-comparisons from the conceptual population.
+model}. This model assumes that researchers have access to comparison
+data from a conceptual population, that is, data representing all
+repeated judgments made by every available judge for each pair of
+stimuli produced by each pair of individuals in the population, hence
+its name. Conversely, Section~\ref{sec-theory-theoretical_SC} integrates
+the assessment design features into what we call the
+\emph{sample-comparison model}. This model assumes a more realistic
+scenario where researchers only have access to a sample of judges,
+individuals, stimuli, and comparisons from the conceptual population.
 
@@
 SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
-theoretical principles into the conceptual-population model. This SCM
-maps the relationship between the conceptual-population outcome,
-\((O^{cp}_{abhijk})\), and several related variables. The subscripts
-\(a\) and \(b\) represent the compared texts (stimuli), while the
-indices \(h\) and \(i\) label the students who authored them
-(individuals). Moreover, the index \(j\) refers to the judge, and \(k\)
-specifies the judgment number, accounting for experimental conditions
-where judges compare the same texts written by the same students
-multiple times, i.e., a \emph{repeated measures design}
-\citep[pp.~366-376]{Lawson_2015}. It is important to note that the
-indexing system allows for comparisons between different texts written
-by the same student (\(h = i\), \(a \neq b\)) and between texts written
-by distinct students (\(h \neq i\), where \(a = b\) is possible).
-However, it excludes cases where judges compare a student's text to
-itself (\(h = i\), \(a = b\)), as such a comparison lacks practical
-meaning.
-
-The other variables in SCM~\ref{fig-cj03_scm} include the texts'
+theoretical principles into the conceptual-population model of the
+motivating example. This SCM maps the relationship between the
+conceptual-population outcome \((O^{cp}_{abhijk})\) and several related
+variables. The subscripts \(a\) and \(b\) represent the compared texts
+(stimuli), while the indices \(h\) and \(i\) label the students who
+authored them (individuals). Moreover, the index \(j\) refers to the
+judge, and \(k\) specifies the judgment number, accounting for
+experimental conditions where judges compare the same texts written by
+the same students multiple times, i.e., a \emph{repeated measures
+design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system
+allow comparisons between different texts written by the same student
+(\(h = i\), \(a \neq b\)) and between texts written by distinct students
+(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
+where judges compare a student's text to itself (\(h = i\), \(a = b\)),
+as such a comparison lacks practical meaning within the CJ framework.
+
+The other variables present in SCM~\ref{fig-cj03_scm} include the texts'
 discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
@@
 recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
-the SCM incorporates the judges' biases \((B_{kj})\). Together with the
-outcome, these variables form the preliminary set of endogenous
-variables,
+the conceptual-population model incorporates the judges' biases
+\((B_{kj})\). Together with the outcome, these variables form the
+preliminary set of endogenous variables,
 \(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
@@
 
-\caption{\label{fig-cj03}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
+\caption{\label{fig-cj03}Conceptual-population model, preliminary
+integration of theoretical principles.}
 
@@
 Figure~\ref{fig-cj03_dag} displays the DAG associated with
-SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
+SCM~\ref{fig-cj03_scm}. This graph illustrates the direction of causal
 influences among variables, reflecting the assumptions of Thurstone's
-theory and CJ's experiments. It also distinguishes between observable
-endogenous variables, such as the outcome, and variables considered
-latent according to the theory and practice of CJ, such as the text's
-discriminal processes, their discriminal difference, and the judges'
-biases.
-
-Although SCMs and DAGs do not necessitate specifying the data structure
-of the modeled system, defining these structures can sometimes enhance
-their clarity and simplify their notation. Thus, we organize certain
-indices into vectors or matrices. First, we define the matrices \(IA\)
-and \(JK\) as follows:
+theory and the practical aspects of CJ's experiments. It also
+distinguishes between observable endogenous variables, such as the
+outcome, and variables considered latent within the CJ framework, such
+as the text's discriminal processes, their discriminal difference, and
+the judges' biases.
+
+Although specifying a system's data structure is not necessary
+requirement for using SCMs and DAGs, defining one can improve clarity
+and simplify their notation. Thus, we define specific data structures to
+facilitate the description of the CJ system. First, we define the
+matrices \(IA\) and \(JK\) as follows:
 
@@
 individual. Thus, the matrix \(IA\) is composed of natural numbers
-\(\mathbb{N}\), with \(n_{I} \cdot n_{A}\) rows and \(2\) columns.
+\(\mathbb{N}\), with \((n_{I} \cdot n_{A})\) rows and \(2\) columns.
 Similarly, each row of \(JK\) specifies a unique pairing of a judge
@@
 judge performs. Therefore, the matrix \(JK\) is also composed of natural
-numbers \(\mathbb{N},\) with \(n_{J} \cdot n_{K}\) rows and \(2\)
+numbers \(\mathbb{N},\) with \((n_{J} \cdot n_{K})\) rows and \(2\)
 columns.
@@
 
-Here, the \(RPUV\) matrix is also composed of natural numbers
-\(\mathbb{N}\), and has a dimension of
+The \(RPUV\) matrix is also composed of natural numbers \(\mathbb{N}\),
+and has a dimension of
 \({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
@@
 conceptual-population model's assumption that researchers have access to
-data representing all repeated comparisons made by every available judge
+data representing all repeated judgments made by every available judge
 for each pair of stimuli produced by each pair of individuals in the
-population. Here, the term \({n_{I} \cdot n_{a} \choose 2}\) represents
-the Binomial coefficient, which quantifies the number of
-individual-stimuli comparison pairs, and it is defined as:
+population. Here, the term \({n_{I} \cdot n_{A} \choose 2}\) represents
+the Binomial coefficient, which quantifies the total number of
+individual-stimuli comparison pairs in the conceptual population. The
+coefficient is defined as:
 
@@
 
-Using the motivating example outlined in Section~\ref{sec-theoretical},
-we can visualize the shape of these structures. For illustrative
-purposes, assume the conceptual population consists of \(n_{I} = 5\)
-individuals, each with \(n_{A} = 2\) available stimuli. Additionally,
-assume that each pair of stimuli is evaluated by \(n_{J} = 3\) judges,
-with each judge making \(n_{K} = 3\) comparative judgments. Given this
-setup, the matrices in (\ref{eq-mat12}) and (\ref{eq-mat3}) are
-represented as follows:
+Readers can better visualize the structure of these matrices by
+considering relatively small values for the parameter corresponding to
+number of individuals \((n_{I} = 5)\), the number of available stimuli
+per individual \((n_{A} = 2)\), the total number of judges
+\((n_{J} = 3)\), and the number of judgments each judge performs
+\((n_{K} = 3)\). With these values, the matrices described in equations
+(\ref{eq-mat12}) and (\ref{eq-mat3}) are represented as follows:
 
@@
 
-With these matrices we can express an equivalent version of
-SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in a vectorized form.
+Thus, with the data structure defined by the matrices in equations
+(\ref{eq-mat12}) and (\ref{eq-mat3}), we can now re-express the
+SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
+vectorized form, as shown in Figure~\ref{fig-cj04}. In this form, the
+outcome \(O^{cp}_{RPUV}\), the texts' discriminal difference
+\(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the judges'
+biases \(B_{JK}\) are no longer scalar values but are instead vectors.
+These vectors encompass all the values from the conceptual population.
+Specifically, \(O^{cp}_{RPUV}\) and \(D_{RPUV}\) are vectors with a
+dimension of \({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\)
+values each, while \(T_{IA}\) and \(B_{JK}\) have dimensions of
+\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\) values,
+respectively.
 
@@
 
-\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
+\caption{\label{fig-cj04}Conceptual-population model, vectorized form of
+SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag}.}
 
 \end{figure}%
+
+Finally, building on the arguments developed in
+Section~\ref{sec-theory-issue2} and the principles of Structural
+Equation Modeling (SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item
+Response Theory (IRT)
+\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
+the conceptual-population model includes two (hierarchical) structural
+components. Each component defines how different manifest or latent
+variables influence the primary latent variable of interest
+\citep{Everitt_et_al_2010}. This addition enables analyses that
+simultaneously account for the traits and their associated
+uncertainties, linking the trait measurement process with hypothesis
+testing, thereby reducing bias and enhancing the measurement precision.
+
+Figure~\ref{fig-cj09} displays the two (hierarchical) structural
+components in the motivating example. Specifically, the
+DAG~\ref{fig-cj09_dag} shows that the first component describes how the
+students' overall writing quality trait \(T_{I}\) and a set of
+\emph{relevant} text-related variables \(X_{IA}\) causally influence the
+texts' written quality trait \(T_{IA}\). The DAG also highlights the
+idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
+quality that neither the students' traits nor other relevant
+text-related variables can explain. In this description, \(T_{I}\) is a
+vectors of dimension \(n_{I}\), \(e_{IA}\) is a vector of dimension
+\((n_{I} \cdot n_{A})\), while \(X_{IA}\) is a matrix with
+\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.
+
+At a higher level in the hierarchy, the DAG~\ref{fig-cj09_dag} also
+describes how a set of \emph{relevant} student-related variables
+\(X_{I}\) causally influence the students' overall writing quality trait
+\(T_{I}\). Additionally, the DAG accounts for idiosyncratic errors
+\(e_{I}\) that capture variations in the students' writing quality that
+the relevant student-related variables cannot explain. Here, \(e_{I}\)
+is a vector of dimension \(n_{I}\), and \(X_{I}\) is a matrix with
+\(n_{I}\) rows and \(p_{I}\) columns.
+
+Similarly, the DAG~\ref{fig-cj09_dag} shows that the second component
+describes how the judges' overall bias \(B_{J}\) and a set of
+\emph{relevant} judgment-related variables \(Z_{JK}\) causally influence
+the biases associated with each text \(B_{JK}\). The DAG also highlights
+the idiosyncratic errors \(e_{JK}\) that capture variations in the
+texts' biases that neither the judges' bias nor other relevant
+judgment-related variables can explain. In this description, \(B_{J}\)
+is a vectors of dimension \(n_{J}\), \(e_{JK}\) is a vector of dimension
+\((n_{J} \cdot n_{K})\), while \(Z_{JK}\) is a matrix with
+\((n_{J} \cdot n_{K})\) rows and \(p_{JK}\) columns.
+
+Furthermore, at a higher level, the DAG~\ref{fig-cj09_dag} describes how
+a set of \emph{relevant} judge-related variables \(Z_{J}\) causally
+influence the judges' overall bias \(B_{J}\). The DAG also accounts for
+idiosyncratic errors \(e_{J}\), which capture the variations in the
+judges' biases that the relevant judge-related variables cannot explain.
+Here, \(e_{J}\) is a vector of dimension \(n_{J}\), and \(Z_{J}\) is a
+matrix with \(n_{J}\) rows and \(p_{J}\) columns.
 
@@
 
-\caption{\label{fig-cj09}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
+\caption{\label{fig-cj09}Conceptual-population model, full integration
+of theoretical principles.}
 
 \end{figure}%
+
+A cautious reader will notice that these descriptions rely on several
+sets of relevant variables. In this study, a \emph{relevant} or a
+\emph{sufficient adjustment} set of variables (potentially an empty set)
+refer to those that satisfy the \emph{backdoor criterion}
+\citep[pp.~37]{Neal_2020}, that is, variables that block all non-causal
+paths without introducing new ones
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
+
+Incorporating these variables into the conceptual-population model
+enables researchers to identify different estimands of interest and
+their causal effects, even with observational data. This identification
+works because conditioning on these variables ensures that the estimator
+meets several critical properties, including confounding elimination
+\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
+the statistical analysis to a subset of the population defined by the
+values of a given variable \citep[pp.~32]{Neal_2020}.
+
+Naturally, the validity of causal claims now hinges on the assumption
+that these variables serve as a sufficient adjustment set, a property
+that cannot be guaranteed for any variable. Nevertheless, as
+\citet[pp.~150]{Kohler_et_al_2019} emphasized, drawing conclusions about
+the real world from observed data inevitably requires assumptions, and
+this requirement holds true for both observational and experimental
+data.
 

### 2025-02-13 paper2.qmd
section:theoretical; process:write,figures

@@
 @sec-theory-issues discusses the identified issues in detail.
-@sec-theoretical extends Thurstone's general form to address these challenges.
-The new model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-@sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of pairwise comparison data.
+@sec-theoretical develops Thurstone's theory to address these challenges.
+The development integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+@sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data.
 @sec-discussion discusses the implications of the findings and explores avenues for future research.
@@
 
-# Extending Thurstone's general form {#sec-theoretical}
+# Developing Thurstone's theory {#sec-theoretical}
 
@@
 This example will appear throughout the document to illustrate various aspects of the proposed development.
-In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researchers could design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
-Judges would then compare pairs of students’ written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait).
-Based on this setup, researchers could reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-Finally, following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
+In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researcher design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+Judges then compare pairs of students’ written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait).
+Based on this setup, researchers reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
+Finally, following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
 
 <!-- 4. But why identification is so important? -->
-However, @sec-theory-issues presents compelling evidence that the BTL model has several statistical and measurement limitations.
+However, @sec-theory-issues presents compelling evidence that Thurstone's Case V, and by extension the BTL model, suffers from several statistical and measurement limitations.
 These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the motivating example.
@@
 Without identification, consistency cannot be achieved, even with "infinite" and error-free data.
-Consequently, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023, pp.
+Thus, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023, pp.
 5].
@@
 <!-- 6. What the section will do? -->
-Thus, this section addresses the identified issues in @sec-theory-issues by extending Thurstone’s general form using the structural approach to causal inference.
+Thus, this section addresses the issues identified in @sec-theory-issues by developing  Thurstone’s theory using the structural approach to causal inference.
 Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
@@
 [SCM -@fig-cj03_scm] illustrates the preliminary integration of theoretical principles into the conceptual-population model of the motivating example.
-This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{abhijk})$ and several related variables.
-The subscripts $a$ and $b$ represent the compared texts (stimuli), while the indices $h$ and $i$ label the students who authored them (individuals).
-Moreover, the index $j$ refers to the judge, and $k$ specifies the judgment number, accounting for experimental conditions where judges compare the same texts written by the same students multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
+This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
+The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli).
+Moreover, the index $j$ refers to the judge, and $k$ specifies the judgment number, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
 366-376].
-Notably, the indexing system allow comparisons between different texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
-However, it excludes cases where judges compare a student’s text to itself ($h = i$, $a = b$), as such a comparison lacks practical meaning within the CJ framework.
-
-The other variables present in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory).
-Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the conceptual-population model incorporates the judges' biases $(B_{kj})$.
-Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}$.
+Notably, the indexing system allows comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
+However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical meaning within the CJ framework.
+
+In line with Thurstone's theory, the other variables present in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
+Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the [SCM -@fig-cj03_scm] incorporates the judges' biases $(B_{kj})$.
+Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
 Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
@@
 \begin{aligned}
-  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
-  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
+  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
+  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
 \end{aligned}
@@
 A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables.
-The term *directed* indicates that edges or arrows extend from one node to another showing the direction of causal influence.
+The term *directed* indicates that edges or arrows extend from one node to another, showing the direction of causal influence.
 The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
-DAGs depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
-The *standard representation* of DAGs typically omits exogenous variables for simplicity, while the *magnified representation* includes them to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
-This section adopts the magnified representation to enhance clarity.
+DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
+Moreover, they often omit exogenous variables for simplicity, following the *standard representation*.
+However, this section adopts the *magnified representation* of DAGS, as including exogenous variables can enhance clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 @fig-cj03_dag displays the DAG associated with [SCM -@fig-cj03_scm].
-This graph illustrates the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and the practical aspects of CJ's experiments.
+This graph illustrates the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and the practical aspects of CJ experiments.
 It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent within the CJ framework, such as the text's discriminal processes, their discriminal difference, and the judges' biases.
 
-<!-- 6. The problem of continuing this way -->
-Although specifying a system's data structure is not necessary requirement for using SCMs and DAGs, defining one can improve clarity and simplify their notation.
-Thus, we define specific data structures to facilitate the description of the CJ system.
-First, we define the matrices $IA$ and $JK$ as follows:
-
-$$
+<!-- 6. Why is useful to declare the data structure? -->
+Although specifying a system's data structure is not mandatory for using SCMs and DAGs, defining one can sometimes enhance clarity and facilitate the description of the system.
+Thus, to simplify the description of the CJ system, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
+
+$$
+I_{n_{I}} = \begin{bmatrix}
+1 \\
+2 \\
+\vdots \\
+h \\
+\vdots \\
+i \\
+\vdots \\
+n_{I}
+\end{bmatrix} ; \;
 IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
@@
 n_{I} & n_{A} \\
-\end{bmatrix} \in \mathbb{N} \; ; \quad
+\end{bmatrix} ; \;
+J_{n_{J}} = \begin{bmatrix}
+1 \\
+2 \\
+\vdots \\
+j \\
+\vdots \\
+n_{J}
+\end{bmatrix} ; \;
 JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
@@
 n_{J} & n_{K} \\
-\end{bmatrix} \in \mathbb{N}
+\end{bmatrix}
 $$ {#eq-mat12}
 
-where each row of $IA$ corresponds to a unique pairing of individuals $h, i$ with stimuli $a, b$.
+<!-- 7. what is the data structure? -->
+where an element of $I$ corresponds to a unique individual $h$ or $i$, while each row of $IA$ corresponds to a unique pairing of individuals $h, i$ with stimuli $a, b$.
 Here, $n_{I}$ represents the number of individuals, while $n_{A}$ denotes the number of stimuli available per individual.
-Thus, the matrix $IA$ is composed of natural numbers $\mathbb{N}$, with $(n_{I} \cdot n_{A})$ rows and $2$ columns.
-Similarly, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment $k$.
+Thus, vector $I$ has a length of $n_{I}$, and matrix $IA$ has a dimension of $(n_{I} \cdot n_{A})$ rows and $2$ columns.
+The vector and the matrix consist of scalars from the set of natural numbers $\mathbb{N}$.
+
+Similarly, an element of $J$ corresponds to a unique judge $j$.
+Moreover, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment number $k$.
 In this case, $n_{J}$ indicates the total number of judges, and $n_{K}$ represents the number of judgments each judge performs.
-Therefore, the matrix $JK$ is also composed of natural numbers $\mathbb{N},$ with $(n_{J} \cdot n_{K})$ rows and $2$ columns.
-
-Importantly, since the discriminal difference of the stimuli becomes an observable outcome only through judges' perceptions, it is necessary to define the matrix $RPUV$, where each row of the matrix uniquely pairs the corresponding rows from the matrices $IA$ and $JK$ as follows:
+As a result, vector $J$ has a length of $n_{J}$, and matrix $JK$ has a dimension of $(n_{J} \cdot n_{K})$ rows and $2$ columns.
+Like the previous structures, the vector and the matrix consist of scalars from the set of natural numbers $\mathbb{N}$.
+
+Additionally, we define the matrix $RPUV$, where each row uniquely pairs the corresponding rows from the matrices $IA$ and $JK$.
+This matrix reflects the fundamental idea that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions.
+Thus, the matrix is defined as follows:
 
@@
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-h & a & i & b & j & k \\
+h & i & a & b & j & k \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
-n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 2 \\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}-1 \\
-n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K} \\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
 \end{bmatrix} \in \mathbb{N}
@@
 
-The $RPUV$ matrix is also composed of natural numbers $\mathbb{N}$, and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
-The dimension of the matrix results from the conceptual-population model's assumption that researchers have access to data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
-Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs in the conceptual population.
-The coefficient is defined as:
+This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
+The matrix's dimension results from the conceptual-population model's assumption that researchers have access to data capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs within the conceptual population.
+The coefficient is defined as follows:
 
@@
 
-Readers can better visualize the structure of these matrices by considering relatively small values for the parameter corresponding to number of individuals $(n_{I} = 5)$, the number of available stimuli per individual $(n_{A} = 2)$, the total number of judges $(n_{J} = 3)$, and the number of judgments each judge performs $(n_{K} = 3)$.
-With these values, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) are represented as follows:
-
-$$
+Readers can better visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals $(n_{I} = 5)$, the number of available stimuli per individual $(n_{A} = 2)$, the total number of judges $(n_{J} = 3)$, and the number of judgments each judge performs $(n_{K} = 3)$.
+Using these values, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:
+
+$$
+I_{5} = \begin{bmatrix}
+1 \\
+2 \\
+3 \\
+4 \\
+5 \\
+\end{bmatrix} ; \;
 IA_{10 \times 2} = \begin{bmatrix}
@@
 5 & 2 \\
-\end{bmatrix}\; ; \quad
+\end{bmatrix} ; \;
+J_{3} = \begin{bmatrix}
+1 \\
+2 \\
+3 \\
+\end{bmatrix} ; \;
 JK_{9 \times 2} = \begin{bmatrix}
@@
 3 & 3 \\
-\end{bmatrix} \; ; \quad
+\end{bmatrix} ; \;
 RPUV_{405  \times 6} = \begin{bmatrix}
@@
 
-Thus, with the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can now re-express the [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
-In this form, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are no longer scalar values but are instead vectors.
-These vectors encompass all the values from the conceptual population.
-Specifically, $O^{cp}_{RPUV}$ and $D_{RPUV}$ are vectors with a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ values each, while $T_{IA}$ and $B_{JK}$ have dimensions of $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$ values, respectively.
+<!-- 8. how de we use that data structure? -->
+Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can now re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
+In this representation, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are no longer scalar values but vectors.
+These vectors capture all the values from the conceptual population.
+Specifically, $O^{cp}_{RPUV}$ and $D_{RPUV}$ are vectors of length ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, while $T_{IA}$ and $B_{JK}$ are vectors of length $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.
 
@@
 
-Finally, building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+<!-- 9. what else? -->
+Building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
 138] and Item Response Theory (IRT) [@Fox_2010, chap.
 6; @vanderLinden_et_al_2017_I, chap.
-24], the conceptual-population model includes two (hierarchical) structural components.
-Each component defines how different manifest or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
-This addition enables analyses that simultaneously account for the traits and their associated uncertainties, linking the trait measurement process with hypothesis testing, thereby reducing bias and enhancing the measurement precision.
+24], the conceptual-population model then includes two hierarchical structural components.
+Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
+This addition facilitates analyses that integrate the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing the measurement precision.
 
@@
 
-@fig-cj09 displays the two (hierarchical) structural components in the motivating example.
-Specifically, the [DAG -@fig-cj09_dag] shows that the first component describes how the students' overall writing quality trait $T_{I}$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written quality trait $T_{IA}$.
-The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' quality that neither the students' traits nor other relevant text-related variables can explain.
-In this description, $T_{I}$ is a vectors of dimension $n_{I}$, $e_{IA}$ is a vector of dimension $(n_{I} \cdot n_{A})$, while $X_{IA}$ is a matrix with $(n_{I} \cdot n_{A})$ rows and $p_{IA}$ columns.
-
-At a higher level in the hierarchy, the [DAG -@fig-cj09_dag] also describes how a set of *relevant* student-related variables $X_{I}$ causally influence the students' overall writing quality trait $T_{I}$.
-Additionally, the DAG accounts for idiosyncratic errors $e_{I}$ that capture variations in the students' writing quality that the relevant student-related variables cannot explain.
-Here, $e_{I}$ is a vector of dimension $n_{I}$, and $X_{I}$ is a matrix with $n_{I}$ rows and $p_{I}$ columns.
-
-Similarly, the [DAG -@fig-cj09_dag] shows that the second component describes how the judges' overall bias $B_{J}$ and a set of *relevant* judgment-related variables $Z_{JK}$ causally influence the biases associated with each text $B_{JK}$.
-The DAG also highlights the idiosyncratic errors $e_{JK}$ that capture variations in the texts' biases that neither the judges' bias nor other relevant judgment-related variables can explain.
-In this description, $B_{J}$ is a vectors of dimension $n_{J}$, $e_{JK}$ is a vector of dimension $(n_{J} \cdot n_{K})$, while $Z_{JK}$ is a matrix with $(n_{J} \cdot n_{K})$ rows and $p_{JK}$ columns.
-
-Furthermore, at a higher level, the [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $Z_{J}$ causally influence the judges' overall bias $B_{J}$.
-The DAG also accounts for idiosyncratic errors $e_{J}$, which capture the variations in the judges' biases that the relevant judge-related variables cannot explain.
-Here, $e_{J}$ is a vector of dimension $n_{J}$, and $Z_{J}$ is a matrix with $n_{J}$ rows and $p_{J}$ columns.
+@fig-cj09 displays the two hierarchical structural components in the motivating example.
+Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component.
+This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written-quality trait $T_{IA}$.
+The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain.
+Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively.
+Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $p_{IA}$ columns.
+
+Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $X_{I}$ causally influence the students' writing-quality trait $T_{I}$.
+The DAG also includes the idiosyncratic errors $e_{I}$, which capture variations in the the students' traits that the relevant student-related variables cannot explain.
+Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ is an observed matrix with $n_{I}$ rows and $p_{I}$ columns.
 
@@
 
-A cautious reader will notice that these descriptions rely on several sets of relevant variables.
-In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refer to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component.
+This component describes how the judges' bias $B_{J}$ and a set of *relevant* judgment-related variables $Z_{JK}$ causally influence the biases associated with each text $B_{JK}$.
+The DAG also highlights the idiosyncratic errors $e_{JK}$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain.
+In this description, $B_{J}$ and $e_{JK}$ are latent vectors of length $n_{J}$ and $(n_{J} \cdot n_{K})$, respectively.
+Additionally, $Z_{JK}$ is an observed matrix with a dimension of $(n_{J} \cdot n_{K})$ rows and $p_{JK}$ columns.
+
+Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $Z_{J}$ causally influence the judges' bias $B_{J}$.
+The DAG also accounts for idiosyncratic errors $e_{J}$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain.
+Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $p_{J}$ columns.
+Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] constitute the set of endogenous variables $V$ and structural equations $F$ for the conceptual-population model.
+
+A cautious reader will notice that these descriptions rely on several relevant variables.
+In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
 37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
@@
 Nevertheless, as @Kohler_et_al_2019 [pp.
-150] emphasized, drawing conclusions about the real world from observed data inevitably requires assumptions, and this requirement holds true for both observational and experimental data.
+150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
 
@@
 150] -->
+
+In summary,
 

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

--- 2025-02-10 _tex/paper2.tex
@@
 theory. Section~\ref{sec-theory-issues} discusses the identified issues
-in detail. Section~\ref{sec-theoretical} extends Thurstone's general
-form to address these challenges. The new model integrates the
-theoretical core principles alongside key assessment design features
-relevant to CJ experiments, such as the selection of judges, stimuli,
-and comparisons. Section~\ref{sec-statistical} translates these
-theoretical and practical elements into a probabilistic statistical
-model for the analysis of pairwise comparison data.
-Section~\ref{sec-discussion} discusses the implications of the findings
-and explores avenues for future research. Finally,
-Section~\ref{sec-conclusion} summarizes the study's key insights.
+in detail. Section~\ref{sec-theoretical} develops Thurstone's theory to
+address these challenges. The development integrates core theoretical
+principles alongside key assessment design features relevant to CJ
+experiments, such as the selection of judges, stimuli, and comparisons.
+Section~\ref{sec-statistical} translates these theoretical and practical
+elements into a probabilistic statistical model for the analysis of
+dichotomous pairwise comparison data. Section~\ref{sec-discussion}
+discusses the implications of the findings and explores avenues for
+future research. Finally, Section~\ref{sec-conclusion} summarizes the
+study's key insights.
 
@@
 
-\section{Extending Thurstone's general form}\label{sec-theoretical}
+\section{Developing Thurstone's theory}\label{sec-theoretical}
 
@@
 ability to produce high-quality written texts?'' To investigate this,
-researchers could design a CJ experiment by randomly assigning students
+researcher design a CJ experiment by randomly assigning students
 (individuals) to two groups, each receiving a different teaching method.
-Judges would then compare pairs of students' written texts (stimuli), to
+Judges then compare pairs of students' written texts (stimuli), to
 produce a dichotomous outcome that reflects the relative quality of each
-text (trait). Based on this setup, researchers could reformulate the
-research question as the estimand: ``\emph{On average}, is there a
-difference in the ability to produce high-quality written texts between
-the two groups of students?''. Finally, following current CJ practices,
-researchers would then rely on estimates from the BTL model, or its
-transformations, to approximate this estimand.
+text (trait). Based on this setup, researchers reformulate the research
+question as the estimand: ``\emph{On average}, is there a difference in
+the ability to produce high-quality written texts between the two groups
+of students?''. Finally, following current CJ practices, researchers
+then rely on estimates from the BTL model, or its transformations, to
+approximate this estimand.
 
 However, Section~\ref{sec-theory-issues} presents compelling evidence
-that the BTL model has several statistical and measurement limitations.
-These limitations hinder the model's ability to identify various
-estimands relevant to CJ inquiries, including the one described in the
-motivating example. Identification is crucial because it is a necessary
-condition for ensuring consistent estimators. \emph{Consistency} refers
-to the property of an estimator whose estimates converge to the ``true''
-value of the estimand as the data size approaches infinity
+that Thurstone's Case V, and by extension the BTL model, suffers from
+several statistical and measurement limitations. These limitations
+hinder the model's ability to identify various estimands relevant to CJ
+inquiries, including the one described in the motivating example.
+Identification is crucial because it is a necessary condition for
+ensuring consistent estimators. \emph{Consistency} refers to the
+property of an estimator whose estimates converge to the ``true'' value
+of the estimand as the data size approaches infinity
 \citep{Everitt_et_al_2010}. Without identification, consistency cannot
-be achieved, even with ``infinite'' and error-free data. Consequently,
-deriving meaningful insights from finite data becomes impossible
+be achieved, even with ``infinite'' and error-free data. Thus, deriving
+meaningful insights from finite data becomes impossible
 \citep[pp.~5]{Schuessler_et_al_2023}.
@@
 
-Thus, this section addresses the identified issues in
-Section~\ref{sec-theory-issues} by extending Thurstone's general form
-using the structural approach to causal inference. Specifically, it
-leverages the capabilities of this approach to formalize the combination
-of the core theoretical principles outlined in
+Thus, this section addresses the issues identified in
+Section~\ref{sec-theory-issues} by developing Thurstone's theory using
+the structural approach to causal inference. Specifically, it leverages
+the capabilities of this approach to formalize the combination of the
+core theoretical principles outlined in
 Section~\ref{sec-thurstone_theory} with key assessment design features
@@
 
-Figure~\ref{fig-cj04} illustrates the preliminary integration of
-theoretical principles into the conceptual-population model.
-Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
+SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
+theoretical principles into the conceptual-population model of the
 motivating example. This SCM maps the relationship between the
-conceptual-population outcome \((O^{cp}_{abhijk})\) to several other
-variables. In this SCM, the subscripts \(\{a,b\}\) represent the
-compared texts (stimuli), with \(a \neq b\). The indices \(\{h,i\}\)
-label the students who wrote the texts (individuals), where \(h = i\) is
-possible. Moreover, the index \(j\) denotes the judge, and \(k\) denotes
-the judgment number, which accounts for experimental cases where judges
-perform multiple judgments of the same texts written by the same
-students, i.e., a \emph{repeated measures design}
-\citep[pp.~366-376]{Lawson_2015}.
-
-The other variables in SCM~\ref{fig-cj04_scm} include the texts'
-discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
-difference \((D_{abhijk})\), in line with Thurstone's theory (see
-Section~\ref{sec-thurstone_theory}). Additionally, based on the evidence
-presented in Section~\ref{sec-theory-issue1b} and the recommendations of
-\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, we incorporate the
-judges' biases \((B_{kj})\) into the model. Together with the outcome,
-these variables form the preliminary set of endogenous variables,
-\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
-Finally, the SCM shows the preliminary set of structural equations,
+conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
+variables. The subscripts \(h\) and \(i\) label the students who
+authored the texts (individuals), while the indices \(a\) and \(b\)
+represent the compared texts (stimuli). Moreover, the index \(j\) refers
+to the judge, and \(k\) specifies the judgment number, accounting for
+experimental conditions where judges perform the same judgment multiple
+times, i.e., a \emph{repeated measures design}
+\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system allows
+comparisons between texts written by the same student (\(h = i\),
+\(a \neq b\)) and between texts written by distinct students
+(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
+where judges compare a student's text to itself (\(h = i\), \(a = b\)),
+as such comparison lacks practical meaning within the CJ framework.
+
+In line with Thurstone's theory, the other variables present in
+SCM~\ref{fig-cj03_scm} include the texts' discriminal processes
+\((T_{ha}, T_{ib})\) and their discriminal difference \((D_{hiabjk})\)
+(see Section~\ref{sec-thurstone_theory}). Additionally, based on the
+arguments presented in Section~\ref{sec-theory-issue1b} and the
+recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
+the SCM~\ref{fig-cj03_scm} incorporates the judges' biases \((B_{kj})\).
+Together with the outcome, these variables form the preliminary set of
+endogenous variables,
+\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
+the SCM shows the preliminary set of structural equations,
 \(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
@@
 \begin{aligned}
-  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
-  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
+  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
+  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
 \end{aligned}
@@
 \citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
-consisting of nodes connected by edges, where the nodes represent random
-variables. The term \emph{directed} means that the edges extend from one
-node to another, with arrows indicating the direction of causal
-influence. The term \emph{acyclic} implies that the causal influences do
+consisting of nodes connected by edges, where nodes represent random
+variables. The term \emph{directed} indicates that edges or arrows
+extend from one node to another, showing the direction of causal
+influence. The term \emph{acyclic} means that the causal influences do
 not form loops, ensuring the influences do not cycle back on themselves
-\citep{McElreath_2020}. DAGs represent observed variables as solid black
-circles, while they use open circles for unobserved (latent) variables
-\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
-DAGs typically omits exogenous variables for simplicity, the
-\emph{magnified representation} includes them, offering the advantage of
-highlighting potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}. Thus, this study uses the magnified
-representation.
-
-Figure~\ref{fig-cj04_dag} displays the DAG associated with
-SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
+\citep{McElreath_2020}. DAGs conventionally depict observed variables as
+solid black circles and unobserved (latent) variables as open circles
+\citep{Morgan_et_al_2014}. Moreover, they often omit exogenous variables
+for simplicity, following the \emph{standard representation}. However,
+this section adopts the \emph{magnified representation} of DAGS, as
+including exogenous variables can enhance clarity and reveal potential
+issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}.
+
+Figure~\ref{fig-cj03_dag} displays the DAG associated with
+SCM~\ref{fig-cj03_scm}. This graph illustrates the direction of causal
 influences among variables, reflecting the assumptions of Thurstone's
-theory and CJ experiments. It also distinguishes between observable
-endogenous variables, such as the outcome, and variables considered
-latent according to the theory and practice of CJ, such as the the
-text's discriminal processes, their discriminal difference, and the
-judges' biases.
-
-It is easy to see that representing the CJ system using indices for a
-single pair of individuals and stimuli, along with the indices for the
-judgment number and the judge responsible for the comparison, can
-quickly become intractable and cumbersome. Thus, to maintain generality
-and simplify the notation, certain indices are grouped into vectors or
-matrices. First, we define the matrices \(IA\) and \(JK\) as:
-
-\[
-IA = \begin{bmatrix}
+theory and the practical aspects of CJ experiments. It also
+distinguishes between observable endogenous variables, such as the
+outcome, and variables considered latent within the CJ framework, such
+as the text's discriminal processes, their discriminal difference, and
+the judges' biases.
+
+Although specifying a system's data structure is not mandatory for using
+SCMs and DAGs, defining one can sometimes enhance clarity and facilitate
+the description of the system. Thus, to simplify the description of the
+CJ system, we first define the vectors \(I\) and \(J\), along with the
+matrices \(IA\) and \(JK\), as follows:
+
+\begin{equation}\phantomsection\label{eq-mat12}{
+I_{n_{I}} = \begin{bmatrix}
+1 \\
+2 \\
+\vdots \\
+h \\
+\vdots \\
+i \\
+\vdots \\
+n_{I}
+\end{bmatrix} ; \;
+IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
 1 & 1 \\
 1 & 2 \\
+\vdots & \vdots \\
+1 & n_{A}-1 \\
+1 & n_{A} \\
 \vdots & \vdots \\
@@
 n_{I} & n_{A} \\
-\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
-JK = \begin{bmatrix}
+\end{bmatrix} ; \;
+J_{n_{J}} = \begin{bmatrix}
+1 \\
+2 \\
+\vdots \\
+j \\
+\vdots \\
+n_{J}
+\end{bmatrix} ; \;
+JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
 1 & 1 \\
 1 & 2 \\
+\vdots & \vdots \\
+1 & n_{K}-1 \\
+1 & n_{K} \\
 \vdots & \vdots \\
@@
 n_{J} & n_{K}-1 \\
-n_{I} & n_{A} \\
-\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
+n_{J} & n_{K} \\
+\end{bmatrix}
+}\end{equation}
+
+where an element of \(I\) corresponds to a unique individual \(h\) or
+\(i\), while each row of \(IA\) corresponds to a unique pairing of
+individuals \(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents
+the number of individuals, while \(n_{A}\) denotes the number of stimuli
+available per individual. Thus, vector \(I\) has a length of \(n_{I}\),
+and matrix \(IA\) has a dimension of \((n_{I} \cdot n_{A})\) rows and
+\(2\) columns. The vector and the matrix consist of scalars from the set
+of natural numbers \(\mathbb{N}\).
+
+Similarly, an element of \(J\) corresponds to a unique judge \(j\).
+Moreover, each row of \(JK\) specifies a unique pairing of a judge \(j\)
+with a judgment number \(k\). In this case, \(n_{J}\) indicates the
+total number of judges, and \(n_{K}\) represents the number of judgments
+each judge performs. As a result, vector \(J\) has a length of
+\(n_{J}\), and matrix \(JK\) has a dimension of \((n_{J} \cdot n_{K})\)
+rows and \(2\) columns. Like the previous structures, the vector and the
+matrix consist of scalars from the set of natural numbers
+\(\mathbb{N}\).
+
+Additionally, we define the matrix \(RPUV\), where each row uniquely
+pairs the corresponding rows from the matrices \(IA\) and \(JK\). This
+matrix reflects the fundamental idea that the discriminal difference of
+the stimuli becomes an observable outcome only through the judges'
+perceptions. Thus, the matrix is defined as follows:
+
+\begin{equation}\phantomsection\label{eq-mat3}{
+RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
+1 & 1 & 1 & 2 & 1 & 1 \\
+1 & 1 & 1 & 2 & 1 & 2 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
+1 & 1 & 1 & 2 & 1 & n_{K} \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+h & i & a & b & j & k \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
+n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
+\end{bmatrix} \in \mathbb{N}
+}\end{equation}
+
+This matrix also consists of scalars from the set of natural numbers
+\(\mathbb{N}\) and has a dimension of
+\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
+columns. The matrix's dimension results from the conceptual-population
+model's assumption that researchers have access to data capturing all
+repeated judgments made by every available judge for each pair of
+stimuli produced by each pair of individuals in the population. Here,
+the term \({n_{I} \cdot n_{A} \choose 2}\) represents the Binomial
+coefficient, which quantifies the total number of individual-stimuli
+comparison pairs within the conceptual population. The coefficient is
+defined as follows:
+
+\[
+{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
 \]
 
-Each row of \(IA\) represents a unique pairing of an individual
-\(\{h,i\}\) with a stimulus \(\{a,b\}\). Here, \(n_{I}\) represents the
-number of individuals and \(n_{A}\) the number of stimulus within
-individuals in the population. Similarly, each row of \(JK\) represents
-a unique pairing of a judge \(j\) with a judgment \(k\). Here, \(n_{J}\)
-represents the number of judges and \(n_{K}\) the number of judgments
-performed by each judge in the population.
+Readers can better visualize the structure of these vectors and matrices
+by considering small values for the parameters corresponding to number
+of individuals \((n_{I} = 5)\), the number of available stimuli per
+individual \((n_{A} = 2)\), the total number of judges \((n_{J} = 3)\),
+and the number of judgments each judge performs \((n_{K} = 3)\). Using
+these values, the matrices described in equations (\ref{eq-mat12}) and
+(\ref{eq-mat3}) take the following form:
+
+\[
+I_{5} = \begin{bmatrix}
+1 \\
+2 \\
+3 \\
+4 \\
+5 \\
+\end{bmatrix} ; \;
+IA_{10 \times 2} = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+2 & 1 \\
+2 & 2 \\
+3 & 1 \\
+3 & 2 \\
+4 & 1 \\
+4 & 2 \\
+5 & 1 \\
+5 & 2 \\
+\end{bmatrix} ; \;
+J_{3} = \begin{bmatrix}
+1 \\
+2 \\
+3 \\
+\end{bmatrix} ; \;
+JK_{9 \times 2} = \begin{bmatrix}
+1 & 1 \\
+1 & 2 \\
+1 & 3 \\
+2 & 1 \\
+2 & 2 \\
+2 & 3 \\
+3 & 1 \\
+3 & 2 \\
+3 & 3 \\
+\end{bmatrix} ; \;
+RPUV_{405  \times 6} = \begin{bmatrix}
+1 & 1 & 1 & 2 & 1 & 1 \\
+1 & 1 & 1 & 2 & 1 & 2 \\
+1 & 1 & 1 & 2 & 1 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 5 & 1 & 1 & 1 \\
+1 & 1 & 5 & 1 & 1 & 2 \\
+1 & 1 & 5 & 1 & 1 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+1 & 1 & 5 & 1 & 3 & 1 \\
+1 & 1 & 5 & 1 & 3 & 2 \\
+1 & 1 & 5 & 1 & 3 & 3 \\
+\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+5 & 1 & 5 & 2 & 3 & 1 \\
+5 & 1 & 5 & 2 & 3 & 2 \\
+5 & 1 & 5 & 2 & 3 & 3 \\
+\end{bmatrix}
+\]
+
+Thus, using the data structure defined by the matrices in equations
+(\ref{eq-mat12}) and (\ref{eq-mat3}), we can now re-express
+SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
+vectorized form, as shown in Figure~\ref{fig-cj04}. In this
+representation, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
+difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
+judges' biases \(B_{JK}\) are no longer scalar values but vectors. These
+vectors capture all the values from the conceptual population.
+Specifically, \(O^{cp}_{RPUV}\) and \(D_{RPUV}\) are vectors of length
+\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
+\(T_{IA}\) and \(B_{JK}\) are vectors of length \((n_{I} \cdot n_{A})\)
+and \((n_{J} \cdot n_{K})\), respectively.
 
@@
 
-\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
+\caption{\label{fig-cj04}Conceptual-population model, vectorized form of
+SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag}.}
 
 \end{figure}%
+
+Building on the arguments developed in Section~\ref{sec-theory-issue2}
+and the principles of Structural Equation Modeling (SEM)
+\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
+\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
+the conceptual-population model then includes two hierarchical
+structural components. Each component defines how different observed or
+latent variables influence the primary latent variable of interest
+\citep{Everitt_et_al_2010}. This addition facilitates analyses that
+integrate the trait measurement and hypothesis testing processes,
+simultaneously accounting for the traits and their associated
+uncertainties, thereby reducing bias and enhancing the measurement
+precision.
+
+Figure~\ref{fig-cj09} displays the two hierarchical structural
+components in the motivating example. Specifically, the top branch of
+DAG~\ref{fig-cj09_dag} depicts the first component. This component
+describes how the students' writing-quality trait \((T_{I})\) and a set
+of \emph{relevant} text-related variables \(X_{IA}\) causally influence
+the texts' written-quality trait \(T_{IA}\). The DAG also highlights the
+idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
+trait that neither the students' trait nor other relevant text-related
+variables can explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors
+of length \(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively.
+Additionally, \(X_{IA}\) is an observed matrix with a dimension of
+\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.
+
+Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
+set of \emph{relevant} student-related variables \(X_{I}\) causally
+influence the students' writing-quality trait \(T_{I}\). The DAG also
+includes the idiosyncratic errors \(e_{I}\), which capture variations in
+the the students' traits that the relevant student-related variables
+cannot explain. Here, \(e_{I}\) is a latent vector of length \(n_{I}\),
+and \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(p_{I}\)
+columns.
 
@@
 
-\caption{\label{fig-cj09}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
+\caption{\label{fig-cj09}Conceptual-population model, full integration
+of theoretical principles.}
 
 \end{figure}%
+
+Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
+second component. This component describes how the judges' bias
+\(B_{J}\) and a set of \emph{relevant} judgment-related variables
+\(Z_{JK}\) causally influence the biases associated with each text
+\(B_{JK}\). The DAG also highlights the idiosyncratic errors \(e_{JK}\),
+which capture variations in the texts' bias that neither the judges'
+bias nor other relevant judgment-related variables can explain. In this
+description, \(B_{J}\) and \(e_{JK}\) are latent vectors of length
+\(n_{J}\) and \((n_{J} \cdot n_{K})\), respectively. Additionally,
+\(Z_{JK}\) is an observed matrix with a dimension of
+\((n_{J} \cdot n_{K})\) rows and \(p_{JK}\) columns.
+
+Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
+set of \emph{relevant} judge-related variables \(Z_{J}\) causally
+influence the judges' bias \(B_{J}\). The DAG also accounts for
+idiosyncratic errors \(e_{J}\), capturing the variations in the judges'
+bias that relevant judge-related variables cannot explain. Here,
+\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
+observed matrix with \(n_{J}\) rows and \(p_{J}\) columns. Notably, all
+variables and functions shown in SCM~\ref{fig-cj09_scm} and
+DAG~\ref{fig-cj09_dag} constitute the set of endogenous variables \(V\)
+and structural equations \(F\) for the conceptual-population model.
+
+A cautious reader will notice that these descriptions rely on several
+relevant variables. In this study, a \emph{relevant} or a
+\emph{sufficient adjustment} set of variables (potentially an empty set)
+refers to those that satisfy the \emph{backdoor criterion}
+\citep[pp.~37]{Neal_2020}, that is, variables that block all non-causal
+paths without introducing new ones
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
+
+Incorporating these variables into the conceptual-population model
+enables researchers to identify different estimands of interest and
+their causal effects, even with observational data. This identification
+works because conditioning on these variables ensures that the estimator
+meets several critical properties, including confounding elimination
+\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
+the statistical analysis to a subset of the population defined by the
+values of a given variable \citep[pp.~32]{Neal_2020}.
+
+Naturally, the validity of causal claims now hinges on the assumption
+that these variables serve as a sufficient adjustment set, a property
+that cannot be guaranteed for any variable. Nevertheless, as
+\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
+real world from observed data inevitably requires assumptions, and these
+requirements hold true for both observational and experimental data.
+
+In summary,
 

### 2025-02-13 paper2.qmd
section:theoretical; process:write,figures

@@
 6; @vanderLinden_et_al_2017_I, chap.
-24], the conceptual-population model then includes two hierarchical structural components.
+24], the conceptual-population model includes two hierarchical structural components.
 Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
@@
 
-@fig-cj09 displays the two hierarchical structural components in the motivating example.
+@fig-cj09 displays the two hierarchical structural components for the motivating example.
 Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component.
-This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written-quality trait $T_{IA}$.
+This component describes how the students' writing-quality trait $T_{I}$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written-quality trait $T_{IA}$.
 The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain.
@@
 Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] constitute the set of endogenous variables $V$ and structural equations $F$ for the conceptual-population model.
-
-A cautious reader will notice that these descriptions rely on several relevant variables.
-In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
-37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-
-Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
-This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
-Here, *conditioning* means restricting the statistical analysis to a subset of the population defined by the values of a given variable [@Neal_2020, pp.
-32].
-
-Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
-Nevertheless, as @Kohler_et_al_2019 [pp.
-150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
-
-<!-- Statements about the real world from observed data require assumptions.
-This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
-The validity of our statements therefore critically hinges on the validity of those assumptions.
-If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
-[@Kohler_et_al_2019, pp.
-150] -->
 
@@
 ## Limitations and further research {#sec-discussion-limitations}
+
+A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
+In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
+37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+
+Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
+This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
+Here, *conditioning* means restricting the statistical analysis to a subset of the population defined by the values of a given variable [@Neal_2020, pp.
+32].
+
+Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
+Nevertheless, as @Kohler_et_al_2019 [pp.
+150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
 

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

@@
 \citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
-the conceptual-population model then includes two hierarchical
-structural components. Each component defines how different observed or
-latent variables influence the primary latent variable of interest
+the conceptual-population model includes two hierarchical structural
+components. Each component defines how different observed or latent
+variables influence the primary latent variable of interest
 \citep{Everitt_et_al_2010}. This addition facilitates analyses that
@@
 Figure~\ref{fig-cj09} displays the two hierarchical structural
-components in the motivating example. Specifically, the top branch of
+components for the motivating example. Specifically, the top branch of
 DAG~\ref{fig-cj09_dag} depicts the first component. This component
-describes how the students' writing-quality trait \((T_{I})\) and a set
-of \emph{relevant} text-related variables \(X_{IA}\) causally influence
-the texts' written-quality trait \(T_{IA}\). The DAG also highlights the
+describes how the students' writing-quality trait \(T_{I}\) and a set of
+\emph{relevant} text-related variables \(X_{IA}\) causally influence the
+texts' written-quality trait \(T_{IA}\). The DAG also highlights the
 idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
@@
 and structural equations \(F\) for the conceptual-population model.
-
-A cautious reader will notice that these descriptions rely on several
-relevant variables. In this study, a \emph{relevant} or a
-\emph{sufficient adjustment} set of variables (potentially an empty set)
-refers to those that satisfy the \emph{backdoor criterion}
-\citep[pp.~37]{Neal_2020}, that is, variables that block all non-causal
-paths without introducing new ones
-\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
-
-Incorporating these variables into the conceptual-population model
-enables researchers to identify different estimands of interest and
-their causal effects, even with observational data. This identification
-works because conditioning on these variables ensures that the estimator
-meets several critical properties, including confounding elimination
-\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
-the statistical analysis to a subset of the population defined by the
-values of a given variable \citep[pp.~32]{Neal_2020}.
-
-Naturally, the validity of causal claims now hinges on the assumption
-that these variables serve as a sufficient adjustment set, a property
-that cannot be guaranteed for any variable. Nevertheless, as
-\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
-real world from observed data inevitably requires assumptions, and these
-requirements hold true for both observational and experimental data.
 
@@
 research}\label{sec-discussion-limitations}
+
+A cautious reader will notice that the model descriptions of
+Section~\ref{sec-theory-theoretical_P} rely on several relevant
+variables. In this study, a \emph{relevant} or a \emph{sufficient
+adjustment} set of variables (potentially an empty set) refers to those
+that satisfy the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020},
+that is, variables that block all non-causal paths without introducing
+new ones \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
+
+Incorporating these variables into the conceptual-population model
+enables researchers to identify different estimands of interest and
+their causal effects, even with observational data. This identification
+works because conditioning on these variables ensures that the estimator
+meets several critical properties, including confounding elimination
+\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
+the statistical analysis to a subset of the population defined by the
+values of a given variable \citep[pp.~32]{Neal_2020}.
+
+Naturally, the validity of causal claims now hinges on the assumption
+that these variables serve as a sufficient adjustment set, a property
+that cannot be guaranteed for any variable. Nevertheless, as
+\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
+real world from observed data inevitably requires assumptions, and these
+requirements hold true for both observational and experimental data.
 

### 2025-02-13 paper2.qmd
section:theoretical; process:write,figures

@@
 @sec-theory-issues discusses the identified issues in detail.
-@sec-theoretical develops Thurstone's theory to address these challenges.
-The development integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+@sec-theoretical extends Thurstone's theory to address these challenges.
+The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
 @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data.
@@
 
-# Developing Thurstone's theory {#sec-theoretical}
+# Extending Thurstone's theory {#sec-theoretical}
 
@@
 A motivating example helps to clarify these concepts.
-This example will appear throughout the document to illustrate various aspects of the proposed development.
+This example will appear throughout the document to illustrate various aspects of the proposed extension.
 In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researcher design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
@@
 <!-- 6. What the section will do? -->
-Thus, this section addresses the issues identified in @sec-theory-issues by developing  Thurstone’s theory using the structural approach to causal inference.
+Thus, this section addresses the issues identified in @sec-theory-issues by extending  Thurstone’s theory using the structural approach to causal inference.
 Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
@@
 
-In line with Thurstone's theory, the other variables present in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
-Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the [SCM -@fig-cj03_scm] incorporates the judges' biases $(B_{kj})$.
+In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
+Additionally, based on the arguments in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
 Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
@@
 
-Conceptual-population model, preliminary integration of theoretical principles.
+Conceptual-population model, scalar form.
 :::
@@
 DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
-Moreover, they often omit exogenous variables for simplicity, following the *standard representation*.
-However, this section adopts the *magnified representation* of DAGS, as including exogenous variables can enhance clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+Moreover, they often omit exogenous variables for simplicity, following a *standard representation*.
+However, this section adopts the *magnified representation* of DAGs, as including exogenous variables in their representation can enhance clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
 <!-- 5. the preliminary DAG for CJ -->
-@fig-cj03_dag displays the DAG associated with [SCM -@fig-cj03_scm].
-This graph illustrates the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and the practical aspects of CJ experiments.
-It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent within the CJ framework, such as the text's discriminal processes, their discriminal difference, and the judges' biases.
+@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the causal relationships among variables based on Thurstone's theory and the practical aspects of CJ experiments.
+Particularly, the graph shows how the texts' discriminal processes $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome.
+It also depicts the judges' biases $(B_{kj})$ as an additional factor influencing the texts' discriminal difference.
+Furthermore, the DAG helps distinguish between observed endogenous variables, such as the outcome, and latent endogenous variables, which include the texts' discriminal processes, their discriminal difference, and the judges' biases.
 
@@
 Although specifying a system's data structure is not mandatory for using SCMs and DAGs, defining one can sometimes enhance clarity and facilitate the description of the system.
-Thus, to simplify the description of the CJ system, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
+Thus, to simplify the description of a CJ experiment, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
 
@@
 
-Readers can better visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals $(n_{I} = 5)$, the number of available stimuli per individual $(n_{A} = 2)$, the total number of judges $(n_{J} = 3)$, and the number of judgments each judge performs $(n_{K} = 3)$.
-Using these values, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:
+Readers can better visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals, the number of available stimuli per individual, the total number of judges, and the number of judgments each judge performs.
+For instance, assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:
 
@@
 <!-- 8. how de we use that data structure? -->
-Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can now re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
-In this representation, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are no longer scalar values but vectors.
-These vectors capture all the values from the conceptual population.
-Specifically, $O^{cp}_{RPUV}$ and $D_{RPUV}$ are vectors of length ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, while $T_{IA}$ and $B_{JK}$ are vectors of length $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.
+Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
+In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values.
+These vectors encompass all values from the conceptual population.
+Specifically, $O^{cp}_{RPUV}$ is an observed vector of length ${n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, while $D_{RPUV}$ is a latent vector of the same dimension.
+Additionally, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.
 
@@
 
-Conceptual-population model, vectorized form of [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag].
+Conceptual-population model, vectorized form.
 :::
@@
 <!-- 9. what else? -->
-Building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+Moreover, building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
 138] and Item Response Theory (IRT) [@Fox_2010, chap.
@@
 Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
-This addition facilitates analyses that integrate the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing the measurement precision.
+This addition facilitates analyses integrating the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing measurement precision.
 
@@
 Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component.
-This component describes how the students' writing-quality trait $T_{I}$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written-quality trait $T_{IA}$.
-The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain.
+This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$ causally influence the texts' written-quality trait $(T_{IA})$.
+The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain.
 Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively.
-Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $p_{IA}$ columns.
-
-Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $X_{I}$ causally influence the students' writing-quality trait $T_{I}$.
-The DAG also includes the idiosyncratic errors $e_{I}$, which capture variations in the the students' traits that the relevant student-related variables cannot explain.
-Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ is an observed matrix with $n_{I}$ rows and $p_{I}$ columns.
+Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns.
+
+Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $(X_{I})$ causally influence the students' writing-quality trait $(T_{I})$.
+The DAG also includes the idiosyncratic errors $(e_{I})$, which capture variations in the students' traits that the relevant student-related variables cannot explain.
+Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns.
 
@@
 
-Conceptual-population model, full integration of theoretical principles.
+Conceptual-population model, final vectorized form.
 :::
@@
 Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component.
-This component describes how the judges' bias $B_{J}$ and a set of *relevant* judgment-related variables $Z_{JK}$ causally influence the biases associated with each text $B_{JK}$.
-The DAG also highlights the idiosyncratic errors $e_{JK}$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain.
+This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$ causally influence the biases associated with each text $B_{JK}$.
+The DAG also highlights the idiosyncratic errors $(e_{JK})$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain.
 In this description, $B_{J}$ and $e_{JK}$ are latent vectors of length $n_{J}$ and $(n_{J} \cdot n_{K})$, respectively.
-Additionally, $Z_{JK}$ is an observed matrix with a dimension of $(n_{J} \cdot n_{K})$ rows and $p_{JK}$ columns.
-
-Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $Z_{J}$ causally influence the judges' bias $B_{J}$.
-The DAG also accounts for idiosyncratic errors $e_{J}$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain.
-Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $p_{J}$ columns.
+Additionally, $Z_{JK}$ is an observed matrix with a dimension of $(n_{J} \cdot n_{K})$ rows and $q_{JK}$ columns.
+
+Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $(Z_{J})$ causally influence the judges' bias $(B_{J})$.
+The DAG also accounts for idiosyncratic errors $(e_{J})$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain.
+Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns.
 Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] constitute the set of endogenous variables $V$ and structural equations $F$ for the conceptual-population model.
 
-In summary,
+Overall, the conceptual-population model extends Thurstone's theory by introducing key innovations to address the limitations discussed in @sec-theory-issues.
+These enhancements include accounting for judges' biases and integrating hierarchical structural components.
+Nevertheless, despite its statistical and measurement advantages, the model still relies on the unrealistic assumption that researchers can access data capturing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
+Since this assumption is rarely met in practice, a more realistic scenario must be considered.
 
@@
 
-A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
+<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
 In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
 37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-
-Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
+-->
+
+<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
 This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
@@
 32].
-
-Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
+-->
+
+<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
 Nevertheless, as @Kohler_et_al_2019 [pp.
 150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
+-->
 

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

@@
 theory. Section~\ref{sec-theory-issues} discusses the identified issues
-in detail. Section~\ref{sec-theoretical} develops Thurstone's theory to
-address these challenges. The development integrates core theoretical
+in detail. Section~\ref{sec-theoretical} extends Thurstone's theory to
+address these challenges. The extension integrates core theoretical
 principles alongside key assessment design features relevant to CJ
@@
 
-\section{Developing Thurstone's theory}\label{sec-theoretical}
+\section{Extending Thurstone's theory}\label{sec-theoretical}
 
@@
 appear throughout the document to illustrate various aspects of the
-proposed development. In this example, researchers aim to determine:
-``To what extent do different teaching methods influence students'
-ability to produce high-quality written texts?'' To investigate this,
-researcher design a CJ experiment by randomly assigning students
-(individuals) to two groups, each receiving a different teaching method.
-Judges then compare pairs of students' written texts (stimuli), to
-produce a dichotomous outcome that reflects the relative quality of each
-text (trait). Based on this setup, researchers reformulate the research
+proposed extension. In this example, researchers aim to determine: ``To
+what extent do different teaching methods influence students' ability to
+produce high-quality written texts?'' To investigate this, researcher
+design a CJ experiment by randomly assigning students (individuals) to
+two groups, each receiving a different teaching method. Judges then
+compare pairs of students' written texts (stimuli), to produce a
+dichotomous outcome that reflects the relative quality of each text
+(trait). Based on this setup, researchers reformulate the research
 question as the estimand: ``\emph{On average}, is there a difference in
@@
 Thus, this section addresses the issues identified in
-Section~\ref{sec-theory-issues} by developing Thurstone's theory using
+Section~\ref{sec-theory-issues} by extending Thurstone's theory using
 the structural approach to causal inference. Specifically, it leverages
@@
 
-In line with Thurstone's theory, the other variables present in
-SCM~\ref{fig-cj03_scm} include the texts' discriminal processes
-\((T_{ha}, T_{ib})\) and their discriminal difference \((D_{hiabjk})\)
-(see Section~\ref{sec-thurstone_theory}). Additionally, based on the
-arguments presented in Section~\ref{sec-theory-issue1b} and the
-recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
-the SCM~\ref{fig-cj03_scm} incorporates the judges' biases \((B_{kj})\).
-Together with the outcome, these variables form the preliminary set of
-endogenous variables,
+In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
+texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
+difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
+Additionally, based on the arguments in Section~\ref{sec-theory-issue1b}
+and the recommendations of \citet{Andrich_1978} and
+\citet{Wainer_et_al_1978}, the SCM incorporates the judges' biases
+\((B_{kj})\). Together with the outcome, these variables form the
+preliminary set of endogenous variables,
 \(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
@@
 
-\caption{\label{fig-cj03}Conceptual-population model, preliminary
-integration of theoretical principles.}
+\caption{\label{fig-cj03}Conceptual-population model, scalar form.}
 
@@
 \citep{Morgan_et_al_2014}. Moreover, they often omit exogenous variables
-for simplicity, following the \emph{standard representation}. However,
-this section adopts the \emph{magnified representation} of DAGS, as
-including exogenous variables can enhance clarity and reveal potential
-issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}.
-
-Figure~\ref{fig-cj03_dag} displays the DAG associated with
-SCM~\ref{fig-cj03_scm}. This graph illustrates the direction of causal
-influences among variables, reflecting the assumptions of Thurstone's
-theory and the practical aspects of CJ experiments. It also
-distinguishes between observable endogenous variables, such as the
-outcome, and variables considered latent within the CJ framework, such
-as the text's discriminal processes, their discriminal difference, and
-the judges' biases.
+for simplicity, following a \emph{standard representation}. However,
+this section adopts the \emph{magnified representation} of DAGs, as
+including exogenous variables in their representation can enhance
+clarity and reveal potential issues related to conditioning and
+confounding \citep{Cinelli_et_al_2020}.
+
+Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
+SCM~\ref{fig-cj03_scm}, illustrating the causal relationships among
+variables based on Thurstone's theory and the practical aspects of CJ
+experiments. Particularly, the graph shows how the texts' discriminal
+processes \((T_{ha}, T_{ib})\) influence their discriminal difference
+\((D_{hiabjk})\), which in turn determines the outcome. It also depicts
+the judges' biases \((B_{kj})\) as an additional factor influencing the
+texts' discriminal difference. Furthermore, the DAG helps distinguish
+between observed endogenous variables, such as the outcome, and latent
+endogenous variables, which include the texts' discriminal processes,
+their discriminal difference, and the judges' biases.
 
@@
 SCMs and DAGs, defining one can sometimes enhance clarity and facilitate
-the description of the system. Thus, to simplify the description of the
-CJ system, we first define the vectors \(I\) and \(J\), along with the
+the description of the system. Thus, to simplify the description of a CJ
+experiment, we first define the vectors \(I\) and \(J\), along with the
 matrices \(IA\) and \(JK\), as follows:
@@
 by considering small values for the parameters corresponding to number
-of individuals \((n_{I} = 5)\), the number of available stimuli per
-individual \((n_{A} = 2)\), the total number of judges \((n_{J} = 3)\),
-and the number of judgments each judge performs \((n_{K} = 3)\). Using
-these values, the matrices described in equations (\ref{eq-mat12}) and
+of individuals, the number of available stimuli per individual, the
+total number of judges, and the number of judgments each judge performs.
+For instance, assuming \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and
+\(n_{K} = 3\), the matrices described in equations (\ref{eq-mat12}) and
 (\ref{eq-mat3}) take the following form:
@@
 Thus, using the data structure defined by the matrices in equations
-(\ref{eq-mat12}) and (\ref{eq-mat3}), we can now re-express
+(\ref{eq-mat12}) and (\ref{eq-mat3}), we can re-express
 SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
-vectorized form, as shown in Figure~\ref{fig-cj04}. In this
-representation, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
-difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
-judges' biases \(B_{JK}\) are no longer scalar values but vectors. These
-vectors capture all the values from the conceptual population.
-Specifically, \(O^{cp}_{RPUV}\) and \(D_{RPUV}\) are vectors of length
-\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
-\(T_{IA}\) and \(B_{JK}\) are vectors of length \((n_{I} \cdot n_{A})\)
-and \((n_{J} \cdot n_{K})\), respectively.
+vectorized form, as shown in Figure~\ref{fig-cj04}. In this depiction,
+the outcome \(O^{cp}_{RPUV}\), the texts' discriminal difference
+\(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the judges'
+biases \(B_{JK}\) are represented as vectors rather than scalar values.
+These vectors encompass all values from the conceptual population.
+Specifically, \(O^{cp}_{RPUV}\) is an observed vector of length
+\({n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
+\(D_{RPUV}\) is a latent vector of the same dimension. Additionally,
+\(T_{IA}\) and \(B_{JK}\) are latent vectors of lengths
+\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\), respectively.
 
@@
 
-\caption{\label{fig-cj04}Conceptual-population model, vectorized form of
-SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag}.}
+\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}
 
@@
 
-Building on the arguments developed in Section~\ref{sec-theory-issue2}
-and the principles of Structural Equation Modeling (SEM)
-\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
+Moreover, building on the arguments developed in
+Section~\ref{sec-theory-issue2} and the principles of Structural
+Equation Modeling (SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item
+Response Theory (IRT)
 \citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
@@
 variables influence the primary latent variable of interest
-\citep{Everitt_et_al_2010}. This addition facilitates analyses that
-integrate the trait measurement and hypothesis testing processes,
+\citep{Everitt_et_al_2010}. This addition facilitates analyses
+integrating the trait measurement and hypothesis testing processes,
 simultaneously accounting for the traits and their associated
-uncertainties, thereby reducing bias and enhancing the measurement
+uncertainties, thereby reducing bias and enhancing measurement
 precision.
@@
 DAG~\ref{fig-cj09_dag} depicts the first component. This component
-describes how the students' writing-quality trait \(T_{I}\) and a set of
-\emph{relevant} text-related variables \(X_{IA}\) causally influence the
-texts' written-quality trait \(T_{IA}\). The DAG also highlights the
-idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
-trait that neither the students' trait nor other relevant text-related
-variables can explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors
-of length \(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively.
-Additionally, \(X_{IA}\) is an observed matrix with a dimension of
-\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.
+describes how the students' writing-quality trait \((T_{I})\) and a set
+of \emph{relevant} text-related variables \((X_{IA})\) causally
+influence the texts' written-quality trait \((T_{IA})\). The DAG also
+highlights the idiosyncratic errors \((e_{IA})\), which capture
+variations in the texts' trait that neither the students' trait nor
+other relevant text-related variables can explain. Here, \(T_{I}\) and
+\(e_{IA}\) are latent vectors of length \(n_{I}\) and
+\((n_{I} \cdot n_{A})\), respectively. Additionally, \(X_{IA}\) is an
+observed matrix with a dimension of \((n_{I} \cdot n_{A})\) rows and
+\(q_{IA}\) columns.
 
 Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
-set of \emph{relevant} student-related variables \(X_{I}\) causally
-influence the students' writing-quality trait \(T_{I}\). The DAG also
-includes the idiosyncratic errors \(e_{I}\), which capture variations in
-the the students' traits that the relevant student-related variables
+set of \emph{relevant} student-related variables \((X_{I})\) causally
+influence the students' writing-quality trait \((T_{I})\). The DAG also
+includes the idiosyncratic errors \((e_{I})\), which capture variations
+in the students' traits that the relevant student-related variables
 cannot explain. Here, \(e_{I}\) is a latent vector of length \(n_{I}\),
-and \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(p_{I}\)
+and \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
 columns.
@@
 
-\caption{\label{fig-cj09}Conceptual-population model, full integration
-of theoretical principles.}
+\caption{\label{fig-cj09}Conceptual-population model, final vectorized
+form.}
 
@@
 second component. This component describes how the judges' bias
-\(B_{J}\) and a set of \emph{relevant} judgment-related variables
-\(Z_{JK}\) causally influence the biases associated with each text
-\(B_{JK}\). The DAG also highlights the idiosyncratic errors \(e_{JK}\),
-which capture variations in the texts' bias that neither the judges'
-bias nor other relevant judgment-related variables can explain. In this
-description, \(B_{J}\) and \(e_{JK}\) are latent vectors of length
-\(n_{J}\) and \((n_{J} \cdot n_{K})\), respectively. Additionally,
-\(Z_{JK}\) is an observed matrix with a dimension of
-\((n_{J} \cdot n_{K})\) rows and \(p_{JK}\) columns.
+\((B_{J})\) and a set of \emph{relevant} judgment-related variables
+\((Z_{JK})\) causally influence the biases associated with each text
+\(B_{JK}\). The DAG also highlights the idiosyncratic errors
+\((e_{JK})\), which capture variations in the texts' bias that neither
+the judges' bias nor other relevant judgment-related variables can
+explain. In this description, \(B_{J}\) and \(e_{JK}\) are latent
+vectors of length \(n_{J}\) and \((n_{J} \cdot n_{K})\), respectively.
+Additionally, \(Z_{JK}\) is an observed matrix with a dimension of
+\((n_{J} \cdot n_{K})\) rows and \(q_{JK}\) columns.
 
 Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
-set of \emph{relevant} judge-related variables \(Z_{J}\) causally
-influence the judges' bias \(B_{J}\). The DAG also accounts for
-idiosyncratic errors \(e_{J}\), capturing the variations in the judges'
-bias that relevant judge-related variables cannot explain. Here,
+set of \emph{relevant} judge-related variables \((Z_{J})\) causally
+influence the judges' bias \((B_{J})\). The DAG also accounts for
+idiosyncratic errors \((e_{J})\), capturing the variations in the
+judges' bias that relevant judge-related variables cannot explain. Here,
 \(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
-observed matrix with \(n_{J}\) rows and \(p_{J}\) columns. Notably, all
+observed matrix with \(n_{J}\) rows and \(q_{J}\) columns. Notably, all
 variables and functions shown in SCM~\ref{fig-cj09_scm} and
@@
 
-In summary,
+Overall, the conceptual-population model extends Thurstone's theory by
+introducing key innovations to address the limitations discussed in
+Section~\ref{sec-theory-issues}. These enhancements include accounting
+for judges' biases and integrating hierarchical structural components.
+Nevertheless, despite its statistical and measurement advantages, the
+model still relies on the unrealistic assumption that researchers can
+access data capturing all repeated judgments from every judge for each
+pair of stimuli produced by every individual in the population. Since
+this assumption is rarely met in practice, a more realistic scenario
+must be considered.
 
@@
 research}\label{sec-discussion-limitations}
-
-A cautious reader will notice that the model descriptions of
-Section~\ref{sec-theory-theoretical_P} rely on several relevant
-variables. In this study, a \emph{relevant} or a \emph{sufficient
-adjustment} set of variables (potentially an empty set) refers to those
-that satisfy the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020},
-that is, variables that block all non-causal paths without introducing
-new ones \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
-
-Incorporating these variables into the conceptual-population model
-enables researchers to identify different estimands of interest and
-their causal effects, even with observational data. This identification
-works because conditioning on these variables ensures that the estimator
-meets several critical properties, including confounding elimination
-\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
-the statistical analysis to a subset of the population defined by the
-values of a given variable \citep[pp.~32]{Neal_2020}.
-
-Naturally, the validity of causal claims now hinges on the assumption
-that these variables serve as a sufficient adjustment set, a property
-that cannot be guaranteed for any variable. Nevertheless, as
-\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
-real world from observed data inevitably requires assumptions, and these
-requirements hold true for both observational and experimental data.
 

### 2025-02-13 paper2.qmd
section:theoretical; process:write

@@
 A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables.
-The term *directed* indicates that edges or arrows extend from one node to another, showing the direction of causal influence.
+The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence, while the absence of an edge implies no relationship between the nodes.
 The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
-DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
-Moreover, they often omit exogenous variables for simplicity, following a *standard representation*.
-However, this section adopts the *magnified representation* of DAGs, as including exogenous variables in their representation can enhance clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+DAGs conventionally represent observed variables with solid black circles and unobserved (latent) variables with open circles [@Morgan_et_al_2014].
+They also display conditioned variables by surrounding them with a square.
+Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp.
+32].
+While DAGs often omit exogenous variables for simplicity, following a *standard representation*, this section adopts the *magnified representation* of DAGs, as including exogenous variables can improve clarity and help reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 It also depicts the judges' biases $(B_{kj})$ as an additional factor influencing the texts' discriminal difference.
-Furthermore, the DAG helps distinguish between observed endogenous variables, such as the outcome, and latent endogenous variables, which include the texts' discriminal processes, their discriminal difference, and the judges' biases.
+Furthermore, the DAG distinguish between observed endogenous variables, such as the outcome, and latent endogenous variables, which include the texts' discriminal processes, their discriminal difference, and the judges' biases.
 
@@
   B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
+  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
+  e_{JK} & \dsep e_{J} \\
 \end{aligned}
@@
 Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns.
-Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] constitute the set of endogenous variables $V$ and structural equations $F$ for the conceptual-population model.
+
+Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] define the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
+Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.
 
@@
 These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its statistical and measurement advantages, the model still relies on the unrealistic assumption that researchers can access data capturing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
-Since this assumption is rarely met in practice, a more realistic scenario must be considered.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
+Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
+The extensions to the conceptual-population model, derived from this realistic scenario, are concisely described in the sample-comparison model.
 
@@
 
-<!-- comparison mechanism as missing mechanism -->
 <!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
@@
 
-Considering the sampling mechanism
+In the sample-comparison model, researchers have access to a limited number of judgments ($n^{s}{K}$) from a sample of judges ($n^{s}{J}$), along with a specific number of texts ($n^{s}{A}$) from a sample of students ($n^{s}{I}$) drawn from the conceptual population.
+
+@fig-cj11 illustrates the sample-comparison model, incorporating the sample mechanism.
+Following the convention of @McElreath_2020 [pp.
+499-516], the [DAG -@fig-cj11_dag] shows several conditioned variables $(S_{R}$, $S_{P}$, $S_{U}$, and $S_{V})$ that causally influence the sample-comparison outcome $O^{sc}_{RPUV}$, reflecting how the outcome is a sample drawn from the conceptual population.
+The DAG also depicts the conceptual-population outcome $O^{cp}_{RPUV}$ is now an unobserved variable, emphasizing the researchers' inability to access such an outcome.
 
@@
 \begin{aligned}
-  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
   O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
@@
 
-CJ's outcome, discriminal difference, discriminal processes and judges' biases
+Sample-comparison model including sampling mechanism
 :::
@@
   O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
-  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
   O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
@@
 
-CJ's outcome, discriminal difference, discriminal processes and judges' biases
-:::
-
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
-267] -->
+Sample-comparison model including sample and comparison mechanisms
+:::
 
@@
 
+<!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
 <!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:56:10 -->
@@
 
+<!-- sampling and comparisons mechanism are effective generators of missing data. -->
+
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
 ## Limitations and further research {#sec-discussion-limitations}
@@
 This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
-Here, *conditioning* means restricting the statistical analysis to a subset of the population defined by the values of a given variable [@Neal_2020, pp.
-32].
 -->

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

@@
 variables. The term \emph{directed} indicates that edges or arrows
-extend from one node to another, showing the direction of causal
-influence. The term \emph{acyclic} means that the causal influences do
+extend from one node to another, indicating the direction of causal
+influence, while the absence of an edge implies no relationship between
+the nodes. The term \emph{acyclic} means that the causal influences do
 not form loops, ensuring the influences do not cycle back on themselves
-\citep{McElreath_2020}. DAGs conventionally depict observed variables as
-solid black circles and unobserved (latent) variables as open circles
-\citep{Morgan_et_al_2014}. Moreover, they often omit exogenous variables
-for simplicity, following a \emph{standard representation}. However,
-this section adopts the \emph{magnified representation} of DAGs, as
-including exogenous variables in their representation can enhance
-clarity and reveal potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}.
+\citep{McElreath_2020}. DAGs conventionally represent observed variables
+with solid black circles and unobserved (latent) variables with open
+circles \citep{Morgan_et_al_2014}. They also display conditioned
+variables by surrounding them with a square. Here, \emph{conditioning}
+refers to restricting the statistical analysis to a subset of the
+population defined by the values of the given variable
+\citep[pp.~32]{Neal_2020}. While DAGs often omit exogenous variables for
+simplicity, following a \emph{standard representation}, this section
+adopts the \emph{magnified representation} of DAGs, as including
+exogenous variables can improve clarity and help reveal potential issues
+related to conditioning and confounding \citep{Cinelli_et_al_2020}.
 
@@
 the judges' biases \((B_{kj})\) as an additional factor influencing the
-texts' discriminal difference. Furthermore, the DAG helps distinguish
-between observed endogenous variables, such as the outcome, and latent
+texts' discriminal difference. Furthermore, the DAG distinguish between
+observed endogenous variables, such as the outcome, and latent
 endogenous variables, which include the texts' discriminal processes,
@@
   B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) 
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
+  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
+  e_{JK} & \:\bot\:e_{J} \\
 \end{aligned}
@@
 \(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
-observed matrix with \(n_{J}\) rows and \(q_{J}\) columns. Notably, all
-variables and functions shown in SCM~\ref{fig-cj09_scm} and
-DAG~\ref{fig-cj09_dag} constitute the set of endogenous variables \(V\)
-and structural equations \(F\) for the conceptual-population model.
+observed matrix with \(n_{J}\) rows and \(q_{J}\) columns.
+
+Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
+DAG~\ref{fig-cj09_dag} define the set of endogenous variables \(V\),
+structural equations \(F\), and exogenous variables \(E\) for the
+conceptual-population model. Additionally, the figures demonstrate that
+all exogenous variables are independent of one another, as indicated by
+the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
+\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).
 
@@
 for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its statistical and measurement advantages, the
-model still relies on the unrealistic assumption that researchers can
-access data capturing all repeated judgments from every judge for each
-pair of stimuli produced by every individual in the population. Since
-this assumption is rarely met in practice, a more realistic scenario
-must be considered.
+Nevertheless, despite its promise of enhancing measurement accuracy and
+precision, the model still depends on the unrealistic assumption that
+researchers have access to data encompassing all repeated judgments from
+every judge for each pair of stimuli produced by every individual in the
+population. Since researchers rarely meet this assumption in practice,
+they must consider a more realistic scenario. The extensions to the
+conceptual-population model, derived from this realistic scenario, are
+concisely described in the sample-comparison model.
 
@@
 
-Considering the sampling mechanism
+In the sample-comparison model, researchers have access to a limited
+number of judgments (\(n^{s}{K}\)) from a sample of judges
+(\(n^{s}{J}\)), along with a specific number of texts (\(n^{s}{A}\))
+from a sample of students (\(n^{s}{I}\)) drawn from the conceptual
+population.
+
+Figure~\ref{fig-cj11} illustrates the sample-comparison model,
+incorporating the sample mechanism. Following the convention of
+\citet[pp.~499-516]{McElreath_2020}, the DAG~\ref{fig-cj11_dag} shows
+several conditioned variables \((S_{R}\), \(S_{P}\), \(S_{U}\), and
+\(S_{V})\) that causally influence the sample-comparison outcome
+\(O^{sc}_{RPUV}\), reflecting how the outcome is a sample drawn from the
+conceptual population. The DAG also depicts the conceptual-population
+outcome \(O^{cp}_{RPUV}\) is now an unobserved variable, emphasizing the
+researchers' inability to access such an outcome.
 
@@
 
-\caption{\label{fig-cj11}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
+\caption{\label{fig-cj11}Sample-comparison model including sampling
+mechanism}
 
@@
 
-\caption{\label{fig-cj13}CJ's outcome, discriminal difference,
-discriminal processes and judges' biases}
+\caption{\label{fig-cj13}Sample-comparison model including sample and
+comparison mechanisms}
 

### 2025-02-14 paper2.qmd
section:theoretical; process:write;figures,code

@@
 $$
-RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
+RPUV_{n \times 6} = \begin{bmatrix}
 1 & 1 & 1 & 2 & 1 & 1 \\
@@
 
-This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
+This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
 The matrix's dimension results from the conceptual-population model's assumption that researchers have access to data capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
@@
   D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
@@
   e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
-  e_{JK} & \dsep e_{J} \\
+  e_{JK} & \dsep e_{J} 
 \end{aligned}
@@
 
-@fig-cj11 illustrates the sample-comparison model, incorporating the sample mechanism.
+@fig-cj12 illustrates the sample-comparison model, incorporating the sample mechanism.
 Following the convention of @McElreath_2020 [pp.
-499-516], the [DAG -@fig-cj11_dag] shows several conditioned variables $(S_{R}$, $S_{P}$, $S_{U}$, and $S_{V})$ that causally influence the sample-comparison outcome $O^{sc}_{RPUV}$, reflecting how the outcome is a sample drawn from the conceptual population.
+499-516], the [DAG -@fig-cj12_dag] shows several conditioned variables $(S_{R}$, $S_{P}$, $S_{U}$, and $S_{V})$ that causally influence the sample-comparison outcome $O^{sc}_{RPUV}$, reflecting how the outcome is a sample drawn from the conceptual population.
 The DAG also depicts the conceptual-population outcome $O^{cp}_{RPUV}$ is now an unobserved variable, emphasizing the researchers' inability to access such an outcome.
 
-::: {#fig-cj11 layout-nrow=2}
-
-::: {#fig-cj11_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
+Concatenating the matrices
+
+::: {#fig-cj12 layout-nrow=2}
+
+::: {#fig-cj12_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
   O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
   D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
@@
 
-![DAG](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=84%}
+![DAG](/images/png/CJ_TM_12.png){#fig-cj12_dag fig-align="center" fig-pos="H" width=84%}
 
@@
 
-::: {#fig-cj13 layout-nrow=2}
-
-::: {#fig-cj13_scm fig-align="center" fig-pos="H"}
+::: {#fig-cj14 layout-nrow=2}
+
+::: {#fig-cj14_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
+  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
+  e_{JK} & \dsep e_{J} 
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_14.png){#fig-cj14_dag fig-align="center" fig-pos="H" width=100%}
+
+Sample-comparison model including sample and comparison mechanisms
+:::
+
+# Abandoning the BTL model {#sec-statistical}
+
+<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
+
+::: {#fig-cj15 layout-ncol=2}
+
+::: {#fig-cj15_stat fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
+  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S_{RPUV} \\
+  O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
+  D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
+  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
+  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
+  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
+  e_{I} & \sim \text{Normal}(0,1) \\
+  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\
+  e_{J} & \sim \text{Normal}(0,1)
+\end{aligned}
+$$
+
+Statistical model
+:::
+
+::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
 $$
@@
   e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
-  e_{JK} & \dsep e_{J} \\
+  e_{JK} & \dsep e_{J}
 \end{aligned}
@@
 
-![DAG](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=100%}
-
 Sample-comparison model including sample and comparison mechanisms
 :::
-
-# Abandoning the BTL model {#sec-statistical}
-
-<!-- recording: Sven 24.10.04; time: 00:46:00 - 01:01:00 -->
-<!-- recording: Sven and Tine 24.11.25; time: 00:24:50 - 00:30:35 -->
-<!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
 

### 2025-02-14 _tex/paper2.tex
Built site for gh-pages

@@
 \begin{equation}\phantomsection\label{eq-mat3}{
-RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
+RPUV_{n \times 6} = \begin{bmatrix}
 1 & 1 & 1 & 2 & 1 & 1 \\
@@
 \(\mathbb{N}\) and has a dimension of
-\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
-columns. The matrix's dimension results from the conceptual-population
-model's assumption that researchers have access to data capturing all
-repeated judgments made by every available judge for each pair of
-stimuli produced by each pair of individuals in the population. Here,
-the term \({n_{I} \cdot n_{A} \choose 2}\) represents the Binomial
-coefficient, which quantifies the total number of individual-stimuli
-comparison pairs within the conceptual population. The coefficient is
-defined as follows:
+\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
+\(6\) columns. The matrix's dimension results from the
+conceptual-population model's assumption that researchers have access to
+data capturing all repeated judgments made by every available judge for
+each pair of stimuli produced by each pair of individuals in the
+population. Here, the term \({n_{I} \cdot n_{A} \choose 2}\) represents
+the Binomial coefficient, which quantifies the total number of
+individual-stimuli comparison pairs within the conceptual population.
+The coefficient is defined as follows:
 
@@
   e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
-  e_{JK} & \:\bot\:e_{J} \\
+  e_{JK} & \:\bot\:e_{J} 
 \end{aligned}
@@
 
-Figure~\ref{fig-cj11} illustrates the sample-comparison model,
+Figure~\ref{fig-cj12} illustrates the sample-comparison model,
 incorporating the sample mechanism. Following the convention of
-\citet[pp.~499-516]{McElreath_2020}, the DAG~\ref{fig-cj11_dag} shows
+\citet[pp.~499-516]{McElreath_2020}, the DAG~\ref{fig-cj12_dag} shows
 several conditioned variables \((S_{R}\), \(S_{P}\), \(S_{U}\), and
@@
 
+Concatenating the matrices
+
 \begin{figure}[H]
@@
 \begin{aligned}
-  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
   O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
   D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
@@
 
-\subcaption{\label{fig-cj11_scm}SCM}
+\subcaption{\label{fig-cj12_scm}SCM}
 
@@
 
-\includegraphics[width=0.84\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}
-
-}
-
-\subcaption{\label{fig-cj11_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj11}Sample-comparison model including sampling
+\includegraphics[width=0.84\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}
+
+}
+
+\subcaption{\label{fig-cj12_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj12}Sample-comparison model including sampling
 mechanism}
@@
 \begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
+  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
+  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
+  e_{JK} & \:\bot\:e_{J} 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj14_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}
+
+}
+
+\subcaption{\label{fig-cj14_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj14}Sample-comparison model including sample and
+comparison mechanisms}
+
+\end{figure}%
+
+\section{Abandoning the BTL model}\label{sec-statistical}
+
+\begin{figure}[H]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
+  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S_{RPUV} \\
+  O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
+  D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
+  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
+  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
+  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
+  e_{I} & \sim \text{Normal}(0,1) \\
+  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\
+  e_{J} & \sim \text{Normal}(0,1)
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj15_stat}Statistical model}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
 
@@
   e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
-  e_{JK} & \:\bot\:e_{J} \\
+  e_{JK} & \:\bot\:e_{J}
 \end{aligned}
@@
 
-\subcaption{\label{fig-cj13_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}
-
-}
-
-\subcaption{\label{fig-cj13_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj13}Sample-comparison model including sample and
+\subcaption{\label{fig-cj15_scm}SCM}
+
+\end{minipage}%
+
+\caption{\label{fig-cj15}Sample-comparison model including sample and
 comparison mechanisms}
@@
 \end{figure}%
-
-\section{Abandoning the BTL model}\label{sec-statistical}
 

### 2025-02-14 code/CJmod_add.R

@@ -28,16 +28,16 @@ data{
     array[nsI*nsA] int<lower=1, upper=nsA> IA2s;    // stimuli
     array[nsI*nsA] real XIc;                        // individual predictor (continuous)
     array[nsI*nsA] int<lower=1, upper=sup_dXI> XId; // individual predictor (discrete)
-    array[nsI*nsA] real XAc;                        // stimuli predictor (continuous)
-    array[nsI*nsA] int<lower=1, upper=sup_dXA> XAd; // stimuli predictor (discrete)
+    array[nsI*nsA] real XIAc;                       // stimuli predictor (continuous)
+    array[nsI*nsA] int<lower=1, upper=sup_dXA> XIAd;// stimuli predictor (discrete)
     
     // juges-repeated comparisons data
     array[nsJ*nsK] int<lower=1, upper=nsJ> JK1s;    // judges
     array[nsJ*nsK] int<lower=1, upper=nsK> JK2s;    // judgments
     array[nsJ*nsK] real ZJc;                        // judges predictor (continuous)
     array[nsJ*nsK] int<lower=1, upper=sup_dZJ> ZJd; // judges predictor (discrete)
-    array[nsJ*nsK] real ZKc;                        // judgments predictor (continuous)
-    array[nsJ*nsK] int<lower=1, upper=sup_dZK> ZKd; // judgments predictor (discrete)
+    array[nsJ*nsK] real ZJKc;                       // judgments predictor (continuous)
+    array[nsJ*nsK] int<lower=1, upper=sup_dZK> ZJKd;// judgments predictor (discrete)
     
 }
 parameters{
@@ -80,7 +80,7 @@ transformed parameters{
     eIA = pIA * zeIA;    // identification as a prop. of previous level
     for( ia in 1:(nsI*nsA) ){
       TI[ IA1s[ia] ] = bXIc*XIc[ ia ] + bXId[ XId[ia] ] + eI[ IA1s[ia] ];
[…]

### 2025-02-25 paper2.qmd
section:code links; process:write,figures,math
section:all; process:write,figures,code

@@
 Meanwhile, research on the validity of CJ scores indicates their capacity to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
-Moreover, research on CJ's practical applicability highlights its versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Moreover, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-The study then addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on causal and Bayesian inference methods.
+This study addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on Causal and Bayesian inference methods.
 
@@
 @sec-theory-issues discusses the identified issues in detail.
-@sec-theoretical extends Thurstone's theory to address these challenges.
+@sec-theoretical extends Thurstone's general form to address these challenges.
 The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
 @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data.
-@sec-discussion discusses the implications of the findings and explores avenues for future research.
+@sec-discussion discusses the findings, limitations, challenges and explores avenues for future research.
 Finally, @sec-conclusion summarizes the study's key insights.
@@
 Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome.
-Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
+Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978, pp.
+452].
 In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.
@@
 
-# Two prominent issues in CJ literature {#sec-theory-issues}
-
-This section identifies and discusses the two prominent issues that can undermine the reliability and validity of CJ's trait estimates.
-@sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-@sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-
-## The Case V and the statistical analysis of CJ data {#sec-theory-issue1}
+# Two prominent issues in CJ practice {#sec-theory-issues}
+
+<!-- 1. What is the problem with the general form -->
+Thurstone noted from the outset that his general form, described in @sec-thurstone_theory, led to a *trait scaling problem*.
+Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b, pp.
+267].
+For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases).
+However, a single judge could provide only ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.
+
+<!-- 2. Thurstone's Cases and their simplifying assumptions -->
+To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions into the model.
+In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons.
+In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge.
+In Case III, he posited that there was no correlation between stimuli.
+In Case IV, he assumed that the stimuli exhibited similar dispersions.
+Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions.
+@tbl-thurstone_cases summarizes the assumptions of the general form and the five cases.
+For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
+248–253].
+
+<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->
+
+![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}
+
+<!-- 3. But are our specific problems? -->
+It is evident that although Thurstone developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+269] and cautioned that its use "should not be made without experimental test" [@Thurstone_1927b, pp.
+270], the CJ literature has nevertheless relied the most on it in practice.
+This preference is particularly notable given that Case V incorporates the most extensive set of simplifying assumptions among all the cases [@Bramley_2008, pp.
+253; @Kelly_et_al_2022, pp.
+677] (see @tbl-thurstone_cases).
+
+Furthermore, since Thurstone's primary goal was to define a psychological scale and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp.
+269], he did not provide guidance on how to use these measurements when, for instance, researchers aim to test hypotheses related to the measured traits.
+Although the CJ practice has circumvented this issue by using scores' point estimates or their transformations for such analyses, a fundamental question remains: Is this method appropriate for addressing these inquiries?
+
+<!-- 4. Section division -->
+Thus, this section discusses these two prominent issues, as they could undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015, pp.
+2].
+Specifically, @sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+Moreover, @sec-theory-issue2 focuses on the apparent disconnect between the approaches to trait measurement and hypothesis testing in CJ.
+
+## The overreliance on Case V in the statistical analysis of CJ data {#sec-theory-issue1}
 
@@
 
-<!-- 1. What is the problem with the general form -->
-Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a trait scaling problem.
-Specifically, the model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp.
-267].
-To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions into the model.
-
-<!-- 2. Thurstone's Cases and their simplifying assumptions -->
-In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons.
-In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge.
-In Case III, he posited that there was no correlation between stimuli.
-In Case IV, he assumed that the stimuli exhibited similar dispersions.
-Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions.
-@tbl-thurstone_cases summarizes the assumptions of the general form and the five cases.
-For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
-248–253].
-
-<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->
-
-![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}
-
-<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
-Notably, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp.
+<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
+As previously discussed, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp.
 253; @Kelly_et_al_2022, pp.
@@
 This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
+The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
 254] (see @tbl-thurstone_cases).
-This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp.
-16].
+As noted by @vanderLinden_et_al_2017_I [pp.
+16], this substitution has little impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation.
+Furthermore, this minimal impact is reinforced by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.
 
@@
 
-<!-- 4. But what are limitations of Case V (and the BTL model)? -->
-However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+<!-- 2. But what are limitations of Case V (and the BTL model)? -->
+However, as mentioned earlier, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
@@
 
-<!-- 1. What does the equal dispersion between stimuli imply? -->
+<!-- 1. But does it hold?  -->
+Thurstone argued that the assumption of equal dispersions may not apply when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+Indeed, the CJ literature may already provide evidence of this violation through "misfit" statistics.
+For example, labeling texts as *misfits* indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp.
+11; @Pollitt_2012b, pp.
+290].
+These discrepancies imply that the discriminal differences for *misfit* texts have broader distributions, further suggesting that their discriminal processes exhibit more dispersion than those of other texts.
+This idea can be illustrated using @fig-discriminal_process, where text B shows a wider discriminal dispersion than text A.
+As a result, the texts' discriminal processes overlap more, making it difficult for a judge to determine which text possesses more of the trait.
+
+A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves [@Pollitt_2012a, pp.
+164-165; @Pollitt_2012b, pp.
+289-290; @vanDaal_et_al_2016, pp.
+4; @Goossens_et_al_2018, pp.
+20].
+These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.
+
+<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
+Thurstone clearly did not expect this to hold for any but the most simple stimuli.
+The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
+-->
+
+<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
+In measurements of the type known as judgment scales the discriminal dispersion on the psychological continuum becomes one of the unknowns to be determined as well as the scale value of the specimen.
+Every specimen in such a series presents two unknown values to be determined, namely, the scale value of its modal discriminal process on the psychological continuum and its discriminal dispersion.
+[@Thurstone_1927a, pp.
+376] -->
+
+<!-- For instance, in evaluating creativity, a student might complete an assignment in a highly creative but unexpected manner.
+Judges, unprepared for such originality, may produce more divergent evaluations of the assignment.
+These increased discrepancies in judgments suggest that the discriminal difference associated with the assignment has a broader dispersion, indicating greater variation in the discriminal process for this assignment compared to others.
+-->
+
+<!-- "weak" writers consistently produce texts of low quality in all aspects that determine the quality of a text, such as grammatical accuracy, organization, and language use.
+In such cases, judges typically agree on their inferior quality, leading to a clear disfavoring of these texts compared to others.
+This agreement ultimately imply narrow dispersions for the discriminal differences related to these texts and, by extension, narrow discriminal processes for them.
+However, as texts improve along the trait continuum, with certain aspects showing improvement and others reaming unaltered, these variations introduce uncertainty into the judgments.
+Now, since judges often rely on various intricate characteristics of the texts to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], these changes lead to greater discrepancies in their evaluations.
+This divergence ultimately imply broader discriminal differences and, by extension, more varied discriminal processes for the texts.
+-->
+
+<!-- So it seem that the extreme of the discriminal process distribution of the stimuli have higher reliability than the center? -->
+
+<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
+[@Pollitt_2012a, pp.
+165] -->
+
+<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
+
+<!-- a misfit statistic.
+Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
+It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
+12] -->
+
+<!-- 2. What does the equal dispersion between stimuli imply? -->
 According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons.
-A thought experiment can illustrate this concept.
-In this experiment, the researcher observes the discriminal processes for the texts depicted in @fig-discriminal_process.
+A thought experiment can illustrate this idea.
+In this experiment, researchers can observe the discriminal processes for the texts depicted in @fig-discriminal_process.
 Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
-@fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference.
-This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome.
-Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ.
-Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area)
+@fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B compared to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference.
+This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B>A)$, ultimately influencing the comparison outcome.
+Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ.
+Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area).
 
@@
 
-The effect of dispersion discrepancies and stimulus correlation on the distribution of the discriminal difference.
+The effect of dispersion discrepancies and stimuli correlation on the distribution of the discriminal difference.
 :::
@@
 In experimental practice, however, the thought experiment occurs in reverse.
-Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp.
+Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp.
 373].
-Therefore, the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
+Consequently, the ability of the outcome to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
 150], particularly the assumption of equal dispersions.
-For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts.
-This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
-The accuracy of the discriminal difference then ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.
+For example, when researchers correctly assume equal dispersions between texts, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between them.
+This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
+The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.
 
@@
 
-<!-- 3. But does it hold?  -->
-However, Thurstone argued that the assumption of equal dispersions may not be applicable when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-Indeed, evidence of this violation may already be present in the CJ literature in the form of misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp.
-12; @Goossens_et_al_2018, pp.
-20].
-For example, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
-These discrepancies, in turn, suggest that the discriminal differences for "misfit" texts have broader distributions, indicating that their discriminal processes may also exhibit more variation than that of other texts.
-A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.
-These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.
-
-<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
-Thurstone clearly did not expect this to hold for any but the most simple stimuli.
-The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
--->
-
-<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
-In measurements of the type known as judgment scales the discriminal dispersion on the psychological continuum becomes one of the unknowns to be determined as well as the scale value of the specimen.
-Every specimen in such a series presents two unknown values to be determined, namely, the scale value of its modal discriminal process on the psychological continuum and its discriminal dispersion.
-[@Thurstone_1927a, pp.
-376] -->
-
-<!-- For instance, in evaluating creativity, a student might complete an assignment in a highly creative but unexpected manner.
-Judges, unprepared for such originality, may produce more divergent evaluations of the assignment.
-These increased discrepancies in judgments suggest that the discriminal difference associated with the assignment has a broader dispersion, indicating greater variation in the discriminal process for this assignment compared to others.
--->
-
-<!-- "weak" writers consistently produce texts of low quality in all aspects that determine the quality of a text, such as grammatical accuracy, organization, and language use.
-In such cases, judges typically agree on their inferior quality, leading to a clear disfavoring of these texts compared to others.
-This agreement ultimately imply narrow dispersions for the discriminal differences related to these texts and, by extension, narrow discriminal processes for them.
-However, as texts improve along the trait continuum, with certain aspects showing improvement and others reaming unaltered, these variations introduce uncertainty into the judgments.
-Now, since judges often rely on various intricate characteristics of the texts to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], these changes lead to greater discrepancies in their evaluations.
-This divergence ultimately imply broader discriminal differences and, by extension, more varied discriminal processes for the texts.
--->
-
-<!-- So it seem that the extreme of the discriminal process distribution of the stimuli have higher reliability than the center? -->
-
-<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
-[@Pollitt_2012a, pp.
-165] -->
-
-<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
-
-<!-- a misfit statistic.
-Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
-It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
-12] -->
-
-<!-- 4. So, what happens when the assumption does not hold? -->
-Thus, erroneously assuming equal dispersions between stimuli, can give rise to significant statistical and measurement issues.
-For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli.
-This overestimation can result in researchers drawing spurious conclusions about these differences [@McElreath_2020, pp.
+<!-- 2. So, what happens when the assumption does not hold? -->
+However, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues.
+Specifically, the model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
+This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp.
 370] and, by extension, about the underlying discriminal processes of stimuli.
-@fig-dispersion also illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
+@fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
 Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
@@
 
-<!-- 1. What does the zero correlation between stimuli imply? -->
-The correlation, represented by the symbol $\rho$, measures how much a judge's perception of a specific trait in one stimulus depends on their perception of the same trait in another.
+<!-- 1. But does it hold?  -->
+In contrast, Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b, pp.
+268].
+@Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
+However, the assumption of zero correlation is not met in practice in at least two scenarios.
+First, when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions, generating biases that resist cancellation.
+Second, when an additional hierarchical structure is relevant to the stimuli.
+
+<!-- 2. The first scenario where the assumption does not hold -->
+In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+These additional relevant characteristics are unlikely to be equally weighted or opposing.
+As a result, they may exert an uneven influence on judges' perceptions, creating biases in their judgments that resist cancellation.
+For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments.
+Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
+346].
+While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
+
+<!-- If the test taker’s position on one latent trait is fixed, the assumption of local stochastic independence requires the association between the responses to the items to vanish.
+Suppose the assumption is violated and more than one dimension is necessary to describe the test takers’ position in the latent space.
+The association between the responses to the items given one proficiency parameter does then not vanish.
+[@vanderLinden_et_al_2017_II, pp.
+346].
+-->
+
+<!-- research shows that, in the presence of these traits and stimuli, the accuracy of the judges' trait perception is influenced by the rank-order distance of the stimuli [@vanDaal_et_al_2017; @Gijsen_et_al_2021], a coarse measure of the relative positioning of the stimuli within the trait continuum.
+-->
+
+<!-- 3. The second scenario where the assumption does not hold -->
+In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010].
+Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient.
+For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
+
+<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
+One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
+In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
+Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
+-->
+
+<!-- 4. What does the zero correlation between stimuli imply? -->
+According to the theory, the correlation $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another.
 As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
-A similar thought experiment, as the one in @sec-theory-issue1a, can illustrate this concept.
+A similar thought experiment, as the one depicted in @sec-theory-issue1a, can illustrate this idea.
 The experiment now assumes that the discriminal dispersions for both texts remain constant.
@@
 
-<!-- 2. But this is not the scenario we observe -->
+<!-- 5. But this is not the scenario we observe -->
 Again, in experimental practice, researchers approach this process in reverse.
-They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli.
-Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts.
-This scenario is also illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$.
-Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.
+They begin by observing the outcomes and use the BTL model to estimate the stimuli' discriminal difference and discriminal processes.
+Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, the model estimates a discriminal difference distribution that accurately reflects the *true* discriminal difference between the texts.
+This scenario is illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the *true* distribution, represented by the thick continuous line corresponding to $\rho=0$.
+Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts.
 
@@
 
-<!-- 2. But does it hold?  -->
-Notably, Thurstone attributed the lack of correlation between stimuli to the cancellation of potential judges' biases.
-He argued that this cancellation resulted from two opposing and equally weighted effects occurring during pairwise comparisons [@Thurstone_1927b, pp.
-268].
-@Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
-However, it is easy to imagine at least two scenarios in which the zero correlation assumption is almost certainly invalid: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.
-
-<!-- 3. The first scenario where the assumption does not hold -->
-In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation.
-Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These additional relevant characteristics are unlikely to be equally weighted or opposing.
-As a result, they may exert an uneven influence on judges' perceptions, creating biases in their judgments.
-Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases may ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
-346].
-For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments.
-While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
-
-<!-- If the test taker’s position on one latent trait is fixed, the assumption of local stochastic independence requires the association between the responses to the items to vanish.
-Suppose the assumption is violated and more than one dimension is necessary to describe the test takers’ position in the latent space.
-The association between the responses to the items given one proficiency parameter does then not vanish.
-[@vanderLinden_et_al_2017_II, pp.
-346].
--->
-
-<!-- research shows that, in the presence of these traits and stimuli, the accuracy of the judges' trait perception is influenced by the rank-order distance of the stimuli [@vanDaal_et_al_2017; @Gijsen_et_al_2021], a coarse measure of the relative positioning of the stimuli within the trait continuum.
--->
-
-<!-- 4. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, creating a statistical phenomenon known as clustering [@Everitt_et_al_2010].
-Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical handling of this additional source of dependence between stimuli has been inadequate.
-For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-This approach, however, can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
-
-<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
-One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
-In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
-Although this example is somewhat contrived, it effectively illustrates how assuming equal dispersions across texts can overlook meaningful differences in the reliability of text quality across groups or individuals.
--->
-
-<!-- 5. But what are the challenges when this assumption is violated? -->
-In any case, similar to @sec-theory-issue1a, assuming zero correlation between stimuli by neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures can cause significant statistical and measurement issues.
-In general, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli.
-Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes.
-This scenario is illustrated by @fig-correlation, when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.
-
-In particular, neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp.
-341] or, worse, introducing bias into the trait's estimates [@Ackerman_1989].
-Excluding judges based on misfit statistics risks discarding valuable information, which may further bias the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap.
+<!-- 6. But what are the challenges when this assumption is violated? -->
+However, erroneously assuming zero correlation between stimuli can lead to significant statistical and measurement issues.
+Specifically, neglecting judges' biases or hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Hoyle_et_al_2023, pp.
+341, 482].
+In other words, the outcome may fail to accurately reflect the *true* discriminal differences between stimuli.
+These inaccuracies can result in spurious conclusions about these differences [@McElreath_2020, pp.
+370] and, by extension, the underlying discriminal processes  [@Ackerman_1989].
+This issue is illustrated in @fig-correlation when the model discriminal difference distribution follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $(\rho \neq 0)$.
+Conversely, excluding judges based on misfit statistics risks discarding valuable information, potentially biasing the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
-Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially overestimating the trait's reliability [@Hoyle_et_al_2023, pp.
-482].
+Again, the direction and magnitude of these biases are often unpredictable, as they depend on which judges are excluded from the analysis.
 
@@
 
+<!-- 1. What happens when you want to test hypotheses -->
+<!-- Although the CJ practice has found a way to circumvent this 'inconvenience', by using estimated scores or their transformations to conduct additional analyses and tests, a critical question remains: Are these scores an appropriate method for addressing such inquiries?
+-->
+
 <!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is clear that, despite its limitations, the BTL model is commonly used as a measurement model in CJ assessments.
-A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010].
+Building on the previous section, it is clear that, despite its limitations, the BTL model commonly serves as the measurement model for CJ assessments.
+Specifically, the model provides a method for transforming manifest into latent variables (psychological scales) and allocating the compared stimuli on this continuum [@Thurstone_1927b, pp.
+269; @Everitt_et_al_2010].
 For example, when evaluating writing quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying level of writing quality (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
@@
 <!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Researchers then typically use these estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests.
+Researchers then typically use the BTL scores' point estimates, or their transformations, to conduct additional analyses or hypothesis tests.
 For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 <!-- 4. segway for next section -->
-In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of the trait estimates.
+In aggregate, excessive reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates.
 This overreliance may also undermine their validity [@Perron_et_al_2015, pp.
-2], especially when combined with the disconnect between CJ's approach to trait measurement and hypothesis testing.
-However, the structural approach to causal inference can address these issues by providing a systematic and integrated framework that enhances statistical accuracy while strengthening measurement reliability and validity.
+2], particularly when coupled with the disconnect between CJ's approach to trait measurement and hypothesis testing.
+However, the structural approach to causal inference can address these issues by offering a systematic and integrated framework that strengthens measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.
 
@@
 
-# Extending Thurstone's theory {#sec-theoretical}
+# Extending Thurstone's general form {#sec-theoretical}
 
@@
 The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments.
-
-<!-- 2. How does SCMs and DAGs are used? -->
 Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
@@
 *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].
-
 <!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
@@
 <!-- 3. Example of a CJ estimand and estimator, and its limitations -->
-A motivating example helps to clarify these concepts.
-This example will appear throughout the document to illustrate various aspects of the proposed extension.
+A motivating example that will appear throughout the document clarifies these concepts.
 In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researcher design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
-Judges then compare pairs of students’ written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait).
+Judges then compare pairs of students' written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait).
 Based on this setup, researchers reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
@@
 However, @sec-theory-issues presents compelling evidence that Thurstone's Case V, and by extension the BTL model, suffers from several statistical and measurement limitations.
-These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the motivating example.
+These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the example.
 Identification is crucial because it is a necessary condition for ensuring consistent estimators.
@@
 <!-- 6. What the section will do? -->
-Thus, this section addresses the issues identified in @sec-theory-issues by extending  Thurstone’s theory using the structural approach to causal inference.
-Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone’s general form using the structural approach to causal inference.
+Specifically, it formalizes the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
 In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
@@
 Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we call the *conceptual-population model*.
-This model assumes that researchers have access to comparison data from a conceptual population, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population, hence its name.
+This model assumes that researchers have access to a *census* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population, hence its name.
 Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we call the *sample-comparison model*.
@@
 
+### Integrating initial theoretical principles {#sec-theory-theoretical_P1}
+
 <!-- 1. The SCMs -->
-Before incorporating the theoretical principles into the *conceptual-population model*, it is essential to define SCMs.
+Before incorporating the theoretical principles into the *conceptual-population model*, it is essential to further define SCMs.
 SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
@@
 <!-- 2. the preliminary SCM for CJ -->
-[SCM -@fig-cj03_scm] illustrates the preliminary integration of theoretical principles into the conceptual-population model of the motivating example.
+[SCM -@fig-cj03_scm] presents the initial integration of theoretical principles into the conceptual-population model for the example, applied to examine the effect of different teaching methods on students' writing ability.
 This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
@@
 In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
-Additionally, based on the arguments in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
+Additionally, based on the arguments developed under the first scenario of @sec-theory-issue1b,  and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
 Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
@@
 <!-- 5. the preliminary DAG for CJ -->
-@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the causal relationships among variables based on Thurstone's theory and the practical aspects of CJ experiments.
-Particularly, the graph shows how the texts' discriminal processes $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome.
+@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the causal relationships among variables based on Thurstone's theory.
+Particularly, the graph shows how the texts' discriminal processes $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$.
 It also depicts the judges' biases $(B_{kj})$ as an additional factor influencing the texts' discriminal difference.
@@
 
-<!-- 6. Why is useful to declare the data structure? -->
-Although specifying a system's data structure is not mandatory for using SCMs and DAGs, defining one can sometimes enhance clarity and facilitate the description of the system.
-Thus, to simplify the description of a CJ experiment, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
+### The comparative census data structure {#sec-theory-theoretical_P2}
+
+<!-- 1. Why is useful to declare the data structure? -->
+Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one can improve clarity and facilitate the description of the system.
+Thus, to re-express the scalar form of the CJ system, outlined in the previous section, into an simpler equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
 
@@
 n_{I}
+\end{bmatrix} ; \;
+J_{n_{J}} = \begin{bmatrix}
+1 \\
+2 \\
+\vdots \\
+j \\
+\vdots \\
+n_{J}
 \end{bmatrix} ; \;
@@
 \end{bmatrix} ; \;
-J_{n_{J}} = \begin{bmatrix}
-1 \\
-2 \\
-\vdots \\
-j \\
-\vdots \\
-n_{J}
-\end{bmatrix} ; \;
 JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
@@
 
-<!-- 7. what is the data structure? -->
+<!-- 2. what is the data structure? -->
 where an element of $I$ corresponds to a unique individual $h$ or $i$, while each row of $IA$ corresponds to a unique pairing of individuals $h, i$ with stimuli $a, b$.
@@
 
-Similarly, an element of $J$ corresponds to a unique judge $j$.
+Similarly, an element of the vector $J$ corresponds to a unique judge $j$.
 Moreover, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment number $k$.
@@
 n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
-\end{bmatrix} \in \mathbb{N}
+\end{bmatrix}
 $$ {#eq-mat3}
@@
 This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
-The matrix's dimension results from the conceptual-population model's assumption that researchers have access to data capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+The matrix's dimension results from the conceptual-population model's assumption that researchers have access to a *census* of comparative data, capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
 Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs within the conceptual population.
@@
 
-Readers can better visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals, the number of available stimuli per individual, the total number of judges, and the number of judgments each judge performs.
+We visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals, the number of available stimuli per individual, the total number of judges, and the number of judgments each judge performs.
 For instance, assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:
@@
 
-<!-- 8. how de we use that data structure? -->
-Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
+<!-- 3. how de we use that data structure? -->
+Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in a simpler equivalent vectorized form, as shown in @fig-cj04.
 In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values.
@@
 
-<!-- 9. what else? -->
-Moreover, building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+### Integrating hierarchical structural components {#sec-theory-theoretical_P3}
+
+<!-- 1. What are we going to do? -->
+Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
 138] and Item Response Theory (IRT) [@Fox_2010, chap.
 6; @vanderLinden_et_al_2017_I, chap.
-24], the conceptual-population model includes two hierarchical structural components.
-Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
-This addition facilitates analyses integrating the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing measurement precision.
+24], the conceptual-population model incorporates two *hierarchical structural components*.
+Each component describes how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
+This addition addresses the problems arising when an additional hierarchical structure is relevant to the stimuli, as outlined in the second scenario of @sec-theory-issue1b.
+Furthermore, it facilitates analyses that integrate both trait measurement and hypothesis testing processes, thereby accounting for both traits and their associated uncertainties, reducing bias and improving measurement precision (see @sec-theory-issue2).
 
@@
 
-@fig-cj09 displays the two hierarchical structural components for the motivating example.
+<!-- 2 Explaining what we did -->
+@fig-cj09 displays the two hierarchical structural components for the example, applied to examine the effect of different teaching methods on students' writing ability.
 Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component.
-This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$ causally influence the texts' written-quality trait $(T_{IA})$.
+This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$, such as text length, causally influence the texts' written-quality trait $(T_{IA})$.
 The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain.
@@
 
-Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $(X_{I})$ causally influence the students' writing-quality trait $(T_{I})$.
+Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $(X_{I})$, including teaching methods, causally influence the students' writing-quality trait $(T_{I})$.
 The DAG also includes the idiosyncratic errors $(e_{I})$, which capture variations in the students' traits that the relevant student-related variables cannot explain.
@@
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
-  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
-  e_{JK} & \dsep e_{J} 
+  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \dsep e_{JK} 
 \end{aligned}
@@
 Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component.
-This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$ causally influence the biases associated with each text $B_{JK}$.
+This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$, such as number of judgments or time for judgment, causally influence the biases associated with each text $B_{JK}$.
 The DAG also highlights the idiosyncratic errors $(e_{JK})$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain.
@@
 
-Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $(Z_{J})$ causally influence the judges' bias $(B_{J})$.
+Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $(Z_{J})$, including judgment expertise, causally influence the judges' bias $(B_{J})$.
 The DAG also accounts for idiosyncratic errors $(e_{J})$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain.
@@
 
-Overall, the conceptual-population model extends Thurstone's theory by introducing key innovations to address the limitations discussed in @sec-theory-issues.
+<!-- 3. segway for next section -->
+Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issues.
 These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *census* of comparative data, encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
 Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
@@
 
-In the sample-comparison model, researchers have access to a limited number of judgments ($n^{s}{K}$) from a sample of judges ($n^{s}{J}$), along with a specific number of texts ($n^{s}{A}$) from a sample of students ($n^{s}{I}$) drawn from the conceptual population.
-
-@fig-cj12 illustrates the sample-comparison model, incorporating the sample mechanism.
+### The sample mechanism {#sec-theory-theoretical_SC1}
+
+<!-- 1. What is the more realistic scenario: the sample mechanism -->
+In the sample-comparison model, we assume a more realistic scenario compared to the conceptual-population model.
+Specifically, we assume that researchers have access to *all comparisons* from a limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$, along with a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$ drawn from the conceptual population.
+@fig-cj12 illustrates this model, incorporating only this sample mechanism.
+
+<!-- 2. What is the effect of the sampling mechanism -->
 Following the convention of @McElreath_2020 [pp.
-499-516], the [DAG -@fig-cj12_dag] shows several conditioned variables $(S_{R}$, $S_{P}$, $S_{U}$, and $S_{V})$ that causally influence the sample-comparison outcome $O^{sc}_{RPUV}$, reflecting how the outcome is a sample drawn from the conceptual population.
-The DAG also depicts the conceptual-population outcome $O^{cp}_{RPUV}$ is now an unobserved variable, emphasizing the researchers' inability to access such an outcome.
-
-Concatenating the matrices
+499-516], the [DAG -@fig-cj12_dag] depicts the conceptual-population outcome $O^{cp}_{RPUV}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the sampling mechanism.
+Moreover, the DAG shows the conditioned *sample design* variable $S$, which causally influence the sample-comparison outcome $O^{sc}_{RPUV}$.
+This variable reflects how this sample-comparison outcome is a sample drawn from the conceptual-population outcome $O^{cp}_{RPUV}$.
+Specifically, the variable is a vector $S = \{ s_{1}, \dots, s_{n} \}$, where $n$ corresponds to the number of rows in the matrix $RPUV$ (refer to @sec-theory-theoretical_P2), and each $s_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the conceptual-population outcome $O^{cp}_{RPUV}$.
+Specifically,
+
+$$
+s_i = \begin{cases} 
+1 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is observed} \\
+0 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is missing}
+\end{cases}
+$$
 
@@
 \begin{aligned}
-  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
+  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
   O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
@@
 
+Notice,
 <!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
 
-Considering comparison mechanisms
+### The comparison mechanism {#sec-theory-theoretical_SC2}
+
+<!-- 1. What is the comparison mechanism -->
+Conversely, we make the scenario depicted in @sec-theory-theoretical_SC1 more realistic by integrating the comparison mechanism, that is, we assume that researchers do not have access to *all comparisons* from the sample, but rather a sufficient number of comparisons $(n_{C})$ so that a proportion, $P(B>A)$, may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267].
+@fig-cj14 illustrates this model, incorporating both the sample and comparison mechanisms.
+
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
+<!-- 2. What is the effect of the sampling mechanism -->
+Following the convention of @McElreath_2020 [pp.
+499-516], the [DAG -@fig-cj14_dag] depicts the sample-comparison outcome $O^{sc}_{RPUV}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the comparison mechanism.
+Moreover, the DAG shows the conditioned *comparison design* variable $C$, which causally influence the observed outcome $O_{RPUV}$.
+This variable reflects how this observed outcome is a sample drawn from the sample-comparison outcome $O^{sp}_{RPUV}$.
+Specifically, the variable is a vector $C = \{ c_{1}, \dots, c_{n} \}$, where $n$ corresponds to the number of rows in the matrix $RPUV$ (refer to @sec-theory-theoretical_P2), and each $c_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the sample-comparison outcome $O^{sp}_{RPUV}$.
+Specifically,
+
+$$
+c_i = \begin{cases} 
+1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{RPUV} \text{ is observed} \\
+0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{RPUV} \text{ is missing}
+\end{cases}
+$$
 
@@
   O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
-  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
+  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
   O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
@@
 :::
+
+Notice,
 
@@
   O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
-  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S_{RPUV} \\
+  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S \\
   O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
@@
   B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
+  e_{I} & \sim \text{Normal}(0,1) \\
+  e_{J} & \sim \text{Normal}(0,1) \\
   e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
-  e_{I} & \sim \text{Normal}(0,1) \\
-  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\
-  e_{J} & \sim \text{Normal}(0,1)
+  e_{JK} & \sim \text{Normal}(0, p_{JK})
 \end{aligned}
@@
   O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
-  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
+  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
   O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
@@
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
-  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
-  e_{JK} & \dsep e_{J}
+  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \dsep e_{JK}
 \end{aligned}
@@
 
-Sample-comparison model including sample and comparison mechanisms
+Sample-comparison model, including sample and comparison mechanisms and assuming equal discriminal dispersions for the student's traits
+:::
+
+::: {#fig-cj16 layout-ncol=2}
+
+::: {#fig-cj16_stat fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
+  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S \\
+  O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
+  D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
+  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
+  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
+  e_{I} & \sim \text{Normal}(0,s_{XI}) \\
+  e_{J} & \sim \text{Normal}(0,s_{ZJ}) \\
+  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
+  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\ \\
+  & \textbf{Contraints:} \\
+  & \sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1 ; \quad \sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1
+\end{aligned}
+$$
+
+Statistical model
+:::
+
+::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
+  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & := f_{e}(X_{I}) \\
+  e_{J} & := f_{e}(Z_{J}) \\
+  e_{IA} & \dsep e_{JK}
+\end{aligned}
+$$
+
+SCM
+:::
+
+Sample-comparison model, including sample and comparison mechanisms and assuming different discriminal dispersions for the student's traits
 :::
@@
 
+<!-- @Bramley_2008 proposes ranking instead of comparisons -->
+
 # Discussion {#sec-discussion}
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:56:10 -->
+
+<!-- Thus, this section addresses the issues identified in @sec-theory-issues by extending  Thurstone’s theory using the structural approach to causal inference.
+Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
+First, it clarifies the interactions among all actors and processes involved in CJ experiments.
+Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.
+-->
 
@@
 267] -->
-
-## Limitations and further research {#sec-discussion-limitations}
-
-<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
-In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
-37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
--->
-
-<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
-This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
--->
-
-<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
-Nevertheless, as @Kohler_et_al_2019 [pp.
-150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
--->
-
-<!-- Statements about the real world from observed data require assumptions.
-This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
-The validity of our statements therefore critically hinges on the validity of those assumptions.
-If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
-[@Kohler_et_al_2019, pp.
-150] -->
 
@@
 
-<!-- @Bramley_2008 proposes ranking instead of comparisons -->
+## Limitations and further research {#sec-discussion-limitations}
+
+<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
+In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
+37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+-->
+
+<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
+This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
+-->
+
+<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
+Nevertheless, as @Kohler_et_al_2019 [pp.
+150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
+-->
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
 

### 2025-02-25 code/CJmod_add.R

@@ -1,3 +1,5 @@
+# equal variances ####
+
 mcmc_code = "
 data{
 
@@ -54,45 +56,54 @@ parameters{
     
     
     // error parameters 
-    vector[nsI] eI;       // individuals errors (identification)
-    real<lower=0> pIA;    // proportion of variability (identification)
+    vector[nsI*nsA] zeI;  // individual errors (non-centered)
+    real<lower=0> pIA;    // stimuli proportion of variability
     matrix[nsI,nsA] zeIA; // stimuli errors (non-centered)
-    
-    vector[nsJ] eJ;       // judges errors (identification)
-    real<lower=0> pJK;    // proportion of variability (identification)
+
+    vector[nsJ*nsK] zeJ;  // judges errors (identification)
+    real<lower=0> pJK;    // judgments proportion of variability
     matrix[nsJ,nsK] zeJK; // judgments errors (non-centered)
-    
+
 }
 transformed parameters{
     
     // declaring
     vector[nsI] TI;       // individuals' trait
+    vector[nsI*nsA] eI;   // individuals errors (identification)
     matrix[nsI,nsA] TIA;  // stimuli trait
     matrix[nsI,nsA] eIA;  // stimuli errors
     
     vector[nsJ] BJ;       // judges' trait
+    vector[nsJ*nsK] eJ;   // judges' errors (identification)
     matrix[nsJ,nsK] BJK;  // judges-judgments trait
     matrix[nsJ,nsK] eJK;  // judgments errors
     
     
     // individuals-stimuli
     // error calculation
-    eIA = pIA * zeIA;    // identification as a prop. of previous level
+    eI = zeI;             // identification: sI = 1
+    eIA = pIA * zeIA;     // identification: pIA prop. of previous level
+    
+    // trait calculation
     for( ia in 1:(nsI*nsA) ){
-      TI[ IA1s[ia] ] = bXIc*XIc[ ia ] + bXId[ XId[ia] ] + eI[ IA1s[ia] ];
+      TI[ IA1s[ia] ] = bXIc*XIc[ ia ] + bXId[ XId[ia] ] + eI[ ia ];
[…]

### 2025-02-25 _tex/paper2.tex
Built site for gh-pages

@@
 \citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
-Moreover, research on CJ's practical applicability highlights its
+Moreover, research on its practical applicability highlights CJ's
 versatility across both educational and non-educational contexts
@@
 The second stems from the apparent disconnect between CJ's approach to
-trait measurement and hypothesis testing. The study then addresses these
+trait measurement and hypothesis testing. This study addresses these
 issues by extending Thurstone's general form through a systematic and
-integrated approach based on causal and Bayesian inference methods.
+integrated approach based on Causal and Bayesian inference methods.
 
@@
 theory. Section~\ref{sec-theory-issues} discusses the identified issues
-in detail. Section~\ref{sec-theoretical} extends Thurstone's theory to
-address these challenges. The extension integrates core theoretical
-principles alongside key assessment design features relevant to CJ
-experiments, such as the selection of judges, stimuli, and comparisons.
-Section~\ref{sec-statistical} translates these theoretical and practical
-elements into a probabilistic statistical model for the analysis of
-dichotomous pairwise comparison data. Section~\ref{sec-discussion}
-discusses the implications of the findings and explores avenues for
-future research. Finally, Section~\ref{sec-conclusion} summarizes the
-study's key insights.
+in detail. Section~\ref{sec-theoretical} extends Thurstone's general
+form to address these challenges. The extension integrates core
+theoretical principles alongside key assessment design features relevant
+to CJ experiments, such as the selection of judges, stimuli, and
+comparisons. Section~\ref{sec-statistical} translates these theoretical
+and practical elements into a probabilistic statistical model for the
+analysis of dichotomous pairwise comparison data.
+Section~\ref{sec-discussion} discusses the findings, limitations,
+challenges and explores avenues for future research. Finally,
+Section~\ref{sec-conclusion} summarizes the study's key insights.
 
@@
 distribution on the continuum, the discriminal difference will also
-conform to a Normal distribution \citep{Andrich_1978}. In this
+conform to a Normal distribution \citep[pp.~452]{Andrich_1978}. In this
 distribution, the mode (mean) represents the relative separation between
@@
 
-\section{Two prominent issues in CJ literature}\label{sec-theory-issues}
-
-This section identifies and discusses the two prominent issues that can
-undermine the reliability and validity of CJ's trait estimates.
-Section~\ref{sec-theory-issue1} examines the the excessive reliance on
-Thurstone's Case V assumptions in the statistical analysis of CJ data.
-Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
-between CJ's approach to trait measurement and hypothesis testing.
-
-\subsection{The Case V and the statistical analysis of CJ
-data}\label{sec-theory-issue1}
-
-Thurstone noted from the outset that the general form of the theory, as
-outlined in Section~\ref{sec-thurstone_theory}, gave rise to a trait
-scaling problem. Specifically, the model required estimating more
-``unknown'' parameters than the available pairwise comparisons
-\citep[pp.~267]{Thurstone_1927b}. To address this issue and facilitate
-the practical implementation of the theory, he developed five cases
-derived from this general form, each progressively incorporating
-additional simplifying assumptions into the model.
-
-In Case I, Thurstone postulated that pairs of stimuli would maintain a
-constant correlation across all comparisons. In Case II, he allowed
-multiple judges to undertake comparisons instead of confining
+\section{Two prominent issues in CJ practice}\label{sec-theory-issues}
+
+Thurstone noted from the outset that his general form, described in
+Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
+problem}. Specifically, the model required estimating more ``unknown''
+parameters than the number of available pairwise comparisons
+\citep[pp.~267]{Thurstone_1927b}. For instance, in a CJ experiment with
+five texts, the general form would require estimating \(20\) parameters:
+five modal discriminal processes, five discriminal dispersions, and
+\(10\) correlations --one per comparison (see
+Table~\ref{tbl-thurstone_cases}). However, a single judge could provide
+only \({5 \choose 2} = 10\) unique comparisons, an insufficient data set
+to estimate the required parameters.
+
+To address this issue and facilitate the practical implementation of the
+theory, Thurstone developed five cases derived from this general form,
+each progressively incorporating additional simplifying assumptions into
+the model. In Case I, Thurstone postulated that pairs of stimuli would
+maintain a constant correlation across all comparisons. In Case II, he
+allowed multiple judges to undertake comparisons instead of confining
 evaluations to a single judge. In Case III, he posited that there was no
@@
 
-Notably, despite relying on the most extensive set of simplifying
-assumptions
+It is evident that although Thurstone developed Case V to provide a
+``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b} and
+cautioned that its use ``should not be made without experimental test''
+\citep[pp.~270]{Thurstone_1927b}, the CJ literature has nevertheless
+relied the most on it in practice. This preference is particularly
+notable given that Case V incorporates the most extensive set of
+simplifying assumptions among all the cases
+\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
+(see Table~\ref{tbl-thurstone_cases}).
+
+Furthermore, since Thurstone's primary goal was to define a
+psychological scale and ``allocate the compared stimuli on this
+continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not provide
+guidance on how to use these measurements when, for instance,
+researchers aim to test hypotheses related to the measured traits.
+Although the CJ practice has circumvented this issue by using scores'
+point estimates or their transformations for such analyses, a
+fundamental question remains: Is this method appropriate for addressing
+these inquiries?
+
+Thus, this section discusses these two prominent issues, as they could
+undermine the reliability and validity of CJ's trait estimates
+\citep[pp.~2]{Perron_et_al_2015}. Specifically,
+Section~\ref{sec-theory-issue1} examines the the excessive reliance on
+Thurstone's Case V assumptions in the statistical analysis of CJ data.
+Moreover, Section~\ref{sec-theory-issue2} focuses on the apparent
+disconnect between the approaches to trait measurement and hypothesis
+testing in CJ.
+
+\subsection{The overreliance on Case V in the statistical analysis of CJ
+data}\label{sec-theory-issue1}
+
+As previously discussed, despite relying on the most extensive set of
+simplifying assumptions
 \citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
@@
 assumptions of Case V, with one notable distinction: whereas Case V
-assumes a Normal distribution for the stimuli's discriminal processes,
-the BTL model uses the more mathematically tractable Logistic
-distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
-Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
-the model's estimation or interpretation, as the Normal and Logistic
-distributions exhibit analogous statistical properties, differing only
-by a scaling factor of approximately \(1.7\)
-\citep[pp.~16]{vanderLinden_et_al_2017_I}.
-
-However, Thurstone originally developed Case V to provide a ``rather
-coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
-prioritizing statistical simplicity over precision in trait measurement
-\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
-untested application, stating that its use ``should not be made without
-(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}. Furthermore,
-he acknowledged that some assumptions could prove problematic when
-researchers assess complex traits or heterogeneous stimuli
-\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
-applications frequently involve such traits and stimuli, two main
-assumptions of Case V and, by extension, of the BTL model may not
-consistently hold in theory or practice, namely the assumption of equal
-dispersion and zero correlation between stimuli.
+assumes a Normal distribution for the stimuli discriminal processes, the
+BTL model uses the more mathematically tractable Logistic distribution
+\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
+Table~\ref{tbl-thurstone_cases}). As noted by
+\citet[pp.~16]{vanderLinden_et_al_2017_I}, this substitution has little
+impact on the model's estimation or interpretation because the
+discriminal process scale is arbitrary up to a non-monotonic
+transformation. Furthermore, this minimal impact is reinforced by the
+fact that the Normal and Logistic distributions exhibit analogous
+statistical properties, differing only by a scaling factor of
+approximately \(1.7\).
+
+However, as mentioned earlier, Thurstone originally developed Case V to
+provide a ``rather coarse scaling'' of traits
+\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
+over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
+He explicitly warned against its untested application, stating that its
+use ``should not be made without (an) experimental test''
+\citep[pp.~270]{Thurstone_1927b}. Furthermore, he acknowledged that some
+assumptions could prove problematic when researchers assess complex
+traits or heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}.
+Consequently, given that modern CJ applications frequently involve such
+traits and stimuli, two main assumptions of Case V and, by extension, of
+the BTL model may not consistently hold in theory or practice, namely
+the assumption of equal dispersion and zero correlation between stimuli.
 
@@
 stimuli}\label{sec-theory-issue1a}
+
+Thurstone argued that the assumption of equal dispersions may not apply
+when researchers assess complex traits or heterogeneous stimuli
+\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
+introduce judgment discrepancies due to their unique characteristics
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
+Indeed, the CJ literature may already provide evidence of this violation
+through ``misfit'' statistics. For example, labeling texts as
+\emph{misfits} indicates that comparisons involving these texts result
+in more judgment discrepancies than those involving other texts
+\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
+These discrepancies imply that the discriminal differences for
+\emph{misfit} texts have broader distributions, further suggesting that
+their discriminal processes exhibit more dispersion than those of other
+texts. This idea can be illustrated using
+Figure~\ref{fig-discriminal_process}, where text B shows a wider
+discriminal dispersion than text A. As a result, the texts' discriminal
+processes overlap more, making it difficult for a judge to determine
+which text possesses more of the trait.
+
+A similar line of reasoning applies to the concept of ``misfit'' judges,
+whose evaluations deviate substantially from the shared consensus due to
+the unique characteristics of the stimuli or the judges themselves
+\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.
+These ``misfit'' judges and their associated deviations can give rise to
+additional statistical and measurement issues, which we discuss in more
+detail in Section~\ref{sec-theory-issue1b}.
 
@@
 direct influence on the outcome of pairwise comparisons. A thought
-experiment can illustrate this concept. In this experiment, the
-researcher observes the discriminal processes for the texts depicted in
+experiment can illustrate this idea. In this experiment, researchers can
+observe the discriminal processes for the texts depicted in
 Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
@@
 Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
-associated with the perception of Text B in comparison to Text A,
-\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
+associated with the perception of Text B compared to Text A
+\((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
 discriminal difference. This broadening affects the probability area
 where the discriminal difference distinctly favors Text B over Text A,
-expressed as \(P(B > A)\), ultimately influencing the comparison
-outcome. Additionally, the figure reveals that when the discriminal
-dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
-discriminal difference is more narrow compared to situations where their
-dispersions differ. Consequently, the discriminal difference is more
-likely to favor Text B over Text A (shaded gray area)
+expressed as \(P(B>A)\), ultimately influencing the comparison outcome.
+Additionally, the figure reveals that when the discriminal dispersions
+of the texts are equal \((\sigma_{B} - \sigma_{A}=0)\), the discriminal
+difference is more narrow compared to situations where their dispersions
+differ. Consequently, the discriminal difference is more likely to favor
+Text B over Text A (shaded gray area).
 
@@
 \caption{\label{fig-casev_issues}The effect of dispersion discrepancies
-and stimulus correlation on the distribution of the discriminal
+and stimuli correlation on the distribution of the discriminal
 difference.}
@@
 reverse. Researchers first observe the comparison outcome and then use
-the BTL model to infer the discriminal difference between the stimuli
-and their respective discriminal processes
-\citep[pp.~373]{Thurstone_1927a}. Therefore, the outcome's ability to
-reflect the ``true'' differences between stimuli largely depends on the
-validity of the model's assumptions \citep[pp.~150]{Kohler_et_al_2019},
-particularly the assumption of equal dispersions. For instance, when
-researchers observe a sample of outcomes favoring Text B over Text A and
-correctly assume equal dispersions between the texts, the BTL model
-estimates a discriminal difference distribution that accurately
-represents the ``true'' discriminal difference of the texts. This
-scenario is illustrated in Figure~\ref{fig-dispersion}, when the model's
-discriminal difference distribution aligns with the ``true''
-distribution, represented by the thick continuous line corresponding to
-\(\sigma_{B}-\sigma_{A}=0\). The accuracy of the discriminal difference
-then ensures reliable estimates for the texts' discriminal processes
-{(citation needed?)}.
-
-However, Thurstone argued that the assumption of equal dispersions may
-not be applicable when researchers assess complex traits or
-heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
-and stimuli can introduce judgment discrepancies due to their unique
-features
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
-Indeed, evidence of this violation may already be present in the CJ
-literature in the form of misfit statistics, which measure judgment
-discrepancies associated with specific stimuli
-\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
-For example, labeling texts as ``misfits'' indicates that comparisons
-involving these texts result in more judgment discrepancies than those
-involving other texts
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
-These discrepancies, in turn, suggest that the discriminal differences
-for ``misfit'' texts have broader distributions, indicating that their
-discriminal processes may also exhibit more variation than that of other
-texts. A similar line of reasoning applies to the concept of ``misfit''
-judges, whose evaluations deviate substantially from the shared
-consensus due to the unique characteristics of the stimuli or the judges
-themselves. These ``misfit'' judges and their associated deviations can
-give rise to additional statistical and measurement issues, which we
-discuss in more detail in Section~\ref{sec-theory-issue1b}.
-
-Thus, erroneously assuming equal dispersions between stimuli, can give
-rise to significant statistical and measurement issues. For instance,
-the model may overestimate the degree to which the outcome accurately
-reflects the ``true'' discriminal differences between stimuli. This
-overestimation can result in researchers drawing spurious conclusions
-about these differences \citep[pp.~370]{McElreath_2020} and, by
-extension, about the underlying discriminal processes of stimuli.
-Figure~\ref{fig-dispersion} also illustrates this issue when the model's
-discriminal difference distribution aligns with the thick continuous
-line for \(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal
+the BTL model to infer the discriminal difference between stimuli and
+their respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
+Consequently, the ability of the outcome to reflect \emph{true}
+differences between stimuli largely depends on the validity of the
+model's assumptions \citep[pp.~150]{Kohler_et_al_2019}, particularly the
+assumption of equal dispersions. For example, when researchers correctly
+assume equal dispersions between texts, the BTL model estimates a
+discriminal difference distribution that accurately represents the
+\emph{true} discriminal difference between them. This scenario is
+illustrated in Figure~\ref{fig-dispersion}, when the model's discriminal
+difference distribution aligns with the \emph{true} discriminal
+difference distribution, represented by the thick continuous line
+corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
+discriminal difference then ensures reliable estimates for the texts'
+discriminal processes.
+
+However, erroneously assuming equal dispersions between stimuli can give
+rise to significant statistical and measurement issues. Specifically,
+the model may overestimate the trait's reliability, that is, the degree
+to which the outcome accurately reflects the \emph{true} discriminal
+differences between stimuli. This overestimation, in turn, results in
+spurious conclusions about these differences
+\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
+discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
+illustrates this scenario when the model's discriminal difference
+distribution aligns with the thick continuous line for
+\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
 difference follows any discontinuous line where
@@
 
-The correlation, represented by the symbol \(\rho\), measures how much a
-judge's perception of a specific trait in one stimulus depends on their
+In contrast, Thurstone assumed that stimuli were uncorrelated because
+judges' biases, arising from two opposing and equally weighted effects
+occurring during the pairwise comparisons, canceled each other out
+\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
+mathematical demonstration of this cancellation using the BTL model
+under the assumption of discriminal processes with additive biases.
+However, the assumption of zero correlation is not met in practice in at
+least two scenarios. First, when intricate aspects of multidimensional,
+complex traits or heterogeneous stimuli influence judges' perceptions,
+generating biases that resist cancellation. Second, when an additional
+hierarchical structure is relevant to the stimuli.
+
+In the first scenario, research on text quality suggests that when
+judges evaluate multidimensional, complex traits or heterogeneous
+stimuli, they often rely on various intricate characteristics of the
+stimuli to form their judgments
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
+additional relevant characteristics are unlikely to be equally weighted
+or opposing. As a result, they may exert an uneven influence on judges'
+perceptions, creating biases in their judgments that resist
+cancellation. For example, this could occur when a judge assessing the
+argumentative quality of a text places more weight on its grammatical
+accuracy than other judges, thereby favoring texts with fewer errors but
+weaker arguments. Furthermore, since the discriminal difference of the
+stimuli becomes an observable outcome only through the judges'
+perceptions, these biases could ultimately introduce dependencies
+between the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While
+direct evidence for this particular scenario is lacking, studies such as
+\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
+supporting the notion that the factors influencing pairwise comparisons
+may not always cancel out.
+
+In the second scenario, the shared context or inherent connections
+introduced by additional hierarchical structures may create dependencies
+between stimuli, a statistical phenomenon known as clustering
+\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
+existence of such hierarchical structures, the statistical approaches to
+account for this additional source of dependence have been insufficient.
+For instance, when CJ data incorporates multiple samples of stimuli from
+the same individuals, researchers frequently rely on (averaged) point
+estimates of the BTL scores to conduct subsequent analyses and tests at
+the individual hierarchical level
+\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+However, this approach can introduce additional statistical and
+measurement issues, which we discuss in greater detail in
+Section~\ref{sec-theory-issue2}.
+
+According to the theory, the correlation \(\rho\) measures how much the
+judges' perception of a specific trait in one stimulus depends on their
 perception of the same trait in another. As with the discriminal
@@
 difference, directly impacting the outcomes of pairwise comparisons. A
-similar thought experiment, as the one in
-Section~\ref{sec-theory-issue1a}, can illustrate this concept. The
+similar thought experiment, as the one depicted in
+Section~\ref{sec-theory-issue1a}, can illustrate this idea. The
 experiment now assumes that the discriminal dispersions for both texts
@@
 Again, in experimental practice, researchers approach this process in
-reverse. They begin by observing the sample of outcomes favoring Text B
-over Text A and then use the BTL model to estimate the discriminal
-difference and the discriminal processes of the stimuli. Given that the
-BTL model assumes independent discriminal processes across comparisons,
-if this assumption holds, then the model estimates a discriminal
-difference distribution that accurately reflects the ``true''
-discriminal difference of the texts. This scenario is also illustrated
-in Figure~\ref{fig-correlation} when the discriminal difference
-distribution of the model aligns with the ``true'' distribution,
-represented by the thick continuous line corresponding to \(\rho=0\).
-Once more, the accuracy of the discriminal difference ensures reliable
-estimates for the discriminal processes of the texts {(citation
-needed?)}.
-
-Notably, Thurstone attributed the lack of correlation between stimuli to
-the cancellation of potential judges' biases. He argued that this
-cancellation resulted from two opposing and equally weighted effects
-occurring during pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
-\citet{Andrich_1978} provided a mathematical demonstration of this
-cancellation using the BTL model under the assumption of discriminal
-processes with additive biases. However, it is easy to imagine at least
-two scenarios in which the zero correlation assumption is almost
-certainly invalid: when the pairwise comparison involves
-multidimensional, complex traits with heterogeneous stimuli and when an
-additional hierarchical structure is relevant to the stimuli.
-
-In the first scenario, the intricate aspects of multidimensional,
-complex traits may introduce dependencies between the stimuli due to
-certain judges' biases that resist cancellation. Research on text
-quality suggests that when judges evaluate these traits, they often rely
-on various intricate characteristics of the stimuli to form their
-judgments
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
-additional relevant characteristics are unlikely to be equally weighted
-or opposing. As a result, they may exert an uneven influence on judges'
-perceptions, creating biases in their judgments. Furthermore, since the
-discriminal difference of the stimuli becomes an observable outcome only
-through the judges' perceptions, these biases may ultimately introduce
-dependencies between the stimuli
-\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
-occur when a judge assessing the argumentative quality of a text places
-more weight on its grammatical accuracy than other judges, thereby
-favoring texts with fewer errors but weaker arguments. While direct
-evidence for this particular scenario is lacking, studies such as
-\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
-supporting the notion that the factors influencing pairwise comparisons
-may not always cancel out.
-
-In the second scenario, the shared context or inherent connections
-created by additional hierarchical structures may further introduce
-dependencies between stimuli, creating a statistical phenomenon known as
-clustering \citep{Everitt_et_al_2010}. Although the CJ literature
-acknowledges the existence of such hierarchical structures, the
-statistical handling of this additional source of dependence between
-stimuli has been inadequate. For instance, when CJ data incorporates
-multiple samples of stimuli from the same individuals, researchers
-frequently rely on (averaged) estimated BTL scores to conduct subsequent
-analyses and tests at the individual hierarchical level
-\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-This approach, however, can introduce additional statistical and
-measurement issues, which we discuss in greater detail in
-Section~\ref{sec-theory-issue2}.
-
-In any case, similar to Section~\ref{sec-theory-issue1a}, assuming zero
-correlation between stimuli by neglecting additional relevant traits,
-excluding judges based on misfit statistics, or ignoring hierarchical
-(grouping) structures can cause significant statistical and measurement
-issues. In general, the model may over- or underestimate how accurately
-the outcome reflects the ``true'' discriminal differences between
-stimuli. Such inaccuracies can result in spurious inferences about these
-differences and, by extension, about the stimuli's discriminal
-processes. This scenario is illustrated by Figure~\ref{fig-correlation},
-when the model's discriminal difference distribution aligns with the
-thick continuous line for \(\rho=0\), while the ``true'' discriminal
-difference follows any discontinuous line where \(\rho \neq 0\).
-
-In particular, neglecting relevant traits, such as judges' biases, can
-cause dimensional mismatches in the BTL model, artificially inflating
-the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
-introducing bias into the trait's estimates \citep{Ackerman_1989}.
-Excluding judges based on misfit statistics risks discarding valuable
-information, which may further bias the trait's estimates
-\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Finally, ignoring
-hierarchical structures may reduce the precision of model parameter
-estimates, potentially overestimating the trait's reliability
-\citep[pp.~482]{Hoyle_et_al_2023}.
+reverse. They begin by observing the outcomes and use the BTL model to
+estimate the stimuli' discriminal difference and discriminal processes.
+Given that the BTL model assumes independent discriminal processes
+across comparisons, if this assumption holds, the model estimates a
+discriminal difference distribution that accurately reflects the
+\emph{true} discriminal difference between the texts. This scenario is
+illustrated in Figure~\ref{fig-correlation} when the discriminal
+difference distribution of the model aligns with the \emph{true}
+distribution, represented by the thick continuous line corresponding to
+\(\rho=0\). Once more, the accuracy of the discriminal difference
+ensures reliable estimates for the discriminal processes of the texts.
+
+However, erroneously assuming zero correlation between stimuli can lead
+to significant statistical and measurement issues. Specifically,
+neglecting judges' biases or hierarchical structures can create
+dimensional mismatches in the model, leading to the over- or
+underestimation of trait reliability \citep[pp.~341,
+482]{Hoyle_et_al_2023}. In other words, the outcome may fail to
+accurately reflect the \emph{true} discriminal differences between
+stimuli. These inaccuracies can result in spurious conclusions about
+these differences \citep[pp.~370]{McElreath_2020} and, by extension, the
+underlying discriminal processes \citep{Ackerman_1989}. This issue is
+illustrated in Figure~\ref{fig-correlation} when the model discriminal
+difference distribution follows the thick continuous line
+\((\rho = 0)\), while the \emph{true} discriminal difference follows any
+discontinuous line where \((\rho \neq 0)\). Conversely, excluding judges
+based on misfit statistics risks discarding valuable information,
+potentially biasing the trait estimates
+\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Again, the direction
+and magnitude of these biases are often unpredictable, as they depend on
+which judges are excluded from the analysis.
 
@@
 Building on the previous section, it is clear that, despite its
-limitations, the BTL model is commonly used as a measurement model in CJ
-assessments. A measurement model specifies how manifest variables
-contribute to the estimation of latent variables
-\citep{Everitt_et_al_2010}. For example, when evaluating writing
-quality, researchers use the BTL model to process the dichotomous
-outcomes resulting from the pairwise comparisons (the manifest
-variables) to estimate scores that reflect the underlying level of
-writing quality (the latent variable)
+limitations, the BTL model commonly serves as the measurement model for
+CJ assessments. Specifically, the model provides a method for
+transforming manifest into latent variables (psychological scales) and
+allocating the compared stimuli on this continuum
+\citetext{\citealp[pp.~269]{Thurstone_1927b}; \citealp{Everitt_et_al_2010}}.
+For example, when evaluating writing quality, researchers use the BTL
+model to process the dichotomous outcomes resulting from the pairwise
+comparisons (the manifest variables) to estimate scores that reflect the
+underlying level of writing quality (the latent variable)
 \citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
 
-Researchers then typically use these estimated BTL scores, or their
+Researchers then typically use the BTL scores' point estimates, or their
 transformations, to conduct additional analyses or hypothesis tests. For
@@
 
-In aggregate, the excessive reliance on Thurstone's Case V assumptions
-in the statistical analysis of comparative data can compromise the
-reliability of the trait estimates. This overreliance may also undermine
-their validity \citep[pp.~2]{Perron_et_al_2015}, especially when
-combined with the disconnect between CJ's approach to trait measurement
+In aggregate, excessive reliance on Thurstone's Case V assumptions in
+the statistical analysis of comparative data can compromise the
+reliability of trait estimates. This overreliance may also undermine
+their validity \citep[pp.~2]{Perron_et_al_2015}, particularly when
+coupled with the disconnect between CJ's approach to trait measurement
 and hypothesis testing. However, the structural approach to causal
-inference can address these issues by providing a systematic and
-integrated framework that enhances statistical accuracy while
-strengthening measurement reliability and validity.
-
-\section{Extending Thurstone's theory}\label{sec-theoretical}
+inference can address these issues by offering a systematic and
+integrated framework that strengthens measurement reliability and
+validity while enhancing the statistical accuracy of hypothesis tests.
+
+\section{Extending Thurstone's general form}\label{sec-theoretical}
 
@@
 formally and graphically represent the assumed causal structure of a
-system, such as the one found in CJ experiments.
-
-Essentially, SCMs and DAGs function as \emph{conceptual models} on which
-identification analysis rests \citep[pp.~4]{Schuessler_et_al_2023}.
+system, such as the one found in CJ experiments. Essentially, SCMs and
+DAGs function as \emph{conceptual models} on which identification
+analysis rests \citep[pp.~4]{Schuessler_et_al_2023}.
 \emph{Identification analysis} helps researchers to determine whether an
@@
 
-A motivating example helps to clarify these concepts. This example will
-appear throughout the document to illustrate various aspects of the
-proposed extension. In this example, researchers aim to determine: ``To
-what extent do different teaching methods influence students' ability to
+A motivating example that will appear throughout the document clarifies
+these concepts. In this example, researchers aim to determine: ``To what
+extent do different teaching methods influence students' ability to
 produce high-quality written texts?'' To investigate this, researcher
@@
 hinder the model's ability to identify various estimands relevant to CJ
-inquiries, including the one described in the motivating example.
-Identification is crucial because it is a necessary condition for
-ensuring consistent estimators. \emph{Consistency} refers to the
-property of an estimator whose estimates converge to the ``true'' value
-of the estimand as the data size approaches infinity
-\citep{Everitt_et_al_2010}. Without identification, consistency cannot
-be achieved, even with ``infinite'' and error-free data. Thus, deriving
-meaningful insights from finite data becomes impossible
-\citep[pp.~5]{Schuessler_et_al_2023}.
+inquiries, including the one described in the example. Identification is
+crucial because it is a necessary condition for ensuring consistent
+estimators. \emph{Consistency} refers to the property of an estimator
+whose estimates converge to the ``true'' value of the estimand as the
+data size approaches infinity \citep{Everitt_et_al_2010}. Without
+identification, consistency cannot be achieved, even with ``infinite''
+and error-free data. Thus, deriving meaningful insights from finite data
+becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}.
 
@@
 Thus, this section addresses the issues identified in
-Section~\ref{sec-theory-issues} by extending Thurstone's theory using
-the structural approach to causal inference. Specifically, it leverages
-the capabilities of this approach to formalize the combination of the
-core theoretical principles outlined in
-Section~\ref{sec-thurstone_theory} with key assessment design features
-relevant to CJ experiments, such as the selection of judges, stimuli,
-and comparisons. In addition to enhancing statistical accuracy and
-strengthening measurement reliability and validity, the approach offers
-two key advantages. First, it clarifies the interactions among all
-actors and processes involved in CJ experiments. Second, it shifts the
-current comparative data analysis paradigm from passively accepting the
-model assumptions to actively testing whether those assumptions fit the
-data under analysis.
+Section~\ref{sec-theory-issues} by extending Thurstone's general form
+using the structural approach to causal inference. Specifically, it
+formalizes the combination of the core theoretical principles outlined
+in Section~\ref{sec-thurstone_theory} with key assessment design
+features relevant to CJ experiments, such as the selection of judges,
+stimuli, and comparisons. In addition to enhancing statistical accuracy
+and strengthening measurement reliability and validity, the approach
+offers two key advantages. First, it clarifies the interactions among
+all actors and processes involved in CJ experiments. Second, it shifts
+the current comparative data analysis paradigm from passively accepting
+the model assumptions to actively testing whether those assumptions fit
+the data under analysis.
 
@@
 theoretical principles into what we call the \emph{conceptual-population
-model}. This model assumes that researchers have access to comparison
-data from a conceptual population, that is, data representing all
+model}. This model assumes that researchers have access to a
+\emph{census} of comparative data, that is, data representing all
 repeated judgments made by every available judge for each pair of
@@
 
+\subsubsection{Integrating initial theoretical
+principles}\label{sec-theory-theoretical_P1}
+
 Before incorporating the theoretical principles into the
-\emph{conceptual-population model}, it is essential to define SCMs. SCMs
-are formal mathematical models characterized by a set of
+\emph{conceptual-population model}, it is essential to further define
+SCMs. SCMs are formal mathematical models characterized by a set of
 \emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
@@
 
-SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
-theoretical principles into the conceptual-population model of the
-motivating example. This SCM maps the relationship between the
+SCM~\ref{fig-cj03_scm} presents the initial integration of theoretical
+principles into the conceptual-population model for the example, applied
+to examine the effect of different teaching methods on students' writing
+ability. This SCM maps the relationship between the
 conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
@@
 difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
-Additionally, based on the arguments in Section~\ref{sec-theory-issue1b}
-and the recommendations of \citet{Andrich_1978} and
-\citet{Wainer_et_al_1978}, the SCM incorporates the judges' biases
-\((B_{kj})\). Together with the outcome, these variables form the
-preliminary set of endogenous variables,
+Additionally, based on the arguments developed under the first scenario
+of Section~\ref{sec-theory-issue1b}, and the recommendations of
+\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
+the judges' biases \((B_{kj})\). Together with the outcome, these
+variables form the preliminary set of endogenous variables,
 \(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
@@
 SCM~\ref{fig-cj03_scm}, illustrating the causal relationships among
-variables based on Thurstone's theory and the practical aspects of CJ
-experiments. Particularly, the graph shows how the texts' discriminal
-processes \((T_{ha}, T_{ib})\) influence their discriminal difference
-\((D_{hiabjk})\), which in turn determines the outcome. It also depicts
-the judges' biases \((B_{kj})\) as an additional factor influencing the
-texts' discriminal difference. Furthermore, the DAG distinguish between
-observed endogenous variables, such as the outcome, and latent
-endogenous variables, which include the texts' discriminal processes,
-their discriminal difference, and the judges' biases.
-
-Although specifying a system's data structure is not mandatory for using
-SCMs and DAGs, defining one can sometimes enhance clarity and facilitate
-the description of the system. Thus, to simplify the description of a CJ
-experiment, we first define the vectors \(I\) and \(J\), along with the
-matrices \(IA\) and \(JK\), as follows:
+variables based on Thurstone's theory. Particularly, the graph shows how
+the texts' discriminal processes \((T_{ha}, T_{ib})\) influence their
+discriminal difference \((D_{hiabjk})\), which in turn determines the
+outcome \((O^{cp}_{hiabjk})\). It also depicts the judges' biases
+\((B_{kj})\) as an additional factor influencing the texts' discriminal
+difference. Furthermore, the DAG distinguish between observed endogenous
+variables, such as the outcome, and latent endogenous variables, which
+include the texts' discriminal processes, their discriminal difference,
+and the judges' biases.
+
+\subsubsection{The comparative census data
+structure}\label{sec-theory-theoretical_P2}
+
+Although specifying a system's data structure is not mandatory when
+using SCMs and DAGs, defining one can improve clarity and facilitate the
+description of the system. Thus, to re-express the scalar form of the CJ
+system, outlined in the previous section, into an simpler equivalent
+vectorized form, we first define the vectors \(I\) and \(J\), along with
+the matrices \(IA\) and \(JK\), as follows:
 
@@
 
-Similarly, an element of \(J\) corresponds to a unique judge \(j\).
-Moreover, each row of \(JK\) specifies a unique pairing of a judge \(j\)
-with a judgment number \(k\). In this case, \(n_{J}\) indicates the
-total number of judges, and \(n_{K}\) represents the number of judgments
-each judge performs. As a result, vector \(J\) has a length of
-\(n_{J}\), and matrix \(JK\) has a dimension of \((n_{J} \cdot n_{K})\)
-rows and \(2\) columns. Like the previous structures, the vector and the
-matrix consist of scalars from the set of natural numbers
-\(\mathbb{N}\).
+Similarly, an element of the vector \(J\) corresponds to a unique judge
+\(j\). Moreover, each row of \(JK\) specifies a unique pairing of a
+judge \(j\) with a judgment number \(k\). In this case, \(n_{J}\)
+indicates the total number of judges, and \(n_{K}\) represents the
+number of judgments each judge performs. As a result, vector \(J\) has a
+length of \(n_{J}\), and matrix \(JK\) has a dimension of
+\((n_{J} \cdot n_{K})\) rows and \(2\) columns. Like the previous
+structures, the vector and the matrix consist of scalars from the set of
+natural numbers \(\mathbb{N}\).
 
@@
 n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
-\end{bmatrix} \in \mathbb{N}
+\end{bmatrix}
 }\end{equation}
@@
 conceptual-population model's assumption that researchers have access to
-data capturing all repeated judgments made by every available judge for
-each pair of stimuli produced by each pair of individuals in the
-population. Here, the term \({n_{I} \cdot n_{A} \choose 2}\) represents
-the Binomial coefficient, which quantifies the total number of
-individual-stimuli comparison pairs within the conceptual population.
-The coefficient is defined as follows:
+a \emph{census} of comparative data, capturing all repeated judgments
+made by every available judge for each pair of stimuli produced by each
+pair of individuals in the population. Here, the term
+\({n_{I} \cdot n_{A} \choose 2}\) represents the Binomial coefficient,
+which quantifies the total number of individual-stimuli comparison pairs
+within the conceptual population. The coefficient is defined as follows:
 
@@
 
-Readers can better visualize the structure of these vectors and matrices
-by considering small values for the parameters corresponding to number
-of individuals, the number of available stimuli per individual, the
-total number of judges, and the number of judgments each judge performs.
-For instance, assuming \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and
-\(n_{K} = 3\), the matrices described in equations (\ref{eq-mat12}) and
-(\ref{eq-mat3}) take the following form:
+We visualize the structure of these vectors and matrices by considering
+small values for the parameters corresponding to number of individuals,
+the number of available stimuli per individual, the total number of
+judges, and the number of judgments each judge performs. For instance,
+assuming \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
+the matrices described in equations (\ref{eq-mat12}) and (\ref{eq-mat3})
+take the following form:
 
@@
 (\ref{eq-mat12}) and (\ref{eq-mat3}), we can re-express
-SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
-vectorized form, as shown in Figure~\ref{fig-cj04}. In this depiction,
-the outcome \(O^{cp}_{RPUV}\), the texts' discriminal difference
-\(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the judges'
-biases \(B_{JK}\) are represented as vectors rather than scalar values.
-These vectors encompass all values from the conceptual population.
-Specifically, \(O^{cp}_{RPUV}\) is an observed vector of length
-\({n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
+SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in a simpler
+equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
+depiction, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
+difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
+judges' biases \(B_{JK}\) are represented as vectors rather than scalar
+values. These vectors encompass all values from the conceptual
+population. Specifically, \(O^{cp}_{RPUV}\) is an observed vector of
+length \({n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
 \(D_{RPUV}\) is a latent vector of the same dimension. Additionally,
@@
 
-Moreover, building on the arguments developed in
-Section~\ref{sec-theory-issue2} and the principles of Structural
-Equation Modeling (SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item
-Response Theory (IRT)
+\subsubsection{Integrating hierarchical structural
+components}\label{sec-theory-theoretical_P3}
+
+Building on the principles of Structural Equation Modeling (SEM)
+\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
 \citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
-the conceptual-population model includes two hierarchical structural
-components. Each component defines how different observed or latent
-variables influence the primary latent variable of interest
-\citep{Everitt_et_al_2010}. This addition facilitates analyses
-integrating the trait measurement and hypothesis testing processes,
-simultaneously accounting for the traits and their associated
-uncertainties, thereby reducing bias and enhancing measurement
-precision.
+the conceptual-population model incorporates two \emph{hierarchical
+structural components}. Each component describes how different observed
+or latent variables influence the primary latent variable of interest
+\citep{Everitt_et_al_2010}. This addition addresses the problems arising
+when an additional hierarchical structure is relevant to the stimuli, as
+outlined in the second scenario of Section~\ref{sec-theory-issue1b}.
+Furthermore, it facilitates analyses that integrate both trait
+measurement and hypothesis testing processes, thereby accounting for
+both traits and their associated uncertainties, reducing bias and
+improving measurement precision (see Section~\ref{sec-theory-issue2}).
 
 Figure~\ref{fig-cj09} displays the two hierarchical structural
-components for the motivating example. Specifically, the top branch of
-DAG~\ref{fig-cj09_dag} depicts the first component. This component
-describes how the students' writing-quality trait \((T_{I})\) and a set
-of \emph{relevant} text-related variables \((X_{IA})\) causally
-influence the texts' written-quality trait \((T_{IA})\). The DAG also
-highlights the idiosyncratic errors \((e_{IA})\), which capture
-variations in the texts' trait that neither the students' trait nor
-other relevant text-related variables can explain. Here, \(T_{I}\) and
-\(e_{IA}\) are latent vectors of length \(n_{I}\) and
-\((n_{I} \cdot n_{A})\), respectively. Additionally, \(X_{IA}\) is an
-observed matrix with a dimension of \((n_{I} \cdot n_{A})\) rows and
-\(q_{IA}\) columns.
+components for the example, applied to examine the effect of different
+teaching methods on students' writing ability. Specifically, the top
+branch of DAG~\ref{fig-cj09_dag} depicts the first component. This
+component describes how the students' writing-quality trait \((T_{I})\)
+and a set of \emph{relevant} text-related variables \((X_{IA})\), such
+as text length, causally influence the texts' written-quality trait
+\((T_{IA})\). The DAG also highlights the idiosyncratic errors
+\((e_{IA})\), which capture variations in the texts' trait that neither
+the students' trait nor other relevant text-related variables can
+explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors of length
+\(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively. Additionally,
+\(X_{IA}\) is an observed matrix with a dimension of
+\((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns.
 
 Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
-set of \emph{relevant} student-related variables \((X_{I})\) causally
-influence the students' writing-quality trait \((T_{I})\). The DAG also
-includes the idiosyncratic errors \((e_{I})\), which capture variations
-in the students' traits that the relevant student-related variables
-cannot explain. Here, \(e_{I}\) is a latent vector of length \(n_{I}\),
-and \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
-columns.
+set of \emph{relevant} student-related variables \((X_{I})\), including
+teaching methods, causally influence the students' writing-quality trait
+\((T_{I})\). The DAG also includes the idiosyncratic errors \((e_{I})\),
+which capture variations in the students' traits that the relevant
+student-related variables cannot explain. Here, \(e_{I}\) is a latent
+vector of length \(n_{I}\), and \(X_{I}\) is an observed matrix with
+\(n_{I}\) rows and \(q_{I}\) columns.
 
@@
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
-  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
-  e_{JK} & \:\bot\:e_{J} 
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
 \end{aligned}
@@
 \((B_{J})\) and a set of \emph{relevant} judgment-related variables
-\((Z_{JK})\) causally influence the biases associated with each text
-\(B_{JK}\). The DAG also highlights the idiosyncratic errors
-\((e_{JK})\), which capture variations in the texts' bias that neither
-the judges' bias nor other relevant judgment-related variables can
-explain. In this description, \(B_{J}\) and \(e_{JK}\) are latent
-vectors of length \(n_{J}\) and \((n_{J} \cdot n_{K})\), respectively.
-Additionally, \(Z_{JK}\) is an observed matrix with a dimension of
-\((n_{J} \cdot n_{K})\) rows and \(q_{JK}\) columns.
+\((Z_{JK})\), such as number of judgments or time for judgment, causally
+influence the biases associated with each text \(B_{JK}\). The DAG also
+highlights the idiosyncratic errors \((e_{JK})\), which capture
+variations in the texts' bias that neither the judges' bias nor other
+relevant judgment-related variables can explain. In this description,
+\(B_{J}\) and \(e_{JK}\) are latent vectors of length \(n_{J}\) and
+\((n_{J} \cdot n_{K})\), respectively. Additionally, \(Z_{JK}\) is an
+observed matrix with a dimension of \((n_{J} \cdot n_{K})\) rows and
+\(q_{JK}\) columns.
 
 Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
-set of \emph{relevant} judge-related variables \((Z_{J})\) causally
-influence the judges' bias \((B_{J})\). The DAG also accounts for
-idiosyncratic errors \((e_{J})\), capturing the variations in the
-judges' bias that relevant judge-related variables cannot explain. Here,
-\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
-observed matrix with \(n_{J}\) rows and \(q_{J}\) columns.
+set of \emph{relevant} judge-related variables \((Z_{J})\), including
+judgment expertise, causally influence the judges' bias \((B_{J})\). The
+DAG also accounts for idiosyncratic errors \((e_{J})\), capturing the
+variations in the judges' bias that relevant judge-related variables
+cannot explain. Here, \(e_{J}\) is a latent vector of length \(n_{J}\),
+and \(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
+columns.
 
@@
 
-Overall, the conceptual-population model extends Thurstone's theory by
-introducing key innovations to address the limitations discussed in
-Section~\ref{sec-theory-issues}. These enhancements include accounting
-for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and
-precision, the model still depends on the unrealistic assumption that
-researchers have access to data encompassing all repeated judgments from
-every judge for each pair of stimuli produced by every individual in the
-population. Since researchers rarely meet this assumption in practice,
-they must consider a more realistic scenario. The extensions to the
+Overall, the conceptual-population model extends Thurstone's general
+form by introducing key innovations to address the limitations discussed
+in Section~\ref{sec-theory-issues}. These enhancements include
+accounting for judges' biases and integrating hierarchical structural
+components. Nevertheless, despite its promise of enhancing measurement
+accuracy and precision, the model still depends on the unrealistic
+assumption that researchers have access to a \emph{census} of
+comparative data, encompassing all repeated judgments from every judge
+for each pair of stimuli produced by every individual in the population.
+Since researchers rarely meet this assumption in practice, they must
+consider a more realistic scenario. The extensions to the
 conceptual-population model, derived from this realistic scenario, are
@@
 
-In the sample-comparison model, researchers have access to a limited
-number of judgments (\(n^{s}{K}\)) from a sample of judges
-(\(n^{s}{J}\)), along with a specific number of texts (\(n^{s}{A}\))
-from a sample of students (\(n^{s}{I}\)) drawn from the conceptual
-population.
-
-Figure~\ref{fig-cj12} illustrates the sample-comparison model,
-incorporating the sample mechanism. Following the convention of
-\citet[pp.~499-516]{McElreath_2020}, the DAG~\ref{fig-cj12_dag} shows
-several conditioned variables \((S_{R}\), \(S_{P}\), \(S_{U}\), and
-\(S_{V})\) that causally influence the sample-comparison outcome
-\(O^{sc}_{RPUV}\), reflecting how the outcome is a sample drawn from the
-conceptual population. The DAG also depicts the conceptual-population
-outcome \(O^{cp}_{RPUV}\) is now an unobserved variable, emphasizing the
-researchers' inability to access such an outcome.
-
-Concatenating the matrices
+\subsubsection{The sample mechanism}\label{sec-theory-theoretical_SC1}
+
+In the sample-comparison model, we assume a more realistic scenario
+compared to the conceptual-population model. Specifically, we assume
+that researchers have access to \emph{all comparisons} from a limited
+number of judgments \((n^{s}_{K})\) from a sample of judges
+\((n^{s}_{J})\), along with a specific number of texts \((n^{s}_{A})\)
+from a sample of students \((n^{s}_{I})\) drawn from the conceptual
+population. Figure~\ref{fig-cj12} illustrates this model, incorporating
+only this sample mechanism.
+
+Following the convention of \citet[pp.~499-516]{McElreath_2020}, the
+DAG~\ref{fig-cj12_dag} depicts the conceptual-population outcome
+\(O^{cp}_{RPUV}\) as an unobserved variable, emphasizing the
+researchers' inability to access such an outcome due to the sampling
+mechanism. Moreover, the DAG shows the conditioned \emph{sample design}
+variable \(S\), which causally influence the sample-comparison outcome
+\(O^{sc}_{RPUV}\). This variable reflects how this sample-comparison
+outcome is a sample drawn from the conceptual-population outcome
+\(O^{cp}_{RPUV}\). Specifically, the variable is a vector
+\(S = \{ s_{1}, \dots, s_{n} \}\), where \(n\) corresponds to the number
+of rows in the matrix \(RPUV\) (refer to
+Section~\ref{sec-theory-theoretical_P2}), and each \(s_{i}\) is a binary
+variable indicating the presence \((1)\) or absence \((0)\) of data in
+the corresponding positions of the conceptual-population outcome
+\(O^{cp}_{RPUV}\). Specifically,
+
+\[
+s_i = \begin{cases} 
+1 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is observed} \\
+0 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is missing}
+\end{cases}
+\]
 
@@
 
-Considering comparison mechanisms
+Notice,
+
+\subsubsection{The comparison
+mechanism}\label{sec-theory-theoretical_SC2}
+
+Conversely, we make the scenario depicted in
+Section~\ref{sec-theory-theoretical_SC1} more realistic by integrating
+the comparison mechanism, that is, we assume that researchers do not
+have access to \emph{all comparisons} from the sample, but rather a
+sufficient number of comparisons \((n_{C})\) so that a proportion,
+\(P(B>A)\), may be determined for each pair of stimuli
+\citep[pp.~267]{Thurstone_1927b}. Figure~\ref{fig-cj14} illustrates this
+model, incorporating both the sample and comparison mechanisms.
+
+Following the convention of \citet[pp.~499-516]{McElreath_2020}, the
+DAG~\ref{fig-cj14_dag} depicts the sample-comparison outcome
+\(O^{sc}_{RPUV}\) as an unobserved variable, emphasizing the
+researchers' inability to access such an outcome due to the comparison
+mechanism. Moreover, the DAG shows the conditioned \emph{comparison
+design} variable \(C\), which causally influence the observed outcome
+\(O_{RPUV}\). This variable reflects how this observed outcome is a
+sample drawn from the sample-comparison outcome \(O^{sp}_{RPUV}\).
+Specifically, the variable is a vector
+\(C = \{ c_{1}, \dots, c_{n} \}\), where \(n\) corresponds to the number
+of rows in the matrix \(RPUV\) (refer to
+Section~\ref{sec-theory-theoretical_P2}), and each \(c_{i}\) is a binary
+variable indicating the presence \((1)\) or absence \((0)\) of data in
+the corresponding positions of the sample-comparison outcome
+\(O^{sp}_{RPUV}\). Specifically,
+
+\[
+c_i = \begin{cases} 
+1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{RPUV} \text{ is observed} \\
+0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{RPUV} \text{ is missing}
+\end{cases}
+\]
 
@@
 \end{figure}%
+
+Notice,
 
@@
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
-  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
-  e_{JK} & \:\bot\:e_{J}
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK}
 \end{aligned}
@@
 
-\caption{\label{fig-cj15}Sample-comparison model including sample and
-comparison mechanisms}
+\caption{\label{fig-cj15}Sample-comparison model, including sample and
+comparison mechanisms and assuming equal discriminal dispersions for the
+student's traits}
+
+\end{figure}%
+
+\begin{figure}[H]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
+  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S \\
+  O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
+  D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
+  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
+  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
+  e_{I} & \sim \text{Normal}(0,s_{XI}) \\
+  e_{J} & \sim \text{Normal}(0,s_{ZJ}) \\
+  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
+  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\ \\
+  & \textbf{Contraints:} \\
+  & \sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1 ; \quad \sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj16_stat}Statistical model}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
+  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
+  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
+  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & := f_{e}(X_{I}) \\
+  e_{J} & := f_{e}(Z_{J}) \\
+  e_{IA} & \:\bot\:e_{JK}
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj16_scm}SCM}
+
+\end{minipage}%
+
+\caption{\label{fig-cj16}Sample-comparison model, including sample and
+comparison mechanisms and assuming different discriminal dispersions for
+the student's traits}
 

### 2025-02-25 code/CJfig.R

@@ -3,6 +3,9 @@ rm(list=ls())
 
 main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript/images/png/'
 
+require(rethinking)
+require(tidyverse)
+
 
 
 # functions ####
@@ -10,6 +13,28 @@ colorArea <- function(from, to, density, ..., col="lightgray", dens=NULL){
   y_seq = seq(from, to, length.out=500)
   d = c(0, density(y_seq, ...), 0)
   polygon(c(from, y_seq, to), d, col=col, density=dens, border=NA)
+}
+
+sim_me = function( n=100, b=0.3, s1=1, s2=0.8){
+  
+  # # test
+  # n = 100
+  # b = 0.2
+  # s1 = 1
+  # s2 = 0.8
+  
+  # simulation
+  x = rnorm( n=n, mean=0, sd=1 )
+  mu_y = b*x
+  y = rnorm( n=n, mean=mu_y, sd=s1 )
+  y_tilde = rnorm( n=n, mean=y, sd=s2 )
+  
+  # saving coefficients
+  res = c( coef( lm( y ~ x ) )[2], coef( lm( y_tilde ~ x ) )[2] )
+  
+  # return object
+  return(res)
+  
 }
 
 
@@ -97,7 +122,7 @@ lines( x=c(dd[1], dd[1]+2*dd[2]), y=rep( pplotD[3], 2), lty=2, lwd=0.7 )
 text( x=dd[1]+1, y=pplotD[3]-0.02, cex=0.8, expression( sigma[BA] ) ) 
 text( x=dd[1]+1.7, y=0.25, cex=0.8, 'where:' )
 text( x=dd[1]+2.4, y=0.22, cex=0.8,
-      expression( sigma[BA] == sqrt( sigma[B]^2 + sigma[A]^2 - rho*sigma[B]*sigma[A]) ) )
+      expression( sigma[BA] == sqrt( sigma[B]^2 + sigma[A]^2 - 2*rho*sigma[B]*sigma[A]) ) )
 
 dev.off()
 
@@ -241,5 +266,76 @@ text( x=dd[[1]]+1.7, y=0.26, cex=0.8, 'where:' )
 text( x=dd[[1]]+2.45, y=0.22, cex=0.8,
[…]

### 2025-03-20 paper2.qmd
section:all; process:write,figures,code

@@
 abstract: |
-  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches.
-Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data.
-Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation.
-The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data.
-The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences.
-This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
+  This study revisits Thurstone's law of comparative judgment (CJ), focusing on two prominent issues of traditional approaches.
+First, it critiques the heavy reliance on Thurstone's Case V assumptions and, by extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
+Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli.
+While these assumptions simplify the trait measurement process, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
+Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes.
+Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.
+
+  To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods.
+This extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.
+
+  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, particularly stressing the need for bespoke CJ models tailored to the specific assumptions of the experiments and data.
+It also lays the foundation for broader applications, encouraging researchers across the social sciences to adopt more robust and interpretable methodologies.
 key-points:
@@
 In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021].
-Meanwhile, research on the validity of CJ scores indicates their capacity to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
-Moreover, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
+Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
 
@@
 Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
-This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates.
-The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.
-This study addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on Causal and Bayesian inference methods.
+This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates  [@Perron_et_al_2015, pp.
+2].
+First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data.
+Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli.
+While these assumptions simplify statistical modeling, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
+Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes.
+Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.
 
 <!-- 4. what is the organization of the study? -->
+To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods.
 As a result, the study divides its content into six main sections.
@@
 The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-@sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data.
-@sec-discussion discusses the findings, limitations, challenges and explores avenues for future research.
-Finally, @sec-conclusion summarizes the study's key insights.
+@sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data.
+@sec-discussion discusses the findings, limitations, and challenges and explores avenues for future research.
+Finally, @sec-conclusion summarizes the study's conclusions.
 
@@
 251].
-This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons.
+This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum.
 Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome.
@@
 The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
-This conclusion is supported by two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
+Two key observations support this conclusion: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
 As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
 
-# Two prominent issues in CJ practice {#sec-theory-issues}
+# Two Prominent Issues in Traditional CJ Practice {#sec-theory-issues}
 
@@
 For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases).
-However, a single judge could provide only ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.
+However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.
 
@@
 <!-- 3. But are our specific problems? -->
-It is evident that although Thurstone developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
-269] and cautioned that its use "should not be made without experimental test" [@Thurstone_1927b, pp.
-270], the CJ literature has nevertheless relied the most on it in practice.
-This preference is particularly notable given that Case V incorporates the most extensive set of simplifying assumptions among all the cases [@Bramley_2008, pp.
+However, Thurstone designed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+269], prioritizing statistical simplicity over precise trait measurement.
+He cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+270] because it imposes the most extensive set of simplifying assumptions among all the proposed cases [@Bramley_2008, pp.
 253; @Kelly_et_al_2022, pp.
 677] (see @tbl-thurstone_cases).
+Therefore, it is surprising that, despite these limitations, CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp.
+677].
+
+<!-- Adding assumptions simplifies the formula and improves its tractability, but one cannot place as much confidence in the outputs of the formula if those assumptions are not met [@Kelly_et_al_2022, pp.
+677].
+-->
 
 Furthermore, since Thurstone's primary goal was to define a psychological scale and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp.
-269], he did not provide guidance on how to use these measurements when, for instance, researchers aim to test hypotheses related to the measured traits.
-Although the CJ practice has circumvented this issue by using scores' point estimates or their transformations for such analyses, a fundamental question remains: Is this method appropriate for addressing these inquiries?
+269], he did not provide specific guidance on how to use the measurement scores from Case V to conduct hypothesis tests related to the measured traits.
+Although the CJ practice has circumvented this issue by using the point estimates of the scores or their transformations for such analyses, a fundamental question remains: Is this method appropriate for addressing these inquiries?
 
 <!-- 4. Section division -->
-Thus, this section discusses these two prominent issues, as they could undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015, pp.
-2].
-Specifically, @sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-Moreover, @sec-theory-issue2 focuses on the apparent disconnect between the approaches to trait measurement and hypothesis testing in CJ.
-
-## The overreliance on Case V in the statistical analysis of CJ data {#sec-theory-issue1}
+Thus, this section discusses these two prominent issues.
+Specifically, @sec-theory-issue1 examines the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
+Conversely, @sec-theory-issue2 focuses on the apparent disconnect between the approaches to trait measurement and hypothesis testing in CJ.
+
+## The Case V and the statistical analysis of CJ data {#sec-theory-issue1}
 
@@
 <!-- 1. The Case V and its simpler statistical representation: the BTL model -->
-As previously discussed, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp.
-253; @Kelly_et_al_2022, pp.
-677], Case V remains the most widely used case in the CJ literature.
-This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959].
-The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
+Despite its limitations, Case V remains the most widely used model in CJ literature.
+This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case.
+The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
 254] (see @tbl-thurstone_cases).
-As noted by @vanderLinden_et_al_2017_I [pp.
-16], this substitution has little impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation.
-Furthermore, this minimal impact is reinforced by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.
+However, as @vanderLinden_et_al_2017_I [pp.
+16] noted, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation.
+Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.
 
@@
 <!-- 2. But what are limitations of Case V (and the BTL model)? -->
-However, as mentioned earlier, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
-269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp.
-677].
-He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270].
-Furthermore, he acknowledged that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+However, Thurstone acknowledged that some assumptions of Case V could be problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376].
-Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice, namely the assumption of equal dispersion and zero correlation between stimuli.
-
-<!-- Adding assumptions simplifies the formula and improves its tractability, but one cannot place as much confidence in the outputs of the formula if those assumptions are not met [@Kelly_et_al_2022, pp.
-677].
--->
+Thus, given that modern CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice.
+These assumptions are the equal dispersion and zero correlation between stimuli.
 
@@
 
-<!-- 1. But does it hold?  -->
-Thurstone argued that the assumption of equal dispersions may not apply when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-Indeed, the CJ literature may already provide evidence of this violation through "misfit" statistics.
-For example, labeling texts as *misfits* indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp.
+<!-- 1. What does the equal dispersion between stimuli imply? -->
+According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, directly influencing the outcome of pairwise comparisons.
+A thought experiment can help illustrate this idea.
+In this experiment, researchers observe the discriminal processes for two texts, similar to those shown in @fig-discriminal_process.
+Additionally, the discriminal dispersion for Text A remains constant, and the texts are uncorrelated  $(\rho=0)$.
+@fig-dispersion demonstrates that an increase in the uncertainty associated with the perception of Text B relative to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference.
+This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B>A)$, ultimately influencing the comparison outcome.
+Additionally, the figure reveals that when the discriminal dispersions of the texts are equal, as in the BTL model $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference is more narrow than when the dispersions differ.
+As a result, the discriminal difference is more likely to favor Text B over Text A, as it is represented by the shaded gray area.
+
+::: {#fig-casev_issues layout-ncol=2}
+
+![Discrepancies in the dispersions of stimuli](/images/png/dispersion.png){#fig-dispersion width=100%}
+
+![Correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}
+
+The effect of dispersion discrepancies and stimuli correlation on the distribution of the discriminal difference.
+:::
+
+<!-- 2. But this is not the scenario we observe -->
+In experimental practice, however, the thought experiment occurs in reverse.
+Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp.
+373].
+Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
+150], in this case, the assumption of equal dispersions.
+For instance, when the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts.
+This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
+The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.
+
+<!-- In experimental practice the procedure is the reverse of this hypothesis because the frequencies are known first experimentally, and from these frequencies we construct the psychological continuum.
+[@Thurstone_1927a, pp.
+373] -->
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
+
+<!-- 3. But does it hold in practice?  -->
+Notably, while the assumption of equal dispersions simplifies the trait measurement process, evidence from the CJ literature suggests that the BTL model may fail to capture the complexity of modern CJ data.
+Specifically, the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+376; @Bramley_2008, pp.
+273; @Kelly_et_al_2022, pp.
+678], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+Indeed, the CJ literature may already provide evidence of such discrepancies, particularly in the form of "misfit" statistics.
+For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp.
 11; @Pollitt_2012b, pp.
 290].
-These discrepancies imply that the discriminal differences for *misfit* texts have broader distributions, further suggesting that their discriminal processes exhibit more dispersion than those of other texts.
-This idea can be illustrated using @fig-discriminal_process, where text B shows a wider discriminal dispersion than text A.
-As a result, the texts' discriminal processes overlap more, making it difficult for a judge to determine which text possesses more of the trait.
-
-A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves [@Pollitt_2012a, pp.
-164-165; @Pollitt_2012b, pp.
-289-290; @vanDaal_et_al_2016, pp.
-4; @Goossens_et_al_2018, pp.
-20].
-These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.
-
-<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
-Thurstone clearly did not expect this to hold for any but the most simple stimuli.
-The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008].
--->
 
@@
 376] -->
+
+<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal.
+Thurstone clearly did not expect this to hold for any but the most simple stimuli.
+The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008, pp.
+273].
+-->
+
+<!-- It also seems unclear whether Thurstone’s work really can be applied to the stimuli typically used in assessment.
+While Thurstone used his Law to create measurement scales for purely psychological or subjective entities, he did so only with stimuli whose values could be evaluated relatively instantaneously, such as short statements and handwriting samples.
+He was also sceptical about its use for very heterogeneous stimuli (Thurstone, 1927b) [@Kelly_et_al_2022, pp.
+678].
+-->
 
@@
 
+<!-- 4. How these judgment discrepancies can happen? -->
+These discrepancies may arise from two possible causes.
+First, the discriminal processes of misfit texts may substantially overlap with those of non-misfit texts.
+Second, misfit texts may represent outlying observations with characteristics that make them difficult to compare and position on the trait scale.
+Regarding the first cause, the overlap can occur when misfit texts share the same trait space and have equal dispersion as non-misfit texts or when they exhibit larger discriminal dispersions.
+Although the BTL model can distinguish the first scenario, it is inherently agnostic to the second.
+@fig-discriminal_process illustrates how an increased discriminal dispersion can widen the overlap between two texts, leading to more judgment discrepancies.
+In the figure, text B has a broader discriminal dispersion than text A, expanding the overlap between their discriminal processes, where judges struggle to determine which text possesses more of the trait.
+
+<!-- A similar challenge arises in the first scenario, where overlapping trait spaces contribute to judgment inconsistencies. -->
+
+<!-- 5. So, what happens when the assumption does not hold? -->
+Thus, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues.
+Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
+This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp.
+370] and, by extension, about the underlying discriminal processes of stimuli.
+@fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
+
+<!-- But ignoring over-dispersion can also lead to all of the same problems as ignoring any predictor variable.
+Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
+[@McElreath_2020, pp.
+370] -->
+
+Finally, regarding the second cause, if researchers recognize that misfit statistics may represent outlying observations, the conventional CJ practice of  excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+12].
+The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
+
+<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
+
 <!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
@@
 
-<!-- 2. What does the equal dispersion between stimuli imply? -->
-According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons.
-A thought experiment can illustrate this idea.
-In this experiment, researchers can observe the discriminal processes for the texts depicted in @fig-discriminal_process.
-Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$.
-@fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B compared to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference.
-This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B>A)$, ultimately influencing the comparison outcome.
-Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ.
-Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area).
-
-::: {#fig-casev_issues layout-ncol=2}
-
-![Discrepancies in the dispersions of stimuli](/images/png/dispersion.png){#fig-dispersion width=100%}
-
-![Correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}
-
-The effect of dispersion discrepancies and stimuli correlation on the distribution of the discriminal difference.
-:::
+### The assumption of zero correlation between stimuli {#sec-theory-issue1b}
+
+<!-- recording: Sven and Tine 24.12.17; time: 00:08:50 - 00:10:30 -->
+
+<!-- 1. What does the zero correlation between stimuli imply? -->
+The correlation $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another.
+As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
+A similar thought experiment, as the one depicted in @sec-theory-issue1a, can illustrate this idea.
+The experiment only assumes that the discriminal dispersions for both texts remain constant.
+@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
+This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
+Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are positively correlated.
+As a result, the discriminal difference is less likely to favor Text B over Text A, as it is represented by the shaded gray area.
+
+<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
+268].
+-->
 
 <!-- 2. But this is not the scenario we observe -->
-In experimental practice, however, the thought experiment occurs in reverse.
-Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp.
-373].
-Consequently, the ability of the outcome to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
-150], particularly the assumption of equal dispersions.
-For example, when researchers correctly assume equal dispersions between texts, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between them.
-This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
-The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.
-
-<!-- In experimental practice the procedure is the reverse of this hypothesis because the frequencies are known first experimentally, and from these frequencies we construct the psychological continuum.
-[@Thurstone_1927a, pp.
-373] -->
-
-<!-- Statements about the real world from observed data require assumptions.
-This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
-The validity of our statements therefore critically hinges on the validity of those assumptions.
-If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
-[@Kohler_et_al_2019, pp.
-150] -->
-
-<!-- 2. So, what happens when the assumption does not hold? -->
-However, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues.
-Specifically, the model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
-This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp.
-370] and, by extension, about the underlying discriminal processes of stimuli.
-@fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
-12].
-The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.
-
-<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
-
-<!-- But ignoring over-dispersion can also lead to all of the same problems as ignoring any predictor variable.
-Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
-[@McElreath_2020, pp.
-370] -->
-
-### The assumption of zero correlation between stimuli {#sec-theory-issue1b}
-
-<!-- recording: Sven and Tine 24.12.17; time: 00:08:50 - 00:10:30 -->
-
-<!-- 1. But does it hold?  -->
-In contrast, Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b, pp.
+As outlined in the previous section, researchers approach this process in reverse.
+They begin by observing the outcomes and using the BTL model to estimate the stimuli' discriminal differences and discriminal processes.
+Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds in the data, the model estimates a discriminal difference distribution that accurately reflects the *true* discriminal difference between the texts.
+Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts.
+
+<!-- It then follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the assumption of zero correlation. -->
+
+<!-- 3. But does it hold?  -->
+Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b, pp.
 268].
 @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
-However, the assumption of zero correlation is not met in practice in at least two scenarios.
-First, when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions, generating biases that resist cancellation.
-Second, when an additional hierarchical structure is relevant to the stimuli.
+However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.
 
@@
 In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These additional relevant characteristics are unlikely to be equally weighted or opposing.
-As a result, they may exert an uneven influence on judges' perceptions, creating biases in their judgments that resist cancellation.
+These additional relevant characteristics, when assessed, are unlikely to be equally weighted or opposing.
+As a result, they may exert an uneven influence on judges' perceptions, creating biases that resist cancellation.
 For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments.
-Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
+Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
 346].
@@
 Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient.
-For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
@@
 
-<!-- 4. What does the zero correlation between stimuli imply? -->
-According to the theory, the correlation $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another.
-As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
-A similar thought experiment, as the one depicted in @sec-theory-issue1a, can illustrate this idea.
-The experiment now assumes that the discriminal dispersions for both texts remain constant.
-@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
-This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
-Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are highly correlated.
-Consequently, the discriminal difference is less likely to favor Text B over Text A (shaded gray area).
-
-<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
-268].
--->
-
-<!-- 5. But this is not the scenario we observe -->
-Again, in experimental practice, researchers approach this process in reverse.
-They begin by observing the outcomes and use the BTL model to estimate the stimuli' discriminal difference and discriminal processes.
-Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, the model estimates a discriminal difference distribution that accurately reflects the *true* discriminal difference between the texts.
-This scenario is illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the *true* distribution, represented by the thick continuous line corresponding to $\rho=0$.
-Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts.
-
-<!-- It then follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the assumption of zero correlation. -->
-
-<!-- 6. But what are the challenges when this assumption is violated? -->
-However, erroneously assuming zero correlation between stimuli can lead to significant statistical and measurement issues.
-Specifically, neglecting judges' biases or hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Hoyle_et_al_2023, pp.
+<!-- 3. But what are the challenges when this assumption is violated? -->
+Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues.
+Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023, pp.
 341, 482].
-In other words, the outcome may fail to accurately reflect the *true* discriminal differences between stimuli.
-These inaccuracies can result in spurious conclusions about these differences [@McElreath_2020, pp.
-370] and, by extension, the underlying discriminal processes  [@Ackerman_1989].
-This issue is illustrated in @fig-correlation when the model discriminal difference distribution follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $(\rho \neq 0)$.
-Conversely, excluding judges based on misfit statistics risks discarding valuable information, potentially biasing the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
-12].
-Again, the direction and magnitude of these biases are often unpredictable, as they depend on which judges are excluded from the analysis.
+These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020, pp.
+370] and, by extension, the underlying discriminal processes of the stimuli.
+This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL model follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $(\rho \neq 0)$.
 
@@
 
+Finally, as discussed in the previous section, removing judges based on misfit statistics risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+12].
+The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis.
+Researchers define *misfit judges* as those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a, pp.
+164-165; @Pollitt_2012b, pp.
+289-290; @vanDaal_et_al_2016, pp.
+4; @Goossens_et_al_2018, pp.
+20].
+
+<!-- These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b. -->
+
 ## The disconnect between trait measurement and hypothesis testing {#sec-theory-issue2}
@@
 
-<!-- 1. What happens when you want to test hypotheses -->
-<!-- Although the CJ practice has found a way to circumvent this 'inconvenience', by using estimated scores or their transformations to conduct additional analyses and tests, a critical question remains: Are these scores an appropriate method for addressing such inquiries?
--->
-
-<!-- 1. The BTL model as a measurement model for estimating latent traits -->
-Building on the previous section, it is clear that, despite its limitations, the BTL model commonly serves as the measurement model for CJ assessments.
-Specifically, the model provides a method for transforming manifest into latent variables (psychological scales) and allocating the compared stimuli on this continuum [@Thurstone_1927b, pp.
+<!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
+Building on the previous section, it is clear that, despite its limitations, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along such trait continuum [@Thurstone_1927b, pp.
 269; @Everitt_et_al_2010].
-For example, when evaluating writing quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying level of writing quality (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
-
-<!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Researchers then typically use the BTL scores' point estimates, or their transformations, to conduct additional analyses or hypothesis tests.
-For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests.
+For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
 <!-- 3. But what are the challenges in using BTL scores for further analysis? -->
-However, the statistical literature advises caution when using estimated scores for additional analyses and tests.
+Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions.
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
@@
 137].
-Finally, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
+Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
+@fig-measurement_error illustrates these issues, showing how neglecting the uncertainty associated with BTL scores can reduce the precision of an effect linked to the predictor.
 
@@
 
+![The effect of outcome uncertainty on the estimation of an effect $\beta$ linked to the predictor.
+Uncertainty increases from left to right.](/images/png/measurement_error.png){#fig-measurement_error width=100%}
+
 <!-- 4. segway for next section -->
-In aggregate, excessive reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates.
+In aggregate, the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates.
 This overreliance may also undermine their validity [@Perron_et_al_2015, pp.
-2], particularly when coupled with the disconnect between CJ's approach to trait measurement and hypothesis testing.
+2], particularly when coupled with the disconnect between the trait measurement and hypothesis testing processes.
 However, the structural approach to causal inference can address these issues by offering a systematic and integrated framework that strengthens measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.
@@
 Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone’s general form using the structural approach to causal inference.
-Specifically, it formalizes the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
 In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
@@
 
+<!-- Develop how you get the vectors S and C and sampling weights when you know the extension of the population. (See https://mc-stan.org/docs/stan-users-guide/poststratification.html ) -->
+
 ### The sample mechanism {#sec-theory-theoretical_SC1}
@@
 @fig-cj12 illustrates this model, incorporating only this sample mechanism.
+
+<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
 
@@
 267] -->
+
+<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
+2, Deffner et al (2022) -->
+
+<!-- Consider algorithm used, as a child of the traits of judges and stimuli.
+It allows to talk about independence assumption between them and the perception of difference.
+Consider algorithm is a categorical variable.
+-->
 
@@
 
+<!-- To build the statistical part, start with the joint probability of observing the sample data P(Y, Z), then partition this joint probability in smaller conditional probabilities derived from the DAG.
+-->
+
+<!-- The difficulty and discrimination parameters for the questions should then be given a diffuse, or ideally a hierarchical prior, which will identify these parameters by scaling and locating relative to the student ability parameters.
+-->
+<!-- (https://mc-stan.org/docs/stan-users-guide/regression.html#priors-for-identification.section) -->
+
 ::: {#fig-cj15 layout-ncol=2}
@@
   O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S \\
-  O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
+  O^{cp}_{RPUV} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
   D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
@@
   B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
-  e_{I} & \sim \text{Normal}(0,1) \\
-  e_{J} & \sim \text{Normal}(0,1) \\
-  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \sim \text{Normal}(0, p_{JK})
+  e_{I} & \overset{iid}{\sim} \text{Normal}(0,1) \\
+  e_{J} & \overset{iid}{\sim} \text{Normal}(0,1) \\
+  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
+  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
 \end{aligned}
@@
   B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
-  e_{I} & \sim \text{Normal}(0,s_{XI}) \\
-  e_{J} & \sim \text{Normal}(0,s_{ZJ}) \\
-  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\ \\
+  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
+  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
+  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
+  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK}) \\ \\
   & \textbf{Contraints:} \\
@@
 
+<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
+Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
+[@Kelly_et_al_2022, pp.
+683] -->
+
 <!-- Such differences may not be detected through analyses of bias and misfit.
@@
 681] -->
+
+<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
 
@@
 
+<!-- Show the logic of assuming an equivalent class for the graph that you are proposing. -->
+
 ## Limitations and further research {#sec-discussion-limitations}
@@
 150] -->
+
+<!-- (do not do this yet!) Develop how you multiply the predictions by the inverse of the sample probability to get the estimates.
+Should I use the probability itself?: Remember that in a random sample the sample probability is n/N, so the the inverse of the sample probability is N/n.
+So, multiplying the estimate by the n * N/n gives you the prediction for the population?, yes.
+-->
+<!-- (See https://timss2023.org/methods/, and pdf in mail, and https://discourse.mc-stan.org/t/survey-weighted-regression/1654 ) -->
+
+<!-- (do not do this yet!) Then to get the differences between groups you need to aggregate the latent variable of interest using the variable of interest without considering any other variable (marginalization) (see https://www.r-bloggers.com/2017/06/sampling-weights-and-multilevel-modeling-in-r/ ) -->
+
+<!-- Add to future research how would be to use the model when you do not know the weights of the sample, but you can use external data to approximate them. -->
+
+<!-- Talk about Bayesian Workflow, and how it would fit the process to arrive to the proposed model. -->
+
+<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
+However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
+
+<!-- When you want to investigate objects, the experiment has an experimental unit: the trial on the objects, but you only have one object per subject.
+Consequently, objects and subjects are confounded.
+-->
 

### 2025-03-20 code/CJfig.R

@@ -5,6 +5,7 @@ main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/p
 
 require(rethinking)
 require(tidyverse)
+require(metafor)
 
 
 
@@ -22,15 +23,20 @@ sim_me = function( n=100, b=0.3, s1=1, s2=0.8){
   # b = 0.2
   # s1 = 1
   # s2 = 0.8
-  
+
   # simulation
   x = rnorm( n=n, mean=0, sd=1 )
   mu_y = b*x
   y = rnorm( n=n, mean=mu_y, sd=s1 )
-  y_tilde = rnorm( n=n, mean=y, sd=s2 )
+  y_star = rnorm( n=n, mean=y, sd=s2 )
+  d = data.frame( obs=1:n, y, y_star, y_sd=s2, x)
+  
+  # regressions
+  reg1 = lm( y ~ x, data=d )
+  reg2 = lm( y_star ~ x, data=d )
   
   # saving coefficients
-  res = c( coef( lm( y ~ x ) )[2], coef( lm( y_tilde ~ x ) )[2] )
+  res = c( coef( reg1 )[2], coef( reg2 )[2] )
   
   # return object
   return(res)
@@ -38,13 +44,16 @@ sim_me = function( n=100, b=0.3, s1=1, s2=0.8){
 }
 
 
-
-# figure 1 ####
-
 # distributional parameters
 dpA = c(0, 0.5) # distribution text A (mean, sd)
 dpB = c(1, 1) # distribution text B (mean, sd)
 rAB = 0 # correlation between discriminal processes
+
+
+
+
+
+# figure 1 ####
 
 # figure
 png( filename=file.path(main_dir, 'discriminal_process.png'), 
@@ -278,30 +287,33 @@ dev.off()
 
 # simulation
 b = 0.2
-s2 = c(0.5, 1, 1.5)
+s2 = c(0.3, 0.6, 1)
 dlist = list()
 for( s in seq_along(s2) ){
   dlist[[s]] = t( replicate( n=2000, sim_me( b=b, s2=s2[s]) ) )
-  dlist[[s]] = data.frame(dlist[[s]])
+  dlist[[s]] = data.frame( dlist[[s]] )
   names(dlist[[s]]) = paste0( c('wo','w'), '_me')
[…]

### 2025-03-20 _tex/paper2.tex
Built site for gh-pages

@@
 \begin{abstract}
-This study revisits Thurstone's law of comparative judgments (CJ) by
-addressing two key limitations in traditional approaches. Firstly, it
-addresses the overreliance on the assumptions of Thurstone's Case V in
-the statistical analysis of CJ data. Secondly, it addresses the apparent
-disconnect between CJ's approach to trait measurement and hypothesis
-testing. We put forward a systematic approach based on causal analysis
-and Bayesian statistical methods, which results in a model that
-facilitates a more comprehensive understanding of the factors
-influencing CJ experiments while offering a robust statistical
-translation. The new model accommodates unequal dispersions and
-correlations between stimuli, enhancing the reliability and validity of
-CJ's trait estimation, thereby ensuring the accurate measurement and
-interpretation of comparative data. The paper highlights the relevance
-of this updated framework for modern empirical research, particularly in
-education and social sciences. This contribution advances current
-research methodologies, providing a robust foundation for future
-applications in diverse fields.
+This study revisits Thurstone's law of comparative judgment (CJ),
+focusing on two prominent issues of traditional approaches. First, it
+critiques the heavy reliance on Thurstone's Case V assumptions and, by
+extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
+Specifically, the study raises concerns about the assumptions of equal
+discriminal dispersions and zero correlation between the stimuli. While
+these assumptions simplify the trait measurement process, they may fail
+to capture the complexity of CJ data, potentially leading to unreliable
+and inaccurate trait estimates. Second, the study highlights the
+apparent disconnect between CJ's trait measurement and hypothesis
+testing processes. Although separating these processes simplifies the
+analysis of CJ data, it may also undermine the reliability of various
+statistical results derived from the measurement.
+
+To address these issues, the study extends Thurstone's general form
+using a systematic and integrated approach based on Causal and Bayesian
+inference methods. This extension integrates core theoretical principles
+alongside key assessment design features relevant to CJ experiments,
+such as the selection of judges, stimuli, and comparisons. It then
+translates these elements into a probabilistic statistical model for
+analyzing dichotomous CJ data, overcoming the rigid assumptions of Case
+V and the BTL model.
+
+Finally, the study emphasizes the relevance of this extension for
+contemporary empirical CJ research, particularly stressing the need for
+bespoke CJ models tailored to the specific assumptions of the
+experiments and data. It also lays the foundation for broader
+applications, encouraging researchers across the social sciences to
+adopt more robust and interpretable methodologies.
 \end{abstract}
@@
 \citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
-Meanwhile, research on the validity of CJ scores indicates their
-capacity to represent accurately the traits under measurement
+Meanwhile, research on validity indicates the capacity of CJ scores to
+represent accurately the traits under measurement
 \citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
-Moreover, research on its practical applicability highlights CJ's
+Lastly, research on its practical applicability highlights CJ's
 versatility across both educational and non-educational contexts
@@
 critical issues unresolved. This study identifies and discusses two
-prominent issues that can undermine the reliability and validity of CJ's
-trait estimates. The first issue arises from the excessive reliance on
-Thurstone's Case V assumptions in the statistical analysis of CJ data.
-The second stems from the apparent disconnect between CJ's approach to
-trait measurement and hypothesis testing. This study addresses these
-issues by extending Thurstone's general form through a systematic and
-integrated approach based on Causal and Bayesian inference methods.
-
-As a result, the study divides its content into six main sections.
-Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
-theory. Section~\ref{sec-theory-issues} discusses the identified issues
-in detail. Section~\ref{sec-theoretical} extends Thurstone's general
-form to address these challenges. The extension integrates core
-theoretical principles alongside key assessment design features relevant
-to CJ experiments, such as the selection of judges, stimuli, and
-comparisons. Section~\ref{sec-statistical} translates these theoretical
-and practical elements into a probabilistic statistical model for the
-analysis of dichotomous pairwise comparison data.
-Section~\ref{sec-discussion} discusses the findings, limitations,
+prominent issues of traditional approaches that can undermine the
+reliability and validity of CJ's trait estimates
+\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
+on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
+extension, the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
+Specifically, the study raises concerns about the assumptions of equal
+discriminal dispersions and zero correlation between the stimuli. While
+these assumptions simplify statistical modeling, they may fail to
+capture the complexity of CJ data, potentially leading to unreliable and
+inaccurate trait estimates. Second, the study highlights the apparent
+disconnect between CJ's trait measurement and hypothesis testing
+processes. Although separating these processes simplifies the analysis
+of CJ data, it may also undermine the reliability of various statistical
+results derived from the measurement.
+
+To address these issues, the study extends Thurstone's general form
+using a systematic and integrated approach based on Causal and Bayesian
+inference methods. As a result, the study divides its content into six
+main sections. Section~\ref{sec-thurstone_theory} provides an overview
+of Thurstone's theory. Section~\ref{sec-theory-issues} discusses the
+identified issues in detail. Section~\ref{sec-theoretical} extends
+Thurstone's general form to address these challenges. The extension
+integrates core theoretical principles alongside key assessment design
+features relevant to CJ experiments, such as the selection of judges,
+stimuli, and comparisons. Section~\ref{sec-statistical} translates these
+theoretical and practical elements into a probabilistic statistical
+model to analyze dichotomous pairwise comparison data.
+Section~\ref{sec-discussion} discusses the findings, limitations, and
 challenges and explores avenues for future research. Finally,
-Section~\ref{sec-conclusion} summarizes the study's key insights.
+Section~\ref{sec-conclusion} summarizes the study's conclusions.
 
@@
 further along the trait continuum as possessing more of the trait
-\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
-between stimuli, rather than their absolute positions on the continuum,
-likely defines the outcome of pairwise comparisons. Indeed, the theory
-assumes that the difference between the underlying discriminal processes
-of the stimuli, referred to as the \emph{discriminal difference},
-determines the observed dichotomous outcome. Furthermore, the theory
-assumes that because the individual discriminal processes form a Normal
-distribution on the continuum, the discriminal difference will also
-conform to a Normal distribution \citep[pp.~452]{Andrich_1978}. In this
-distribution, the mode (mean) represents the relative separation between
-the stimuli, and its dispersion indicates the variability of that
-separation.
+\citep[pp.~251]{Bramley_2008}. This suggests that pairwise comparison
+outcomes depend on the relative distance between stimuli, not their
+absolute positions on the continuum. Indeed, the theory assumes that the
+difference between the underlying discriminal processes of the stimuli,
+referred to as the \emph{discriminal difference}, determines the
+observed dichotomous outcome. Furthermore, the theory assumes that
+because the individual discriminal processes form a Normal distribution
+on the continuum, the discriminal difference will also conform to a
+Normal distribution \citep[pp.~452]{Andrich_1978}. In this distribution,
+the mode (mean) represents the relative separation between the stimuli,
+and its dispersion indicates the variability of that separation.
 
@@
 judge perceives Text B as having significantly higher quality than Text
-A. This conclusion is supported by two key observations: the positive
-difference between their modal discriminal processes
-\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
-difference distinctly favors Text B over Text A, represented by the
-shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
-outcome of this comparison is more likely to favor Text B over Text A.
-
-\section{Two prominent issues in CJ practice}\label{sec-theory-issues}
+A. Two key observations support this conclusion: the positive difference
+between their modal discriminal processes \((T_{B} - T_{A} > 0)\) and
+the probability area where the discriminal difference distinctly favors
+Text B over Text A, represented by the shaded gray area denoted as
+\(P(B > A)\). As a result, the dichotomous outcome of this comparison is
+more likely to favor Text B over Text A.
+
+\section{Two Prominent Issues in Traditional CJ
+Practice}\label{sec-theory-issues}
 
@@
 \(10\) correlations --one per comparison (see
-Table~\ref{tbl-thurstone_cases}). However, a single judge could provide
-only \({5 \choose 2} = 10\) unique comparisons, an insufficient data set
-to estimate the required parameters.
+Table~\ref{tbl-thurstone_cases}). However, a single judge could only
+provide \({5 \choose 2} = 10\) unique comparisons, an insufficient data
+set to estimate the required parameters.
 
@@
 
-It is evident that although Thurstone developed Case V to provide a
-``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b} and
-cautioned that its use ``should not be made without experimental test''
-\citep[pp.~270]{Thurstone_1927b}, the CJ literature has nevertheless
-relied the most on it in practice. This preference is particularly
-notable given that Case V incorporates the most extensive set of
-simplifying assumptions among all the cases
+However, Thurstone designed Case V to provide a ``rather coarse
+scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
+statistical simplicity over precise trait measurement. He cautioned that
+its use ``should not be made without (an) experimental test''
+\citep[pp.~270]{Thurstone_1927b} because it imposes the most extensive
+set of simplifying assumptions among all the proposed cases
 \citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
-(see Table~\ref{tbl-thurstone_cases}).
+(see Table~\ref{tbl-thurstone_cases}). Therefore, it is surprising that,
+despite these limitations, CJ research has predominantly relied on Case
+V to measure different traits, raising significant concerns about the
+reliability and validity of such measurements in contexts where the
+case's assumptions may not hold \citep[pp.~677]{Kelly_et_al_2022}.
 
@@
 continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not provide
-guidance on how to use these measurements when, for instance,
-researchers aim to test hypotheses related to the measured traits.
-Although the CJ practice has circumvented this issue by using scores'
-point estimates or their transformations for such analyses, a
-fundamental question remains: Is this method appropriate for addressing
-these inquiries?
-
-Thus, this section discusses these two prominent issues, as they could
-undermine the reliability and validity of CJ's trait estimates
-\citep[pp.~2]{Perron_et_al_2015}. Specifically,
-Section~\ref{sec-theory-issue1} examines the the excessive reliance on
+specific guidance on how to use the measurement scores from Case V to
+conduct hypothesis tests related to the measured traits. Although the CJ
+practice has circumvented this issue by using the point estimates of the
+scores or their transformations for such analyses, a fundamental
+question remains: Is this method appropriate for addressing these
+inquiries?
+
+Thus, this section discusses these two prominent issues. Specifically,
+Section~\ref{sec-theory-issue1} examines the heavy reliance on
 Thurstone's Case V assumptions in the statistical analysis of CJ data.
-Moreover, Section~\ref{sec-theory-issue2} focuses on the apparent
+Conversely, Section~\ref{sec-theory-issue2} focuses on the apparent
 disconnect between the approaches to trait measurement and hypothesis
@@
 
-\subsection{The overreliance on Case V in the statistical analysis of CJ
+\subsection{The Case V and the statistical analysis of CJ
 data}\label{sec-theory-issue1}
 
-As previously discussed, despite relying on the most extensive set of
-simplifying assumptions
-\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
-Case V remains the most widely used case in the CJ literature. This
-popularity stems mainly from its simplified statistical representation
-in the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
-assumptions of Case V, with one notable distinction: whereas Case V
-assumes a Normal distribution for the stimuli discriminal processes, the
-BTL model uses the more mathematically tractable Logistic distribution
-\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
-Table~\ref{tbl-thurstone_cases}). As noted by
-\citet[pp.~16]{vanderLinden_et_al_2017_I}, this substitution has little
-impact on the model's estimation or interpretation because the
+Despite its limitations, Case V remains the most widely used model in CJ
+literature. This preference primarily stems from the BTL model, which
+provides a simplified statistical representation of the case. The BTL
+model mirrors the assumptions of Case V, with one notable distinction:
+while Case V assumes a Normal distribution for the stimuli' discriminal
+processes, the BTL model uses the more mathematically tractable Logistic
+distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
+Table~\ref{tbl-thurstone_cases}). However, as
+\citet[pp.~16]{vanderLinden_et_al_2017_I} noted, this substitution has
+minimal impact on the model's estimation or interpretation because the
 discriminal process scale is arbitrary up to a non-monotonic
-transformation. Furthermore, this minimal impact is reinforced by the
+transformation. Furthermore, this limited impact is supported by the
 fact that the Normal and Logistic distributions exhibit analogous
@@
 
-However, as mentioned earlier, Thurstone originally developed Case V to
-provide a ``rather coarse scaling'' of traits
-\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
-over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
-He explicitly warned against its untested application, stating that its
-use ``should not be made without (an) experimental test''
-\citep[pp.~270]{Thurstone_1927b}. Furthermore, he acknowledged that some
-assumptions could prove problematic when researchers assess complex
-traits or heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}.
-Consequently, given that modern CJ applications frequently involve such
-traits and stimuli, two main assumptions of Case V and, by extension, of
-the BTL model may not consistently hold in theory or practice, namely
-the assumption of equal dispersion and zero correlation between stimuli.
+However, Thurstone acknowledged that some assumptions of Case V could be
+problematic when researchers assess complex traits or heterogeneous
+stimuli \citep[pp.~376]{Thurstone_1927a}. Thus, given that modern CJ
+applications often involve such traits and stimuli, two key assumptions
+of Case V, and by extension, the BTL model, may not always hold in
+theory or practice. These assumptions are the equal dispersion and zero
+correlation between stimuli.
 
@@
 
-Thurstone argued that the assumption of equal dispersions may not apply
-when researchers assess complex traits or heterogeneous stimuli
-\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
-introduce judgment discrepancies due to their unique characteristics
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
-Indeed, the CJ literature may already provide evidence of this violation
-through ``misfit'' statistics. For example, labeling texts as
-\emph{misfits} indicates that comparisons involving these texts result
-in more judgment discrepancies than those involving other texts
-\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
-These discrepancies imply that the discriminal differences for
-\emph{misfit} texts have broader distributions, further suggesting that
-their discriminal processes exhibit more dispersion than those of other
-texts. This idea can be illustrated using
-Figure~\ref{fig-discriminal_process}, where text B shows a wider
-discriminal dispersion than text A. As a result, the texts' discriminal
-processes overlap more, making it difficult for a judge to determine
-which text possesses more of the trait.
-
-A similar line of reasoning applies to the concept of ``misfit'' judges,
-whose evaluations deviate substantially from the shared consensus due to
-the unique characteristics of the stimuli or the judges themselves
-\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.
-These ``misfit'' judges and their associated deviations can give rise to
-additional statistical and measurement issues, which we discuss in more
-detail in Section~\ref{sec-theory-issue1b}.
-
 According to the theory, discrepancies in the discriminal dispersions of
-stimuli shape the distribution of the discriminal difference, exerting a
-direct influence on the outcome of pairwise comparisons. A thought
-experiment can illustrate this idea. In this experiment, researchers can
-observe the discriminal processes for the texts depicted in
-Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
-assumes that the discriminal dispersion for Text A remains constant and
-that the texts are uncorrelated \((\rho=0)\).
-Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
-associated with the perception of Text B compared to Text A
-\((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
+stimuli shape the distribution of the discriminal difference, directly
+influencing the outcome of pairwise comparisons. A thought experiment
+can help illustrate this idea. In this experiment, researchers observe
+the discriminal processes for two texts, similar to those shown in
+Figure~\ref{fig-discriminal_process}. Additionally, the discriminal
+dispersion for Text A remains constant, and the texts are uncorrelated
+\((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that an increase
+in the uncertainty associated with the perception of Text B relative to
+Text A \((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
 discriminal difference. This broadening affects the probability area
@@
 Additionally, the figure reveals that when the discriminal dispersions
-of the texts are equal \((\sigma_{B} - \sigma_{A}=0)\), the discriminal
-difference is more narrow compared to situations where their dispersions
-differ. Consequently, the discriminal difference is more likely to favor
-Text B over Text A (shaded gray area).
+of the texts are equal, as in the BTL model
+\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference is more
+narrow than when the dispersions differ. As a result, the discriminal
+difference is more likely to favor Text B over Text A, as it is
+represented by the shaded gray area.
 
@@
 their respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
-Consequently, the ability of the outcome to reflect \emph{true}
-differences between stimuli largely depends on the validity of the
-model's assumptions \citep[pp.~150]{Kohler_et_al_2019}, particularly the
-assumption of equal dispersions. For example, when researchers correctly
-assume equal dispersions between texts, the BTL model estimates a
-discriminal difference distribution that accurately represents the
-\emph{true} discriminal difference between them. This scenario is
+Consequently, the outcome's ability to reflect \emph{true} differences
+between stimuli largely depends on the validity of the model's
+assumptions \citep[pp.~150]{Kohler_et_al_2019}, in this case, the
+assumption of equal dispersions. For instance, when the assumption
+accurately captures the complexity of the data, the BTL model estimates
+a discriminal difference distribution that accurately represents the
+\emph{true} discriminal difference between the texts. This scenario is
 illustrated in Figure~\ref{fig-dispersion}, when the model's discriminal
@@
 
-However, erroneously assuming equal dispersions between stimuli can give
+Notably, while the assumption of equal dispersions simplifies the trait
+measurement process, evidence from the CJ literature suggests that the
+BTL model may fail to capture the complexity of modern CJ data.
+Specifically, the assumption of equal dispersions may not hold when
+researchers assess complex traits or heterogeneous stimuli
+\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
+as these traits and stimuli can introduce judgment discrepancies due to
+their unique characteristics
+\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
+Indeed, the CJ literature may already provide evidence of such
+discrepancies, particularly in the form of ``misfit'' statistics. For
+instance, \emph{misfit texts} are those whose comparisons result in more
+judgment discrepancies than those involving other texts
+\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
+
+These discrepancies may arise from two possible causes. First, the
+discriminal processes of misfit texts may substantially overlap with
+those of non-misfit texts. Second, misfit texts may represent outlying
+observations with characteristics that make them difficult to compare
+and position on the trait scale. Regarding the first cause, the overlap
+can occur when misfit texts share the same trait space and have equal
+dispersion as non-misfit texts or when they exhibit larger discriminal
+dispersions. Although the BTL model can distinguish the first scenario,
+it is inherently agnostic to the second.
+Figure~\ref{fig-discriminal_process} illustrates how an increased
+discriminal dispersion can widen the overlap between two texts, leading
+to more judgment discrepancies. In the figure, text B has a broader
+discriminal dispersion than text A, expanding the overlap between their
+discriminal processes, where judges struggle to determine which text
+possesses more of the trait.
+
+Thus, erroneously assuming equal dispersions between stimuli can give
 rise to significant statistical and measurement issues. Specifically,
-the model may overestimate the trait's reliability, that is, the degree
-to which the outcome accurately reflects the \emph{true} discriminal
-differences between stimuli. This overestimation, in turn, results in
-spurious conclusions about these differences
+the BTL model may overestimate the trait's reliability, that is, the
+degree to which the outcome accurately reflects the \emph{true}
+discriminal differences between stimuli. This overestimation, in turn,
+results in spurious conclusions about these differences
 \citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
@@
 difference follows any discontinuous line where
-\(\sigma_{B}-\sigma_{A} \neq 0\). Additionally, if researchers recognize
-that misfit statistics highlight these critical differences in
-dispersions, the conventional CJ practice of excluding stimuli based on
-these statistics
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
-can unintentionally discard valuable information, introducing bias into
-trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
-direction and magnitude of these biases are often unpredictable, as they
-depend on which stimuli are excluded from the analysis.
+\(\sigma_{B}-\sigma_{A} \neq 0\).
+
+Finally, regarding the second cause, if researchers recognize that
+misfit statistics may represent outlying observations, the conventional
+CJ practice of excluding stimuli based on these statistics
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
+may unintentionally discard valuable information, introducing bias into
+the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
+The direction and magnitude of these biases remain unpredictable, as
+they depend on which stimuli researchers exclude from the analysis.
 
@@
 
-In contrast, Thurstone assumed that stimuli were uncorrelated because
-judges' biases, arising from two opposing and equally weighted effects
-occurring during the pairwise comparisons, canceled each other out
+The correlation \(\rho\) measures how much the judges' perception of a
+specific trait in one stimulus depends on their perception of the same
+trait in another. As with the discriminal dispersions, this correlation
+shapes the distribution of the discriminal difference, directly
+impacting the outcomes of pairwise comparisons. A similar thought
+experiment, as the one depicted in Section~\ref{sec-theory-issue1a}, can
+illustrate this idea. The experiment only assumes that the discriminal
+dispersions for both texts remain constant. Figure~\ref{fig-correlation}
+reveals that as the correlation between the texts increases, the
+distribution of their discriminal difference becomes narrower. This
+narrowing affects the area under the curve where the discriminal
+difference distinctly favors Text B over Text A, denoted as
+\(P(B > A)\), thus influencing the comparison outcome. Furthermore, the
+figure shows that when two texts are independent or uncorrelated
+\((\rho=0)\), their discriminal difference is less narrow compared to
+scenarios where the texts are positively correlated. As a result, the
+discriminal difference is less likely to favor Text B over Text A, as it
+is represented by the shaded gray area.
+
+As outlined in the previous section, researchers approach this process
+in reverse. They begin by observing the outcomes and using the BTL model
+to estimate the stimuli' discriminal differences and discriminal
+processes. Given that the BTL model assumes independent discriminal
+processes across comparisons, if this assumption holds in the data, the
+model estimates a discriminal difference distribution that accurately
+reflects the \emph{true} discriminal difference between the texts. Once
+more, the accuracy of the discriminal difference ensures reliable
+estimates for the discriminal processes of the texts.
+
+Thurstone assumed that stimuli were uncorrelated because judges' biases,
+arising from two opposing and equally weighted effects occurring during
+the pairwise comparisons, canceled each other out
 \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
@@
 under the assumption of discriminal processes with additive biases.
-However, the assumption of zero correlation is not met in practice in at
-least two scenarios. First, when intricate aspects of multidimensional,
-complex traits or heterogeneous stimuli influence judges' perceptions,
-generating biases that resist cancellation. Second, when an additional
-hierarchical structure is relevant to the stimuli.
+However, evidence from the CJ literature indicates that the assumption
+of zero correlation does not hold in practice in at least two scenarios:
+when intricate aspects of multidimensional, complex traits or
+heterogeneous stimuli influence judges' perceptions or when additional
+hierarchical structures are relevant to the stimuli.
 
@@
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
-additional relevant characteristics are unlikely to be equally weighted
-or opposing. As a result, they may exert an uneven influence on judges'
-perceptions, creating biases in their judgments that resist
+additional relevant characteristics, when assessed, are unlikely to be
+equally weighted or opposing. As a result, they may exert an uneven
+influence on judges' perceptions, creating biases that resist
 cancellation. For example, this could occur when a judge assessing the
@@
 stimuli becomes an observable outcome only through the judges'
-perceptions, these biases could ultimately introduce dependencies
-between the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While
-direct evidence for this particular scenario is lacking, studies such as
+perceptions, these biases could introduce dependencies between the
+stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While direct
+evidence for this particular scenario is lacking, studies such as
 \citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
@@
 estimates of the BTL scores to conduct subsequent analyses and tests at
-the individual hierarchical level
+the individual level
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 
-According to the theory, the correlation \(\rho\) measures how much the
-judges' perception of a specific trait in one stimulus depends on their
-perception of the same trait in another. As with the discriminal
-dispersions, this correlation shapes the distribution of the discriminal
-difference, directly impacting the outcomes of pairwise comparisons. A
-similar thought experiment, as the one depicted in
-Section~\ref{sec-theory-issue1a}, can illustrate this idea. The
-experiment now assumes that the discriminal dispersions for both texts
-remain constant. Figure~\ref{fig-correlation} reveals that as the
-correlation between the texts increases, the distribution of their
-discriminal difference becomes narrower. This narrowing affects the area
-under the curve where the discriminal difference distinctly favors Text
-B over Text A, denoted as \(P(B > A)\), thus influencing the comparison
-outcome. Furthermore, the figure shows that when two texts are
-independent or uncorrelated \((\rho=0)\), their discriminal difference
-is less narrow compared to scenarios where the texts are highly
-correlated. Consequently, the discriminal difference is less likely to
-favor Text B over Text A (shaded gray area).
-
-Again, in experimental practice, researchers approach this process in
-reverse. They begin by observing the outcomes and use the BTL model to
-estimate the stimuli' discriminal difference and discriminal processes.
-Given that the BTL model assumes independent discriminal processes
-across comparisons, if this assumption holds, the model estimates a
-discriminal difference distribution that accurately reflects the
-\emph{true} discriminal difference between the texts. This scenario is
-illustrated in Figure~\ref{fig-correlation} when the discriminal
-difference distribution of the model aligns with the \emph{true}
-distribution, represented by the thick continuous line corresponding to
-\(\rho=0\). Once more, the accuracy of the discriminal difference
-ensures reliable estimates for the discriminal processes of the texts.
-
-However, erroneously assuming zero correlation between stimuli can lead
-to significant statistical and measurement issues. Specifically,
-neglecting judges' biases or hierarchical structures can create
+Thus, erroneously assuming zero correlation between stimuli can also
+lead to significant statistical and measurement issues. Specifically,
+neglecting judges' biases or relevant hierarchical structures can create
 dimensional mismatches in the model, leading to the over- or
 underestimation of trait reliability \citep[pp.~341,
-482]{Hoyle_et_al_2023}. In other words, the outcome may fail to
-accurately reflect the \emph{true} discriminal differences between
-stimuli. These inaccuracies can result in spurious conclusions about
-these differences \citep[pp.~370]{McElreath_2020} and, by extension, the
-underlying discriminal processes \citep{Ackerman_1989}. This issue is
-illustrated in Figure~\ref{fig-correlation} when the model discriminal
-difference distribution follows the thick continuous line
+482]{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result in
+spurious conclusions about the discriminal differences
+\citep[pp.~370]{McElreath_2020} and, by extension, the underlying
+discriminal processes of the stimuli. This issue is illustrated in
+Figure~\ref{fig-correlation} when the discriminal difference
+distribution of the BTL model follows the thick continuous line
 \((\rho = 0)\), while the \emph{true} discriminal difference follows any
-discontinuous line where \((\rho \neq 0)\). Conversely, excluding judges
-based on misfit statistics risks discarding valuable information,
-potentially biasing the trait estimates
-\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Again, the direction
-and magnitude of these biases are often unpredictable, as they depend on
-which judges are excluded from the analysis.
+discontinuous line where \((\rho \neq 0)\).
+
+Finally, as discussed in the previous section, removing judges based on
+misfit statistics risks discarding valuable information and even
+introduce bias into the trait estimates
+\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
+magnitude of these biases remain unpredictable because they depend on
+which judges researchers exclude from the analysis. Researchers define
+\emph{misfit judges} as those whose evaluations deviate substantially
+from the shared consensus due to the unique characteristics of either
+the stimuli or the judges themselves
+\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.
 
@@
 Building on the previous section, it is clear that, despite its
-limitations, the BTL model commonly serves as the measurement model for
-CJ assessments. Specifically, the model provides a method for
-transforming manifest into latent variables (psychological scales) and
-allocating the compared stimuli on this continuum
+limitations, researchers typically rely on the BTL model to measure a
+trait and place the compared stimuli along such trait continuum
 \citetext{\citealp[pp.~269]{Thurstone_1927b}; \citealp{Everitt_et_al_2010}}.
-For example, when evaluating writing quality, researchers use the BTL
-model to process the dichotomous outcomes resulting from the pairwise
-comparisons (the manifest variables) to estimate scores that reflect the
-underlying level of writing quality (the latent variable)
-\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
-
-Researchers then typically use the BTL scores' point estimates, or their
-transformations, to conduct additional analyses or hypothesis tests. For
-example, these scores have been used to identify `misfit' judges and
-stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
-detect biases in judges' ratings
-\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
-other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
-or test hypotheses related to the underlying trait of interest
+Additionally, the CJ literature shows that researchers frequently use
+point estimates of BTL scores or their transformations to conduct
+further analyses or hypothesis tests. For example, researchers have used
+these scores to identify `misfit' judges and stimuli
+\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
+biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
+calculate correlations with other assessment methods
+\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
+related to the underlying trait of interest
 \citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
 
-However, the statistical literature advises caution when using estimated
-scores for additional analyses and tests. A key consideration is that
-BTL scores are parameter estimates that inherently carry uncertainty.
-Ignoring this uncertainty can bias the analysis and reduce the precision
-of hypothesis tests. Notably, the direction and magnitude of such biases
-are often unpredictable. Results may be attenuated, exaggerated, or
-remain unaffected depending on the degree of uncertainty in the scores
-and the actual effects being tested
+Nevertheless, while separating the trait measurement and hypothesis
+testing processes simplifies the analysis of CJ data, the statistical
+literature cautions against relying solely on the point estimates of BTL
+scores to conduct further analyses or hypothesis tests, as this practice
+can undermine the resulting statistical conclusions. A key consideration
+is that BTL scores are parameter estimates that inherently carry
+uncertainty. Ignoring this uncertainty can bias the analysis and reduce
+the precision of hypothesis tests. Notably, the direction and magnitude
+of such biases are often unpredictable. Results may be attenuated,
+exaggerated, or remain unaffected depending on the degree of uncertainty
+in the scores and the actual effects being tested
 \citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
-Finally, the reduced precision in hypothesis tests diminishes their
+Furthermore, the reduced precision in hypothesis tests diminishes their
 statistical power, increasing the likelihood of committing type-I or
 type-II errors \citep{McElreath_2020}.
-
-In aggregate, excessive reliance on Thurstone's Case V assumptions in
+Figure~\ref{fig-measurement_error} illustrates these issues, showing how
+neglecting the uncertainty associated with BTL scores can reduce the
+precision of an effect linked to the predictor.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/measurement_error.png}
+
+}
+
+\caption{\label{fig-measurement_error}The effect of outcome uncertainty
+on the estimation of an effect \(\beta\) linked to the predictor.
+Uncertainty increases from left to right.}
+
+\end{figure}%
+
+In aggregate, the heavy reliance on Thurstone's Case V assumptions in
 the statistical analysis of comparative data can compromise the
@@
 their validity \citep[pp.~2]{Perron_et_al_2015}, particularly when
-coupled with the disconnect between CJ's approach to trait measurement
-and hypothesis testing. However, the structural approach to causal
-inference can address these issues by offering a systematic and
-integrated framework that strengthens measurement reliability and
-validity while enhancing the statistical accuracy of hypothesis tests.
+coupled with the disconnect between the trait measurement and hypothesis
+testing processes. However, the structural approach to causal inference
+can address these issues by offering a systematic and integrated
+framework that strengthens measurement reliability and validity while
+enhancing the statistical accuracy of hypothesis tests.
 
@@
 using the structural approach to causal inference. Specifically, it
-formalizes the combination of the core theoretical principles outlined
-in Section~\ref{sec-thurstone_theory} with key assessment design
-features relevant to CJ experiments, such as the selection of judges,
-stimuli, and comparisons. In addition to enhancing statistical accuracy
-and strengthening measurement reliability and validity, the approach
-offers two key advantages. First, it clarifies the interactions among
-all actors and processes involved in CJ experiments. Second, it shifts
-the current comparative data analysis paradigm from passively accepting
-the model assumptions to actively testing whether those assumptions fit
-the data under analysis.
+combines the core theoretical principles outlined in
+Section~\ref{sec-thurstone_theory} with key assessment design features
+relevant to CJ experiments, such as the selection of judges, stimuli,
+and comparisons. In addition to enhancing statistical accuracy and
+strengthening measurement reliability and validity, the approach offers
+two key advantages. First, it clarifies the interactions among all
+actors and processes involved in CJ experiments. Second, it shifts the
+current comparative data analysis paradigm from passively accepting the
+model assumptions to actively testing whether those assumptions fit the
+data under analysis.
 

### 2025-03-20 code/CJmod_add.R

@@ -11,10 +11,10 @@ data{
     int nsK;    // max. number of judgments
     
     // categories discrete variables
-    int sup_dXI;  // max. number of categories in XI
-    int sup_dXA;  // max. number of categories in XA
-    int sup_dZJ;  // max. number of categories in ZJ
-    int sup_dZK;  // max. number of categories in ZK
+    int sup_XId;  // max. number of categories in XI
+    int sup_XAd;  // max. number of categories in XA
+    int sup_ZJd;  // max. number of categories in ZJ
+    int sup_ZKd;  // max. number of categories in ZK
     
     // comparison data
     array[ns] int<lower=1, upper=nsI> Rs1;   // individual (left)
@@ -29,82 +29,93 @@ data{
     array[nsI*nsA] int<lower=1, upper=nsI> IA1s;    // individuals
     array[nsI*nsA] int<lower=1, upper=nsA> IA2s;    // stimuli
     array[nsI*nsA] real XIc;                        // individual predictor (continuous)
-    array[nsI*nsA] int<lower=1, upper=sup_dXI> XId; // individual predictor (discrete)
+    array[nsI*nsA] int<lower=1, upper=sup_XId> XId; // individual predictor (discrete)
     array[nsI*nsA] real XIAc;                       // stimuli predictor (continuous)
-    array[nsI*nsA] int<lower=1, upper=sup_dXA> XIAd;// stimuli predictor (discrete)
+    array[nsI*nsA] int<lower=1, upper=sup_XAd> XIAd;// stimuli predictor (discrete)
     
     // juges-repeated comparisons data
     array[nsJ*nsK] int<lower=1, upper=nsJ> JK1s;    // judges
     array[nsJ*nsK] int<lower=1, upper=nsK> JK2s;    // judgments
     array[nsJ*nsK] real ZJc;                        // judges predictor (continuous)
-    array[nsJ*nsK] int<lower=1, upper=sup_dZJ> ZJd; // judges predictor (discrete)
+    array[nsJ*nsK] int<lower=1, upper=sup_ZJd> ZJd; // judges predictor (discrete)
     array[nsJ*nsK] real ZJKc;                       // judgments predictor (continuous)
-    array[nsJ*nsK] int<lower=1, upper=sup_dZK> ZJKd;// judgments predictor (discrete)
[…]

### 2025-03-21 paper2.qmd
section:4; process:write

@@
 While these assumptions simplify statistical modeling, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
-Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes.
+Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes.
 Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.
 
+<!-- 5. So, what are we doing? -->
+To address these issues, this study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods.
+In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
+First, it clarifies the interactions among all actors and processes involved in CJ experiments.
+Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.
+
+<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014).
+For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
+[@Deffner_et_al_2022, pp.
+2] -->
+
+<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
+
 <!-- 4. what is the organization of the study? -->
-To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods.
 As a result, the study divides its content into six main sections.
@@
 254] (see @tbl-thurstone_cases).
-However, as @vanderLinden_et_al_2017_I [pp.
-16] noted, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation.
+However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I, pp.
+16; @McElreath_2022].
 Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.
@@
 <!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Building on the previous section, it is clear that, despite its limitations, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along such trait continuum [@Thurstone_1927b, pp.
-269; @Everitt_et_al_2010].
+Building on the previous section, it is clear that, despite its limitations, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp.
+269].
 Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests.
@@
 Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions.
-A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty.
+A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error).
 Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests.
-Notably, the direction and magnitude of such biases are often unpredictable.
+The direction and magnitude of such biases are often unpredictable.
 Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp.
@@
 Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
-@fig-measurement_error illustrates these issues, showing how neglecting the uncertainty associated with BTL scores can reduce the precision of an effect linked to the predictor.
+@fig-measurement_error illustrates these issues, showing how neglecting measurement error in an outcome can reduce the precision of a predictor' estimated effect.
 
@@
 *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].
+
 <!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
@@
 A motivating example that will appear throughout the document clarifies these concepts.
-In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researcher design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
-Judges then compare pairs of students' written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait).
-Based on this setup, researchers reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-Finally, following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
+In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait).
+Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
+Following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
 
@@
 <!-- 6. What the section will do? -->
-Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone’s general form using the structural approach to causal inference.
+Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to Causal inference.
 Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
-In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
+In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
 First, it clarifies the interactions among all actors and processes involved in CJ experiments.
@@
 <!-- 7. How the section is divided? -->
-Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we call the *conceptual-population model*.
-This model assumes that researchers have access to a *census* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population, hence its name.
-Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we call the *sample-comparison model*.
+Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we refer to as the *conceptual-population model*.
+This model assumes that researchers have access to a "conceptual population" of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we refer to as the *sample-comparison model*.
 This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
@@
 
-### Integrating initial theoretical principles {#sec-theory-theoretical_P1}
-
 <!-- 1. The SCMs -->
@@
 
-<!-- 2. the preliminary SCM for CJ -->
-[SCM -@fig-cj03_scm] presents the initial integration of theoretical principles into the conceptual-population model for the example, applied to examine the effect of different teaching methods on students' writing ability.
-This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
+<!-- 2. The DAGs -->
+Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
+A DAG is a *graph* consisting of nodes connected by edges, where nodes represent random variables.
+The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence.
+The absence of an edge implies no direct relationship between the nodes.
+The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
+DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
+They also display conditioned variables by surrounding them with a square.
+Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp.
+32].
+Although DAGs often omit exogenous variables for simplicity, this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+
+<!-- the *standard representation* omit exogenous variables for simplicity -->
+<!-- the *magnified representation* includes exogenous variables -->
+
+### Integrating the first theoretical principles {#sec-theory-theoretical_P1}
+
+<!-- 1. the preliminary SCM for CJ -->
+[SCM -@fig-cj03_scm] illustrates the first theoretical principles of the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability.
+This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
 The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli).
-Moreover, the index $j$ refers to the judge, and $k$ specifies the judgment number, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
+The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
 366-376].
-Notably, the indexing system allows comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
-However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical meaning within the CJ framework.
+Notably, the indexing system facilitates comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
+However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical relevance within the CJ framework.
 
 In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
-Additionally, based on the arguments developed under the first scenario of @sec-theory-issue1b,  and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
-Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
-Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
+Additionally, based on the arguments developed in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
+Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
+Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
 
@@
 
-<!-- 4. The DAGs -->
-Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
-A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables.
-The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence, while the absence of an edge implies no relationship between the nodes.
-The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
-DAGs conventionally represent observed variables with solid black circles and unobserved (latent) variables with open circles [@Morgan_et_al_2014].
-They also display conditioned variables by surrounding them with a square.
-Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp.
-32].
-While DAGs often omit exogenous variables for simplicity, following a *standard representation*, this section adopts the *magnified representation* of DAGs, as including exogenous variables can improve clarity and help reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
-
-<!-- 5. the preliminary DAG for CJ -->
-@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the causal relationships among variables based on Thurstone's theory.
-Particularly, the graph shows how the texts' discriminal processes $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$.
-It also depicts the judges' biases $(B_{kj})$ as an additional factor influencing the texts' discriminal difference.
-Furthermore, the DAG distinguish between observed endogenous variables, such as the outcome, and latent endogenous variables, which include the texts' discriminal processes, their discriminal difference, and the judges' biases.
-
-### The comparative census data structure {#sec-theory-theoretical_P2}
+<!-- 2. the preliminary DAG for CJ -->
+Moreover, @fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the expected causal outlined in Thurstone's theory.
+The graph shows that the discriminal processes of the texts $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$.
+It also highlights the influence of judges' biases $(B_{kj})$ on the discriminal difference.
+Additionally, the DAG differentiates between observed endogenous variables, such as the outcome, and latent endogenous variables, including the texts' discriminal processes, their discriminal difference, and the judges' biases.
+
+<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
+<!-- 1.
+No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
+-->
+<!-- 2.
+No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
+-->
+<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
+Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
+[@Lash_et_al_2021, pp.
+113] -->
+
+### The conceptual-population data structure {#sec-theory-theoretical_P2}
 
@@
 Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one can improve clarity and facilitate the description of the system.
-Thus, to re-express the scalar form of the CJ system, outlined in the previous section, into an simpler equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
+Thus, to re-express the scalar form of the CJ system outlined in the previous section into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
 
@@
 <!-- 2. what is the data structure? -->
-where an element of $I$ corresponds to a unique individual $h$ or $i$, while each row of $IA$ corresponds to a unique pairing of individuals $h, i$ with stimuli $a, b$.
-Here, $n_{I}$ represents the number of individuals, while $n_{A}$ denotes the number of stimuli available per individual.
-Thus, vector $I$ has a length of $n_{I}$, and matrix $IA$ has a dimension of $(n_{I} \cdot n_{A})$ rows and $2$ columns.
-The vector and the matrix consist of scalars from the set of natural numbers $\mathbb{N}$.
-
-Similarly, an element of the vector $J$ corresponds to a unique judge $j$.
-Moreover, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment number $k$.
-In this case, $n_{J}$ indicates the total number of judges, and $n_{K}$ represents the number of judgments each judge performs.
-As a result, vector $J$ has a length of $n_{J}$, and matrix $JK$ has a dimension of $(n_{J} \cdot n_{K})$ rows and $2$ columns.
-Like the previous structures, the vector and the matrix consist of scalars from the set of natural numbers $\mathbb{N}$.
-
-Additionally, we define the matrix $RPUV$, where each row uniquely pairs the corresponding rows from the matrices $IA$ and $JK$.
-This matrix reflects the fundamental idea that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions.
-Thus, the matrix is defined as follows:
+Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals.
+Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges.
+Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$.
+As a result, the matrix $IA$ contains $(n_{I} \cdot n_{A})$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
+Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$, where $n_{K}$ indicates the number of judgments each judge makes.
+Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns.
+
+Additionally, we define the matrix $RPUV$ that assigns each row from the $IA$ matrix to a corresponding row from the $JK$ matrix.
+Thus, the matrix consists of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
+Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of repeated judgments each judge makes for every pair of stimuli generated by each pair of individuals in the population.
+The matrix is defined as follows:
+
+<!-- $$ -->
+<!-- {n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! } -->
+<!-- $$ -->
 
@@
 
-This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
-The matrix's dimension results from the conceptual-population model's assumption that researchers have access to a *census* of comparative data, capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
-Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs within the conceptual population.
-The coefficient is defined as follows:
-
-$$
-{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
-$$
-
-We visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals, the number of available stimuli per individual, the total number of judges, and the number of judgments each judge performs.
-For instance, assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:
+It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$.
+In this simple case, the vectors and matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:
 
@@
 <!-- 3. how de we use that data structure? -->
-Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in a simpler equivalent vectorized form, as shown in @fig-cj04.
+Then, using equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
 In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values.
-These vectors encompass all values from the conceptual population.
-Specifically, $O^{cp}_{RPUV}$ is an observed vector of length ${n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, while $D_{RPUV}$ is a latent vector of the same dimension.
-Additionally, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.
+These vectors capture all the observations from the conceptual population.
+Specifically, $O^{cp}_{RPUV}$ is an observed vector of length $n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, and $D_{RPUV}$ is a latent vector of the same dimension.
+Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.
 
@@
 24], the conceptual-population model incorporates two *hierarchical structural components*.
-Each component describes how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
-This addition addresses the problems arising when an additional hierarchical structure is relevant to the stimuli, as outlined in the second scenario of @sec-theory-issue1b.
-Furthermore, it facilitates analyses that integrate both trait measurement and hypothesis testing processes, thereby accounting for both traits and their associated uncertainties, reducing bias and improving measurement precision (see @sec-theory-issue2).
-
-<!-- principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
-138] and IRT [@Fox_2010, chap.
-6; @vanderLinden_et_al_2017_I, chap.
-24] recommend conducting these analyses and tests within a structural model.
-A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
-This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
--->
+A structural component describes how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
+Moreover, their hierarchical nature enable analyses when the stimuli exhibit a hierarchical structure (refer to @sec-theory-issue1b) and when researchers aim to test hypotheses while accounting for the uncertainties associated with the traits (see @sec-theory-issue2).
 
 <!-- 2 Explaining what we did -->
-@fig-cj09 displays the two hierarchical structural components for the example, applied to examine the effect of different teaching methods on students' writing ability.
-Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component.
+@fig-cj09 displays these components applied to examine the effect of different teaching methods on students' writing ability.
+The top branch of [DAG -@fig-cj09_dag] depicts the first component.
 This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$, such as text length, causally influence the texts' written-quality trait $(T_{IA})$.
@@
 These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *census* of comparative data, encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population* of comparative data, encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
 Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
@@
 Following the convention of @McElreath_2020 [pp.
-499-516], the [DAG -@fig-cj12_dag] depicts the conceptual-population outcome $O^{cp}_{RPUV}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the sampling mechanism.
-Moreover, the DAG shows the conditioned *sample design* variable $S$, which causally influence the sample-comparison outcome $O^{sc}_{RPUV}$.
+499-516] and @Deffner_et_al_2022 [pp.
+6], the [DAG -@fig-cj12_dag] depicts the conceptual-population outcome $O^{cp}_{RPUV}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the sampling mechanism.
+Moreover, the DAG shows the conditioned *sample design* variable $S$, which causally influence the sample-comparison outcome $O^{sc}_{RPUV}$ .
 This variable reflects how this sample-comparison outcome is a sample drawn from the conceptual-population outcome $O^{cp}_{RPUV}$.
@@
 
-<!-- Pollitt (2012a) argues that assessors in pairwise comparisons need less training, given intuitive comparison and decision making.
-However, there is a distinct lack of research on the role of training.
--->
-
-<!-- Due to efficiency of scoring, most of research only collects only one text.
-This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
-However, this cannot be known ad-hoc.
--->
-
 <!-- (Is this a validity contention?) -->
@@
 683] -->
+
+<!-- Samples from different sites often differ in terms of their demographic profiles (here, their age and gender distribution), and these demographic variables might in turn affect the distribution of the trait of interest.
+How should researchers deal with these differences?
+The answer depends on the processes that generated the observed disparities.
+Demographic disparities among samples may result from (a) differences in the actual populations from which the samples are taken or (b) sampling procedures that differ among sites.
+[@Deffner_et_al_2022, pp.
+6] -->
+
+<!-- Pollitt (2012a) argues that assessors in pairwise comparisons need less training, given intuitive comparison and decision making.
+However, there is a distinct lack of research on the role of training.
+[?] -->
+
+<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
+
+<!-- These examples demonstrate that the generalizability of experimental effects does not depend on the presence of population differences per se but on the exact mechanisms by which populations differ.
+While some differences—especially those concerning the independent variable—are inconsequential for intended generalizations, differences concerning effect modifiers or mediators require statistical adjustment.
+[@Deffner_et_al_2022, pp.
+11] -->
+
+<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
+If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
+...
+It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
+681] -->
 
@@
 
-<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
-If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
-...
-It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
-681] -->
-
-<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
-
 <!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
@@
 
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
-267] -->
+<!-- Due to efficiency of scoring, most of research only collects only one text.
+This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
+However, this cannot be known ad-hoc.
+-->
 
@@
 
-<!-- Show the logic of assuming an equivalent class for the graph that you are proposing. -->
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
+<!-- DAGs are not about adding assumptions—they are about revealing the -->
+<!-- assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
+
+<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
+
+<!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
+Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
+12].
+-->
+
+<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
+[@Schuessler_et_al_2023, pp.
+3] -->
+
+<!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
+[@Deffner_et_al_2022, pp.
+4] -->
+
+<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
+But an incomplete model is still an improvement over no model at all.
+[@Deffner_et_al_2022, pp.
+5] -->
+
+<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
+
+<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+<!-- This approach makes strong causal assumptions about the nature of confounding and our ability to measure shared history.
+However, strong assumptions are always necessary in observational settings.
+What is important is that the assumptions are transparent and logically connected to data analysis.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+<!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
+But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
+[@Deffner_et_al_2022, pp.
+14] -->
+
+<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
+[@Deffner_et_al_2022, pp.
+14] -->
 
@@
 150] -->
+
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
 

### 2025-03-21 _tex/paper2.tex
Built site for gh-pages

@@
 capture the complexity of CJ data, potentially leading to unreliable and
-inaccurate trait estimates. Second, the study highlights the apparent
-disconnect between CJ's trait measurement and hypothesis testing
-processes. Although separating these processes simplifies the analysis
-of CJ data, it may also undermine the reliability of various statistical
-results derived from the measurement.
-
-To address these issues, the study extends Thurstone's general form
+inaccurate trait estimates. Second, the study highlights the disconnect
+between CJ's trait measurement and hypothesis testing processes.
+Although separating these processes simplifies the analysis of CJ data,
+it may also undermine the reliability of various statistical results
+derived from the measurement.
+
+To address these issues, this study extends Thurstone's general form
 using a systematic and integrated approach based on Causal and Bayesian
-inference methods. As a result, the study divides its content into six
-main sections. Section~\ref{sec-thurstone_theory} provides an overview
-of Thurstone's theory. Section~\ref{sec-theory-issues} discusses the
-identified issues in detail. Section~\ref{sec-theoretical} extends
-Thurstone's general form to address these challenges. The extension
-integrates core theoretical principles alongside key assessment design
-features relevant to CJ experiments, such as the selection of judges,
-stimuli, and comparisons. Section~\ref{sec-statistical} translates these
-theoretical and practical elements into a probabilistic statistical
-model to analyze dichotomous pairwise comparison data.
-Section~\ref{sec-discussion} discusses the findings, limitations, and
-challenges and explores avenues for future research. Finally,
-Section~\ref{sec-conclusion} summarizes the study's conclusions.
+inference methods. In addition to improving statistical accuracy and
+strengthening measurement reliability and validity, the approach offers
+two key advantages. First, it clarifies the interactions among all
+actors and processes involved in CJ experiments. Second, it shifts the
+current comparative data analysis paradigm from passively accepting the
+BTL model assumptions to actively testing whether those assumptions fit
+the data under analysis.
+
+As a result, the study divides its content into six main sections.
+Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
+theory. Section~\ref{sec-theory-issues} discusses the identified issues
+in detail. Section~\ref{sec-theoretical} extends Thurstone's general
+form to address these challenges. The extension integrates core
+theoretical principles alongside key assessment design features relevant
+to CJ experiments, such as the selection of judges, stimuli, and
+comparisons. Section~\ref{sec-statistical} translates these theoretical
+and practical elements into a probabilistic statistical model to analyze
+dichotomous pairwise comparison data. Section~\ref{sec-discussion}
+discusses the findings, limitations, and challenges and explores avenues
+for future research. Finally, Section~\ref{sec-conclusion} summarizes
+the study's conclusions.
 
@@
 distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
-Table~\ref{tbl-thurstone_cases}). However, as
-\citet[pp.~16]{vanderLinden_et_al_2017_I} noted, this substitution has
-minimal impact on the model's estimation or interpretation because the
+Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
+impact on the model's estimation or interpretation because the
 discriminal process scale is arbitrary up to a non-monotonic
-transformation. Furthermore, this limited impact is supported by the
-fact that the Normal and Logistic distributions exhibit analogous
-statistical properties, differing only by a scaling factor of
-approximately \(1.7\).
+transformation
+\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2022}}.
+Furthermore, this limited impact is supported by the fact that the
+Normal and Logistic distributions exhibit analogous statistical
+properties, differing only by a scaling factor of approximately \(1.7\).
 
@@
 limitations, researchers typically rely on the BTL model to measure a
-trait and place the compared stimuli along such trait continuum
-\citetext{\citealp[pp.~269]{Thurstone_1927b}; \citealp{Everitt_et_al_2010}}.
-Additionally, the CJ literature shows that researchers frequently use
-point estimates of BTL scores or their transformations to conduct
-further analyses or hypothesis tests. For example, researchers have used
-these scores to identify `misfit' judges and stimuli
+trait and place the compared stimuli along its continuum
+\citep[pp.~269]{Thurstone_1927b}. Additionally, the CJ literature shows
+that researchers frequently use point estimates of BTL scores or their
+transformations to conduct further analyses or hypothesis tests. For
+example, researchers have used these scores to identify `misfit' judges
+and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
@@
 is that BTL scores are parameter estimates that inherently carry
-uncertainty. Ignoring this uncertainty can bias the analysis and reduce
-the precision of hypothesis tests. Notably, the direction and magnitude
-of such biases are often unpredictable. Results may be attenuated,
-exaggerated, or remain unaffected depending on the degree of uncertainty
-in the scores and the actual effects being tested
+uncertainty (measurement error). Ignoring this uncertainty can bias the
+analysis and reduce the precision of hypothesis tests. The direction and
+magnitude of such biases are often unpredictable. Results may be
+attenuated, exaggerated, or remain unaffected depending on the degree of
+uncertainty in the scores and the actual effects being tested
 \citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
@@
 Figure~\ref{fig-measurement_error} illustrates these issues, showing how
-neglecting the uncertainty associated with BTL scores can reduce the
-precision of an effect linked to the predictor.
+neglecting measurement error in an outcome can reduce the precision of a
+predictor' estimated effect.
 
@@
 extent do different teaching methods influence students' ability to
-produce high-quality written texts?'' To investigate this, researcher
-design a CJ experiment by randomly assigning students (individuals) to
+produce high-quality written texts?'' To investigate this, a researcher
+designs a CJ experiment by randomly assigning students (individuals) to
 two groups, each receiving a different teaching method. Judges then
-compare pairs of students' written texts (stimuli), to produce a
-dichotomous outcome that reflects the relative quality of each text
-(trait). Based on this setup, researchers reformulate the research
+compare pairs of students' written texts (stimuli) to produce a
+dichotomous outcome reflecting the relative quality of each text
+(trait). Based on this setup, researchers can reformulate the research
 question as the estimand: ``\emph{On average}, is there a difference in
 the ability to produce high-quality written texts between the two groups
-of students?''. Finally, following current CJ practices, researchers
-then rely on estimates from the BTL model, or its transformations, to
-approximate this estimand.
+of students?''. Following current CJ practices, researchers then rely on
+estimates from the BTL model, or its transformations, to approximate
+this estimand.
 
@@
 Section~\ref{sec-theory-issues} by extending Thurstone's general form
-using the structural approach to causal inference. Specifically, it
+using the structural approach to Causal inference. Specifically, it
 combines the core theoretical principles outlined in
@@
 relevant to CJ experiments, such as the selection of judges, stimuli,
-and comparisons. In addition to enhancing statistical accuracy and
+and comparisons. In addition to improving statistical accuracy and
 strengthening measurement reliability and validity, the approach offers
@@
 Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
-theoretical principles into what we call the \emph{conceptual-population
-model}. This model assumes that researchers have access to a
-\emph{census} of comparative data, that is, data representing all
-repeated judgments made by every available judge for each pair of
-stimuli produced by each pair of individuals in the population, hence
-its name. Conversely, Section~\ref{sec-theory-theoretical_SC} integrates
-the assessment design features into what we call the
+theoretical principles into what we refer to as the
+\emph{conceptual-population model}. This model assumes that researchers
+have access to a ``conceptual population'' of comparative data, that is,
+data representing all repeated judgments made by every available judge
+for each pair of stimuli produced by each pair of individuals in the
+population. Conversely, Section~\ref{sec-theory-theoretical_SC}
+integrates the assessment design features into what we refer to as the
 \emph{sample-comparison model}. This model assumes a more realistic
@@
 model}\label{sec-theory-theoretical_P}
-
-\subsubsection{Integrating initial theoretical
-principles}\label{sec-theory-theoretical_P1}
 
@@
 
-SCM~\ref{fig-cj03_scm} presents the initial integration of theoretical
-principles into the conceptual-population model for the example, applied
-to examine the effect of different teaching methods on students' writing
-ability. This SCM maps the relationship between the
-conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
-variables. The subscripts \(h\) and \(i\) label the students who
-authored the texts (individuals), while the indices \(a\) and \(b\)
-represent the compared texts (stimuli). Moreover, the index \(j\) refers
-to the judge, and \(k\) specifies the judgment number, accounting for
-experimental conditions where judges perform the same judgment multiple
-times, i.e., a \emph{repeated measures design}
-\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system allows
-comparisons between texts written by the same student (\(h = i\),
-\(a \neq b\)) and between texts written by distinct students
+Notably, every SCM has an associated DAG
+\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a \emph{graph}
+consisting of nodes connected by edges, where nodes represent random
+variables. The term \emph{directed} indicates that edges or arrows
+extend from one node to another, indicating the direction of causal
+influence. The absence of an edge implies no direct relationship between
+the nodes. The term \emph{acyclic} means that the causal influences do
+not form loops, ensuring the influences do not cycle back on themselves
+\citep{McElreath_2020}. DAGs conventionally depict observed variables as
+solid black circles and unobserved (latent) variables as open circles
+\citep{Morgan_et_al_2014}. They also display conditioned variables by
+surrounding them with a square. Here, \emph{conditioning} refers to
+restricting the statistical analysis to a subset of the population
+defined by the values of the given variable \citep[pp.~32]{Neal_2020}.
+Although DAGs often omit exogenous variables for simplicity, this
+section includes exogenous variables to improve clarity and reveal
+potential issues related to conditioning and confounding
+\citep{Cinelli_et_al_2020}.
+
+\subsubsection{Integrating the first theoretical
+principles}\label{sec-theory-theoretical_P1}
+
+SCM~\ref{fig-cj03_scm} illustrates the first theoretical principles of
+the conceptual-population model designed to examine the impact of
+different teaching methods on students' writing ability. This SCM
+outlines the relationship between the conceptual-population outcome
+\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
+\(h\) and \(i\) label the students who authored the texts (individuals),
+while the indices \(a\) and \(b\) represent the compared texts
+(stimuli). The index \(j\) refers to the judge, and \(k\) represents the
+judgment index, accounting for experimental conditions where judges
+perform the same judgment multiple times, i.e., a \emph{repeated
+measures design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing
+system facilitates comparisons between texts written by the same student
+(\(h = i\), \(a \neq b\)) and between texts written by distinct students
 (\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
 where judges compare a student's text to itself (\(h = i\), \(a = b\)),
-as such comparison lacks practical meaning within the CJ framework.
+as such comparison lacks practical relevance within the CJ framework.
 
@@
 difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
-Additionally, based on the arguments developed under the first scenario
-of Section~\ref{sec-theory-issue1b}, and the recommendations of
+Additionally, based on the arguments developed in
+Section~\ref{sec-theory-issue1b} and the recommendations of
 \citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
 the judges' biases \((B_{kj})\). Together with the outcome, these
-variables form the preliminary set of endogenous variables,
+variables constitute the preliminary set of endogenous variables,
 \(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
-the SCM shows the preliminary set of structural equations,
+the SCM presents the preliminary set of structural equations,
 \(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
@@
 
-Notably, every SCM has an associated DAG
-\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
-consisting of nodes connected by edges, where nodes represent random
-variables. The term \emph{directed} indicates that edges or arrows
-extend from one node to another, indicating the direction of causal
-influence, while the absence of an edge implies no relationship between
-the nodes. The term \emph{acyclic} means that the causal influences do
-not form loops, ensuring the influences do not cycle back on themselves
-\citep{McElreath_2020}. DAGs conventionally represent observed variables
-with solid black circles and unobserved (latent) variables with open
-circles \citep{Morgan_et_al_2014}. They also display conditioned
-variables by surrounding them with a square. Here, \emph{conditioning}
-refers to restricting the statistical analysis to a subset of the
-population defined by the values of the given variable
-\citep[pp.~32]{Neal_2020}. While DAGs often omit exogenous variables for
-simplicity, following a \emph{standard representation}, this section
-adopts the \emph{magnified representation} of DAGs, as including
-exogenous variables can improve clarity and help reveal potential issues
-related to conditioning and confounding \citep{Cinelli_et_al_2020}.
-
-Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
-SCM~\ref{fig-cj03_scm}, illustrating the causal relationships among
-variables based on Thurstone's theory. Particularly, the graph shows how
-the texts' discriminal processes \((T_{ha}, T_{ib})\) influence their
-discriminal difference \((D_{hiabjk})\), which in turn determines the
-outcome \((O^{cp}_{hiabjk})\). It also depicts the judges' biases
-\((B_{kj})\) as an additional factor influencing the texts' discriminal
-difference. Furthermore, the DAG distinguish between observed endogenous
-variables, such as the outcome, and latent endogenous variables, which
-include the texts' discriminal processes, their discriminal difference,
-and the judges' biases.
-
-\subsubsection{The comparative census data
+Moreover, Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
+SCM~\ref{fig-cj03_scm}, illustrating the expected causal outlined in
+Thurstone's theory. The graph shows that the discriminal processes of
+the texts \((T_{ha}, T_{ib})\) influence their discriminal difference
+\((D_{hiabjk})\), which in turn determines the outcome
+\((O^{cp}_{hiabjk})\). It also highlights the influence of judges'
+biases \((B_{kj})\) on the discriminal difference. Additionally, the DAG
+differentiates between observed endogenous variables, such as the
+outcome, and latent endogenous variables, including the texts'
+discriminal processes, their discriminal difference, and the judges'
+biases.
+
+\subsubsection{The conceptual-population data
 structure}\label{sec-theory-theoretical_P2}
@@
 description of the system. Thus, to re-express the scalar form of the CJ
-system, outlined in the previous section, into an simpler equivalent
-vectorized form, we first define the vectors \(I\) and \(J\), along with
-the matrices \(IA\) and \(JK\), as follows:
+system outlined in the previous section into an equivalent vectorized
+form, we first define the vectors \(I\) and \(J\), along with the
+matrices \(IA\) and \(JK\), as follows:
 
@@
 
-where an element of \(I\) corresponds to a unique individual \(h\) or
-\(i\), while each row of \(IA\) corresponds to a unique pairing of
-individuals \(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents
-the number of individuals, while \(n_{A}\) denotes the number of stimuli
-available per individual. Thus, vector \(I\) has a length of \(n_{I}\),
-and matrix \(IA\) has a dimension of \((n_{I} \cdot n_{A})\) rows and
-\(2\) columns. The vector and the matrix consist of scalars from the set
-of natural numbers \(\mathbb{N}\).
-
-Similarly, an element of the vector \(J\) corresponds to a unique judge
-\(j\). Moreover, each row of \(JK\) specifies a unique pairing of a
-judge \(j\) with a judgment number \(k\). In this case, \(n_{J}\)
-indicates the total number of judges, and \(n_{K}\) represents the
-number of judgments each judge performs. As a result, vector \(J\) has a
-length of \(n_{J}\), and matrix \(JK\) has a dimension of
-\((n_{J} \cdot n_{K})\) rows and \(2\) columns. Like the previous
-structures, the vector and the matrix consist of scalars from the set of
-natural numbers \(\mathbb{N}\).
-
-Additionally, we define the matrix \(RPUV\), where each row uniquely
-pairs the corresponding rows from the matrices \(IA\) and \(JK\). This
-matrix reflects the fundamental idea that the discriminal difference of
-the stimuli becomes an observable outcome only through the judges'
-perceptions. Thus, the matrix is defined as follows:
+Here, each element of \(I\) represents a unique individual \(h\) or
+\(i\), where \(n_{I}\) denotes the total number of individuals.
+Similarly, each element of \(J\) corresponds to a unique judge \(j\),
+with \(n_{J}\) indicating the total number of judges. Moreover, each row
+of \(IA\) represents a unique pairing of individuals \(h, i\) with
+stimuli \(a, b\). As a result, the matrix \(IA\) contains
+\((n_{I} \cdot n_{A})\) rows and \(2\) columns, where \(n_{A}\)
+specifies the number of stimuli available per individual. Likewise, each
+row of \(JK\) associates a judge \(j\) with a judgment index \(k\),
+where \(n_{K}\) indicates the number of judgments each judge makes.
+Consequently, the matrix \(JK\) has \((n_{J} \cdot n_{K})\) rows and
+\(2\) columns.
+
+Additionally, we define the matrix \(RPUV\) that assigns each row from
+the \(IA\) matrix to a corresponding row from the \(JK\) matrix. Thus,
+the matrix consists of
+\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
+\(6\) columns. Here, the term \({n_{I} \cdot n_{A} \choose 2}\)
+represents the Binomial coefficient, which quantifies the total number
+of repeated judgments each judge makes for every pair of stimuli
+generated by each pair of individuals in the population. The matrix is
+defined as follows:
 
@@
 
-This matrix also consists of scalars from the set of natural numbers
-\(\mathbb{N}\) and has a dimension of
-\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
-\(6\) columns. The matrix's dimension results from the
-conceptual-population model's assumption that researchers have access to
-a \emph{census} of comparative data, capturing all repeated judgments
-made by every available judge for each pair of stimuli produced by each
-pair of individuals in the population. Here, the term
-\({n_{I} \cdot n_{A} \choose 2}\) represents the Binomial coefficient,
-which quantifies the total number of individual-stimuli comparison pairs
-within the conceptual population. The coefficient is defined as follows:
-
-\[
-{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
-\]
-
-We visualize the structure of these vectors and matrices by considering
-small values for the parameters corresponding to number of individuals,
-the number of available stimuli per individual, the total number of
-judges, and the number of judgments each judge performs. For instance,
-assuming \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
-the matrices described in equations (\ref{eq-mat12}) and (\ref{eq-mat3})
+It is easier to visualize the structure of these vectors and matrices by
+considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
+\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
+matrices described in equations (\ref{eq-mat12}) and (\ref{eq-mat3})
 take the following form:
@@
 
-Thus, using the data structure defined by the matrices in equations
-(\ref{eq-mat12}) and (\ref{eq-mat3}), we can re-express
-SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in a simpler
+Then, using equations (\ref{eq-mat12}) and (\ref{eq-mat3}), we can
+re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
 equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
@@
 judges' biases \(B_{JK}\) are represented as vectors rather than scalar
-values. These vectors encompass all values from the conceptual
+values. These vectors capture all the observations from the conceptual
 population. Specifically, \(O^{cp}_{RPUV}\) is an observed vector of
-length \({n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
-\(D_{RPUV}\) is a latent vector of the same dimension. Additionally,
+length \(n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), and
+\(D_{RPUV}\) is a latent vector of the same dimension. Moreover,
 \(T_{IA}\) and \(B_{JK}\) are latent vectors of lengths
@@
 the conceptual-population model incorporates two \emph{hierarchical
-structural components}. Each component describes how different observed
-or latent variables influence the primary latent variable of interest
-\citep{Everitt_et_al_2010}. This addition addresses the problems arising
-when an additional hierarchical structure is relevant to the stimuli, as
-outlined in the second scenario of Section~\ref{sec-theory-issue1b}.
-Furthermore, it facilitates analyses that integrate both trait
-measurement and hypothesis testing processes, thereby accounting for
-both traits and their associated uncertainties, reducing bias and
-improving measurement precision (see Section~\ref{sec-theory-issue2}).
-
-Figure~\ref{fig-cj09} displays the two hierarchical structural
-components for the example, applied to examine the effect of different
-teaching methods on students' writing ability. Specifically, the top
-branch of DAG~\ref{fig-cj09_dag} depicts the first component. This
+structural components}. A structural component describes how different
+observed or latent variables influence the primary latent variable of
+interest \citep{Everitt_et_al_2010}. Moreover, their hierarchical nature
+enable analyses when the stimuli exhibit a hierarchical structure (refer
+to Section~\ref{sec-theory-issue1b}) and when researchers aim to test
+hypotheses while accounting for the uncertainties associated with the
+traits (see Section~\ref{sec-theory-issue2}).
+
+Figure~\ref{fig-cj09} displays these components applied to examine the
+effect of different teaching methods on students' writing ability. The
+top branch of DAG~\ref{fig-cj09_dag} depicts the first component. This
 component describes how the students' writing-quality trait \((T_{I})\)
@@
 accuracy and precision, the model still depends on the unrealistic
-assumption that researchers have access to a \emph{census} of
-comparative data, encompassing all repeated judgments from every judge
-for each pair of stimuli produced by every individual in the population.
-Since researchers rarely meet this assumption in practice, they must
-consider a more realistic scenario. The extensions to the
-conceptual-population model, derived from this realistic scenario, are
-concisely described in the sample-comparison model.
+assumption that researchers have access to a \emph{conceptual
+population} of comparative data, encompassing all repeated judgments
+from every judge for each pair of stimuli produced by every individual
+in the population. Since researchers rarely meet this assumption in
+practice, they must consider a more realistic scenario. The extensions
+to the conceptual-population model, derived from this realistic
+scenario, are concisely described in the sample-comparison model.
 
@@
 
-Following the convention of \citet[pp.~499-516]{McElreath_2020}, the
-DAG~\ref{fig-cj12_dag} depicts the conceptual-population outcome
-\(O^{cp}_{RPUV}\) as an unobserved variable, emphasizing the
-researchers' inability to access such an outcome due to the sampling
-mechanism. Moreover, the DAG shows the conditioned \emph{sample design}
-variable \(S\), which causally influence the sample-comparison outcome
-\(O^{sc}_{RPUV}\). This variable reflects how this sample-comparison
-outcome is a sample drawn from the conceptual-population outcome
-\(O^{cp}_{RPUV}\). Specifically, the variable is a vector
-\(S = \{ s_{1}, \dots, s_{n} \}\), where \(n\) corresponds to the number
-of rows in the matrix \(RPUV\) (refer to
+Following the convention of \citet[pp.~499-516]{McElreath_2020} and
+\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} depicts
+the conceptual-population outcome \(O^{cp}_{RPUV}\) as an unobserved
+variable, emphasizing the researchers' inability to access such an
+outcome due to the sampling mechanism. Moreover, the DAG shows the
+conditioned \emph{sample design} variable \(S\), which causally
+influence the sample-comparison outcome \(O^{sc}_{RPUV}\) . This
+variable reflects how this sample-comparison outcome is a sample drawn
+from the conceptual-population outcome \(O^{cp}_{RPUV}\). Specifically,
+the variable is a vector \(S = \{ s_{1}, \dots, s_{n} \}\), where \(n\)
+corresponds to the number of rows in the matrix \(RPUV\) (refer to
 Section~\ref{sec-theory-theoretical_P2}), and each \(s_{i}\) is a binary

### 2025-03-22 paper2.qmd
section:4; process:write,figures

@@
 He cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270] because it imposes the most extensive set of simplifying assumptions among all the proposed cases [@Bramley_2008, pp.
+270], as it imposes the most extensive set of simplifying assumptions among all the proposed cases [@Bramley_2008, pp.
 253; @Kelly_et_al_2022, pp.
@@
 <!-- 1. The Case V and its simpler statistical representation: the BTL model -->
-Despite its limitations, Case V remains the most widely used model in CJ literature.
+As previously discussed, Case V remains the most widely used model in CJ literature.
 This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case.
@@
 However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I, pp.
-16; @McElreath_2022].
+16; @McElreath_2021].
 Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.
@@
 <!-- Model misspecification happens when the set of probability distributions considered by the statistician does not include the distribution that generated the observed data. -->
+
+<!-- the scale of the latent variable is arbitrary, so there is no loss of generality [@McElreath_2021] 02:40:30 - 02:40:45  -->
 
@@
 Second, misfit texts may represent outlying observations with characteristics that make them difficult to compare and position on the trait scale.
-Regarding the first cause, the overlap can occur when misfit texts share the same trait space and have equal dispersion as non-misfit texts or when they exhibit larger discriminal dispersions.
+Regarding the first cause, the overlap can occur when misfit texts have equal dispersion and share the same trait space as non-misfit texts or when they exhibit larger discriminal dispersions.
 Although the BTL model can distinguish the first scenario, it is inherently agnostic to the second.
 @fig-discriminal_process illustrates how an increased discriminal dispersion can widen the overlap between two texts, leading to more judgment discrepancies.
-In the figure, text B has a broader discriminal dispersion than text A, expanding the overlap between their discriminal processes, where judges struggle to determine which text possesses more of the trait.
+In the figure, text B has a broader discriminal dispersion than text A, expanding the overlap between their discriminal processes, a space where judges struggle to determine which text possesses more of the trait.
 
@@
 <!-- 5. So, what happens when the assumption does not hold? -->
-Thus, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues.
-Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
+When the assumption of equal dispersions between stimuli does not hold, significant statistical and measurement issues can arise Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
 This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp.
@@
 The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis.
-Researchers define *misfit judges* as those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a, pp.
+*Misfit judges* are those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a, pp.
 164-165; @Pollitt_2012b, pp.
@@
 <!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Building on the previous section, it is clear that, despite its limitations, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp.
+Building on the previous section, it is clear that, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp.
 269].
@@
 Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
-@fig-measurement_error illustrates these issues, showing how neglecting measurement error in an outcome can reduce the precision of a predictor' estimated effect.
+@fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.
 
@@
 
-<!-- People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. -->
+<!-- "If Thurstone’s Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes." [@Kelly_et_al_2022, pp.
+678] -->
 
@@
 <!-- 3. Example of a CJ estimand and estimator, and its limitations -->
-A motivating example that will appear throughout the document clarifies these concepts.
+A motivating example that will appear in the rest of the document clarifies these concepts.
 In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
@@
 This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
-Second, they depict causal relationships in an interactive, non-parametric way.
+Second, they depict causal relationships in a non-parametric way.
 This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp.
@@
 Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we refer to as the *conceptual-population model*.
-This model assumes that researchers have access to a "conceptual population" of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+This model assumes that researchers have access to a *conceptual population* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
 Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we refer to as the *sample-comparison model*.
@@
 
+<!-- 1. What is the assumption on the conceptual-population model -->
+In the conceptual-population model, we assume an idealized scenario where researchers have access to a *conceptual population* of comparative data --data representing all repeated judgments made by each judge for every stimulus pair produced by each pair of individuals in the population.
+This assumption allows us to integrate Thurstone's theoretical principles and propose innovations to address some of the issues raised in @sec-theory-issues.
+
+### Integrating the first theoretical principles {#sec-theory-theoretical_P1}
+
 <!-- 1. The SCMs -->
@@
 
-<!-- 2. The DAGs -->
+<!-- 2. the preliminary SCM for CJ -->
+[SCM -@fig-cj03_scm] illustrates the first theoretical principles of the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability.
+This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
+The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli).
+The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
+366-376].
+Notably, the indexing system facilitates comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
+However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical relevance within the CJ framework.
+
+In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
+Additionally, based on the arguments developed in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
+Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
+Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
+
+<!-- @Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
+Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
+Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
+26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap.
+15], depending on the theoretical and statistical treatment of judges' biases.
+-->
+
+::: {#fig-cj03 layout-nrow=2}
+
+::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
+  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=35%}
+
+Conceptual-population model, scalar form.
+:::
+
+<!-- 3. The DAGs -->
 Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
@@
 
-### Integrating the first theoretical principles {#sec-theory-theoretical_P1}
-
-<!-- 1. the preliminary SCM for CJ -->
-[SCM -@fig-cj03_scm] illustrates the first theoretical principles of the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability.
-This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
-The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli).
-The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
-366-376].
-Notably, the indexing system facilitates comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
-However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical relevance within the CJ framework.
-
-In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
-Additionally, based on the arguments developed in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
-Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
-Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
-
-<!-- @Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
-Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
-Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
-26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap.
-15], depending on the theoretical and statistical treatment of judges' biases.
--->
-
-::: {#fig-cj03 layout-nrow=2}
-
-::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
-  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=35%}
-
-Conceptual-population model, scalar form.
-:::
-
-<!-- 2. the preliminary DAG for CJ -->
-Moreover, @fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the expected causal outlined in Thurstone's theory.
+<!-- 4. the preliminary DAG for CJ -->
+@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the expected causal relationships outlined in Thurstone's theory.
 The graph shows that the discriminal processes of the texts $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$.
 It also highlights the influence of judges' biases $(B_{kj})$ on the discriminal difference.
-Additionally, the DAG differentiates between observed endogenous variables, such as the outcome, and latent endogenous variables, including the texts' discriminal processes, their discriminal difference, and the judges' biases.
+Additionally, the DAG differentiates between observed endogenous variables, such as the outcome (solid black circle), and latent endogenous variables, including the texts' discriminal processes, their discriminal difference, and the judges' biases (open circles).
 
@@
 
-### The conceptual-population data structure {#sec-theory-theoretical_P2}
+### The *conceptual population* data structure {#sec-theory-theoretical_P2}
 
 <!-- 1. Why is useful to declare the data structure? -->
-Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one can improve clarity and facilitate the description of the system.
-Thus, to re-express the scalar form of the CJ system outlined in the previous section into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
-
-$$
-I_{n_{I}} = \begin{bmatrix}
+Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system.
+Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
+
+$$
+I = \begin{bmatrix}
 1 \\
@@
 \end{bmatrix} ; \;
-J_{n_{J}} = \begin{bmatrix}
+J = \begin{bmatrix}
 1 \\
@@
 \end{bmatrix} ; \;
-IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
+IA = \begin{bmatrix}
 1 & 1 \\
@@
 \end{bmatrix} ; \;
-JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
+JK = \begin{bmatrix}
 1 & 1 \\
@@
 \end{bmatrix}
-$$ {#eq-mat12}
+$$ {#eq-mat11}
 
@@
 As a result, the matrix $IA$ contains $(n_{I} \cdot n_{A})$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
-Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$, where $n_{K}$ indicates the number of judgments each judge makes.
-Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns.
-
-Additionally, we define the matrix $RPUV$ that assigns each row from the $IA$ matrix to a corresponding row from the $JK$ matrix.
-Thus, the matrix consists of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns.
-Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of repeated judgments each judge makes for every pair of stimuli generated by each pair of individuals in the population.
-The matrix is defined as follows:
+Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$.
+Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.
+
+Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
+Thus, this matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
+Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
+Whe define the matrix as follows:
 
@@
 $$
-RPUV_{n \times 6} = \begin{bmatrix}
+R = \begin{bmatrix}
 1 & 1 & 1 & 2 & 1 & 1 \\
@@
 \end{bmatrix}
-$$ {#eq-mat3}
+$$ {#eq-mat12}
 
 It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$.
-In this simple case, the vectors and matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:
-
-$$
-I_{5} = \begin{bmatrix}
+In this simple case, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
+
+$$
+I = \begin{bmatrix}
 1 \\
@@
 \end{bmatrix} ; \;
-IA_{10 \times 2} = \begin{bmatrix}
+J = \begin{bmatrix}
+1 \\
+2 \\
+3 \\
+\end{bmatrix} ; \;
+IA = \begin{bmatrix}
 1 & 1 \\
@@
 \end{bmatrix} ; \;
-J_{3} = \begin{bmatrix}
-1 \\
-2 \\
-3 \\
-\end{bmatrix} ; \;
-JK_{9 \times 2} = \begin{bmatrix}
+JK = \begin{bmatrix}
 1 & 1 \\
@@
 \end{bmatrix} ; \;
-RPUV_{405  \times 6} = \begin{bmatrix}
+R = \begin{bmatrix}
 1 & 1 & 1 & 2 & 1 & 1 \\
@@
 <!-- 3. how de we use that data structure? -->
-Then, using equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
-In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values.
+Now, using equations ([-@eq-mat11]) and ([-@eq-mat12]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
+In this depiction, the outcome $O^{cp}_{R}$, the texts' discriminal difference $D_{R}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values.
 These vectors capture all the observations from the conceptual population.
-Specifically, $O^{cp}_{RPUV}$ is an observed vector of length $n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, and $D_{RPUV}$ is a latent vector of the same dimension.
+Specifically, $O^{cp}_{R}$ and $D_{R}$ are observed and latent vectors of length $n$, respectively.
 Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.
@@
 \begin{aligned}
-  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
-  D_{RPUV} & := f_{D}(T_{IA}, B_{JK})
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
 \end{aligned}
@@
 <!-- 1. What are we going to do? -->
-Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+The conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability.
+This approach builds on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
 138] and Item Response Theory (IRT) [@Fox_2010, chap.
 6; @vanderLinden_et_al_2017_I, chap.
-24], the conceptual-population model incorporates two *hierarchical structural components*.
-A structural component describes how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010].
-Moreover, their hierarchical nature enable analyses when the stimuli exhibit a hierarchical structure (refer to @sec-theory-issue1b) and when researchers aim to test hypotheses while accounting for the uncertainties associated with the traits (see @sec-theory-issue2).
+24].
+Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
+Their hierarchical nature enables researchers to test hypotheses while accounting for uncertainties associated with trait estimation (see @sec-theory-issue2), particularly when stimuli exhibit a hierarchical structure (see @sec-theory-issue1b).
 
 <!-- 2 Explaining what we did -->
-@fig-cj09 displays these components applied to examine the effect of different teaching methods on students' writing ability.
 The top branch of [DAG -@fig-cj09_dag] depicts the first component.
-This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$, such as text length, causally influence the texts' written-quality trait $(T_{IA})$.
+This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$.
 The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain.
-Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively.
-Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns.
-
-Furthermore, the top branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* student-related variables $(X_{I})$, including teaching methods, causally influence the students' writing-quality trait $(T_{I})$.
+In this structure, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively.
+Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns (variables).
+
+Additionally, the DAG's top branch illustrates how a set of *relevant* student-related variables $(X_{I})$, such as the teaching method, causally influence the students' writing-quality trait $(T_{I})$.
 The DAG also includes the idiosyncratic errors $(e_{I})$, which capture variations in the students' traits that the relevant student-related variables cannot explain.
-Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns.
+Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ can be an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables).
+
+Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
 
@@
 \begin{aligned}
-  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
-  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
   T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
@@
 Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component.
-This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$, such as number of judgments or time for judgment, causally influence the biases associated with each text $B_{JK}$.
+This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$.
 The DAG also highlights the idiosyncratic errors $(e_{JK})$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain.
@@
 
-Moreover, the bottom branch of [DAG -@fig-cj09_dag] describes how a set of *relevant* judge-related variables $(Z_{J})$, including judgment expertise, causally influence the judges' bias $(B_{J})$.
+Moreover, the DAG's bottom branch describes how a set of *relevant* judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the judges' bias $(B_{J})$.
 The DAG also accounts for idiosyncratic errors $(e_{J})$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain.
@@
 <!-- 3. segway for next section -->
-Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issues.
+Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
 These enhancements include accounting for judges' biases and integrating hierarchical structural components.
@@
 
+<!-- 1. What is the more realistic scenario: the sample mechanism -->
+In the sample-comparison model, we assume a more realistic scenario than the conceptual-population model.
+Specifically, we assume researchers have access to a data sample consisting of a limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, drawn from the conceptual population.
+Additionally, we assume that judges do not make *all possible comparisons* allowed by the sample, but rather only $n_{C}$ comparisons.
+
+### The sample mechanism {#sec-theory-theoretical_SC1}
+
 <!-- Develop how you get the vectors S and C and sampling weights when you know the extension of the population. (See https://mc-stan.org/docs/stan-users-guide/poststratification.html ) -->
 
-### The sample mechanism {#sec-theory-theoretical_SC1}
-
-<!-- 1. What is the more realistic scenario: the sample mechanism -->
-In the sample-comparison model, we assume a more realistic scenario compared to the conceptual-population model.
-Specifically, we assume that researchers have access to *all comparisons* from a limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$, along with a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$ drawn from the conceptual population.
-@fig-cj12 illustrates this model, incorporating only this sample mechanism.
-
-<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
+<!-- 1. start by defining the sampling vector and matrices  -->
+To facilitate the interpretation of the sample-comparison model, we first define the data sample using several binary vector variables: $S_{I}$, $S_{J}$, $S_{IA}$, $S_{JK}$, and $S$ as follows.
+
+$$
+S_{I} = \begin{bmatrix}
+i_{(1)} \\
+i_{(2)} \\
+\vdots \\
+i_{(h)} \\
+\vdots \\
+i_{(i)} \\
+\vdots \\
+i_{(nI)}
+\end{bmatrix} ; \;
+S_{J} = \begin{bmatrix}
+j_{(1)} \\
+j_{(2)} \\
+\vdots \\
+j_{(j)} \\
+\vdots \\
+j_{(nJ)}
+\end{bmatrix} ; \;
+S_{IA} = \begin{bmatrix}
+ia_{(1)} \\
+ia_{(2)} \\
+\vdots \\
+ia_{(ia)} \\
+\vdots \\
+ia_{(nI \cdot nA)} \\
+\end{bmatrix} ; \;
+S_{JK} = \begin{bmatrix}
+jk_{(1)} \\
+jk_{(2)} \\
+\vdots \\
+jk_{(jk)} \\
+\vdots \\
+jk_{(nJ \cdot nK)} \\
+\end{bmatrix}; \;
+S = \begin{bmatrix}
+s_{(1)} \\
+s_{(2)} \\
+\vdots \\
+s_{(s)} \\
+\vdots \\
+s_{(n)} \\
+\end{bmatrix}
+$$ {#eq-mat21}
+
+Each element of the vectors $S_{I}$ and $S_{J}$ is a binary value that indicates the presence $(1)$ or absence $(0)$ of data in the $I$ and $J$ vectors, as defined in equation ([-@eq-mat11]).
+Therefore, the vector $S_{I}$ contains $n_{I}$ elements, and the vector $S_{J}$ contains $n_{J}$ elements.
+For example, we define each element of the vector $S_{I}$ as in equation ([-@eq-mat22]), and we apply a similar definition to $S_{J}$.
+
+$$
+i_{(i)} = \begin{cases} 
+1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
+0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
+\end{cases}
+$$ {#eq-mat22}
+
+Each element of the vectors $S_{IA}$ and $S_{JK}$ is defined like $i_{(i)}$ with vector $I$, but applied to data in the matrices $IA$ and $JK$, as defined in equation ([-@eq-mat11]).
+Therefore, the vector $S_{IA}$ contains $n_{I} \cdot n_{A}$ elements, and the vector $S_{JK}$ contains $n_{J} \cdot n_{K}$ elements.
+For example, we define each element of the vector $S_{IA}$ as in equation ([-@eq-mat23]), and we apply a similar definition to $S_{JK}$.
+
+$$
+ia_{(ia)} = \begin{cases} 
+1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
+0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
+\end{cases}
+$$ {#eq-mat23}
+
+The equation shows that the observed values in the $S_{I}$ and $S_{J}$ vectors determine the binary values in the $S_{IA}$ and $S_{JK}$ vectors, through the $IA$ and $JK$ matrices, respectively.
+Specifically, researchers cannot observe stimuli and judgments for individuals and judges they did not sample initially.
+
+Finally, the vector $S$ in equation ([-@eq-mat21]) contains $n = \binom{n_{I} \cdot n_{A}}{2} \cdot n_{J} \cdot n_{K}$ elements, with each element defined in equation ([-@eq-mat24]).
+This equation shows that the observed values in $S_{IA}$ and $S_{JK}$ determine the binary values in $S$ through the $R$ matrix.
+Specifically, researchers can only observe a subset of judges' comparisons for a subset of stimuli generated by a subset of individuals.
+
+$$
+s_{(s)} = \begin{cases} 
+1 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are observed} \\
+0 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are missing}
+\end{cases}
+$$ {#eq-mat24}
 
 <!-- 2. What is the effect of the sampling mechanism -->
+With the definition of $S$, we can now integrate the sample mechanism into the conceptual-population model.
+@fig-cj12 illustrates this model.
 Following the convention of @McElreath_2020 [pp.
 499-516] and @Deffner_et_al_2022 [pp.
-6], the [DAG -@fig-cj12_dag] depicts the conceptual-population outcome $O^{cp}_{RPUV}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the sampling mechanism.
-Moreover, the DAG shows the conditioned *sample design* variable $S$, which causally influence the sample-comparison outcome $O^{sc}_{RPUV}$ .
-This variable reflects how this sample-comparison outcome is a sample drawn from the conceptual-population outcome $O^{cp}_{RPUV}$.
-Specifically, the variable is a vector $S = \{ s_{1}, \dots, s_{n} \}$, where $n$ corresponds to the number of rows in the matrix $RPUV$ (refer to @sec-theory-theoretical_P2), and each $s_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the conceptual-population outcome $O^{cp}_{RPUV}$.
-Specifically,
-
-$$
-s_i = \begin{cases} 
-1 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is observed} \\
-0 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is missing}
-\end{cases}
-$$
+6], the [DAG -@fig-cj12_dag] depicts the conceptual-population outcome $O^{cp}_{R}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the sampling mechanism.
+Additionally, the DAG depicts the conditioned *sample design* variable $S$ causally influencing the sample-comparison outcome $O^{sc}_{R}$.
+This variable shows that the sample-comparison outcome is a subset drawn from the conceptual-population outcome $O^{cp}_{R}$.
+
+<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
 
@@
 \begin{aligned}
-  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
-  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
-  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
   T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
@@
 Following the convention of @McElreath_2020 [pp.
-499-516], the [DAG -@fig-cj14_dag] depicts the sample-comparison outcome $O^{sc}_{RPUV}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the comparison mechanism.
-Moreover, the DAG shows the conditioned *comparison design* variable $C$, which causally influence the observed outcome $O_{RPUV}$.
-This variable reflects how this observed outcome is a sample drawn from the sample-comparison outcome $O^{sp}_{RPUV}$.
-Specifically, the variable is a vector $C = \{ c_{1}, \dots, c_{n} \}$, where $n$ corresponds to the number of rows in the matrix $RPUV$ (refer to @sec-theory-theoretical_P2), and each $c_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the sample-comparison outcome $O^{sp}_{RPUV}$.
+499-516], the [DAG -@fig-cj14_dag] depicts the sample-comparison outcome $O^{sc}_{R}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the comparison mechanism.
+Moreover, the DAG shows the conditioned *comparison design* variable $C$, which causally influence the observed outcome $O_{R}$.
+This variable reflects how this observed outcome is a sample drawn from the sample-comparison outcome $O^{sp}_{R}$.
+Specifically, the variable is a vector $C = \{ c_{1}, \dots, c_{n} \}$, where $n$ corresponds to the number of rows in the matrix $R$ (refer to @sec-theory-theoretical_P2), and each $c_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the sample-comparison outcome $O^{sp}_{R}$.
 Specifically,
@@
 c_i = \begin{cases} 
-1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{RPUV} \text{ is observed} \\
-0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{RPUV} \text{ is missing}
+1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is observed} \\
+0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is missing}
 \end{cases}
@@
 \begin{aligned}
-  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
-  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
-  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
-  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
+  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
   T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
@@
 \begin{aligned}
-  O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
-  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S \\
-  O^{cp}_{RPUV} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
-  D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
+  O_{R} & = O^{sc}_{R} \odot C \\
+  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
+  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
+  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
   T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
@@
 \begin{aligned}
-  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
-  O^{sc}_{RPUV} & := f_{S}(O^{cp}_{RPUV}, S) \\
-  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
-  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
+  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
+  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
   T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
@@
 
-<!-- DAGs are not about adding assumptions—they are about revealing the -->
-<!-- assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
-
-<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
-
-<!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
-Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
-12].
--->
-
-<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
-[@Schuessler_et_al_2023, pp.
-3] -->
-
-<!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
-[@Deffner_et_al_2022, pp.
-4] -->
-
-<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
-But an incomplete model is still an improvement over no model at all.
-[@Deffner_et_al_2022, pp.
-5] -->
-
-<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
-
-<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
-[@Deffner_et_al_2022, pp.
-13] -->
-
-<!-- This approach makes strong causal assumptions about the nature of confounding and our ability to measure shared history.
-However, strong assumptions are always necessary in observational settings.
-What is important is that the assumptions are transparent and logically connected to data analysis.
-[@Deffner_et_al_2022, pp.
-13] -->
-
-<!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
-But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
-[@Deffner_et_al_2022, pp.
-14] -->
-
-<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
-[@Deffner_et_al_2022, pp.
-14] -->
-
-## Limitations and further research {#sec-discussion-limitations}
-
-<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
-In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
-37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
--->
-
-<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
-This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
--->
-
-<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
-Nevertheless, as @Kohler_et_al_2019 [pp.
-150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
--->
-
 <!-- Statements about the real world from observed data require assumptions.
@@
 
+<!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
+But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
+[@Deffner_et_al_2022, pp.
+14] -->
+
+<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
+[@Deffner_et_al_2022, pp.
+14] -->
+
+<!-- This approach makes strong causal assumptions about the nature of confounding and our ability to measure shared history.
+However, strong assumptions are always necessary in observational settings.
+What is important is that the assumptions are transparent and logically connected to data analysis.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+<!-- DAGs are not about adding assumptions—they are about revealing the -->
+<!-- assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
+
+<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
+
+<!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
+Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
+12].
+-->
+
+<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
+[@Schuessler_et_al_2023, pp.
+3] -->
+
+<!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
+[@Deffner_et_al_2022, pp.
+4] -->
+
+<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
+But an incomplete model is still an improvement over no model at all.
+[@Deffner_et_al_2022, pp.
+5] -->
+
+<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
+
+<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+## Limitations and further research {#sec-discussion-limitations}
+
+<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
+In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
+37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+-->
+
+<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
+This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
+-->
+
+<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
+Nevertheless, as @Kohler_et_al_2019 [pp.
+150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
+-->
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
+
 <!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
@@
 -->
+
+<!-- Brought Rubin's missing data framework to the field. -->
 

### 2025-03-22 _tex/paper2.tex
Built site for gh-pages

@@
 its use ``should not be made without (an) experimental test''
-\citep[pp.~270]{Thurstone_1927b} because it imposes the most extensive
-set of simplifying assumptions among all the proposed cases
+\citep[pp.~270]{Thurstone_1927b}, as it imposes the most extensive set
+of simplifying assumptions among all the proposed cases
 \citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
@@
 
-Despite its limitations, Case V remains the most widely used model in CJ
+As previously discussed, Case V remains the most widely used model in CJ
 literature. This preference primarily stems from the BTL model, which
@@
 transformation
-\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2022}}.
+\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2021}}.
 Furthermore, this limited impact is supported by the fact that the
@@
 and position on the trait scale. Regarding the first cause, the overlap
-can occur when misfit texts share the same trait space and have equal
-dispersion as non-misfit texts or when they exhibit larger discriminal
+can occur when misfit texts have equal dispersion and share the same
+trait space as non-misfit texts or when they exhibit larger discriminal
 dispersions. Although the BTL model can distinguish the first scenario,
@@
 discriminal dispersion than text A, expanding the overlap between their
-discriminal processes, where judges struggle to determine which text
-possesses more of the trait.
-
-Thus, erroneously assuming equal dispersions between stimuli can give
-rise to significant statistical and measurement issues. Specifically,
+discriminal processes, a space where judges struggle to determine which
+text possesses more of the trait.
+
+When the assumption of equal dispersions between stimuli does not hold,
+significant statistical and measurement issues can arise Specifically,
 the BTL model may overestimate the trait's reliability, that is, the
@@
 magnitude of these biases remain unpredictable because they depend on
-which judges researchers exclude from the analysis. Researchers define
-\emph{misfit judges} as those whose evaluations deviate substantially
-from the shared consensus due to the unique characteristics of either
-the stimuli or the judges themselves
+which judges researchers exclude from the analysis. \emph{Misfit judges}
+are those whose evaluations deviate substantially from the shared
+consensus due to the unique characteristics of either the stimuli or the
+judges themselves
 \citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.
@@
 
-Building on the previous section, it is clear that, despite its
-limitations, researchers typically rely on the BTL model to measure a
-trait and place the compared stimuli along its continuum
-\citep[pp.~269]{Thurstone_1927b}. Additionally, the CJ literature shows
-that researchers frequently use point estimates of BTL scores or their
-transformations to conduct further analyses or hypothesis tests. For
-example, researchers have used these scores to identify `misfit' judges
-and stimuli
+Building on the previous section, it is clear that, researchers
+typically rely on the BTL model to measure a trait and place the
+compared stimuli along its continuum \citep[pp.~269]{Thurstone_1927b}.
+Additionally, the CJ literature shows that researchers frequently use
+point estimates of BTL scores or their transformations to conduct
+further analyses or hypothesis tests. For example, researchers have used
+these scores to identify `misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
@@
 type-II errors \citep{McElreath_2020}.
-Figure~\ref{fig-measurement_error} illustrates these issues, showing how
-neglecting measurement error in an outcome can reduce the precision of a
-predictor' estimated effect.
+Figure~\ref{fig-measurement_error} illustrates these issues,
+demonstrating how neglecting measurement error by relying only on
+outcome averages can reduce the precision of a predictor's estimated
+effect.
 
@@
 
-A motivating example that will appear throughout the document clarifies
-these concepts. In this example, researchers aim to determine: ``To what
-extent do different teaching methods influence students' ability to
-produce high-quality written texts?'' To investigate this, a researcher
-designs a CJ experiment by randomly assigning students (individuals) to
-two groups, each receiving a different teaching method. Judges then
-compare pairs of students' written texts (stimuli) to produce a
-dichotomous outcome reflecting the relative quality of each text
-(trait). Based on this setup, researchers can reformulate the research
-question as the estimand: ``\emph{On average}, is there a difference in
-the ability to produce high-quality written texts between the two groups
-of students?''. Following current CJ practices, researchers then rely on
-estimates from the BTL model, or its transformations, to approximate
-this estimand.
+A motivating example that will appear in the rest of the document
+clarifies these concepts. In this example, researchers aim to determine:
+``To what extent do different teaching methods influence students'
+ability to produce high-quality written texts?'' To investigate this, a
+researcher designs a CJ experiment by randomly assigning students
+(individuals) to two groups, each receiving a different teaching method.
+Judges then compare pairs of students' written texts (stimuli) to
+produce a dichotomous outcome reflecting the relative quality of each
+text (trait). Based on this setup, researchers can reformulate the
+research question as the estimand: ``\emph{On average}, is there a
+difference in the ability to produce high-quality written texts between
+the two groups of students?''. Following current CJ practices,
+researchers then rely on estimates from the BTL model, or its
+transformations, to approximate this estimand.
 
@@
 their analysis \citep{McElreath_2020}. Second, they depict causal
-relationships in an interactive, non-parametric way. This flexibility
-enables feasible identification strategies without requiring
-specification of the types of variables, the functional forms relating
-them, or the parameters of those functional forms
-\citep[pp.~35]{Pearl_et_al_2016}.
+relationships in a non-parametric way. This flexibility enables feasible
+identification strategies without requiring specification of the types
+of variables, the functional forms relating them, or the parameters of
+those functional forms \citep[pp.~35]{Pearl_et_al_2016}.
 
@@
 \emph{conceptual-population model}. This model assumes that researchers
-have access to a ``conceptual population'' of comparative data, that is,
-data representing all repeated judgments made by every available judge
-for each pair of stimuli produced by each pair of individuals in the
-population. Conversely, Section~\ref{sec-theory-theoretical_SC}
+have access to a \emph{conceptual population} of comparative data, that
+is, data representing all repeated judgments made by every available
+judge for each pair of stimuli produced by each pair of individuals in
+the population. Conversely, Section~\ref{sec-theory-theoretical_SC}
 integrates the assessment design features into what we refer to as the
@@
 model}\label{sec-theory-theoretical_P}
+
+In the conceptual-population model, we assume an idealized scenario
+where researchers have access to a \emph{conceptual population} of
+comparative data --data representing all repeated judgments made by each
+judge for every stimulus pair produced by each pair of individuals in
+the population. This assumption allows us to integrate Thurstone's
+theoretical principles and propose innovations to address some of the
+issues raised in Section~\ref{sec-theory-issues}.
+
+\subsubsection{Integrating the first theoretical
+principles}\label{sec-theory-theoretical_P1}
 
@@
 (conditional) independence.
+
+SCM~\ref{fig-cj03_scm} illustrates the first theoretical principles of
+the conceptual-population model designed to examine the impact of
+different teaching methods on students' writing ability. This SCM
+outlines the relationship between the conceptual-population outcome
+\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
+\(h\) and \(i\) label the students who authored the texts (individuals),
+while the indices \(a\) and \(b\) represent the compared texts
+(stimuli). The index \(j\) refers to the judge, and \(k\) represents the
+judgment index, accounting for experimental conditions where judges
+perform the same judgment multiple times, i.e., a \emph{repeated
+measures design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing
+system facilitates comparisons between texts written by the same student
+(\(h = i\), \(a \neq b\)) and between texts written by distinct students
+(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
+where judges compare a student's text to itself (\(h = i\), \(a = b\)),
+as such comparison lacks practical relevance within the CJ framework.
+
+In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
+texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
+difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
+Additionally, based on the arguments developed in
+Section~\ref{sec-theory-issue1b} and the recommendations of
+\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
+the judges' biases \((B_{kj})\). Together with the outcome, these
+variables constitute the preliminary set of endogenous variables,
+\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
+the SCM presents the preliminary set of structural equations,
+\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
+among these variables.
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
+  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj03_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
+
+}
+
+\subcaption{\label{fig-cj03_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj03}Conceptual-population model, scalar form.}
+
+\end{figure}%
 
@@
 
-\subsubsection{Integrating the first theoretical
-principles}\label{sec-theory-theoretical_P1}
-
-SCM~\ref{fig-cj03_scm} illustrates the first theoretical principles of
-the conceptual-population model designed to examine the impact of
-different teaching methods on students' writing ability. This SCM
-outlines the relationship between the conceptual-population outcome
-\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
-\(h\) and \(i\) label the students who authored the texts (individuals),
-while the indices \(a\) and \(b\) represent the compared texts
-(stimuli). The index \(j\) refers to the judge, and \(k\) represents the
-judgment index, accounting for experimental conditions where judges
-perform the same judgment multiple times, i.e., a \emph{repeated
-measures design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing
-system facilitates comparisons between texts written by the same student
-(\(h = i\), \(a \neq b\)) and between texts written by distinct students
-(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
-where judges compare a student's text to itself (\(h = i\), \(a = b\)),
-as such comparison lacks practical relevance within the CJ framework.
-
-In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
-texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
-difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
-Additionally, based on the arguments developed in
-Section~\ref{sec-theory-issue1b} and the recommendations of
-\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
-the judges' biases \((B_{kj})\). Together with the outcome, these
-variables constitute the preliminary set of endogenous variables,
-\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
-the SCM presents the preliminary set of structural equations,
-\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
-among these variables.
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
-  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj03_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
-
-}
-
-\subcaption{\label{fig-cj03_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj03}Conceptual-population model, scalar form.}
-
-\end{figure}%
-
-Moreover, Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
-SCM~\ref{fig-cj03_scm}, illustrating the expected causal outlined in
-Thurstone's theory. The graph shows that the discriminal processes of
-the texts \((T_{ha}, T_{ib})\) influence their discriminal difference
-\((D_{hiabjk})\), which in turn determines the outcome
+Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
+SCM~\ref{fig-cj03_scm}, illustrating the expected causal relationships
+outlined in Thurstone's theory. The graph shows that the discriminal
+processes of the texts \((T_{ha}, T_{ib})\) influence their discriminal
+difference \((D_{hiabjk})\), which in turn determines the outcome
 \((O^{cp}_{hiabjk})\). It also highlights the influence of judges'
@@
 differentiates between observed endogenous variables, such as the
-outcome, and latent endogenous variables, including the texts'
-discriminal processes, their discriminal difference, and the judges'
-biases.
-
-\subsubsection{The conceptual-population data
-structure}\label{sec-theory-theoretical_P2}
+outcome (solid black circle), and latent endogenous variables, including
+the texts' discriminal processes, their discriminal difference, and the
+judges' biases (open circles).
+
+\subsubsection{\texorpdfstring{The \emph{conceptual population} data
+structure}{The conceptual population data structure}}\label{sec-theory-theoretical_P2}
 
 Although specifying a system's data structure is not mandatory when
-using SCMs and DAGs, defining one can improve clarity and facilitate the
-description of the system. Thus, to re-express the scalar form of the CJ
-system outlined in the previous section into an equivalent vectorized
-form, we first define the vectors \(I\) and \(J\), along with the
-matrices \(IA\) and \(JK\), as follows:
-
-\begin{equation}\phantomsection\label{eq-mat12}{
-I_{n_{I}} = \begin{bmatrix}
+using SCMs and DAGs, defining one in this case can improve clarity and
+facilitate the description of the system. Thus, to re-express the scalar
+form of the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
+vectorized form, we first define the vectors \(I\) and \(J\), along with
+the matrices \(IA\) and \(JK\), as follows:
+
+\begin{equation}\phantomsection\label{eq-mat11}{
+I = \begin{bmatrix}
 1 \\
@@
 specifies the number of stimuli available per individual. Likewise, each
-row of \(JK\) associates a judge \(j\) with a judgment index \(k\),
-where \(n_{K}\) indicates the number of judgments each judge makes.
+row of \(JK\) associates a judge \(j\) with a judgment index \(k\).
 Consequently, the matrix \(JK\) has \((n_{J} \cdot n_{K})\) rows and
-\(2\) columns.
-
-Additionally, we define the matrix \(RPUV\) that assigns each row from
-the \(IA\) matrix to a corresponding row from the \(JK\) matrix. Thus,
-the matrix consists of
-\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
-\(6\) columns. Here, the term \({n_{I} \cdot n_{A} \choose 2}\)
-represents the Binomial coefficient, which quantifies the total number
-of repeated judgments each judge makes for every pair of stimuli
-generated by each pair of individuals in the population. The matrix is
-defined as follows:
-
-\begin{equation}\phantomsection\label{eq-mat3}{
-RPUV_{n \times 6} = \begin{bmatrix}
+\(2\) columns, where \(n_{K}\) indicates the number of judgments each
+judge makes.
+
+Additionally, we construct the matrix \(R\) to map each row of the
+\(IA\) matrix with a corresponding row from the \(JK\) matrix. Thus,
+this matrix has \(n\) rows and \(6\) columns, where
+\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
+term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
+coefficient, which quantifies the total number of unique comparisons
+possible between every pair of stimuli generated by each pair of
+individuals in the population. Whe define the matrix as follows:
+
+\begin{equation}\phantomsection\label{eq-mat12}{
+R = \begin{bmatrix}
 1 & 1 & 1 & 2 & 1 & 1 \\
@@
 \(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
-matrices described in equations (\ref{eq-mat12}) and (\ref{eq-mat3})
+matrices described in equations (\ref{eq-mat11}) and (\ref{eq-mat12})
 take the following form:
@@
 \[
-I_{5} = \begin{bmatrix}
+I = \begin{bmatrix}
 1 \\
@@
 
-Then, using equations (\ref{eq-mat12}) and (\ref{eq-mat3}), we can
+Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
 re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
 equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
-depiction, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
-difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
-judges' biases \(B_{JK}\) are represented as vectors rather than scalar
-values. These vectors capture all the observations from the conceptual
-population. Specifically, \(O^{cp}_{RPUV}\) is an observed vector of
-length \(n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), and
-\(D_{RPUV}\) is a latent vector of the same dimension. Moreover,
-\(T_{IA}\) and \(B_{JK}\) are latent vectors of lengths
-\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\), respectively.
+depiction, the outcome \(O^{cp}_{R}\), the texts' discriminal difference
+\(D_{R}\), their discriminal processes \(T_{IA}\), and the judges'
+biases \(B_{JK}\) are represented as vectors rather than scalar values.
+These vectors capture all the observations from the conceptual
+population. Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and
+latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
+\(B_{JK}\) are latent vectors of lengths \((n_{I} \cdot n_{A})\) and
+\((n_{J} \cdot n_{K})\), respectively.
 
@@
 
-Building on the principles of Structural Equation Modeling (SEM)
+The conceptual-population model integrates two \emph{hierarchical
+structural components} to examine how different teaching methods
+influence students' writing ability. This approach builds on the
+principles of Structural Equation Modeling (SEM)
 \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
-\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
-the conceptual-population model incorporates two \emph{hierarchical
-structural components}. A structural component describes how different
-observed or latent variables influence the primary latent variable of
-interest \citep{Everitt_et_al_2010}. Moreover, their hierarchical nature
-enable analyses when the stimuli exhibit a hierarchical structure (refer
-to Section~\ref{sec-theory-issue1b}) and when researchers aim to test
-hypotheses while accounting for the uncertainties associated with the
-traits (see Section~\ref{sec-theory-issue2}).
-
-Figure~\ref{fig-cj09} displays these components applied to examine the
-effect of different teaching methods on students' writing ability. The
-top branch of DAG~\ref{fig-cj09_dag} depicts the first component. This
-component describes how the students' writing-quality trait \((T_{I})\)
-and a set of \emph{relevant} text-related variables \((X_{IA})\), such
-as text length, causally influence the texts' written-quality trait
-\((T_{IA})\). The DAG also highlights the idiosyncratic errors
-\((e_{IA})\), which capture variations in the texts' trait that neither
-the students' trait nor other relevant text-related variables can
-explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors of length
-\(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively. Additionally,
-\(X_{IA}\) is an observed matrix with a dimension of
-\((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns.
-
-Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
-set of \emph{relevant} student-related variables \((X_{I})\), including
-teaching methods, causally influence the students' writing-quality trait
+\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}.
+Each structural component defines how observed or latent variables
+affect the primary latent variable of interest
+\citep{Everitt_et_al_2010}. Their hierarchical nature enables
+researchers to test hypotheses while accounting for uncertainties
+associated with trait estimation (see Section~\ref{sec-theory-issue2}),
+particularly when stimuli exhibit a hierarchical structure (see
+Section~\ref{sec-theory-issue1b}).
+
+The top branch of DAG~\ref{fig-cj09_dag} depicts the first component.
+This component describes how the students' writing-quality trait
+\((T_{I})\) and a set of \emph{relevant} text-related variables
+\((X_{IA})\) (e.g., text length) causally influence the texts'
+written-quality trait \((T_{IA})\). The DAG also highlights the
+idiosyncratic errors \((e_{IA})\), which capture variations in the
+texts' trait that neither the students' trait nor other relevant
+text-related variables can explain. In this structure, \(T_{I}\) and
+\(e_{IA}\) are latent vectors of length \(n_{I}\) and
+\((n_{I} \cdot n_{A})\), respectively. Additionally, \(X_{IA}\) is an
+observed matrix with a dimension of \((n_{I} \cdot n_{A})\) rows and
+\(q_{IA}\) columns (variables).
+
+Additionally, the DAG's top branch illustrates how a set of
+\emph{relevant} student-related variables \((X_{I})\), such as the
+teaching method, causally influence the students' writing-quality trait
 \((T_{I})\). The DAG also includes the idiosyncratic errors \((e_{I})\),
@@
 student-related variables cannot explain. Here, \(e_{I}\) is a latent
-vector of length \(n_{I}\), and \(X_{I}\) is an observed matrix with
-\(n_{I}\) rows and \(q_{I}\) columns.
+vector of length \(n_{I}\), and \(X_{I}\) can be an observed matrix with
+\(n_{I}\) rows and \(q_{I}\) columns (variables).
+
+Here, \emph{relevant} refers to variables that satisfy the
+\emph{backdoor criterion} \citep[pp 37]{Neal_2020}, that is, they belong
+to a \emph{sufficient adjustment set}
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
+\emph{sufficient} set (potentially empty) blocks all non-causal paths
+between a predictor and an outcome without opening new ones
+\citep{Pearl_2009}.
 
@@
 \((B_{J})\) and a set of \emph{relevant} judgment-related variables
-\((Z_{JK})\), such as number of judgments or time for judgment, causally
+\((Z_{JK})\) (e.g., the number of judgments a judge makes) causally
 influence the biases associated with each text \(B_{JK}\). The DAG also
@@
 
-Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
-set of \emph{relevant} judge-related variables \((Z_{J})\), including
-judgment expertise, causally influence the judges' bias \((B_{J})\). The
-DAG also accounts for idiosyncratic errors \((e_{J})\), capturing the
-variations in the judges' bias that relevant judge-related variables
-cannot explain. Here, \(e_{J}\) is a latent vector of length \(n_{J}\),
-and \(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
-columns.
+Moreover, the DAG's bottom branch describes how a set of \emph{relevant}
+judge-related variables \((Z_{J})\), such as judgment expertise,
+causally influence the judges' bias \((B_{J})\). The DAG also accounts
+for idiosyncratic errors \((e_{J})\), capturing the variations in the
+judges' bias that relevant judge-related variables cannot explain. Here,
+\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
+observed matrix with \(n_{J}\) rows and \(q_{J}\) columns.
 
@@
 form by introducing key innovations to address the limitations discussed
-in Section~\ref{sec-theory-issues}. These enhancements include
-accounting for judges' biases and integrating hierarchical structural
-components. Nevertheless, despite its promise of enhancing measurement
-accuracy and precision, the model still depends on the unrealistic
-assumption that researchers have access to a \emph{conceptual
-population} of comparative data, encompassing all repeated judgments
-from every judge for each pair of stimuli produced by every individual
-in the population. Since researchers rarely meet this assumption in
-practice, they must consider a more realistic scenario. The extensions
-to the conceptual-population model, derived from this realistic
-scenario, are concisely described in the sample-comparison model.
+in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
+These enhancements include accounting for judges' biases and integrating
+hierarchical structural components. Nevertheless, despite its promise of
+enhancing measurement accuracy and precision, the model still depends on
+the unrealistic assumption that researchers have access to a
+\emph{conceptual population} of comparative data, encompassing all
+repeated judgments from every judge for each pair of stimuli produced by
+every individual in the population. Since researchers rarely meet this
+assumption in practice, they must consider a more realistic scenario.
+The extensions to the conceptual-population model, derived from this
+realistic scenario, are concisely described in the sample-comparison
+model.
 
@@
 
+In the sample-comparison model, we assume a more realistic scenario than
+the conceptual-population model. Specifically, we assume researchers
+have access to a data sample consisting of a limited number of judgments
+\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
+number of texts \((n^{s}_{A})\) from a sample of students
+\((n^{s}_{I})\), drawn from the conceptual population. Additionally, we
+assume that judges do not make \emph{all possible comparisons} allowed
+by the sample, but rather only \(n_{C}\) comparisons.
+
 \subsubsection{The sample mechanism}\label{sec-theory-theoretical_SC1}
 
-In the sample-comparison model, we assume a more realistic scenario
-compared to the conceptual-population model. Specifically, we assume
-that researchers have access to \emph{all comparisons} from a limited
-number of judgments \((n^{s}_{K})\) from a sample of judges
-\((n^{s}_{J})\), along with a specific number of texts \((n^{s}_{A})\)
-from a sample of students \((n^{s}_{I})\) drawn from the conceptual
-population. Figure~\ref{fig-cj12} illustrates this model, incorporating
-only this sample mechanism.
-
-Following the convention of \citet[pp.~499-516]{McElreath_2020} and
+To facilitate the interpretation of the sample-comparison model, we
+first define the data sample using several binary vector variables:
+\(S_{I}\), \(S_{J}\), \(S_{IA}\), \(S_{JK}\), and \(S\) as follows.
+
+\begin{equation}\phantomsection\label{eq-mat21}{
+S_{I} = \begin{bmatrix}
+i_{(1)} \\
+i_{(2)} \\
+\vdots \\
+i_{(h)} \\
+\vdots \\
+i_{(i)} \\
+\vdots \\
+i_{(nI)}
+\end{bmatrix} ; \;
+S_{J} = \begin{bmatrix}
+j_{(1)} \\
+j_{(2)} \\
+\vdots \\
+j_{(j)} \\
+\vdots \\
+j_{(nJ)}
+\end{bmatrix} ; \;
+S_{IA} = \begin{bmatrix}
+ia_{(1)} \\
+ia_{(2)} \\
+\vdots \\
+ia_{(ia)} \\
+\vdots \\
+ia_{(nI \cdot nA)} \\
+\end{bmatrix} ; \;
+S_{JK} = \begin{bmatrix}
+jk_{(1)} \\
+jk_{(2)} \\
+\vdots \\
+jk_{(jk)} \\
+\vdots \\
+jk_{(nJ \cdot nK)} \\
+\end{bmatrix}; \;
+S = \begin{bmatrix}
+s_{(1)} \\
+s_{(2)} \\
+\vdots \\
+s_{(s)} \\
+\vdots \\
+s_{(n)} \\
+\end{bmatrix}
+}\end{equation}
+
+Each element of the vectors \(S_{I}\) and \(S_{J}\) is a binary value
+that indicates the presence \((1)\) or absence \((0)\) of data in the
+\(I\) and \(J\) vectors, as defined in equation (\ref{eq-mat11}).
+Therefore, the vector \(S_{I}\) contains \(n_{I}\) elements, and the
+vector \(S_{J}\) contains \(n_{J}\) elements. For example, we define
+each element of the vector \(S_{I}\) as in equation (\ref{eq-mat22}),
+and we apply a similar definition to \(S_{J}\).
+
+\begin{equation}\phantomsection\label{eq-mat22}{
+i_{(i)} = \begin{cases} 
+1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
+0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
+\end{cases}
+}\end{equation}
+
+Each element of the vectors \(S_{IA}\) and \(S_{JK}\) is defined like
+\(i_{(i)}\) with vector \(I\), but applied to data in the matrices
+\(IA\) and \(JK\), as defined in equation (\ref{eq-mat11}). Therefore,
+the vector \(S_{IA}\) contains \(n_{I} \cdot n_{A}\) elements, and the
+vector \(S_{JK}\) contains \(n_{J} \cdot n_{K}\) elements. For example,
+we define each element of the vector \(S_{IA}\) as in equation
+(\ref{eq-mat23}), and we apply a similar definition to \(S_{JK}\).
+
+\begin{equation}\phantomsection\label{eq-mat23}{
+ia_{(ia)} = \begin{cases} 
+1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
+0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
+\end{cases}
+}\end{equation}
+
+The equation shows that the observed values in the \(S_{I}\) and
+\(S_{J}\) vectors determine the binary values in the \(S_{IA}\) and
+\(S_{JK}\) vectors, through the \(IA\) and \(JK\) matrices,
+respectively. Specifically, researchers cannot observe stimuli and
+judgments for individuals and judges they did not sample initially.
+
+Finally, the vector \(S\) in equation (\ref{eq-mat21}) contains
+\(n = \binom{n_{I} \cdot n_{A}}{2} \cdot n_{J} \cdot n_{K}\) elements,
+with each element defined in equation (\ref{eq-mat24}). This equation
+shows that the observed values in \(S_{IA}\) and \(S_{JK}\) determine
+the binary values in \(S\) through the \(R\) matrix. Specifically,
+researchers can only observe a subset of judges' comparisons for a
+subset of stimuli generated by a subset of individuals.
+
+\begin{equation}\phantomsection\label{eq-mat24}{
+s_{(s)} = \begin{cases} 
+1 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are observed} \\
+0 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are missing}
+\end{cases}
+}\end{equation}
+
+With the definition of \(S\), we can now integrate the sample mechanism
+into the conceptual-population model. Figure~\ref{fig-cj12} illustrates
+this model. Following the convention of
+\citet[pp.~499-516]{McElreath_2020} and
 \citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} depicts
-the conceptual-population outcome \(O^{cp}_{RPUV}\) as an unobserved
+the conceptual-population outcome \(O^{cp}_{R}\) as an unobserved
 variable, emphasizing the researchers' inability to access such an
-outcome due to the sampling mechanism. Moreover, the DAG shows the
-conditioned \emph{sample design} variable \(S\), which causally
-influence the sample-comparison outcome \(O^{sc}_{RPUV}\) . This
-variable reflects how this sample-comparison outcome is a sample drawn
-from the conceptual-population outcome \(O^{cp}_{RPUV}\). Specifically,
-the variable is a vector \(S = \{ s_{1}, \dots, s_{n} \}\), where \(n\)
-corresponds to the number of rows in the matrix \(RPUV\) (refer to
-Section~\ref{sec-theory-theoretical_P2}), and each \(s_{i}\) is a binary
-variable indicating the presence \((1)\) or absence \((0)\) of data in
-the corresponding positions of the conceptual-population outcome
-\(O^{cp}_{RPUV}\). Specifically,
-
-\[
-s_i = \begin{cases} 
-1 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is observed} \\
-0 & \text{if the data for the } i^{th} \text{ element of } O^{cp}_{RPUV} \text{ is missing}
-\end{cases}
-\]
+outcome due to the sampling mechanism. Additionally, the DAG depicts the
+conditioned \emph{sample design} variable \(S\) causally influencing the
+sample-comparison outcome \(O^{sc}_{R}\). This variable shows that the
+sample-comparison outcome is a subset drawn from the
+conceptual-population outcome \(O^{cp}_{R}\).
 
@@
 DAG~\ref{fig-cj14_dag} depicts the sample-comparison outcome
-\(O^{sc}_{RPUV}\) as an unobserved variable, emphasizing the
-researchers' inability to access such an outcome due to the comparison
-mechanism. Moreover, the DAG shows the conditioned \emph{comparison
-design} variable \(C\), which causally influence the observed outcome
-\(O_{RPUV}\). This variable reflects how this observed outcome is a
-sample drawn from the sample-comparison outcome \(O^{sp}_{RPUV}\).
-Specifically, the variable is a vector
-\(C = \{ c_{1}, \dots, c_{n} \}\), where \(n\) corresponds to the number
-of rows in the matrix \(RPUV\) (refer to
+\(O^{sc}_{R}\) as an unobserved variable, emphasizing the researchers'
+inability to access such an outcome due to the comparison mechanism.
+Moreover, the DAG shows the conditioned \emph{comparison design}
+variable \(C\), which causally influence the observed outcome \(O_{R}\).
+This variable reflects how this observed outcome is a sample drawn from
+the sample-comparison outcome \(O^{sp}_{R}\). Specifically, the variable
+is a vector \(C = \{ c_{1}, \dots, c_{n} \}\), where \(n\) corresponds
+to the number of rows in the matrix \(R\) (refer to
 Section~\ref{sec-theory-theoretical_P2}), and each \(c_{i}\) is a binary
@@
 the corresponding positions of the sample-comparison outcome
-\(O^{sp}_{RPUV}\). Specifically,
+\(O^{sp}_{R}\). Specifically,
 

### 2025-03-27 paper2.qmd
section:all; process:write

@@
       non-dropping-particle: van
-    orcid: https://orcid.org/0000-0001-9398-9775
+    orcid: 0000-0001-9398-9775
     url: https://www.uantwerpen.be/en/staff/tine-vandaal/
@@
 Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli.
-While these assumptions simplify the trait measurement process, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
+While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
 Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes.
-Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.
+Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.
 
@@
 
-  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, particularly stressing the need for bespoke CJ models tailored to the specific assumptions of the experiments and data.
+  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, particularly stressing the need for bespoke CJ models tailored to the experiments and data assumptions.
 It also lays the foundation for broader applications, encouraging researchers across the social sciences to adopt more robust and interpretable methodologies.
@@
 Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli.
-While these assumptions simplify statistical modeling, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
+While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
 Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes.
-Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.
+Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.
 
@@
 <!-- 2. Thurstone's Cases and their simplifying assumptions -->
-To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions into the model.
+To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions.
 In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons.
@@
 <!-- 3. But does it hold in practice?  -->
-Notably, while the assumption of equal dispersions simplifies the trait measurement process, evidence from the CJ literature suggests that the BTL model may fail to capture the complexity of modern CJ data.
+Notably, while the assumption of equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that the BTL model may fail to capture the complexity of modern CJ data.
 Specifically, the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
@@
 @fig-discriminal_process illustrates how an increased discriminal dispersion can widen the overlap between two texts, leading to more judgment discrepancies.
-In the figure, text B has a broader discriminal dispersion than text A, expanding the overlap between their discriminal processes, a space where judges struggle to determine which text possesses more of the trait.
+In the figure, Text B has a broader discriminal dispersion than Text A, expanding the overlap between their discriminal processes, a space where judges struggle to determine which text possesses more of the trait.
 
@@
 <!-- 5. So, what happens when the assumption does not hold? -->
-When the assumption of equal dispersions between stimuli does not hold, significant statistical and measurement issues can arise Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
+Significant statistical and measurement issues can arise when the assumption of equal dispersions between stimuli does not hold.
+Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
 This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp.
@@
 370] and, by extension, the underlying discriminal processes of the stimuli.
-This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL model follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $(\rho \neq 0)$.
+This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$.
 
@@
 
-Finally, as discussed in the previous section, removing judges based on misfit statistics risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Finally, as discussed in the previous section, removing *misfit* judges based risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
@@
 Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
-@fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.
+@fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error $(\sigma_{T})$ by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.
 
@@
 
-![The effect of outcome uncertainty on the estimation of an effect $\beta$ linked to the predictor.
-Uncertainty increases from left to right.](/images/png/measurement_error.png){#fig-measurement_error width=100%}
+![The effect of outcome uncertainty $\sigma_{T}$ on the estimation of an effect $\beta_{X}$ linked to the predictor.
+The example assumes a sample size $n=100$ and an uncertainty that increases from left to right.](/images/png/measurement_error.png){#fig-measurement_error width=100%}
 
@@
 Luckily, SCMs and DAGs support identification analysis through two key advantages.
-First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
+First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks.
+This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@Neal_2020; @McElreath_2020].
 Second, they depict causal relationships in a non-parametric way.
@@
 <!-- 1. The SCMs -->
-Before incorporating the theoretical principles into the *conceptual-population model*, it is essential to further define SCMs.
+Before incorporating the first theoretical principles of Thurstone's theory, it is essential to further define SCMs.
 SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
@@
 <!-- 2. the preliminary SCM for CJ -->
-[SCM -@fig-cj03_scm] illustrates the first theoretical principles of the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability.
+[SCM -@fig-cj03_scm] illustrates the inclusion of the first theoretical principles into the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability.
 This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
@@
 366-376].
-Notably, the indexing system facilitates comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
-However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical relevance within the CJ framework.
+Notably, the indexing system facilitates comparisons between texts written by the same student $(h = i, a \neq b)$ and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
+However, it excludes cases where judges compare a student's text to itself $(h = i, a = b)$, as such comparison lacks practical relevance within the CJ framework.
 
@@
 32].
-Although DAGs often omit exogenous variables for simplicity, this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+Although DAGs often omit exogenous variables for simplicity, the DAGs presented in this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
-Whe define the matrix as follows:
+Thus, we define the matrix as follows:
 

### 2025-03-27 _tex/paper2.tex
Built site for gh-pages

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   1p]{elsarticle}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-harv}
 
@@
 \journal{Psychometrika}
-
-\usepackage[]{natbib}
-\bibliographystyle{elsarticle-harv}
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 discriminal dispersions and zero correlation between the stimuli. While
-these assumptions simplify the trait measurement process, they may fail
-to capture the complexity of CJ data, potentially leading to unreliable
-and inaccurate trait estimates. Second, the study highlights the
-apparent disconnect between CJ's trait measurement and hypothesis
-testing processes. Although separating these processes simplifies the
-analysis of CJ data, it may also undermine the reliability of various
-statistical results derived from the measurement.
+these assumptions simplify the trait measurement model, they may fail to
+capture the complexity of CJ data, potentially leading to unreliable and
+inaccurate trait estimates. Second, the study highlights the apparent
+disconnect between CJ's trait measurement and hypothesis testing
+processes. Although separating these processes simplifies the analysis
+of CJ data, it may also undermine the reliability of various statistical
+results derived from these processes.
 
@@
 contemporary empirical CJ research, particularly stressing the need for
-bespoke CJ models tailored to the specific assumptions of the
-experiments and data. It also lays the foundation for broader
-applications, encouraging researchers across the social sciences to
-adopt more robust and interpretable methodologies.
+bespoke CJ models tailored to the experiments and data assumptions. It
+also lays the foundation for broader applications, encouraging
+researchers across the social sciences to adopt more robust and
+interpretable methodologies.
 \end{abstract}
@@
 discriminal dispersions and zero correlation between the stimuli. While
-these assumptions simplify statistical modeling, they may fail to
+these assumptions simplify the trait measurement model, they may fail to
 capture the complexity of CJ data, potentially leading to unreliable and
@@
 it may also undermine the reliability of various statistical results
-derived from the measurement.
+derived from these processes.
 
@@
 theory, Thurstone developed five cases derived from this general form,
-each progressively incorporating additional simplifying assumptions into
-the model. In Case I, Thurstone postulated that pairs of stimuli would
-maintain a constant correlation across all comparisons. In Case II, he
-allowed multiple judges to undertake comparisons instead of confining
+each progressively incorporating additional simplifying assumptions. In
+Case I, Thurstone postulated that pairs of stimuli would maintain a
+constant correlation across all comparisons. In Case II, he allowed
+multiple judges to undertake comparisons instead of confining
 evaluations to a single judge. In Case III, he posited that there was no
@@
 Notably, while the assumption of equal dispersions simplifies the trait
-measurement process, evidence from the CJ literature suggests that the
-BTL model may fail to capture the complexity of modern CJ data.
+measurement model, evidence from the CJ literature suggests that the BTL
+model may fail to capture the complexity of modern CJ data.
 Specifically, the assumption of equal dispersions may not hold when
@@
 discriminal dispersion can widen the overlap between two texts, leading
-to more judgment discrepancies. In the figure, text B has a broader
-discriminal dispersion than text A, expanding the overlap between their
+to more judgment discrepancies. In the figure, Text B has a broader
+discriminal dispersion than Text A, expanding the overlap between their
 discriminal processes, a space where judges struggle to determine which
@@
 
-When the assumption of equal dispersions between stimuli does not hold,
-significant statistical and measurement issues can arise Specifically,
-the BTL model may overestimate the trait's reliability, that is, the
-degree to which the outcome accurately reflects the \emph{true}
-discriminal differences between stimuli. This overestimation, in turn,
-results in spurious conclusions about these differences
-\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
-discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
-illustrates this scenario when the model's discriminal difference
+Significant statistical and measurement issues can arise when the
+assumption of equal dispersions between stimuli does not hold.
+Specifically, the BTL model may overestimate the trait's reliability,
+that is, the degree to which the outcome accurately reflects the
+\emph{true} discriminal differences between stimuli. This
+overestimation, in turn, results in spurious conclusions about these
+differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
+underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
+also illustrates this scenario when the model's discriminal difference
 distribution aligns with the thick continuous line for
@@
 Figure~\ref{fig-correlation} when the discriminal difference
-distribution of the BTL model follows the thick continuous line
+distribution of the BTL scores follows the thick continuous line
 \((\rho = 0)\), while the \emph{true} discriminal difference follows any
-discontinuous line where \((\rho \neq 0)\).
-
-Finally, as discussed in the previous section, removing judges based on
-misfit statistics risks discarding valuable information and even
-introduce bias into the trait estimates
+discontinuous line where \(\rho \neq 0\).
+
+Finally, as discussed in the previous section, removing \emph{misfit}
+judges based risks discarding valuable information and even introduce
+bias into the trait estimates
 \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
@@
 Figure~\ref{fig-measurement_error} illustrates these issues,
-demonstrating how neglecting measurement error by relying only on
-outcome averages can reduce the precision of a predictor's estimated
-effect.
+demonstrating how neglecting measurement error \((\sigma_{T})\) by
+relying only on outcome averages can reduce the precision of a
+predictor's estimated effect.
 
@@
 \caption{\label{fig-measurement_error}The effect of outcome uncertainty
-on the estimation of an effect \(\beta\) linked to the predictor.
-Uncertainty increases from left to right.}
+\(\sigma_{T}\) on the estimation of an effect \(\beta_{X}\) linked to
+the predictor. The example assumes a sample size \(n=100\) and an
+uncertainty that increases from left to right.}
 
@@
 advantages. First, regardless of complexity, they can represent various
-causal structures using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
-decompose complex structures into manageable components, facilitating
-their analysis \citep{McElreath_2020}. Second, they depict causal
+causal structures using only five fundamental building blocks. This
+feature allows researchers to decompose complex structures into
+manageable components, facilitating their analysis
+\citep{Neal_2020, McElreath_2020}. Second, they depict causal
 relationships in a non-parametric way. This flexibility enables feasible
@@
 
-Before incorporating the theoretical principles into the
-\emph{conceptual-population model}, it is essential to further define
-SCMs. SCMs are formal mathematical models characterized by a set of
-\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
-\(E\), and a set of functions \(F\)
-\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
-whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
-In contrast, exogenous variables represent \emph{errors} or
-\emph{disturbances} arising from omitted factors that the investigator
-chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
-the functions, referred to as \emph{structural equations}, express the
-endogenous variables as non-parametric functions of other endogenous and
-exogenous variables. These functions use the symbol `\(:=\)' to denote
-the asymmetrical causal dependence between variables and the symbol
-`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
-(conditional) independence.
-
-SCM~\ref{fig-cj03_scm} illustrates the first theoretical principles of
-the conceptual-population model designed to examine the impact of
-different teaching methods on students' writing ability. This SCM
-outlines the relationship between the conceptual-population outcome
-\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
-\(h\) and \(i\) label the students who authored the texts (individuals),
-while the indices \(a\) and \(b\) represent the compared texts
-(stimuli). The index \(j\) refers to the judge, and \(k\) represents the
-judgment index, accounting for experimental conditions where judges
-perform the same judgment multiple times, i.e., a \emph{repeated
-measures design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing
-system facilitates comparisons between texts written by the same student
-(\(h = i\), \(a \neq b\)) and between texts written by distinct students
+Before incorporating the first theoretical principles of Thurstone's
+theory, it is essential to further define SCMs. SCMs are formal
+mathematical models characterized by a set of \emph{endogenous}
+variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
+functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
+variables are those whose causal mechanisms a researcher chooses to
+model \citep{Neal_2020}. In contrast, exogenous variables represent
+\emph{errors} or \emph{disturbances} arising from omitted factors that
+the investigator chooses not to model explicitly
+\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
+\emph{structural equations}, express the endogenous variables as
+non-parametric functions of other endogenous and exogenous variables.
+These functions use the symbol `\(:=\)' to denote the asymmetrical
+causal dependence between variables and the symbol `\(\:\bot\:\)' to
+represent \emph{d-separation}, a concept akin to (conditional)
+independence.
+
+SCM~\ref{fig-cj03_scm} illustrates the inclusion of the first
+theoretical principles into the conceptual-population model designed to
+examine the impact of different teaching methods on students' writing
+ability. This SCM outlines the relationship between the
+conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
+variables. The subscripts \(h\) and \(i\) label the students who
+authored the texts (individuals), while the indices \(a\) and \(b\)
+represent the compared texts (stimuli). The index \(j\) refers to the
+judge, and \(k\) represents the judgment index, accounting for
+experimental conditions where judges perform the same judgment multiple
+times, i.e., a \emph{repeated measures design}
+\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system
+facilitates comparisons between texts written by the same student
+\((h = i, a \neq b)\) and between texts written by distinct students
 (\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
-where judges compare a student's text to itself (\(h = i\), \(a = b\)),
-as such comparison lacks practical relevance within the CJ framework.
+where judges compare a student's text to itself \((h = i, a = b)\), as
+such comparison lacks practical relevance within the CJ framework.
 
@@
 defined by the values of the given variable \citep[pp.~32]{Neal_2020}.
-Although DAGs often omit exogenous variables for simplicity, this
-section includes exogenous variables to improve clarity and reveal
-potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}.
+Although DAGs often omit exogenous variables for simplicity, the DAGs
+presented in this section includes exogenous variables to improve
+clarity and reveal potential issues related to conditioning and
+confounding \citep{Cinelli_et_al_2020}.
 
@@
 possible between every pair of stimuli generated by each pair of
-individuals in the population. Whe define the matrix as follows:
+individuals in the population. Thus, we define the matrix as follows:
 

### 2025-03-28 paper2.qmd
section:4.2; process:write

@@
 DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
-They also display conditioned variables by surrounding them with a square.
-Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp.
-32].
 Although DAGs often omit exogenous variables for simplicity, the DAGs presented in this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
@@
 ### The *conceptual population* data structure {#sec-theory-theoretical_P2}
+
+<!-- recording: Sven and Tine 25.03.28; time: 00:10:00 - 00:11:00 -->
 
@@
 \end{bmatrix}
-$$
+$$ {#eq-mat13}
 
@@
 
+<!-- recording: Sven and Tine 25.03.28; time: 00:12:45 - 00:33:00 -->
+
 <!-- 1. What are we going to do? -->
-The conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability.
-This approach builds on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
+Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
 138] and Item Response Theory (IRT) [@Fox_2010, chap.
 6; @vanderLinden_et_al_2017_I, chap.
-24].
+24], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability.
 Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
@@
 <!-- 2 Explaining what we did -->
-The top branch of [DAG -@fig-cj09_dag] depicts the first component.
-This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$.
-The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain.
-In this structure, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively.
-Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns (variables).
-
-Additionally, the DAG's top branch illustrates how a set of *relevant* student-related variables $(X_{I})$, such as the teaching method, causally influence the students' writing-quality trait $(T_{I})$.
-The DAG also includes the idiosyncratic errors $(e_{I})$, which capture variations in the students' traits that the relevant student-related variables cannot explain.
-Here, $e_{I}$ is a latent vector of length $n_{I}$, and $X_{I}$ can be an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables).
-
-Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as the teaching method, causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
+It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the primary latent variable of interest.
+Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures unexplained variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$.
+Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
+Similarly, $X_{IA}$ is an observed matrix with dimension $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
+
+[^1]: Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
 A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
+While these topics are beyond the scope of this study, readers seeking a deeper understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, as well as introductory books like @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful.
+For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020, are recommended.
+
+[^2]: refer to footnote 1.
 
@@
 
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component.
-This component describes how the judges' bias $(B_{J})$ and a set of *relevant* judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$.
-The DAG also highlights the idiosyncratic errors $(e_{JK})$, which capture variations in the texts' bias that neither the judges' bias nor other relevant judgment-related variables can explain.
-In this description, $B_{J}$ and $e_{JK}$ are latent vectors of length $n_{J}$ and $(n_{J} \cdot n_{K})$, respectively.
-Additionally, $Z_{JK}$ is an observed matrix with a dimension of $(n_{J} \cdot n_{K})$ rows and $q_{JK}$ columns.
-
-Moreover, the DAG's bottom branch describes how a set of *relevant* judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the judges' bias $(B_{J})$.
-The DAG also accounts for idiosyncratic errors $(e_{J})$, capturing the variations in the judges' bias that relevant judge-related variables cannot explain.
-Here, $e_{J}$ is a latent vector of length $n_{J}$, and $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns.
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$.
+It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the primary latent variable of interest.
+Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures unexplained variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
+Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
+Similarly, $Z_{JK}$ is an observed matrix with dimension $(n_{J} \cdot n_{K})$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
+
+[^3]: refer to footnote 1.
+[^4]: refer to footnote 1.
 
@@
 
+<!-- recording: Sven and Tine 25.03.28; time: 00:09:00 - 00:10:00 -->
+<!-- recording: Sven and Tine 25.03.28; time: 00:11:30 - 00:12:25 -->
+
 <!-- Develop how you get the vectors S and C and sampling weights when you know the extension of the population. (See https://mc-stan.org/docs/stan-users-guide/poststratification.html ) -->
@@
 <!-- 1. start by defining the sampling vector and matrices  -->
-To facilitate the interpretation of the sample-comparison model, we first define the data sample using several binary vector variables: $S_{I}$, $S_{J}$, $S_{IA}$, $S_{JK}$, and $S$ as follows.
+To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the data sampling process using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:
 
@@
 jk_{(nJ \cdot nK)} \\
-\end{bmatrix}; \;
+\end{bmatrix}
+$$ {#eq-mat21}
+
+Where each element of the vectors $S_{I}$ and $S_{J}$ is a binary value indicating the presence or absence of data in the $I$ and $J$ vectors (see equation ([-@eq-mat11])).
+Thus, the vectors $S_{I}$ and $S_{J}$ contains $n_{I}$ and $n_{J}$ elements, respectively.
+Specifically, we define the vector $S_{I}$ as in equation ([-@eq-mat22]) and apply a similar definition to $S_{J}$.
+
+$$
+i_{(i)} = \begin{cases} 
+1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
+0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
+\end{cases}
+$$ {#eq-mat22}
+
+Similarly, each element of the vectors $S_{IA}$ and $S_{JK}$ is a binary value indicating the presence or absence of data in the matrices $IA$ and $JK$ (see equation ([-@eq-mat11])).
+Thus, the vectors $S_{IA}$ and $S_{JK}$ contains $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$ elements, respectively.
+Specifically, we define the vector $S_{IA}$ as in equation ([-@eq-mat23]) and apply a similar definition to $S_{JK}$.
+
+$$
+ia_{(ia)} = \begin{cases} 
+1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
+0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
+\end{cases}
+$$ {#eq-mat23}
+
+We can visualize these vectors more clearly using the example in equation ([-@eq-mat13]).
+Suppose the sample excludes the second student and the third judge.
+Additionally, assume that the second text from each individual and the third judgment from each judge are also missing.
+Under these conditions, the vectors have the following structure:
+
+$$
+S_{I} = \begin{bmatrix}
+1 \\
+0 \\
+1 \\
+1 \\
+1
+\end{bmatrix} ; \;
+S_{J} = \begin{bmatrix}
+1 \\
+1 \\
+0
+\end{bmatrix} ; \;
+S_{IA} = \begin{bmatrix}
+1 \\
+0 \\
+0 \\
+0 \\
+1 \\
+0 \\
+1 \\
+0 \\
+1 \\
+0 
+\end{bmatrix} ; \;
+S_{JK} = \begin{bmatrix}
+1 \\
+1 \\
+0 \\
+1 \\
+1 \\
+0 \\
+0 \\
+0 \\
+0 
+\end{bmatrix}
+$$ {#eq-mat24}
+
+Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$ vectors.
+In other words, researchers can only observe texts and judgments from students and judges initially included in the sample.
+Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$.
+Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all judges $(n^{s}_{J} \cdot n^{s}_{K})$.
+
+Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every component of $S_{JK}$.
+Essentially, this vector serves as the sampled equivalent of the $R$ matrix defined in equation ([-@eq-mat12]), where the sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
+
+$$
 S = \begin{bmatrix}
@@
 \end{bmatrix}
-$$ {#eq-mat21}
-
-Each element of the vectors $S_{I}$ and $S_{J}$ is a binary value that indicates the presence $(1)$ or absence $(0)$ of data in the $I$ and $J$ vectors, as defined in equation ([-@eq-mat11]).
-Therefore, the vector $S_{I}$ contains $n_{I}$ elements, and the vector $S_{J}$ contains $n_{J}$ elements.
-For example, we define each element of the vector $S_{I}$ as in equation ([-@eq-mat22]), and we apply a similar definition to $S_{J}$.
-
-$$
-i_{(i)} = \begin{cases} 
-1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
-0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
-\end{cases}
-$$ {#eq-mat22}
-
-Each element of the vectors $S_{IA}$ and $S_{JK}$ is defined like $i_{(i)}$ with vector $I$, but applied to data in the matrices $IA$ and $JK$, as defined in equation ([-@eq-mat11]).
-Therefore, the vector $S_{IA}$ contains $n_{I} \cdot n_{A}$ elements, and the vector $S_{JK}$ contains $n_{J} \cdot n_{K}$ elements.
-For example, we define each element of the vector $S_{IA}$ as in equation ([-@eq-mat23]), and we apply a similar definition to $S_{JK}$.
-
-$$
-ia_{(ia)} = \begin{cases} 
-1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
-0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
-\end{cases}
-$$ {#eq-mat23}
-
-The equation shows that the observed values in the $S_{I}$ and $S_{J}$ vectors determine the binary values in the $S_{IA}$ and $S_{JK}$ vectors, through the $IA$ and $JK$ matrices, respectively.
-Specifically, researchers cannot observe stimuli and judgments for individuals and judges they did not sample initially.
-
-Finally, the vector $S$ in equation ([-@eq-mat21]) contains $n = \binom{n_{I} \cdot n_{A}}{2} \cdot n_{J} \cdot n_{K}$ elements, with each element defined in equation ([-@eq-mat24]).
-This equation shows that the observed values in $S_{IA}$ and $S_{JK}$ determine the binary values in $S$ through the $R$ matrix.
-Specifically, researchers can only observe a subset of judges' comparisons for a subset of stimuli generated by a subset of individuals.
-
-$$
-s_{(s)} = \begin{cases} 
-1 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are observed} \\
-0 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are missing}
-\end{cases}
-$$ {#eq-mat24}
+$$ {#eq-mat25}
 
 <!-- 2. What is the effect of the sampling mechanism -->
-With the definition of $S$, we can now integrate the sample mechanism into the conceptual-population model.
-@fig-cj12 illustrates this model.
+With the definition of $S$, we incorporate the sample mechanism into the conceptual-population model.
 Following the convention of @McElreath_2020 [pp.
 499-516] and @Deffner_et_al_2022 [pp.
-6], the [DAG -@fig-cj12_dag] depicts the conceptual-population outcome $O^{cp}_{R}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the sampling mechanism.
-Additionally, the DAG depicts the conditioned *sample design* variable $S$ causally influencing the sample-comparison outcome $O^{sc}_{R}$.
-This variable shows that the sample-comparison outcome is a subset drawn from the conceptual-population outcome $O^{cp}_{R}$.
+6], the [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.
+
+Additionally, the DAG depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
+A square encloses this vector, indicating that it is a conditioned variable.
+In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a given variable [@Neal_2020, pp.
+32].
+Thus, the vector $S$ is responsible for selecting the subset of observation from the conceptual-population outcome $O^{cp}_{R}$, generating the sample-comparison outcome $O^{sc}_{R}$.
 
@@
 :::
-
-Notice,
-<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
 
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:49:30 - 00:53:30 -->
+<!-- recording: Sven and Tine 25.03.28; time: 00:33:55 - 00:37:50 -->
 
@@
 
+when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification.
+
 ::: {#fig-cj15 layout-ncol=2}
@@
 ::: {#fig-cj15_stat fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{R} & = O^{sc}_{R} \odot C \\
-  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
-  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
-  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
-  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
-  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
-  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
-  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
-  e_{I} & \overset{iid}{\sim} \text{Normal}(0,1) \\
-  e_{J} & \overset{iid}{\sim} \text{Normal}(0,1) \\
-  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
-\end{aligned}
-$$
-
-Statistical model
-:::
-
-::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
-  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \dsep e_{JK}
-\end{aligned}
-$$
-
-SCM
-:::
-
-Sample-comparison model, including sample and comparison mechanisms and assuming equal discriminal dispersions for the student's traits
-:::
-
-::: {#fig-cj16 layout-ncol=2}
-
-::: {#fig-cj16_stat fig-align="center" fig-pos="H"}
 $$
@@
 
-::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
+::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
 $$
@@
 
+<!-- Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable.
+Relevant variables depends on the context, you really have to think through which variables you have to put in.
+-->
+
 <!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need?
@@
 
+<!-- remember to mention fidelity of the theory in the model -->
+
 <!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
@@
 
+<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
+4-5] -->
+
 ## Limitations and further research {#sec-discussion-limitations}
 
-<!-- A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables.
-In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp.
-37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+<!-- *Relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
+Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable.
+Relevant variables depends on the context, you really have to think through which variables you have to put in.
 -->

### 2025-03-28 code/CJmod_add.R

@@ -1,4 +1,4 @@
-# equal variances ####
+# equal discriminal dispersions ####
 
 mcmc_code = "
 data{
@@ -178,7 +178,7 @@ writeLines(mcmc_code, con=file.path(dir, 'code', model_nam) )
 
 
 
-# unequal variances ####
+# different discriminal dispersions ####
 
 mcmc_code = "
 data{

### 2025-03-28 _tex/paper2.tex
Built site for gh-pages

@@
 solid black circles and unobserved (latent) variables as open circles
-\citep{Morgan_et_al_2014}. They also display conditioned variables by
-surrounding them with a square. Here, \emph{conditioning} refers to
-restricting the statistical analysis to a subset of the population
-defined by the values of the given variable \citep[pp.~32]{Neal_2020}.
-Although DAGs often omit exogenous variables for simplicity, the DAGs
-presented in this section includes exogenous variables to improve
-clarity and reveal potential issues related to conditioning and
-confounding \citep{Cinelli_et_al_2020}.
+\citep{Morgan_et_al_2014}. Although DAGs often omit exogenous variables
+for simplicity, the DAGs presented in this section includes exogenous
+variables to improve clarity and reveal potential issues related to
+conditioning and confounding \citep{Cinelli_et_al_2020}.
 
@@
 
-\[
+\begin{equation}\phantomsection\label{eq-mat13}{
 I = \begin{bmatrix}
@@
 \end{bmatrix}
-\]
+}\end{equation}
 
@@
 
-The conceptual-population model integrates two \emph{hierarchical
+Building on the principles of Structural Equation Modeling (SEM)
+\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
+\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
+the conceptual-population model integrates two \emph{hierarchical
 structural components} to examine how different teaching methods
-influence students' writing ability. This approach builds on the
-principles of Structural Equation Modeling (SEM)
-\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
-\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}.
-Each structural component defines how observed or latent variables
-affect the primary latent variable of interest
-\citep{Everitt_et_al_2010}. Their hierarchical nature enables
+influence students' writing ability. Each structural component defines
+how observed or latent variables affect the primary latent variable of
+interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
 researchers to test hypotheses while accounting for uncertainties
@@
 
-The top branch of DAG~\ref{fig-cj09_dag} depicts the first component.
-This component describes how the students' writing-quality trait
-\((T_{I})\) and a set of \emph{relevant} text-related variables
-\((X_{IA})\) (e.g., text length) causally influence the texts'
-written-quality trait \((T_{IA})\). The DAG also highlights the
-idiosyncratic errors \((e_{IA})\), which capture variations in the
-texts' trait that neither the students' trait nor other relevant
-text-related variables can explain. In this structure, \(T_{I}\) and
-\(e_{IA}\) are latent vectors of length \(n_{I}\) and
-\((n_{I} \cdot n_{A})\), respectively. Additionally, \(X_{IA}\) is an
-observed matrix with a dimension of \((n_{I} \cdot n_{A})\) rows and
-\(q_{IA}\) columns (variables).
-
-Additionally, the DAG's top branch illustrates how a set of
-\emph{relevant} student-related variables \((X_{I})\), such as the
-teaching method, causally influence the students' writing-quality trait
-\((T_{I})\). The DAG also includes the idiosyncratic errors \((e_{I})\),
-which capture variations in the students' traits that the relevant
-student-related variables cannot explain. Here, \(e_{I}\) is a latent
-vector of length \(n_{I}\), and \(X_{I}\) can be an observed matrix with
-\(n_{I}\) rows and \(q_{I}\) columns (variables).
-
-Here, \emph{relevant} refers to variables that satisfy the
-\emph{backdoor criterion} \citep[pp 37]{Neal_2020}, that is, they belong
-to a \emph{sufficient adjustment set}
-\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
+The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
+component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
+  to variables that satisfy the \emph{backdoor criterion} \citep[pp
+  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
+  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
   \emph{sufficient} set (potentially empty) blocks all non-causal paths
   between a predictor and an outcome without opening new ones
-\citep{Pearl_2009}.
+  \citep{Pearl_2009}. While these topics are beyond the scope of this
+  study, readers seeking a deeper understanding can refer to
+  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
+  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
+  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}
+  and \citet{McElreath_2020} useful. For more advanced study, seminal
+  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
+  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
+  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
+  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
+\((X_{I})\), such as the teaching method, causally influence the latent
+variable representing students' writing-quality trait \((T_{I})\). It
+also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
+  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
+influence the texts' written-quality trait \((T_{IA})\), the primary
+latent variable of interest. Additionally, the branch accounts for
+idiosyncratic errors: \(e_{I}\) captures unexplained variations in
+students' traits unexplained by \(X_{I}\), while \(e_{IA}\) captures
+variations in texts' traits unexplained by \(T_{I}\) or \(X_{IA}\).
+Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
+columns (variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors
+of length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
+dimension \((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns
+(variables), while \(e_{IA}\) and \(T_{IA}\) are latent vectors of
+length \(n_{I} \cdot n_{A}\).
 
@@
 Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
-second component. This component describes how the judges' bias
-\((B_{J})\) and a set of \emph{relevant} judgment-related variables
-\((Z_{JK})\) (e.g., the number of judgments a judge makes) causally
-influence the biases associated with each text \(B_{JK}\). The DAG also
-highlights the idiosyncratic errors \((e_{JK})\), which capture
-variations in the texts' bias that neither the judges' bias nor other
-relevant judgment-related variables can explain. In this description,
-\(B_{J}\) and \(e_{JK}\) are latent vectors of length \(n_{J}\) and
-\((n_{J} \cdot n_{K})\), respectively. Additionally, \(Z_{JK}\) is an
-observed matrix with a dimension of \((n_{J} \cdot n_{K})\) rows and
-\(q_{JK}\) columns.
-
-Moreover, the DAG's bottom branch describes how a set of \emph{relevant}
+second component, where \emph{relevant} \footnote{refer to footnote 1.}
 judge-related variables \((Z_{J})\), such as judgment expertise,
-causally influence the judges' bias \((B_{J})\). The DAG also accounts
-for idiosyncratic errors \((e_{J})\), capturing the variations in the
-judges' bias that relevant judge-related variables cannot explain. Here,
-\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
-observed matrix with \(n_{J}\) rows and \(q_{J}\) columns.
+causally influence the latent variable representing judges' bias
+\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
+  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
+number of judgments a judge makes) causally influence the biases
+associated with each text \(B_{JK}\), the primary latent variable of
+interest. Additionally, the branch accounts for idiosyncratic errors:
+\(e_{J}\) captures unexplained variations in judges' bias unexplained by
+\(Z_{J}\), while \(e_{JK}\) captures variations in judgments unexplained
+by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
+\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
+\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
+is an observed matrix with dimension \((n_{J} \cdot n_{K})\) rows and
+\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
+latent vectors of length \(n_{J} \cdot n_{K}\).
 
@@
 
-To facilitate the interpretation of the sample-comparison model, we
-first define the data sample using several binary vector variables:
-\(S_{I}\), \(S_{J}\), \(S_{IA}\), \(S_{JK}\), and \(S\) as follows.
+To incorporate the sampling mechanism and facilitate the interpretation
+of the sample-comparison model, we first define the data sampling
+process using the binary vector variables \(S_{I}\), \(S_{J}\),
+\(S_{IA}\), and \(S_{JK}\) as follows:
 
@@
 jk_{(nJ \cdot nK)} \\
-\end{bmatrix}; \;
+\end{bmatrix}
+}\end{equation}
+
+Where each element of the vectors \(S_{I}\) and \(S_{J}\) is a binary
+value indicating the presence or absence of data in the \(I\) and \(J\)
+vectors (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{I}\) and
+\(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements, respectively.
+Specifically, we define the vector \(S_{I}\) as in equation
+(\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).
+
+\begin{equation}\phantomsection\label{eq-mat22}{
+i_{(i)} = \begin{cases} 
+1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
+0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
+\end{cases}
+}\end{equation}
+
+Similarly, each element of the vectors \(S_{IA}\) and \(S_{JK}\) is a
+binary value indicating the presence or absence of data in the matrices
+\(IA\) and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors
+\(S_{IA}\) and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and
+\(n_{J} \cdot n_{K}\) elements, respectively. Specifically, we define
+the vector \(S_{IA}\) as in equation (\ref{eq-mat23}) and apply a
+similar definition to \(S_{JK}\).
+
+\begin{equation}\phantomsection\label{eq-mat23}{
+ia_{(ia)} = \begin{cases} 
+1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
+0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
+\end{cases}
+}\end{equation}
+
+We can visualize these vectors more clearly using the example in
+equation (\ref{eq-mat13}). Suppose the sample excludes the second
+student and the third judge. Additionally, assume that the second text
+from each individual and the third judgment from each judge are also
+missing. Under these conditions, the vectors have the following
+structure:
+
+\begin{equation}\phantomsection\label{eq-mat24}{
+S_{I} = \begin{bmatrix}
+1 \\
+0 \\
+1 \\
+1 \\
+1
+\end{bmatrix} ; \;
+S_{J} = \begin{bmatrix}
+1 \\
+1 \\
+0
+\end{bmatrix} ; \;
+S_{IA} = \begin{bmatrix}
+1 \\
+0 \\
+0 \\
+0 \\
+1 \\
+0 \\
+1 \\
+0 \\
+1 \\
+0 
+\end{bmatrix} ; \;
+S_{JK} = \begin{bmatrix}
+1 \\
+1 \\
+0 \\
+1 \\
+1 \\
+0 \\
+0 \\
+0 \\
+0 
+\end{bmatrix}
+}\end{equation}
+
+Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
+\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
+\(S_{JK}\) vectors. In other words, researchers can only observe texts
+and judgments from students and judges initially included in the sample.
+Additionally, it shows that the sum of observed elements in \(S_{I}\)
+equals the sampled students \((n^{s}_{I})\) and that a similar sum in
+vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
+the sum of observed elements in \(S_{IA}\) represents the total sampled
+texts across all students \((n^{s}_{I} \cdot n^{s}_{A})\), while a
+similar sum in vector \(S_{JK}\) represents the total sampled judgments
+across all judges \((n^{s}_{J} \cdot n^{s}_{K})\).
+
+Finally, we define the \emph{sample design} vector \(S\) in equation
+(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every
+component of \(S_{JK}\). Essentially, this vector serves as the sampled
+equivalent of the \(R\) matrix defined in equation (\ref{eq-mat12}),
+where the sum of its elements represents the total data sample
+\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
+
+\begin{equation}\phantomsection\label{eq-mat25}{
 S = \begin{bmatrix}
@@
 
-Each element of the vectors \(S_{I}\) and \(S_{J}\) is a binary value
-that indicates the presence \((1)\) or absence \((0)\) of data in the
-\(I\) and \(J\) vectors, as defined in equation (\ref{eq-mat11}).
-Therefore, the vector \(S_{I}\) contains \(n_{I}\) elements, and the
-vector \(S_{J}\) contains \(n_{J}\) elements. For example, we define
-each element of the vector \(S_{I}\) as in equation (\ref{eq-mat22}),
-and we apply a similar definition to \(S_{J}\).
-
-\begin{equation}\phantomsection\label{eq-mat22}{
-i_{(i)} = \begin{cases} 
-1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
-0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
-\end{cases}
-}\end{equation}
-
-Each element of the vectors \(S_{IA}\) and \(S_{JK}\) is defined like
-\(i_{(i)}\) with vector \(I\), but applied to data in the matrices
-\(IA\) and \(JK\), as defined in equation (\ref{eq-mat11}). Therefore,
-the vector \(S_{IA}\) contains \(n_{I} \cdot n_{A}\) elements, and the
-vector \(S_{JK}\) contains \(n_{J} \cdot n_{K}\) elements. For example,
-we define each element of the vector \(S_{IA}\) as in equation
-(\ref{eq-mat23}), and we apply a similar definition to \(S_{JK}\).
-
-\begin{equation}\phantomsection\label{eq-mat23}{
-ia_{(ia)} = \begin{cases} 
-1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
-0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
-\end{cases}
-}\end{equation}
-
-The equation shows that the observed values in the \(S_{I}\) and
-\(S_{J}\) vectors determine the binary values in the \(S_{IA}\) and
-\(S_{JK}\) vectors, through the \(IA\) and \(JK\) matrices,
-respectively. Specifically, researchers cannot observe stimuli and
-judgments for individuals and judges they did not sample initially.
-
-Finally, the vector \(S\) in equation (\ref{eq-mat21}) contains
-\(n = \binom{n_{I} \cdot n_{A}}{2} \cdot n_{J} \cdot n_{K}\) elements,
-with each element defined in equation (\ref{eq-mat24}). This equation
-shows that the observed values in \(S_{IA}\) and \(S_{JK}\) determine
-the binary values in \(S\) through the \(R\) matrix. Specifically,
-researchers can only observe a subset of judges' comparisons for a
-subset of stimuli generated by a subset of individuals.
-
-\begin{equation}\phantomsection\label{eq-mat24}{
-s_{(s)} = \begin{cases} 
-1 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are observed} \\
-0 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are missing}
-\end{cases}
-}\end{equation}
-
-With the definition of \(S\), we can now integrate the sample mechanism
-into the conceptual-population model. Figure~\ref{fig-cj12} illustrates
-this model. Following the convention of
+With the definition of \(S\), we incorporate the sample mechanism into
+the conceptual-population model. Following the convention of
 \citet[pp.~499-516]{McElreath_2020} and
-\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} depicts
-the conceptual-population outcome \(O^{cp}_{R}\) as an unobserved
-variable, emphasizing the researchers' inability to access such an
-outcome due to the sampling mechanism. Additionally, the DAG depicts the
-conditioned \emph{sample design} variable \(S\) causally influencing the
-sample-comparison outcome \(O^{sc}_{R}\). This variable shows that the
-sample-comparison outcome is a subset drawn from the
-conceptual-population outcome \(O^{cp}_{R}\).
+\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} represents
+the conceptual-population outcome \(O^{cp}_{R}\) as unobserved,
+emphasizing that researchers cannot directly access this outcome due to
+the sampling mechanism.
+
+Additionally, the DAG depicts the \emph{sample design} vector \(S\) as a
+causal factor influencing the sample-comparison outcome \(O^{sc}_{R}\).
+A square encloses this vector, indicating that it is a conditioned
+variable. In this context, \emph{conditioning} means restricting the
+statistical analysis to a specific subset of the population based on the
+values of a given variable \citep[pp.~32]{Neal_2020}. Thus, the vector
+\(S\) is responsible for selecting the subset of observation from the
+conceptual-population outcome \(O^{cp}_{R}\), generating the
+sample-comparison outcome \(O^{sc}_{R}\).
 
@@
 \end{figure}%
-
-Notice,
 
@@
 
-\begin{figure}[H]
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O_{R} & = O^{sc}_{R} \odot C \\
-  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
-  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
-  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
-  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
-  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
-  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
-  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
-  e_{I} & \overset{iid}{\sim} \text{Normal}(0,1) \\
-  e_{J} & \overset{iid}{\sim} \text{Normal}(0,1) \\
-  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj15_stat}Statistical model}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
-  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj15_scm}SCM}
-
-\end{minipage}%
-
-\caption{\label{fig-cj15}Sample-comparison model, including sample and
-comparison mechanisms and assuming equal discriminal dispersions for the
-student's traits}
-
-\end{figure}%
+when you talk about the variability, refer that the equal dispersion is
+just assuming 1 for identification.
 
@@
 
-\subcaption{\label{fig-cj16_stat}Statistical model}
+\subcaption{\label{fig-cj15_stat}Statistical model}
 
@@
 
-\subcaption{\label{fig-cj16_scm}SCM}
-
-\end{minipage}%
-
-\caption{\label{fig-cj16}Sample-comparison model, including sample and
+\subcaption{\label{fig-cj15_scm}SCM}
+
+\end{minipage}%
+
+\caption{\label{fig-cj15}Sample-comparison model, including sample and
 comparison mechanisms and assuming different discriminal dispersions for

### 2025-03-29 paper2.qmd
section:4.2; process:write

@@
 <!-- 1. What is the assumption on the conceptual-population model -->
-In the conceptual-population model, we assume an idealized scenario where researchers have access to a *conceptual population* of comparative data --data representing all repeated judgments made by each judge for every stimulus pair produced by each pair of individuals in the population.
+In the conceptual-population model, we assume an idealized scenario where researchers have access to a *conceptual population* of comparative data --data representing *all repeated judgments made by each judge for every stimulus pair produced by each pair of individuals* in the population.
 This assumption allows us to integrate Thurstone's theoretical principles and propose innovations to address some of the issues raised in @sec-theory-issues.
@@
 Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
-Their hierarchical nature enables researchers to test hypotheses while accounting for uncertainties associated with trait estimation (see @sec-theory-issue2), particularly when stimuli exhibit a hierarchical structure (see @sec-theory-issue1b).
+Their hierarchical nature enables researchers to test hypotheses while accounting for uncertainties associated with trait estimation (see @sec-theory-issue2), and when stimuli exhibit a hierarchical structure (see @sec-theory-issue1b).
 
@@
 It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the primary latent variable of interest.
-Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures unexplained variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$.
+Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$.
 Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
-Similarly, $X_{IA}$ is an observed matrix with dimension $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
+Similarly, $X_{IA}$ is an observed matrix with dimension $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
 
@@
 A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
-While these topics are beyond the scope of this study, readers seeking a deeper understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, as well as introductory books like @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful.
-For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020, are recommended.
+These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, as well as introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 useful.
+For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020, are recommended.
 
@@
 It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the primary latent variable of interest.
-Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures unexplained variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
+Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
 Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Similarly, $Z_{JK}$ is an observed matrix with dimension $(n_{J} \cdot n_{K})$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
+Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
 
@@
 Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
-The extensions to the conceptual-population model, derived from this realistic scenario, are concisely described in the sample-comparison model.
+The extensions to the conceptual-population model, derived from this realistic scenario, are described in the sample-comparison model.
 
@@
 In the sample-comparison model, we assume a more realistic scenario than the conceptual-population model.
-Specifically, we assume researchers have access to a data sample consisting of a limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, drawn from the conceptual population.
-Additionally, we assume that judges do not make *all possible comparisons* allowed by the sample, but rather only $n_{C}$ comparisons.
+Specifically, we assume researchers have access to a data sample consisting of a *limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$*, drawn from the conceptual population.
+In a subsequent step, we assume that judges *do not make all possible comparisons* allowed by the sample, but rather only $n_{C}$ comparisons.
 
@@
 
-Where each element of the vectors $S_{I}$ and $S_{J}$ is a binary value indicating the presence or absence of data in the $I$ and $J$ vectors (see equation ([-@eq-mat11])).
+Where each element of $S_{I}$ and $S_{J}$ is a binary value indicating the presence or absence of data elements in the $I$ and $J$ vectors (see equation ([-@eq-mat11])).
 Thus, the vectors $S_{I}$ and $S_{J}$ contains $n_{I}$ and $n_{J}$ elements, respectively.
@@
 
-Similarly, each element of the vectors $S_{IA}$ and $S_{JK}$ is a binary value indicating the presence or absence of data in the matrices $IA$ and $JK$ (see equation ([-@eq-mat11])).
+Similarly, each element of $S_{IA}$ and $S_{JK}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$ and $JK$ (see equation ([-@eq-mat11])).
 Thus, the vectors $S_{IA}$ and $S_{JK}$ contains $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$ elements, respectively.
@@
 We can visualize these vectors more clearly using the example in equation ([-@eq-mat13]).
-Suppose the sample excludes the second student and the third judge.
-Additionally, assume that the second text from each individual and the third judgment from each judge are also missing.
-Under these conditions, the vectors have the following structure:
+Suppose researchers exclude the second student and the third judge.
+Additionally, they do not include the second text from each individual and the third judgment from each judge.
+Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, these vectors take the following structure:
 
@@
 
-Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$ vectors.
+Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$.
 In other words, researchers can only observe texts and judgments from students and judges initially included in the sample.
@@
 
-Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every component of $S_{JK}$.
-Essentially, this vector serves as the sampled equivalent of the $R$ matrix defined in equation ([-@eq-mat12]), where the sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
+Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
+The sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
 
@@
 6], the [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.
-
-Additionally, the DAG depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
-A square encloses this vector, indicating that it is a conditioned variable.
-In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a given variable [@Neal_2020, pp.
-32].
-Thus, the vector $S$ is responsible for selecting the subset of observation from the conceptual-population outcome $O^{cp}_{R}$, generating the sample-comparison outcome $O^{sc}_{R}$.
-
-<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
 
@@
 
+The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
+A square encloses this vector, indicating that it is a conditioned variable.
+In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp.
+32].
+Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector determines *which repeated comparisons a subset of judges makes for a subset of stimuli produced by a subset of individuals* from the conceptual-population, ultimately generating the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
+
+Notably, the DAG shows that $S$ is independent of all other variables in the model.
+This implies that the conceptual model represented in [DAG -@fig-cj12_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019, pp.
+150].
+In these designs, each judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp.
+16-112].
+
+<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
+
+However, because CJ experiments rarely exhaustively pair sampled judgments, judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
+5]--a realistic scenario must account for judges comparing only specific stimuli from certain individuals.
+
 ### The comparison mechanism {#sec-theory-theoretical_SC2}
@@
 <!-- 1. What is the comparison mechanism -->
-Conversely, we make the scenario depicted in @sec-theory-theoretical_SC1 more realistic by integrating the comparison mechanism, that is, we assume that researchers do not have access to *all comparisons* from the sample, but rather a sufficient number of comparisons $(n_{C})$ so that a proportion, $P(B>A)$, may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+<!-- Conversely, we make the scenario depicted in @sec-theory-theoretical_SC1 more realistic by integrating the comparison mechanism, that is, we assume that researchers do not have access to *all comparisons* from the sample, but rather a sufficient number of comparisons $(n_{C})$ so that a proportion, $P(B>A)$, may be determined for each pair of stimuli [@Thurstone_1927b, pp.
 267].
 @fig-cj14 illustrates this model, incorporating both the sample and comparison mechanisms.
+-->
 
@@
 <!-- 2. What is the effect of the sampling mechanism -->
-Following the convention of @McElreath_2020 [pp.
+<!-- Following the convention of @McElreath_2020 [pp.
 499-516], the [DAG -@fig-cj14_dag] depicts the sample-comparison outcome $O^{sc}_{R}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the comparison mechanism.
@@
 Specifically, the variable is a vector $C = \{ c_{1}, \dots, c_{n} \}$, where $n$ corresponds to the number of rows in the matrix $R$ (refer to @sec-theory-theoretical_P2), and each $c_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the sample-comparison outcome $O^{sp}_{R}$.
-Specifically,
+Specifically, -->
 
@@
 
-Notice,
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
 
@@
 
+<!-- Thus Thurstone’s model (in both its Case 5 and Rasch formulation) achieves the goal of sample-free calibration (Wright, 1977) in the sense that the estimate of the distance between A and B does not depend on which other objects they are compared with.
+[@Bramley_2008, pp.
+256] -->
+
 <!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).

### 2025-03-29 _tex/paper2.tex
Built site for gh-pages

@@
 where researchers have access to a \emph{conceptual population} of
-comparative data --data representing all repeated judgments made by each
-judge for every stimulus pair produced by each pair of individuals in
-the population. This assumption allows us to integrate Thurstone's
-theoretical principles and propose innovations to address some of the
-issues raised in Section~\ref{sec-theory-issues}.
+comparative data --data representing \emph{all repeated judgments made
+by each judge for every stimulus pair produced by each pair of
+individuals} in the population. This assumption allows us to integrate
+Thurstone's theoretical principles and propose innovations to address
+some of the issues raised in Section~\ref{sec-theory-issues}.
 
@@
 associated with trait estimation (see Section~\ref{sec-theory-issue2}),
-particularly when stimuli exhibit a hierarchical structure (see
+and when stimuli exhibit a hierarchical structure (see
 Section~\ref{sec-theory-issue1b}).
@@
   between a predictor and an outcome without opening new ones
-  \citep{Pearl_2009}. While these topics are beyond the scope of this
-  study, readers seeking a deeper understanding can refer to
+  \citep{Pearl_2009}. These topics are beyond the scope of this study,
+  thus, readers seeking a more profound understanding can refer to
   introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
   \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
-  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}
+  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020},
   and \citet{McElreath_2020} useful. For more advanced study, seminal
@@
   \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
-  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
+  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
   \citet{Hernan_et_al_2020}, are recommended.} student-related variables
@@
 latent variable of interest. Additionally, the branch accounts for
-idiosyncratic errors: \(e_{I}\) captures unexplained variations in
-students' traits unexplained by \(X_{I}\), while \(e_{IA}\) captures
-variations in texts' traits unexplained by \(T_{I}\) or \(X_{IA}\).
-Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
-columns (variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors
-of length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
-dimension \((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns
-(variables), while \(e_{IA}\) and \(T_{IA}\) are latent vectors of
-length \(n_{I} \cdot n_{A}\).
+idiosyncratic errors: \(e_{I}\) captures variations in students' traits
+unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in texts'
+traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is an
+observed matrix with \(n_{I}\) rows and \(q_{I}\) columns (variables),
+and both \(e_{I}\) and \(T_{I}\) are latent vectors of length \(n_{I}\).
+Similarly, \(X_{IA}\) is an observed matrix with dimension
+\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables), while
+\(e_{IA}\) and \(T_{IA}\) are latent vectors of length
+\(n_{I} \cdot n_{A}\).
 
@@
 interest. Additionally, the branch accounts for idiosyncratic errors:
-\(e_{J}\) captures unexplained variations in judges' bias unexplained by
-\(Z_{J}\), while \(e_{JK}\) captures variations in judgments unexplained
-by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
+\(e_{J}\) captures variations in judges' bias unexplained by \(Z_{J}\),
+while \(e_{JK}\) captures variations in judgments unexplained by
+\(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
 \(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
 \(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
-is an observed matrix with dimension \((n_{J} \cdot n_{K})\) rows and
+is an observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and
 \(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
@@
 The extensions to the conceptual-population model, derived from this
-realistic scenario, are concisely described in the sample-comparison
-model.
+realistic scenario, are described in the sample-comparison model.
 
@@
 the conceptual-population model. Specifically, we assume researchers
-have access to a data sample consisting of a limited number of judgments
-\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
-number of texts \((n^{s}_{A})\) from a sample of students
-\((n^{s}_{I})\), drawn from the conceptual population. Additionally, we
-assume that judges do not make \emph{all possible comparisons} allowed
-by the sample, but rather only \(n_{C}\) comparisons.
+have access to a data sample consisting of a \emph{limited number of
+judgments \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a
+specific number of texts \((n^{s}_{A})\) from a sample of students
+\((n^{s}_{I})\)}, drawn from the conceptual population. In a subsequent
+step, we assume that judges \emph{do not make all possible comparisons}
+allowed by the sample, but rather only \(n_{C}\) comparisons.
 
@@
 
-Where each element of the vectors \(S_{I}\) and \(S_{J}\) is a binary
-value indicating the presence or absence of data in the \(I\) and \(J\)
-vectors (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{I}\) and
-\(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements, respectively.
-Specifically, we define the vector \(S_{I}\) as in equation
-(\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).
+Where each element of \(S_{I}\) and \(S_{J}\) is a binary value
+indicating the presence or absence of data elements in the \(I\) and
+\(J\) vectors (see equation (\ref{eq-mat11})). Thus, the vectors
+\(S_{I}\) and \(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements,
+respectively. Specifically, we define the vector \(S_{I}\) as in
+equation (\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).
 
@@
 
-Similarly, each element of the vectors \(S_{IA}\) and \(S_{JK}\) is a
-binary value indicating the presence or absence of data in the matrices
-\(IA\) and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors
-\(S_{IA}\) and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and
-\(n_{J} \cdot n_{K}\) elements, respectively. Specifically, we define
-the vector \(S_{IA}\) as in equation (\ref{eq-mat23}) and apply a
-similar definition to \(S_{JK}\).
+Similarly, each element of \(S_{IA}\) and \(S_{JK}\) is a binary value
+indicating the presence or absence of data rows in the matrices \(IA\)
+and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{IA}\)
+and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
+elements, respectively. Specifically, we define the vector \(S_{IA}\) as
+in equation (\ref{eq-mat23}) and apply a similar definition to
+\(S_{JK}\).
 
@@
 We can visualize these vectors more clearly using the example in
-equation (\ref{eq-mat13}). Suppose the sample excludes the second
-student and the third judge. Additionally, assume that the second text
-from each individual and the third judgment from each judge are also
-missing. Under these conditions, the vectors have the following
-structure:
+equation (\ref{eq-mat13}). Suppose researchers exclude the second
+student and the third judge. Additionally, they do not include the
+second text from each individual and the third judgment from each judge.
+Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
+these vectors take the following structure:
 
@@
 \(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
-\(S_{JK}\) vectors. In other words, researchers can only observe texts
-and judgments from students and judges initially included in the sample.
+\(S_{JK}\). In other words, researchers can only observe texts and
+judgments from students and judges initially included in the sample.
 Additionally, it shows that the sum of observed elements in \(S_{I}\)
@@
 Finally, we define the \emph{sample design} vector \(S\) in equation
-(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every
-component of \(S_{JK}\). Essentially, this vector serves as the sampled
-equivalent of the \(R\) matrix defined in equation (\ref{eq-mat12}),
-where the sum of its elements represents the total data sample
+(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
+of \(S_{JK}\). The sum of its elements represents the total data sample
 \(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
@@
 the sampling mechanism.
-
-Additionally, the DAG depicts the \emph{sample design} vector \(S\) as a
-causal factor influencing the sample-comparison outcome \(O^{sc}_{R}\).
-A square encloses this vector, indicating that it is a conditioned
-variable. In this context, \emph{conditioning} means restricting the
-statistical analysis to a specific subset of the population based on the
-values of a given variable \citep[pp.~32]{Neal_2020}. Thus, the vector
-\(S\) is responsible for selecting the subset of observation from the
-conceptual-population outcome \(O^{cp}_{R}\), generating the
-sample-comparison outcome \(O^{sc}_{R}\).
 
@@
 
+The DAG also depicts the \emph{sample design} vector \(S\) as a causal
+factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
+square encloses this vector, indicating that it is a conditioned
+variable. In this context, \emph{conditioning} means restricting the
+statistical analysis to a specific subset of the population based on the
+values of a conditioned variable \citep[pp.~32]{Neal_2020}. Essentially,
+\(S\) serves as the sample equivalent for the \(R\) matrix (see equation
+(\ref{eq-mat12})), that is, this vector determines \emph{which repeated
+comparisons a subset of judges makes for a subset of stimuli produced by
+a subset of individuals} from the conceptual-population, ultimately
+generating the sample-comparison outcome \(O^{sc}_{R}\) from the
+conceptual-population outcome \(O^{cp}_{R}\).
+
+Notably, the DAG shows that \(S\) is independent of all other variables
+in the model. This implies that the conceptual model represented in
+DAG~\ref{fig-cj12_dag} applies exclusively to Simple Random Sampling
+(SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
+judgment, judge, stimulus, and individual has the same probability of
+being included in the sample as any other observation within their
+respective groups \citep[pp.~16-112]{Lawson_2015}.
+
+However, because CJ experiments rarely exhaustively pair sampled
+judgments, judges, stimuli, and individuals--due to concerns about the
+practical feasibility of the comparison task, like in
+\citet[p.~5]{Boonen_et_al_2020}--a realistic scenario must account for
+judges comparing only specific stimuli from certain individuals.
+
 \subsubsection{The comparison
 mechanism}\label{sec-theory-theoretical_SC2}
-
-Conversely, we make the scenario depicted in
-Section~\ref{sec-theory-theoretical_SC1} more realistic by integrating
-the comparison mechanism, that is, we assume that researchers do not
-have access to \emph{all comparisons} from the sample, but rather a
-sufficient number of comparisons \((n_{C})\) so that a proportion,
-\(P(B>A)\), may be determined for each pair of stimuli
-\citep[pp.~267]{Thurstone_1927b}. Figure~\ref{fig-cj14} illustrates this
-model, incorporating both the sample and comparison mechanisms.
-
-Following the convention of \citet[pp.~499-516]{McElreath_2020}, the
-DAG~\ref{fig-cj14_dag} depicts the sample-comparison outcome
-\(O^{sc}_{R}\) as an unobserved variable, emphasizing the researchers'
-inability to access such an outcome due to the comparison mechanism.
-Moreover, the DAG shows the conditioned \emph{comparison design}
-variable \(C\), which causally influence the observed outcome \(O_{R}\).
-This variable reflects how this observed outcome is a sample drawn from
-the sample-comparison outcome \(O^{sp}_{R}\). Specifically, the variable
-is a vector \(C = \{ c_{1}, \dots, c_{n} \}\), where \(n\) corresponds
-to the number of rows in the matrix \(R\) (refer to
-Section~\ref{sec-theory-theoretical_P2}), and each \(c_{i}\) is a binary
-variable indicating the presence \((1)\) or absence \((0)\) of data in
-the corresponding positions of the sample-comparison outcome
-\(O^{sp}_{R}\). Specifically,
 

### 2025-03-31 paper2.qmd
section:4.2; process:write

@@
 Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-Following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.
+Following current CJ practices, researchers rely on estimates from the BTL model, or its transformations, to approximate this estimand.
 
@@
 <!-- 1. Why is useful to declare the data structure? -->
-Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system.
+Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system.
 Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
@@
 Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$.
-As a result, the matrix $IA$ contains $(n_{I} \cdot n_{A})$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
+As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
 Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$.
-Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.
+Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.
 
@@
 Specifically, $O^{cp}_{R}$ and $D_{R}$ are observed and latent vectors of length $n$, respectively.
-Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.
+Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$, respectively.
 
@@
 Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
-Their hierarchical nature enables researchers to test hypotheses while accounting for uncertainties associated with trait estimation (see @sec-theory-issue2), and when stimuli exhibit a hierarchical structure (see @sec-theory-issue1b).
+Their hierarchical nature enables researchers to test hypotheses considering the hierarchical structure of stimuli (see @sec-theory-issue1b) and the uncertainties in trait estimation (see @sec-theory-issue2).
 
@@
 The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as the teaching method, causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
-It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the primary latent variable of interest.
+It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest.
 Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$.
@@
 Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$.
-It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the primary latent variable of interest.
+It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
 Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
@@
 
-Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] define the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
+Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
 Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.
@@
 Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$.
-Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all judges $(n^{s}_{J} \cdot n^{s}_{K})$.
+Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.
 
@@
 The sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
+Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals from the population.
 
@@
 In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp.
-32].
-Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector determines *which repeated comparisons a subset of judges makes for a subset of stimuli produced by a subset of individuals* from the conceptual-population, ultimately generating the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
+32], in this case $S$.
+Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector determines *which repeated comparisons the subset of judges makes for the subset of stimuli produced by the subset of individuals* sampled from the conceptual-population.
+This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
 
@@
 
-However, because CJ experiments rarely exhaustively pair sampled judgments, judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
-5]--a realistic scenario must account for judges comparing only specific stimuli from certain individuals.
+However, because CJ experiments rarely use the exhaustive pairing of the sampled judgments, judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
+5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
 

### 2025-03-31 _tex/paper2.tex
Built site for gh-pages

@@
 the two groups of students?''. Following current CJ practices,
-researchers then rely on estimates from the BTL model, or its
+researchers rely on estimates from the BTL model, or its
 transformations, to approximate this estimand.
@@
 
-Although specifying a system's data structure is not mandatory when
-using SCMs and DAGs, defining one in this case can improve clarity and
-facilitate the description of the system. Thus, to re-express the scalar
-form of the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
+Although specifying a data structure is not mandatory when using SCMs
+and DAGs, defining one in this case can improve clarity and facilitate
+the description of the system. Thus, to re-express the scalar form of
+the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
 vectorized form, we first define the vectors \(I\) and \(J\), along with
@@
 stimuli \(a, b\). As a result, the matrix \(IA\) contains
-\((n_{I} \cdot n_{A})\) rows and \(2\) columns, where \(n_{A}\)
-specifies the number of stimuli available per individual. Likewise, each
-row of \(JK\) associates a judge \(j\) with a judgment index \(k\).
-Consequently, the matrix \(JK\) has \((n_{J} \cdot n_{K})\) rows and
-\(2\) columns, where \(n_{K}\) indicates the number of judgments each
-judge makes.
+\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
+the number of stimuli available per individual. Likewise, each row of
+\(JK\) associates a judge \(j\) with a judgment index \(k\).
+Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
+columns, where \(n_{K}\) indicates the number of judgments each judge
+makes.
 
@@
 latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
-\(B_{JK}\) are latent vectors of lengths \((n_{I} \cdot n_{A})\) and
-\((n_{J} \cdot n_{K})\), respectively.
+\(B_{JK}\) are latent vectors of lengths \(n_{I} \cdot n_{A}\) and
+\(n_{J} \cdot n_{K}\), respectively.
 
@@
 interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
-researchers to test hypotheses while accounting for uncertainties
-associated with trait estimation (see Section~\ref{sec-theory-issue2}),
-and when stimuli exhibit a hierarchical structure (see
-Section~\ref{sec-theory-issue1b}).
+researchers to test hypotheses considering the hierarchical structure of
+stimuli (see Section~\ref{sec-theory-issue1b}) and the uncertainties in
+trait estimation (see Section~\ref{sec-theory-issue2}).
 
@@
   1.} text-related variables \((X_{IA})\) (e.g., text length) causally
-influence the texts' written-quality trait \((T_{IA})\), the primary
-latent variable of interest. Additionally, the branch accounts for
-idiosyncratic errors: \(e_{I}\) captures variations in students' traits
-unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in texts'
-traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is an
-observed matrix with \(n_{I}\) rows and \(q_{I}\) columns (variables),
-and both \(e_{I}\) and \(T_{I}\) are latent vectors of length \(n_{I}\).
-Similarly, \(X_{IA}\) is an observed matrix with dimension
-\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables), while
-\(e_{IA}\) and \(T_{IA}\) are latent vectors of length
+influence the texts' written-quality trait \((T_{IA})\), the first
+primary latent variable of interest. Additionally, the branch accounts
+for idiosyncratic errors: \(e_{I}\) captures variations in students'
+traits unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in
+texts' traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is
+an observed matrix with \(n_{I}\) rows and \(q_{I}\) columns
+(variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors of
+length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
+dimension \(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables),
+while \(e_{IA}\) and \(T_{IA}\) are latent vectors of length
 \(n_{I} \cdot n_{A}\).
@@
 number of judgments a judge makes) causally influence the biases
-associated with each text \(B_{JK}\), the primary latent variable of
-interest. Additionally, the branch accounts for idiosyncratic errors:
+associated with each text \(B_{JK}\), the second primary latent variable
+of interest. Additionally, the branch accounts for idiosyncratic errors:
 \(e_{J}\) captures variations in judges' bias unexplained by \(Z_{J}\),
@@
 Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
-DAG~\ref{fig-cj09_dag} define the set of endogenous variables \(V\),
-structural equations \(F\), and exogenous variables \(E\) for the
+DAG~\ref{fig-cj09_dag} are part of the set of endogenous variables
+\(V\), structural equations \(F\), and exogenous variables \(E\) for the
 conceptual-population model. Additionally, the figures demonstrate that
@@
 the sum of observed elements in \(S_{IA}\) represents the total sampled
-texts across all students \((n^{s}_{I} \cdot n^{s}_{A})\), while a
-similar sum in vector \(S_{JK}\) represents the total sampled judgments
-across all judges \((n^{s}_{J} \cdot n^{s}_{K})\).
+texts across all sampled students \((n^{s}_{I} \cdot n^{s}_{A})\), while
+a similar sum in vector \(S_{JK}\) represents the total sampled
+judgments across all sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\).
 
@@
 \(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
+Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
+binomial coefficient, which quantifies the total number of unique
+comparisons possible between every pair of sampled stimuli generated by
+each pair of sampled individuals from the population.
 
@@
 statistical analysis to a specific subset of the population based on the
-values of a conditioned variable \citep[pp.~32]{Neal_2020}. Essentially,
-\(S\) serves as the sample equivalent for the \(R\) matrix (see equation
-(\ref{eq-mat12})), that is, this vector determines \emph{which repeated
-comparisons a subset of judges makes for a subset of stimuli produced by
-a subset of individuals} from the conceptual-population, ultimately
-generating the sample-comparison outcome \(O^{sc}_{R}\) from the
-conceptual-population outcome \(O^{cp}_{R}\).
+values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
+\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
+matrix (see equation (\ref{eq-mat12})), that is, this vector determines
+\emph{which repeated comparisons the subset of judges makes for the
+subset of stimuli produced by the subset of individuals} sampled from
+the conceptual-population. This process ultimately generates the
+sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
+outcome \(O^{cp}_{R}\).
 
@@
 
-However, because CJ experiments rarely exhaustively pair sampled
-judgments, judges, stimuli, and individuals--due to concerns about the
-practical feasibility of the comparison task, like in
-\citet[p.~5]{Boonen_et_al_2020}--a realistic scenario must account for
-judges comparing only specific stimuli from certain individuals.
+However, because CJ experiments rarely use the exhaustive pairing of the
+sampled judgments, judges, stimuli, and individuals --due to concerns
+about the practical feasibility of the comparison task, like in
+\citet[p.~5]{Boonen_et_al_2020}-- a realistic scenario must also account
+for judges comparing only specific stimuli from certain individuals.
 

### 2025-03-31 paper2.qmd
section:4.2; process:write

@@
 
+::: {#fig-cj04 layout-nrow=2}
+
+::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
+
+Conceptual-population model, vectorized form.
+:::
+
 <!-- 3. how de we use that data structure? -->
@@
 Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$, respectively.
-
-::: {#fig-cj04 layout-nrow=2}
-
-::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
-
-Conceptual-population model, vectorized form.
-:::
 
@@
 i_{(1)} \\
-i_{(2)} \\
 \vdots \\
@@
 j_{(1)} \\
-j_{(2)} \\
 \vdots \\
@@
 ia_{(1)} \\
-ia_{(2)} \\
 \vdots \\
@@
 jk_{(1)} \\
-jk_{(2)} \\
 \vdots \\
@@
 s_{(1)} \\
-s_{(2)} \\
 \vdots \\
@@
 
-However, because CJ experiments rarely use the exhaustive pairing of the sampled judgments, judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
+However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
 5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.

### 2025-03-31 _tex/paper2.tex
Built site for gh-pages

@@
 
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj04_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+
+}
+
+\subcaption{\label{fig-cj04_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}
+
+\end{figure}%
+
 Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
@@
 \(n_{J} \cdot n_{K}\), respectively.
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj04_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
-
-}
-
-\subcaption{\label{fig-cj04_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}
-
-\end{figure}%
 
@@
 
-However, because CJ experiments rarely use the exhaustive pairing of the
-sampled judgments, judges, stimuli, and individuals --due to concerns
-about the practical feasibility of the comparison task, like in
+However, because CJ experiments rarely use the exhaustive pairing of
+sampled judges, stimuli, and individuals --due to concerns about the
+practical feasibility of the comparison task, like in
 \citet[p.~5]{Boonen_et_al_2020}-- a realistic scenario must also account

### 2025-03-31 paper2.qmd
section:4.2; process:write,finish

@@
 <!-- 1. What is the more realistic scenario: the sample mechanism -->
-In the sample-comparison model, we assume a more realistic scenario than the conceptual-population model.
-Specifically, we assume researchers have access to a data sample consisting of a *limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$*, drawn from the conceptual population.
-In a subsequent step, we assume that judges *do not make all possible comparisons* allowed by the sample, but rather only $n_{C}$ comparisons.
+The sample-comparison model presents a more realistic scenario than the conceptual-population model.
+Expressly, we assume researchers work with a data sample consisting of a *limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$*, all drawn from the conceptual population.
+In a subsequent step, we assume that judges do not perform *all possible repeated comparisons* within the data sample.
+Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.
 
@@
 <!-- 1. start by defining the sampling vector and matrices  -->
-To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the data sampling process using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:
+To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the *data sampling process* using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:
 
@@
 i_{(i)} = \begin{cases} 
-1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
-0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
+1 & \text{if data element } i \text{ from } I \text{ is observed} \\
+0 & \text{if data element } i \text{ from } I \text{ is missing}
 \end{cases}
@@
 ia_{(ia)} = \begin{cases} 
-1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
-0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
+1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
+0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
 \end{cases}
@@
 Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
-The sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
-Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals from the population.
+Thus, the vector contains $n$ elements, matching the number of rows in the $R$ matrix, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
+Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
 
@@
 32], in this case $S$.
-Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector determines *which repeated comparisons the subset of judges makes for the subset of stimuli produced by the subset of individuals* sampled from the conceptual-population.
+Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated comparisons a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
 This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
@@
 150].
-In these designs, each judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp.
+In these designs, each repeated judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp.
 16-112].
@@
 
+<!-- 3. segway for next section -->
 However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
@@
 
-<!-- 1. What is the comparison mechanism -->
-<!-- Conversely, we make the scenario depicted in @sec-theory-theoretical_SC1 more realistic by integrating the comparison mechanism, that is, we assume that researchers do not have access to *all comparisons* from the sample, but rather a sufficient number of comparisons $(n_{C})$ so that a proportion, $P(B>A)$, may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+<!-- 1. start by defining the comparison vector  -->
+As in the previous section, we first define the *comparison design* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model and integrate the comparison mechanism.
+
+$$
+C = \begin{bmatrix}
+c_{(1)} \\
+\vdots \\
+c_{(c)} \\
+\vdots \\
+c_{(n)} \\
+\end{bmatrix}
+$$ {#eq-mat26}
+
+Where each element of $C$ is a binary value indicating the presence or absence of data rows in the $R$ matrix (see equation ([-@eq-mat12])):
+
+$$
+c_{c} = \begin{cases} 
+1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
+0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
+\end{cases}
+$$
+
+Thus, the *comparison design* vector $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, and the sum of its elements represents the selected comparisons, $n_{C}$.
+This definition assumes that judges do not perform *all possible repeated comparisons* within the data sample.
+Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp.
 267].
-@fig-cj14 illustrates this model, incorporating both the sample and comparison mechanisms.
--->
 
@@
 
+<!-- 2. What is the effect of the comparison mechanism -->
+With the definition of $C$, the [DAG -@fig-cj14_dag] incorporates the sample and comparison mechanism into the conceptual-population model.
+The DAG now also represents the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
+The DAG also represents the *comparison design* vector $C$ as a causal factor influencing the final comparative outcome $O_{R}$.
+Equivalent to the sampling mechanism, a square encloses this vector, indicating that $C$ determines *which repeated comparisons the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+This process ultimately derives the final comparative outcome $O_{R}$ from the sample-comparison outcome $O^{sc}_{R}$.
+
+Notably, the DAG shows that $C$ is also independent of all other variables in the model.
+This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015, pp.
+8] or Incomplete Block Designs [@Lawson_2015, pp.
+259-271].
+In these designs, every repeated judgment has an equal probability of being included in the sample.
+
 <!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
@@
 
-<!-- Consider algorithm used, as a child of the traits of judges and stimuli.
-It allows to talk about independence assumption between them and the perception of difference.
-Consider algorithm is a categorical variable.
--->
-
-<!-- 2. What is the effect of the sampling mechanism -->
-<!-- Following the convention of @McElreath_2020 [pp.
-499-516], the [DAG -@fig-cj14_dag] depicts the sample-comparison outcome $O^{sc}_{R}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the comparison mechanism.
-Moreover, the DAG shows the conditioned *comparison design* variable $C$, which causally influence the observed outcome $O_{R}$.
-This variable reflects how this observed outcome is a sample drawn from the sample-comparison outcome $O^{sp}_{R}$.
-Specifically, the variable is a vector $C = \{ c_{1}, \dots, c_{n} \}$, where $n$ corresponds to the number of rows in the matrix $R$ (refer to @sec-theory-theoretical_P2), and each $c_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the sample-comparison outcome $O^{sp}_{R}$.
-Specifically, -->
-
-$$
-c_i = \begin{cases} 
-1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is observed} \\
-0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is missing}
-\end{cases}
-$$
+In summary, extending Thurstone's general form with causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2).
+However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a.
+Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we will develop in the next section.
 
@@
 
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
-267] -->
-
 # Abandoning the BTL model {#sec-statistical}
@@
 
-when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification.
+<!-- when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification. -->
 
@@
   e_{J} & := f_{e}(Z_{J}) \\
-  e_{IA} & \dsep e_{JK}
+  e_{IA} & \dsep e_{JK} \\ 
+  \\ 
 \end{aligned}
@@
 
+<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
+(2002).
+Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
+[@Bramley_2008, pp.
+274] -->
+<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges.  -->
+
 <!-- (Is this a validity contention?) -->
@@
 11] -->
-
-<!-- Thus Thurstone’s model (in both its Case 5 and Rasch formulation) achieves the goal of sample-free calibration (Wright, 1977) in the sense that the estimate of the distance between A and B does not depend on which other objects they are compared with.
-[@Bramley_2008, pp.
-256] -->
 
@@
 
-<!-- Due to efficiency of scoring, most of research only collects only one text.
-This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
-However, this cannot be known ad-hoc.
--->
-
-<!-- Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable.
-Relevant variables depends on the context, you really have to think through which variables you have to put in.
--->
+<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
+
+<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
+
+<!-- design matters -->
 
@@
 267] -->
+
+<!-- Thus Thurstone’s model (in both its Case 5 and Rasch formulation) achieves the goal of sample-free calibration (Wright, 1977) in the sense that the estimate of the distance between A and B does not depend on which other objects they are compared with.
+The practical importance of this is that it is not necessary to have a completely crossed or balanced design of paired comparisons because the estimation process naturally handles missing data.
+[@Bramley_2008, pp.
+256] -->
+
+<!-- if the data fit the Case 5 Thurstone model, the estimate of the separation between any two scripts does not depend on which other scripts they are compared with.
+This means that data can be missing in a non-random way without affecting the results.
+However, the precision (standard error) of each script’s estimate depends largely on how many comparisons it has been involved in, so some effort needs to be made to ensure that each script is compared a similar number of times in total.
+[@Bramley_2008, pp.
+260] -->
+
+<!-- Since the analysis can cope with non-random missing data (see earlier), the collection of data can be very flexible – tailored to the needs of the particular situation.
+Because the comparative judgements are all independent, data can be accumulated as needed.
+The investigator is thus not (in principle) bound by a prior design.
+At any point more data could be collected and added to the existing data to recover from accidents, to replace an unacceptable judge, to add extra scripts, to resolve conflicts, to investigate possible bias or to deepen the study of any aspect that proves especially interesting.
+[@Bramley_2008, pp.
+261] -->
+
+<!-- Due to efficiency of scoring, most of research only collects only one text.
+This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
+However, this cannot be known ad-hoc.
+-->
+
+<!-- Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable.
+Relevant variables depends on the context, you really have to think through which variables you have to put in.
+-->
 

### 2025-03-31 _tex/paper2.tex
Built site for gh-pages

@@
 
-In the sample-comparison model, we assume a more realistic scenario than
-the conceptual-population model. Specifically, we assume researchers
-have access to a data sample consisting of a \emph{limited number of
-judgments \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a
-specific number of texts \((n^{s}_{A})\) from a sample of students
-\((n^{s}_{I})\)}, drawn from the conceptual population. In a subsequent
-step, we assume that judges \emph{do not make all possible comparisons}
-allowed by the sample, but rather only \(n_{C}\) comparisons.
+The sample-comparison model presents a more realistic scenario than the
+conceptual-population model. Expressly, we assume researchers work with
+a data sample consisting of a \emph{limited number of judgments
+\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
+number of texts \((n^{s}_{A})\) from a sample of students
+\((n^{s}_{I})\)}, all drawn from the conceptual population. In a
+subsequent step, we assume that judges do not perform \emph{all possible
+repeated comparisons} within the data sample. Instead, they conduct a
+sufficient number of comparisons, \(n_{C}\), to ensure an accurate
+estimation of the proportion \(P(B>A)\).
 
@@
 To incorporate the sampling mechanism and facilitate the interpretation
-of the sample-comparison model, we first define the data sampling
-process using the binary vector variables \(S_{I}\), \(S_{J}\),
+of the sample-comparison model, we first define the \emph{data sampling
+process} using the binary vector variables \(S_{I}\), \(S_{J}\),
 \(S_{IA}\), and \(S_{JK}\) as follows:
@@
 (\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
-of \(S_{JK}\). The sum of its elements represents the total data sample
+of \(S_{JK}\). Thus, the vector contains \(n\) elements, matching the
+number of rows in the \(R\) matrix, and the sum of its elements
+represents the total data sample:
 \(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
@@
 comparisons possible between every pair of sampled stimuli generated by
-each pair of sampled individuals from the population.
+each pair of sampled individuals.
 
@@
 \(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
-matrix (see equation (\ref{eq-mat12})), that is, this vector determines
-\emph{which repeated comparisons the subset of judges makes for the
-subset of stimuli produced by the subset of individuals} sampled from
-the conceptual-population. This process ultimately generates the
-sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
-outcome \(O^{cp}_{R}\).
+matrix (see equation (\ref{eq-mat12})), that is, this vector selects
+\emph{all repeated comparisons a subset of judges makes for the subset
+of stimuli produced by the sampled individuals}. This process ultimately
+generates the sample-comparison outcome \(O^{sc}_{R}\) from the
+conceptual-population outcome \(O^{cp}_{R}\).
 
@@
 (SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
-judgment, judge, stimulus, and individual has the same probability of
-being included in the sample as any other observation within their
-respective groups \citep[pp.~16-112]{Lawson_2015}.
+repeated judgment, judge, stimulus, and individual has the same
+probability of being included in the sample as any other observation
+within their respective groups \citep[pp.~16-112]{Lawson_2015}.
 
@@
 
+As in the previous section, we first define the \emph{comparison design}
+using the binary vector variable \(C\) to facilitate the interpretation
+of the sample-comparison model and integrate the comparison mechanism.
+
+\begin{equation}\phantomsection\label{eq-mat26}{
+C = \begin{bmatrix}
+c_{(1)} \\
+\vdots \\
+c_{(c)} \\
+\vdots \\
+c_{(n)} \\
+\end{bmatrix}
+}\end{equation}
+
+Where each element of \(C\) is a binary value indicating the presence or
+absence of data rows in the \(R\) matrix (see equation
+(\ref{eq-mat12})):
+
 \[
-c_i = \begin{cases} 
-1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is observed} \\
-0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is missing}
+c_{c} = \begin{cases} 
+1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
+0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
 \end{cases}
 \]
+
+Thus, the \emph{comparison design} vector \(C\) contains \(n\) elements
+corresponding to the number of rows in the \(R\) matrix, and the sum of
+its elements represents the selected comparisons, \(n_{C}\). This
+definition assumes that judges do not perform \emph{all possible
+repeated comparisons} within the data sample. Instead, they conduct a
+sufficient number of comparisons, \(n_{C}\), to ensure the accurate
+estimation of the proportion \(P(B>A)\) for each pair of stimuli
+\citep[pp.~267]{Thurstone_1927b}.
+
+With the definition of \(C\), the DAG~\ref{fig-cj14_dag} incorporates
+the sample and comparison mechanism into the conceptual-population
+model. The DAG now also represents the sample-comparison outcome
+\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
+directly access this outcome due to the comparison mechanism. The DAG
+also represents the \emph{comparison design} vector \(C\) as a causal
+factor influencing the final comparative outcome \(O_{R}\). Equivalent
+to the sampling mechanism, a square encloses this vector, indicating
+that \(C\) determines \emph{which repeated comparisons the subset of
+judges makes for the subset of stimuli produced by the sampled
+individuals}. This process ultimately derives the final comparative
+outcome \(O_{R}\) from the sample-comparison outcome \(O^{sc}_{R}\).
+
+Notably, the DAG shows that \(C\) is also independent of all other
+variables in the model. This independence implies that the conceptual
+model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
+Random Allocation Comparative Designs \citep[pp.~8]{Bramley_2015} or
+Incomplete Block Designs \citep[pp.~259-271]{Lawson_2015}. In these
+designs, every repeated judgment has an equal probability of being
+included in the sample.
+
+In summary, extending Thurstone's general form with causal inference
+allows researchers to address BTL modeling issues, such as judge biases
+(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
+stimuli (see Section~\ref{sec-theory-issue1b}), and measurement errors
+in trait estimation and hypothesis testing (see
+Section~\ref{sec-theory-issue2}). However, the model does not address
+concerns about the assumption of equal dispersion discussed in
+Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
+assumptions regarding the discriminal process distribution, it requires
+a statistical model, which we will develop in the next section.
 
@@
 \section{Abandoning the BTL model}\label{sec-statistical}
-
-when you talk about the variability, refer that the equal dispersion is
-just assuming 1 for identification.
 
@@
   e_{J} & := f_{e}(Z_{J}) \\
-  e_{IA} & \:\bot\:e_{JK}
+  e_{IA} & \:\bot\:e_{JK} \\ 
+  \\ 
 \end{aligned}

### 2025-04-04 paper2.qmd
section:all; process:write,figures

@@
 <!-- 3. But are our specific problems? -->
-However, Thurstone designed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
-269], prioritizing statistical simplicity over precise trait measurement.
-He cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], as it imposes the most extensive set of simplifying assumptions among all the proposed cases [@Bramley_2008, pp.
+However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use its measurement scores for hypothesis testing.
+Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008, pp.
 253; @Kelly_et_al_2022, pp.
 677] (see @tbl-thurstone_cases).
-Therefore, it is surprising that, despite these limitations, CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp.
+Moreover, because Thurstone's primary goal was to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
+269] and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp.
+269], he did not address how to use the resulting measurement scores for hypothesis testing.
+Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp.
 677].
+Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?
 
@@
 -->
-
-Furthermore, since Thurstone's primary goal was to define a psychological scale and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp.
-269], he did not provide specific guidance on how to use the measurement scores from Case V to conduct hypothesis tests related to the measured traits.
-Although the CJ practice has circumvented this issue by using the point estimates of the scores or their transformations for such analyses, a fundamental question remains: Is this method appropriate for addressing these inquiries?
 
@@
 
-<!-- ![Cummulative probability](/images/png/cummulative.png){#fig-cumulative width=70%} -->
+<!-- ![Cumulative probability](/images/png/cumulative.png){#fig-cumulative width=70%} -->
 
@@
 <!-- recording: Sven and Tine 24.12.17; time: 00:06:10 - 00:08:50 -->
+<!-- recording: Sven and Tine 25.03.31; time: 00:01:20 - 00:34:00 -->
 
@@
 A thought experiment can help illustrate this idea.
-In this experiment, researchers observe the discriminal processes for two texts, similar to those shown in @fig-discriminal_process.
-Additionally, the discriminal dispersion for Text A remains constant, and the texts are uncorrelated  $(\rho=0)$.
+In it, researchers observe the discriminal processes for two texts, A and B, assuming that the dispersion for Text A remains constant and that the two texts are uncorrelated $(\rho=0)$.
 @fig-dispersion demonstrates that an increase in the uncertainty associated with the perception of Text B relative to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference.
 This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B>A)$, ultimately influencing the comparison outcome.
-Additionally, the figure reveals that when the discriminal dispersions of the texts are equal, as in the BTL model $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference is more narrow than when the dispersions differ.
+Additionally, the figure reveals that when the discriminal dispersions of the texts are equal, as in the BTL model $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference distribution is more narrow than when the dispersions differ.
 As a result, the discriminal difference is more likely to favor Text B over Text A, as it is represented by the shaded gray area.
@@
 
-![Discrepancies in the dispersions of stimuli](/images/png/dispersion.png){#fig-dispersion width=100%}
-
-![Correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}
+![Discriminal Difference distribution under varying discrepancies in stimuli dispersions](/images/png/dispersion.png){#fig-dispersion width=100%}
+
+![Discriminal Difference distribution under varying levels of correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}
 
@@
 <!-- 3. But does it hold in practice? -->
-Notably, while the assumption of equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that the BTL model may fail to capture the complexity of modern CJ data.
-Specifically, the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
+Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data.
+In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
 376; @Bramley_2008, pp.
@@
 678], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-Indeed, the CJ literature may already provide evidence of such discrepancies, particularly in the form of "misfit" statistics.
+Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics.
 For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp.
@@
 290].
+These misfit texts may exhibit larger discriminal dispersions or represent outliers --texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+In either case, comparing misfit texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them.
 
@@
 While Thurstone used his Law to create measurement scales for purely psychological or subjective entities, he did so only with stimuli whose values could be evaluated relatively instantaneously, such as short statements and handwriting samples.
-He was also sceptical about its use for very heterogeneous stimuli (Thurstone, 1927b) [@Kelly_et_al_2022, pp.
+He was also skeptical about its use for very heterogeneous stimuli (Thurstone, 1927b) [@Kelly_et_al_2022, pp.
 678].
@@
 
-<!-- 4. How these judgment discrepancies can happen? -->
-These discrepancies may arise from two possible causes.
-First, the discriminal processes of misfit texts may substantially overlap with those of non-misfit texts.
-Second, misfit texts may represent outlying observations with characteristics that make them difficult to compare and position on the trait scale.
-Regarding the first cause, the overlap can occur when misfit texts have equal dispersion and share the same trait space as non-misfit texts or when they exhibit larger discriminal dispersions.
-Although the BTL model can distinguish the first scenario, it is inherently agnostic to the second.
-@fig-discriminal_process illustrates how an increased discriminal dispersion can widen the overlap between two texts, leading to more judgment discrepancies.
-In the figure, Text B has a broader discriminal dispersion than Text A, expanding the overlap between their discriminal processes, a space where judges struggle to determine which text possesses more of the trait.
-
-<!-- A similar challenge arises in the first scenario, where overlapping trait spaces contribute to judgment inconsistencies. -->
+<!-- 4. How these judgment discrepancies happen?  -->
+<!-- These misfit texts likely arise from two underlying mechanisms.
+First, a text may exhibit a larger discriminal dispersion, increasing the overlap between its discriminal process and those of multiple other texts.
+This greater overlap leads to more judgment discrepancies, increasing the likelihood of classifying the text as a misfit.
+@fig-discriminal_process illustrates this scenario.
+In the figure, Text B has a wider discriminal dispersion than Text A, enlarging the region where their discriminal processes intersect.
+This larger overlap represents an area where judges face greater difficulty in determining which text exhibits a higher level of the trait.
+Second, a text may represent an outlier --one with distinct characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+Such texts are harder to compare and place on the trait scale, resulting in more judgment discrepancies and a misfit classification.
+In either case, the BTL model is inherently agnostic to these mechanisms and does not provide any means to address them.
+-->
 
@@
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
+Furthermore, if researchers acknowledge that misfit statistics may represent outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+12].
+The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
 
@@
 370] -->
-
-Finally, regarding the second cause, if researchers recognize that misfit statistics may represent outlying observations, the conventional CJ practice of  excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
-12].
-The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
 
@@
 <!-- 2. But this is not the scenario we observe -->
-As outlined in the previous section, researchers approach this process in reverse.
+<!-- As outlined in the previous section, researchers approach this process in reverse.
 They begin by observing the outcomes and using the BTL model to estimate the stimuli' discriminal differences and discriminal processes.
@@
 Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts.
+-->
 
@@
 346].
-While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
+While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 and @vanDaal_et_al_2016 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
 
@@
 <!-- 3. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010].
+In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010, pp.
+86].
+For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or general quality.
 Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient.
@@
 Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we refer to as the *conceptual-population model*.
-This model assumes that researchers have access to a *conceptual population* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+This model assumes an idealized scenario where researchers have access to a *conceptual population* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
 Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we refer to as the *sample-comparison model*.
@@
 <!-- recording: Sven and Tine 25.02.07; time: 00:00:00 - 00:02:20 -->
+<!-- recording: Sven and Tine 25.03.31; time: 00:47:05 - 00:48:40 -->
 
 <!-- 1. What is the assumption on the conceptual-population model -->
-In the conceptual-population model, we assume an idealized scenario where researchers have access to a *conceptual population* of comparative data --data representing *all repeated judgments made by each judge for every stimulus pair produced by each pair of individuals* in the population.
-This assumption allows us to integrate Thurstone's theoretical principles and propose innovations to address some of the issues raised in @sec-theory-issues.
+In the conceptual-population model, the idealized scenario of a *conceptual population* of comparative data enables the integration of Thurstone's theoretical principles and provides a foundation for proposing innovations aimed at addressing some of the issues discussed in @sec-theory-issues.
 
@@
 h \\
-\vdots \\
 i \\
@@
 \vdots & \vdots \\
-1 & n_{A}-1 \\
 1 & n_{A} \\
@@
 \vdots & \vdots \\
-n_{I} & n_{A}-1 \\
+n_{I} & 1 \\
+\vdots & \vdots \\
 n_{I} & n_{A} \\
@@
 \vdots & \vdots \\
-1 & n_{K}-1 \\
 1 & n_{K} \\
@@
 \vdots & \vdots \\
-n_{J} & n_{K}-1 \\
+n_{J} & 1 \\
+\vdots & \vdots \\
 n_{J} & n_{K} \\
@@
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
 1 & 1 & 1 & 2 & 1 & n_{K} \\
@@
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
 n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
@@
 <!-- 2 Explaining what we did -->
-The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as the teaching method, causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
-It also shows how $T_{I}$ and *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length) causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest.
-Additionally, the branch accounts for idiosyncratic errors: $e_{I}$ captures variations in students' traits unexplained by $X_{I}$, while $e_{IA}$ captures variations in texts' traits unexplained by $T_{I}$ or $X_{IA}$.
-Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
-Similarly, $X_{IA}$ is an observed matrix with dimension $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
+The top branch of [DAG - @fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
+The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$.
+This branch also shows how $T_{I}$, along with *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' writing-quality trait $(T_{IA})$, the first primary latent variable of interest.
+The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
+In this model, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
+Similarly, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
 
@@
 [^2]: refer to footnote 1.
+
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$.
+It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
+Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
+Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
+
+[^3]: refer to footnote 1.
+[^4]: refer to footnote 1.
+
+Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
+Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.
 
@@
 
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$.
-It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
-Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
-Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
-
-[^3]: refer to footnote 1.
-[^4]: refer to footnote 1.
-
-Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
-Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.
-
 <!-- 3. segway for next section -->
@@
 <!-- recording: Sven 24.10.04; time: 00:34:10 - 00:41:00 -->
+<!-- recording: Sven 25.03.31; time: 00:34:00 - 00:38:50 -->
 
@@
 
+The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
+A square encloses this vector, indicating that it is a conditioned variable.
+In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp.
+32], in this case $S$.
+Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated comparisons a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
+
+Notably, the DAG shows that $S$ is independent of all other variables in the model.
+This implies that the conceptual model represented in [DAG -@fig-cj12_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019, pp.
+150].
+In these designs, each repeated judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp.
+16-112].
+
+<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
+
 ::: {#fig-cj12 layout-nrow=2}
@@
 
-The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
-A square encloses this vector, indicating that it is a conditioned variable.
-In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp.
-32], in this case $S$.
-Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated comparisons a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
-
-Notably, the DAG shows that $S$ is independent of all other variables in the model.
-This implies that the conceptual model represented in [DAG -@fig-cj12_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019, pp.
-150].
-In these designs, each repeated judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp.
-16-112].
-
-<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
-
 <!-- 3. segway for next section -->
@@
 This process ultimately derives the final comparative outcome $O_{R}$ from the sample-comparison outcome $O^{sc}_{R}$.
-
-Notably, the DAG shows that $C$ is also independent of all other variables in the model.
-This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015, pp.
-8] or Incomplete Block Designs [@Lawson_2015, pp.
-259-271].
-In these designs, every repeated judgment has an equal probability of being included in the sample.
-
-<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
-2, Deffner et al (2022) -->
-
-In summary, extending Thurstone's general form with causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2).
-However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a.
-Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we will develop in the next section.
 
@@
 
+Notably, the DAG shows that $C$ is also independent of all other variables in the model.
+This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015, pp.
+8] or Incomplete Block Designs [@Lawson_2015, pp.
+259-271].
+In these designs, every repeated judgment has an equal probability of being included in the sample.
+
+<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
+2, [@Deffner_et_al_2022] -->
+
+In summary, extending Thurstone's general form with causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2).
+However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a.
+Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we will develop in the next section.
+
 # Abandoning the BTL model {#sec-statistical}
@@
 <!-- recording: Sven and Tine 25.03.28; time: 00:33:55 - 00:37:50 -->
+<!-- recording: Sven and Tine 25.03.31; time: 00:38:50 - 00:47:00 -->
+
+<!-- make it short, move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
+-->
+
+<!-- if I have the distribution of the population outcome, and it is a bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
 
@@
 
+<!-- 4. How these judgment discrepancies happen?  -->
+<!-- These misfit texts likely arise from two underlying mechanisms.
+First, a text may exhibit a larger discriminal dispersion, increasing the overlap between its discriminal process and those of multiple other texts.
+This greater overlap leads to more judgment discrepancies, increasing the likelihood of classifying the text as a misfit.
+@fig-discriminal_process illustrates this scenario.
+In the figure, Text B has a wider discriminal dispersion than Text A, enlarging the region where their discriminal processes intersect.
+This larger overlap represents an area where judges face greater difficulty in determining which text exhibits a higher level of the trait.
+Second, a text may represent an outlier --one with distinct characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+Such texts are harder to compare and place on the trait scale, resulting in more judgment discrepancies and a misfit classification.
+In either case, the BTL model is inherently agnostic to these mechanisms and does not provide any means to address them.
+-->
+
 <!-- (Is this a validity contention?) -->
@@
 
-**CRediT authorship contribution statement:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review and editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
+**CRediT authorship contribution statement:** *Conceptualization:* J.M.R.E, T.vD., S.DM., and S.G.; *Methodology:* J.M.R.E, T.vD., and S.DM.; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* T.vD.
+and S.DM.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review and editing:* T.vD., S.DM., and S.G.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
 and S.DM.; *Project administration:* S.G.

### 2025-04-04 _tex/paper2.tex
Built site for gh-pages

@@
 
-However, Thurstone designed Case V to provide a ``rather coarse
-scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
-statistical simplicity over precise trait measurement. He cautioned that
-its use ``should not be made without (an) experimental test''
-\citep[pp.~270]{Thurstone_1927b}, as it imposes the most extensive set
-of simplifying assumptions among all the proposed cases
+However, Thurstone developed Case V prioritizing statistical simplicity
+over precise trait measurement, and without providing clear guidance on
+how to use its measurement scores for hypothesis testing. Specifically,
+he cautioned that its use ``should not be made without (an) experimental
+test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
+extensive set of simplifying assumptions
 \citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
-(see Table~\ref{tbl-thurstone_cases}). Therefore, it is surprising that,
-despite these limitations, CJ research has predominantly relied on Case
-V to measure different traits, raising significant concerns about the
-reliability and validity of such measurements in contexts where the
-case's assumptions may not hold \citep[pp.~677]{Kelly_et_al_2022}.
-
-Furthermore, since Thurstone's primary goal was to define a
-psychological scale and ``allocate the compared stimuli on this
-continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not provide
-specific guidance on how to use the measurement scores from Case V to
-conduct hypothesis tests related to the measured traits. Although the CJ
-practice has circumvented this issue by using the point estimates of the
-scores or their transformations for such analyses, a fundamental
-question remains: Is this method appropriate for addressing these
-inquiries?
+(see Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's
+primary goal was to provide a ``rather coarse scaling'' of traits
+\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
+this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
+how to use the resulting measurement scores for hypothesis testing.
+Thus, given these limitations, it is surprising that CJ research has
+predominantly relied on Case V to measure different traits, raising
+significant concerns about the reliability and validity of such
+measurements in contexts where the case's assumptions may not hold
+\citep[pp.~677]{Kelly_et_al_2022}. Furthermore, although the CJ practice
+has attempted to address the issue of hypothesis testing by using the
+scores' point estimates or their transformations, a critical question
+remains: Is this approach suitable for addressing these inquiries?
 
@@
 influencing the outcome of pairwise comparisons. A thought experiment
-can help illustrate this idea. In this experiment, researchers observe
-the discriminal processes for two texts, similar to those shown in
-Figure~\ref{fig-discriminal_process}. Additionally, the discriminal
-dispersion for Text A remains constant, and the texts are uncorrelated
-\((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that an increase
-in the uncertainty associated with the perception of Text B relative to
-Text A \((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
-discriminal difference. This broadening affects the probability area
-where the discriminal difference distinctly favors Text B over Text A,
-expressed as \(P(B>A)\), ultimately influencing the comparison outcome.
-Additionally, the figure reveals that when the discriminal dispersions
-of the texts are equal, as in the BTL model
-\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference is more
-narrow than when the dispersions differ. As a result, the discriminal
-difference is more likely to favor Text B over Text A, as it is
-represented by the shaded gray area.
+can help illustrate this idea. In it, researchers observe the
+discriminal processes for two texts, A and B, assuming that the
+dispersion for Text A remains constant and that the two texts are
+uncorrelated \((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that
+an increase in the uncertainty associated with the perception of Text B
+relative to Text A \((\sigma_{B} - \sigma_{A})\), broadens the
+distribution of their discriminal difference. This broadening affects
+the probability area where the discriminal difference distinctly favors
+Text B over Text A, expressed as \(P(B>A)\), ultimately influencing the
+comparison outcome. Additionally, the figure reveals that when the
+discriminal dispersions of the texts are equal, as in the BTL model
+\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference distribution
+is more narrow than when the dispersions differ. As a result, the
+discriminal difference is more likely to favor Text B over Text A, as it
+is represented by the shaded gray area.
 
@@
 
-\subcaption{\label{fig-dispersion}Discrepancies in the dispersions of
-stimuli}
+\subcaption{\label{fig-dispersion}Discriminal Difference distribution
+under varying discrepancies in stimuli dispersions}
 
@@
 
-\subcaption{\label{fig-correlation}Correlation between stimuli}
+\subcaption{\label{fig-correlation}Discriminal Difference distribution
+under varying levels of correlation between stimuli}
 
@@
 
-Notably, while the assumption of equal dispersions simplifies the trait
-measurement model, evidence from the CJ literature suggests that the BTL
-model may fail to capture the complexity of modern CJ data.
-Specifically, the assumption of equal dispersions may not hold when
-researchers assess complex traits or heterogeneous stimuli
+Notably, while assuming equal dispersions simplifies the trait
+measurement model, evidence from the CJ literature suggests that this
+assumption may fail to capture the complexity of modern CJ data. In
+particular, the assumption may not hold when researchers assess complex
+traits or heterogeneous stimuli
 \citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
@@
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
-Indeed, the CJ literature may already provide evidence of such
+Indeed, the CJ literature may already provide indications of such
 discrepancies, particularly in the form of ``misfit'' statistics. For
@@
 \citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
-
-These discrepancies may arise from two possible causes. First, the
-discriminal processes of misfit texts may substantially overlap with
-those of non-misfit texts. Second, misfit texts may represent outlying
-observations with characteristics that make them difficult to compare
-and position on the trait scale. Regarding the first cause, the overlap
-can occur when misfit texts have equal dispersion and share the same
-trait space as non-misfit texts or when they exhibit larger discriminal
-dispersions. Although the BTL model can distinguish the first scenario,
-it is inherently agnostic to the second.
-Figure~\ref{fig-discriminal_process} illustrates how an increased
-discriminal dispersion can widen the overlap between two texts, leading
-to more judgment discrepancies. In the figure, Text B has a broader
-discriminal dispersion than Text A, expanding the overlap between their
-discriminal processes, a space where judges struggle to determine which
-text possesses more of the trait.
+These misfit texts may exhibit larger discriminal dispersions or
+represent outliers --texts with distinctive characteristics that deviate
+markedly from the rest of the sample \citep{Grubbs_1969}. In either
+case, comparing misfit texts results in more judgment discrepancies, yet
+the BTL model neither accounts for these cases nor offers any means to
+address them.
 
@@
 difference follows any discontinuous line where
-\(\sigma_{B}-\sigma_{A} \neq 0\).
-
-Finally, regarding the second cause, if researchers recognize that
-misfit statistics may represent outlying observations, the conventional
-CJ practice of excluding stimuli based on these statistics
+\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
+acknowledge that misfit statistics may represent outlying observations,
+the common CJ practice of excluding stimuli based on these statistics
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
@@
 
-As outlined in the previous section, researchers approach this process
-in reverse. They begin by observing the outcomes and using the BTL model
-to estimate the stimuli' discriminal differences and discriminal
-processes. Given that the BTL model assumes independent discriminal
-processes across comparisons, if this assumption holds in the data, the
-model estimates a discriminal difference distribution that accurately
-reflects the \emph{true} discriminal difference between the texts. Once
-more, the accuracy of the discriminal difference ensures reliable
-estimates for the discriminal processes of the texts.
-
 Thurstone assumed that stimuli were uncorrelated because judges' biases,
@@
 evidence for this particular scenario is lacking, studies such as
-\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
-supporting the notion that the factors influencing pairwise comparisons
-may not always cancel out.
+\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
+the presence of such biases, supporting the notion that the factors
+influencing pairwise comparisons may not always cancel out.
 
@@
 between stimuli, a statistical phenomenon known as clustering
-\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
-existence of such hierarchical structures, the statistical approaches to
-account for this additional source of dependence have been insufficient.
-For instance, when CJ data incorporates multiple samples of stimuli from
-the same individuals, researchers frequently rely on (averaged) point
-estimates of the BTL scores to conduct subsequent analyses and tests at
-the individual level
+\citep[pp.~86]{Everitt_et_al_2010}. For instance, when multiple texts
+are produced by the same individual, these texts are likely to share
+common features, such as writing style or general quality. Although the
+CJ literature acknowledges the existence of such hierarchical
+structures, the statistical approaches to account for this additional
+source of dependence have been insufficient. For instance, when CJ data
+incorporates multiple samples of stimuli from the same individuals,
+researchers frequently rely on (averaged) point estimates of the BTL
+scores to conduct subsequent analyses and tests at the individual level
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 theoretical principles into what we refer to as the
-\emph{conceptual-population model}. This model assumes that researchers
-have access to a \emph{conceptual population} of comparative data, that
-is, data representing all repeated judgments made by every available
-judge for each pair of stimuli produced by each pair of individuals in
-the population. Conversely, Section~\ref{sec-theory-theoretical_SC}
-integrates the assessment design features into what we refer to as the
-\emph{sample-comparison model}. This model assumes a more realistic
-scenario where researchers only have access to a sample of judges,
-individuals, stimuli, and comparisons from the conceptual population.
+\emph{conceptual-population model}. This model assumes an idealized
+scenario where researchers have access to a \emph{conceptual population}
+of comparative data, that is, data representing all repeated judgments
+made by every available judge for each pair of stimuli produced by each
+pair of individuals in the population. Conversely,
+Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
+features into what we refer to as the \emph{sample-comparison model}.
+This model assumes a more realistic scenario where researchers only have
+access to a sample of judges, individuals, stimuli, and comparisons from
+the conceptual population.
 
@@
 
-In the conceptual-population model, we assume an idealized scenario
-where researchers have access to a \emph{conceptual population} of
-comparative data --data representing \emph{all repeated judgments made
-by each judge for every stimulus pair produced by each pair of
-individuals} in the population. This assumption allows us to integrate
-Thurstone's theoretical principles and propose innovations to address
-some of the issues raised in Section~\ref{sec-theory-issues}.
+In the conceptual-population model, the idealized scenario of a
+\emph{conceptual population} of comparative data enables the integration
+of Thurstone's theoretical principles and provides a foundation for
+proposing innovations aimed at addressing some of the issues discussed
+in Section~\ref{sec-theory-issues}.
 
@@
 
-The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
+The top branch of DAG -~\ref{fig-cj09_dag} illustrates the first
 component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
@@
   \citet{Hernan_et_al_2020}, are recommended.} student-related variables
-\((X_{I})\), such as the teaching method, causally influence the latent
-variable representing students' writing-quality trait \((T_{I})\). It
-also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
-  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
-influence the texts' written-quality trait \((T_{IA})\), the first
-primary latent variable of interest. Additionally, the branch accounts
-for idiosyncratic errors: \(e_{I}\) captures variations in students'
-traits unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in
-texts' traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is
-an observed matrix with \(n_{I}\) rows and \(q_{I}\) columns
-(variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors of
-length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
-dimension \(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables),
-while \(e_{IA}\) and \(T_{IA}\) are latent vectors of length
-\(n_{I} \cdot n_{A}\).
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} 
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj09_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
-
-}
-
-\subcaption{\label{fig-cj09_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj09}Conceptual-population model, final vectorized
-form.}
-
-\end{figure}%
+\((X_{I})\), such as teaching method, and students' idiosyncratic errors
+\(e_{I}\) causally influence the latent variable representing students'
+writing-quality trait \((T_{I})\). The error term \(e_{I}\) captures
+variations in students' traits unexplained by \(X_{I}\). This branch
+also shows how \(T_{I}\), along with \emph{relevant} \footnote{refer to
+  footnote 1.} text-related variables \((X_{IA})\) (e.g., text length),
+and texts' idiosyncratic errors \(e_{IA}\) causally influence the texts'
+writing-quality trait \((T_{IA})\), the first primary latent variable of
+interest. The error term \(e_{IA}\) captures variations in the texts'
+traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). In this
+model, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
+columns (variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors
+of length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
+dimensions \(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns
+(variables), while \(e_{IA}\) and \(T_{IA}\) are latent vectors of
+length \(n_{I} \cdot n_{A}\).
 
@@
 \(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj09_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
+
+}
+
+\subcaption{\label{fig-cj09_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj09}Conceptual-population model, final vectorized
+form.}
+
+\end{figure}%
 
@@
 
+The DAG also depicts the \emph{sample design} vector \(S\) as a causal
+factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
+square encloses this vector, indicating that it is a conditioned
+variable. In this context, \emph{conditioning} means restricting the
+statistical analysis to a specific subset of the population based on the
+values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
+\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
+matrix (see equation (\ref{eq-mat12})), that is, this vector selects
+\emph{all repeated comparisons a subset of judges makes for the subset
+of stimuli produced by the sampled individuals}. This process ultimately
+generates the sample-comparison outcome \(O^{sc}_{R}\) from the
+conceptual-population outcome \(O^{cp}_{R}\).
+
+Notably, the DAG shows that \(S\) is independent of all other variables
+in the model. This implies that the conceptual model represented in
+DAG~\ref{fig-cj12_dag} applies exclusively to Simple Random Sampling
+(SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
+repeated judgment, judge, stimulus, and individual has the same
+probability of being included in the sample as any other observation
+within their respective groups \citep[pp.~16-112]{Lawson_2015}.
+
 \begin{figure}[H]
@@
 
-The DAG also depicts the \emph{sample design} vector \(S\) as a causal
-factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
-square encloses this vector, indicating that it is a conditioned
-variable. In this context, \emph{conditioning} means restricting the
-statistical analysis to a specific subset of the population based on the
-values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
-\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
-matrix (see equation (\ref{eq-mat12})), that is, this vector selects
-\emph{all repeated comparisons a subset of judges makes for the subset
-of stimuli produced by the sampled individuals}. This process ultimately
-generates the sample-comparison outcome \(O^{sc}_{R}\) from the
-conceptual-population outcome \(O^{cp}_{R}\).
-
-Notably, the DAG shows that \(S\) is independent of all other variables
-in the model. This implies that the conceptual model represented in
-DAG~\ref{fig-cj12_dag} applies exclusively to Simple Random Sampling
-(SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
-repeated judgment, judge, stimulus, and individual has the same
-probability of being included in the sample as any other observation
-within their respective groups \citep[pp.~16-112]{Lawson_2015}.
-
 However, because CJ experiments rarely use the exhaustive pairing of
@@
 outcome \(O_{R}\) from the sample-comparison outcome \(O^{sc}_{R}\).
-
-Notably, the DAG shows that \(C\) is also independent of all other
-variables in the model. This independence implies that the conceptual
-model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
-Random Allocation Comparative Designs \citep[pp.~8]{Bramley_2015} or
-Incomplete Block Designs \citep[pp.~259-271]{Lawson_2015}. In these
-designs, every repeated judgment has an equal probability of being
-included in the sample.
-
-In summary, extending Thurstone's general form with causal inference
-allows researchers to address BTL modeling issues, such as judge biases
-(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
-stimuli (see Section~\ref{sec-theory-issue1b}), and measurement errors
-in trait estimation and hypothesis testing (see
-Section~\ref{sec-theory-issue2}). However, the model does not address
-concerns about the assumption of equal dispersion discussed in
-Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
-assumptions regarding the discriminal process distribution, it requires
-a statistical model, which we will develop in the next section.
 
@@
 \end{figure}%
+
+Notably, the DAG shows that \(C\) is also independent of all other
+variables in the model. This independence implies that the conceptual
+model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
+Random Allocation Comparative Designs \citep[pp.~8]{Bramley_2015} or
+Incomplete Block Designs \citep[pp.~259-271]{Lawson_2015}. In these
+designs, every repeated judgment has an equal probability of being
+included in the sample.
+
+In summary, extending Thurstone's general form with causal inference
+allows researchers to address BTL modeling issues, such as judge biases
+(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
+stimuli (see Section~\ref{sec-theory-issue1b}), and measurement errors
+in trait estimation and hypothesis testing (see
+Section~\ref{sec-theory-issue2}). However, the model does not address
+concerns about the assumption of equal dispersion discussed in
+Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
+assumptions regarding the discriminal process distribution, it requires
+a statistical model, which we will develop in the next section.
 
@@
 \textbf{CRediT authorship contribution statement:}
-\emph{Conceptualization:} S.G., S.DM., T.vD., and J.M.R.E;
-\emph{Methodology:} S.DM., T.vD., and J.M.R.E; \emph{Software:}
+\emph{Conceptualization:} J.M.R.E, T.vD., S.DM., and S.G.;
+\emph{Methodology:} J.M.R.E, T.vD., and S.DM.; \emph{Software:}
 J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal Analysis:} J.M.R.E.;
-\emph{Investigation:} J.M.R.E; \emph{Resources:} S.G., S.DM., and T.vD.;
+\emph{Investigation:} J.M.R.E; \emph{Resources:} T.vD. and S.DM.;
 \emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
-J.M.R.E.; \emph{Writing - review and editing:} S.G., S.DM., and T.vD.;
+J.M.R.E.; \emph{Writing - review and editing:} T.vD., S.DM., and S.G.;
 \emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;

### 2025-04-04 paper2.qmd
section:all; process:write,figures

@@
 <!-- 5. So, how do SCMs and DAGs support identification? -->
-Luckily, SCMs and DAGs support identification analysis through two key advantages.
-First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks.
-This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@Neal_2020; @McElreath_2020].
+Fortunately, SCMs and DAGs support identification analysis through two key advantages.
+First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [see @Neal_2020; @McElreath_2020].
+This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis.
 Second, they depict causal relationships in a non-parametric way.
@@
 <!-- 2. the preliminary SCM for CJ -->
-[SCM -@fig-cj03_scm] illustrates the inclusion of the first theoretical principles into the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability.
+[SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the conceptual-population model, which evaluates the impact of different teaching methods on students' writing ability.
 This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
-The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli).
-The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp.
-366-376].
-Notably, the indexing system facilitates comparisons between texts written by the same student $(h = i, a \neq b)$ and between texts written by distinct students ($h \neq i$, where $a = b$ is possible).
-However, it excludes cases where judges compare a student's text to itself $(h = i, a = b)$, as such comparison lacks practical relevance within the CJ framework.
+The subscripts $h$ and $i$ identify the students who authored the texts (i.e., the individuals).
+The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli).
+The index $j$ indicates the judge conducting the comparison, while the index $k$ captures *repeated measures designs* [@Lawson_2015, pp.
+366-376], accounting for experimental conditions where a judge compares the same pair of stimuli multiple times.
+Notably, the indexing system supports comparisons between different texts written by the same student $(h = i;$ $a \neq b)$ and between texts written by distinct students $(h \neq i;$ $a = b \; \text{is permitted})$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K)$.
+However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(h = i;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K})$, as such comparison lacks practical relevance within the CJ framework.
+Here, $n_{J}$ indicates the total number of judges, and $n_{K}$ denotes the number of repeated judgments each judge performs.
 
@@
 Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
-
 $$
@@
 h \\
+\vdots \\
 i \\
@@
 n_{J} & 1 \\
+n_{J} & 2 \\
 \vdots & \vdots \\
@@
 Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$.
-Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.
+Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
 
@@
 Thus, we define the matrix as follows:
-
-<!-- $$ -->
-<!-- {n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! } -->
-<!-- $$ -->
 
@@
 In this simple case, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
-
 $$
@@
 
-::: {#fig-cj04 layout-nrow=2}
-
-::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
-
-Conceptual-population model, vectorized form.
-:::
-
 <!-- 3. how de we use that data structure? -->
@@
 Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$, respectively.
+
+::: {#fig-cj04 layout-nrow=2}
+
+::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
+
+Conceptual-population model, vectorized form.
+:::
 
@@
 
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$.
-It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
-Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $(B_{J})$.
+The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
+This branch also shows how $B_{J}$, along with *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
 Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
@@
 These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population* of comparative data, encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population* of comparative data.
 Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
@@
 The sample-comparison model presents a more realistic scenario than the conceptual-population model.
-Expressly, we assume researchers work with a data sample consisting of a *limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$*, all drawn from the conceptual population.
-In a subsequent step, we assume that judges do not perform *all possible repeated comparisons* within the data sample.
-Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.
+Expressly, it assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
+In a subsequent step, it assumes that judges do not perform *all possible stimuli comparisons* within the data sample.
+Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.
 
@@
 To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the *data sampling process* using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:
-
 $$
@@
 Specifically, we define the vector $S_{I}$ as in equation ([-@eq-mat22]) and apply a similar definition to $S_{J}$.
-
 $$
@@
 Specifically, we define the vector $S_{IA}$ as in equation ([-@eq-mat23]) and apply a similar definition to $S_{JK}$.
-
 $$
@@
 Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, these vectors take the following structure:
-
 $$
@@
 Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$.
-Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.
-
-Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
+Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled repeated judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.
+
+Finally, we define the *sample mechanism* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
 Thus, the vector contains $n$ elements, matching the number of rows in the $R$ matrix, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
 Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
-
 $$
@@
 499-516] and @Deffner_et_al_2022 [pp.
-6], the [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.
-
+6], [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.
 The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
@@
 32], in this case $S$.
-Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated comparisons a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
 This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
@@
 <!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
+
+<!-- 3. segway for next section -->
+However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
+5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
 
@@
 
-<!-- 3. segway for next section -->
-However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
-5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
-
 ### The comparison mechanism {#sec-theory-theoretical_SC2}
@@
 <!-- 1. start by defining the comparison vector  -->
-As in the previous section, we first define the *comparison design* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model and integrate the comparison mechanism.
-
+As in the previous section, to facilitate the interpretation of the sample-comparison model, we first define the *comparison mechanism* using the binary vector variable $C$ as follows: 
 $$
@@
 
-Where each element of $C$ is a binary value indicating the presence or absence of data rows in the $R$ matrix (see equation ([-@eq-mat12])):
-
-$$
-c_{c} = \begin{cases} 
+Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element being a binary value indicating the presence or absence of data rows in the $R$ matrix (see equation ([-@eq-mat12])):
+$$
+c_{(c)} = \begin{cases} 
 1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
@@
 
-Thus, the *comparison design* vector $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, and the sum of its elements represents the selected comparisons, $n_{C}$.
-This definition assumes that judges do not perform *all possible repeated comparisons* within the data sample.
-Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp.
+This definition assumes that the subset of judges *do not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals.
+Instead, they conduct a "sufficient number of comparisons", $n_{C}$, to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp.
 267].
@@
 This process ultimately derives the final comparative outcome $O_{R}$ from the sample-comparison outcome $O^{sc}_{R}$.
+
+Notably, the DAG shows that $C$ is also independent of all other variables in the model.
+This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015, pp.
+8] or Incomplete Block Designs [@Lawson_2015, pp.
+259-271].
+In these designs, every repeated judgment has an equal probability of being included in the sample.
+
+<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
+2, [@Deffner_et_al_2022] -->
 
@@
 
-Notably, the DAG shows that $C$ is also independent of all other variables in the model.
-This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015, pp.
-8] or Incomplete Block Designs [@Lawson_2015, pp.
-259-271].
-In these designs, every repeated judgment has an equal probability of being included in the sample.
-
-<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
-2, [@Deffner_et_al_2022] -->
-
-In summary, extending Thurstone's general form with causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2).
+In summary, extending Thurstone's general form using Causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2).
 However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a.

### 2025-04-04 _tex/paper2.tex
Built site for gh-pages

@@
 
-Luckily, SCMs and DAGs support identification analysis through two key
-advantages. First, regardless of complexity, they can represent various
-causal structures using only five fundamental building blocks. This
-feature allows researchers to decompose complex structures into
-manageable components, facilitating their analysis
-\citep{Neal_2020, McElreath_2020}. Second, they depict causal
-relationships in a non-parametric way. This flexibility enables feasible
-identification strategies without requiring specification of the types
-of variables, the functional forms relating them, or the parameters of
-those functional forms \citep[pp.~35]{Pearl_et_al_2016}.
+Fortunately, SCMs and DAGs support identification analysis through two
+key advantages. First, regardless of complexity, they can represent
+various causal structures using only five fundamental building blocks
+\citep[see][]{Neal_2020, McElreath_2020}. This feature allows
+researchers to decompose complex structures into manageable components,
+facilitating their analysis. Second, they depict causal relationships in
+a non-parametric way. This flexibility enables feasible identification
+strategies without requiring specification of the types of variables,
+the functional forms relating them, or the parameters of those
+functional forms \citep[pp.~35]{Pearl_et_al_2016}.
 
@@
 
-SCM~\ref{fig-cj03_scm} illustrates the inclusion of the first
-theoretical principles into the conceptual-population model designed to
-examine the impact of different teaching methods on students' writing
-ability. This SCM outlines the relationship between the
-conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
-variables. The subscripts \(h\) and \(i\) label the students who
-authored the texts (individuals), while the indices \(a\) and \(b\)
-represent the compared texts (stimuli). The index \(j\) refers to the
-judge, and \(k\) represents the judgment index, accounting for
-experimental conditions where judges perform the same judgment multiple
-times, i.e., a \emph{repeated measures design}
-\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system
-facilitates comparisons between texts written by the same student
-\((h = i, a \neq b)\) and between texts written by distinct students
-(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
-where judges compare a student's text to itself \((h = i, a = b)\), as
-such comparison lacks practical relevance within the CJ framework.
+SCM~\ref{fig-cj03_scm} presents the first theoretical principles
+embedded in the conceptual-population model, which evaluates the impact
+of different teaching methods on students' writing ability. This SCM
+outlines the relationship between the conceptual-population outcome
+\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
+\(h\) and \(i\) identify the students who authored the texts (i.e., the
+individuals). The indices \(a\) and \(b\) represent the texts under
+comparison (i.e., the stimuli). The index \(j\) indicates the judge
+conducting the comparison, while the index \(k\) captures \emph{repeated
+measures designs} \citep[pp.~366-376]{Lawson_2015}, accounting for
+experimental conditions where a judge compares the same pair of stimuli
+multiple times. Notably, the indexing system supports comparisons
+between different texts written by the same student \((h = i;\)
+\(a \neq b)\) and between texts written by distinct students
+\((h \neq i;\) \(a = b \; \text{is permitted})\), each judged once or
+repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K)\).
+However, it excludes cases where a judge compares a student's text to
+itself, whether once or multiple times \((h = i;\) \(a = b;\)
+\(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K})\), as such comparison lacks
+practical relevance within the CJ framework. Here, \(n_{J}\) indicates
+the total number of judges, and \(n_{K}\) denotes the number of repeated
+judgments each judge performs.
 
@@
 the matrices \(IA\) and \(JK\), as follows:
-
 \begin{equation}\phantomsection\label{eq-mat11}{
@@
 Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
-columns, where \(n_{K}\) indicates the number of judgments each judge
-makes.
+columns, where \(n_{K}\) indicates the number of repeated judgments each
+judge makes.
 
@@
 take the following form:
-
 \begin{equation}\phantomsection\label{eq-mat13}{
@@
 
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj04_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
-
-}
-
-\subcaption{\label{fig-cj04_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}
-
-\end{figure}%
-
 Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
@@
 \(n_{J} \cdot n_{K}\), respectively.
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj04_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+
+}
+
+\subcaption{\label{fig-cj04_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}
+
+\end{figure}%
 
@@
 second component, where \emph{relevant} \footnote{refer to footnote 1.}
-judge-related variables \((Z_{J})\), such as judgment expertise,
-causally influence the latent variable representing judges' bias
-\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
+judge-related variables \((Z_{J})\), such as judgment expertise, and
+judges' idiosyncratic errors \(e_{J}\) causally influence the latent
+variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
+captures variations in judges' bias unexplained by \(Z_{J}\). This
+branch also shows how \(B_{J}\), along with \emph{relevant} \footnote{refer
   to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
-number of judgments a judge makes) causally influence the biases
-associated with each text \(B_{JK}\), the second primary latent variable
-of interest. Additionally, the branch accounts for idiosyncratic errors:
-\(e_{J}\) captures variations in judges' bias unexplained by \(Z_{J}\),
-while \(e_{JK}\) captures variations in judgments unexplained by
-\(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
-\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
-\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
-is an observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and
-\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
-latent vectors of length \(n_{J} \cdot n_{K}\).
+number of judgments a judge makes), and judgments' idiosyncratic errors
+\(e_{JK}\) causally influence the biases associated with each text
+\(B_{JK}\), the second primary latent variable of interest. The error
+\(e_{JK}\) captures variations in judgments unexplained by \(B_{J}\) or
+\(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with \(n_{J}\) rows
+and \(q_{J}\) columns (variables), and both \(e_{J}\) and \(B_{J}\) are
+latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\) is an observed
+matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\) columns
+(variables), while \(e_{JK}\) and \(B_{JK}\) are latent vectors of
+length \(n_{J} \cdot n_{K}\).
 
@@
 the unrealistic assumption that researchers have access to a
-\emph{conceptual population} of comparative data, encompassing all
-repeated judgments from every judge for each pair of stimuli produced by
-every individual in the population. Since researchers rarely meet this
-assumption in practice, they must consider a more realistic scenario.
-The extensions to the conceptual-population model, derived from this
-realistic scenario, are described in the sample-comparison model.
+\emph{conceptual population} of comparative data. Since researchers
+rarely meet this assumption in practice, they must consider a more
+realistic scenario. The extensions to the conceptual-population model,
+derived from this realistic scenario, are described in the
+sample-comparison model.
 
@@
 The sample-comparison model presents a more realistic scenario than the
-conceptual-population model. Expressly, we assume researchers work with
-a data sample consisting of a \emph{limited number of judgments
+conceptual-population model. Expressly, it assumes researchers work with
+a data sample consisting of a limited number of repeated judgments
 \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
 number of texts \((n^{s}_{A})\) from a sample of students
-\((n^{s}_{I})\)}, all drawn from the conceptual population. In a
-subsequent step, we assume that judges do not perform \emph{all possible
-repeated comparisons} within the data sample. Instead, they conduct a
-sufficient number of comparisons, \(n_{C}\), to ensure an accurate
-estimation of the proportion \(P(B>A)\).
+\((n^{s}_{I})\), all drawn from the conceptual population. In a
+subsequent step, it assumes that judges do not perform \emph{all
+possible stimuli comparisons} within the data sample. Instead, they
+conduct a sufficient number of stimuli comparisons, \(n_{C}\), to ensure
+an accurate estimation of the proportion \(P(B>A)\).
 
@@
 \(S_{IA}\), and \(S_{JK}\) as follows:
-
 \begin{equation}\phantomsection\label{eq-mat21}{
@@
 equation (\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).
-
 \begin{equation}\phantomsection\label{eq-mat22}{
@@
 in equation (\ref{eq-mat23}) and apply a similar definition to
-\(S_{JK}\).
-
-\begin{equation}\phantomsection\label{eq-mat23}{
+\(S_{JK}\). \begin{equation}\phantomsection\label{eq-mat23}{
 ia_{(ia)} = \begin{cases} 
@@
 these vectors take the following structure:
-
 \begin{equation}\phantomsection\label{eq-mat24}{
@@
 texts across all sampled students \((n^{s}_{I} \cdot n^{s}_{A})\), while
-a similar sum in vector \(S_{JK}\) represents the total sampled
+a similar sum in vector \(S_{JK}\) represents the total sampled repeated
 judgments across all sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\).
 
-Finally, we define the \emph{sample design} vector \(S\) in equation
+Finally, we define the \emph{sample mechanism} vector \(S\) in equation
 (\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
@@
 each pair of sampled individuals.
-
 \begin{equation}\phantomsection\label{eq-mat25}{
@@
 \citet[pp.~499-516]{McElreath_2020} and
-\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} represents
-the conceptual-population outcome \(O^{cp}_{R}\) as unobserved,
-emphasizing that researchers cannot directly access this outcome due to
-the sampling mechanism.
-
-The DAG also depicts the \emph{sample design} vector \(S\) as a causal
-factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
-square encloses this vector, indicating that it is a conditioned
-variable. In this context, \emph{conditioning} means restricting the
-statistical analysis to a specific subset of the population based on the
-values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
-\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
-matrix (see equation (\ref{eq-mat12})), that is, this vector selects
-\emph{all repeated comparisons a subset of judges makes for the subset
-of stimuli produced by the sampled individuals}. This process ultimately
-generates the sample-comparison outcome \(O^{sc}_{R}\) from the
-conceptual-population outcome \(O^{cp}_{R}\).
+\citet[pp.~6]{Deffner_et_al_2022}, DAG~\ref{fig-cj12_dag} represents the
+conceptual-population outcome \(O^{cp}_{R}\) as unobserved, emphasizing
+that researchers cannot directly access this outcome due to the sampling
+mechanism. The DAG also depicts the \emph{sample design} vector \(S\) as
+a causal factor influencing the sample-comparison outcome
+\(O^{sc}_{R}\). A square encloses this vector, indicating that it is a
+conditioned variable. In this context, \emph{conditioning} means
+restricting the statistical analysis to a specific subset of the
+population based on the values of a conditioned variable
+\citep[pp.~32]{Neal_2020}, in this case \(S\). Essentially, \(S\) serves
+as the sample equivalent for the \(R\) matrix (see equation
+(\ref{eq-mat12})), that is, this vector selects \emph{all repeated
+judgments a subset of judges makes for the subset of stimuli produced by
+the sampled individuals}. This process ultimately generates the
+sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
+outcome \(O^{cp}_{R}\).
 
@@
 within their respective groups \citep[pp.~16-112]{Lawson_2015}.
+
+However, because CJ experiments rarely use the exhaustive pairing of
+sampled judges, stimuli, and individuals --due to concerns about the
+practical feasibility of the comparison task, like in
+\citet[p.~5]{Boonen_et_al_2020}-- a realistic scenario must also account
+for judges comparing only specific stimuli from certain individuals.
 
@@
 
-However, because CJ experiments rarely use the exhaustive pairing of
-sampled judges, stimuli, and individuals --due to concerns about the
-practical feasibility of the comparison task, like in
-\citet[p.~5]{Boonen_et_al_2020}-- a realistic scenario must also account
-for judges comparing only specific stimuli from certain individuals.
-
 \subsubsection{The comparison
@@
 
-As in the previous section, we first define the \emph{comparison design}
-using the binary vector variable \(C\) to facilitate the interpretation
-of the sample-comparison model and integrate the comparison mechanism.
-
+As in the previous section, to facilitate the interpretation of the
+sample-comparison model, we first define the \emph{comparison mechanism}
+using the binary vector variable \(C\) as follows:
 \begin{equation}\phantomsection\label{eq-mat26}{
@@
 
-Where each element of \(C\) is a binary value indicating the presence or
-absence of data rows in the \(R\) matrix (see equation
-(\ref{eq-mat12})):
-
-\[
-c_{c} = \begin{cases} 
+Where \(C\) contains \(n\) elements corresponding to the number of rows
+in the \(R\) matrix, with each element being a binary value indicating
+the presence or absence of data rows in the \(R\) matrix (see equation
+(\ref{eq-mat12})): \[
+c_{(c)} = \begin{cases} 
 1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
@@
 
-Thus, the \emph{comparison design} vector \(C\) contains \(n\) elements
-corresponding to the number of rows in the \(R\) matrix, and the sum of
-its elements represents the selected comparisons, \(n_{C}\). This
-definition assumes that judges do not perform \emph{all possible
-repeated comparisons} within the data sample. Instead, they conduct a
-sufficient number of comparisons, \(n_{C}\), to ensure the accurate
-estimation of the proportion \(P(B>A)\) for each pair of stimuli
+This definition assumes that the subset of judges \emph{do not} perform
+all repeated judgments for the subset of stimuli produced by the sampled
+individuals. Instead, they conduct a ``sufficient number of
+comparisons'', \(n_{C}\), to ensure the accurate estimation of the
+proportion \(P(B>A)\) for each pair of stimuli
 \citep[pp.~267]{Thurstone_1927b}.
@@
 outcome \(O_{R}\) from the sample-comparison outcome \(O^{sc}_{R}\).
+
+Notably, the DAG shows that \(C\) is also independent of all other
+variables in the model. This independence implies that the conceptual
+model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
+Random Allocation Comparative Designs \citep[pp.~8]{Bramley_2015} or
+Incomplete Block Designs \citep[pp.~259-271]{Lawson_2015}. In these
+designs, every repeated judgment has an equal probability of being
+included in the sample.
 
@@
 
-Notably, the DAG shows that \(C\) is also independent of all other
-variables in the model. This independence implies that the conceptual
-model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
-Random Allocation Comparative Designs \citep[pp.~8]{Bramley_2015} or
-Incomplete Block Designs \citep[pp.~259-271]{Lawson_2015}. In these
-designs, every repeated judgment has an equal probability of being
-included in the sample.
-
-In summary, extending Thurstone's general form with causal inference
+In summary, extending Thurstone's general form using Causal inference
 allows researchers to address BTL modeling issues, such as judge biases

### 2025-04-07 paper2.qmd
section:all; process:write,figures
section:all; process:write,figures,code

@@
 Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
-This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates  [@Perron_et_al_2015, pp.
-2].
+This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015].
 First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data.
@@
 Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.
+
+<!-- [@Perron_et_al_2015, pp. 2] -->
 
@@
 <!-- 1. Thurstone's Theory of CJ: discriminal process -->
-In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b, pp.
-267].
+In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b].
 The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
 The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait.
-The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp.
-266].
+The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b].
 The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.
+
+<!-- [@Thurstone_1927b, pp. 267] -->
 
@@
 However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
-This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp.
-251].
+This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008].
 This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum.
@@
 In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.
+
+<!-- [@Bramley_2008, pp. 251] -->
 
@@
 Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008, pp.
-253; @Kelly_et_al_2022, pp.
-677] (see @tbl-thurstone_cases).
+270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases).
 Moreover, because Thurstone's primary goal was to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
@@
 269], he did not address how to use the resulting measurement scores for hypothesis testing.
-Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp.
-677].
+Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022].
 Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?
+
+<!-- [@Bramley_2008, pp. 253] -->
 
@@
 This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case.
-The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp.
-254] (see @tbl-thurstone_cases).
-However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I, pp.
-16; @McElreath_2021].
+The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases).
+However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021].
 Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.
+
+<!-- [@Bramley_2008, pp. 254] -->
+
+<!-- [@vanderLinden_et_al_2017_I, pp. 16] -->
 
@@
 <!-- 2. But what are limitations of Case V (and the BTL model)? -->
-However, Thurstone acknowledged that some assumptions of Case V could be problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376].
+However, Thurstone acknowledged that some assumptions of Case V could be problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a].
 Thus, given that modern CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice.
 These assumptions are the equal dispersion and zero correlation between stimuli.
+
+<!-- [@Thurstone_1927a, pp. 376] -->
 
@@
 In experimental practice, however, the thought experiment occurs in reverse.
-Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp.
-373].
-Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp.
-150], in this case, the assumption of equal dispersions.
+Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a].
+Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019], in this case, the assumption of equal dispersions.
 For instance, when the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts.
@@
 Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data.
-In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp.
-376; @Bramley_2008, pp.
-273; @Kelly_et_al_2022, pp.
-678], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Bramley_2008; @Kelly_et_al_2022], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
 Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics.
-For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp.
-11; @Pollitt_2012b, pp.
-290].
+For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b].
 These misfit texts may exhibit larger discriminal dispersions or represent outliers --texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
@@
 <!-- So it seem that the extreme of the discriminal process distribution of the stimuli have higher reliability than the center? -->
+
+<!-- [@Pollitt_2004, pp. 11]  -->
+<!-- [@Pollitt_2012b, pp. 290] -->
 
@@
 Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
-This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp.
-370] and, by extension, about the underlying discriminal processes of stimuli.
+This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020] and, by extension, about the underlying discriminal processes of stimuli.
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
@@
 <!-- 3. But does it hold?  -->
-Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b, pp.
-268].
+Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b].
 @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
 However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.
+
+<!-- [@Thurstone_1927b, pp. 268] -->
 
@@
 For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments.
-Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp.
-346].
+Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
 While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 and @vanDaal_et_al_2016 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
@@
 <!-- 3. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010, pp.
-86].
+In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010].
 For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or general quality.
@@
 
+<!-- [@Everitt_et_al_2010, pp. 86] -->
+
 <!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
@@
 Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues.
-Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023, pp.
-341, 482].
-These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020, pp.
-370] and, by extension, the underlying discriminal processes of the stimuli.
+Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023].
+These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli.
 This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$.
@@
 
+<!-- [@McElreath_2020, pp. 370] -->
+
 Finally, as discussed in the previous section, removing *misfit* judges based risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
@@
 The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis.
-*Misfit judges* are those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a, pp.
-164-165; @Pollitt_2012b, pp.
-289-290; @vanDaal_et_al_2016, pp.
-4; @Goossens_et_al_2018, pp.
-20].
+*Misfit judges* are those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
 
 <!-- These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b. -->
+
+<!-- [@Pollitt_2012a, pp. 164-165]  -->
+<!-- [@Pollitt_2012b, pp. 289-290] -->
+<!-- [@vanDaal_et_al_2016, pp. 4] -->
+<!-- [@Goossens_et_al_2018, pp. 20] -->
 
@@
 <!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Building on the previous section, it is clear that, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp.
-269].
+Building on the previous section, it is clear that, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b].
 Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests.
 For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+
+<!-- [@Thurstone_1927b, pp. 269] -->
 
@@
 The direction and magnitude of such biases are often unpredictable.
-Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp.
-25; @Hoyle_et_al_2023, pp.
-137].
+Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023; @Hoyle_et_al_2023].
 Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
@@
 
+<!-- [@Kline_et_al_2023, pp. 25]  -->
+<!-- [@Hoyle_et_al_2023, pp. 137] -->
+
 ![The effect of outcome uncertainty $\sigma_{T}$ on the estimation of an effect $\beta_{X}$ linked to the predictor.
@@
 In aggregate, the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates.
-This overreliance may also undermine their validity [@Perron_et_al_2015, pp.
-2], particularly when coupled with the disconnect between the trait measurement and hypothesis testing processes.
+This overreliance may also undermine their validity [@Perron_et_al_2015], particularly when coupled with the disconnect between the trait measurement and hypothesis testing processes.
 However, the structural approach to causal inference can address these issues by offering a systematic and integrated framework that strengthens measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.
@@
 The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments.
-Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
-4].
+Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests.
 *Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
@@
 Without identification, consistency cannot be achieved, even with "infinite" and error-free data.
-Thus, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023, pp.
-5].
+Thus, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023].
 
@@
 <!-- 5. So, how do SCMs and DAGs support identification? -->
-Fortunately, SCMs and DAGs support identification analysis through two key advantages.
-First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [see @Neal_2020; @McElreath_2020].
+Fortunately, SCMs and DAGs support identification analysis through two key advantages[^1].
+First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
 This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis.
 Second, they depict causal relationships in a non-parametric way.
-This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp.
-35].
+This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016].
+
+<!-- [@Pearl_et_al_2016, pp. 35] -->
+
+[^1]: These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, and introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 are useful.
+For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020, are recommended.
 
@@
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
-In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
-27,68].
+In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009].
 Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other endogenous and exogenous variables.
@@
 
+<!-- [@Pearl_2009, pp. 27,68] -->
+
 <!-- 2. the preliminary SCM for CJ -->
 [SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the conceptual-population model, which evaluates the impact of different teaching methods on students' writing ability.
-This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables.
+This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{haibjk})$ and several related variables.
 The subscripts $h$ and $i$ identify the students who authored the texts (i.e., the individuals).
@@
 366-376], accounting for experimental conditions where a judge compares the same pair of stimuli multiple times.
-Notably, the indexing system supports comparisons between different texts written by the same student $(h = i;$ $a \neq b)$ and between texts written by distinct students $(h \neq i;$ $a = b \; \text{is permitted})$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K)$.
+Thus, the indexing system supports comparisons between different texts written by the same student $(h = i;$ $a \neq b)$ and between texts written by distinct students $(h \neq i;$ $a = b \; \text{is permitted})$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K)$.
 However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(h = i;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K})$, as such comparison lacks practical relevance within the CJ framework.
@@
 
-In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory).
+In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{haibjk})$ (see @sec-thurstone_theory).
 Additionally, based on the arguments developed in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
-Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$.
+Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{haibjk}, D_{haibjk}, T_{ha}, T_{ib}, B_{kj} \}$.
 Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
@@
 \begin{aligned}
-  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
-  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
+  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
+  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
 \end{aligned}
@@
 Additionally, the DAG differentiates between observed endogenous variables, such as the outcome (solid black circle), and latent endogenous variables, including the texts' discriminal processes, their discriminal difference, and the judges' biases (open circles).
-
-<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
-<!-- 1.
-No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
--->
-<!-- 2.
-No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
--->
-<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
-Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
-[@Lash_et_al_2021, pp.
-113] -->
 
@@
 As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
-Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$.
+Likewise, each row of $JK$ associates a judge $j$ with a (repeated) judgment index $k$.
 Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
@@
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-h & i & a & b & j & k \\
+h & a & i & b & j & k \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
-n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 2 \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K} \\
 \end{bmatrix}
@@
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-1 & 1 & 5 & 1 & 1 & 1 \\
-1 & 1 & 5 & 1 & 1 & 2 \\
-1 & 1 & 5 & 1 & 1 & 3 \\
+1 & 1 & 5 & 2 & 1 & 1 \\
+1 & 1 & 5 & 2 & 1 & 2 \\
+1 & 1 & 5 & 2 & 1 & 3 \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-1 & 1 & 5 & 1 & 3 & 1 \\
-1 & 1 & 5 & 1 & 3 & 2 \\
-1 & 1 & 5 & 1 & 3 & 3 \\
-\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
+4 & 2 & 5 & 2 & 3 & 1 \\
+4 & 2 & 5 & 2 & 3 & 2 \\
+4 & 2 & 5 & 2 & 3 & 3 \\
 5 & 1 & 5 & 2 & 3 & 1 \\
@@
 
-Conceptual-population model, vectorized form.
+Conceptual-population model, initial vectorized form.
 :::
@@
 <!-- 1. What are we going to do? -->
-Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
-138] and Item Response Theory (IRT) [@Fox_2010, chap.
-6; @vanderLinden_et_al_2017_I, chap.
-24], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability.
+Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023] and Item Response Theory (IRT) [@Fox_2010; @vanderLinden_et_al_2017_I], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability.
 Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
@@
 
+<!-- [@Hoyle_et_al_2023, pp. 138] -->
+<!-- [@Fox_2010, chap. 6]  -->
+<!-- [@vanderLinden_et_al_2017_I, chap. 24] -->
+
 <!-- 2 Explaining what we did -->
-The top branch of [DAG - @fig-cj09_dag] illustrates the first component, where *relevant* [^1] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
+The top branch of [DAG - @fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
 The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$.
-This branch also shows how $T_{I}$, along with *relevant* [^2] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' writing-quality trait $(T_{IA})$, the first primary latent variable of interest.
+This branch also shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' writing-quality trait $(T_{IA})$, the first primary latent variable of interest.
 The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
@@
 
-[^1]: Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+[^2]: Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
 A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
-These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, as well as introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 useful.
-For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020, are recommended.
-
-[^2]: refer to footnote 1.
-
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $(B_{J})$.
+Refer also to footnote 1.
+
+[^3]: refer to footnote 2.
+
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $(Z_{J})$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $(B_{J})$.
 The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
-This branch also shows how $B_{J}$, along with *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+This branch also shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
@@
 
-[^3]: refer to footnote 1.
-[^4]: refer to footnote 1.
+[^4]: refer to footnote 2.
+[^5]: refer to footnote 2.
 
@@
 Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.
+
+<!-- 3. segway for next section -->
+Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
+These enhancements include accounting for judges' biases and integrating hierarchical structural components.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population*.
+Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
 
@@
 
-<!-- 3. segway for next section -->
-Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
-These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population* of comparative data.
-Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
-The extensions to the conceptual-population model, derived from this realistic scenario, are described in the sample-comparison model.
-
 ## The sample-comparison model {#sec-theory-theoretical_SC}
@@
 Expressly, it assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
-In a subsequent step, it assumes that judges do not perform *all possible stimuli comparisons* within the data sample.
+In a subsequent step, it assumes that judges do not perform *all repeated judgments* within the data sample.
 Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.
@@
 i_{(i)} = \begin{cases} 
-1 & \text{if data element } i \text{ from } I \text{ is observed} \\
+1 & \text{if data element } i \text{ from } I \text{ is sampled} \\
 0 & \text{if data element } i \text{ from } I \text{ is missing}
@@
 ia_{(ia)} = \begin{cases} 
-1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
+1 & \text{if data elements } i,a \text{ from } IA \text{ are sampled} \\
 0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
@@
 Suppose researchers exclude the second student and the third judge.
-Additionally, they do not include the second text from each individual and the third judgment from each judge.
+They also omit the second text from each individual and the third repeated judgment from each judge.
 Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, these vectors take the following structure:
@@
 
-Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$.
+Equation ([-@eq-mat24]) shows that missing observations in the vectors $S_{I}$ and $S_{J}$--which represent unsampled students and judges--directly determine which observations are missing in $S_{IA}$ and $S_{JK}$.
 In other words, researchers can only observe texts and judgments from students and judges initially included in the sample.
-Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$.
+The equation also shows that the sum of observed elements in $S_{I}$ equals the number of sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$.
 Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled repeated judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.
-
-Finally, we define the *sample mechanism* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
-Thus, the vector contains $n$ elements, matching the number of rows in the $R$ matrix, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
-Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
+Notice that because the design systematically excludes every third repeated judgment, researchers can also express $S_{JK}$ using $n_{K} = n^{s}_{K} = 2$.
 $$
@@
 
-<!-- 2. What is the effect of the sampling mechanism -->
+$$
+s_{(s)} = \begin{cases} 
+1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
+0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
+\end{cases}
+$$ {#eq-mat26}
+
+Finally, we define the *sample mechanism* $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
+Each element $s_{(s)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism (see equation ([-@eq-mat26])).
+Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
+Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
+
+<!-- 2. Integrate the sample mechanism -->
 With the definition of $S$, we incorporate the sample mechanism into the conceptual-population model.
-Following the convention of @McElreath_2020 [pp.
-499-516] and @Deffner_et_al_2022 [pp.
-6], [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.
+Following the convention of @McElreath_2020 and @Deffner_et_al_2022, [DAG -@fig-cj14_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.
 The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
 A square encloses this vector, indicating that it is a conditioned variable.
-In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp.
-32], in this case $S$.
-Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+Here, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp.
+32].
+Essentially, $S$ serves as the sample equivalent for the $R$ matrix, that is, this vector selects *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
 This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
 
+<!-- [@McElreath_2020, pp. 499-516] -->
+<!-- [@Deffner_et_al_2022, pp. 6] -->
+
 Notably, the DAG shows that $S$ is independent of all other variables in the model.
-This implies that the conceptual model represented in [DAG -@fig-cj12_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019, pp.
-150].
-In these designs, each repeated judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp.
-16-112].
+This implies that [DAG -@fig-cj14_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019].
+In these designs, each repeated judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015].
+
+<!-- [@Kohler_et_al_2019, pp. 150] -->
+<!-- [@Lawson_2015, pp. 16-112] -->
 
@@
 <!-- 3. segway for next section -->
-However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p.
-5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
-
-::: {#fig-cj12 layout-nrow=2}
-
-::: {#fig-cj12_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \dsep e_{JK} 
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_12.png){#fig-cj12_dag fig-align="center" fig-pos="H" width=84%}
-
-Sample-comparison model including sampling mechanism
-:::
-
-### The comparison mechanism {#sec-theory-theoretical_SC2}
-
-<!-- 1. start by defining the comparison vector  -->
-As in the previous section, to facilitate the interpretation of the sample-comparison model, we first define the *comparison mechanism* using the binary vector variable $C$ as follows: 
-$$
-C = \begin{bmatrix}
-c_{(1)} \\
-\vdots \\
-c_{(c)} \\
-\vdots \\
-c_{(n)} \\
-\end{bmatrix}
-$$ {#eq-mat26}
-
-Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element being a binary value indicating the presence or absence of data rows in the $R$ matrix (see equation ([-@eq-mat12])):
-$$
-c_{(c)} = \begin{cases} 
-1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
-0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
-\end{cases}
-$$
-
-This definition assumes that the subset of judges *do not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals.
-Instead, they conduct a "sufficient number of comparisons", $n_{C}$, to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp.
-267].
-
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
-267] -->
-
-<!-- 2. What is the effect of the comparison mechanism -->
-With the definition of $C$, the [DAG -@fig-cj14_dag] incorporates the sample and comparison mechanism into the conceptual-population model.
-The DAG now also represents the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
-The DAG also represents the *comparison design* vector $C$ as a causal factor influencing the final comparative outcome $O_{R}$.
-Equivalent to the sampling mechanism, a square encloses this vector, indicating that $C$ determines *which repeated comparisons the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-This process ultimately derives the final comparative outcome $O_{R}$ from the sample-comparison outcome $O^{sc}_{R}$.
-
-Notably, the DAG shows that $C$ is also independent of all other variables in the model.
-This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015, pp.
-8] or Incomplete Block Designs [@Lawson_2015, pp.
-259-271].
-In these designs, every repeated judgment has an equal probability of being included in the sample.
-
-<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
-2, [@Deffner_et_al_2022] -->
+However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
+
+<!-- [@Boonen_et_al_2020, p. 5] -->
 
@@
 
-Sample-comparison model including sample and comparison mechanisms
-:::
-
-In summary, extending Thurstone's general form using Causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2).
-However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a.
-Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we will develop in the next section.
+Sample-comparison model, final vectorized form
+:::
+
+### The comparison mechanism {#sec-theory-theoretical_SC2}
+
+<!-- 1. start by defining the comparison vector  -->
+As in the previous section, to facilitate the interpretation of the sample-comparison model, we first define the *comparison mechanism* using the binary vector variable $C$ as follows: 
+$$
+C = \begin{bmatrix}
+c_{(1)} \\
+\vdots \\
+c_{(c)} \\
+\vdots \\
+c_{(n)} \\
+\end{bmatrix}
+$$ {#eq-mat27}
+
+Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(c)}$ being a binary value indicating the presence or absence of data rows in the $R$ similar to equation ([-@eq-mat26]).
+This definition assumes that the subset of judges *does not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals.
+Instead, they conduct a "sufficient number of comparisons" $n_{C}$ to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp.
+267].
+
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
+<!-- 2. What is the effect of the comparison mechanism -->
+[DAG -@fig-cj14_dag] also incorporates the comparison mechanism $C$ into the conceptual-population model.
+The DAG represents the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
+Moreover, it represents the *comparison mechanism* $C$ as a causal factor influencing the final comparative outcome $O_{R}$.
+Equivalent to the sampling mechanism, a square encloses this vector, indicating that $C$ is a conditioned variable that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+This process ultimately derives the final comparative outcome $O_{R}$ from the sample-comparison outcome $O^{sc}_{R}$.
+
+Notably, the DAG shows that $C$ is also independent of all other variables in the model.
+This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015] or Incomplete Block Designs [@Lawson_2015].
+In these designs, every repeated judgment has an equal probability of being included in the sample.
+
+<!-- [@Bramley_2015, pp. 8] -->
+<!-- [@Lawson_2015, pp. 259-271] -->
+
+<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
+2, [@Deffner_et_al_2022] -->
+
+Finally, without loss of generality, researchers can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
+This reformulation rests on two principles.
+First, because the sampling and comparison mechanism vectors $S$ and $C$ are binary, researchers can integrate them into the model using element-wise products.
+This approach eliminates the need to define the structural equations $f_{S}$ and $f_{C}$.
+Second, since only the final comparative outcome $O_{R}$ is observed, it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$, defined by the structural equation $f_{O}$, also applies to this outcome $O_{R}$.
+This assumption removes the necessity of describing the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$.
+
+<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
+For example, if I have the distribution of the population outcome, and it is a bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
+
+::: {#fig-cj15 layout-nrow=2}
+
+::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{R} & := f_{O}(D_{R}, S, C) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \dsep e_{JK} 
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_15.png){#fig-cj15_dag fig-align="center" fig-pos="H" width=72%}
+
+Comparative judgment model
+:::
+
+In summary, extending Thurstone's general form allows researchers to address issues with the BTL model, such as accounting for judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2).
+However, it does not address concerns about stimuli' equal dispersions discussed in @sec-theory-issue1a.
+Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we develop in the next section.
 
@@
 
-<!-- make it short, move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
--->
-
-<!-- if I have the distribution of the population outcome, and it is a bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
+Using [SCM -@fig-cj15_scm] we finally define the
+
+$\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1$ $\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1$
 
@@
 
-::: {#fig-cj15 layout-ncol=2}
-
-::: {#fig-cj15_stat fig-align="center" fig-pos="H"}
+::: {#fig-cj16 layout-ncol=2}
+
+::: {#fig-cj16_stat fig-align="center" fig-pos="H"}
 $$
 \begin{aligned}
-  O_{R} & = O^{sc}_{R} \odot C \\
-  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
-  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
-  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
-  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
+  D_{R} & = \left( T_{IA}[R_{1},R_{2}] - T_{IA}[R_{3},R_{4}] \right) + B_{JK}[R_{5},R_{6}] \\
+  T_{IA} & = T_{I}[IA_{1}] + \beta_{XA} X_{IA} + e_{IA} \\
   T_{I} & = \beta_{XI} X_{I} + e_{I} \\
-  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{JK} & = B_{J}[JK_{1}] + \beta_{ZK} Z_{JK} + e_{JK} \\
   B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
@@
   e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK}) \\ \\
-  & \textbf{Contraints:} \\
-  & \sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1 ; \quad \sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1
+  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
 \end{aligned}
@@
 
-::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
+::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
 $$
 \begin{aligned}
-  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
-  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  O_{R} & := f_{O}(D_{R}, S, C) \\
   D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & := f_{e}(X_{I}) \\
-  e_{J} & := f_{e}(Z_{J}) \\
-  e_{IA} & \dsep e_{JK} \\ 
-  \\ 
+  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \dsep e_{JK} 
 \end{aligned}
@@
 
-Sample-comparison model, including sample and comparison mechanisms and assuming different discriminal dispersions for the student's traits
+Comparative judgment model, SCM and statistical model assuming different discriminal dispersions for the student's traits
 :::
@@
 
+<!-- Brought Rubin's missing data framework to the field. -->
+<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are are random samples, they fulfill ignorability.  -->
+
 <!-- Statements about the real world from observed data require assumptions.
@@
 
+<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
+<!-- 1.
+No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
+-->
+<!-- 2.
+No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
+-->
+<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
+Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
+[@Lash_et_al_2021, pp.
+113] -->
+
 ## Limitations and further research {#sec-discussion-limitations}
@@
 -->
-
-<!-- Brought Rubin's missing data framework to the field. -->
 

### 2025-04-07 _tex/paper2.tex
Built site for gh-pages

@@
 reliability and validity of CJ's trait estimates
-\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
-on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
+\citep{Perron_et_al_2015}. First, it critiques the heavy reliance on
+Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
 extension, the Bradley-Terry-Luce (BTL) model
@@
 comparisons wherein a single judge evaluates multiple stimuli
-\citep[pp.~267]{Thurstone_1927b}. The theory posits that two key factors
+\citep{Thurstone_1927b}. The theory posits that two key factors
 determine the dichotomous outcome of these comparisons: the discriminal
@@
 given stimulus forms a Normal distribution along the trait continuum
-\citep[pp.~266]{Thurstone_1927b}. The mode (mean) of this distribution,
-known as the \emph{modal discriminal process}, indicates the stimulus
-position on this continuum, while its dispersion, referred to as the
+\citep{Thurstone_1927b}. The mode (mean) of this distribution, known as
+the \emph{modal discriminal process}, indicates the stimulus position on
+this continuum, while its dispersion, referred to as the
 \emph{discriminal dispersion}, reflects variability in the perceived
@@
 further along the trait continuum as possessing more of the trait
-\citep[pp.~251]{Bramley_2008}. This suggests that pairwise comparison
-outcomes depend on the relative distance between stimuli, not their
-absolute positions on the continuum. Indeed, the theory assumes that the
+\citep{Bramley_2008}. This suggests that pairwise comparison outcomes
+depend on the relative distance between stimuli, not their absolute
+positions on the continuum. Indeed, the theory assumes that the
 difference between the underlying discriminal processes of the stimuli,
@@
 extensive set of simplifying assumptions
-\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
-(see Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's
-primary goal was to provide a ``rather coarse scaling'' of traits
+\citep{Bramley_2008, Kelly_et_al_2022} (see
+Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
+goal was to provide a ``rather coarse scaling'' of traits
 \citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
@@
 measurements in contexts where the case's assumptions may not hold
-\citep[pp.~677]{Kelly_et_al_2022}. Furthermore, although the CJ practice
-has attempted to address the issue of hypothesis testing by using the
+\citep{Kelly_et_al_2022}. Furthermore, although the CJ practice has
+attempted to address the issue of hypothesis testing by using the
 scores' point estimates or their transformations, a critical question
@@
 processes, the BTL model uses the more mathematically tractable Logistic
-distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
+distribution \citep{Andrich_1978, Bramley_2008} (see
 Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
@@
 discriminal process scale is arbitrary up to a non-monotonic
-transformation
-\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2021}}.
+transformation \citep{vanderLinden_et_al_2017_I, McElreath_2021}.
 Furthermore, this limited impact is supported by the fact that the
@@
 problematic when researchers assess complex traits or heterogeneous
-stimuli \citep[pp.~376]{Thurstone_1927a}. Thus, given that modern CJ
-applications often involve such traits and stimuli, two key assumptions
-of Case V, and by extension, the BTL model, may not always hold in
-theory or practice. These assumptions are the equal dispersion and zero
+stimuli \citep{Thurstone_1927a}. Thus, given that modern CJ applications
+often involve such traits and stimuli, two key assumptions of Case V,
+and by extension, the BTL model, may not always hold in theory or
+practice. These assumptions are the equal dispersion and zero
 correlation between stimuli.
@@
 the BTL model to infer the discriminal difference between stimuli and
-their respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
+their respective discriminal processes \citep{Thurstone_1927a}.
 Consequently, the outcome's ability to reflect \emph{true} differences
 between stimuli largely depends on the validity of the model's
-assumptions \citep[pp.~150]{Kohler_et_al_2019}, in this case, the
-assumption of equal dispersions. For instance, when the assumption
-accurately captures the complexity of the data, the BTL model estimates
-a discriminal difference distribution that accurately represents the
-\emph{true} discriminal difference between the texts. This scenario is
-illustrated in Figure~\ref{fig-dispersion}, when the model's discriminal
-difference distribution aligns with the \emph{true} discriminal
-difference distribution, represented by the thick continuous line
-corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
-discriminal difference then ensures reliable estimates for the texts'
-discriminal processes.
+assumptions \citep{Kohler_et_al_2019}, in this case, the assumption of
+equal dispersions. For instance, when the assumption accurately captures
+the complexity of the data, the BTL model estimates a discriminal
+difference distribution that accurately represents the \emph{true}
+discriminal difference between the texts. This scenario is illustrated
+in Figure~\ref{fig-dispersion}, when the model's discriminal difference
+distribution aligns with the \emph{true} discriminal difference
+distribution, represented by the thick continuous line corresponding to
+\(\sigma_{B}-\sigma_{A}=0\). The accuracy of this discriminal difference
+then ensures reliable estimates for the texts' discriminal processes.
 
@@
 traits or heterogeneous stimuli
-\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
-as these traits and stimuli can introduce judgment discrepancies due to
-their unique characteristics
+\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
+and stimuli can introduce judgment discrepancies due to their unique
+characteristics
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
@@
 judgment discrepancies than those involving other texts
-\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
-These misfit texts may exhibit larger discriminal dispersions or
-represent outliers --texts with distinctive characteristics that deviate
-markedly from the rest of the sample \citep{Grubbs_1969}. In either
-case, comparing misfit texts results in more judgment discrepancies, yet
-the BTL model neither accounts for these cases nor offers any means to
-address them.
+\citep{Pollitt_2004, Pollitt_2012b}. These misfit texts may exhibit
+larger discriminal dispersions or represent outliers --texts with
+distinctive characteristics that deviate markedly from the rest of the
+sample \citep{Grubbs_1969}. In either case, comparing misfit texts
+results in more judgment discrepancies, yet the BTL model neither
+accounts for these cases nor offers any means to address them.
 
@@
 overestimation, in turn, results in spurious conclusions about these
-differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
+differences \citep{McElreath_2020} and, by extension, about the
 underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
@@
 the pairwise comparisons, canceled each other out
-\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
-mathematical demonstration of this cancellation using the BTL model
-under the assumption of discriminal processes with additive biases.
-However, evidence from the CJ literature indicates that the assumption
-of zero correlation does not hold in practice in at least two scenarios:
-when intricate aspects of multidimensional, complex traits or
-heterogeneous stimuli influence judges' perceptions or when additional
-hierarchical structures are relevant to the stimuli.
+\citep{Thurstone_1927b}. \citet{Andrich_1978} provided a mathematical
+demonstration of this cancellation using the BTL model under the
+assumption of discriminal processes with additive biases. However,
+evidence from the CJ literature indicates that the assumption of zero
+correlation does not hold in practice in at least two scenarios: when
+intricate aspects of multidimensional, complex traits or heterogeneous
+stimuli influence judges' perceptions or when additional hierarchical
+structures are relevant to the stimuli.
 
@@
 perceptions, these biases could introduce dependencies between the
-stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While direct
-evidence for this particular scenario is lacking, studies such as
+stimuli \citep{vanderLinden_et_al_2017_II}. While direct evidence for
+this particular scenario is lacking, studies such as
 \citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
@@
 between stimuli, a statistical phenomenon known as clustering
-\citep[pp.~86]{Everitt_et_al_2010}. For instance, when multiple texts
-are produced by the same individual, these texts are likely to share
-common features, such as writing style or general quality. Although the
-CJ literature acknowledges the existence of such hierarchical
-structures, the statistical approaches to account for this additional
-source of dependence have been insufficient. For instance, when CJ data
+\citep{Everitt_et_al_2010}. For instance, when multiple texts are
+produced by the same individual, these texts are likely to share common
+features, such as writing style or general quality. Although the CJ
+literature acknowledges the existence of such hierarchical structures,
+the statistical approaches to account for this additional source of
+dependence have been insufficient. For instance, when CJ data
 incorporates multiple samples of stimuli from the same individuals,
@@
 dimensional mismatches in the model, leading to the over- or
-underestimation of trait reliability \citep[pp.~341,
-482]{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result in
-spurious conclusions about the discriminal differences
-\citep[pp.~370]{McElreath_2020} and, by extension, the underlying
-discriminal processes of the stimuli. This issue is illustrated in
+underestimation of trait reliability
+\citep{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result
+in spurious conclusions about the discriminal differences
+\citep{McElreath_2020} and, by extension, the underlying discriminal
+processes of the stimuli. This issue is illustrated in
 Figure~\ref{fig-correlation} when the discriminal difference
@@
 judges themselves
-\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
 
@@
 typically rely on the BTL model to measure a trait and place the
-compared stimuli along its continuum \citep[pp.~269]{Thurstone_1927b}.
+compared stimuli along its continuum \citep{Thurstone_1927b}.
 Additionally, the CJ literature shows that researchers frequently use
@@
 uncertainty in the scores and the actual effects being tested
-\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
-Furthermore, the reduced precision in hypothesis tests diminishes their
-statistical power, increasing the likelihood of committing type-I or
-type-II errors \citep{McElreath_2020}.
-Figure~\ref{fig-measurement_error} illustrates these issues,
-demonstrating how neglecting measurement error \((\sigma_{T})\) by
-relying only on outcome averages can reduce the precision of a
-predictor's estimated effect.
+\citep{Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore, the reduced
+precision in hypothesis tests diminishes their statistical power,
+increasing the likelihood of committing type-I or type-II errors
+\citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
+these issues, demonstrating how neglecting measurement error
+\((\sigma_{T})\) by relying only on outcome averages can reduce the
+precision of a predictor's estimated effect.
 
@@
 reliability of trait estimates. This overreliance may also undermine
-their validity \citep[pp.~2]{Perron_et_al_2015}, particularly when
-coupled with the disconnect between the trait measurement and hypothesis
-testing processes. However, the structural approach to causal inference
-can address these issues by offering a systematic and integrated
-framework that strengthens measurement reliability and validity while
-enhancing the statistical accuracy of hypothesis tests.
+their validity \citep{Perron_et_al_2015}, particularly when coupled with
+the disconnect between the trait measurement and hypothesis testing
+processes. However, the structural approach to causal inference can
+address these issues by offering a systematic and integrated framework
+that strengthens measurement reliability and validity while enhancing
+the statistical accuracy of hypothesis tests.
 
@@
 DAGs function as \emph{conceptual models} on which identification
-analysis rests \citep[pp.~4]{Schuessler_et_al_2023}.
-\emph{Identification analysis} helps researchers to determine whether an
-estimator can accurately compute an estimand based solely on its
-(causal) assumptions, regardless of random variability
+analysis rests. \emph{Identification analysis} helps researchers to
+determine whether an estimator can accurately compute an estimand based
+solely on its (causal) assumptions, regardless of random variability
 \citep[pp.~4]{Schuessler_et_al_2023}. Here, \emph{estimands} represent
@@
 and error-free data. Thus, deriving meaningful insights from finite data
-becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}.
+becomes impossible \citep{Schuessler_et_al_2023}.
 
 Fortunately, SCMs and DAGs support identification analysis through two
-key advantages. First, regardless of complexity, they can represent
-various causal structures using only five fundamental building blocks
-\citep[see][]{Neal_2020, McElreath_2020}. This feature allows
-researchers to decompose complex structures into manageable components,
-facilitating their analysis. Second, they depict causal relationships in
-a non-parametric way. This flexibility enables feasible identification
-strategies without requiring specification of the types of variables,
-the functional forms relating them, or the parameters of those
-functional forms \citep[pp.~35]{Pearl_et_al_2016}.
+key advantages\footnote{These topics are beyond the scope of this study,
+  thus, readers seeking a more profound understanding can refer to
+  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
+  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, and introductory
+  books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
+  \citet{McElreath_2020} are useful. For more advanced study, seminal
+  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
+  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
+  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
+  \citet{Hernan_et_al_2020}, are recommended.}. First, regardless of
+complexity, they can represent various causal structures using only five
+fundamental building blocks \citep{Neal_2020, McElreath_2020}. This
+feature allows researchers to decompose complex structures into
+manageable components, facilitating their analysis. Second, they depict
+causal relationships in a non-parametric way. This flexibility enables
+feasible identification strategies without requiring specification of
+the types of variables, the functional forms relating them, or the
+parameters of those functional forms \citep{Pearl_et_al_2016}.
 
@@
 \emph{errors} or \emph{disturbances} arising from omitted factors that
-the investigator chooses not to model explicitly
-\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
-\emph{structural equations}, express the endogenous variables as
-non-parametric functions of other endogenous and exogenous variables.
-These functions use the symbol `\(:=\)' to denote the asymmetrical
-causal dependence between variables and the symbol `\(\:\bot\:\)' to
-represent \emph{d-separation}, a concept akin to (conditional)
-independence.
+the investigator chooses not to model explicitly \citep{Pearl_2009}.
+Lastly, the functions, referred to as \emph{structural equations},
+express the endogenous variables as non-parametric functions of other
+endogenous and exogenous variables. These functions use the symbol
+`\(:=\)' to denote the asymmetrical causal dependence between variables
+and the symbol `\(\:\bot\:\)' to represent \emph{d-separation}, a
+concept akin to (conditional) independence.
 
@@
 outlines the relationship between the conceptual-population outcome
-\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
+\((O^{cp}_{haibjk})\) and several related variables. The subscripts
 \(h\) and \(i\) identify the students who authored the texts (i.e., the
@@
 experimental conditions where a judge compares the same pair of stimuli
-multiple times. Notably, the indexing system supports comparisons
-between different texts written by the same student \((h = i;\)
-\(a \neq b)\) and between texts written by distinct students
-\((h \neq i;\) \(a = b \; \text{is permitted})\), each judged once or
-repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K)\).
-However, it excludes cases where a judge compares a student's text to
-itself, whether once or multiple times \((h = i;\) \(a = b;\)
+multiple times. Thus, the indexing system supports comparisons between
+different texts written by the same student \((h = i;\) \(a \neq b)\)
+and between texts written by distinct students \((h \neq i;\)
+\(a = b \; \text{is permitted})\), each judged once or repeatedly by all
+judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K)\). However, it
+excludes cases where a judge compares a student's text to itself,
+whether once or multiple times \((h = i;\) \(a = b;\)
 \(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K})\), as such comparison lacks
@@
 texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
-difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
+difference \((D_{haibjk})\) (see Section~\ref{sec-thurstone_theory}).
 Additionally, based on the arguments developed in
@@
 variables constitute the preliminary set of endogenous variables,
-\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
+\(V = \{ O_{haibjk}, D_{haibjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
 the SCM presents the preliminary set of structural equations,
@@
 the number of stimuli available per individual. Likewise, each row of
-\(JK\) associates a judge \(j\) with a judgment index \(k\).
+\(JK\) associates a judge \(j\) with a (repeated) judgment index \(k\).
 Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
@@
 
-\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}
+\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
+form.}
 
@@
 Building on the principles of Structural Equation Modeling (SEM)
-\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
-\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
-the conceptual-population model integrates two \emph{hierarchical
-structural components} to examine how different teaching methods
-influence students' writing ability. Each structural component defines
-how observed or latent variables affect the primary latent variable of
-interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
+\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
+\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
+model integrates two \emph{hierarchical structural components} to
+examine how different teaching methods influence students' writing
+ability. Each structural component defines how observed or latent
+variables affect the primary latent variable of interest
+\citep{Everitt_et_al_2010}. Their hierarchical nature enables
 researchers to test hypotheses considering the hierarchical structure of
@@
   between a predictor and an outcome without opening new ones
-  \citep{Pearl_2009}. These topics are beyond the scope of this study,
-  thus, readers seeking a more profound understanding can refer to
-  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
-  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
-  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020},
-  and \citet{McElreath_2020} useful. For more advanced study, seminal
-  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
-  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
-  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
-  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
-\((X_{I})\), such as teaching method, and students' idiosyncratic errors
-\(e_{I}\) causally influence the latent variable representing students'
-writing-quality trait \((T_{I})\). The error term \(e_{I}\) captures
-variations in students' traits unexplained by \(X_{I}\). This branch
-also shows how \(T_{I}\), along with \emph{relevant} \footnote{refer to
-  footnote 1.} text-related variables \((X_{IA})\) (e.g., text length),
-and texts' idiosyncratic errors \(e_{IA}\) causally influence the texts'
-writing-quality trait \((T_{IA})\), the first primary latent variable of
-interest. The error term \(e_{IA}\) captures variations in the texts'
-traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). In this
-model, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
-columns (variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors
-of length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
-dimensions \(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns
-(variables), while \(e_{IA}\) and \(T_{IA}\) are latent vectors of
-length \(n_{I} \cdot n_{A}\).
+  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
+variables \((X_{I})\), such as teaching method, and students'
+idiosyncratic errors \(e_{I}\) causally influence the latent variable
+representing students' writing-quality trait \((T_{I})\). The error term
+\(e_{I}\) captures variations in students' traits unexplained by
+\(X_{I}\). This branch also shows how \(T_{I}\), along with
+\emph{relevant} \footnote{refer to footnote 2.} text-related variables
+\((X_{IA})\) (e.g., text length), and texts' idiosyncratic errors
+\(e_{IA}\) causally influence the texts' writing-quality trait
+\((T_{IA})\), the first primary latent variable of interest. The error
+term \(e_{IA}\) captures variations in the texts' traits that remain
+unexplained by \(T_{I}\) or \(X_{IA}\). In this model, \(X_{I}\) is an
+observed matrix with \(n_{I}\) rows and \(q_{I}\) columns (variables),
+and both \(e_{I}\) and \(T_{I}\) are latent vectors of length \(n_{I}\).
+Similarly, \(X_{IA}\) is an observed matrix with dimensions
+\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables), while
+\(e_{IA}\) and \(T_{IA}\) are latent vectors of length
+\(n_{I} \cdot n_{A}\).
 
 Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
-second component, where \emph{relevant} \footnote{refer to footnote 1.}
+second component, where \emph{relevant} \footnote{refer to footnote 2.}
 judge-related variables \((Z_{J})\), such as judgment expertise, and
@@
 branch also shows how \(B_{J}\), along with \emph{relevant} \footnote{refer
-  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
+  to footnote 2.} judgment-related variables \((Z_{JK})\) (e.g., the
 number of judgments a judge makes), and judgments' idiosyncratic errors
@@
 \(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).
+
+Overall, the conceptual-population model extends Thurstone's general
+form by introducing key innovations to address the limitations discussed
+in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
+These enhancements include accounting for judges' biases and integrating
+hierarchical structural components. Nevertheless, despite its promise of
+enhancing measurement accuracy and precision, the model still depends on
+the unrealistic assumption that researchers have access to a
+\emph{conceptual population}. Since researchers rarely meet this
+assumption in practice, they must consider a more realistic scenario.
 
@@
 
-Overall, the conceptual-population model extends Thurstone's general
-form by introducing key innovations to address the limitations discussed
-in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
-These enhancements include accounting for judges' biases and integrating
-hierarchical structural components. Nevertheless, despite its promise of
-enhancing measurement accuracy and precision, the model still depends on
-the unrealistic assumption that researchers have access to a
-\emph{conceptual population} of comparative data. Since researchers
-rarely meet this assumption in practice, they must consider a more
-realistic scenario. The extensions to the conceptual-population model,
-derived from this realistic scenario, are described in the
-sample-comparison model.
-
 \subsection{The sample-comparison
@@
 subsequent step, it assumes that judges do not perform \emph{all
-possible stimuli comparisons} within the data sample. Instead, they
-conduct a sufficient number of stimuli comparisons, \(n_{C}\), to ensure
-an accurate estimation of the proportion \(P(B>A)\).
+repeated judgments} within the data sample. Instead, they conduct a
+sufficient number of stimuli comparisons, \(n_{C}\), to ensure an
+accurate estimation of the proportion \(P(B>A)\).
 
@@
 equation (\ref{eq-mat13}). Suppose researchers exclude the second
-student and the third judge. Additionally, they do not include the
-second text from each individual and the third judgment from each judge.
-Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
-these vectors take the following structure:
+student and the third judge. They also omit the second text from each
+individual and the third repeated judgment from each judge. Given
+\(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
+vectors take the following structure:
 \begin{equation}\phantomsection\label{eq-mat24}{
@@
 
-Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
-\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
-\(S_{JK}\). In other words, researchers can only observe texts and
-judgments from students and judges initially included in the sample.
-Additionally, it shows that the sum of observed elements in \(S_{I}\)
-equals the sampled students \((n^{s}_{I})\) and that a similar sum in
-vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
-the sum of observed elements in \(S_{IA}\) represents the total sampled
-texts across all sampled students \((n^{s}_{I} \cdot n^{s}_{A})\), while
-a similar sum in vector \(S_{JK}\) represents the total sampled repeated
-judgments across all sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\).
-
-Finally, we define the \emph{sample mechanism} vector \(S\) in equation
-(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
-of \(S_{JK}\). Thus, the vector contains \(n\) elements, matching the
-number of rows in the \(R\) matrix, and the sum of its elements
-represents the total data sample:
-\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
-Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
-binomial coefficient, which quantifies the total number of unique
-comparisons possible between every pair of sampled stimuli generated by
-each pair of sampled individuals.
+Equation (\ref{eq-mat24}) shows that missing observations in the vectors
+\(S_{I}\) and \(S_{J}\)--which represent unsampled students and
+judges--directly determine which observations are missing in \(S_{IA}\)
+and \(S_{JK}\). In other words, researchers can only observe texts and
+judgments from students and judges initially included in the sample. The
+equation also shows that the sum of observed elements in \(S_{I}\)
+equals the number of sampled students \((n^{s}_{I})\) and that a similar
+sum in vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\).
+Conversely, the sum of observed elements in \(S_{IA}\) represents the
+total sampled texts across all sampled students
+\((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
+\(S_{JK}\) represents the total sampled repeated judgments across all
+sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that because the
+design systematically excludes every third repeated judgment,
+researchers can also express \(S_{JK}\) using \(n_{K} = n^{s}_{K} = 2\).
 \begin{equation}\phantomsection\label{eq-mat25}{
@@
 
+\begin{equation}\phantomsection\label{eq-mat26}{
+s_{(s)} = \begin{cases} 
+1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
+0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
+\end{cases}
+}\end{equation}
+
+Finally, we define the \emph{sample mechanism} \(S\) in equation
+(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
+of \(S_{JK}\). Each element \(s_{(s)}\) is a binary value indicating the
+presence or absence of data rows in the matrix \(R\) resulting from the
+sample mechanism (see equation (\ref{eq-mat26})). Thus, the vector
+contains \(n\) elements, matching the number of rows in \(R\), and the
+sum of its elements represents the total data sample:
+\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
+Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
+binomial coefficient, which quantifies the total number of unique
+comparisons possible between every pair of sampled stimuli generated by
+each pair of sampled individuals.
+
 With the definition of \(S\), we incorporate the sample mechanism into
 the conceptual-population model. Following the convention of
-\citet[pp.~499-516]{McElreath_2020} and
-\citet[pp.~6]{Deffner_et_al_2022}, DAG~\ref{fig-cj12_dag} represents the
-conceptual-population outcome \(O^{cp}_{R}\) as unobserved, emphasizing
-that researchers cannot directly access this outcome due to the sampling
-mechanism. The DAG also depicts the \emph{sample design} vector \(S\) as
-a causal factor influencing the sample-comparison outcome
-\(O^{sc}_{R}\). A square encloses this vector, indicating that it is a
-conditioned variable. In this context, \emph{conditioning} means
-restricting the statistical analysis to a specific subset of the
-population based on the values of a conditioned variable
-\citep[pp.~32]{Neal_2020}, in this case \(S\). Essentially, \(S\) serves
-as the sample equivalent for the \(R\) matrix (see equation
-(\ref{eq-mat12})), that is, this vector selects \emph{all repeated
-judgments a subset of judges makes for the subset of stimuli produced by
-the sampled individuals}. This process ultimately generates the
-sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
-outcome \(O^{cp}_{R}\).
+\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
+DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
+\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
+directly access this outcome due to the sampling mechanism. The DAG also
+depicts the \emph{sample design} vector \(S\) as a causal factor
+influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
+encloses this vector, indicating that it is a conditioned variable.
+Here, \emph{conditioning} means restricting the statistical analysis to
+a specific subset of the population based on the values of a conditioned
+variable \citep[pp.~32]{Neal_2020}. Essentially, \(S\) serves as the
+sample equivalent for the \(R\) matrix, that is, this vector selects
+\emph{all repeated judgments a subset of judges makes for the subset of
+stimuli produced by the sampled individuals}. This process ultimately
+generates the sample-comparison outcome \(O^{sc}_{R}\) from the
+conceptual-population outcome \(O^{cp}_{R}\).
 
 Notably, the DAG shows that \(S\) is independent of all other variables
-in the model. This implies that the conceptual model represented in
-DAG~\ref{fig-cj12_dag} applies exclusively to Simple Random Sampling
-(SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
-repeated judgment, judge, stimulus, and individual has the same
-probability of being included in the sample as any other observation
-within their respective groups \citep[pp.~16-112]{Lawson_2015}.
+in the model. This implies that DAG~\ref{fig-cj14_dag} applies
+exclusively to Simple Random Sampling (SRS) designs
+\citep{Kohler_et_al_2019}. In these designs, each repeated judgment,
+judge, stimulus, and individual has the same probability of being
+included in the sample as any other observation within their respective
+groups \citep{Lawson_2015}.
 
 However, because CJ experiments rarely use the exhaustive pairing of
-sampled judges, stimuli, and individuals --due to concerns about the
-practical feasibility of the comparison task, like in
-\citet[p.~5]{Boonen_et_al_2020}-- a realistic scenario must also account
-for judges comparing only specific stimuli from certain individuals.
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} 
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj12_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.84\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}
-
-}
-
-\subcaption{\label{fig-cj12_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj12}Sample-comparison model including sampling
-mechanism}
-
-\end{figure}%
-
-\subsubsection{The comparison
-mechanism}\label{sec-theory-theoretical_SC2}
-
-As in the previous section, to facilitate the interpretation of the
-sample-comparison model, we first define the \emph{comparison mechanism}
-using the binary vector variable \(C\) as follows:
-\begin{equation}\phantomsection\label{eq-mat26}{
-C = \begin{bmatrix}
-c_{(1)} \\
-\vdots \\
-c_{(c)} \\
-\vdots \\
-c_{(n)} \\
-\end{bmatrix}
-}\end{equation}
-
-Where \(C\) contains \(n\) elements corresponding to the number of rows
-in the \(R\) matrix, with each element being a binary value indicating
-the presence or absence of data rows in the \(R\) matrix (see equation
-(\ref{eq-mat12})): \[
-c_{(c)} = \begin{cases} 
-1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
-0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
-\end{cases}
-\]
-
-This definition assumes that the subset of judges \emph{do not} perform
-all repeated judgments for the subset of stimuli produced by the sampled
-individuals. Instead, they conduct a ``sufficient number of
-comparisons'', \(n_{C}\), to ensure the accurate estimation of the
-proportion \(P(B>A)\) for each pair of stimuli
-\citep[pp.~267]{Thurstone_1927b}.
-
-With the definition of \(C\), the DAG~\ref{fig-cj14_dag} incorporates
-the sample and comparison mechanism into the conceptual-population
-model. The DAG now also represents the sample-comparison outcome
-\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
-directly access this outcome due to the comparison mechanism. The DAG
-also represents the \emph{comparison design} vector \(C\) as a causal
-factor influencing the final comparative outcome \(O_{R}\). Equivalent
-to the sampling mechanism, a square encloses this vector, indicating
-that \(C\) determines \emph{which repeated comparisons the subset of
-judges makes for the subset of stimuli produced by the sampled
-individuals}. This process ultimately derives the final comparative
-outcome \(O_{R}\) from the sample-comparison outcome \(O^{sc}_{R}\).
-
-Notably, the DAG shows that \(C\) is also independent of all other
-variables in the model. This independence implies that the conceptual
-model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
-Random Allocation Comparative Designs \citep[pp.~8]{Bramley_2015} or
-Incomplete Block Designs \citep[pp.~259-271]{Lawson_2015}. In these
-designs, every repeated judgment has an equal probability of being
-included in the sample.
+sampled judges, stimuli, and individuals--due to concerns about the
+practical feasibility of the comparison task \citep{Boonen_et_al_2020}--
+a realistic scenario must also account for judges comparing only
+specific stimuli from certain individuals.
 
@@
 
-\caption{\label{fig-cj14}Sample-comparison model including sample and
-comparison mechanisms}
+\caption{\label{fig-cj14}Sample-comparison model, final vectorized form}
 
@@
 
-In summary, extending Thurstone's general form using Causal inference
-allows researchers to address BTL modeling issues, such as judge biases
+\subsubsection{The comparison
+mechanism}\label{sec-theory-theoretical_SC2}
+
+As in the previous section, to facilitate the interpretation of the
+sample-comparison model, we first define the \emph{comparison mechanism}
+using the binary vector variable \(C\) as follows:
+\begin{equation}\phantomsection\label{eq-mat27}{
+C = \begin{bmatrix}
+c_{(1)} \\
+\vdots \\
+c_{(c)} \\
+\vdots \\
+c_{(n)} \\
+\end{bmatrix}
+}\end{equation}
+
+Where \(C\) contains \(n\) elements corresponding to the number of rows
+in the \(R\) matrix, with each element \(c_{(c)}\) being a binary value
+indicating the presence or absence of data rows in the \(R\) similar to
+equation (\ref{eq-mat26}). This definition assumes that the subset of
+judges \emph{does not} perform all repeated judgments for the subset of
+stimuli produced by the sampled individuals. Instead, they conduct a
+``sufficient number of comparisons'' \(n_{C}\) to ensure the accurate
+estimation of the proportion \(P(B>A)\) for each pair of stimuli
+\citep[pp.~267]{Thurstone_1927b}.
+
+DAG~\ref{fig-cj14_dag} also incorporates the comparison mechanism \(C\)
+into the conceptual-population model. The DAG represents the
+sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
+that researchers cannot directly access this outcome due to the
+comparison mechanism. Moreover, it represents the \emph{comparison
+mechanism} \(C\) as a causal factor influencing the final comparative
+outcome \(O_{R}\). Equivalent to the sampling mechanism, a square
+encloses this vector, indicating that \(C\) is a conditioned variable
+that determines \emph{which repeated judgments the subset of judges
+makes for the subset of stimuli produced by the sampled individuals}.
+This process ultimately derives the final comparative outcome \(O_{R}\)
+from the sample-comparison outcome \(O^{sc}_{R}\).
+
+Notably, the DAG shows that \(C\) is also independent of all other
+variables in the model. This independence implies that the conceptual
+model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
+Random Allocation Comparative Designs \citep{Bramley_2015} or Incomplete
+Block Designs \citep{Lawson_2015}. In these designs, every repeated
+judgment has an equal probability of being included in the sample.
+
+Finally, without loss of generality, researchers can reformulate the
+sample-comparison model in Figure~\ref{fig-cj14} into the equivalent
+form shown in Figure~\ref{fig-cj15}. This reformulation rests on two
+principles. First, because the sampling and comparison mechanism vectors
+\(S\) and \(C\) are binary, researchers can integrate them into the
+model using element-wise products. This approach eliminates the need to
+define the structural equations \(f_{S}\) and \(f_{C}\). Second, since
+only the final comparative outcome \(O_{R}\) is observed, it is standard
+to assume that the distribution of the conceptual-population outcome
+\(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
+applies to this outcome \(O_{R}\). This assumption removes the necessity
+of describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{R} & := f_{O}(D_{R}, S, C) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj15_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}
+
+}
+
+\subcaption{\label{fig-cj15_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj15}Comparative judgment model}
+
+\end{figure}%
+
+In summary, extending Thurstone's general form allows researchers to
+address issues with the BTL model, such as accounting for judge biases
 (see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
-stimuli (see Section~\ref{sec-theory-issue1b}), and measurement errors
-in trait estimation and hypothesis testing (see
-Section~\ref{sec-theory-issue2}). However, the model does not address
-concerns about the assumption of equal dispersion discussed in
+stimuli (see Section~\ref{sec-theory-issue1b}), and measurement error in
+trait estimation and hypothesis testing (see
+Section~\ref{sec-theory-issue2}). However, it does not address concerns
+about stimuli' equal dispersions discussed in
 Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
 assumptions regarding the discriminal process distribution, it requires
-a statistical model, which we will develop in the next section.
+a statistical model, which we develop in the next section.
 
 \section{Abandoning the BTL model}\label{sec-statistical}
+
+Using SCM~\ref{fig-cj15_scm} we finally define the
+
+\(\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1\)
+\(\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1\)
 
@@
 \begin{aligned}
-  O_{R} & = O^{sc}_{R} \odot C \\
-  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
-  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
-  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
-  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
+  D_{R} & = \left( T_{IA}[R_{1},R_{2}] - T_{IA}[R_{3},R_{4}] \right) + B_{JK}[R_{5},R_{6}] \\
+  T_{IA} & = T_{I}[IA_{1}] + \beta_{XA} X_{IA} + e_{IA} \\
   T_{I} & = \beta_{XI} X_{I} + e_{I} \\
-  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{JK} & = B_{J}[JK_{1}] + \beta_{ZK} Z_{JK} + e_{JK} \\
   B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
@@
 
-\subcaption{\label{fig-cj15_stat}Statistical model}
+\subcaption{\label{fig-cj16_stat}Statistical model}
 
@@
 \begin{aligned}
-  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
-  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  O_{R} & := f_{O}(D_{R}, S, C) \\
   D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
   T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
   B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & := f_{e}(X_{I}) \\
-  e_{J} & := f_{e}(Z_{J}) \\
-  e_{IA} & \:\bot\:e_{JK} \\ 
-  \\ 
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
 \end{aligned}
@@
 
-\subcaption{\label{fig-cj15_scm}SCM}
-
-\end{minipage}%
-
-\caption{\label{fig-cj15}Sample-comparison model, including sample and
-comparison mechanisms and assuming different discriminal dispersions for
-the student's traits}
+\subcaption{\label{fig-cj16_scm}SCM}
+
+\end{minipage}%
+
+\caption{\label{fig-cj16}Comparative judgment model, SCM and statistical
+model assuming different discriminal dispersions for the student's
+traits}
 

### 2025-04-07 code/CJmod_add.R

@@ -17,13 +17,13 @@ data{
     int sup_ZKd;  // max. number of categories in ZK
     
     // comparison data
-    array[ns] int<lower=1, upper=nsI> Rs1;   // individual (left)
-    array[ns] int<lower=1, upper=nsA> Ps1;   // stimuli (left)
-    array[ns] int<lower=1, upper=nsI> Rs2;   // individual (right)
-    array[ns] int<lower=1, upper=nsA> Ps2;   // stimuli (right)
-    array[ns] int<lower=1, upper=nsJ> Us;    // judges
-    array[ns] int<lower=1, upper=nsK> Vs;    // judgments
-    array[ns] int<lower=0, upper=1> ORPUV;   // outcome (dichotomous)
+    array[ns] int<lower=1, upper=nsI> R1;   // individual (left)
+    array[ns] int<lower=1, upper=nsA> R2;   // stimuli (left)
+    array[ns] int<lower=1, upper=nsI> R3;   // individual (right)
+    array[ns] int<lower=1, upper=nsA> R4;   // stimuli (right)
+    array[ns] int<lower=1, upper=nsJ> R5;    // judges
+    array[ns] int<lower=1, upper=nsK> R6;    // judgments
+    array[ns] int<lower=0, upper=1> OR;   // outcome (dichotomous)
     
     // individual-stimuli data
     array[nsI*nsA] int<lower=1, upper=nsI> IA1s;    // individuals
@@ -98,7 +98,7 @@ transformed parameters{
       
         // error calculation
         eJ[ JK1s[jk] ] = zeJ[ JK1s[jk] ];                             // identification: sJ = 1
-        eJK[ JK1s[jk], JK2s[jk] ] = pJK * zeJK[ JK1s[jk], JK2s[jk] ]; // identification: pJK prop. of previous level
+        eJK[ JK1s[jk], JK2s[jk] ] = pJK * zeJK[ JK1s[jk], JK2s[jk] ]; // identification: pJK prop. of 1
         
         // trait calculation
         BJ[ JK1s[jk] ] = bZJc*ZJc[ jk ] + bZJd[ ZJd[jk] ] + eJ[ JK1s[jk] ];
@@ -110,7 +110,7 @@ transformed parameters{
[…]

### 2025-04-08 paper2.qmd
section:all; process:write

@@
 <!-- rubrics [@Coertjens_et_al_2017; @Goossens_et_al_2018]  -->
+
 <!-- holistic benchmark ratings [@Bouwer_et_al_2023] -->
 
-<!-- Adaptive comparison algorithms dynamically present stimuli to judges, based on the results of previous comparisons.
-They aim to enhance the informativeness of comparisons while optimizing judges' time [@Bramley_2015].
+<!-- The A for ‘adaptivity’ in ACJ means that the choice of which objects are presented to the judges depends on the outcomes of judgments made so far – the idea being to use judge time efficiently by not presenting objects to compare where the result of the comparison is almost certain.
+[@Bramley_2015, pp.
+4].
 -->
@@
 
-<!-- [@Perron_et_al_2015, pp. 2] -->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, pp. 2] -->
 
@@
 
-<!-- [@Thurstone_1927b, pp. 267] -->
+<!-- It applies fundamentally to the judgments of a single observer who compares a series of stimuli by the method of paired comparison when no 'equal' judgments are allowed [@Thurstone_1927b, pp.
+267] -->
 
@@
 Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome.
-Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978, pp.
-452].
+Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
 In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.
-
-<!-- [@Bramley_2008, pp. 251] -->
 
@@
 -->
+
+<!-- The discriminal mode and dispersion corresponding to a single stimulus are inaccessible to observation.
+They can only be estimated when two objects are compared.
+Thurstone assumed that when two objects are compared with respect to a specified attribute, the object evoking the discriminal process further along the psychological continuum would be judged as possessing more of the attribute.
+...
+The outcome of the paired comparison judgement is therefore related to the distribution of the difference between the two distributions of discriminal processes for essay A and essay B.
+[@Bramley_2008, pp.
+251] -->
+
+<!-- When subject v compares two statements, j and i, he/she reports that item j reflects more of the property of D_{vj}, - D_{vi}, > 0.
+Since D_{vj} and D_{vi}, are assumed normally distributed, the difference process ...
+is normally distributed with mean a_{j} - a_{i}, and variance s^{2}_{ji} [@Andrich_1978, pp.
+452] -->
 
@@
 
-<!-- [@Bramley_2008, pp. 253] -->
+<!-- Case V - This is the simplest formulation of the law and it involves, in addition to the previous assumptions, the assumption that all the discriminal dispersions are equal.
+This assumption should not be made without experimental test.
+[@Thurstone_1927b, pp.
+270] -->
+
+<!-- Case 5 makes the greatest simplification by assuming that the discriminal dispersions are all equal [@Bramley_2008, pp. 253] -->
 
@@
 -->
+
+<!-- This is a simple observation equation which may be used for rather coarse scaling .
+....
+This law defines the psychological scale or continuum.
+It allocates the compared stimuli on the continuum [@Thurstone_1927b, pp.
+269] -->
 
@@
 
-<!-- [@Bramley_2008, pp. 254] -->
-
-<!-- [@vanderLinden_et_al_2017_I, pp. 16] -->
-
-<!-- the mathematical function linking dichotomous outcomes to trait scores -->
-
 <!-- Model misspecification happens when the set of probability distributions considered by the statistician does not include the distribution that generated the observed data. -->
+
+<!-- It will perhaps not come as a great surprise to the reader to discover that it is the Case 5 version of Thurstone’s law that has usually been used in cross-moderation exercises.
+The only difference is in the mathematical function linking the observed proportions to the difference in scale values – the more tractable logistic function is used rather than the cumulative Normal.
+[@Bramley_2008, pp.
+254] -->
+
+<!-- Finally, it is still not uncommon to find a version of the model with a scale constant D = 1.7 added to it, giving it 1.7a_{i}(θ_{p} − b_{i}) as its core structure.
+The original purpose of this practice was to bring the shape of the response functions close to that of the normal-ogive model, generally believed to be the “true model” that would give us a scale with “equal units of measurement” in the early days of test theory (Volume One, Chapter 1).
+However, as discussed below, the scale for the θp parameters is arbitrary up to a non monotonic transformation, and the practice of bringing it close to the scale of a normal ogive is a relict from a distant past, bound to create communication problems between users of item response theory (IRT) rather than add to their understanding of it.
+[@vanderLinden_et_al_2017_I, pp.
+16] -->
 
@@
 
-<!-- [@Thurstone_1927a, pp. 376] -->
+<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
+[@Thurstone_1927a, pp.
+376] -->
 
@@
 Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics.
-For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b].
+For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018].
 These misfit texts may exhibit larger discriminal dispersions or represent outliers --texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
@@
 -->
+
+<!-- As well as an estimate of its value parameter and standard error the analysis will report, for each script, a misfit statistic.
+Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
+[@Pollitt_2004, pp.
+11]  -->
+
+<!-- A similar statistic is calculated for every script.
+A misfitting script is one that judges find difficult for some reason, one that some judges value more than others do.
+[@Pollitt_2012b, pp.
+290] -->
+
+<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
+[@Pollitt_2012a, pp.
+165] -->
+
+<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
+
+<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, pp. 1] -->
 
@@
 <!-- So it seem that the extreme of the discriminal process distribution of the stimuli have higher reliability than the center? -->
-
-<!-- [@Pollitt_2004, pp. 11]  -->
-<!-- [@Pollitt_2012b, pp. 290] -->
 
@@
 
-<!-- Moreover, excluding data using ad hoc univariate procedures can compound these issues by discarding potentially valuable information, further exacerbating the bias [@Zimmerman_1994; @McElreath_2020] -->
-
-<!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
-[@Pollitt_2012a, pp.
-165] -->
-
-<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
-
-<!-- a misfit statistic.
-Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
-It can act as a flag identifying scripts that judges find difficult to judge [@Pollitt_2004, pp.
-12] -->
+<!-- The results that I have described in the present article reveal that outliers in mixed-normal distributions influence both parametric and nonparametric tests.
+The influence on the parametric test is greater, but the direction of the changes in probability of Type 1 and Type I1 errors is the same for both.
+The typical power function is depressed.
+irrespective of whether the analysis is based on parametric or nonparametric methods [@Zimmerman_1994, pp.
+397-398] -->
+
+<!-- What can be done about this?
+There is a tradition of dropping outliers.
+People sometimes drop outliers even before a model is fit, based only on standard deviations from the mean outcome value.
+You should never do that—a point can only be unexpected and highly influential in light of a model.
+After you fit a model, the picture changes.
+If there are only a few outliers, and you are sure to report results both with and without them, dropping outliers might be okay.
+But if there are several outliers and we really need to model them, what then?
+...
+One way to both use these extreme observations and reduce their influence is to employ some kind of robust regression.
+[@McElreath_2020, pp.
+232-233] -->
 
@@
 
-<!-- [@Thurstone_1927b, pp. 268] -->
+<!-- If we accept that correlation as zero, we are really assuming that the degree of excellence which an observer perceives in one of the specimens has no influence on the degree of excellence that he perceives in the comparison specimen.
+There are two effects that may be operative here and which are agnostic to each other.
+...
+mood and the opposite effect of simultaneous contrast ....
+[@Thurstone_1927b, pp.
+268] -->
 
@@
 
+<!-- research shows that, in the presence of these traits and stimuli, the accuracy of the judges' trait perception is influenced by the rank-order distance of the stimuli [@vanDaal_et_al_2017; @Gijsen_et_al_2021], a coarse measure of the relative positioning of the stimuli within the trait continuum.
+-->
+
 <!-- If the test taker’s position on one latent trait is fixed, the assumption of local stochastic independence requires the association between the responses to the items to vanish.
@@
 
-<!-- research shows that, in the presence of these traits and stimuli, the accuracy of the judges' trait perception is influenced by the rank-order distance of the stimuli [@vanDaal_et_al_2017; @Gijsen_et_al_2021], a coarse measure of the relative positioning of the stimuli within the trait continuum.
--->
-
 <!-- 3. The second scenario where the assumption does not hold -->
 In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010].
-For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or general quality.
+For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or even their general quality.
 Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient.
@@
 
-<!-- [@Everitt_et_al_2010, pp. 86] -->
-
-<!-- no study explicitly proposes that this assumption could also be violated due to the presence of an additional hierarchical (grouping) structure relevant to the texts.
-One such scenario might arise, for example, when comparing texts produced by university and secondary school students.
+<!-- for example, when comparing texts produced by university and secondary school students.
 In this case, university students may consistently (or more precisely) produce higher-quality texts, while secondary school students, who exhibit a broader range of writing abilities, would show greater variability in the quality of their texts.
@@
 -->
+
+<!-- Clustered data: A term applied to both data in which the sampling units are grouped into clusters sharing some common feature, for example, animal litters, families or geographical regions, and longitudinal data in which a cluster is deﬁned by a set of repeated measures on the same unit.
+[@Everitt_et_al_2010, pp.
+86] -->
 
@@
 
+<!-- The whole paper [@Ackerman_1989] -->
+
 <!-- To show the practical consequences, in Figure 18.1, we present the test information for the two models; clearly, the information is positively biased when the multidimensional data are forced into a unidimensional IRT model.
@@
 
-<!-- Accounting for measurement error in defining latent variables improves the accuracy of the model’s estimated structural parameters between the latent constructs, which in turn can provide a stronger test of a proposed theoretical model.
-[@Hoyle_et_al_2023, pp.
-482] -->
-
-<!-- [@McElreath_2020, pp. 370] -->
+<!-- But ignoring over-dispersion can also lead to all of the same problems as ignoring any predictor variable.
+Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
+[@McElreath_2020, pp.
+370] -->
 
@@
 
-<!-- [@Pollitt_2012a, pp. 164-165]  -->
-<!-- [@Pollitt_2012b, pp. 289-290] -->
-<!-- [@vanDaal_et_al_2016, pp. 4] -->
-<!-- [@Goossens_et_al_2018, pp. 20] -->
+<!-- When a judge compares two portfolios, Eq. 2 can be used to measure the degree of ‘surprise’ in their decision. ... [@Pollitt_2012a, pp. 164-165]  -->
+
+<!-- Conventionally, any ﬁgure greater than the mean plus two standard deviations – 1.14 here – is taken to indicate signiﬁcant misﬁt.
+In this case, only Judge ‘46 N-M 2’ exceeded the criterion.
+To a signiﬁcant degree, this judge interpreted the trait ‘good writing’ differently from the others.
+[@Pollitt_2012b, pp.
+289-290] -->
+
+<!-- Furthermore, in previous studies of CJ, judges are identified as taking decisions that deviate from the shared consensus (Pollitt, 2012b).
+These judges are labelled as ‘misfit’ and statistically identified by examining the extent to which judges make unexpected decisions (Pollitt, 2012b.
+See Pollitt, 2012a, 2012b for an extended explanation of the identification of misfitting judges).
+A misfitting judge can either be judging too inconsistently compared to the average judge by taking too many unexpected decisions or judging too consistently compared to the average judge.
+However, within CJ, misfit statistics have to be interpreted relatively.
+Misfit statistics quantify the extent to which judges are misfit compared to the whole of judges (Pollitt, 2012b).
+It is suggested that judges with the highest misfit statistic may have a diverging conceptualisation of the competence to be assessed (Bramley, 2007; Whitehouse & Pollitt, 2012).
+[@vanDaal_et_al_2016, pp.
+4] -->
+
+<!-- Pollitt [9] states that a large inﬁt for raters suggests that they consistently judge away from the group consensus.
+Those raters are called misfits.
+As can be seen in Fig.
+3 no rater misfits in this assessment.
+Representations with a large infit are representations which lead to more inconsistent judgments [33].
+[@Goossens_et_al_2018, pp.
+20] -->
 
@@
 
-<!-- [@Thurstone_1927b, pp. 269] -->
+<!-- This is a simple observation equation which may be used for rather coarse scaling .
+....
+This law defines the psychological scale or continuum.
+It allocates the compared stimuli on the continuum [@Thurstone_1927b, pp.
+269] -->
 
@@
 The direction and magnitude of such biases are often unpredictable.
-Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023; @Hoyle_et_al_2023].
+Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
@@
 
-<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->
-
-<!-- [@Kline_et_al_2023, pp. 25]  -->
-<!-- [@Hoyle_et_al_2023, pp. 137] -->
+<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted. [@Everitt_et_al_2010] -->
+
+<!-- Measurement error in the criterion only does not bias unstandardized regression coefficients, if that error is unrelated to the predictors.
+But there is downward bias in the standardized coefficients, which also makes the value of R2 decrease as error in the criterion increases under the same condition (Williams et al., 2013).
+Things are more complicated when there is error in the predictors: Measurement error in a single predictor that is not shared with any other variable will attenuate the absolute value of the regression coefficient for that predictor.
+But regression coefficients for the other, error-free predictors can also be affected.
+Whether those other coefficients are attenuated, exaggerated, or unbiased depends on the values and signs of (1) correlations between the underlying latent variable imperfectly measured by a predictor and all other explanatory variables, and (2) the effect of the conceptual variable just mentioned on the criterion (Bollen, 1989; Kenny, 1979).
+Values of these correlations and their potential distorting effects are rarely known in practice, so it difficult to anticipate the magnitudes and directions of this propagation of measurement error in a single predictor.
+[@Kline_et_al_2023, pp.
+25]  -->
+
+<!-- In manifest variable path models, it is assumed that scores on exogenous variables, such as X1 and X2 in Figures 7.1(a) and 7.1(b), are perfectly reliable.
+Unlike endogenous variables, which have error terms that can “absorb” random measurement error—see Y1 and Y2 in the figures—there is no place in the model for measurement error in exogenous variables to “go” except downstream.
+This means that unmodeled measurement error can bias estimates of path coefficients or disturbance variances (Cole & Preacher, 2014).
+Specifically, values of absolute path coefficients for multiple exogenous causes of the same outcome are negatively biased (i.e., the estimates are too small) when measurement error is uncorrelated (independent) over
+those causes.
+But when measurement error is correlated over multiple predictors, that bias can be either negative or positive (i.e., the estimates are too large).
+If scores on exogenous variables are perfectly precise, then measurement error in endogenous variables results in truncated values of R2, or proportions of explained variation, and also truncated absolute values of standardized—but not unstandardized—regression coefficients, assuming this measurement error is unrelated to the exogenous variables (Williams, Grajales, & Kurkiewicz, 2013).
+[@Hoyle_et_al_2023, pp.
+137] -->
+
+<!-- Probably the majority of regressions are really using proxies like Dobs , because most variables are measurements with some error.
+But even though it doesn’t necessarily open a non-causal path, using a proxy can introduce systematic bias, distorting the estimates [@McElreath_2020, pp.
+492].
+-->
+
+<!-- The effect that measurement error has depends upon the context.
+Sometimes it exaggerates effects, as in this example.
+Other times it hides them.
+But you can’t safely assume that measurement error makes estimates conservative.
+[@McElreath_2020, pp.
+494] -->
+
+<!-- The big take home point for this section is that when you have a distribution of values, don't reduce it down to a single value to use in a regression.
+Instead, use the entire distribution.
+Anytime we use an average value, discarding the uncertainty around that average, we risk overconfidence and spurious inference.
+This doesn't only apply to measurement error, but also to cases in which data are averaged before analysis.
+[@McElreath_2020, pp.
+497] -->
 
@@
 
-<!-- "If Thurstone’s Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes." [@Kelly_et_al_2022, pp.
-678] -->
-
 <!-- ## The diverse assessment design features and their role on reliability and validity {#sec-theory-issue3} -->
@@
 Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests.
-*Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
-4].
+*Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023].
 Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010].
@@
 
-<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
+<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest.
+...
+The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
 Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
@@
 
-<!-- In general terms, causal graphs encode a researcher’s assumptions about the causal process that generated the data in a population of interest [@Schuessler_et_al_2023, pp. 4] -->
+<!-- Estimand: The quantity (usually a parameter) to be estimated using an estimator. -->
+<!-- Estimate: See estimation. -->
+<!-- Estimation: The process of providing a numerical value for a population parameter on the basis of information collected from a sample.
+If a single ﬁgure is calculated for the unknown parameter the process is called point estimation.
+If an interval is calculated which is likely to contain the parameter, then the procedure is called interval estimation.
+-->
+<!-- Estimator: A statistic used to provide an estimate for a parameter. The sample mean, for example, is an unbiased estimator of the population mean. -->
+<!-- [@Everitt_et_al_2010, pp. 154] -->
 
@@
 
+<!-- Consistency: An estimator is said to be consistent if it converges to its estimand as the sample size tends to infinity. [@Everitt_et_al_2010, pp. 100] -->
+
 <!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
@@
 Fortunately, SCMs and DAGs support identification analysis through two key advantages[^1].
-First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
+First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2024].
 This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis.
@@
 
-<!-- [@Pearl_et_al_2016, pp. 35] -->
+<!-- We can understand this flow in general DAGs by understanding the flow in the minimal building blocks of graphs.
+These minimal building blocks consist of chains (Figure 3.9a), forks (Figure 3.9b), immoralities (Figure 3.9c), two unconnected nodes (Figure 3.10), and two connected nodes (Figure 3.11).
+[@Neal_2020, pp.
+23]  -->
+
+<!-- Four elemental confounds [@McElreath_2024] -->
+
+<!-- The net result is that a researcher who has scientific knowledge in the form of a structural equation model is able to predict patterns of independencies in the data, based solely on the structure of the model's graph, without relying on any quantitative information carried by the equations or by the distributions of the errors.
+Conversely, it means that observing patterns of independencies in the data enables us to say something about whether a hypothesized model is correct.
+[@Pearl_et_al_2016, pp.
+35] -->
 
@@
 
-<!-- [@Pearl_2009, pp. 27,68] -->
+<!-- Consider the SCM M shown in Figure 13(a).
+The variables V = {Z, X, Y } are called the endogenous variables, and stand for those variables that the investigator chose to model their cause-effect relationships; the variables U = {Uz , Ux , Uy } are called the exogenous variables and represent everything else that the investigator chose not to explicitly model (these are also usually called disturbances).
+The functions F = {fz , fx , fy } are called structural equations, and each function represents a causal process that assigns to its respective endogenous variable a value based on the values of the other variables.
+[@Cinelli_et_al_2020, pp.
+15] -->
+
+<!-- A graph consists of a set V of vertices (or nodes) and a set E of edges (or links) that connect some pairs of vertices.
+The vertices in our graphs will correspond to variables (whence the common symbol V), and the edges will denote a certain relationship that holds in pairs of variables, the interpretation of which will vary with the application.
+[@Pearl_2009, pp.
+42] -->
+
+<!-- where pa_{i} (connoting parents) stands for the set of variables that directly determine the value of X_{i} and where the U_{i} represent errors (or "disturbances") due to omitted factors.
+[@Pearl_2009, pp.
+65] -->
+
+<!-- These disturbances reflect "hidden" or unmeasurable conditions that Nature governs by some undisclosed probability function. [@Pearl_2009, pp. 89] -->
 
@@
 366-376], accounting for experimental conditions where a judge compares the same pair of stimuli multiple times.
-Thus, the indexing system supports comparisons between different texts written by the same student $(h = i;$ $a \neq b)$ and between texts written by distinct students $(h \neq i;$ $a = b \; \text{is permitted})$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K)$.
-However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(h = i;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K})$, as such comparison lacks practical relevance within the CJ framework.
+Thus, the indexing system supports comparisons between different texts written by the same student $(h = i;$ $a \neq b)$ and between texts written by distinct students $(h \neq i;$ where $a = b$ is permitted$)$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$.
+However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(h = i;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K};$ where $n_{J}>1$ and $n_{K}\geq1)$, as such comparison lacks practical relevance within the CJ framework.
 Here, $n_{J}$ indicates the total number of judges, and $n_{K}$ denotes the number of repeated judgments each judge performs.
@@
 DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014].
-Although DAGs often omit exogenous variables for simplicity, the DAGs presented in this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
+Although DAGs conventionally omit exogenous variables for simplicity, the DAGs presented in this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
 
@@
 <!-- the *magnified representation* includes exogenous variables -->
+
+<!-- DAG, short for directed acyclic graph.
+Graph means it is nodes and connections.
+Directed means the connections have arrows that indicate directions of causal influence.
+And acyclic means that causes do not eventually flow back on themselves.
+[@McElreath_2020, pp.
+128]  -->
+
+<!-- Nodes that are represented by a solid circle are observed random variables, whereas nodes that are represented by a hollow circle are unobserved random variables.
+Causal effects are represented by directed edges (i.e., single-headed arrows), [@Morgan_et_al_2014, pp.
+79] -->
+
+<!-- Here we explicitly show the exogenous variables, but, conventionally, these are omitted from the graph for brevity. [@Cinelli_et_al_2020] -->
 
@@
 1 \\
-2 \\
 \vdots \\
@@
 1 & 1 \\
-1 & 2 \\
 \vdots & \vdots \\
@@
 \vdots & \vdots \\
-n_{I} & n_{A} \\
+n_{I} & n_{A}
 \end{bmatrix} ; \;
@@
 n_{J} & 1 \\
-n_{J} & 2 \\
 \vdots & \vdots \\
-n_{J} & n_{K} \\
+n_{J} & n_{K}
 \end{bmatrix}
@@
 1 & 1 & 1 & 2 & 1 & 1 \\
-1 & 1 & 1 & 2 & 1 & 2 \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
@@
 n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
-n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 2 \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K} \\
+n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
 \end{bmatrix}
@@
 4 \\
-5 \\
+5 
 \end{bmatrix} ; \;
@@
 2 \\
-3 \\
+3 
 \end{bmatrix} ; \;
@@
 5 & 1 \\
-5 & 2 \\
+5 & 2 
 \end{bmatrix} ; \;
@@
 3 & 2 \\
-3 & 3 \\
+3 & 3 
 \end{bmatrix} ; \;
@@
 5 & 1 & 5 & 2 & 3 & 2 \\
-5 & 1 & 5 & 2 & 3 & 3 \\
+5 & 1 & 5 & 2 & 3 & 3 
 \end{bmatrix}
@@
 
-<!-- [@Hoyle_et_al_2023, pp. 138] -->
-<!-- [@Fox_2010, chap. 6]  -->
-<!-- [@vanderLinden_et_al_2017_I, chap. 24] -->
+<!-- The assumption of no measurement error in exogenous variables is generally unrealistic, such as for self-report measures of psychological traits such as anxiety, extraversion, or leadership.
+An alternative is to respecify a manifest variable path model as a latent variable model with single indicators such that values of score reliability coefficients for all variables are explicitly represented in the model.
+Doing so relaxes the requirement for perfect measurement of exogenous variables while also controlling for measurement error in endogenous variables.
+[@Hoyle_et_al_2023, pp.
+138] -->
+
+<!-- Accounting for measurement error in defining latent variables improves the accuracy of the model’s estimated structural parameters between the latent constructs, which in turn can provide a stronger test of a proposed theoretical model.
+[@Hoyle_et_al_2023, pp.
+482] -->
+
+<!-- This is a whole section [@Fox_2010, chap. 6] -->
+
+<!-- This is a whole section  [@vanderLinden_et_al_2017_I, chap. 24] -->
 
 <!-- 2 Explaining what we did -->
-The top branch of [DAG - @fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
+The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
 The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$.
-This branch also shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' writing-quality trait $(T_{IA})$, the first primary latent variable of interest.
+Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
+Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' writing-quality trait $(T_{IA})$, the first primary latent variable of interest.
 The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
-In this model, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
-Similarly, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
-
-[^2]: Here, *relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+Here, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
+
+[^2]: *Relevant variables* are those that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
 A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
@@
 The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
-This branch also shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
+Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
-Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
+Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
 
@@
 The sample-comparison model presents a more realistic scenario than the conceptual-population model.
-Expressly, it assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
-In a subsequent step, it assumes that judges do not perform *all repeated judgments* within the data sample.
+In @sec-theory-theoretical_SC1, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
+Moreover, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample.
 Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.
+
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
 
@@
 <!-- recording: Sven and Tine 25.03.28; time: 00:11:30 - 00:12:25 -->
-
-<!-- Develop how you get the vectors S and C and sampling weights when you know the extension of the population. (See https://mc-stan.org/docs/stan-users-guide/poststratification.html ) -->
 
@@
 S_{IA} = \begin{bmatrix}
-ia_{(1)} \\
-\vdots \\
-ia_{(ia)} \\
-\vdots \\
-ia_{(nI \cdot nA)} \\
+ia_{(1,1)} \\
+\vdots \\
+ia_{(1,n_{A})} \\
+\vdots \\
+ia_{(h,a)} \\
+\vdots \\
+ia_{(i,b)} \\
+\vdots \\
+ia_{(nI,1)} \\
+\vdots \\
+ia_{(nI,nA)}
 \end{bmatrix} ; \;
 S_{JK} = \begin{bmatrix}
-jk_{(1)} \\
-\vdots \\
-jk_{(jk)} \\
-\vdots \\
-jk_{(nJ \cdot nK)} \\
+jk_{(1,1)} \\
+\vdots \\
+jk_{(1,n_{K})} \\
+\vdots \\
+jk_{(j,k)} \\
+\vdots \\
+jk_{(nJ,1)} \\
+\vdots \\
+jk_{(nJ,nK)}
 \end{bmatrix}
@@
 
-Where each element of $S_{I}$ and $S_{J}$ is a binary value indicating the presence or absence of data elements in the $I$ and $J$ vectors (see equation ([-@eq-mat11])).
+Where each element of $S_{I}$ is a binary value indicating the presence or absence of corresponding elements in the vector $I$, as in equation ([-@eq-mat22]).
+We apply the same logic to $S_{J}$, based on the vector $J$ (not shown).
 Thus, the vectors $S_{I}$ and $S_{J}$ contains $n_{I}$ and $n_{J}$ elements, respectively.
-Specifically, we define the vector $S_{I}$ as in equation ([-@eq-mat22]) and apply a similar definition to $S_{J}$.
 $$
@@
 
-Similarly, each element of $S_{IA}$ and $S_{JK}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$ and $JK$ (see equation ([-@eq-mat11])).
+Similarly, each element of $S_{IA}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$, as defined in equation ([-@eq-mat23]).
+We apply the same logic to $S_{JK}$, based on the matrix $JK$ (not shown).
 Thus, the vectors $S_{IA}$ and $S_{JK}$ contains $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$ elements, respectively.
-Specifically, we define the vector $S_{IA}$ as in equation ([-@eq-mat23]) and apply a similar definition to $S_{JK}$.
-$$
-ia_{(ia)} = \begin{cases} 
+$$
+ia_{(i,a)} = \begin{cases} 
 1 & \text{if data elements } i,a \text{ from } IA \text{ are sampled} \\
@@
 Notice that because the design systematically excludes every third repeated judgment, researchers can also express $S_{JK}$ using $n_{K} = n^{s}_{K} = 2$.
+
+Finally, we define the *sample mechanism* $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
+Each element $s_{(h,a,i,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in equation ([-@eq-mat26]).
+Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
+Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
+
 $$
 S = \begin{bmatrix}
-s_{(1)} \\
-\vdots \\
-s_{(s)} \\
-\vdots \\
-s_{(n)} \\
+s_{(1,1,1,2,1,1)} \\
+\vdots \\
+s_{(1,1,1,2,1,n_{K})} \\
+\vdots \\
+s_{(h,a,i,b,j,k)} \\
+\vdots \\
+s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
+\vdots \\
+s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
 \end{bmatrix}
@@
 $$
-s_{(s)} = \begin{cases} 
+s_{(h,i,a,b,j,k)} = \begin{cases} 
 1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
@@
 $$ {#eq-mat26}
-
-Finally, we define the *sample mechanism* $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
-Each element $s_{(s)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism (see equation ([-@eq-mat26])).
-Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
-Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
 
@@
 A square encloses this vector, indicating that it is a conditioned variable.
-Here, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp.
-32].
+Here, *conditioning* means that researchers restrict their focus to the elements of $O^{cp}_{R}$ that satisfy $s_{(h,i,a,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
 Essentially, $S$ serves as the sample equivalent for the $R$ matrix, that is, this vector selects *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
-
-<!-- [@McElreath_2020, pp. 499-516] -->
-<!-- [@Deffner_et_al_2022, pp. 6] -->
+This process generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
+
+<!-- Follow conventions of @McElreath_2020 NOT @Schuessler_et_al_2023 regarding selection variables and missing data. -->
+
+<!-- This is a whole section [@McElreath_2020, pp. 499-516] -->
+
+<!-- This is a whole section [@Deffner_et_al_2022, pp. 6-7] -->
+
+<!-- What does “conditioning” mean? Informally, conditioning on a variable Z means learning its value and then asking if X adds any additional information about Y. [@McElreath_2020, pp. 130] -->
+
+<!-- In the regular notation for probability, we have conditioning, but that isn’t the same as intervening.
+Conditioning on T=t just means that we are restricting our focus to the subset of the population to those who received treatment t.
+In contrast, an intervention would be to take the whole population and give everyone treatment t.
+[@Neal_2020, pp.
+32] -->
 
 Notably, the DAG shows that $S$ is independent of all other variables in the model.
-This implies that [DAG -@fig-cj14_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019].
+This implies that [DAG -@fig-cj14_dag] applies exclusively to Simple Random Sampling (SRSg) designs.
 In these designs, each repeated judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015].
 
-<!-- [@Kohler_et_al_2019, pp. 150] -->
-<!-- [@Lawson_2015, pp. 16-112] -->
-
-<!-- Follow conventions of McElreath (2020) NOT Schuessler et al (2023) regarding selection variables and missing data. -->
-
-<!-- 3. segway for next section -->
-However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
-
-<!-- [@Boonen_et_al_2020, p. 5] -->
+<!-- This is a whole section [@Lawson_2015, pp. 16-112] -->
 
@@
 
+<!-- 3. segway for next section -->
+However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
+
+<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
+Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
+However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
+More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
+[@Boonen_et_al_2020, p.
+5] -->
+
 ### The comparison mechanism {#sec-theory-theoretical_SC2}
@@
 C = \begin{bmatrix}
-c_{(1)} \\
-\vdots \\
-c_{(c)} \\
-\vdots \\
-c_{(n)} \\
+c_{(1,1,1,2,1,1)} \\
+\vdots \\
+c_{(1,1,1,2,1,n_{K})} \\
+\vdots \\
+c_{(h,a,i,b,j,k)} \\
+\vdots \\
+c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
+\vdots \\
+c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
 \end{bmatrix}
@@
 
-Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(c)}$ being a binary value indicating the presence or absence of data rows in the $R$ similar to equation ([-@eq-mat26]).
+Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(h,a,i,b,j,k)}$ being a binary value indicating the presence or absence of data rows in the $R$ similar to equation ([-@eq-mat26]).
 This definition assumes that the subset of judges *does not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals.
@@
 
-<!-- [@Bramley_2015, pp. 8] -->
-<!-- [@Lawson_2015, pp. 259-271] -->
-
 <!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
@@
 
-Finally, without loss of generality, researchers can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
-This reformulation rests on two principles.
-First, because the sampling and comparison mechanism vectors $S$ and $C$ are binary, researchers can integrate them into the model using element-wise products.
-This approach eliminates the need to define the structural equations $f_{S}$ and $f_{C}$.
-Second, since only the final comparative outcome $O_{R}$ is observed, it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$, defined by the structural equation $f_{O}$, also applies to this outcome $O_{R}$.
-This assumption removes the necessity of describing the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$.
-
-<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
-For example, if I have the distribution of the population outcome, and it is a bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
+<!-- The different analyses were based on the same two underlying datasets.
+First of all, observations from these datasets were sampled in order to mimic either random allocation of pairs of scripts to judges, or adaptive allocation (where the probability of being sampled depends on the outcome of earlier comparisons).
+In all cases the selection of observations was carefully planned to ensure that i) each script was compared the same number of times; and ii) the complete set of judgments formed a fully linked design whereby each script was compared directly or indirectly with every other script5.
+For the ‘random only’ analyses, this could be achieved by systematically sampling the required observations.
+[@Bramley_2015, pp.
+8] -->
+
+<!-- This is a whole section [@Lawson_2015, pp. 259-271] -->
 
@@
 
+<!-- 3. preparing to transform the conceptual model into an statistical one -->
+Finally, without loss of generality, researchers can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
+This reformulation rests on two principles.
+First, because the sampling and comparison mechanism vectors $S$ and $C$ are binary, researchers can integrate them into the model using element-wise products.
+This approach eliminates the need to define the structural equations $f_{S}$ and $f_{C}$.
+Second, since only the final comparative outcome $O_{R}$ is observed, it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$, defined by the structural equation $f_{O}$, also applies to this outcome $O_{R}$.
+This assumption removes the necessity of describing the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$.
+
+<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
+For example, if I have the distribution of the population outcome, and it is a bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
+
+<!-- 4. segway for next section -->
 In summary, extending Thurstone's general form allows researchers to address issues with the BTL model, such as accounting for judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2).
@@
 
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
+
 <!-- design matters -->
@@
 
+<!-- DAGs are heuristic—they are not detailed statistical models.
+But they allow us to deduce which statistical models can provide valid causal inferences, assuming the DAG is true.
+[@McElreath_2020, pp.
+17] -->
+
 <!-- DAGs are not about adding assumptions—they are about revealing the -->
@@
 -->
+
+<!-- Develop how you get the vectors S and C and sampling weights when you know the extension of the population. (See https://mc-stan.org/docs/stan-users-guide/poststratification.html ) -->
 

### 2025-04-08 _tex/paper2.tex
Built site for gh-pages

@@
 on the continuum, the discriminal difference will also conform to a
-Normal distribution \citep[pp.~452]{Andrich_1978}. In this distribution,
-the mode (mean) represents the relative separation between the stimuli,
-and its dispersion indicates the variability of that separation.
+Normal distribution \citep{Andrich_1978}. In this distribution, the mode
+(mean) represents the relative separation between the stimuli, and its
+dispersion indicates the variability of that separation.
 
@@
 judgment discrepancies than those involving other texts
-\citep{Pollitt_2004, Pollitt_2012b}. These misfit texts may exhibit
-larger discriminal dispersions or represent outliers --texts with
-distinctive characteristics that deviate markedly from the rest of the
-sample \citep{Grubbs_1969}. In either case, comparing misfit texts
-results in more judgment discrepancies, yet the BTL model neither
-accounts for these cases nor offers any means to address them.
+\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}.
+These misfit texts may exhibit larger discriminal dispersions or
+represent outliers --texts with distinctive characteristics that deviate
+markedly from the rest of the sample \citep{Grubbs_1969}. In either
+case, comparing misfit texts results in more judgment discrepancies, yet
+the BTL model neither accounts for these cases nor offers any means to
+address them.
 
@@
 produced by the same individual, these texts are likely to share common
-features, such as writing style or general quality. Although the CJ
-literature acknowledges the existence of such hierarchical structures,
-the statistical approaches to account for this additional source of
-dependence have been insufficient. For instance, when CJ data
+features, such as writing style or even their general quality. Although
+the CJ literature acknowledges the existence of such hierarchical
+structures, the statistical approaches to account for this additional
+source of dependence have been insufficient. For instance, when CJ data
 incorporates multiple samples of stimuli from the same individuals,
@@
 uncertainty in the scores and the actual effects being tested
-\citep{Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore, the reduced
-precision in hypothesis tests diminishes their statistical power,
-increasing the likelihood of committing type-I or type-II errors
+\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore,
+the reduced precision in hypothesis tests diminishes their statistical
+power, increasing the likelihood of committing type-I or type-II errors
 \citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
@@
 solely on its (causal) assumptions, regardless of random variability
-\citep[pp.~4]{Schuessler_et_al_2023}. Here, \emph{estimands} represent
-the specific quantities researchers aim to determine
+\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
+specific quantities researchers aim to determine
 \citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
@@
 complexity, they can represent various causal structures using only five
-fundamental building blocks \citep{Neal_2020, McElreath_2020}. This
+fundamental building blocks \citep{Neal_2020, McElreath_2024}. This
 feature allows researchers to decompose complex structures into
@@
 different texts written by the same student \((h = i;\) \(a \neq b)\)
-and between texts written by distinct students \((h \neq i;\)
-\(a = b \; \text{is permitted})\), each judged once or repeatedly by all
-judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K)\). However, it
-excludes cases where a judge compares a student's text to itself,
-whether once or multiple times \((h = i;\) \(a = b;\)
-\(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K})\), as such comparison lacks
-practical relevance within the CJ framework. Here, \(n_{J}\) indicates
-the total number of judges, and \(n_{K}\) denotes the number of repeated
-judgments each judge performs.
+and between texts written by distinct students \((h \neq i;\) where
+\(a = b\) is permitted\()\), each judged once or repeatedly by all
+judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\) where \(n_{J}>1\)
+and \(n_{K}\geq1)\). However, it excludes cases where a judge compares a
+student's text to itself, whether once or multiple times \((h = i;\)
+\(a = b;\) \(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K};\) where
+\(n_{J}>1\) and \(n_{K}\geq1)\), as such comparison lacks practical
+relevance within the CJ framework. Here, \(n_{J}\) indicates the total
+number of judges, and \(n_{K}\) denotes the number of repeated judgments
+each judge performs.
 
@@
 solid black circles and unobserved (latent) variables as open circles
-\citep{Morgan_et_al_2014}. Although DAGs often omit exogenous variables
-for simplicity, the DAGs presented in this section includes exogenous
-variables to improve clarity and reveal potential issues related to
-conditioning and confounding \citep{Cinelli_et_al_2020}.
+\citep{Morgan_et_al_2014}. Although DAGs conventionally omit exogenous
+variables for simplicity, the DAGs presented in this section includes
+exogenous variables to improve clarity and reveal potential issues
+related to conditioning and confounding \citep{Cinelli_et_al_2020}.
 
@@
 
-The top branch of DAG -~\ref{fig-cj09_dag} illustrates the first
-component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
-  to variables that satisfy the \emph{backdoor criterion} \citep[pp
+The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
+component, where \emph{relevant} \footnote{\emph{Relevant variables} are
+  those that satisfy the \emph{backdoor criterion} \citep[pp
   37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
@@
 \(e_{I}\) captures variations in students' traits unexplained by
-\(X_{I}\). This branch also shows how \(T_{I}\), along with
-\emph{relevant} \footnote{refer to footnote 2.} text-related variables
-\((X_{IA})\) (e.g., text length), and texts' idiosyncratic errors
-\(e_{IA}\) causally influence the texts' writing-quality trait
-\((T_{IA})\), the first primary latent variable of interest. The error
-term \(e_{IA}\) captures variations in the texts' traits that remain
-unexplained by \(T_{I}\) or \(X_{IA}\). In this model, \(X_{I}\) is an
-observed matrix with \(n_{I}\) rows and \(q_{I}\) columns (variables),
-and both \(e_{I}\) and \(T_{I}\) are latent vectors of length \(n_{I}\).
-Similarly, \(X_{IA}\) is an observed matrix with dimensions
-\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables), while
-\(e_{IA}\) and \(T_{IA}\) are latent vectors of length
-\(n_{I} \cdot n_{A}\).
+\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
+\(q_{I}\) columns (variables), and both \(e_{I}\) and \(T_{I}\) are
+latent vectors of length \(n_{I}\). Additionally, this branch shows how
+\(T_{I}\), along with \emph{relevant} \footnote{refer to footnote 2.}
+text-related variables \((X_{IA})\) (e.g., text length), and texts'
+idiosyncratic errors \(e_{IA}\) causally influence the texts'
+writing-quality trait \((T_{IA})\), the first primary latent variable of
+interest. The error term \(e_{IA}\) captures variations in the texts'
+traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
+\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
+rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
+are latent vectors of length \(n_{I} \cdot n_{A}\).
 
@@
 variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
-captures variations in judges' bias unexplained by \(Z_{J}\). This
-branch also shows how \(B_{J}\), along with \emph{relevant} \footnote{refer
-  to footnote 2.} judgment-related variables \((Z_{JK})\) (e.g., the
-number of judgments a judge makes), and judgments' idiosyncratic errors
-\(e_{JK}\) causally influence the biases associated with each text
-\(B_{JK}\), the second primary latent variable of interest. The error
-\(e_{JK}\) captures variations in judgments unexplained by \(B_{J}\) or
-\(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with \(n_{J}\) rows
-and \(q_{J}\) columns (variables), and both \(e_{J}\) and \(B_{J}\) are
-latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\) is an observed
-matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\) columns
-(variables), while \(e_{JK}\) and \(B_{JK}\) are latent vectors of
-length \(n_{J} \cdot n_{K}\).
+captures variations in judges' bias unexplained by \(Z_{J}\). Here,
+\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
+columns (variables), and both \(e_{J}\) and \(B_{J}\) are latent vectors
+of length \(n_{J}\). Furthermore, the branch shows how \(B_{J}\), along
+with \emph{relevant} \footnote{refer to footnote 2.} judgment-related
+variables \((Z_{JK})\) (e.g., the number of judgments a judge makes),
+and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
+biases associated with each text \(B_{JK}\), the second primary latent
+variable of interest. The error \(e_{JK}\) captures variations in
+judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
+observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
+columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent vectors
+of length \(n_{J} \cdot n_{K}\).
 
@@
 The sample-comparison model presents a more realistic scenario than the
-conceptual-population model. Expressly, it assumes researchers work with
-a data sample consisting of a limited number of repeated judgments
-\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
-number of texts \((n^{s}_{A})\) from a sample of students
-\((n^{s}_{I})\), all drawn from the conceptual population. In a
-subsequent step, it assumes that judges do not perform \emph{all
-repeated judgments} within the data sample. Instead, they conduct a
-sufficient number of stimuli comparisons, \(n_{C}\), to ensure an
-accurate estimation of the proportion \(P(B>A)\).
+conceptual-population model. In
+Section~\ref{sec-theory-theoretical_SC1}, it explicitly assumes
+researchers work with a data sample consisting of a limited number of
+repeated judgments \((n^{s}_{K})\) from a sample of judges
+\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
+sample of students \((n^{s}_{I})\), all drawn from the conceptual
+population. Moreover, in Section~\ref{sec-theory-theoretical_SC2}, the
+model assumes that judges do not perform \emph{all repeated judgments}
+within the data sample. Instead, they conduct a sufficient number of
+stimuli comparisons, \(n_{C}\), to ensure an accurate estimation of the
+proportion \(P(B>A)\).
 
@@
 S_{IA} = \begin{bmatrix}
-ia_{(1)} \\
+ia_{(1,1)} \\
 \vdots \\
-ia_{(ia)} \\
+ia_{(1,n_{A})} \\
 \vdots \\
-ia_{(nI \cdot nA)} \\
+ia_{(h,a)} \\
+\vdots \\
+ia_{(i,b)} \\
+\vdots \\
+ia_{(nI,1)} \\
+\vdots \\
+ia_{(nI,nA)}
 \end{bmatrix} ; \;
 S_{JK} = \begin{bmatrix}
-jk_{(1)} \\
+jk_{(1,1)} \\
 \vdots \\
-jk_{(jk)} \\
+jk_{(1,n_{K})} \\
 \vdots \\
-jk_{(nJ \cdot nK)} \\
+jk_{(j,k)} \\
+\vdots \\
+jk_{(nJ,1)} \\
+\vdots \\
+jk_{(nJ,nK)}
 \end{bmatrix}
@@
 
-Where each element of \(S_{I}\) and \(S_{J}\) is a binary value
-indicating the presence or absence of data elements in the \(I\) and
-\(J\) vectors (see equation (\ref{eq-mat11})). Thus, the vectors
-\(S_{I}\) and \(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements,
-respectively. Specifically, we define the vector \(S_{I}\) as in
-equation (\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).
+Where each element of \(S_{I}\) is a binary value indicating the
+presence or absence of corresponding elements in the vector \(I\), as in
+equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\), based
+on the vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and
+\(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements, respectively.
 \begin{equation}\phantomsection\label{eq-mat22}{
@@
 
-Similarly, each element of \(S_{IA}\) and \(S_{JK}\) is a binary value
-indicating the presence or absence of data rows in the matrices \(IA\)
-and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{IA}\)
-and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
-elements, respectively. Specifically, we define the vector \(S_{IA}\) as
-in equation (\ref{eq-mat23}) and apply a similar definition to
-\(S_{JK}\). \begin{equation}\phantomsection\label{eq-mat23}{
-ia_{(ia)} = \begin{cases} 
+Similarly, each element of \(S_{IA}\) is a binary value indicating the
+presence or absence of data rows in the matrices \(IA\), as defined in
+equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\), based
+on the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
+\(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
+elements, respectively. \begin{equation}\phantomsection\label{eq-mat23}{
+ia_{(i,a)} = \begin{cases} 
 1 & \text{if data elements } i,a \text{ from } IA \text{ are sampled} \\
@@
 researchers can also express \(S_{JK}\) using \(n_{K} = n^{s}_{K} = 2\).
-\begin{equation}\phantomsection\label{eq-mat25}{
-S = \begin{bmatrix}
-s_{(1)} \\
-\vdots \\
-s_{(s)} \\
-\vdots \\
-s_{(n)} \\
-\end{bmatrix}
-}\end{equation}
-
-\begin{equation}\phantomsection\label{eq-mat26}{
-s_{(s)} = \begin{cases} 
-1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
-0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
-\end{cases}
-}\end{equation}
 
@@
 (\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
-of \(S_{JK}\). Each element \(s_{(s)}\) is a binary value indicating the
-presence or absence of data rows in the matrix \(R\) resulting from the
-sample mechanism (see equation (\ref{eq-mat26})). Thus, the vector
-contains \(n\) elements, matching the number of rows in \(R\), and the
-sum of its elements represents the total data sample:
+of \(S_{JK}\). Each element \(s_{(h,a,i,b,j,k)}\) is a binary value
+indicating the presence or absence of data rows in the matrix \(R\)
+resulting from the sample mechanism, as in equation (\ref{eq-mat26}).
+Thus, the vector contains \(n\) elements, matching the number of rows in
+\(R\), and the sum of its elements represents the total data sample:
 \(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
@@
 each pair of sampled individuals.
+
+\begin{equation}\phantomsection\label{eq-mat25}{
+S = \begin{bmatrix}
+s_{(1,1,1,2,1,1)} \\
+\vdots \\
+s_{(1,1,1,2,1,n_{K})} \\
+\vdots \\
+s_{(h,a,i,b,j,k)} \\
+\vdots \\
+s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
+\vdots \\
+s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
+\end{bmatrix}
+}\end{equation}
+
+\begin{equation}\phantomsection\label{eq-mat26}{
+s_{(h,i,a,b,j,k)} = \begin{cases} 
+1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
+0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
+\end{cases}
+}\end{equation}
 
@@
 encloses this vector, indicating that it is a conditioned variable.
-Here, \emph{conditioning} means restricting the statistical analysis to
-a specific subset of the population based on the values of a conditioned
-variable \citep[pp.~32]{Neal_2020}. Essentially, \(S\) serves as the
+Here, \emph{conditioning} means that researchers restrict their focus to
+the elements of \(O^{cp}_{R}\) that satisfy \(s_{(h,i,a,b,j,k)}=1\)
+\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
 sample equivalent for the \(R\) matrix, that is, this vector selects
 \emph{all repeated judgments a subset of judges makes for the subset of
-stimuli produced by the sampled individuals}. This process ultimately
-generates the sample-comparison outcome \(O^{sc}_{R}\) from the
-conceptual-population outcome \(O^{cp}_{R}\).
+stimuli produced by the sampled individuals}. This process generates the
+sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
+outcome \(O^{cp}_{R}\).
 
@@
 in the model. This implies that DAG~\ref{fig-cj14_dag} applies
-exclusively to Simple Random Sampling (SRS) designs
-\citep{Kohler_et_al_2019}. In these designs, each repeated judgment,
-judge, stimulus, and individual has the same probability of being
-included in the sample as any other observation within their respective
-groups \citep{Lawson_2015}.
-
-However, because CJ experiments rarely use the exhaustive pairing of
-sampled judges, stimuli, and individuals--due to concerns about the
-practical feasibility of the comparison task \citep{Boonen_et_al_2020}--
-a realistic scenario must also account for judges comparing only
-specific stimuli from certain individuals.
+exclusively to Simple Random Sampling (SRSg) designs. In these designs,
+each repeated judgment, judge, stimulus, and individual has the same
+probability of being included in the sample as any other observation
+within their respective groups \citep{Lawson_2015}.
 
@@
 
+However, because CJ experiments rarely use the exhaustive pairing of
+sampled judges, stimuli, and individuals--due to concerns about the
+practical feasibility of the comparison task \citep{Boonen_et_al_2020}--
+a realistic scenario must also account for judges comparing only
+specific stimuli from certain individuals.
+
 \subsubsection{The comparison
@@
 C = \begin{bmatrix}
-c_{(1)} \\
+c_{(1,1,1,2,1,1)} \\
 \vdots \\
-c_{(c)} \\
+c_{(1,1,1,2,1,n_{K})} \\
 \vdots \\
-c_{(n)} \\
+c_{(h,a,i,b,j,k)} \\
+\vdots \\
+c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
+\vdots \\
+c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
 \end{bmatrix}
@@
 Where \(C\) contains \(n\) elements corresponding to the number of rows
-in the \(R\) matrix, with each element \(c_{(c)}\) being a binary value
-indicating the presence or absence of data rows in the \(R\) similar to
-equation (\ref{eq-mat26}). This definition assumes that the subset of
-judges \emph{does not} perform all repeated judgments for the subset of
-stimuli produced by the sampled individuals. Instead, they conduct a
-``sufficient number of comparisons'' \(n_{C}\) to ensure the accurate
-estimation of the proportion \(P(B>A)\) for each pair of stimuli
-\citep[pp.~267]{Thurstone_1927b}.
+in the \(R\) matrix, with each element \(c_{(h,a,i,b,j,k)}\) being a
+binary value indicating the presence or absence of data rows in the
+\(R\) similar to equation (\ref{eq-mat26}). This definition assumes that
+the subset of judges \emph{does not} perform all repeated judgments for
+the subset of stimuli produced by the sampled individuals. Instead, they
+conduct a ``sufficient number of comparisons'' \(n_{C}\) to ensure the
+accurate estimation of the proportion \(P(B>A)\) for each pair of
+stimuli \citep[pp.~267]{Thurstone_1927b}.
 
@@
 judgment has an equal probability of being included in the sample.
-
-Finally, without loss of generality, researchers can reformulate the
-sample-comparison model in Figure~\ref{fig-cj14} into the equivalent
-form shown in Figure~\ref{fig-cj15}. This reformulation rests on two
-principles. First, because the sampling and comparison mechanism vectors
-\(S\) and \(C\) are binary, researchers can integrate them into the
-model using element-wise products. This approach eliminates the need to
-define the structural equations \(f_{S}\) and \(f_{C}\). Second, since
-only the final comparative outcome \(O_{R}\) is observed, it is standard
-to assume that the distribution of the conceptual-population outcome
-\(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
-applies to this outcome \(O_{R}\). This assumption removes the necessity
-of describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).
 
@@
 \end{figure}%
+
+Finally, without loss of generality, researchers can reformulate the
+sample-comparison model in Figure~\ref{fig-cj14} into the equivalent
+form shown in Figure~\ref{fig-cj15}. This reformulation rests on two
+principles. First, because the sampling and comparison mechanism vectors
+\(S\) and \(C\) are binary, researchers can integrate them into the
+model using element-wise products. This approach eliminates the need to
+define the structural equations \(f_{S}\) and \(f_{C}\). Second, since
+only the final comparative outcome \(O_{R}\) is observed, it is standard
+to assume that the distribution of the conceptual-population outcome
+\(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
+applies to this outcome \(O_{R}\). This assumption removes the necessity
+of describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).
 

### 2025-04-08 paper2.qmd
section:5; process:write,formulas

@@
 The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
-Here, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent vectors of length $n_{I} \cdot n_{A}$.
+Here, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent matrices with $n_{I}$ rows and $n_{A}$ columns.
 
@@
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
-Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.
+Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
 
@@
 In @sec-theory-theoretical_SC1, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
-Moreover, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample.
-Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.
+Furthermore, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample.
+Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$, as proposed by @Thurstone_1927b.
 
@@
 Where each element of $S_{I}$ is a binary value indicating the presence or absence of corresponding elements in the vector $I$, as in equation ([-@eq-mat22]).
-We apply the same logic to $S_{J}$, based on the vector $J$ (not shown).
+We apply the same logic to $S_{J}$ using vector $J$ (not shown).
 Thus, the vectors $S_{I}$ and $S_{J}$ contains $n_{I}$ and $n_{J}$ elements, respectively.
@@
 Similarly, each element of $S_{IA}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$, as defined in equation ([-@eq-mat23]).
-We apply the same logic to $S_{JK}$, based on the matrix $JK$ (not shown).
+We apply the same logic to $S_{JK}$ using the matrix $JK$ (not shown).
 Thus, the vectors $S_{IA}$ and $S_{JK}$ contains $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$ elements, respectively.
@@
 We can visualize these vectors more clearly using the example in equation ([-@eq-mat13]).
-Suppose researchers exclude the second student and the third judge.
-They also omit the second text from each individual and the third repeated judgment from each judge.
+Suppose researchers exclude the second student and the third judge, and they also omit the second text from each individual and the third repeated judgment from each judge.
 Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, these vectors take the following structure:
@@
 Here, *conditioning* means that researchers restrict their focus to the elements of $O^{cp}_{R}$ that satisfy $s_{(h,i,a,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
-Essentially, $S$ serves as the sample equivalent for the $R$ matrix, that is, this vector selects *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-This process generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.
+Essentially, $S$ serves as the vector selecting *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+This process generates the sample-comparison outcome $O^{sc}_{R}$.
 
@@
 <!-- 3. segway for next section -->
-However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
+However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020]--a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
 
@@
 Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(h,a,i,b,j,k)}$ being a binary value indicating the presence or absence of data rows in the $R$ similar to equation ([-@eq-mat26]).
-This definition assumes that the subset of judges *does not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals.
+This definition assumes that the sampled judges *do not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals.
 Instead, they conduct a "sufficient number of comparisons" $n_{C}$ to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp.
@@
 Equivalent to the sampling mechanism, a square encloses this vector, indicating that $C$ is a conditioned variable that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-This process ultimately derives the final comparative outcome $O_{R}$ from the sample-comparison outcome $O^{sc}_{R}$.
+This process ultimately derives the final comparative outcome $O_{R}$.
 
@@
 This approach eliminates the need to define the structural equations $f_{S}$ and $f_{C}$.
-Second, since only the final comparative outcome $O_{R}$ is observed, it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$, defined by the structural equation $f_{O}$, also applies to this outcome $O_{R}$.
+Second, since only the final comparative outcome $O_{R}$ is observed, it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$, defined by the structural equation $f_{O}$, also applies to this outcome.
 This assumption removes the necessity of describing the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$.
@@
 
-Using [SCM -@fig-cj15_scm] we finally define the
-
-$\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1$ $\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1$
-
-<!-- To build the statistical part, start with the joint probability of observing the sample data P(Y, Z), then partition this joint probability in smaller conditional probabilities derived from the DAG.
--->
-
-<!-- The difficulty and discrimination parameters for the questions should then be given a diffuse, or ideally a hierarchical prior, which will identify these parameters by scaling and locating relative to the student ability parameters.
--->
-<!-- (https://mc-stan.org/docs/stan-users-guide/regression.html#priors-for-identification.section) -->
-
-<!-- when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification. -->
+<!-- Deriving a statistical model to address violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a) becomes straightforward when using SCMs.
+Because the structural equations $F$ in the [SCM -@fig-cj15_scm] encode probabilistic information, the SCM can represent the joint distribution of the variables more efficiently [@Pearl_et_al_2016, pp.
+29].
+Specifically, it expresses the joint distribution as a product of conditional probability distributions (CPDs) of the type $P(child \cond parents)$.
+This property is formally known as the *Bayesian Network factorization* (BNF) [@Pearl_et_al_2016, pp.
+29; @Neal_2020, pp.
+21].
+Equation [-@eq-mat28] presents the joint distribution of the sample data and partitions it into smaller CPDs derived from the SCM structure.
+-->
+
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!--   P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\ -->
+<!--   &= P(O_{R} \;|\; S, C, D_{R}) \cdot P(D_{R} \;|\; T_{IA}, B_{JK}) \cdot P(T_{IA} \;|\; T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \;|\; X_{I}, e_{I}) \\ -->
+<!--   & \quad \cdot P(B_{JK} \;|\; B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \;|\; Z_{J}, e_{J}) \cdot P(S) \cdot P(C) \cdot P(X_{IA}) \cdot P(X_{I}) \\  -->
+<!--   & \quad \cdot P(Z_{JK}) \cdot P(Z_{J}) \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J}) -->
+<!-- \end{aligned} -->
+<!-- $$ {#eq-mat28} -->
+
+<!-- 1. How to derive an stat model from an SCM? aboutn the outcome distribution -->
+Researchers can easily derive a statistical model to address violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a) using only [SCM -@fig-cj15_scm].
+This derivation is possible because the structural equations $F$ encode probabilistic information [@Pearl_et_al_2016], allowing researchers to replace them with suitable probabilistic assumptions.
+As a first step, given the dichotomous nature of CJ outcomes, the model [-@fig-cj16_stat] assumes that $O_{R}$ follows a Bernoulli distribution[^6].
+Next, following the conventions of Generalized Linear Latent And Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the model links the observed outcome to the latent discriminal difference vector $D_{R}$ through an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
+Notice that the sampling and comparison mechanisms, $S$ and $C$, act as element-wise multipliers of the discriminal difference (denoted by the symbol $\odot$), incorporating the design structure into the statistical formulation.
+Researchers can omit these vectors when only the observed CJ sample is available.
+
+<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
+...
+For any model whose graph is acyclic, the joint distribution of the variables in the model is given by the product of the conditional distributions P(child|parents) over all the "families" in the graph.
+[@Pearl_et_al_2016, pp.
+29] -->
+
+[^6]: The binomial distribution--including its special case, the Bernoulli distribution-- "represents the maximum entropy way to count binary events" [@McElreath_2020, pp.
+34].
+This means that when only two un-ordered outcomes are possible, and the expected numbers of each type of event are assumed to be constant, then the binomial distribution is the one that is most consistent with these constraints [@McElreath_2020, pp.
+310].
+For a detailed discussion of the binomial as a maximum entropy distribution, see @McElreath_2020 [chap.
+10.1.2].
+
+<!-- The binomial distribution is rather special, because it represents the maximum entropy way to count binary events. [@McElreath_2020, pp. 34] -->
+
+<!-- Choosing the distribution with the largest entropy means spreading probability as evenly as possible, while still remaining consistent with anything we think we know about a process.
+[@McElreath_2020, pp.
+299] -->
+
+<!-- The principle of maximum entropy helps us choose likelihood functions, by providing a way to use stated assumptions about constraints on the outcome variable to choose the likelihood function that is the most conservative distribution compatible with the known constraints.
+[@McElreath_2020, pp.
+300] -->
+
+<!-- There are a few conceptual lessons to take away from this example.
+First, hopefully it reinforces the maximum entropy nature of the binomial distribution.
+When only two un-ordered outcomes are possible—such as blue and white marbles—and the expected numbers of each type of event are assumed to be constant, then the distribution that is most consistent with these constraints is the binomial distribution.
+-->
+<!-- Second, of course usually we do not know the expected value, but wish to estimate it.
+If only two un-ordered outcomes are possible and you think the process generating them is invariant in time—so that the expected value remains constant at each combination of predictor values—then the distribution that is most conservative is the binomial.
+-->
+<!-- Third, back in Chapter 2, we derived the binomial distribution just by counting how many paths through the garden of forking data were consistent with our assumptions.
+...
+The likelihoods derived in that way turn out to be exactly the same as the likelihoods we get by maximizing entropy.
+This is not a coincidence.
+Entropy counts up the number of different ways a process can produce a particular result, according to our assumptions.
+[@McElreath_2020, pp.
+310-311, chap.
+10.1.2] -->
 
@@
   O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
-  D_{R} & = \left( T_{IA}[R_{1},R_{2}] - T_{IA}[R_{3},R_{4}] \right) + B_{JK}[R_{5},R_{6}] \\
-  T_{IA} & = T_{I}[IA_{1}] + \beta_{XA} X_{IA} + e_{IA} \\
-  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
-  B_{JK} & = B_{J}[JK_{1}] + \beta_{ZK} Z_{JK} + e_{JK} \\
-  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
-  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
-  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
-  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
+  D_{R} & = \left( T_{IA}[h,a] - T_{IA}[i,b] \right) + B_{JK}[j,k] \\
+  T_{IA}[i,b] & = T_{I}[i] + \beta_{XA} X_{IA}[i,] + e_{IA}[i,b] \\
+  T_{I}[i] & = \beta_{XI} X_{I}[i,] + e_{I}[i] \\
+  B_{JK}[j,k] & = B_{J}[j] + \beta_{ZK} Z_{JK}[j,] + e_{JK}[j,k] \\
+  B_{J}[j] & = \beta_{ZJ} Z_{J}[j,] + e_{J}[j] \\
+  e_{I}[i] & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
+  e_{J}[j] & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
+  e_{IA}[i,b] & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
+  e_{JK}[j,k] & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
 \end{aligned}
@@
 
-SCM
+[SCM -@fig-cj15_scm]
 :::
@@
 :::
+
+<!-- 2. Continuing with the probabilistic assumptions -->
+Furthermore, the statistical model defines the discriminal difference $D_{R}$ as the difference between the writing-quality traits of two distinct texts, $T_{IA}[h,a]$ and $T_{IA}[i,b]$, along with the corresponding repeated judge biases, $B_{JK}[j,k]$.
+The square brackets indicate the indices of the elements within their respective trait vectors and matrices, as in equation ([-@eq-mat11]) and ([-@eq-mat12]).
+
+<!-- The difficulty and discrimination parameters for the questions should then be given a diffuse, or ideally a hierarchical prior, which will identify these parameters by scaling and locating relative to the student ability parameters.
+-->
+<!-- (https://mc-stan.org/docs/stan-users-guide/regression.html#priors-for-identification.section) -->
+
+<!-- when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification. -->
+
+$\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1$ $\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1$
 
@@
 In this expression, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.
-
 $$

### 2025-04-08 _tex/paper2.tex
Built site for gh-pages

@@
 rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
-are latent vectors of length \(n_{I} \cdot n_{A}\).
+are latent matrices with \(n_{I}\) rows and \(n_{A}\) columns.
 
@@
 observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
-columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent vectors
-of length \(n_{J} \cdot n_{K}\).
+columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent
+matrices with \(n_{J}\) rows and \(n_{K}\) columns
 
@@
 sample of students \((n^{s}_{I})\), all drawn from the conceptual
-population. Moreover, in Section~\ref{sec-theory-theoretical_SC2}, the
-model assumes that judges do not perform \emph{all repeated judgments}
-within the data sample. Instead, they conduct a sufficient number of
-stimuli comparisons, \(n_{C}\), to ensure an accurate estimation of the
-proportion \(P(B>A)\).
+population. Furthermore, in Section~\ref{sec-theory-theoretical_SC2},
+the model assumes that judges do not perform \emph{all repeated
+judgments} within the data sample. Instead, they conduct a sufficient
+number of stimuli comparisons, \(n_{C}\), to ensure an accurate
+estimation of the proportion \(P(B>A)\), as proposed by
+\citet{Thurstone_1927b}.
 
@@
 presence or absence of corresponding elements in the vector \(I\), as in
-equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\), based
-on the vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and
-\(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements, respectively.
+equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\) using
+vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and \(S_{J}\)
+contains \(n_{I}\) and \(n_{J}\) elements, respectively.
 \begin{equation}\phantomsection\label{eq-mat22}{
@@
 presence or absence of data rows in the matrices \(IA\), as defined in
-equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\), based
-on the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
+equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\) using
+the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
 \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
@@
 equation (\ref{eq-mat13}). Suppose researchers exclude the second
-student and the third judge. They also omit the second text from each
-individual and the third repeated judgment from each judge. Given
+student and the third judge, and they also omit the second text from
+each individual and the third repeated judgment from each judge. Given
 \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
@@
 \citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
-sample equivalent for the \(R\) matrix, that is, this vector selects
-\emph{all repeated judgments a subset of judges makes for the subset of
-stimuli produced by the sampled individuals}. This process generates the
-sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
-outcome \(O^{cp}_{R}\).
+vector selecting \emph{all repeated judgments a subset of judges makes
+for the subset of stimuli produced by the sampled individuals}. This
+process generates the sample-comparison outcome \(O^{sc}_{R}\).
 
@@
 sampled judges, stimuli, and individuals--due to concerns about the
-practical feasibility of the comparison task \citep{Boonen_et_al_2020}--
-a realistic scenario must also account for judges comparing only
-specific stimuli from certain individuals.
+practical feasibility of the comparison task
+\citep{Boonen_et_al_2020}--a realistic scenario must also account for
+judges comparing only specific stimuli from certain individuals.
 
@@
 \(R\) similar to equation (\ref{eq-mat26}). This definition assumes that
-the subset of judges \emph{does not} perform all repeated judgments for
-the subset of stimuli produced by the sampled individuals. Instead, they
+the sampled judges \emph{do not} perform all repeated judgments for the
+subset of stimuli produced by the sampled individuals. Instead, they
 conduct a ``sufficient number of comparisons'' \(n_{C}\) to ensure the
@@
 makes for the subset of stimuli produced by the sampled individuals}.
-This process ultimately derives the final comparative outcome \(O_{R}\)
-from the sample-comparison outcome \(O^{sc}_{R}\).
+This process ultimately derives the final comparative outcome \(O_{R}\).
 
@@
 \(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
-applies to this outcome \(O_{R}\). This assumption removes the necessity
-of describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).
+applies to this outcome. This assumption removes the necessity of
+describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).
 
@@
 
-Using SCM~\ref{fig-cj15_scm} we finally define the
-
-\(\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1\)
-\(\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1\)
+Researchers can easily derive a statistical model to address violations
+of the equal dispersion assumption in CJ experiments (see
+Section~\ref{sec-theory-issue1a}) using only SCM~\ref{fig-cj15_scm}.
+This derivation is possible because the structural equations \(F\)
+encode probabilistic information \citep{Pearl_et_al_2016}, allowing
+researchers to replace them with suitable probabilistic assumptions. As
+a first step, given the dichotomous nature of CJ outcomes, the model
+\ref{fig-cj16_stat} assumes that \(O_{R}\) follows a Bernoulli
+distribution\footnote{The binomial distribution--including its special
+  case, the Bernoulli distribution-- ``represents the maximum entropy
+  way to count binary events'' \citep[pp.~34]{McElreath_2020}. This
+  means that when only two un-ordered outcomes are possible, and the
+  expected numbers of each type of event are assumed to be constant,
+  then the binomial distribution is the one that is most consistent with
+  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
+  discussion of the binomial as a maximum entropy distribution, see
+  \citet[chap.~10.1.2]{McElreath_2020}.}. Next, following the
+conventions of Generalized Linear Latent And Mixed Models (GLLAMMs)
+\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
+model links the observed outcome to the latent discriminal difference
+vector \(D_{R}\) through an inverse-logit function:
+\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Notice that the sampling
+and comparison mechanisms, \(S\) and \(C\), act as element-wise
+multipliers of the discriminal difference (denoted by the symbol
+\(\odot\)), incorporating the design structure into the statistical
+formulation. Researchers can omit these vectors when only the observed
+CJ sample is available.
 
@@
 
-\subcaption{\label{fig-cj16_scm}SCM}
+\subcaption{\label{fig-cj16_scm}SCM~\ref{fig-cj15_scm}}
 
@@
 \end{figure}%
+
+Furthermore, the statistical model defines the discriminal difference
+\(D_{R}\) as the difference between the writing-quality traits of two
+distinct texts, \(T_{IA}[h,a]\) and \(T_{IA}[i,b]\), along with the
+corresponding repeated judge biases, \(B_{JK}[j,k]\). The square
+brackets indicate the indices of the elements within their respective
+trait vectors and matrices, as in equation (\ref{eq-mat11}) and
+(\ref{eq-mat12}).
+
+\(\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1\)
+\(\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1\)
 
@@
 the parents of \(X_{i}\).
-
 \begin{equation}\phantomsection\label{eq-net_factor}{

### 2025-04-09 paper2.qmd
section:5; process:write,formulas

@@
 269], he did not address how to use the resulting measurement scores for hypothesis testing.
-Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022].
+Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978].
 Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?
@@
 269] -->
+
+<!-- Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates.
+In this sense the models are also "sample-free."
+However, this property of the model is true in empirical data only if the model holds in the data.
+Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
+[@Andrich_1978, pp.459] -->
 
@@
 <!-- Model misspecification happens when the set of probability distributions considered by the statistician does not include the distribution that generated the observed data. -->
+
+<!-- An alternative to the normal density function is the logistic.
+The substitution of the logistic for the normal is commonly made in circumstances such as the above (cf.
+Birnbaum, 1968; Bock, 1975) because it is more tractable.
+[@Andrich_1978, pp.
+455] -->
 
@@
 
+<!-- The correlation $\rho$, considered above is not the kind of correlation considered by Bock (1958) in studies of preference and choice of objects.
+The Bock correlation arises from a specific interaction between a subject and an item because the subject has a particular "preference", such as a taste preference, for the object.
+This preference for the object is then reflected in its comparisons with all other objects.
+However, it is less likely that subjects would have such particular preferences for some items in a test and, in any case, the assumption is open to empirical checking [@Andrich_1978, pp.
+455].
+-->
+
 <!-- 2. The first scenario where the assumption does not hold -->
@@
 -->
+
+<!-- The elimination of the subject parameters in the Rasch model has given rise to such expressions "sample-free" and "population-free" In this sense, then, so is the Thurstone Case V model.
+Although these phrases suggest desirable properties, they are not always fully appreciated; therefore, these terms have been avoided so far.
+A clarification seems in order and possible in this context.
+Because subject parameters are eliminated explicitly, their values (individually or collectively) need not be known.
+As a result, no assumptions need be made about the distribution of these parameters in the population, for example, that they are normally distributed.
+This does not mean, of course, that the "empirical" subject population need not be specified or described or that the scale values will be the same in different empirical populations.
+In fact, it is generally vital that the empirical populations be described.
+But because it does not matter what the actual values of the subject parameters are in the population, the models are, in the statistical distribution sense of population,
+"population-free" This is a property of the model and independent of empirical data.
+It is important to note that this holds only with respect to the estimation of the relative scale values of the items.
+It does not matter if the subject parameters are large, small, or otherwise; the subjects' impressions of the relative positions of the items should be the same.
+That is, items which appear more intense to one group should appear similarly more intense to the other groups.
+Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates.
+In this sense the models are also "sample-free."
+However, this property of the model is true in empirical data only if the model holds in the data.
+Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
+Conversely, if the relative scale values are not sample-free, the model does not hold across these samples.
+This feature is often used in checks on the Rasch model, and the usual sample formation is according to total scores on the items.
+Therefore, the check is with respect to an interaction between relative scale values and the level of the subject parameters.
+This is one of the best checks since it is then known that the subject parameter values are different.
+However, other samples may also be formed (e.g., on the basis of sex); then the check is with respect to a sex by relative scale value interaction.
+If there is no interaction, then it can be inferred that the items are working in the same relative way in both groups.
+As a result, generalizability of the scale across the groups is demonstrated.
+[@Andrich_1978, pp.459] -->
 
@@
 [SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the conceptual-population model, which evaluates the impact of different teaching methods on students' writing ability.
-This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{haibjk})$ and several related variables.
-The subscripts $h$ and $i$ identify the students who authored the texts (i.e., the individuals).
+This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{iahbjk})$ and several related variables.
+The subscripts $i$ and $h$ identify the students who authored the texts (i.e., the individuals).
 The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli).
@@
 366-376], accounting for experimental conditions where a judge compares the same pair of stimuli multiple times.
-Thus, the indexing system supports comparisons between different texts written by the same student $(h = i;$ $a \neq b)$ and between texts written by distinct students $(h \neq i;$ where $a = b$ is permitted$)$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$.
-However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(h = i;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K};$ where $n_{J}>1$ and $n_{K}\geq1)$, as such comparison lacks practical relevance within the CJ framework.
+Thus, the indexing system supports comparisons between different texts written by the same student $(i = h;$ $a \neq b)$ and between texts written by distinct students $(i \neq h;$ where $a = b$ is permitted$)$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$.
+However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(i = h;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K};$ where $n_{J}>1$ and $n_{K}\geq1)$, as such comparison lacks practical relevance within the CJ framework.
 Here, $n_{J}$ indicates the total number of judges, and $n_{K}$ denotes the number of repeated judgments each judge performs.
 
-In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{haibjk})$ (see @sec-thurstone_theory).
-Additionally, based on the arguments developed in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$.
-Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{haibjk}, D_{haibjk}, T_{ha}, T_{ib}, B_{kj} \}$.
+In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory).
+Additionally, based on the arguments developed in @sec-theory-issue1b, the SCM incorporates the judges' biases $(B_{kj})$.
+Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
 Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
 
-<!-- @Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
-Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
+<!-- Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
 Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
@@
 
+<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem for latent variables?. -->
+
 ::: {#fig-cj03 layout-nrow=2}
@@
 \begin{aligned}
-  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
-  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
+  O^{cp}_{iahbjk} & := f_{O}(D_{iahbjk}) \\
+  D_{iahbjk} & := f_{D}(T_{ia}, T_{hb}, B_{jk})
 \end{aligned}
@@
 @fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the expected causal relationships outlined in Thurstone's theory.
-The graph shows that the discriminal processes of the texts $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$.
+The graph shows that the discriminal processes of the texts $(T_{ia}, T_{hb})$ influence their discriminal difference $(D_{iahbjk})$, which in turn determines the outcome $(O^{cp}_{iahbjk})$.
 It also highlights the influence of judges' biases $(B_{kj})$ on the discriminal difference.
@@
 \vdots \\
+i \\
+\vdots \\
 h \\
-\vdots \\
-i \\
 \vdots \\
@@
 \vdots & \vdots \\
-h & a \\
+i & a \\
 \vdots & \vdots \\
-i & b \\
+h & b \\
 \vdots & \vdots \\
@@
 <!-- 2. what is the data structure? -->
-Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals.
+Here, each element of $I$ represents a unique individual $i$ or $h$, where $n_{I}$ denotes the total number of individuals.
 Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges.
-Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$.
+Moreover, each row of $IA$ represents a unique pairing of individuals $i, h$ with stimuli $a, b$.
 As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
@@
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
-h & a & i & b & j & k \\
+i & a & h & b & j & k \\
 \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
@@
 Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
-Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' writing-quality trait $(T_{IA})$, the first primary latent variable of interest.
+Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest.
 The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
@@
 Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
@@
 \vdots \\
+i_{(i)} \\
+\vdots \\
 i_{(h)} \\
-\vdots \\
-i_{(i)} \\
 \vdots \\
@@
 \vdots \\
-ia_{(h,a)} \\
-\vdots \\
-ia_{(i,b)} \\
+ia_{(i,a)} \\
+\vdots \\
+ia_{(h,b)} \\
 \vdots \\
@@
 Finally, we define the *sample mechanism* $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
-Each element $s_{(h,a,i,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in equation ([-@eq-mat26]).
+Each element $s_{(i,a,h,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in equation ([-@eq-mat26]).
 Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
@@
 \vdots \\
-s_{(h,a,i,b,j,k)} \\
+s_{(i,a,h,b,j,k)} \\
 \vdots \\
@@
 $$
-s_{(h,i,a,b,j,k)} = \begin{cases} 
-1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
+s_{(i,a,h,b,j,k)} = \begin{cases} 
+1 & \text{if data elements } i,a,h,b,j,k \text{ from } R \text{ are sampled} \\
 0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
@@
 A square encloses this vector, indicating that it is a conditioned variable.
-Here, *conditioning* means that researchers restrict their focus to the elements of $O^{cp}_{R}$ that satisfy $s_{(h,i,a,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
+Here, *conditioning* means that researchers restrict their focus to the elements of $O^{cp}_{R}$ that satisfy $s_{(i,a,h,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
 Essentially, $S$ serves as the vector selecting *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
@@
 <!-- 1. start by defining the comparison vector  -->
-As in the previous section, to facilitate the interpretation of the sample-comparison model, we first define the *comparison mechanism* using the binary vector variable $C$ as follows: 
+As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model (see equation ([-@eq-mat27])).
+The vector contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $S$ in equation ([-@eq-mat26]).
 $$
@@
 \vdots \\
-c_{(h,a,i,b,j,k)} \\
+c_{(i,a,h,b,j,k)} \\
 \vdots \\
@@
 
-Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(h,a,i,b,j,k)}$ being a binary value indicating the presence or absence of data rows in the $R$ similar to equation ([-@eq-mat26]).
-This definition assumes that the sampled judges *do not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals.
-Instead, they conduct a "sufficient number of comparisons" $n_{C}$ to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp.
+<!-- 2. What is the effect of the comparison mechanism -->
+The [DAG -@fig-cj14_dag] also integrates the comparison mechanism $C$ into the conceptual-population model, which ultimately derives the final comparative outcome $O_{R}$.
+First, it shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
+Second, it depicts the *comparison mechanism* $C$ as a causal factor influencing the final comparative outcome $O_{R}$.
+As with the sampling mechanism, a square encloses $C$ to indicate that it is a conditioned variable--that is, one that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli.
+Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp.
 267].
@@
 
-<!-- 2. What is the effect of the comparison mechanism -->
-[DAG -@fig-cj14_dag] also incorporates the comparison mechanism $C$ into the conceptual-population model.
-The DAG represents the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
-Moreover, it represents the *comparison mechanism* $C$ as a causal factor influencing the final comparative outcome $O_{R}$.
-Equivalent to the sampling mechanism, a square encloses this vector, indicating that $C$ is a conditioned variable that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-This process ultimately derives the final comparative outcome $O_{R}$.
-
-Notably, the DAG shows that $C$ is also independent of all other variables in the model.
-This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015] or Incomplete Block Designs [@Lawson_2015].
+Notably, [DAG -@fig-cj14_dag] shows that $C$ is independent of all other variables in the model.
+This independence implies that the conceptual model represented by the DAG applies exclusively to Random Allocation Comparative Designs [@Bramley_2015] or Incomplete Block Designs [@Lawson_2015].
 In these designs, every repeated judgment has an equal probability of being included in the sample.
@@
 Researchers can easily derive a statistical model to address violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a) using only [SCM -@fig-cj15_scm].
-This derivation is possible because the structural equations $F$ encode probabilistic information [@Pearl_et_al_2016], allowing researchers to replace them with suitable probabilistic assumptions.
-As a first step, given the dichotomous nature of CJ outcomes, the model [-@fig-cj16_stat] assumes that $O_{R}$ follows a Bernoulli distribution[^6].
-Next, following the conventions of Generalized Linear Latent And Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the model links the observed outcome to the latent discriminal difference vector $D_{R}$ through an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
-Notice that the sampling and comparison mechanisms, $S$ and $C$, act as element-wise multipliers of the discriminal difference (denoted by the symbol $\odot$), incorporating the design structure into the statistical formulation.
+This derivation is possible because the structural equations $F$ encode probabilistic information [@Pearl_et_al_2016], which researchers can replace with suitable probabilistic assumptions.
+As a starting point, given the dichotomous nature of CJ outcomes, [model -@fig-cj16_stat] assumes that $O_{R}$ follows a Bernoulli distribution[^6].
+
+<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
+This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
+12].
+-->
+
+Next, following the conventions of Generalized Linear Latent And Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the model links the observed outcome to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
+Notice that the sampling and comparison mechanisms, $S$ and $C$, act as element-wise multipliers of $D_{R}$ (denoted by the symbol $\odot$), directly incorporating key assessment design features relevant to CJ experiments into the statistical formulation.
 Researchers can omit these vectors when only the observed CJ sample is available.
@@
 
-[^6]: The binomial distribution--including its special case, the Bernoulli distribution-- "represents the maximum entropy way to count binary events" [@McElreath_2020, pp.
+[^6]: The binomial distribution--including its special case, the Bernoulli distribution-- represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
 34].
@@
   O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
-  D_{R} & = \left( T_{IA}[h,a] - T_{IA}[i,b] \right) + B_{JK}[j,k] \\
-  T_{IA}[i,b] & = T_{I}[i] + \beta_{XA} X_{IA}[i,] + e_{IA}[i,b] \\
-  T_{I}[i] & = \beta_{XI} X_{I}[i,] + e_{I}[i] \\
-  B_{JK}[j,k] & = B_{J}[j] + \beta_{ZK} Z_{JK}[j,] + e_{JK}[j,k] \\
-  B_{J}[j] & = \beta_{ZJ} Z_{J}[j,] + e_{J}[j] \\
-  e_{I}[i] & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
-  e_{J}[j] & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
-  e_{IA}[i,b] & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
-  e_{JK}[j,k] & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
+  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
+  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
+  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
+  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
+  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
+  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
+  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
 \end{aligned}
@@
 <!-- 2. Continuing with the probabilistic assumptions -->
-Furthermore, the statistical model defines the discriminal difference $D_{R}$ as the difference between the writing-quality traits of two distinct texts, $T_{IA}[h,a]$ and $T_{IA}[i,b]$, along with the corresponding repeated judge biases, $B_{JK}[j,k]$.
-The square brackets indicate the indices of the elements within their respective trait vectors and matrices, as in equation ([-@eq-mat11]) and ([-@eq-mat12]).
+Then, [model -@fig-cj16_stat] defines $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, which represent their underlying written-quality trait.
+This difference also includes the corresponding repeated judge biases, $B_{JK}[j, k]$.
+Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, then we can re-write the discriminal difference as:
+$$
+\begin{aligned}
+D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] ] \\
+&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
+& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
+\end{aligned}
+$$ {#eq-mat51}
+This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022].
+In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
+For clarity, the square brackets in $D_{R}$ only indicate the relevant indices within each trait vector; they do not represent any data subsetting.
+
+<!-- Weinberg[4] proposed a further diagram to capture ‘pure’ interaction in which neither G nor E has a direct effect on D and they only have an effect when present together (see Figure 4a).
+We suggest that this ‘pure’ interaction could be captured by the DAG shown in Figure 4b.
+An example of this is phenylketonuria in which phenylalanine in the diet (E) and the absence of the enzyme to break it down (phenylalanine hydroxylase, G) are both needed in order for a disease to manifest (intellectual disability, D).
+[@Attia_et_al_2022, pp.
+1052] -->
+
+<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder?
+-->
+
+Next, the model defines $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length), and the texts' idiosyncratic errors $e_{IA}$.
+The model assumes that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$, defined as a proportion of $1$.
+This standardization establishes the scale of the latent trait, as it is required by latent variable models [@Depaoli_2021].
+
+<!-- Within CFA, it is also necessary to resolve a series of indeterminacies, which are largely linked to the fact that the latent factors do not have a scale of their own.
+Indeterminacies exist in terms of the scale, location, and orientation of the factors.
+In order to handle these indeterminacies, we must set the scale of each of the latent factors.
+This can be handled in one of two ways.
+The first method that can be used is to standardize the latent factors such that the variances are set to 1 and the latent factor means are set to 0.
+The second method, and the one employed in the examples below, is to fix an otherwise free factor loading to a value of 1.
+[@Depaoli_2021, pp.
+93] -->
 
 <!-- The difficulty and discrimination parameters for the questions should then be given a diffuse, or ideally a hierarchical prior, which will identify these parameters by scaling and locating relative to the student ability parameters.
--->
-<!-- (https://mc-stan.org/docs/stan-users-guide/regression.html#priors-for-identification.section) -->
+(https://mc-stan.org/docs/stan-users-guide/regression.html#priors-for-identification.section) -->
+
+Additionally, the model defines $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$.
+As before, each element of $e_{I}$ is $iid$ following a Normal distribution with a mean of zero.
+However, the standard deviation, $s_{XI}$, varies depending on the teaching method to which each student belongs.
+Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$  This constraint effectively addresses concerns about the equal dispersion assumption raised in @sec-theory-issue1a.
 
@@
 
-$\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1$ $\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1$
-
-<!-- Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data.
--->
-
-<!-- These estimates are obtained through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010]. -->
+Similarly, the model defines $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., the number of judgments a judge makes), and judgment-specific idiosyncratic errors $e_{JK}$.
+The model assumes that each element of $e_{JK}$ is $iid$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].
+
+The model also defines $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$.
+Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs.
+For instance, if $Z_{J} = 1, \dots, 3$ represents three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
+
+<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->
+
+<!-- [@deAyala_2009, p.
+291-29] think about indeterminancies in MIRT.
+to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
+
+Finally, we apply *Bayesian inference methods* to turn [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data.
+The **Declarations** section of this document provides a link to this model, along with an alternative specification that assumes equal discriminal dispersions.
+We implemented both models in `Stan` [version $2.26.1$, @Stan_2020].
 
@@
 
-<!-- Statements about the real world from observed data require assumptions.
-This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
-The validity of our statements therefore critically hinges on the validity of those assumptions.
-If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are [@Kohler_et_al_2019, pp.
-150] -->
-
 <!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
-
-<!-- 6. So, how do we mitigate these risks? -->
-<!-- Fortunately, the statistical literature offers a solution to address the abovementioned issues.
-This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
-12].
--->
-
-<!-- 11. How to solve the violation of the assumption? -->
-<!-- Fortunately, the same literature offers solutions for addressing these issues.
-@Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model.
-Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies.
-Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap.
-26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap.
-15], depending on the theoretical and statistical treatment of judges' biases.
--->
-
-<!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem for latent variables?. -->
-
-<!-- 4. So, how do we mitigate these risks? -->
-<!-- To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp.
-138] and IRT [@Fox_2010, chap.
-6; @vanderLinden_et_al_2017_I, chap.
-24] recommend conducting these analyses and tests within a structural model.
-A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010].
-This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
--->
-
-<!-- Remember: the "mitigate" paragraph have the purpose of being the design principles that will guide the development of the theoretical and statistical model. -->
-
-<!-- In other words, researchers observe the CJ outcome only when texts' discriminal processes and the judges biases create a *pure interaction effect*, meaning that neither of them has a direct effect on the outcome, but their interaction produces an effect [@Attia_et_al_2022].
--->
-
-<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->
-
-<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->
-
-<!-- [@deAyala_2009, p.
-291-29] think about indeterminancies in MIRT.
-to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
-
-<!-- The correlation $\rho$, considered above is not the kind of correlation considered by Bock (1958) in studies of preference and choice of objects.
-The Bock correlation arises from a specific interaction between a subject and an item because the subject has a particular "preference", such as a taste preference, for the object.
-This preference for the object is then reflected in its comparisons with all other objects.
-However, it is less likely that subjects would have such particular preferences for some items in a test and, in any case, the assumption is open to empirical checking [@Andrich_1978, pp.
-455].
--->
 
@@
 678] -->
+
+<!-- These procedures also bear on a larger topic of discussion in social science.
+The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
+[@Wainer_et_al_1978, pp.
+332] -->
 

### 2025-04-09 _tex/paper2.tex
Built site for gh-pages

@@
 measurements in contexts where the case's assumptions may not hold
-\citep{Kelly_et_al_2022}. Furthermore, although the CJ practice has
-attempted to address the issue of hypothesis testing by using the
-scores' point estimates or their transformations, a critical question
-remains: Is this approach suitable for addressing these inquiries?
+\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
+practice has attempted to address the issue of hypothesis testing by
+using the scores' point estimates or their transformations, a critical
+question remains: Is this approach suitable for addressing these
+inquiries?
 
@@
 outlines the relationship between the conceptual-population outcome
-\((O^{cp}_{haibjk})\) and several related variables. The subscripts
-\(h\) and \(i\) identify the students who authored the texts (i.e., the
+\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
+\(i\) and \(h\) identify the students who authored the texts (i.e., the
 individuals). The indices \(a\) and \(b\) represent the texts under
@@
 multiple times. Thus, the indexing system supports comparisons between
-different texts written by the same student \((h = i;\) \(a \neq b)\)
-and between texts written by distinct students \((h \neq i;\) where
+different texts written by the same student \((i = h;\) \(a \neq b)\)
+and between texts written by distinct students \((i \neq h;\) where
 \(a = b\) is permitted\()\), each judged once or repeatedly by all
@@
 and \(n_{K}\geq1)\). However, it excludes cases where a judge compares a
-student's text to itself, whether once or multiple times \((h = i;\)
+student's text to itself, whether once or multiple times \((i = h;\)
 \(a = b;\) \(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K};\) where
@@
 In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
-texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
-difference \((D_{haibjk})\) (see Section~\ref{sec-thurstone_theory}).
+texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
+difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
 Additionally, based on the arguments developed in
-Section~\ref{sec-theory-issue1b} and the recommendations of
-\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
-the judges' biases \((B_{kj})\). Together with the outcome, these
-variables constitute the preliminary set of endogenous variables,
-\(V = \{ O_{haibjk}, D_{haibjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
+Section~\ref{sec-theory-issue1b}, the SCM incorporates the judges'
+biases \((B_{kj})\). Together with the outcome, these variables
+constitute the preliminary set of endogenous variables,
+\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,
 the SCM presents the preliminary set of structural equations,
@@
 outlined in Thurstone's theory. The graph shows that the discriminal
-processes of the texts \((T_{ha}, T_{ib})\) influence their discriminal
-difference \((D_{hiabjk})\), which in turn determines the outcome
-\((O^{cp}_{hiabjk})\). It also highlights the influence of judges'
+processes of the texts \((T_{ia}, T_{hb})\) influence their discriminal
+difference \((D_{iahbjk})\), which in turn determines the outcome
+\((O^{cp}_{iahbjk})\). It also highlights the influence of judges'
 biases \((B_{kj})\) on the discriminal difference. Additionally, the DAG
@@
 
-Here, each element of \(I\) represents a unique individual \(h\) or
-\(i\), where \(n_{I}\) denotes the total number of individuals.
+Here, each element of \(I\) represents a unique individual \(i\) or
+\(h\), where \(n_{I}\) denotes the total number of individuals.
 Similarly, each element of \(J\) corresponds to a unique judge \(j\),
 with \(n_{J}\) indicating the total number of judges. Moreover, each row
-of \(IA\) represents a unique pairing of individuals \(h, i\) with
+of \(IA\) represents a unique pairing of individuals \(i, h\) with
 stimuli \(a, b\). As a result, the matrix \(IA\) contains
@@
 idiosyncratic errors \(e_{IA}\) causally influence the texts'
-writing-quality trait \((T_{IA})\), the first primary latent variable of
+written-quality trait \((T_{IA})\), the first primary latent variable of
 interest. The error term \(e_{IA}\) captures variations in the texts'
@@
 and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
-biases associated with each text \(B_{JK}\), the second primary latent
-variable of interest. The error \(e_{JK}\) captures variations in
+judges' biases associated with each text \(B_{JK}\), the second primary
+latent variable of interest. The error \(e_{JK}\) captures variations in
 judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
@@
 \vdots \\
-ia_{(h,a)} \\
+ia_{(i,a)} \\
 \vdots \\
-ia_{(i,b)} \\
+ia_{(h,b)} \\
 \vdots \\
@@
 (\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
-of \(S_{JK}\). Each element \(s_{(h,a,i,b,j,k)}\) is a binary value
+of \(S_{JK}\). Each element \(s_{(i,a,h,b,j,k)}\) is a binary value
 indicating the presence or absence of data rows in the matrix \(R\)
@@
 \begin{equation}\phantomsection\label{eq-mat26}{
-s_{(h,i,a,b,j,k)} = \begin{cases} 
-1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
+s_{(i,a,h,b,j,k)} = \begin{cases} 
+1 & \text{if data elements } i,a,h,b,j,k \text{ from } R \text{ are sampled} \\
 0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
@@
 Here, \emph{conditioning} means that researchers restrict their focus to
-the elements of \(O^{cp}_{R}\) that satisfy \(s_{(h,i,a,b,j,k)}=1\)
+the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
 \citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
@@
 
-As in the previous section, to facilitate the interpretation of the
-sample-comparison model, we first define the \emph{comparison mechanism}
-using the binary vector variable \(C\) as follows:
+As in the previous section, we begin defining the \emph{comparison
+mechanism} using the binary vector variable \(C\) to facilitate the
+interpretation of the sample-comparison model (see equation
+(\ref{eq-mat27})). The vector contains \(n\) elements corresponding to
+the number of rows in the \(R\) matrix, with each element
+\(c_{(i,a,h,b,j,k)}\) being a binary value indicating the presence or
+absence of data rows in \(R\), a definition similar to that of \(S\) in
+equation (\ref{eq-mat26}).
 \begin{equation}\phantomsection\label{eq-mat27}{
@@
 
-Where \(C\) contains \(n\) elements corresponding to the number of rows
-in the \(R\) matrix, with each element \(c_{(h,a,i,b,j,k)}\) being a
-binary value indicating the presence or absence of data rows in the
-\(R\) similar to equation (\ref{eq-mat26}). This definition assumes that
-the sampled judges \emph{do not} perform all repeated judgments for the
-subset of stimuli produced by the sampled individuals. Instead, they
-conduct a ``sufficient number of comparisons'' \(n_{C}\) to ensure the
-accurate estimation of the proportion \(P(B>A)\) for each pair of
-stimuli \citep[pp.~267]{Thurstone_1927b}.
-
-DAG~\ref{fig-cj14_dag} also incorporates the comparison mechanism \(C\)
-into the conceptual-population model. The DAG represents the
+The DAG~\ref{fig-cj14_dag} also integrates the comparison mechanism
+\(C\) into the conceptual-population model, which ultimately derives the
+final comparative outcome \(O_{R}\). First, it shows the
 sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
 that researchers cannot directly access this outcome due to the
-comparison mechanism. Moreover, it represents the \emph{comparison
-mechanism} \(C\) as a causal factor influencing the final comparative
-outcome \(O_{R}\). Equivalent to the sampling mechanism, a square
-encloses this vector, indicating that \(C\) is a conditioned variable
-that determines \emph{which repeated judgments the subset of judges
-makes for the subset of stimuli produced by the sampled individuals}.
-This process ultimately derives the final comparative outcome \(O_{R}\).
-
-Notably, the DAG shows that \(C\) is also independent of all other
-variables in the model. This independence implies that the conceptual
-model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
-Random Allocation Comparative Designs \citep{Bramley_2015} or Incomplete
-Block Designs \citep{Lawson_2015}. In these designs, every repeated
-judgment has an equal probability of being included in the sample.
+comparison mechanism. Second, it depicts the \emph{comparison mechanism}
+\(C\) as a causal factor influencing the final comparative outcome
+\(O_{R}\). As with the sampling mechanism, a square encloses \(C\) to
+indicate that it is a conditioned variable--that is, one that determines
+\emph{which repeated judgments the subset of judges makes for the subset
+of stimuli produced by the sampled individuals}. In essence, \(C\)
+reflects the assumption that judges \emph{do not} perform all possible
+repeated judgments for the selected stimuli. Instead, they carry out a
+``sufficient number of comparisons'' \(n_{C}\) to allow for accurate
+estimation of the proportion \(P(B>A)\) for each stimulus pair
+\citep[pp.~267]{Thurstone_1927b}.
+
+Notably, DAG~\ref{fig-cj14_dag} shows that \(C\) is independent of all
+other variables in the model. This independence implies that the
+conceptual model represented by the DAG applies exclusively to Random
+Allocation Comparative Designs \citep{Bramley_2015} or Incomplete Block
+Designs \citep{Lawson_2015}. In these designs, every repeated judgment
+has an equal probability of being included in the sample.
 
@@
 This derivation is possible because the structural equations \(F\)
-encode probabilistic information \citep{Pearl_et_al_2016}, allowing
-researchers to replace them with suitable probabilistic assumptions. As
-a first step, given the dichotomous nature of CJ outcomes, the model
-\ref{fig-cj16_stat} assumes that \(O_{R}\) follows a Bernoulli
+encode probabilistic information \citep{Pearl_et_al_2016}, which
+researchers can replace with suitable probabilistic assumptions. As a
+starting point, given the dichotomous nature of CJ outcomes,
+model~\ref{fig-cj16_stat} assumes that \(O_{R}\) follows a Bernoulli
 distribution\footnote{The binomial distribution--including its special
-  case, the Bernoulli distribution-- ``represents the maximum entropy
-  way to count binary events'' \citep[pp.~34]{McElreath_2020}. This
+  case, the Bernoulli distribution-- represent a maximum entropy
+  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
   means that when only two un-ordered outcomes are possible, and the
@@
   discussion of the binomial as a maximum entropy distribution, see
-  \citet[chap.~10.1.2]{McElreath_2020}.}. Next, following the
-conventions of Generalized Linear Latent And Mixed Models (GLLAMMs)
+  \citet[chap.~10.1.2]{McElreath_2020}.}.
+
+Next, following the conventions of Generalized Linear Latent And Mixed
+Models (GLLAMMs)
 \citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
 model links the observed outcome to the latent discriminal difference
-vector \(D_{R}\) through an inverse-logit function:
+vector \(D_{R}\) using an inverse-logit function:
 \(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Notice that the sampling
 and comparison mechanisms, \(S\) and \(C\), act as element-wise
-multipliers of the discriminal difference (denoted by the symbol
-\(\odot\)), incorporating the design structure into the statistical
-formulation. Researchers can omit these vectors when only the observed
-CJ sample is available.
+multipliers of \(D_{R}\) (denoted by the symbol \(\odot\)), directly
+incorporating key assessment design features relevant to CJ experiments
+into the statistical formulation. Researchers can omit these vectors
+when only the observed CJ sample is available.
 
@@
 
-Furthermore, the statistical model defines the discriminal difference
-\(D_{R}\) as the difference between the writing-quality traits of two
-distinct texts, \(T_{IA}[h,a]\) and \(T_{IA}[i,b]\), along with the
-corresponding repeated judge biases, \(B_{JK}[j,k]\). The square
-brackets indicate the indices of the elements within their respective
-trait vectors and matrices, as in equation (\ref{eq-mat11}) and
-(\ref{eq-mat12}).
-
-\(\sum_{g=1}^{G_{XI}} s_{XI}/G_{XI} = 1\)
-\(\sum_{g=1}^{G_{ZJ}} s_{ZJ}/G_{ZJ} = 1\)
+Then, model~\ref{fig-cj16_stat} defines \(D_{R}\) as the difference
+between the discriminal processes of two texts, \(T_{IA}[i, a]\) and
+\(T_{IA}[h, b]\), which represent their underlying written-quality
+trait. This difference also includes the corresponding repeated judge
+biases, \(B_{JK}[j, k]\). Note that if we assume that \(B_{JK}[j,k]\)
+reflects the difference in stimulus-specific biases, i.e.,
+\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), then we can
+re-write the discriminal difference as:
+\begin{equation}\phantomsection\label{eq-mat51}{
+\begin{aligned}
+D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] ] \\
+&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
+& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
+\end{aligned}
+}\end{equation} This formulation reveals that the discriminal difference
+captures a \emph{pure interaction effect}, in which neither the texts'
+discriminal processes nor the judges' biases alone determine the
+outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
+terms, the discriminal processes of the stimuli become an observable
+outcome only through the judges' perceptions (biases). For clarity, the
+square brackets in \(D_{R}\) only indicate the relevant indices within
+each trait vector; they do not represent any data subsetting.
+
+Next, the model defines \(T_{IA}\) as a linear combination of the
+students' underlying writing-quality traits \(T_{I}\), the effects of
+relevant text-related variables \(\beta_{XA}X_{IA}\) (e.g., text
+length), and the texts' idiosyncratic errors \(e_{IA}\). The model
+assumes that each element of \(e_{IA}\) is independent and identically
+distributed \((iid)\) according to a Normal distribution, with a mean of
+zero and a standard deviation \(p_{IA}\), defined as a proportion of
+\(1\). This standardization establishes the scale of the latent trait,
+as it is required by latent variable models \citep{Depaoli_2021}.
+
+Additionally, the model defines \(T_{I}\) as a linear combination of the
+effects of relevant student-related variables, \(\beta_{XI} X_{I}\), and
+students' idiosyncratic errors, \(e_{I}\). As before, each element of
+\(e_{I}\) is \(iid\) following a Normal distribution with a mean of
+zero. However, the standard deviation, \(s_{XI}\), varies depending on
+the teaching method to which each student belongs. Given the example in
+Section~\ref{sec-theoretical}, where the teaching method
+\(X_{I} = 1,2\), the model sets the following constraint to anchor the
+scale of the latent trait \citep{Depaoli_2021}:
+\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\) This constraint effectively addresses
+concerns about the equal dispersion assumption raised in
+Section~\ref{sec-theory-issue1a}.
+
+Similarly, the model defines \(B_{JK}\) as a linear combination of the
+judges' individual bias \(B_{J}\), the effects of relevant
+judgment-related variables \(\beta_{ZK}Z_{JK}\) (e.g., the number of
+judgments a judge makes), and judgment-specific idiosyncratic errors
+\(e_{JK}\). The model assumes that each element of \(e_{JK}\) is \(iid\)
+according to a Normal distribution, with a mean of zero and a standard
+deviation \(p_{JK}\), defined as a proportion of \(1\), to anchor the
+scale of the latent trait \citep{Depaoli_2021}.
+
+The model also defines \(B_{J}\) as a linear combination of the effects
+of relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
+expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
+element of \(e_{J}\) is likewise assumed to be \(iid\) Normal with mean
+zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
+which each judge belongs. For instance, if \(Z_{J} = 1, \dots, 3\)
+represents three groups of judges with varying expertise, the model
+imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
+scale of the latent trait \citep{Depaoli_2021}.
+
+Finally, we apply \emph{Bayesian inference methods} to turn
+model~\ref{fig-cj16_stat} into a practical statistical tool for
+analyzing paired comparison data. The \textbf{Declarations} section of
+this document provides a link to this model, along with an alternative
+specification that assumes equal discriminal dispersions. We implemented
+both models in \texttt{Stan} \citep[version \(2.26.1\),][]{Stan_2020}.
 

### 2025-04-11 paper2.qmd
section:5; process:write,formulas,code

@@
 \newcommand{\ndsep}{\:\not\bot\:}
-\newcommand{\cond}{\:|\:}
 
@@
 Before incorporating the first theoretical principles of Thurstone's theory, it is essential to further define SCMs.
-SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020].
+SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Pearl_et_al_2016; @Cinelli_et_al_2020].
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
@@
 <!-- These disturbances reflect "hidden" or unmeasurable conditions that Nature governs by some undisclosed probability function. [@Pearl_2009, pp. 89] -->
+
+<!-- Formally, a structural causal model consists of two sets of variables U and V, and a set of functions f that assigns each variable in V a value based on the values of the other variables in the model.
+...
+The variables in U are called exogenous variables, meaning, roughly, that they are external to the model; we choose, for whatever reason, not to explain how they are caused.
+The variables in V are endogenous.
+Every endogenous variable in a model is a descendant of at least one exogenous variable.
+Exogenous variables cannot be descendants of any other variables, and in particular, cannot be a descendant of an endogenous variable; they have no ancestors and are represented as root nodes in graphs.
+If we know the value of every exogenous variable, then using the functions in f , we can determine with perfect certainty the value of very endogenous variable.
+[@Pearl_et_al_2016, pp.
+26-27] -->
 
@@
 <!-- the *magnified representation* includes exogenous variables -->
+
+<!-- Every SCM is associated with a graphical causal model, referred to informally as a “graphical model” or simply “graph.” Graphical models consist of a set of nodes representing the variables in U and V, and a set of edges between the nodes representing the functions in f.
+[@Pearl_et_al_2016, pp.
+27] -->
+
+<!-- Every SCM has an associated graph G, usually called its causal diagram.
+In the types of models we consider here, which do not exhibit cycles, the causal diagram will be a directed acyclic graph (DAG).
+[@Cinelli_et_al_2020, pp.
+15] -->
 
@@
 
+<!-- Do not depict the variances, nor the sample sizes. Or if you do explain that it is with illustrative purposes and then remove them to make the reading of the DAGs easier. -->
+
 ### The sample mechanism {#sec-theory-theoretical_SC1}
@@
 <!-- recording: Sven and Tine 25.03.28; time: 00:11:30 - 00:12:25 -->
+
+<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
+C -> {Y}
+e_{A} -> {C, Y} 
+this means that e_{A} is a confounder?,  
+then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?.
+Does this mean that random algorithms per se will always be unbiased?.
+-->
 
@@
 
+<!-- recording: Sven and Tine 25.04.10; time: 00:16:30 - 00:19:00 -->
+
 <!-- 1. start by defining the comparison vector  -->
 As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model (see equation ([-@eq-mat27])).
-The vector contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $S$ in equation ([-@eq-mat26]).
+The vector contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in equation ([-@eq-mat26]).
 $$
@@
 <!-- 2. What is the effect of the comparison mechanism -->
-The [DAG -@fig-cj14_dag] also integrates the comparison mechanism $C$ into the conceptual-population model, which ultimately derives the final comparative outcome $O_{R}$.
-First, it shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
-Second, it depicts the *comparison mechanism* $C$ as a causal factor influencing the final comparative outcome $O_{R}$.
+The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$.
+It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
+Moreover, it depicts $C$ as a causal factor influencing $O_{R}$.
 As with the sampling mechanism, a square encloses $C$ to indicate that it is a conditioned variable--that is, one that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
@@
 
-Notably, [DAG -@fig-cj14_dag] shows that $C$ is independent of all other variables in the model.
+Notably, [DAG -@fig-cj14_dag] also shows that $C$ is independent of all other variables in the model.
 This independence implies that the conceptual model represented by the DAG applies exclusively to Random Allocation Comparative Designs [@Bramley_2015] or Incomplete Block Designs [@Lawson_2015].
 In these designs, every repeated judgment has an equal probability of being included in the sample.
-
-<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
-2, [@Deffner_et_al_2022] -->
 
@@
 <!-- This is a whole section [@Lawson_2015, pp. 259-271] -->
+
+<!-- Actually algorithm can be thought as the observation mechanism that "transforms" the outcome in the population of comparisons to the outcome in the sample of comparisons, see fig.
+2, [@Deffner_et_al_2022] -->
 
@@
 <!-- 3. preparing to transform the conceptual model into an statistical one -->
-Finally, without loss of generality, researchers can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
-This reformulation rests on two principles.
-First, because the sampling and comparison mechanism vectors $S$ and $C$ are binary, researchers can integrate them into the model using element-wise products.
-This approach eliminates the need to define the structural equations $f_{S}$ and $f_{C}$.
-Second, since only the final comparative outcome $O_{R}$ is observed, it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$, defined by the structural equation $f_{O}$, also applies to this outcome.
-This assumption removes the necessity of describing the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$.
+Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
+This reformulation produces a model that applies directly to a sample of comparative data.
+In this version, the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$ are omitted, and $O_{R}$ inherits the structural equation $f_{O}$ that originally defined $O^{cp}_{R}$.
+Moreover, the definition of $O_{R}$ now reflects its direct dependence on the discriminal difference $D_{R}$ and the sample and comparison mechanisms, $S$ and $C$.
 
 <!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
-For example, if I have the distribution of the population outcome, and it is a bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
+For example, if I have the distribution of the population outcome, and it is a Bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
 
 <!-- 4. segway for next section -->
-In summary, extending Thurstone's general form allows researchers to address issues with the BTL model, such as accounting for judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2).
-However, it does not address concerns about stimuli' equal dispersions discussed in @sec-theory-issue1a.
-Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we develop in the next section.
-
-# Abandoning the BTL model {#sec-statistical}
+In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model.
+These extensions account for judge biases (see @sec-theory-issue1b), reflect the hierarchical structure of stimuli (see @sec-theory-issue1b), and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2).
+However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a.
+Because this issue involves statistical assumptions about the distribution of the discriminal process, it requires a statistical model, which we develop in the next section.
+
+# From SCMs to a statistical model  {#sec-statistical}
 
@@
 <!-- recording: Sven and Tine 25.03.31; time: 00:38:50 - 00:47:00 -->
-
-<!-- Deriving a statistical model to address violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a) becomes straightforward when using SCMs.
-Because the structural equations $F$ in the [SCM -@fig-cj15_scm] encode probabilistic information, the SCM can represent the joint distribution of the variables more efficiently [@Pearl_et_al_2016, pp.
-29].
-Specifically, it expresses the joint distribution as a product of conditional probability distributions (CPDs) of the type $P(child \cond parents)$.
-This property is formally known as the *Bayesian Network factorization* (BNF) [@Pearl_et_al_2016, pp.
-29; @Neal_2020, pp.
-21].
-Equation [-@eq-mat28] presents the joint distribution of the sample data and partitions it into smaller CPDs derived from the SCM structure.
--->
-
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!--   P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\ -->
-<!--   &= P(O_{R} \;|\; S, C, D_{R}) \cdot P(D_{R} \;|\; T_{IA}, B_{JK}) \cdot P(T_{IA} \;|\; T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \;|\; X_{I}, e_{I}) \\ -->
-<!--   & \quad \cdot P(B_{JK} \;|\; B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \;|\; Z_{J}, e_{J}) \cdot P(S) \cdot P(C) \cdot P(X_{IA}) \cdot P(X_{I}) \\  -->
-<!--   & \quad \cdot P(Z_{JK}) \cdot P(Z_{J}) \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J}) -->
-<!-- \end{aligned} -->
-<!-- $$ {#eq-mat28} -->
-
-<!-- 1. How to derive an stat model from an SCM? aboutn the outcome distribution -->
-Researchers can easily derive a statistical model to address violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a) using only [SCM -@fig-cj15_scm].
-This derivation is possible because the structural equations $F$ encode probabilistic information [@Pearl_et_al_2016], which researchers can replace with suitable probabilistic assumptions.
-As a starting point, given the dichotomous nature of CJ outcomes, [model -@fig-cj16_stat] assumes that $O_{R}$ follows a Bernoulli distribution[^6].
-
-<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
-This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
-12].
--->
-
-Next, following the conventions of Generalized Linear Latent And Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the model links the observed outcome to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
-Notice that the sampling and comparison mechanisms, $S$ and $C$, act as element-wise multipliers of $D_{R}$ (denoted by the symbol $\odot$), directly incorporating key assessment design features relevant to CJ experiments into the statistical formulation.
-Researchers can omit these vectors when only the observed CJ sample is available.
+<!-- recording: Sven and Tine 25.04.10; time: 00:00:00 - 00:01:10 -->
+<!-- recording: Sven and Tine 25.04.10; time: 00:01:10 - 00:01:35 -->
+<!-- recording: Sven and Tine 25.04.10; time: 00:01:35 - 00:16:30 -->
+<!-- recording: Sven and Tine 25.04.10; time: 00:19:00 - 00:26:40 -->
+<!-- recording: Sven and Tine 25.04.10; time: 00:52:00 - 00:53:40 -->
+
+<!-- 1. How to derive an stat model from an SCM? -->
+With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a).
+This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with appropriate functions and probabilistic assumptions [@Pearl_et_al_2016].
+Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]).
+For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent.
+Here, $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
+$$
+\begin{aligned}
+  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
+  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
+  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
+  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
+  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
+  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
+\end{aligned}
+$$ {#eq-mat51}
 
 <!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution.
-...
 For any model whose graph is acyclic, the joint distribution of the variables in the model is given by the product of the conditional distributions P(child|parents) over all the "families" in the graph.
@@
 
-[^6]: The binomial distribution--including its special case, the Bernoulli distribution-- represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
+<!-- The functions F = {fz, fx, fy} are called structural equations, and each function represents a causal process that assigns to its respective endogenous variable a value based on the values of the other variables.
+[@Cinelli_et_al_2020, pp.
+15] -->
+
+[^6]: This re-expression is possible because the *chain rule* of probability and the *Bayesian Network Factorization (BNF)* property. For further details, see @Pearl_et_al_2016 and @Neal_2020.
+
+<!-- We will denote the parents of a node X with pa(X).
+We’ll use an even simpler shorthand when the nodes are ordered so that we can denote the i^{th} node by X_{i}; in that case, we will also denote the parents of X_{i} by pa_{i} [@Neal_2020, pp.
+19] Definition 3.1 (Bayesian Network Factorization) Given a probability distribution P and a DAG G, P factorizes according to G if: P(x_{1}, \dots, x_{n}) = \prod_{i}P(x_{i} | pa_{i}).
+[@Neal_2020, pp.
+21] -->
+
+<!-- For any model whose graph is acyclic, the joint distribution of the variables in the model is given by the product of the conditional distributions P(child | parents) over all the "families" in the graph.
+Formally, we write this rule as P(x_{1}, \dots, x_{n}) = \prod_{i}P(x_{i} | pa_{i}) where pa_{i} stands for the values of the parents of variable X_{i}, and the product i runs over all i, from 1 to n.
+[@Pearl_et_al_2016, pp.
+29] -->
+
+<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
+Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline below.
+First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
+Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
+Note that the probability distributions for the sample and comparison mechanisms, $P(S)$ and $P(C)$, included in the probabilistic model [-@fig-cj16_prob], are omitted from the statistical model [-@fig-cj16_stat].
+We exclude these probabilities from the statistical definition because, although they are part of the joint distribution of the data, they do not influence the dependence between $O_{R}$ and $D_{R}$, since $S$ and $C$ are governed by random selection designs[^8].
+
+[^7]: The binomial distribution--including its special case, the Bernoulli distribution-- represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
 34].
@@
 
-::: {#fig-cj16 layout-ncol=2}
+[^8]: Randomization ensures that data—and, by extension, an estimator—satisfies several key identification properties, such as common support, no interference, and consistency.
+The most critical property, however, is the elimination of confounding.
+*Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $S$), resulting in spurious associations between the two [@Everitt_et_al_2010].
+Randomization addresses this issue by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself.
+For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.
+
+<!-- 2. Continuing with the assumptions -->
+Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
+The discriminal processes represent the underlying written-quality traits of the texts.
+Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
+$$
+\begin{aligned}
+D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
+&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
+& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
+\end{aligned}
+$$ {#eq-mat52}
+This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022].
+In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
+For clarity, the square brackets in $D_{R}$ only indicate the relevant indices within each trait vector; they do not represent any data subsetting.
+
+<!-- Weinberg[4] proposed a further diagram to capture ‘pure’ interaction in which neither G nor E has a direct effect on D and they only have an effect when present together (see Figure 4a).
+We suggest that this ‘pure’ interaction could be captured by the DAG shown in Figure 4b.
+An example of this is phenylketonuria in which phenylalanine in the diet (E) and the absence of the enzyme to break it down (phenylalanine hydroxylase, G) are both needed in order for a disease to manifest (intellectual disability, D).
+[@Attia_et_al_2022, pp.
+1052] -->
+
+<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder?
+-->
+
+<!-- Judges discrimination ability can be integrated in the models if we assume that the bias is not additive (as in Bramley explanation) but multiplicative.
+In the latter case the bias is not eliminated by taking just the difference.
+True, but it is false to think that because the biases are additive they get just eliminated.
+For instance, if we assume that the bias is additive but different for each stimuli, this means that the DAG should be: {T_{IA}, B_{JK}} -> O_{R}, but mathematically you can write it as it is above  -->
+
+::: {#fig-cj16 layout-ncol=3}
+
+::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O_{R} & := f_{O}(D_{R}, S, C) \\ 
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \dsep e_{JK} \\ \\
+\end{aligned}
+$$
+
+SCM
+:::
+
+::: {#fig-cj16_prob fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
+  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
+  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
+  & P( T_{I} \mid X_{I}, e_{I} ) \\
+  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
+  & P( B_{J} \mid Z_{J}, e_{J} ) \\
+  & P( e_{I} ) \\
+  & P( e_{J} ) \\
+  & P( e_{IA} ) \\  
+  & P( e_{JK} )
+\end{aligned}
+$$
+
+Probabilistic model
+:::
 
@@
 \begin{aligned}
-  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
+  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
   D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
@@
 
-::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O_{R} & := f_{O}(D_{R}, S, C) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \dsep e_{JK} 
-\end{aligned}
-$$
-
-[SCM -@fig-cj15_scm]
-:::
-
-Comparative judgment model, SCM and statistical model assuming different discriminal dispersions for the student's traits
-:::
-
-<!-- 2. Continuing with the probabilistic assumptions -->
-Then, [model -@fig-cj16_stat] defines $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, which represent their underlying written-quality trait.
-This difference also includes the corresponding repeated judge biases, $B_{JK}[j, k]$.
-Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, then we can re-write the discriminal difference as:
-$$
-\begin{aligned}
-D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] ] \\
-&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
-& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
-\end{aligned}
-$$ {#eq-mat51}
-This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022].
-In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
-For clarity, the square brackets in $D_{R}$ only indicate the relevant indices within each trait vector; they do not represent any data subsetting.
-
-<!-- Weinberg[4] proposed a further diagram to capture ‘pure’ interaction in which neither G nor E has a direct effect on D and they only have an effect when present together (see Figure 4a).
-We suggest that this ‘pure’ interaction could be captured by the DAG shown in Figure 4b.
-An example of this is phenylketonuria in which phenylalanine in the diet (E) and the absence of the enzyme to break it down (phenylalanine hydroxylase, G) are both needed in order for a disease to manifest (intellectual disability, D).
-[@Attia_et_al_2022, pp.
-1052] -->
-
-<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder?
--->
-
-Next, the model defines $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length), and the texts' idiosyncratic errors $e_{IA}$.
-The model assumes that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$, defined as a proportion of $1$.
-This standardization establishes the scale of the latent trait, as it is required by latent variable models [@Depaoli_2021].
+Comparative judgment model, SCM, probabilistic and statistical model assuming different discriminal dispersions for the student's traits
+:::
+
+Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$.
+We assume that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$.
+The parameter $p_{IA}$ is defined as a proportion of 1 to establish the scale of the latent trait, which is required in any latent variable model [@Depaoli_2021].
+Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed.
+This result indicates that our model reverts to the original assumptions for the discriminal processes of stimuli proposed by @Thurstone_1927b (see @tbl-thurstone_cases).
 
@@
 
-Additionally, the model defines $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$.
-As before, each element of $e_{I}$ is $iid$ following a Normal distribution with a mean of zero.
-However, the standard deviation, $s_{XI}$, varies depending on the teaching method to which each student belongs.
-Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$  This constraint effectively addresses concerns about the equal dispersion assumption raised in @sec-theory-issue1a.
+Fourth, we define $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$.
+As before, we assume that each element of $e_{I}$ is $iid$ Normal with a mean of zero.
+However, we consider the standard deviation, $s_{XI}$, varies depending on the teaching method group to which each student belongs.
+Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$.
+Note that this parametrization of $e_{I}$ relaxes the equal dispersion assumption for the stimuli, effectively addressing the concerns raised in @sec-theory-issue1a.
+
+<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
+This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
+12].
+-->
 
@@
 
-Similarly, the model defines $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., the number of judgments a judge makes), and judgment-specific idiosyncratic errors $e_{JK}$.
-The model assumes that each element of $e_{JK}$ is $iid$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].
-
-The model also defines $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$.
+Fifth, we define $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments a judge makes affect the assessment of quality), and judgment-specific idiosyncratic errors $e_{JK}$.
+We assume that each element of $e_{JK}$ is $iid$ Normal with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].
+
+Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$.
 Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs.
@@
 
-Finally, we apply *Bayesian inference methods* to turn [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data.
-The **Declarations** section of this document provides a link to this model, along with an alternative specification that assumes equal discriminal dispersions.
-We implemented both models in `Stan` [version $2.26.1$, @Stan_2020].
+Finally, to transform [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data, we use *Bayesian inference methods*.
+The **Declarations** section of this document includes a link to the model, along with an alternative specification that assumes equal discriminal dispersions.
+We implemented both versions of the model using `Stan` [version $2.26.1$, @Stan_2020].
+Although the implementation was successful, we do not report the results here, as they fall outside the scope of this document.
+
+<!-- Proposal: Make the model to be bX %*% X 
+Decision: Although this makes it easier for the software, the way it is declared is more explicit. -->
+
+<!-- Proposal: Make e_{I} and e_{J} fully orthogonal, a\
+Decision: Although I can see the benefit of this, it would add several layers of explanations to the document or the code for that matter. I think this paper has already a lot of info. -->
+
+<!-- Proposal: Add new way of declaring non-centered parametrization and optional variables (See section 'Affinely transformed real' and 'Declaring optional variables' sections in https://mc-stan.org/docs/reference-manual/types.html)
+Decision: interesting, but I think I like the model as it is, explicit! -->
 
@@
 
-## Findings {#sec-discussion-finding}
-
 <!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
 <!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:56:10 -->
+<!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->
 
@@
 
+<!-- recording: Sven and Tine 25.04.10; time: 00:12:00 - 00:14:00 -->
+<!-- Consider how to use the S and C, when the designs are more complexes. -->
+<!-- Normally, S and C are not typically mentioned  -->
+
 <!-- Statements about the real world from observed data require assumptions.
@@
 $$
-E[Y_{i} \cond T_{i}=1] - E[Y_{i} \cond T_{i}=2]
+E[Y_{i} \mid T_{i}=1] - E[Y_{i} \mid T_{i}=2]
 $$ {#eq-group_diff}
@@
 <!-- If both assumptions are met, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
-The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \cond T_{i}=1]$ and $E[Y_{i} \cond T_{i}=2]$, respectively.
+The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \mid T_{i}=1]$ and $E[Y_{i} \mid T_{i}=2]$, respectively.
 The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.
@@
 $$
-\tau_{i} = Y_{i} \cond do(T_{i}=1) - Y_{i} \cond do(T_{i}=2)
+\tau_{i} = Y_{i} \mid do(T_{i}=1) - Y_{i} \mid do(T_{i}=2)
 $$ {#eq-ICE}
 
-where $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \cond do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \cond do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
+where $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \mid do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \mid do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
 Here, an *intervention* involves assigning a constant value to the treatment variable for each student's potential outcomes.
@@
 \tau & = E[\tau_{i}] \\
-  & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)]
+  & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)]
 \end{aligned}
@@
 This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones.
-When such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
+When such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \mid X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
 Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
@@
 $$
-CACE_{t} = E[Y_{i} \cond T_{i}=t, X]
+CACE_{t} = E[Y_{i} \mid T_{i}=t, X]
 $$ {#eq-CACE}
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)] \\
+  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
   & = E_{X}[CACE_{1} - CACE_{2}] \\
-  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
+  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right]
 \end{aligned}
@@
 \begin{aligned}
-  \tau &= E[Y_{i} \cond do(T_{i}=t)] \\
-  & = d E_{X}\left[ E[Y_{i} \cond T_{i}=t, X]\right]/ dt
+  \tau &= E[Y_{i} \mid do(T_{i}=t)] \\
+  & = d E_{X}\left[ E[Y_{i} \mid T_{i}=t, X]\right]/ dt
   \end{aligned}
@@
 29].
-This expression takes the form of a product of conditional probability distributions (CPDs) of the type $P(child \cond parents)$.
+This expression takes the form of a product of conditional probability distributions (CPDs) of the type $P(child \mid parents)$.
 This property is formally known as the *Bayesian Network factorization* (BNF, @eq-net_factor) [@Pearl_et_al_2016, pp.
@@
 \begin{aligned}
-P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \cond X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
-& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \cond pa(X_{i}) ) & (\small{\text{by BNF}})
+P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
+& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}) ) & (\small{\text{by BNF}})
 \end{aligned}
@@
 24].
-Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$.
+Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \mid X_{1}) P(X_{1})$.
 Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendixB31 and the CPDs.
@@
 36].
-For instance, applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] allow researchers to represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}) P(X_{3} \cond X_{2})$.
-This expression implies that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
-Moreover, in the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \cond X_{2}) P(X_{2}) P(X_{3} \cond X_{2})$.
-Similar to the chain structure, this expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$.
-Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}, X_{3}) P(X_{3})$.
-This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent $(X_{1} \dsep X_{3})$, but conditionally dependent when controlling for $X_{2}$ $(X_{1} \ndsep X_{3} \cond X_{2})$.
+For instance, applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] allow researchers to represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \mid X_{1}) P(X_{3} \mid X_{2})$.
+This expression implies that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \mid X_{2})$.
+Moreover, in the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \mid X_{2}) P(X_{2}) P(X_{3} \mid X_{2})$.
+Similar to the chain structure, this expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \mid X_{2})$.
+Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \mid X_{1}, X_{3}) P(X_{3})$.
+This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent $(X_{1} \dsep X_{3})$, but conditionally dependent when controlling for $X_{2}$ $(X_{1} \ndsep X_{3} \mid X_{2})$.
 The authors @Pearl_et_al_2016 [pp.
@@
 Using these additional probabilistic insights, researchers can revisit the scenario in @sec-appendixB2.
-In this context, applying the BNF to the SCM [-@fig-scm_example2] structure, enables the representation of the joint probability distribution of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$.
+In this context, applying the BNF to the SCM [-@fig-scm_example2] structure, enables the representation of the joint probability distribution of the observed variables as $P(Y, T, X) =$ $P(Y \mid T, X) P(T \mid X) P(X)$.
 From this expression, researchers can infer that the outcome $Y$ is unconditionally dependent on the teaching method $T$ $(Y \ndsep T)$.
@@
 <!-- 2. additional necessary concepts  -->
-The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions $P(Y_{i} \cond do(T_{i}=t))$ within SCMs and DAGs.
+The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions $P(Y_{i} \mid do(T_{i}=t))$ within SCMs and DAGs.
 *Manipulated graphs* simulate physical interventions by removing specific edges from a DAG, while preserving the remaining structure unchanged [@Neal_2020, pp.
@@
 $$
-P(X_{1}, X_{2}, \dots, X_{P} \cond do(S)) =
+P(X_{1}, X_{2}, \dots, X_{P} \mid do(S)) =
 \begin{cases}
-  \prod P(X_{p} \cond pa(X_{p}) ) & \text{if} \: p \not\in S \\
+  \prod P(X_{p} \mid pa(X_{p}) ) & \text{if} \: p \not\in S \\
   1 \quad & \text{otherwise}
@@
 The sufficient adjustment set (potentially empty) must block all non-causal paths between $X_{a}$ and $X_{b}$ without introducing new paths.
-If such a set exists, then $X_{a}$ and $X_{b}$ are *d-separated* by $X_{p}$ ($X_{a} \dsep X_{b} \cond X_{p}$) [@Pearl_2009], and $X_{p}$ satisfies the *backdoor criterion* [@Neal_2020, pp.
+If such a set exists, then $X_{a}$ and $X_{b}$ are *d-separated* by $X_{p}$ ($X_{a} \dsep X_{b} \mid X_{p}$) [@Pearl_2009], and $X_{p}$ satisfies the *backdoor criterion* [@Neal_2020, pp.
 37].
@@
 $$
-P(X_{a} \cond do(X_{b}=x)) = \sum_{Xp} P(X_{a} \cond X_{b}=x, X_{p}) P(X_{p})
+P(X_{a} \mid do(X_{b}=x)) = \sum_{Xp} P(X_{a} \mid X_{b}=x, X_{p}) P(X_{p})
 $$ {#eq-backdoor}
@@
 \begin{aligned}
-\tau & = E[X_{a} \cond do(X_{b}=1)]- E[X_{a} \cond do(X_{b}=2)] \\
-  & = E_{Xp}\left[ E[X_{a} \cond do(X_{b}=1), X_{p}]- E[X_{a} \cond do(X_{b}=2), X_{p}] \right] \\
-  & = \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=2, X_{p}) \cdot P(X_{p})
+\tau & = E[X_{a} \mid do(X_{b}=1)]- E[X_{a} \mid do(X_{b}=2)] \\
+  & = E_{Xp}\left[ E[X_{a} \mid do(X_{b}=1), X_{p}]- E[X_{a} \mid do(X_{b}=2), X_{p}] \right] \\
+  & = \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=2, X_{p}) \cdot P(X_{p})
 \end{aligned}
@@
 This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure.
-Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.
+Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \mid X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.
 Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right] \\
-  & = \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \cond T_{i}=2, X) \cdot P(X)
+  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right] \\
+  & = \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=2, X) \cdot P(X)
 \end{aligned}
@@
 Ultimately, researchers can use Bayesian inference methods to estimate the ACE.
-The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \cond \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
+The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \mid \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
 After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule in @eq-bayes_rule [@Jeffreys_1998]:
@@
 $$
-P(\theta \cond X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \cond \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
+P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
 $$ {#eq-bayes_rule}
@@
 $$
-P(\theta \cond X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\cond \theta) \cdot P(\theta)
+P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\mid \theta) \cdot P(\theta)
 $$ {#eq-prop_rule}
@@
 $$
-P(X_{1}, X_{2}, \dots, X_{P} \cond \theta) = P(X_{1} \cond \theta) \prod^{P}_{p=2} P(X_{i} \cond X_{i-1}, \dots, X_{1}, \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}, \theta )
 $$ {#eq-like_chain}
@@
 $$
-P(X_{1}, X_{2}, \dots, X_{P} \cond \theta) = P(X_{1} \cond \theta) \prod^{P}_{p=2} P(X_{i} \cond pa(X_{i}), \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}), \theta )
 $$ {#eq-like_BNF}

### 2025-04-11 code/CJfig.R

@@ -44,16 +44,14 @@ sim_me = function( n=100, b=0.3, s1=1, s2=0.8){
 }
 
 
+
+# figure 1 ####
+
 # distributional parameters
 dpA = c(0, 0.5) # distribution text A (mean, sd)
 dpB = c(1, 1) # distribution text B (mean, sd)
 rAB = 0 # correlation between discriminal processes
 
-
-
-
-
-# figure 1 ####
 
 # figure
 png( filename=file.path(main_dir, 'discriminal_process.png'), 

### 2025-04-11 _tex/paper2.tex
Built site for gh-pages

@@
 variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
-functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
+functions \(F\)
+\citep{Pearl_2009, Pearl_et_al_2016, Cinelli_et_al_2020}. Endogenous
 variables are those whose causal mechanisms a researcher chooses to
@@
 \(c_{(i,a,h,b,j,k)}\) being a binary value indicating the presence or
-absence of data rows in \(R\), a definition similar to that of \(S\) in
-equation (\ref{eq-mat26}).
+absence of data rows in \(R\), a definition similar to that of
+\(s_{(i,a,h,b,j,k)}\) in equation (\ref{eq-mat26}).
 \begin{equation}\phantomsection\label{eq-mat27}{
@@
 
-The DAG~\ref{fig-cj14_dag} also integrates the comparison mechanism
-\(C\) into the conceptual-population model, which ultimately derives the
-final comparative outcome \(O_{R}\). First, it shows the
-sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
-that researchers cannot directly access this outcome due to the
-comparison mechanism. Second, it depicts the \emph{comparison mechanism}
-\(C\) as a causal factor influencing the final comparative outcome
-\(O_{R}\). As with the sampling mechanism, a square encloses \(C\) to
-indicate that it is a conditioned variable--that is, one that determines
-\emph{which repeated judgments the subset of judges makes for the subset
-of stimuli produced by the sampled individuals}. In essence, \(C\)
-reflects the assumption that judges \emph{do not} perform all possible
-repeated judgments for the selected stimuli. Instead, they carry out a
-``sufficient number of comparisons'' \(n_{C}\) to allow for accurate
-estimation of the proportion \(P(B>A)\) for each stimulus pair
+The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
+mechanism} \(C\) into the conceptual-population model, ultimately
+deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
+\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
+directly access this outcome due to the comparison mechanism. Moreover,
+it depicts \(C\) as a causal factor influencing \(O_{R}\). As with the
+sampling mechanism, a square encloses \(C\) to indicate that it is a
+conditioned variable--that is, one that determines \emph{which repeated
+judgments the subset of judges makes for the subset of stimuli produced
+by the sampled individuals}. In essence, \(C\) reflects the assumption
+that judges \emph{do not} perform all possible repeated judgments for
+the selected stimuli. Instead, they carry out a ``sufficient number of
+comparisons'' \(n_{C}\) to allow for accurate estimation of the
+proportion \(P(B>A)\) for each stimulus pair
 \citep[pp.~267]{Thurstone_1927b}.
 
-Notably, DAG~\ref{fig-cj14_dag} shows that \(C\) is independent of all
-other variables in the model. This independence implies that the
+Notably, DAG~\ref{fig-cj14_dag} also shows that \(C\) is independent of
+all other variables in the model. This independence implies that the
 conceptual model represented by the DAG applies exclusively to Random
@@
 
-Finally, without loss of generality, researchers can reformulate the
-sample-comparison model in Figure~\ref{fig-cj14} into the equivalent
-form shown in Figure~\ref{fig-cj15}. This reformulation rests on two
-principles. First, because the sampling and comparison mechanism vectors
-\(S\) and \(C\) are binary, researchers can integrate them into the
-model using element-wise products. This approach eliminates the need to
-define the structural equations \(f_{S}\) and \(f_{C}\). Second, since
-only the final comparative outcome \(O_{R}\) is observed, it is standard
-to assume that the distribution of the conceptual-population outcome
-\(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
-applies to this outcome. This assumption removes the necessity of
-describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).
-
-In summary, extending Thurstone's general form allows researchers to
-address issues with the BTL model, such as accounting for judge biases
-(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
-stimuli (see Section~\ref{sec-theory-issue1b}), and measurement error in
-trait estimation and hypothesis testing (see
-Section~\ref{sec-theory-issue2}). However, it does not address concerns
-about stimuli' equal dispersions discussed in
-Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
-assumptions regarding the discriminal process distribution, it requires
-a statistical model, which we develop in the next section.
-
-\section{Abandoning the BTL model}\label{sec-statistical}
-
-Researchers can easily derive a statistical model to address violations
-of the equal dispersion assumption in CJ experiments (see
-Section~\ref{sec-theory-issue1a}) using only SCM~\ref{fig-cj15_scm}.
-This derivation is possible because the structural equations \(F\)
-encode probabilistic information \citep{Pearl_et_al_2016}, which
-researchers can replace with suitable probabilistic assumptions. As a
-starting point, given the dichotomous nature of CJ outcomes,
-model~\ref{fig-cj16_stat} assumes that \(O_{R}\) follows a Bernoulli
-distribution\footnote{The binomial distribution--including its special
-  case, the Bernoulli distribution-- represent a maximum entropy
+Finally, since it is standard to assume that the distribution of the
+conceptual-population outcome \(O^{cp}_{R}\) also holds for
+\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the sample-comparison
+model in Figure~\ref{fig-cj14} into the equivalent form shown in
+Figure~\ref{fig-cj15}. This reformulation produces a model that applies
+directly to a sample of comparative data. In this version, the
+unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\) are omitted, and
+\(O_{R}\) inherits the structural equation \(f_{O}\) that originally
+defined \(O^{cp}_{R}\). Moreover, the definition of \(O_{R}\) now
+reflects its direct dependence on the discriminal difference \(D_{R}\)
+and the sample and comparison mechanisms, \(S\) and \(C\).
+
+In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
+Thurstone's general form to address several limitations of the BTL
+model. These extensions account for judge biases (see
+Section~\ref{sec-theory-issue1b}), reflect the hierarchical structure of
+stimuli (see Section~\ref{sec-theory-issue1b}), and incorporate
+measurement error in trait estimation and hypothesis testing (see
+Section~\ref{sec-theory-issue2}). However, they do not resolve concerns
+about the assumption of equal dispersions among stimuli discussed in
+Section~\ref{sec-theory-issue1a}. Because this issue involves
+statistical assumptions about the distribution of the discriminal
+process, it requires a statistical model, which we develop in the next
+section.
+
+\section{From SCMs to a statistical model}\label{sec-statistical}
+
+With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
+statistical model that addresses violations of the equal dispersion
+assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
+This derivation is possible because a fully specified SCM encodes
+functional and probabilistic information, which we can replace with
+appropriate functions and probabilistic assumptions
+\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
+to express the joint distribution of our CJ system as a product of
+simpler conditional probability distributions (CPDs)\footnote{This
+  re-expression is possible because the \emph{chain rule} of probability
+  and the \emph{Bayesian Network Factorization (BNF)} property. For
+  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
+as shown in equation (\ref{eq-mat51}). For clarity, we treat expressions
+such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
+as equivalent. Here, \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
+CPD of \(Y\) given \(X\).
+\begin{equation}\phantomsection\label{eq-mat51}{
+\begin{aligned}
+  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
+  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
+  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
+  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
+  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
+  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
+\end{aligned}
+}\end{equation}
+
+Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
+which we outline below. First, we assume that \(O_{R}\) follows a
+Bernoulli distribution\footnote{The binomial distribution--including its
+  special case, the Bernoulli distribution-- represent a maximum entropy
   distribution for binary events \citep[pp.~34]{McElreath_2020}. This
@@
   discussion of the binomial as a maximum entropy distribution, see
-  \citet[chap.~10.1.2]{McElreath_2020}.}.
-
-Next, following the conventions of Generalized Linear Latent And Mixed
-Models (GLLAMMs)
+  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
+of CJ outcomes. Furthermore, following the conventions of Generalized
+Linear Latent and Mixed Models (GLLAMMs)
 \citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
-model links the observed outcome to the latent discriminal difference
-vector \(D_{R}\) using an inverse-logit function:
-\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Notice that the sampling
-and comparison mechanisms, \(S\) and \(C\), act as element-wise
-multipliers of \(D_{R}\) (denoted by the symbol \(\odot\)), directly
-incorporating key assessment design features relevant to CJ experiments
-into the statistical formulation. Researchers can omit these vectors
-when only the observed CJ sample is available.
+distribution links \(O_{R}\) to the latent discriminal difference vector
+\(D_{R}\) using an inverse-logit function:
+\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that the probability
+distributions for the sample and comparison mechanisms, \(P(S)\) and
+\(P(C)\), included in the probabilistic model \ref{fig-cj16_prob}, are
+omitted from the statistical model \ref{fig-cj16_stat}. We exclude these
+probabilities from the statistical definition because, although they are
+part of the joint distribution of the data, they do not influence the
+dependence between \(O_{R}\) and \(D_{R}\), since \(S\) and \(C\) are
+governed by random selection designs\footnote{Randomization ensures that
+  data---and, by extension, an estimator---satisfies several key
+  identification properties, such as common support, no interference,
+  and consistency. The most critical property, however, is the
+  elimination of confounding. \emph{Confounding} occurs when an external
+  variable, such as \(X_{I}\), simultaneously influences both the
+  outcome (e.g., \(O_{R}\)) and a variable of interest (e.g., \(S\)),
+  resulting in spurious associations between the two
+  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
+  effectively decoupling the association between the variable of
+  interest and any other variable, except for the outcome itself. For a
+  more detailed discussion on the benefits of randomization, see
+  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
+  \citet{Hernan_et_al_2020}.}.
+
+Second, we define \(D_{R}\) as the difference between the discriminal
+processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), plus the
+corresponding repeated judge biases, \(B_{JK}[j, k]\). The discriminal
+processes represent the underlying written-quality traits of the texts.
+Note that if we assume that \(B_{JK}[j,k]\) reflects the difference in
+stimulus-specific biases, i.e.,
+\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
+discriminal difference as:
+\begin{equation}\phantomsection\label{eq-mat52}{
+\begin{aligned}
+D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
+&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
+& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
+\end{aligned}
+}\end{equation} This formulation reveals that the discriminal difference
+captures a \emph{pure interaction effect}, in which neither the texts'
+discriminal processes nor the judges' biases alone determine the
+outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
+terms, the discriminal processes of the stimuli become an observable
+outcome only through the judges' perceptions (biases). For clarity, the
+square brackets in \(D_{R}\) only indicate the relevant indices within
+each trait vector; they do not represent any data subsetting.
 
@@
 \begin{aligned}
-  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
+  O_{R} & := f_{O}(D_{R}, S, C) \\ 
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} \\ \\
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj16_scm}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
+  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
+  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
+  & P( T_{I} \mid X_{I}, e_{I} ) \\
+  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
+  & P( B_{J} \mid Z_{J}, e_{J} ) \\
+  & P( e_{I} ) \\
+  & P( e_{J} ) \\
+  & P( e_{IA} ) \\  
+  & P( e_{JK} )
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj16_prob}Probabilistic model}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
   D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
@@
 \end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O_{R} & := f_{O}(D_{R}, S, C) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} 
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj16_scm}SCM~\ref{fig-cj15_scm}}
-
-\end{minipage}%
-
-\caption{\label{fig-cj16}Comparative judgment model, SCM and statistical
-model assuming different discriminal dispersions for the student's
-traits}
+
+\caption{\label{fig-cj16}Comparative judgment model, SCM, probabilistic
+and statistical model assuming different discriminal dispersions for the
+student's traits}
 
@@
 
-Then, model~\ref{fig-cj16_stat} defines \(D_{R}\) as the difference
-between the discriminal processes of two texts, \(T_{IA}[i, a]\) and
-\(T_{IA}[h, b]\), which represent their underlying written-quality
-trait. This difference also includes the corresponding repeated judge
-biases, \(B_{JK}[j, k]\). Note that if we assume that \(B_{JK}[j,k]\)
-reflects the difference in stimulus-specific biases, i.e.,
-\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), then we can
-re-write the discriminal difference as:
-\begin{equation}\phantomsection\label{eq-mat51}{
-\begin{aligned}
-D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] ] \\
-&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
-& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
-\end{aligned}
-}\end{equation} This formulation reveals that the discriminal difference
-captures a \emph{pure interaction effect}, in which neither the texts'
-discriminal processes nor the judges' biases alone determine the
-outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
-terms, the discriminal processes of the stimuli become an observable
-outcome only through the judges' perceptions (biases). For clarity, the
-square brackets in \(D_{R}\) only indicate the relevant indices within
-each trait vector; they do not represent any data subsetting.
-
-Next, the model defines \(T_{IA}\) as a linear combination of the
-students' underlying writing-quality traits \(T_{I}\), the effects of
-relevant text-related variables \(\beta_{XA}X_{IA}\) (e.g., text
-length), and the texts' idiosyncratic errors \(e_{IA}\). The model
-assumes that each element of \(e_{IA}\) is independent and identically
-distributed \((iid)\) according to a Normal distribution, with a mean of
-zero and a standard deviation \(p_{IA}\), defined as a proportion of
-\(1\). This standardization establishes the scale of the latent trait,
-as it is required by latent variable models \citep{Depaoli_2021}.
-
-Additionally, the model defines \(T_{I}\) as a linear combination of the
-effects of relevant student-related variables, \(\beta_{XI} X_{I}\), and
-students' idiosyncratic errors, \(e_{I}\). As before, each element of
-\(e_{I}\) is \(iid\) following a Normal distribution with a mean of
-zero. However, the standard deviation, \(s_{XI}\), varies depending on
-the teaching method to which each student belongs. Given the example in
+Third, we define \(T_{IA}\) as a linear combination of the students'
+underlying writing-quality traits \(T_{I}\), the effects of relevant
+text-related variables \(\beta_{XA}X_{IA}\) (e.g., text length on
+quality), and the texts' idiosyncratic errors \(e_{IA}\). We assume that
+each element of \(e_{IA}\) is independent and identically distributed
+\((iid)\) according to a Normal distribution, with a mean of zero and a
+standard deviation \(p_{IA}\). The parameter \(p_{IA}\) is defined as a
+proportion of 1 to establish the scale of the latent trait, which is
+required in any latent variable model \citep{Depaoli_2021}. Note that
+because \(e_{IA}\) follows a Normal distribution, \(T_{IA}\) is also
+normally distributed. This result indicates that our model reverts to
+the original assumptions for the discriminal processes of stimuli
+proposed by \citet{Thurstone_1927b} (see
+Table~\ref{tbl-thurstone_cases}).
+
+Fourth, we define \(T_{I}\) as a linear combination of the effects of
+relevant student-related variables, \(\beta_{XI} X_{I}\), and students'
+idiosyncratic errors, \(e_{I}\). As before, we assume that each element
+of \(e_{I}\) is \(iid\) Normal with a mean of zero. However, we consider
+the standard deviation, \(s_{XI}\), varies depending on the teaching
+method group to which each student belongs. Given the example in
 Section~\ref{sec-theoretical}, where the teaching method
@@
 scale of the latent trait \citep{Depaoli_2021}:
-\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\) This constraint effectively addresses
-concerns about the equal dispersion assumption raised in
+\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\). Note that this parametrization of
+\(e_{I}\) relaxes the equal dispersion assumption for the stimuli,
+effectively addressing the concerns raised in
 Section~\ref{sec-theory-issue1a}.
 
-Similarly, the model defines \(B_{JK}\) as a linear combination of the
-judges' individual bias \(B_{J}\), the effects of relevant
-judgment-related variables \(\beta_{ZK}Z_{JK}\) (e.g., the number of
-judgments a judge makes), and judgment-specific idiosyncratic errors
-\(e_{JK}\). The model assumes that each element of \(e_{JK}\) is \(iid\)
-according to a Normal distribution, with a mean of zero and a standard
+Fifth, we define \(B_{JK}\) as a linear combination of the judges'
+individual bias \(B_{J}\), the effects of relevant judgment-related
+variables \(\beta_{ZK}Z_{JK}\) (e.g., how the number of judgments a
+judge makes affect the assessment of quality), and judgment-specific
+idiosyncratic errors \(e_{JK}\). We assume that each element of
+\(e_{JK}\) is \(iid\) Normal with a mean of zero and a standard
 deviation \(p_{JK}\), defined as a proportion of \(1\), to anchor the
@@
 
-The model also defines \(B_{J}\) as a linear combination of the effects
-of relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
+Sixth, we define \(B_{J}\) as a linear combination of the effects of
+relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
 expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
@@
 
-Finally, we apply \emph{Bayesian inference methods} to turn
-model~\ref{fig-cj16_stat} into a practical statistical tool for
-analyzing paired comparison data. The \textbf{Declarations} section of
-this document provides a link to this model, along with an alternative
+Finally, to transform model~\ref{fig-cj16_stat} into a practical
+statistical tool for analyzing paired comparison data, we use
+\emph{Bayesian inference methods}. The \textbf{Declarations} section of
+this document includes a link to the model, along with an alternative
 specification that assumes equal discriminal dispersions. We implemented
-both models in \texttt{Stan} \citep[version \(2.26.1\),][]{Stan_2020}.
+both versions of the model using \texttt{Stan} \citep[version
+\(2.26.1\),][]{Stan_2020}. Although the implementation was successful,
+we do not report the results here, as they fall outside the scope of
+this document.
 
 \section{Discussion}\label{sec-discussion}
-
-\subsection{Findings}\label{sec-discussion-finding}
 
@@
 \begin{equation}\phantomsection\label{eq-group_diff}{
-E[Y_{i} \:|\:T_{i}=1] - E[Y_{i} \:|\:T_{i}=2]
+E[Y_{i} \mid T_{i}=1] - E[Y_{i} \mid T_{i}=2]
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-ICE}{
-\tau_{i} = Y_{i} \:|\:do(T_{i}=1) - Y_{i} \:|\:do(T_{i}=2)
+\tau_{i} = Y_{i} \mid do(T_{i}=1) - Y_{i} \mid do(T_{i}=2)
 }\end{equation}
@@
 where \(do(T_{i}=t)\) represents the intervention operator,
-\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
-intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
+\(Y_{i} \mid do(T_{i}=1)\) represents the potential outcome under
+intervention \(T_{i}=1\), and \(Y_{i} \mid do(T_{i}=1)\) represents the
 potential outcome under intervention \(T_{i}=2\). Here, an
@@
 \tau & = E[\tau_{i}] \\
-  & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
+  & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)]
 \end{aligned}
@@
 exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
-(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
+(\(T \:\bot\:Y \mid X\)) \citep{Pearl_2009}, and \(X\) satisfies the
 \emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
@@
 \begin{equation}\phantomsection\label{eq-CACE}{
-CACE_{t} = E[Y_{i} \:|\:T_{i}=t, X]
+CACE_{t} = E[Y_{i} \mid T_{i}=t, X]
 }\end{equation}
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
+  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
   & = E_{X}[CACE_{1} - CACE_{2}] \\
-  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
+  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right]
 \end{aligned}
@@
 \begin{aligned}
-  \tau &= E[Y_{i} \:|\:do(T_{i}=t)] \\
-  & = d E_{X}\left[ E[Y_{i} \:|\:T_{i}=t, X]\right]/ dt
+  \tau &= E[Y_{i} \mid do(T_{i}=t)] \\
+  & = d E_{X}\left[ E[Y_{i} \mid T_{i}=t, X]\right]/ dt
   \end{aligned}
@@
 product of conditional probability distributions (CPDs) of the type
-\(P(child \:|\:parents)\). This property is formally known as the
+\(P(child \mid parents)\). This property is formally known as the
 \emph{Bayesian Network factorization} (BNF,
@@
 \begin{aligned}
-P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
-& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:pa(X_{i}) ) & (\small{\text{by BNF}})
+P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
+& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}) ) & (\small{\text{by BNF}})
 \end{aligned}
@@
 and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
-distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
+distribution as \(P(X_{1}, X_{3}) = P(X_{3} \mid X_{1}) P(X_{1})\).
 Notably, these descriptions demonstrate the clear correspondence between
@@
 \(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
+\(P(X_{1}) P(X_{2} \mid X_{1}) P(X_{3} \mid X_{2})\). This expression
 implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
 \((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
-controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\).
+controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \mid X_{2})\).
 Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
@@
 of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). Similar to the
+\(P(X_{1} \mid X_{2}) P(X_{2}) P(X_{3} \mid X_{2})\). Similar to the
 chain structure, this expression allows researchers to further infer
@@
 \((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
-controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally,
+controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \mid X_{2})\). Finally,
 researchers analyzing the collider structure illustrated in the SCM
@@
 distribution of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
+\(P(X_{1}) P(X_{2} \mid X_{1}, X_{3}) P(X_{3})\). This representation
 allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
@@
 dependent when controlling for \(X_{2}\)
-\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
+\((X_{1} \:\not\bot\:X_{3} \mid X_{2})\). The authors \citet[pp.~37, 40,
 41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
@@
 representation of the joint probability distribution of the observed
-variables as \(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From
+variables as \(P(Y, T, X) =\) \(P(Y \mid T, X) P(T \mid X) P(X)\). From
 this expression, researchers can infer that the outcome \(Y\) is
@@
 and Truncated Factorization, which are essential for representing
-interventions \(P(Y_{i} \:|\:do(T_{i}=t))\) within SCMs and DAGs.
+interventions \(P(Y_{i} \mid do(T_{i}=t))\) within SCMs and DAGs.
 \emph{Manipulated graphs} simulate physical interventions by removing
@@
 \begin{equation}\phantomsection\label{eq-trunc_factor}{
-P(X_{1}, X_{2}, \dots, X_{P} \:|\:do(S)) =
+P(X_{1}, X_{2}, \dots, X_{P} \mid do(S)) =
 \begin{cases}
-  \prod P(X_{p} \:|\:pa(X_{p}) ) & \text{if} \: p \not\in S \\
+  \prod P(X_{p} \mid pa(X_{p}) ) & \text{if} \: p \not\in S \\
   1 \quad & \text{otherwise}
@@
 \(X_{b}\) are \emph{d-separated} by \(X_{p}\)
-(\(X_{a} \:\bot\:X_{b} \:|\:X_{p}\)) \citep{Pearl_2009}, and \(X_{p}\)
+(\(X_{a} \:\bot\:X_{b} \mid X_{p}\)) \citep{Pearl_2009}, and \(X_{p}\)
 satisfies the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020}.
@@
 \begin{equation}\phantomsection\label{eq-backdoor}{
-P(X_{a} \:|\:do(X_{b}=x)) = \sum_{Xp} P(X_{a} \:|\:X_{b}=x, X_{p}) P(X_{p})
+P(X_{a} \mid do(X_{b}=x)) = \sum_{Xp} P(X_{a} \mid X_{b}=x, X_{p}) P(X_{p})
 }\end{equation}
@@
 \begin{aligned}
-\tau & = E[X_{a} \:|\:do(X_{b}=1)]- E[X_{a} \:|\:do(X_{b}=2)] \\
-  & = E_{Xp}\left[ E[X_{a} \:|\:do(X_{b}=1), X_{p}]- E[X_{a} \:|\:do(X_{b}=2), X_{p}] \right] \\
-  & = \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \:|\:X_{b}=2, X_{p}) \cdot P(X_{p})
+\tau & = E[X_{a} \mid do(X_{b}=1)]- E[X_{a} \mid do(X_{b}=2)] \\
+  & = E_{Xp}\left[ E[X_{a} \mid do(X_{b}=1), X_{p}]- E[X_{a} \mid do(X_{b}=2), X_{p}] \right] \\
+  & = \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=2, X_{p}) \cdot P(X_{p})
 \end{aligned}
@@
 \(Y\) remains dependent on \(T\) even after conditioning
-\((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
+\((Y \:\not\bot\:T \mid X)\), this dependency can only be attributed to
 the direct causal effect \(T \rightarrow Y\). Notably, for the purpose
@@
 \begin{aligned}
-  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right] \\
-  & = \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \:|\:T_{i}=2, X) \cdot P(X)
+  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
+  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right] \\
+  & = \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=2, X) \cdot P(X)
 \end{aligned}
@@
 the likelihood of the data,
-\(P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta)\), and the prior
+\(P(X_{1}, X_{2}, \dots, X_{P} \mid \theta)\), and the prior
 distribution, \(P(\theta)\) \citep{Everitt_et_al_2010}, where \(X_{P}\)
@@
 \begin{equation}\phantomsection\label{eq-bayes_rule}{
-P(\theta \:|\:X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
+P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-prop_rule}{
-P(\theta \:|\:X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\:|\:\theta) \cdot P(\theta)
+P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\mid \theta) \cdot P(\theta)
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-like_chain}{
-P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta) = P(X_{1} \:|\:\theta) \prod^{P}_{p=2} P(X_{i} \:|\:X_{i-1}, \dots, X_{1}, \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}, \theta )
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-like_BNF}{
-P(X_{1}, X_{2}, \dots, X_{P} \:|\:\theta) = P(X_{1} \:|\:\theta) \prod^{P}_{p=2} P(X_{i} \:|\:pa(X_{i}), \theta )
+P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}), \theta )
 }\end{equation}

### 2025-04-11 paper2.qmd
section:5; process:write

@@
 <!-- 3. segway for next section -->
-However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals--due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020]--a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by certain individuals.
 
@@
 The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$.
-It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism.
-Moreover, it depicts $C$ as a causal factor influencing $O_{R}$.
-As with the sampling mechanism, a square encloses $C$ to indicate that it is a conditioned variable--that is, one that determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism.
+Moreover, the DAG depicts $C$ as a causal factor influencing $O_{R}$, enclosing it in a square to indicate that it is a conditioned variable.
+Specifically, $C$ determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
 In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli.
@@
 <!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
-Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline below.
+Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm].
 First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
 Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
-Note that the probability distributions for the sample and comparison mechanisms, $P(S)$ and $P(C)$, included in the probabilistic model [-@fig-cj16_prob], are omitted from the statistical model [-@fig-cj16_stat].
-We exclude these probabilities from the statistical definition because, although they are part of the joint distribution of the data, they do not influence the dependence between $O_{R}$ and $D_{R}$, since $S$ and $C$ are governed by random selection designs[^8].
+Note that while the probabilistic model [-@fig-cj16_prob] includes the probability distributions for the sampling and comparison mechanisms, $P(S)$ and $P(C)$, we omit these from the statistical model [-@fig-cj16_stat].
+We do so because, although these distributions contribute to the joint distribution of the data, they do not affect the relationship between $O_{R}$ and $D_{R}$.
+This lack of influence is due to the random selection designs that govern $S$ and $C$[^8].
 
@@
 Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
-The discriminal processes represent the underlying written-quality traits of the texts.
+The discriminal processes represent the underlying written-quality trait of the texts.
 Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
@@
 In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
-For clarity, the square brackets in $D_{R}$ only indicate the relevant indices within each trait vector; they do not represent any data subsetting.
+It is important to clarify that the square brackets in $D_{R}$ merely indicate the relevant indices for each trait vector, without implying any subsetting of the data.
 
@@
 Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed.
-This result indicates that our model reverts to the original assumptions for the discriminal processes of stimuli proposed by @Thurstone_1927b (see @tbl-thurstone_cases).
+This result implies that our model reverts to assuming Normal discriminal processes for the stimuli, as originally proposed by @Thurstone_1927b (see @tbl-thurstone_cases).
 
@@
 Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs.
-For instance, if $Z_{J} = 1, \dots, 3$ represents three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
+For instance, if $Z_{J}$ is a variable that can take on three values representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
 

### 2025-04-11 _tex/paper2.tex
Built site for gh-pages

@@
 
-However, because CJ experiments rarely use the exhaustive pairing of
-sampled judges, stimuli, and individuals--due to concerns about the
-practical feasibility of the comparison task
-\citep{Boonen_et_al_2020}--a realistic scenario must also account for
-judges comparing only specific stimuli from certain individuals.
+However, due to concerns about the practical feasibility of the
+comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
+implement an exhaustive pairings of sampled judges, stimuli, and
+individuals. Thus, a realistic scenario must account for the fact that
+judges typically compare only a subset of stimuli authored by certain
+individuals.
 
@@
 \(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
-directly access this outcome due to the comparison mechanism. Moreover,
-it depicts \(C\) as a causal factor influencing \(O_{R}\). As with the
-sampling mechanism, a square encloses \(C\) to indicate that it is a
-conditioned variable--that is, one that determines \emph{which repeated
-judgments the subset of judges makes for the subset of stimuli produced
-by the sampled individuals}. In essence, \(C\) reflects the assumption
-that judges \emph{do not} perform all possible repeated judgments for
-the selected stimuli. Instead, they carry out a ``sufficient number of
+directly access to \(O^{sc}_{R}\) due to the comparison mechanism.
+Moreover, the DAG depicts \(C\) as a causal factor influencing
+\(O_{R}\), enclosing it in a square to indicate that it is a conditioned
+variable. Specifically, \(C\) determines \emph{which repeated judgments
+the subset of judges makes for the subset of stimuli produced by the
+sampled individuals}. In essence, \(C\) reflects the assumption that
+judges \emph{do not} perform all possible repeated judgments for the
+selected stimuli. Instead, they carry out a ``sufficient number of
 comparisons'' \(n_{C}\) to allow for accurate estimation of the
@@
 Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
-which we outline below. First, we assume that \(O_{R}\) follows a
-Bernoulli distribution\footnote{The binomial distribution--including its
-  special case, the Bernoulli distribution-- represent a maximum entropy
-  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
-  means that when only two un-ordered outcomes are possible, and the
-  expected numbers of each type of event are assumed to be constant,
-  then the binomial distribution is the one that is most consistent with
-  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
-  discussion of the binomial as a maximum entropy distribution, see
+which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
+\ref{fig-cj16_scm}, and the statistical model \ref{fig-cj16_scm}. First,
+we assume that \(O_{R}\) follows a Bernoulli distribution\footnote{The
+  binomial distribution--including its special case, the Bernoulli
+  distribution-- represent a maximum entropy distribution for binary
+  events \citep[pp.~34]{McElreath_2020}. This means that when only two
+  un-ordered outcomes are possible, and the expected numbers of each
+  type of event are assumed to be constant, then the binomial
+  distribution is the one that is most consistent with these constraints
+  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
+  binomial as a maximum entropy distribution, see
   \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
@@
 \(D_{R}\) using an inverse-logit function:
-\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that the probability
-distributions for the sample and comparison mechanisms, \(P(S)\) and
-\(P(C)\), included in the probabilistic model \ref{fig-cj16_prob}, are
-omitted from the statistical model \ref{fig-cj16_stat}. We exclude these
-probabilities from the statistical definition because, although they are
-part of the joint distribution of the data, they do not influence the
-dependence between \(O_{R}\) and \(D_{R}\), since \(S\) and \(C\) are
-governed by random selection designs\footnote{Randomization ensures that
-  data---and, by extension, an estimator---satisfies several key
-  identification properties, such as common support, no interference,
-  and consistency. The most critical property, however, is the
-  elimination of confounding. \emph{Confounding} occurs when an external
-  variable, such as \(X_{I}\), simultaneously influences both the
-  outcome (e.g., \(O_{R}\)) and a variable of interest (e.g., \(S\)),
-  resulting in spurious associations between the two
+\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that while the
+probabilistic model \ref{fig-cj16_prob} includes the probability
+distributions for the sampling and comparison mechanisms, \(P(S)\) and
+\(P(C)\), we omit these from the statistical model \ref{fig-cj16_stat}.
+We do so because, although these distributions contribute to the joint
+distribution of the data, they do not affect the relationship between
+\(O_{R}\) and \(D_{R}\). This lack of influence is due to the random
+selection designs that govern \(S\) and \(C\)\footnote{Randomization
+  ensures that data---and, by extension, an estimator---satisfies
+  several key identification properties, such as common support, no
+  interference, and consistency. The most critical property, however, is
+  the elimination of confounding. \emph{Confounding} occurs when an
+  external variable, such as \(X_{I}\), simultaneously influences both
+  the outcome (e.g., \(O_{R}\)) and a variable of interest (e.g.,
+  \(S\)), resulting in spurious associations between the two
   \citep{Everitt_et_al_2010}. Randomization addresses this issue by
@@
 corresponding repeated judge biases, \(B_{JK}[j, k]\). The discriminal
-processes represent the underlying written-quality traits of the texts.
+processes represent the underlying written-quality trait of the texts.
 Note that if we assume that \(B_{JK}[j,k]\) reflects the difference in
@@
 terms, the discriminal processes of the stimuli become an observable
-outcome only through the judges' perceptions (biases). For clarity, the
-square brackets in \(D_{R}\) only indicate the relevant indices within
-each trait vector; they do not represent any data subsetting.
+outcome only through the judges' perceptions (biases). It is important
+to clarify that the square brackets in \(D_{R}\) merely indicate the
+relevant indices for each trait vector, without implying any subsetting
+of the data.
 
@@
 because \(e_{IA}\) follows a Normal distribution, \(T_{IA}\) is also
-normally distributed. This result indicates that our model reverts to
-the original assumptions for the discriminal processes of stimuli
+normally distributed. This result implies that our model reverts to
+assuming Normal discriminal processes for the stimuli, as originally
 proposed by \citet{Thurstone_1927b} (see
@@
 zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
-which each judge belongs. For instance, if \(Z_{J} = 1, \dots, 3\)
-represents three groups of judges with varying expertise, the model
-imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
-scale of the latent trait \citep{Depaoli_2021}.
+which each judge belongs. For instance, if \(Z_{J}\) is a variable that
+can take on three values representing three groups of judges with
+varying expertise, the model imposes the constraint
+\(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the scale of the latent trait
+\citep{Depaoli_2021}.
 

### 2025-04-11 paper2.qmd
section:5; process:write

@@
 
+<!-- 3. segway for next section -->
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by certain individuals.
+
+<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
+Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
+However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
+More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
+[@Boonen_et_al_2020, p.
+5] -->
+
 ::: {#fig-cj14 layout-nrow=2}
@@
 
-<!-- 3. segway for next section -->
-However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
-Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by certain individuals.
-
-<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
-Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
-However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
-More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
-[@Boonen_et_al_2020, p.
-5] -->
-
 ### The comparison mechanism {#sec-theory-theoretical_SC2}
@@
 
-# From SCMs to a statistical model  {#sec-statistical}
+# From SCM to a statistical model  {#sec-statistical}
 
@@
 This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with appropriate functions and probabilistic assumptions [@Pearl_et_al_2016].
-Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]).
+Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]).
 For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent.

### 2025-04-11 _tex/paper2.tex
Built site for gh-pages

@@
 
+However, due to concerns about the practical feasibility of the
+comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
+implement an exhaustive pairings of sampled judges, stimuli, and
+individuals. Thus, a realistic scenario must account for the fact that
+judges typically compare only a subset of stimuli authored by certain
+individuals.
+
 \begin{figure}[H]
@@
 \end{figure}%
-
-However, due to concerns about the practical feasibility of the
-comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
-implement an exhaustive pairings of sampled judges, stimuli, and
-individuals. Thus, a realistic scenario must account for the fact that
-judges typically compare only a subset of stimuli authored by certain
-individuals.
 
@@
 
-\section{From SCMs to a statistical model}\label{sec-statistical}
+\section{From SCM to a statistical model}\label{sec-statistical}
 
@@
 \citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
-to express the joint distribution of our CJ system as a product of
-simpler conditional probability distributions (CPDs)\footnote{This
+to express the joint distribution of our complex CJ system as a product
+of simpler conditional probability distributions (CPDs)\footnote{This
   re-expression is possible because the \emph{chain rule} of probability

### 2025-04-17 paper2.qmd
section:4,5; process:write,formulas

@@
 Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests.
-*Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023].
+*Identification analysis* determines whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023].
 Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010].
@@
 Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
-Their hierarchical nature enables researchers to test hypotheses considering the hierarchical structure of stimuli (see @sec-theory-issue1b) and the uncertainties in trait estimation (see @sec-theory-issue2).
+The model's hierarchical design allows researchers to formulate and test hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for further discussion of these considerations).
 
@@
 The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$.
-Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
+Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ independent columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
 Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest.
 The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
-Here, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ columns (variables), while $e_{IA}$ and $T_{IA}$ are latent matrices with $n_{I}$ rows and $n_{A}$ columns.
+Here, $X_{IA}$ is an observed matrix with dimensions $n_{I} \cdot n_{A}$ rows and $q_{IA}$ independent columns (variables), while $e_{IA}$ and $T_{IA}$ are latent matrices with $n_{I}$ rows and $n_{A}$ columns.
 
@@
 The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
-Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
+Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
 Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
-Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
+Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
 
@@
 The sample-comparison model presents a more realistic scenario than the conceptual-population model.
-In @sec-theory-theoretical_SC1, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
-Furthermore, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample.
+First, in @sec-theory-theoretical_SC1, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
+Second, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample.
 Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$, as proposed by @Thurstone_1927b.
@@
 <!-- 1. start by defining the comparison vector  -->
-As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model (see equation ([-@eq-mat27])).
-The vector contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in equation ([-@eq-mat26]).
+As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model.
+Equation ([-@eq-mat27]) shows that $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in equation ([-@eq-mat26]).
 $$
@@
 The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$.
-It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism.
+It shows the sample-comparison outcome as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism.
 Moreover, the DAG depicts $C$ as a causal factor influencing $O_{R}$, enclosing it in a square to indicate that it is a conditioned variable.
@@
 In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model.
-These extensions account for judge biases (see @sec-theory-issue1b), reflect the hierarchical structure of stimuli (see @sec-theory-issue1b), and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-issue2).
+These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli (see @sec-theory-theoretical_P3), incorporate measurement error in trait estimation and hypothesis testing (also see @sec-theory-theoretical_P3), and clarify the role of the sample and comparison mechanisms in CJ experiments (see @sec-theory-theoretical_SC).
 However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a.
-Because this issue involves statistical assumptions about the distribution of the discriminal process, it requires a statistical model, which we develop in the next section.
+Because this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.
 
@@
 With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a).
-This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with appropriate functions and probabilistic assumptions [@Pearl_et_al_2016].
+This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
 Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]).
@@
 Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
-Note that while the probabilistic model [-@fig-cj16_prob] includes the probability distributions for the sampling and comparison mechanisms, $P(S)$ and $P(C)$, we omit these from the statistical model [-@fig-cj16_stat].
-We do so because, although these distributions contribute to the joint distribution of the data, they do not affect the relationship between $O_{R}$ and $D_{R}$.
-This lack of influence is due to the random selection designs that govern $S$ and $C$[^8].
 
@@
 10.1.2] -->
-
-[^8]: Randomization ensures that data—and, by extension, an estimator—satisfies several key identification properties, such as common support, no interference, and consistency.
-The most critical property, however, is the elimination of confounding.
-*Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $S$), resulting in spurious associations between the two [@Everitt_et_al_2010].
-Randomization addresses this issue by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself.
-For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.
-
-<!-- 2. Continuing with the assumptions -->
-Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
-The discriminal processes represent the underlying written-quality trait of the texts.
-Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
-$$
-\begin{aligned}
-D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
-&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
-& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
-\end{aligned}
-$$ {#eq-mat52}
-This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022].
-In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
-It is important to clarify that the square brackets in $D_{R}$ merely indicate the relevant indices for each trait vector, without implying any subsetting of the data.
-
-<!-- Weinberg[4] proposed a further diagram to capture ‘pure’ interaction in which neither G nor E has a direct effect on D and they only have an effect when present together (see Figure 4a).
-We suggest that this ‘pure’ interaction could be captured by the DAG shown in Figure 4b.
-An example of this is phenylketonuria in which phenylalanine in the diet (E) and the absence of the enzyme to break it down (phenylalanine hydroxylase, G) are both needed in order for a disease to manifest (intellectual disability, D).
-[@Attia_et_al_2022, pp.
-1052] -->
-
-<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder?
--->
-
-<!-- Judges discrimination ability can be integrated in the models if we assume that the bias is not additive (as in Bramley explanation) but multiplicative.
-In the latter case the bias is not eliminated by taking just the difference.
-True, but it is false to think that because the biases are additive they get just eliminated.
-For instance, if we assume that the bias is additive but different for each stimuli, this means that the DAG should be: {T_{IA}, B_{JK}} -> O_{R}, but mathematically you can write it as it is above  -->
 
@@
 \begin{aligned}
-  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
+  & P( O_{R} \mid D_{R}, S, C ) \\
   & P( D_{R} \mid T_{IA}, B_{JK} ) \\
@@
 
+Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16.
+We make this omission because, although these distributions contribute to the overall joint distribution of the data, they do not influence the relationship between $O_{R}$ and $D_{R}$.
+This independence stems from the random selection designs that govern $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$[^8].
+
+[^8]: Randomization ensures that data—and, by extension, an estimator—satisfies several key identification properties, such as common support, no interference, and consistency.
+The most critical property, however, is the elimination of confounding.
+*Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $X_{I}$ or $S$), resulting in spurious associations between the two [@Everitt_et_al_2010].
+Randomization addresses this issue by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself.
+For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.
+
+<!-- 2. Continuing with the assumptions -->
+Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
+The discriminal processes represent the underlying written-quality trait of the texts.
+Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
+$$
+\begin{aligned}
+D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
+&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
+& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
+\end{aligned}
+$$ {#eq-mat52}
+This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022].
+In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
+It is important to clarify that the square brackets in $D_{R}$ merely indicate the relevant indices for each trait vector, without implying any subsetting of the data.
+
+<!-- Weinberg[4] proposed a further diagram to capture ‘pure’ interaction in which neither G nor E has a direct effect on D and they only have an effect when present together (see Figure 4a).
+We suggest that this ‘pure’ interaction could be captured by the DAG shown in Figure 4b.
+An example of this is phenylketonuria in which phenylalanine in the diet (E) and the absence of the enzyme to break it down (phenylalanine hydroxylase, G) are both needed in order for a disease to manifest (intellectual disability, D).
+[@Attia_et_al_2022, pp.
+1052] -->
+
+<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder?
+-->
+
+<!-- Judges discrimination ability can be integrated in the models if we assume that the bias is not additive (as in Bramley explanation) but multiplicative.
+In the latter case the bias is not eliminated by taking just the difference.
+True, but it is false to think that because the biases are additive they get just eliminated.
+For instance, if we assume that the bias is additive but different for each stimuli, this means that the DAG should be: {T_{IA}, B_{JK}} -> O_{R}, but mathematically you can write it as it is above  -->
+
 Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$.
@@
 
-Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., judgment expertise) and judge-specific idiosyncratic errors $e_{J}$.
+Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., the effects of judgment expertise on the quality assessment) and judge-specific idiosyncratic errors $e_{J}$.
 Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs.
@@
 
-Finally, to transform [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data, we use *Bayesian inference methods*.
-The **Declarations** section of this document includes a link to the model, along with an alternative specification that assumes equal discriminal dispersions.
-We implemented both versions of the model using `Stan` [version $2.26.1$, @Stan_2020].
-Although the implementation was successful, we do not report the results here, as they fall outside the scope of this document.
+Finally, using *Bayesian inference methods* to transform the [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data.
+Bayesian inference offers three key advantages.
+First, it is well-suited for estimating complex and overparameterized models, where the number of parameters may exceed the number of observations [@Baker_1998; @Kim_et_al_1999].
+Second, it allows researchers to incorporate prior information, which constrains parameter estimates within plausible bounds.
+This feature mitigates common estimation issues, such as non-convergence or improper solutions, often encountered in complex models under frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
+Third, Bayesian inference is particularly effective for drawing conclusions from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
+The **Declarations** section of this document provides a link to the model code, along with an alternative specification that assumes equal discriminal dispersions.
+We implemented both versions of the model using `Stan` [@Stan_2020, version 2.26.1].
+Although the implementations were successful, we do not report the results here, as they fall outside the scope of this document.
 
@@
 Decision: interesting, but I think I like the model as it is, explicit! -->
-
-<!-- 2. What are the benefits of Bayesian inference?-->
-<!-- Bayesian inference procedures offer three key advantages.
-First, they are well-suited to handling complex and overparameterized models, enabling researchers to estimate models where the number of parameters exceeds the number of observations for estimation [@Baker_1998; @Kim_et_al_1999].
-Second, they allow researchers to incorporate prior information, which helps constrain parameters within specified bounds.
-This capability addresses challenges such as non-convergence or improper parameter estimation that often arise in complex models when analyzed with frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
-Finally, Bayesian methods are particularly effective at drawing inferences from small sample sizes, where relying on the asymptotic properties of frequentist approaches may not be justified [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
--->
 

### 2025-04-17 _tex/paper2.tex
Built site for gh-pages

@@
 DAGs function as \emph{conceptual models} on which identification
-analysis rests. \emph{Identification analysis} helps researchers to
-determine whether an estimator can accurately compute an estimand based
-solely on its (causal) assumptions, regardless of random variability
+analysis rests. \emph{Identification analysis} determines whether an
+estimator can accurately compute an estimand based solely on its
+(causal) assumptions, regardless of random variability
 \citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
@@
 variables affect the primary latent variable of interest
-\citep{Everitt_et_al_2010}. Their hierarchical nature enables
-researchers to test hypotheses considering the hierarchical structure of
-stimuli (see Section~\ref{sec-theory-issue1b}) and the uncertainties in
-trait estimation (see Section~\ref{sec-theory-issue2}).
+\citep{Everitt_et_al_2010}. The model's hierarchical design allows
+researchers to formulate and test hypotheses that account for both the
+nested structure of stimuli and the uncertainties inherent in trait
+estimation (see Section~\ref{sec-theory-issue1b} and
+Section~\ref{sec-theory-issue2} for further discussion of these
+considerations).
 
@@
 \(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
-\(q_{I}\) columns (variables), and both \(e_{I}\) and \(T_{I}\) are
-latent vectors of length \(n_{I}\). Additionally, this branch shows how
-\(T_{I}\), along with \emph{relevant} \footnote{refer to footnote 2.}
-text-related variables \((X_{IA})\) (e.g., text length), and texts'
-idiosyncratic errors \(e_{IA}\) causally influence the texts'
-written-quality trait \((T_{IA})\), the first primary latent variable of
-interest. The error term \(e_{IA}\) captures variations in the texts'
-traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
-\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
-rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
-are latent matrices with \(n_{I}\) rows and \(n_{A}\) columns.
+\(q_{I}\) independent columns (variables), and both \(e_{I}\) and
+\(T_{I}\) are latent vectors of length \(n_{I}\). Additionally, this
+branch shows how \(T_{I}\), along with \emph{relevant} \footnote{refer
+  to footnote 2.} text-related variables \((X_{IA})\) (e.g., text
+length), and texts' idiosyncratic errors \(e_{IA}\) causally influence
+the texts' written-quality trait \((T_{IA})\), the first primary latent
+variable of interest. The error term \(e_{IA}\) captures variations in
+the texts' traits that remain unexplained by \(T_{I}\) or \(X_{IA}\).
+Here, \(X_{IA}\) is an observed matrix with dimensions
+\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) independent columns
+(variables), while \(e_{IA}\) and \(T_{IA}\) are latent matrices with
+\(n_{I}\) rows and \(n_{A}\) columns.
 
@@
 \(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
-columns (variables), and both \(e_{J}\) and \(B_{J}\) are latent vectors
-of length \(n_{J}\). Furthermore, the branch shows how \(B_{J}\), along
-with \emph{relevant} \footnote{refer to footnote 2.} judgment-related
-variables \((Z_{JK})\) (e.g., the number of judgments a judge makes),
-and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
-judges' biases associated with each text \(B_{JK}\), the second primary
-latent variable of interest. The error \(e_{JK}\) captures variations in
-judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
-observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
-columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent
-matrices with \(n_{J}\) rows and \(n_{K}\) columns
+independent columns (variables), and both \(e_{J}\) and \(B_{J}\) are
+latent vectors of length \(n_{J}\). Furthermore, the branch shows how
+\(B_{J}\), along with \emph{relevant} \footnote{refer to footnote 2.}
+judgment-related variables \((Z_{JK})\) (e.g., the number of judgments a
+judge makes), and judgments' idiosyncratic errors \(e_{JK}\) causally
+influence the judges' biases associated with each text \(B_{JK}\), the
+second primary latent variable of interest. The error \(e_{JK}\)
+captures variations in judgments unexplained by \(B_{J}\) or \(Z_{JK}\).
+Here, \(Z_{JK}\) is an observed matrix with dimension
+\(n_{J} \cdot n_{K}\) rows and \(q_{JK}\) independent columns
+(variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent matrices
+with \(n_{J}\) rows and \(n_{K}\) columns
 
@@
 The sample-comparison model presents a more realistic scenario than the
-conceptual-population model. In
+conceptual-population model. First, in
 Section~\ref{sec-theory-theoretical_SC1}, it explicitly assumes
@@
 sample of students \((n^{s}_{I})\), all drawn from the conceptual
-population. Furthermore, in Section~\ref{sec-theory-theoretical_SC2},
-the model assumes that judges do not perform \emph{all repeated
-judgments} within the data sample. Instead, they conduct a sufficient
-number of stimuli comparisons, \(n_{C}\), to ensure an accurate
-estimation of the proportion \(P(B>A)\), as proposed by
-\citet{Thurstone_1927b}.
+population. Second, in Section~\ref{sec-theory-theoretical_SC2}, the
+model assumes that judges do not perform \emph{all repeated judgments}
+within the data sample. Instead, they conduct a sufficient number of
+stimuli comparisons, \(n_{C}\), to ensure an accurate estimation of the
+proportion \(P(B>A)\), as proposed by \citet{Thurstone_1927b}.
 
@@
 mechanism} using the binary vector variable \(C\) to facilitate the
-interpretation of the sample-comparison model (see equation
-(\ref{eq-mat27})). The vector contains \(n\) elements corresponding to
-the number of rows in the \(R\) matrix, with each element
-\(c_{(i,a,h,b,j,k)}\) being a binary value indicating the presence or
-absence of data rows in \(R\), a definition similar to that of
-\(s_{(i,a,h,b,j,k)}\) in equation (\ref{eq-mat26}).
-\begin{equation}\phantomsection\label{eq-mat27}{
+interpretation of the sample-comparison model. Equation (\ref{eq-mat27})
+shows that \(C\) contains \(n\) elements corresponding to the number of
+rows in the \(R\) matrix, with each element \(c_{(i,a,h,b,j,k)}\) being
+a binary value indicating the presence or absence of data rows in \(R\),
+a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in equation
+(\ref{eq-mat26}). \begin{equation}\phantomsection\label{eq-mat27}{
 C = \begin{bmatrix}
@@
 deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
-\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
-directly access to \(O^{sc}_{R}\) due to the comparison mechanism.
-Moreover, the DAG depicts \(C\) as a causal factor influencing
-\(O_{R}\), enclosing it in a square to indicate that it is a conditioned
-variable. Specifically, \(C\) determines \emph{which repeated judgments
-the subset of judges makes for the subset of stimuli produced by the
-sampled individuals}. In essence, \(C\) reflects the assumption that
-judges \emph{do not} perform all possible repeated judgments for the
-selected stimuli. Instead, they carry out a ``sufficient number of
-comparisons'' \(n_{C}\) to allow for accurate estimation of the
-proportion \(P(B>A)\) for each stimulus pair
+as unobserved, highlighting that researchers cannot directly access to
+\(O^{sc}_{R}\) due to the comparison mechanism. Moreover, the DAG
+depicts \(C\) as a causal factor influencing \(O_{R}\), enclosing it in
+a square to indicate that it is a conditioned variable. Specifically,
+\(C\) determines \emph{which repeated judgments the subset of judges
+makes for the subset of stimuli produced by the sampled individuals}. In
+essence, \(C\) reflects the assumption that judges \emph{do not} perform
+all possible repeated judgments for the selected stimuli. Instead, they
+carry out a ``sufficient number of comparisons'' \(n_{C}\) to allow for
+accurate estimation of the proportion \(P(B>A)\) for each stimulus pair
 \citep[pp.~267]{Thurstone_1927b}.
@@
 model. These extensions account for judge biases (see
-Section~\ref{sec-theory-issue1b}), reflect the hierarchical structure of
-stimuli (see Section~\ref{sec-theory-issue1b}), and incorporate
-measurement error in trait estimation and hypothesis testing (see
-Section~\ref{sec-theory-issue2}). However, they do not resolve concerns
-about the assumption of equal dispersions among stimuli discussed in
-Section~\ref{sec-theory-issue1a}. Because this issue involves
-statistical assumptions about the distribution of the discriminal
-process, it requires a statistical model, which we develop in the next
-section.
+Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
+structure of stimuli (see Section~\ref{sec-theory-theoretical_P3}),
+incorporate measurement error in trait estimation and hypothesis testing
+(also see Section~\ref{sec-theory-theoretical_P3}), and clarify the role
+of the sample and comparison mechanisms in CJ experiments (see
+Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
+concerns about the assumption of equal dispersions among stimuli
+discussed in Section~\ref{sec-theory-issue1a}. Because this concern
+relates to the statistical assumption underlying the distribution of the
+discriminal process, we develop a formal statistical model to address it
+in the next section.
 
@@
 functional and probabilistic information, which we can replace with
-appropriate functions and probabilistic assumptions
+suitable functions and probabilistic assumptions
 \citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
@@
 \(D_{R}\) using an inverse-logit function:
-\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that while the
-probabilistic model \ref{fig-cj16_prob} includes the probability
-distributions for the sampling and comparison mechanisms, \(P(S)\) and
-\(P(C)\), we omit these from the statistical model \ref{fig-cj16_stat}.
-We do so because, although these distributions contribute to the joint
-distribution of the data, they do not affect the relationship between
-\(O_{R}\) and \(D_{R}\). This lack of influence is due to the random
-selection designs that govern \(S\) and \(C\)\footnote{Randomization
-  ensures that data---and, by extension, an estimator---satisfies
-  several key identification properties, such as common support, no
-  interference, and consistency. The most critical property, however, is
-  the elimination of confounding. \emph{Confounding} occurs when an
-  external variable, such as \(X_{I}\), simultaneously influences both
-  the outcome (e.g., \(O_{R}\)) and a variable of interest (e.g.,
-  \(S\)), resulting in spurious associations between the two
+\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).
+
+\begin{figure}[H]
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{R} & := f_{O}(D_{R}, S, C) \\ 
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} \\ \\
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj16_scm}SCM}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  & P( O_{R} \mid D_{R}, S, C ) \\
+  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
+  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
+  & P( T_{I} \mid X_{I}, e_{I} ) \\
+  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
+  & P( B_{J} \mid Z_{J}, e_{J} ) \\
+  & P( e_{I} ) \\
+  & P( e_{J} ) \\
+  & P( e_{IA} ) \\  
+  & P( e_{JK} )
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj16_prob}Probabilistic model}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
+  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
+  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
+  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
+  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
+  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
+  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
+  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
+  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
+  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj16_stat}Statistical model}
+
+\end{minipage}%
+
+\caption{\label{fig-cj16}Comparative judgment model, SCM, probabilistic
+and statistical model assuming different discriminal dispersions for the
+student's traits}
+
+\end{figure}%
+
+Note that while the joint distribution in Equation (\ref{eq-mat51})
+includes the probability distributions of the sampling and comparison
+mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
+variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
+\(P(Z_{J})\)--we omit these probabilities from the models presented in
+Figure~\ref{fig-cj16}. We make this omission because, although these
+distributions contribute to the overall joint distribution of the data,
+they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
+This independence stems from the random selection designs that govern
+\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
+\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
+  an estimator---satisfies several key identification properties, such
+  as common support, no interference, and consistency. The most critical
+  property, however, is the elimination of confounding.
+  \emph{Confounding} occurs when an external variable, such as
+  \(X_{I}\), simultaneously influences both the outcome (e.g.,
+  \(O_{R}\)) and a variable of interest (e.g., \(X_{I}\) or \(S\)),
+  resulting in spurious associations between the two
   \citep{Everitt_et_al_2010}. Randomization addresses this issue by
@@
 
-\begin{figure}[H]
-
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O_{R} & := f_{O}(D_{R}, S, C) \\ 
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} \\ \\
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj16_scm}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
-  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
-  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
-  & P( T_{I} \mid X_{I}, e_{I} ) \\
-  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
-  & P( B_{J} \mid Z_{J}, e_{J} ) \\
-  & P( e_{I} ) \\
-  & P( e_{J} ) \\
-  & P( e_{IA} ) \\  
-  & P( e_{JK} )
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj16_prob}Probabilistic model}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
-  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
-  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
-  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
-  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
-  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
-  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
-  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
-  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj16_stat}Statistical model}
-
-\end{minipage}%
-
-\caption{\label{fig-cj16}Comparative judgment model, SCM, probabilistic
-and statistical model assuming different discriminal dispersions for the
-student's traits}
-
-\end{figure}%
-
 Third, we define \(T_{IA}\) as a linear combination of the students'
@@
 Sixth, we define \(B_{J}\) as a linear combination of the effects of
-relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
-expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
-element of \(e_{J}\) is likewise assumed to be \(iid\) Normal with mean
-zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
-which each judge belongs. For instance, if \(Z_{J}\) is a variable that
-can take on three values representing three groups of judges with
-varying expertise, the model imposes the constraint
-\(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the scale of the latent trait
-\citep{Depaoli_2021}.
-
-Finally, to transform model~\ref{fig-cj16_stat} into a practical
-statistical tool for analyzing paired comparison data, we use
-\emph{Bayesian inference methods}. The \textbf{Declarations} section of
-this document includes a link to the model, along with an alternative
-specification that assumes equal discriminal dispersions. We implemented
-both versions of the model using \texttt{Stan} \citep[version
-\(2.26.1\),][]{Stan_2020}. Although the implementation was successful,
-we do not report the results here, as they fall outside the scope of
-this document.
+relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., the effects of
+judgment expertise on the quality assessment) and judge-specific
+idiosyncratic errors \(e_{J}\). Each element of \(e_{J}\) is likewise
+assumed to be \(iid\) Normal with mean zero and a standard deviation
+\(s_{ZJ}\) that depends on the groups to which each judge belongs. For
+instance, if \(Z_{J}\) is a variable that can take on three values
+representing three groups of judges with varying expertise, the model
+imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
+scale of the latent trait \citep{Depaoli_2021}.
+
+Finally, using \emph{Bayesian inference methods} to transform the
+model~\ref{fig-cj16_stat} into a practical statistical tool for
+analyzing paired comparison data. Bayesian inference offers three key
+advantages. First, it is well-suited for estimating complex and
+overparameterized models, where the number of parameters may exceed the
+number of observations \citep{Baker_1998, Kim_et_al_1999}. Second, it
+allows researchers to incorporate prior information, which constrains
+parameter estimates within plausible bounds. This feature mitigates
+common estimation issues, such as non-convergence or improper solutions,
+often encountered in complex models under frequentist methods
+\citep{Martin_et_al_1975, Seaman_et_al_2011}. Third, Bayesian inference
+is particularly effective for drawing conclusions from small samples,
+where the asymptotic properties underlying frequentist methods are less
+reliable \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
+The \textbf{Declarations} section of this document provides a link to
+the model code, along with an alternative specification that assumes
+equal discriminal dispersions. We implemented both versions of the model
+using \texttt{Stan} \citep[version 2.26.1]{Stan_2020}. Although the
+implementations were successful, we do not report the results here, as
+they fall outside the scope of this document.
 

### 2025-04-18 paper2.qmd
section:5; process:write

@@
 
+<!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
 <!-- recording: Sven and Tine 25.04.10; time: 00:16:30 - 00:19:00 -->
@@
 
-<!-- 2. assumptions about the outcome, and the sample and comparison mechanism -->
+<!-- 2. assumptions about the outcome -->
 Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm].
@@
 
+<!-- 3. Note about the sample and comparison mechanism, as well as the predictor's distribution -->
 Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16.
@@
 
-<!-- 2. Continuing with the assumptions -->
+<!-- 4. functional form for D_{R} -->
 Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
@@
 
+<!-- 5. Functional form and assumptions for T_{IA} -->
 Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$.
@@
 
+<!-- 6. Functional form and assumptions for T_{I} -->
 Fourth, we define $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$.
@@
 
+<!-- 7. Functional form and assumptions for B_{JK} -->
 Fifth, we define $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments a judge makes affect the assessment of quality), and judgment-specific idiosyncratic errors $e_{JK}$.
@@
 
+<!-- 8. Functional form and assumptions for B_{J} -->
 Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., the effects of judgment expertise on the quality assessment) and judge-specific idiosyncratic errors $e_{J}$.
@@
 
-Finally, using *Bayesian inference methods* to transform the [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data.
-Bayesian inference offers three key advantages.
-First, it is well-suited for estimating complex and overparameterized models, where the number of parameters may exceed the number of observations [@Baker_1998; @Kim_et_al_1999].
-Second, it allows researchers to incorporate prior information, which constrains parameter estimates within plausible bounds.
-This feature mitigates common estimation issues, such as non-convergence or improper solutions, often encountered in complex models under frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
-Third, Bayesian inference is particularly effective for drawing conclusions from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
+<!-- 9. Computational implementation -->
+Finally, we use *Bayesian inference methods* to convert [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data.
+Bayesian inference offers three key advantages for this purpose.
+First, it handles complex and overparameterized models--such as the one we implement--where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999].
+Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
+In our implementation, we use priors to define the error distribution, which sets the scale of the latent variables [@Depaoli_2014].
+Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
+This feature is particularly relevant in CJ experiments, where researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.
+
 The **Declarations** section of this document provides a link to the model code, along with an alternative specification that assumes equal discriminal dispersions.
@@
 
-<!-- recording: Sven 24.10.04; time: 00:09:10 - 00:14:30 -->
-<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:56:10 -->
+<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:58:20 -->
 <!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->
 
-<!-- Thus, this section addresses the issues identified in @sec-theory-issues by extending  Thurstone’s theory using the structural approach to causal inference.
-Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+<!-- make a summary of what you have accomplished so far -->
+
+<!-- change the paradigm -->
+
+<!-- opportunities open  -->
+
+<!-- Now that we have this model we have a principled way to make some power analysis which will helps us to design studies, rather than just relying in simple rules of thumbs. -->
+
+<!-- Thus, this study addresses the issues identified in @sec-theory-issues by extending  Thurstone’s general form using the structural approach to causal inference and Bayesian inference methods.
+Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons, i.e, the sampleand compariosn mechanisms.
 In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.

### 2025-04-18 _tex/paper2.tex
Built site for gh-pages

@@
 
-Finally, using \emph{Bayesian inference methods} to transform the
+Finally, we use \emph{Bayesian inference methods} to convert
 model~\ref{fig-cj16_stat} into a practical statistical tool for
 analyzing paired comparison data. Bayesian inference offers three key
-advantages. First, it is well-suited for estimating complex and
-overparameterized models, where the number of parameters may exceed the
-number of observations \citep{Baker_1998, Kim_et_al_1999}. Second, it
-allows researchers to incorporate prior information, which constrains
-parameter estimates within plausible bounds. This feature mitigates
-common estimation issues, such as non-convergence or improper solutions,
-often encountered in complex models under frequentist methods
-\citep{Martin_et_al_1975, Seaman_et_al_2011}. Third, Bayesian inference
-is particularly effective for drawing conclusions from small samples,
-where the asymptotic properties underlying frequentist methods are less
-reliable \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
+advantages for this purpose. First, it handles complex and
+overparameterized models--such as the one we implement--where the number
+of parameters exceeds the number of observations
+\citep{Baker_1998, Kim_et_al_1999}. Second, it incorporates prior
+information to constrain parameter estimates within plausible bounds,
+thereby mitigating estimation issues like non-convergence or improper
+solutions that often affect frequentist methods
+\citep{Martin_et_al_1975, Seaman_et_al_2011}. In our implementation, we
+use priors to define the error distribution, which sets the scale of the
+latent variables \citep{Depaoli_2014}. Third, Bayesian inference
+supports robust inferences from small samples, where the asymptotic
+properties underlying frequentist methods are less reliable
+\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
+feature is particularly relevant in CJ experiments, where researchers
+often collect large volumes of paired comparisons but work with
+relatively small samples of judges, stimuli, and individuals to test
+hypotheses.
+
 The \textbf{Declarations} section of this document provides a link to

### 2025-04-21 paper2.qmd
section:6, process:write

@@
 <!-- 5. So, what are we doing? -->
-To address these issues, this study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods.
-In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
+To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines Causal and Bayesian inference methods.
+In addition to potentially enhancing measurement reliability and validity [@Perron_et_al_2015], and improving statistical accuracy in hypothesis testing, this approach offers two key advantages.
 First, it clarifies the interactions among all actors and processes involved in CJ experiments.
 Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, pp. 2] -->
 
@@
 @sec-theoretical extends Thurstone's general form to address these challenges.
-The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+The extension integrates core theoretical principles alongside key CJ experimental design features, such as the selection of judges, stimuli, and comparisons.
 @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data.
@@
 For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018].
-These misfit texts may exhibit larger discriminal dispersions or represent outliers --texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, comparing misfit texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them.
+These misfit texts may exhibit larger discriminal dispersions or represent outliers--texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+In either case, the comparison of these texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them, except their elimination from the analysis.
 
@@
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Furthermore, if researchers acknowledge that misfit statistics may represent outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
+Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
 12].
@@
 In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory).
-Additionally, based on the arguments developed in @sec-theory-issue1b, the SCM incorporates the judges' biases $(B_{kj})$.
-Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
+Additionally, building on the arguments developed in @sec-theory-issue1b, the SCM incorporates a key design features of CJ experiments: the judges' biases $(B_{kj})$.
+These variables, along with the outcome $(O_{iahbjk})$, constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
 Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
@@
 <!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:58:20 -->
+<!-- recording: Sven and Tine 25.04.10; time: 00:12:00 - 00:14:30 -->
 <!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->
 
-<!-- make a summary of what you have accomplished so far -->
-
-<!-- change the paradigm -->
-
-<!-- opportunities open  -->
-
-<!-- Now that we have this model we have a principled way to make some power analysis which will helps us to design studies, rather than just relying in simple rules of thumbs. -->
-
-<!-- Thus, this study addresses the issues identified in @sec-theory-issues by extending  Thurstone’s general form using the structural approach to causal inference and Bayesian inference methods.
-Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons, i.e, the sampleand compariosn mechanisms.
-In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
+<!-- 1. General summary of what we have done. What are the first benefits? -->
+This study extends Thurstone's general form through a systematic, integrated approach that combines Causal and Bayesian inference methods.
+It begins by formulating the extension as a conceptual model, represented by an SCM and a DAG.
+This formulation integrates Thurstone's core theoretical principles, including the distribution of the discriminal process, with key CJ experimental design features, such as the sampling and comparison mechanisms.
+The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model.
+Specifically, the statistical model accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error in trait estimation and hypothesis testing, and accommodates violations of the equal dispersion assumption.
+These improvements potentially enhance measurement reliability and validity [@Perron_et_al_2015] and improve the statistical accuracy of hypothesis testing.
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, pp. 2] -->
+
+<!-- 2. Are there any additional advantages? -->
+Beyond these potential benefits, the approach offers two key advantages.
 First, it clarifies the interactions among all actors and processes involved in CJ experiments.
-Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.
--->
-
-<!-- "If Thurstone’s Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes." [@Kelly_et_al_2022, pp.
+Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data.
+Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs--one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.
+
+<!-- If Thurstone's Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes.
+[@Kelly_et_al_2022, pp.
 678] -->
 
-<!-- These procedures also bear on a larger topic of discussion in social science.
-The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
-[@Wainer_et_al_1978, pp.
-332] -->
-
-<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
-(2002).
-Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
-[@Bramley_2008, pp.
-274] -->
-<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges.  -->
+<!-- We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used.
+However, work to derive some general principles for good practice would be valuable.
+[@Kelly_et_al_2022, pp.
+683] -->
+
+<!-- 3. First research avenue -->
+One promising research avenue involves investigating how the hierarchical structure of stimuli affects CJ experiment designs and hypotheses testing.
+One such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy.
+Due to concerns about the feasibility of the comparison task [@Boonen_et_al_2020], many CJ studies assess traits using only one text per individual [@citation].
+This experimental design implicitly assumes that within-individual variability is small relative to between-individual differences--that is, texts produced by the same individual are relatively homogeneous.
+However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates.
+
+Relatedly, another study could investigate hypotheses about the characteristics of the texts themselves.
+Previously, the literature lacked a direct and integrated method for evaluating these types of hypotheses.
+Now, by collecting data on specific stimulus features, researchers can assess whether these variables produce the expected effects.
+
+<!-- power analysis to know the implications of certain sampling is nice to have it explicitly. -->
+
+<!-- One of these avenues could be about the implications of the hierarchical structure of the stimuli, not only for the design of CJ experiments, but also for the test of hypothesis.  -->
+<!-- The avenues of research span the following:  -->
+<!-- - assessing if the number of texts a researcher collects is sufficient.
+Due to efficiency of scoring, most of research only collects only one text.
+This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
+However, this cannot be known ad-hoc.
+The frameworks open the avenue to test this assumption -->
+<!-- - Assessing hypothesis regarding the characteristics of the texts.
+Before the literature did not have a direct and integrated way to assess these types of hypothesis, now you just need to collect data on the particular stimuli, and test if these variables reflect the expected effect -->
+
+<!-- Derived from this other avenues of research that are open are -->
+<!-- - assessing if the number of individuals a researcher collects is sufficient.
+Most of the literature focuses on ensuring that the number of comparisons between texts are sufficient, but no literature actively engages in determining if the number of individuals are sufficient to test hypothesis at that level.
+-->
+<!-- - Moreover, if certain characteristics of the individuals explain the trait locations of their texts -->
 
@@
 
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
+
+<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
+However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
+
+<!-- When you want to investigate objects, the experiment has an experimental unit: the trial on the objects, but you only have one object per subject.
+Consequently, objects and subjects are confounded.
+-->
+
+<!-- presence of judges biases in the comparative judgment procedure. From this it derives research on:  -->
+<!-- - does the judges have biases, and if so, do ad-hoc sta -->
+
+<!-- Given that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, this study has argued that judges biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
+-->
+
+<!-- This hierarchical nature open paths for research in, for instance, the benefits of *repeated measures designs* [@Lawson_2015, pp.
+366-376] in CJ experiments.
+This is when the experimental conditions accounts for judges comparing the same pair of stimuli multiple times.
+Or research that determines the minimum number of judges and judgments required to ensure an acceptable level of reliability and validity under particular assumptions of the data [@Kelly_et_al_2022, pp.
+682].
+-->
+
 <!-- (Is this a validity contention?) -->
@@
 
-<!-- Pollitt (2012a) argues that assessors in pairwise comparisons need less training, given intuitive comparison and decision making.
-However, there is a distinct lack of research on the role of training.
-[?] -->
-
-<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
+<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology.
+We do not yet know how widely it can be, or should be, applied.
+[@Pollitt_2012b, pp.
+297]-->
+
+<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
+[@Kelly_et_al_2022, pp.
+681] -->
 
@@
 11] -->
+
+<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
+--->
 
@@
 
-<!-- sampling and comparisons mechanism are effective generators of missing data. -->
-
 <!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
@@
 
-<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
-
-<!-- design matters -->
+<!-- clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
+Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
+-->
+
+<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
+Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
+It also accounts for potential judge biases in the comparative judgment process.
+-->
+
+<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
 
@@
 
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
-267] -->
+<!-- Continuing with the first advantage, now we clarify two additional processes of CJ experiments that previous literature do not typically mention: that the sample and comparison mechanism are generators of missing data--that is, they determine which observations do we get to see in our data sample.
+This opens the path to use Rubin's missing data framework [@Rubin_1976] into CJ applications to investigate more complex CJ designs, particularly the ones that are not random samples as the ones outlined here, which could be characterized in Rubin's structure parlor as missing at random (MAR) and missing not at random (MNAR) [@Rubin_1976].
+Among these cases are adaptive comparative algorithm designs, for which the causal structure is not clear, and thus, its impact on trait measurement and hypothesis testing is also not clear.
+Another  -->
+
+<!-- normally S and C are not explicitly mentioned  -->
+
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
+
+<!-- Brought Rubin's missing data framework to the field. -->
+<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are random samples, they fulfill ignorability. -->
 
@@
 
-<!-- Due to efficiency of scoring, most of research only collects only one text.
-This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
-However, this cannot be known ad-hoc.
--->
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
+<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
+(2002).
+Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
+[@Bramley_2008, pp.
+274] -->
+<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges. -->
 
@@
 
-<!-- Brought Rubin's missing data framework to the field. -->
-<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are are random samples, they fulfill ignorability.  -->
-
-<!-- recording: Sven and Tine 25.04.10; time: 00:12:00 - 00:14:00 -->
-<!-- Consider how to use the S and C, when the designs are more complexes. -->
-<!-- Normally, S and C are not typically mentioned  -->
+<!-- These procedures also bear on a larger topic of discussion in social science.
+The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
+[@Wainer_et_al_1978, pp.
+332] -->
 
@@
 
+<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014).
+For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
+[@Deffner_et_al_2022, pp.
+2] -->
+
+<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
+
 <!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
@@
 113] -->
-
-## Limitations and further research {#sec-discussion-limitations}
 
@@
 
-<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
+<!-- Talk about Bayesian Workflow, and how it would fit the process to arrive to the proposed model. -->
 
@@
 <!-- Add to future research how would be to use the model when you do not know the weights of the sample, but you can use external data to approximate them. -->
-
-<!-- Talk about Bayesian Workflow, and how it would fit the process to arrive to the proposed model. -->
-
-<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
-However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
-
-<!-- When you want to investigate objects, the experiment has an experimental unit: the trial on the objects, but you only have one object per subject.
-Consequently, objects and subjects are confounded.
--->
 

### 2025-04-21 _tex/paper2.tex
Built site for gh-pages

@@
 To address these issues, this study extends Thurstone's general form
-using a systematic and integrated approach based on Causal and Bayesian
-inference methods. In addition to improving statistical accuracy and
-strengthening measurement reliability and validity, the approach offers
-two key advantages. First, it clarifies the interactions among all
-actors and processes involved in CJ experiments. Second, it shifts the
-current comparative data analysis paradigm from passively accepting the
-BTL model assumptions to actively testing whether those assumptions fit
-the data under analysis.
+through a systematic and integrated approach that combines Causal and
+Bayesian inference methods. In addition to potentially enhancing
+measurement reliability and validity \citep{Perron_et_al_2015}, and
+improving statistical accuracy in hypothesis testing, this approach
+offers two key advantages. First, it clarifies the interactions among
+all actors and processes involved in CJ experiments. Second, it shifts
+the current comparative data analysis paradigm from passively accepting
+the BTL model assumptions to actively testing whether those assumptions
+fit the data under analysis.
 
@@
 form to address these challenges. The extension integrates core
-theoretical principles alongside key assessment design features relevant
-to CJ experiments, such as the selection of judges, stimuli, and
-comparisons. Section~\ref{sec-statistical} translates these theoretical
-and practical elements into a probabilistic statistical model to analyze
-dichotomous pairwise comparison data. Section~\ref{sec-discussion}
-discusses the findings, limitations, and challenges and explores avenues
-for future research. Finally, Section~\ref{sec-conclusion} summarizes
-the study's conclusions.
+theoretical principles alongside key CJ experimental design features,
+such as the selection of judges, stimuli, and comparisons.
+Section~\ref{sec-statistical} translates these theoretical and practical
+elements into a probabilistic statistical model to analyze dichotomous
+pairwise comparison data. Section~\ref{sec-discussion} discusses the
+findings, limitations, and challenges and explores avenues for future
+research. Finally, Section~\ref{sec-conclusion} summarizes the study's
+conclusions.
 
@@
 These misfit texts may exhibit larger discriminal dispersions or
-represent outliers --texts with distinctive characteristics that deviate
+represent outliers--texts with distinctive characteristics that deviate
 markedly from the rest of the sample \citep{Grubbs_1969}. In either
-case, comparing misfit texts results in more judgment discrepancies, yet
-the BTL model neither accounts for these cases nor offers any means to
-address them.
+case, the comparison of these texts results in more judgment
+discrepancies, yet the BTL model neither accounts for these cases nor
+offers any means to address them, except their elimination from the
+analysis.
 
@@
 \(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
-acknowledge that misfit statistics may represent outlying observations,
-the common CJ practice of excluding stimuli based on these statistics
+acknowledge that misfit statistics may represent texts with different
+dispersions or outlying observations, the common CJ practice of
+excluding stimuli based on these statistics
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
@@
 difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
-Additionally, based on the arguments developed in
-Section~\ref{sec-theory-issue1b}, the SCM incorporates the judges'
-biases \((B_{kj})\). Together with the outcome, these variables
-constitute the preliminary set of endogenous variables,
+Additionally, building on the arguments developed in
+Section~\ref{sec-theory-issue1b}, the SCM incorporates a key design
+features of CJ experiments: the judges' biases \((B_{kj})\). These
+variables, along with the outcome \((O_{iahbjk})\), constitute the
+preliminary set of endogenous variables,
 \(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,
@@
 
-\subsection{Limitations and further
-research}\label{sec-discussion-limitations}
+This study extends Thurstone's general form through a systematic,
+integrated approach that combines Causal and Bayesian inference methods.
+It begins by formulating the extension as a conceptual model,
+represented by an SCM and a DAG. This formulation integrates Thurstone's
+core theoretical principles, including the distribution of the
+discriminal process, with key CJ experimental design features, such as
+the sampling and comparison mechanisms. The study then translates the
+SCM into a bespoke statistical model that addresses key limitations of
+the BTL model. Specifically, the statistical model accounts for judge
+biases, captures the hierarchical structure of stimuli, incorporates
+measurement error in trait estimation and hypothesis testing, and
+accommodates violations of the equal dispersion assumption. These
+improvements potentially enhance measurement reliability and validity
+\citep{Perron_et_al_2015} and improve the statistical accuracy of
+hypothesis testing.
+
+Beyond these potential benefits, the approach offers two key advantages.
+First, it clarifies the interactions among all actors and processes
+involved in CJ experiments. Second, it shifts the analytic paradigm from
+passively accepting the assumptions of the BTL model to actively testing
+their fit with observed data. Together, these advantages create a
+principled framework for evaluating best practices in CJ experimental
+designs--one that better aligns with the demands of contemporary CJ
+assessment contexts \citep{Kelly_et_al_2022} and opens new avenues for
+research.
+
+One promising research avenue involves investigating how the
+hierarchical structure of stimuli affects CJ experiment designs and
+hypotheses testing. One such study could examine whether collecting a
+single text per individual provides sufficient information to produce
+reliable trait estimates at different levels of the hierarchy. Due to
+concerns about the feasibility of the comparison task
+\citep{Boonen_et_al_2020}, many CJ studies assess traits using only one
+text per individual \citep{citation}. This experimental design
+implicitly assumes that within-individual variability is small relative
+to between-individual differences--that is, texts produced by the same
+individual are relatively homogeneous. However, researchers rarely test
+this assumption, even though it directly influences decisions about the
+number of texts and participants needed for the experiment, ultimately
+affecting the reliability of the trait estimates.
+
+Relatedly, another study could investigate hypotheses about the
+characteristics of the texts themselves. Previously, the literature
+lacked a direct and integrated method for evaluating these types of
+hypotheses. Now, by collecting data on specific stimulus features,
+researchers can assess whether these variables produce the expected
+effects.
 

### 2025-04-29 paper2.qmd
section:4.1.1; process:write

@@
 
-Finally, as discussed in the previous section, removing *misfit* judges based risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
-12].
-The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis.
-*Misfit judges* are those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].
+Finally, as discussed in the previous section, researchers risk discarding valuable information and introducing bias into trait estimates when they remove *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], considered outliers under the BTL model [@Grubbs_1969].
+The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
 
@@
 
-<!-- Conventionally, any ﬁgure greater than the mean plus two standard deviations – 1.14 here – is taken to indicate signiﬁcant misﬁt.
+<!-- Conventionally, any figure greater than the mean plus two standard deviations – 1.14 here – is taken to indicate signiﬁcant misfit.
 In this case, only Judge ‘46 N-M 2’ exceeded the criterion.
@@
 20] -->
+
+<!-- @Zimmerman_1994 the whole paper -->
+
+<!-- Furthermore, these points tend to be outliers with unlikely values, according to the model.
+As a result, they are highly influential and make it difficult to estimate out-of-sample predictive accuracy.
+Why?
+Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
+[@McElreath_2020, pp.
+231] -->
+
+<!-- What can be done about this?
+There is a tradition of dropping outliers.
+People sometimes drop outliers even before a model is fit, based only on standard deviations from the mean outcome value.
+You should never do that—a point can only be unexpected and highly influential in light of a model.
+After you fit a model, the picture changes.
+If there are only a few outliers, and you are sure to report results both with and without them, dropping outliers might be okay.
+But if there are several outliers and we really need to model them, what then?
+[@McElreath_2020, pp.
+232] -->
+
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
+
+<!-- Consider, for example, the simplest location and scale problem-inference for a normal distribution with unknown mean \theta and variance \sigma^{2}.
+If we look at the marginal posterior distribution for \theta we know that if \sigma^{2} is known this posterior exhibits outlier-resistance, and we might expect similar behavior for unknown \sigma^{2}, but a simple example refutes this.
+Let the prior distribution give probability p (0<p<1) to (\theta = 1, \sigma^{2} = 1) and 1-p to (\theta = 0, \sigma^{2} = 4).
+Then the posterior mean of \theta is a decreasing function of any sufficiently large observation.
+[@OHagan_2018, pp.
+366] -->
 
@@
 In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory).
-Additionally, building on the arguments developed in @sec-theory-issue1b, the SCM incorporates a key design features of CJ experiments: the judges' biases $(B_{kj})$.
-These variables, along with the outcome $(O_{iahbjk})$, constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
+Additionally, the SCM incorporates a key design feature of CJ experiments: the judges' biases $(B_{kj})$.
+This extension builds on the arguments presented in @sec-theory-issue1b, contending that the discriminal difference becomes an observable outcome only through judges' perceptions.
+Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons.
+
+Notably, all these variables and the outcome $(O_{iahbjk})$ constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
 Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

### 2025-04-29 code/CJfig.R

@@ -95,6 +95,72 @@ dev.off()
 
 
 
+
+
+## figure 1a ####
+
+# distributional parameters
+dpA = c(-0.9, 0.5) # distribution text A (mean, sd)
+dpB = c(0, 1) # distribution text B (mean, sd)
+dpC = c(0.9, 0.5) # distribution text B (mean, sd)
+rAB = 0 # correlation between discriminal processes
+
+
+# figure
+png( filename=file.path(main_dir, 'diff_discriminal_dispersion.png'),
+     height=10, width=15, units='cm', res=400  )
+
+# initial figure
+par( mar=c(2,0,0.5,0) )
+plot( NULL, xlim=c(-3,3), ylim=c(0,0.9), 
+      xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
+axis( side=1, labels=F, lwd.ticks=0 )
+axis( side=1, at=c(dpA[1], dpB[1], dpC[1]), tick=F, 
+      label=c( expression( T[A] ), expression( T[B] ), expression( T[C] ) ) )
+
+# figure parameters
+pplotA = c( dnorm( dpA[1], mean=dpA[1], sd=dpA[2] ),
+            dnorm( dpA[1], mean=dpA[1], sd=dpA[2] ) + 0.05,
+            dnorm( dpA[1] + dpA[2], mean=dpA[1], sd=dpA[2]),
+            dnorm( dpA[1]+0.25, mean=dpA[1], sd=dpA[2] )-0.25 )
+
+pplotB = c( dnorm( dpB[1], mean=dpB[1], sd=dpB[2] ),
+            dnorm( dpB[1], mean=dpB[1], sd=dpB[2] ) + 0.05,
+            dnorm( dpB[1] + dpB[2], mean=dpB[1], sd=dpB[2]),
+            dnorm( dpB[1]+0.45, mean=dpB[1], sd=dpB[2] )-0.15 )
+
+pplotC = c( dnorm( dpC[1], mean=dpC[1], sd=dpC[2] ),
+            dnorm( dpC[1], mean=dpC[1], sd=dpC[2] ) + 0.05,
+            dnorm( dpC[1] + dpC[2], mean=dpC[1], sd=dpC[2]),
[…]

### 2025-04-29 _tex/paper2.tex
Built site for gh-pages

@@
 
-Finally, as discussed in the previous section, removing \emph{misfit}
-judges based risks discarding valuable information and even introduce
-bias into the trait estimates
-\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
-magnitude of these biases remain unpredictable because they depend on
-which judges researchers exclude from the analysis. \emph{Misfit judges}
-are those whose evaluations deviate substantially from the shared
-consensus due to the unique characteristics of either the stimuli or the
-judges themselves
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
+Finally, as discussed in the previous section, researchers risk
+discarding valuable information and introducing bias into trait
+estimates when they remove \emph{misfit} judges--that is, judges whose
+assessments deviate markedly from the shared consensus
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
+considered outliers under the BTL model \citep{Grubbs_1969}. The
+direction and magnitude of these biases remain unpredictable, as they
+depend on which judges researchers exclude from the analysis
+\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
 
@@
 difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
-Additionally, building on the arguments developed in
-Section~\ref{sec-theory-issue1b}, the SCM incorporates a key design
-features of CJ experiments: the judges' biases \((B_{kj})\). These
-variables, along with the outcome \((O_{iahbjk})\), constitute the
-preliminary set of endogenous variables,
+Additionally, the SCM incorporates a key design feature of CJ
+experiments: the judges' biases \((B_{kj})\). This extension builds on
+the arguments presented in Section~\ref{sec-theory-issue1b}, contending
+that the discriminal difference becomes an observable outcome only
+through judges' perceptions. Given that such perceptions may be
+imperfect--and that each judge may carry some degree of bias
+\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
+that judges' perceptions (bias) should be treated as an integral
+component of the CJ system from the outset, as this leads to a more
+accurate representation of the data-generating process underlying the
+pairwise comparisons.
+
+Notably, all these variables and the outcome \((O_{iahbjk})\) constitute
+the preliminary set of endogenous variables,
 \(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,

### 2025-04-30 paper2.qmd
section:all; process:write

@@
 Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
-This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015].
+This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates.
 First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data.
@@
 To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines Causal and Bayesian inference methods.
-In addition to potentially enhancing measurement reliability and validity [@Perron_et_al_2015], and improving statistical accuracy in hypothesis testing, this approach offers two key advantages.
+In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages.
 First, it clarifies the interactions among all actors and processes involved in CJ experiments.
@@
 Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
-In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.
+In this distribution, the mode (mean) represents the average relative separation between the stimuli, and its dispersion indicates the variability of that separation.
 
@@
 <!-- 4. Example of discriminal difference -->
-@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process.
+@fig-discriminal_difference illustrates the distribution of the discriminal difference for the two hypothetical texts.
 The figure indicates that the judge perceives Text B as having significantly higher quality than Text A.
@@
 <!-- 3. But are our specific problems? -->
-However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use its measurement scores for hypothesis testing.
+However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use Case V's measurement scores for hypothesis testing.
 Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
@@
 These misfit texts may exhibit larger discriminal dispersions or represent outliers--texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, the comparison of these texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them, except their elimination from the analysis.
+In either case, the BTL model neither accounts for these cases nor offers any means to address them, except their elimination from the analysis.
 
@@
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap.
-12].
+Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
 The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
-
-<!-- But ignoring over-dispersion can also lead to all of the same problems as ignoring any predictor variable.
-Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
-[@McElreath_2020, pp.
-370] -->
 
@@
 
+<!-- But ignoring over-dispersion can also lead to all of the same problems as ignoring any predictor variable.
+Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
+[@McElreath_2020, pp.
+370] -->
+
 ### The assumption of zero correlation between stimuli {#sec-theory-issue1b}
@@
 <!-- 1. What does the zero correlation between stimuli imply? -->
-The correlation $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another.
-As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
-A similar thought experiment, as the one depicted in @sec-theory-issue1a, can illustrate this idea.
-The experiment only assumes that the discriminal dispersions for both texts remain constant.
-@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower.
-This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome.
-Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are positively correlated.
-As a result, the discriminal difference is less likely to favor Text B over Text A, as it is represented by the shaded gray area.
+The correlation $\rho$, illustrated in @fig-discriminal_difference, measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another.
+Similar to the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
+A thought experiment, akin to the one presented in @sec-theory-issue1a, can illustrate this idea.
+Assuming that the discriminal dispersions for both texts remain constant, @fig-correlation shows that as the correlation between the two texts increases, the distribution of their discriminal difference becomes narrower.
+This narrowing, in turn, affects the probability that the discriminal difference distinctly favors Text B over Text A--denoted as $P(B > A)$--and thus directly influences the comparison outcome.
+Furthermore, the figure shows that when two texts are independent or uncorrelated, as assumed in the BTL model $(\rho=0)$, the distribution of their discriminal difference is less narrow than in scenarios where the texts are positively correlated.
+As a result, it becomes less likely for the comparison to favor Text B over Text A, as indicated by the larger shaded area.
 
@@
 <!-- 3. But does it hold?  -->
-Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b].
-@Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
+Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, experimental designs often adopt the assumption of no correlation between stimuli based on an old theoretical justification.
+Specifically, @Thurstone_1927b argued that stimuli could be treated as uncorrelated because judges' biases--arising from two opposing and equally weighted effects occurring during the pairwise comparisons--would cancel each other out.
+This idea was later formalized by @Andrich_1978, who provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
 However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.
@@
 <!-- 3. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010].
-For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or even their general quality.
-Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient.
+In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli--a statistical phenomenon known as clustering [@Everitt_et_al_2010].
+For instance, when the same individual produces multiple texts, those texts often exhibit common features, such as writing style or overall quality, that judges can easily recognize.
+In this regard, although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient.
 For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.
+However, this approach can introduce additional statistical and measurement issues, similar to the ones discussed in @sec-theory-issue2.
 
@@
 
-Finally, as discussed in the previous section, researchers risk discarding valuable information and introducing bias into trait estimates when they remove *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], considered outliers under the BTL model [@Grubbs_1969].
+Finally, as discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969]--risk discarding valuable information and introducing bias into trait estimates.
 The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
@@
 Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
-@fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error $(\sigma_{T})$ by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.
+@fig-measurement_error illustrates these issues, showing that neglecting measurement error $(\sigma_{T})$--by relying solely on outcome averages--reduces the precision of a predictor's estimated effect $\beta_{X}$, as indicated by the solid lines.
 
@@
 Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons.
-
-Notably, all these variables and the outcome $(O_{iahbjk})$ constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
+All these variables, together with the outcome $(O_{iahbjk})$, constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
 Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
@@
 
-### The *conceptual population* data structure {#sec-theory-theoretical_P2}
+### The *conceptual-population* data structure {#sec-theory-theoretical_P2}
 
@@
 
-<!-- 3. segway for next section -->
-However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
-Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by certain individuals.
-
-<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
-Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
-However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
-More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
-[@Boonen_et_al_2020, p.
-5] -->
-
 ::: {#fig-cj14 layout-nrow=2}
@@
 :::
+
+<!-- 3. segway for next section -->
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by certain individuals.
+
+<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
+Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
+However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
+More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
+[@Boonen_et_al_2020, p.
+5] -->
 

### 2025-04-30 _tex/paper2.tex
Built site for gh-pages

@@
 prominent issues of traditional approaches that can undermine the
-reliability and validity of CJ's trait estimates
-\citep{Perron_et_al_2015}. First, it critiques the heavy reliance on
-Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
-extension, the Bradley-Terry-Luce (BTL) model
-\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
+reliability and validity of CJ's trait estimates. First, it critiques
+the heavy reliance on Thurstone's Case V assumptions
+\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
+model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
 Specifically, the study raises concerns about the assumptions of equal
@@
 Bayesian inference methods. In addition to potentially enhancing
-measurement reliability and validity \citep{Perron_et_al_2015}, and
-improving statistical accuracy in hypothesis testing, this approach
-offers two key advantages. First, it clarifies the interactions among
-all actors and processes involved in CJ experiments. Second, it shifts
-the current comparative data analysis paradigm from passively accepting
-the BTL model assumptions to actively testing whether those assumptions
-fit the data under analysis.
+measurement reliability and validity, and improving statistical accuracy
+in hypothesis testing, this approach offers two key advantages. First,
+it clarifies the interactions among all actors and processes involved in
+CJ experiments. Second, it shifts the current comparative data analysis
+paradigm from passively accepting the BTL model assumptions to actively
+testing whether those assumptions fit the data under analysis.
 
@@
 Normal distribution \citep{Andrich_1978}. In this distribution, the mode
-(mean) represents the relative separation between the stimuli, and its
-dispersion indicates the variability of that separation.
+(mean) represents the average relative separation between the stimuli,
+and its dispersion indicates the variability of that separation.
 
 Figure~\ref{fig-discriminal_difference} illustrates the distribution of
-the discriminal difference for the hypothetical texts depicted in
-Figure~\ref{fig-discriminal_process}. The figure indicates that the
-judge perceives Text B as having significantly higher quality than Text
-A. Two key observations support this conclusion: the positive difference
-between their modal discriminal processes \((T_{B} - T_{A} > 0)\) and
-the probability area where the discriminal difference distinctly favors
-Text B over Text A, represented by the shaded gray area denoted as
-\(P(B > A)\). As a result, the dichotomous outcome of this comparison is
-more likely to favor Text B over Text A.
+the discriminal difference for the two hypothetical texts. The figure
+indicates that the judge perceives Text B as having significantly higher
+quality than Text A. Two key observations support this conclusion: the
+positive difference between their modal discriminal processes
+\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
+difference distinctly favors Text B over Text A, represented by the
+shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
+outcome of this comparison is more likely to favor Text B over Text A.
 
@@
 over precise trait measurement, and without providing clear guidance on
-how to use its measurement scores for hypothesis testing. Specifically,
-he cautioned that its use ``should not be made without (an) experimental
-test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
-extensive set of simplifying assumptions
+how to use Case V's measurement scores for hypothesis testing.
+Specifically, he cautioned that its use ``should not be made without
+(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as the case
+imposes the most extensive set of simplifying assumptions
 \citep{Bramley_2008, Kelly_et_al_2022} (see
@@
 markedly from the rest of the sample \citep{Grubbs_1969}. In either
-case, the comparison of these texts results in more judgment
-discrepancies, yet the BTL model neither accounts for these cases nor
-offers any means to address them, except their elimination from the
-analysis.
+case, the BTL model neither accounts for these cases nor offers any
+means to address them, except their elimination from the analysis.
 
@@
 may unintentionally discard valuable information, introducing bias into
-the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
-The direction and magnitude of these biases remain unpredictable, as
-they depend on which stimuli researchers exclude from the analysis.
+the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
+direction and magnitude of these biases remain unpredictable, as they
+depend on which stimuli researchers exclude from the analysis.
 
@@
 
-The correlation \(\rho\) measures how much the judges' perception of a
-specific trait in one stimulus depends on their perception of the same
-trait in another. As with the discriminal dispersions, this correlation
-shapes the distribution of the discriminal difference, directly
-impacting the outcomes of pairwise comparisons. A similar thought
-experiment, as the one depicted in Section~\ref{sec-theory-issue1a}, can
-illustrate this idea. The experiment only assumes that the discriminal
-dispersions for both texts remain constant. Figure~\ref{fig-correlation}
-reveals that as the correlation between the texts increases, the
-distribution of their discriminal difference becomes narrower. This
-narrowing affects the area under the curve where the discriminal
-difference distinctly favors Text B over Text A, denoted as
-\(P(B > A)\), thus influencing the comparison outcome. Furthermore, the
-figure shows that when two texts are independent or uncorrelated
-\((\rho=0)\), their discriminal difference is less narrow compared to
-scenarios where the texts are positively correlated. As a result, the
-discriminal difference is less likely to favor Text B over Text A, as it
-is represented by the shaded gray area.
-
-Thurstone assumed that stimuli were uncorrelated because judges' biases,
-arising from two opposing and equally weighted effects occurring during
-the pairwise comparisons, canceled each other out
-\citep{Thurstone_1927b}. \citet{Andrich_1978} provided a mathematical
-demonstration of this cancellation using the BTL model under the
-assumption of discriminal processes with additive biases. However,
-evidence from the CJ literature indicates that the assumption of zero
-correlation does not hold in practice in at least two scenarios: when
-intricate aspects of multidimensional, complex traits or heterogeneous
-stimuli influence judges' perceptions or when additional hierarchical
-structures are relevant to the stimuli.
+The correlation \(\rho\), illustrated in
+Figure~\ref{fig-discriminal_difference}, measures how much the judges'
+perception of a specific trait in one stimulus depends on their
+perception of the same trait in another. Similar to the discriminal
+dispersions, this correlation shapes the distribution of the discriminal
+difference, directly impacting the outcomes of pairwise comparisons. A
+thought experiment, akin to the one presented in
+Section~\ref{sec-theory-issue1a}, can illustrate this idea. Assuming
+that the discriminal dispersions for both texts remain constant,
+Figure~\ref{fig-correlation} shows that as the correlation between the
+two texts increases, the distribution of their discriminal difference
+becomes narrower. This narrowing, in turn, affects the probability that
+the discriminal difference distinctly favors Text B over Text A--denoted
+as \(P(B > A)\)--and thus directly influences the comparison outcome.
+Furthermore, the figure shows that when two texts are independent or
+uncorrelated, as assumed in the BTL model \((\rho=0)\), the distribution
+of their discriminal difference is less narrow than in scenarios where
+the texts are positively correlated. As a result, it becomes less likely
+for the comparison to favor Text B over Text A, as indicated by the
+larger shaded area.
+
+Despite these notable differences in the distribution of the discriminal
+difference under various correlational assumptions, in practice,
+experimental designs often adopt the assumption of no correlation
+between stimuli based on an old theoretical justification. Specifically,
+\citet{Thurstone_1927b} argued that stimuli could be treated as
+uncorrelated because judges' biases--arising from two opposing and
+equally weighted effects occurring during the pairwise
+comparisons--would cancel each other out. This idea was later formalized
+by \citet{Andrich_1978}, who provided a mathematical demonstration of
+this cancellation using the BTL model under the assumption of
+discriminal processes with additive biases. However, evidence from the
+CJ literature indicates that the assumption of zero correlation does not
+hold in practice in at least two scenarios: when intricate aspects of
+multidimensional, complex traits or heterogeneous stimuli influence
+judges' perceptions or when additional hierarchical structures are
+relevant to the stimuli.
 
@@
 introduced by additional hierarchical structures may create dependencies
-between stimuli, a statistical phenomenon known as clustering
-\citep{Everitt_et_al_2010}. For instance, when multiple texts are
-produced by the same individual, these texts are likely to share common
-features, such as writing style or even their general quality. Although
-the CJ literature acknowledges the existence of such hierarchical
-structures, the statistical approaches to account for this additional
-source of dependence have been insufficient. For instance, when CJ data
-incorporates multiple samples of stimuli from the same individuals,
-researchers frequently rely on (averaged) point estimates of the BTL
-scores to conduct subsequent analyses and tests at the individual level
+between stimuli--a statistical phenomenon known as clustering
+\citep{Everitt_et_al_2010}. For instance, when the same individual
+produces multiple texts, those texts often exhibit common features, such
+as writing style or overall quality, that judges can easily recognize.
+In this regard, although the CJ literature acknowledges the existence of
+such hierarchical structures, the statistical approaches to account for
+this additional source of dependence have been insufficient. For
+instance, when CJ data incorporates multiple samples of stimuli from the
+same individuals, researchers frequently rely on (averaged) point
+estimates of the BTL scores to conduct subsequent analyses and tests at
+the individual level
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
 However, this approach can introduce additional statistical and
-measurement issues, which we discuss in greater detail in
+measurement issues, similar to the ones discussed in
 Section~\ref{sec-theory-issue2}.
@@
 
-Finally, as discussed in the previous section, researchers risk
+Finally, as discussed in the previous section, removing \emph{misfit}
+judges--that is, judges whose assessments deviate markedly from the
+shared consensus
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
+and considered outliers under the BTL model \citep{Grubbs_1969}--risk
 discarding valuable information and introducing bias into trait
-estimates when they remove \emph{misfit} judges--that is, judges whose
-assessments deviate markedly from the shared consensus
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
-considered outliers under the BTL model \citep{Grubbs_1969}. The
-direction and magnitude of these biases remain unpredictable, as they
-depend on which judges researchers exclude from the analysis
-\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
+estimates. The direction and magnitude of these biases remain
+unpredictable, as they depend on which judges researchers exclude from
+the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
 
@@
 \citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
-these issues, demonstrating how neglecting measurement error
-\((\sigma_{T})\) by relying only on outcome averages can reduce the
-precision of a predictor's estimated effect.
+these issues, showing that neglecting measurement error
+\((\sigma_{T})\)--by relying solely on outcome averages--reduces the
+precision of a predictor's estimated effect \(\beta_{X}\), as indicated
+by the solid lines.
 
@@
 accurate representation of the data-generating process underlying the
-pairwise comparisons.
-
-Notably, all these variables and the outcome \((O_{iahbjk})\) constitute
-the preliminary set of endogenous variables,
-\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,
-the SCM presents the preliminary set of structural equations,
+pairwise comparisons. All these variables, together with the outcome
+\((O_{iahbjk})\), constitute the preliminary set of endogenous
+variables, \(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\).
+Finally, the SCM presents the preliminary set of structural equations,
 \(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
@@
 
-\subsubsection{\texorpdfstring{The \emph{conceptual population} data
-structure}{The conceptual population data structure}}\label{sec-theory-theoretical_P2}
+\subsubsection{\texorpdfstring{The \emph{conceptual-population} data
+structure}{The conceptual-population data structure}}\label{sec-theory-theoretical_P2}
 
@@
 
-However, due to concerns about the practical feasibility of the
-comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
-implement an exhaustive pairings of sampled judges, stimuli, and
-individuals. Thus, a realistic scenario must account for the fact that
-judges typically compare only a subset of stimuli authored by certain
-individuals.
-
 \begin{figure}[H]
@@
 \end{figure}%
+
+However, due to concerns about the practical feasibility of the
+comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
+implement an exhaustive pairings of sampled judges, stimuli, and
+individuals. Thus, a realistic scenario must account for the fact that
+judges typically compare only a subset of stimuli authored by certain
+individuals.
 

### 2025-04-30 code/CJfig.R

@@ -34,7 +34,7 @@ sim_me = function( n=100, b=0.3, s1=1, s2=0.8){
   # regressions
   reg1 = lm( y ~ x, data=d )
   reg2 = lm( y_star ~ x, data=d )
-  
+
   # saving coefficients
   res = c( coef( reg1 )[2], coef( reg2 )[2] )
   
@@ -47,13 +47,14 @@ sim_me = function( n=100, b=0.3, s1=1, s2=0.8){
 
 # figure 1 ####
 
+## figure 1a ####
+
 # distributional parameters
 dpA = c(0, 0.5) # distribution text A (mean, sd)
 dpB = c(1, 1) # distribution text B (mean, sd)
 rAB = 0 # correlation between discriminal processes
 
 
-# figure
 png( filename=file.path(main_dir, 'discriminal_process.png'), 
      height=10, width=15, units='cm', res=400  )
 
@@ -95,9 +96,331 @@ dev.off()
 
 
 
+## figure 1b ####
+
+# distributional parameters
+dd = c( dpB[1]-dpA[1], sqrt( dpB[2]^2 + dpA[2]^2 - 2*rAB*dpB[2]*dpA[2] ) )
+
+# figure
+png( filename=file.path(main_dir, 'discriminal_difference.png'), 
+     height=10, width=15, units='cm', res=400  )
+
+# initial figure
+par( mar=c(2,0,0.5,0) )
+plot( NULL, xlim=c(-2,5), ylim=c(0,0.4), 
+      xaxt='n', yaxt='n', xlab='', ylab='', axes=F )
+axis( side=1, labels=F, lwd.ticks=0 )
+axis( side=1, at=c(0, dd[1]), tick=F, 
+      label=c(0, expression( T[B] - T[A] ) ) )
+
+# figure parameters
+pplotD = c( dnorm( 0, mean=dd[1], sd=dd[2] ),
+            dnorm( dd[1], mean=dd[1], sd=dd[2] ),
+            dnorm( dd[1]+2*dd[2], mean=dd[1], sd=dd[2]),
+            dnorm( dd[1], mean=dd[1], sd=dd[2] ) + 0.05,
[…]

### 2025-04-30 paper2.qmd
section:all; process:write

@@
 Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other endogenous and exogenous variables.
-These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
+These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to statistical (conditional) independence.
 
@@
 The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli).
-The index $j$ indicates the judge conducting the comparison, while the index $k$ captures *repeated measures designs* [@Lawson_2015, pp.
-366-376], accounting for experimental conditions where a judge compares the same pair of stimuli multiple times.
-Thus, the indexing system supports comparisons between different texts written by the same student $(i = h;$ $a \neq b)$ and between texts written by distinct students $(i \neq h;$ where $a = b$ is permitted$)$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$.
+The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for experimental conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, pp.
+366-376].
+Thus, the indexing system supports comparisons between different texts written by the same student $(i = h;$ $a \neq b)$ and between texts written by distinct students $(i \neq h;$ where $a = b$ is permitted$)$, each compared once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$.
 However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(i = h;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K};$ where $n_{J}>1$ and $n_{K}\geq1)$, as such comparison lacks practical relevance within the CJ framework.
@@
 Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons.
-All these variables, together with the outcome $(O_{iahbjk})$, constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$.
-Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.
+All these variables constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$, while the preliminary set of structural equations, $F = { f_{O}, f_{D} }$, captures the non-parametric dependencies among them.
 
@@
 
-It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$.
-In this simple case, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
+It is easier to visualize the structure of these vectors and matrices by considering an example.
+Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
 $$
@@
 Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
-The model's hierarchical design allows researchers to formulate and test hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for further discussion of these considerations).
+The model's hierarchical design allows researchers to formulate and test hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for a discussion of these considerations).
 
@@
 <!-- 2 Explaining what we did -->
-The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $(X_{I})$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $(T_{I})$.
+The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $X_{I}$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $T_{I}$.
 The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$.
 Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ independent columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
-Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $(X_{IA})$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $(T_{IA})$, the first primary latent variable of interest.
+Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $X_{IA}$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $T_{IA}$, the first primary latent variable of interest.
 The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
@@
 
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $(Z_{J})$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $(B_{J})$.
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $Z_{J}$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $B_{J}$.
 The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
 Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $Z_{JK}$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
@@
 Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
-Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$.
+Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$ and the absence of connecting arrows.
 
@@
 Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled repeated judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.
-Notice that because the design systematically excludes every third repeated judgment, researchers can also express $S_{JK}$ using $n_{K} = n^{s}_{K} = 2$.
+Notice that in this example, because the design systematically excludes every third repeated judgment, researchers can also express $S_{JK}$ using $n_{K} = n^{s}_{K} = 2$.
 
@@
 Essentially, $S$ serves as the vector selecting *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-This process generates the sample-comparison outcome $O^{sc}_{R}$.
 
@@
 <!-- 2. What is the effect of the comparison mechanism -->
-The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model, ultimately deriving the outcome $O_{R}$.
+The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model.
 It shows the sample-comparison outcome as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism.
@@
 In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli.
-Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow for accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp.
+Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow the accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp.
 267].
@@
 Notably, [DAG -@fig-cj14_dag] also shows that $C$ is independent of all other variables in the model.
-This independence implies that the conceptual model represented by the DAG applies exclusively to Random Allocation Comparative Designs [@Bramley_2015] or Incomplete Block Designs [@Lawson_2015].
-In these designs, every repeated judgment has an equal probability of being included in the sample.
+This independence implies that the conceptual model represented by the DAG applies exclusively to Random Allocation Comparative Designs [@Bramley_2015], or Incomplete Block Designs [@Lawson_2015], where every repeated judgment has an equal probability of being included in the sample.
 
@@
 
-# From SCM to a statistical model  {#sec-statistical}
+# From SCM to statistical model  {#sec-statistical}
 
@@
 <!-- 2. assumptions about the outcome -->
-Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_scm], and the statistical model [-@fig-cj16_scm].
+Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_prob], and the statistical model [-@fig-cj16_stat].
 First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
@@
 
-[^7]: The binomial distribution--including its special case, the Bernoulli distribution-- represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
+[^7]: The binomial distribution--including its special case, the Bernoulli distribution--represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
 34].
@@
 310].
-For a detailed discussion of the binomial as a maximum entropy distribution, see @McElreath_2020 [chap.
+For a detailed discussion of the binomial as a maximum entropy distribution, see @McElreath_2020 [sec.
 10.1.2].
@@
 *Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $X_{I}$ or $S$), resulting in spurious associations between the two [@Everitt_et_al_2010].
-Randomization addresses this issue by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself.
+Randomization ensure the absence of confounding by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself.
 For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.
@@
 <!-- 4. functional form for D_{R} -->
-Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
-The discriminal processes represent the underlying written-quality trait of the texts.
+Second, we define $D_{R}$ as the difference between the discriminal processes of two texts--$T_{IA}[i, a]$ and $T_{IA}[h, b]$--representing the underlying written-quality trait of the texts, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
 Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
@@
 In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
-It is important to clarify that the square brackets in $D_{R}$ merely indicate the relevant indices for each trait vector, without implying any subsetting of the data.
+To clarify, the square brackets in $D_{R}$ simply indicate the relevant indices for each trait vector and do not imply any subsetting of the data.
 
@@
 However, we consider the standard deviation, $s_{XI}$, varies depending on the teaching method group to which each student belongs.
-Given the example in @sec-theoretical, where the teaching method $X_{I} = 1,2$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$.
+Given the example in @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$.
 Note that this parametrization of $e_{I}$ relaxes the equal dispersion assumption for the stimuli, effectively addressing the concerns raised in @sec-theory-issue1a.
@@
 Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs.
-For instance, if $Z_{J}$ is a variable that can take on three values representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
+For instance, if $Z_{J} = \{1,2,3\}$ representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
 
@@
 Bayesian inference offers three key advantages for this purpose.
-First, it handles complex and overparameterized models--such as the one we implement--where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999].
+First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999]--such as the one we implement.
 Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
@@
 The **Declarations** section of this document provides a link to the model code, along with an alternative specification that assumes equal discriminal dispersions.
-We implemented both versions of the model using `Stan` [@Stan_2020, version 2.26.1].
-Although the implementations were successful, we do not report the results here, as they fall outside the scope of this document.
+We tested both versions of the model with success using `Stan` [@Stan_2020, version 2.26.1].
 

### 2025-04-30 _tex/paper2.tex
Built site for gh-pages

@@
 and the symbol `\(\:\bot\:\)' to represent \emph{d-separation}, a
-concept akin to (conditional) independence.
+concept akin to statistical (conditional) independence.
 
@@
 comparison (i.e., the stimuli). The index \(j\) indicates the judge
-conducting the comparison, while the index \(k\) captures \emph{repeated
-measures designs} \citep[pp.~366-376]{Lawson_2015}, accounting for
+conducting the comparison, while the index \(k\) accounts for
 experimental conditions where a judge compares the same pair of stimuli
-multiple times. Thus, the indexing system supports comparisons between
-different texts written by the same student \((i = h;\) \(a \neq b)\)
-and between texts written by distinct students \((i \neq h;\) where
-\(a = b\) is permitted\()\), each judged once or repeatedly by all
-judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\) where \(n_{J}>1\)
-and \(n_{K}\geq1)\). However, it excludes cases where a judge compares a
-student's text to itself, whether once or multiple times \((i = h;\)
-\(a = b;\) \(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K};\) where
-\(n_{J}>1\) and \(n_{K}\geq1)\), as such comparison lacks practical
-relevance within the CJ framework. Here, \(n_{J}\) indicates the total
-number of judges, and \(n_{K}\) denotes the number of repeated judgments
-each judge performs.
+multiple times, i.e., a \emph{repeated measures designs}
+\citep[pp.~366-376]{Lawson_2015}. Thus, the indexing system supports
+comparisons between different texts written by the same student
+\((i = h;\) \(a \neq b)\) and between texts written by distinct students
+\((i \neq h;\) where \(a = b\) is permitted\()\), each compared once or
+repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\)
+where \(n_{J}>1\) and \(n_{K}\geq1)\). However, it excludes cases where
+a judge compares a student's text to itself, whether once or multiple
+times \((i = h;\) \(a = b;\) \(j = 1,\dots,n_{J};\)
+\(k = 1,\dots,n_{K};\) where \(n_{J}>1\) and \(n_{K}\geq1)\), as such
+comparison lacks practical relevance within the CJ framework. Here,
+\(n_{J}\) indicates the total number of judges, and \(n_{K}\) denotes
+the number of repeated judgments each judge performs.
 
@@
 accurate representation of the data-generating process underlying the
-pairwise comparisons. All these variables, together with the outcome
-\((O_{iahbjk})\), constitute the preliminary set of endogenous
-variables, \(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\).
-Finally, the SCM presents the preliminary set of structural equations,
-\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
-among these variables.
+pairwise comparisons. All these variables constitute the preliminary set
+of endogenous variables,
+\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), while the
+preliminary set of structural equations, \(F = { f_{O}, f_{D} }\),
+captures the non-parametric dependencies among them.
 
@@
 It is easier to visualize the structure of these vectors and matrices by
-considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
-\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
-matrices described in equations (\ref{eq-mat11}) and (\ref{eq-mat12})
-take the following form:
+considering an example. Assuming \(n_{I} = 5\), \(n_{A} = 2\),
+\(n_{J} = 3\), and \(n_{K} = 3\), the vectors and matrices described in
+equations (\ref{eq-mat11}) and (\ref{eq-mat12}) take the following form:
 \begin{equation}\phantomsection\label{eq-mat13}{
@@
 estimation (see Section~\ref{sec-theory-issue1b} and
-Section~\ref{sec-theory-issue2} for further discussion of these
+Section~\ref{sec-theory-issue2} for a discussion of these
 considerations).
@@
   \citep{Pearl_2009}. Refer also to footnote 1.} student-related
-variables \((X_{I})\), such as teaching method, and students'
+variables \(X_{I}\), such as teaching method, and students'
 idiosyncratic errors \(e_{I}\) causally influence the latent variable
-representing students' writing-quality trait \((T_{I})\). The error term
+representing students' writing-quality trait \(T_{I}\). The error term
 \(e_{I}\) captures variations in students' traits unexplained by
@@
 branch shows how \(T_{I}\), along with \emph{relevant} \footnote{refer
-  to footnote 2.} text-related variables \((X_{IA})\) (e.g., text
-length), and texts' idiosyncratic errors \(e_{IA}\) causally influence
-the texts' written-quality trait \((T_{IA})\), the first primary latent
-variable of interest. The error term \(e_{IA}\) captures variations in
-the texts' traits that remain unexplained by \(T_{I}\) or \(X_{IA}\).
-Here, \(X_{IA}\) is an observed matrix with dimensions
-\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) independent columns
-(variables), while \(e_{IA}\) and \(T_{IA}\) are latent matrices with
-\(n_{I}\) rows and \(n_{A}\) columns.
+  to footnote 2.} text-related variables \(X_{IA}\) (e.g., text length),
+and texts' idiosyncratic errors \(e_{IA}\) causally influence the texts'
+written-quality trait \(T_{IA}\), the first primary latent variable of
+interest. The error term \(e_{IA}\) captures variations in the texts'
+traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
+\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
+rows and \(q_{IA}\) independent columns (variables), while \(e_{IA}\)
+and \(T_{IA}\) are latent matrices with \(n_{I}\) rows and \(n_{A}\)
+columns.
 
@@
 second component, where \emph{relevant} \footnote{refer to footnote 2.}
-judge-related variables \((Z_{J})\), such as judgment expertise, and
+judge-related variables \(Z_{J}\), such as judgment expertise, and
 judges' idiosyncratic errors \(e_{J}\) causally influence the latent
-variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
+variable representing judges' bias \(B_{J}\). The error \(e_{J}\)
 captures variations in judges' bias unexplained by \(Z_{J}\). Here,
@@
 \(B_{J}\), along with \emph{relevant} \footnote{refer to footnote 2.}
-judgment-related variables \((Z_{JK})\) (e.g., the number of judgments a
+judgment-related variables \(Z_{JK}\) (e.g., the number of judgments a
 judge makes), and judgments' idiosyncratic errors \(e_{JK}\) causally
@@
 the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
-\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).
+\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\) and
+the absence of connecting arrows.
 
@@
 \(S_{JK}\) represents the total sampled repeated judgments across all
-sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that because the
-design systematically excludes every third repeated judgment,
-researchers can also express \(S_{JK}\) using \(n_{K} = n^{s}_{K} = 2\).
+sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that in this
+example, because the design systematically excludes every third repeated
+judgment, researchers can also express \(S_{JK}\) using
+\(n_{K} = n^{s}_{K} = 2\).
 
@@
 vector selecting \emph{all repeated judgments a subset of judges makes
-for the subset of stimuli produced by the sampled individuals}. This
-process generates the sample-comparison outcome \(O^{sc}_{R}\).
+for the subset of stimuli produced by the sampled individuals}.
 
@@
 The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
-mechanism} \(C\) into the conceptual-population model, ultimately
-deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
-as unobserved, highlighting that researchers cannot directly access to
-\(O^{sc}_{R}\) due to the comparison mechanism. Moreover, the DAG
-depicts \(C\) as a causal factor influencing \(O_{R}\), enclosing it in
-a square to indicate that it is a conditioned variable. Specifically,
-\(C\) determines \emph{which repeated judgments the subset of judges
-makes for the subset of stimuli produced by the sampled individuals}. In
-essence, \(C\) reflects the assumption that judges \emph{do not} perform
-all possible repeated judgments for the selected stimuli. Instead, they
-carry out a ``sufficient number of comparisons'' \(n_{C}\) to allow for
-accurate estimation of the proportion \(P(B>A)\) for each stimulus pair
+mechanism} \(C\) into the conceptual-population model. It shows the
+sample-comparison outcome as unobserved, highlighting that researchers
+cannot directly access to \(O^{sc}_{R}\) due to the comparison
+mechanism. Moreover, the DAG depicts \(C\) as a causal factor
+influencing \(O_{R}\), enclosing it in a square to indicate that it is a
+conditioned variable. Specifically, \(C\) determines \emph{which
+repeated judgments the subset of judges makes for the subset of stimuli
+produced by the sampled individuals}. In essence, \(C\) reflects the
+assumption that judges \emph{do not} perform all possible repeated
+judgments for the selected stimuli. Instead, they carry out a
+``sufficient number of comparisons'' \(n_{C}\) to allow the accurate
+estimation of the proportion \(P(B>A)\) for each stimulus pair
 \citep[pp.~267]{Thurstone_1927b}.
@@
 conceptual model represented by the DAG applies exclusively to Random
-Allocation Comparative Designs \citep{Bramley_2015} or Incomplete Block
-Designs \citep{Lawson_2015}. In these designs, every repeated judgment
-has an equal probability of being included in the sample.
+Allocation Comparative Designs \citep{Bramley_2015}, or Incomplete Block
+Designs \citep{Lawson_2015}, where every repeated judgment has an equal
+probability of being included in the sample.
 
@@
 
-\section{From SCM to a statistical model}\label{sec-statistical}
+\section{From SCM to statistical model}\label{sec-statistical}
 
@@
 which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
-\ref{fig-cj16_scm}, and the statistical model \ref{fig-cj16_scm}. First,
-we assume that \(O_{R}\) follows a Bernoulli distribution\footnote{The
-  binomial distribution--including its special case, the Bernoulli
-  distribution-- represent a maximum entropy distribution for binary
-  events \citep[pp.~34]{McElreath_2020}. This means that when only two
-  un-ordered outcomes are possible, and the expected numbers of each
-  type of event are assumed to be constant, then the binomial
-  distribution is the one that is most consistent with these constraints
-  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
-  binomial as a maximum entropy distribution, see
-  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
-of CJ outcomes. Furthermore, following the conventions of Generalized
+\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}.
+First, we assume that \(O_{R}\) follows a Bernoulli
+distribution\footnote{The binomial distribution--including its special
+  case, the Bernoulli distribution--represent a maximum entropy
+  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
+  means that when only two un-ordered outcomes are possible, and the
+  expected numbers of each type of event are assumed to be constant,
+  then the binomial distribution is the one that is most consistent with
+  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
+  discussion of the binomial as a maximum entropy distribution, see
+  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
+CJ outcomes. Furthermore, following the conventions of Generalized
 Linear Latent and Mixed Models (GLLAMMs)
@@
   resulting in spurious associations between the two
-  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
-  effectively decoupling the association between the variable of
-  interest and any other variable, except for the outcome itself. For a
-  more detailed discussion on the benefits of randomization, see
-  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
-  \citet{Hernan_et_al_2020}.}.
+  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
+  confounding by effectively decoupling the association between the
+  variable of interest and any other variable, except for the outcome
+  itself. For a more detailed discussion on the benefits of
+  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
+  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.
 
 Second, we define \(D_{R}\) as the difference between the discriminal
-processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), plus the
-corresponding repeated judge biases, \(B_{JK}[j, k]\). The discriminal
-processes represent the underlying written-quality trait of the texts.
-Note that if we assume that \(B_{JK}[j,k]\) reflects the difference in
-stimulus-specific biases, i.e.,
+processes of two texts--\(T_{IA}[i, a]\) and
+\(T_{IA}[h, b]\)--representing the underlying written-quality trait of
+the texts, plus the corresponding repeated judge biases,
+\(B_{JK}[j, k]\). Note that if we assume that \(B_{JK}[j,k]\) reflects
+the difference in stimulus-specific biases, i.e.,
 \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
@@
 terms, the discriminal processes of the stimuli become an observable
-outcome only through the judges' perceptions (biases). It is important
-to clarify that the square brackets in \(D_{R}\) merely indicate the
-relevant indices for each trait vector, without implying any subsetting
-of the data.
+outcome only through the judges' perceptions (biases). To clarify, the
+square brackets in \(D_{R}\) simply indicate the relevant indices for
+each trait vector and do not imply any subsetting of the data.
 
@@
 Section~\ref{sec-theoretical}, where the teaching method
-\(X_{I} = 1,2\), the model sets the following constraint to anchor the
-scale of the latent trait \citep{Depaoli_2021}:
+\(X_{I} = \{1,2\}\), the model sets the following constraint to anchor
+the scale of the latent trait \citep{Depaoli_2021}:
 \(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\). Note that this parametrization of
@@
 \(s_{ZJ}\) that depends on the groups to which each judge belongs. For
-instance, if \(Z_{J}\) is a variable that can take on three values
-representing three groups of judges with varying expertise, the model
-imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
-scale of the latent trait \citep{Depaoli_2021}.
+instance, if \(Z_{J} = \{1,2,3\}\) representing three groups of judges
+with varying expertise, the model imposes the constraint
+\(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the scale of the latent trait
+\citep{Depaoli_2021}.
 
@@
 advantages for this purpose. First, it handles complex and
-overparameterized models--such as the one we implement--where the number
-of parameters exceeds the number of observations
-\citep{Baker_1998, Kim_et_al_1999}. Second, it incorporates prior
-information to constrain parameter estimates within plausible bounds,
-thereby mitigating estimation issues like non-convergence or improper
-solutions that often affect frequentist methods
-\citep{Martin_et_al_1975, Seaman_et_al_2011}. In our implementation, we
-use priors to define the error distribution, which sets the scale of the
-latent variables \citep{Depaoli_2014}. Third, Bayesian inference
-supports robust inferences from small samples, where the asymptotic
-properties underlying frequentist methods are less reliable
-\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
-feature is particularly relevant in CJ experiments, where researchers
-often collect large volumes of paired comparisons but work with
-relatively small samples of judges, stimuli, and individuals to test
-hypotheses.
+overparameterized models, where the number of parameters exceeds the
+number of observations \citep{Baker_1998, Kim_et_al_1999}--such as the
+one we implement. Second, it incorporates prior information to constrain
+parameter estimates within plausible bounds, thereby mitigating
+estimation issues like non-convergence or improper solutions that often
+affect frequentist methods \citep{Martin_et_al_1975, Seaman_et_al_2011}.
+In our implementation, we use priors to define the error distribution,
+which sets the scale of the latent variables \citep{Depaoli_2014}.
+Third, Bayesian inference supports robust inferences from small samples,
+where the asymptotic properties underlying frequentist methods are less
+reliable \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
+This feature is particularly relevant in CJ experiments, where
+researchers often collect large volumes of paired comparisons but work
+with relatively small samples of judges, stimuli, and individuals to
+test hypotheses.
 
@@
 the model code, along with an alternative specification that assumes
-equal discriminal dispersions. We implemented both versions of the model
-using \texttt{Stan} \citep[version 2.26.1]{Stan_2020}. Although the
-implementations were successful, we do not report the results here, as
-they fall outside the scope of this document.
+equal discriminal dispersions. We tested both versions of the model with
+success using \texttt{Stan} \citep[version 2.26.1]{Stan_2020}.
 

### 2025-05-07 paper2.qmd
section:6; process:write

@@
 @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data.
-@sec-discussion discusses the findings, limitations, and challenges and explores avenues for future research.
-Finally, @sec-conclusion summarizes the study's conclusions.
+Finally, @sec-discussion discusses the findings, explores avenues for future research, and detail the challenges for future researchers.
 
@@
 The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.
+
+<!-- The object of this paper is to describe a new psychophysical law which may be called the law of comparative judgment and to show some of its special applications in the measurement of psychological values.
+[@Thurstone_1927b, pp.
+266] -->
 
@@
 Thurstone noted from the outset that his general form, described in @sec-thurstone_theory, led to a *trait scaling problem*.
-Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b, pp.
-267].
+Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
 For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases).
@@
 
+<!-- According to this equation every pair of stimuli presents the possibility of a different correlation between the discriminal deviations.
+If this degree of freedom is allowed, the problem of psychological scaling would be insoluble because every observation equation would introduce a new unknown and the number of unknowns would then always be greater that the number of observation equations.
+In order to make the problem soluble it is necessary to make at least one assumption, namely that the correlation between discriminal deviations is practically constant throughout the stimulus series and for the single observer.
+[@Thurstone_1927b, pp.
+267] -->
+
 <!-- 2. Thurstone's Cases and their simplifying assumptions -->
-To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions.
+To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions [@Thurstone_1927b].
 In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons.
@@
 
+<!-- For the practical application of the law of comparative judgment we shall consider five cases which differ in assumptions, approximations, and degree of simplification.
+The more assumptions we care to make, the simpler will be the observation equations.
+[@Thurstone_1927b, pp.
+267] -->
+
 <!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->
@@
 <!-- 3. But are our specific problems? -->
-However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use Case V's measurement scores for hypothesis testing.
-Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
-270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases).
-Moreover, because Thurstone's primary goal was to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp.
-269] and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp.
-269], he did not address how to use the resulting measurement scores for hypothesis testing.
-Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978].
-Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?
+However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement and offering no guidance on how to use its trait estimates for statistical inference or hypothesis testing.
+Specifically, Thurstone cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+270], as it imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases).
+Moreover, because Thurstone's primary goal was to produce a "rather coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp.
+269], his theory did not support formal statistical inference.
+Despite these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, which raises significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978].
+Furthermore, although the CJ tradition has attempted to address the gap of hypothesis testing by using the point estimates of the traits--or their transformations--a critical question remains: Does this approach provide a valid foundation for statistical inference?
 
@@
 As previously discussed, Case V remains the most widely used model in CJ literature.
-This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case.
-The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases).
-However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021].
-Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.
+This preference largely stems from the widespread adoption of the BTL model, which provides a simplified statistical representation of the case.
+The BTL model mirrors most of Case V's assumptions, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases).
+However, this substitution has minimal impact on trait estimation or model interpretation because the scale of the discriminal process (i.e., the latent trait) is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021].
+That is, as long as the substitution (transformation) preserves the data rank order, the choice of distribution for the discriminal processes is inconsequential.
+This condition is satisfied in this case, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I].
 
@@
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
+Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics, as seen in @Pollitt_2012a, @Pollitt_2012b, @vanDaal_et_al_2016, and @Goossens_et_al_2018, may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
 The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
@@
 <!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Building on the previous section, it is clear that, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b].
-Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests.
+Building on the previous section, it is clear that, researchers in CJ studies typically use the BTL model to measure traits and position the compared stimuli along a latent continuum [@Thurstone_1927b].
+The CJ literature also show that they frequently rely on point estimates of these traits (i.e, the BTL scores)--or their transformations--to conduct statistical inference or hypothesis testing.
 For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 <!-- 3. But what are the challenges in using BTL scores for further analysis? -->
-Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions.
+Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct statistical inference or hypothesis tests, as this practice can undermine the resulting statistical conclusions.
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error).
@@
 Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].
-@fig-measurement_error illustrates these issues, showing that neglecting measurement error $(\sigma_{T})$--by relying solely on outcome averages--reduces the precision of a predictor's estimated effect $\beta_{X}$, as indicated by the solid lines.
 
@@
 
-![The effect of outcome uncertainty $\sigma_{T}$ on the estimation of an effect $\beta_{X}$ linked to the predictor.
-The example assumes a sample size $n=100$ and an uncertainty that increases from left to right.](/images/png/measurement_error.png){#fig-measurement_error width=100%}
+<!-- @fig-measurement_error illustrates these issues, showing that neglecting measurement error $(\sigma_{T})$--by relying solely on outcome averages--reduces the precision of a predictor's estimated effect $\beta_{X}$, as indicated by the solid lines.
+-->
+
+<!-- ![The effect of outcome uncertainty $\sigma_{T}$ on the estimation of an effect $\beta_{X}$ linked to the predictor.
+The example assumes a sample size $n=100$ and an uncertainty that increases from left to right.](/images/png/measurement_error.png){#fig-measurement_error width=100%} -->
 
@@
 Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons.
-All these variables constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$, while the preliminary set of structural equations, $F = { f_{O}, f_{D} }$, captures the non-parametric dependencies among them.
+This model defines the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$, and the preliminary set of structural equations, $F = { f_{O}, f_{D} }$, which capture the non-parametric dependencies among these variables.
 
@@
 Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
-Thus, this matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
+This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
 Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
@@
 
-<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
-C -> {Y}
-e_{A} -> {C, Y} 
-this means that e_{A} is a confounder?,  
-then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?.
-Does this mean that random algorithms per se will always be unbiased?.
--->
-
 <!-- 1. start by defining the sampling vector and matrices  -->
@@
 In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model.
-These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli (see @sec-theory-theoretical_P3), incorporate measurement error in trait estimation and hypothesis testing (also see @sec-theory-theoretical_P3), and clarify the role of the sample and comparison mechanisms in CJ experiments (see @sec-theory-theoretical_SC).
+These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and clarify the role of the sample and comparison mechanisms in CJ experiments (see @sec-theory-theoretical_SC).
 However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a.
-Because this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.
+Since this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.
 
@@
 The most critical property, however, is the elimination of confounding.
-*Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $X_{I}$ or $S$), resulting in spurious associations between the two [@Everitt_et_al_2010].
+*Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $S$), resulting in spurious associations between the latter two [@Everitt_et_al_2010].
 Randomization ensure the absence of confounding by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself.
@@
 <!-- 4. functional form for D_{R} -->
-Second, we define $D_{R}$ as the difference between the discriminal processes of two texts--$T_{IA}[i, a]$ and $T_{IA}[h, b]$--representing the underlying written-quality trait of the texts, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
+Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the texts, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
 Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
@@
 However, we consider the standard deviation, $s_{XI}$, varies depending on the teaching method group to which each student belongs.
-Given the example in @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$.
+Given the example in @sec-theoretical where the teaching method $X_{I} = \{1,2\}$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$.
 Note that this parametrization of $e_{I}$ relaxes the equal dispersion assumption for the stimuli, effectively addressing the concerns raised in @sec-theory-issue1a.
@@
 Bayesian inference offers three key advantages for this purpose.
-First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999]--such as the one we implement.
+First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999]--such as the model we try to implement.
 Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
@@
 <!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->
-
-<!-- 1. General summary of what we have done. What are the first benefits? -->
-This study extends Thurstone's general form through a systematic, integrated approach that combines Causal and Bayesian inference methods.
-It begins by formulating the extension as a conceptual model, represented by an SCM and a DAG.
-This formulation integrates Thurstone's core theoretical principles, including the distribution of the discriminal process, with key CJ experimental design features, such as the sampling and comparison mechanisms.
+<!-- recording: Sven and Tine 25.04.22; time: 00:00:00 - 00:13:15 -->
+
+<!-- Follow the structure of the paper: -->
+<!-- 1. Talk about Thurstone -->
+<!-- 2. Talk about the issues with the model of Thurstone -->
+<!-- 3. Talk about what you did -->
+<!-- 4. Potential benefits and extras that you can do now -->
+<!-- 5. Challenges -->
+<!-- 6. Conclusion: what we did and why is a great paper -->
+
+<!-- "This explicit part of the model it open ups the idea of doing more throughout research on ..." -->
+
+<!-- How about power?, you have a framework that allows to do this and more. -->
+
+<!-- talk about S and C -->
+
+<!-- 1. Summary of Thurstone's general theory -->
+Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b].
+In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli.
+Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level.
+However, Thurstone identified one key issue in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
+To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from the general form, each progressively incorporating several simplifying assumptions.
+
+<!-- 2. Summary of the why the literature uses Case V -->
+Of these five cases, the CJ literature has predominantly relied on Case V to measure various psychological traits, as shown in studies like @Kimbell_2012, @Jones_et_al_2015, and @Boonen_et_al_2020.
+This preference largely stems from the widespread adoption of the BTL model, which offers a simplified statistical formulation of the case.
+The BTL model mirrors the assumptions of Case V--namely, equal discriminal dispersions and zero correlation for the stimuli' discriminal processes--with one notable distinction: while Case V assumes normally distributed discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008].
+
+<!-- 3. The issues with Case V, and the BTL model -->
+Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b].
+In particular, he noted that although these assumptions simplify the trait measurement model, they could fail to capture the complexity of CJ data [@Kelly_et_al_2022; @Andrich_1978], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
+
+Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and to "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp.
+269], his theory did not specify how to use the resulting trait estimates for statistical inference or hypothesis testing.
+The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing.
+Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct hypothesis tests.
+However, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+
+<!-- 4. General summary of what we have done. -->
+To address these issues, this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
+It begins by formulating a conceptual model represented by an SCM and a DAG.
+This model integrates Thurstone's core theoretical principles--such as the discriminal processes (i.e., traits) of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms.
 The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model.
-Specifically, the statistical model accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error in trait estimation and hypothesis testing, and accommodates violations of the equal dispersion assumption.
-These improvements potentially enhance measurement reliability and validity [@Perron_et_al_2015] and improve the statistical accuracy of hypothesis testing.
+In particular, it enables the analysis of CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when researchers seek to conduct statistical inference or hypothesis testing (refer to @sec-theory-issue2).
+These methodological innovations potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical conclusions.
 
@@
 
+For instance,
+
+<!-- accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error in trait estimation and hypothesis testing, and accommodates violations of the equal dispersion assumption.
+-->
+
 <!-- 3. First research avenue -->
-One promising research avenue involves investigating how the hierarchical structure of stimuli affects CJ experiment designs and hypotheses testing.
+<!-- One promising research avenue involves investigating how the hierarchical structure of stimuli affects CJ experiment designs and hypotheses testing.
 One such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy.
@@
 However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates.
-
-Relatedly, another study could investigate hypotheses about the characteristics of the texts themselves.
+-->
+
+<!-- Relatedly, another study could investigate hypotheses about the characteristics of the texts themselves.
 Previously, the literature lacked a direct and integrated method for evaluating these types of hypotheses.
 Now, by collecting data on specific stimulus features, researchers can assess whether these variables produce the expected effects.
+-->
 
@@
 
+<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
+C -> {Y}
+e_{A} -> {C, Y} 
+this means that e_{A} is a confounder?,  
+then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?.
+Does this mean that random algorithms per se will always be unbiased?.
+-->
+
 <!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
@@
 256] -->
+
+<!-- Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates.
+In this sense the models are also "sample-free."
+However, this property of the model is true in empirical data only if the model holds in the data.
+Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
+[@Andrich_1978, pp.459] -->
 
@@
 
+<!-- Challenges -->
+
+Despite its potential, this study presents several challenges that researchers must navigate.
+First, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate.
+In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest?
+However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces.
+Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.
+
+Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts.
+We have argued that the model's ability to estimate traits accurately and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
+That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.
+
+Third, applying this approach requires a solid understanding of Causal inference.
+For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity.
+Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.
+
+Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods.
+This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand.
+
 <!-- (do not do this yet!) Develop how you multiply the predictions by the inverse of the sample probability to get the estimates.
@@
 <!-- Develop how you get the vectors S and C and sampling weights when you know the extension of the population. (See https://mc-stan.org/docs/stan-users-guide/poststratification.html ) -->
-
-# Conclusion {#sec-conclusion}
 

### 2025-05-07 _tex/paper2.tex
Built site for gh-pages

@@
 elements into a probabilistic statistical model to analyze dichotomous
-pairwise comparison data. Section~\ref{sec-discussion} discusses the
-findings, limitations, and challenges and explores avenues for future
-research. Finally, Section~\ref{sec-conclusion} summarizes the study's
-conclusions.
+pairwise comparison data. Finally, Section~\ref{sec-discussion}
+discusses the findings, explores avenues for future research, and detail
+the challenges for future researchers.
 
@@
 parameters than the number of available pairwise comparisons
-\citep[pp.~267]{Thurstone_1927b}. For instance, in a CJ experiment with
-five texts, the general form would require estimating \(20\) parameters:
-five modal discriminal processes, five discriminal dispersions, and
-\(10\) correlations --one per comparison (see
-Table~\ref{tbl-thurstone_cases}). However, a single judge could only
-provide \({5 \choose 2} = 10\) unique comparisons, an insufficient data
-set to estimate the required parameters.
+\citep{Thurstone_1927b}. For instance, in a CJ experiment with five
+texts, the general form would require estimating \(20\) parameters: five
+modal discriminal processes, five discriminal dispersions, and \(10\)
+correlations --one per comparison (see Table~\ref{tbl-thurstone_cases}).
+However, a single judge could only provide \({5 \choose 2} = 10\) unique
+comparisons, an insufficient data set to estimate the required
+parameters.
 
@@
 theory, Thurstone developed five cases derived from this general form,
-each progressively incorporating additional simplifying assumptions. In
-Case I, Thurstone postulated that pairs of stimuli would maintain a
-constant correlation across all comparisons. In Case II, he allowed
-multiple judges to undertake comparisons instead of confining
-evaluations to a single judge. In Case III, he posited that there was no
-correlation between stimuli. In Case IV, he assumed that the stimuli
-exhibited similar dispersions. Finally, in Case V, he replaced this
-assumption with the condition that stimuli had equal discriminal
-dispersions. Table~\ref{tbl-thurstone_cases} summarizes the assumptions
-of the general form and the five cases. For a detailed discussion of
-these cases and their progression, refer to \citet{Thurstone_1927b} and
-\citet[pp.~248--253]{Bramley_2008}.
+each progressively incorporating additional simplifying assumptions
+\citep{Thurstone_1927b}. In Case I, Thurstone postulated that pairs of
+stimuli would maintain a constant correlation across all comparisons. In
+Case II, he allowed multiple judges to undertake comparisons instead of
+confining evaluations to a single judge. In Case III, he posited that
+there was no correlation between stimuli. In Case IV, he assumed that
+the stimuli exhibited similar dispersions. Finally, in Case V, he
+replaced this assumption with the condition that stimuli had equal
+discriminal dispersions. Table~\ref{tbl-thurstone_cases} summarizes the
+assumptions of the general form and the five cases. For a detailed
+discussion of these cases and their progression, refer to
+\citet{Thurstone_1927b} and \citet[pp.~248--253]{Bramley_2008}.
 
@@
 However, Thurstone developed Case V prioritizing statistical simplicity
-over precise trait measurement, and without providing clear guidance on
-how to use Case V's measurement scores for hypothesis testing.
-Specifically, he cautioned that its use ``should not be made without
-(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as the case
+over precise trait measurement and offering no guidance on how to use
+its trait estimates for statistical inference or hypothesis testing.
+Specifically, Thurstone cautioned that its use ``should not be made
+without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as it
 imposes the most extensive set of simplifying assumptions
@@
 Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
-goal was to provide a ``rather coarse scaling'' of traits
-\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
-this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
-how to use the resulting measurement scores for hypothesis testing.
-Thus, given these limitations, it is surprising that CJ research has
-predominantly relied on Case V to measure different traits, raising
-significant concerns about the reliability and validity of such
-measurements in contexts where the case's assumptions may not hold
-\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
-practice has attempted to address the issue of hypothesis testing by
-using the scores' point estimates or their transformations, a critical
-question remains: Is this approach suitable for addressing these
-inquiries?
+goal was to produce a ``rather coarse scaling'' of traits and ``allocate
+the compared stimuli on this continuum''
+\citep[pp.~269]{Thurstone_1927b}, his theory did not support formal
+statistical inference. Despite these limitations, it is surprising that
+CJ research has predominantly relied on Case V to measure different
+traits, which raises significant concerns about the reliability and
+validity of such measurements in contexts where the case's assumptions
+may not hold \citep{Kelly_et_al_2022, Andrich_1978}. Furthermore,
+although the CJ tradition has attempted to address the gap of hypothesis
+testing by using the point estimates of the traits--or their
+transformations--a critical question remains: Does this approach provide
+a valid foundation for statistical inference?
 
@@
 As previously discussed, Case V remains the most widely used model in CJ
-literature. This preference primarily stems from the BTL model, which
-provides a simplified statistical representation of the case. The BTL
-model mirrors the assumptions of Case V, with one notable distinction:
-while Case V assumes a Normal distribution for the stimuli' discriminal
-processes, the BTL model uses the more mathematically tractable Logistic
-distribution \citep{Andrich_1978, Bramley_2008} (see
+literature. This preference largely stems from the widespread adoption
+of the BTL model, which provides a simplified statistical representation
+of the case. The BTL model mirrors most of Case V's assumptions, with
+one notable distinction: while Case V assumes a Normal distribution for
+the stimuli' discriminal processes, the BTL model uses the more
+mathematically tractable Logistic distribution
+\citep{Andrich_1978, Bramley_2008} (see
 Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
-impact on the model's estimation or interpretation because the
-discriminal process scale is arbitrary up to a non-monotonic
-transformation \citep{vanderLinden_et_al_2017_I, McElreath_2021}.
-Furthermore, this limited impact is supported by the fact that the
-Normal and Logistic distributions exhibit analogous statistical
-properties, differing only by a scaling factor of approximately \(1.7\).
+impact on trait estimation or model interpretation because the scale of
+the discriminal process (i.e., the latent trait) is arbitrary up to a
+non-monotonic transformation
+\citep{vanderLinden_et_al_2017_I, McElreath_2021}. That is, as long as
+the substitution (transformation) preserves the data rank order, the
+choice of distribution for the discriminal processes is inconsequential.
+This condition is satisfied in this case, as the Normal and Logistic
+distributions exhibit analogous statistical properties, differing only
+by a scaling factor of approximately \(1.7\)
+\citep{vanderLinden_et_al_2017_I}.
 
@@
 dispersions or outlying observations, the common CJ practice of
-excluding stimuli based on these statistics
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
-may unintentionally discard valuable information, introducing bias into
-the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
-direction and magnitude of these biases remain unpredictable, as they
-depend on which stimuli researchers exclude from the analysis.
+excluding stimuli based on these statistics, as seen in
+\citet{Pollitt_2012a}, \citet{Pollitt_2012b},
+\citet{vanDaal_et_al_2016}, and \citet{Goossens_et_al_2018}, may
+unintentionally discard valuable information, introducing bias into the
+trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
+and magnitude of these biases remain unpredictable, as they depend on
+which stimuli researchers exclude from the analysis.
 
@@
 
-Building on the previous section, it is clear that, researchers
-typically rely on the BTL model to measure a trait and place the
-compared stimuli along its continuum \citep{Thurstone_1927b}.
-Additionally, the CJ literature shows that researchers frequently use
-point estimates of BTL scores or their transformations to conduct
-further analyses or hypothesis tests. For example, researchers have used
-these scores to identify `misfit' judges and stimuli
+Building on the previous section, it is clear that, researchers in CJ
+studies typically use the BTL model to measure traits and position the
+compared stimuli along a latent continuum \citep{Thurstone_1927b}. The
+CJ literature also show that they frequently rely on point estimates of
+these traits (i.e, the BTL scores)--or their transformations--to conduct
+statistical inference or hypothesis testing. For example, researchers
+have used these scores to identify `misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
@@
 literature cautions against relying solely on the point estimates of BTL
-scores to conduct further analyses or hypothesis tests, as this practice
-can undermine the resulting statistical conclusions. A key consideration
-is that BTL scores are parameter estimates that inherently carry
-uncertainty (measurement error). Ignoring this uncertainty can bias the
-analysis and reduce the precision of hypothesis tests. The direction and
-magnitude of such biases are often unpredictable. Results may be
-attenuated, exaggerated, or remain unaffected depending on the degree of
-uncertainty in the scores and the actual effects being tested
+scores to conduct statistical inference or hypothesis tests, as this
+practice can undermine the resulting statistical conclusions. A key
+consideration is that BTL scores are parameter estimates that inherently
+carry uncertainty (measurement error). Ignoring this uncertainty can
+bias the analysis and reduce the precision of hypothesis tests. The
+direction and magnitude of such biases are often unpredictable. Results
+may be attenuated, exaggerated, or remain unaffected depending on the
+degree of uncertainty in the scores and the actual effects being tested
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore,
@@
 power, increasing the likelihood of committing type-I or type-II errors
-\citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
-these issues, showing that neglecting measurement error
-\((\sigma_{T})\)--by relying solely on outcome averages--reduces the
-precision of a predictor's estimated effect \(\beta_{X}\), as indicated
-by the solid lines.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/measurement_error.png}
-
-}
-
-\caption{\label{fig-measurement_error}The effect of outcome uncertainty
-\(\sigma_{T}\) on the estimation of an effect \(\beta_{X}\) linked to
-the predictor. The example assumes a sample size \(n=100\) and an
-uncertainty that increases from left to right.}
-
-\end{figure}%
+\citep{McElreath_2020}.
 
@@
 accurate representation of the data-generating process underlying the
-pairwise comparisons. All these variables constitute the preliminary set
-of endogenous variables,
-\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), while the
-preliminary set of structural equations, \(F = { f_{O}, f_{D} }\),
-captures the non-parametric dependencies among them.
+pairwise comparisons. This model defines the preliminary set of
+endogenous variables,
+\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\), and the
+preliminary set of structural equations, \(F = { f_{O}, f_{D} }\), which
+capture the non-parametric dependencies among these variables.
 
@@
 Additionally, we construct the matrix \(R\) to map each row of the
-\(IA\) matrix with a corresponding row from the \(JK\) matrix. Thus,
-this matrix has \(n\) rows and \(6\) columns, where
+\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
+matrix has \(n\) rows and \(6\) columns, where
 \(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
@@
 Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
-structure of stimuli (see Section~\ref{sec-theory-theoretical_P3}),
-incorporate measurement error in trait estimation and hypothesis testing
-(also see Section~\ref{sec-theory-theoretical_P3}), and clarify the role
-of the sample and comparison mechanisms in CJ experiments (see
+structure of stimuli and incorporate measurement error in trait
+estimation and hypothesis testing (see
+Section~\ref{sec-theory-theoretical_P3}), and clarify the role of the
+sample and comparison mechanisms in CJ experiments (see
 Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
 concerns about the assumption of equal dispersions among stimuli
-discussed in Section~\ref{sec-theory-issue1a}. Because this concern
+discussed in Section~\ref{sec-theory-issue1a}. Since this concern
 relates to the statistical assumption underlying the distribution of the
@@
   \(X_{I}\), simultaneously influences both the outcome (e.g.,
-  \(O_{R}\)) and a variable of interest (e.g., \(X_{I}\) or \(S\)),
-  resulting in spurious associations between the two
+  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
+  spurious associations between the latter two
   \citep{Everitt_et_al_2010}. Randomization ensure the absence of
@@
 Second, we define \(D_{R}\) as the difference between the discriminal
-processes of two texts--\(T_{IA}[i, a]\) and
-\(T_{IA}[h, b]\)--representing the underlying written-quality trait of
-the texts, plus the corresponding repeated judge biases,
-\(B_{JK}[j, k]\). Note that if we assume that \(B_{JK}[j,k]\) reflects
-the difference in stimulus-specific biases, i.e.,
-\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
-discriminal difference as:
+processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\),
+representing the underlying written-quality trait of the texts, plus the
+corresponding repeated judge biases, \(B_{JK}[j, k]\). Note that if we
+assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
+biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
+can re-write the discriminal difference as:
 \begin{equation}\phantomsection\label{eq-mat52}{
@@
 method group to which each student belongs. Given the example in
-Section~\ref{sec-theoretical}, where the teaching method
+Section~\ref{sec-theoretical} where the teaching method
 \(X_{I} = \{1,2\}\), the model sets the following constraint to anchor
@@
 number of observations \citep{Baker_1998, Kim_et_al_1999}--such as the
-one we implement. Second, it incorporates prior information to constrain
-parameter estimates within plausible bounds, thereby mitigating
-estimation issues like non-convergence or improper solutions that often
-affect frequentist methods \citep{Martin_et_al_1975, Seaman_et_al_2011}.
-In our implementation, we use priors to define the error distribution,
-which sets the scale of the latent variables \citep{Depaoli_2014}.
-Third, Bayesian inference supports robust inferences from small samples,
-where the asymptotic properties underlying frequentist methods are less
-reliable \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
-This feature is particularly relevant in CJ experiments, where
-researchers often collect large volumes of paired comparisons but work
-with relatively small samples of judges, stimuli, and individuals to
-test hypotheses.
+model we try to implement. Second, it incorporates prior information to
+constrain parameter estimates within plausible bounds, thereby
+mitigating estimation issues like non-convergence or improper solutions
+that often affect frequentist methods
+\citep{Martin_et_al_1975, Seaman_et_al_2011}. In our implementation, we
+use priors to define the error distribution, which sets the scale of the
+latent variables \citep{Depaoli_2014}. Third, Bayesian inference
+supports robust inferences from small samples, where the asymptotic
+properties underlying frequentist methods are less reliable
+\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
+feature is particularly relevant in CJ experiments, where researchers
+often collect large volumes of paired comparisons but work with
+relatively small samples of judges, stimuli, and individuals to test
+hypotheses.
 
@@
 
-This study extends Thurstone's general form through a systematic,
-integrated approach that combines Causal and Bayesian inference methods.
-It begins by formulating the extension as a conceptual model,
-represented by an SCM and a DAG. This formulation integrates Thurstone's
-core theoretical principles, including the distribution of the
-discriminal process, with key CJ experimental design features, such as
-the sampling and comparison mechanisms. The study then translates the
-SCM into a bespoke statistical model that addresses key limitations of
-the BTL model. Specifically, the statistical model accounts for judge
-biases, captures the hierarchical structure of stimuli, incorporates
-measurement error in trait estimation and hypothesis testing, and
-accommodates violations of the equal dispersion assumption. These
-improvements potentially enhance measurement reliability and validity
-\citep{Perron_et_al_2015} and improve the statistical accuracy of
-hypothesis testing.
+Thurstone introduced the Law of Comparative Judgment to measure
+psychological traits of stimuli through pairwise comparisons
+\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
+theory models single-judge comparisons across multiple, potentially
+correlated stimuli. Each comparison produces a dichotomous outcome
+indicating which stimulus the judge perceives as having a higher trait
+level. However, Thurstone identified one key issue in this model: it
+required estimating more ``unknown'' parameters than the number of
+available pairwise comparisons \citep{Thurstone_1927b}. To address this
+issue and facilitate the theory's practical applicability, he formulated
+five cases derived from the general form, each progressively
+incorporating several simplifying assumptions.
+
+Of these five cases, the CJ literature has predominantly relied on Case
+V to measure various psychological traits, as shown in studies like
+\citet{Kimbell_2012}, \citet{Jones_et_al_2015}, and
+\citet{Boonen_et_al_2020}. This preference largely stems from the
+widespread adoption of the BTL model, which offers a simplified
+statistical formulation of the case. The BTL model mirrors the
+assumptions of Case V--namely, equal discriminal dispersions and zero
+correlation for the stimuli' discriminal processes--with one notable
+distinction: while Case V assumes normally distributed discriminal
+processes, the BTL model uses the more mathematically tractable Logistic
+distribution \citep{Andrich_1978, Bramley_2008}.
+
+Although replacing the Normal with the Logistic distribution has a
+minimal impact on trait estimation or model interpretation
+\citep{vanderLinden_et_al_2017_I, McElreath_2021}, Thurstone
+acknowledged that the two remaining assumptions of Case V--and, by
+extension, of the BTL model--could pose significant problems when
+researchers assess complex traits or heterogeneous stimuli
+\citep{Thurstone_1927b}. In particular, he noted that although these
+assumptions simplify the trait measurement model, they could fail to
+capture the complexity of CJ data
+\citep{Kelly_et_al_2022, Andrich_1978}, potentially leading to
+unreliable and inaccurate trait estimates
+\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.
+
+Furthermore, because Thurstone's primary goal was to produce a ``coarse
+scaling'' of traits and to ``allocate the compared stimuli on this
+continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
+how to use the resulting trait estimates for statistical inference or
+hypothesis testing. The CJ tradition has attempted to address this gap
+by separating trait estimation from hypothesis testing. Specifically, CJ
+studies often rely on point estimates of the traits (i.e., the BTL
+scores)--or their transformations--to conduct hypothesis tests. However,
+the statistical literature cautions against this practice, as it can
+introduce bias into the analysis and compromise the reliability of the
+resulting conclusions
+\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
+
+To address these issues, this study extends Thurstone's general form
+using a systematic, integrated approach that combines Causal and
+Bayesian inference methods. It begins by formulating a conceptual model
+represented by an SCM and a DAG. This model integrates Thurstone's core
+theoretical principles--such as the discriminal processes (i.e., traits)
+of stimuli--alongside key CJ experimental design features, including
+judges' bias, sampling procedures, and comparison mechanisms. The study
+then translates the SCM into a bespoke statistical model that addresses
+key limitations of the BTL model. In particular, it enables the analysis
+of CJ data when the assumptions of equal dispersion and zero correlation
+do not hold (see Section~\ref{sec-theory-issue1a} and
+Section~\ref{sec-theory-issue1b}), and when researchers seek to conduct
+statistical inference or hypothesis testing (refer to
+Section~\ref{sec-theory-issue2}). These methodological innovations
+potentially enhance the reliability and validity of trait measurement
+\citep{Perron_et_al_2015} and improve the accuracy of statistical
+conclusions.
 
@@
 
-One promising research avenue involves investigating how the
-hierarchical structure of stimuli affects CJ experiment designs and
-hypotheses testing. One such study could examine whether collecting a
-single text per individual provides sufficient information to produce
-reliable trait estimates at different levels of the hierarchy. Due to
-concerns about the feasibility of the comparison task
-\citep{Boonen_et_al_2020}, many CJ studies assess traits using only one
-text per individual \citep{citation}. This experimental design
-implicitly assumes that within-individual variability is small relative
-to between-individual differences--that is, texts produced by the same
-individual are relatively homogeneous. However, researchers rarely test
-this assumption, even though it directly influences decisions about the
-number of texts and participants needed for the experiment, ultimately
-affecting the reliability of the trait estimates.
-
-Relatedly, another study could investigate hypotheses about the
-characteristics of the texts themselves. Previously, the literature
-lacked a direct and integrated method for evaluating these types of
-hypotheses. Now, by collecting data on specific stimulus features,
-researchers can assess whether these variables produce the expected
-effects.
-
-\section{Conclusion}\label{sec-conclusion}
+For instance,
+
+Despite its potential, this study presents several challenges that
+researchers must navigate. First, our approach does not establish a
+clear epistemological connection between the theoretical latent trait
+and its quantitative estimate. In other words, the approach does not
+provide a definitive answer to the fundamental question: How do we know
+that the estimated trait reflects the theoretical trait of interest?
+However, this limitation is not unique to our approach; it reflects a
+fundamental challenge all research involving latent constructs faces.
+Nevertheless, we argue that our approach offers methodological tools
+that enable researchers to design CJ experiments under conditions that
+support reliable and valid trait measurement.
+
+Second, although initial simulation tests of the statistical model have
+yielded promising results (not shown), these findings may not generalize
+across research contexts. We have argued that the model's ability to
+estimate traits accurately and support valid statistical inferences
+depends heavily on the specific research context--particularly on
+whether the data and inferential goals align with the model's underlying
+assumptions. That said, our approach provides tools for researchers to
+test and evaluate these context-dependent CJ models directly.
+
+Third, applying this approach requires a solid understanding of Causal
+inference. For instance, throughout the study, we assume that predictor
+variables must be ``relevant'' to the research context, interpreting
+relevance as the inclusion of the predictor variables in a ``sufficient
+adjustment set.'' However, we do not fully explore what this entails or
+its implications for model specification and validity. Nonetheless, we
+provide key references throughout the study to help motivated
+researchers explore these foundational concepts in greater depth.
+
+Fourth and final, researchers must also be proficient in Probabilistic
+Programming Languages (PPLs) and Bayesian inference methods. This
+expertise is essential for translating context-specific conceptual
+models into bespoke statistical models that meaningfully address the
+research questions at hand.
 

### 2025-05-12 paper2.qmd
section:6; process:write

@@
 
-<!-- Proposal: Make e_{I} and e_{J} fully orthogonal, a\
+<!-- Proposal: Make e_{I} and e_{J} fully orthogonal,
 Decision: Although I can see the benefit of this, it would add several layers of explanations to the document or the code for that matter. I think this paper has already a lot of info. -->
@@
 Decision: interesting, but I think I like the model as it is, explicit! -->
-
-<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
-
-<!-- @Bramley_2008 proposes ranking instead of comparisons -->
 
@@
 Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level.
-However, Thurstone identified one key issue in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
+However, Thurstone identified one key challenge in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
 To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from the general form, each progressively incorporating several simplifying assumptions.
@@
 Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and to "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp.
-269], his theory did not specify how to use the resulting trait estimates for statistical inference or hypothesis testing.
-The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing.
-Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct hypothesis tests.
-However, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+269], his theory did not specify how to use the resulting trait estimates for statistical inference.
+The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process.
+Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference.
+Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
 <!-- 4. General summary of what we have done. -->
-To address these issues, this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
+Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
 It begins by formulating a conceptual model represented by an SCM and a DAG.
@@
 <!-- 2. Are there any additional advantages? -->
-Beyond these potential benefits, the approach offers two key advantages.
+Beyond these potential benefits, the approach offers two additional advantages.
 First, it clarifies the interactions among all actors and processes involved in CJ experiments.
@@
 
+<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
+
+<!-- @Bramley_2008 proposes ranking instead of comparisons -->
+
 <!-- Challenges -->
-
+<!-- 1. Epistemiology  -->
 Despite its potential, this study presents several challenges that researchers must navigate.
@@
 
+<!-- 2. Research context is important  -->
 Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts.
-We have argued that the model's ability to estimate traits accurately and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
+We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
 That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.
 
+<!-- 3. Causal inference knowledge -->
 Third, applying this approach requires a solid understanding of Causal inference.
@@
 
+<!-- 4. PPL and Bayesian Inference -->
 Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods.
 This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand.
+While such translation is not exclusive to Bayesian methods--frequentist approaches can also be used--we argue that Bayesian methods offer a more accessible pathway,  as they reduce the need for additional expertise in optimization theory and techniques.
 

### 2025-05-12 _tex/paper2.tex
Built site for gh-pages

@@
 indicating which stimulus the judge perceives as having a higher trait
-level. However, Thurstone identified one key issue in this model: it
+level. However, Thurstone identified one key challenge in this model: it
 required estimating more ``unknown'' parameters than the number of
@@
 continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
-how to use the resulting trait estimates for statistical inference or
-hypothesis testing. The CJ tradition has attempted to address this gap
-by separating trait estimation from hypothesis testing. Specifically, CJ
-studies often rely on point estimates of the traits (i.e., the BTL
-scores)--or their transformations--to conduct hypothesis tests. However,
-the statistical literature cautions against this practice, as it can
+how to use the resulting trait estimates for statistical inference. The
+CJ tradition has attempted to address this gap by separating trait
+estimation from the hypothesis testing process. Specifically, CJ studies
+often rely on point estimates of the traits (i.e., the BTL scores)--or
+their transformations--to conduct statistical inference. Although
+separating these processes simplifies the analysis of CJ data, the
+statistical literature cautions against this practice, as it can
 introduce bias into the analysis and compromise the reliability of the
@@
 
-To address these issues, this study extends Thurstone's general form
-using a systematic, integrated approach that combines Causal and
-Bayesian inference methods. It begins by formulating a conceptual model
-represented by an SCM and a DAG. This model integrates Thurstone's core
-theoretical principles--such as the discriminal processes (i.e., traits)
-of stimuli--alongside key CJ experimental design features, including
-judges' bias, sampling procedures, and comparison mechanisms. The study
-then translates the SCM into a bespoke statistical model that addresses
-key limitations of the BTL model. In particular, it enables the analysis
-of CJ data when the assumptions of equal dispersion and zero correlation
-do not hold (see Section~\ref{sec-theory-issue1a} and
-Section~\ref{sec-theory-issue1b}), and when researchers seek to conduct
-statistical inference or hypothesis testing (refer to
-Section~\ref{sec-theory-issue2}). These methodological innovations
-potentially enhance the reliability and validity of trait measurement
-\citep{Perron_et_al_2015} and improve the accuracy of statistical
-conclusions.
-
-Beyond these potential benefits, the approach offers two key advantages.
-First, it clarifies the interactions among all actors and processes
-involved in CJ experiments. Second, it shifts the analytic paradigm from
-passively accepting the assumptions of the BTL model to actively testing
-their fit with observed data. Together, these advantages create a
-principled framework for evaluating best practices in CJ experimental
-designs--one that better aligns with the demands of contemporary CJ
-assessment contexts \citep{Kelly_et_al_2022} and opens new avenues for
-research.
+Thus, to address the limitations of Thurstone's Case V and the BTL
+model--particularly their strong assumptions and the disconnect between
+trait measurement and hypothesis testing--this study extends Thurstone's
+general form using a systematic, integrated approach that combines
+Causal and Bayesian inference methods. It begins by formulating a
+conceptual model represented by an SCM and a DAG. This model integrates
+Thurstone's core theoretical principles--such as the discriminal
+processes (i.e., traits) of stimuli--alongside key CJ experimental
+design features, including judges' bias, sampling procedures, and
+comparison mechanisms. The study then translates the SCM into a bespoke
+statistical model that addresses key limitations of the BTL model. In
+particular, it enables the analysis of CJ data when the assumptions of
+equal dispersion and zero correlation do not hold (see
+Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b}),
+and when researchers seek to conduct statistical inference or hypothesis
+testing (refer to Section~\ref{sec-theory-issue2}). These methodological
+innovations potentially enhance the reliability and validity of trait
+measurement \citep{Perron_et_al_2015} and improve the accuracy of
+statistical conclusions.
+
+Beyond these potential benefits, the approach offers two additional
+advantages. First, it clarifies the interactions among all actors and
+processes involved in CJ experiments. Second, it shifts the analytic
+paradigm from passively accepting the assumptions of the BTL model to
+actively testing their fit with observed data. Together, these
+advantages create a principled framework for evaluating best practices
+in CJ experimental designs--one that better aligns with the demands of
+contemporary CJ assessment contexts \citep{Kelly_et_al_2022} and opens
+new avenues for research.
 
@@
 across research contexts. We have argued that the model's ability to
-estimate traits accurately and support valid statistical inferences
+accurately estimate traits and support valid statistical inferences
 depends heavily on the specific research context--particularly on
@@
 models into bespoke statistical models that meaningfully address the
-research questions at hand.
+research questions at hand. While such translation is not exclusive to
+Bayesian methods--frequentist approaches can also be used--we argue that
+Bayesian methods offer a more accessible pathway, as they reduce the
+need for additional expertise in optimization theory and techniques.
 

### 2025-05-12 paper2.qmd
section:6; process:write

@@
 Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes.
-Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.
+Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes.
 
@@
 Two key observations support this conclusion: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$.
-As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.
+As a result, the dichotomous outcome of this comparison is more likely to favor Text B over A.
 
@@
 
+<!-- @fig-measurement_error illustrates these issues, showing that neglecting measurement error $(\sigma_{T})$--by relying solely on outcome averages--reduces the precision of a predictor's estimated effect $\beta_{X}$, as indicated by the solid lines.
+-->
+
+<!-- ![The effect of outcome uncertainty $\sigma_{T}$ on the estimation of an effect $\beta_{X}$ linked to the predictor.
+The example assumes a sample size $n=100$ and an uncertainty that increases from left to right.](/images/png/measurement_error.png){#fig-measurement_error width=100%} -->
+
 <!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted. [@Everitt_et_al_2010] -->
@@
 497] -->
-
-<!-- @fig-measurement_error illustrates these issues, showing that neglecting measurement error $(\sigma_{T})$--by relying solely on outcome averages--reduces the precision of a predictor's estimated effect $\beta_{X}$, as indicated by the solid lines.
--->
-
-<!-- ![The effect of outcome uncertainty $\sigma_{T}$ on the estimation of an effect $\beta_{X}$ linked to the predictor.
-The example assumes a sample size $n=100$ and an uncertainty that increases from left to right.](/images/png/measurement_error.png){#fig-measurement_error width=100%} -->
 
@@
 However, Thurstone identified one key challenge in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
-To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from the general form, each progressively incorporating several simplifying assumptions.
+To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from this general form, each progressively incorporating several simplifying assumptions.
 
@@
 
-Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and to "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp.
+Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp.
 269], his theory did not specify how to use the resulting trait estimates for statistical inference.
@@
 Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
-It begins by formulating a conceptual model represented by an SCM and a DAG.
+The approach begins by formulating a conceptual model represented by an SCM and a DAG.
 This model integrates Thurstone's core theoretical principles--such as the discriminal processes (i.e., traits) of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms.
+
 The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model.
-In particular, it enables the analysis of CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when researchers seek to conduct statistical inference or hypothesis testing (refer to @sec-theory-issue2).
-These methodological innovations potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical conclusions.
+This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2).
+In particular, the model accounts for judge biases, captures the hierarchical structure of stimuli, accommodates heterogeneity in discriminal dispersions, and incorporates measurement error in the hypothesis testing process.
+These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.
 
@@
 
-<!-- 2. Are there any additional advantages? -->
+<!-- 5. Are there any additional advantages? -->
 Beyond these potential benefits, the approach offers two additional advantages.
@@
 Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data.
-Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs--one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.
+Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.
 
@@
 
-For instance,
-
-<!-- accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error in trait estimation and hypothesis testing, and accommodates violations of the equal dispersion assumption.
--->
-
-<!-- 3. First research avenue -->
-<!-- One promising research avenue involves investigating how the hierarchical structure of stimuli affects CJ experiment designs and hypotheses testing.
-One such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy.
-Due to concerns about the feasibility of the comparison task [@Boonen_et_al_2020], many CJ studies assess traits using only one text per individual [@citation].
-This experimental design implicitly assumes that within-individual variability is small relative to between-individual differences--that is, texts produced by the same individual are relatively homogeneous.
-However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates.
--->
-
-<!-- Relatedly, another study could investigate hypotheses about the characteristics of the texts themselves.
-Previously, the literature lacked a direct and integrated method for evaluating these types of hypotheses.
-Now, by collecting data on specific stimulus features, researchers can assess whether these variables produce the expected effects.
--->
+<!-- 6. Which are these new avenues for research? -->
+Five research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
+First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect the robustness of statistical inferences.
+Second, further work is needed to understand how judges' biases impact the number of comparisons required to achieve reliable trait estimates.
+Third, researchers should re-evaluate existing methods for identifying and excluding 'misfitting' judges and stimuli to ensure their effectiveness.
+Fourth, studies should explore how the hierarchical structure of stimuli influences both experimental design and hypothesis testing in CJ contexts.
+Finally,  studies should explore principled approaches for selecting contextually relevant variables.
+
+<!-- 6.2 About the implications of the sampling and comparison mechanisms -->
+
+[About the implications of the sampling and comparison mechanisms]{style="color:blue;"}
+
+<!-- Continuing with the first advantage, now we clarify two additional processes of CJ experiments that previous literature do not typically mention: that the sample and comparison mechanism are generators of missing data--that is, they determine which observations do we get to see in our data sample.
+This opens the path to use Rubin's missing data framework [@Rubin_1976] into CJ applications to investigate more complex CJ designs, particularly the ones that are not random samples as the ones outlined here, which could be characterized in Rubin's structure parlor as missing at random (MAR) and missing not at random (MNAR) [@Rubin_1976].
+Among these cases are adaptive comparative algorithm designs, for which the causal structure is not clear, and thus, its impact on trait measurement and hypothesis testing is also not clear.
+Another  -->
+
+<!-- normally S and C are not explicitly mentioned  -->
+
+<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
+C -> {Y}
+e_{A} -> {C, Y} 
+this means that e_{A} is a confounder?,  
+then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?.
+Does this mean that random algorithms per se will always be unbiased?.
+-->
+
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
+
+<!-- Brought Rubin's missing data framework to the field. -->
+<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are random samples, they fulfill ignorability. -->
+
+<!-- Thus Thurstone’s model (in both its Case 5 and Rasch formulation) achieves the goal of sample-free calibration (Wright, 1977) in the sense that the estimate of the distance between A and B does not depend on which other objects they are compared with.
+The practical importance of this is that it is not necessary to have a completely crossed or balanced design of paired comparisons because the estimation process naturally handles missing data.
+[@Bramley_2008, pp.
+256] -->
+
+<!-- Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates.
+In this sense the models are also "sample-free."
+However, this property of the model is true in empirical data only if the model holds in the data.
+Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
+[@Andrich_1978, pp.459] -->
+
+<!-- if the data fit the Case 5 Thurstone model, the estimate of the separation between any two scripts does not depend on which other scripts they are compared with.
+This means that data can be missing in a non-random way without affecting the results.
+However, the precision (standard error) of each script’s estimate depends largely on how many comparisons it has been involved in, so some effort needs to be made to ensure that each script is compared a similar number of times in total.
+[@Bramley_2008, pp.
+260] -->
+
+<!-- Since the analysis can cope with non-random missing data (see earlier), the collection of data can be very flexible – tailored to the needs of the particular situation.
+Because the comparative judgements are all independent, data can be accumulated as needed.
+The investigator is thus not (in principle) bound by a prior design.
+At any point more data could be collected and added to the existing data to recover from accidents, to replace an unacceptable judge, to add extra scripts, to resolve conflicts, to investigate possible bias or to deepen the study of any aspect that proves especially interesting.
+[@Bramley_2008, pp.
+261] -->
+
+<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
+However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
 
@@
 
-<!-- One of these avenues could be about the implications of the hierarchical structure of the stimuli, not only for the design of CJ experiments, but also for the test of hypothesis.  -->
-<!-- The avenues of research span the following:  -->
-<!-- - assessing if the number of texts a researcher collects is sufficient.
-Due to efficiency of scoring, most of research only collects only one text.
-This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
-However, this cannot be known ad-hoc.
-The frameworks open the avenue to test this assumption -->
-<!-- - Assessing hypothesis regarding the characteristics of the texts.
-Before the literature did not have a direct and integrated way to assess these types of hypothesis, now you just need to collect data on the particular stimuli, and test if these variables reflect the expected effect -->
-
-<!-- Derived from this other avenues of research that are open are -->
-<!-- - assessing if the number of individuals a researcher collects is sufficient.
-Most of the literature focuses on ensuring that the number of comparisons between texts are sufficient, but no literature actively engages in determining if the number of individuals are sufficient to test hypothesis at that level.
--->
-<!-- - Moreover, if certain characteristics of the individuals explain the trait locations of their texts -->
+<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
+Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp.
+683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement.
+As @Kelly_et_al_2022 noted [pp.
+683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->
+
+<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
+--->
+
+<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
+Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
+[@Kelly_et_al_2022, pp.
+683] -->
+
+<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
+[@Kelly_et_al_2022, pp.
+681] -->
+
+<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology.
+We do not yet know how widely it can be, or should be, applied.
+[@Pollitt_2012b, pp.
+297]-->
+
+<!-- Samples from different sites often differ in terms of their demographic profiles (here, their age and gender distribution), and these demographic variables might in turn affect the distribution of the trait of interest.
+How should researchers deal with these differences?
+The answer depends on the processes that generated the observed disparities.
+Demographic disparities among samples may result from (a) differences in the actual populations from which the samples are taken or (b) sampling procedures that differ among sites.
+[@Deffner_et_al_2022, pp.
+6] -->
+
+<!-- These examples demonstrate that the generalizability of experimental effects does not depend on the presence of population differences per se but on the exact mechanisms by which populations differ.
+While some differences—especially those concerning the independent variable—are inconsequential for intended generalizations, differences concerning effect modifiers or mediators require statistical adjustment.
+[@Deffner_et_al_2022, pp.
+11] -->
+
+<!-- 6.3 About judges biases, judgment discrepancies, and the number of judges and judgments -->
+
+[About judges biases, judgment discrepancies, and the number of judges and judgments]{style="color:blue;"}
+
+<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
+Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
+It also accounts for potential judge biases in the comparative judgment process.
+-->
+
+<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
+
+<!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need?
+Who should judge?
+How many judgments should they make?
+And crucially, how can we be confident that they are basing their judgments on acceptable criteria?
+If judges are not using construct-relevant criteria, then the argument that comparative judgments are necessarily valid cannot be upheld.
+We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used.
+However, work to derive some general principles for good practice would be valuable.
+[@Kelly_et_al_2022, pp.
+683] -->
+
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
+<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
+(2002).
+Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
+[@Bramley_2008, pp.
+274] -->
+<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges. -->
+
+<!-- presence of judges biases in the comparative judgment procedure. From this it derives research on:  -->
+<!-- - does the judges have biases, and if so, do ad-hoc methods can identify them -->
+
+<!-- Given that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, this study has argued that judges biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
+-->
+
+<!-- This hierarchical nature open paths for research in, for instance, the benefits of *repeated measures designs* [@Lawson_2015, pp.
+366-376] in CJ experiments.
+This is when the experimental conditions accounts for judges comparing the same pair of stimuli multiple times.
+Or research that determines the minimum number of judges and judgments required to ensure an acceptable level of reliability and validity under particular assumptions of the data [@Kelly_et_al_2022, pp.
+682].
+-->
+
+<!-- 6.4 About removing misfit judges and stimuli  -->
+
+[About removing misfit judges and stimuli]{style="color:blue;"}
+
+<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
+If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
+...
+It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
+681] -->
+
+<!-- Such differences may not be detected through analyses of bias and misfit.
+McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
+Where students prioritized factual recall, teachers prioritized scientific explanation.
+Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
+Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
+[@Kelly_et_al_2022, pp.
+683] -->
 
@@
 
-<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
-However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
+<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
+
+<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
+
+<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
+
+<!-- 6.1 About the hierarchical structure of stimuli, number and predictors -->
+
+[About the hierarchical structure of stimuli]{style="color:blue;"}
+
+<!-- About the hierarchical structure of stimuli and its impact on CJ experiment design and hypothesis testing, one such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy.
+It has been highlighted that due to concerns about the feasibility of the comparison task [@Boonen_et_al_2020], many CJ studies assess traits using only one text per individual [@citation].
+Although the literature have examples of collecting more than one stimuli per individual, their statistical handling still separates the trait measurement from the statistical inference process.
+This experimental design implicitly assumes that within-individual variability is small relative to between-individual differences--that is, texts produced by the same individual are relatively homogeneous.
+However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates.
+-->
+
+<!-- Relatedly, another study could investigate hypotheses about the characteristics of the texts themselves.
+Previously, the literature lacked a direct and integrated method for evaluating these types of hypotheses.
+Now, by collecting data on specific stimulus features, researchers can assess whether these variables produce the expected effects.
+-->
+
+<!-- clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
+Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
+-->
+
+<!-- The avenues of research span the following:  -->
+<!-- - assessing if the number of texts a researcher collects is sufficient.
+Due to efficiency of scoring, most of research only collects only one text.
+This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
+However, this cannot be known ad-hoc.
+The frameworks open the avenue to test this assumption -->
+<!-- - Assessing hypothesis regarding the characteristics of the texts.
+Before the literature did not have a direct and integrated way to assess these types of hypothesis, now you just need to collect data on the particular stimuli, and test if these variables reflect the expected effect -->
+
+<!-- Derived from this other avenues of research that are open are -->
+<!-- - assessing if the number of individuals a researcher collects is sufficient.
+Most of the literature focuses on ensuring that the number of comparisons between texts are sufficient, but no literature actively engages in determining if the number of individuals are sufficient to test hypothesis at that level.
+-->
+<!-- - Moreover, if certain characteristics of the individuals explain the trait locations of their texts -->
 
@@
 
-<!-- presence of judges biases in the comparative judgment procedure. From this it derives research on:  -->
-<!-- - does the judges have biases, and if so, do ad-hoc sta -->
-
-<!-- Given that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, this study has argued that judges biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
--->
-
-<!-- This hierarchical nature open paths for research in, for instance, the benefits of *repeated measures designs* [@Lawson_2015, pp.
-366-376] in CJ experiments.
-This is when the experimental conditions accounts for judges comparing the same pair of stimuli multiple times.
-Or research that determines the minimum number of judges and judgments required to ensure an acceptable level of reliability and validity under particular assumptions of the data [@Kelly_et_al_2022, pp.
-682].
--->
-
-<!-- (Is this a validity contention?) -->
-<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
-Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp.
-683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement.
-As @Kelly_et_al_2022 noted [pp.
-683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->
-
-<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
-Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
-[@Kelly_et_al_2022, pp.
-683] -->
-
-<!-- Samples from different sites often differ in terms of their demographic profiles (here, their age and gender distribution), and these demographic variables might in turn affect the distribution of the trait of interest.
-How should researchers deal with these differences?
-The answer depends on the processes that generated the observed disparities.
-Demographic disparities among samples may result from (a) differences in the actual populations from which the samples are taken or (b) sampling procedures that differ among sites.
-[@Deffner_et_al_2022, pp.
-6] -->
-
-<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology.
-We do not yet know how widely it can be, or should be, applied.
-[@Pollitt_2012b, pp.
-297]-->
-
-<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
-[@Kelly_et_al_2022, pp.
-681] -->
-
-<!-- These examples demonstrate that the generalizability of experimental effects does not depend on the presence of population differences per se but on the exact mechanisms by which populations differ.
-While some differences—especially those concerning the independent variable—are inconsequential for intended generalizations, differences concerning effect modifiers or mediators require statistical adjustment.
-[@Deffner_et_al_2022, pp.
-11] -->
-
-<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
---->
-
-<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
-If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
-...
-It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
-681] -->
-
-<!-- Such differences may not be detected through analyses of bias and misfit.
-McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
-Where students prioritized factual recall, teachers prioritized scientific explanation.
-Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
-Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
-[@Kelly_et_al_2022, pp.
-683] -->
-
-<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
-
-<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
-
-<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
-
-<!-- clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
-Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
--->
-
-<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
-Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
-It also accounts for potential judge biases in the comparative judgment process.
--->
-
-<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
-
-<!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need?
-Who should judge?
-How many judgments should they make?
-And crucially, how can we be confident that they are basing their judgments on acceptable criteria?
-If judges are not using construct-relevant criteria, then the argument that comparative judgments are necessarily valid cannot be upheld.
-We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used.
-However, work to derive some general principles for good practice would be valuable.
-[@Kelly_et_al_2022, pp.
-683] -->
-
-<!-- Continuing with the first advantage, now we clarify two additional processes of CJ experiments that previous literature do not typically mention: that the sample and comparison mechanism are generators of missing data--that is, they determine which observations do we get to see in our data sample.
-This opens the path to use Rubin's missing data framework [@Rubin_1976] into CJ applications to investigate more complex CJ designs, particularly the ones that are not random samples as the ones outlined here, which could be characterized in Rubin's structure parlor as missing at random (MAR) and missing not at random (MNAR) [@Rubin_1976].
-Among these cases are adaptive comparative algorithm designs, for which the causal structure is not clear, and thus, its impact on trait measurement and hypothesis testing is also not clear.
-Another  -->
-
-<!-- normally S and C are not explicitly mentioned  -->
-
-<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
-C -> {Y}
-e_{A} -> {C, Y} 
-this means that e_{A} is a confounder?,  
-then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?.
-Does this mean that random algorithms per se will always be unbiased?.
--->
-
-<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
-
-<!-- Brought Rubin's missing data framework to the field. -->
-<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are random samples, they fulfill ignorability. -->
-
-<!-- Thus Thurstone’s model (in both its Case 5 and Rasch formulation) achieves the goal of sample-free calibration (Wright, 1977) in the sense that the estimate of the distance between A and B does not depend on which other objects they are compared with.
-The practical importance of this is that it is not necessary to have a completely crossed or balanced design of paired comparisons because the estimation process naturally handles missing data.
-[@Bramley_2008, pp.
-256] -->
-
-<!-- Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates.
-In this sense the models are also "sample-free."
-However, this property of the model is true in empirical data only if the model holds in the data.
-Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
-[@Andrich_1978, pp.459] -->
-
-<!-- if the data fit the Case 5 Thurstone model, the estimate of the separation between any two scripts does not depend on which other scripts they are compared with.
-This means that data can be missing in a non-random way without affecting the results.
-However, the precision (standard error) of each script’s estimate depends largely on how many comparisons it has been involved in, so some effort needs to be made to ensure that each script is compared a similar number of times in total.
-[@Bramley_2008, pp.
-260] -->
-
-<!-- Since the analysis can cope with non-random missing data (see earlier), the collection of data can be very flexible – tailored to the needs of the particular situation.
-Because the comparative judgements are all independent, data can be accumulated as needed.
-The investigator is thus not (in principle) bound by a prior design.
-At any point more data could be collected and added to the existing data to recover from accidents, to replace an unacceptable judge, to add extra scripts, to resolve conflicts, to investigate possible bias or to deepen the study of any aspect that proves especially interesting.
-[@Bramley_2008, pp.
-261] -->
-
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
-267] -->
-
-<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
-(2002).
-Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
-[@Bramley_2008, pp.
-274] -->
-<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges. -->
+<!-- 6.5 About using Causal inference to carefully the most relevant variables for specific research contexts -->
+
+[About revisit the approaches for selecting contextually relevant variables]{style="color:blue;"}
 
@@
 
-<!-- Challenges -->
-<!-- 1. Epistemiology  -->
+<!-- 7. Challenges -->
+<!-- 7.1 Epistemiology  -->
 Despite its potential, this study presents several challenges that researchers must navigate.
@@
 
-<!-- 2. Research context is important  -->
+<!-- 7.2 Research context is important  -->
 Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts.
@@
 
-<!-- 3. Causal inference knowledge -->
+<!-- 7.3 Causal inference knowledge -->
 Third, applying this approach requires a solid understanding of Causal inference.
@@
 
-<!-- 4. PPL and Bayesian Inference -->
+<!-- 7.4 PPL and Bayesian Inference -->
 Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods.

### 2025-05-12 _tex/paper2.tex
Built site for gh-pages

@@
 Although separating these processes simplifies the analysis of CJ data,
-it may also undermine the reliability of various statistical results
+it may also undermine the reliability of various statistical inferences
 derived from these processes.
@@
 shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
-outcome of this comparison is more likely to favor Text B over Text A.
+outcome of this comparison is more likely to favor Text B over A.
 
@@
 issue and facilitate the theory's practical applicability, he formulated
-five cases derived from the general form, each progressively
+five cases derived from this general form, each progressively
 incorporating several simplifying assumptions.
@@
 Furthermore, because Thurstone's primary goal was to produce a ``coarse
-scaling'' of traits and to ``allocate the compared stimuli on this
+scaling'' of traits and ``allocate the compared stimuli on this
 continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
@@
 general form using a systematic, integrated approach that combines
-Causal and Bayesian inference methods. It begins by formulating a
-conceptual model represented by an SCM and a DAG. This model integrates
-Thurstone's core theoretical principles--such as the discriminal
-processes (i.e., traits) of stimuli--alongside key CJ experimental
-design features, including judges' bias, sampling procedures, and
-comparison mechanisms. The study then translates the SCM into a bespoke
-statistical model that addresses key limitations of the BTL model. In
-particular, it enables the analysis of CJ data when the assumptions of
-equal dispersion and zero correlation do not hold (see
-Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b}),
-and when researchers seek to conduct statistical inference or hypothesis
-testing (refer to Section~\ref{sec-theory-issue2}). These methodological
-innovations potentially enhance the reliability and validity of trait
-measurement \citep{Perron_et_al_2015} and improve the accuracy of
-statistical conclusions.
+Causal and Bayesian inference methods. The approach begins by
+formulating a conceptual model represented by an SCM and a DAG. This
+model integrates Thurstone's core theoretical principles--such as the
+discriminal processes (i.e., traits) of stimuli--alongside key CJ
+experimental design features, including judges' bias, sampling
+procedures, and comparison mechanisms.
+
+The study then translates the SCM into a bespoke statistical model that
+addresses key limitations of the BTL model. This model allows
+researchers to analyze CJ data when the assumptions of equal dispersion
+and zero correlation do not hold (see Section~\ref{sec-theory-issue1a}
+and Section~\ref{sec-theory-issue1b}), and when they need to perform
+statistical inference (refer to Section~\ref{sec-theory-issue2}). In
+particular, the model accounts for judge biases, captures the
+hierarchical structure of stimuli, accommodates heterogeneity in
+discriminal dispersions, and incorporates measurement error in the
+hypothesis testing process. These methodological innovations can
+potentially enhance the reliability and validity of trait measurement
+\citep{Perron_et_al_2015} and improve the accuracy of statistical
+inferences.
 
@@
 advantages create a principled framework for evaluating best practices
-in CJ experimental designs--one that better aligns with the demands of
+in CJ experimental designs, one that better aligns with the demands of
 contemporary CJ assessment contexts \citep{Kelly_et_al_2022} and opens
@@
 
-For instance,
+Five research avenues deserve particular attention due to their direct
+impact on the reliability and validity of trait estimates and the
+accuracy of statistical inferences. First, researchers should
+investigate how sampling and comparison mechanisms influence CJ
+experimental outcomes and affect the robustness of statistical
+inferences. Second, further work is needed to understand how judges'
+biases impact the number of comparisons required to achieve reliable
+trait estimates. Third, researchers should re-evaluate existing methods
+for identifying and excluding `misfitting' judges and stimuli to ensure
+their effectiveness. Fourth, studies should explore how the hierarchical
+structure of stimuli influences both experimental design and hypothesis
+testing in CJ contexts. Finally, studies should explore principled
+approaches for selecting contextually relevant variables.
+
+{About the implications of the sampling and comparison mechanisms}
+
+{About judges biases, judgment discrepancies, and the number of judges
+and judgments}
+
+{About removing misfit judges and stimuli}
+
+{About the hierarchical structure of stimuli}
+
+{About revisit the approaches for selecting contextually relevant
+variables}
 

### 2025-05-14 paper2.qmd
section:5; process:write,equation,code

@@
 It is easier to visualize the structure of these vectors and matrices by considering an example.
-Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
+Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
 $$
@@
 <!-- 3. how de we use that data structure? -->
-Now, using equations ([-@eq-mat11]) and ([-@eq-mat12]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
+Now, using Equations ([-@eq-mat11]) and ([-@eq-mat12]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
 In this depiction, the outcome $O^{cp}_{R}$, the texts' discriminal difference $D_{R}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values.
@@
 
-Where each element of $S_{I}$ is a binary value indicating the presence or absence of corresponding elements in the vector $I$, as in equation ([-@eq-mat22]).
+Where each element of $S_{I}$ is a binary value indicating the presence or absence of corresponding elements in the vector $I$, as in Equation ([-@eq-mat22]).
 We apply the same logic to $S_{J}$ using vector $J$ (not shown).
@@
 
-Similarly, each element of $S_{IA}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$, as defined in equation ([-@eq-mat23]).
+Similarly, each element of $S_{IA}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$, as defined in Equation ([-@eq-mat23]).
 We apply the same logic to $S_{JK}$ using the matrix $JK$ (not shown).
@@
 
-We can visualize these vectors more clearly using the example in equation ([-@eq-mat13]).
+We can visualize these vectors more clearly using the example in Equation ([-@eq-mat13]).
 Suppose researchers exclude the second student and the third judge, and they also omit the second text from each individual and the third repeated judgment from each judge.
@@
 
-Finally, we define the *sample mechanism* $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
-Each element $s_{(i,a,h,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in equation ([-@eq-mat26]).
+Finally, we define the *sample mechanism* $S$ in Equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$.
+Each element $s_{(i,a,h,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in Equation ([-@eq-mat26]).
 Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$.
@@
 As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model.
-Equation ([-@eq-mat27]) shows that $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in equation ([-@eq-mat26]).
+Equation ([-@eq-mat27]) shows that $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in Equation ([-@eq-mat26]).
 $$
@@
 This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
-Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in equation ([-@eq-mat51]).
-For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent.
-Here, $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
+Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in Equation ([-@eq-mat51]).
+For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent, where $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
 $$
@@
 <!-- 2. assumptions about the outcome -->
-Each CPD in equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_prob], and the statistical model [-@fig-cj16_stat].
-First, we assume that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
-Furthermore, following the conventions of Generalized Linear Latent and Mixed Models (GLLAMMs) [@Rabe_et_al_2004a; @Skrondal_et_al_2004a; @Rabe_et_al_2012b], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
+Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_prob], and the statistical model [-@fig-cj16_stat].
+We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
+Furthermore, following the conventions of Generalized Linear Models (GLMs) [@Nelder_et_al_1983; @Nelder_et_al_1996; @Agresti_2015], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
 
@@
 34].
-This means that when only two un-ordered outcomes are possible, and the expected numbers of each type of event are assumed to be constant, then the binomial distribution is the one that is most consistent with these constraints [@McElreath_2020, pp.
+This means that the Bernoulli distribution is the most consistent alternative when only two un-ordered outcomes are possible and their expected frequencies are assumed to be constant [@McElreath_2020, pp.
 310].
@@
 
-::: {#fig-cj16 layout-ncol=3}
-
-::: {#fig-cj16_scm fig-align="center" fig-pos="H"}
+::: {#fig-cj16 layout-ncol=3 layout-valign="top"}
+
+::: {#fig-cj16_scm}
 $$
@@
   B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\ \\
   e_{I} & \dsep \{ e_{J}, e_{IA}, e_{JK} \} \\
   e_{J} & \dsep \{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \dsep e_{JK} \\ \\
+  e_{IA} & \dsep e_{JK}
 \end{aligned}
@@
 
-::: {#fig-cj16_prob fig-align="center" fig-pos="H"}
+::: {#fig-cj16_prob}
 $$
@@
   & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
-  & P( B_{J} \mid Z_{J}, e_{J} ) \\
-  & P( e_{I} ) \\
-  & P( e_{J} ) \\
-  & P( e_{IA} ) \\  
-  & P( e_{JK} )
+  & P( B_{J} \mid Z_{J}, e_{J} ) \\ \\
+  & P( e_{I} ) P( e_{IA} ) P( e_{J} ) P( e_{JK} ) \\ \\ \\
 \end{aligned}
@@
 
-::: {#fig-cj16_stat fig-align="center" fig-pos="H"}
+::: {#fig-cj16_stat}
 $$
@@
   B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
-  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
-  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
-  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
-  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
-  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
+  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\ \\
+  \boldsymbol{e} & \sim \text{Multi-Normal}( \boldsymbol{\mu}, \boldsymbol{\Sigma} )
+  \\
+  \boldsymbol{\Sigma} &= \boldsymbol{D} \boldsymbol{R} \boldsymbol{D} \\ \\
 \end{aligned}
@@
 
-[^8]: Randomization ensures that data—and, by extension, an estimator—satisfies several key identification properties, such as common support, no interference, and consistency.
+[^8]: Randomization ensures that data--and, by extension, an estimator--satisfies several key identification properties, such as common support, no interference, and consistency.
 The most critical property, however, is the elimination of confounding.
@@
 <!-- 4. functional form for D_{R} -->
-Second, we define $D_{R}$ as the difference between the discriminal processes of two texts, $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the texts, plus the corresponding repeated judge biases, $B_{JK}[j, k]$.
+Next we define $D_{R}$ as the difference between the discriminal processes $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the texts, plus the corresponding repeated judge biases $B_{JK}[j, k]$.
 Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
@@
 $$ {#eq-mat52}
+
 This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022].
 In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
-To clarify, the square brackets in $D_{R}$ simply indicate the relevant indices for each trait vector and do not imply any subsetting of the data.
+To clarify, the square brackets in $D_{R}$ indicate the relevant indices for each trait vector; they do not imply any subsetting of the data.
 
@@
 
-<!-- 5. Functional form and assumptions for T_{IA} -->
-Third, we define $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables $\beta_{XA}X_{IA}$ (e.g., text length on quality), and the texts' idiosyncratic errors $e_{IA}$.
-We assume that each element of $e_{IA}$ is independent and identically distributed $(iid)$ according to a Normal distribution, with a mean of zero and a standard deviation $p_{IA}$.
-The parameter $p_{IA}$ is defined as a proportion of 1 to establish the scale of the latent trait, which is required in any latent variable model [@Depaoli_2021].
-Note that because $e_{IA}$ follows a Normal distribution, $T_{IA}$ is also normally distributed.
-This result implies that our model reverts to assuming Normal discriminal processes for the stimuli, as originally proposed by @Thurstone_1927b (see @tbl-thurstone_cases).
+<!-- 5. Functional form and assumptions -->
+We now specify the functional forms for $T_{IA}$, $T_{I}$, $B_{JK}$, and $B_{J}$.
+We model $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables on quality assessment $\beta_{XA}X_{IA}$ (such as the influence of text length), and the text-specific idiosyncratic errors $e_{IA}$.
+Similarly, we express $T_{I}$ as a linear combination of relevant student-related variables affecting the quality assessment $\beta_{XI} X_{I}$, and student-specific idiosyncratic errors $e_{I}$.
+For the judge-specific terms, we model $B_{JK}$ as a linear combination of the judge's individual bias $B_{J}$, the influence of relevant judgment-related variables on quality assessment $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments affect the evaluation), and judgment-specific idiosyncratic errors $e_{JK}$.
+Finally, we define $B_{J}$ as a linear combination of relevant judge-level variables influencing the quality assessment $\beta_{ZJ}Z_{J}$ (such as judgment expertise) and judge-specific idiosyncratic errors $e_{J}$.
+
+<!-- 6. error assumptions -->
+Next, we specify the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$.
+These error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables.
+We must resolve these indeterminacies to identify the model [@Depaoli_2021; @deAyala_2009].
+Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix.
+To address the *location* indeterminacy, we set the errors' mean vector to zero:
+$$
+\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
+$$ {#eq-mat53}
+
+Following [SCM -@fig-cj16_scm], we assume uncorrelated errors to resolve the *orientation* indeterminacy. Thus, we define the correlation matrix $\boldsymbol{R}$ as:
+$$
+\boldsymbol{R} = \begin{bmatrix}
+    1 & 0 & 0 & 0 \\
+    0 & 1 & 0 & 0 \\
+    0 & 0 & 1 & 0 \\
+    0 & 0 & 0 & 1 
+  \end{bmatrix}
+$$ {#eq-mat54}
+
+To resolve the *scale* indeterminacy, we define the diagonal matrix $\boldsymbol{D}$ as:
+$$
+\boldsymbol{D} = \begin{bmatrix}
+    s_{XI} & 0 & 0 & 0 \\
+    0 & p_{IA} & 0 & 0 \\
+    0 & 0 & s_{ZJ} & 0 \\
+    0 & 0 & 0 & p_{JK} 
+  \end{bmatrix}
+$$ {#eq-mat55}
+
+Here, $s_{XI}$ represents the standard deviation for the individuals, $p_{IA}$ for the stimuli, $s_{ZJ}$ for the judges, and $p_{JK}$ for the judgments.
+We assume $s_{XI}$ varies depending on the teaching method group to which each student belongs.
+In the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation [-@eq-mat56].
+This constraint anchors the scale of the individuals' latent trait while relaxing the assumption of equal dispersion for the stimuli, thereby addressing the concerns raised in @sec-theory-issue1a.
+$$
+\sum_{g=1}^{2} s_{XI}[g]/2 = 1
+$$ {#eq-mat56}
+
+In contrast, we define $p_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait.
+Note that, since the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, $e_{IA}$ follows a univariate Normal distribution by the properties of the distribution.
+As a result, $T_{IA}$ is also normally distributed, which effectively reverts to assuming Normal discriminal processes for the stimuli, an assumption originally proposed by @Thurstone_1927b (see @tbl-thurstone_cases)
+
+Similarly, we assume $s_{ZJ}$ varies depending on the groups to which each judge belongs.
+For instance, if $Z_{J} = \{1,2,3\}$ represents three groups of judges with varying expertise, the model sets the constraint:
+$$
+\sum_{g=1}^{2} s_{ZJ}[g]/3 = 1
+$$ {#eq-mat56}
+
+This constraint anchors the scale of the judges' latent trait and relaxes the assumption of equal dispersion for the judgments.
+Conversely, $p_{JK}$ is defined as a proportion of $1$ to establish the scale of the judgments' latent trait relative to the scale of the judges' trait.
+
+<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->
+
+<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
+This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
+12].
+-->
+
+<!-- when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification. -->
 
@@
 
-<!-- The difficulty and discrimination parameters for the questions should then be given a diffuse, or ideally a hierarchical prior, which will identify these parameters by scaling and locating relative to the student ability parameters.
-(https://mc-stan.org/docs/stan-users-guide/regression.html#priors-for-identification.section) -->
-
-<!-- 6. Functional form and assumptions for T_{I} -->
-Fourth, we define $T_{I}$ as a linear combination of the effects of relevant student-related variables, $\beta_{XI} X_{I}$, and students' idiosyncratic errors, $e_{I}$.
-As before, we assume that each element of $e_{I}$ is $iid$ Normal with a mean of zero.
-However, we consider the standard deviation, $s_{XI}$, varies depending on the teaching method group to which each student belongs.
-Given the example in @sec-theoretical where the teaching method $X_{I} = \{1,2\}$, the model sets the following constraint to anchor the scale of the latent trait [@Depaoli_2021]: $\sum_{g=1}^{2} s_{XI}[g]/2 = 1$.
-Note that this parametrization of $e_{I}$ relaxes the equal dispersion assumption for the stimuli, effectively addressing the concerns raised in @sec-theory-issue1a.
-
-<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
-This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
-12].
--->
-
-<!-- when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification. -->
-
-<!-- 7. Functional form and assumptions for B_{JK} -->
-Fifth, we define $B_{JK}$ as a linear combination of the judges' individual bias $B_{J}$, the effects of relevant judgment-related variables $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments a judge makes affect the assessment of quality), and judgment-specific idiosyncratic errors $e_{JK}$.
-We assume that each element of $e_{JK}$ is $iid$ Normal with a mean of zero and a standard deviation $p_{JK}$, defined as a proportion of $1$, to anchor the scale of the latent trait [@Depaoli_2021].
-
-<!-- 8. Functional form and assumptions for B_{J} -->
-Sixth, we define $B_{J}$ as a linear combination of the effects of relevant judge-level variables $\beta_{ZJ}Z_{J}$ (e.g., the effects of judgment expertise on the quality assessment) and judge-specific idiosyncratic errors $e_{J}$.
-Each element of $e_{J}$ is likewise assumed to be $iid$ Normal with mean zero and a standard deviation $s_{ZJ}$ that depends on the groups to which each judge belongs.
-For instance, if $Z_{J} = \{1,2,3\}$ representing three groups of judges with varying expertise, the model imposes the constraint $\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1$ to fix the scale of the latent trait [@Depaoli_2021].
-
-<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->
-
 <!-- [@deAyala_2009, p.
@@
 to solve metric indeterminancy you commonly fix a mu=0, and s=1, but for rotational indeterminancy, you need to define (maybe) orthogonality between units traits and judges bias -->
+
+<!-- The difficulty and discrimination parameters for the questions should then be given a diffuse, or ideally a hierarchical prior, which will identify these parameters by scaling and locating relative to the student ability parameters.
+(https://mc-stan.org/docs/stan-users-guide/regression.html#priors-for-identification.section) -->
 
@@
 
-<!-- Proposal: Make e_{I} and e_{J} fully orthogonal,
-Decision: Although I can see the benefit of this, it would add several layers of explanations to the document or the code for that matter. I think this paper has already a lot of info. -->
-
 <!-- Proposal: Add new way of declaring non-centered parametrization and optional variables (See section 'Affinely transformed real' and 'Declaring optional variables' sections in https://mc-stan.org/docs/reference-manual/types.html)
@@
 Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b].
-In particular, he noted that although these assumptions simplify the trait measurement model, they could fail to capture the complexity of CJ data [@Kelly_et_al_2022; @Andrich_1978], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
+In particular, he noted that while assuming equal discriminal dispersions and zero correlation for the stimuli' discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Kelly_et_al_2022; @Andrich_1978].
+As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
 

### 2025-05-14 code/CJmod_add.R

@@ -21,9 +21,9 @@ data{
     array[ns] int<lower=1, upper=nsA> R2;   // stimuli (left)
     array[ns] int<lower=1, upper=nsI> R3;   // individual (right)
     array[ns] int<lower=1, upper=nsA> R4;   // stimuli (right)
-    array[ns] int<lower=1, upper=nsJ> R5;    // judges
-    array[ns] int<lower=1, upper=nsK> R6;    // judgments
-    array[ns] int<lower=0, upper=1> OR;   // outcome (dichotomous)
+    array[ns] int<lower=1, upper=nsJ> R5;   // judges
+    array[ns] int<lower=1, upper=nsK> R6;   // judgments
+    array[ns] int<lower=0, upper=1> OR;     // outcome (dichotomous)
     
     // individual-stimuli data
     array[nsI*nsA] int<lower=1, upper=nsI> IA1s;    // individuals
@@ -56,61 +56,70 @@ parameters{
     
     
     // error parameters 
-    vector[nsI] zeI;            // individual errors (non-centered)
-    real<lower=0,upper=1> pIA;  // stimuli proportion of variability
-    matrix[nsI,nsA] zeIA;       // stimuli errors (non-centered)
-
-    vector[nsJ] zeJ;            // judges errors (identification)
-    real<lower=0,upper=1> pJK;  // judgments proportion of variability
-    matrix[nsJ,nsK] zeJK;       // judgments errors (non-centered)
+    real<lower=0,upper=1> pIA;                  // stimuli proportion of variability
+    real<lower=0,upper=1> pJK;                  // judgments proportion of variability
+    array[max(nsI*nsA, nsJ*nsK)] vector[4] ze;  // errors (non-centered)
 
 }
 transformed parameters{
     
     // declaring
-    vector[nsI] TI;       // individuals' trait
     vector[nsI] eI;       // individuals errors (identification)
-    matrix[nsI,nsA] TIA;  // stimuli trait
     matrix[nsI,nsA] eIA;  // stimuli errors
-    
-    vector[nsJ] BJ;       // judges' trait
[…]

### 2025-05-14 _tex/paper2.tex
Built site for gh-pages

@@
 \(n_{J} = 3\), and \(n_{K} = 3\), the vectors and matrices described in
-equations (\ref{eq-mat11}) and (\ref{eq-mat12}) take the following form:
+Equations (\ref{eq-mat11}) and (\ref{eq-mat12}) take the following form:
 \begin{equation}\phantomsection\label{eq-mat13}{
@@
 
-Now, using equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
+Now, using Equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
 re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
@@
 presence or absence of corresponding elements in the vector \(I\), as in
-equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\) using
+Equation (\ref{eq-mat22}). We apply the same logic to \(S_{J}\) using
 vector \(J\) (not shown). Thus, the vectors \(S_{I}\) and \(S_{J}\)
@@
 presence or absence of data rows in the matrices \(IA\), as defined in
-equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\) using
+Equation (\ref{eq-mat23}). We apply the same logic to \(S_{JK}\) using
 the matrix \(JK\) (not shown). Thus, the vectors \(S_{IA}\) and
@@
 We can visualize these vectors more clearly using the example in
-equation (\ref{eq-mat13}). Suppose researchers exclude the second
+Equation (\ref{eq-mat13}). Suppose researchers exclude the second
 student and the third judge, and they also omit the second text from
@@
 
-Finally, we define the \emph{sample mechanism} \(S\) in equation
+Finally, we define the \emph{sample mechanism} \(S\) in Equation
 (\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
@@
 indicating the presence or absence of data rows in the matrix \(R\)
-resulting from the sample mechanism, as in equation (\ref{eq-mat26}).
+resulting from the sample mechanism, as in Equation (\ref{eq-mat26}).
 Thus, the vector contains \(n\) elements, matching the number of rows in
@@
 a binary value indicating the presence or absence of data rows in \(R\),
-a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in equation
+a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in Equation
 (\ref{eq-mat26}). \begin{equation}\phantomsection\label{eq-mat27}{
@@
   further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
-as shown in equation (\ref{eq-mat51}). For clarity, we treat expressions
+as shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
 such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
-as equivalent. Here, \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
+as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
 CPD of \(Y\) given \(X\).
@@
 
-Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
-which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
-\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}.
-First, we assume that \(O_{R}\) follows a Bernoulli
+Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
+which we outline in SCM~\ref{fig-cj16_scm}, the probabilistic model
+\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}. We
+begin by assuming that \(O_{R}\) follows a Bernoulli
 distribution\footnote{The binomial distribution--including its special
@@
   distribution for binary events \citep[pp.~34]{McElreath_2020}. This
-  means that when only two un-ordered outcomes are possible, and the
-  expected numbers of each type of event are assumed to be constant,
-  then the binomial distribution is the one that is most consistent with
-  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
-  discussion of the binomial as a maximum entropy distribution, see
+  means that the Bernoulli distribution is the most consistent
+  alternative when only two un-ordered outcomes are possible and their
+  expected frequencies are assumed to be constant
+  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
+  binomial as a maximum entropy distribution, see
   \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
 CJ outcomes. Furthermore, following the conventions of Generalized
-Linear Latent and Mixed Models (GLLAMMs)
-\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
+Linear Models (GLMs)
+\citep{Nelder_et_al_1983, Nelder_et_al_1996, Agresti_2015}, the
 distribution links \(O_{R}\) to the latent discriminal difference vector
@@
 
-\begin{figure}[H]
+\begin{figure}
 
@@
   B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\ \\
   e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
   e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} \\ \\
+  e_{IA} & \:\bot\:e_{JK}
 \end{aligned}
@@
 \(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
-\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
-  an estimator---satisfies several key identification properties, such
-  as common support, no interference, and consistency. The most critical
+\(Z_{J}\)\footnote{Randomization ensures that data--and, by extension,
+  an estimator--satisfies several key identification properties, such as
+  common support, no interference, and consistency. The most critical
   property, however, is the elimination of confounding.
@@
 
-Second, we define \(D_{R}\) as the difference between the discriminal
-processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\),
-representing the underlying written-quality trait of the texts, plus the
-corresponding repeated judge biases, \(B_{JK}[j, k]\). Note that if we
-assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
-biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
-can re-write the discriminal difference as:
+Next we define \(D_{R}\) as the difference between the discriminal
+processes \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), representing the
+underlying written-quality trait of the texts, plus the corresponding
+repeated judge biases \(B_{JK}[j, k]\). Note that if we assume that
+\(B_{JK}[j,k]\) reflects the difference in stimulus-specific biases,
+i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can
+re-write the discriminal difference as:
 \begin{equation}\phantomsection\label{eq-mat52}{
@@
 \end{aligned}
-}\end{equation} This formulation reveals that the discriminal difference
-captures a \emph{pure interaction effect}, in which neither the texts'
-discriminal processes nor the judges' biases alone determine the
-outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
-terms, the discriminal processes of the stimuli become an observable
-outcome only through the judges' perceptions (biases). To clarify, the
-square brackets in \(D_{R}\) simply indicate the relevant indices for
-each trait vector and do not imply any subsetting of the data.
-
-Third, we define \(T_{IA}\) as a linear combination of the students'
-underlying writing-quality traits \(T_{I}\), the effects of relevant
-text-related variables \(\beta_{XA}X_{IA}\) (e.g., text length on
-quality), and the texts' idiosyncratic errors \(e_{IA}\). We assume that
-each element of \(e_{IA}\) is independent and identically distributed
-\((iid)\) according to a Normal distribution, with a mean of zero and a
-standard deviation \(p_{IA}\). The parameter \(p_{IA}\) is defined as a
-proportion of 1 to establish the scale of the latent trait, which is
-required in any latent variable model \citep{Depaoli_2021}. Note that
-because \(e_{IA}\) follows a Normal distribution, \(T_{IA}\) is also
-normally distributed. This result implies that our model reverts to
-assuming Normal discriminal processes for the stimuli, as originally
-proposed by \citet{Thurstone_1927b} (see
-Table~\ref{tbl-thurstone_cases}).
-
-Fourth, we define \(T_{I}\) as a linear combination of the effects of
-relevant student-related variables, \(\beta_{XI} X_{I}\), and students'
-idiosyncratic errors, \(e_{I}\). As before, we assume that each element
-of \(e_{I}\) is \(iid\) Normal with a mean of zero. However, we consider
-the standard deviation, \(s_{XI}\), varies depending on the teaching
-method group to which each student belongs. Given the example in
-Section~\ref{sec-theoretical} where the teaching method
-\(X_{I} = \{1,2\}\), the model sets the following constraint to anchor
-the scale of the latent trait \citep{Depaoli_2021}:
-\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\). Note that this parametrization of
-\(e_{I}\) relaxes the equal dispersion assumption for the stimuli,
-effectively addressing the concerns raised in
+}\end{equation}
+
+This formulation reveals that the discriminal difference captures a
+\emph{pure interaction effect}, in which neither the texts' discriminal
+processes nor the judges' biases alone determine the outcome, but their
+interaction does \citep{Attia_et_al_2022}. In simple terms, the
+discriminal processes of the stimuli become an observable outcome only
+through the judges' perceptions (biases). To clarify, the square
+brackets in \(D_{R}\) indicate the relevant indices for each trait
+vector; they do not imply any subsetting of the data.
+
+We now specify the functional forms for \(T_{IA}\), \(T_{I}\),
+\(B_{JK}\), and \(B_{J}\). We model \(T_{IA}\) as a linear combination
+of the students' underlying writing-quality traits \(T_{I}\), the
+effects of relevant text-related variables on quality assessment
+\(\beta_{XA}X_{IA}\) (such as the influence of text length), and the
+text-specific idiosyncratic errors \(e_{IA}\). Similarly, we express
+\(T_{I}\) as a linear combination of relevant student-related variables
+affecting the quality assessment \(\beta_{XI} X_{I}\), and
+student-specific idiosyncratic errors \(e_{I}\). For the judge-specific
+terms, we model \(B_{JK}\) as a linear combination of the judge's
+individual bias \(B_{J}\), the influence of relevant judgment-related
+variables on quality assessment \(\beta_{ZK}Z_{JK}\) (e.g., how the
+number of judgments affect the evaluation), and judgment-specific
+idiosyncratic errors \(e_{JK}\). Finally, we define \(B_{J}\) as a
+linear combination of relevant judge-level variables influencing the
+quality assessment \(\beta_{ZJ}Z_{J}\) (such as judgment expertise) and
+judge-specific idiosyncratic errors \(e_{J}\).
+
+Next, we specify the probabilistic assumptions for the idiosyncratic
+errors \(e_{I}\), \(e_{IA}\), \(e_{J}\), and \(e_{JK}\). These error
+terms exhibit indeterminacies in their \emph{location},
+\emph{orientation}, and \emph{scale} due to the lack of an inherent
+scale in the associated latent variables. We must resolve these
+indeterminacies to identify the model
+\citep{Depaoli_2021, deAyala_2009}. Drawing on principles from SEM
+\citep{Hoyle_et_al_2023}, we assume that the vector of idiosyncratic
+errors \(\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}\), follows
+a Multivariate Normal distribution with mean vector \(\boldsymbol{\mu}\)
+and a covariance matrix
+\(\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}\),
+with \(\boldsymbol{D}\) denoting a diagonal matrix of standard
+deviations and \(\boldsymbol{R}\) a correlation matrix. To address the
+\emph{location} indeterminacy, we set the errors' mean vector to zero:
+\begin{equation}\phantomsection\label{eq-mat53}{
+\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
+}\end{equation}
+
+Following SCM~\ref{fig-cj16_scm}, we assume uncorrelated errors to
+resolve the \emph{orientation} indeterminacy. Thus, we define the
+correlation matrix \(\boldsymbol{R}\) as:
+\begin{equation}\phantomsection\label{eq-mat54}{
+\boldsymbol{R} = \begin{bmatrix}
+    1 & 0 & 0 & 0 \\
+    0 & 1 & 0 & 0 \\
+    0 & 0 & 1 & 0 \\
+    0 & 0 & 0 & 1 
+  \end{bmatrix}
+}\end{equation}
+
+To resolve the \emph{scale} indeterminacy, we define the diagonal matrix
+\(\boldsymbol{D}\) as: \begin{equation}\phantomsection\label{eq-mat55}{
+\boldsymbol{D} = \begin{bmatrix}
+    s_{XI} & 0 & 0 & 0 \\
+    0 & p_{IA} & 0 & 0 \\
+    0 & 0 & s_{ZJ} & 0 \\
+    0 & 0 & 0 & p_{JK} 
+  \end{bmatrix}
+}\end{equation}
+
+Here, \(s_{XI}\) represents the standard deviation for the individuals,
+\(p_{IA}\) for the stimuli, \(s_{ZJ}\) for the judges, and \(p_{JK}\)
+for the judgments. We assume \(s_{XI}\) varies depending on the teaching
+method group to which each student belongs. In the example from
+Section~\ref{sec-theoretical}, where the teaching method
+\(X_{I} = \{1,2\}\), the model sets the constraint according to Equation
+\ref{eq-mat56}. This constraint anchors the scale of the individuals'
+latent trait while relaxing the assumption of equal dispersion for the
+stimuli, thereby addressing the concerns raised in
 Section~\ref{sec-theory-issue1a}.
-
-Fifth, we define \(B_{JK}\) as a linear combination of the judges'
-individual bias \(B_{J}\), the effects of relevant judgment-related
-variables \(\beta_{ZK}Z_{JK}\) (e.g., how the number of judgments a
-judge makes affect the assessment of quality), and judgment-specific
-idiosyncratic errors \(e_{JK}\). We assume that each element of
-\(e_{JK}\) is \(iid\) Normal with a mean of zero and a standard
-deviation \(p_{JK}\), defined as a proportion of \(1\), to anchor the
-scale of the latent trait \citep{Depaoli_2021}.
-
-Sixth, we define \(B_{J}\) as a linear combination of the effects of
-relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., the effects of
-judgment expertise on the quality assessment) and judge-specific
-idiosyncratic errors \(e_{J}\). Each element of \(e_{J}\) is likewise
-assumed to be \(iid\) Normal with mean zero and a standard deviation
-\(s_{ZJ}\) that depends on the groups to which each judge belongs. For
-instance, if \(Z_{J} = \{1,2,3\}\) representing three groups of judges
-with varying expertise, the model imposes the constraint
-\(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the scale of the latent trait
-\citep{Depaoli_2021}.
+\begin{equation}\phantomsection\label{eq-mat56}{
+\sum_{g=1}^{2} s_{XI}[g]/2 = 1
+}\end{equation}
+
+In contrast, we define \(p_{IA}\) as a proportion of \(1\) to establish
+the scale of the stimuli' latent trait relative to the scale of the
+individuals' trait. Note that, since the error vector \(\boldsymbol{e}\)
+follows an uncorrelated Multivariate Normal distribution, \(e_{IA}\)
+follows a univariate Normal distribution by the properties of the
+distribution. As a result, \(T_{IA}\) is also normally distributed,
+which effectively reverts to assuming Normal discriminal processes for
+the stimuli, an assumption originally proposed by
+\citet{Thurstone_1927b} (see Table~\ref{tbl-thurstone_cases})
+
+Similarly, we assume \(s_{ZJ}\) varies depending on the groups to which
+each judge belongs. For instance, if \(Z_{J} = \{1,2,3\}\) represents
+three groups of judges with varying expertise, the model sets the
+constraint: \begin{equation}\phantomsection\label{eq-mat56}{
+\sum_{g=1}^{2} s_{ZJ}[g]/3 = 1
+}\end{equation}
+
+This constraint anchors the scale of the judges' latent trait and
+relaxes the assumption of equal dispersion for the judgments.
+Conversely, \(p_{JK}\) is defined as a proportion of \(1\) to establish
+the scale of the judgments' latent trait relative to the scale of the
+judges' trait.
 
@@
 researchers assess complex traits or heterogeneous stimuli
-\citep{Thurstone_1927b}. In particular, he noted that although these
-assumptions simplify the trait measurement model, they could fail to
-capture the complexity of CJ data
-\citep{Kelly_et_al_2022, Andrich_1978}, potentially leading to
+\citep{Thurstone_1927b}. In particular, he noted that while assuming
+equal discriminal dispersions and zero correlation for the stimuli'
+discriminal processes simplifies the trait measurement model, these
+assumptions may fail to reflect the complexity of CJ data
+\citep{Kelly_et_al_2022, Andrich_1978}. As a result, they can lead to
 unreliable and inaccurate trait estimates

### 2025-05-14 paper2.qmd
section:5; process:write,equation,code

@@
 These error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables.
-We must resolve these indeterminacies to identify the model [@Depaoli_2021; @deAyala_2009].
+Thus, we must resolve these indeterminacies to identify the latent variable model [@Depaoli_2021; @deAyala_2009].
 Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix.
@@
 
-Following [SCM -@fig-cj16_scm], we assume uncorrelated errors to resolve the *orientation* indeterminacy. Thus, we define the correlation matrix $\boldsymbol{R}$ as:
+Following [SCM -@fig-cj16_scm], we resolve the *orientation* indeterminacy by assuming that the errors are uncorrelated.
+This assumption leads us to specify a correlation matrix for the errors, $\boldsymbol{R}$, as:
 $$
@@
 We assume $s_{XI}$ varies depending on the teaching method group to which each student belongs.
-In the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation [-@eq-mat56].
+In the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation ([-@eq-mat56]).
 This constraint anchors the scale of the individuals' latent trait while relaxing the assumption of equal dispersion for the stimuli, thereby addressing the concerns raised in @sec-theory-issue1a.
@@
 Similarly, we assume $s_{ZJ}$ varies depending on the groups to which each judge belongs.
-For instance, if $Z_{J} = \{1,2,3\}$ represents three groups of judges with varying expertise, the model sets the constraint:
+For instance, if $Z_{J} = \{1,2,3\}$ represents three groups of judges with varying expertise, the model sets the constraint according to Equation ([-@eq-mat57]).
+This constraint anchors the scale of the judges' latent trait and relaxes the assumption of equal dispersion for the judgments.
 $$
 \sum_{g=1}^{2} s_{ZJ}[g]/3 = 1
-$$ {#eq-mat56}
-
-This constraint anchors the scale of the judges' latent trait and relaxes the assumption of equal dispersion for the judgments.
+$$ {#eq-mat57}
+
 Conversely, $p_{JK}$ is defined as a proportion of $1$ to establish the scale of the judgments' latent trait relative to the scale of the judges' trait.

### 2025-05-14 _tex/paper2.tex
Built site for gh-pages

@@
 \emph{orientation}, and \emph{scale} due to the lack of an inherent
-scale in the associated latent variables. We must resolve these
-indeterminacies to identify the model
+scale in the associated latent variables. Thus, we must resolve these
+indeterminacies to identify the latent variable model
 \citep{Depaoli_2021, deAyala_2009}. Drawing on principles from SEM
@@
 
-Following SCM~\ref{fig-cj16_scm}, we assume uncorrelated errors to
-resolve the \emph{orientation} indeterminacy. Thus, we define the
-correlation matrix \(\boldsymbol{R}\) as:
-\begin{equation}\phantomsection\label{eq-mat54}{
+Following SCM~\ref{fig-cj16_scm}, we resolve the \emph{orientation}
+indeterminacy by assuming that the errors are uncorrelated. This
+assumption leads us to specify a correlation matrix for the errors,
+\(\boldsymbol{R}\), as: \begin{equation}\phantomsection\label{eq-mat54}{
 \boldsymbol{R} = \begin{bmatrix}
@@
 \(X_{I} = \{1,2\}\), the model sets the constraint according to Equation
-\ref{eq-mat56}. This constraint anchors the scale of the individuals'
+(\ref{eq-mat56}). This constraint anchors the scale of the individuals'
 latent trait while relaxing the assumption of equal dispersion for the
@@
 three groups of judges with varying expertise, the model sets the
-constraint: \begin{equation}\phantomsection\label{eq-mat56}{
+constraint according to Equation (\ref{eq-mat57}). This constraint
+anchors the scale of the judges' latent trait and relaxes the assumption
+of equal dispersion for the judgments.
+\begin{equation}\phantomsection\label{eq-mat57}{
 \sum_{g=1}^{2} s_{ZJ}[g]/3 = 1
@@
 
-This constraint anchors the scale of the judges' latent trait and
-relaxes the assumption of equal dispersion for the judgments.
 Conversely, \(p_{JK}\) is defined as a proportion of \(1\) to establish

### 2025-05-15 paper2.qmd
section:5,6; process:write

@@
 We assume $s_{XI}$ varies depending on the teaching method group to which each student belongs.
-In the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation ([-@eq-mat56]).
+Using the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation ([-@eq-mat56]).
 This constraint anchors the scale of the individuals' latent trait while relaxing the assumption of equal dispersion for the stimuli, thereby addressing the concerns raised in @sec-theory-issue1a.
@@
 
-In contrast, we define $p_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait.
-Note that, since the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, $e_{IA}$ follows a univariate Normal distribution by the properties of the distribution.
-As a result, $T_{IA}$ is also normally distributed, which effectively reverts to assuming Normal discriminal processes for the stimuli, an assumption originally proposed by @Thurstone_1927b (see @tbl-thurstone_cases)
+Because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$.
+We set $P_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait.
+As a result, $T_{IA}$ is also normally distributed.
+This configuration effectively reinstates Thurstone’s original assumption of Normal discriminal processes for the stimuli (see @tbl-thurstone_cases).
 
@@
 <!-- 6. Which are these new avenues for research? -->
-Five research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
+Five research avenues merit particular attention for their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
 First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect the robustness of statistical inferences.
@@
 Fourth, studies should explore how the hierarchical structure of stimuli influences both experimental design and hypothesis testing in CJ contexts.
-Finally,  studies should explore principled approaches for selecting contextually relevant variables.
+Finally, researchers should explore principled approaches for selecting contextually relevant variables.
 
@@
 This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand.
-While such translation is not exclusive to Bayesian methods--frequentist approaches can also be used--we argue that Bayesian methods offer a more accessible pathway,  as they reduce the need for additional expertise in optimization theory and techniques.
+While such translations are not exclusive to Bayesian methods, we argue that Bayesian approaches offer a more accessible path.
+They reduce, for instance, the need for additional expertise in optimization techniques and theory often required by frequentist methods, thereby lowering the technical barrier for many applied researchers.
 

### 2025-05-15 _tex/paper2.tex
Built site for gh-pages

@@
 for the judgments. We assume \(s_{XI}\) varies depending on the teaching
-method group to which each student belongs. In the example from
+method group to which each student belongs. Using the example from
 Section~\ref{sec-theoretical}, where the teaching method
@@
 
-In contrast, we define \(p_{IA}\) as a proportion of \(1\) to establish
-the scale of the stimuli' latent trait relative to the scale of the
-individuals' trait. Note that, since the error vector \(\boldsymbol{e}\)
-follows an uncorrelated Multivariate Normal distribution, \(e_{IA}\)
-follows a univariate Normal distribution by the properties of the
-distribution. As a result, \(T_{IA}\) is also normally distributed,
-which effectively reverts to assuming Normal discriminal processes for
-the stimuli, an assumption originally proposed by
-\citet{Thurstone_1927b} (see Table~\ref{tbl-thurstone_cases})
+Because the error vector \(\boldsymbol{e}\) follows an uncorrelated
+Multivariate Normal distribution, the marginal distribution of
+\(e_{IA}\) is a univariate Normal distribution with mean zero and
+standard deviation \(p_{IA}\). We set \(P_{IA}\) as a proportion of
+\(1\) to establish the scale of the stimuli' latent trait relative to
+the scale of the individuals' trait. As a result, \(T_{IA}\) is also
+normally distributed. This configuration effectively reinstates
+Thurstone's original assumption of Normal discriminal processes for the
+stimuli (see Table~\ref{tbl-thurstone_cases}).
 
@@
 
-Five research avenues deserve particular attention due to their direct
-impact on the reliability and validity of trait estimates and the
-accuracy of statistical inferences. First, researchers should
-investigate how sampling and comparison mechanisms influence CJ
-experimental outcomes and affect the robustness of statistical
-inferences. Second, further work is needed to understand how judges'
-biases impact the number of comparisons required to achieve reliable
-trait estimates. Third, researchers should re-evaluate existing methods
-for identifying and excluding `misfitting' judges and stimuli to ensure
-their effectiveness. Fourth, studies should explore how the hierarchical
-structure of stimuli influences both experimental design and hypothesis
-testing in CJ contexts. Finally, studies should explore principled
-approaches for selecting contextually relevant variables.
+Five research avenues merit particular attention for their direct impact
+on the reliability and validity of trait estimates and the accuracy of
+statistical inferences. First, researchers should investigate how
+sampling and comparison mechanisms influence CJ experimental outcomes
+and affect the robustness of statistical inferences. Second, further
+work is needed to understand how judges' biases impact the number of
+comparisons required to achieve reliable trait estimates. Third,
+researchers should re-evaluate existing methods for identifying and
+excluding `misfitting' judges and stimuli to ensure their effectiveness.
+Fourth, studies should explore how the hierarchical structure of stimuli
+influences both experimental design and hypothesis testing in CJ
+contexts. Finally, researchers should explore principled approaches for
+selecting contextually relevant variables.
 
@@
 models into bespoke statistical models that meaningfully address the
-research questions at hand. While such translation is not exclusive to
-Bayesian methods--frequentist approaches can also be used--we argue that
-Bayesian methods offer a more accessible pathway, as they reduce the
-need for additional expertise in optimization theory and techniques.
+research questions at hand. While such translations are not exclusive to
+Bayesian methods, we argue that Bayesian approaches offer a more
+accessible path. They reduce, for instance, the need for additional
+expertise in optimization techniques and theory often required by
+frequentist methods, thereby lowering the technical barrier for many
+applied researchers.
 

### 2025-05-20 paper2.qmd
section:6; process:write

@@
 <!-- and spoken language [@Boonen_et_al_2020].
-Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_et_al_2015], engineering design skills [@Bartholomew_et_al_2018], build real-time web-based portfolios of performance [@@Kimbell_2012], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020].
+Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_et_al_2015], engineering design skills [@Bartholomew_et_al_2018], build real-time web-based portfolios of performance [@@Kimbell_2012], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020b].
 -->
@@
 Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
-Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].
+Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020b; @Boonen_et_al_2020].
 
@@
 Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
-While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 and @vanDaal_et_al_2016 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
+While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003, @vanDaal_et_al_2016 and @Bartholomew_et_al_2020a demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
 
@@
 <!-- 1. How to derive an stat model from an SCM? -->
-With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a).
+With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption (see @sec-theory-issue1a).
 This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
@@
 <!-- 2. assumptions about the outcome -->
-Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_prob], and the statistical model [-@fig-cj16_stat].
+Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in @fig-cj16.
 We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
@@
 Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16.
-We make this omission because, although these distributions contribute to the overall joint distribution of the data, they do not influence the relationship between $O_{R}$ and $D_{R}$.
-This independence stems from the random selection designs that govern $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$[^8].
+We make this omission because, although these distributions contribute to the overall joint distribution of the data, the variables  $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$ are observed and independent of any other variable in the model.
+As observed variables, they do not require distributional assumptions--unlike the idiosyncratic errors.
+Moreover, their independence stems from the random selection designs that govern the variables[^8].
 
@@
 <!-- 4. functional form for D_{R} -->
-Next we define $D_{R}$ as the difference between the discriminal processes $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the texts, plus the corresponding repeated judge biases $B_{JK}[j, k]$.
+Next we define $D_{R}$ as the difference between the discriminal processes $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the compared texts, plus the corresponding repeated judge bias $B_{JK}[j, k]$.
 Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
@@
 This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022].
-In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases).
-To clarify, the square brackets in $D_{R}$ indicate the relevant indices for each trait vector; they do not imply any subsetting of the data.
+Put simply, this mathematical description captures the idea that the stimuli' discriminal processes become an observable outcome only through the lens of judges' perceptions (i.e., their biases).
+For clarity, the square brackets in $D_{R}$ indicate the relevant indices for each trait vector; they do not imply any subsetting of the data.
 
@@
 Next, we specify the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$.
-These error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables.
-Thus, we must resolve these indeterminacies to identify the latent variable model [@Depaoli_2021; @deAyala_2009].
+Unlike other variables in the model, these error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables $T_{I}$, $T_{IA}$, $B_{J}$, and $B_{JK}$.
+Thus, to identify the latent variable model, we must resolve these indeterminacies [@Depaoli_2021; @deAyala_2009].
 Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix.
@@
 
-Because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$.
+Note that because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$.
 We set $P_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait.
@@
 
-<!-- "This explicit part of the model it open ups the idea of doing more throughout research on ..." -->
-
-<!-- How about power?, you have a framework that allows to do this and more. -->
-
-<!-- talk about S and C -->
-
 <!-- 1. Summary of Thurstone's general theory -->
@@
 Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level.
-However, Thurstone identified one key challenge in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
+However, Thurstone identified one key challenge with its measurement model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
 To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from this general form, each progressively incorporating several simplifying assumptions.
@@
 Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b].
-In particular, he noted that while assuming equal discriminal dispersions and zero correlation for the stimuli' discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Kelly_et_al_2022; @Andrich_1978].
+In particular, he noted that while assuming equal discriminal dispersions and zero correlation for discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Thurstone_1927a; @Andrich_1978; @Kelly_et_al_2022].
 As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
 
-Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp.
+Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp.
 269], his theory did not specify how to use the resulting trait estimates for statistical inference.
@@
 Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference.
-Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the hypothesis test [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
@@
 
-The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model.
+The study then translates the SCM into a bespoke statistical model that addresses key limitations of Case V.
 This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2).
-In particular, the model accounts for judge biases, captures the hierarchical structure of stimuli, accommodates heterogeneity in discriminal dispersions, and incorporates measurement error in the hypothesis testing process.
+In particular, the model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli (refer to @sec-theory-theoretical_P3), incorporates measurement error in the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical).
 These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.
@@
 <!-- 6. Which are these new avenues for research? -->
-Five research avenues merit particular attention for their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
-First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect the robustness of statistical inferences.
-Second, further work is needed to understand how judges' biases impact the number of comparisons required to achieve reliable trait estimates.
-Third, researchers should re-evaluate existing methods for identifying and excluding 'misfitting' judges and stimuli to ensure their effectiveness.
-Fourth, studies should explore how the hierarchical structure of stimuli influences both experimental design and hypothesis testing in CJ contexts.
-Finally, researchers should explore principled approaches for selecting contextually relevant variables.
-
-<!-- 6.2 About the implications of the sampling and comparison mechanisms -->
-
-[About the implications of the sampling and comparison mechanisms]{style="color:blue;"}
-
-<!-- Continuing with the first advantage, now we clarify two additional processes of CJ experiments that previous literature do not typically mention: that the sample and comparison mechanism are generators of missing data--that is, they determine which observations do we get to see in our data sample.
-This opens the path to use Rubin's missing data framework [@Rubin_1976] into CJ applications to investigate more complex CJ designs, particularly the ones that are not random samples as the ones outlined here, which could be characterized in Rubin's structure parlor as missing at random (MAR) and missing not at random (MNAR) [@Rubin_1976].
-Among these cases are adaptive comparative algorithm designs, for which the causal structure is not clear, and thus, its impact on trait measurement and hypothesis testing is also not clear.
-Another  -->
-
-<!-- normally S and C are not explicitly mentioned  -->
-
-<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
-C -> {Y}
-e_{A} -> {C, Y} 
-this means that e_{A} is a confounder?,  
-then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?.
-Does this mean that random algorithms per se will always be unbiased?.
--->
-
-<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
-
-<!-- Brought Rubin's missing data framework to the field. -->
-<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are random samples, they fulfill ignorability. -->
+Five research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
+First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect inference robustness.
+Second, studies should assess the impact of judges' bias on the number of comparisons needed to achieve reliable trait estimates.
+Third, future work should re-evaluate current methods for identifying 'misfitting' judges and stimuli.
+Fourth, researchers should analyze how the hierarchical structure of stimuli affects both experimental design and hypothesis testing.
+Finally, studies should explore how modeling non-dichotomous outcomes or incorporating multidimensional traits affects the reliability and validity of the trait estimates.
+
+<!-- 6.1 About the implications of the sampling and comparison mechanisms -->
+The explicit integration of the sampling and comparison mechanisms within our conceptual model offers a novel perspective on the first research avenue.
+Specifically, it shows how these mechanisms influence the data-generating process by acting as missing data sources, determining which observations are part of the data sample.
+This new perspective invites the application of Little and Rubin's principled missing data framework [@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex experimental setups.
+
+<!-- --unlike prior literature, which merely acknowledged their presence in the experimental design. -->
+
+One prominent claim in the CJ literature is that Thurstone's model naturally handles non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].
+However, it is easy to imagine scenarios where this assumption may not hold.
+For instance, selecting a particular group of judges could result in an incomplete representation of the trait's full range and complexity, as judgment accuracy varies with demographic factors [@Bartholomew_et_al_2020a; @Kelly_et_al_2022] and even between individuals [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020].
+Consequently, judges' selection could affect the formation of a "shared consensus", potentially biasing trait estimates and distorting the observed trait distribution [@Deffner_et_al_2022].
+It is therefore essential to determine whether such sampling biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences.
+Our framework offers a rigorous setting to test the scenario.
+
+[another example: For instance, consider a CJ experiment assessing speech quality using a case-control design [@Everitt_et_al_2010] that samples individuals with normal hearing (NH) and those with hearing impairment (HI).
+Here, the sampling mechanism depends on hearing status--a variable likely to influence the trait of interest, as HI individuals may exhibit lower speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024].
+Suppose a subgroup within the HI population, such as individuals with genetically induced hearing loss, is difficult to reach.
+If this group is systematically excluded and happens to over- or under-perform in speech quality, the resulting trait estimates may be biased.
+It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"}
+
+[another example: For instance, consider a high-stakes non-mandatory summative CJ assessment, where the results the individuals have in this assessment have some implications for them, however, their participation is not mandatory.
+In this case, it not hard to imagine that some of these individuals will not participate in the CJ assessment, thus affecting the sample, and ultimately the trait estimates: now a probable under performing group will not available in the data sample, thus biasing the resulting trait estimates.
+It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"}
 
@@
 256] -->
-
-<!-- Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates.
-In this sense the models are also "sample-free."
-However, this property of the model is true in empirical data only if the model holds in the data.
-Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
-[@Andrich_1978, pp.459] -->
 
@@
 
-<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
-However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
-
-<!-- power analysis to know the implications of certain sampling is nice to have it explicitly. -->
-
-<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
-Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp.
-683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement.
-As @Kelly_et_al_2022 noted [pp.
-683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->
-
-<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait.
---->
-
-<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
+<!-- The logic is that in aggregating the varied judgments, one comes to an overall view of the community of practice.
+However, this raises important questions about diversity of the judging pool.
+If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
 Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
+For example, Bartholomew et al.
+(2020) found differences between what judges from the UK, Sweden, and the USA valued when assessing design portfolios, providing evidence to support divergence along demographic lines.
 [@Kelly_et_al_2022, pp.
 683] -->
-
-<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
-[@Kelly_et_al_2022, pp.
-681] -->
-
-<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology.
-We do not yet know how widely it can be, or should be, applied.
-[@Pollitt_2012b, pp.
-297]-->
 
@@
 
-<!-- 6.3 About judges biases, judgment discrepancies, and the number of judges and judgments -->
-
-[About judges biases, judgment discrepancies, and the number of judges and judgments]{style="color:blue;"}
-
-<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
-Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
-It also accounts for potential judge biases in the comparative judgment process.
--->
+Moreover, when designing and evaluating more complex experimental setups, our framework highlights one key principle: the sampling and comparison mechanisms directly influence the CJ outcome.
+In this study, we deliberately structured these mechanisms to produce data that is *missing completely at random* (MCAR)--that is, the missingness is unrelated to any observed or unobserved variables, including the outcome [@Little_et_al_2020].
+This design offers one key advantage: it generates simple random samples that satisfy the condition of ignorability, enabling researchers to legitimately ignore missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019].
+
+However, many modern CJ applications involve more complex forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) mechanism [@Little_et_al_2020]--that arise from the chosen sampling and comparison methods.
+Two notable examples are adaptive comparative judgment designs [@Pollitt_2012a; @Pollitt_2012b] and case-control studies [@Everitt_et_al_2010].
+In adaptive designs, an algorithm defines the comparison mechanism by dynamically selecting which stimulus pairs judges evaluate.
+These designs have produced mixed evidence: some studies report improved trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others suggest these gains may be artificially inflated [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
+In contrast, case-control studies rely on sampling mechanisms linked to non-assignable variables that likely influence the trait of interest.
+For instance, @Boonen_et_al_2020 selected children based on hearing status, a non-assignable variable that likely affects speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024].
+Nevertheless, regardless of their missingness mechanism, both designs could benefit from explicitly articulating their underlying causal structures.
+Doing so would enable researchers to assess and test their impact on trait estimation and statistical inference.
+Our framework provides a rigorous foundation for such articulations.
+
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
+
+<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
+However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
+
+<!-- Comment for myself: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
+{S, C} -> {Y}
+e_{A} -> {S, C, Y} 
+this means that e_{A} is a confounder?,  
+then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C, S} by taking a random sample of stimuli?.
+Does this mean that random algorithms per se will always be unbiased?.
+-->
+
+<!-- 6.2 About judges biases, judgment discrepancies, and the number of judges and judgments -->
+Additionally, by explicitly incorporating judges' biases, our framework provides a novel approach to investigate the second research avenue.
+Specifically, it enables researchers to determine whether such biases occur and, if so, to assess their impact on the reliability and validity of trait estimation.
+Although some anecdotal evidence already suggests judges' biases exist [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], further research is needed to address several key questions.
+For instance: Can CJ data be analyzed under the assumption of "sample-free" trait calibration--that is, the estimated trait difference for any two scripts does not depend on which other scripts they are compared with, implying an absence of judges' bias?
+[@Bramley_2008; @Andrich_1978] If this assumption does not hold, how many judges and judgments are required to ensure acceptable levels of trait reliability and validity?
+Do training and expertise influence judgment biases, and to what extent?
+[@Kelly_et_al_2022] Finally, what are the benefits of *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times?
+[@Lawson_2015]
 
@@
 
+<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
+[@Kelly_et_al_2022, pp.
+681] -->
+
+<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology.
+We do not yet know how widely it can be, or should be, applied.
+[@Pollitt_2012b, pp.
+297]-->
+
 <!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
@@
 
-<!-- presence of judges biases in the comparative judgment procedure. From this it derives research on:  -->
-<!-- - does the judges have biases, and if so, do ad-hoc methods can identify them -->
-
-<!-- Given that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, this study has argued that judges biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
--->
-
-<!-- This hierarchical nature open paths for research in, for instance, the benefits of *repeated measures designs* [@Lawson_2015, pp.
-366-376] in CJ experiments.
-This is when the experimental conditions accounts for judges comparing the same pair of stimuli multiple times.
-Or research that determines the minimum number of judges and judgments required to ensure an acceptable level of reliability and validity under particular assumptions of the data [@Kelly_et_al_2022, pp.
-682].
--->
-
-<!-- 6.4 About removing misfit judges and stimuli  -->
+<!-- The elimination of the subject parameters in the Rasch model has given rise to such expressions as "sample-free"" and "population-free." In this sense, then, so is the Thurstone Case V model.
+Although these phrases suggest desirable properties, they are not always fully appreciated; therefore, these terms have been avoided so far.
+A clarification seems in order and possible in this context.
+Because subject parameters are eliminated explicitly, their values (individually or collectively) need not be known.
+As a result, no assumptions need be made about the distribution of these parameters in the population, for example, that they are normally distributed.
+This does not mean, of course, that the "empirical" subject population need not be specified or described or that the scale values will be the same in different empirical populations.
+In fact, it is generally vital that the empirical populations be described.
+But because it does not matter what the actual values of the subject parameters are in the population, the models are, in the statistical distribution sense of population, "population-free." This is a property of the model and independent of empirical data.
+It is important to note that this holds only with respect to the estimation of the relative scale values of the items.
+It does not matter if the subject parameters are large, small, or otherwise; the subjects’ impressions of the relative positions of the items should be the same.
+That is, items which appear more intense to one group should appear similarly more intense to the other groups.
+Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates.
+In this sense the models are also "sample-free."
+However, this property of the model is true in empirical data only if the model holds in the data.
+Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
+...
+Note that it would be missing a crucial point to choose two random samples from which to check "sample-freeness" and that it would not be putting the models to any special test.
+Two random samples should give the same relative scale values because it is an inherent assumption of all statistical models that parameter estimates are "random sample-free." For the greatest benefits, the samples should be systematically different.
+[@Andrich_1978, pp.459] -->
+
+<!-- 6.3 About removing misfit judges and stimuli  -->
 
 [About removing misfit judges and stimuli]{style="color:blue;"}
+
+<!-- "This explicit part of the model it open ups the idea of doing more throughout research on ..." -->
 
@@
 
-<!-- 6.1 About the hierarchical structure of stimuli, number and predictors -->
+<!-- 6.4 About the hierarchical structure of stimuli, number and predictors -->
 
 [About the hierarchical structure of stimuli]{style="color:blue;"}
+
+<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
+Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
+-->
 
@@
 
-<!-- 6.5 About using Causal inference to carefully the most relevant variables for specific research contexts -->
-
-[About revisit the approaches for selecting contextually relevant variables]{style="color:blue;"}
+<!-- 6.5 About non-dichotomous outcomes and multidimensional traits -->
+
+[About non-dichotomous outcomes and multidimensional traits]{style="color:blue;"}
+
+<!-- Talk about Bayesian Workflow, and how it would fit the process to arrive to the proposed model. -->
+
+<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
+
+<!-- @Bramley_2008 proposes ranking instead of comparisons -->
+
+While a natural step would be to formally derive answers to all these research avenues--similar to approaches used in classical models like linear regression--we argue that the complexity of the CJ system makes this route impractical.
+Instead, simulation-based methods, like power analysis, offer a more feasible alternative, allowing researchers to investigate these issues without relying on cumbersome mathematical proofs.
+
+<!-- 7. Challenges -->
+
+<!-- 7.1 PPL and Bayesian Inference -->
+Despite its potential, this study presents several challenges that researchers must navigate.
+First, researchers need to be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods.
+This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand.
+While such translations are not exclusive to Bayesian methods, we argue that Bayesian approaches offer a more accessible path.
+They reduce, for instance, the need for additional expertise in optimization techniques and theory often required by frequentist methods, thereby lowering the technical barrier for many applied researchers.
+
+<!-- 7.2 Research context is important  -->
+Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts.
+We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
+That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.
+
+<!-- 7.3 Faithfulness  -->
+Third,
+
+<!-- remember to mention fidelity of the theory in the model -->
 
@@
 
-<!-- remember to mention fidelity of the theory in the model -->
-
 <!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
@@
 
-<!-- Talk about Bayesian Workflow, and how it would fit the process to arrive to the proposed model. -->
-
-<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
-
-<!-- @Bramley_2008 proposes ranking instead of comparisons -->
-
-<!-- 7. Challenges -->
-<!-- 7.1 Epistemiology  -->
-Despite its potential, this study presents several challenges that researchers must navigate.
-First, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate.
+<!-- 7.4 Causal inference knowledge -->
+Fourth, applying this approach requires a solid understanding of Causal inference.
+For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity.
+Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.
+
+<!-- 7.5 Epistemiology  -->
+Fifth and final, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate.
 In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest?
@@
 Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.
-
-<!-- 7.2 Research context is important  -->
-Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts.
-We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
-That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.
-
-<!-- 7.3 Causal inference knowledge -->
-Third, applying this approach requires a solid understanding of Causal inference.
-For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity.
-Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.
-
-<!-- 7.4 PPL and Bayesian Inference -->
-Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods.
-This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand.
-While such translations are not exclusive to Bayesian methods, we argue that Bayesian approaches offer a more accessible path.
-They reduce, for instance, the need for additional expertise in optimization techniques and theory often required by frequentist methods, thereby lowering the technical barrier for many applied researchers.
 

### 2025-05-20 _tex/paper2.tex
Built site for gh-pages

@@
 versatility across both educational and non-educational contexts
-\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
+\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020b, Boonen_et_al_2020}.
 
@@
 this particular scenario is lacking, studies such as
-\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
-the presence of such biases, supporting the notion that the factors
-influencing pairwise comparisons may not always cancel out.
+\citet{Pollitt_et_al_2003}, \citet{vanDaal_et_al_2016} and
+\citet{Bartholomew_et_al_2020a} demonstrate the presence of such biases,
+supporting the notion that the factors influencing pairwise comparisons
+may not always cancel out.
 
@@
 statistical model that addresses violations of the equal dispersion
-assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
-This derivation is possible because a fully specified SCM encodes
-functional and probabilistic information, which we can replace with
-suitable functions and probabilistic assumptions
-\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
-to express the joint distribution of our complex CJ system as a product
-of simpler conditional probability distributions (CPDs)\footnote{This
-  re-expression is possible because the \emph{chain rule} of probability
-  and the \emph{Bayesian Network Factorization (BNF)} property. For
-  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
-as shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
-such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
-as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
-CPD of \(Y\) given \(X\).
-\begin{equation}\phantomsection\label{eq-mat51}{
+assumption (see Section~\ref{sec-theory-issue1a}). This derivation is
+possible because a fully specified SCM encodes functional and
+probabilistic information, which we can replace with suitable functions
+and probabilistic assumptions \citep{Pearl_et_al_2016}. Specifically,
+SCM~\ref{fig-cj15_scm} allows us to express the joint distribution of
+our complex CJ system as a product of simpler conditional probability
+distributions (CPDs)\footnote{This re-expression is possible because the
+  \emph{chain rule} of probability and the \emph{Bayesian Network
+  Factorization (BNF)} property. For further details, see
+  \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as shown in Equation
+(\ref{eq-mat51}). For clarity, we treat expressions such as
+\(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\) as
+equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the CPD
+of \(Y\) given \(X\). \begin{equation}\phantomsection\label{eq-mat51}{
 \begin{aligned}
@@
 Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
-which we outline in SCM~\ref{fig-cj16_scm}, the probabilistic model
-\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}. We
-begin by assuming that \(O_{R}\) follows a Bernoulli
-distribution\footnote{The binomial distribution--including its special
-  case, the Bernoulli distribution--represent a maximum entropy
-  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
-  means that the Bernoulli distribution is the most consistent
-  alternative when only two un-ordered outcomes are possible and their
-  expected frequencies are assumed to be constant
-  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
-  binomial as a maximum entropy distribution, see
+which we outline in Figure~\ref{fig-cj16}. We begin by assuming that
+\(O_{R}\) follows a Bernoulli distribution\footnote{The binomial
+  distribution--including its special case, the Bernoulli
+  distribution--represent a maximum entropy distribution for binary
+  events \citep[pp.~34]{McElreath_2020}. This means that the Bernoulli
+  distribution is the most consistent alternative when only two
+  un-ordered outcomes are possible and their expected frequencies are
+  assumed to be constant \citep[pp.~310]{McElreath_2020}. For a detailed
+  discussion of the binomial as a maximum entropy distribution, see
   \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
@@
 distributions contribute to the overall joint distribution of the data,
-they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
-This independence stems from the random selection designs that govern
-\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
-\(Z_{J}\)\footnote{Randomization ensures that data--and, by extension,
+the variables \(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
+\(Z_{J}\) are observed and independent of any other variable in the
+model. As observed variables, they do not require distributional
+assumptions--unlike the idiosyncratic errors. Moreover, their
+independence stems from the random selection designs that govern the
+variables\footnote{Randomization ensures that data--and, by extension,
   an estimator--satisfies several key identification properties, such as
@@
 processes \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), representing the
-underlying written-quality trait of the texts, plus the corresponding
-repeated judge biases \(B_{JK}[j, k]\). Note that if we assume that
-\(B_{JK}[j,k]\) reflects the difference in stimulus-specific biases,
-i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can
-re-write the discriminal difference as:
+underlying written-quality trait of the compared texts, plus the
+corresponding repeated judge bias \(B_{JK}[j, k]\). Note that if we
+assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
+biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
+can re-write the discriminal difference as:
 \begin{equation}\phantomsection\label{eq-mat52}{
@@
 processes nor the judges' biases alone determine the outcome, but their
-interaction does \citep{Attia_et_al_2022}. In simple terms, the
-discriminal processes of the stimuli become an observable outcome only
-through the judges' perceptions (biases). To clarify, the square
-brackets in \(D_{R}\) indicate the relevant indices for each trait
-vector; they do not imply any subsetting of the data.
+interaction does \citep{Attia_et_al_2022}. Put simply, this mathematical
+description captures the idea that the stimuli' discriminal processes
+become an observable outcome only through the lens of judges'
+perceptions (i.e., their biases). For clarity, the square brackets in
+\(D_{R}\) indicate the relevant indices for each trait vector; they do
+not imply any subsetting of the data.
 
@@
 Next, we specify the probabilistic assumptions for the idiosyncratic
-errors \(e_{I}\), \(e_{IA}\), \(e_{J}\), and \(e_{JK}\). These error
-terms exhibit indeterminacies in their \emph{location},
-\emph{orientation}, and \emph{scale} due to the lack of an inherent
-scale in the associated latent variables. Thus, we must resolve these
-indeterminacies to identify the latent variable model
+errors \(e_{I}\), \(e_{IA}\), \(e_{J}\), and \(e_{JK}\). Unlike other
+variables in the model, these error terms exhibit indeterminacies in
+their \emph{location}, \emph{orientation}, and \emph{scale} due to the
+lack of an inherent scale in the associated latent variables \(T_{I}\),
+\(T_{IA}\), \(B_{J}\), and \(B_{JK}\). Thus, to identify the latent
+variable model, we must resolve these indeterminacies
 \citep{Depaoli_2021, deAyala_2009}. Drawing on principles from SEM
@@
 
-Because the error vector \(\boldsymbol{e}\) follows an uncorrelated
-Multivariate Normal distribution, the marginal distribution of
-\(e_{IA}\) is a univariate Normal distribution with mean zero and
+Note that because the error vector \(\boldsymbol{e}\) follows an
+uncorrelated Multivariate Normal distribution, the marginal distribution
+of \(e_{IA}\) is a univariate Normal distribution with mean zero and
 standard deviation \(p_{IA}\). We set \(P_{IA}\) as a proportion of
@@
 indicating which stimulus the judge perceives as having a higher trait
-level. However, Thurstone identified one key challenge in this model: it
-required estimating more ``unknown'' parameters than the number of
-available pairwise comparisons \citep{Thurstone_1927b}. To address this
-issue and facilitate the theory's practical applicability, he formulated
-five cases derived from this general form, each progressively
-incorporating several simplifying assumptions.
+level. However, Thurstone identified one key challenge with its
+measurement model: it required estimating more ``unknown'' parameters
+than the number of available pairwise comparisons
+\citep{Thurstone_1927b}. To address this issue and facilitate the
+theory's practical applicability, he formulated five cases derived from
+this general form, each progressively incorporating several simplifying
+assumptions.
 
@@
 \citep{Thurstone_1927b}. In particular, he noted that while assuming
-equal discriminal dispersions and zero correlation for the stimuli'
-discriminal processes simplifies the trait measurement model, these
-assumptions may fail to reflect the complexity of CJ data
-\citep{Kelly_et_al_2022, Andrich_1978}. As a result, they can lead to
-unreliable and inaccurate trait estimates
+equal discriminal dispersions and zero correlation for discriminal
+processes simplifies the trait measurement model, these assumptions may
+fail to reflect the complexity of CJ data
+\citep{Thurstone_1927a, Andrich_1978, Kelly_et_al_2022}. As a result,
+they can lead to unreliable and inaccurate trait estimates
 \citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.
 
-Furthermore, because Thurstone's primary goal was to produce a ``coarse
-scaling'' of traits and ``allocate the compared stimuli on this
+Furthermore, because Thurstone aimed primarily to produce a trait's
+``coarse scaling'' and ``allocate the compared stimuli on this
 continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
@@
 introduce bias into the analysis and compromise the reliability of the
-resulting conclusions
+hypothesis test
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
@@
 The study then translates the SCM into a bespoke statistical model that
-addresses key limitations of the BTL model. This model allows
-researchers to analyze CJ data when the assumptions of equal dispersion
-and zero correlation do not hold (see Section~\ref{sec-theory-issue1a}
-and Section~\ref{sec-theory-issue1b}), and when they need to perform
+addresses key limitations of Case V. This model allows researchers to
+analyze CJ data when the assumptions of equal dispersion and zero
+correlation do not hold (see Section~\ref{sec-theory-issue1a} and
+Section~\ref{sec-theory-issue1b}), and when they need to perform
 statistical inference (refer to Section~\ref{sec-theory-issue2}). In
-particular, the model accounts for judge biases, captures the
-hierarchical structure of stimuli, accommodates heterogeneity in
-discriminal dispersions, and incorporates measurement error in the
-hypothesis testing process. These methodological innovations can
+particular, the model accounts for judge biases (see
+Section~\ref{sec-theory-theoretical_P1}), captures the hierarchical
+structure of stimuli (refer to Section~\ref{sec-theory-theoretical_P3}),
+incorporates measurement error in the hypothesis testing process (refer
+to Section~\ref{sec-theory-theoretical_P3}), and accommodates
+heterogeneity in discriminal dispersions (see
+Section~\ref{sec-statistical}). These methodological innovations can
 potentially enhance the reliability and validity of trait measurement
@@
 
-Five research avenues merit particular attention for their direct impact
-on the reliability and validity of trait estimates and the accuracy of
-statistical inferences. First, researchers should investigate how
-sampling and comparison mechanisms influence CJ experimental outcomes
-and affect the robustness of statistical inferences. Second, further
-work is needed to understand how judges' biases impact the number of
-comparisons required to achieve reliable trait estimates. Third,
-researchers should re-evaluate existing methods for identifying and
-excluding `misfitting' judges and stimuli to ensure their effectiveness.
-Fourth, studies should explore how the hierarchical structure of stimuli
-influences both experimental design and hypothesis testing in CJ
-contexts. Finally, researchers should explore principled approaches for
-selecting contextually relevant variables.
-
-{About the implications of the sampling and comparison mechanisms}
-
-{About judges biases, judgment discrepancies, and the number of judges
-and judgments}
+Five research avenues deserve particular attention due to their direct
+impact on the reliability and validity of trait estimates and the
+accuracy of statistical inferences. First, researchers should
+investigate how sampling and comparison mechanisms influence CJ
+experimental outcomes and affect inference robustness. Second, studies
+should assess the impact of judges' bias on the number of comparisons
+needed to achieve reliable trait estimates. Third, future work should
+re-evaluate current methods for identifying `misfitting' judges and
+stimuli. Fourth, researchers should analyze how the hierarchical
+structure of stimuli affects both experimental design and hypothesis
+testing. Finally, studies should explore how modeling non-dichotomous
+outcomes or incorporating multidimensional traits affects the
+reliability and validity of the trait estimates.
+
+The explicit integration of the sampling and comparison mechanisms
+within our conceptual model offers a novel perspective on the first
+research avenue. Specifically, it shows how these mechanisms influence
+the data-generating process by acting as missing data sources,
+determining which observations are part of the data sample. This new
+perspective invites the application of Little and Rubin's principled
+missing data framework \citep{Little_et_al_2020}, enabling a more
+rigorous evaluation of existing claims about these mechanisms, their
+influence on CJ outcomes, and their implications for designing and
+evaluating more complex experimental setups.
+
+One prominent claim in the CJ literature is that Thurstone's model
+naturally handles non-random missing data without compromising the
+reliability or validity of trait estimates \citep{Bramley_2008}.
+However, it is easy to imagine scenarios where this assumption may not
+hold. For instance, selecting a particular group of judges could result
+in an incomplete representation of the trait's full range and
+complexity, as judgment accuracy varies with demographic factors
+\citep{Bartholomew_et_al_2020a, Kelly_et_al_2022} and even between
+individuals
+\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
+Consequently, judges' selection could affect the formation of a ``shared
+consensus'', potentially biasing trait estimates and distorting the
+observed trait distribution \citep{Deffner_et_al_2022}. It is therefore
+essential to determine whether such sampling biases occur--and if so,
+whether they undermine the validity of trait estimates and the
+robustness of statistical inferences. Our framework offers a rigorous
+setting to test the scenario.
+
+{another example: For instance, consider a CJ experiment assessing
+speech quality using a case-control design \citep{Everitt_et_al_2010}
+that samples individuals with normal hearing (NH) and those with hearing
+impairment (HI). Here, the sampling mechanism depends on hearing
+status--a variable likely to influence the trait of interest, as HI
+individuals may exhibit lower speech quality
+\citep{Boonen_et_al_2023, Rivera_et_al_2024}. Suppose a subgroup within
+the HI population, such as individuals with genetically induced hearing
+loss, is difficult to reach. If this group is systematically excluded
+and happens to over- or under-perform in speech quality, the resulting
+trait estimates may be biased. It is therefore essential to determine
+whether such biases occur--and if so, whether they undermine the
+validity of trait estimates and the robustness of statistical
+inferences--and our framework provides a rigorous setting to test this
+claim.}
+
+{another example: For instance, consider a high-stakes non-mandatory
+summative CJ assessment, where the results the individuals have in this
+assessment have some implications for them, however, their participation
+is not mandatory. In this case, it not hard to imagine that some of
+these individuals will not participate in the CJ assessment, thus
+affecting the sample, and ultimately the trait estimates: now a probable
+under performing group will not available in the data sample, thus
+biasing the resulting trait estimates. It is therefore essential to
+determine whether such biases occur--and if so, whether they undermine
+the validity of trait estimates and the robustness of statistical
+inferences--and our framework provides a rigorous setting to test this
+claim.}
+
+Moreover, when designing and evaluating more complex experimental
+setups, our framework highlights one key principle: the sampling and
+comparison mechanisms directly influence the CJ outcome. In this study,
+we deliberately structured these mechanisms to produce data that is
+\emph{missing completely at random} (MCAR)--that is, the missingness is
+unrelated to any observed or unobserved variables, including the outcome
+\citep{Little_et_al_2020}. This design offers one key advantage: it
+generates simple random samples that satisfy the condition of
+ignorability, enabling researchers to legitimately ignore missing data
+during analysis without introducing bias
+\citep{Everitt_et_al_2010, Kohler_et_al_2019}.
+
+However, many modern CJ applications involve more complex forms of
+missingness--such as \emph{missing at random} (MAR) or \emph{missing not
+at random} (MNAR) mechanism \citep{Little_et_al_2020}--that arise from
+the chosen sampling and comparison methods. Two notable examples are
+adaptive comparative judgment designs
+\citep{Pollitt_2012a, Pollitt_2012b} and case-control studies
+\citep{Everitt_et_al_2010}. In adaptive designs, an algorithm defines
+the comparison mechanism by dynamically selecting which stimulus pairs
+judges evaluate. These designs have produced mixed evidence: some
+studies report improved trait reliability
+\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
+suggest these gains may be artificially inflated
+\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
+In contrast, case-control studies rely on sampling mechanisms linked to
+non-assignable variables that likely influence the trait of interest.
+For instance, \citet{Boonen_et_al_2020} selected children based on
+hearing status, a non-assignable variable that likely affects speech
+quality \citep{Boonen_et_al_2023, Rivera_et_al_2024}. Nevertheless,
+regardless of their missingness mechanism, both designs could benefit
+from explicitly articulating their underlying causal structures. Doing
+so would enable researchers to assess and test their impact on trait
+estimation and statistical inference. Our framework provides a rigorous
+foundation for such articulations.
+
+Additionally, by explicitly incorporating judges' biases, our framework
+provides a novel approach to investigate the second research avenue.
+Specifically, it enables researchers to determine whether such biases
+occur and, if so, to assess their impact on the reliability and validity
+of trait estimation. Although some anecdotal evidence already suggests
+judges' biases exist
+\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
+further research is needed to address several key questions. For
+instance: Can CJ data be analyzed under the assumption of
+``sample-free'' trait calibration--that is, the estimated trait
+difference for any two scripts does not depend on which other scripts
+they are compared with, implying an absence of judges' bias?
+\citep{Bramley_2008, Andrich_1978} If this assumption does not hold, how
+many judges and judgments are required to ensure acceptable levels of
+trait reliability and validity? Do training and expertise influence
+judgment biases, and to what extent? \citep{Kelly_et_al_2022} Finally,
+what are the benefits of \emph{repeated measures designs}, in which
+judges evaluate the same stimulus pairs multiple times?
+\citep{Lawson_2015}
 
@@
 
-{About revisit the approaches for selecting contextually relevant
-variables}
+{About non-dichotomous outcomes and multidimensional traits}
+
+While a natural step would be to formally derive answers to all these
+research avenues--similar to approaches used in classical models like
+linear regression--we argue that the complexity of the CJ system makes
+this route impractical. Instead, simulation-based methods, like power
+analysis, offer a more feasible alternative, allowing researchers to
+investigate these issues without relying on cumbersome mathematical
+proofs.
 
 Despite its potential, this study presents several challenges that
-researchers must navigate. First, our approach does not establish a
-clear epistemological connection between the theoretical latent trait
-and its quantitative estimate. In other words, the approach does not
-provide a definitive answer to the fundamental question: How do we know
-that the estimated trait reflects the theoretical trait of interest?
-However, this limitation is not unique to our approach; it reflects a
-fundamental challenge all research involving latent constructs faces.
-Nevertheless, we argue that our approach offers methodological tools
-that enable researchers to design CJ experiments under conditions that
-support reliable and valid trait measurement.
+researchers must navigate. First, researchers need to be proficient in
+Probabilistic Programming Languages (PPLs) and Bayesian inference
+methods. This expertise is essential for translating context-specific
+conceptual models into bespoke statistical models that meaningfully
+address the research questions at hand. While such translations are not
+exclusive to Bayesian methods, we argue that Bayesian approaches offer a
+more accessible path. They reduce, for instance, the need for additional
+expertise in optimization techniques and theory often required by
+frequentist methods, thereby lowering the technical barrier for many
+applied researchers.
 
@@
 
-Third, applying this approach requires a solid understanding of Causal
+Third,
+
+Fourth, applying this approach requires a solid understanding of Causal
 inference. For instance, throughout the study, we assume that predictor
@@
 
-Fourth and final, researchers must also be proficient in Probabilistic
-Programming Languages (PPLs) and Bayesian inference methods. This
-expertise is essential for translating context-specific conceptual
-models into bespoke statistical models that meaningfully address the
-research questions at hand. While such translations are not exclusive to
-Bayesian methods, we argue that Bayesian approaches offer a more
-accessible path. They reduce, for instance, the need for additional
-expertise in optimization techniques and theory often required by
-frequentist methods, thereby lowering the technical barrier for many
-applied researchers.
+Fifth and final, our approach does not establish a clear epistemological
+connection between the theoretical latent trait and its quantitative
+estimate. In other words, the approach does not provide a definitive
+answer to the fundamental question: How do we know that the estimated
+trait reflects the theoretical trait of interest? However, this
+limitation is not unique to our approach; it reflects a fundamental
+challenge all research involving latent constructs faces. Nevertheless,
+we argue that our approach offers methodological tools that enable
+researchers to design CJ experiments under conditions that support
+reliable and valid trait measurement.
 

### 2025-05-21 paper2.qmd
section:6; process:write

@@
 Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b].
-In particular, he noted that while assuming equal discriminal dispersions and zero correlation for discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Thurstone_1927a; @Andrich_1978; @Kelly_et_al_2022].
+Specifically, while assuming assuming equal discriminal dispersions and zero correlation for discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Thurstone_1927a; @Andrich_1978; @Kelly_et_al_2022].
 As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
@@
 269], his theory did not specify how to use the resulting trait estimates for statistical inference.
-The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process.
-Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference.
-Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the hypothesis test [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing.
+Specifically, CJ studies often rely on point estimates of the traits--typically the BTL scores or their transformations--to conduct statistical inference.
+Although this separation simplifies CJ data analysis, the statistical literature warns that this practice can introduce bias into the analysis and compromise the reliability of the hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
@@
 Specifically, it shows how these mechanisms influence the data-generating process by acting as missing data sources, determining which observations are part of the data sample.
-This new perspective invites the application of Little and Rubin's principled missing data framework [@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex experimental setups.
+This new perspective invites the application of Little and Rubin's [-@Little_et_al_2020] principled missing data framework, enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex experimental setups.
 
@@
 Consequently, judges' selection could affect the formation of a "shared consensus", potentially biasing trait estimates and distorting the observed trait distribution [@Deffner_et_al_2022].
-It is therefore essential to determine whether such sampling biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences.
-Our framework offers a rigorous setting to test the scenario.
+It is therefore essential to determine whether such sampling biases undermine the validity of trait estimates and the robustness of statistical inferences.
+Our framework offers a rigorous setting to test this scenario.
 
@@
 In this study, we deliberately structured these mechanisms to produce data that is *missing completely at random* (MCAR)--that is, the missingness is unrelated to any observed or unobserved variables, including the outcome [@Little_et_al_2020].
-This design offers one key advantage: it generates simple random samples that satisfy the condition of ignorability, enabling researchers to legitimately ignore missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019].
+This design offers one key advantage: it generates simple random samples that satisfy the condition of ignorability, enabling researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019].
 
@@
 Doing so would enable researchers to assess and test their impact on trait estimation and statistical inference.
-Our framework provides a rigorous foundation for such articulations.
+Our framework, again, provides a rigorous foundation for such articulations.
 
@@
 Specifically, it enables researchers to determine whether such biases occur and, if so, to assess their impact on the reliability and validity of trait estimation.
-Although some anecdotal evidence already suggests judges' biases exist [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], further research is needed to address several key questions.
-For instance: Can CJ data be analyzed under the assumption of "sample-free" trait calibration--that is, the estimated trait difference for any two scripts does not depend on which other scripts they are compared with, implying an absence of judges' bias?
-[@Bramley_2008; @Andrich_1978] If this assumption does not hold, how many judges and judgments are required to ensure acceptable levels of trait reliability and validity?
+Although some anecdotal evidence already suggests judges' biases exist [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], further research is needed to address several critical questions.
+For instance, can CJ data be analyzed under the assumption of "sample-free" trait calibration, which implies an absence of judges' bias?
+[@Bramley_2008; @Andrich_1978] If this assumption does not hold, how many judges and judgments per judge are required to ensure acceptable levels of trait reliability and validity?
 Do training and expertise influence judgment biases, and to what extent?
-[@Kelly_et_al_2022] Finally, what are the benefits of *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times?
+[@Kelly_et_al_2022] Finally, what are the benefits of *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times, in assessing judgment accuracy?
 [@Lawson_2015]

### 2025-05-21 _tex/paper2.tex
Built site for gh-pages

@@
 researchers assess complex traits or heterogeneous stimuli
-\citep{Thurstone_1927b}. In particular, he noted that while assuming
-equal discriminal dispersions and zero correlation for discriminal
-processes simplifies the trait measurement model, these assumptions may
-fail to reflect the complexity of CJ data
+\citep{Thurstone_1927b}. Specifically, while assuming assuming equal
+discriminal dispersions and zero correlation for discriminal processes
+simplifies the trait measurement model, these assumptions may fail to
+reflect the complexity of CJ data
 \citep{Thurstone_1927a, Andrich_1978, Kelly_et_al_2022}. As a result,
@@
 CJ tradition has attempted to address this gap by separating trait
-estimation from the hypothesis testing process. Specifically, CJ studies
-often rely on point estimates of the traits (i.e., the BTL scores)--or
-their transformations--to conduct statistical inference. Although
-separating these processes simplifies the analysis of CJ data, the
-statistical literature cautions against this practice, as it can
-introduce bias into the analysis and compromise the reliability of the
-hypothesis test
+estimation from hypothesis testing. Specifically, CJ studies often rely
+on point estimates of the traits--typically the BTL scores or their
+transformations--to conduct statistical inference. Although this
+separation simplifies CJ data analysis, the statistical literature warns
+that this practice can introduce bias into the analysis and compromise
+the reliability of the hypothesis tests
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
@@
 determining which observations are part of the data sample. This new
-perspective invites the application of Little and Rubin's principled
-missing data framework \citep{Little_et_al_2020}, enabling a more
-rigorous evaluation of existing claims about these mechanisms, their
-influence on CJ outcomes, and their implications for designing and
-evaluating more complex experimental setups.
+perspective invites the application of Little and Rubin's
+\citeyearpar{Little_et_al_2020} principled missing data framework,
+enabling a more rigorous evaluation of existing claims about these
+mechanisms, their influence on CJ outcomes, and their implications for
+designing and evaluating more complex experimental setups.
 
@@
 observed trait distribution \citep{Deffner_et_al_2022}. It is therefore
-essential to determine whether such sampling biases occur--and if so,
-whether they undermine the validity of trait estimates and the
-robustness of statistical inferences. Our framework offers a rigorous
-setting to test the scenario.
+essential to determine whether such sampling biases undermine the
+validity of trait estimates and the robustness of statistical
+inferences. Our framework offers a rigorous setting to test this
+scenario.
 
@@
 generates simple random samples that satisfy the condition of
-ignorability, enabling researchers to legitimately ignore missing data
-during analysis without introducing bias
+ignorability, enabling researchers to legitimately \emph{ignore} missing
+data during analysis without introducing bias
 \citep{Everitt_et_al_2010, Kohler_et_al_2019}.
@@
 so would enable researchers to assess and test their impact on trait
-estimation and statistical inference. Our framework provides a rigorous
-foundation for such articulations.
+estimation and statistical inference. Our framework, again, provides a
+rigorous foundation for such articulations.
 
@@
 \citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
-further research is needed to address several key questions. For
-instance: Can CJ data be analyzed under the assumption of
-``sample-free'' trait calibration--that is, the estimated trait
-difference for any two scripts does not depend on which other scripts
-they are compared with, implying an absence of judges' bias?
-\citep{Bramley_2008, Andrich_1978} If this assumption does not hold, how
-many judges and judgments are required to ensure acceptable levels of
-trait reliability and validity? Do training and expertise influence
-judgment biases, and to what extent? \citep{Kelly_et_al_2022} Finally,
-what are the benefits of \emph{repeated measures designs}, in which
-judges evaluate the same stimulus pairs multiple times?
+further research is needed to address several critical questions. For
+instance, can CJ data be analyzed under the assumption of
+``sample-free'' trait calibration, which implies an absence of judges'
+bias? \citep{Bramley_2008, Andrich_1978} If this assumption does not
+hold, how many judges and judgments per judge are required to ensure
+acceptable levels of trait reliability and validity? Do training and
+expertise influence judgment biases, and to what extent?
+\citep{Kelly_et_al_2022} Finally, what are the benefits of
+\emph{repeated measures designs}, in which judges evaluate the same
+stimulus pairs multiple times, in assessing judgment accuracy?
 \citep{Lawson_2015}

### 2025-05-29 paper2.qmd
section:6; process:write

@@
 This preference largely stems from the widespread adoption of the BTL model, which provides a simplified statistical representation of the case.
-The BTL model mirrors most of Case V's assumptions, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases).
+The BTL model mirrors most of Case V's assumptions, with one notable distinction.
+While Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases).
 However, this substitution has minimal impact on trait estimation or model interpretation because the scale of the discriminal process (i.e., the latent trait) is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021].
@@
 These assumptions are the equal dispersion and zero correlation between stimuli.
-
-<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs.
-[@Thurstone_1927a, pp.
-376] -->
 
@@
 <!-- 3. But does it hold in practice? -->
-Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data.
-In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Bramley_2008; @Kelly_et_al_2022], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics.
-For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018].
-These misfit texts may exhibit larger discriminal dispersions or represent outliers--texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, the BTL model neither accounts for these cases nor offers any means to address them, except their elimination from the analysis.
+Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of some traits or account for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
+Indeed, the presence of the so-called *misfit* texts may already signal these limitations, as these are texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018], possibly due to larger discriminal dispersions or because they are genuine outliers--meaning, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+In either case, the BTL model neither accounts for these anomalies nor provides tools for addressing them, apart from excluding the "problematic" texts from analysis.
 
@@
 
-<!-- 4. How these judgment discrepancies happen?  -->
-<!-- These misfit texts likely arise from two underlying mechanisms.
-First, a text may exhibit a larger discriminal dispersion, increasing the overlap between its discriminal process and those of multiple other texts.
-This greater overlap leads to more judgment discrepancies, increasing the likelihood of classifying the text as a misfit.
-@fig-discriminal_process illustrates this scenario.
-In the figure, Text B has a wider discriminal dispersion than Text A, enlarging the region where their discriminal processes intersect.
-This larger overlap represents an area where judges face greater difficulty in determining which text exhibits a higher level of the trait.
-Second, a text may represent an outlier --one with distinct characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-Such texts are harder to compare and place on the trait scale, resulting in more judgment discrepancies and a misfit classification.
-In either case, the BTL model is inherently agnostic to these mechanisms and does not provide any means to address them.
--->
-
 <!-- 5. So, what happens when the assumption does not hold? -->
@@
 Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli.
-This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020] and, by extension, about the underlying discriminal processes of stimuli.
+This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020; @Wu_et_al_2022] and, by extension, about the underlying discriminal processes of stimuli.
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
@@
 <!-- 2. The first scenario where the assumption does not hold -->
-In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These additional relevant characteristics, when assessed, are unlikely to be equally weighted or opposing.
-As a result, they may exert an uneven influence on judges' perceptions, creating biases that resist cancellation.
-For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments.
-Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
-While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003, @vanDaal_et_al_2016 and @Bartholomew_et_al_2020a demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.
+In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
+These characteristics, regardless of their relevance, are unlikely to be equally weighted or to oppose one another across comparisons.
+As a result, they may exert a disproportionate influence on judges' perceptions, generating biases that persist rather than cancel out.
+For example, this could occur when a judge assessing the argumentative quality of a text may place disproportionate emphasis on handwriting clarity, thereby favoring neatly written texts despite their weaker arguments.
+Moreover, because the discriminal process of stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
+Evidence from traditional marking contexts confirms a similar phenomenon [@Morin_et_al_2018], and additional studies [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a] support the presence of these biases in CJ--reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.
 
@@
 Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues.
-Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023].
+Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023] and even biases [@Wu_et_al_2022].
 These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli.
@@
 
-Finally, as discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969]--risk discarding valuable information and introducing bias into trait estimates.
+Finally, as discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969; @Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates.
 The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
@@
 
-<!-- Conventionally, any figure greater than the mean plus two standard deviations – 1.14 here – is taken to indicate signiﬁcant misfit.
+<!-- Conventionally, any figure greater than the mean plus two standard deviations – 1.14 here – is taken to indicate significant misfit.
 In this case, only Judge ‘46 N-M 2’ exceeded the criterion.
-To a signiﬁcant degree, this judge interpreted the trait ‘good writing’ differently from the others.
+To a significant degree, this judge interpreted the trait ‘good writing’ differently from the others.
 [@Pollitt_2012b, pp.
@@
 
-<!-- Pollitt [9] states that a large inﬁt for raters suggests that they consistently judge away from the group consensus.
+<!-- Pollitt [9] states that a large infit for raters suggests that they consistently judge away from the group consensus.
 Those raters are called misfits.
@@
 Building on the previous section, it is clear that, researchers in CJ studies typically use the BTL model to measure traits and position the compared stimuli along a latent continuum [@Thurstone_1927b].
-The CJ literature also show that they frequently rely on point estimates of these traits (i.e, the BTL scores)--or their transformations--to conduct statistical inference or hypothesis testing.
+The CJ literature also show that they frequently rely on point estimates of these traits--typically the BTL scores or its transformations--to conduct statistical inference or hypothesis testing.
 For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 
-<!-- general development -->
 <!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
@@
 These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population*.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
 Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
@@
 
-We can visualize these vectors more clearly using the example in Equation ([-@eq-mat13]).
-Suppose researchers exclude the second student and the third judge, and they also omit the second text from each individual and the third repeated judgment from each judge.
-Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, these vectors take the following structure:
+We can illustrate the structure of these vectors more clearly with an example.
+Suppose researchers exclude the second student, the second text from each student, and the third judge from the setup shown in Equation ([-@eq-mat13]).
+Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the resulting vectors would have the following structure:
 $$
@@
 1 \\
-0 \\
 1 \\
 1 \\
-0 \\
+1 \\
+1 \\
 0 \\
@@
 
-Equation ([-@eq-mat24]) shows that missing observations in the vectors $S_{I}$ and $S_{J}$--which represent unsampled students and judges--directly determine which observations are missing in $S_{IA}$ and $S_{JK}$.
+Notably, Equation ([-@eq-mat24]) shows that missing observations in the vectors $S_{I}$ and $S_{J}$--which represent unsampled students and judges--directly determine which observations are missing in $S_{IA}$ and $S_{JK}$.
 In other words, researchers can only observe texts and judgments from students and judges initially included in the sample.
@@
 The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
-A square encloses this vector, indicating that it is a conditioned variable.
-Here, *conditioning* means that researchers restrict their focus to the elements of $O^{cp}_{R}$ that satisfy $s_{(i,a,h,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
-Essentially, $S$ serves as the vector selecting *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*.
+A square encloses $S$, indicating that it is a conditioned variable.
+In this context, *conditioning* means that researchers restrict their focus to the elements of $O^{cp}_{R}$ that satisfy $s_{(i,a,h,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
+In essence, $S$ is a vector that selects *all repeated judgments made by a subset of judges for a subset of stimuli produced by the sampled individuals*.
 
@@
 <!-- 3. segway for next section -->
-However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
-Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by certain individuals.
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
 
@@
 <!-- 2. What is the effect of the comparison mechanism -->
-The [DAG -@fig-cj14_dag] also integrates the *comparison mechanism* $C$ into the conceptual-population model.
-It shows the sample-comparison outcome as unobserved, highlighting that researchers cannot directly access to $O^{sc}_{R}$ due to the comparison mechanism.
-Moreover, the DAG depicts $C$ as a causal factor influencing $O_{R}$, enclosing it in a square to indicate that it is a conditioned variable.
-Specifically, $C$ determines *which repeated judgments the subset of judges makes for the subset of stimuli produced by the sampled individuals*.
-In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments for the selected stimuli.
-Instead, they carry out a "sufficient number of comparisons" $n_{C}$ to allow the accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp.
+The [DAG -@fig-cj14_dag] also incorporates the *comparison mechanism* $C$ into the conceptual-population model.
+It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, emphasizing that researchers cannot directly access this variable because of the comparison mechanism.
+The DAG further shows $C$ as a conditioned variable (enclosed in a square) that causally influences the observed outcome $O_{R}$.
+This structure implies that $C$ determines *which repeated judgments judges make for the stimuli produced by the individuals*.
+In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments but instead complete a sufficient number, $n_{C}$, to enable the accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp.
 267].
@@
 <!-- recording: Sven and Tine 25.04.10; time: 00:52:00 - 00:53:40 -->
+<!-- recording: Sven and Tine 25.05.21; time: 00:42:20 - 00:46:10 -->
 
 <!-- 1. How to derive an stat model from an SCM? -->
-With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption (see @sec-theory-issue1a).
+Using the structural causal model (SCM) [-@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption (see @sec-theory-issue1a).
 This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
@@
 <!-- 2. assumptions about the outcome -->
-Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in @fig-cj16.
+Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the statistical model [-@fig-cj16_stat].
 We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
@@
   \\
-  \boldsymbol{\Sigma} &= \boldsymbol{D} \boldsymbol{R} \boldsymbol{D} \\ \\
+  \boldsymbol{\Sigma} &= \boldsymbol{V} \boldsymbol{Q} \boldsymbol{V} \\ \\
 \end{aligned}
@@
 <!-- 3. Note about the sample and comparison mechanism, as well as the predictor's distribution -->
-Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16.
-We make this omission because, although these distributions contribute to the overall joint distribution of the data, the variables  $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$ are observed and independent of any other variable in the model.
-As observed variables, they do not require distributional assumptions--unlike the idiosyncratic errors.
-Moreover, their independence stems from the random selection designs that govern the variables[^8].
+Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit all of these probabilities from the statistical model [-@fig-cj16_stat].
+This omission is justified because, while these distributions contribute to the overall joint distribution of the data, the variables $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$ are observed and independent of any other variable in the model.
+As observed variables, they do not require distributional assumptions in the same way the idiosyncratic errors do.
+Furthermore, their independence follows from the underlying random selection procedures that govern the variables[^8].
 
@@
 Thus, to identify the latent variable model, we must resolve these indeterminacies [@Depaoli_2021; @deAyala_2009].
-Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix.
+Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{V} \boldsymbol{Q} \boldsymbol{V}$, with $\boldsymbol{V}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{Q}$ a correlation matrix.
 To address the *location* indeterminacy, we set the errors' mean vector to zero:
@@
 Following [SCM -@fig-cj16_scm], we resolve the *orientation* indeterminacy by assuming that the errors are uncorrelated.
-This assumption leads us to specify a correlation matrix for the errors, $\boldsymbol{R}$, as:
-$$
-\boldsymbol{R} = \begin{bmatrix}
+This assumption leads us to define the error correlation matrix, $\boldsymbol{Q}$, as the identity matrix:
+$$
+\boldsymbol{Q} = \begin{bmatrix}
     1 & 0 & 0 & 0 \\
@@
 
-To resolve the *scale* indeterminacy, we define the diagonal matrix $\boldsymbol{D}$ as:
-$$
-\boldsymbol{D} = \begin{bmatrix}
+To resolve the *scale* indeterminacy, we define the diagonal matrix $\boldsymbol{V}$ as:
+$$
+\boldsymbol{V} = \begin{bmatrix}
     s_{XI} & 0 & 0 & 0 \\
@@
 
-Note that because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$.
+Because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$.
 We set $P_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait.
-As a result, $T_{IA}$ is also normally distributed.
+Note that as a result, $T_{IA}$ is also normally distributed.
 This configuration effectively reinstates Thurstone’s original assumption of Normal discriminal processes for the stimuli (see @tbl-thurstone_cases).
@@
 <!-- 9. Computational implementation -->
-Finally, we use *Bayesian inference methods* to convert [model -@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data.
-Bayesian inference offers three key advantages for this purpose.
-First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999]--such as the model we try to implement.
+Finally, we use *Bayesian inference methods* to convert the statistical model [-@fig-cj16_stat] into a practical statistical tool for analyzing paired comparison data.
+Bayesian inference offers three main advantages in this context.
+First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999].
+This capability is essential for our implementation, as the model is indeed overparameterized.
 Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
-In our implementation, we use priors to define the error distribution, which sets the scale of the latent variables [@Depaoli_2014].
+In our implementation, we use mainly use priors to define the error distribution, setting the scale of the latent variables [@Depaoli_2014].
 Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
-This feature is particularly relevant in CJ experiments, where researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.
+This feature is particularly relevant in CJ experiments, as researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.
 
@@
 <!-- recording: Sven and Tine 25.04.22; time: 00:00:00 - 00:13:15 -->
+<!-- recording: Sven and Tine 25.05.21; time: 00:00:00 - 00:03:30 -->
+<!-- recording: Sven and Tine 25.05.21; time: 00:32:55 - 00:33:50 -->
 
@@
 
-<!-- 2. Summary of the why the literature uses Case V -->
+<!-- 2. Summary of the why the literature uses Case V, and its problems -->
 Of these five cases, the CJ literature has predominantly relied on Case V to measure various psychological traits, as shown in studies like @Kimbell_2012, @Jones_et_al_2015, and @Boonen_et_al_2020.
 This preference largely stems from the widespread adoption of the BTL model, which offers a simplified statistical formulation of the case.
-The BTL model mirrors the assumptions of Case V--namely, equal discriminal dispersions and zero correlation for the stimuli' discriminal processes--with one notable distinction: while Case V assumes normally distributed discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008].
-
-<!-- 3. The issues with Case V, and the BTL model -->
-Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b].
-Specifically, while assuming assuming equal discriminal dispersions and zero correlation for discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Thurstone_1927a; @Andrich_1978; @Kelly_et_al_2022].
+The BTL model mirrors the assumptions of Case V--namely, equal discriminal dispersions and zero correlation for the stimuli' discriminal processes--with one notable distinction.
+While Case V assumes normally distributed discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008].
+Although this substitution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that assuming equal discriminal dispersions and zero correlation among stimuli' discriminal processes could introduce significant problems.
+In particular, he recognized that while these assumptions simplify the trait measurement model, they may not capture the complexity of some traits or account for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
 As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
 
-Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp.
+Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and allocate the compared stimuli along this continuum [@Thurstone_1927a, pp.
 269], his theory did not specify how to use the resulting trait estimates for statistical inference.
 The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing.
-Specifically, CJ studies often rely on point estimates of the traits--typically the BTL scores or their transformations--to conduct statistical inference.
-Although this separation simplifies CJ data analysis, the statistical literature warns that this practice can introduce bias into the analysis and compromise the reliability of the hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+Specifically, CJ studies often rely on point estimates of the traits--typically the BTL scores or its transformations--to conduct statistical inference.
+Although this separation simplifies CJ data analysis, the statistical literature warns that this practice can introduce bias into the analysis and compromise the reliability of statistical inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
@@
 Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
-The approach begins by formulating a conceptual model represented by an SCM and a DAG.
-This model integrates Thurstone's core theoretical principles--such as the discriminal processes (i.e., traits) of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms.
+The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG).
+This model integrates Thurstone's core theoretical principles--such as the discriminal processes of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.
 
@@
 This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2).
-In particular, the model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli (refer to @sec-theory-theoretical_P3), incorporates measurement error in the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical).
+In particular, the model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli and incorporates measurement error in the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical).
 These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.
@@
 Beyond these potential benefits, the approach offers two additional advantages.
-First, it clarifies the interactions among all actors and processes involved in CJ experiments.
+First, it clarifies the roles and interactions of all actors and processes involved in CJ experiments.
 Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data.
-Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.
+Together, these advantages establish a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.
 
@@
 
-<!-- 6. Which are these new avenues for research? -->
-Five research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
-First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect inference robustness.
-Second, studies should assess the impact of judges' bias on the number of comparisons needed to achieve reliable trait estimates.
-Third, future work should re-evaluate current methods for identifying 'misfitting' judges and stimuli.
-Fourth, researchers should analyze how the hierarchical structure of stimuli affects both experimental design and hypothesis testing.
-Finally, studies should explore how modeling non-dichotomous outcomes or incorporating multidimensional traits affects the reliability and validity of the trait estimates.
-
-<!-- 6.1 About the implications of the sampling and comparison mechanisms -->
-The explicit integration of the sampling and comparison mechanisms within our conceptual model offers a novel perspective on the first research avenue.
-Specifically, it shows how these mechanisms influence the data-generating process by acting as missing data sources, determining which observations are part of the data sample.
-This new perspective invites the application of Little and Rubin's [-@Little_et_al_2020] principled missing data framework, enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex experimental setups.
-
-<!-- --unlike prior literature, which merely acknowledged their presence in the experimental design. -->
-
-One prominent claim in the CJ literature is that Thurstone's model naturally handles non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].
-However, it is easy to imagine scenarios where this assumption may not hold.
-For instance, selecting a particular group of judges could result in an incomplete representation of the trait's full range and complexity, as judgment accuracy varies with demographic factors [@Bartholomew_et_al_2020a; @Kelly_et_al_2022] and even between individuals [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020].
-Consequently, judges' selection could affect the formation of a "shared consensus", potentially biasing trait estimates and distorting the observed trait distribution [@Deffner_et_al_2022].
-It is therefore essential to determine whether such sampling biases undermine the validity of trait estimates and the robustness of statistical inferences.
-Our framework offers a rigorous setting to test this scenario.
-
-[another example: For instance, consider a CJ experiment assessing speech quality using a case-control design [@Everitt_et_al_2010] that samples individuals with normal hearing (NH) and those with hearing impairment (HI).
+Five research avenues--current and new--deserve particular attention.
+The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
+Notably, while addressing these questions through formal derivations--following the tradition of classical statistical models like linear regression--may seem a natural step, we argue that the complexity of the CJ system makes this approach initially impractical.
+Instead, simulation-based methods, such as power analysis, offer a more feasible and flexible alternative, enabling researchers to investigate these questions without relying on cumbersome mathematical proofs.
+Nonetheless, developing formal mathematical proofs remains an essential goal for future work.
+
+<!-- due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. -->
+
+## The impact of sampling and comparison mechanisms on CJ outcomes and inference robustness {#sec-discussion_RA1}
+<!-- recording: Sven and Tine 25.05.21; time: 00:03:30 - 00:24:00 -->
+
+Sampling and comparison mechanisms play a central role in modern CJ experimental design.
+However, most of the CJ literature has examined these mechanisms within a limited scope.
+Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022].
+Although these studies offer valuable insights, we argue that they may have overlooked the broader role these mechanisms play within the CJ system.
+Moreover, we suggest that this oversight largely stems from a lack of conceptual clarity about how the mechanisms function within the system.
+To address this gap, this study decided to integrate these mechanisms into CJ's conceptual model.
+
+The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process.
+In particular, it clarifies their role in the data-generating process by framing them as sources of missing data--that is,  as mechanisms that determine which observations are missing from the final data sample.
+This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and assessing more complex experimental setups.
+
+Notably, this study circumvents the need to apply this missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome.
+In other words, we designed these mechanisms so they produce data that are *missing completely at random* (MCAR) [@Little_et_al_2020].
+This design offers one key advantage: it generates simple random samples that satisfy the condition of *ignorability*, allowing researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019; @Neal_2020].
+
+<!-- Well, what assumption(s) would make it so that the ATE is simply the associational difference?
+This is equivalent to saying “what makes it valid to calculate the ATE by taking the average of the Y(0) column, ignoring the question marks, and subtracting that from the average of the Y(1) column, ignoring the question marks?” This ignoring of the question marks (missing data) is known as ignorability.
+Assuming ignorability is like ignoring how people ended up selecting the treatment they selected and just assuming they were randomly assigned their treatment; we depict this graphically in Figure 2.2 by the lack of a causal arrow from X to T.
+[@Neal_2020, pp.
+9] -->
+
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
+
+However, many modern CJ applications rely on more complex experimental designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
+A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
+This outcome-informed selection suggests that the comparison mechanism may be outcome-dependent, potentially classifying ACJ as a generator of MNAR data.
+This classification may, in turn, facilitate a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities.
+While some studies report improvements in trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b],  others argue that these gains are artificially inflated by the method [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
+
+<!-- The A for ‘adaptivity’ in ACJ means that the choice of which objects are presented to the judges depends on the outcomes of judgments made so far – the idea being to use judge time efficiently by not presenting objects to compare where the result of the comparison is almost certain.
+The analogy is with computerised adaptive testing (CAT) where the choice of the next item to present is based on the examinee’s success or failure on items presented so far.
+[@Bramley_2015, pp.
+4].
+-->
+
+<!-- In contrast, case-control studies [@Everitt_et_al_2010] use sampling mechanisms tied to non-assignable variables that likely influence--or are influenced by--the trait of interest.
+For example, @Boonen_et_al_2020 formed comparison groups based on children’s hearing status, a non-assignable variable likely affecting speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024].
+Notably, although the authors carefully constructed comparable groups, they noted that some subgroups--such as children with hearing aids--were not representative of their broader population.
+This is a common challenge in such designs, care must be taken to ensure that the excluded individuals are not systematically different in trait performance.
+Otherwise, the resulting trait estimates or the statistical inferences may be biased.
+-->
+
+<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
+However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
+
+<!-- [another example: For instance, consider a CJ experiment assessing speech quality using a case-control design [@Everitt_et_al_2010] that samples individuals with normal hearing (NH) and those with hearing impairment (HI).
 Here, the sampling mechanism depends on hearing status--a variable likely to influence the trait of interest, as HI individuals may exhibit lower speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024].
@@
 If this group is systematically excluded and happens to over- or under-perform in speech quality, the resulting trait estimates may be biased.
-It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"}
-
-[another example: For instance, consider a high-stakes non-mandatory summative CJ assessment, where the results the individuals have in this assessment have some implications for them, however, their participation is not mandatory.
+It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"} -->
+
+<!-- [another example: For instance, consider a high-stakes non-mandatory summative CJ assessment, where the results the individuals have in this assessment have some implications for them, however, their participation is not mandatory.
 In this case, it not hard to imagine that some of these individuals will not participate in the CJ assessment, thus affecting the sample, and ultimately the trait estimates: now a probable under performing group will not available in the data sample, thus biasing the resulting trait estimates.
-It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"}
+It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"} -->
+
+<!-- Comment for myself: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
+{S, C} -> {Y}
+e_{A} -> {S, C, Y} 
+this means that e_{A} is a confounder?,  
+then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C, S} by taking a random sample of stimuli?.
+Does this mean that random algorithms per se will always be unbiased?.
+-->
+
+Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions--an approach supported by our framework.
+This clarity enables researchers to evaluate, for instance, how sampling and comparison mechanisms affect trait estimation and statistical inference within each design.
+Such assessments are particularly relevant given the common misconception in the CJ literature that Thurstone's model can naturally handle even non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].
 
@@
 
-<!-- The logic is that in aggregating the varied judgments, one comes to an overall view of the community of practice.
-However, this raises important questions about diversity of the judging pool.
-If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
-Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
-For example, Bartholomew et al.
-(2020) found differences between what judges from the UK, Sweden, and the USA valued when assessing design portfolios, providing evidence to support divergence along demographic lines.
-[@Kelly_et_al_2022, pp.
-683] -->
-
-<!-- Samples from different sites often differ in terms of their demographic profiles (here, their age and gender distribution), and these demographic variables might in turn affect the distribution of the trait of interest.
-How should researchers deal with these differences?
-The answer depends on the processes that generated the observed disparities.
-Demographic disparities among samples may result from (a) differences in the actual populations from which the samples are taken or (b) sampling procedures that differ among sites.
-[@Deffner_et_al_2022, pp.
-6] -->
-
-<!-- These examples demonstrate that the generalizability of experimental effects does not depend on the presence of population differences per se but on the exact mechanisms by which populations differ.
-While some differences—especially those concerning the independent variable—are inconsequential for intended generalizations, differences concerning effect modifiers or mediators require statistical adjustment.
-[@Deffner_et_al_2022, pp.
-11] -->
-
-Moreover, when designing and evaluating more complex experimental setups, our framework highlights one key principle: the sampling and comparison mechanisms directly influence the CJ outcome.
-In this study, we deliberately structured these mechanisms to produce data that is *missing completely at random* (MCAR)--that is, the missingness is unrelated to any observed or unobserved variables, including the outcome [@Little_et_al_2020].
-This design offers one key advantage: it generates simple random samples that satisfy the condition of ignorability, enabling researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019].
-
-However, many modern CJ applications involve more complex forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) mechanism [@Little_et_al_2020]--that arise from the chosen sampling and comparison methods.
-Two notable examples are adaptive comparative judgment designs [@Pollitt_2012a; @Pollitt_2012b] and case-control studies [@Everitt_et_al_2010].
-In adaptive designs, an algorithm defines the comparison mechanism by dynamically selecting which stimulus pairs judges evaluate.
-These designs have produced mixed evidence: some studies report improved trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others suggest these gains may be artificially inflated [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
-In contrast, case-control studies rely on sampling mechanisms linked to non-assignable variables that likely influence the trait of interest.
-For instance, @Boonen_et_al_2020 selected children based on hearing status, a non-assignable variable that likely affects speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024].
-Nevertheless, regardless of their missingness mechanism, both designs could benefit from explicitly articulating their underlying causal structures.
-Doing so would enable researchers to assess and test their impact on trait estimation and statistical inference.
-Our framework, again, provides a rigorous foundation for such articulations.
-
-<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
-
-<!-- (Done) It is more likely that we can decide the characteristics of the objects we want to include in the study (independent variables), ensuring groups that are more balanced.
-However, in the case of subjects sometimes the independent variables cannot be assigned (e.g., hearing status) -->
-
-<!-- Comment for myself: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
-{S, C} -> {Y}
-e_{A} -> {S, C, Y} 
-this means that e_{A} is a confounder?,  
-then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C, S} by taking a random sample of stimuli?.
-Does this mean that random algorithms per se will always be unbiased?.
--->
-
-<!-- 6.2 About judges biases, judgment discrepancies, and the number of judges and judgments -->
-Additionally, by explicitly incorporating judges' biases, our framework provides a novel approach to investigate the second research avenue.
-Specifically, it enables researchers to determine whether such biases occur and, if so, to assess their impact on the reliability and validity of trait estimation.
-Although some anecdotal evidence already suggests judges' biases exist [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], further research is needed to address several critical questions.
-For instance, can CJ data be analyzed under the assumption of "sample-free" trait calibration, which implies an absence of judges' bias?
-[@Bramley_2008; @Andrich_1978] If this assumption does not hold, how many judges and judgments per judge are required to ensure acceptable levels of trait reliability and validity?
-Do training and expertise influence judgment biases, and to what extent?
-[@Kelly_et_al_2022] Finally, what are the benefits of *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times, in assessing judgment accuracy?
-[@Lawson_2015]
-
-<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
-
-<!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need?
-Who should judge?
-How many judgments should they make?
-And crucially, how can we be confident that they are basing their judgments on acceptable criteria?
-If judges are not using construct-relevant criteria, then the argument that comparative judgments are necessarily valid cannot be upheld.
-We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used.
-However, work to derive some general principles for good practice would be valuable.
-[@Kelly_et_al_2022, pp.
-683] -->
-
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
-267] -->
-
-<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
-[@Kelly_et_al_2022, pp.
-681] -->
-
-<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology.
-We do not yet know how widely it can be, or should be, applied.
-[@Pollitt_2012b, pp.
-297]-->
-
-<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
-(2002).
-Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
-[@Bramley_2008, pp.
-274] -->
-<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges. -->
+## The effects of judges' bias on the reliability of traits {#sec-discussion_RA2}
+<!-- recording: Sven and Tine 25.05.21; time: 00:24:00 - 00:31:45 -->
+
+Despite the growing notion that various stimulus-related factors can influence judges' perceptions during pairwise comparisons--and that these influences may not always cancel each other out--few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
+We argue that this gap persists not due to a lack of interest or research but because researchers continue to rely on ad-hoc detection methods, such as 'misfit' statistics, that are poorly suited for the task [@Kelly_et_al_2022].
+To address this limitation, this study decided to treat judges' biases as an integral component of the CJ system from the outset.
+This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons--one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.
+
+The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions.
+For instance, can researchers validly analyze CJ data under the assumption of "sample-free" trait calibration--specifically, under the assumption that judges exhibit no systematic bias?
+This question is especially relevant because researchers often treat "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases.
+Furthermore, if judges' biases reflect additional stimulus characteristics that should not influence assessment, to what extent can training or expertise mitigate these biases?
+[@Kelly_et_al_2022] More specifically, as discussed in @sec-theory-issue1b, can training or expertise help prevent judges from prioritizing irrelevant features--such as handwriting--over more central criteria like argumentative quality in writing assessments?
+Addressing these questions may also shed light on what it truly means to be an "expert" in the CJ context [@Kelly_et_al_2022].
 
@@
 
-<!-- 6.3 About removing misfit judges and stimuli  -->
-
-[About removing misfit judges and stimuli]{style="color:blue;"}
-
-<!-- "This explicit part of the model it open ups the idea of doing more throughout research on ..." -->
-
-<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
-If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
-...
-It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
+<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al.
+(2002).
+Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
+[@Bramley_2008, pp.
+274] -->
+<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges. -->
+
+<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
+[@Kelly_et_al_2022, pp.
 681] -->
+
+<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology.
+We do not yet know how widely it can be, or should be, applied.
+[@Pollitt_2012b, pp.
+297]-->
+
+Moreover, if we accept that judges may value different aspects of a stimulus, it follows that their judgments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020].
+Building on this and the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
+Furthermore, if such characteristics indeed undermine the "sample-freeness" assumption, then it becomes essential to explore strategies for controlling their influence.
+This includes investigating how many judges--and how many judgments per judge--are required to produce reliable trait estimates, as well as whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
+
+<!-- The logic is that in aggregating the varied judgments, one comes to an overall view of the community of practice.
+However, this raises important questions about diversity of the judging pool.
+If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on.
+Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?
+For example, Bartholomew et al.
+(2020) found differences between what judges from the UK, Sweden, and the USA valued when assessing design portfolios, providing evidence to support divergence along demographic lines.
+[@Kelly_et_al_2022, pp.
+683] -->
 
@@
 
-<!-- 4. How these judgment discrepancies happen? -->
-<!-- These misfit texts likely arise from two underlying mechanisms.
+<!-- Samples from different sites often differ in terms of their demographic profiles (here, their age and gender distribution), and these demographic variables might in turn affect the distribution of the trait of interest.
+How should researchers deal with these differences?
+The answer depends on the processes that generated the observed disparities.
+Demographic disparities among samples may result from (a) differences in the actual populations from which the samples are taken or (b) sampling procedures that differ among sites.
+[@Deffner_et_al_2022, pp.
+6] -->
+
+<!-- These examples demonstrate that the generalizability of experimental effects does not depend on the presence of population differences per se but on the exact mechanisms by which populations differ.
+While some differences—especially those concerning the independent variable—are inconsequential for intended generalizations, differences concerning effect modifiers or mediators require statistical adjustment.
+[@Deffner_et_al_2022, pp.
+11] -->
+
+<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
+
+<!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need?
+Who should judge?
+How many judgments should they make?
+And crucially, how can we be confident that they are basing their judgments on acceptable criteria?
+If judges are not using construct-relevant criteria, then the argument that comparative judgments are necessarily valid cannot be upheld.
+We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used.
+However, work to derive some general principles for good practice would be valuable.
+[@Kelly_et_al_2022, pp.
+683] -->
+
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+267] -->
+
+## The identification of 'misfitting' judges and stimuli {#sec-discussion_RA3}
+<!-- recording: Sven and Tine 25.05.21; time: 00:31:45 - 00:32:20 -->
+
+[About the identification of 'misfitting' judges and stimuli]{style="color:blue;"}
+
+<!-- Indeed, the presence of the so-called *misfit* texts may already signal these limitations, as these are texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018], possibly due to larger discriminal dispersions or because they are genuine outliers--meaning, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+In either case, the BTL model neither accounts for these anomalies nor provides tools for addressing them, apart from excluding the "problematic" texts from analysis.
+-->
+
+<!-- removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969; @Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates.
+The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
+-->
+
+<!-- Second, and complementary to the previous point, the statistical literature consistently highlights the risks of using ad hoc procedures to identify "outliers" when the model does not accurately represent the data-generating process.
+For instance, McElreath (2020, pp.
+232) expressed that: "There is a tradition of dropping outliers.
+People sometimes drop outliers even before a model is fit, based only on standard deviations from the mean outcome value.
+You should never do that--a point can only be unexpected and highly influential in light of a model.
+After you fit a model, the picture changes.
+If there are only a few outliers, and you are sure to report results both with and without them, dropping outliers might be okay.
+But if there are several outliers, we really need to model them".
+Similarly, O'Hagan (2018, pp.
+366) makes a small example explaining how the average of a parameter--much like a trait score you could get from the BTL model--depends on outliers.
+Miller (2023), in psychology, explores a similar topic.
+-->
+
+<!-- You might notice I use the term "outlier" here to describe biased judges.
+This is because, like statistical outliers, these judges deviate significantly from the rest of the sample in which they occur (Grubbs, 1969).
+However, the key reason for using this terminology lies in that any ad hoc method for flagging biased judges based on the BTL model--such as misfit statistics--effectively acts as an outlier detection tool.
+This is due to the fact that the BTL model relies on several strong assumptions, which limit its capacity to capture other features of the data unless supplemented by additional procedures--procedures that, in essence, resemble outlier detection.
+-->
+
+<!-- Moreover, from McElreath's text, you might recognize that the practice of removing outliers mirrors the act of discarding biased judges in CJ using misfit statistics, where these statistics are essentially transformations of residual probabilities linked to trait estimates, or as McElreath puts it "standard deviations from the mean outcome values." But what are the potential consequences of wrongfully identifying and discarding misfit judges, you might ask?.
+Well, this practice risks removing valuable information and introducing bias into trait estimates [Zimmerman, 1994; O'Hagan, 2018; McElreath, 2020].
+The direction and magnitude of these biases are unpredictable, depending on which judges are excluded from the analysis.
+-->
+
+<!-- In the previous subsection we argue that ad-hoc detection methods, such as 'misfit' statistics, could be poorly suited for the task of detecting judges bias [@Kelly_et_al_2022].
+This hinges in the assumption that the CJ data does have judges biases, but the method does not manages to identify them, leading to false negatives?
+or lack of sensitivity in the tests.
+However, there is a related problem which is: these ad-hoc method could be detecting judges' biases that are not there.
+as in @Wu_et_al_2022  -->
+
+<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
+If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
+...
+It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
+681] -->
+
+<!-- The same could be said for the stimuli.
+These misfit texts likely arise from two underlying mechanisms.
 First, a text may exhibit a larger discriminal dispersion, increasing the overlap between its discriminal process and those of multiple other texts.
@@
 
+<!-- Such differences may not be detected through analyses of bias and misfit.
+McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
+Where students prioritized factual recall, teachers prioritized scientific explanation.
+Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
+Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
+[@Kelly_et_al_2022, pp.
+683] -->
+
+<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
+
+<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
+
+<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
+
+<!-- Furthermore, these points tend to be outliers with unlikely values, according to the model.
+As a result, they are highly influential and make it difficult to estimate out-of-sample predictive accuracy.
+Why?
+Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
+[@McElreath_2020, pp.
+231] -->
+
+<!-- What can be done about this?
+There is a tradition of dropping outliers.
+People sometimes drop outliers even before a model is fit, based only on standard deviations from the mean outcome value.
+You should never do that—a point can only be unexpected and highly influential in light of a model.
+After you fit a model, the picture changes.
+If there are only a few outliers, and you are sure to report results both with and without them, dropping outliers might be okay.
+But if there are several outliers and we really need to model them, what then?
+[@McElreath_2020, pp.
+232] -->
+
 <!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
 
-<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
-
-<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
-
-<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
-
-<!-- 6.4 About the hierarchical structure of stimuli, number and predictors -->
+<!-- Consider, for example, the simplest location and scale problem-inference for a normal distribution with unknown mean \theta and variance \sigma^{2}.
+If we look at the marginal posterior distribution for \theta we know that if \sigma^{2} is known this posterior exhibits outlier-resistance, and we might expect similar behavior for unknown \sigma^{2}, but a simple example refutes this.
+Let the prior distribution give probability p (0<p<1) to (\theta = 1, \sigma^{2} = 1) and 1-p to (\theta = 0, \sigma^{2} = 4).
+Then the posterior mean of \theta is a decreasing function of any sufficiently large observation.
+[@OHagan_2018, pp.
+366] -->
+
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
+
+## The effects of hierarchies on experimental design and inference {#sec-discussion_RA4}
+<!-- recording: Sven and Tine 25.05.21; time: 00:32:25 - 00:33:00 -->
 
 [About the hierarchical structure of stimuli]{style="color:blue;"}
+
+<!-- "This explicit part of the model it open ups the idea of doing more throughout research on ..." -->
 
@@
 Now, by collecting data on specific stimulus features, researchers can assess whether these variables produce the expected effects.
--->
-
-<!-- clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
-Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
 -->
@@
 
+## The effects of non-dichotomous outcomes and multidimensional traits on CJ experiments {#sec-discussion_RA5}
+
 <!-- 6.5 About non-dichotomous outcomes and multidimensional traits -->
@@
 
-While a natural step would be to formally derive answers to all these research avenues--similar to approaches used in classical models like linear regression--we argue that the complexity of the CJ system makes this route impractical.
-Instead, simulation-based methods, like power analysis, offer a more feasible alternative, allowing researchers to investigate these issues without relying on cumbersome mathematical proofs.
-
-<!-- 7. Challenges -->
+## Challenges {#sec-discussion_challenges}
+<!-- recording: Sven and Tine 25.05.21; time: 00:33:50 - 00:42:20 -->
 
 <!-- 7.1 PPL and Bayesian Inference -->
-Despite its potential, this study presents several challenges that researchers must navigate.
+<!-- Despite its potential, this study presents several challenges that researchers must navigate.
 First, researchers need to be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods.
@@
 They reduce, for instance, the need for additional expertise in optimization techniques and theory often required by frequentist methods, thereby lowering the technical barrier for many applied researchers.
+-->
+
+<!-- make these four into one general challenge, "something you can always say about of almost every study".
+Now the challenge need a thorough knowledge about the CJ system.
+No easy route now, they have to think about their research context, assumptions in the experiment, model.
+-->
+
+<!-- our framework is not designed to work for everything, is just a starting point to think about which model applies to your research context  -->
+
+<!-- "although in this paper we have outlined the causal process behind the CJ system, this does not refrain other research to thoroughly think their causal structure in their studies" -->
 
 <!-- 7.2 Research context is important  -->
-Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts.
-We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
+<!-- We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
 That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.
+-->
+
+<!-- 7.4 Causal inference knowledge -->
+<!-- Fourth, applying this approach requires a solid understanding of Causal inference.
+For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity.
+Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.
+-->
+
+<!-- 7.5 Epistemiology  -->
+<!-- Fifth and final, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate.
+In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest?
+However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces.
+Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.
+-->
 
 <!-- 7.3 Faithfulness  -->
-Third,
 
@@
 150] -->
-
-<!-- 7.4 Causal inference knowledge -->
-Fourth, applying this approach requires a solid understanding of Causal inference.
-For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity.
-Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.
-
-<!-- 7.5 Epistemiology  -->
-Fifth and final, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate.
-In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest?
-However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces.
-Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.
 

### 2025-05-29 _tex/paper2.tex
Built site for gh-pages

--- 2025-05-15 _tex/paper2.tex
@@
 of the case. The BTL model mirrors most of Case V's assumptions, with
-one notable distinction: while Case V assumes a Normal distribution for
+one notable distinction. While Case V assumes a Normal distribution for
 the stimuli' discriminal processes, the BTL model uses the more
@@
 measurement model, evidence from the CJ literature suggests that this
-assumption may fail to capture the complexity of modern CJ data. In
-particular, the assumption may not hold when researchers assess complex
-traits or heterogeneous stimuli
-\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
-and stimuli can introduce judgment discrepancies due to their unique
-characteristics
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
-Indeed, the CJ literature may already provide indications of such
-discrepancies, particularly in the form of ``misfit'' statistics. For
-instance, \emph{misfit texts} are those whose comparisons result in more
-judgment discrepancies than those involving other texts
-\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}.
-These misfit texts may exhibit larger discriminal dispersions or
-represent outliers--texts with distinctive characteristics that deviate
-markedly from the rest of the sample \citep{Grubbs_1969}. In either
-case, the BTL model neither accounts for these cases nor offers any
-means to address them, except their elimination from the analysis.
+assumption may fail to capture the complexity of some traits or account
+for heterogeneous stimuli, such as handwritten texts or English
+compositions
+\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
+Indeed, the presence of the so-called \emph{misfit} texts may already
+signal these limitations, as these are texts that elicit more judgment
+discrepancies than others
+\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018},
+possibly due to larger discriminal dispersions or because they are
+genuine outliers--meaning, texts with distinctive characteristics that
+deviate markedly from the rest of the sample \citep{Grubbs_1969}. In
+either case, the BTL model neither accounts for these anomalies nor
+provides tools for addressing them, apart from excluding the
+``problematic'' texts from analysis.
 
@@
 overestimation, in turn, results in spurious conclusions about these
-differences \citep{McElreath_2020} and, by extension, about the
-underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
-also illustrates this scenario when the model's discriminal difference
-distribution aligns with the thick continuous line for
-\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
-difference follows any discontinuous line where
+differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
+about the underlying discriminal processes of stimuli.
+Figure~\ref{fig-dispersion} also illustrates this scenario when the
+model's discriminal difference distribution aligns with the thick
+continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
+discriminal difference follows any discontinuous line where
 \(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
@@
 judges evaluate multidimensional, complex traits or heterogeneous
-stimuli, they often rely on various intricate characteristics of the
-stimuli to form their judgments
+stimuli, they often rely on a variety of intricate stimulus
+characteristics to inform their judgments
 \citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
-additional relevant characteristics, when assessed, are unlikely to be
-equally weighted or opposing. As a result, they may exert an uneven
-influence on judges' perceptions, creating biases that resist
-cancellation. For example, this could occur when a judge assessing the
-argumentative quality of a text places more weight on its grammatical
-accuracy than other judges, thereby favoring texts with fewer errors but
-weaker arguments. Furthermore, since the discriminal difference of the
-stimuli becomes an observable outcome only through the judges'
-perceptions, these biases could introduce dependencies between the
-stimuli \citep{vanderLinden_et_al_2017_II}. While direct evidence for
-this particular scenario is lacking, studies such as
-\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
-the presence of such biases, supporting the notion that the factors
-influencing pairwise comparisons may not always cancel out.
+characteristics, regardless of their relevance, are unlikely to be
+equally weighted or to oppose one another across comparisons. As a
+result, they may exert a disproportionate influence on judges'
+perceptions, generating biases that persist rather than cancel out. For
+example, this could occur when a judge assessing the argumentative
+quality of a text may place disproportionate emphasis on handwriting
+clarity, thereby favoring neatly written texts despite their weaker
+arguments. Moreover, because the discriminal process of stimuli becomes
+an observable outcome only through the judges' perceptions, these biases
+could introduce dependencies between the stimuli
+\citep{vanderLinden_et_al_2017_II}. Evidence from traditional marking
+contexts confirms a similar phenomenon \citep{Morin_et_al_2018}, and
+additional studies
+\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}
+support the presence of these biases in CJ--reinforcing the argument
+that the factors influencing pairwise comparisons do not always cancel
+each other out.
 
@@
 underestimation of trait reliability
-\citep{Ackerman_1989, Hoyle_et_al_2023}. These inaccuracies can result
-in spurious conclusions about the discriminal differences
-\citep{McElreath_2020} and, by extension, the underlying discriminal
-processes of the stimuli. This issue is illustrated in
-Figure~\ref{fig-correlation} when the discriminal difference
-distribution of the BTL scores follows the thick continuous line
-\((\rho = 0)\), while the \emph{true} discriminal difference follows any
-discontinuous line where \(\rho \neq 0\).
+\citep{Ackerman_1989, Hoyle_et_al_2023} and even biases
+\citep{Wu_et_al_2022}. These inaccuracies can result in spurious
+conclusions about the discriminal differences \citep{McElreath_2020}
+and, by extension, the underlying discriminal processes of the stimuli.
+This issue is illustrated in Figure~\ref{fig-correlation} when the
+discriminal difference distribution of the BTL scores follows the thick
+continuous line \((\rho = 0)\), while the \emph{true} discriminal
+difference follows any discontinuous line where \(\rho \neq 0\).
 
@@
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
-and considered outliers under the BTL model \citep{Grubbs_1969}--risk
-discarding valuable information and introducing bias into trait
-estimates. The direction and magnitude of these biases remain
-unpredictable, as they depend on which judges researchers exclude from
-the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
+and considered outliers under the BTL model
+\citep{Grubbs_1969, Wu_et_al_2022}--risk discarding valuable information
+and introducing bias into trait estimates. The direction and magnitude
+of these biases remain unpredictable, as they depend on which judges
+researchers exclude from the analysis
+\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
 
@@
 CJ literature also show that they frequently rely on point estimates of
-these traits (i.e, the BTL scores)--or their transformations--to conduct
-statistical inference or hypothesis testing. For example, researchers
-have used these scores to identify `misfit' judges and stimuli
-\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
-biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
-calculate correlations with other assessment methods
-\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
-related to the underlying trait of interest
+these traits--typically the BTL scores or its transformations--to
+conduct statistical inference or hypothesis testing. For example,
+researchers have used these scores to identify `misfit' judges and
+stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
+detect biases in judges' ratings
+\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
+other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
+or test hypotheses related to the underlying trait of interest
 \citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 enhancing measurement accuracy and precision, the model still depends on
-the unrealistic assumption that researchers have access to a
+the unrealistic assumption that researchers have access to data from the
 \emph{conceptual population}. Since researchers rarely meet this
@@
 
-We can visualize these vectors more clearly using the example in
-Equation (\ref{eq-mat13}). Suppose researchers exclude the second
-student and the third judge, and they also omit the second text from
-each individual and the third repeated judgment from each judge. Given
-\(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), these
-vectors take the following structure:
+We can illustrate the structure of these vectors more clearly with an
+example. Suppose researchers exclude the second student, the second text
+from each student, and the third judge from the setup shown in Equation
+(\ref{eq-mat13}). Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and
+\(n_{K} = 3\), the resulting vectors would have the following structure:
 \begin{equation}\phantomsection\label{eq-mat24}{
@@
 
-Equation (\ref{eq-mat24}) shows that missing observations in the vectors
-\(S_{I}\) and \(S_{J}\)--which represent unsampled students and
-judges--directly determine which observations are missing in \(S_{IA}\)
-and \(S_{JK}\). In other words, researchers can only observe texts and
-judgments from students and judges initially included in the sample. The
-equation also shows that the sum of observed elements in \(S_{I}\)
-equals the number of sampled students \((n^{s}_{I})\) and that a similar
-sum in vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\).
-Conversely, the sum of observed elements in \(S_{IA}\) represents the
-total sampled texts across all sampled students
+Notably, Equation (\ref{eq-mat24}) shows that missing observations in
+the vectors \(S_{I}\) and \(S_{J}\)--which represent unsampled students
+and judges--directly determine which observations are missing in
+\(S_{IA}\) and \(S_{JK}\). In other words, researchers can only observe
+texts and judgments from students and judges initially included in the
+sample. The equation also shows that the sum of observed elements in
+\(S_{I}\) equals the number of sampled students \((n^{s}_{I})\) and that
+a similar sum in vector \(S_{J}\) equals the sampled judges
+\((n^{s}_{J})\). Conversely, the sum of observed elements in \(S_{IA}\)
+represents the total sampled texts across all sampled students
 \((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
@@
 influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
-encloses this vector, indicating that it is a conditioned variable.
-Here, \emph{conditioning} means that researchers restrict their focus to
-the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
-\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
-vector selecting \emph{all repeated judgments a subset of judges makes
-for the subset of stimuli produced by the sampled individuals}.
+encloses \(S\), indicating that it is a conditioned variable. In this
+context, \emph{conditioning} means that researchers restrict their focus
+to the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
+\citep{Neal_2020, McElreath_2020}. In essence, \(S\) is a vector that
+selects \emph{all repeated judgments made by a subset of judges for a
+subset of stimuli produced by the sampled individuals}.
 
@@
 However, due to concerns about the practical feasibility of the
-comparison task \citep{Boonen_et_al_2020}, CJ experiments rarely
+comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ experiments rarely
 implement an exhaustive pairings of sampled judges, stimuli, and
 individuals. Thus, a realistic scenario must account for the fact that
-judges typically compare only a subset of stimuli authored by certain
-individuals.
+judges typically compare only a subset of stimuli authored by a sample
+of individuals.
 
@@
 
-The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
+The DAG~\ref{fig-cj14_dag} also incorporates the \emph{comparison
 mechanism} \(C\) into the conceptual-population model. It shows the
-sample-comparison outcome as unobserved, highlighting that researchers
-cannot directly access to \(O^{sc}_{R}\) due to the comparison
-mechanism. Moreover, the DAG depicts \(C\) as a causal factor
-influencing \(O_{R}\), enclosing it in a square to indicate that it is a
-conditioned variable. Specifically, \(C\) determines \emph{which
-repeated judgments the subset of judges makes for the subset of stimuli
-produced by the sampled individuals}. In essence, \(C\) reflects the
-assumption that judges \emph{do not} perform all possible repeated
-judgments for the selected stimuli. Instead, they carry out a
-``sufficient number of comparisons'' \(n_{C}\) to allow the accurate
+sample-comparison outcome \(O^{sc}_{R}\) as unobserved, emphasizing that
+researchers cannot directly access this variable because of the
+comparison mechanism. The DAG further shows \(C\) as a conditioned
+variable (enclosed in a square) that causally influences the observed
+outcome \(O_{R}\). This structure implies that \(C\) determines
+\emph{which repeated judgments judges make for the stimuli produced by
+the individuals}. In essence, \(C\) reflects the assumption that judges
+\emph{do not} perform all possible repeated judgments but instead
+complete a sufficient number, \(n_{C}\), to enable the accurate
 estimation of the proportion \(P(B>A)\) for each stimulus pair
@@
 
-With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
-statistical model that addresses violations of the equal dispersion
-assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
-This derivation is possible because a fully specified SCM encodes
-functional and probabilistic information, which we can replace with
-suitable functions and probabilistic assumptions
-\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
-to express the joint distribution of our complex CJ system as a product
-of simpler conditional probability distributions (CPDs)\footnote{This
-  re-expression is possible because the \emph{chain rule} of probability
-  and the \emph{Bayesian Network Factorization (BNF)} property. For
-  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
-as shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
+Using the structural causal model (SCM) \ref{fig-cj15_scm}, we can
+derive a statistical model that addresses violations of the equal
+dispersion assumption (see Section~\ref{sec-theory-issue1a}). This
+derivation is possible because a fully specified SCM encodes functional
+and probabilistic information, which we can replace with suitable
+functions and probabilistic assumptions \citep{Pearl_et_al_2016}.
+Specifically, SCM~\ref{fig-cj15_scm} allows us to express the joint
+distribution of our complex CJ system as a product of simpler
+conditional probability distributions (CPDs)\footnote{This re-expression
+  is possible because the \emph{chain rule} of probability and the
+  \emph{Bayesian Network Factorization (BNF)} property. For further
+  details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as
+shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
 such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
@@
 Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
-which we outline in SCM~\ref{fig-cj16_scm}, the probabilistic model
-\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}. We
-begin by assuming that \(O_{R}\) follows a Bernoulli
-distribution\footnote{The binomial distribution--including its special
-  case, the Bernoulli distribution--represent a maximum entropy
-  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
-  means that the Bernoulli distribution is the most consistent
-  alternative when only two un-ordered outcomes are possible and their
-  expected frequencies are assumed to be constant
-  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
-  binomial as a maximum entropy distribution, see
+which we outline in the statistical model \ref{fig-cj16_stat}. We begin
+by assuming that \(O_{R}\) follows a Bernoulli distribution\footnote{The
+  binomial distribution--including its special case, the Bernoulli
+  distribution--represent a maximum entropy distribution for binary
+  events \citep[pp.~34]{McElreath_2020}. This means that the Bernoulli
+  distribution is the most consistent alternative when only two
+  un-ordered outcomes are possible and their expected frequencies are
+  assumed to be constant \citep[pp.~310]{McElreath_2020}. For a detailed
+  discussion of the binomial as a maximum entropy distribution, see
   \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
@@
 variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
-\(P(Z_{J})\)--we omit these probabilities from the models presented in
-Figure~\ref{fig-cj16}. We make this omission because, although these
-distributions contribute to the overall joint distribution of the data,
-they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
-This independence stems from the random selection designs that govern
-\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
-\(Z_{J}\)\footnote{Randomization ensures that data--and, by extension,
-  an estimator--satisfies several key identification properties, such as
-  common support, no interference, and consistency. The most critical
-  property, however, is the elimination of confounding.
-  \emph{Confounding} occurs when an external variable, such as
-  \(X_{I}\), simultaneously influences both the outcome (e.g.,
-  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
-  spurious associations between the latter two
+\(P(Z_{J})\)--we omit all of these probabilities from the statistical
+model \ref{fig-cj16_stat}. This omission is justified because, while
+these distributions contribute to the overall joint distribution of the
+data, the variables \(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
+\(Z_{J}\) are observed and independent of any other variable in the
+model. As observed variables, they do not require distributional
+assumptions in the same way the idiosyncratic errors do. Furthermore,
+their independence follows from the underlying random selection
+procedures that govern the variables\footnote{Randomization ensures that
+  data--and, by extension, an estimator--satisfies several key
+  identification properties, such as common support, no interference,
+  and consistency. The most critical property, however, is the
+  elimination of confounding. \emph{Confounding} occurs when an external
+  variable, such as \(X_{I}\), simultaneously influences both the
+  outcome (e.g., \(O_{R}\)) and a variable of interest (e.g., \(S\)),
+  resulting in spurious associations between the latter two
   \citep{Everitt_et_al_2010}. Randomization ensure the absence of
@@
 and a covariance matrix
-\(\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}\),
-with \(\boldsymbol{D}\) denoting a diagonal matrix of standard
-deviations and \(\boldsymbol{R}\) a correlation matrix. To address the
+\(\boldsymbol{\Sigma} = \boldsymbol{V} \boldsymbol{Q} \boldsymbol{V}\),
+with \(\boldsymbol{V}\) denoting a diagonal matrix of standard
+deviations and \(\boldsymbol{Q}\) a correlation matrix. To address the
 \emph{location} indeterminacy, we set the errors' mean vector to zero:
@@
 indeterminacy by assuming that the errors are uncorrelated. This
-assumption leads us to specify a correlation matrix for the errors,
-\(\boldsymbol{R}\), as: \begin{equation}\phantomsection\label{eq-mat54}{
-\boldsymbol{R} = \begin{bmatrix}
+assumption leads us to define the error correlation matrix,
+\(\boldsymbol{Q}\), as the identity matrix:
+\begin{equation}\phantomsection\label{eq-mat54}{
+\boldsymbol{Q} = \begin{bmatrix}
     1 & 0 & 0 & 0 \\
@@
 To resolve the \emph{scale} indeterminacy, we define the diagonal matrix
-\(\boldsymbol{D}\) as: \begin{equation}\phantomsection\label{eq-mat55}{
-\boldsymbol{D} = \begin{bmatrix}
+\(\boldsymbol{V}\) as: \begin{equation}\phantomsection\label{eq-mat55}{
+\boldsymbol{V} = \begin{bmatrix}
     s_{XI} & 0 & 0 & 0 \\
@@
 \(1\) to establish the scale of the stimuli' latent trait relative to
-the scale of the individuals' trait. As a result, \(T_{IA}\) is also
-normally distributed. This configuration effectively reinstates
+the scale of the individuals' trait. Note that as a result, \(T_{IA}\)
+is also normally distributed. This configuration effectively reinstates
 Thurstone's original assumption of Normal discriminal processes for the
@@
 
-Finally, we use \emph{Bayesian inference methods} to convert
-model~\ref{fig-cj16_stat} into a practical statistical tool for
-analyzing paired comparison data. Bayesian inference offers three key
-advantages for this purpose. First, it handles complex and
+Finally, we use \emph{Bayesian inference methods} to convert the
+statistical model \ref{fig-cj16_stat} into a practical statistical tool
+for analyzing paired comparison data. Bayesian inference offers three
+main advantages in this context. First, it handles complex and
 overparameterized models, where the number of parameters exceeds the
-number of observations \citep{Baker_1998, Kim_et_al_1999}--such as the
-model we try to implement. Second, it incorporates prior information to
+number of observations \citep{Baker_1998, Kim_et_al_1999}. This
+capability is essential for our implementation, as the model is indeed
+overparameterized. Second, it incorporates prior information to
 constrain parameter estimates within plausible bounds, thereby
@@
 \citep{Martin_et_al_1975, Seaman_et_al_2011}. In our implementation, we
-use priors to define the error distribution, which sets the scale of the
-latent variables \citep{Depaoli_2014}. Third, Bayesian inference
-supports robust inferences from small samples, where the asymptotic
-properties underlying frequentist methods are less reliable
+use mainly use priors to define the error distribution, setting the
+scale of the latent variables \citep{Depaoli_2014}. Third, Bayesian
+inference supports robust inferences from small samples, where the
+asymptotic properties underlying frequentist methods are less reliable
 \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
-feature is particularly relevant in CJ experiments, where researchers
-often collect large volumes of paired comparisons but work with
-relatively small samples of judges, stimuli, and individuals to test
-hypotheses.
+feature is particularly relevant in CJ experiments, as researchers often
+collect large volumes of paired comparisons but work with relatively
+small samples of judges, stimuli, and individuals to test hypotheses.
 
@@
 correlation for the stimuli' discriminal processes--with one notable
-distinction: while Case V assumes normally distributed discriminal
+distinction. While Case V assumes normally distributed discriminal
 processes, the BTL model uses the more mathematically tractable Logistic
-distribution \citep{Andrich_1978, Bramley_2008}.
-
-Although replacing the Normal with the Logistic distribution has a
-minimal impact on trait estimation or model interpretation
-\citep{vanderLinden_et_al_2017_I, McElreath_2021}, Thurstone
-acknowledged that the two remaining assumptions of Case V--and, by
-extension, of the BTL model--could pose significant problems when
-researchers assess complex traits or heterogeneous stimuli
-\citep{Thurstone_1927b}. In particular, he noted that while assuming
-equal discriminal dispersions and zero correlation for the stimuli'
-discriminal processes simplifies the trait measurement model, these
-assumptions may fail to reflect the complexity of CJ data
-\citep{Kelly_et_al_2022, Andrich_1978}. As a result, they can lead to
-unreliable and inaccurate trait estimates
+distribution \citep{Andrich_1978, Bramley_2008}. Although this
+substitution has a minimal impact on trait estimation or model
+interpretation \citep{vanderLinden_et_al_2017_I, McElreath_2021},
+Thurstone acknowledged that assuming equal discriminal dispersions and
+zero correlation among stimuli' discriminal processes could introduce
+significant problems. In particular, he recognized that while these
+assumptions simplify the trait measurement model, they may not capture
+the complexity of some traits or account for heterogeneous stimuli, such
+as handwritten texts or English compositions
+\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
+As a result, they can lead to unreliable and inaccurate trait estimates
 \citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.
 
-Furthermore, because Thurstone's primary goal was to produce a ``coarse
-scaling'' of traits and ``allocate the compared stimuli on this
-continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
+Furthermore, because Thurstone aimed primarily to produce a trait's
+``coarse scaling'' and allocate the compared stimuli along this
+continuum \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
 how to use the resulting trait estimates for statistical inference. The
 CJ tradition has attempted to address this gap by separating trait
-estimation from the hypothesis testing process. Specifically, CJ studies
-often rely on point estimates of the traits (i.e., the BTL scores)--or
-their transformations--to conduct statistical inference. Although
-separating these processes simplifies the analysis of CJ data, the
-statistical literature cautions against this practice, as it can
-introduce bias into the analysis and compromise the reliability of the
-resulting conclusions
+estimation from hypothesis testing. Specifically, CJ studies often rely
+on point estimates of the traits--typically the BTL scores or its
+transformations--to conduct statistical inference. Although this
+separation simplifies CJ data analysis, the statistical literature warns
+that this practice can introduce bias into the analysis and compromise
+the reliability of statistical inferences
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
@@
 Causal and Bayesian inference methods. The approach begins by
-formulating a conceptual model represented by an SCM and a DAG. This
-model integrates Thurstone's core theoretical principles--such as the
-discriminal processes (i.e., traits) of stimuli--alongside key CJ
-experimental design features, including judges' bias, sampling
-procedures, and comparison mechanisms.
+formulating a conceptual model represented by a Structural Causal Model
+(SCM) and a Directed Acyclic Graph (DAG). This model integrates
+Thurstone's core theoretical principles--such as the discriminal
+processes of stimuli--alongside key CJ experimental design features,
+including judges' bias, sampling procedures, and comparison mechanisms,
+thereby disentangling the causal processes in the CJ system.
 
 The study then translates the SCM into a bespoke statistical model that
-addresses key limitations of the BTL model. This model allows
-researchers to analyze CJ data when the assumptions of equal dispersion
-and zero correlation do not hold (see Section~\ref{sec-theory-issue1a}
-and Section~\ref{sec-theory-issue1b}), and when they need to perform
+addresses key limitations of Case V. This model allows researchers to
+analyze CJ data when the assumptions of equal dispersion and zero
+correlation do not hold (see Section~\ref{sec-theory-issue1a} and
+Section~\ref{sec-theory-issue1b}), and when they need to perform
 statistical inference (refer to Section~\ref{sec-theory-issue2}). In
-particular, the model accounts for judge biases, captures the
-hierarchical structure of stimuli, accommodates heterogeneity in
-discriminal dispersions, and incorporates measurement error in the
-hypothesis testing process. These methodological innovations can
-potentially enhance the reliability and validity of trait measurement
-\citep{Perron_et_al_2015} and improve the accuracy of statistical
-inferences.
+particular, the model accounts for judge biases (see
+Section~\ref{sec-theory-theoretical_P1}), captures the hierarchical
+structure of stimuli and incorporates measurement error in the
+hypothesis testing process (refer to
+Section~\ref{sec-theory-theoretical_P3}), and accommodates heterogeneity
+in discriminal dispersions (see Section~\ref{sec-statistical}). These
+methodological innovations can potentially enhance the reliability and
+validity of trait measurement \citep{Perron_et_al_2015} and improve the
+accuracy of statistical inferences.
 
 Beyond these potential benefits, the approach offers two additional
-advantages. First, it clarifies the interactions among all actors and
-processes involved in CJ experiments. Second, it shifts the analytic
+advantages. First, it clarifies the roles and interactions of all actors
+and processes involved in CJ experiments. Second, it shifts the analytic
 paradigm from passively accepting the assumptions of the BTL model to
 actively testing their fit with observed data. Together, these
-advantages create a principled framework for evaluating best practices
-in CJ experimental designs, one that better aligns with the demands of
-contemporary CJ assessment contexts \citep{Kelly_et_al_2022} and opens
-new avenues for research.
-
-Five research avenues merit particular attention for their direct impact
-on the reliability and validity of trait estimates and the accuracy of
-statistical inferences. First, researchers should investigate how
-sampling and comparison mechanisms influence CJ experimental outcomes
-and affect the robustness of statistical inferences. Second, further
-work is needed to understand how judges' biases impact the number of
-comparisons required to achieve reliable trait estimates. Third,
-researchers should re-evaluate existing methods for identifying and
-excluding `misfitting' judges and stimuli to ensure their effectiveness.
-Fourth, studies should explore how the hierarchical structure of stimuli
-influences both experimental design and hypothesis testing in CJ
-contexts. Finally, researchers should explore principled approaches for
-selecting contextually relevant variables.
-
-{About the implications of the sampling and comparison mechanisms}
-
-{About judges biases, judgment discrepancies, and the number of judges
-and judgments}
-
-{About removing misfit judges and stimuli}
+advantages establish a principled framework for evaluating best
+practices in CJ experimental designs, one that better aligns with the
+demands of contemporary CJ assessment contexts \citep{Kelly_et_al_2022},
+offering new insights into existing research and opening promising
+avenues for future inquiry.
+
+Five research avenues--current and new--deserve particular attention.
+The following sections outline these avenues and explain how our model
+provides a rigorous research framework to investigate them. Notably,
+while addressing these questions through formal derivations--following
+the tradition of classical statistical models like linear
+regression--may seem a natural step, we argue that the complexity of the
+CJ system makes this approach initially impractical. Instead,
+simulation-based methods, such as power analysis, offer a more feasible
+and flexible alternative, enabling researchers to investigate these
+questions without relying on cumbersome mathematical proofs.
+Nonetheless, developing formal mathematical proofs remains an essential
+goal for future work.
+
+\subsection{The impact of sampling and comparison mechanisms on CJ
+outcomes and inference robustness}\label{sec-discussion_RA1}
+
+Sampling and comparison mechanisms play a central role in modern CJ
+experimental design. However, most of the CJ literature has examined
+these mechanisms within a limited scope. Researchers have primarily
+investigated the effects of adaptive comparative judgment (ACJ) designs
+on trait reliability
+\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
+or proposed practical guidelines for the number of comparisons judges
+should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}.
+Although these studies offer valuable insights, we argue that they may
+have overlooked the broader role these mechanisms play within the CJ
+system. Moreover, we suggest that this oversight largely stems from a
+lack of conceptual clarity about how the mechanisms function within the
+system. To address this gap, this study decided to integrate these
+mechanisms into CJ's conceptual model.
+
+The explicit integration of the sampling and comparison mechanisms
+offers a new perspective on how these mechanisms shape the CJ process.
+In particular, it clarifies their role in the data-generating process by
+framing them as sources of missing data--that is, as mechanisms that
+determine which observations are missing from the final data sample.
+This new perspective encourages the application of Little and Rubin's
+principled missing data framework \citeyearpar{Little_et_al_2020},
+enabling a more rigorous evaluation of existing claims about these
+mechanisms, their influence on CJ outcomes, and their implications for
+designing and assessing more complex experimental setups.
+
+Notably, this study circumvents the need to apply this missing data
+framework by deliberately structuring the sampling and comparison
+mechanisms to be independent of any observed or unobserved variables,
+including the outcome. In other words, we designed these mechanisms so
+they produce data that are \emph{missing completely at random} (MCAR)
+\citep{Little_et_al_2020}. This design offers one key advantage: it
+generates simple random samples that satisfy the condition of
+\emph{ignorability}, allowing researchers to legitimately \emph{ignore}
+missing data during analysis without introducing bias
+\citep{Everitt_et_al_2010, Kohler_et_al_2019, Neal_2020}.
+
+However, many modern CJ applications rely on more complex experimental
+designs in which the sampling and comparison mechanisms introduce more
+intricate forms of missingness--such as \emph{missing at random} (MAR)
+or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
+prominent example is the previously discussed ACJ designs, where prior
+judgment outcomes inform the selection of stimulus pairs for subsequent
+comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
+outcome-informed selection suggests that the comparison mechanism may be
+outcome-dependent, potentially classifying ACJ as a generator of MNAR
+data. This classification may, in turn, facilitate a more nuanced
+understanding of the mixed evidence surrounding ACJ's capabilities.
+While some studies report improvements in trait reliability
+\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
+that these gains are artificially inflated by the method
+\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
+
+Nevertheless, regardless of the underlying missingness mechanisms, any
+CJ experimental design would benefit from explicitly defining its
+assumptions--an approach supported by our framework. This clarity
+enables researchers to evaluate, for instance, how sampling and
+comparison mechanisms affect trait estimation and statistical inference
+within each design. Such assessments are particularly relevant given the
+common misconception in the CJ literature that Thurstone's model can
+naturally handle even non-random missing data without compromising the
+reliability or validity of trait estimates \citep{Bramley_2008}.
+
+\subsection{The effects of judges' bias on the reliability of
+traits}\label{sec-discussion_RA2}
+
+Despite the growing notion that various stimulus-related factors can
+influence judges' perceptions during pairwise comparisons--and that
+these influences may not always cancel each other out--few studies in
+the CJ literature provide empirical evidence for judges' biases
+\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
+We argue that this gap persists not due to a lack of interest or
+research but because researchers continue to rely on ad-hoc detection
+methods, such as `misfit' statistics, that are poorly suited for the
+task \citep{Kelly_et_al_2022}. To address this limitation, this study
+decided to treat judges' biases as an integral component of the CJ
+system from the outset. This approach offers one key advantage: it
+provides a more accurate representation of the data-generating process
+behind pairwise comparisons--one that acknowledges that the discriminal
+processes of stimuli become an observable outcome only through judges'
+perceptions, which may exhibit bias.
+
+The explicit integration of judges' bias into CJ's conceptual model then
+paves the way for investigating several relevant research questions. For
+instance, can researchers validly analyze CJ data under the assumption
+of ``sample-free'' trait calibration--specifically, under the assumption
+that judges exhibit no systematic bias? This question is especially
+relevant because researchers often treat ``sample-freeness'' as an
+inherent property of the BTL model \citep{Bramley_2008, Andrich_1978}
+despite growing evidence of persistent biases. Furthermore, if judges'
+biases reflect additional stimulus characteristics that should not
+influence assessment, to what extent can training or expertise mitigate
+these biases? \citep{Kelly_et_al_2022} More specifically, as discussed
+in Section~\ref{sec-theory-issue1b}, can training or expertise help
+prevent judges from prioritizing irrelevant features--such as
+handwriting--over more central criteria like argumentative quality in
+writing assessments? Addressing these questions may also shed light on
+what it truly means to be an ``expert'' in the CJ context
+\citep{Kelly_et_al_2022}.
+
+Moreover, if we accept that judges may value different aspects of a
+stimulus, it follows that their judgments might also vary with
+individual characteristics such as gender, age, culture, income,
+education, training, or expertise
+\citep{Kelly_et_al_2022, Bartholomew_et_al_2020a}, and even from one
+individual to another
+\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
+Building on this and the discussion in Section~\ref{sec-discussion_RA1},
+researchers could investigate how the selection of judges influences the
+formation of a ``shared consensus'' and whether these individual
+characteristics introduce systematic biases or distortions in the
+observed trait distribution \citep{Deffner_et_al_2022}. Furthermore, if
+such characteristics indeed undermine the ``sample-freeness''
+assumption, then it becomes essential to explore strategies for
+controlling their influence. This includes investigating how many
+judges--and how many judgments per judge--are required to produce
+reliable trait estimates, as well as whether \emph{repeated measures
+designs}, in which judges evaluate the same stimulus pairs multiple
+times \citep{Lawson_2015}, can improve judgment consistency and
+accuracy.
+
+\subsection{The identification of `misfitting' judges and
+stimuli}\label{sec-discussion_RA3}
+
+{About the identification of `misfitting' judges and stimuli}
+
+\subsection{The effects of hierarchies on experimental design and
+inference}\label{sec-discussion_RA4}
 
@@
 
-{About revisit the approaches for selecting contextually relevant
-variables}
-
-Despite its potential, this study presents several challenges that
-researchers must navigate. First, our approach does not establish a
-clear epistemological connection between the theoretical latent trait
-and its quantitative estimate. In other words, the approach does not
-provide a definitive answer to the fundamental question: How do we know
-that the estimated trait reflects the theoretical trait of interest?
-However, this limitation is not unique to our approach; it reflects a
-fundamental challenge all research involving latent constructs faces.
-Nevertheless, we argue that our approach offers methodological tools
-that enable researchers to design CJ experiments under conditions that
-support reliable and valid trait measurement.
-
-Second, although initial simulation tests of the statistical model have
-yielded promising results (not shown), these findings may not generalize
-across research contexts. We have argued that the model's ability to
-accurately estimate traits and support valid statistical inferences
-depends heavily on the specific research context--particularly on
-whether the data and inferential goals align with the model's underlying
-assumptions. That said, our approach provides tools for researchers to
-test and evaluate these context-dependent CJ models directly.
-
-Third, applying this approach requires a solid understanding of Causal
-inference. For instance, throughout the study, we assume that predictor
-variables must be ``relevant'' to the research context, interpreting
-relevance as the inclusion of the predictor variables in a ``sufficient
-adjustment set.'' However, we do not fully explore what this entails or
-its implications for model specification and validity. Nonetheless, we
-provide key references throughout the study to help motivated
-researchers explore these foundational concepts in greater depth.
-
-Fourth and final, researchers must also be proficient in Probabilistic
-Programming Languages (PPLs) and Bayesian inference methods. This
-expertise is essential for translating context-specific conceptual
-models into bespoke statistical models that meaningfully address the
-research questions at hand. While such translations are not exclusive to
-Bayesian methods, we argue that Bayesian approaches offer a more
-accessible path. They reduce, for instance, the need for additional
-expertise in optimization techniques and theory often required by
-frequentist methods, thereby lowering the technical barrier for many
-applied researchers.
+\subsection{The effects of non-dichotomous outcomes and multidimensional
+traits on CJ experiments}\label{sec-discussion_RA5}
+
+{About non-dichotomous outcomes and multidimensional traits}
+
+\subsection{Challenges}\label{sec-discussion_challenges}
 

### 2025-05-29 paper2.qmd
section:6; process:write

@@
 [^1]: These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, and introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 are useful.
-For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020, are recommended.
+For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2025, are recommended.
 
@@
 Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
-
-Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
-This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
-Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
-Thus, we define the matrix as follows:
-
 $$
@@
 
-It is easier to visualize the structure of these vectors and matrices by considering an example.
-Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
+Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
+This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
+Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
+Thus, we define the matrix as in Equation ([-@eq-mat12]).
+
+It is easier to visualize the structure of the previously defined vectors and matrices by considering an example.
+Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the form as in Equation ([-@eq-mat13]).
 $$
@@
 Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
-
 $$
@@
 Randomization ensure the absence of confounding by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself.
-For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.
+For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2025.
 
@@
 However, Thurstone identified one key challenge with its measurement model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
-To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from this general form, each progressively incorporating several simplifying assumptions.
+To address this issue and facilitate the theory's practical applicability, he formulated five cases, each progressively incorporating several simplifying assumptions.
 
 <!-- 2. Summary of the why the literature uses Case V, and its problems -->
-Of these five cases, the CJ literature has predominantly relied on Case V to measure various psychological traits, as shown in studies like @Kimbell_2012, @Jones_et_al_2015, and @Boonen_et_al_2020.
-This preference largely stems from the widespread adoption of the BTL model, which offers a simplified statistical formulation of the case.
-The BTL model mirrors the assumptions of Case V--namely, equal discriminal dispersions and zero correlation for the stimuli' discriminal processes--with one notable distinction.
-While Case V assumes normally distributed discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008].
-Although this substitution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that assuming equal discriminal dispersions and zero correlation among stimuli' discriminal processes could introduce significant problems.
-In particular, he recognized that while these assumptions simplify the trait measurement model, they may not capture the complexity of some traits or account for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
-As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
-
-Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and allocate the compared stimuli along this continuum [@Thurstone_1927a, pp.
-269], his theory did not specify how to use the resulting trait estimates for statistical inference.
-The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing.
-Specifically, CJ studies often rely on point estimates of the traits--typically the BTL scores or its transformations--to conduct statistical inference.
-Although this separation simplifies CJ data analysis, the statistical literature warns that this practice can introduce bias into the analysis and compromise the reliability of statistical inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+Among these, Case V has become the most widely used model in empirical CJ research, primarily due to the widespread adoption of the BTL model.
+The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008].
+Although the substitution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021] and simplifies the measurement model, the other two core assumptions of the model can still lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023], especially when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
+
+Furthermore, because Thurstone primarily aimed to produce a trait's "coarse scaling" and allocate the compared stimuli along this continuum [@Thurstone_1927a, pp.
+269], his theory did not specify how to use the trait estimates for statistical inference.
+The CJ tradition has attempted to fill this gap by separating trait estimation from hypothesis testing.
+Specifically, researchers often rely on point estimates of the traits, such as the BTL scores or their transformations, for inference.
+Although this approach simplifies analysis, the statistical literature cautions that it can introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
 <!-- 4. General summary of what we have done. -->
-Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
+Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
 The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG).
-This model integrates Thurstone's core theoretical principles--such as the discriminal processes of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.
+This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.
 
@@
 
+<!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
 Sampling and comparison mechanisms play a central role in modern CJ experimental design.
@@
 
+<!-- 2. The missing data perspective on sampling and comparison mechanisms -->
 The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process.
@@
 
+<!-- 3. Leveraging MCAR assumptions to simplify CJ analysis in this study -->
 Notably, this study circumvents the need to apply this missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome.
@@
 
+<!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
 However, many modern CJ applications rely on more complex experimental designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
@@
 
+<!-- 5. The importance of clarifying assumptions behind CJ experimental designs -->
 Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions--an approach supported by our framework.
@@
 
+<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
 Despite the growing notion that various stimulus-related factors can influence judges' perceptions during pairwise comparisons--and that these influences may not always cancel each other out--few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
@@
 
+<!-- 2. Relevant research question paved by recognizing judges' bias (part I) -->
 The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions.
@@
 
+<!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
 Moreover, if we accept that judges may value different aspects of a stimulus, it follows that their judgments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020].
@@
 
-<!-- 7.1 PPL and Bayesian Inference -->
-<!-- Despite its potential, this study presents several challenges that researchers must navigate.
-First, researchers need to be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods.
-This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand.
-While such translations are not exclusive to Bayesian methods, we argue that Bayesian approaches offer a more accessible path.
-They reduce, for instance, the need for additional expertise in optimization techniques and theory often required by frequentist methods, thereby lowering the technical barrier for many applied researchers.
--->
-
-<!-- make these four into one general challenge, "something you can always say about of almost every study".
-Now the challenge need a thorough knowledge about the CJ system.
-No easy route now, they have to think about their research context, assumptions in the experiment, model.
--->
-
-<!-- our framework is not designed to work for everything, is just a starting point to think about which model applies to your research context  -->
-
-<!-- "although in this paper we have outlined the causal process behind the CJ system, this does not refrain other research to thoroughly think their causal structure in their studies" -->
-
-<!-- 7.2 Research context is important  -->
-<!-- We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
+<!-- 1. Challenges for the framework: background knowledge and research context -->
+Despite its promise, this study presents several challenges that researchers must navigate --challenges common to many quantitative studies.
+These challenges fall into two main categories.
+The first concerns the background knowledge researchers need to apply the proposed framework effectively.
+The second involves the increased attention researchers must now devote to their specific CJ research context.
+
+<!-- 2. Challenge 1: background knowledge -->
+In terms of background knowledge, researchers face two challenges: mastering Causal inference and developing proficiency in Bayesian inference methods.
+One example illustrating the need to master Causal inference is our assumption that predictor variables are "relevant" to the research context--interpreting relevance as the inclusion of these variables in a "sufficient adjustment set" [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+However, we do not fully explore what this entails or its implications for model specification, estimation, and inference.
+As a result, researchers must engage deeply with these complex ideas--and with others, such as how SCMs encode functional and probabilistic information--to apply the framework rigorously.
+To assist this effort, we provide key references throughout the study to guide motivated readers toward a deeper understanding of these foundational concepts[^9].
+
+<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014).
+For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
+[@Deffner_et_al_2022, pp.
+2] -->
+
+<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
+
+<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
+4-5] -->
+
+<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
+
+[^9]: refer to footnote 1, 2, and the detailed online document referenced in the Declarations section
+
+Bayesian inference presents a second challenge.
+Researchers must acquire sufficient expertise to translate context-specific conceptual models into bespoke statistical models that meaningfully address their research questions.
+Although such translations are not unique to Bayesian approaches, we argue that Bayesian methods offer a more accessible route, as they reduce the need for specialized knowledge in optimization theory and techniques--often required by frequentist methods--thereby lowering the technical barriers for many applied researchers.
+
+<!-- 3. Challenge 2: research context -->
+In terms of the attention researchers must now devote to their specific CJ research context, researchers face two challenges: ensuring the conceptual model is a *faithful* representation for the CJ system, and verifying whether this model can accurately estimate the intended estimand from real data.
+
+<!-- See fine point 6.1 and 6.2 [@Hernan_et_al_2025, pp. 79-81] -->
+
+<!-- we are confident that these causal structure is faithful to a design -->
+
+<!-- Ideally, we would like to assess this property through an *identification analysis* [@Schuessler_et_al_2023], where researchers assess how accurate a model computes an estimand based solely on its (causal) assumptions, regardless of random variability.
+As one can imagine these types of analysis requires formal proofs, nevertheless, we argue that simulation-based methods can provide preliminary assessments akin to an identification analysis.
+-->
+
+<!-- In this regard, we argue that researchers face the challenge that need a thorough knowledge about the CJ system, particularly regarding experimental design, the assumptions underlying the data analysis, and appropriate model specification.
+This is neccesary because the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
+For instance, the selection of relevant variables largely depends on the research context, and thus you really have to think through which variables you have to put in in your analysis.
+-->
+
+<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
+Nevertheless, as @Kohler_et_al_2019 [pp.
+150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
+-->
+
+<!-- Nevertheless, although in this paper we have outlined the causal process behind the CJ system, our framework is not designed to work every CJ design or data, is just a starting point for researchers to think thoroughly about the causal structure and statistical model that applies for their research context.
 That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.
@@
 
-<!-- 7.4 Causal inference knowledge -->
-<!-- Fourth, applying this approach requires a solid understanding of Causal inference.
-For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity.
-Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.
--->
-
-<!-- 7.5 Epistemiology  -->
 <!-- Fifth and final, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate.
@@
 
-<!-- 7.3 Faithfulness  -->
-
-<!-- remember to mention fidelity of the theory in the model -->
-
-<!-- Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable.
-Relevant variables depends on the context, you really have to think through which variables you have to put in.
--->
-
-<!-- These procedures also bear on a larger topic of discussion in social science.
-The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
-[@Wainer_et_al_1978, pp.
-332] -->
+<!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
+[@Deffner_et_al_2022, pp.
+4] -->
+
+<!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
+But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
+[@Deffner_et_al_2022, pp.
+14] -->
+
+<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
+[@Deffner_et_al_2022, pp.
+14] -->
+
+<!-- This approach makes strong causal assumptions about the nature of confounding and our ability to measure shared history.
+However, strong assumptions are always necessary in observational settings.
+What is important is that the assumptions are transparent and logically connected to data analysis.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+<!-- DAGs are heuristic—they are not detailed statistical models.
+But they allow us to deduce which statistical models can provide valid causal inferences, assuming the DAG is true.
+[@McElreath_2020, pp.
+17] -->
+
+<!-- DAGs are not about adding assumptions—they are about revealing the assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
+
+<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
+
+<!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
+Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
+12].
+-->
+
+<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
+[@Schuessler_et_al_2023, pp.
+3] -->
+
+<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
+But an incomplete model is still an improvement over no model at all.
+[@Deffner_et_al_2022, pp.
+5] -->
+
+<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
+<!-- 1.
+No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
+-->
+<!-- 2.
+No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
+-->
+<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
+Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
+[@Lash_et_al_2021, pp.
+113] -->
+
+<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
+This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
+-->
+
+<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
+Nevertheless, as @Kohler_et_al_2019 [pp.
+150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
+-->
 
@@
 
-<!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
-But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
-[@Deffner_et_al_2022, pp.
-14] -->
-
-<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
-[@Deffner_et_al_2022, pp.
-14] -->
-
-<!-- This approach makes strong causal assumptions about the nature of confounding and our ability to measure shared history.
-However, strong assumptions are always necessary in observational settings.
-What is important is that the assumptions are transparent and logically connected to data analysis.
-[@Deffner_et_al_2022, pp.
-13] -->
-
-<!-- DAGs are heuristic—they are not detailed statistical models.
-But they allow us to deduce which statistical models can provide valid causal inferences, assuming the DAG is true.
-[@McElreath_2020, pp.
-17] -->
-
-<!-- DAGs are not about adding assumptions—they are about revealing the -->
-<!-- assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
-
-<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
-
-<!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
-Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
-12].
--->
-
-<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
-[@Schuessler_et_al_2023, pp.
-3] -->
-
-<!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
-[@Deffner_et_al_2022, pp.
-4] -->
-
-<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
-But an incomplete model is still an improvement over no model at all.
-[@Deffner_et_al_2022, pp.
-5] -->
-
-<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
-
-<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
-[@Deffner_et_al_2022, pp.
-13] -->
-
-<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
-4-5] -->
-
-<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014).
-For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
-[@Deffner_et_al_2022, pp.
-2] -->
-
-<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
-
-<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
-<!-- 1.
-No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
--->
-<!-- 2.
-No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
--->
-<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
-Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
-[@Lash_et_al_2021, pp.
-113] -->
-
-<!-- *Relevant* refers to variables that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
-Use your example, teaching practice and Aptitude-Treatment Interaction (ATI), certain interventions do not have the same effect for all pupils, treatment is a random intercept and slope variable.
-Relevant variables depends on the context, you really have to think through which variables you have to put in.
--->
-
-<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
-This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
--->
-
-<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
-Nevertheless, as @Kohler_et_al_2019 [pp.
-150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
--->
-
-<!-- Statements about the real world from observed data require assumptions.
-This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
-The validity of our statements therefore critically hinges on the validity of those assumptions.
-If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
-[@Kohler_et_al_2019, pp.
-150] -->
+<!-- These procedures also bear on a larger topic of discussion in social science.
+The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
+[@Wainer_et_al_1978, pp.
+332] -->
 
@@
 They may also find it helpful to consult introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020.
-For more advanced study, readers may refer to seminal intermediate papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020.
+For more advanced study, readers may refer to seminal intermediate papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2025.
 

### 2025-05-29 _tex/paper2.tex
Built site for gh-pages

@@
   such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
-  \citet{Hernan_et_al_2020}, are recommended.}. First, regardless of
+  \citet{Hernan_et_al_2025}, are recommended.}. First, regardless of
 complexity, they can represent various causal structures using only five
@@
 columns, where \(n_{K}\) indicates the number of repeated judgments each
-judge makes.
-
-Additionally, we construct the matrix \(R\) to map each row of the
-\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
-matrix has \(n\) rows and \(6\) columns, where
-\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
-term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
-coefficient, which quantifies the total number of unique comparisons
-possible between every pair of stimuli generated by each pair of
-individuals in the population. Thus, we define the matrix as follows:
-
-\begin{equation}\phantomsection\label{eq-mat12}{
+judge makes. \begin{equation}\phantomsection\label{eq-mat12}{
 R = \begin{bmatrix}
@@
 
-It is easier to visualize the structure of these vectors and matrices by
-considering an example. Assuming \(n_{I} = 5\), \(n_{A} = 2\),
-\(n_{J} = 3\), and \(n_{K} = 3\), the vectors and matrices described in
-Equations (\ref{eq-mat11}) and (\ref{eq-mat12}) take the following form:
+Additionally, we construct the matrix \(R\) to map each row of the
+\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
+matrix has \(n\) rows and \(6\) columns, where
+\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
+term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
+coefficient, which quantifies the total number of unique comparisons
+possible between every pair of stimuli generated by each pair of
+individuals in the population. Thus, we define the matrix as in Equation
+(\ref{eq-mat12}).
+
+It is easier to visualize the structure of the previously defined
+vectors and matrices by considering an example. Assuming \(n_{I} = 5\),
+\(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), the vectors and
+matrices described in Equations (\ref{eq-mat11}) and (\ref{eq-mat12})
+take the form as in Equation (\ref{eq-mat13}).
 \begin{equation}\phantomsection\label{eq-mat13}{
@@
   randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
-  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.
+  \citet{Neal_2020}, and \citet{Hernan_et_al_2025}.}.
 
@@
 \citep{Thurstone_1927b}. To address this issue and facilitate the
-theory's practical applicability, he formulated five cases derived from
-this general form, each progressively incorporating several simplifying
-assumptions.
-
-Of these five cases, the CJ literature has predominantly relied on Case
-V to measure various psychological traits, as shown in studies like
-\citet{Kimbell_2012}, \citet{Jones_et_al_2015}, and
-\citet{Boonen_et_al_2020}. This preference largely stems from the
-widespread adoption of the BTL model, which offers a simplified
-statistical formulation of the case. The BTL model mirrors the
-assumptions of Case V--namely, equal discriminal dispersions and zero
-correlation for the stimuli' discriminal processes--with one notable
-distinction. While Case V assumes normally distributed discriminal
-processes, the BTL model uses the more mathematically tractable Logistic
-distribution \citep{Andrich_1978, Bramley_2008}. Although this
-substitution has a minimal impact on trait estimation or model
-interpretation \citep{vanderLinden_et_al_2017_I, McElreath_2021},
-Thurstone acknowledged that assuming equal discriminal dispersions and
-zero correlation among stimuli' discriminal processes could introduce
-significant problems. In particular, he recognized that while these
-assumptions simplify the trait measurement model, they may not capture
-the complexity of some traits or account for heterogeneous stimuli, such
-as handwritten texts or English compositions
+theory's practical applicability, he formulated five cases, each
+progressively incorporating several simplifying assumptions.
+
+Among these, Case V has become the most widely used model in empirical
+CJ research, primarily due to the widespread adoption of the BTL model.
+The BTL model mirrors the core assumptions of Case V--namely, equal
+discriminal dispersions and zero correlation among stimuli' discriminal
+processes--but replaces the processes' normal distribution with the more
+mathematically tractable logistic distribution
+\citep{Andrich_1978, Bramley_2008}. Although the substitution has a
+minimal impact on trait estimation or model interpretation
+\citep{vanderLinden_et_al_2017_I, McElreath_2021} and simplifies the
+measurement model, the other two core assumptions of the model can still
+lead to unreliable and inaccurate trait estimates
+\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023},
+especially when researchers assess complex traits or heterogeneous
+stimuli
 \citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
-As a result, they can lead to unreliable and inaccurate trait estimates
-\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.
-
-Furthermore, because Thurstone aimed primarily to produce a trait's
+
+Furthermore, because Thurstone primarily aimed to produce a trait's
 ``coarse scaling'' and allocate the compared stimuli along this
 continuum \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
-how to use the resulting trait estimates for statistical inference. The
-CJ tradition has attempted to address this gap by separating trait
-estimation from hypothesis testing. Specifically, CJ studies often rely
-on point estimates of the traits--typically the BTL scores or its
-transformations--to conduct statistical inference. Although this
-separation simplifies CJ data analysis, the statistical literature warns
-that this practice can introduce bias into the analysis and compromise
-the reliability of statistical inferences
+how to use the trait estimates for statistical inference. The CJ
+tradition has attempted to fill this gap by separating trait estimation
+from hypothesis testing. Specifically, researchers often rely on point
+estimates of the traits, such as the BTL scores or their
+transformations, for inference. Although this approach simplifies
+analysis, the statistical literature cautions that it can introduce bias
+and compromise the reliability of the resulting inferences
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
 
-Thus, to address the limitations of Thurstone's Case V and the BTL
-model--particularly their strong assumptions and the disconnect between
-trait measurement and hypothesis testing--this study extends Thurstone's
-general form using a systematic, integrated approach that combines
-Causal and Bayesian inference methods. The approach begins by
-formulating a conceptual model represented by a Structural Causal Model
-(SCM) and a Directed Acyclic Graph (DAG). This model integrates
-Thurstone's core theoretical principles--such as the discriminal
-processes of stimuli--alongside key CJ experimental design features,
-including judges' bias, sampling procedures, and comparison mechanisms,
-thereby disentangling the causal processes in the CJ system.
+Thus, to address these limitations of Thurstone's Case V and the BTL
+model, this study extends Thurstone's general form using a systematic,
+integrated approach that combines Causal and Bayesian inference methods.
+The approach begins by formulating a conceptual model represented by a
+Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This
+model integrates Thurstone's core theoretical principles, such as the
+discriminal processes of stimuli, alongside key CJ experimental design
+features, including judges' bias, sampling procedures, and comparison
+mechanisms, thereby disentangling the causal processes in the CJ system.
 
@@
 
+Despite its promise, this study presents several challenges that
+researchers must navigate --challenges common to many quantitative
+studies. These challenges fall into two main categories. The first
+concerns the background knowledge researchers need to apply the proposed
+framework effectively. The second involves the increased attention
+researchers must now devote to their specific CJ research context.
+
+In terms of background knowledge, researchers face two challenges:
+mastering Causal inference and developing proficiency in Bayesian
+inference methods. One example illustrating the need to master Causal
+inference is our assumption that predictor variables are ``relevant'' to
+the research context--interpreting relevance as the inclusion of these
+variables in a ``sufficient adjustment set''
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
+not fully explore what this entails or its implications for model
+specification, estimation, and inference. As a result, researchers must
+engage deeply with these complex ideas--and with others, such as how
+SCMs encode functional and probabilistic information--to apply the
+framework rigorously. To assist this effort, we provide key references
+throughout the study to guide motivated readers toward a deeper
+understanding of these foundational concepts\footnote{refer to footnote
+  1, 2, and the detailed online document referenced in the Declarations
+  section}.
+
+Bayesian inference presents a second challenge. Researchers must acquire
+sufficient expertise to translate context-specific conceptual models
+into bespoke statistical models that meaningfully address their research
+questions. Although such translations are not unique to Bayesian
+approaches, we argue that Bayesian methods offer a more accessible
+route, as they reduce the need for specialized knowledge in optimization
+theory and techniques--often required by frequentist methods--thereby
+lowering the technical barriers for many applied researchers.
+
+In terms of the attention researchers must now devote to their specific
+CJ research context, researchers face two challenges: ensuring the
+conceptual model is a \emph{faithful} representation for the CJ system,
+and verifying whether this model can accurately estimate the intended
+estimand from real data.
+
 \newpage{}
@@
 as well as books such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
-and \citet{Hernan_et_al_2020}.
+and \citet{Hernan_et_al_2025}.
 

### 2025-05-30 paper2.qmd
section:6; process:write

@@
 Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli.
-While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
+While these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates.
 Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes.
@@
 <!-- 1. What is the problem with the general form -->
-Thurstone noted from the outset that his general form, described in @sec-thurstone_theory, led to a *trait scaling problem*.
+Thurstone noted from the outset that his general formulation, described in @sec-thurstone_theory, led to a *trait scaling problem*.
 Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
-For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases).
+For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases).
 However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.
@@
 Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level.
-However, Thurstone identified one key challenge with its measurement model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
+However, Thurstone identified one key challenge in this general formulation: the measurement model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
 To address this issue and facilitate the theory's practical applicability, he formulated five cases, each progressively incorporating several simplifying assumptions.
@@
 <!-- 2. Summary of the why the literature uses Case V, and its problems -->
-Among these, Case V has become the most widely used model in empirical CJ research, primarily due to the widespread adoption of the BTL model.
+Among these, Case V remains the most widely used model in empirical CJ research, mainly due to the widespread adoption of the BTL model.
 The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008].
-Although the substitution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021] and simplifies the measurement model, the other two core assumptions of the model can still lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023], especially when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
-
-Furthermore, because Thurstone primarily aimed to produce a trait's "coarse scaling" and allocate the compared stimuli along this continuum [@Thurstone_1927a, pp.
-269], his theory did not specify how to use the trait estimates for statistical inference.
-The CJ tradition has attempted to fill this gap by separating trait estimation from hypothesis testing.
-Specifically, researchers often rely on point estimates of the traits, such as the BTL scores or their transformations, for inference.
-Although this approach simplifies analysis, the statistical literature cautions that it can introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+Notably, although these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
+
+Moreover, because Thurstone's original goal was to produce a "coarse scaling" of  traits and allocate stimuli along this continuum [@Thurstone_1927a, pp.
+269], his theory offered no guidance on how to use trait estimates for statistical inference.
+To address this gap, researchers in the CJ tradition have often separated trait estimation from hypothesis testing, relying on point estimates of the traits--such as BTL scores or their transformations--for inference.
+Notably, while this approach simplifies analysis, it can also introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
@@
 
-The study then translates the SCM into a bespoke statistical model that addresses key limitations of Case V.
-This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2).
-In particular, the model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli and incorporates measurement error in the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical).
-These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.
+The study then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur (see @sec-theory-issue1a and @sec-theory-issue1b) and when statistical inference is necessary (refer to @sec-theory-issue2).
+In particular, this model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical).
+By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.
 
@@
 
-Five research avenues--current and new--deserve particular attention.
+Five research avenues, both current and new, deserve particular attention.
 The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
-Notably, while addressing these questions through formal derivations--following the tradition of classical statistical models like linear regression--may seem a natural step, we argue that the complexity of the CJ system makes this approach initially impractical.
-Instead, simulation-based methods, such as power analysis, offer a more feasible and flexible alternative, enabling researchers to investigate these questions without relying on cumbersome mathematical proofs.
+Notably, although it may seem like a natural next step to address these questions through formal derivations, following the tradition of classical statistical models like linear regression, we contend that the complexity of the CJ system makes this approach impractical from the outset.
+Instead, simulation-based methods, such as power analysis, offer a more feasible and flexible alternative, enabling researchers to investigate these questions without depending on cumbersome mathematical proofs.
 Nonetheless, developing formal mathematical proofs remains an essential goal for future work.
@@
 <!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
-Sampling and comparison mechanisms play a central role in modern CJ experimental design.
-However, most of the CJ literature has examined these mechanisms within a limited scope.
+Although sampling and comparison mechanisms are central to modern CJ experimental design, it is striking that most of the CJ literature has examined them only within a limited scope.
 Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022].
-Although these studies offer valuable insights, we argue that they may have overlooked the broader role these mechanisms play within the CJ system.
+Although these studies offer valuable insights, we argue that they have overlooked the broader role these mechanisms play within the CJ system.
 Moreover, we suggest that this oversight largely stems from a lack of conceptual clarity about how the mechanisms function within the system.
@@
 This classification may, in turn, facilitate a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities.
-While some studies report improvements in trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b],  others argue that these gains are artificially inflated by the method [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
+While some studies report that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], others argue that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
 
@@
 <!-- 5. The importance of clarifying assumptions behind CJ experimental designs -->
-Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions--an approach supported by our framework.
+Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions, an approach supported by our framework.
 This clarity enables researchers to evaluate, for instance, how sampling and comparison mechanisms affect trait estimation and statistical inference within each design.
@@
 <!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
-Despite the growing notion that various stimulus-related factors can influence judges' perceptions during pairwise comparisons--and that these influences may not always cancel each other out--few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
+Despite the growing notion that various stimulus-related factors can influence judges' perceptions [@Morin_et_al_2018] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
 We argue that this gap persists not due to a lack of interest or research but because researchers continue to rely on ad-hoc detection methods, such as 'misfit' statistics, that are poorly suited for the task [@Kelly_et_al_2022].
 To address this limitation, this study decided to treat judges' biases as an integral component of the CJ system from the outset.
-This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons--one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.
+This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.
 
@@
 <!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
-Moreover, if we accept that judges may value different aspects of a stimulus, it follows that their judgments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020].
-Building on this and the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
-Furthermore, if such characteristics indeed undermine the "sample-freeness" assumption, then it becomes essential to explore strategies for controlling their influence.
-This includes investigating how many judges--and how many judgments per judge--are required to produce reliable trait estimates, as well as whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
+Moreover, if we accept that judges may prioritize different aspects of a stimulus, it follows that their assessments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020].
+Thus, building on the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
+Furthermore, if such characteristics indeed undermine the assumption of "sample-freeness," then it becomes essential to explore strategies to control their influence and to determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts.
+Additionally, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
 

### 2025-05-30 _tex/paper2.tex
Built site for gh-pages

@@
 these assumptions simplify the trait measurement model, they may fail to
-capture the complexity of CJ data, potentially leading to unreliable and
-inaccurate trait estimates. Second, the study highlights the disconnect
-between CJ's trait measurement and hypothesis testing processes.
-Although separating these processes simplifies the analysis of CJ data,
-it may also undermine the reliability of various statistical inferences
-derived from these processes.
+capture the complexity of some traits or account for heterogeneous
+stimuli, potentially leading to unreliable and inaccurate trait
+estimates. Second, the study highlights the disconnect between CJ's
+trait measurement and hypothesis testing processes. Although separating
+these processes simplifies the analysis of CJ data, it may also
+undermine the reliability of various statistical inferences derived from
+these processes.
 
@@
 
-Thurstone noted from the outset that his general form, described in
-Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
+Thurstone noted from the outset that his general formulation, described
+in Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
 problem}. Specifically, the model required estimating more ``unknown''
@@
 modal discriminal processes, five discriminal dispersions, and \(10\)
-correlations --one per comparison (see Table~\ref{tbl-thurstone_cases}).
+correlations--one per comparison (see Table~\ref{tbl-thurstone_cases}).
 However, a single judge could only provide \({5 \choose 2} = 10\) unique
@@
 indicating which stimulus the judge perceives as having a higher trait
-level. However, Thurstone identified one key challenge with its
-measurement model: it required estimating more ``unknown'' parameters
-than the number of available pairwise comparisons
+level. However, Thurstone identified one key challenge in this general
+formulation: the measurement model required estimating more ``unknown''
+parameters than the number of available pairwise comparisons
 \citep{Thurstone_1927b}. To address this issue and facilitate the
@@
 
-Among these, Case V has become the most widely used model in empirical
-CJ research, primarily due to the widespread adoption of the BTL model.
-The BTL model mirrors the core assumptions of Case V--namely, equal
+Among these, Case V remains the most widely used model in empirical CJ
+research, mainly due to the widespread adoption of the BTL model. The
+BTL model mirrors the core assumptions of Case V--namely, equal
 discriminal dispersions and zero correlation among stimuli' discriminal
@@
 mathematically tractable logistic distribution
-\citep{Andrich_1978, Bramley_2008}. Although the substitution has a
-minimal impact on trait estimation or model interpretation
-\citep{vanderLinden_et_al_2017_I, McElreath_2021} and simplifies the
-measurement model, the other two core assumptions of the model can still
-lead to unreliable and inaccurate trait estimates
-\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023},
-especially when researchers assess complex traits or heterogeneous
-stimuli
-\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
-
-Furthermore, because Thurstone primarily aimed to produce a trait's
-``coarse scaling'' and allocate the compared stimuli along this
-continuum \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
-how to use the trait estimates for statistical inference. The CJ
-tradition has attempted to fill this gap by separating trait estimation
-from hypothesis testing. Specifically, researchers often rely on point
-estimates of the traits, such as the BTL scores or their
-transformations, for inference. Although this approach simplifies
-analysis, the statistical literature cautions that it can introduce bias
-and compromise the reliability of the resulting inferences
+\citep{Andrich_1978, Bramley_2008}. Notably, although these assumptions
+simplify the trait measurement model, they may fail to capture the
+complexity of some traits or account for heterogeneous stimuli
+\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022},
+potentially leading to unreliable and inaccurate trait estimates
+\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.
+
+Moreover, because Thurstone's original goal was to produce a ``coarse
+scaling'' of traits and allocate stimuli along this continuum
+\citep[pp.~269]{Thurstone_1927a}, his theory offered no guidance on how
+to use trait estimates for statistical inference. To address this gap,
+researchers in the CJ tradition have often separated trait estimation
+from hypothesis testing, relying on point estimates of the traits--such
+as BTL scores or their transformations--for inference. Notably, while
+this approach simplifies analysis, it can also introduce bias and
+compromise the reliability of the resulting inferences
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
@@
 The study then translates the SCM into a bespoke statistical model that
-addresses key limitations of Case V. This model allows researchers to
-analyze CJ data when the assumptions of equal dispersion and zero
-correlation do not hold (see Section~\ref{sec-theory-issue1a} and
-Section~\ref{sec-theory-issue1b}), and when they need to perform
-statistical inference (refer to Section~\ref{sec-theory-issue2}). In
-particular, the model accounts for judge biases (see
-Section~\ref{sec-theory-theoretical_P1}), captures the hierarchical
-structure of stimuli and incorporates measurement error in the
-hypothesis testing process (refer to
+allows researchers to analyze CJ data when violations to the assumptions
+of equal dispersion and zero correlation occur (see
+Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b})
+and when statistical inference is necessary (refer to
+Section~\ref{sec-theory-issue2}). In particular, this model accounts for
+judge biases (see Section~\ref{sec-theory-theoretical_P1}), captures the
+hierarchical structure of stimuli, incorporates measurement error into
+the hypothesis testing process (refer to
 Section~\ref{sec-theory-theoretical_P3}), and accommodates heterogeneity
-in discriminal dispersions (see Section~\ref{sec-statistical}). These
-methodological innovations can potentially enhance the reliability and
-validity of trait measurement \citep{Perron_et_al_2015} and improve the
-accuracy of statistical inferences.
+in discriminal dispersions (see Section~\ref{sec-statistical}). By
+addressing all these issues, these methodological innovations have the
+potential to enhance the reliability and validity of trait measurement
+\citep{Perron_et_al_2015} and improve the accuracy of statistical
+inferences.
 
@@
 
-Five research avenues--current and new--deserve particular attention.
-The following sections outline these avenues and explain how our model
-provides a rigorous research framework to investigate them. Notably,
-while addressing these questions through formal derivations--following
-the tradition of classical statistical models like linear
-regression--may seem a natural step, we argue that the complexity of the
-CJ system makes this approach initially impractical. Instead,
-simulation-based methods, such as power analysis, offer a more feasible
-and flexible alternative, enabling researchers to investigate these
-questions without relying on cumbersome mathematical proofs.
-Nonetheless, developing formal mathematical proofs remains an essential
-goal for future work.
+Five research avenues, both current and new, deserve particular
+attention. The following sections outline these avenues and explain how
+our model provides a rigorous research framework to investigate them.
+Notably, although it may seem like a natural next step to address these
+questions through formal derivations, following the tradition of
+classical statistical models like linear regression, we contend that the
+complexity of the CJ system makes this approach impractical from the
+outset. Instead, simulation-based methods, such as power analysis, offer
+a more feasible and flexible alternative, enabling researchers to
+investigate these questions without depending on cumbersome mathematical
+proofs. Nonetheless, developing formal mathematical proofs remains an
+essential goal for future work.
 
@@
 
-Sampling and comparison mechanisms play a central role in modern CJ
-experimental design. However, most of the CJ literature has examined
-these mechanisms within a limited scope. Researchers have primarily
+Although sampling and comparison mechanisms are central to modern CJ
+experimental design, it is striking that most of the CJ literature has
+examined them only within a limited scope. Researchers have primarily
 investigated the effects of adaptive comparative judgment (ACJ) designs
@@
 should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}.
-Although these studies offer valuable insights, we argue that they may
-have overlooked the broader role these mechanisms play within the CJ
-system. Moreover, we suggest that this oversight largely stems from a
-lack of conceptual clarity about how the mechanisms function within the
-system. To address this gap, this study decided to integrate these
-mechanisms into CJ's conceptual model.
+Although these studies offer valuable insights, we argue that they have
+overlooked the broader role these mechanisms play within the CJ system.
+Moreover, we suggest that this oversight largely stems from a lack of
+conceptual clarity about how the mechanisms function within the system.
+To address this gap, this study decided to integrate these mechanisms
+into CJ's conceptual model.
 
@@
 understanding of the mixed evidence surrounding ACJ's capabilities.
-While some studies report improvements in trait reliability
+While some studies report that the method improves trait reliability
 \citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
-that these gains are artificially inflated by the method
+that it artificially inflates these gains
 \citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
@@
 CJ experimental design would benefit from explicitly defining its
-assumptions--an approach supported by our framework. This clarity
+assumptions, an approach supported by our framework. This clarity
 enables researchers to evaluate, for instance, how sampling and
@@
 Despite the growing notion that various stimulus-related factors can
-influence judges' perceptions during pairwise comparisons--and that
-these influences may not always cancel each other out--few studies in
-the CJ literature provide empirical evidence for judges' biases
+influence judges' perceptions \citep{Morin_et_al_2018} and that these
+influences may not always cancel each other out, few studies in the CJ
+literature provide empirical evidence for judges' biases
 \citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
@@
 provides a more accurate representation of the data-generating process
-behind pairwise comparisons--one that acknowledges that the discriminal
+behind pairwise comparisons, one that acknowledges that the discriminal
 processes of stimuli become an observable outcome only through judges'
@@
 
-Moreover, if we accept that judges may value different aspects of a
-stimulus, it follows that their judgments might also vary with
+Moreover, if we accept that judges may prioritize different aspects of a
+stimulus, it follows that their assessments might also vary with
 individual characteristics such as gender, age, culture, income,
@@
 \citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
-Building on this and the discussion in Section~\ref{sec-discussion_RA1},
+Thus, building on the discussion in Section~\ref{sec-discussion_RA1},
 researchers could investigate how the selection of judges influences the
@@
 observed trait distribution \citep{Deffner_et_al_2022}. Furthermore, if
-such characteristics indeed undermine the ``sample-freeness''
-assumption, then it becomes essential to explore strategies for
-controlling their influence. This includes investigating how many
-judges--and how many judgments per judge--are required to produce
-reliable trait estimates, as well as whether \emph{repeated measures
-designs}, in which judges evaluate the same stimulus pairs multiple
-times \citep{Lawson_2015}, can improve judgment consistency and
-accuracy.
+such characteristics indeed undermine the assumption of
+``sample-freeness,'' then it becomes essential to explore strategies to
+control their influence and to determine how many judges--and how many
+judgments per judge--are needed to produce reliable trait estimates in
+these contexts. Additionally, it is worth considering whether
+\emph{repeated measures designs}, in which judges evaluate the same
+stimulus pairs multiple times \citep{Lawson_2015}, can improve judgment
+consistency and accuracy.
 

### 2025-06-02 paper2.qmd
section:6; process:write

@@
 <!-- 5. So, what are we doing? -->
-To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines Causal and Bayesian inference methods.
+To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods.
 In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages.
@@
 <!-- 3. Example of a CJ estimand and estimator, and its limitations -->
-A motivating example that will appear in the rest of the document clarifies these concepts.
-In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+A motivating example that will appear throughout this document clarifies these concepts.
+In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
 Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait).
 Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-Following current CJ practices, researchers rely on estimates from the BTL model, or its transformations, to approximate this estimand.
-
-<!-- 4. But why identification is so important? -->
-However, @sec-theory-issues presents compelling evidence that Thurstone's Case V, and by extension the BTL model, suffers from several statistical and measurement limitations.
-These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the example.
-Identification is crucial because it is a necessary condition for ensuring consistent estimators.
-*Consistency* refers to the property of an estimator whose estimates converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010].
-Without identification, consistency cannot be achieved, even with "infinite" and error-free data.
-Thus, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023].
-
-<!-- Consistency: An estimator is said to be consistent if it converges to its estimand as the sample size tends to infinity. [@Everitt_et_al_2010, pp. 100] -->
-
-<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
-So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
-5].
--->
+
+Following current CJ practices, researchers rely on estimates from the BTL model--or its transformations--to approximate this estimand.
+However, as discussed in @sec-theory-issues, Thurstone's Case V and the BTL model exhibit several statistical and measurement limitations.
+These limitations hinder the model's ability to identify various estimands relevant to CJ inquiries, including the one described in the example.
 
@@
 <!-- 6. What the section will do? -->
-Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to Causal inference.
+Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to causal inference.
 Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
@@
 
+::: {#fig-cj03 layout-nrow=2}
+
+::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{iahbjk} & := f_{O}(D_{iahbjk}) \\
+  D_{iahbjk} & := f_{D}(T_{ia}, T_{hb}, B_{jk})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=35%}
+
+Conceptual-population model, scalar form.
+:::
+
 In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory).
@@
 <!-- Brennen generalizability theory, generalize conclusions over groups. To generalize over judges, you need to estimate them. Akin to Frisch-Wough-Lovel theorem for latent variables?. -->
-
-::: {#fig-cj03 layout-nrow=2}
-
-::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{cp}_{iahbjk} & := f_{O}(D_{iahbjk}) \\
-  D_{iahbjk} & := f_{D}(T_{ia}, T_{hb}, B_{jk})
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=35%}
-
-Conceptual-population model, scalar form.
-:::
 
@@
 Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system.
-Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
+Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as in Equation ([-@eq-mat11]).
+Here, each element of $I$ represents a unique individual $i$ or $h$, where $n_{I}$ denotes the total number of individuals.
+Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges.
+Moreover, each row of $IA$ represents a unique pairing of individuals $i, h$ with stimuli $a, b$.
+As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
+Likewise, each row of $JK$ associates a judge $j$ with a (repeated) judgment index $k$.
+Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
 $$
@@
 
-<!-- 2. what is the data structure? -->
-Here, each element of $I$ represents a unique individual $i$ or $h$, where $n_{I}$ denotes the total number of individuals.
-Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges.
-Moreover, each row of $IA$ represents a unique pairing of individuals $i, h$ with stimuli $a, b$.
-As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual.
-Likewise, each row of $JK$ associates a judge $j$ with a (repeated) judgment index $k$.
-Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
+Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
+This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
+Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
+Thus, we define the matrix as in Equation ([-@eq-mat12]).
 $$
@@
 
-Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
-This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
-Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
-Thus, we define the matrix as in Equation ([-@eq-mat12]).
-
 It is easier to visualize the structure of the previously defined vectors and matrices by considering an example.
 Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the form as in Equation ([-@eq-mat13]).
+
 $$
@@
 
+::: {#fig-cj04 layout-nrow=2}
+
+::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
+
+Conceptual-population model, initial vectorized form.
+:::
+
 <!-- 3. how de we use that data structure? -->
@@
 Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$, respectively.
-
-::: {#fig-cj04 layout-nrow=2}
-
-::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
-
-Conceptual-population model, initial vectorized form.
-:::
 
@@
 [^3]: refer to footnote 2.
-
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $Z_{J}$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $B_{J}$.
-The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
-Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $Z_{JK}$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
-The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
-Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
-
-[^4]: refer to footnote 2.
-[^5]: refer to footnote 2.
-
-Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
-Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$ and the absence of connecting arrows.
-
-<!-- 3. segway for next section -->
-Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
-These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
-Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
 
@@
 :::
+
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $Z_{J}$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $B_{J}$.
+The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
+Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
+Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $Z_{JK}$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
+Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
+
+[^4]: refer to footnote 2.
+[^5]: refer to footnote 2.
+
+Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
+Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$ and the absence of connecting arrows.
+
+<!-- 3. segway for next section -->
+Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
+These enhancements include accounting for judges' biases and integrating hierarchical structural components.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
+Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
 
@@
 Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the resulting vectors would have the following structure:
+
 $$
@@
 Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
+
+$$
+s_{(i,a,h,b,j,k)} = \begin{cases} 
+1 & \text{if data elements } i,a,h,b,j,k \text{ from } R \text{ are sampled} \\
+0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
+\end{cases}
+$$ {#eq-mat26}
+
 $$
@@
 
-$$
-s_{(i,a,h,b,j,k)} = \begin{cases} 
-1 & \text{if data elements } i,a,h,b,j,k \text{ from } R \text{ are sampled} \\
-0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
-\end{cases}
-$$ {#eq-mat26}
-
 <!-- 2. Integrate the sample mechanism -->
@@
 <!-- This is a whole section [@Lawson_2015, pp. 16-112] -->
+
+<!-- 3. segway for next section -->
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
+
+<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
+Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
+However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
+More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
+[@Boonen_et_al_2020, p.
+5] -->
 
@@
 
-<!-- 3. segway for next section -->
-However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
-Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
-
-<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
-Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
-However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
-More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
-[@Boonen_et_al_2020, p.
-5] -->
-
 ### The comparison mechanism {#sec-theory-theoretical_SC2}
@@
 
+<!-- 3. preparing to transform the conceptual model into an statistical one -->
+Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
+This reformulation produces a model that applies directly to a sample of comparative data.
+In this version, the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$ are omitted, and $O_{R}$ inherits the structural equation $f_{O}$ that originally defined $O^{cp}_{R}$.
+Moreover, the definition of $O_{R}$ now reflects its direct dependence on the discriminal difference $D_{R}$ and the sample and comparison mechanisms, $S$ and $C$.
+
+<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
+For example, if I have the distribution of the population outcome, and it is a Bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
+
 ::: {#fig-cj15 layout-nrow=2}
@@
 
-<!-- 3. preparing to transform the conceptual model into an statistical one -->
-Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
-This reformulation produces a model that applies directly to a sample of comparative data.
-In this version, the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$ are omitted, and $O_{R}$ inherits the structural equation $f_{O}$ that originally defined $O^{cp}_{R}$.
-Moreover, the definition of $O_{R}$ now reflects its direct dependence on the discriminal difference $D_{R}$ and the sample and comparison mechanisms, $S$ and $C$.
-
-<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
-For example, if I have the distribution of the population outcome, and it is a Bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
-
 <!-- 4. segway for next section -->
@@
 <!-- 4. General summary of what we have done. -->
-Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods.
+Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods.
 The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG).
@@
 Together, these advantages establish a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.
+Five current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
+The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
 
@@
 
-Five research avenues, both current and new, deserve particular attention.
-The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
-Notably, although it may seem like a natural next step to address these questions through formal derivations, following the tradition of classical statistical models like linear regression, we contend that the complexity of the CJ system makes this approach impractical from the outset.
-Instead, simulation-based methods, such as power analysis, offer a more feasible and flexible alternative, enabling researchers to investigate these questions without depending on cumbersome mathematical proofs.
-Nonetheless, developing formal mathematical proofs remains an essential goal for future work.
-
-<!-- due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. -->
+<!-- These procedures also bear on a larger topic of discussion in social science.
+The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
+[@Wainer_et_al_1978, pp.
+332] -->
 
@@
 <!-- 1. Challenges for the framework: background knowledge and research context -->
-Despite its promise, this study presents several challenges that researchers must navigate --challenges common to many quantitative studies.
-These challenges fall into two main categories.
-The first concerns the background knowledge researchers need to apply the proposed framework effectively.
-The second involves the increased attention researchers must now devote to their specific CJ research context.
-
-<!-- 2. Challenge 1: background knowledge -->
-In terms of background knowledge, researchers face two challenges: mastering Causal inference and developing proficiency in Bayesian inference methods.
-One example illustrating the need to master Causal inference is our assumption that predictor variables are "relevant" to the research context--interpreting relevance as the inclusion of these variables in a "sufficient adjustment set" [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-However, we do not fully explore what this entails or its implications for model specification, estimation, and inference.
-As a result, researchers must engage deeply with these complex ideas--and with others, such as how SCMs encode functional and probabilistic information--to apply the framework rigorously.
-To assist this effort, we provide key references throughout the study to guide motivated readers toward a deeper understanding of these foundational concepts[^9].
+Despite its promise, this study presents several challenges that researchers must navigate--challenges common to many quantitative studies.
+They fall into two main categories.
+The first concerns the foundational knowledge researchers need to apply the proposed framework effectively.
+The second involves the increased attention researchers must now devote to their specific research context.
+
+### Foundational knowledge {#sec-discussion_challenges1}
+
+Researchers must develop foundational knowledge in two critical areas: causal and Bayesian inference methods.
+One example that illustrates the importance of mastering causal inference is our common assumption that predictor variables are "relevant" to the research context--interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference.
+As a result, to apply the framework rigorously, researchers must engage deeply with these and other complex ideas, such as how SCMs encode functional and probabilistic information.
+To assist this effort, we have provided key references throughout the study to guide motivated readers toward a deeper understanding of these foundational concepts[^9].
+
+[^9]: refer to footnote 1, 2, and the detailed online document referenced in the Declarations section
+
+Additionally, researchers must develop the expertise to translate context-specific conceptual models into bespoke statistical models using Bayesian inference methods.
+While Bayesian approaches introduce specific technical demands--such as familiarity with probabilistic programming languages (PPLs) and concepts like probability distributions and convergence--we argue they offer a more accessible path for many applied researchers.
+This is because Bayesian methods significantly reduce the need for specialized knowledge in areas like optimization theory and techniques, which are often necessary in frequentist approaches.
+
+### Attention to the research context {#sec-discussion_challenges2}
+
+When applying our framework to a specific CJ research context, researchers face two additional challenges.
+First, they must verify that the conceptual model provides a *faithful* [^10] representation of the CJ system under study.
+Second, they must evaluate whether the model's statistical translation can accurately estimate the intended estimands (i.e., parameters) from empirical data.
+Notably, addressing these challenges requires careful attention because causal inference relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
+Consequently, the framework does not deliver automatic answers; instead, it encourages researchers to formulate precise questions and explicitly state their assumptions to build a generalizable understanding of the phenomenon under study, in this case, the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022].
+This theory-driven foundation then supports the empirical testing of causal assumptions [@Deffner_et_al_2022].
+
+[^10]: Avoid confusing this term with the *faithfulness assumption* described in the causal inference literature. For further details, see footnote 1.
+
+<!-- Causal inference generally requires expert knowledge and untestable assumptions about the causal network linking treatment, outcome, and other variables. ...
+Because the scenarios we considered were so simple, there was really no need to make the causal network explicit.
+As we start to turn our attention towards more complex situations, however, it will become crucial to be explicit about what we know and what we assume about the variables relevant to our particular causal inference problem.
+[@Hernan_et_al_2025, pp.
+73] -->
+
+<!-- Causal DAGs are of no practical use unless we make an assumption linking the causal structure represented by the DAG to the data obtained in a study.
+This assumption, referred to as the causal Markov assumption, states that, conditional on its direct causes, a variable Vj is independent of any variable for which it is not a cause.
+That is, conditional on its parents, Vj is independent of its non-descendants; hence, a causal DAG is Markov with respect to the DAG G.
+[@Hernan_et_al_2025, pp.
+74] -->
+
+<!-- Causal diagrams are a simple way to encode our subject-matter knowledge, and our assumptions, about the qualitative causal structure of a problem.
+But, as described in the next sections, causal diagrams also encode information about potential associations between the variables in the causal network.
+It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool.
+[@Hernan_et_al_2025, pp.
+75] -->
 
@@
 2] -->
-
-<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
-[@Deffner_et_al_2022, pp.
-13] -->
-
-<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
-
-<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
-4-5] -->
-
-<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
-
-[^9]: refer to footnote 1, 2, and the detailed online document referenced in the Declarations section
-
-Bayesian inference presents a second challenge.
-Researchers must acquire sufficient expertise to translate context-specific conceptual models into bespoke statistical models that meaningfully address their research questions.
-Although such translations are not unique to Bayesian approaches, we argue that Bayesian methods offer a more accessible route, as they reduce the need for specialized knowledge in optimization theory and techniques--often required by frequentist methods--thereby lowering the technical barriers for many applied researchers.
-
-<!-- 3. Challenge 2: research context -->
-In terms of the attention researchers must now devote to their specific CJ research context, researchers face two challenges: ensuring the conceptual model is a *faithful* representation for the CJ system, and verifying whether this model can accurately estimate the intended estimand from real data.
-
-<!-- See fine point 6.1 and 6.2 [@Hernan_et_al_2025, pp. 79-81] -->
-
-<!-- we are confident that these causal structure is faithful to a design -->
-
-<!-- Ideally, we would like to assess this property through an *identification analysis* [@Schuessler_et_al_2023], where researchers assess how accurate a model computes an estimand based solely on its (causal) assumptions, regardless of random variability.
-As one can imagine these types of analysis requires formal proofs, nevertheless, we argue that simulation-based methods can provide preliminary assessments akin to an identification analysis.
--->
-
-<!-- In this regard, we argue that researchers face the challenge that need a thorough knowledge about the CJ system, particularly regarding experimental design, the assumptions underlying the data analysis, and appropriate model specification.
-This is neccesary because the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions.
-For instance, the selection of relevant variables largely depends on the research context, and thus you really have to think through which variables you have to put in in your analysis.
--->
-
-<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
-Nevertheless, as @Kohler_et_al_2019 [pp.
-150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
--->
-
-<!-- Nevertheless, although in this paper we have outlined the causal process behind the CJ system, our framework is not designed to work every CJ design or data, is just a starting point for researchers to think thoroughly about the causal structure and statistical model that applies for their research context.
-That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.
--->
-
-<!-- Fifth and final, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate.
-In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest?
-However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces.
-Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.
--->
-
-<!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
-[@Deffner_et_al_2022, pp.
-4] -->
-
-<!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
-But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
-[@Deffner_et_al_2022, pp.
-14] -->
-
-<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
-[@Deffner_et_al_2022, pp.
-14] -->
 
@@
 
-<!-- DAGs are heuristic—they are not detailed statistical models.
-But they allow us to deduce which statistical models can provide valid causal inferences, assuming the DAG is true.
-[@McElreath_2020, pp.
-17] -->
-
-<!-- DAGs are not about adding assumptions—they are about revealing the assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
-
-<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
-
-<!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
-Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
-12].
--->
-
-<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
-[@Schuessler_et_al_2023, pp.
-3] -->
-
-<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
-But an incomplete model is still an improvement over no model at all.
+<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
 [@Deffner_et_al_2022, pp.
-5] -->
-
-<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
-<!-- 1.
-No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
--->
-<!-- 2.
-No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
--->
-<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
-Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
-[@Lash_et_al_2021, pp.
-113] -->
-
-<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
-This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
--->
-
-<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable.
-Nevertheless, as @Kohler_et_al_2019 [pp.
-150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data.
--->
+13] -->
+
+<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
+4-5] -->
+
+<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
+[@Deffner_et_al_2022, pp.
+14] -->
+
+Thus, regarding the *faithful* representation of the CJ system, we advise researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets.
+This consideration means that researchers must face the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification.
+This careful adaptation is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019].
+Nevertheless, as @Kohler_et_al_2019 points out, drawing conclusions about the real world from observed data always requires assumptions, regardless of whether the data are observational or experimental.
+In this context, our approach provides a transparent framework for researchers to clearly articulate these assumptions and design CJ experiments that support reliable and valid trait measurement.
 
@@
 
+<!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
+But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
+[@Deffner_et_al_2022, pp.
+14] -->
+
+<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
+
+<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
+
+<!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
+[@Deffner_et_al_2022, pp.
+4] -->
+
+<!-- DAGs are heuristic—they are not detailed statistical models.
+But they allow us to deduce which statistical models can provide valid causal inferences, assuming the DAG is true.
+[@McElreath_2020, pp.
+17] -->
+
+<!-- DAGs are not about adding assumptions—they are about revealing the assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
+
+<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
+
+<!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
+Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
+12].
+-->
+
+<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
+[@Schuessler_et_al_2023, pp.
+3] -->
+
+<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
+But an incomplete model is still an improvement over no model at all.
+[@Deffner_et_al_2022, pp.
+5] -->
+
+<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
+<!-- 1.
+No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
+-->
+<!-- 2.
+No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
+-->
+<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
+Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
+[@Lash_et_al_2021, pp.
+113] -->
+
+<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
+This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
+-->
+
 <!-- These procedures also bear on a larger topic of discussion in social science.
@@
 332] -->
+
+Finally, regarding the estimation capabilities of the statistical model, researchers must tackle the challenge of performing identification analysis.
+As outlined in @sec-theoretical, *identification analysis* determines whether a statistical model can accurately compute a given estimand (e.g., a parameter) based solely on its (causal) assumptions, independent of random variability [@Schuessler_et_al_2023].
+Identification is crucial because it is a necessary condition for consistency.
+*Consistency* is the property of an estimator (or model) whose estimates converge to the "true" value of an estimand as data size approaches infinity [@Everitt_et_al_2010].
+Logically, if researchers fail to identify a model, it cannot achieve consistency--even with infinite, error-free data.
+As a result, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023].
+
+Although performing identification analysis through formal derivations may appear to be a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset.
+Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs.
+Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research.
+
+<!-- Consistency: An estimator is said to be consistent if it converges to its estimand as the sample size tends to infinity. [@Everitt_et_al_2010, pp. 100] -->
+
+<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+5].
+-->
 
@@
 <!-- 1. How the SCMS and DAGs encode probabilistic information? -->
-Beyond their graphical capabilities, SCMs and DAGs can encode the probabilistic information embedded within a causal structure.
-They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges assumption.
-The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020, pp.
-20].
-Meanwhile, the *minimality assumption* encodes probabilistic dependencies among variables by stating that every pair of adjacent nodes exhibits a dependency [@Neal_2020, pp.
-21].
-Finally, the *causal edges assumption* encodes causal relationships between variables by declaring that each parent node acts as a direct cause of its children [@Neal_2020, pp.
-22].
-@fig-ACflow illustrates how these assumptions influence the statistical and causal interpretations of graphs.
-
-![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%}
+Beyond their graphical capabilities, researchers can translate SCMs and DAGs into statistical models because these conceptual models encode functional and probabilistic information, which they can then replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
+Notably, this encoding relies on two fundamental assumptions: the causal Markov and the faithfulness assumption.
+The *causal Markov assumption* links the graph structure to the joint probability distribution of variables by asserting that any variable in a graph is independent of all its non-descendants conditional on its parents [@Neal_2020, pp.
+20; @Hernan_et_al_2025, pp.
+80].
+The *faithfulness assumption*, in turn, states that the observed statistical (in)dependencies in the data reflect the underlying causal structure among variables--except in rare degenerate cases [@Neal_2020, pp.
+100; @Hernan_et_al_2025, pp.
+81].
+
+<!-- Causal DAGs are of no practical use unless we make an assumption linking the causal structure represented by the DAG to the data obtained in a study.
+This assumption, referred to as the causal Markov assumption, states that, conditional on its direct causes, a variable Vj is independent of any variable for which it is not a cause.
+That is, conditional on its parents, Vj is independent of its non-descendants; hence, a causal DAG is Markov with respect to the DAG G.
+[@Hernan_et_al_2025, pp.
+74] -->
+
+<!-- causal diagrams also encode information about potential associations between the variables in the causal network.
+It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool.
+[@Hernan_et_al_2025, pp.
+75] -->
+
+<!-- This and the previous section review three structural reasons why two variables may be associated: one causes the other, they share common causes, or they share a common effect and the analysis is restricted to certain level of that common effect (or of its descendants).
+Along the way we introduced a number of graphical rules that can be applied to any causal diagram to determine whether two variables are (conditionally) independent.
+The arguments we used to support these graphical rules were heuristic and relied on our causal intuitions.
+These arguments, however, have been formalized and mathematically proven.
+See Fine Point 6.1 for a systematic summary of the graphical rules, and Fine Point 6.2 for an introduction to the concept of faithfulness.
+[@Hernan_et_al_2025, pp.
+79] -->
+
+<!-- (from probability to graph): The relationship between statistical independence and the purely graphical concept of d-separation relies on the causal Markov assumption (Technical Point 6.1): In a causal DAG, any variable is independent of its non-descendants conditional on its parents.
+Pearl (1988) proved the following fundamental theorem: The causal Markov assumption implies that, given any three disjoint sets A, B, C of variables, if A is d-separated from B conditional on C, then A is statistically independent of B given C.
+The assumption that the converse holds, i.e., that A is d-separated from B conditional on C if A is statistically independent of B given C, is a separate assumption—the faithfulness assumption described in Fine Point 6.2.
+[@Hernan_et_al_2025, pp.
+80] -->
+
+<!-- (from graph to probability): Formally, faithfulness is the assumption that, for three disjoint sets A, B, C on a causal DAG (where C may be the empty set), A independent of B given C implies A is d-separated from B given C.
+When, as in our example, the causal diagram makes us expect a non-null association that does not actually exist in the data, we say that the joint distribution of the data is not faithful to the causal DAG.
+In our example the unfaithfulness was the result of effect modification (by sex) with opposite effects of exactly equal magnitude in each half of the population.
+Such perfect cancellation of effects is rare, and thus we will assume faithfulness throughout this book.
+Because unfaithful distributions are rare, in practice lack of d-separation (See Fine Point 6.1) can be almost always equated to non-zero association.
+There are, however, instances in which faithfulness is violated by design.
+For example, consider the prospective study in Section 4.5.
+The average causal effect of A on Y was computed after matching on L.
+In the matched population, L and A are not associated because the distribution of L is the same in the treated and the untreated.
+That is, individuals are selected into the matched population because they have a particular combination of values of L and A.
+The causal diagram in Figure 6.9 represents the setting of a matched study in which selection S (1: yes, 0: no) is determined by both A and L.
+The box around S indicates that the analysis is restricted to those selected into the matched cohort (S = 1).
+According to d-separation rules, there are two open paths between A and L when conditioning on S: L -> A and L -> S <- A.
+Thus one would expect L and A to be associated conditionally on S.
+However, matching ensures that L and A are not associated (see Chapter 4).
+Why the discrepancy?
+Matching creates an association via the path L -> S <- A that is of equal magnitude, but opposite direction, as the association via the path L -> A.
+The net result is a perfect cancellation of the associations.
+Matching leads to unfaithfulness.
+[@Hernan_et_al_2025, pp.
+81] -->
+
+<!-- ![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%} -->
 

### 2025-06-02 _tex/paper2.tex
Built site for gh-pages

@@
 To address these issues, this study extends Thurstone's general form
-through a systematic and integrated approach that combines Causal and
+through a systematic and integrated approach that combines causal and
 Bayesian inference methods. In addition to potentially enhancing
@@
 
-A motivating example that will appear in the rest of the document
-clarifies these concepts. In this example, researchers aim to determine:
+A motivating example that will appear throughout this document clarifies
+these concepts. In this example, researchers aim to answer the question:
 ``To what extent do different teaching methods influence students'
@@
 difference in the ability to produce high-quality written texts between
-the two groups of students?''. Following current CJ practices,
-researchers rely on estimates from the BTL model, or its
-transformations, to approximate this estimand.
-
-However, Section~\ref{sec-theory-issues} presents compelling evidence
-that Thurstone's Case V, and by extension the BTL model, suffers from
-several statistical and measurement limitations. These limitations
-hinder the model's ability to identify various estimands relevant to CJ
-inquiries, including the one described in the example. Identification is
-crucial because it is a necessary condition for ensuring consistent
-estimators. \emph{Consistency} refers to the property of an estimator
-whose estimates converge to the ``true'' value of the estimand as the
-data size approaches infinity \citep{Everitt_et_al_2010}. Without
-identification, consistency cannot be achieved, even with ``infinite''
-and error-free data. Thus, deriving meaningful insights from finite data
-becomes impossible \citep{Schuessler_et_al_2023}.
+the two groups of students?''.
+
+Following current CJ practices, researchers rely on estimates from the
+BTL model--or its transformations--to approximate this estimand.
+However, as discussed in Section~\ref{sec-theory-issues}, Thurstone's
+Case V and the BTL model exhibit several statistical and measurement
+limitations. These limitations hinder the model's ability to identify
+various estimands relevant to CJ inquiries, including the one described
+in the example.
 
@@
 Section~\ref{sec-theory-issues} by extending Thurstone's general form
-using the structural approach to Causal inference. Specifically, it
+using the structural approach to causal inference. Specifically, it
 combines the core theoretical principles outlined in
@@
 
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{iahbjk} & := f_{O}(D_{iahbjk}) \\
+  D_{iahbjk} & := f_{D}(T_{ia}, T_{hb}, B_{jk})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj03_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
+
+}
+
+\subcaption{\label{fig-cj03_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj03}Conceptual-population model, scalar form.}
+
+\end{figure}%
+
 In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
@@
 
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{iahbjk} & := f_{O}(D_{iahbjk}) \\
-  D_{iahbjk} & := f_{D}(T_{ia}, T_{hb}, B_{jk})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj03_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
-
-}
-
-\subcaption{\label{fig-cj03_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj03}Conceptual-population model, scalar form.}
-
-\end{figure}%
-
 Notably, every SCM has an associated DAG
@@
 vectorized form, we first define the vectors \(I\) and \(J\), along with
-the matrices \(IA\) and \(JK\), as follows:
+the matrices \(IA\) and \(JK\), as in Equation (\ref{eq-mat11}). Here,
+each element of \(I\) represents a unique individual \(i\) or \(h\),
+where \(n_{I}\) denotes the total number of individuals. Similarly, each
+element of \(J\) corresponds to a unique judge \(j\), with \(n_{J}\)
+indicating the total number of judges. Moreover, each row of \(IA\)
+represents a unique pairing of individuals \(i, h\) with stimuli
+\(a, b\). As a result, the matrix \(IA\) contains \(n_{I} \cdot n_{A}\)
+rows and \(2\) columns, where \(n_{A}\) specifies the number of stimuli
+available per individual. Likewise, each row of \(JK\) associates a
+judge \(j\) with a (repeated) judgment index \(k\). Consequently, the
+matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\) columns, where
+\(n_{K}\) indicates the number of repeated judgments each judge makes.
 \begin{equation}\phantomsection\label{eq-mat11}{
@@
 
-Here, each element of \(I\) represents a unique individual \(i\) or
-\(h\), where \(n_{I}\) denotes the total number of individuals.
-Similarly, each element of \(J\) corresponds to a unique judge \(j\),
-with \(n_{J}\) indicating the total number of judges. Moreover, each row
-of \(IA\) represents a unique pairing of individuals \(i, h\) with
-stimuli \(a, b\). As a result, the matrix \(IA\) contains
-\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
-the number of stimuli available per individual. Likewise, each row of
-\(JK\) associates a judge \(j\) with a (repeated) judgment index \(k\).
-Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
-columns, where \(n_{K}\) indicates the number of repeated judgments each
-judge makes. \begin{equation}\phantomsection\label{eq-mat12}{
+Additionally, we construct the matrix \(R\) to map each row of the
+\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
+matrix has \(n\) rows and \(6\) columns, where
+\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
+term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
+coefficient, which quantifies the total number of unique comparisons
+possible between every pair of stimuli generated by each pair of
+individuals in the population. Thus, we define the matrix as in Equation
+(\ref{eq-mat12}). \begin{equation}\phantomsection\label{eq-mat12}{
 R = \begin{bmatrix}
@@
 
-Additionally, we construct the matrix \(R\) to map each row of the
-\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
-matrix has \(n\) rows and \(6\) columns, where
-\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
-term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
-coefficient, which quantifies the total number of unique comparisons
-possible between every pair of stimuli generated by each pair of
-individuals in the population. Thus, we define the matrix as in Equation
-(\ref{eq-mat12}).
-
 It is easier to visualize the structure of the previously defined
@@
 take the form as in Equation (\ref{eq-mat13}).
+
 \begin{equation}\phantomsection\label{eq-mat13}{
@@
 
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj04_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+
+}
+
+\subcaption{\label{fig-cj04_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
+form.}
+
+\end{figure}%
+
 Now, using Equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
@@
 \(n_{J} \cdot n_{K}\), respectively.
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj04_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
-
-}
-
-\subcaption{\label{fig-cj04_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
-form.}
-
-\end{figure}%
 
@@
 
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj09_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
+
+}
+
+\subcaption{\label{fig-cj09_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj09}Conceptual-population model, final vectorized
+form.}
+
+\end{figure}%
+
 Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
@@
 assumption in practice, they must consider a more realistic scenario.
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} 
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj09_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
-
-}
-
-\subcaption{\label{fig-cj09_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj09}Conceptual-population model, final vectorized
-form.}
-
-\end{figure}%
 
@@
 \(n_{K} = 3\), the resulting vectors would have the following structure:
+
 \begin{equation}\phantomsection\label{eq-mat24}{
@@
 each pair of sampled individuals.
+
+\begin{equation}\phantomsection\label{eq-mat26}{
+s_{(i,a,h,b,j,k)} = \begin{cases} 
+1 & \text{if data elements } i,a,h,b,j,k \text{ from } R \text{ are sampled} \\
+0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
+\end{cases}
+}\end{equation}
+
 \begin{equation}\phantomsection\label{eq-mat25}{
@@
 \end{bmatrix}
-}\end{equation}
-
-\begin{equation}\phantomsection\label{eq-mat26}{
-s_{(i,a,h,b,j,k)} = \begin{cases} 
-1 & \text{if data elements } i,a,h,b,j,k \text{ from } R \text{ are sampled} \\
-0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
-\end{cases}
 }\end{equation}
@@
 within their respective groups \citep{Lawson_2015}.
+
+However, due to concerns about the practical feasibility of the
+comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ experiments rarely
+implement an exhaustive pairings of sampled judges, stimuli, and
+individuals. Thus, a realistic scenario must account for the fact that
+judges typically compare only a subset of stimuli authored by a sample
+of individuals.
 
@@
 \end{figure}%
-
-However, due to concerns about the practical feasibility of the
-comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ experiments rarely
-implement an exhaustive pairings of sampled judges, stimuli, and
-individuals. Thus, a realistic scenario must account for the fact that
-judges typically compare only a subset of stimuli authored by a sample
-of individuals.
 
@@
 
+Finally, since it is standard to assume that the distribution of the
+conceptual-population outcome \(O^{cp}_{R}\) also holds for
+\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the sample-comparison
+model in Figure~\ref{fig-cj14} into the equivalent form shown in
+Figure~\ref{fig-cj15}. This reformulation produces a model that applies
+directly to a sample of comparative data. In this version, the
+unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\) are omitted, and
+\(O_{R}\) inherits the structural equation \(f_{O}\) that originally
+defined \(O^{cp}_{R}\). Moreover, the definition of \(O_{R}\) now
+reflects its direct dependence on the discriminal difference \(D_{R}\)
+and the sample and comparison mechanisms, \(S\) and \(C\).
+
 \begin{figure}[H]
@@
 \end{figure}%
-
-Finally, since it is standard to assume that the distribution of the
-conceptual-population outcome \(O^{cp}_{R}\) also holds for
-\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the sample-comparison
-model in Figure~\ref{fig-cj14} into the equivalent form shown in
-Figure~\ref{fig-cj15}. This reformulation produces a model that applies
-directly to a sample of comparative data. In this version, the
-unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\) are omitted, and
-\(O_{R}\) inherits the structural equation \(f_{O}\) that originally
-defined \(O^{cp}_{R}\). Moreover, the definition of \(O_{R}\) now
-reflects its direct dependence on the discriminal difference \(D_{R}\)
-and the sample and comparison mechanisms, \(S\) and \(C\).
 
@@
 model, this study extends Thurstone's general form using a systematic,
-integrated approach that combines Causal and Bayesian inference methods.
+integrated approach that combines causal and Bayesian inference methods.
 The approach begins by formulating a conceptual model represented by a
@@
 offering new insights into existing research and opening promising
-avenues for future inquiry.
-
-Five research avenues, both current and new, deserve particular
-attention. The following sections outline these avenues and explain how
-our model provides a rigorous research framework to investigate them.
-Notably, although it may seem like a natural next step to address these
-questions through formal derivations, following the tradition of
-classical statistical models like linear regression, we contend that the
-complexity of the CJ system makes this approach impractical from the
-outset. Instead, simulation-based methods, such as power analysis, offer
-a more feasible and flexible alternative, enabling researchers to
-investigate these questions without depending on cumbersome mathematical
-proofs. Nonetheless, developing formal mathematical proofs remains an
-essential goal for future work.
+avenues for future inquiry. Five current and new research avenues
+deserve particular attention due to their direct impact on the
+reliability and validity of trait estimates and the accuracy of
+statistical inferences. The following sections outline these avenues and
+explain how our model provides a rigorous research framework to
+investigate them.
 
@@
 Despite its promise, this study presents several challenges that
-researchers must navigate --challenges common to many quantitative
-studies. These challenges fall into two main categories. The first
-concerns the background knowledge researchers need to apply the proposed
-framework effectively. The second involves the increased attention
-researchers must now devote to their specific CJ research context.
-
-In terms of background knowledge, researchers face two challenges:
-mastering Causal inference and developing proficiency in Bayesian
-inference methods. One example illustrating the need to master Causal
-inference is our assumption that predictor variables are ``relevant'' to
-the research context--interpreting relevance as the inclusion of these
-variables in a ``sufficient adjustment set''
+researchers must navigate--challenges common to many quantitative
+studies. They fall into two main categories. The first concerns the
+foundational knowledge researchers need to apply the proposed framework
+effectively. The second involves the increased attention researchers
+must now devote to their specific research context.
+
+\subsubsection{Foundational knowledge}\label{sec-discussion_challenges1}
+
+Researchers must develop foundational knowledge in two critical areas:
+causal and Bayesian inference methods. One example that illustrates the
+importance of mastering causal inference is our common assumption that
+predictor variables are ``relevant'' to the research
+context--interpreting this relevance as their inclusion in a
+\emph{sufficient adjustment set}
 \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
-not fully explore what this entails or its implications for model
-specification, estimation, and inference. As a result, researchers must
-engage deeply with these complex ideas--and with others, such as how
-SCMs encode functional and probabilistic information--to apply the
-framework rigorously. To assist this effort, we provide key references
-throughout the study to guide motivated readers toward a deeper
-understanding of these foundational concepts\footnote{refer to footnote
-  1, 2, and the detailed online document referenced in the Declarations
-  section}.
-
-Bayesian inference presents a second challenge. Researchers must acquire
-sufficient expertise to translate context-specific conceptual models
-into bespoke statistical models that meaningfully address their research
-questions. Although such translations are not unique to Bayesian
-approaches, we argue that Bayesian methods offer a more accessible
-route, as they reduce the need for specialized knowledge in optimization
-theory and techniques--often required by frequentist methods--thereby
-lowering the technical barriers for many applied researchers.
-
-In terms of the attention researchers must now devote to their specific
-CJ research context, researchers face two challenges: ensuring the
-conceptual model is a \emph{faithful} representation for the CJ system,
-and verifying whether this model can accurately estimate the intended
-estimand from real data.
+not fully explore what this assumption entails or its implications for
+model specification, estimation, and inference. As a result, to apply
+the framework rigorously, researchers must engage deeply with these and
+other complex ideas, such as how SCMs encode functional and
+probabilistic information. To assist this effort, we have provided key
+references throughout the study to guide motivated readers toward a
+deeper understanding of these foundational concepts\footnote{refer to
+  footnote 1, 2, and the detailed online document referenced in the
+  Declarations section}.
+
+Additionally, researchers must develop the expertise to translate
+context-specific conceptual models into bespoke statistical models using
+Bayesian inference methods. While Bayesian approaches introduce specific
+technical demands--such as familiarity with probabilistic programming
+languages (PPLs) and concepts like probability distributions and
+convergence--we argue they offer a more accessible path for many applied
+researchers. This is because Bayesian methods significantly reduce the
+need for specialized knowledge in areas like optimization theory and
+techniques, which are often necessary in frequentist approaches.
+
+\subsubsection{Attention to the research
+context}\label{sec-discussion_challenges2}
+
+When applying our framework to a specific CJ research context,
+researchers face two additional challenges. First, they must verify that
+the conceptual model provides a \emph{faithful} \footnote{Avoid
+  confusing this term with the \emph{faithfulness assumption} described
+  in the causal inference literature. For further details, see footnote
+  1.} representation of the CJ system under study. Second, they must
+evaluate whether the model's statistical translation can accurately
+estimate the intended estimands (i.e., parameters) from empirical data.
+Notably, addressing these challenges requires careful attention because
+causal inference relies on expert knowledge and assumptions about the
+variables' causal structure that are often untestable at the outset
+\citep{Hernan_et_al_2025}. Consequently, the framework does not deliver
+automatic answers; instead, it encourages researchers to formulate
+precise questions and explicitly state their assumptions to build a
+generalizable understanding of the phenomenon under study, in this case,
+the CJ system \citep{Rohrer_et_al_2022, Deffner_et_al_2022}. This
+theory-driven foundation then supports the empirical testing of causal
+assumptions \citep{Deffner_et_al_2022}.
+
+Thus, regarding the \emph{faithful} representation of the CJ system, we
+advise researchers to treat our conceptual and statistical models as
+starting points rather than universal solutions for all CJ designs or
+datasets. This consideration means that researchers must face the
+challenge of adapting these models to their specific research contexts,
+paying close attention to causal assumptions, experimental design, and
+model specification. This careful adaptation is essential because a
+model's ability to produce accurate estimates and valid inferences
+depends heavily on how well the data and inferential goals align with
+its underlying assumptions \citep{Kohler_et_al_2019}. Nevertheless, as
+\citet{Kohler_et_al_2019} points out, drawing conclusions about the real
+world from observed data always requires assumptions, regardless of
+whether the data are observational or experimental. In this context, our
+approach provides a transparent framework for researchers to clearly
+articulate these assumptions and design CJ experiments that support
+reliable and valid trait measurement.
+
+Finally, regarding the estimation capabilities of the statistical model,
+researchers must tackle the challenge of performing identification
+analysis. As outlined in Section~\ref{sec-theoretical},
+\emph{identification analysis} determines whether a statistical model
+can accurately compute a given estimand (e.g., a parameter) based solely
+on its (causal) assumptions, independent of random variability
+\citep{Schuessler_et_al_2023}. Identification is crucial because it is a
+necessary condition for consistency. \emph{Consistency} is the property
+of an estimator (or model) whose estimates converge to the ``true''
+value of an estimand as data size approaches infinity
+\citep{Everitt_et_al_2010}. Logically, if researchers fail to identify a
+model, it cannot achieve consistency--even with infinite, error-free
+data. As a result, deriving meaningful insights from finite data becomes
+impossible \citep{Schuessler_et_al_2023}.
+
+Although performing identification analysis through formal derivations
+may appear to be a natural next step, we argue that the complexity of
+the CJ system renders such an approach impractical at the outset.
+Instead, simulation-based methods, such as power analysis, provide a
+more practical and flexible alternative, enabling researchers to examine
+the consistency of estimates without relying on cumbersome mathematical
+proofs. Nonetheless, the development of formal identification proofs
+remains a significant goal for future CJ research.
 
@@
 
-Beyond their graphical capabilities, SCMs and DAGs can encode the
-probabilistic information embedded within a causal structure. They
-achieve this encoding by relying on three fundamental assumptions: the
-local Markov, the minimality, the causal edges assumption. The
-\emph{local Markov assumption} encodes probabilistic independencies
-between variables by declaring that nodes in a graph are independent of
-all its non-descendants, given its parents \citep[pp.~20]{Neal_2020}.
-Meanwhile, the \emph{minimality assumption} encodes probabilistic
-dependencies among variables by stating that every pair of adjacent
-nodes exhibits a dependency \citep[pp.~21]{Neal_2020}. Finally, the
-\emph{causal edges assumption} encodes causal relationships between
-variables by declaring that each parent node acts as a direct cause of
-its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
-illustrates how these assumptions influence the statistical and causal
-interpretations of graphs.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{images/png/ACflow.png}
-
-}
-
-\caption{\label{fig-ACflow}The flow of association and causation in
-graphs. Extracted and slightly modified from \citet[pp.~31]{Neal_2020}}
-
-\end{figure}%
+Beyond their graphical capabilities, researchers can translate SCMs and
+DAGs into statistical models because these conceptual models encode
+functional and probabilistic information, which they can then replace
+with suitable functions and probabilistic assumptions
+\citep{Pearl_et_al_2016}. Notably, this encoding relies on two
+fundamental assumptions: the causal Markov and the faithfulness
+assumption. The \emph{causal Markov assumption} links the graph
+structure to the joint probability distribution of variables by
+asserting that any variable in a graph is independent of all its
+non-descendants conditional on its parents
+\citetext{\citealp[pp.~20]{Neal_2020}; \citealp[pp.~80]{Hernan_et_al_2025}}.
+The \emph{faithfulness assumption}, in turn, states that the observed
+statistical (in)dependencies in the data reflect the underlying causal
+structure among variables--except in rare degenerate cases
+\citetext{\citealp[pp.~100]{Neal_2020}; \citealp[pp.~81]{Hernan_et_al_2025}}.
 

### 2025-06-02 paper2.qmd
section:6; process:write

@@
 Together, these advantages establish a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.
-Five current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
-The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
 
@@
 
-## The impact of sampling and comparison mechanisms on CJ outcomes and inference robustness {#sec-discussion_RA1}
+## Future research {#sec-discussion_RA}
+
+Five current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
+The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
+
+### The impact of sampling and comparison mechanisms {#sec-discussion_RA1}
 <!-- recording: Sven and Tine 25.05.21; time: 00:03:30 - 00:24:00 -->
@@
 
-## The effects of judges' bias on the reliability of traits {#sec-discussion_RA2}
+### The effects of judges' bias on the reliability of traits {#sec-discussion_RA2}
 <!-- recording: Sven and Tine 25.05.21; time: 00:24:00 - 00:31:45 -->
@@
 
-## The identification of 'misfitting' judges and stimuli {#sec-discussion_RA3}
+### The identification of 'misfitting' judges and stimuli {#sec-discussion_RA3}
 <!-- recording: Sven and Tine 25.05.21; time: 00:31:45 - 00:32:20 -->
@@
 
-## The effects of hierarchies on experimental design and inference {#sec-discussion_RA4}
+### The effects of hierarchies on experimental design and inference {#sec-discussion_RA4}
 <!-- recording: Sven and Tine 25.05.21; time: 00:32:25 - 00:33:00 -->
@@
 
-## The effects of non-dichotomous outcomes and multidimensional traits on CJ experiments {#sec-discussion_RA5}
+### The effects of non-dichotomous outcomes and multidimensional traits on CJ experiments {#sec-discussion_RA5}
 

### 2025-06-02 _tex/paper2.tex
Built site for gh-pages

@@
 offering new insights into existing research and opening promising
-avenues for future inquiry. Five current and new research avenues
-deserve particular attention due to their direct impact on the
-reliability and validity of trait estimates and the accuracy of
-statistical inferences. The following sections outline these avenues and
-explain how our model provides a rigorous research framework to
-investigate them.
-
-\subsection{The impact of sampling and comparison mechanisms on CJ
-outcomes and inference robustness}\label{sec-discussion_RA1}
+avenues for future inquiry.
+
+\subsection{Future research}\label{sec-discussion_RA}
+
+Five current and new research avenues deserve particular attention due
+to their direct impact on the reliability and validity of trait
+estimates and the accuracy of statistical inferences. The following
+sections outline these avenues and explain how our model provides a
+rigorous research framework to investigate them.
+
+\subsubsection{The impact of sampling and comparison
+mechanisms}\label{sec-discussion_RA1}
 
@@
 
-\subsection{The effects of judges' bias on the reliability of
+\subsubsection{The effects of judges' bias on the reliability of
 traits}\label{sec-discussion_RA2}
@@
 
-\subsection{The identification of `misfitting' judges and
+\subsubsection{The identification of `misfitting' judges and
 stimuli}\label{sec-discussion_RA3}
@@
 
-\subsection{The effects of hierarchies on experimental design and
+\subsubsection{The effects of hierarchies on experimental design and
 inference}\label{sec-discussion_RA4}
@@
 
-\subsection{The effects of non-dichotomous outcomes and multidimensional
-traits on CJ experiments}\label{sec-discussion_RA5}
+\subsubsection{The effects of non-dichotomous outcomes and
+multidimensional traits on CJ experiments}\label{sec-discussion_RA5}
 

### 2025-06-03 paper2.qmd
section:6; process:write

@@
 *Identification analysis* determines whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023].
-Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010].
-*Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].
+Here, *estimands* represent the specific quantities researchers aim to determine (i.e., a parameter) [@Everitt_et_al_2010].
+*Estimators* denote the methods or functions that transform data into an estimate (e.g., a statistical model), while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].
 
@@
 
-Following current CJ practices, researchers rely on estimates from the BTL model--or its transformations--to approximate this estimand.
+Following standard CJ practices, researchers typically use estimates from the BTL model--or its transformations--to approximate this estimand.
 However, as discussed in @sec-theory-issues, Thurstone's Case V and the BTL model exhibit several statistical and measurement limitations.
@@
 <!-- 1. What are we going to do? -->
-Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023] and Item Response Theory (IRT) [@Fox_2010; @vanderLinden_et_al_2017_I], the conceptual-population model integrates two *hierarchical structural components* to examine how different teaching methods influence students' writing ability.
-Each structural component defines how observed or latent variables affect the primary latent variable of interest [@Everitt_et_al_2010].
-The model's hierarchical design allows researchers to formulate and test hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for a discussion of these considerations).
+Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023] and Item Response Theory (IRT) [@Fox_2010; @vanderLinden_et_al_2017_I], the conceptual-population model integrates two *hierarchical structural components* to examine how different *relevant* [^2] variables--whether observed or latent--affect the primary latent variable of interest [@Everitt_et_al_2010].
+This hierarchical design enables researchers to formulate and test hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for a discussion of these considerations).
+
+[^2]: *Relevant variables* are those that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
+Refer also to footnote 1.
 
@@
 <!-- 2 Explaining what we did -->
-The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^2] student-related variables $X_{I}$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $T_{I}$.
+The top branch of [DAG -@fig-cj09_dag] illustrates the first component, where *relevant* [^3] student-related variables $X_{I}$, such as teaching method, and students' idiosyncratic errors $e_{I}$ causally influence the latent variable representing students' writing-quality trait $T_{I}$.
 The error term $e_{I}$ captures variations in students' traits unexplained by $X_{I}$.
 Here, $X_{I}$ is an observed matrix with $n_{I}$ rows and $q_{I}$ independent columns (variables), and both $e_{I}$ and $T_{I}$ are latent vectors of length $n_{I}$.
-Additionally, this branch shows how $T_{I}$, along with *relevant* [^3] text-related variables $X_{IA}$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $T_{IA}$, the first primary latent variable of interest.
+Additionally, this branch shows how $T_{I}$, along with *relevant* [^4] text-related variables $X_{IA}$ (e.g., text length), and texts' idiosyncratic errors $e_{IA}$ causally influence the texts' written-quality trait $T_{IA}$, the first primary latent variable of interest.
 The error term $e_{IA}$ captures variations in the texts' traits that remain unexplained by $T_{I}$ or $X_{IA}$.
@@
 
-[^2]: *Relevant variables* are those that satisfy the *backdoor criterion* [@Neal_2020, pp 37], that is, they belong to a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-A *sufficient* set (potentially empty) blocks all non-causal paths between a predictor and an outcome without opening new ones [@Pearl_2009].
-Refer also to footnote 1.
-
 [^3]: refer to footnote 2.
+[^4]: refer to footnote 2.
 
@@
 
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^4] judge-related variables $Z_{J}$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $B_{J}$.
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^5] judge-related variables $Z_{J}$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $B_{J}$.
 The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
 Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Furthermore, the branch shows how $B_{J}$, along with *relevant* [^5] judgment-related variables $Z_{JK}$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+Furthermore, the branch shows how $B_{J}$, along with *relevant* [^6] judgment-related variables $Z_{JK}$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
@@
 
-[^4]: refer to footnote 2.
 [^5]: refer to footnote 2.
+[^6]: refer to footnote 2.
 
@@
 This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
-Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in Equation ([-@eq-mat51]).
+Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^7], as shown in Equation ([-@eq-mat51]).
 For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent, where $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
@@
 
-[^6]: This re-expression is possible because the *chain rule* of probability and the *Bayesian Network Factorization (BNF)* property. For further details, see @Pearl_et_al_2016 and @Neal_2020.
+[^7]: This re-expression is possible because the *chain rule* of probability and the *Bayesian Network Factorization (BNF)* property. For further details, see @Pearl_et_al_2016 and @Neal_2020.
 
@@
 Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the statistical model [-@fig-cj16_stat].
-We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes.
+We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^8], reflecting the binary nature of CJ outcomes.
 Furthermore, following the conventions of Generalized Linear Models (GLMs) [@Nelder_et_al_1983; @Nelder_et_al_1996; @Agresti_2015], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
 
-[^7]: The binomial distribution--including its special case, the Bernoulli distribution--represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
+[^8]: The binomial distribution--including its special case, the Bernoulli distribution--represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
 34].
@@
 As observed variables, they do not require distributional assumptions in the same way the idiosyncratic errors do.
-Furthermore, their independence follows from the underlying random selection procedures that govern the variables[^8].
-
-[^8]: Randomization ensures that data--and, by extension, an estimator--satisfies several key identification properties, such as common support, no interference, and consistency.
+Furthermore, their independence follows from the underlying random selection procedures that govern the variables[^9].
+
+[^9]: Randomization ensures that data--and, by extension, an estimator--satisfies several key identification properties, such as common support, no interference, and consistency.
 The most critical property, however, is the elimination of confounding.
@@
 First, it handles complex and overparameterized models, where the number of parameters exceeds the number of observations [@Baker_1998; @Kim_et_al_1999].
-This capability is essential for our implementation, as the model is indeed overparameterized.
+This feature is essential for our implementation, as the proposed model is indeed overparameterized.
 Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
-In our implementation, we use mainly use priors to define the error distribution, setting the scale of the latent variables [@Depaoli_2014].
+We primarily use this capability to define the error distribution and set the scale of latent variables [@Depaoli_2014].
 Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
@@
 
-## Future research {#sec-discussion_RA}
-
-Five current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
+## Further research {#sec-discussion_RA}
+
+Four current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
 The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
@@
 
+{{< pagebreak >}}
+
 ### The effects of hierarchies on experimental design and inference {#sec-discussion_RA4}
@@
 
-### The effects of non-dichotomous outcomes and multidimensional traits on CJ experiments {#sec-discussion_RA5}
-
-<!-- 6.5 About non-dichotomous outcomes and multidimensional traits -->
-
-[About non-dichotomous outcomes and multidimensional traits]{style="color:blue;"}
-
-<!-- Talk about Bayesian Workflow, and how it would fit the process to arrive to the proposed model. -->
-
-<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->
-
-<!-- @Bramley_2008 proposes ranking instead of comparisons -->
-
-## Challenges {#sec-discussion_challenges}
+{{< pagebreak >}}
+
+## Limitations and challenges {#sec-discussion_challenges}
 <!-- recording: Sven and Tine 25.05.21; time: 00:33:50 - 00:42:20 -->
 
-<!-- 1. Challenges for the framework: background knowledge and research context -->
-Despite its promise, this study presents several challenges that researchers must navigate--challenges common to many quantitative studies.
-They fall into two main categories.
-The first concerns the foundational knowledge researchers need to apply the proposed framework effectively.
-The second involves the increased attention researchers must now devote to their specific research context.
-
-### Foundational knowledge {#sec-discussion_challenges1}
-
-Researchers must develop foundational knowledge in two critical areas: causal and Bayesian inference methods.
-One example that illustrates the importance of mastering causal inference is our common assumption that predictor variables are "relevant" to the research context--interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference.
-As a result, to apply the framework rigorously, researchers must engage deeply with these and other complex ideas, such as how SCMs encode functional and probabilistic information.
-To assist this effort, we have provided key references throughout the study to guide motivated readers toward a deeper understanding of these foundational concepts[^9].
-
-[^9]: refer to footnote 1, 2, and the detailed online document referenced in the Declarations section
-
-Additionally, researchers must develop the expertise to translate context-specific conceptual models into bespoke statistical models using Bayesian inference methods.
-While Bayesian approaches introduce specific technical demands--such as familiarity with probabilistic programming languages (PPLs) and concepts like probability distributions and convergence--we argue they offer a more accessible path for many applied researchers.
-This is because Bayesian methods significantly reduce the need for specialized knowledge in areas like optimization theory and techniques, which are often necessary in frequentist approaches.
-
-### Attention to the research context {#sec-discussion_challenges2}
-
-When applying our framework to a specific CJ research context, researchers face two additional challenges.
-First, they must verify that the conceptual model provides a *faithful* [^10] representation of the CJ system under study.
-Second, they must evaluate whether the model's statistical translation can accurately estimate the intended estimands (i.e., parameters) from empirical data.
-Notably, addressing these challenges requires careful attention because causal inference relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
-Consequently, the framework does not deliver automatic answers; instead, it encourages researchers to formulate precise questions and explicitly state their assumptions to build a generalizable understanding of the phenomenon under study, in this case, the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022].
-This theory-driven foundation then supports the empirical testing of causal assumptions [@Deffner_et_al_2022].
-
-[^10]: Avoid confusing this term with the *faithfulness assumption* described in the causal inference literature. For further details, see footnote 1.
-
-<!-- Causal inference generally requires expert knowledge and untestable assumptions about the causal network linking treatment, outcome, and other variables. ...
-Because the scenarios we considered were so simple, there was really no need to make the causal network explicit.
-As we start to turn our attention towards more complex situations, however, it will become crucial to be explicit about what we know and what we assume about the variables relevant to our particular causal inference problem.
-[@Hernan_et_al_2025, pp.
-73] -->
-
-<!-- Causal DAGs are of no practical use unless we make an assumption linking the causal structure represented by the DAG to the data obtained in a study.
-This assumption, referred to as the causal Markov assumption, states that, conditional on its direct causes, a variable Vj is independent of any variable for which it is not a cause.
-That is, conditional on its parents, Vj is independent of its non-descendants; hence, a causal DAG is Markov with respect to the DAG G.
-[@Hernan_et_al_2025, pp.
-74] -->
-
-<!-- Causal diagrams are a simple way to encode our subject-matter knowledge, and our assumptions, about the qualitative causal structure of a problem.
-But, as described in the next sections, causal diagrams also encode information about potential associations between the variables in the causal network.
-It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool.
-[@Hernan_et_al_2025, pp.
-75] -->
-
-<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014).
-For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
-[@Deffner_et_al_2022, pp.
-2] -->
-
-<!-- This approach makes strong causal assumptions about the nature of confounding and our ability to measure shared history.
-However, strong assumptions are always necessary in observational settings.
-What is important is that the assumptions are transparent and logically connected to data analysis.
-[@Deffner_et_al_2022, pp.
-13] -->
-
-<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
-[@Deffner_et_al_2022, pp.
-13] -->
-
-<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
-4-5] -->
-
-<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
-[@Deffner_et_al_2022, pp.
-14] -->
-
-Thus, regarding the *faithful* representation of the CJ system, we advise researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets.
-This consideration means that researchers must face the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification.
-This careful adaptation is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019].
-Nevertheless, as @Kohler_et_al_2019 points out, drawing conclusions about the real world from observed data always requires assumptions, regardless of whether the data are observational or experimental.
-In this context, our approach provides a transparent framework for researchers to clearly articulate these assumptions and design CJ experiments that support reliable and valid trait measurement.
+Drawing conclusions about the real world from observed data always requires assumptions, regardless of whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022].
+This study is no exception to this principle.
+By adopting causal inference as the theoretical backbone of our proposed framework, we necessarily rely on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
+As such, our framework does not yield automatic answers.
+Instead, it encourages researchers to formulate precise questions and to explicitly articulate their assumptions to develop a generalizable understanding of the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024].
+This clarity is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019]--a property we cannot yet guarantee for our models.
+However, this theory-driven approach offers a key benefit: its lays the foundation for empirically testing the proposed causal assumptions [@Deffner_et_al_2022].
 
@@
 
+<!-- This approach makes strong causal assumptions about the nature of confounding and our ability to measure shared history.
+However, strong assumptions are always necessary in observational settings.
+What is important is that the assumptions are transparent and logically connected to data analysis.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+<!-- Causal inference generally requires expert knowledge and untestable assumptions about the causal network linking treatment, outcome, and other variables. ...
+Because the scenarios we considered were so simple, there was really no need to make the causal network explicit.
+As we start to turn our attention towards more complex situations, however, it will become crucial to be explicit about what we know and what we assume about the variables relevant to our particular causal inference problem.
+[@Hernan_et_al_2025, pp.
+73] -->
+
+<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014).
+For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
+[@Deffner_et_al_2022, pp.
+2] -->
+
+<!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
+[@Deffner_et_al_2022, pp.
+13] -->
+
+<!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
+[@Deffner_et_al_2022, pp.
+14] -->
+
 <!-- An explicit causal-inference framework makes it (at times painfully) transparent how strong the assumptions are that we need to arrive at substantive conclusions and how little we collectively know about many real-world phenomena.
@@
 
-<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
+<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
+4-5] -->
 
 <!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
+
+<!-- 1. Challenges for the framework: background knowledge and research context -->
+In light of these limitation, this study presents several challenges that researchers must navigate--challenges common to many quantitative studies.
+These challenges fall into two main categories.
+The first concerns the foundational knowledge researchers need to apply our proposed framework effectively.
+The second involves the increased attention researchers must now devote to their specific research context.
+
+### Foundational knowledge {#sec-discussion_challenges1}
+
+In terms of foundational knowledge, researchers must face two core challenges to use our framework effectively.
+First, they need a solid understanding of causal inference principles.
+Second, they must learn to translate the functional and probabilistic content of conceptual models into bespoke statistical models.
+One example that illustrates the importance of a sound knowledge of causal inference is the recurrent assumption that predictor variables are "relevant" to the research context—interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference.
+As a result, researchers must engage deeply with these and other complex ideas, such as how SCMs encode functional and probabilistic information.
+To assist this effort, we include key references throughout the study to guide motivated readers toward deeper understanding[^10].
+
+[^10]: refer to footnote 1, 2, and the detailed online document referenced in the Declarations section
+
+Moreover, developing the skills to translate functional and probabilistic information into bespoke statistical models has its challenges.
+Although we argue that Bayesian inference methods offer a more accessible path for many applied researchers to develop these skills--since they reduce the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--these methods still demand familiarity with technical concepts, such as probabilistic programming languages (PPLs), probability distributions, and convergence, that researchers must acquire.
+Nevertheless, to support researchers in building these skills, we include links to model code and alternative specifications in the **Declarations** section of this document.
+
+### Attention to the research context {#sec-discussion_challenges2}
+
+Regarding attention to the research context, researchers face two additional challenges.
+First, they must verify that the conceptual model provides a *faithful* [^11] representation of the CJ system under study.
+Second, they must assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data.
+To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets.
+As such, researchers must tackle the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification.
+Nonetheless, our approach offers a transparent framework that enables researchers to explicitly articulate any new assumptions and design CJ experiments that support reliable and valid trait measurement.
+
+[^11]: Avoid confusing this term with the *faithfulness assumption* described in the causal inference literature. For further details, see footnote 1.
+
+<!-- Statements about the real world from observed data require assumptions.
+This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
+The validity of our statements therefore critically hinges on the validity of those assumptions.
+If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
+[@Kohler_et_al_2019, pp.
+150] -->
+
+<!-- These procedures also bear on a larger topic of discussion in social science.
+The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
+[@Wainer_et_al_1978, pp.
+332] -->
+
+<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
+<!-- 1.
+No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
+-->
+<!-- 2.
+No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
+-->
+<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
+Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
+[@Lash_et_al_2021, pp.
+113] -->
 
@@
 
-<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
-[@Schuessler_et_al_2023, pp.
-3] -->
-
 <!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
@@
 
-<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
-<!-- 1.
-No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured.
--->
-<!-- 2.
-No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured.
--->
-<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both.
-Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
-[@Lash_et_al_2021, pp.
-113] -->
-
-<!-- Incorporating these variables into the conceptual-population model enables researchers to identify different estimands of interest and their causal effects, even with observational data.
-This identification works because conditioning on these variables ensures that the estimator meets several critical properties, including confounding elimination [@Morgan_et_al_2014].
--->
-
-<!-- These procedures also bear on a larger topic of discussion in social science.
-The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
-[@Wainer_et_al_1978, pp.
-332] -->
-
-Finally, regarding the estimation capabilities of the statistical model, researchers must tackle the challenge of performing identification analysis.
+Conversely, to evaluate the estimation capabilities of the statistical model, researchers must tackle the challenge of performing identification analysis.
 As outlined in @sec-theoretical, *identification analysis* determines whether a statistical model can accurately compute a given estimand (e.g., a parameter) based solely on its (causal) assumptions, independent of random variability [@Schuessler_et_al_2023].
 Identification is crucial because it is a necessary condition for consistency.
-*Consistency* is the property of an estimator (or model) whose estimates converge to the "true" value of an estimand as data size approaches infinity [@Everitt_et_al_2010].
-Logically, if researchers fail to identify a model, it cannot achieve consistency--even with infinite, error-free data.
-As a result, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023].
+*Consistency* is the property of an estimator (e.g., a statistical model) whose estimates converge to the "true" value of an estimand as data size approaches infinity [@Everitt_et_al_2010].
+Without identification, consistency is impossible--even with infinite, error-free data--and thus, meaningful inference from finite samples cannot be achieved [@Schuessler_et_al_2023].
 
@@
 Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research.
+Notably, our framework supports both approaches, providing a foundation for formal derivations as well as simulation-based methods.
 
@@
 -->
+
+<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
+
+<!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
+[@Schuessler_et_al_2023, pp.
+3] -->
 
@@
 -->
+
+<!-- Causal diagrams are a simple way to encode our subject-matter knowledge, and our assumptions, about the qualitative causal structure of a problem.
+But, as described in the next sections, causal diagrams also encode information about potential associations between the variables in the causal network.
+It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool.
+[@Hernan_et_al_2025, pp.
+75] -->
 

### 2025-06-03 _tex/paper2.tex
Built site for gh-pages

@@
 \citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
-specific quantities researchers aim to determine
+specific quantities researchers aim to determine (i.e., a parameter)
 \citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
-functions that transform data into an estimate, while \emph{estimates}
-are the numerical values approximating the estimand
-\citep{Neal_2020, Everitt_et_al_2010}.
+functions that transform data into an estimate (e.g., a statistical
+model), while \emph{estimates} are the numerical values approximating
+the estimand \citep{Neal_2020, Everitt_et_al_2010}.
 
@@
 
-Following current CJ practices, researchers rely on estimates from the
-BTL model--or its transformations--to approximate this estimand.
-However, as discussed in Section~\ref{sec-theory-issues}, Thurstone's
-Case V and the BTL model exhibit several statistical and measurement
-limitations. These limitations hinder the model's ability to identify
-various estimands relevant to CJ inquiries, including the one described
-in the example.
+Following standard CJ practices, researchers typically use estimates
+from the BTL model--or its transformations--to approximate this
+estimand. However, as discussed in Section~\ref{sec-theory-issues},
+Thurstone's Case V and the BTL model exhibit several statistical and
+measurement limitations. These limitations hinder the model's ability to
+identify various estimands relevant to CJ inquiries, including the one
+described in the example.
 
@@
 model integrates two \emph{hierarchical structural components} to
-examine how different teaching methods influence students' writing
-ability. Each structural component defines how observed or latent
-variables affect the primary latent variable of interest
-\citep{Everitt_et_al_2010}. The model's hierarchical design allows
-researchers to formulate and test hypotheses that account for both the
-nested structure of stimuli and the uncertainties inherent in trait
-estimation (see Section~\ref{sec-theory-issue1b} and
+examine how different \emph{relevant} \footnote{\emph{Relevant
+  variables} are those that satisfy the \emph{backdoor criterion}
+  \citep[pp 37]{Neal_2020}, that is, they belong to a \emph{sufficient
+  adjustment set}
+  \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
+  \emph{sufficient} set (potentially empty) blocks all non-causal paths
+  between a predictor and an outcome without opening new ones
+  \citep{Pearl_2009}. Refer also to footnote 1.} variables--whether
+observed or latent--affect the primary latent variable of interest
+\citep{Everitt_et_al_2010}. This hierarchical design enables researchers
+to formulate and test hypotheses that account for both the nested
+structure of stimuli and the uncertainties inherent in trait estimation
+(see Section~\ref{sec-theory-issue1b} and
 Section~\ref{sec-theory-issue2} for a discussion of these
@@
 The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
-component, where \emph{relevant} \footnote{\emph{Relevant variables} are
-  those that satisfy the \emph{backdoor criterion} \citep[pp
-  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
-  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
-  \emph{sufficient} set (potentially empty) blocks all non-causal paths
-  between a predictor and an outcome without opening new ones
-  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
-variables \(X_{I}\), such as teaching method, and students'
-idiosyncratic errors \(e_{I}\) causally influence the latent variable
-representing students' writing-quality trait \(T_{I}\). The error term
-\(e_{I}\) captures variations in students' traits unexplained by
-\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
-\(q_{I}\) independent columns (variables), and both \(e_{I}\) and
+component, where \emph{relevant} \footnote{refer to footnote 2.}
+student-related variables \(X_{I}\), such as teaching method, and
+students' idiosyncratic errors \(e_{I}\) causally influence the latent
+variable representing students' writing-quality trait \(T_{I}\). The
+error term \(e_{I}\) captures variations in students' traits unexplained
+by \(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows
+and \(q_{I}\) independent columns (variables), and both \(e_{I}\) and
 \(T_{I}\) are latent vectors of length \(n_{I}\). Additionally, this
@@
 overparameterized models, where the number of parameters exceeds the
-number of observations \citep{Baker_1998, Kim_et_al_1999}. This
-capability is essential for our implementation, as the model is indeed
+number of observations \citep{Baker_1998, Kim_et_al_1999}. This feature
+is essential for our implementation, as the proposed model is indeed
 overparameterized. Second, it incorporates prior information to
@@
 that often affect frequentist methods
-\citep{Martin_et_al_1975, Seaman_et_al_2011}. In our implementation, we
-use mainly use priors to define the error distribution, setting the
-scale of the latent variables \citep{Depaoli_2014}. Third, Bayesian
-inference supports robust inferences from small samples, where the
-asymptotic properties underlying frequentist methods are less reliable
+\citep{Martin_et_al_1975, Seaman_et_al_2011}. We primarily use this
+capability to define the error distribution and set the scale of latent
+variables \citep{Depaoli_2014}. Third, Bayesian inference supports
+robust inferences from small samples, where the asymptotic properties
+underlying frequentist methods are less reliable
 \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
@@
 
-\subsection{Future research}\label{sec-discussion_RA}
-
-Five current and new research avenues deserve particular attention due
+\subsection{Further research}\label{sec-discussion_RA}
+
+Four current and new research avenues deserve particular attention due
 to their direct impact on the reliability and validity of trait
@@
 
+\newpage{}
+
 \subsubsection{The effects of hierarchies on experimental design and
@@
 
-\subsubsection{The effects of non-dichotomous outcomes and
-multidimensional traits on CJ experiments}\label{sec-discussion_RA5}
-
-{About non-dichotomous outcomes and multidimensional traits}
-
-\subsection{Challenges}\label{sec-discussion_challenges}
-
-Despite its promise, this study presents several challenges that
-researchers must navigate--challenges common to many quantitative
-studies. They fall into two main categories. The first concerns the
-foundational knowledge researchers need to apply the proposed framework
-effectively. The second involves the increased attention researchers
-must now devote to their specific research context.
+\newpage{}
+
+\subsection{Limitations and challenges}\label{sec-discussion_challenges}
+
+Drawing conclusions about the real world from observed data always
+requires assumptions, regardless of whether the data are observational
+or experimental \citep{Kohler_et_al_2019, Deffner_et_al_2022}. This
+study is no exception to this principle. By adopting causal inference as
+the theoretical backbone of our proposed framework, we necessarily rely
+on expert knowledge and assumptions about the variables' causal
+structure that are often untestable at the outset
+\citep{Hernan_et_al_2025}. As such, our framework does not yield
+automatic answers. Instead, it encourages researchers to formulate
+precise questions and to explicitly articulate their assumptions to
+develop a generalizable understanding of the CJ system
+\citep{Rohrer_et_al_2022, Deffner_et_al_2022, Sterner_et_al_2024}. This
+clarity is essential because a model's ability to produce accurate
+estimates and valid inferences depends heavily on how well the data and
+inferential goals align with its underlying assumptions
+\citep{Kohler_et_al_2019}--a property we cannot yet guarantee for our
+models. However, this theory-driven approach offers a key benefit: its
+lays the foundation for empirically testing the proposed causal
+assumptions \citep{Deffner_et_al_2022}.
+
+In light of these limitation, this study presents several challenges
+that researchers must navigate--challenges common to many quantitative
+studies. These challenges fall into two main categories. The first
+concerns the foundational knowledge researchers need to apply our
+proposed framework effectively. The second involves the increased
+attention researchers must now devote to their specific research
+context.
 
@@
 
-Researchers must develop foundational knowledge in two critical areas:
-causal and Bayesian inference methods. One example that illustrates the
-importance of mastering causal inference is our common assumption that
-predictor variables are ``relevant'' to the research
-context--interpreting this relevance as their inclusion in a
+In terms of foundational knowledge, researchers must face two core
+challenges to use our framework effectively. First, they need a solid
+understanding of causal inference principles. Second, they must learn to
+translate the functional and probabilistic content of conceptual models
+into bespoke statistical models. One example that illustrates the
+importance of a sound knowledge of causal inference is the recurrent
+assumption that predictor variables are ``relevant'' to the research
+context---interpreting this relevance as their inclusion in a
 \emph{sufficient adjustment set}
@@
 not fully explore what this assumption entails or its implications for
-model specification, estimation, and inference. As a result, to apply
-the framework rigorously, researchers must engage deeply with these and
-other complex ideas, such as how SCMs encode functional and
-probabilistic information. To assist this effort, we have provided key
-references throughout the study to guide motivated readers toward a
-deeper understanding of these foundational concepts\footnote{refer to
-  footnote 1, 2, and the detailed online document referenced in the
-  Declarations section}.
-
-Additionally, researchers must develop the expertise to translate
-context-specific conceptual models into bespoke statistical models using
-Bayesian inference methods. While Bayesian approaches introduce specific
-technical demands--such as familiarity with probabilistic programming
-languages (PPLs) and concepts like probability distributions and
-convergence--we argue they offer a more accessible path for many applied
-researchers. This is because Bayesian methods significantly reduce the
-need for specialized knowledge in areas like optimization theory and
-techniques, which are often necessary in frequentist approaches.
+model specification, estimation, and inference. As a result, researchers
+must engage deeply with these and other complex ideas, such as how SCMs
+encode functional and probabilistic information. To assist this effort,
+we include key references throughout the study to guide motivated
+readers toward deeper understanding\footnote{refer to footnote 1, 2, and
+  the detailed online document referenced in the Declarations section}.
+
+Moreover, developing the skills to translate functional and
+probabilistic information into bespoke statistical models has its
+challenges. Although we argue that Bayesian inference methods offer a
+more accessible path for many applied researchers to develop these
+skills--since they reduce the need for specialized knowledge in areas
+like optimization theory, which frequentist methods often require--these
+methods still demand familiarity with technical concepts, such as
+probabilistic programming languages (PPLs), probability distributions,
+and convergence, that researchers must acquire. Nevertheless, to support
+researchers in building these skills, we include links to model code and
+alternative specifications in the \textbf{Declarations} section of this
+document.
 
@@
 
-When applying our framework to a specific CJ research context,
-researchers face two additional challenges. First, they must verify that
-the conceptual model provides a \emph{faithful} \footnote{Avoid
-  confusing this term with the \emph{faithfulness assumption} described
-  in the causal inference literature. For further details, see footnote
-  1.} representation of the CJ system under study. Second, they must
-evaluate whether the model's statistical translation can accurately
-estimate the intended estimands (i.e., parameters) from empirical data.
-Notably, addressing these challenges requires careful attention because
-causal inference relies on expert knowledge and assumptions about the
-variables' causal structure that are often untestable at the outset
-\citep{Hernan_et_al_2025}. Consequently, the framework does not deliver
-automatic answers; instead, it encourages researchers to formulate
-precise questions and explicitly state their assumptions to build a
-generalizable understanding of the phenomenon under study, in this case,
-the CJ system \citep{Rohrer_et_al_2022, Deffner_et_al_2022}. This
-theory-driven foundation then supports the empirical testing of causal
-assumptions \citep{Deffner_et_al_2022}.
-
-Thus, regarding the \emph{faithful} representation of the CJ system, we
-advise researchers to treat our conceptual and statistical models as
-starting points rather than universal solutions for all CJ designs or
-datasets. This consideration means that researchers must face the
-challenge of adapting these models to their specific research contexts,
-paying close attention to causal assumptions, experimental design, and
-model specification. This careful adaptation is essential because a
-model's ability to produce accurate estimates and valid inferences
-depends heavily on how well the data and inferential goals align with
-its underlying assumptions \citep{Kohler_et_al_2019}. Nevertheless, as
-\citet{Kohler_et_al_2019} points out, drawing conclusions about the real
-world from observed data always requires assumptions, regardless of
-whether the data are observational or experimental. In this context, our
-approach provides a transparent framework for researchers to clearly
-articulate these assumptions and design CJ experiments that support
-reliable and valid trait measurement.
-
-Finally, regarding the estimation capabilities of the statistical model,
-researchers must tackle the challenge of performing identification
-analysis. As outlined in Section~\ref{sec-theoretical},
+Regarding attention to the research context, researchers face two
+additional challenges. First, they must verify that the conceptual model
+provides a \emph{faithful} \footnote{Avoid confusing this term with the
+  \emph{faithfulness assumption} described in the causal inference
+  literature. For further details, see footnote 1.} representation of
+the CJ system under study. Second, they must assess whether the
+statistical translation of the model can accurately estimate the
+intended estimands (i.e., parameters) from empirical data. To ensure a
+faithful representation of the CJ system, we encourage researchers to
+treat our conceptual and statistical models as starting points rather
+than universal solutions for all CJ designs or datasets. As such,
+researchers must tackle the challenge of adapting these models to their
+specific research contexts, paying close attention to causal
+assumptions, experimental design, and model specification. Nonetheless,
+our approach offers a transparent framework that enables researchers to
+explicitly articulate any new assumptions and design CJ experiments that
+support reliable and valid trait measurement.
+
+Conversely, to evaluate the estimation capabilities of the statistical
+model, researchers must tackle the challenge of performing
+identification analysis. As outlined in Section~\ref{sec-theoretical},
 \emph{identification analysis} determines whether a statistical model
@@
 necessary condition for consistency. \emph{Consistency} is the property
-of an estimator (or model) whose estimates converge to the ``true''
-value of an estimand as data size approaches infinity
-\citep{Everitt_et_al_2010}. Logically, if researchers fail to identify a
-model, it cannot achieve consistency--even with infinite, error-free
-data. As a result, deriving meaningful insights from finite data becomes
-impossible \citep{Schuessler_et_al_2023}.
+of an estimator (e.g., a statistical model) whose estimates converge to
+the ``true'' value of an estimand as data size approaches infinity
+\citep{Everitt_et_al_2010}. Without identification, consistency is
+impossible--even with infinite, error-free data--and thus, meaningful
+inference from finite samples cannot be achieved
+\citep{Schuessler_et_al_2023}.
 
@@
 proofs. Nonetheless, the development of formal identification proofs
-remains a significant goal for future CJ research.
+remains a significant goal for future CJ research. Notably, our
+framework supports both approaches, providing a foundation for formal
+derivations as well as simulation-based methods.
 

### 2025-06-04 paper2.qmd
section:6; process:write

@@
 Instead, it encourages researchers to formulate precise questions and to explicitly articulate their assumptions to develop a generalizable understanding of the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024].
-This clarity is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019]--a property we cannot yet guarantee for our models.
-However, this theory-driven approach offers a key benefit: its lays the foundation for empirically testing the proposed causal assumptions [@Deffner_et_al_2022].
+This clarity is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019]--a condition we cannot yet guarantee for our proposed models.
+Nevertheless, a key strength of this theory-driven approach is that it lays the foundation for empirically testing the proposed causal assumptions [@Deffner_et_al_2022], as discussed in @sec-discussion_RA.
 
@@
 <!-- 1. Challenges for the framework: background knowledge and research context -->
-In light of these limitation, this study presents several challenges that researchers must navigate--challenges common to many quantitative studies.
-These challenges fall into two main categories.
+This theoretical commitment to causal inference also introduces several challenges that researchers must navigate--challenges common to many quantitative studies.
+These fall into two main categories.
 The first concerns the foundational knowledge researchers need to apply our proposed framework effectively.
@@
 
-In terms of foundational knowledge, researchers must face two core challenges to use our framework effectively.
+To effectively apply our framework, researchers must tackle two main challenges related to foundational knowledge.
 First, they need a solid understanding of causal inference principles.
 Second, they must learn to translate the functional and probabilistic content of conceptual models into bespoke statistical models.
-One example that illustrates the importance of a sound knowledge of causal inference is the recurrent assumption that predictor variables are "relevant" to the research context—interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+One example that illustrates the importance of a sound knowledge of causal inference is the recurrent assumption that predictor variables are "relevant" to the research context--interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
 However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference.
-As a result, researchers must engage deeply with these and other complex ideas, such as how SCMs encode functional and probabilistic information.
-To assist this effort, we include key references throughout the study to guide motivated readers toward deeper understanding[^10].
+Consequently, researchers must deeply engage with these and other complex ideas, including how SCMs encode functional and probabilistic information.
+To assist this effort, we include key references throughout the study to guide motivated readers toward a deeper understanding of these concepts [^10].
 
@@
 
-Moreover, developing the skills to translate functional and probabilistic information into bespoke statistical models has its challenges.
-Although we argue that Bayesian inference methods offer a more accessible path for many applied researchers to develop these skills--since they reduce the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--these methods still demand familiarity with technical concepts, such as probabilistic programming languages (PPLs), probability distributions, and convergence, that researchers must acquire.
-Nevertheless, to support researchers in building these skills, we include links to model code and alternative specifications in the **Declarations** section of this document.
+Developing the skills to translate conceptual models into bespoke statistical models also presents challenges.
+Although Bayesian inference methods offer a more accessible path for many applied researchers to develop these skills--by reducing the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--they still demand familiarity with other technical concepts such as probabilistic programming languages (PPLs), probability distributions, and convergence, which researchers may find hard to master.
+Thus, to support this skill development, we include links to the statistical model code and alternative specifications in the **Declarations** section of this document.
 
@@
 
-Regarding attention to the research context, researchers face two additional challenges.
+Researchers face two additional challenges when applying our framework to their specific research context.
 First, they must verify that the conceptual model provides a *faithful* [^11] representation of the CJ system under study.
-Second, they must assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data.
+Second, they need to assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data.
 To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets.
-As such, researchers must tackle the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification.
-Nonetheless, our approach offers a transparent framework that enables researchers to explicitly articulate any new assumptions and design CJ experiments that support reliable and valid trait measurement.
-
-[^11]: Avoid confusing this term with the *faithfulness assumption* described in the causal inference literature. For further details, see footnote 1.
-
-<!-- Statements about the real world from observed data require assumptions.
-This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
-The validity of our statements therefore critically hinges on the validity of those assumptions.
-If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
-[@Kohler_et_al_2019, pp.
-150] -->
+This consideration means researchers must tackle the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification.
+Our approach supports this process by offering a transparent framework that helps researchers explicitly articulate these new assumptions and supports the design of CJ experiments that can yield reliable and valid inferences.
+
+[^11]: Avoid confusing this term with the *faithfulness assumption* [@Neal_2020; @Hernan_et_al_2025] described in the causal inference literature. For further details, see footnote 1.
+
+<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
+But an incomplete model is still an improvement over no model at all.
+[@Deffner_et_al_2022, pp.
+5] -->
+
+<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
 
@@
 -->
-
-<!-- Our DAG is, of course, incomplete and possibly wrong, in particular when it comes to the nodes that have not been experimentally manipulated.
-But an incomplete model is still an improvement over no model at all.
-[@Deffner_et_al_2022, pp.
-5] -->
 
@@
 Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research.
-Notably, our framework supports both approaches, providing a foundation for formal derivations as well as simulation-based methods.
+Importantly, our framework supports both approaches, as it provides the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
 
@@
 -->
-
-<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
 

### 2025-06-04 _tex/paper2.tex
Built site for gh-pages

@@
 inferential goals align with its underlying assumptions
-\citep{Kohler_et_al_2019}--a property we cannot yet guarantee for our
-models. However, this theory-driven approach offers a key benefit: its
-lays the foundation for empirically testing the proposed causal
-assumptions \citep{Deffner_et_al_2022}.
-
-In light of these limitation, this study presents several challenges
-that researchers must navigate--challenges common to many quantitative
-studies. These challenges fall into two main categories. The first
+\citep{Kohler_et_al_2019}--a condition we cannot yet guarantee for our
+proposed models. Nevertheless, a key strength of this theory-driven
+approach is that it lays the foundation for empirically testing the
+proposed causal assumptions \citep{Deffner_et_al_2022}, as discussed in
+Section~\ref{sec-discussion_RA}.
+
+This theoretical commitment to causal inference also introduces several
+challenges that researchers must navigate--challenges common to many
+quantitative studies. These fall into two main categories. The first
 concerns the foundational knowledge researchers need to apply our
@@
 
-In terms of foundational knowledge, researchers must face two core
-challenges to use our framework effectively. First, they need a solid
+To effectively apply our framework, researchers must tackle two main
+challenges related to foundational knowledge. First, they need a solid
 understanding of causal inference principles. Second, they must learn to
@@
 assumption that predictor variables are ``relevant'' to the research
-context---interpreting this relevance as their inclusion in a
+context--interpreting this relevance as their inclusion in a
 \emph{sufficient adjustment set}
@@
 not fully explore what this assumption entails or its implications for
-model specification, estimation, and inference. As a result, researchers
-must engage deeply with these and other complex ideas, such as how SCMs
-encode functional and probabilistic information. To assist this effort,
-we include key references throughout the study to guide motivated
-readers toward deeper understanding\footnote{refer to footnote 1, 2, and
-  the detailed online document referenced in the Declarations section}.
-
-Moreover, developing the skills to translate functional and
-probabilistic information into bespoke statistical models has its
-challenges. Although we argue that Bayesian inference methods offer a
-more accessible path for many applied researchers to develop these
-skills--since they reduce the need for specialized knowledge in areas
-like optimization theory, which frequentist methods often require--these
-methods still demand familiarity with technical concepts, such as
-probabilistic programming languages (PPLs), probability distributions,
-and convergence, that researchers must acquire. Nevertheless, to support
-researchers in building these skills, we include links to model code and
-alternative specifications in the \textbf{Declarations} section of this
-document.
+model specification, estimation, and inference. Consequently,
+researchers must deeply engage with these and other complex ideas,
+including how SCMs encode functional and probabilistic information. To
+assist this effort, we include key references throughout the study to
+guide motivated readers toward a deeper understanding of these concepts
+\footnote{refer to footnote 1, 2, and the detailed online document
+  referenced in the Declarations section}.
+
+Developing the skills to translate conceptual models into bespoke
+statistical models also presents challenges. Although Bayesian inference
+methods offer a more accessible path for many applied researchers to
+develop these skills--by reducing the need for specialized knowledge in
+areas like optimization theory, which frequentist methods often
+require--they still demand familiarity with other technical concepts
+such as probabilistic programming languages (PPLs), probability
+distributions, and convergence, which researchers may find hard to
+master. Thus, to support this skill development, we include links to the
+statistical model code and alternative specifications in the
+\textbf{Declarations} section of this document.
 
@@
 
-Regarding attention to the research context, researchers face two
-additional challenges. First, they must verify that the conceptual model
-provides a \emph{faithful} \footnote{Avoid confusing this term with the
-  \emph{faithfulness assumption} described in the causal inference
+Researchers face two additional challenges when applying our framework
+to their specific research context. First, they must verify that the
+conceptual model provides a \emph{faithful} \footnote{Avoid confusing
+  this term with the \emph{faithfulness assumption}
+  \citep{Neal_2020, Hernan_et_al_2025} described in the causal inference
   literature. For further details, see footnote 1.} representation of
-the CJ system under study. Second, they must assess whether the
+the CJ system under study. Second, they need to assess whether the
 statistical translation of the model can accurately estimate the
@@
 treat our conceptual and statistical models as starting points rather
-than universal solutions for all CJ designs or datasets. As such,
-researchers must tackle the challenge of adapting these models to their
-specific research contexts, paying close attention to causal
-assumptions, experimental design, and model specification. Nonetheless,
-our approach offers a transparent framework that enables researchers to
-explicitly articulate any new assumptions and design CJ experiments that
-support reliable and valid trait measurement.
+than universal solutions for all CJ designs or datasets. This
+consideration means researchers must tackle the challenge of adapting
+these models to their specific research contexts, paying close attention
+to causal assumptions, experimental design, and model specification. Our
+approach supports this process by offering a transparent framework that
+helps researchers explicitly articulate these new assumptions and
+supports the design of CJ experiments that can yield reliable and valid
+inferences.
 
@@
 proofs. Nonetheless, the development of formal identification proofs
-remains a significant goal for future CJ research. Notably, our
-framework supports both approaches, providing a foundation for formal
-derivations as well as simulation-based methods.
+remains a significant goal for future CJ research. Importantly, our
+framework supports both approaches, as it provides the probabilistic
+foundation for formal derivations and the necessary statistical
+structure for simulation-based methods.
 

### 2025-06-04 paper2.qmd
section:6; process:write

@@
 
-## Further research {#sec-discussion_RA}
-
-Four current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
+## Future research {#sec-discussion_RA}
+
+Among the many potential directions for future research, four avenues, both current and new, deserve particular attention due to their direct impact on the reliability and validity of trait estimates, as well as on the accuracy of statistical inferences.
 The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.

### 2025-06-04 _tex/paper2.tex
Built site for gh-pages

@@
 
-\subsection{Further research}\label{sec-discussion_RA}
-
-Four current and new research avenues deserve particular attention due
-to their direct impact on the reliability and validity of trait
-estimates and the accuracy of statistical inferences. The following
-sections outline these avenues and explain how our model provides a
-rigorous research framework to investigate them.
+\subsection{Future research}\label{sec-discussion_RA}
+
+Among the many potential directions for future research, four avenues,
+both current and new, deserve particular attention due to their direct
+impact on the reliability and validity of trait estimates, as well as on
+the accuracy of statistical inferences. The following sections outline
+these avenues and explain how our model provides a rigorous research
+framework to investigate them.
 

### 2025-06-05 paper2.qmd
section:6; process:write

--- 2025-06-04 paper2.qmd
@@
   To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods.
-This extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+This extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
 It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.
@@
 In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages.
-First, it clarifies the interactions among all actors and processes involved in CJ experiments.
+First, it clarifies the interactions among all actors and processes involved in CJ assessments.
 Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.
@@
 @sec-theoretical extends Thurstone's general form to address these challenges.
-The extension integrates core theoretical principles alongside key CJ experimental design features, such as the selection of judges, stimuli, and comparisons.
+The extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
 @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data.
@@
 Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
-For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases).
+For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases).
 However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.
@@
 <!-- 3. But does it hold in practice? -->
-Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of some traits or account for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
-Indeed, the presence of the so-called *misfit* texts may already signal these limitations, as these are texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018], possibly due to larger discriminal dispersions or because they are genuine outliers--meaning, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, the BTL model neither accounts for these anomalies nor provides tools for addressing them, apart from excluding the "problematic" texts from analysis.
+Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail accommodate for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
+Indeed, the presence of the so-called *misfit texts*--texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]--may signal this limitation, as such discrepancies can arise from larger discriminal dispersions due to stimulus heterogeneity or because the texts are genuine outliers--that is, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+In either case, the BTL model's assumptions prevent it from adequately accounting for or addressing these anomalies, leaving exclusion of the "problematic" texts as the primary remedy.
 
@@
 However, as texts improve along the trait continuum, with certain aspects showing improvement and others reaming unaltered, these variations introduce uncertainty into the judgments.
-Now, since judges often rely on various intricate characteristics of the texts to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], these changes lead to greater discrepancies in their evaluations.
+Now, since judges often rely on various intricate characteristics of the texts to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], these changes lead to greater discrepancies in their evaluations.
 This divergence ultimately imply broader discriminal differences and, by extension, more varied discriminal processes for the texts.
@@
 <!-- 3. But does it hold?  -->
-Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, experimental designs often adopt the assumption of no correlation between stimuli based on an old theoretical justification.
+Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, assessment designs often adopt the assumption of no correlation between stimuli based on an old theoretical justification.
 Specifically, @Thurstone_1927b argued that stimuli could be treated as uncorrelated because judges' biases--arising from two opposing and equally weighted effects occurring during the pairwise comparisons--would cancel each other out.
@@
 <!-- 2. The first scenario where the assumption does not hold -->
-In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022].
-These characteristics, regardless of their relevance, are unlikely to be equally weighted or to oppose one another across comparisons.
+In the first scenario, research on text quality suggests that when judges evaluate complex, multidimensional traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022].
+Regardless of their relevance, these characteristics may not receive equal weight or consistently oppose one another across comparisons.
 As a result, they may exert a disproportionate influence on judges' perceptions, generating biases that persist rather than cancel out.
@@
 Moreover, because the discriminal process of stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
-Evidence from traditional marking contexts confirms a similar phenomenon [@Morin_et_al_2018], and additional studies [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a] support the presence of these biases in CJ--reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.
+While direct evidence for this exact scenario is limited, existing studies document the presence of judge bias in CJ [@Pollitt_et_al_2003, @vanDaal_et_al_2016, @Bartholomew_et_al_2020a], reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.
 
@@
 Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023] and even biases [@Wu_et_al_2022].
-These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli.
 This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$.
+Moreover, these inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli--potentially causing the model to incorrectly classify stimuli or judges as *misfits*.
 
@@
 
-Finally, as discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969; @Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates.
+Finally, similar to what we discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and may appear as outliers under the BTL model [@Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates.
 The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
-
-<!-- These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b. -->
 
@@
 
+<!-- Under the Bradley–Terry model, the object residual ri asymptotically follows a standard normal distribution, and is independent of \beta_i.
+Note that the asymptotic distribution of Rij holds when Vij grows large, so it is important to exclude residuals show Rij for which Vij = 0.
+If the absolute value of ri is very large, object i is regarded as an outlier.
+By definition, this happens if object i unexpectedly wins over a popular object j (i.e., \beta_j is positive) and Rij is large and positive, or if object i unexpectedly loses to an unpopular object j (\beta_j negative) and Rij is large and negative.
+Too many outlying objects may imply that the model suffers from a lack of fit.
+As Liner and Amin (2004) pointed out, this is more likely to happen if some of the objects are involved in only a few comparisons with most of the others.
+[@Wu_et_al_2022, pp.
+466] -->
+
 <!-- @Zimmerman_1994 the whole paper -->
@@
 The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data.
-The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments.
+The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ assessments.
 Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests.
@@
 A motivating example that will appear throughout this document clarifies these concepts.
-In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ assessment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
 Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait).
@@
 Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to causal inference.
-Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.
+Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
 In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
-First, it clarifies the interactions among all actors and processes involved in CJ experiments.
+First, it clarifies the interactions among all actors and processes involved in CJ assessments.
 Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.
@@
 The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli).
-The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for experimental conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, pp.
+The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for assessment conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, pp.
 366-376].
@@
 In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory).
-Additionally, the SCM incorporates a key design feature of CJ experiments: the judges' biases $(B_{kj})$.
+Additionally, the SCM incorporates a key CJ design feature: the judges' biases $(B_{kj})$.
 This extension builds on the arguments presented in @sec-theory-issue1b, contending that the discriminal difference becomes an observable outcome only through judges' perceptions.
@@
 <!-- 3. segway for next section -->
-However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ experiments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ assessments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
 Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
@@
 In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model.
-These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and clarify the role of the sample and comparison mechanisms in CJ experiments (see @sec-theory-theoretical_SC).
+These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and even clarify the role of the sample and comparison mechanisms in CJ assessments (see @sec-theory-theoretical_SC).
 However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a.
@@
 
-<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
-This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
-12].
--->
-
 <!-- when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification. -->
@@
 Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
-This feature is particularly relevant in CJ experiments, as researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.
+This feature is particularly relevant in CJ assessments, as researchers often collect large volumes of paired comparisons but work with relatively small samples of judges, stimuli, and individuals to test hypotheses.
 
@@
 <!-- recording: Sven and Tine 25.05.21; time: 00:32:55 - 00:33:50 -->
-
-<!-- Follow the structure of the paper: -->
-<!-- 1. Talk about Thurstone -->
-<!-- 2. Talk about the issues with the model of Thurstone -->
-<!-- 3. Talk about what you did -->
-<!-- 4. Potential benefits and extras that you can do now -->
-<!-- 5. Challenges -->
-<!-- 6. Conclusion: what we did and why is a great paper -->
+<!-- recording: Sven and Tine 25.06.04; time: 00:00:00 - 00:05:40 -->
 
@@
 The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008].
-Notably, although these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
+Although this substitution has minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], the simplifying assumptions of the BTL model--and by extension, of Case V--may fail to capture the complexity of some traits or account for heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
 
@@
 269], his theory offered no guidance on how to use trait estimates for statistical inference.
-To address this gap, researchers in the CJ tradition have often separated trait estimation from hypothesis testing, relying on point estimates of the traits--such as BTL scores or their transformations--for inference.
+The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing, relying on point estimates--such as BTL scores or their transformations--for inference.
 Notably, while this approach simplifies analysis, it can also introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
@@
 <!-- 4. General summary of what we have done. -->
-Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods.
+Thus, to address the limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods.
 The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG).
-This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.
-
-The study then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur (see @sec-theory-issue1a and @sec-theory-issue1b) and when statistical inference is necessary (refer to @sec-theory-issue2).
+This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.
+
+The approach then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur (see @sec-theory-issue1a and @sec-theory-issue1b) and when statistical inference is necessary (refer to @sec-theory-issue2).
 In particular, this model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical).
-By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.
+By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement in CJ [@Perron_et_al_2015], while also improving the accuracy of statistical inferences.
 
@@
 Beyond these potential benefits, the approach offers two additional advantages.
-First, it clarifies the roles and interactions of all actors and processes involved in CJ experiments.
+First, it clarifies the roles and interactions of all actors and processes involved in CJ assessments.
 Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data.
-Together, these advantages establish a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.
+Together, these advantages establish a principled framework for evaluating best practices in designing CJ assessments, one that better aligns with the demands of contemporary CJ contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.
 
@@
 
-## Further research {#sec-discussion_RA}
-
-Four current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences.
-The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.
+## Future research directions based on this study's approach {#sec-discussion_RA}
+<!-- recording: Sven and Tine 25.06.04; time: 00:05:40 - 00:07:25 -->
+
+Among the many potential directions for future research--both current and new--three avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates, as well as on the accuracy of statistical inferences.
+The following sections outline these avenues and explain how our approach provides a rigorous research framework to investigate them.
 
@@
 <!-- recording: Sven and Tine 25.05.21; time: 00:03:30 - 00:24:00 -->
+<!-- recording: Sven and Tine 25.06.04; time: 00:07:25 - 00:11:55 -->
 
 <!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
-Although sampling and comparison mechanisms are central to modern CJ experimental design, it is striking that most of the CJ literature has examined them only within a limited scope.
+Although sampling and comparison mechanisms are central to modern CJ assessments, it is striking that most of the CJ literature has examined them only within a limited scope.
 Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022].
-Although these studies offer valuable insights, we argue that they have overlooked the broader role these mechanisms play within the CJ system.
-Moreover, we suggest that this oversight largely stems from a lack of conceptual clarity about how the mechanisms function within the system.
-To address this gap, this study decided to integrate these mechanisms into CJ's conceptual model.
+While these studies offer valuable insights, they often overlook the broader role these mechanisms play within the CJ system.
+Moreover, as this oversight likely stems from a broader lack of conceptual clarity about how these mechanisms function within the system, this study formally incorporates these mechanisms into CJ's conceptual model.
 
@@
 In particular, it clarifies their role in the data-generating process by framing them as sources of missing data--that is, as mechanisms that determine which observations are missing from the final data sample.
-This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and assessing more complex experimental setups.
+This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and assessing more complex assessments setups.
 
@@
 <!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
-However, many modern CJ applications rely on more complex experimental designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
+However, many modern CJ applications rely on more complex assessment designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
 A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
-This outcome-informed selection suggests that the comparison mechanism may be outcome-dependent, potentially classifying ACJ as a generator of MNAR data.
-This classification may, in turn, facilitate a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities.
-While some studies report that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], others argue that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
+This outcome-informed pair selection suggests that the comparison mechanism is outcome-dependent, potentially classifying ACJ as a generator of MNAR data.
+This, in turn, facilitates a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities.
+While some studies report that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], others argue that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022], potentially as a result of outcome-dependent missingness.
 
@@
 
-<!-- 5. The importance of clarifying assumptions behind CJ experimental designs -->
-Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions, an approach supported by our framework.
-This clarity enables researchers to evaluate, for instance, how sampling and comparison mechanisms affect trait estimation and statistical inference within each design.
+<!-- 5. The importance of clarifying assumptions behind CJ assessment designs -->
+Nevertheless, regardless of the underlying missingness mechanisms, any CJ assessment design would benefit from explicitly defining its assumptions--a practice supported by our proposed approach.
+This clarity enables researchers to evaluate how the sampling and comparison mechanisms affect trait estimation and statistical inference within each design.
 Such assessments are particularly relevant given the common misconception in the CJ literature that Thurstone's model can naturally handle even non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].
@@
 <!-- recording: Sven and Tine 25.05.21; time: 00:24:00 - 00:31:45 -->
+<!-- recording: Sven and Tine 25.06.04; time: 00:11:55 - 00:16:50 -->
 
 <!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
-Despite the growing notion that various stimulus-related factors can influence judges' perceptions [@Morin_et_al_2018] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
-We argue that this gap persists not due to a lack of interest or research but because researchers continue to rely on ad-hoc detection methods, such as 'misfit' statistics, that are poorly suited for the task [@Kelly_et_al_2022].
-To address this limitation, this study decided to treat judges' biases as an integral component of the CJ system from the outset.
+Despite the growing notion that various stimulus-related factors can influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
+This gap likely persists not due to a lack of interest or research but because researchers often rely on ad-hoc detection methods--such as 'misfit' statistics--that are poorly suited for the task [@Kelly_et_al_2022].
+To address this limitation, this study treats judges' biases as an integral component of the CJ system from the outset.
 This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.
+
+<!-- Such differences may not be detected through analyses of bias and misfit.
+McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
+Where students prioritized factual recall, teachers prioritized scientific explanation.
+Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
+Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
+[@Kelly_et_al_2022, pp.
+683] -->
 
@@
 The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions.
-For instance, can researchers validly analyze CJ data under the assumption of "sample-free" trait calibration--specifically, under the assumption that judges exhibit no systematic bias?
-This question is especially relevant because researchers often treat "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases.
-Furthermore, if judges' biases reflect additional stimulus characteristics that should not influence assessment, to what extent can training or expertise mitigate these biases?
-[@Kelly_et_al_2022] More specifically, as discussed in @sec-theory-issue1b, can training or expertise help prevent judges from prioritizing irrelevant features--such as handwriting--over more central criteria like argumentative quality in writing assessments?
-Addressing these questions may also shed light on what it truly means to be an "expert" in the CJ context [@Kelly_et_al_2022].
+One key question is whether researchers can validly analyze CJ data under the assumption of "sample-free" trait calibration, specifically under the hypothesis that judges exhibit no systematic bias.
+This question is particularly relevant because many researchers still regard "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases.
+Another critical question, discussed in @sec-theory-issue1b, is whether training or expertise can help judges avoid focusing on irrelevant stimulus features--such as handwriting--over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022].
+Notably, exploring these questions may also provide insights into what it truly means to be an "expert" within the CJ context [@Kelly_et_al_2022].
 
@@
 <!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
-Moreover, if we accept that judges may prioritize different aspects of a stimulus, it follows that their assessments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020].
-Thus, building on the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
-Furthermore, if such characteristics indeed undermine the assumption of "sample-freeness," then it becomes essential to explore strategies to control their influence and to determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts.
+Moreover, since judges draw on a variety of intricate stimulus characteristics when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these characteristics account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that judges' assessments also vary with individual attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a].
+Thus, building on the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual attributes introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
+Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," then it becomes essential to explore strategies to control their influence and to determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts.
 Additionally, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
@@
 
-### The identification of 'misfitting' judges and stimuli {#sec-discussion_RA3}
+### The identification of 'misfitting' stimuli and judges {#sec-discussion_RA3}
 <!-- recording: Sven and Tine 25.05.21; time: 00:31:45 - 00:32:20 -->
-
-[About the identification of 'misfitting' judges and stimuli]{style="color:blue;"}
+<!-- recording: Sven and Tine 25.06.04; time: 00:33:55 - 00:53:15 -->
+
+[LAST THING TO DO]{style="color:blue;"}
+
+<!-- Although the CJ literature clearly defines *misfit judges* as those whose assessments deviate markedly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], there has been little discussion about how these misfit judges relate to Thurstonian theory.
+In particular, it remains unclear which aspects of Thurstone's theory might explain the occurrence of misfit judges.
+To address this, we propose three statistical perspectives: (1) misfit judges are simply judges who exhibit bias; (2) misfit judges are those whose judgments display greater variability than their peers, and (2) misfit judges are indeed outliers.
+-->
+
+<!-- Regarding the first perspective, @sec-theory-issue1b explains in detail how, due to the assumption of zero correlation, judges' biases can be miss classified as *misfits*.
+Meanwhile, @sec-discussion_RA2 discusses how our approach explicitly account for such biases, aiming to reduce the reliance on ad-hoc detection methods--such as misfit statistics--which are poorly suited to the task [@Kelly_et_al_2022].
+-->
+
+<!-- Regarding the second perspective -->
+
+<!-- Conversely, regarding the second perspective, the very definition of *misfit judges* leads to their identification as outliers--that is, observations that deviate significantly from the rest of the sample in which they occur [@Grubbs_1969].
+-->
+
+<!-- Such differences may not be detected through analyses of bias and misfit.
+McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
+Where students prioritized factual recall, teachers prioritized scientific explanation.
+Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
+Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
+[@Kelly_et_al_2022, pp.
+683] -->
+
+<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, pp. 1] -->
+
+<!-- In the previous subsection we argue that ad-hoc detection methods, such as 'misfit' statistics, could be poorly suited for the task of detecting judges bias [@Kelly_et_al_2022].
+This hinges in the assumption that the CJ data does have judges biases, but the method does not manages to identify them, leading to false negatives?
+or lack of sensitivity in the tests.
+However, there is a related problem which is: these ad-hoc method could be detecting judges' biases that are not there.
+as in @Wu_et_al_2022  -->
+
+<!-- These procedures also bear on a larger topic of discussion in social science.
+The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
+[@Wainer_et_al_1978, pp.
+332] -->
+
+<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
+If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
+...
+It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
+681] -->
+
+<!-- A similar point can be made for *misfit stimuli*--that is, those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018].
+It remains unclear which aspects of Thurstone's theory might explain the occurrence of misfit stimuli.
+To address this, we propose two statistical perspectives: (1) misfit stimuli are those whose discriminal processes exhibit greater variability compared to other stimuli, and (2) are statistical outliers.
+-->
+
+<!-- Regarding the first perspective, @sec-theory-issue1a  -->
+
+<!-- , possibly due to larger discriminal dispersions or because they are genuine outliers--meaning, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+In either case, the BTL model neither accounts for these anomalies nor provides tools for addressing them, apart from excluding the "problematic" texts from analysis.
+-->
+
+<!-- The same could be said for the stimuli.
+These misfit texts likely arise from two underlying mechanisms.
+First, a text may exhibit a larger discriminal dispersion, increasing the overlap between its discriminal process and those of multiple other texts.
+This greater overlap leads to more judgment discrepancies, increasing the likelihood of classifying the text as a misfit.
+@fig-discriminal_process illustrates this scenario.
+In the figure, Text B has a wider discriminal dispersion than Text A, enlarging the region where their discriminal processes intersect.
+This larger overlap represents an area where judges face greater difficulty in determining which text exhibits a higher level of the trait.
+Second, a text may represent an outlier --one with distinct characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+Such texts are harder to compare and place on the trait scale, resulting in more judgment discrepancies and a misfit classification.
+In either case, the BTL model is inherently agnostic to these mechanisms and does not provide any means to address them.
+-->
 
@@
 
-<!-- removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969; @Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates.
-The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
--->
+<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
+
+<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
+
+<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
+
+<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
+This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
+12].
+-->
+
+<!-- Furthermore, these points tend to be outliers with unlikely values, according to the model.
+As a result, they are highly influential and make it difficult to estimate out-of-sample predictive accuracy.
+Why?
+Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
+[@McElreath_2020, pp.
+231] -->
+
+<!-- What can be done about this?
+There is a tradition of dropping outliers.
+People sometimes drop outliers even before a model is fit, based only on standard deviations from the mean outcome value.
+You should never do that—a point can only be unexpected and highly influential in light of a model.
+After you fit a model, the picture changes.
+If there are only a few outliers, and you are sure to report results both with and without them, dropping outliers might be okay.
+But if there are several outliers and we really need to model them, what then?
+[@McElreath_2020, pp.
+232] -->
+
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
+
+<!-- Consider, for example, the simplest location and scale problem-inference for a normal distribution with unknown mean \theta and variance \sigma^{2}.
+If we look at the marginal posterior distribution for \theta we know that if \sigma^{2} is known this posterior exhibits outlier-resistance, and we might expect similar behavior for unknown \sigma^{2}, but a simple example refutes this.
+Let the prior distribution give probability p (0<p<1) to (\theta = 1, \sigma^{2} = 1) and 1-p to (\theta = 0, \sigma^{2} = 4).
+Then the posterior mean of \theta is a decreasing function of any sufficiently large observation.
+[@OHagan_2018, pp.
+366] -->
+
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
 
@@
 
-<!-- In the previous subsection we argue that ad-hoc detection methods, such as 'misfit' statistics, could be poorly suited for the task of detecting judges bias [@Kelly_et_al_2022].
-This hinges in the assumption that the CJ data does have judges biases, but the method does not manages to identify them, leading to false negatives?
-or lack of sensitivity in the tests.
-However, there is a related problem which is: these ad-hoc method could be detecting judges' biases that are not there.
-as in @Wu_et_al_2022  -->
-
-<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
-If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
-...
-It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
-681] -->
-
-<!-- The same could be said for the stimuli.
-These misfit texts likely arise from two underlying mechanisms.
-First, a text may exhibit a larger discriminal dispersion, increasing the overlap between its discriminal process and those of multiple other texts.
-This greater overlap leads to more judgment discrepancies, increasing the likelihood of classifying the text as a misfit.
-@fig-discriminal_process illustrates this scenario.
-In the figure, Text B has a wider discriminal dispersion than Text A, enlarging the region where their discriminal processes intersect.
-This larger overlap represents an area where judges face greater difficulty in determining which text exhibits a higher level of the trait.
-Second, a text may represent an outlier --one with distinct characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-Such texts are harder to compare and place on the trait scale, resulting in more judgment discrepancies and a misfit classification.
-In either case, the BTL model is inherently agnostic to these mechanisms and does not provide any means to address them.
--->
-
-<!-- Such differences may not be detected through analyses of bias and misfit.
-McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment.
-Where students prioritized factual recall, teachers prioritized scientific explanation.
-Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers).
-Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
-[@Kelly_et_al_2022, pp.
-683] -->
-
-<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
-
-<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
-
-<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
-
-<!-- Furthermore, these points tend to be outliers with unlikely values, according to the model.
-As a result, they are highly influential and make it difficult to estimate out-of-sample predictive accuracy.
-Why?
-Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
-[@McElreath_2020, pp.
-231] -->
-
-<!-- What can be done about this?
-There is a tradition of dropping outliers.
-People sometimes drop outliers even before a model is fit, based only on standard deviations from the mean outcome value.
-You should never do that—a point can only be unexpected and highly influential in light of a model.
-After you fit a model, the picture changes.
-If there are only a few outliers, and you are sure to report results both with and without them, dropping outliers might be okay.
-But if there are several outliers and we really need to model them, what then?
-[@McElreath_2020, pp.
-232] -->
-
-<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
-
-<!-- Consider, for example, the simplest location and scale problem-inference for a normal distribution with unknown mean \theta and variance \sigma^{2}.
-If we look at the marginal posterior distribution for \theta we know that if \sigma^{2} is known this posterior exhibits outlier-resistance, and we might expect similar behavior for unknown \sigma^{2}, but a simple example refutes this.
-Let the prior distribution give probability p (0<p<1) to (\theta = 1, \sigma^{2} = 1) and 1-p to (\theta = 0, \sigma^{2} = 4).
-Then the posterior mean of \theta is a decreasing function of any sufficiently large observation.
-[@OHagan_2018, pp.
-366] -->
-
-<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
-
-{{< pagebreak >}}
-
-### The effects of hierarchies on experimental design and inference {#sec-discussion_RA4}
-<!-- recording: Sven and Tine 25.05.21; time: 00:32:25 - 00:33:00 -->
-
-[About the hierarchical structure of stimuli]{style="color:blue;"}
-
-<!-- "This explicit part of the model it open ups the idea of doing more throughout research on ..." -->
-
-<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments.
-Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges.
--->
-
-<!-- About the hierarchical structure of stimuli and its impact on CJ experiment design and hypothesis testing, one such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy.
-It has been highlighted that due to concerns about the feasibility of the comparison task [@Boonen_et_al_2020], many CJ studies assess traits using only one text per individual [@citation].
-Although the literature have examples of collecting more than one stimuli per individual, their statistical handling still separates the trait measurement from the statistical inference process.
-This experimental design implicitly assumes that within-individual variability is small relative to between-individual differences--that is, texts produced by the same individual are relatively homogeneous.
-However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates.
--->
-
-<!-- Relatedly, another study could investigate hypotheses about the characteristics of the texts themselves.
-Previously, the literature lacked a direct and integrated method for evaluating these types of hypotheses.
-Now, by collecting data on specific stimulus features, researchers can assess whether these variables produce the expected effects.
--->
-
-<!-- The avenues of research span the following:  -->
-<!-- - assessing if the number of texts a researcher collects is sufficient.
-Due to efficiency of scoring, most of research only collects only one text.
-This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability.
-However, this cannot be known ad-hoc.
-The frameworks open the avenue to test this assumption -->
-<!-- - Assessing hypothesis regarding the characteristics of the texts.
-Before the literature did not have a direct and integrated way to assess these types of hypothesis, now you just need to collect data on the particular stimuli, and test if these variables reflect the expected effect -->
-
-<!-- Derived from this other avenues of research that are open are -->
-<!-- - assessing if the number of individuals a researcher collects is sufficient.
-Most of the literature focuses on ensuring that the number of comparisons between texts are sufficient, but no literature actively engages in determining if the number of individuals are sufficient to test hypothesis at that level.
--->
-<!-- - Moreover, if certain characteristics of the individuals explain the trait locations of their texts -->
-
-<!-- When you want to investigate objects, the experiment has an experimental unit: the trial on the objects, but you only have one object per subject.
-Consequently, objects and subjects are confounded.
--->
-
-{{< pagebreak >}}
-
-## Limitations and challenges {#sec-discussion_challenges}
+## Study limitations and practical challenges for applied CJ researchers {#sec-discussion_challenges}
 <!-- recording: Sven and Tine 25.05.21; time: 00:33:50 - 00:42:20 -->
-
-Drawing conclusions about the real world from observed data always requires assumptions, regardless of whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022].
-This study is no exception to this principle.
-By adopting causal inference as the theoretical backbone of our proposed framework, we necessarily rely on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
-As such, our framework does not yield automatic answers.
-Instead, it encourages researchers to formulate precise questions and to explicitly articulate their assumptions to develop a generalizable understanding of the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024].
-This clarity is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019]--a condition we cannot yet guarantee for our proposed models.
-Nevertheless, a key strength of this theory-driven approach is that it lays the foundation for empirically testing the proposed causal assumptions [@Deffner_et_al_2022], as discussed in @sec-discussion_RA.
+<!-- recording: Sven and Tine 25.06.04; time: 00:18:45 - 00:33:55 -->
+
+The process of deriving conclusions from observed data always requires assumptions, whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022].
+Our approach is not an exception to this fundamental principle.
+As with all approaches grounded on causal inference, ours has the limitation of relying on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
+As such, the approach does not seek to yield automatic answers when applied to a given CJ dataset.
+Instead, it aims to encourage researchers to formulate precise questions and to make their assumptions explicit, fostering a generalizable understanding of the CJ system under study [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024].
+This clarity is critical because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019].
+Although this alignment remains empirically untested for the proposed models, the theory-driven nature of our approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], which are well supported by both theory and existing evidence (see @sec-theory-issues and @sec-theoretical).
 
@@
 
-<!-- 1. Challenges for the framework: background knowledge and research context -->
-This theoretical commitment to causal inference also introduces several challenges that researchers must navigate--challenges common to many quantitative studies.
-These fall into two main categories.
-The first concerns the foundational knowledge researchers need to apply our proposed framework effectively.
-The second involves the increased attention researchers must now devote to their specific research context.
-
-### Foundational knowledge {#sec-discussion_challenges1}
-
-To effectively apply our framework, researchers must tackle two main challenges related to foundational knowledge.
+<!-- 1. Challenges for the approach: background knowledge and research context -->
+Conversely, this theoretical commitment to causal inference also introduces several practical challenges that applied CJ researchers must navigate.
+These fall into two main categories: first, acquiring the foundational knowledge necessary to apply the approach effectively; and second, dedicating greater attention to both conceptual and statistical modeling.
+
+### Required foundational knowledge {#sec-discussion_challenges1}
+<!-- recording: Sven and Tine 25.06.04; time: 00:24:30 - 00:25:15 -->
+<!-- recording: Sven and Tine 25.06.04; time: 00:26:05 - 00:26:30 -->
+
+To apply our approach effectively, CJ researchers must tackle two main challenges related to the required foundational knowledge.
 First, they need a solid understanding of causal inference principles.
@@
 However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference.
-Consequently, researchers must deeply engage with these and other complex ideas, including how SCMs encode functional and probabilistic information.
-To assist this effort, we include key references throughout the study to guide motivated readers toward a deeper understanding of these concepts [^10].
+As a result, CJ researchers must deeply engage with these and other complex ideas, including how SCMs encode functional and probabilistic information.
+To assist this effort, this study includes key references throughout the study to guide motivated aaplied CJ researchers toward a deeper understanding of these concepts [^10].
 
@@
 Developing the skills to translate conceptual models into bespoke statistical models also presents challenges.
-Although Bayesian inference methods offer a more accessible path for many applied researchers to develop these skills--by reducing the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--they still demand familiarity with other technical concepts such as probabilistic programming languages (PPLs), probability distributions, and convergence, which researchers may find hard to master.
-Thus, to support this skill development, we include links to the statistical model code and alternative specifications in the **Declarations** section of this document.
-
-### Attention to the research context {#sec-discussion_challenges2}
-
-Researchers face two additional challenges when applying our framework to their specific research context.
-First, they must verify that the conceptual model provides a *faithful* [^11] representation of the CJ system under study.
+Although Bayesian inference methods offer a more accessible path for many applied CJ researchers to develop these skills--by reducing the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--they still demand a working knowledge of other technical concepts such as probabilistic programming languages (PPLs), probability distributions, and convergence.
+These requirements can be non-trivial to master.
+To support CJ researchers in developing these skills, this study provides a link to the statistical model code and alternative model specifications in the **Declarations** section of this document [^11].
+
+[^11]: Seminal texts on Bayesian inference methods, such as @Gelman_et_al_2014 and @McElreath_2020, offer valuable support for developing a deeper understanding of these models.
+
+### Attention to conceptual and statistical modeling {#sec-discussion_challenges2}
+<!-- recording: Sven and Tine 25.06.04; time: 00:26:30 - 00:33:55 -->
+
+Even after acquiring the necessary foundational knowledge, CJ researchers face two additional challenges when applying our approach to their specific research context.
+First, they must verify whether the conceptual model provides a *faithful* [^12] representation of the CJ system under study.
 Second, they need to assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data.
 To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets.
-This consideration means researchers must tackle the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification.
-Our approach supports this process by offering a transparent framework that helps researchers explicitly articulate these new assumptions and supports the design of CJ experiments that can yield reliable and valid inferences.
-
-[^11]: Avoid confusing this term with the *faithfulness assumption* [@Neal_2020; @Hernan_et_al_2025] described in the causal inference literature. For further details, see footnote 1.
+While these models may prove adequate in some contexts, researchers cannot assume this apriori.
+Instead, they need to adapt these models to their specific CJ contexts, paying close attention to assessment design features and causal assumptions.
+As discussed in @sec-discussion_challenges, our approach facilitates this process by offering a transparent framework for articulating new assumptions and guiding the design of CJ assessments.
+
+[^12]: Avoid confusing this term with the *faithfulness assumption* [@Neal_2020; @Hernan_et_al_2025] described in the causal inference literature. For further details, see footnote 1.
 
@@
 <!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
-
-<!-- These procedures also bear on a larger topic of discussion in social science.
-The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
-[@Wainer_et_al_1978, pp.
-332] -->
 
@@
 
-Although performing identification analysis through formal derivations may appear to be a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset.
+Although performing identification analysis through formal derivations may seem like a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset.
 Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs.
 Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research.
-Importantly, our framework supports both approaches, as it provides the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
+Importantly, our approach supports both strategies by providing the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
 

### 2025-06-05 _tex/paper2.tex
Built site for gh-pages

--- 2025-06-04 _tex/paper2.tex
@@
 inference methods. This extension integrates core theoretical principles
-alongside key assessment design features relevant to CJ experiments,
-such as the selection of judges, stimuli, and comparisons. It then
-translates these elements into a probabilistic statistical model for
-analyzing dichotomous CJ data, overcoming the rigid assumptions of Case
-V and the BTL model.
+alongside key CJ assessment design features, such as the selection of
+judges, stimuli, and comparisons. It then translates these elements into
+a probabilistic statistical model for analyzing dichotomous CJ data,
+overcoming the rigid assumptions of Case V and the BTL model.
 
@@
 it clarifies the interactions among all actors and processes involved in
-CJ experiments. Second, it shifts the current comparative data analysis
+CJ assessments. Second, it shifts the current comparative data analysis
 paradigm from passively accepting the BTL model assumptions to actively
@@
 form to address these challenges. The extension integrates core
-theoretical principles alongside key CJ experimental design features,
-such as the selection of judges, stimuli, and comparisons.
+theoretical principles alongside key CJ assessment design features, such
+as the selection of judges, stimuli, and comparisons.
 Section~\ref{sec-statistical} translates these theoretical and practical
@@
 parameters than the number of available pairwise comparisons
-\citep{Thurstone_1927b}. For instance, in a CJ experiment with five
+\citep{Thurstone_1927b}. For instance, in a CJ assessment with five
 texts, the general form would require estimating \(20\) parameters: five
@@
 measurement model, evidence from the CJ literature suggests that this
-assumption may fail to capture the complexity of some traits or account
-for heterogeneous stimuli, such as handwritten texts or English
-compositions
+assumption may fail accommodate for heterogeneous stimuli, such as
+handwritten texts or English compositions
 \citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
-Indeed, the presence of the so-called \emph{misfit} texts may already
-signal these limitations, as these are texts that elicit more judgment
-discrepancies than others
-\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018},
-possibly due to larger discriminal dispersions or because they are
-genuine outliers--meaning, texts with distinctive characteristics that
-deviate markedly from the rest of the sample \citep{Grubbs_1969}. In
-either case, the BTL model neither accounts for these anomalies nor
-provides tools for addressing them, apart from excluding the
-``problematic'' texts from analysis.
+Indeed, the presence of the so-called \emph{misfit texts}--texts that
+elicit more judgment discrepancies than others
+\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}--may
+signal this limitation, as such discrepancies can arise from larger
+discriminal dispersions due to stimulus heterogeneity or because the
+texts are genuine outliers--that is, texts with distinctive
+characteristics that deviate markedly from the rest of the sample
+\citep{Grubbs_1969}. In either case, the BTL model's assumptions prevent
+it from adequately accounting for or addressing these anomalies, leaving
+exclusion of the ``problematic'' texts as the primary remedy.
 
@@
 difference under various correlational assumptions, in practice,
-experimental designs often adopt the assumption of no correlation
-between stimuli based on an old theoretical justification. Specifically,
+assessment designs often adopt the assumption of no correlation between
+stimuli based on an old theoretical justification. Specifically,
 \citet{Thurstone_1927b} argued that stimuli could be treated as
@@
 In the first scenario, research on text quality suggests that when
-judges evaluate multidimensional, complex traits or heterogeneous
+judges evaluate complex, multidimensional traits or heterogeneous
 stimuli, they often rely on a variety of intricate stimulus
 characteristics to inform their judgments
-\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
-characteristics, regardless of their relevance, are unlikely to be
-equally weighted or to oppose one another across comparisons. As a
+\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
+Regardless of their relevance, these characteristics may not receive
+equal weight or consistently oppose one another across comparisons. As a
 result, they may exert a disproportionate influence on judges'
@@
 could introduce dependencies between the stimuli
-\citep{vanderLinden_et_al_2017_II}. Evidence from traditional marking
-contexts confirms a similar phenomenon \citep{Morin_et_al_2018}, and
-additional studies
-\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}
-support the presence of these biases in CJ--reinforcing the argument
-that the factors influencing pairwise comparisons do not always cancel
-each other out.
+\citep{vanderLinden_et_al_2017_II}. While direct evidence for this exact
+scenario is limited, existing studies document the presence of judge
+bias in CJ \citep[\citet{vanDaal_et_al_2016},
+\citet{Bartholomew_et_al_2020a}]{Pollitt_et_al_2003}, reinforcing the
+argument that the factors influencing pairwise comparisons do not always
+cancel each other out.
 
@@
 \citep{Ackerman_1989, Hoyle_et_al_2023} and even biases
-\citep{Wu_et_al_2022}. These inaccuracies can result in spurious
-conclusions about the discriminal differences \citep{McElreath_2020}
-and, by extension, the underlying discriminal processes of the stimuli.
-This issue is illustrated in Figure~\ref{fig-correlation} when the
-discriminal difference distribution of the BTL scores follows the thick
-continuous line \((\rho = 0)\), while the \emph{true} discriminal
-difference follows any discontinuous line where \(\rho \neq 0\).
-
-Finally, as discussed in the previous section, removing \emph{misfit}
-judges--that is, judges whose assessments deviate markedly from the
-shared consensus
-\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
-and considered outliers under the BTL model
-\citep{Grubbs_1969, Wu_et_al_2022}--risk discarding valuable information
-and introducing bias into trait estimates. The direction and magnitude
-of these biases remain unpredictable, as they depend on which judges
+\citep{Wu_et_al_2022}. This issue is illustrated in
+Figure~\ref{fig-correlation} when the discriminal difference
+distribution of the BTL scores follows the thick continuous line
+\((\rho = 0)\), while the \emph{true} discriminal difference follows any
+discontinuous line where \(\rho \neq 0\). Moreover, these inaccuracies
+can result in spurious conclusions about the discriminal differences
+\citep{McElreath_2020} and, by extension, the underlying discriminal
+processes of the stimuli--potentially causing the model to incorrectly
+classify stimuli or judges as \emph{misfits}.
+
+Finally, similar to what we discussed in the previous section, removing
+\emph{misfit} judges--that is, judges whose assessments deviate markedly
+from the shared consensus
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
+and may appear as outliers under the BTL model
+\citep{Wu_et_al_2022}--risk discarding valuable information and
+introducing bias into trait estimates. The direction and magnitude of
+these biases remain unpredictable, as they depend on which judges
 researchers exclude from the analysis
@@
 formally and graphically represent the assumed causal structure of a
-system, such as the one found in CJ experiments. Essentially, SCMs and
+system, such as the one found in CJ assessments. Essentially, SCMs and
 DAGs function as \emph{conceptual models} on which identification
@@
 ability to produce high-quality written texts?'' To investigate this, a
-researcher designs a CJ experiment by randomly assigning students
+researcher designs a CJ assessment by randomly assigning students
 (individuals) to two groups, each receiving a different teaching method.
@@
 combines the core theoretical principles outlined in
-Section~\ref{sec-thurstone_theory} with key assessment design features
-relevant to CJ experiments, such as the selection of judges, stimuli,
-and comparisons. In addition to improving statistical accuracy and
-strengthening measurement reliability and validity, the approach offers
-two key advantages. First, it clarifies the interactions among all
-actors and processes involved in CJ experiments. Second, it shifts the
-current comparative data analysis paradigm from passively accepting the
-model assumptions to actively testing whether those assumptions fit the
-data under analysis.
+Section~\ref{sec-thurstone_theory} with key CJ assessment design
+features, such as the selection of judges, stimuli, and comparisons. In
+addition to improving statistical accuracy and strengthening measurement
+reliability and validity, the approach offers two key advantages. First,
+it clarifies the interactions among all actors and processes involved in
+CJ assessments. Second, it shifts the current comparative data analysis
+paradigm from passively accepting the model assumptions to actively
+testing whether those assumptions fit the data under analysis.
 
@@
 comparison (i.e., the stimuli). The index \(j\) indicates the judge
-conducting the comparison, while the index \(k\) accounts for
-experimental conditions where a judge compares the same pair of stimuli
-multiple times, i.e., a \emph{repeated measures designs}
+conducting the comparison, while the index \(k\) accounts for assessment
+conditions where a judge compares the same pair of stimuli multiple
+times, i.e., a \emph{repeated measures designs}
 \citep[pp.~366-376]{Lawson_2015}. Thus, the indexing system supports
@@
 difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
-Additionally, the SCM incorporates a key design feature of CJ
-experiments: the judges' biases \((B_{kj})\). This extension builds on
-the arguments presented in Section~\ref{sec-theory-issue1b}, contending
-that the discriminal difference becomes an observable outcome only
-through judges' perceptions. Given that such perceptions may be
-imperfect--and that each judge may carry some degree of bias
+Additionally, the SCM incorporates a key CJ design feature: the judges'
+biases \((B_{kj})\). This extension builds on the arguments presented in
+Section~\ref{sec-theory-issue1b}, contending that the discriminal
+difference becomes an observable outcome only through judges'
+perceptions. Given that such perceptions may be imperfect--and that each
+judge may carry some degree of bias
 \citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
@@
 However, due to concerns about the practical feasibility of the
-comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ experiments rarely
+comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ assessments rarely
 implement an exhaustive pairings of sampled judges, stimuli, and
@@
 estimation and hypothesis testing (see
-Section~\ref{sec-theory-theoretical_P3}), and clarify the role of the
-sample and comparison mechanisms in CJ experiments (see
+Section~\ref{sec-theory-theoretical_P3}), and even clarify the role of
+the sample and comparison mechanisms in CJ assessments (see
 Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
@@
 \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
-feature is particularly relevant in CJ experiments, as researchers often
+feature is particularly relevant in CJ assessments, as researchers often
 collect large volumes of paired comparisons but work with relatively
@@
 mathematically tractable logistic distribution
-\citep{Andrich_1978, Bramley_2008}. Notably, although these assumptions
-simplify the trait measurement model, they may fail to capture the
-complexity of some traits or account for heterogeneous stimuli
+\citep{Andrich_1978, Bramley_2008}. Although this substitution has
+minimal impact on trait estimation or model interpretation
+\citep{vanderLinden_et_al_2017_I, McElreath_2021}, the simplifying
+assumptions of the BTL model--and by extension, of Case V--may fail to
+capture the complexity of some traits or account for heterogeneous
+stimuli
 \citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022},
@@
 \citep[pp.~269]{Thurstone_1927a}, his theory offered no guidance on how
-to use trait estimates for statistical inference. To address this gap,
-researchers in the CJ tradition have often separated trait estimation
-from hypothesis testing, relying on point estimates of the traits--such
-as BTL scores or their transformations--for inference. Notably, while
-this approach simplifies analysis, it can also introduce bias and
-compromise the reliability of the resulting inferences
+to use trait estimates for statistical inference. The CJ tradition has
+attempted to address this gap by separating trait estimation from
+hypothesis testing, relying on point estimates--such as BTL scores or
+their transformations--for inference. Notably, while this approach
+simplifies analysis, it can also introduce bias and compromise the
+reliability of the resulting inferences
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
 
-Thus, to address these limitations of Thurstone's Case V and the BTL
+Thus, to address the limitations of Thurstone's Case V and the BTL
 model, this study extends Thurstone's general form using a systematic,
@@
 model integrates Thurstone's core theoretical principles, such as the
-discriminal processes of stimuli, alongside key CJ experimental design
+discriminal processes of stimuli, alongside key CJ assessment design
 features, including judges' bias, sampling procedures, and comparison
@@
 
-The study then translates the SCM into a bespoke statistical model that
-allows researchers to analyze CJ data when violations to the assumptions
-of equal dispersion and zero correlation occur (see
+The approach then translates the SCM into a bespoke statistical model
+that allows researchers to analyze CJ data when violations to the
+assumptions of equal dispersion and zero correlation occur (see
 Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b})
@@
 potential to enhance the reliability and validity of trait measurement
-\citep{Perron_et_al_2015} and improve the accuracy of statistical
-inferences.
+in CJ \citep{Perron_et_al_2015}, while also improving the accuracy of
+statistical inferences.
 
@@
 advantages. First, it clarifies the roles and interactions of all actors
-and processes involved in CJ experiments. Second, it shifts the analytic
+and processes involved in CJ assessments. Second, it shifts the analytic
 paradigm from passively accepting the assumptions of the BTL model to
@@
 advantages establish a principled framework for evaluating best
-practices in CJ experimental designs, one that better aligns with the
-demands of contemporary CJ assessment contexts \citep{Kelly_et_al_2022},
-offering new insights into existing research and opening promising
-avenues for future inquiry.
-
-\subsection{Further research}\label{sec-discussion_RA}
-
-Four current and new research avenues deserve particular attention due
-to their direct impact on the reliability and validity of trait
-estimates and the accuracy of statistical inferences. The following
-sections outline these avenues and explain how our model provides a
-rigorous research framework to investigate them.
+practices in designing CJ assessments, one that better aligns with the
+demands of contemporary CJ contexts \citep{Kelly_et_al_2022}, offering
+new insights into existing research and opening promising avenues for
+future inquiry.
+
+\subsection{Future research directions based on this study's
+approach}\label{sec-discussion_RA}
+
+Among the many potential directions for future research--both current
+and new--three avenues deserve particular attention due to their direct
+impact on the reliability and validity of trait estimates, as well as on
+the accuracy of statistical inferences. The following sections outline
+these avenues and explain how our approach provides a rigorous research
+framework to investigate them.
 
@@
 Although sampling and comparison mechanisms are central to modern CJ
-experimental design, it is striking that most of the CJ literature has
-examined them only within a limited scope. Researchers have primarily
+assessments, it is striking that most of the CJ literature has examined
+them only within a limited scope. Researchers have primarily
 investigated the effects of adaptive comparative judgment (ACJ) designs
@@
 or proposed practical guidelines for the number of comparisons judges
-should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}.
-Although these studies offer valuable insights, we argue that they have
-overlooked the broader role these mechanisms play within the CJ system.
-Moreover, we suggest that this oversight largely stems from a lack of
-conceptual clarity about how the mechanisms function within the system.
-To address this gap, this study decided to integrate these mechanisms
-into CJ's conceptual model.
+should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}. While
+these studies offer valuable insights, they often overlook the broader
+role these mechanisms play within the CJ system. Moreover, as this
+oversight likely stems from a broader lack of conceptual clarity about
+how these mechanisms function within the system, this study formally
+incorporates these mechanisms into CJ's conceptual model.
 
@@
 enabling a more rigorous evaluation of existing claims about these
-mechanisms, their influence on CJ outcomes, and their implications for
-designing and assessing more complex experimental setups.
+missing data mechanisms, their influence on CJ outcomes, and their
+implications for designing and assessing more complex assessments
+setups.
 
@@
 
-However, many modern CJ applications rely on more complex experimental
+However, many modern CJ applications rely on more complex assessment
 designs in which the sampling and comparison mechanisms introduce more
@@
 comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
-outcome-informed selection suggests that the comparison mechanism may be
-outcome-dependent, potentially classifying ACJ as a generator of MNAR
-data. This classification may, in turn, facilitate a more nuanced
-understanding of the mixed evidence surrounding ACJ's capabilities.
-While some studies report that the method improves trait reliability
+outcome-informed pair selection suggests that the comparison mechanism
+is outcome-dependent, potentially classifying ACJ as a generator of MNAR
+data. This, in turn, facilitates a more nuanced understanding of the
+mixed evidence surrounding ACJ's capabilities. While some studies report
+that the method improves trait reliability
 \citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
 that it artificially inflates these gains
-\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
+\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022},
+potentially as a result of outcome-dependent missingness.
 
 Nevertheless, regardless of the underlying missingness mechanisms, any
-CJ experimental design would benefit from explicitly defining its
-assumptions, an approach supported by our framework. This clarity
-enables researchers to evaluate, for instance, how sampling and
-comparison mechanisms affect trait estimation and statistical inference
-within each design. Such assessments are particularly relevant given the
-common misconception in the CJ literature that Thurstone's model can
-naturally handle even non-random missing data without compromising the
-reliability or validity of trait estimates \citep{Bramley_2008}.
+CJ assessment design would benefit from explicitly defining its
+assumptions--a practice supported by our proposed approach. This clarity
+enables researchers to evaluate how the sampling and comparison
+mechanisms affect trait estimation and statistical inference within each
+design. Such assessments are particularly relevant given the common
+misconception in the CJ literature that Thurstone's model can naturally
+handle even non-random missing data without compromising the reliability
+or validity of trait estimates \citep{Bramley_2008}.
 
@@
 Despite the growing notion that various stimulus-related factors can
-influence judges' perceptions \citep{Morin_et_al_2018} and that these
-influences may not always cancel each other out, few studies in the CJ
-literature provide empirical evidence for judges' biases
+influence judges' perceptions
+\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}
+and that these influences may not always cancel each other out, few
+studies in the CJ literature provide empirical evidence for judges'
+biases
 \citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
-We argue that this gap persists not due to a lack of interest or
-research but because researchers continue to rely on ad-hoc detection
-methods, such as `misfit' statistics, that are poorly suited for the
-task \citep{Kelly_et_al_2022}. To address this limitation, this study
-decided to treat judges' biases as an integral component of the CJ
-system from the outset. This approach offers one key advantage: it
-provides a more accurate representation of the data-generating process
-behind pairwise comparisons, one that acknowledges that the discriminal
-processes of stimuli become an observable outcome only through judges'
-perceptions, which may exhibit bias.
+This gap likely persists not due to a lack of interest or research but
+because researchers often rely on ad-hoc detection methods--such as
+`misfit' statistics--that are poorly suited for the task
+\citep{Kelly_et_al_2022}. To address this limitation, this study treats
+judges' biases as an integral component of the CJ system from the
+outset. This approach offers one key advantage: it provides a more
+accurate representation of the data-generating process behind pairwise
+comparisons, one that acknowledges that the discriminal processes of
+stimuli become an observable outcome only through judges' perceptions,
+which may exhibit bias.
 
 The explicit integration of judges' bias into CJ's conceptual model then
-paves the way for investigating several relevant research questions. For
-instance, can researchers validly analyze CJ data under the assumption
-of ``sample-free'' trait calibration--specifically, under the assumption
-that judges exhibit no systematic bias? This question is especially
-relevant because researchers often treat ``sample-freeness'' as an
-inherent property of the BTL model \citep{Bramley_2008, Andrich_1978}
-despite growing evidence of persistent biases. Furthermore, if judges'
-biases reflect additional stimulus characteristics that should not
-influence assessment, to what extent can training or expertise mitigate
-these biases? \citep{Kelly_et_al_2022} More specifically, as discussed
-in Section~\ref{sec-theory-issue1b}, can training or expertise help
-prevent judges from prioritizing irrelevant features--such as
+paves the way for investigating several relevant research questions. One
+key question is whether researchers can validly analyze CJ data under
+the assumption of ``sample-free'' trait calibration, specifically under
+the hypothesis that judges exhibit no systematic bias. This question is
+particularly relevant because many researchers still regard
+``sample-freeness'' as an inherent property of the BTL model
+\citep{Bramley_2008, Andrich_1978} despite growing evidence of
+persistent biases. Another critical question, discussed in
+Section~\ref{sec-theory-issue1b}, is whether training or expertise can
+help judges avoid focusing on irrelevant stimulus features--such as
 handwriting--over more central criteria like argumentative quality in
-writing assessments? Addressing these questions may also shed light on
-what it truly means to be an ``expert'' in the CJ context
-\citep{Kelly_et_al_2022}.
-
-Moreover, if we accept that judges may prioritize different aspects of a
-stimulus, it follows that their assessments might also vary with
-individual characteristics such as gender, age, culture, income,
-education, training, or expertise
-\citep{Kelly_et_al_2022, Bartholomew_et_al_2020a}, and even from one
-individual to another
-\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
-Thus, building on the discussion in Section~\ref{sec-discussion_RA1},
-researchers could investigate how the selection of judges influences the
-formation of a ``shared consensus'' and whether these individual
-characteristics introduce systematic biases or distortions in the
-observed trait distribution \citep{Deffner_et_al_2022}. Furthermore, if
-such characteristics indeed undermine the assumption of
-``sample-freeness,'' then it becomes essential to explore strategies to
-control their influence and to determine how many judges--and how many
-judgments per judge--are needed to produce reliable trait estimates in
-these contexts. Additionally, it is worth considering whether
-\emph{repeated measures designs}, in which judges evaluate the same
-stimulus pairs multiple times \citep{Lawson_2015}, can improve judgment
-consistency and accuracy.
-
-\subsubsection{The identification of `misfitting' judges and
-stimuli}\label{sec-discussion_RA3}
-
-{About the identification of `misfitting' judges and stimuli}
-
-\newpage{}
-
-\subsubsection{The effects of hierarchies on experimental design and
-inference}\label{sec-discussion_RA4}
-
-{About the hierarchical structure of stimuli}
-
-\newpage{}
-
-\subsection{Limitations and challenges}\label{sec-discussion_challenges}
-
-Drawing conclusions about the real world from observed data always
-requires assumptions, regardless of whether the data are observational
-or experimental \citep{Kohler_et_al_2019, Deffner_et_al_2022}. This
-study is no exception to this principle. By adopting causal inference as
-the theoretical backbone of our proposed framework, we necessarily rely
-on expert knowledge and assumptions about the variables' causal
-structure that are often untestable at the outset
-\citep{Hernan_et_al_2025}. As such, our framework does not yield
-automatic answers. Instead, it encourages researchers to formulate
-precise questions and to explicitly articulate their assumptions to
-develop a generalizable understanding of the CJ system
+writing assessments \citep{Kelly_et_al_2022}. Notably, exploring these
+questions may also provide insights into what it truly means to be an
+``expert'' within the CJ context \citep{Kelly_et_al_2022}.
+
+Moreover, since judges draw on a variety of intricate stimulus
+characteristics when evaluating complex, multidimensional traits
+\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022},
+and these characteristics account for variation in judgment accuracy
+\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020, Gijsen_et_al_2021},
+it is reasonable to expect that judges' assessments also vary with
+individual attributes such as gender, age, culture, income, education,
+training, or expertise
+\citep{Kelly_et_al_2022, Bartholomew_et_al_2020a}. Thus, building on the
+discussion in Section~\ref{sec-discussion_RA1}, researchers could
+investigate how the selection of judges influences the formation of a
+``shared consensus'' and whether these individual attributes introduce
+systematic biases or distortions in the observed trait distribution
+\citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
+undermine the assumption of ``sample-freeness,'' then it becomes
+essential to explore strategies to control their influence and to
+determine how many judges--and how many judgments per judge--are needed
+to produce reliable trait estimates in these contexts. Additionally, it
+is worth considering whether \emph{repeated measures designs}, in which
+judges evaluate the same stimulus pairs multiple times
+\citep{Lawson_2015}, can improve judgment consistency and accuracy.
+
+\subsubsection{The identification of `misfitting' stimuli and
+judges}\label{sec-discussion_RA3}
+
+{LAST THING TO DO}
+
+\subsection{Study limitations and practical challenges for applied CJ
+researchers}\label{sec-discussion_challenges}
+
+The process of deriving conclusions from observed data always requires
+assumptions, whether the data are observational or experimental
+\citep{Kohler_et_al_2019, Deffner_et_al_2022}. Our approach is not an
+exception to this fundamental principle. As with all approaches grounded
+on causal inference, ours has the limitation of relying on expert
+knowledge and assumptions about the variables' causal structure that are
+often untestable at the outset \citep{Hernan_et_al_2025}. As such, the
+approach does not seek to yield automatic answers when applied to a
+given CJ dataset. Instead, it aims to encourage researchers to formulate
+precise questions and to make their assumptions explicit, fostering a
+generalizable understanding of the CJ system under study
 \citep{Rohrer_et_al_2022, Deffner_et_al_2022, Sterner_et_al_2024}. This
-clarity is essential because a model's ability to produce accurate
+clarity is critical because a model's ability to produce accurate
 estimates and valid inferences depends heavily on how well the data and
 inferential goals align with its underlying assumptions
-\citep{Kohler_et_al_2019}--a condition we cannot yet guarantee for our
-proposed models. Nevertheless, a key strength of this theory-driven
-approach is that it lays the foundation for empirically testing the
-proposed causal assumptions \citep{Deffner_et_al_2022}, as discussed in
-Section~\ref{sec-discussion_RA}.
-
-This theoretical commitment to causal inference also introduces several
-challenges that researchers must navigate--challenges common to many
-quantitative studies. These fall into two main categories. The first
-concerns the foundational knowledge researchers need to apply our
-proposed framework effectively. The second involves the increased
-attention researchers must now devote to their specific research
-context.
-
-\subsubsection{Foundational knowledge}\label{sec-discussion_challenges1}
-
-To effectively apply our framework, researchers must tackle two main
-challenges related to foundational knowledge. First, they need a solid
-understanding of causal inference principles. Second, they must learn to
-translate the functional and probabilistic content of conceptual models
-into bespoke statistical models. One example that illustrates the
-importance of a sound knowledge of causal inference is the recurrent
-assumption that predictor variables are ``relevant'' to the research
-context--interpreting this relevance as their inclusion in a
-\emph{sufficient adjustment set}
+\citep{Kohler_et_al_2019}. Although this alignment remains empirically
+untested for the proposed models, the theory-driven nature of our
+approach provides a solid foundation for future empirical evaluation of
+its causal assumptions \citep{Deffner_et_al_2022}, which are well
+supported by both theory and existing evidence (see
+Section~\ref{sec-theory-issues} and Section~\ref{sec-theoretical}).
+
+Conversely, this theoretical commitment to causal inference also
+introduces several practical challenges that applied CJ researchers must
+navigate. These fall into two main categories: first, acquiring the
+foundational knowledge necessary to apply the approach effectively; and
+second, dedicating greater attention to both conceptual and statistical
+modeling.
+
+\subsubsection{Required foundational
+knowledge}\label{sec-discussion_challenges1}
+
+To apply our approach effectively, CJ researchers must tackle two main
+challenges related to the required foundational knowledge. First, they
+need a solid understanding of causal inference principles. Second, they
+must learn to translate the functional and probabilistic content of
+conceptual models into bespoke statistical models. One example that
+illustrates the importance of a sound knowledge of causal inference is
+the recurrent assumption that predictor variables are ``relevant'' to
+the research context--interpreting this relevance as their inclusion in
+a \emph{sufficient adjustment set}
 \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
 not fully explore what this assumption entails or its implications for
-model specification, estimation, and inference. Consequently,
+model specification, estimation, and inference. As a result, CJ
 researchers must deeply engage with these and other complex ideas,
 including how SCMs encode functional and probabilistic information. To
-assist this effort, we include key references throughout the study to
-guide motivated readers toward a deeper understanding of these concepts
-\footnote{refer to footnote 1, 2, and the detailed online document
-  referenced in the Declarations section}.
+assist this effort, this study includes key references throughout the
+study to guide motivated aaplied CJ researchers toward a deeper
+understanding of these concepts \footnote{refer to footnote 1, 2, and
+  the detailed online document referenced in the Declarations section}.
 
@@
 statistical models also presents challenges. Although Bayesian inference
-methods offer a more accessible path for many applied researchers to
+methods offer a more accessible path for many applied CJ researchers to
 develop these skills--by reducing the need for specialized knowledge in
 areas like optimization theory, which frequentist methods often
-require--they still demand familiarity with other technical concepts
-such as probabilistic programming languages (PPLs), probability
-distributions, and convergence, which researchers may find hard to
-master. Thus, to support this skill development, we include links to the
-statistical model code and alternative specifications in the
-\textbf{Declarations} section of this document.
-
-\subsubsection{Attention to the research
-context}\label{sec-discussion_challenges2}
-
-Researchers face two additional challenges when applying our framework
-to their specific research context. First, they must verify that the
+require--they still demand a working knowledge of other technical
+concepts such as probabilistic programming languages (PPLs), probability
+distributions, and convergence. These requirements can be non-trivial to
+master. To support CJ researchers in developing these skills, this study
+provides a link to the statistical model code and alternative model
+specifications in the \textbf{Declarations} section of this document
+\footnote{Seminal texts on Bayesian inference methods, such as
+  \citet{Gelman_et_al_2014} and \citet{McElreath_2020}, offer valuable
+  support for developing a deeper understanding of these models.}.
+
+\subsubsection{Attention to conceptual and statistical
+modeling}\label{sec-discussion_challenges2}
+
+Even after acquiring the necessary foundational knowledge, CJ
+researchers face two additional challenges when applying our approach to
+their specific research context. First, they must verify whether the
 conceptual model provides a \emph{faithful} \footnote{Avoid confusing
@@
 treat our conceptual and statistical models as starting points rather
-than universal solutions for all CJ designs or datasets. This
-consideration means researchers must tackle the challenge of adapting
-these models to their specific research contexts, paying close attention
-to causal assumptions, experimental design, and model specification. Our
-approach supports this process by offering a transparent framework that
-helps researchers explicitly articulate these new assumptions and
-supports the design of CJ experiments that can yield reliable and valid
-inferences.
+than universal solutions for all CJ designs or datasets. While these
+models may prove adequate in some contexts, researchers cannot assume
+this apriori. Instead, they need to adapt these models to their specific
+CJ contexts, paying close attention to assessment design features and
+causal assumptions. As discussed in
+Section~\ref{sec-discussion_challenges}, our approach facilitates this
+process by offering a transparent framework for articulating new
+assumptions and guiding the design of CJ assessments.
 
@@
 Although performing identification analysis through formal derivations
-may appear to be a natural next step, we argue that the complexity of
-the CJ system renders such an approach impractical at the outset.
-Instead, simulation-based methods, such as power analysis, provide a
-more practical and flexible alternative, enabling researchers to examine
-the consistency of estimates without relying on cumbersome mathematical
+may seem like a natural next step, we argue that the complexity of the
+CJ system renders such an approach impractical at the outset. Instead,
+simulation-based methods, such as power analysis, provide a more
+practical and flexible alternative, enabling researchers to examine the
+consistency of estimates without relying on cumbersome mathematical
 proofs. Nonetheless, the development of formal identification proofs
 remains a significant goal for future CJ research. Importantly, our
-framework supports both approaches, as it provides the probabilistic
+approach supports both strategies by providing the probabilistic
 foundation for formal derivations and the necessary statistical

### 2025-06-06 paper2.qmd
section:6; process:write

@@
 First, it clarifies the interactions among all actors and processes involved in CJ assessments.
-Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.
+Second, it shifts the current comparative data analysis paradigm from passively accepting Case V and the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.
 
@@
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics, as seen in @Pollitt_2012a, @Pollitt_2012b, @vanDaal_et_al_2016, and @Goossens_et_al_2018, may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
+Furthermore, if researchers acknowledge that *misfit statistics* may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics may unintentionally discard valuable information [@Miller_2023], introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
 The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
@@
 
-Finally, similar to what we discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and may appear as outliers under the BTL model [@Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates.
+Finally, similar to what we discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and may appear as outliers under the BTL model [@Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates [@Miller_2023].
 The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
@@
 Thus, to address the limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods.
-The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG).
-This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.
+The approach begins with the development of a conceptual model, represented as a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
+This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes underlying the CJ system.
 
@@
 First, it clarifies the roles and interactions of all actors and processes involved in CJ assessments.
-Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data.
+Second, it shifts the analytic paradigm from passively accepting the assumptions of Case V and the BTL model to actively testing their fit with observed data.
 Together, these advantages establish a principled framework for evaluating best practices in designing CJ assessments, one that better aligns with the demands of contemporary CJ contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.
@@
 
-## Future research directions based on this study's approach {#sec-discussion_RA}
+## Future research directions using our approach {#sec-discussion_RA}
 <!-- recording: Sven and Tine 25.06.04; time: 00:05:40 - 00:07:25 -->
 
-Among the many potential directions for future research--both current and new--three avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates, as well as on the accuracy of statistical inferences.
-The following sections outline these avenues and explain how our approach provides a rigorous research framework to investigate them.
+Among the many potential directions for future research--both current and new--three avenues deserve particular attention due to their direct impact on the reliability and validity of CJ trait estimates, as well as on the accuracy of statistical inferences.
+The following sections outline these avenues and explain how our approach facilitates their investigation.
 
@@
 While these studies offer valuable insights, they often overlook the broader role these mechanisms play within the CJ system.
-Moreover, as this oversight likely stems from a broader lack of conceptual clarity about how these mechanisms function within the system, this study formally incorporates these mechanisms into CJ's conceptual model.
+As this oversight likely stems from a broader lack of conceptual clarity about how these mechanisms function within the system, this study decided to formally incorporate these mechanisms into CJ's conceptual model.
 
@@
 <!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
-However, many modern CJ applications rely on more complex assessment designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
+However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness — such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
 A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
-This outcome-informed pair selection suggests that the comparison mechanism is outcome-dependent, potentially classifying ACJ as a generator of MNAR data.
-This, in turn, facilitates a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities.
-While some studies report that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], others argue that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022], potentially as a result of outcome-dependent missingness.
+This pair selection process suggests that ACJ's comparison mechanism is outcome-dependent, potentially classifying the method as a generator of MNAR data.
+If this classification holds, the mixed findings on ACJ's capabilities become more comprehensible: some studies find that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others contend that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
 
@@
 <!-- 5. The importance of clarifying assumptions behind CJ assessment designs -->
-Nevertheless, regardless of the underlying missingness mechanisms, any CJ assessment design would benefit from explicitly defining its assumptions--a practice supported by our proposed approach.
+Nevertheless, regardless of the underlying missingness mechanisms, any CJ assessment design would benefit from explicitly defining its assumptions--a practice supported by our approach.
 This clarity enables researchers to evaluate how the sampling and comparison mechanisms affect trait estimation and statistical inference within each design.
@@
 <!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
-Moreover, since judges draw on a variety of intricate stimulus characteristics when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these characteristics account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that judges' assessments also vary with individual attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a].
-Thus, building on the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual attributes introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
-Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," then it becomes essential to explore strategies to control their influence and to determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts.
+Moreover, since judges draw on a variety of intricate stimulus characteristics when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these characteristics account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that judges' assessments also vary with attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022].
+Prior studies support this view [@Bartholomew_et_al_2020a; @McMahon_et_al_2015].
+Thus, building on the discussion in @sec-discussion_RA1, researchers could further explore how judges' selection influences the formation of a "shared consensus" and whether these attributes introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
+Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," researchers should explore ways to control their influence and determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts.
 Additionally, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
+Our approach provides the necessary structure to rigorously investigate these questions.
 
@@
 
-[LAST THING TO DO]{style="color:blue;"}
-
-<!-- Although the CJ literature clearly defines *misfit judges* as those whose assessments deviate markedly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], there has been little discussion about how these misfit judges relate to Thurstonian theory.
-In particular, it remains unclear which aspects of Thurstone's theory might explain the occurrence of misfit judges.
-To address this, we propose three statistical perspectives: (1) misfit judges are simply judges who exhibit bias; (2) misfit judges are those whose judgments display greater variability than their peers, and (2) misfit judges are indeed outliers.
--->
-
-<!-- Regarding the first perspective, @sec-theory-issue1b explains in detail how, due to the assumption of zero correlation, judges' biases can be miss classified as *misfits*.
-Meanwhile, @sec-discussion_RA2 discusses how our approach explicitly account for such biases, aiming to reduce the reliance on ad-hoc detection methods--such as misfit statistics--which are poorly suited to the task [@Kelly_et_al_2022].
--->
-
-<!-- Regarding the second perspective -->
-
-<!-- Conversely, regarding the second perspective, the very definition of *misfit judges* leads to their identification as outliers--that is, observations that deviate significantly from the rest of the sample in which they occur [@Grubbs_1969].
--->
+<!-- without disentangling all the potential processes the concept of misfits is strange, because it is not defined from Thustonian theory.
+As it is how is treated in CJ literature, misfit are outliers.
+People have theories on what makes misfits, but the approach now allows to test some hypothesis that might generate outliers under the BTL model, but no longer considered outliers in our approach  -->
+
+<!-- the approach makes it possible to have hypothesis about why judges are labeled as misfits.  -->
+
+Although the CJ literature clearly defines *misfit* judges and stimuli, researchers have rarely examined how these observations relate to Thurstonian theory.
+In particular, they have not identified which elements of Thurstone's theory might explain the occurrence of misfits.
+This disconnect likely stems from the fact that CJ researchers derive misfit statistics from methods such as residual analysis and outlier detection rather than from Thurstonian principles [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022].
+They typically define *misfit judges* as those whose assessments diverge significantly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and *misfit stimuli* as those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012a; @Pollitt_2012b; @Goossens_et_al_2018].
+Both definitions closely mirror the statistical concept of outliers--that is, observations that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969].
+However, this resemblance extends beyond these definitions.
+CJ researchers also identify misfits using standard outlier detection procedures, such as transforming BTL model residuals into statistics of interest and comparing them against a predefined threshold.
+
+<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
+
+<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
+
+<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, pp. 1] -->
+
+[After identifying misfit observations, researchers often propose various hypotheses to explain their presence.
+Yet, without providing evidence for any of these explanations, they typically exclude these observations from analysis.
+This exclusion process raises three major concerns: one related to the reliability of the methods used to identify misfits, a second concerning the potential impact of their removal on trait estimates and inferences, and a third regarding the hypotheses used to explain the presence and exclusion of misfits.]{style="color:blue;"}
+
+[First, regarding the reliability of identification methods, it is relevant to note that outlying observations can only be defined relative to a specific model [@McElreath_2020].
+Consequently, methods based on models such as the BTL--which rest on strong assumptions--should be treated with caution, as these assumptions may prevent the model from accurately capturing the underlying data-generating process of the CJ system, thereby potentially compromising the identification of misfits.]{style="color:blue;"}
+
+<!-- Furthermore, these points tend to be outliers with unlikely values, according to the model.
+As a result, they are highly influential and make it difficult to estimate out-of-sample predictive accuracy.
+Why?
+Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
+[@McElreath_2020, pp.
+231] -->
 
@@
 
-<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, pp. 1] -->
-
-<!-- In the previous subsection we argue that ad-hoc detection methods, such as 'misfit' statistics, could be poorly suited for the task of detecting judges bias [@Kelly_et_al_2022].
-This hinges in the assumption that the CJ data does have judges biases, but the method does not manages to identify them, leading to false negatives?
-or lack of sensitivity in the tests.
-However, there is a related problem which is: these ad-hoc method could be detecting judges' biases that are not there.
-as in @Wu_et_al_2022  -->
-
-<!-- These procedures also bear on a larger topic of discussion in social science.
-The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
-[@Wainer_et_al_1978, pp.
-332] -->
-
-<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
-If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
-...
-It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
-681] -->
-
-<!-- A similar point can be made for *misfit stimuli*--that is, those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018].
-It remains unclear which aspects of Thurstone's theory might explain the occurrence of misfit stimuli.
-To address this, we propose two statistical perspectives: (1) misfit stimuli are those whose discriminal processes exhibit greater variability compared to other stimuli, and (2) are statistical outliers.
--->
-
-<!-- Regarding the first perspective, @sec-theory-issue1a  -->
-
-<!-- , possibly due to larger discriminal dispersions or because they are genuine outliers--meaning, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, the BTL model neither accounts for these anomalies nor provides tools for addressing them, apart from excluding the "problematic" texts from analysis.
--->
-
-<!-- The same could be said for the stimuli.
-These misfit texts likely arise from two underlying mechanisms.
-First, a text may exhibit a larger discriminal dispersion, increasing the overlap between its discriminal process and those of multiple other texts.
-This greater overlap leads to more judgment discrepancies, increasing the likelihood of classifying the text as a misfit.
-@fig-discriminal_process illustrates this scenario.
-In the figure, Text B has a wider discriminal dispersion than Text A, enlarging the region where their discriminal processes intersect.
-This larger overlap represents an area where judges face greater difficulty in determining which text exhibits a higher level of the trait.
-Second, a text may represent an outlier --one with distinct characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-Such texts are harder to compare and place on the trait scale, resulting in more judgment discrepancies and a misfit classification.
-In either case, the BTL model is inherently agnostic to these mechanisms and does not provide any means to address them.
--->
-
-<!-- Indeed, the presence of the so-called *misfit* texts may already signal these limitations, as these are texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018], possibly due to larger discriminal dispersions or because they are genuine outliers--meaning, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, the BTL model neither accounts for these anomalies nor provides tools for addressing them, apart from excluding the "problematic" texts from analysis.
--->
-
-<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
-
-<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
-
-<!-- We advocate for the no removal of judges, but rather identify them and treat the data with the appropriate model, so these 'misfits' do not affect the measurement of the trait -->
-
-<!-- Fortunately, the statistical literature offers a solution to address violations of the equal dispersion assumption in CJ experiments.
-This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap.
-12].
--->
-
-<!-- Furthermore, these points tend to be outliers with unlikely values, according to the model.
-As a result, they are highly influential and make it difficult to estimate out-of-sample predictive accuracy.
-Why?
-Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
-[@McElreath_2020, pp.
-231] -->
+[Second, concerning the impact of removing misfits, the statistical literature warns that ad-hoc outlier exclusion procedures risk discarding valuable information [@Miller_2023] and introducing bias into trait estimates.
+The direction and magnitude of this bias are often unpredictable, as they depend on which observations researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].]{style="color:blue;"}
 
@@
 
-<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
-
 <!-- Consider, for example, the simplest location and scale problem-inference for a normal distribution with unknown mean \theta and variance \sigma^{2}.
@@
 
+<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
+If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
+...
+It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
+681] -->
+
+[Lastly, regarding the hypotheses explaining the presence and exclusion of misfits, it is important to note that a significant number of these hypotheses could not be tested within the BTL model due to its strong assumptions and limited flexibility.
+This is the reason why there is a lack of supporting evidence for misfit observations.
+In contrast, our approach enables researchers to test these hypotheses.
+For instance, researchers could examine whether misfit judges are simply those who exhibit (an outlying degree of) systematic bias or whether they display greater variability in their judgments compared to their peers.
+Similarly, in the case of misfit stimuli, researchers could investigate whether their status reflects an increased variability in their discriminal processes or stems from their classification as outlying stimuli.]{style="color:blue;"}
+
+[Moreover, since outliers are not necessarily "bad data" [@McElreath_2020], our approach offers a principled alternative that retains misfits in the analysis without compromising trait estimation.
+This alternative involves adapting our proposed models into more sophisticated, robust measurement models [@Wainer_et_al_1978]--a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].]{style="color:blue;"}
+
 <!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
 
-<!-- Second, and complementary to the previous point, the statistical literature consistently highlights the risks of using ad hoc procedures to identify "outliers" when the model does not accurately represent the data-generating process.
-For instance, McElreath (2020, pp.
-232) expressed that: "There is a tradition of dropping outliers.
-People sometimes drop outliers even before a model is fit, based only on standard deviations from the mean outcome value.
-You should never do that--a point can only be unexpected and highly influential in light of a model.
-After you fit a model, the picture changes.
-If there are only a few outliers, and you are sure to report results both with and without them, dropping outliers might be okay.
-But if there are several outliers, we really need to model them".
-Similarly, O'Hagan (2018, pp.
-366) makes a small example explaining how the average of a parameter--much like a trait score you could get from the BTL model--depends on outliers.
-Miller (2023), in psychology, explores a similar topic.
--->
-
-<!-- You might notice I use the term "outlier" here to describe biased judges.
-This is because, like statistical outliers, these judges deviate significantly from the rest of the sample in which they occur (Grubbs, 1969).
-However, the key reason for using this terminology lies in that any ad hoc method for flagging biased judges based on the BTL model--such as misfit statistics--effectively acts as an outlier detection tool.
-This is due to the fact that the BTL model relies on several strong assumptions, which limit its capacity to capture other features of the data unless supplemented by additional procedures--procedures that, in essence, resemble outlier detection.
--->
-
-<!-- Moreover, from McElreath's text, you might recognize that the practice of removing outliers mirrors the act of discarding biased judges in CJ using misfit statistics, where these statistics are essentially transformations of residual probabilities linked to trait estimates, or as McElreath puts it "standard deviations from the mean outcome values." But what are the potential consequences of wrongfully identifying and discarding misfit judges, you might ask?.
-Well, this practice risks removing valuable information and introducing bias into trait estimates [Zimmerman, 1994; O'Hagan, 2018; McElreath, 2020].
-The direction and magnitude of these biases are unpredictable, depending on which judges are excluded from the analysis.
--->
+<!-- These procedures also bear on a larger topic of discussion in social science.
+The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
+[@Wainer_et_al_1978, pp.
+332] -->
 
@@
 <!-- 1. Challenges for the approach: background knowledge and research context -->
-Conversely, this theoretical commitment to causal inference also introduces several practical challenges that applied CJ researchers must navigate.
+Moreover, this theoretical commitment to causal inference also introduces several practical challenges that applied CJ researchers must navigate.
 These fall into two main categories: first, acquiring the foundational knowledge necessary to apply the approach effectively; and second, dedicating greater attention to both conceptual and statistical modeling.

### 2025-06-06 _tex/paper2.tex
Built site for gh-pages

@@
 CJ assessments. Second, it shifts the current comparative data analysis
-paradigm from passively accepting the BTL model assumptions to actively
-testing whether those assumptions fit the data under analysis.
+paradigm from passively accepting Case V and the BTL model assumptions
+to actively testing whether those assumptions fit the data under
+analysis.
 
@@
 \(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
-acknowledge that misfit statistics may represent texts with different
-dispersions or outlying observations, the common CJ practice of
-excluding stimuli based on these statistics, as seen in
-\citet{Pollitt_2012a}, \citet{Pollitt_2012b},
-\citet{vanDaal_et_al_2016}, and \citet{Goossens_et_al_2018}, may
-unintentionally discard valuable information, introducing bias into the
-trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
-and magnitude of these biases remain unpredictable, as they depend on
-which stimuli researchers exclude from the analysis.
+acknowledge that \emph{misfit statistics} may represent texts with
+different dispersions or outlying observations, the common CJ practice
+of excluding stimuli based on these statistics may unintentionally
+discard valuable information \citep{Miller_2023}, introducing bias into
+the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
+direction and magnitude of these biases remain unpredictable, as they
+depend on which stimuli researchers exclude from the analysis.
 
@@
 \citep{Wu_et_al_2022}--risk discarding valuable information and
-introducing bias into trait estimates. The direction and magnitude of
-these biases remain unpredictable, as they depend on which judges
-researchers exclude from the analysis
+introducing bias into trait estimates \citep{Miller_2023}. The direction
+and magnitude of these biases remain unpredictable, as they depend on
+which judges researchers exclude from the analysis
 \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
@@
 integrated approach that combines causal and Bayesian inference methods.
-The approach begins by formulating a conceptual model represented by a
-Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This
+The approach begins with the development of a conceptual model,
+represented as a Structural Causal Model (SCM) and a Directed Acyclic
+Graph (DAG)
+\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. This
 model integrates Thurstone's core theoretical principles, such as the
@@
 features, including judges' bias, sampling procedures, and comparison
-mechanisms, thereby disentangling the causal processes in the CJ system.
+mechanisms, thereby disentangling the causal processes underlying the CJ
+system.
 
@@
 and processes involved in CJ assessments. Second, it shifts the analytic
-paradigm from passively accepting the assumptions of the BTL model to
-actively testing their fit with observed data. Together, these
+paradigm from passively accepting the assumptions of Case V and the BTL
+model to actively testing their fit with observed data. Together, these
 advantages establish a principled framework for evaluating best
@@
 
-\subsection{Future research directions based on this study's
+\subsection{Future research directions using our
 approach}\label{sec-discussion_RA}
@@
 and new--three avenues deserve particular attention due to their direct
-impact on the reliability and validity of trait estimates, as well as on
-the accuracy of statistical inferences. The following sections outline
-these avenues and explain how our approach provides a rigorous research
-framework to investigate them.
+impact on the reliability and validity of CJ trait estimates, as well as
+on the accuracy of statistical inferences. The following sections
+outline these avenues and explain how our approach facilitates their
+investigation.
 
@@
 these studies offer valuable insights, they often overlook the broader
-role these mechanisms play within the CJ system. Moreover, as this
-oversight likely stems from a broader lack of conceptual clarity about
-how these mechanisms function within the system, this study formally
-incorporates these mechanisms into CJ's conceptual model.
+role these mechanisms play within the CJ system. As this oversight
+likely stems from a broader lack of conceptual clarity about how these
+mechanisms function within the system, this study decided to formally
+incorporate these mechanisms into CJ's conceptual model.
 
@@
 However, many modern CJ applications rely on more complex assessment
-designs in which the sampling and comparison mechanisms introduce more
-intricate forms of missingness--such as \emph{missing at random} (MAR)
-or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
-prominent example is the previously discussed ACJ designs, where prior
+designs, in which the sampling and comparison mechanisms introduce more
+intricate forms of missingness --- such as \emph{missing at random}
+(MAR) or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}.
+A prominent example is the previously discussed ACJ designs, where prior
 judgment outcomes inform the selection of stimulus pairs for subsequent
 comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
-outcome-informed pair selection suggests that the comparison mechanism
-is outcome-dependent, potentially classifying ACJ as a generator of MNAR
-data. This, in turn, facilitates a more nuanced understanding of the
-mixed evidence surrounding ACJ's capabilities. While some studies report
-that the method improves trait reliability
-\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
-that it artificially inflates these gains
-\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022},
-potentially as a result of outcome-dependent missingness.
+pair selection process suggests that ACJ's comparison mechanism is
+outcome-dependent, potentially classifying the method as a generator of
+MNAR data. If this classification holds, the mixed findings on ACJ's
+capabilities become more comprehensible: some studies find that the
+method improves trait reliability
+\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
+contend that it artificially inflates these gains
+\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
 
@@
 CJ assessment design would benefit from explicitly defining its
-assumptions--a practice supported by our proposed approach. This clarity
-enables researchers to evaluate how the sampling and comparison
-mechanisms affect trait estimation and statistical inference within each
-design. Such assessments are particularly relevant given the common
+assumptions--a practice supported by our approach. This clarity enables
+researchers to evaluate how the sampling and comparison mechanisms
+affect trait estimation and statistical inference within each design.
+Such assessments are particularly relevant given the common
 misconception in the CJ literature that Thurstone's model can naturally
@@
 it is reasonable to expect that judges' assessments also vary with
-individual attributes such as gender, age, culture, income, education,
-training, or expertise
-\citep{Kelly_et_al_2022, Bartholomew_et_al_2020a}. Thus, building on the
-discussion in Section~\ref{sec-discussion_RA1}, researchers could
-investigate how the selection of judges influences the formation of a
-``shared consensus'' and whether these individual attributes introduce
-systematic biases or distortions in the observed trait distribution
+attributes such as gender, age, culture, income, education, training, or
+expertise \citep{Kelly_et_al_2022}. Prior studies support this view
+\citep{Bartholomew_et_al_2020a, McMahon_et_al_2015}. Thus, building on
+the discussion in Section~\ref{sec-discussion_RA1}, researchers could
+further explore how judges' selection influences the formation of a
+``shared consensus'' and whether these attributes introduce systematic
+biases or distortions in the observed trait distribution
 \citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
-undermine the assumption of ``sample-freeness,'' then it becomes
-essential to explore strategies to control their influence and to
-determine how many judges--and how many judgments per judge--are needed
-to produce reliable trait estimates in these contexts. Additionally, it
-is worth considering whether \emph{repeated measures designs}, in which
-judges evaluate the same stimulus pairs multiple times
-\citep{Lawson_2015}, can improve judgment consistency and accuracy.
+undermine the assumption of ``sample-freeness,'' researchers should
+explore ways to control their influence and determine how many
+judges--and how many judgments per judge--are needed to produce reliable
+trait estimates in these contexts. Additionally, it is worth considering
+whether \emph{repeated measures designs}, in which judges evaluate the
+same stimulus pairs multiple times \citep{Lawson_2015}, can improve
+judgment consistency and accuracy. Our approach provides the necessary
+structure to rigorously investigate these questions.
 
@@
 
-{LAST THING TO DO}
+Although the CJ literature clearly defines \emph{misfit} judges and
+stimuli, researchers have rarely examined how these observations relate
+to Thurstonian theory. In particular, they have not identified which
+elements of Thurstone's theory might explain the occurrence of misfits.
+This disconnect likely stems from the fact that CJ researchers derive
+misfit statistics from methods such as residual analysis and outlier
+detection rather than from Thurstonian principles
+\citep{Pollitt_2012a, Pollitt_2012b, Wu_et_al_2022}. They typically
+define \emph{misfit judges} as those whose assessments diverge
+significantly from the ``shared consensus''
+\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
+and \emph{misfit stimuli} as those that elicit more judgment
+discrepancies than others
+\citep{Pollitt_2004, Pollitt_2012a, Pollitt_2012b, Goossens_et_al_2018}.
+Both definitions closely mirror the statistical concept of
+outliers--that is, observations that deviate markedly from the rest of
+the sample in which they occur \citep{Grubbs_1969}. However, this
+resemblance extends beyond these definitions. CJ researchers also
+identify misfits using standard outlier detection procedures, such as
+transforming BTL model residuals into statistics of interest and
+comparing them against a predefined threshold.
+
+{After identifying misfit observations, researchers often propose
+various hypotheses to explain their presence. Yet, without providing
+evidence for any of these explanations, they typically exclude these
+observations from analysis. This exclusion process raises three major
+concerns: one related to the reliability of the methods used to identify
+misfits, a second concerning the potential impact of their removal on
+trait estimates and inferences, and a third regarding the hypotheses
+used to explain the presence and exclusion of misfits.}
+
+{First, regarding the reliability of identification methods, it is
+relevant to note that outlying observations can only be defined relative
+to a specific model \citep{McElreath_2020}. Consequently, methods based
+on models such as the BTL--which rest on strong assumptions--should be
+treated with caution, as these assumptions may prevent the model from
+accurately capturing the underlying data-generating process of the CJ
+system, thereby potentially compromising the identification of misfits.}
+
+{Second, concerning the impact of removing misfits, the statistical
+literature warns that ad-hoc outlier exclusion procedures risk
+discarding valuable information \citep{Miller_2023} and introducing bias
+into trait estimates. The direction and magnitude of this bias are often
+unpredictable, as they depend on which observations researchers exclude
+from the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.}
+
+{Lastly, regarding the hypotheses explaining the presence and exclusion
+of misfits, it is important to note that a significant number of these
+hypotheses could not be tested within the BTL model due to its strong
+assumptions and limited flexibility. This is the reason why there is a
+lack of supporting evidence for misfit observations. In contrast, our
+approach enables researchers to test these hypotheses. For instance,
+researchers could examine whether misfit judges are simply those who
+exhibit (an outlying degree of) systematic bias or whether they display
+greater variability in their judgments compared to their peers.
+Similarly, in the case of misfit stimuli, researchers could investigate
+whether their status reflects an increased variability in their
+discriminal processes or stems from their classification as outlying
+stimuli.}
+
+{Moreover, since outliers are not necessarily ``bad data''
+\citep{McElreath_2020}, our approach offers a principled alternative
+that retains misfits in the analysis without compromising trait
+estimation. This alternative involves adapting our proposed models into
+more sophisticated, robust measurement models
+\citep{Wainer_et_al_1978}--a broad class of procedures designed to
+reduce the sensitivity of parameter estimates to mild or moderate
+departures from model assumptions \citep{Everitt_et_al_2010}.}
 
@@
 
-Conversely, this theoretical commitment to causal inference also
+Moreover, this theoretical commitment to causal inference also
 introduces several practical challenges that applied CJ researchers must

### 2025-06-07 paper2.qmd
section:6; process:write

@@
 
-Among the many potential directions for future research--both current and new--three avenues deserve particular attention due to their direct impact on the reliability and validity of CJ trait estimates, as well as on the accuracy of statistical inferences.
+Among the many potential directions for future research, three avenues deserve particular attention due to their direct impact on the reliability and validity of CJ trait estimates, as well as on the accuracy of statistical inferences.
 The following sections outline these avenues and explain how our approach facilitates their investigation.
@@
 <!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
-Although sampling and comparison mechanisms are central to modern CJ assessments, it is striking that most of the CJ literature has examined them only within a limited scope.
+Although sampling and comparison mechanisms are central to modern CJ assessments, it is striking that most CJ literature has examined them within a limited scope.
 Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022].
-While these studies offer valuable insights, they often overlook the broader role these mechanisms play within the CJ system.
-As this oversight likely stems from a broader lack of conceptual clarity about how these mechanisms function within the system, this study decided to formally incorporate these mechanisms into CJ's conceptual model.
+While these studies offer valuable insights, they also overlook the broader role that these mechanisms play within the CJ system.
+As this oversight likely stems from a more fundamental lack of conceptual clarity about how these mechanisms function within the system, this study decided to integrate these mechanisms into the conceptual model of CJ.
 
@@
 The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process.
-In particular, it clarifies their role in the data-generating process by framing them as sources of missing data--that is, as mechanisms that determine which observations are missing from the final data sample.
-This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and assessing more complex assessments setups.
+Specifically, it clarifies their role as sources of missing data in CJ's data-generating process--that is, as mechanisms that determine which observations are missing from the final data sample.
+This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], allowing a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex assessments setups.
 
@@
 <!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
-However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness — such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
+However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
 A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
@@
 <!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
-Despite the growing notion that various stimulus-related factors can influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
-This gap likely persists not due to a lack of interest or research but because researchers often rely on ad-hoc detection methods--such as 'misfit' statistics--that are poorly suited for the task [@Kelly_et_al_2022].
+Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
+This gap likely persists not due to a lack of interest or research but because researchers often rely on ad-hoc detection methods, such as 'misfit' statistics, that may not be well-suited for the task [@Kelly_et_al_2022].
 To address this limitation, this study treats judges' biases as an integral component of the CJ system from the outset.
@@
 This question is particularly relevant because many researchers still regard "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases.
-Another critical question, discussed in @sec-theory-issue1b, is whether training or expertise can help judges avoid focusing on irrelevant stimulus features--such as handwriting--over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022].
+Another critical question, discussed in @sec-theory-issue1b, is whether training or expertise can help judges avoid focusing on irrelevant stimulus features, such as handwriting, over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022].
 Notably, exploring these questions may also provide insights into what it truly means to be an "expert" within the CJ context [@Kelly_et_al_2022].
@@
 <!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
-Moreover, since judges draw on a variety of intricate stimulus characteristics when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these characteristics account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that judges' assessments also vary with attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022].
+Moreover, since judges rely on these stimulus-related factors when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these factors account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that assessments also vary according to judge-specific attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022].
 Prior studies support this view [@Bartholomew_et_al_2020a; @McMahon_et_al_2015].
 Thus, building on the discussion in @sec-discussion_RA1, researchers could further explore how judges' selection influences the formation of a "shared consensus" and whether these attributes introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
-Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," researchers should explore ways to control their influence and determine how many judges--and how many judgments per judge--are needed to produce reliable trait estimates in these contexts.
-Additionally, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
-Our approach provides the necessary structure to rigorously investigate these questions.
+Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," it becomes essential to explore strategies for mitigating their influence and to determine how many judges (and how many judgments per judge) are needed to produce reliable trait estimates under these conditions.
+In addition, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
+As anticipated, our approach provides the necessary structure to investigate rigorously these questions.
 
@@
 
-### The identification of 'misfitting' stimuli and judges {#sec-discussion_RA3}
+### The identification of 'misfitting' judges and stimuli {#sec-discussion_RA3}
 <!-- recording: Sven and Tine 25.05.21; time: 00:31:45 - 00:32:20 -->
@@
 
-<!-- without disentangling all the potential processes the concept of misfits is strange, because it is not defined from Thustonian theory.
-As it is how is treated in CJ literature, misfit are outliers.
-People have theories on what makes misfits, but the approach now allows to test some hypothesis that might generate outliers under the BTL model, but no longer considered outliers in our approach  -->
-
-<!-- the approach makes it possible to have hypothesis about why judges are labeled as misfits.  -->
-
-Although the CJ literature clearly defines *misfit* judges and stimuli, researchers have rarely examined how these observations relate to Thurstonian theory.
-In particular, they have not identified which elements of Thurstone's theory might explain the occurrence of misfits.
-This disconnect likely stems from the fact that CJ researchers derive misfit statistics from methods such as residual analysis and outlier detection rather than from Thurstonian principles [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022].
-They typically define *misfit judges* as those whose assessments diverge significantly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and *misfit stimuli* as those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012a; @Pollitt_2012b; @Goossens_et_al_2018].
+Although the CJ literature clearly defines *misfit* judges and stimuli, CJ researchers have rarely examined how these observations relate to Thurstonian theory.
+In particular, they have not identified which elements of Thurstone's theory might account for the occurrence of misfits.
+This disconnect likely stems from the fact that CJ researchers derive misfit statistics from residual analysis and outlier detection methods rather than from Thurstonian principles.
+Specifically, *misfit judges* are typically defined as those whose assessments diverge significantly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], while *misfit stimuli* are those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012a; @Pollitt_2012b; @Goossens_et_al_2018].
 Both definitions closely mirror the statistical concept of outliers--that is, observations that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969].
-However, this resemblance extends beyond these definitions.
-CJ researchers also identify misfits using standard outlier detection procedures, such as transforming BTL model residuals into statistics of interest and comparing them against a predefined threshold.
+But this resemblance extends beyond the definitions themselves, as CJ researchers often identify misfits using conventional outlier detection procedures, such as transforming BTL model residuals into diagnostic statistics and comparing them against predefined thresholds [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022].
 
@@
 
-[After identifying misfit observations, researchers often propose various hypotheses to explain their presence.
-Yet, without providing evidence for any of these explanations, they typically exclude these observations from analysis.
-This exclusion process raises three major concerns: one related to the reliability of the methods used to identify misfits, a second concerning the potential impact of their removal on trait estimates and inferences, and a third regarding the hypotheses used to explain the presence and exclusion of misfits.]{style="color:blue;"}
-
-[First, regarding the reliability of identification methods, it is relevant to note that outlying observations can only be defined relative to a specific model [@McElreath_2020].
-Consequently, methods based on models such as the BTL--which rest on strong assumptions--should be treated with caution, as these assumptions may prevent the model from accurately capturing the underlying data-generating process of the CJ system, thereby potentially compromising the identification of misfits.]{style="color:blue;"}
-
-<!-- Furthermore, these points tend to be outliers with unlikely values, according to the model.
-As a result, they are highly influential and make it difficult to estimate out-of-sample predictive accuracy.
-Why?
-Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
-[@McElreath_2020, pp.
-231] -->
+Nevertheless, it is not the classification of misfits as outliers that raises concerns for trait measurement and inference, but rather the prevailing CJ practice of identifying them through these ad-hoc detection procedures and then excluding them from analysis, often without providing empirical evidence for the various hypotheses proposed to explain their occurrence in the first place.
+In this regard, it is essential to recognize that outliers can only be defined relative to a specific model [@McElreath_2020].
+Thus, identification procedures based on models like the BTL, which relies on strong assumptions, should be approached with caution, as they may not be well-suited for the task [@Kelly_et_al_2022].
+Specifically, because the BTL model may not always accurately reflect the underlying data-generating process of the CJ system (see @sec-theory-issues), detection procedures grounded in this model risk compromising the accurate identification of misfits.
+Moreover, the statistical literature cautions that excluding outliers risks discarding valuable information [@Miller_2023] and introducing bias into trait estimates.
+The direction and magnitude of this bias are often unpredictable, as they depend on which observations researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
+Lastly, regarding the hypotheses proposed to explain the presence of misfits, it is evident that the BTL model lacks the flexibility needed to adequately test many of them, largely due to its strong assumptions.
 
@@
 683] -->
-
-[Second, concerning the impact of removing misfits, the statistical literature warns that ad-hoc outlier exclusion procedures risk discarding valuable information [@Miller_2023] and introducing bias into trait estimates.
-The direction and magnitude of this bias are often unpredictable, as they depend on which observations researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].]{style="color:blue;"}
 
@@
 
-[Lastly, regarding the hypotheses explaining the presence and exclusion of misfits, it is important to note that a significant number of these hypotheses could not be tested within the BTL model due to its strong assumptions and limited flexibility.
-This is the reason why there is a lack of supporting evidence for misfit observations.
-In contrast, our approach enables researchers to test these hypotheses.
-For instance, researchers could examine whether misfit judges are simply those who exhibit (an outlying degree of) systematic bias or whether they display greater variability in their judgments compared to their peers.
-Similarly, in the case of misfit stimuli, researchers could investigate whether their status reflects an increased variability in their discriminal processes or stems from their classification as outlying stimuli.]{style="color:blue;"}
-
-[Moreover, since outliers are not necessarily "bad data" [@McElreath_2020], our approach offers a principled alternative that retains misfits in the analysis without compromising trait estimation.
-This alternative involves adapting our proposed models into more sophisticated, robust measurement models [@Wainer_et_al_1978]--a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].]{style="color:blue;"}
+In contrast, our approach provides a rigorous framework that enables the examination of several relevant hypotheses.
+For instance, researchers could investigate whether misfit judges are those who exhibit (an outlying degree of) systematic bias or greater variability in their judgments compared to their peers.
+Similarly, researchers might explore whether misfit stimuli exhibit more variable discriminal processes relative to other stimuli or if they are genuinely outlying cases.
+Moreover, since outliers are not necessarily "bad data" [@McElreath_2020], our approach also offers a principled alternative that retains misfits in the analysis without compromising trait estimation or inference.
+This alternative involves adapting our proposed models into robust measurement models [@McElreath_2020]--a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].
+This alternative also relates to a broader discussion in social science, which is: that when researchers face low predictive capacity from their models, they should not only search for "new" variables or alternative procedures but also consider employing more sophisticated measurement models [@Wainer_et_al_1978].
 
@@
 
-Even after acquiring the necessary foundational knowledge, CJ researchers face two additional challenges when applying our approach to their specific research context.
+Even after acquiring the necessary foundational knowledge, CJ researchers still face two additional challenges when applying our approach to their specific research context.
 First, they must verify whether the conceptual model provides a *faithful* [^12] representation of the CJ system under study.
@@
 While these models may prove adequate in some contexts, researchers cannot assume this apriori.
-Instead, they need to adapt these models to their specific CJ contexts, paying close attention to assessment design features and causal assumptions.
+Instead, they need to adapt these models to their specific contexts, paying close attention to assessment design features and causal assumptions.
 As discussed in @sec-discussion_challenges, our approach facilitates this process by offering a transparent framework for articulating new assumptions and guiding the design of CJ assessments.
@@
 Without identification, consistency is impossible--even with infinite, error-free data--and thus, meaningful inference from finite samples cannot be achieved [@Schuessler_et_al_2023].
-
-Although performing identification analysis through formal derivations may seem like a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset.
+Although performing identification analysis through formal derivations may seem like a natural next step, the complexity of the CJ system renders such an approach impractical at the outset.
 Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs.

### 2025-06-07 paper2.qmd
1st submission version

@@
 
-# Appendix {#sec-appendix .appendix appendix-style=plain}
-
-## Statistical and Causal inference {#sec-appendixB .appendix appendix-style=plain}
-
-<!-- recording: Sven and Tine 25.01.22; time: 00:00:00 - 00:05:40 -->
-
-This section introduces fundamental statistical and causal inference concepts necessary for understanding the core theoretical principles described in this document.
-It does not, however, offer a comprehensive overview of causal inference methods.
-Readers seeking more in-depth understanding may wish to explore introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020.
-They may also find it helpful to consult introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020.
-For more advanced study, readers may refer to seminal intermediate papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2025.
-
-### Empirical research and randomized experiments {#sec-appendixB1 .appendix appendix-style=plain}
-
-<!-- 1. What does empirical research do first? -->
-Empirical research uses evidence from observation and experimentation to address real-world challenges.
-In this context, researchers typically formulate their research questions as *estimands* or *targets of inference*, i.e., the specific quantities they seek to determine [@Everitt_et_al_2010].
-For instance, researchers might be interested in answering the following question: "To what extent do different teaching methods $(T)$ influence students' ability to produce high-quality written texts $(Y)$?" To investigate this, researchers could randomly assign students to two groups, each exposed to a different teaching method $(T_{i} = \{1,2\})$.
-Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability.
-In this scenario, the research question can be rephrased as the estimand, "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?" and this estimand can be mathematically represented by the random associational quantity in @eq-group_diff, where $E[\cdot]$ denotes the expected value.
-
-$$
-E[Y_{i} \mid T_{i}=1] - E[Y_{i} \mid T_{i}=2]
-$$ {#eq-group_diff}
-
-<!-- 2. What do they do next? -->
-Researchers then proceed to identify the estimands.
-*Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
-4].
-An *estimator* refers to a method or function that transforms data into an estimate [@Neal_2020].
-*Estimates* are numerical values that approximate the estimand derived through the process of *estimation*, which integrates data with an estimator [@Everitt_et_al_2010].
-The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], shown in @fig-IEflow, visually represents the transition from estimands to estimates.
-
-<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
-Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
-4].
--->
-
-![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}
-
-<!-- 3. the importance of identification -->
-Identification is a necessary condition to ensure *consistent* estimators.
-An estimator achieves *consistency* when it converges to the "true" value of an estimand as the data size approaches infinity [@Everitt_et_al_2010].
-Without identification, researchers cannot achieve consistency, even with "infinite" and error-free data.
-As a result, deriving meaningful insights about an estimand from finite data becomes impossible [@Schuessler_et_al_2023, pp.
-5].
-Therefore, to ensure accurate and reliable estimates, researchers prioritize estimators with desirable identification properties.
-For instance, the Z-test is a widely used estimator for comparing group proportions, yielding accurate estimates when its underlying assumptions are satisfied [@Kanji_2006].
-Furthermore, researchers can interpret estimates from the Z-test as causal, provided the data is collected through a randomized experiment.
-
-<!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
-So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
-5].
--->
-
-<!-- If both assumptions are met, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$.
-The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \mid T_{i}=1]$ and $E[Y_{i} \mid T_{i}=2]$, respectively.
-The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.
--->
-
-<!-- the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the comparison of two groups, as long as the assumptions of the model are met.
-Usually these estimates are expressed in terms of the proportions of groups  -->
-
-<!-- However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply estimate associations. -->
-
-<!-- 4. Why does causality stem from randomize experiments? -->
-Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014].
-This recognition stems from their ability to enable researchers interpret associational estimates as causal.
-They achieve this by ensuring data, and by extension an estimator, satisfies several key identification properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020].
-The most critical property, however, is the elimination of confounding.
-*Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010].
-Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from any other variable $X$ [@Morgan_et_al_2014; @Neal_2020].
-
-<!-- 5. Experiments are not available to everyone, then causal inference comes to the rescue -->
-Nevertheless, researchers often face constraints that limit their ability to conduct randomized experiments.
-These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020].
-In these cases, causal inference offers a valuable alternative for generating causal estimates and understanding the mechanisms underlying specific data.
-In addition, the framework can provide significant theoretical insights that can enhance the design of experimental and observational studies [@McElreath_2020].
-
-### Identification under causal inference {#sec-appendixB2 .appendix appendix-style=plain}
-
-<!-- 1. What is causal inference? -->
-Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, the *causal inference* framework is designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020].
-The framework uses rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" This question introduces the concept of counterfactuals, which are instrumental in defining and identifying causal effects.
-
-<!-- 2. What are counterfactuals? -->
-*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].
-The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals.
-For instance, in the scenario described in @sec-appendixB1, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes, as in @eq-ICE.
-
-$$
-\tau_{i} = Y_{i} \mid do(T_{i}=1) - Y_{i} \mid do(T_{i}=2)
-$$ {#eq-ICE}
-
-where $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \mid do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \mid do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$.
-Here, an *intervention* involves assigning a constant value to the treatment variable for each student's potential outcomes.
-Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable.
-To address this challenge, the structural approach extends the ICE to the *average causal effect* (ACE, @eq-ACE), representing the average difference between the students' observed potential outcomes and their counterfactual counterparts.
-
-$$
-\begin{aligned}
-\tau & = E[\tau_{i}] \\
-  & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)]
-\end{aligned}
-$$ {#eq-ACE}
-
-<!-- [^1]: The potential outcomes approach [@Neyman_et_al_1923; @Rubin_1974] defines the ICE as $\tau_{i} = Y^{1}_{i} - Y^{2}_{i}$, while it defines the ACE as $\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]$.
--->
-
-<!-- 3. But how does it work? -->
-Even though counterfactuals are unobservable, researchers can still identify the ACE from associational estimates by leveraging the structural approach.
-The approach identifies the ACE by statistically conditioning data on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones.
-When such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \mid X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
-Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
-32] (see @eq-CACE).
-
-<!-- Conditioning on T=t just means that we are restricting our focus to the subset of the population to those who received treatment t.
-In contrast, an intervention would be to take the whole population and give everyone treatment t [@Neal_2020, pp.
-32].
--->
-
-Conditioning on a sufficient adjustment set enables researchers to estimate the ACE, even when the data comes from an observational study.
-This process is feasible because such conditioning ensures that the ACE estimator satisfies several critical properties, including confounding elimination [@Morgan_et_al_2014].
-Naturally, the validity of claims about the causal effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set.
-However, as @Kohler_et_al_2019 [pp.
-150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions.
-This requirement holds true for both observational and experimental data.
-
-<!-- Statements about the real world from observed data require assumptions.
-This is true no matter what kind of data we have, and it is explicitly also true for statements based on data stemming from SRSg.
-The validity of our statements therefore critically hinges on the validity of those assumptions.
-If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
-[@Kohler_et_al_2019, pp.
-150] -->
-
-For instance, if researchers cannot conduct the randomized experiments described in @sec-appendixB1 and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion.
-Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)
-
-$$
-CACE_{t} = E[Y_{i} \mid T_{i}=t, X]
-$$ {#eq-CACE}
-
-From the CACE, researchers can identify the ACE from associational quantities as in @eq-mACE1.
-This identification process is commonly known as the *backdoor adjustment*.
-Here, $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].
-
-$$
-\begin{aligned}
-  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
-  & = E_{X}[CACE_{1} - CACE_{2}] \\
-  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right]
-\end{aligned}
-$$ {#eq-mACE1}
-
-Notably, the approach extends the ACE identification for a continuous variable $T$ as in @eq-mACE_cont, ensuring broad applicability across different causal scenarios [@Neal_2020, pp. 45]
-
-$$
-\begin{aligned}
-  \tau &= E[Y_{i} \mid do(T_{i}=t)] \\
-  & = d E_{X}\left[ E[Y_{i} \mid T_{i}=t, X]\right]/ dt
-  \end{aligned}
-$$ {#eq-mACE_cont}
-
-### Diving into the specifics {#sec-appendixB3 .appendix appendix-style=plain}
-
-<!-- 1. The benefits of SCMs and DAGs -->
-The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-Essentially, these tools serve as *conceptual (theoretical) models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
-4].
-Thus, using these tools, researchers can determine which statistical models can identify (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020], thus enabling valid causal inference.
-@fig-IEflow shows the role of theoretical models in the inference process.
-
-<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
-Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
-4].
--->
-
-<!-- Causal diagrams are a simple way to encode our subject-matter knowledge, and our assumptions, about the qualitative causal structure of a problem.
-But, as described in the next sections, causal diagrams also encode information about potential associations between the variables in the causal network.
-It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool.
-[@Hernan_et_al_2025, pp.
-75] -->
-
-SCMs and DAGs support identification analysis through two key advantages.
-First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020].
-This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020].
-Second, they depict causal relationships in a non-parametric and fully interactive way.
-This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp.
-35].
-
-Thus, @sec-appendixB31 and @sec-appendixB32 elaborate on the first advantage, while @sec-appendixB32 and @sec-appendixB33 do so for the second.
-Finally, @sec-appendixB34 explains how researchers use SCMs and DAGs alongside Bayesian inference methods in the estimation process.
-
-<!-- A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. -->
-
-#### The five fundamental block for SCMs and DAGs {#sec-appendixB31 .appendix appendix-style=plain}
-
-<!-- 2. the SCMs -->
-Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display the five fundamental building blocks for SCMs and DAGs.
-The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020].
-Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
-In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
-27,68].
-Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.
-These functions use the symbol '$:=$' to denote the asymmetrical causal dependence of the variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.
-
-<!-- 3. The DAGs -->
-Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020].
-The right panels of the figures display these DAGs.
-A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables.
-The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence.
-The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020].
-DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014].
-Although the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* depicted in the figures offers one key advantage: including exogenous variables can help researchers highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020].
-
-::: {#fig-dags_scms1 layout-ncol=2}
-
-::: {#fig-scm_bb1 fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{3} & := f_{X3}(e_{X3}) \\
-  e_{X1} & \dsep e_{X3}
-\end{aligned}
-$$
-
-SCM
+# References {.unnumbered .appendix appendix-style=plain}
+
+<!-- {.unnumbered} -->
+
+:::{#refs}
+
 :::
-
-![DAG](images/png/mdag_bb1.png){#fig-mdag_bb1 fig-align="center" fig-pos="H" width=54%}
-
-Two unconnected nodes
-:::
-
-::: {#fig-dags_scms2 layout-ncol=2}
-
-::: {#fig-scm_bb2 fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{3} & := f_{X3}(X_{1},e_{X3}) \\
-  e_{X1} & \dsep e_{X3}
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](images/png/mdag_bb2.png){#fig-mdag_bb2 fig-align="center" fig-pos="H" width=54%}
-
-Two connected nodes or descendant
-:::
-
-::: {#fig-dags_scms3 layout-ncol=2}
-
-::: {#fig-scm_bb3 fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{2} & := f_{X2}(X_{1},e_{X2}) \\
-  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
-  e_{X1} & \dsep e_{X2} \\
-  e_{X1} & \dsep e_{X3} \\
-  e_{X2} & \dsep e_{X3}
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](images/png/mdag_bb3.png){#fig-mdag_bb3 fig-align="center" fig-pos="H" width=65%}
-
-Chain or mediator
-:::
-
-::: {#fig-dags_scms4 layout-ncol=2}
-
-::: {#fig-scm_bb4 fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  X_{1} & := f_{X1}(X_{2},e_{X1}) \\
-  X_{2} & := f_{X2}(e_{X2}) \\
-  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
-  e_{X1} & \dsep e_{X2} \\
-  e_{X1} & \dsep e_{X3} \\
-  e_{X2} & \dsep e_{X3}
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](images/png/mdag_bb4.png){#fig-mdag_bb4 fig-align="center" fig-pos="H" width=65%}
-
-Fork or confounder
-:::
-
-::: {#fig-dags_scms5 layout-ncol=2}
-
-::: {#fig-scm_bb5 fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{2} & := f_{X2}(X_{1},X_{3},e_{X2}) \\
-  X_{3} & := f_{X3}(e_{X3}) \\
-  e_{X1} & \dsep e_{X2} \\
-  e_{X1} & \dsep e_{X3} \\
-  e_{X2} & \dsep e_{X3}
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](images/png/mdag_bb5.png){#fig-mdag_bb5 fig-align="center" fig-pos="H" width=65%}
-
-Collider or inmorality
-:::
-
-<!-- 4. Theoretical implications of the basic building block of the SCMS and DAGs -->
-A careful examination of these building blocks highlights the theoretical assumptions underlying their observed variables.
-SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are independent or not causally related.
-SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X_{1}$ exerts a causal influence on a *child* node $X_{3}$.
-In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$.
-Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because there is a *direct path* connecting them.
-SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$.
-In this configuration, $X_{1}$ is a parent node of $X_{2}$, which is a parent node of $X_{3}$.
-This structure creates a *directed path* between $X_{1}$ and $X_{3}$.
-Consequently, $X_{1}$ is an *ancestor* of $X_{3}$, and $X_{2}$ fully *mediates* the relationship between the two.
-SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4] illustrate a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$.
-Here, $X_{2}$ is a parent node that *confounds* the relationship between $X_{1}$ and $X_{3}$.
-Finally, SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] show a *collider*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$.
-In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$ (an *inmorality*).
-Notably, all building blocks assume the errors are independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X1} \dsep e_{X2}$, $e_{X1} \dsep e_{X3}$, and $e_{X2} \dsep e_{X3}$.
-
-<!-- 5. Continuing with the motivating example -->
-Researchers can then use these building blocks to represent the scenario outlined in @sec-appendixB2.
-SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example.
-In this context, the variable $X$ (socio-economic status of the school) is thought to be a confounder in the relationship between the teaching method $T$ and the outcome $Y$.
-The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$.
-They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$.
-Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.
-
-::: {#fig-example1 layout-ncol=2}
-
-::: {#fig-scm_example1 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{T}(X,e_{T}) \\
-  Y & := f_{Y}(T,X,e_{Y}) \\
-  e_{T} & \dsep e_{X} \\
-  e_{T} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Y}
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](images/png/mdag_example1.png){#fig-mdag_example1 fig-align="center" width=65%}
-
-Plausible causal structure the scenario outlined in @sec-appendixB2.
-:::
-
-#### The probabilistic implications of these blocks {#sec-appendixB32 .appendix appendix-style=plain}
-
-<!-- 1. How the SCMS and DAGs encode probabilistic information? -->
-Beyond their graphical capabilities, researchers can translate SCMs and DAGs into statistical models because these conceptual models encode functional and probabilistic information, which they can then replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
-Notably, this encoding relies on two fundamental assumptions: the causal Markov and the faithfulness assumption.
-The *causal Markov assumption* links the graph structure to the joint probability distribution of variables by asserting that any variable in a graph is independent of all its non-descendants conditional on its parents [@Neal_2020, pp.
-20; @Hernan_et_al_2025, pp.
-80].
-The *faithfulness assumption*, in turn, states that the observed statistical (in)dependencies in the data reflect the underlying causal structure among variables--except in rare degenerate cases [@Neal_2020, pp.
-100; @Hernan_et_al_2025, pp.
-81].
-
-<!-- Causal DAGs are of no practical use unless we make an assumption linking the causal structure represented by the DAG to the data obtained in a study.
-This assumption, referred to as the causal Markov assumption, states that, conditional on its direct causes, a variable Vj is independent of any variable for which it is not a cause.
-That is, conditional on its parents, Vj is independent of its non-descendants; hence, a causal DAG is Markov with respect to the DAG G.
-[@Hernan_et_al_2025, pp.
-74] -->
-
-<!-- causal diagrams also encode information about potential associations between the variables in the causal network.
-It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool.
-[@Hernan_et_al_2025, pp.
-75] -->
-
-<!-- This and the previous section review three structural reasons why two variables may be associated: one causes the other, they share common causes, or they share a common effect and the analysis is restricted to certain level of that common effect (or of its descendants).
-Along the way we introduced a number of graphical rules that can be applied to any causal diagram to determine whether two variables are (conditionally) independent.
-The arguments we used to support these graphical rules were heuristic and relied on our causal intuitions.
-These arguments, however, have been formalized and mathematically proven.
-See Fine Point 6.1 for a systematic summary of the graphical rules, and Fine Point 6.2 for an introduction to the concept of faithfulness.
-[@Hernan_et_al_2025, pp.
-79] -->
-
-<!-- (from probability to graph): The relationship between statistical independence and the purely graphical concept of d-separation relies on the causal Markov assumption (Technical Point 6.1): In a causal DAG, any variable is independent of its non-descendants conditional on its parents.
-Pearl (1988) proved the following fundamental theorem: The causal Markov assumption implies that, given any three disjoint sets A, B, C of variables, if A is d-separated from B conditional on C, then A is statistically independent of B given C.
-The assumption that the converse holds, i.e., that A is d-separated from B conditional on C if A is statistically independent of B given C, is a separate assumption—the faithfulness assumption described in Fine Point 6.2.
-[@Hernan_et_al_2025, pp.
-80] -->
-
-<!-- (from graph to probability): Formally, faithfulness is the assumption that, for three disjoint sets A, B, C on a causal DAG (where C may be the empty set), A independent of B given C implies A is d-separated from B given C.
-When, as in our example, the causal diagram makes us expect a non-null association that does not actually exist in the data, we say that the joint distribution of the data is not faithful to the causal DAG.
-In our example the unfaithfulness was the result of effect modification (by sex) with opposite effects of exactly equal magnitude in each half of the population.
-Such perfect cancellation of effects is rare, and thus we will assume faithfulness throughout this book.
-Because unfaithful distributions are rare, in practice lack of d-separation (See Fine Point 6.1) can be almost always equated to non-zero association.
-There are, however, instances in which faithfulness is violated by design.
-For example, consider the prospective study in Section 4.5.
-The average causal effect of A on Y was computed after matching on L.
-In the matched population, L and A are not associated because the distribution of L is the same in the treated and the untreated.
-That is, individuals are selected into the matched population because they have a particular combination of values of L and A.
-The causal diagram in Figure 6.9 represents the setting of a matched study in which selection S (1: yes, 0: no) is determined by both A and L.
-The box around S indicates that the analysis is restricted to those selected into the matched cohort (S = 1).
-According to d-separation rules, there are two open paths between A and L when conditioning on S: L -> A and L -> S <- A.
-Thus one would expect L and A to be associated conditionally on S.
-However, matching ensures that L and A are not associated (see Chapter 4).
-Why the discrepancy?
-Matching creates an association via the path L -> S <- A that is of equal magnitude, but opposite direction, as the association via the path L -> A.
-The net result is a perfect cancellation of the associations.
-Matching leads to unfaithfulness.
-[@Hernan_et_al_2025, pp.
-81] -->
-
-<!-- ![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%} -->
-
-<!-- 2. What are the general implications of this? -->
-A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can represent the joint distribution of variables more efficiently [@Pearl_et_al_2016, pp.
-29].
-This expression takes the form of a product of conditional probability distributions (CPDs) of the type $P(child \mid parents)$.
-This property is formally known as the *Bayesian Network factorization* (BNF, @eq-net_factor) [@Pearl_et_al_2016, pp.
-29; @Neal_2020, pp.
-21].
-In this expression, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.
-$$
-\begin{aligned}
-P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
-& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}) ) & (\small{\text{by BNF}})
-\end{aligned}
-$$ {#eq-net_factor}
-
-<!-- 3. What is the net result of this? -->
-This encoding enables researchers with conceptual (theoretical) knowledge in the form of an SCM and DAG to predict patterns of (in)dependencies in the data.
-As highlighted by @Pearl_et_al_2016 [pp.
-35], these predictions depend solely on the structure of these conceptual models without requiring the quantitative details of the equations or the distributions of the errors.
-Moreover, once researchers observe empirical data, the patterns of (in)dependencies in the data can provide significant insights into the validity of the proposed conceptual model.
-
-<!-- The net result is that a researcher who has scientific knowledge in the form of a structural equation model is able to predict patterns of independencies in the data, based solely on the structure of the model’s graph, without relying on any quantitative information carried by the equations or by the distributions of the errors [@Pearl_et_al_2016, pp.
-35].
--->
-
-<!-- 4. can you illustrate this concept? -->
-The five fundamental building blocks described in @sec-appendixB31 clearly illustrate which (in)dependencies can SMCs and DAGs predict.
-For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] enables researchers to express the joint probability distribution of the observed variables as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, supporting the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$) [@Neal_2020, pp.
-24].
-Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \mid X_{1}) P(X_{1})$.
-Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendixB31 and the CPDs.
-
-Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of(in)dependencies from chains, forks, and colliders.
-These (in)dependencies apply to any data set generated by a causal model with those structures, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, pp.
-36].
-For instance, applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] allow researchers to represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \mid X_{1}) P(X_{3} \mid X_{2})$.
-This expression implies that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \mid X_{2})$.
-Moreover, in the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \mid X_{2}) P(X_{2}) P(X_{3} \mid X_{2})$.
-Similar to the chain structure, this expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \mid X_{2})$.
-Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \mid X_{1}, X_{3}) P(X_{3})$.
-This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent $(X_{1} \dsep X_{3})$, but conditionally dependent when controlling for $X_{2}$ $(X_{1} \ndsep X_{3} \mid X_{2})$.
-The authors @Pearl_et_al_2016 [pp.
-37, 40, 41] and @Neal_2020 [pp.
-25–26] provide the mathematical proofs for these conclusions.
-
-<!-- These (in)dependencies will be true of every data set generated by a causal model with that graphical structure, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, pp.
-36].
--->
-
-<!-- 5. Continuing with the motivating example -->
-Using these additional probabilistic insights, researchers can revisit the scenario in @sec-appendixB2.
-In this context, applying the BNF to the SCM [-@fig-scm_example2] structure, enables the representation of the joint probability distribution of the observed variables as $P(Y, T, X) =$ $P(Y \mid T, X) P(T \mid X) P(X)$.
-From this expression, researchers can infer that the outcome $Y$ is unconditionally dependent on the teaching method $T$ $(Y \ndsep T)$.
-This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path in DAG [-@fig-mdag_example2]), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path in DAG [-@fig-mdag_example2]).
-
-::: {#fig-example2 layout-ncol=2}
-
-::: {#fig-scm_example2 fig-align="center"}
-$$
-\begin{aligned}
-  X & := x \\
-  T & := f_{T}(x,e_{T}) \\
-  Y & := f_{Y}(T,x,e_{Y}) \\
-  e_{T} & \dsep e_{X} \\
-  e_{T} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Y}
-\end{aligned}
-$$
-
-SCM
-:::
-
-![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example2 fig-align="center" width=65%}
-
-Plausible causal structure the scenario outlined in @sec-appendixB2.
-:::
-
-#### From probability to causality {#sec-appendixB33 .appendix appendix-style=plain}
-
-<!-- 1. Transforming probabilistic association into causation?  -->
-The structural approach to causal inference translates probabilistic insights into actionable strategies seeking to identify the ACE from associational quantities.
-The approach achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
-34].
-
-<!-- 2. additional necessary concepts  -->
-The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions $P(Y_{i} \mid do(T_{i}=t))$ within SCMs and DAGs.
-*Manipulated graphs* simulate physical interventions by removing specific edges from a DAG, while preserving the remaining structure unchanged [@Neal_2020, pp.
-34].
-In parallel, *Truncated Factorization* (TF) achieves a similar simulation by removing specific functions from the conceptual model and replacing them with constants, while keeping the rest of the structure unchanged [@Pearl_2010].
-The probabilistic implications of this factorization are formalized in @eq-trunc_factor, where $S$ represents the subset of variables $X_{p}$ directly influenced by the intervention, while an example illustrating these concepts follows below.
-
-$$
-P(X_{1}, X_{2}, \dots, X_{P} \mid do(S)) =
-\begin{cases}
-  \prod P(X_{p} \mid pa(X_{p}) ) & \text{if} \: p \not\in S \\
-  1 \quad & \text{otherwise}
-\end{cases}
-$$ {#eq-trunc_factor}
-
-<!-- 3. How to use these concepts to finally define the ACE -->
-Using the TF, researchers can define the *backdoor adjustment* to identify the ACE.
-This adjustment states that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-The sufficient adjustment set (potentially empty) must block all non-causal paths between $X_{a}$ and $X_{b}$ without introducing new paths.
-If such a set exists, then $X_{a}$ and $X_{b}$ are *d-separated* by $X_{p}$ ($X_{a} \dsep X_{b} \mid X_{p}$) [@Pearl_2009], and $X_{p}$ satisfies the *backdoor criterion* [@Neal_2020, pp.
-37].
-
-$$
-P(X_{a} \mid do(X_{b}=x)) = \sum_{Xp} P(X_{a} \mid X_{b}=x, X_{p}) P(X_{p})
-$$ {#eq-backdoor}
-
-Ultimately, the backdoor adjustment enables researchers to express the ACE as:
-
-$$
-\begin{aligned}
-\tau & = E[X_{a} \mid do(X_{b}=1)]- E[X_{a} \mid do(X_{b}=2)] \\
-  & = E_{Xp}\left[ E[X_{a} \mid do(X_{b}=1), X_{p}]- E[X_{a} \mid do(X_{b}=2), X_{p}] \right] \\
-  & = \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=2, X_{p}) \cdot P(X_{p})
-\end{aligned}
-$$ {#eq-backdoor_adjustment}
-
-<!-- 4. revisiting the motivating example -->
-With these new insights, researchers revisiting the scenario in @sec-appendixB32 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct.
-This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure.
-Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \mid X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.
-Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.
-
-::: {#fig-example3 layout-ncol=2}
-
-::: {#fig-scm_example3 fig-align="center"}
-$$
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := t \\
-  Y & := f_{Y}(t,X,e_{Y}) \\
-  e_{T} & \dsep e_{X} \\
-  e_{T} & \dsep e_{Y} \\
-  e_{X} & \dsep e_{Y}
-\end{aligned}
-$$
-
-SCM
-:::
-
-![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example3 fig-align="center" width=65%}
-
-Plausible causal structure the scenario outlined in @sec-appendixB32.
-:::
-
-Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$.
-They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE.
-This process is expressed in @eq-mACE2 (see @sec-appendixB2).
-
-$$
-\begin{aligned}
-  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right] \\
-  & = \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=2, X) \cdot P(X)
-\end{aligned}
-$$ {#eq-mACE2}
-
-#### The estimation process {#sec-appendixB34 .appendix appendix-style=plain}
-
-<!-- 1. How do we use Bayesian inference? -->
-Ultimately, researchers can use Bayesian inference methods to estimate the ACE.
-The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{P} \mid \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010], where $X_{P}$ represents a random variable, and $\theta$ represents a one-dimensional parameter space for simplicity.
-After observing empirical data, researchers can update the priors to posterior distributions using Bayes' rule in @eq-bayes_rule [@Jeffreys_1998]:
-
-$$
-P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
-$$ {#eq-bayes_rule}
-
-Given that the denominator on the right-hand side of @eq-bayes_rule serves as a normalizing constant independent of the parameter $\theta$, researchers can simplify the posterior updating process into three steps.
-First, they integrate new empirical data through the likelihood.
-Second, they update the parameters' priors to a posterior distribution according to @eq-prop_rule.
-Ultimately, they normalize these results to obtain a valid probability distribution.
-
-$$
-P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\mid \theta) \cdot P(\theta)
-$$ {#eq-prop_rule}
-
-<!-- check: -->
-<!-- https://en.wikipedia.org/wiki/Maximum_a_posteriori_estimation -->
-<!-- https://en.wikipedia.org/wiki/Bayesian_inference -->
-
-<!-- 2. But what is the problem? -->
-
-Temporarily setting aside the definition of prior distributions $P(\theta)$, note that the posterior updating process depends heavily on the assumptions underlying the likelihood of the data.
-However, as the number of random variables, $P$, increases, this joint distribution quickly becomes intractable [@Neal_2020].
-This intractability is evident from @eq-like_chain, where the likelihood distribution is expressed by multiple chained CPDs.
-
-$$
-P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}, \theta )
-$$ {#eq-like_chain}
-
-<!-- 3. How do DAGs can help? -->
-Nevertheless, researchers can manage the complexity of the likelihood by assuming specific local (in)dependencies among variables.
-SCMs and DAGs provide a formal framework to represent these assumptions, as detailed in @sec-appendixB32.
-These assumptions improve model tractability and simplify the estimation process by enabling the derivation of the BNF of the likelihood (@eq-like_BNF).
-With this simplified structure, any probabilistic programming language can model the system and compute the parameter's posterior distribution using @eq-bayes_rule.
-
-$$
-P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}), \theta )
-$$ {#eq-like_BNF}
-
-{{< pagebreak >}}
-
-# References {.unnumbered .appendix appendix-style=plain}
-
-<!-- {.unnumbered} -->
-
-:::{#refs}
-
-:::

### 2025-06-07 _tex/paper2.tex
Built site for gh-pages

@@
 
-Among the many potential directions for future research--both current
-and new--three avenues deserve particular attention due to their direct
-impact on the reliability and validity of CJ trait estimates, as well as
-on the accuracy of statistical inferences. The following sections
-outline these avenues and explain how our approach facilitates their
-investigation.
+Among the many potential directions for future research, three avenues
+deserve particular attention due to their direct impact on the
+reliability and validity of CJ trait estimates, as well as on the
+accuracy of statistical inferences. The following sections outline these
+avenues and explain how our approach facilitates their investigation.
 
@@
 Although sampling and comparison mechanisms are central to modern CJ
-assessments, it is striking that most of the CJ literature has examined
-them only within a limited scope. Researchers have primarily
-investigated the effects of adaptive comparative judgment (ACJ) designs
-on trait reliability
+assessments, it is striking that most CJ literature has examined them
+within a limited scope. Researchers have primarily investigated the
+effects of adaptive comparative judgment (ACJ) designs on trait
+reliability
 \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
@@
 should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}. While
-these studies offer valuable insights, they often overlook the broader
-role these mechanisms play within the CJ system. As this oversight
-likely stems from a broader lack of conceptual clarity about how these
-mechanisms function within the system, this study decided to formally
-incorporate these mechanisms into CJ's conceptual model.
+these studies offer valuable insights, they also overlook the broader
+role that these mechanisms play within the CJ system. As this oversight
+likely stems from a more fundamental lack of conceptual clarity about
+how these mechanisms function within the system, this study decided to
+integrate these mechanisms into the conceptual model of CJ.
 
@@
 offers a new perspective on how these mechanisms shape the CJ process.
-In particular, it clarifies their role in the data-generating process by
-framing them as sources of missing data--that is, as mechanisms that
-determine which observations are missing from the final data sample.
-This new perspective encourages the application of Little and Rubin's
-principled missing data framework \citeyearpar{Little_et_al_2020},
-enabling a more rigorous evaluation of existing claims about these
-missing data mechanisms, their influence on CJ outcomes, and their
-implications for designing and assessing more complex assessments
-setups.
+Specifically, it clarifies their role as sources of missing data in CJ's
+data-generating process--that is, as mechanisms that determine which
+observations are missing from the final data sample. This new
+perspective encourages the application of Little and Rubin's principled
+missing data framework \citeyearpar{Little_et_al_2020}, allowing a more
+rigorous evaluation of existing claims about these missing data
+mechanisms, their influence on CJ outcomes, and their implications for
+designing and evaluating more complex assessments setups.
 
@@
 designs, in which the sampling and comparison mechanisms introduce more
-intricate forms of missingness --- such as \emph{missing at random}
-(MAR) or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}.
-A prominent example is the previously discussed ACJ designs, where prior
+intricate forms of missingness--such as \emph{missing at random} (MAR)
+or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
+prominent example is the previously discussed ACJ designs, where prior
 judgment outcomes inform the selection of stimulus pairs for subsequent
@@
 
-Despite the growing notion that various stimulus-related factors can
+Despite the growing notion that various stimulus-related factors
 influence judges' perceptions
@@
 This gap likely persists not due to a lack of interest or research but
-because researchers often rely on ad-hoc detection methods--such as
-`misfit' statistics--that are poorly suited for the task
+because researchers often rely on ad-hoc detection methods, such as
+`misfit' statistics, that may not be well-suited for the task
 \citep{Kelly_et_al_2022}. To address this limitation, this study treats
@@
 Section~\ref{sec-theory-issue1b}, is whether training or expertise can
-help judges avoid focusing on irrelevant stimulus features--such as
-handwriting--over more central criteria like argumentative quality in
+help judges avoid focusing on irrelevant stimulus features, such as
+handwriting, over more central criteria like argumentative quality in
 writing assessments \citep{Kelly_et_al_2022}. Notably, exploring these
@@
 
-Moreover, since judges draw on a variety of intricate stimulus
-characteristics when evaluating complex, multidimensional traits
+Moreover, since judges rely on these stimulus-related factors when
+evaluating complex, multidimensional traits
 \citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022},
-and these characteristics account for variation in judgment accuracy
+and these factors account for variation in judgment accuracy
 \citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020, Gijsen_et_al_2021},
-it is reasonable to expect that judges' assessments also vary with
-attributes such as gender, age, culture, income, education, training, or
-expertise \citep{Kelly_et_al_2022}. Prior studies support this view
+it is reasonable to expect that assessments also vary according to
+judge-specific attributes such as gender, age, culture, income,
+education, training, or expertise \citep{Kelly_et_al_2022}. Prior
+studies support this view
 \citep{Bartholomew_et_al_2020a, McMahon_et_al_2015}. Thus, building on
@@
 \citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
-undermine the assumption of ``sample-freeness,'' researchers should
-explore ways to control their influence and determine how many
-judges--and how many judgments per judge--are needed to produce reliable
-trait estimates in these contexts. Additionally, it is worth considering
-whether \emph{repeated measures designs}, in which judges evaluate the
-same stimulus pairs multiple times \citep{Lawson_2015}, can improve
-judgment consistency and accuracy. Our approach provides the necessary
-structure to rigorously investigate these questions.
-
-\subsubsection{The identification of `misfitting' stimuli and
-judges}\label{sec-discussion_RA3}
+undermine the assumption of ``sample-freeness,'' it becomes essential to
+explore strategies for mitigating their influence and to determine how
+many judges (and how many judgments per judge) are needed to produce
+reliable trait estimates under these conditions. In addition, it is
+worth considering whether \emph{repeated measures designs}, in which
+judges evaluate the same stimulus pairs multiple times
+\citep{Lawson_2015}, can improve judgment consistency and accuracy. As
+anticipated, our approach provides the necessary structure to
+investigate rigorously these questions.
+
+\subsubsection{The identification of `misfitting' judges and
+stimuli}\label{sec-discussion_RA3}
 
 Although the CJ literature clearly defines \emph{misfit} judges and
-stimuli, researchers have rarely examined how these observations relate
-to Thurstonian theory. In particular, they have not identified which
-elements of Thurstone's theory might explain the occurrence of misfits.
-This disconnect likely stems from the fact that CJ researchers derive
-misfit statistics from methods such as residual analysis and outlier
-detection rather than from Thurstonian principles
-\citep{Pollitt_2012a, Pollitt_2012b, Wu_et_al_2022}. They typically
-define \emph{misfit judges} as those whose assessments diverge
-significantly from the ``shared consensus''
+stimuli, CJ researchers have rarely examined how these observations
+relate to Thurstonian theory. In particular, they have not identified
+which elements of Thurstone's theory might account for the occurrence of
+misfits. This disconnect likely stems from the fact that CJ researchers
+derive misfit statistics from residual analysis and outlier detection
+methods rather than from Thurstonian principles. Specifically,
+\emph{misfit judges} are typically defined as those whose assessments
+diverge significantly from the ``shared consensus''
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
-and \emph{misfit stimuli} as those that elicit more judgment
+while \emph{misfit stimuli} are those that elicit more judgment
 discrepancies than others
@@
 outliers--that is, observations that deviate markedly from the rest of
-the sample in which they occur \citep{Grubbs_1969}. However, this
-resemblance extends beyond these definitions. CJ researchers also
-identify misfits using standard outlier detection procedures, such as
-transforming BTL model residuals into statistics of interest and
-comparing them against a predefined threshold.
-
-{After identifying misfit observations, researchers often propose
-various hypotheses to explain their presence. Yet, without providing
-evidence for any of these explanations, they typically exclude these
-observations from analysis. This exclusion process raises three major
-concerns: one related to the reliability of the methods used to identify
-misfits, a second concerning the potential impact of their removal on
-trait estimates and inferences, and a third regarding the hypotheses
-used to explain the presence and exclusion of misfits.}
-
-{First, regarding the reliability of identification methods, it is
-relevant to note that outlying observations can only be defined relative
-to a specific model \citep{McElreath_2020}. Consequently, methods based
-on models such as the BTL--which rest on strong assumptions--should be
-treated with caution, as these assumptions may prevent the model from
-accurately capturing the underlying data-generating process of the CJ
-system, thereby potentially compromising the identification of misfits.}
-
-{Second, concerning the impact of removing misfits, the statistical
-literature warns that ad-hoc outlier exclusion procedures risk
-discarding valuable information \citep{Miller_2023} and introducing bias
-into trait estimates. The direction and magnitude of this bias are often
-unpredictable, as they depend on which observations researchers exclude
-from the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.}
-
-{Lastly, regarding the hypotheses explaining the presence and exclusion
-of misfits, it is important to note that a significant number of these
-hypotheses could not be tested within the BTL model due to its strong
-assumptions and limited flexibility. This is the reason why there is a
-lack of supporting evidence for misfit observations. In contrast, our
-approach enables researchers to test these hypotheses. For instance,
-researchers could examine whether misfit judges are simply those who
-exhibit (an outlying degree of) systematic bias or whether they display
-greater variability in their judgments compared to their peers.
-Similarly, in the case of misfit stimuli, researchers could investigate
-whether their status reflects an increased variability in their
-discriminal processes or stems from their classification as outlying
-stimuli.}
-
-{Moreover, since outliers are not necessarily ``bad data''
-\citep{McElreath_2020}, our approach offers a principled alternative
-that retains misfits in the analysis without compromising trait
-estimation. This alternative involves adapting our proposed models into
-more sophisticated, robust measurement models
-\citep{Wainer_et_al_1978}--a broad class of procedures designed to
-reduce the sensitivity of parameter estimates to mild or moderate
-departures from model assumptions \citep{Everitt_et_al_2010}.}
+the sample in which they occur \citep{Grubbs_1969}. But this resemblance
+extends beyond the definitions themselves, as CJ researchers often
+identify misfits using conventional outlier detection procedures, such
+as transforming BTL model residuals into diagnostic statistics and
+comparing them against predefined thresholds
+\citep{Pollitt_2012a, Pollitt_2012b, Wu_et_al_2022}.
+
+Nevertheless, it is not the classification of misfits as outliers that
+raises concerns for trait measurement and inference, but rather the
+prevailing CJ practice of identifying them through these ad-hoc
+detection procedures and then excluding them from analysis, often
+without providing empirical evidence for the various hypotheses proposed
+to explain their occurrence in the first place. In this regard, it is
+essential to recognize that outliers can only be defined relative to a
+specific model \citep{McElreath_2020}. Thus, identification procedures
+based on models like the BTL, which relies on strong assumptions, should
+be approached with caution, as they may not be well-suited for the task
+\citep{Kelly_et_al_2022}. Specifically, because the BTL model may not
+always accurately reflect the underlying data-generating process of the
+CJ system (see Section~\ref{sec-theory-issues}), detection procedures
+grounded in this model risk compromising the accurate identification of
+misfits. Moreover, the statistical literature cautions that excluding
+outliers risks discarding valuable information \citep{Miller_2023} and
+introducing bias into trait estimates. The direction and magnitude of
+this bias are often unpredictable, as they depend on which observations
+researchers exclude from the analysis
+\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. Lastly, regarding
+the hypotheses proposed to explain the presence of misfits, it is
+evident that the BTL model lacks the flexibility needed to adequately
+test many of them, largely due to its strong assumptions.
+
+In contrast, our approach provides a rigorous framework that enables the
+examination of several relevant hypotheses. For instance, researchers
+could investigate whether misfit judges are those who exhibit (an
+outlying degree of) systematic bias or greater variability in their
+judgments compared to their peers. Similarly, researchers might explore
+whether misfit stimuli exhibit more variable discriminal processes
+relative to other stimuli or if they are genuinely outlying cases.
+Moreover, since outliers are not necessarily ``bad data''
+\citep{McElreath_2020}, our approach also offers a principled
+alternative that retains misfits in the analysis without compromising
+trait estimation or inference. This alternative involves adapting our
+proposed models into robust measurement models \citep{McElreath_2020}--a
+broad class of procedures designed to reduce the sensitivity of
+parameter estimates to mild or moderate departures from model
+assumptions \citep{Everitt_et_al_2010}. This alternative also relates to
+a broader discussion in social science, which is: that when researchers
+face low predictive capacity from their models, they should not only
+search for ``new'' variables or alternative procedures but also consider
+employing more sophisticated measurement models
+\citep{Wainer_et_al_1978}.
 
@@
 Even after acquiring the necessary foundational knowledge, CJ
-researchers face two additional challenges when applying our approach to
-their specific research context. First, they must verify whether the
-conceptual model provides a \emph{faithful} \footnote{Avoid confusing
-  this term with the \emph{faithfulness assumption}
+researchers still face two additional challenges when applying our
+approach to their specific research context. First, they must verify
+whether the conceptual model provides a \emph{faithful} \footnote{Avoid
+  confusing this term with the \emph{faithfulness assumption}
   \citep{Neal_2020, Hernan_et_al_2025} described in the causal inference
@@
 this apriori. Instead, they need to adapt these models to their specific
-CJ contexts, paying close attention to assessment design features and
+contexts, paying close attention to assessment design features and
 causal assumptions. As discussed in
@@
 inference from finite samples cannot be achieved
-\citep{Schuessler_et_al_2023}.
-
-Although performing identification analysis through formal derivations
-may seem like a natural next step, we argue that the complexity of the
-CJ system renders such an approach impractical at the outset. Instead,
-simulation-based methods, such as power analysis, provide a more
-practical and flexible alternative, enabling researchers to examine the
-consistency of estimates without relying on cumbersome mathematical
-proofs. Nonetheless, the development of formal identification proofs
-remains a significant goal for future CJ research. Importantly, our
-approach supports both strategies by providing the probabilistic
-foundation for formal derivations and the necessary statistical
-structure for simulation-based methods.
+\citep{Schuessler_et_al_2023}. Although performing identification
+analysis through formal derivations may seem like a natural next step,
+the complexity of the CJ system renders such an approach impractical at
+the outset. Instead, simulation-based methods, such as power analysis,
+provide a more practical and flexible alternative, enabling researchers
+to examine the consistency of estimates without relying on cumbersome
+mathematical proofs. Nonetheless, the development of formal
+identification proofs remains a significant goal for future CJ research.
+Importantly, our approach supports both strategies by providing the
+probabilistic foundation for formal derivations and the necessary
+statistical structure for simulation-based methods.
 
@@
 
-\section{Appendix}\label{sec-appendix}
-
-\subsection{Statistical and Causal inference}\label{sec-appendixB}
-
-This section introduces fundamental statistical and causal inference
-concepts necessary for understanding the core theoretical principles
-described in this document. It does not, however, offer a comprehensive
-overview of causal inference methods. Readers seeking more in-depth
-understanding may wish to explore introductory papers such as
-\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
-\citet{Cinelli_et_al_2020}. They may also find it helpful to consult
-introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
-\citet{McElreath_2020}. For more advanced study, readers may refer to
-seminal intermediate papers such as \citet{Neyman_et_al_1923},
-\citet{Rubin_1974}, \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009},
-as well as books such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
-and \citet{Hernan_et_al_2025}.
-
-\subsubsection{Empirical research and randomized
-experiments}\label{sec-appendixB1}
-
-Empirical research uses evidence from observation and experimentation to
-address real-world challenges. In this context, researchers typically
-formulate their research questions as \emph{estimands} or \emph{targets
-of inference}, i.e., the specific quantities they seek to determine
-\citep{Everitt_et_al_2010}. For instance, researchers might be
-interested in answering the following question: ``To what extent do
-different teaching methods \((T)\) influence students' ability to
-produce high-quality written texts \((Y)\)?'' To investigate this,
-researchers could randomly assign students to two groups, each exposed
-to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
-perform pairwise comparisons, generating a dichotomous outcome
-\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
-ability. In this scenario, the research question can be rephrased as the
-estimand, ``\emph{On average}, is there a difference in the ability to
-produce high-quality written texts between the two groups of students?''
-and this estimand can be mathematically represented by the random
-associational quantity in Equation~\ref{eq-group_diff}, where
-\(E[\cdot]\) denotes the expected value.
-
-\begin{equation}\phantomsection\label{eq-group_diff}{
-E[Y_{i} \mid T_{i}=1] - E[Y_{i} \mid T_{i}=2]
-}\end{equation}
-
-Researchers then proceed to identify the estimands.
-\emph{Identification} determines whether an estimator can accurately
-compute the estimand based solely on its assumptions, regardless of
-random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
-\emph{estimator} refers to a method or function that transforms data
-into an estimate \citep{Neal_2020}. \emph{Estimates} are numerical
-values that approximate the estimand derived through the process of
-\emph{estimation}, which integrates data with an estimator
-\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
-\citep{McElreath_2020, Neal_2020}, shown in Figure~\ref{fig-IEflow},
-visually represents the transition from estimands to estimates.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{images/png/IEflow.png}
-
-}
-
-\caption{\label{fig-IEflow}Identification-Estimation flowchart.
-Extracted and slightly modified from \citet[pp.~32]{Neal_2020}}
-
-\end{figure}%
-
-Identification is a necessary condition to ensure \emph{consistent}
-estimators. An estimator achieves \emph{consistency} when it converges
-to the ``true'' value of an estimand as the data size approaches
-infinity \citep{Everitt_et_al_2010}. Without identification, researchers
-cannot achieve consistency, even with ``infinite'' and error-free data.
-As a result, deriving meaningful insights about an estimand from finite
-data becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore,
-to ensure accurate and reliable estimates, researchers prioritize
-estimators with desirable identification properties. For instance, the
-Z-test is a widely used estimator for comparing group proportions,
-yielding accurate estimates when its underlying assumptions are
-satisfied \citep{Kanji_2006}. Furthermore, researchers can interpret
-estimates from the Z-test as causal, provided the data is collected
-through a randomized experiment.
-
-Randomized experiments are widely recognized as the gold standard in
-evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
-recognition stems from their ability to enable researchers interpret
-associational estimates as causal. They achieve this by ensuring data,
-and by extension an estimator, satisfies several key identification
-properties, such as common support, no interference, and consistency
-\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
-however, is the elimination of confounding. \emph{Confounding} occurs
-when an external variable \(X\) simultaneously influences the outcome
-\(Y\) and the variable of interest \(T\), resulting in spurious
-associations \citep{Everitt_et_al_2010}. Randomization addresses this
-issue by decoupling the association between the intervention allocation
-\(T\) from any other variable \(X\)
-\citep{Morgan_et_al_2014, Neal_2020}.
-
-Nevertheless, researchers often face constraints that limit their
-ability to conduct randomized experiments. These constraints include
-ethical concerns, such as the assignment of individuals to potentially
-harmful interventions, and practical limitations, such as the
-infeasibility of, for example, assigning individuals to genetic
-modifications or physical impairments \citep{Neal_2020}. In these cases,
-causal inference offers a valuable alternative for generating causal
-estimates and understanding the mechanisms underlying specific data. In
-addition, the framework can provide significant theoretical insights
-that can enhance the design of experimental and observational studies
-\citep{McElreath_2020}.
-
-\subsubsection{Identification under causal
-inference}\label{sec-appendixB2}
-
-Unlike classical statistical modeling, which focuses primarily on
-summarizing data and inferring associations, the \emph{causal inference}
-framework is designed to identify causes and estimate their effects
-using data \citep{Shaughnessy_et_al_2010, Neal_2020}. The framework uses
-rigorous mathematical techniques to address the \emph{fundamental
-problem of causality}
-\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This problem
-revolves around the question, ``What would have happened `in the world'
-under different circumstances?'' This question introduces the concept of
-counterfactuals, which are instrumental in defining and identifying
-causal effects.
-
-\emph{Counterfactuals} are hypothetical scenarios that are
-\emph{contrary to fact}, where alternative outcomes resulting from a
-given cause are neither observed nor observable
-\citep{Neal_2020, Counterfactual_2024}. The structural approach to
-causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
-framework for defining counterfactuals. For instance, in the scenario
-described in Section~\ref{sec-appendixB1}, the approach begins by
-defining the \emph{individual causal effect} (ICE) as the difference
-between each student's potential outcomes, as in Equation~\ref{eq-ICE}.
-
-\begin{equation}\phantomsection\label{eq-ICE}{
-\tau_{i} = Y_{i} \mid do(T_{i}=1) - Y_{i} \mid do(T_{i}=2)
-}\end{equation}
-
-where \(do(T_{i}=t)\) represents the intervention operator,
-\(Y_{i} \mid do(T_{i}=1)\) represents the potential outcome under
-intervention \(T_{i}=1\), and \(Y_{i} \mid do(T_{i}=1)\) represents the
-potential outcome under intervention \(T_{i}=2\). Here, an
-\emph{intervention} involves assigning a constant value to the treatment
-variable for each student's potential outcomes. Note that if a student
-is assigned to intervention \(T_{i}=1\), the potential outcome under
-\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
-observable. To address this challenge, the structural approach extends
-the ICE to the \emph{average causal effect} (ACE,
-Equation~\ref{eq-ACE}), representing the average difference between the
-students' observed potential outcomes and their counterfactual
-counterparts.
-
-\begin{equation}\phantomsection\label{eq-ACE}{
-\begin{aligned}
-\tau & = E[\tau_{i}] \\
-  & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)]
-\end{aligned}
-}\end{equation}
-
-Even though counterfactuals are unobservable, researchers can still
-identify the ACE from associational estimates by leveraging the
-structural approach. The approach identifies the ACE by statistically
-conditioning data on a \emph{sufficient adjustment set} of variables
-\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
-\emph{sufficient} set (potentially empty) must block all non-causal
-paths between \(T\) to \(Y\) without opening new ones. When such a set
-exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
-(\(T \:\bot\:Y \mid X\)) \citep{Pearl_2009}, and \(X\) satisfies the
-\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
-\emph{conditioning} describes the process of restricting the focus to
-the subset of the population defined by the conditioning variable
-\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).
-
-Conditioning on a sufficient adjustment set enables researchers to
-estimate the ACE, even when the data comes from an observational study.
-This process is feasible because such conditioning ensures that the ACE
-estimator satisfies several critical properties, including confounding
-elimination \citep{Morgan_et_al_2014}. Naturally, the validity of claims
-about the causal effects of \(T\) on \(Y\) now hinges on the assumption
-that \(X\) serves as a sufficient adjustment set. However, as
-\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
-real world from observed data inevitably requires assumptions. This
-requirement holds true for both observational and experimental data.
-
-For instance, if researchers cannot conduct the randomized experiments
-described in Section~\ref{sec-appendixB1} and must instead rely on
-observational data, they can still identify the ACE as long as an
-observed variable \(X\), such as the socio-economic status of the
-school, satisfies the backdoor criterion. Under these circumstances,
-researchers first identify the \emph{conditional average causal effect}
-(CACE, Equation~\ref{eq-CACE})
-
-\begin{equation}\phantomsection\label{eq-CACE}{
-CACE_{t} = E[Y_{i} \mid T_{i}=t, X]
-}\end{equation}
-
-From the CACE, researchers can identify the ACE from associational
-quantities as in Equation~\ref{eq-mACE1}. This identification process is
-commonly known as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\)
-represents the marginal expected value over \(X\)
-\citep{Morgan_et_al_2014}.
-
-\begin{equation}\phantomsection\label{eq-mACE1}{
-\begin{aligned}
-  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
-  & = E_{X}[CACE_{1} - CACE_{2}] \\
-  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right]
-\end{aligned}
-}\end{equation}
-
-Notably, the approach extends the ACE identification for a continuous
-variable \(T\) as in Equation~\ref{eq-mACE_cont}, ensuring broad
-applicability across different causal scenarios
-\citep[pp.~45]{Neal_2020}
-
-\begin{equation}\phantomsection\label{eq-mACE_cont}{
-\begin{aligned}
-  \tau &= E[Y_{i} \mid do(T_{i}=t)] \\
-  & = d E_{X}\left[ E[Y_{i} \mid T_{i}=t, X]\right]/ dt
-  \end{aligned}
-}\end{equation}
-
-\subsubsection{Diving into the specifics}\label{sec-appendixB3}
-
-The structural approach to causal inference uses SCMs and DAGs to
-formally and graphically represent the presumed causal structure
-underlying the ACE
-\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
-Essentially, these tools serve as \emph{conceptual (theoretical) models}
-on which identification analysis rests
-\citep[pp.~4]{Schuessler_et_al_2023}. Thus, using these tools,
-researchers can determine which statistical models can identify (ACE,
-CACE, or other), assuming the depicted causal structure is correct
-\citep{McElreath_2020}, thus enabling valid causal inference.
-Figure~\ref{fig-IEflow} shows the role of theoretical models in the
-inference process.
-
-SCMs and DAGs support identification analysis through two key
-advantages. First, regardless of complexity, they can represent various
-causal structures using only five fundamental building blocks
-\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
-decompose complex structures into manageable components, facilitating
-their analysis \citep{McElreath_2020}. Second, they depict causal
-relationships in a non-parametric and fully interactive way. This
-flexibility enables feasible ACE identification strategies without
-defining the variables' data types, the functional form between them, or
-their parameters \citep[pp.~35]{Pearl_et_al_2016}.
-
-Thus, Section~\ref{sec-appendixB31} and Section~\ref{sec-appendixB32}
-elaborate on the first advantage, while Section~\ref{sec-appendixB32}
-and Section~\ref{sec-appendixB33} do so for the second. Finally,
-Section~\ref{sec-appendixB34} explains how researchers use SCMs and DAGs
-alongside Bayesian inference methods in the estimation process.
-
-\paragraph{The five fundamental block for SCMs and
-DAGs}\label{sec-appendixB31}
-
-Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
-\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
-display the five fundamental building blocks for SCMs and DAGs. The left
-panels of the figures show the formal mathematical models, represented
-by the SCMs, defined in terms of a set of \emph{endogenous} variables
-\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
-\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
-\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
-Endogenous variables are those whose causal mechanisms a researcher
-chooses to model \citep{Neal_2020}. In contrast, exogenous variables
-represent \emph{errors} or \emph{disturbances} arising from omitted
-factors that the investigator chooses not to model explicitly
-\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
-\emph{structural equations}, express the endogenous variables as
-non-parametric functions of other variables. These functions use the
-symbol `\(:=\)' to denote the asymmetrical causal dependence of the
-variables and the symbol `\(\:\bot\:\)' to represent
-\emph{d-separation}, a concept akin to (conditional) independence.
-
-Notably, every SCM has an associated DAG
-\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
-figures display these DAGs. A DAG is a graph consisting of nodes
-connected by edges, where the nodes represent random variables. The term
-\emph{directed} means that the edges extend from one node to another,
-with arrows indicating the direction of causal influence. The term
-\emph{acyclic} implies that the causal influences do not form loops,
-ensuring the influences do not cycle back on themselves
-\citep{McElreath_2020}. DAGs represent observed variables as solid black
-circles, while they use open circles for unobserved (latent) variables
-\citep{Morgan_et_al_2014}. Although the \emph{standard representation}
-of DAGs typically omits exogenous variables for simplicity, the
-\emph{magnified representation} depicted in the figures offers one key
-advantage: including exogenous variables can help researchers highlight
-potential issues related to conditioning and confounding
-\citep{Cinelli_et_al_2020}.
-
-\begin{figure}[H]
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{3} & := f_{X3}(e_{X3}) \\
-  e_{X1} & \:\bot\:e_{X3}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb1}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb1}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-dags_scms1}Two unconnected nodes}
-
-\end{figure}%
-
-\begin{figure}[H]
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{3} & := f_{X3}(X_{1},e_{X3}) \\
-  e_{X1} & \:\bot\:e_{X3}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb2}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb2}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-dags_scms2}Two connected nodes or descendant}
-
-\end{figure}%
-
-\begin{figure}[H]
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{2} & := f_{X2}(X_{1},e_{X2}) \\
-  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
-  e_{X1} & \:\bot\:e_{X2} \\
-  e_{X1} & \:\bot\:e_{X3} \\
-  e_{X2} & \:\bot\:e_{X3}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb3}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb3}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-dags_scms3}Chain or mediator}
-
-\end{figure}%
-
-\begin{figure}[H]
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X_{1} & := f_{X1}(X_{2},e_{X1}) \\
-  X_{2} & := f_{X2}(e_{X2}) \\
-  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
-  e_{X1} & \:\bot\:e_{X2} \\
-  e_{X1} & \:\bot\:e_{X3} \\
-  e_{X2} & \:\bot\:e_{X3}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb4}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb4}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-dags_scms4}Fork or confounder}
-
-\end{figure}%
-
-\begin{figure}[H]
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X_{1} & := f_{X1}(e_{X1}) \\
-  X_{2} & := f_{X2}(X_{1},X_{3},e_{X2}) \\
-  X_{3} & := f_{X3}(e_{X3}) \\
-  e_{X1} & \:\bot\:e_{X2} \\
-  e_{X1} & \:\bot\:e_{X3} \\
-  e_{X2} & \:\bot\:e_{X3}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_bb5}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}
-
-}
-
-\subcaption{\label{fig-mdag_bb5}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-dags_scms5}Collider or inmorality}
-
-\end{figure}%
-
-A careful examination of these building blocks highlights the
-theoretical assumptions underlying their observed variables. SCM
-\ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
-nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
-are independent or not causally related. SCM \ref{fig-scm_bb2} and DAG
-\ref{fig-mdag_bb2} illustrate two connected nodes, representing a
-scenario where a \emph{parent} node \(X_{1}\) exerts a causal influence
-on a \emph{child} node \(X_{3}\). In this setup, \(X_{3}\) is considered
-a \emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
-are described as \emph{adjacent} because there is a \emph{direct path}
-connecting them. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict
-a \emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
-influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
-of \(X_{2}\), which is a parent node of \(X_{3}\). This structure
-creates a \emph{directed path} between \(X_{1}\) and \(X_{3}\).
-Consequently, \(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and
-\(X_{2}\) fully \emph{mediates} the relationship between the two. SCM
-\ref{fig-scm_bb4} and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork},
-where variables \(X_{1}\) and \(X_{3}\) are both influenced by
-\(X_{2}\). Here, \(X_{2}\) is a parent node that \emph{confounds} the
-relationship between \(X_{1}\) and \(X_{3}\). Finally, SCM
-\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} show a \emph{collider},
-where variables \(X_{1}\) and \(X_{3}\) are concurrent causes of
-\(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are not
-causally related to each other but both influence \(X_{2}\) (an
-\emph{inmorality}). Notably, all building blocks assume the errors are
-independent of each other and from all other variables in the graph, as
-evidenced by the pairwise relations \(e_{X1} \:\bot\:e_{X2}\),
-\(e_{X1} \:\bot\:e_{X3}\), and \(e_{X2} \:\bot\:e_{X3}\).
-
-Researchers can then use these building blocks to represent the scenario
-outlined in Section~\ref{sec-appendixB2}. SCM \ref{fig-scm_example1} and
-DAG \ref{fig-mdag_example1} depict the plausible causal structure for
-this example. In this context, the variable \(X\) (socio-economic status
-of the school) is thought to be a confounder in the relationship between
-the teaching method \(T\) and the outcome \(Y\). The figures display
-multiple descendant relationships such as \(X \rightarrow T\),
-\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
-unconnected node pairs, evident from the relationships
-\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
-\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
-\(X \rightarrow \{T, Y\}\), and two colliders:
-\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).
-
-\begin{figure}
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := f_{T}(X,e_{T}) \\
-  Y & := f_{Y}(T,X,e_{Y}) \\
-  e_{T} & \:\bot\:e_{X} \\
-  e_{T} & \:\bot\:e_{Y} \\
-  e_{X} & \:\bot\:e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_example1}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}
-
-}
-
-\subcaption{\label{fig-mdag_example1}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-example1}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendixB2}.}
-
-\end{figure}%
-
-\paragraph{The probabilistic implications of these
-blocks}\label{sec-appendixB32}
-
-Beyond their graphical capabilities, researchers can translate SCMs and
-DAGs into statistical models because these conceptual models encode
-functional and probabilistic information, which they can then replace
-with suitable functions and probabilistic assumptions
-\citep{Pearl_et_al_2016}. Notably, this encoding relies on two
-fundamental assumptions: the causal Markov and the faithfulness
-assumption. The \emph{causal Markov assumption} links the graph
-structure to the joint probability distribution of variables by
-asserting that any variable in a graph is independent of all its
-non-descendants conditional on its parents
-\citetext{\citealp[pp.~20]{Neal_2020}; \citealp[pp.~80]{Hernan_et_al_2025}}.
-The \emph{faithfulness assumption}, in turn, states that the observed
-statistical (in)dependencies in the data reflect the underlying causal
-structure among variables--except in rare degenerate cases
-\citetext{\citealp[pp.~100]{Neal_2020}; \citealp[pp.~81]{Hernan_et_al_2025}}.
-
-A notable implication of the assumptions underlying the probabilistic
-encoding is that any conceptual model described by an SCM and DAG can
-represent the joint distribution of variables more efficiently
-\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
-product of conditional probability distributions (CPDs) of the type
-\(P(child \mid parents)\). This property is formally known as the
-\emph{Bayesian Network factorization} (BNF,
-Equation~\ref{eq-net_factor})
-\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
-In this expression, \(pa(X_{i})\) denotes the set of variables that are
-the parents of \(X_{i}\).
-\begin{equation}\phantomsection\label{eq-net_factor}{
-\begin{aligned}
-P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
-& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}) ) & (\small{\text{by BNF}})
-\end{aligned}
-}\end{equation}
-
-This encoding enables researchers with conceptual (theoretical)
-knowledge in the form of an SCM and DAG to predict patterns of
-(in)dependencies in the data. As highlighted by
-\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
-structure of these conceptual models without requiring the quantitative
-details of the equations or the distributions of the errors. Moreover,
-once researchers observe empirical data, the patterns of
-(in)dependencies in the data can provide significant insights into the
-validity of the proposed conceptual model.
-
-The five fundamental building blocks described in
-Section~\ref{sec-appendixB31} clearly illustrate which (in)dependencies
-can SMCs and DAGs predict. For instance, applying the BNF to the causal
-structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
-enables researchers to express the joint probability distribution of the
-observed variables as \(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\),
-supporting the theoretical assumption that the observed variables
-\(X_{1}\) and \(X_{3}\) are unconditionally independent
-(\(X_{1} \:\bot\:X_{3}\)) \citep[pp.~24]{Neal_2020}. Conversely, when
-\(X_{3}\) is unconditionally dependent on \(X_{1}\)
-(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
-and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
-distribution as \(P(X_{1}, X_{3}) = P(X_{3} \mid X_{1}) P(X_{1})\).
-Notably, these descriptions demonstrate the clear correspondence between
-the structural equations illustrated in Section~\ref{sec-appendixB31}
-and the CPDs.
-
-Beyond the insights gained from two-node structures, researchers can
-uncover more nuanced patterns of(in)dependencies from chains, forks, and
-colliders. These (in)dependencies apply to any data set generated by a
-causal model with those structures, regardless of the specific functions
-attached to the SCM \citep[pp.~36]{Pearl_et_al_2016}. For instance,
-applying the BNF to the chain structure depicted in the SCM
-\ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} allow researchers to
-represent the joint distribution for the observed variables as
-\(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1}) P(X_{2} \mid X_{1}) P(X_{3} \mid X_{2})\). This expression
-implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
-\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
-controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \mid X_{2})\).
-Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
-DAG \ref{fig-mdag_bb4}, researchers can express the joint distribution
-of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1} \mid X_{2}) P(X_{2}) P(X_{3} \mid X_{2})\). Similar to the
-chain structure, this expression allows researchers to further infer
-that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
-\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
-controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \mid X_{2})\). Finally,
-researchers analyzing the collider structure illustrated in the SCM
-\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} can express the joint
-distribution of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
-\(P(X_{1}) P(X_{2} \mid X_{1}, X_{3}) P(X_{3})\). This representation
-allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
-unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
-dependent when controlling for \(X_{2}\)
-\((X_{1} \:\not\bot\:X_{3} \mid X_{2})\). The authors \citet[pp.~37, 40,
-41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
-mathematical proofs for these conclusions.
-
-Using these additional probabilistic insights, researchers can revisit
-the scenario in Section~\ref{sec-appendixB2}. In this context, applying
-the BNF to the SCM \ref{fig-scm_example2} structure, enables the
-representation of the joint probability distribution of the observed
-variables as \(P(Y, T, X) =\) \(P(Y \mid T, X) P(T \mid X) P(X)\). From
-this expression, researchers can infer that the outcome \(Y\) is
-unconditionally dependent on the teaching method \(T\)
-\((Y \:\not\bot\:T)\). This dependency arises from two key structures: a
-direct causal path from the teaching method \(T\) to the outcome \(Y\),
-represented by the two-connected-nodes structure \(T \rightarrow Y\)
-(black path in DAG \ref{fig-mdag_example2}), and a confounding
-non-causal path from the teaching method \(T\) to the outcome \(Y\)
-through the socio-economic status of the school \(X\), represented by
-the fork structure \(T \leftarrow X \rightarrow Y\) (gray path in DAG
-\ref{fig-mdag_example2}).
-
-\begin{figure}
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := x \\
-  T & := f_{T}(x,e_{T}) \\
-  Y & := f_{Y}(T,x,e_{Y}) \\
-  e_{T} & \:\bot\:e_{X} \\
-  e_{T} & \:\bot\:e_{Y} \\
-  e_{X} & \:\bot\:e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_example2}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1a.png}
-
-}
-
-\subcaption{\label{fig-mdag_example2}Conditioned DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-example2}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendixB2}.}
-
-\end{figure}%
-
-\paragraph{From probability to causality}\label{sec-appendixB33}
-
-The structural approach to causal inference translates probabilistic
-insights into actionable strategies seeking to identify the ACE from
-associational quantities. The approach achieves this by relying on the
-\emph{modularity assumption}, which posits that intervening on a node
-alters only the causal mechanism of that node, leaving others unchanged
-\citep[pp.~34]{Neal_2020}.
-
-The modularity assumption underpins the concepts of manipulated graphs
-and Truncated Factorization, which are essential for representing
-interventions \(P(Y_{i} \mid do(T_{i}=t))\) within SCMs and DAGs.
-\emph{Manipulated graphs} simulate physical interventions by removing
-specific edges from a DAG, while preserving the remaining structure
-unchanged \citep[pp.~34]{Neal_2020}. In parallel, \emph{Truncated
-Factorization} (TF) achieves a similar simulation by removing specific
-functions from the conceptual model and replacing them with constants,
-while keeping the rest of the structure unchanged \citep{Pearl_2010}.
-The probabilistic implications of this factorization are formalized in
-Equation~\ref{eq-trunc_factor}, where \(S\) represents the subset of
-variables \(X_{p}\) directly influenced by the intervention, while an
-example illustrating these concepts follows below.
-
-\begin{equation}\phantomsection\label{eq-trunc_factor}{
-P(X_{1}, X_{2}, \dots, X_{P} \mid do(S)) =
-\begin{cases}
-  \prod P(X_{p} \mid pa(X_{p}) ) & \text{if} \: p \not\in S \\
-  1 \quad & \text{otherwise}
-\end{cases}
-}\end{equation}
-
-Using the TF, researchers can define the \emph{backdoor adjustment} to
-identify the ACE. This adjustment states that if a variable
-\(X_{p} \in S\) serves as a \emph{sufficient adjustment set} for the
-effect of \(X_{a}\) on \(X_{b}\), then the ACE can be identified using
-Equation~\ref{eq-backdoor}. The sufficient adjustment set (potentially
-empty) must block all non-causal paths between \(X_{a}\) and \(X_{b}\)
-without introducing new paths. If such a set exists, then \(X_{a}\) and
-\(X_{b}\) are \emph{d-separated} by \(X_{p}\)
-(\(X_{a} \:\bot\:X_{b} \mid X_{p}\)) \citep{Pearl_2009}, and \(X_{p}\)
-satisfies the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020}.
-
-\begin{equation}\phantomsection\label{eq-backdoor}{
-P(X_{a} \mid do(X_{b}=x)) = \sum_{Xp} P(X_{a} \mid X_{b}=x, X_{p}) P(X_{p})
-}\end{equation}
-
-Ultimately, the backdoor adjustment enables researchers to express the
-ACE as:
-
-\begin{equation}\phantomsection\label{eq-backdoor_adjustment}{
-\begin{aligned}
-\tau & = E[X_{a} \mid do(X_{b}=1)]- E[X_{a} \mid do(X_{b}=2)] \\
-  & = E_{Xp}\left[ E[X_{a} \mid do(X_{b}=1), X_{p}]- E[X_{a} \mid do(X_{b}=2), X_{p}] \right] \\
-  & = \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=1, X_{p}) \cdot P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \mid X_{b}=2, X_{p}) \cdot P(X_{p})
-\end{aligned}
-}\end{equation}
-
-With these new insights, researchers revisiting the scenario in
-Section~\ref{sec-appendixB32} can infer that the socio-economic status
-of the school, \(X\), satisfies the backdoor criterion, assuming the
-causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
-\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
-sufficient adjustment set, as it effectively blocks all confounding
-non-causal paths introduced by the fork structure. Nevertheless, since
-\(Y\) remains dependent on \(T\) even after conditioning
-\((Y \:\not\bot\:T \mid X)\), this dependency can only be attributed to
-the direct causal effect \(T \rightarrow Y\). Notably, for the purpose
-of identification, the conditioned DAG \ref{fig-mdag_example2} is
-equivalent to the manipulated DAG \ref{fig-mdag_example3}, because \(X\)
-satisfies the backdoor criterion.
-
-\begin{figure}
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  X & := f_{X}(e_{X}) \\
-  T & := t \\
-  Y & := f_{Y}(t,X,e_{Y}) \\
-  e_{T} & \:\bot\:e_{X} \\
-  e_{T} & \:\bot\:e_{Y} \\
-  e_{X} & \:\bot\:e_{Y}
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-scm_example3}SCM}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1b.png}
-
-}
-
-\subcaption{\label{fig-mdag_example3}Manipulated DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-example3}Plausible causal structure the scenario
-outlined in Section~\ref{sec-appendixB32}.}
-
-\end{figure}%
-
-Researchers can then apply the \emph{backdoor adjustment} to identify
-the ACE of \(T\) on \(Y\). They achieve this by first identifying the
-CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
-this effect over \(X\) to obtain the ACE. This process is expressed in
-Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendixB2}).
-
-\begin{equation}\phantomsection\label{eq-mACE2}{
-\begin{aligned}
-  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
-  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right] \\
-  & = \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=1, X) \cdot P(X) - \sum_{X} Y_{i} \cdot P( Y_{i} \mid T_{i}=2, X) \cdot P(X)
-\end{aligned}
-}\end{equation}
-
-\paragraph{The estimation process}\label{sec-appendixB34}
-
-Ultimately, researchers can use Bayesian inference methods to estimate
-the ACE. The approach begins by defining two probability distributions:
-the likelihood of the data,
-\(P(X_{1}, X_{2}, \dots, X_{P} \mid \theta)\), and the prior
-distribution, \(P(\theta)\) \citep{Everitt_et_al_2010}, where \(X_{P}\)
-represents a random variable, and \(\theta\) represents a
-one-dimensional parameter space for simplicity. After observing
-empirical data, researchers can update the priors to posterior
-distributions using Bayes' rule in Equation~\ref{eq-bayes_rule}:
-
-\begin{equation}\phantomsection\label{eq-bayes_rule}{
-P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) = \frac{P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{P})}
-}\end{equation}
-
-Given that the denominator on the right-hand side of
-Equation~\ref{eq-bayes_rule} serves as a normalizing constant
-independent of the parameter \(\theta\), researchers can simplify the
-posterior updating process into three steps. First, they integrate new
-empirical data through the likelihood. Second, they update the
-parameters' priors to a posterior distribution according to
-Equation~\ref{eq-prop_rule}. Ultimately, they normalize these results to
-obtain a valid probability distribution.
-
-\begin{equation}\phantomsection\label{eq-prop_rule}{
-P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\mid \theta) \cdot P(\theta)
-}\end{equation}
-
-Temporarily setting aside the definition of prior distributions
-\(P(\theta)\), note that the posterior updating process depends heavily
-on the assumptions underlying the likelihood of the data. However, as
-the number of random variables, \(P\), increases, this joint
-distribution quickly becomes intractable \citep{Neal_2020}. This
-intractability is evident from Equation~\ref{eq-like_chain}, where the
-likelihood distribution is expressed by multiple chained CPDs.
-
-\begin{equation}\phantomsection\label{eq-like_chain}{
-P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid X_{i-1}, \dots, X_{1}, \theta )
-}\end{equation}
-
-Nevertheless, researchers can manage the complexity of the likelihood by
-assuming specific local (in)dependencies among variables. SCMs and DAGs
-provide a formal framework to represent these assumptions, as detailed
-in Section~\ref{sec-appendixB32}. These assumptions improve model
-tractability and simplify the estimation process by enabling the
-derivation of the BNF of the likelihood (Equation~\ref{eq-like_BNF}).
-With this simplified structure, any probabilistic programming language
-can model the system and compute the parameter's posterior distribution
-using Equation~\ref{eq-bayes_rule}.
-
-\begin{equation}\phantomsection\label{eq-like_BNF}{
-P(X_{1}, X_{2}, \dots, X_{P} \mid \theta) = P(X_{1} \mid \theta) \prod^{P}_{p=2} P(X_{i} \mid pa(X_{i}), \theta )
-}\end{equation}
-
-\newpage{}
-
 \section*{References}\label{references}

### 2025-06-09 paper2.qmd
review and editing

--- 2025-06-07 paper2.qmd
@@
 Thus, we define the matrix as in Equation ([-@eq-mat12]).
+
 $$
@@
 Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the form as in Equation ([-@eq-mat13]).
-
 $$
@@
 
-::: {#fig-cj04 layout-nrow=2}
-
-::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
-$$
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-$$
-
-SCM
-:::
-
-![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
-
-Conceptual-population model, initial vectorized form.
-:::
-
 <!-- 3. how de we use that data structure? -->
@@
 Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$, respectively.
+
+::: {#fig-cj04 layout-nrow=2}
+
+::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
+$$
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+$$
+
+SCM
+:::
+
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
+
+Conceptual-population model, initial vectorized form.
+:::
 
@@
 [^4]: refer to footnote 2.
+
+Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^5] judge-related variables $Z_{J}$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $B_{J}$.
+The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
+Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
+Furthermore, the branch shows how $B_{J}$, along with *relevant* [^6] judgment-related variables $Z_{JK}$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
+The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
+Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
+
+[^5]: refer to footnote 2.
+[^6]: refer to footnote 2.
+
+Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
+Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$ and the absence of connecting arrows.
+
+<!-- 3. segway for next section -->
+Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
+These enhancements include accounting for judges' biases and integrating hierarchical structural components.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
+Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
 
@@
 :::
-
-Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^5] judge-related variables $Z_{J}$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $B_{J}$.
-The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$.
-Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ independent columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$.
-Furthermore, the branch shows how $B_{J}$, along with *relevant* [^6] judgment-related variables $Z_{JK}$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the judges' biases associated with each text $B_{JK}$, the second primary latent variable of interest.
-The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
-Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
-
-[^5]: refer to footnote 2.
-[^6]: refer to footnote 2.
-
-Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
-Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$ and the absence of connecting arrows.
-
-<!-- 3. segway for next section -->
-Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
-These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
-Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
 
@@
 Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the resulting vectors would have the following structure:
-
 $$
@@
 <!-- This is a whole section [@Lawson_2015, pp. 16-112] -->
-
-<!-- 3. segway for next section -->
-However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020; ], CJ assessments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
-Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
-
-<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
-Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
-However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
-More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
-[@Boonen_et_al_2020, p.
-5] -->
 
@@
 
+<!-- 3. segway for next section -->
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ assessments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
+
+<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
+Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
+However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
+More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
+[@Boonen_et_al_2020, p.
+5] -->
+
 ### The comparison mechanism {#sec-theory-theoretical_SC2}
@@
 
-<!-- 3. preparing to transform the conceptual model into an statistical one -->
-Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
-This reformulation produces a model that applies directly to a sample of comparative data.
-In this version, the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$ are omitted, and $O_{R}$ inherits the structural equation $f_{O}$ that originally defined $O^{cp}_{R}$.
-Moreover, the definition of $O_{R}$ now reflects its direct dependence on the discriminal difference $D_{R}$ and the sample and comparison mechanisms, $S$ and $C$.
-
-<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
-For example, if I have the distribution of the population outcome, and it is a Bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
-
 ::: {#fig-cj15 layout-nrow=2}
@@
 :::
+
+<!-- 3. preparing to transform the conceptual model into an statistical one -->
+Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
+This reformulation produces a model that applies directly to a sample of comparative data.
+In this version, the unobserved outcomes $O^{cp}_{R}$ and $O^{sc}_{R}$ are omitted, and $O_{R}$ inherits the structural equation $f_{O}$ that originally defined $O^{cp}_{R}$.
+Moreover, the definition of $O_{R}$ now reflects its direct dependence on the discriminal difference $D_{R}$ and the sample and comparison mechanisms, $S$ and $C$.
+
+<!-- move the assumption for the conceptual population outcome to the data outcome based on the fact that it is logical to think that the data outcome would have the same distribution as the population outcome, if the data comes from that population.
+For example, if I have the distribution of the population outcome, and it is a Bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
 

### 2025-06-09 _tex/paper2.tex
Built site for gh-pages

@@
 individuals in the population. Thus, we define the matrix as in Equation
-(\ref{eq-mat12}). \begin{equation}\phantomsection\label{eq-mat12}{
+(\ref{eq-mat12}).
+
+\begin{equation}\phantomsection\label{eq-mat12}{
 R = \begin{bmatrix}
@@
 take the form as in Equation (\ref{eq-mat13}).
-
 \begin{equation}\phantomsection\label{eq-mat13}{
@@
 
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK})
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj04_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
-
-}
-
-\subcaption{\label{fig-cj04_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
-form.}
-
-\end{figure}%
-
 Now, using Equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
@@
 \(n_{J} \cdot n_{K}\), respectively.
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK})
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj04_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+
+}
+
+\subcaption{\label{fig-cj04_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
+form.}
+
+\end{figure}%
 
@@
 
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O^{cp}_{R} & := f_{O}(D_{R}) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} 
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj09_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
-
-}
-
-\subcaption{\label{fig-cj09_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj09}Conceptual-population model, final vectorized
-form.}
-
-\end{figure}%
-
 Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
@@
 assumption in practice, they must consider a more realistic scenario.
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O^{cp}_{R} & := f_{O}(D_{R}) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj09_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
+
+}
+
+\subcaption{\label{fig-cj09_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj09}Conceptual-population model, final vectorized
+form.}
+
+\end{figure}%
 
@@
 \(n_{K} = 3\), the resulting vectors would have the following structure:
-
 \begin{equation}\phantomsection\label{eq-mat24}{
@@
 each pair of sampled individuals.
-
 \begin{equation}\phantomsection\label{eq-mat26}{
@@
 within their respective groups \citep{Lawson_2015}.
-
-However, due to concerns about the practical feasibility of the
-comparison task {[}\citet{Boonen_et_al_2020}; {]}, CJ assessments rarely
-implement an exhaustive pairings of sampled judges, stimuli, and
-individuals. Thus, a realistic scenario must account for the fact that
-judges typically compare only a subset of stimuli authored by a sample
-of individuals.
 
@@
 \end{figure}%
+
+However, due to concerns about the practical feasibility of the
+comparison task \citep{Boonen_et_al_2020}, CJ assessments rarely
+implement an exhaustive pairings of sampled judges, stimuli, and
+individuals. Thus, a realistic scenario must account for the fact that
+judges typically compare only a subset of stimuli authored by a sample
+of individuals.
 
@@
 
-Finally, since it is standard to assume that the distribution of the
-conceptual-population outcome \(O^{cp}_{R}\) also holds for
-\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the sample-comparison
-model in Figure~\ref{fig-cj14} into the equivalent form shown in
-Figure~\ref{fig-cj15}. This reformulation produces a model that applies
-directly to a sample of comparative data. In this version, the
-unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\) are omitted, and
-\(O_{R}\) inherits the structural equation \(f_{O}\) that originally
-defined \(O^{cp}_{R}\). Moreover, the definition of \(O_{R}\) now
-reflects its direct dependence on the discriminal difference \(D_{R}\)
-and the sample and comparison mechanisms, \(S\) and \(C\).
-
 \begin{figure}[H]
@@
 \end{figure}%
+
+Finally, since it is standard to assume that the distribution of the
+conceptual-population outcome \(O^{cp}_{R}\) also holds for
+\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the sample-comparison
+model in Figure~\ref{fig-cj14} into the equivalent form shown in
+Figure~\ref{fig-cj15}. This reformulation produces a model that applies
+directly to a sample of comparative data. In this version, the
+unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\) are omitted, and
+\(O_{R}\) inherits the structural equation \(f_{O}\) that originally
+defined \(O^{cp}_{R}\). Moreover, the definition of \(O_{R}\) now
+reflects its direct dependence on the discriminal difference \(D_{R}\)
+and the sample and comparison mechanisms, \(S\) and \(C\).
 

### 2025-06-25 paper2.qmd
review and editing

@@
 It also lays the foundation for broader applications, encouraging researchers across the social sciences to adopt more robust and interpretable methodologies.
-key-points:
-  - (to do)
 date: last-modified
@@
 Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to have a higher trait level.
-For example, when assessing writing quality, judges compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, when assessing writing quality, judges compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Pollitt_2012b; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes.
-Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes.
+Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
@@
 <!-- 4. what is the organization of the study? -->
-As a result, the study divides its content into six main sections.
+The study divides its content into six main sections.
 @sec-thurstone_theory provides an overview of Thurstone's theory.
@@
 
+![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}
+
 <!-- 2. Thurstone's Cases and their simplifying assumptions -->
@@
 @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases.
-For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp.
-248–253].
+For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008.
 
@@
 <!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->
-
-![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}
 
@@
 269], his theory did not support formal statistical inference.
-Despite these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, which raises significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978].
-Furthermore, although the CJ tradition has attempted to address the gap of hypothesis testing by using the point estimates of the traits--or their transformations--a critical question remains: Does this approach provide a valid foundation for statistical inference?
+Despite these limitations, CJ research has predominantly relied on Case V to measure different traits, which raises significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978].
+Furthermore, although the CJ tradition has attempted to address the gap in hypothesis testing by relying on the point estimates of traits--or their transformations--the statistical literature cautions against using these estimates as the sole basis for statistical inference, as such practices introduce bias in the analysis and reduce the precision of hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+Next, both issues are discussed in more depth.
 
@@
 [@Andrich_1978, pp.459] -->
-
-<!-- 4. Section division -->
-Thus, this section discusses these two prominent issues.
-Specifically, @sec-theory-issue1 examines the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data.
-Conversely, @sec-theory-issue2 focuses on the apparent disconnect between the approaches to trait measurement and hypothesis testing in CJ.
 
@@
 Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019], in this case, the assumption of equal dispersions.
-For instance, when the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts.
+When the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts.
 This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
@@
 <!-- 3. But does it hold in practice? -->
-Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail accommodate for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
-Indeed, the presence of the so-called *misfit texts*--texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]--may signal this limitation, as such discrepancies can arise from larger discriminal dispersions due to stimulus heterogeneity or because the texts are genuine outliers--that is, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, the BTL model's assumptions prevent it from adequately accounting for or addressing these anomalies, leaving exclusion of the "problematic" texts as the primary remedy.
+Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to accommodate for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
+The presence of the so-called *misfit texts*--texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]--may signal this limitation, as such discrepancies can arise from larger discriminal dispersions due to stimulus heterogeneity or because the texts are genuine outliers--that is, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
+In either case, the BTL model's assumptions prevent it from adequately accounting for or addressing these anomalies, leaving exclusion of the "problematic" texts as the primary remedy [@Pollitt_2012a; @Pollitt_2012b].
 
@@
 
-<!-- A similar statistic is calculated for every script.
-A misfitting script is one that judges find difficult for some reason, one that some judges value more than others do.
-[@Pollitt_2012b, pp.
-290] -->
+<!-- But from a measurement stance, reliable assessment requires judges that are consistent with each other, and at any time, misﬁtting judges can be stopped, and their share of decisions shared amongst the other judges.
+If necessary, their data can be removed from the file.
+...
+A similar statistic is calculated for every script. A misfitting script is one that judges find difficult for some reason, one that some judges value more than others do. [@Pollitt_2012b, pp. 290] -->
 
@@
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Furthermore, if researchers acknowledge that *misfit statistics* may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics may unintentionally discard valuable information [@Miller_2023], introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
+Furthermore, if researchers acknowledge that *misfit statistics* may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics may unintentionally discard valuable information [@Miller_2023], and introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
 The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
@@
 <!-- 1. What does the zero correlation between stimuli imply? -->
-The correlation $\rho$, illustrated in @fig-discriminal_difference, measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another.
+The correlation between two stimuli $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in other stimulus.
 Similar to the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons.
-A thought experiment, akin to the one presented in @sec-theory-issue1a, can illustrate this idea.
-Assuming that the discriminal dispersions for both texts remain constant, @fig-correlation shows that as the correlation between the two texts increases, the distribution of their discriminal difference becomes narrower.
+Assuming that the discriminal dispersions for a couple of texts remain constant, @fig-correlation shows that as the correlation between the two texts increases, the distribution of their discriminal difference becomes narrower.
 This narrowing, in turn, affects the probability that the discriminal difference distinctly favors Text B over Text A--denoted as $P(B > A)$--and thus directly influences the comparison outcome.
@@
 This idea was later formalized by @Andrich_1978, who provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
-However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.
+However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two cases: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.
 
@@
 <!-- 2. The first scenario where the assumption does not hold -->
-In the first scenario, research on text quality suggests that when judges evaluate complex, multidimensional traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022].
+Research on text quality assessments suggests that when judges evaluate complex, multidimensional traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022].
 Regardless of their relevance, these characteristics may not receive equal weight or consistently oppose one another across comparisons.
@@
 Moreover, because the discriminal process of stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
-While direct evidence for this exact scenario is limited, existing studies document the presence of judge bias in CJ [@Pollitt_et_al_2003, @vanDaal_et_al_2016, @Bartholomew_et_al_2020a], reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.
+While direct evidence for this exact scenario is limited, existing studies document the presence of judge bias in CJ [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.
 
@@
 <!-- 3. The second scenario where the assumption does not hold -->
-In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli--a statistical phenomenon known as clustering [@Everitt_et_al_2010].
-For instance, when the same individual produces multiple texts, those texts often exhibit common features, such as writing style or overall quality, that judges can easily recognize.
-In this regard, although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient.
+In the second case, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli--a statistical phenomenon known as clustering [@Everitt_et_al_2010].
+For instance, when the same individual produces multiple texts, those texts often share several features such as writing style or overall quality.
+Although the CJ literature acknowledges the existence of such hierarchical structures [e.g., @Boonen_et_al_2020], the statistical approaches to account for this additional source of dependence have been insufficient.
 For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
-However, this approach can introduce additional statistical and measurement issues, similar to the ones discussed in @sec-theory-issue2.
 
@@
 Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues.
-Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023] and even biases [@Wu_et_al_2022].
-This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$.
-Moreover, these inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli--potentially causing the model to incorrectly classify stimuli or judges as *misfits*.
+In particular, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023] and even introduce statistical biases [@Wu_et_al_2022].
+These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli.
+One such spurious conclusion could be the incorrect classification of stimuli or judges as *misfits*.
+@fig-correlation illustrates how assuming zero correlation can undermine trait reliability: the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference may correspond to any discontinuous line where $\rho \neq 0$.
 
@@
 
-Finally, similar to what we discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and may appear as outliers under the BTL model [@Wu_et_al_2022]--risk discarding valuable information and introducing bias into trait estimates [@Miller_2023].
+Finally, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and may appear as outliers under the BTL model [@Wu_et_al_2022]--risks discarding valuable information and introducing bias into trait estimates [@Miller_2023].
 The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
@@
 <!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Building on the previous section, it is clear that, researchers in CJ studies typically use the BTL model to measure traits and position the compared stimuli along a latent continuum [@Thurstone_1927b].
-The CJ literature also show that they frequently rely on point estimates of these traits--typically the BTL scores or its transformations--to conduct statistical inference or hypothesis testing.
+Researchers in CJ studies typically use the BTL model to measure traits and position the compared stimuli along a latent continuum [@Thurstone_1927b].
+The CJ literature shows that research frequently relies on point estimates of these traits--typically the BTL scores or its transformations--to conduct statistical inference or hypothesis testing.
 For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 This overreliance may also undermine their validity [@Perron_et_al_2015], particularly when coupled with the disconnect between the trait measurement and hypothesis testing processes.
-However, the structural approach to causal inference can address these issues by offering a systematic and integrated framework that strengthens measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.
+The structural approach to causal inference can address these issues by offering a systematic and integrated framework to extend Thurstone's general form.
+This approach can also strengthen measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.
 
@@
 <!-- 3. Example of a CJ estimand and estimator, and its limitations -->
-A motivating example that will appear throughout this document clarifies these concepts.
+A motivating example that will appear throughout this study clarifies these concepts.
 In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ assessment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
@@
 
-Following standard CJ practices, researchers typically use estimates from the BTL model--or its transformations--to approximate this estimand.
+Following standard CJ practices, researchers would typically use estimates from the BTL model--or its transformations--to approximate this estimand.
 However, as discussed in @sec-theory-issues, Thurstone's Case V and the BTL model exhibit several statistical and measurement limitations.
@@
 
-[^1]: These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, and introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 are useful.
+[^1]: In depth explanation of these topics is beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, and introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 are useful.
 For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2025, are recommended.
@@
 Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
-In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages.
-First, it clarifies the interactions among all actors and processes involved in CJ assessments.
-Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.
-
-<!-- 7. How the section is divided? -->
-Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we refer to as the *conceptual-population model*.
-This model assumes an idealized scenario where researchers have access to a *conceptual population* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
-Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we refer to as the *sample-comparison model*.
-This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
+@sec-theory-theoretical_P introduces the *conceptual-population model*, which incorporates these theoretical principles and assumes an idealized setting where researchers observe a *conceptual population* of comparative judgment data--that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+Conversely, @sec-theory-theoretical_SC presents the *sample-comparison model*, which integrates the assessment design features and reflects a more realistic setting where researchers access only a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
 
@@
 
-::: {#fig-cj03 layout-nrow=2}
+::: {#fig-cj03 layout-ncol=2}
 
@@
 
-![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=35%}
+![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=77%}
 
@@
 <!-- 1. Why is useful to declare the data structure? -->
-Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system.
+Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one improves clarity and facilitates the description of the system.
 Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as in Equation ([-@eq-mat11]).
@@
 Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
+
+Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
+This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
+Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
+Thus, we define the matrix as in Equation ([-@eq-mat11]).
+
 $$
@@
 n_{I}
-\end{bmatrix} ; \;
+\end{bmatrix};
 J = \begin{bmatrix}
@@
 n_{J}
-\end{bmatrix} ; \;
+\end{bmatrix};
 IA = \begin{bmatrix}
@@
 n_{I} & n_{A}
-\end{bmatrix} ; \;
+\end{bmatrix};
 JK = \begin{bmatrix}
@@
 n_{J} & n_{K}
-\end{bmatrix}
-$$ {#eq-mat11}
-
-Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix.
-This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$.
-Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population.
-Thus, we define the matrix as in Equation ([-@eq-mat12]).
-
-$$
+\end{bmatrix};
 R = \begin{bmatrix}
@@
 \end{bmatrix}
-$$ {#eq-mat12}
+$$ {#eq-mat11}
 
 It is easier to visualize the structure of the previously defined vectors and matrices by considering an example.
-Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the form as in Equation ([-@eq-mat13]).
+Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equation ([-@eq-mat11]) take the form as in Equation ([-@eq-mat13]).
 $$
@@
 <!-- 3. how de we use that data structure? -->
-Now, using Equations ([-@eq-mat11]) and ([-@eq-mat12]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
+Now, using Equation ([-@eq-mat11]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04.
 In this depiction, the outcome $O^{cp}_{R}$, the texts' discriminal difference $D_{R}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values.
@@
 
-::: {#fig-cj04 layout-nrow=2}
+::: {#fig-cj04 layout-ncol=2}
 
@@
 
-![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}
+![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=87%}
 
@@
 The sample-comparison model presents a more realistic scenario than the conceptual-population model.
-First, in @sec-theory-theoretical_SC1, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population.
-Second, in @sec-theory-theoretical_SC2, the model assumes that judges do not perform *all repeated judgments* within the data sample.
+First, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population (@sec-theory-theoretical_SC1).
+Second, the model assumes that judges do not perform *all repeated judgments* within the data sample (@sec-theory-theoretical_SC2).
 Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$, as proposed by @Thurstone_1927b.
@@
 Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
+
 $$
@@
 <!-- This is a whole section [@Lawson_2015, pp. 16-112] -->
+
+<!-- 3. segway for next section -->
+However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ assessments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
+Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
+
+<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
+Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
+However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
+More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
+[@Boonen_et_al_2020, p.
+5] -->
 
@@
 
-<!-- 3. segway for next section -->
-However, due to concerns about the practical feasibility of the comparison task [@Boonen_et_al_2020], CJ assessments rarely implement an exhaustive pairings of sampled judges, stimuli, and individuals.
-Thus, a realistic scenario must account for the fact that judges typically compare only a subset of stimuli authored by a sample of individuals.
-
-<!-- An exhaustive pairing of all stimuli would lead to a total of 7875 possible pairs for each listener group.
-Obviously, such a large number of pairs would undermine the practical feasibility of the experiment.
-However, in order to arrive at a reliable ranking, not all possible combinations of stimuli had to be assessed.
-More specifically, each stimulus had to be judged by the listeners of each particular listener group – together, not individually – at least 20 times.
-[@Boonen_et_al_2020, p.
-5] -->
-
 ### The comparison mechanism {#sec-theory-theoretical_SC2}
@@
 <!-- 2. assumptions about the outcome -->
-Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the statistical model [-@fig-cj16_stat].
-We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^8], reflecting the binary nature of CJ outcomes.
+Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in the statistical model presented in @fig-cj16_stat.
+The model starts by assuming that $O_{R}$ follows a Bernoulli distribution[^8], reflecting the binary nature of CJ outcomes.
 Furthermore, following the conventions of Generalized Linear Models (GLMs) [@Nelder_et_al_1983; @Nelder_et_al_1996; @Agresti_2015], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.
@@
 <!-- 3. Note about the sample and comparison mechanism, as well as the predictor's distribution -->
-Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit all of these probabilities from the statistical model [-@fig-cj16_stat].
+While the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--all of these probabilities are omitted from the statistical model [-@fig-cj16_stat].
 This omission is justified because, while these distributions contribute to the overall joint distribution of the data, the variables $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$ are observed and independent of any other variable in the model.
 As observed variables, they do not require distributional assumptions in the same way the idiosyncratic errors do.
-Furthermore, their independence follows from the underlying random selection procedures that govern the variables[^9].
+Their independence follows from the underlying random selection procedures that govern the variables[^9].
 
@@
 <!-- 4. functional form for D_{R} -->
-Next we define $D_{R}$ as the difference between the discriminal processes $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the compared texts, plus the corresponding repeated judge bias $B_{JK}[j, k]$.
-Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as:
+Next $D_{R}$ is defined as the difference between the discriminal processes $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the compared texts, plus the corresponding repeated judge bias $B_{JK}[j, k]$.
+Note that if it is assumed that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, the discriminal difference can be re-written as:
 $$
@@
 <!-- 5. Functional form and assumptions -->
-We now specify the functional forms for $T_{IA}$, $T_{I}$, $B_{JK}$, and $B_{J}$.
-We model $T_{IA}$ as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables on quality assessment $\beta_{XA}X_{IA}$ (such as the influence of text length), and the text-specific idiosyncratic errors $e_{IA}$.
-Similarly, we express $T_{I}$ as a linear combination of relevant student-related variables affecting the quality assessment $\beta_{XI} X_{I}$, and student-specific idiosyncratic errors $e_{I}$.
-For the judge-specific terms, we model $B_{JK}$ as a linear combination of the judge's individual bias $B_{J}$, the influence of relevant judgment-related variables on quality assessment $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments affect the evaluation), and judgment-specific idiosyncratic errors $e_{JK}$.
-Finally, we define $B_{J}$ as a linear combination of relevant judge-level variables influencing the quality assessment $\beta_{ZJ}Z_{J}$ (such as judgment expertise) and judge-specific idiosyncratic errors $e_{J}$.
+Now the functional forms for $T_{IA}$, $T_{I}$, $B_{JK}$, and $B_{J}$ are specified.
+$T_{IA}$ is modeled as a linear combination of the students' underlying writing-quality traits $T_{I}$, the effects of relevant text-related variables on quality assessment $\beta_{XA}X_{IA}$ (such as the influence of text length), and the text-specific idiosyncratic errors $e_{IA}$.
+Similarly, $T_{I}$ is expressed as a linear combination of relevant student-related variables affecting the quality assessment $\beta_{XI} X_{I}$, and student-specific idiosyncratic errors $e_{I}$.
+For the judge-specific terms, $B_{JK}$ is modeled as a linear combination of the judge's individual bias $B_{J}$, the influence of relevant judgment-related variables on quality assessment $\beta_{ZK}Z_{JK}$ (e.g., how the number of judgments affect the evaluation), and judgment-specific idiosyncratic errors $e_{JK}$.
+Finally, $B_{J}$ is defined as a linear combination of relevant judge-level variables influencing the quality assessment $\beta_{ZJ}Z_{J}$ (such as judgment expertise) and judge-specific idiosyncratic errors $e_{J}$.
 
 <!-- 6. error assumptions -->
-Next, we specify the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$.
+Next, the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$ are specified.
 Unlike other variables in the model, these error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables $T_{I}$, $T_{IA}$, $B_{J}$, and $B_{JK}$.
-Thus, to identify the latent variable model, we must resolve these indeterminacies [@Depaoli_2021; @deAyala_2009].
-Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{V} \boldsymbol{Q} \boldsymbol{V}$, with $\boldsymbol{V}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{Q}$ a correlation matrix.
-To address the *location* indeterminacy, we set the errors' mean vector to zero:
+Thus, to identify the latent variable model these indeterminacies must be resolved [@Depaoli_2021; @deAyala_2009].
+Drawing on principles from SEM [@Hoyle_et_al_2023], the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$ are assumed to follow a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{V} \boldsymbol{Q} \boldsymbol{V}$, with $\boldsymbol{V}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{Q}$ a correlation matrix.
+To address the *location* indeterminacy, the errors' mean vector is set to zero:
 $$
@@
 
-Following [SCM -@fig-cj16_scm], we resolve the *orientation* indeterminacy by assuming that the errors are uncorrelated.
-This assumption leads us to define the error correlation matrix, $\boldsymbol{Q}$, as the identity matrix:
+Following [SCM -@fig-cj16_scm], the *orientation* indeterminacy is solved by assuming that the errors are uncorrelated.
+This assumption leads to the definition of the error's correlation matrix, $\boldsymbol{Q}$, as the identity matrix:
 $$
@@
 
-To resolve the *scale* indeterminacy, we define the diagonal matrix $\boldsymbol{V}$ as:
+To resolve the *scale* indeterminacy, the diagonal matrix $\boldsymbol{V}$ is defined as follows:
 $$
@@
 Here, $s_{XI}$ represents the standard deviation for the individuals, $p_{IA}$ for the stimuli, $s_{ZJ}$ for the judges, and $p_{JK}$ for the judgments.
-We assume $s_{XI}$ varies depending on the teaching method group to which each student belongs.
+It is assumed $s_{XI}$ varies depending on the teaching method group to which each student belongs.
 Using the example from @sec-theoretical, where the teaching method $X_{I} = \{1,2\}$, the model sets the constraint according to Equation ([-@eq-mat56]).
@@
 Because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$.
-We set $P_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait.
+Thus, $p_{IA}$ is set as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait.
 Note that as a result, $T_{IA}$ is also normally distributed.
-This configuration effectively reinstates Thurstone’s original assumption of Normal discriminal processes for the stimuli (see @tbl-thurstone_cases).
-
-Similarly, we assume $s_{ZJ}$ varies depending on the groups to which each judge belongs.
+This configuration effectively reinstates Thurstone's original assumption of Normal discriminal processes for the stimuli (see @tbl-thurstone_cases).
+
+Similarly, it is assumed that $s_{ZJ}$ varies depending on the groups to which each judge belongs.
 For instance, if $Z_{J} = \{1,2,3\}$ represents three groups of judges with varying expertise, the model sets the constraint according to Equation ([-@eq-mat57]).
@@
 Second, it incorporates prior information to constrain parameter estimates within plausible bounds, thereby mitigating estimation issues like non-convergence or improper solutions that often affect frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011].
-We primarily use this capability to define the error distribution and set the scale of latent variables [@Depaoli_2014].
+Prior distributions are used to define the error distribution and set the scale of latent variables [@Depaoli_2014].
 Third, Bayesian inference supports robust inferences from small samples, where the asymptotic properties underlying frequentist methods are less reliable [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].
@@
 269], his theory offered no guidance on how to use trait estimates for statistical inference.
-The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing, relying on point estimates--such as BTL scores or their transformations--for inference.
-Notably, while this approach simplifies analysis, it can also introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing, relying on point estimates, such as BTL scores or their transformations, for inference.
+While this approach simplifies analysis, it can also introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
 
 <!-- 4. General summary of what we have done. -->
-Thus, to address the limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods.
+To address the limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods.
 The approach begins with the development of a conceptual model, represented as a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes underlying the CJ system.
-
-The approach then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur (see @sec-theory-issue1a and @sec-theory-issue1b) and when statistical inference is necessary (refer to @sec-theory-issue2).
-In particular, this model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical).
+This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms, to disentangle the causal processes underlying the CJ system.
+
+The approach then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur, and when statistical inference is necessary.
+In particular, this model accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process, and accommodates heterogeneity in discriminal dispersions.
 By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement in CJ [@Perron_et_al_2015], while also improving the accuracy of statistical inferences.
@@
 
-<!-- These procedures also bear on a larger topic of discussion in social science.
-The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
-[@Wainer_et_al_1978, pp.
-332] -->
-
 ## Future research directions using our approach {#sec-discussion_RA}
@@
 While these studies offer valuable insights, they also overlook the broader role that these mechanisms play within the CJ system.
-As this oversight likely stems from a more fundamental lack of conceptual clarity about how these mechanisms function within the system, this study decided to integrate these mechanisms into the conceptual model of CJ.
+As this oversight likely stems from a more fundamental lack of conceptual clarity about how these mechanisms function within the system, this study integrates these mechanisms into the conceptual model of CJ.
 
@@
 The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process.
-Specifically, it clarifies their role as sources of missing data in CJ's data-generating process--that is, as mechanisms that determine which observations are missing from the final data sample.
+Specifically, it clarifies their role as sources of missing data in CJ's data-generating process, that is, as mechanisms that determine which observations are missing from the final data sample.
 This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], allowing a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex assessments setups.
@@
 <!-- 3. Leveraging MCAR assumptions to simplify CJ analysis in this study -->
-Notably, this study circumvents the need to apply this missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome.
-In other words, we designed these mechanisms so they produce data that are *missing completely at random* (MCAR) [@Little_et_al_2020].
+This study circumvents the need to apply this missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome.
+In other words, these mechanisms are designed to produce data that are *missing completely at random* (MCAR) [@Little_et_al_2020].
 This design offers one key advantage: it generates simple random samples that satisfy the condition of *ignorability*, allowing researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019; @Neal_2020].
@@
 <!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
-However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
-A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
+However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
+A prominent example is the previously discussed ACJ design, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
 This pair selection process suggests that ACJ's comparison mechanism is outcome-dependent, potentially classifying the method as a generator of MNAR data.
@@
 <!-- 5. The importance of clarifying assumptions behind CJ assessment designs -->
-Nevertheless, regardless of the underlying missingness mechanisms, any CJ assessment design would benefit from explicitly defining its assumptions--a practice supported by our approach.
+Regardless of the underlying missingness mechanisms, any CJ assessment design would benefit from explicitly defining its assumptions--a practice supported by our approach.
 This clarity enables researchers to evaluate how the sampling and comparison mechanisms affect trait estimation and statistical inference within each design.
@@
 This question is particularly relevant because many researchers still regard "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases.
-Another critical question, discussed in @sec-theory-issue1b, is whether training or expertise can help judges avoid focusing on irrelevant stimulus features, such as handwriting, over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022].
-Notably, exploring these questions may also provide insights into what it truly means to be an "expert" within the CJ context [@Kelly_et_al_2022].
+Another critical question is whether training or expertise can help judges avoid focusing on irrelevant stimulus features, such as handwriting, over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022].
+Exploring these questions may also provide insights into what it truly means to be an "expert" within the CJ context [@Kelly_et_al_2022].
 
@@
 In addition, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
-As anticipated, our approach provides the necessary structure to investigate rigorously these questions.
+As anticipated, the approach presented in this study provides the necessary structure to investigate rigorously these questions.
 
@@
 Specifically, *misfit judges* are typically defined as those whose assessments diverge significantly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], while *misfit stimuli* are those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012a; @Pollitt_2012b; @Goossens_et_al_2018].
-Both definitions closely mirror the statistical concept of outliers--that is, observations that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969].
+Both definitions closely mirror the statistical concept of outliers, that is, observations that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969].
 But this resemblance extends beyond the definitions themselves, as CJ researchers often identify misfits using conventional outlier detection procedures, such as transforming BTL model residuals into diagnostic statistics and comparing them against predefined thresholds [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022].
@@
 
-Nevertheless, it is not the classification of misfits as outliers that raises concerns for trait measurement and inference, but rather the prevailing CJ practice of identifying them through these ad-hoc detection procedures and then excluding them from analysis, often without providing empirical evidence for the various hypotheses proposed to explain their occurrence in the first place.
+Nevertheless, it is not the classification of misfits as outliers that raises concerns for trait measurement and inference.
+Rather, the concern lies in the prevailing CJ practice of classify them through ad-hoc detection procedures and then excluding them from analysis [@Pollitt_2012a; @Pollitt_2012b], often without providing empirical evidence for the various hypotheses proposed to explain their occurrence.
 In this regard, it is essential to recognize that outliers can only be defined relative to a specific model [@McElreath_2020].
-Thus, identification procedures based on models like the BTL, which relies on strong assumptions, should be approached with caution, as they may not be well-suited for the task [@Kelly_et_al_2022].
-Specifically, because the BTL model may not always accurately reflect the underlying data-generating process of the CJ system (see @sec-theory-issues), detection procedures grounded in this model risk compromising the accurate identification of misfits.
-Moreover, the statistical literature cautions that excluding outliers risks discarding valuable information [@Miller_2023] and introducing bias into trait estimates.
+As such, detection procedures based on models like the BTL, which relies on strong assumptions, should be approached with caution, as they may not be well-suited for the task [@Kelly_et_al_2022].
+In particular, because the BTL model may not always accurately reflect the underlying data-generating process of the CJ system, misfit classification based on this model risks misidentifying observations.
+Furthermore, the exclusion of these observations carries additional risks.
+The statistical literature cautions that removing outliers can discard valuable information [@Miller_2023] and introduce bias into trait estimates.
 The direction and magnitude of this bias are often unpredictable, as they depend on which observations researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
-Lastly, regarding the hypotheses proposed to explain the presence of misfits, it is evident that the BTL model lacks the flexibility needed to adequately test many of them, largely due to its strong assumptions.
+Finally, precisely because of its rigid assumptions, the BTL model lacks the flexibility to adequately test many of the hypotheses proposed to explain the presence of misfits.
 
@@
 
-<!-- "Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
+<!-- Another challenge to the intrinsic validity argument is that the aggregate approach and valuing of diversity in expert views (Van Daal et al., 2019) is in tension with the idea of removing misfitting judges (Pollitt, 2012a).
 If someone is considered an expert, and the validity of comparative judgment rests on the collective expertise of a community of practice, the decision to remove them and discard their judgments can only be justifiable on the assumption that those judgments do not reflect their expertise.
 ...
-It is thus unclear whether or not this divergence is welcomed." [@Kelly_et_al_2022, pp.
-681] -->
-
-In contrast, our approach provides a rigorous framework that enables the examination of several relevant hypotheses.
+However, this is not the usual argument proposed in such cases, because as Van Daal et al.
+(2019) note, misfitting judges are generally supposed to diverge in their conceptualisation of the construct (e.g.
+Bramley, 2007; Roose et al., 2019; Whitehouse & Pollitt, 2012).
+It is thus unclear whether or not this divergence is welcomed.
+[@Kelly_et_al_2022, pp.
+681-682] -->
+
+In contrast, the approach presented in this study provides a rigorous framework that enables the examination of several relevant hypotheses.
 For instance, researchers could investigate whether misfit judges are those who exhibit (an outlying degree of) systematic bias or greater variability in their judgments compared to their peers.
@@
 Moreover, since outliers are not necessarily "bad data" [@McElreath_2020], our approach also offers a principled alternative that retains misfits in the analysis without compromising trait estimation or inference.
-This alternative involves adapting our proposed models into robust measurement models [@McElreath_2020]--a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].
+This alternative involves adapting the proposed model into robust measurement models [@McElreath_2020], a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].
 This alternative also relates to a broader discussion in social science, which is: that when researchers face low predictive capacity from their models, they should not only search for "new" variables or alternative procedures but also consider employing more sophisticated measurement models [@Wainer_et_al_1978].
@@
 The process of deriving conclusions from observed data always requires assumptions, whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022].
-Our approach is not an exception to this fundamental principle.
-As with all approaches grounded on causal inference, ours has the limitation of relying on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
+The proposed approach is not an exception to this fundamental principle.
+As with all approaches grounded on causal inference, it relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
 As such, the approach does not seek to yield automatic answers when applied to a given CJ dataset.
@@
 This clarity is critical because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019].
-Although this alignment remains empirically untested for the proposed models, the theory-driven nature of our approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], which are well supported by both theory and existing evidence (see @sec-theory-issues and @sec-theoretical).
+Although this alignment remains empirically untested for the proposed models, the theory-driven nature of our approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], which are well supported by both theory and existing evidence.
 
@@
 
-To apply our approach effectively, CJ researchers must tackle two main challenges related to the required foundational knowledge.
+To apply the approach effectively, CJ researchers require foundational knowledge in two areas.
 First, they need a solid understanding of causal inference principles.
@@
 As a result, CJ researchers must deeply engage with these and other complex ideas, including how SCMs encode functional and probabilistic information.
-To assist this effort, this study includes key references throughout the study to guide motivated aaplied CJ researchers toward a deeper understanding of these concepts [^10].
+To assist this effort, this study includes key references throughout the study to guide applied CJ researchers toward a deeper understanding of these concepts [^10].
 
@@
 Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research.
-Importantly, our approach supports both strategies by providing the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
+Notably, our approach supports both strategies by providing the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
 

### 2025-06-25 _tex/paper2.tex
Built site for gh-pages

@@
 
+\usepackage{float}
+\floatplacement{table}{H}
 \makeatletter
@@
 determine the relative writing quality each text exhibit (the trait)
-\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
+\citep{Pollitt_2012b, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
 
@@
 undermine the reliability of various statistical inferences derived from
-these processes.
+these processes
+\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
 
@@
 
-As a result, the study divides its content into six main sections.
+The study divides its content into six main sections.
 Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
@@
 
+\begin{table}
+
+\caption{\label{tbl-thurstone_cases}Thurstones cases and their
+asumptions}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/thurstone_cases.png}
+
+}
+
+\end{table}%
+
 To address this issue and facilitate the practical implementation of the
@@
 discussion of these cases and their progression, refer to
-\citet{Thurstone_1927b} and \citet[pp.~248--253]{Bramley_2008}.
-
-\begin{table}
-
-\caption{\label{tbl-thurstone_cases}Thurstones cases and their
-asumptions}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/thurstone_cases.png}
-
-}
-
-\end{table}%
+\citet{Thurstone_1927b} and \citet{Bramley_2008}.
 
@@
 \citep[pp.~269]{Thurstone_1927b}, his theory did not support formal
-statistical inference. Despite these limitations, it is surprising that
-CJ research has predominantly relied on Case V to measure different
-traits, which raises significant concerns about the reliability and
-validity of such measurements in contexts where the case's assumptions
-may not hold \citep{Kelly_et_al_2022, Andrich_1978}. Furthermore,
-although the CJ tradition has attempted to address the gap of hypothesis
-testing by using the point estimates of the traits--or their
-transformations--a critical question remains: Does this approach provide
-a valid foundation for statistical inference?
-
-Thus, this section discusses these two prominent issues. Specifically,
-Section~\ref{sec-theory-issue1} examines the heavy reliance on
-Thurstone's Case V assumptions in the statistical analysis of CJ data.
-Conversely, Section~\ref{sec-theory-issue2} focuses on the apparent
-disconnect between the approaches to trait measurement and hypothesis
-testing in CJ.
+statistical inference. Despite these limitations, CJ research has
+predominantly relied on Case V to measure different traits, which raises
+significant concerns about the reliability and validity of such
+measurements in contexts where the case's assumptions may not hold
+\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
+tradition has attempted to address the gap in hypothesis testing by
+relying on the point estimates of traits--or their transformations--the
+statistical literature cautions against using these estimates as the
+sole basis for statistical inference, as such practices introduce bias
+in the analysis and reduce the precision of hypothesis tests
+\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Next, both
+issues are discussed in more depth.
 
@@
 assumptions \citep{Kohler_et_al_2019}, in this case, the assumption of
-equal dispersions. For instance, when the assumption accurately captures
-the complexity of the data, the BTL model estimates a discriminal
-difference distribution that accurately represents the \emph{true}
-discriminal difference between the texts. This scenario is illustrated
-in Figure~\ref{fig-dispersion}, when the model's discriminal difference
+equal dispersions. When the assumption accurately captures the
+complexity of the data, the BTL model estimates a discriminal difference
+distribution that accurately represents the \emph{true} discriminal
+difference between the texts. This scenario is illustrated in
+Figure~\ref{fig-dispersion}, when the model's discriminal difference
 distribution aligns with the \emph{true} discriminal difference
@@
 measurement model, evidence from the CJ literature suggests that this
-assumption may fail accommodate for heterogeneous stimuli, such as
+assumption may fail to accommodate for heterogeneous stimuli, such as
 handwritten texts or English compositions
 \citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
-Indeed, the presence of the so-called \emph{misfit texts}--texts that
-elicit more judgment discrepancies than others
+The presence of the so-called \emph{misfit texts}--texts that elicit
+more judgment discrepancies than others
 \citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}--may
@@
 it from adequately accounting for or addressing these anomalies, leaving
-exclusion of the ``problematic'' texts as the primary remedy.
+exclusion of the ``problematic'' texts as the primary remedy
+\citep{Pollitt_2012a, Pollitt_2012b}.
 
@@
 of excluding stimuli based on these statistics may unintentionally
-discard valuable information \citep{Miller_2023}, introducing bias into
-the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
+discard valuable information \citep{Miller_2023}, and introduce bias
+into the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
 direction and magnitude of these biases remain unpredictable, as they
@@
 
-The correlation \(\rho\), illustrated in
-Figure~\ref{fig-discriminal_difference}, measures how much the judges'
-perception of a specific trait in one stimulus depends on their
-perception of the same trait in another. Similar to the discriminal
-dispersions, this correlation shapes the distribution of the discriminal
-difference, directly impacting the outcomes of pairwise comparisons. A
-thought experiment, akin to the one presented in
-Section~\ref{sec-theory-issue1a}, can illustrate this idea. Assuming
-that the discriminal dispersions for both texts remain constant,
-Figure~\ref{fig-correlation} shows that as the correlation between the
-two texts increases, the distribution of their discriminal difference
-becomes narrower. This narrowing, in turn, affects the probability that
-the discriminal difference distinctly favors Text B over Text A--denoted
-as \(P(B > A)\)--and thus directly influences the comparison outcome.
-Furthermore, the figure shows that when two texts are independent or
-uncorrelated, as assumed in the BTL model \((\rho=0)\), the distribution
-of their discriminal difference is less narrow than in scenarios where
-the texts are positively correlated. As a result, it becomes less likely
-for the comparison to favor Text B over Text A, as indicated by the
-larger shaded area.
+The correlation between two stimuli \(\rho\) measures how much the
+judges' perception of a specific trait in one stimulus depends on their
+perception of the same trait in other stimulus. Similar to the
+discriminal dispersions, this correlation shapes the distribution of the
+discriminal difference, directly impacting the outcomes of pairwise
+comparisons. Assuming that the discriminal dispersions for a couple of
+texts remain constant, Figure~\ref{fig-correlation} shows that as the
+correlation between the two texts increases, the distribution of their
+discriminal difference becomes narrower. This narrowing, in turn,
+affects the probability that the discriminal difference distinctly
+favors Text B over Text A--denoted as \(P(B > A)\)--and thus directly
+influences the comparison outcome. Furthermore, the figure shows that
+when two texts are independent or uncorrelated, as assumed in the BTL
+model \((\rho=0)\), the distribution of their discriminal difference is
+less narrow than in scenarios where the texts are positively correlated.
+As a result, it becomes less likely for the comparison to favor Text B
+over Text A, as indicated by the larger shaded area.
 
@@
 CJ literature indicates that the assumption of zero correlation does not
-hold in practice in at least two scenarios: when intricate aspects of
+hold in practice in at least two cases: when intricate aspects of
 multidimensional, complex traits or heterogeneous stimuli influence
@@
 
-In the first scenario, research on text quality suggests that when
-judges evaluate complex, multidimensional traits or heterogeneous
-stimuli, they often rely on a variety of intricate stimulus
-characteristics to inform their judgments
+Research on text quality assessments suggests that when judges evaluate
+complex, multidimensional traits or heterogeneous stimuli, they often
+rely on a variety of intricate stimulus characteristics to inform their
+judgments
 \citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
@@
 scenario is limited, existing studies document the presence of judge
-bias in CJ \citep[\citet{vanDaal_et_al_2016},
-\citet{Bartholomew_et_al_2020a}]{Pollitt_et_al_2003}, reinforcing the
-argument that the factors influencing pairwise comparisons do not always
-cancel each other out.
-
-In the second scenario, the shared context or inherent connections
+bias in CJ
+\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
+reinforcing the argument that the factors influencing pairwise
+comparisons do not always cancel each other out.
+
+In the second case, the shared context or inherent connections
 introduced by additional hierarchical structures may create dependencies
@@
 \citep{Everitt_et_al_2010}. For instance, when the same individual
-produces multiple texts, those texts often exhibit common features, such
-as writing style or overall quality, that judges can easily recognize.
-In this regard, although the CJ literature acknowledges the existence of
-such hierarchical structures, the statistical approaches to account for
-this additional source of dependence have been insufficient. For
-instance, when CJ data incorporates multiple samples of stimuli from the
-same individuals, researchers frequently rely on (averaged) point
+produces multiple texts, those texts often share several features such
+as writing style or overall quality. Although the CJ literature
+acknowledges the existence of such hierarchical structures
+\citep[e.g.,][]{Boonen_et_al_2020}, the statistical approaches to
+account for this additional source of dependence have been insufficient.
+For instance, when CJ data incorporates multiple samples of stimuli from
+the same individuals, researchers frequently rely on (averaged) point
 estimates of the BTL scores to conduct subsequent analyses and tests at
@@
 \citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
-However, this approach can introduce additional statistical and
-measurement issues, similar to the ones discussed in
-Section~\ref{sec-theory-issue2}.
 
 Thus, erroneously assuming zero correlation between stimuli can also
-lead to significant statistical and measurement issues. Specifically,
+lead to significant statistical and measurement issues. In particular,
 neglecting judges' biases or relevant hierarchical structures can create
@@
 underestimation of trait reliability
-\citep{Ackerman_1989, Hoyle_et_al_2023} and even biases
-\citep{Wu_et_al_2022}. This issue is illustrated in
-Figure~\ref{fig-correlation} when the discriminal difference
-distribution of the BTL scores follows the thick continuous line
-\((\rho = 0)\), while the \emph{true} discriminal difference follows any
-discontinuous line where \(\rho \neq 0\). Moreover, these inaccuracies
-can result in spurious conclusions about the discriminal differences
-\citep{McElreath_2020} and, by extension, the underlying discriminal
-processes of the stimuli--potentially causing the model to incorrectly
-classify stimuli or judges as \emph{misfits}.
-
-Finally, similar to what we discussed in the previous section, removing
-\emph{misfit} judges--that is, judges whose assessments deviate markedly
-from the shared consensus
+\citep{Ackerman_1989, Hoyle_et_al_2023} and even introduce statistical
+biases \citep{Wu_et_al_2022}. These inaccuracies can result in spurious
+conclusions about the discriminal differences \citep{McElreath_2020}
+and, by extension, the underlying discriminal processes of the stimuli.
+One such spurious conclusion could be the incorrect classification of
+stimuli or judges as \emph{misfits}. Figure~\ref{fig-correlation}
+illustrates how assuming zero correlation can undermine trait
+reliability: the discriminal difference distribution of the BTL scores
+follows the thick continuous line \((\rho = 0)\), while the \emph{true}
+discriminal difference may correspond to any discontinuous line where
+\(\rho \neq 0\).
+
+Finally, removing \emph{misfit} judges--that is, judges whose
+assessments deviate markedly from the shared consensus
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
 and may appear as outliers under the BTL model
-\citep{Wu_et_al_2022}--risk discarding valuable information and
+\citep{Wu_et_al_2022}--risks discarding valuable information and
 introducing bias into trait estimates \citep{Miller_2023}. The direction
@@
 
-Building on the previous section, it is clear that, researchers in CJ
-studies typically use the BTL model to measure traits and position the
-compared stimuli along a latent continuum \citep{Thurstone_1927b}. The
-CJ literature also show that they frequently rely on point estimates of
-these traits--typically the BTL scores or its transformations--to
-conduct statistical inference or hypothesis testing. For example,
-researchers have used these scores to identify `misfit' judges and
-stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
-detect biases in judges' ratings
-\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
-other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
-or test hypotheses related to the underlying trait of interest
+Researchers in CJ studies typically use the BTL model to measure traits
+and position the compared stimuli along a latent continuum
+\citep{Thurstone_1927b}. The CJ literature shows that research
+frequently relies on point estimates of these traits--typically the BTL
+scores or its transformations--to conduct statistical inference or
+hypothesis testing. For example, researchers have used these scores to
+identify `misfit' judges and stimuli
+\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
+biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
+calculate correlations with other assessment methods
+\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
+related to the underlying trait of interest
 \citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
@@
 the disconnect between the trait measurement and hypothesis testing
-processes. However, the structural approach to causal inference can
-address these issues by offering a systematic and integrated framework
-that strengthens measurement reliability and validity while enhancing
-the statistical accuracy of hypothesis tests.
+processes. The structural approach to causal inference can address these
+issues by offering a systematic and integrated framework to extend
+Thurstone's general form. This approach can also strengthen measurement
+reliability and validity while enhancing the statistical accuracy of
+hypothesis tests.
 
@@
 
-A motivating example that will appear throughout this document clarifies
+A motivating example that will appear throughout this study clarifies
 these concepts. In this example, researchers aim to answer the question:
@@
 
-Following standard CJ practices, researchers typically use estimates
-from the BTL model--or its transformations--to approximate this
-estimand. However, as discussed in Section~\ref{sec-theory-issues},
+Following standard CJ practices, researchers would typically use
+estimates from the BTL model--or its transformations--to approximate
+this estimand. However, as discussed in Section~\ref{sec-theory-issues},
 Thurstone's Case V and the BTL model exhibit several statistical and
@@
 Fortunately, SCMs and DAGs support identification analysis through two
-key advantages\footnote{These topics are beyond the scope of this study,
-  thus, readers seeking a more profound understanding can refer to
-  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
-  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, and introductory
-  books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
+key advantages\footnote{In depth explanation of these topics is beyond
+  the scope of this study, thus, readers seeking a more profound
+  understanding can refer to introductory papers such as
+  \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
+  \citet{Cinelli_et_al_2020}, and introductory books like
+  \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
   \citet{McElreath_2020} are useful. For more advanced study, seminal
@@
 Section~\ref{sec-thurstone_theory} with key CJ assessment design
-features, such as the selection of judges, stimuli, and comparisons. In
-addition to improving statistical accuracy and strengthening measurement
-reliability and validity, the approach offers two key advantages. First,
-it clarifies the interactions among all actors and processes involved in
-CJ assessments. Second, it shifts the current comparative data analysis
-paradigm from passively accepting the model assumptions to actively
-testing whether those assumptions fit the data under analysis.
-
-Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
-theoretical principles into what we refer to as the
-\emph{conceptual-population model}. This model assumes an idealized
-scenario where researchers have access to a \emph{conceptual population}
-of comparative data, that is, data representing all repeated judgments
-made by every available judge for each pair of stimuli produced by each
-pair of individuals in the population. Conversely,
-Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
-features into what we refer to as the \emph{sample-comparison model}.
-This model assumes a more realistic scenario where researchers only have
-access to a sample of judges, individuals, stimuli, and comparisons from
-the conceptual population.
+features, such as the selection of judges, stimuli, and comparisons.
+Section~\ref{sec-theory-theoretical_P} introduces the
+\emph{conceptual-population model}, which incorporates these theoretical
+principles and assumes an idealized setting where researchers observe a
+\emph{conceptual population} of comparative judgment data--that is, data
+representing all repeated judgments made by every available judge for
+each pair of stimuli produced by each pair of individuals in the
+population. Conversely, Section~\ref{sec-theory-theoretical_SC} presents
+the \emph{sample-comparison model}, which integrates the assessment
+design features and reflects a more realistic setting where researchers
+access only a sample of judges, individuals, stimuli, and comparisons
+from the conceptual population.
 
@@
 
-\begin{minipage}{\linewidth}
+\begin{minipage}{0.50\linewidth}
 
@@
 \end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.77\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
 
@@
 Although specifying a data structure is not mandatory when using SCMs
-and DAGs, defining one in this case can improve clarity and facilitate
-the description of the system. Thus, to re-express the scalar form of
-the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
-vectorized form, we first define the vectors \(I\) and \(J\), along with
-the matrices \(IA\) and \(JK\), as in Equation (\ref{eq-mat11}). Here,
-each element of \(I\) represents a unique individual \(i\) or \(h\),
-where \(n_{I}\) denotes the total number of individuals. Similarly, each
-element of \(J\) corresponds to a unique judge \(j\), with \(n_{J}\)
-indicating the total number of judges. Moreover, each row of \(IA\)
-represents a unique pairing of individuals \(i, h\) with stimuli
-\(a, b\). As a result, the matrix \(IA\) contains \(n_{I} \cdot n_{A}\)
-rows and \(2\) columns, where \(n_{A}\) specifies the number of stimuli
-available per individual. Likewise, each row of \(JK\) associates a
-judge \(j\) with a (repeated) judgment index \(k\). Consequently, the
-matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\) columns, where
-\(n_{K}\) indicates the number of repeated judgments each judge makes.
+and DAGs, defining one improves clarity and facilitates the description
+of the system. Thus, to re-express the scalar form of the CJ system
+shown in Figure~\ref{fig-cj03} into an equivalent vectorized form, we
+first define the vectors \(I\) and \(J\), along with the matrices \(IA\)
+and \(JK\), as in Equation (\ref{eq-mat11}). Here, each element of \(I\)
+represents a unique individual \(i\) or \(h\), where \(n_{I}\) denotes
+the total number of individuals. Similarly, each element of \(J\)
+corresponds to a unique judge \(j\), with \(n_{J}\) indicating the total
+number of judges. Moreover, each row of \(IA\) represents a unique
+pairing of individuals \(i, h\) with stimuli \(a, b\). As a result, the
+matrix \(IA\) contains \(n_{I} \cdot n_{A}\) rows and \(2\) columns,
+where \(n_{A}\) specifies the number of stimuli available per
+individual. Likewise, each row of \(JK\) associates a judge \(j\) with a
+(repeated) judgment index \(k\). Consequently, the matrix \(JK\) has
+\(n_{J} \cdot n_{K}\) rows and \(2\) columns, where \(n_{K}\) indicates
+the number of repeated judgments each judge makes.
+
+Additionally, we construct the matrix \(R\) to map each row of the
+\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
+matrix has \(n\) rows and \(6\) columns, where
+\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
+term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
+coefficient, which quantifies the total number of unique comparisons
+possible between every pair of stimuli generated by each pair of
+individuals in the population. Thus, we define the matrix as in Equation
+(\ref{eq-mat11}).
+
 \begin{equation}\phantomsection\label{eq-mat11}{
@@
 n_{J} & n_{K}
-\end{bmatrix}
-}\end{equation}
-
-Additionally, we construct the matrix \(R\) to map each row of the
-\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
-matrix has \(n\) rows and \(6\) columns, where
-\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
-term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
-coefficient, which quantifies the total number of unique comparisons
-possible between every pair of stimuli generated by each pair of
-individuals in the population. Thus, we define the matrix as in Equation
-(\ref{eq-mat12}).
-
-\begin{equation}\phantomsection\label{eq-mat12}{
+\end{bmatrix};
 R = \begin{bmatrix}
@@
 \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), the vectors and
-matrices described in Equations (\ref{eq-mat11}) and (\ref{eq-mat12})
-take the form as in Equation (\ref{eq-mat13}).
+matrices described in Equation (\ref{eq-mat11}) take the form as in
+Equation (\ref{eq-mat13}).
 \begin{equation}\phantomsection\label{eq-mat13}{
@@
 
-Now, using Equations (\ref{eq-mat11}) and (\ref{eq-mat12}), we can
-re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
-equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
-depiction, the outcome \(O^{cp}_{R}\), the texts' discriminal difference
-\(D_{R}\), their discriminal processes \(T_{IA}\), and the judges'
-biases \(B_{JK}\) are represented as vectors rather than scalar values.
-These vectors capture all the observations from the conceptual
-population. Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and
-latent vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
+Now, using Equation (\ref{eq-mat11}), we can re-express
+SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
+vectorized form, as shown in Figure~\ref{fig-cj04}. In this depiction,
+the outcome \(O^{cp}_{R}\), the texts' discriminal difference \(D_{R}\),
+their discriminal processes \(T_{IA}\), and the judges' biases
+\(B_{JK}\) are represented as vectors rather than scalar values. These
+vectors capture all the observations from the conceptual population.
+Specifically, \(O^{cp}_{R}\) and \(D_{R}\) are observed and latent
+vectors of length \(n\), respectively. Moreover, \(T_{IA}\) and
 \(B_{JK}\) are latent vectors of lengths \(n_{I} \cdot n_{A}\) and
@@
 \end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=0.87\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
 
@@
 The sample-comparison model presents a more realistic scenario than the
-conceptual-population model. First, in
-Section~\ref{sec-theory-theoretical_SC1}, it explicitly assumes
-researchers work with a data sample consisting of a limited number of
-repeated judgments \((n^{s}_{K})\) from a sample of judges
-\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
-sample of students \((n^{s}_{I})\), all drawn from the conceptual
-population. Second, in Section~\ref{sec-theory-theoretical_SC2}, the
-model assumes that judges do not perform \emph{all repeated judgments}
-within the data sample. Instead, they conduct a sufficient number of
-stimuli comparisons, \(n_{C}\), to ensure an accurate estimation of the
-proportion \(P(B>A)\), as proposed by \citet{Thurstone_1927b}.
+conceptual-population model. First, it explicitly assumes researchers
+work with a data sample consisting of a limited number of repeated
+judgments \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a
+specific number of texts \((n^{s}_{A})\) from a sample of students
+\((n^{s}_{I})\), all drawn from the conceptual population
+(Section~\ref{sec-theory-theoretical_SC1}). Second, the model assumes
+that judges do not perform \emph{all repeated judgments} within the data
+sample (Section~\ref{sec-theory-theoretical_SC2}). Instead, they conduct
+a sufficient number of stimuli comparisons, \(n_{C}\), to ensure an
+accurate estimation of the proportion \(P(B>A)\), as proposed by
+\citet{Thurstone_1927b}.
 
@@
 each pair of sampled individuals.
+
 \begin{equation}\phantomsection\label{eq-mat26}{
@@
 within their respective groups \citep{Lawson_2015}.
+
+However, due to concerns about the practical feasibility of the
+comparison task \citep{Boonen_et_al_2020}, CJ assessments rarely
+implement an exhaustive pairings of sampled judges, stimuli, and
+individuals. Thus, a realistic scenario must account for the fact that
+judges typically compare only a subset of stimuli authored by a sample
+of individuals.
 
@@
 \end{figure}%
-
-However, due to concerns about the practical feasibility of the
-comparison task \citep{Boonen_et_al_2020}, CJ assessments rarely
-implement an exhaustive pairings of sampled judges, stimuli, and
-individuals. Thus, a realistic scenario must account for the fact that
-judges typically compare only a subset of stimuli authored by a sample
-of individuals.
 
@@
 Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
-which we outline in the statistical model \ref{fig-cj16_stat}. We begin
-by assuming that \(O_{R}\) follows a Bernoulli distribution\footnote{The
-  binomial distribution--including its special case, the Bernoulli
+which we outline in the statistical model presented in
+Figure~\ref{fig-cj16_stat}. The model starts by assuming that \(O_{R}\)
+follows a Bernoulli distribution\footnote{The binomial
+  distribution--including its special case, the Bernoulli
   distribution--represent a maximum entropy distribution for binary
@@
 
-Note that while the joint distribution in Equation (\ref{eq-mat51})
-includes the probability distributions of the sampling and comparison
-mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
+While the joint distribution in Equation (\ref{eq-mat51}) includes the
+probability distributions of the sampling and comparison mechanisms,
+\(P(S)\) and \(P(C)\), as well as those of the predictor
 variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
-\(P(Z_{J})\)--we omit all of these probabilities from the statistical
-model \ref{fig-cj16_stat}. This omission is justified because, while
-these distributions contribute to the overall joint distribution of the
-data, the variables \(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
-\(Z_{J}\) are observed and independent of any other variable in the
-model. As observed variables, they do not require distributional
-assumptions in the same way the idiosyncratic errors do. Furthermore,
-their independence follows from the underlying random selection
+\(P(Z_{J})\)--all of these probabilities are omitted from the
+statistical model \ref{fig-cj16_stat}. This omission is justified
+because, while these distributions contribute to the overall joint
+distribution of the data, the variables \(S\), \(C\), \(X_{IA}\),
+\(X_{I}\), \(Z_{JK}\), and \(Z_{J}\) are observed and independent of any
+other variable in the model. As observed variables, they do not require
+distributional assumptions in the same way the idiosyncratic errors do.
+Their independence follows from the underlying random selection
 procedures that govern the variables\footnote{Randomization ensures that
@@
 
-Next we define \(D_{R}\) as the difference between the discriminal
+Next \(D_{R}\) is defined as the difference between the discriminal
 processes \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), representing the
 underlying written-quality trait of the compared texts, plus the
-corresponding repeated judge bias \(B_{JK}[j, k]\). Note that if we
-assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
-biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
-can re-write the discriminal difference as:
+corresponding repeated judge bias \(B_{JK}[j, k]\). Note that if it is
+assumed that \(B_{JK}[j,k]\) reflects the difference in
+stimulus-specific biases, i.e.,
+\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), the discriminal
+difference can be re-written as:
 \begin{equation}\phantomsection\label{eq-mat52}{
@@
 
-We now specify the functional forms for \(T_{IA}\), \(T_{I}\),
-\(B_{JK}\), and \(B_{J}\). We model \(T_{IA}\) as a linear combination
+Now the functional forms for \(T_{IA}\), \(T_{I}\), \(B_{JK}\), and
+\(B_{J}\) are specified. \(T_{IA}\) is modeled as a linear combination
 of the students' underlying writing-quality traits \(T_{I}\), the
@@
 \(\beta_{XA}X_{IA}\) (such as the influence of text length), and the
-text-specific idiosyncratic errors \(e_{IA}\). Similarly, we express
-\(T_{I}\) as a linear combination of relevant student-related variables
+text-specific idiosyncratic errors \(e_{IA}\). Similarly, \(T_{I}\) is
+expressed as a linear combination of relevant student-related variables
 affecting the quality assessment \(\beta_{XI} X_{I}\), and
 student-specific idiosyncratic errors \(e_{I}\). For the judge-specific
-terms, we model \(B_{JK}\) as a linear combination of the judge's
+terms, \(B_{JK}\) is modeled as a linear combination of the judge's
 individual bias \(B_{J}\), the influence of relevant judgment-related
@@
 number of judgments affect the evaluation), and judgment-specific
-idiosyncratic errors \(e_{JK}\). Finally, we define \(B_{J}\) as a
+idiosyncratic errors \(e_{JK}\). Finally, \(B_{J}\) is defined as a
 linear combination of relevant judge-level variables influencing the
@@
 
-Next, we specify the probabilistic assumptions for the idiosyncratic
-errors \(e_{I}\), \(e_{IA}\), \(e_{J}\), and \(e_{JK}\). Unlike other
-variables in the model, these error terms exhibit indeterminacies in
-their \emph{location}, \emph{orientation}, and \emph{scale} due to the
-lack of an inherent scale in the associated latent variables \(T_{I}\),
-\(T_{IA}\), \(B_{J}\), and \(B_{JK}\). Thus, to identify the latent
-variable model, we must resolve these indeterminacies
+Next, the probabilistic assumptions for the idiosyncratic errors
+\(e_{I}\), \(e_{IA}\), \(e_{J}\), and \(e_{JK}\) are specified. Unlike
+other variables in the model, these error terms exhibit indeterminacies
+in their \emph{location}, \emph{orientation}, and \emph{scale} due to
+the lack of an inherent scale in the associated latent variables
+\(T_{I}\), \(T_{IA}\), \(B_{J}\), and \(B_{JK}\). Thus, to identify the
+latent variable model these indeterminacies must be resolved
 \citep{Depaoli_2021, deAyala_2009}. Drawing on principles from SEM
-\citep{Hoyle_et_al_2023}, we assume that the vector of idiosyncratic
-errors \(\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}\), follows
-a Multivariate Normal distribution with mean vector \(\boldsymbol{\mu}\)
-and a covariance matrix
+\citep{Hoyle_et_al_2023}, the vector of idiosyncratic errors
+\(\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}\) are assumed to
+follow a Multivariate Normal distribution with mean vector
+\(\boldsymbol{\mu}\) and a covariance matrix
 \(\boldsymbol{\Sigma} = \boldsymbol{V} \boldsymbol{Q} \boldsymbol{V}\),
@@
 deviations and \(\boldsymbol{Q}\) a correlation matrix. To address the
-\emph{location} indeterminacy, we set the errors' mean vector to zero:
+\emph{location} indeterminacy, the errors' mean vector is set to zero:
 \begin{equation}\phantomsection\label{eq-mat53}{
@@
 
-Following SCM~\ref{fig-cj16_scm}, we resolve the \emph{orientation}
-indeterminacy by assuming that the errors are uncorrelated. This
-assumption leads us to define the error correlation matrix,
+Following SCM~\ref{fig-cj16_scm}, the \emph{orientation} indeterminacy
+is solved by assuming that the errors are uncorrelated. This assumption
+leads to the definition of the error's correlation matrix,
 \(\boldsymbol{Q}\), as the identity matrix:
@@
 
-To resolve the \emph{scale} indeterminacy, we define the diagonal matrix
-\(\boldsymbol{V}\) as: \begin{equation}\phantomsection\label{eq-mat55}{
+To resolve the \emph{scale} indeterminacy, the diagonal matrix
+\(\boldsymbol{V}\) is defined as follows:
+\begin{equation}\phantomsection\label{eq-mat55}{
 \boldsymbol{V} = \begin{bmatrix}
@@
 \(p_{IA}\) for the stimuli, \(s_{ZJ}\) for the judges, and \(p_{JK}\)
-for the judgments. We assume \(s_{XI}\) varies depending on the teaching
-method group to which each student belongs. Using the example from
-Section~\ref{sec-theoretical}, where the teaching method
+for the judgments. It is assumed \(s_{XI}\) varies depending on the
+teaching method group to which each student belongs. Using the example
+from Section~\ref{sec-theoretical}, where the teaching method
 \(X_{I} = \{1,2\}\), the model sets the constraint according to Equation
@@
 \(e_{IA}\) is a univariate Normal distribution with mean zero and
-standard deviation \(p_{IA}\). We set \(P_{IA}\) as a proportion of
-\(1\) to establish the scale of the stimuli' latent trait relative to
+standard deviation \(p_{IA}\). Thus, \(p_{IA}\) is set as a proportion
+of \(1\) to establish the scale of the stimuli' latent trait relative to
 the scale of the individuals' trait. Note that as a result, \(T_{IA}\)
@@
 
-Similarly, we assume \(s_{ZJ}\) varies depending on the groups to which
-each judge belongs. For instance, if \(Z_{J} = \{1,2,3\}\) represents
-three groups of judges with varying expertise, the model sets the
-constraint according to Equation (\ref{eq-mat57}). This constraint
+Similarly, it is assumed that \(s_{ZJ}\) varies depending on the groups
+to which each judge belongs. For instance, if \(Z_{J} = \{1,2,3\}\)
+represents three groups of judges with varying expertise, the model sets
+the constraint according to Equation (\ref{eq-mat57}). This constraint
 anchors the scale of the judges' latent trait and relaxes the assumption
@@
 that often affect frequentist methods
-\citep{Martin_et_al_1975, Seaman_et_al_2011}. We primarily use this
-capability to define the error distribution and set the scale of latent
+\citep{Martin_et_al_1975, Seaman_et_al_2011}. Prior distributions are
+used to define the error distribution and set the scale of latent
 variables \citep{Depaoli_2014}. Third, Bayesian inference supports
@@
 attempted to address this gap by separating trait estimation from
-hypothesis testing, relying on point estimates--such as BTL scores or
-their transformations--for inference. Notably, while this approach
-simplifies analysis, it can also introduce bias and compromise the
-reliability of the resulting inferences
+hypothesis testing, relying on point estimates, such as BTL scores or
+their transformations, for inference. While this approach simplifies
+analysis, it can also introduce bias and compromise the reliability of
+the resulting inferences
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
 
-Thus, to address the limitations of Thurstone's Case V and the BTL
-model, this study extends Thurstone's general form using a systematic,
-integrated approach that combines causal and Bayesian inference methods.
-The approach begins with the development of a conceptual model,
-represented as a Structural Causal Model (SCM) and a Directed Acyclic
-Graph (DAG)
+To address the limitations of Thurstone's Case V and the BTL model, this
+study extends Thurstone's general form using a systematic, integrated
+approach that combines causal and Bayesian inference methods. The
+approach begins with the development of a conceptual model, represented
+as a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG)
 \citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. This
@@
 features, including judges' bias, sampling procedures, and comparison
-mechanisms, thereby disentangling the causal processes underlying the CJ
+mechanisms, to disentangle the causal processes underlying the CJ
 system.
@@
 that allows researchers to analyze CJ data when violations to the
-assumptions of equal dispersion and zero correlation occur (see
-Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b})
-and when statistical inference is necessary (refer to
-Section~\ref{sec-theory-issue2}). In particular, this model accounts for
-judge biases (see Section~\ref{sec-theory-theoretical_P1}), captures the
-hierarchical structure of stimuli, incorporates measurement error into
-the hypothesis testing process (refer to
-Section~\ref{sec-theory-theoretical_P3}), and accommodates heterogeneity
-in discriminal dispersions (see Section~\ref{sec-statistical}). By
-addressing all these issues, these methodological innovations have the
-potential to enhance the reliability and validity of trait measurement
-in CJ \citep{Perron_et_al_2015}, while also improving the accuracy of
+assumptions of equal dispersion and zero correlation occur, and when
+statistical inference is necessary. In particular, this model accounts
+for judge biases, captures the hierarchical structure of stimuli,
+incorporates measurement error into the hypothesis testing process, and
+accommodates heterogeneity in discriminal dispersions. By addressing all
+these issues, these methodological innovations have the potential to
+enhance the reliability and validity of trait measurement in CJ
+\citep{Perron_et_al_2015}, while also improving the accuracy of
 statistical inferences.
@@
 likely stems from a more fundamental lack of conceptual clarity about
-how these mechanisms function within the system, this study decided to
-integrate these mechanisms into the conceptual model of CJ.
+how these mechanisms function within the system, this study integrates
+these mechanisms into the conceptual model of CJ.
 
@@
 Specifically, it clarifies their role as sources of missing data in CJ's
-data-generating process--that is, as mechanisms that determine which
+data-generating process, that is, as mechanisms that determine which
 observations are missing from the final data sample. This new
@@
 
-Notably, this study circumvents the need to apply this missing data
-framework by deliberately structuring the sampling and comparison
-mechanisms to be independent of any observed or unobserved variables,
-including the outcome. In other words, we designed these mechanisms so
-they produce data that are \emph{missing completely at random} (MCAR)
+This study circumvents the need to apply this missing data framework by
+deliberately structuring the sampling and comparison mechanisms to be
+independent of any observed or unobserved variables, including the
+outcome. In other words, these mechanisms are designed to produce data
+that are \emph{missing completely at random} (MCAR)
 \citep{Little_et_al_2020}. This design offers one key advantage: it
@@
 designs, in which the sampling and comparison mechanisms introduce more
-intricate forms of missingness--such as \emph{missing at random} (MAR)
-or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
-prominent example is the previously discussed ACJ designs, where prior
+intricate forms of missingness such as \emph{missing at random} (MAR) or
+\emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
+prominent example is the previously discussed ACJ design, where prior
 judgment outcomes inform the selection of stimulus pairs for subsequent
@@
 
-Nevertheless, regardless of the underlying missingness mechanisms, any
-CJ assessment design would benefit from explicitly defining its
-assumptions--a practice supported by our approach. This clarity enables
-researchers to evaluate how the sampling and comparison mechanisms
-affect trait estimation and statistical inference within each design.
-Such assessments are particularly relevant given the common
-misconception in the CJ literature that Thurstone's model can naturally
-handle even non-random missing data without compromising the reliability
-or validity of trait estimates \citep{Bramley_2008}.
+Regardless of the underlying missingness mechanisms, any CJ assessment
+design would benefit from explicitly defining its assumptions--a
+practice supported by our approach. This clarity enables researchers to
+evaluate how the sampling and comparison mechanisms affect trait
+estimation and statistical inference within each design. Such
+assessments are particularly relevant given the common misconception in
+the CJ literature that Thurstone's model can naturally handle even
+non-random missing data without compromising the reliability or validity
+of trait estimates \citep{Bramley_2008}.
 
@@
 \citep{Bramley_2008, Andrich_1978} despite growing evidence of
-persistent biases. Another critical question, discussed in
-Section~\ref{sec-theory-issue1b}, is whether training or expertise can
-help judges avoid focusing on irrelevant stimulus features, such as
-handwriting, over more central criteria like argumentative quality in
-writing assessments \citep{Kelly_et_al_2022}. Notably, exploring these
-questions may also provide insights into what it truly means to be an
-``expert'' within the CJ context \citep{Kelly_et_al_2022}.
+persistent biases. Another critical question is whether training or
+expertise can help judges avoid focusing on irrelevant stimulus
+features, such as handwriting, over more central criteria like
+argumentative quality in writing assessments \citep{Kelly_et_al_2022}.
+Exploring these questions may also provide insights into what it truly
+means to be an ``expert'' within the CJ context
+\citep{Kelly_et_al_2022}.
 
@@
 \citep{Lawson_2015}, can improve judgment consistency and accuracy. As
-anticipated, our approach provides the necessary structure to
-investigate rigorously these questions.
+anticipated, the approach presented in this study provides the necessary
+structure to investigate rigorously these questions.
 
@@
 \citep{Pollitt_2004, Pollitt_2012a, Pollitt_2012b, Goossens_et_al_2018}.
-Both definitions closely mirror the statistical concept of
-outliers--that is, observations that deviate markedly from the rest of
-the sample in which they occur \citep{Grubbs_1969}. But this resemblance
-extends beyond the definitions themselves, as CJ researchers often
-identify misfits using conventional outlier detection procedures, such
-as transforming BTL model residuals into diagnostic statistics and
+Both definitions closely mirror the statistical concept of outliers,
+that is, observations that deviate markedly from the rest of the sample
+in which they occur \citep{Grubbs_1969}. But this resemblance extends
+beyond the definitions themselves, as CJ researchers often identify
+misfits using conventional outlier detection procedures, such as
+transforming BTL model residuals into diagnostic statistics and
 comparing them against predefined thresholds
@@
 Nevertheless, it is not the classification of misfits as outliers that
-raises concerns for trait measurement and inference, but rather the
-prevailing CJ practice of identifying them through these ad-hoc
-detection procedures and then excluding them from analysis, often
-without providing empirical evidence for the various hypotheses proposed
-to explain their occurrence in the first place. In this regard, it is
-essential to recognize that outliers can only be defined relative to a
-specific model \citep{McElreath_2020}. Thus, identification procedures
-based on models like the BTL, which relies on strong assumptions, should
-be approached with caution, as they may not be well-suited for the task
-\citep{Kelly_et_al_2022}. Specifically, because the BTL model may not
-always accurately reflect the underlying data-generating process of the
-CJ system (see Section~\ref{sec-theory-issues}), detection procedures
-grounded in this model risk compromising the accurate identification of
-misfits. Moreover, the statistical literature cautions that excluding
-outliers risks discarding valuable information \citep{Miller_2023} and
-introducing bias into trait estimates. The direction and magnitude of
-this bias are often unpredictable, as they depend on which observations
-researchers exclude from the analysis
-\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. Lastly, regarding
-the hypotheses proposed to explain the presence of misfits, it is
-evident that the BTL model lacks the flexibility needed to adequately
-test many of them, largely due to its strong assumptions.
-
-In contrast, our approach provides a rigorous framework that enables the
-examination of several relevant hypotheses. For instance, researchers
-could investigate whether misfit judges are those who exhibit (an
-outlying degree of) systematic bias or greater variability in their
-judgments compared to their peers. Similarly, researchers might explore
-whether misfit stimuli exhibit more variable discriminal processes
-relative to other stimuli or if they are genuinely outlying cases.
-Moreover, since outliers are not necessarily ``bad data''
-\citep{McElreath_2020}, our approach also offers a principled
+raises concerns for trait measurement and inference. Rather, the concern
+lies in the prevailing CJ practice of classify them through ad-hoc
+detection procedures and then excluding them from analysis
+\citep{Pollitt_2012a, Pollitt_2012b}, often without providing empirical
+evidence for the various hypotheses proposed to explain their
+occurrence. In this regard, it is essential to recognize that outliers
+can only be defined relative to a specific model \citep{McElreath_2020}.
+As such, detection procedures based on models like the BTL, which relies
+on strong assumptions, should be approached with caution, as they may
+not be well-suited for the task \citep{Kelly_et_al_2022}. In particular,
+because the BTL model may not always accurately reflect the underlying
+data-generating process of the CJ system, misfit classification based on
+this model risks misidentifying observations. Furthermore, the exclusion
+of these observations carries additional risks. The statistical
+literature cautions that removing outliers can discard valuable
+information \citep{Miller_2023} and introduce bias into trait estimates.
+The direction and magnitude of this bias are often unpredictable, as
+they depend on which observations researchers exclude from the analysis
+\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. Finally, precisely
+because of its rigid assumptions, the BTL model lacks the flexibility to
+adequately test many of the hypotheses proposed to explain the presence
+of misfits.
+
+In contrast, the approach presented in this study provides a rigorous
+framework that enables the examination of several relevant hypotheses.
+For instance, researchers could investigate whether misfit judges are
+those who exhibit (an outlying degree of) systematic bias or greater
+variability in their judgments compared to their peers. Similarly,
+researchers might explore whether misfit stimuli exhibit more variable
+discriminal processes relative to other stimuli or if they are genuinely
+outlying cases. Moreover, since outliers are not necessarily ``bad
+data'' \citep{McElreath_2020}, our approach also offers a principled
 alternative that retains misfits in the analysis without compromising
-trait estimation or inference. This alternative involves adapting our
-proposed models into robust measurement models \citep{McElreath_2020}--a
+trait estimation or inference. This alternative involves adapting the
+proposed model into robust measurement models \citep{McElreath_2020}, a
 broad class of procedures designed to reduce the sensitivity of
@@
 assumptions, whether the data are observational or experimental
-\citep{Kohler_et_al_2019, Deffner_et_al_2022}. Our approach is not an
-exception to this fundamental principle. As with all approaches grounded
-on causal inference, ours has the limitation of relying on expert
-knowledge and assumptions about the variables' causal structure that are
-often untestable at the outset \citep{Hernan_et_al_2025}. As such, the
+\citep{Kohler_et_al_2019, Deffner_et_al_2022}. The proposed approach is
+not an exception to this fundamental principle. As with all approaches
+grounded on causal inference, it relies on expert knowledge and
+assumptions about the variables' causal structure that are often
+untestable at the outset \citep{Hernan_et_al_2025}. As such, the
 approach does not seek to yield automatic answers when applied to a
@@
 its causal assumptions \citep{Deffner_et_al_2022}, which are well
-supported by both theory and existing evidence (see
-Section~\ref{sec-theory-issues} and Section~\ref{sec-theoretical}).
+supported by both theory and existing evidence.
 
@@
 
-To apply our approach effectively, CJ researchers must tackle two main
-challenges related to the required foundational knowledge. First, they
-need a solid understanding of causal inference principles. Second, they
-must learn to translate the functional and probabilistic content of
-conceptual models into bespoke statistical models. One example that
-illustrates the importance of a sound knowledge of causal inference is
-the recurrent assumption that predictor variables are ``relevant'' to
-the research context--interpreting this relevance as their inclusion in
-a \emph{sufficient adjustment set}
+To apply the approach effectively, CJ researchers require foundational
+knowledge in two areas. First, they need a solid understanding of causal
+inference principles. Second, they must learn to translate the
+functional and probabilistic content of conceptual models into bespoke
+statistical models. One example that illustrates the importance of a
+sound knowledge of causal inference is the recurrent assumption that
+predictor variables are ``relevant'' to the research
+context--interpreting this relevance as their inclusion in a
+\emph{sufficient adjustment set}
 \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
@@
 assist this effort, this study includes key references throughout the
-study to guide motivated aaplied CJ researchers toward a deeper
-understanding of these concepts \footnote{refer to footnote 1, 2, and
-  the detailed online document referenced in the Declarations section}.
+study to guide applied CJ researchers toward a deeper understanding of
+these concepts \footnote{refer to footnote 1, 2, and the detailed online
+  document referenced in the Declarations section}.
 
@@
 identification proofs remains a significant goal for future CJ research.
-Importantly, our approach supports both strategies by providing the
+Notably, our approach supports both strategies by providing the
 probabilistic foundation for formal derivations and the necessary

### 2025-06-25 paper2.qmd
review and editing

@@
 
+# Conclussion {#sec-conclussion}
+
+This paper underscores the need to extend Thurstone's theory to meet the demands of contemporary empirical CJ research.
+It specifically advocates for the development of bespoke CJ models aligned with the underlying data-generating processes of specific CJ assessment designs, as these models have the potential to strengthen the robustness of trait estimates and the clarity of inferences.
+The work also outlines a clear trajectory for advancing CJ research in both theoretical and applied domains.
+Finally, the study also lays the groundwork for broader applications, encouraging researchers across the social sciences to adopt more robust and interpretable methodologies.
+
 {{< pagebreak >}}

### 2025-06-25 _tex/paper2.tex
Built site for gh-pages

@@
 
+\section{Conclussion}\label{sec-conclussion}
+
+This paper underscores the need to extend Thurstone's theory to meet the
+demands of contemporary empirical CJ research. It specifically advocates
+for the development of bespoke CJ models aligned with the underlying
+data-generating processes of specific CJ assessment designs, as these
+models have the potential to strengthen the robustness of trait
+estimates and the clarity of inferences. The work also outlines a clear
+trajectory for advancing CJ research in both theoretical and applied
+domains. Finally, the study also lays the groundwork for broader
+applications, encouraging researchers across the social sciences to
+adopt more robust and interpretable methodologies.
+
 \newpage{}

### 2025-07-03 paper2.qmd
review and editing

@@
 abstract: |
-  This study revisits Thurstone's law of comparative judgment (CJ), focusing on two prominent issues of traditional approaches.
-First, it critiques the heavy reliance on Thurstone's Case V assumptions and, by extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
-Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli.
-While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates.
-Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes.
-Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.
-
-  To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods.
-This extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
+  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two central issues in the CJ literature.
+First, it critiques the widespread reliance on Thurstone's Case V assumptions and, by extension, the Bradley-Terry-Luce (BTL) model in the analysis of CJ data.
+Specifically, this study argues that while the assumptions of equal discriminal dispersions and zero correlation between stimuli simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates.
+Second, the study highlights the apparent disconnect between trait measurement and hypothesis testing procedures in CJ applications.
+It contends that while separating these procedures simplifies the analysis of CJ data, this practice can also undermine the resulting statistical inferences.
+
+  To address these issues, this study aims to extend Thurstone's general form using a systematic and integrated approach based on causal and Bayesian inference methods.
+This extension integrates Thurstone's core theoretical principles alongside key CJ assessment design features.
 It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.
@@
 Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to have a higher trait level.
-For example, when assessing writing quality, judges compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Pollitt_2012b; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
+For example, judges can compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Pollitt_2012b; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
 
@@
 Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade.
-These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability.
+These studies have highlighted three key strengths of the method: its reliability, its validity, and practical applicability.
 Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
@@
 Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023].
-Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020b; @Boonen_et_al_2020].
+Lastly, research on its practical applicability highlights CJ's versatility across educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020b; @Boonen_et_al_2020].
 
@@
 <!-- The A for ‘adaptivity’ in ACJ means that the choice of which objects are presented to the judges depends on the outcomes of judgments made so far – the idea being to use judge time efficiently by not presenting objects to compare where the result of the comparison is almost certain.
-[@Bramley_2015, pp.
+[@Bramley_2015, p.
 4].
@@
 Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved.
-This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates.
-First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data.
-Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli.
-While these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates.
-Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes.
-Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
-
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, pp. 2] -->
+This study identifies and discusses two prominent issues in the CJ literature that can undermine the reliability and validity of CJ's trait estimates.
+First, it critiques the widespread reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] in the analysis of CJ data.
+Specifically, this study argues that while the assumptions of equal discriminal dispersions and zero correlation between stimuli simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates.
+Second, the study highlights the disconnect between trait measurement and hypothesis testing procedures in CJ applications.
+In particular, it contends that while separating these procedures simplifies the analysis of CJ data, this practice can also undermine the resulting statistical inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, p. 2] -->
 
 <!-- 5. So, what are we doing? -->
-To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods.
+To address these issues, this study aims to extend Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods.
+This extension integrates Thurstone's core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
 In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages.
@@
 
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, pp. 2] -->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, p. 2] -->
 
@@
 For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 2] -->
 
-<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
+<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, p. 751] -->
 
 <!-- 4. what is the organization of the study? -->
-The study divides its content into six main sections.
+The remainder of this study is organized into seven sections.
 @sec-thurstone_theory provides an overview of Thurstone's theory.
-@sec-theory-issues discusses the identified issues in detail.
-@sec-theoretical extends Thurstone's general form to address these challenges.
+@sec-theory-issues examines the central issues identified in the CJ literature.
+@sec-theoretical extends Thurstone's general form to address these issues.
 The extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
-@sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data.
-Finally, @sec-discussion discusses the findings, explores avenues for future research, and detail the challenges for future researchers.
+@sec-statistical translates these theoretical and design elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data.
+@sec-discussion reviews the findings, outline directions for future research, discusses the study's limitations and details the challenges that applied researchers may encounter.
+Finally, @sec-conclusion provides the concluding remarks.
 
@@
 <!-- 1. Thurstone's Theory of CJ: discriminal process -->
-In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b].
+In its *general form*, Thurstone’s theory addresses pairwise comparisons of a single judge who evaluates multiple stimuli [@Thurstone_1927a; @Thurstone_1927b].
 The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference.
@@
 <!-- The object of this paper is to describe a new psychophysical law which may be called the law of comparative judgment and to show some of its special applications in the measurement of psychological values.
-[@Thurstone_1927b, pp.
+[@Thurstone_1927b, p.
 266] -->
 
-<!-- It applies fundamentally to the judgments of a single observer who compares a series of stimuli by the method of paired comparison when no 'equal' judgments are allowed [@Thurstone_1927b, pp.
+<!-- It applies fundamentally to the judgments of a single observer who compares a series of stimuli by the method of paired comparison when no 'equal' judgments are allowed [@Thurstone_1927b, p.
 267] -->
@@
 We define the scale in terms of the frequencies of the discriminal processes for any stimulus.
-This artificial construct, the psychological scale, is so spaced off that the frequencies of the discriminal processes for any given stimulus form a normal distribution on the scale.[@Thurstone_1927b, pp.
+This artificial construct, the psychological scale, is so spaced off that the frequencies of the discriminal processes for any given stimulus form a normal distribution on the scale.[@Thurstone_1927b, p.
 266] -->
@@
 <!-- 2. Example of discriminal processes -->
-@fig-discriminal_process illustrates the hypothetical discriminal processes along a quality trait continuum for two written texts.
-The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
-Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
+These ideas become clearer with an example.
+@fig-discriminal_process illustrates the hypothetical discriminal processes for two written texts along a *quality* trait continuum.
+The figure shows that the modal discriminal process for Text B lies further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality.
+The figure also shows that Text B has a broader distribution than Text A, indicating a greater variability in perception due to its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.
 
@@
 
-![Discriminal processes](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}
-
-![Discriminal difference](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}
+![Hypothetical discriminal processes for two written texts](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}
+
+![Hypothetical discriminal difference of two written texts](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}
 
@@
 <!-- 3. Thurstone's Theory of CJ: discriminal difference -->
-However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*.
+However, since the individual discriminal processes of the stimuli are not directly observable [@Thurstone_1927a], the theory introduces the *law of comparative judgment*.
 This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008].
@@
 Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978].
-In this distribution, the mode (mean) represents the average relative separation between the stimuli, and its dispersion indicates the variability of that separation.
+In this distribution, the mode (mean) represents the average relative separation between the stimuli $(T_{B} - T_{A})$, and its dispersion $\sigma_{BA}$ indicates the variability of that separation.
 
@@
 The outcome of the paired comparison judgement is therefore related to the distribution of the difference between the two distributions of discriminal processes for essay A and essay B.
-[@Bramley_2008, pp.
+[@Bramley_2008, p.
 251] -->
@@
 Since D_{vj} and D_{vi}, are assumed normally distributed, the difference process ...
-is normally distributed with mean a_{j} - a_{i}, and variance s^{2}_{ji} [@Andrich_1978, pp.
+is normally distributed with mean a_{j} - a_{i}, and variance s^{2}_{ji} [@Andrich_1978, p.
 452] -->
@@
 Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
-For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases).
-However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.
+For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison.
+However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an "insufficient" data set to estimate the required parameters.
 
@@
 In order to make the problem soluble it is necessary to make at least one assumption, namely that the correlation between discriminal deviations is practically constant throughout the stimulus series and for the single observer.
-[@Thurstone_1927b, pp.
+[@Thurstone_1927b, p.
 267] -->
-
-![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}
 
@@
 @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases.
-For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008.
+For a detailed discussion of these cases and their growing simplification, refer to @Thurstone_1927b and @Bramley_2008.
+
+![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}
 
@@
 The more assumptions we care to make, the simpler will be the observation equations.
-[@Thurstone_1927b, pp.
+[@Thurstone_1927b, p.
 267] -->
 
-<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->
+<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, p. 268]. -->
 
 <!-- 3. But are our specific problems? -->
-However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement and offering no guidance on how to use its trait estimates for statistical inference or hypothesis testing.
-Specifically, Thurstone cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp.
+As the table suggest, Thurstone developed Case V with an emphasis on statistical simplicity, but this simplicity comes at the expense of accurate and precise trait measurement and practical guidance for inference.
+Specifically, Thurstone cautioned that Case V use "should not be made without (an) experimental test" [@Thurstone_1927b, p.
 270], as it imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases).
-Moreover, because Thurstone's primary goal was to produce a "rather coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp.
+Moreover, because Thurstone's primary goal was to produce a "rather coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, p.
 269], his theory did not support formal statistical inference.
-Despite these limitations, CJ research has predominantly relied on Case V to measure different traits, which raises significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978].
-Furthermore, although the CJ tradition has attempted to address the gap in hypothesis testing by relying on the point estimates of traits--or their transformations--the statistical literature cautions against using these estimates as the sole basis for statistical inference, as such practices introduce bias in the analysis and reduce the precision of hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
-Next, both issues are discussed in more depth.
+Nevertheless, despite Case V's limitations, CJ research has predominantly relied on it to measure various traits, raising significant concerns about the reliability and validity of such measurements in contexts where its assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978].
+Furthermore, although the CJ tradition has attempted to address the gap in statistical inference by relying on the point estimates of traits--or their transformations--the statistical literature cautions against using these estimates as the sole basis for inference, as such practices introduce bias and reduce the precision of hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].
+The next sections discuss both issues in greater depth.
 
@@
 This assumption should not be made without experimental test.
-[@Thurstone_1927b, pp.
+[@Thurstone_1927b, p.
 270] -->
 
-<!-- Case 5 makes the greatest simplification by assuming that the discriminal dispersions are all equal [@Bramley_2008, pp. 253] -->
-
-<!-- Adding assumptions simplifies the formula and improves its tractability, but one cannot place as much confidence in the outputs of the formula if those assumptions are not met [@Kelly_et_al_2022, pp.
+<!-- Case 5 makes the greatest simplification by assuming that the discriminal dispersions are all equal [@Bramley_2008, p. 253] -->
+
+<!-- Adding assumptions simplifies the formula and improves its tractability, but one cannot place as much confidence in the outputs of the formula if those assumptions are not met [@Kelly_et_al_2022, p.
 677].
@@
 This law defines the psychological scale or continuum.
-It allocates the compared stimuli on the continuum [@Thurstone_1927b, pp.
+It allocates the compared stimuli on the continuum [@Thurstone_1927b, p.
 269] -->
@@
 Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds.
-[@Andrich_1978, pp.459] -->
+[@Andrich_1978, p.459] -->
 
@@
 <!-- 1. The Case V and its simpler statistical representation: the BTL model -->
-As previously discussed, Case V remains the most widely used model in CJ literature.
+Case V is the most widely used model in the CJ literature.
 This preference largely stems from the widespread adoption of the BTL model, which provides a simplified statistical representation of the case.
-The BTL model mirrors most of Case V's assumptions, with one notable distinction.
+The BTL model incorporates most of Case V's assumptions, with one notable exception.
 While Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases).
 However, this substitution has minimal impact on trait estimation or model interpretation because the scale of the discriminal process (i.e., the latent trait) is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021].
-That is, as long as the substitution (transformation) preserves the data rank order, the choice of distribution for the discriminal processes is inconsequential.
-This condition is satisfied in this case, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I].
+In other words, as long as the substitution (transformation) preserves the rank order of the data, the choice of distribution for the discriminal processes is inconsequential.
+The BTL model satisfies this condition because the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I].
 
@@
 Birnbaum, 1968; Bock, 1975) because it is more tractable.
-[@Andrich_1978, pp.
+[@Andrich_1978, p.
 455] -->
@@
 The only difference is in the mathematical function linking the observed proportions to the difference in scale values – the more tractable logistic function is used rather than the cumulative Normal.
-[@Bramley_2008, pp.
+[@Bramley_2008, p.
 254] -->
@@
 However, as discussed below, the scale for the θp parameters is arbitrary up to a non monotonic transformation, and the practice of bringing it close to the scale of a normal ogive is a relict from a distant past, bound to create communication problems between users of item response theory (IRT) rather than add to their understanding of it.
-[@vanderLinden_et_al_2017_I, pp.
+[@vanderLinden_et_al_2017_I, p.
 16] -->
@@
 <!-- 2. But what are limitations of Case V (and the BTL model)? -->
-However, Thurstone acknowledged that some assumptions of Case V could be problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a].
-Thus, given that modern CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice.
+However, Thurstone acknowledged that some assumptions of Case V are problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a].
+Thus, given that contemporary CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice.
 These assumptions are the equal dispersion and zero correlation between stimuli.
@@
 A thought experiment can help illustrate this idea.
-In it, researchers observe the discriminal processes for two texts, A and B, assuming that the dispersion for Text A remains constant and that the two texts are uncorrelated $(\rho=0)$.
+Suppose researchers observe the discriminal processes for two texts, A and B, assuming that the dispersion for Text A remains constant and that the two texts are uncorrelated $(\rho=0)$.
 @fig-dispersion demonstrates that an increase in the uncertainty associated with the perception of Text B relative to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference.
@@
 Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a].
-Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019], in this case, the assumption of equal dispersions.
+Consequently, the effectiveness of the outcome to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019], in this case, the assumption of equal dispersions.
 When the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts.
 This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$.
-The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.
+The accuracy of this discriminal difference then ensures reliable estimates for the texts' discriminal processes.
 
 <!-- In experimental practice the procedure is the reverse of this hypothesis because the frequencies are known first experimentally, and from these frequencies we construct the psychological continuum.
-[@Thurstone_1927a, pp.
+[@Thurstone_1927a, p.
 373] -->
@@
 If our statistical assumptions are wrong for the analysis at hand, the inferential statements will be incorrect no matter how sophisticated the statistical methods are.
-[@Kohler_et_al_2019, pp.
+[@Kohler_et_al_2019, p.
 150] -->
@@
 <!-- 3. But does it hold in practice? -->
-Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to accommodate for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
-The presence of the so-called *misfit texts*--texts that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]--may signal this limitation, as such discrepancies can arise from larger discriminal dispersions due to stimulus heterogeneity or because the texts are genuine outliers--that is, texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969].
-In either case, the BTL model's assumptions prevent it from adequately accounting for or addressing these anomalies, leaving exclusion of the "problematic" texts as the primary remedy [@Pollitt_2012a; @Pollitt_2012b].
+Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may not hold for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].
+The presence of the so-called misfit texts signals this limitation.
+*Misfit texts* are those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018], and these discrepancies may arise from larger discriminal dispersions caused by the stimulus' heterogeneity or because the texts are genuine outliers--i.e., texts with distinctive characteristics that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969].
+In either case, the BTL model's assumptions prevent it from adequately accounting for or addressing these anomalies, leaving exclusion of such "problematic" texts as the primary remedy [@Pollitt_2012a; @Pollitt_2012b].
 
@@
 Every specimen in such a series presents two unknown values to be determined, namely, the scale value of its modal discriminal process on the psychological continuum and its discriminal dispersion.
-[@Thurstone_1927a, pp.
+[@Thurstone_1927a, p.
 376] -->
@@
 Thurstone clearly did not expect this to hold for any but the most simple stimuli.
-The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008, pp.
+The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008, p.
 273].
@@
 While Thurstone used his Law to create measurement scales for purely psychological or subjective entities, he did so only with stimuli whose values could be evaluated relatively instantaneously, such as short statements and handwriting samples.
-He was also skeptical about its use for very heterogeneous stimuli (Thurstone, 1927b) [@Kelly_et_al_2022, pp.
+He was also skeptical about its use for very heterogeneous stimuli (Thurstone, 1927b) [@Kelly_et_al_2022, p.
 678].
@@
 Essentially this reports the amount of inconsistency in the various judgements that have been made of that script.
-[@Pollitt_2004, pp.
+[@Pollitt_2004, p.
 11]  -->
@@
 ...
-A similar statistic is calculated for every script. A misfitting script is one that judges find difficult for some reason, one that some judges value more than others do. [@Pollitt_2012b, pp. 290] -->
+A similar statistic is calculated for every script. A misfitting script is one that judges find difficult for some reason, one that some judges value more than others do. [@Pollitt_2012b, p. 290] -->
 
 <!-- If a portfolio’s WMS exceeds the criterion of mean plus two standard deviations, this means the judges did not judge it consistently, some considering it ‘better’ than others did.
-[@Pollitt_2012a, pp.
+[@Pollitt_2012a, p.
 165] -->
 
-<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, pp. 20] -->
-
-<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, pp. 1] -->
+<!-- Representations with a large infit are representations which lead to more inconsistent judgments [@Goossens_et_al_2018, p. 20] -->
+
+<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, p. 1] -->
 
@@
 The typical power function is depressed.
-irrespective of whether the analysis is based on parametric or nonparametric methods [@Zimmerman_1994, pp.
+irrespective of whether the analysis is based on parametric or nonparametric methods [@Zimmerman_1994, p.
 397-398] -->
@@
 One way to both use these extreme observations and reduce their influence is to employ some kind of robust regression.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 232-233] -->
@@
 Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 370] -->
@@
 
-<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp.
+<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, p.
 268].
@@
 <!-- 3. But does it hold?  -->
-Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, assessment designs often adopt the assumption of no correlation between stimuli based on an old theoretical justification.
-Specifically, @Thurstone_1927b argued that stimuli could be treated as uncorrelated because judges' biases--arising from two opposing and equally weighted effects occurring during the pairwise comparisons--would cancel each other out.
+Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, assessment designs often adopt the assumption of no correlation between stimuli based on Thurstone's early theoretical justification [@Thurstone_1927b].
+He argued that stimuli could be treated as uncorrelated because judges' biases--arising from two opposing and equally weighted effects occurring during the pairwise comparisons--would cancel each other out.
 This idea was later formalized by @Andrich_1978, who provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases.
-However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two cases: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.
+However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two cases: (1) when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions, (2) when additional hierarchical structures are relevant to the stimuli.
 
@@
 mood and the opposite effect of simultaneous contrast ....
-[@Thurstone_1927b, pp.
+[@Thurstone_1927b, p.
 268] -->
@@
 This preference for the object is then reflected in its comparisons with all other objects.
-However, it is less likely that subjects would have such particular preferences for some items in a test and, in any case, the assumption is open to empirical checking [@Andrich_1978, pp.
+However, it is less likely that subjects would have such particular preferences for some items in a test and, in any case, the assumption is open to empirical checking [@Andrich_1978, p.
 455].
@@
 <!-- 2. The first scenario where the assumption does not hold -->
-Research on text quality assessments suggests that when judges evaluate complex, multidimensional traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022].
-Regardless of their relevance, these characteristics may not receive equal weight or consistently oppose one another across comparisons.
+Regarding the first case, research on text quality assessments suggests that when judges evaluate complex, multidimensional traits or heterogeneous stimuli, they often rely on a variety of intricate stimulus characteristics to inform their judgments [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022].
+Regardless of their relevance, these characteristics may not be equally weighted or consistently opposed across comparisons.
 As a result, they may exert a disproportionate influence on judges' perceptions, generating biases that persist rather than cancel out.
-For example, this could occur when a judge assessing the argumentative quality of a text may place disproportionate emphasis on handwriting clarity, thereby favoring neatly written texts despite their weaker arguments.
+For example, this might occur when a judge assessing the argumentative quality of a text is disproportionately influenced by the clarity of the handwriting, thereby favoring neatly written texts even if their arguments are weaker.
 Moreover, because the discriminal process of stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].
-While direct evidence for this exact scenario is limited, existing studies document the presence of judge bias in CJ [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.
+While direct evidence for this exact scenario is limited, existing studies document the presence of judge's biases in CJ contexts [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], reinforcing the argument that the factors influencing pairwise comparisons do not always cancel each other out.
 
@@
 The association between the responses to the items given one proficiency parameter does then not vanish.
-[@vanderLinden_et_al_2017_II, pp.
+[@vanderLinden_et_al_2017_II, p.
 346].
@@
 As a result, generalizability of the scale across the groups is demonstrated.
-[@Andrich_1978, pp.459] -->
+[@Andrich_1978, p.459] -->
 
@@
 For instance, when the same individual produces multiple texts, those texts often share several features such as writing style or overall quality.
-Although the CJ literature acknowledges the existence of such hierarchical structures [e.g., @Boonen_et_al_2020], the statistical approaches to account for this additional source of dependence have been insufficient.
-For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
+Although some CJ studies acknowledge the presence of such hierarchical structures and account for them [e.g., @Boonen_et_al_2020], the treatment of this additional source of dependence in other research has often been insufficient.
+For instance, when CJ data include multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
 
@@
 <!-- Clustered data: A term applied to both data in which the sampling units are grouped into clusters sharing some common feature, for example, animal litters, families or geographical regions, and longitudinal data in which a cluster is deﬁned by a set of repeated measures on the same unit.
-[@Everitt_et_al_2010, pp.
+[@Everitt_et_al_2010, p.
 86] -->
@@
 These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli.
-One such spurious conclusion could be the incorrect classification of stimuli or judges as *misfits*.
+One such spurious conclusion could be the incorrect classification of stimuli (or judges) as *misfits*.
 @fig-correlation illustrates how assuming zero correlation can undermine trait reliability: the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference may correspond to any discontinuous line where $\rho \neq 0$.
@@
 The estimated implication is that if one applied a unidimensional IRT model, psychometric information is inflated due to the multidimensionality and examinee standard errors, which are the inverse square root of information, are too small.
-[@Hoyle_et_al_2023, pp.
+[@Hoyle_et_al_2023, p.
 340-341] -->
@@
 Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 370] -->
 
-Finally, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], and may appear as outliers under the BTL model [@Wu_et_al_2022]--risks discarding valuable information and introducing bias into trait estimates [@Miller_2023].
+Finally, similar to misfit stimuli, removing *misfit judges* risks discarding valuable information and introducing bias into trait estimates [@Miller_2023].
 The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
-
-<!-- When a judge compares two portfolios, Eq. 2 can be used to measure the degree of ‘surprise’ in their decision. ... [@Pollitt_2012a, pp. 164-165]  -->
+*Misfit* judges are those whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], often appearing as outliers under the BTL model [@Wu_et_al_2022].
+
+<!-- When a judge compares two portfolios, Eq. 2 can be used to measure the degree of ‘surprise’ in their decision. ... [@Pollitt_2012a, p. 164-165]  -->
 
@@
 To a significant degree, this judge interpreted the trait ‘good writing’ differently from the others.
-[@Pollitt_2012b, pp.
+[@Pollitt_2012b, p.
 289-290] -->
@@
 It is suggested that judges with the highest misfit statistic may have a diverging conceptualisation of the competence to be assessed (Bramley, 2007; Whitehouse & Pollitt, 2012).
-[@vanDaal_et_al_2016, pp.
+[@vanDaal_et_al_2016, p.
 4] -->
@@
 Representations with a large infit are representations which lead to more inconsistent judgments [33].
-[@Goossens_et_al_2018, pp.
+[@Goossens_et_al_2018, p.
 20] -->
@@
 As Liner and Amin (2004) pointed out, this is more likely to happen if some of the objects are involved in only a few comparisons with most of the others.
-[@Wu_et_al_2022, pp.
+[@Wu_et_al_2022, p.
 466] -->
@@
 Because any new sample is unlikely to contain these same outliers, and since these outliers were highly influential, they could make out-of-sample predictions worse than expected.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 231] -->
@@
 But if there are several outliers and we really need to model them, what then?
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 232] -->
 
-<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, p. 369] -->
 
@@
 Then the posterior mean of \theta is a decreasing function of any sufficiently large observation.
-[@OHagan_2018, pp.
+[@OHagan_2018, p.
 366] -->
@@
 <!-- 3. But what are the challenges in using BTL scores for further analysis? -->
-Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct statistical inference or hypothesis tests, as this practice can undermine the resulting statistical conclusions.
+Nevertheless, while separating the trait measurement and hypothesis testing procedures simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct statistical inference or hypothesis tests, as this practice can undermine the resulting statistical inferences.
 A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error).
@@
 Values of these correlations and their potential distorting effects are rarely known in practice, so it difficult to anticipate the magnitudes and directions of this propagation of measurement error in a single predictor.
-[@Kline_et_al_2023, pp.
+[@Kline_et_al_2023, p.
 25]  -->
@@
 If scores on exogenous variables are perfectly precise, then measurement error in endogenous variables results in truncated values of R2, or proportions of explained variation, and also truncated absolute values of standardized—but not unstandardized—regression coefficients, assuming this measurement error is unrelated to the exogenous variables (Williams, Grajales, & Kurkiewicz, 2013).
-[@Hoyle_et_al_2023, pp.
+[@Hoyle_et_al_2023, p.
 137] -->
@@
 <!-- Probably the majority of regressions are really using proxies like Dobs , because most variables are measurements with some error.
-But even though it doesn’t necessarily open a non-causal path, using a proxy can introduce systematic bias, distorting the estimates [@McElreath_2020, pp.
+But even though it doesn’t necessarily open a non-causal path, using a proxy can introduce systematic bias, distorting the estimates [@McElreath_2020, p.
 492].
@@
 But you can’t safely assume that measurement error makes estimates conservative.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 494] -->
@@
 This doesn't only apply to measurement error, but also to cases in which data are averaged before analysis.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 497] -->
@@
 In aggregate, the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates.
-This overreliance may also undermine their validity [@Perron_et_al_2015], particularly when coupled with the disconnect between the trait measurement and hypothesis testing processes.
+This overreliance may also undermine their validity [@Perron_et_al_2015], particularly when coupled with the disconnect between the trait measurement and hypothesis testing procedures.
 The structural approach to causal inference can address these issues by offering a systematic and integrated framework to extend Thurstone's general form.
@@
 
-<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. -->
+<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, p. 2]. -->
 
@@
 The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
-Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, p.
 4].
@@
 <!-- Estimator: A statistic used to provide an estimate for a parameter. The sample mean, for example, is an unbiased estimator of the population mean. -->
-<!-- [@Everitt_et_al_2010, pp. 154] -->
+<!-- [@Everitt_et_al_2010, p. 154] -->
 
@@
 These minimal building blocks consist of chains (Figure 3.9a), forks (Figure 3.9b), immoralities (Figure 3.9c), two unconnected nodes (Figure 3.10), and two connected nodes (Figure 3.11).
-[@Neal_2020, pp.
+[@Neal_2020, p.
 23]  -->
@@
 Conversely, it means that observing patterns of independencies in the data enables us to say something about whether a hypothesized model is correct.
-[@Pearl_et_al_2016, pp.
+[@Pearl_et_al_2016, p.
 35] -->
@@
 Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to causal inference.
-Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
+Specifically, it combines Thurstone's core theoretical principles outlined in @sec-thurstone_theory with key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
 @sec-theory-theoretical_P introduces the *conceptual-population model*, which incorporates these theoretical principles and assumes an idealized setting where researchers observe a *conceptual population* of comparative judgment data--that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
@@
 The functions F = {fz , fx , fy } are called structural equations, and each function represents a causal process that assigns to its respective endogenous variable a value based on the values of the other variables.
-[@Cinelli_et_al_2020, pp.
+[@Cinelli_et_al_2020, p.
 15] -->
@@
 The vertices in our graphs will correspond to variables (whence the common symbol V), and the edges will denote a certain relationship that holds in pairs of variables, the interpretation of which will vary with the application.
-[@Pearl_2009, pp.
+[@Pearl_2009, p.
 42] -->
@@
 <!-- where pa_{i} (connoting parents) stands for the set of variables that directly determine the value of X_{i} and where the U_{i} represent errors (or "disturbances") due to omitted factors.
-[@Pearl_2009, pp.
+[@Pearl_2009, p.
 65] -->
 
-<!-- These disturbances reflect "hidden" or unmeasurable conditions that Nature governs by some undisclosed probability function. [@Pearl_2009, pp. 89] -->
+<!-- These disturbances reflect "hidden" or unmeasurable conditions that Nature governs by some undisclosed probability function. [@Pearl_2009, p. 89] -->
 
@@
 If we know the value of every exogenous variable, then using the functions in f , we can determine with perfect certainty the value of very endogenous variable.
-[@Pearl_et_al_2016, pp.
+[@Pearl_et_al_2016, p.
 26-27] -->
@@
 The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli).
-The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for assessment conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, pp.
+The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for assessment conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, p.
 366-376].
@@
 <!-- Every SCM is associated with a graphical causal model, referred to informally as a “graphical model” or simply “graph.” Graphical models consist of a set of nodes representing the variables in U and V, and a set of edges between the nodes representing the functions in f.
-[@Pearl_et_al_2016, pp.
+[@Pearl_et_al_2016, p.
 27] -->
@@
 In the types of models we consider here, which do not exhibit cycles, the causal diagram will be a directed acyclic graph (DAG).
-[@Cinelli_et_al_2020, pp.
+[@Cinelli_et_al_2020, p.
 15] -->
@@
 And acyclic means that causes do not eventually flow back on themselves.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 128]  -->
@@
 <!-- Nodes that are represented by a solid circle are observed random variables, whereas nodes that are represented by a hollow circle are unobserved random variables.
-Causal effects are represented by directed edges (i.e., single-headed arrows), [@Morgan_et_al_2014, pp.
+Causal effects are represented by directed edges (i.e., single-headed arrows), [@Morgan_et_al_2014, p.
 79] -->
@@
 Doing so relaxes the requirement for perfect measurement of exogenous variables while also controlling for measurement error in endogenous variables.
-[@Hoyle_et_al_2023, pp.
+[@Hoyle_et_al_2023, p.
 138] -->
@@
 <!-- Accounting for measurement error in defining latent variables improves the accuracy of the model’s estimated structural parameters between the latent constructs, which in turn can provide a stronger test of a proposed theoretical model.
-[@Hoyle_et_al_2023, pp.
+[@Hoyle_et_al_2023, p.
 482] -->
@@
 
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, p.
 267] -->
@@
 
-<!-- This is a whole section [@McElreath_2020, pp. 499-516] -->
-
-<!-- This is a whole section [@Deffner_et_al_2022, pp. 6-7] -->
-
-<!-- What does “conditioning” mean? Informally, conditioning on a variable Z means learning its value and then asking if X adds any additional information about Y. [@McElreath_2020, pp. 130] -->
+<!-- This is a whole section [@McElreath_2020, p. 499-516] -->
+
+<!-- This is a whole section [@Deffner_et_al_2022, p. 6-7] -->
+
+<!-- What does “conditioning” mean? Informally, conditioning on a variable Z means learning its value and then asking if X adds any additional information about Y. [@McElreath_2020, p. 130] -->
 
@@
 In contrast, an intervention would be to take the whole population and give everyone treatment t.
-[@Neal_2020, pp.
+[@Neal_2020, p.
 32] -->
@@
 
-<!-- This is a whole section [@Lawson_2015, pp. 16-112] -->
+<!-- This is a whole section [@Lawson_2015, p. 16-112] -->
 
@@
 This structure implies that $C$ determines *which repeated judgments judges make for the stimuli produced by the individuals*.
-In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments but instead complete a sufficient number, $n_{C}$, to enable the accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, pp.
+In essence, $C$ reflects the assumption that judges *do not* perform all possible repeated judgments but instead complete a sufficient number, $n_{C}$, to enable the accurate estimation of the proportion $P(B>A)$ for each stimulus pair [@Thurstone_1927b, p.
 267].
 
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, p.
 267] -->
@@
 For the ‘random only’ analyses, this could be achieved by systematically sampling the required observations.
-[@Bramley_2015, pp.
+[@Bramley_2015, p.
 8] -->
 
-<!-- This is a whole section [@Lawson_2015, pp. 259-271] -->
+<!-- This is a whole section [@Lawson_2015, p. 259-271] -->
 
@@
 For any model whose graph is acyclic, the joint distribution of the variables in the model is given by the product of the conditional distributions P(child|parents) over all the "families" in the graph.
-[@Pearl_et_al_2016, pp.
+[@Pearl_et_al_2016, p.
 29] -->
@@
 <!-- The functions F = {fz, fx, fy} are called structural equations, and each function represents a causal process that assigns to its respective endogenous variable a value based on the values of the other variables.
-[@Cinelli_et_al_2020, pp.
+[@Cinelli_et_al_2020, p.
 15] -->
@@
 <!-- We will denote the parents of a node X with pa(X).
-We’ll use an even simpler shorthand when the nodes are ordered so that we can denote the i^{th} node by X_{i}; in that case, we will also denote the parents of X_{i} by pa_{i} [@Neal_2020, pp.
+We’ll use an even simpler shorthand when the nodes are ordered so that we can denote the i^{th} node by X_{i}; in that case, we will also denote the parents of X_{i} by pa_{i} [@Neal_2020, p.
 19] Definition 3.1 (Bayesian Network Factorization) Given a probability distribution P and a DAG G, P factorizes according to G if: P(x_{1}, \dots, x_{n}) = \prod_{i}P(x_{i} | pa_{i}).
-[@Neal_2020, pp.
+[@Neal_2020, p.
 21] -->
@@
 Formally, we write this rule as P(x_{1}, \dots, x_{n}) = \prod_{i}P(x_{i} | pa_{i}) where pa_{i} stands for the values of the parents of variable X_{i}, and the product i runs over all i, from 1 to n.
-[@Pearl_et_al_2016, pp.
+[@Pearl_et_al_2016, p.
 29] -->
@@
 
-[^8]: The binomial distribution--including its special case, the Bernoulli distribution--represent a maximum entropy distribution for binary events [@McElreath_2020, pp.
+[^8]: The binomial distribution--including its special case, the Bernoulli distribution--represent a maximum entropy distribution for binary events [@McElreath_2020, p.
 34].
-This means that the Bernoulli distribution is the most consistent alternative when only two un-ordered outcomes are possible and their expected frequencies are assumed to be constant [@McElreath_2020, pp.
+This means that the Bernoulli distribution is the most consistent alternative when only two un-ordered outcomes are possible and their expected frequencies are assumed to be constant [@McElreath_2020, p.
 310].
@@
 
-<!-- The binomial distribution is rather special, because it represents the maximum entropy way to count binary events. [@McElreath_2020, pp. 34] -->
+<!-- The binomial distribution is rather special, because it represents the maximum entropy way to count binary events. [@McElreath_2020, p. 34] -->
 
 <!-- Choosing the distribution with the largest entropy means spreading probability as evenly as possible, while still remaining consistent with anything we think we know about a process.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 299] -->
@@
 <!-- The principle of maximum entropy helps us choose likelihood functions, by providing a way to use stated assumptions about constraints on the outcome variable to choose the likelihood function that is the most conservative distribution compatible with the known constraints.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 300] -->
@@
 Entropy counts up the number of different ways a process can produce a particular result, according to our assumptions.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 310-311, chap.
@@
 An example of this is phenylketonuria in which phenylalanine in the diet (E) and the absence of the enzyme to break it down (phenylalanine hydroxylase, G) are both needed in order for a disease to manifest (intellectual disability, D).
-[@Attia_et_al_2022, pp.
+[@Attia_et_al_2022, p.
 1052] -->
@@
 The second method, and the one employed in the examples below, is to fix an otherwise free factor loading to a value of 1.
-[@Depaoli_2021, pp.
+[@Depaoli_2021, p.
 93] -->
@@
 
-Moreover, because Thurstone's original goal was to produce a "coarse scaling" of traits and allocate stimuli along this continuum [@Thurstone_1927a, pp.
+Moreover, because Thurstone's original goal was to produce a "coarse scaling" of traits and allocate stimuli along this continuum [@Thurstone_1927a, p.
 269], his theory offered no guidance on how to use trait estimates for statistical inference.
@@
 <!-- If Thurstone's Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes.
-[@Kelly_et_al_2022, pp.
+[@Kelly_et_al_2022, p.
 678] -->
@@
 However, work to derive some general principles for good practice would be valuable.
-[@Kelly_et_al_2022, pp.
+[@Kelly_et_al_2022, p.
 683] -->
@@
 Assuming ignorability is like ignoring how people ended up selecting the treatment they selected and just assuming they were randomly assigned their treatment; we depict this graphically in Figure 2.2 by the lack of a causal arrow from X to T.
-[@Neal_2020, pp.
+[@Neal_2020, p.
 9] -->
 
-<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, pp. 160] -->
+<!-- The condition of case selection to be independent from Y is always met in experiments because the Y is not realized before the experiment starts. [@Kohler_et_al_2019, p. 160] -->
 
@@
 A prominent example is the previously discussed ACJ design, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
-This pair selection process suggests that ACJ's comparison mechanism is outcome-dependent, potentially classifying the method as a generator of MNAR data.
+This pair selection procedure suggests that ACJ's comparison mechanism is outcome-dependent, potentially classifying the method as a generator of MNAR data.
 If this classification holds, the mixed findings on ACJ's capabilities become more comprehensible: some studies find that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others contend that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
@@
 The analogy is with computerised adaptive testing (CAT) where the choice of the next item to present is based on the examinee’s success or failure on items presented so far.
-[@Bramley_2015, pp.
+[@Bramley_2015, p.
 4].
@@
 The practical importance of this is that it is not necessary to have a completely crossed or balanced design of paired comparisons because the estimation process naturally handles missing data.
-[@Bramley_2008, pp.
+[@Bramley_2008, p.
 256] -->
@@
 However, the precision (standard error) of each script’s estimate depends largely on how many comparisons it has been involved in, so some effort needs to be made to ensure that each script is compared a similar number of times in total.
-[@Bramley_2008, pp.
+[@Bramley_2008, p.
 260] -->
@@
 At any point more data could be collected and added to the existing data to recover from accidents, to replace an unacceptable judge, to add extra scripts, to resolve conflicts, to investigate possible bias or to deepen the study of any aspect that proves especially interesting.
-[@Bramley_2008, pp.
+[@Bramley_2008, p.
 261] -->
@@
 Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked.
-[@Kelly_et_al_2022, pp.
+[@Kelly_et_al_2022, p.
 683] -->
@@
 Two random samples should give the same relative scale values because it is an inherent assumption of all statistical models that parameter estimates are "random sample-free." For the greatest benefits, the samples should be systematically different.
-[@Andrich_1978, pp.459] -->
+[@Andrich_1978, p.459] -->
 
@@
 Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained).
-[@Bramley_2008, pp.
+[@Bramley_2008, p.
 274] -->
@@
 <!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well.
-[@Kelly_et_al_2022, pp.
+[@Kelly_et_al_2022, p.
 681] -->
@@
 We do not yet know how widely it can be, or should be, applied.
-[@Pollitt_2012b, pp.
+[@Pollitt_2012b, p.
 297]-->
@@
 (2020) found differences between what judges from the UK, Sweden, and the USA valued when assessing design portfolios, providing evidence to support divergence along demographic lines.
-[@Kelly_et_al_2022, pp.
+[@Kelly_et_al_2022, p.
 683] -->
@@
 Demographic disparities among samples may result from (a) differences in the actual populations from which the samples are taken or (b) sampling procedures that differ among sites.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 6] -->
@@
 While some differences—especially those concerning the independent variable—are inconsequential for intended generalizations, differences concerning effect modifiers or mediators require statistical adjustment.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 11] -->
 
-<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, pp. 682]. -->
+<!-- To date, there is no evidence to support a minimum number of judges and judgments required to ensure an acceptable level of validity [@Kelly_et_al_2022, p. 682]. -->
 
@@
 However, work to derive some general principles for good practice would be valuable.
-[@Kelly_et_al_2022, pp.
+[@Kelly_et_al_2022, p.
 683] -->
 
-<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp.
+<!-- It is assumed that the single observer compares each pair of stimuli a "sufficient number of times" so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, p.
 267] -->
@@
 
-<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->
-
-<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->
-
-<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, pp. 1] -->
+<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, p. 260] -->
+
+<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, p. 7]. -->
+
+<!-- An outlying observation, or "outlier," is one that appears to deviate markedly from other members of the sample in which it occurs. [@Grubbs_1969, p. 1] -->
 
@@
 But if there are several outliers and we really need to model them, what then?
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 232] -->
@@
 It is thus unclear whether or not this divergence is welcomed.
-[@Kelly_et_al_2022, pp.
+[@Kelly_et_al_2022, p.
 681-682] -->
@@
 
-<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, pp. 369] -->
+<!-- The problem isn’t necessarily that “outliers” are bad data. Rather processes are often variable mixtures and this results in thicker tails. [@McElreath_2020, p. 369] -->
 
@@
 The proper response, when faced with low predictive capacity of presumed independent variables, is not only to search for new independent variables and/or alternative theoretical models; the proper recourse may also be to use more sophisticated measurement models.
-[@Wainer_et_al_1978, pp.
+[@Wainer_et_al_1978, p.
 332] -->
@@
 What is important is that the assumptions are transparent and logically connected to data analysis.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 13] -->
@@
 As we start to turn our attention towards more complex situations, however, it will become crucial to be explicit about what we know and what we assume about the variables relevant to our particular causal inference problem.
-[@Hernan_et_al_2025, pp.
+[@Hernan_et_al_2025, p.
 73] -->
@@
 For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 2] -->
@@
 <!-- A generalizable understanding of a given phenomenon, therefore, cannot be based only on the accumulation of data but requires the theory-driven testing of causal assumptions.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 13] -->
@@
 <!-- A structural causal framework encourages researchers to explicitly spell out their assumptions, removing verbal ambiguity and facilitating communication, and it calls for a cumulative approach to science as one study’s findings become the scaffolding assumptions of the next.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 14] -->
@@
 But this is no reason to embrace the status quo that often avoids causal language (Grosz et al., 2020)—assumptions do not disappear just because we ignore them.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 14] -->
 
-<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, pp.
+<!-- Causal inference frameworks do not magically guarantee value-free answers, but they force us to be precise about the questions we ask, and to be transparent about the assumptions that we are willing to make [@Rohrer_et_al_2022, p.
 4-5] -->
 
-<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->
+<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, p. 751] -->
 
@@
 But an incomplete model is still an improvement over no model at all.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 5] -->
 
-<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, pp. 11] -->
+<!-- A causal framework is not only useful for analysis but also aids research design [@Deffner_et_al_2022, p. 11] -->
 
@@
 Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical.
-[@Lash_et_al_2021, pp.
+[@Lash_et_al_2021, p.
 113] -->
@@
 <!-- If we are willing to assume that the depicted DAG is a causal DAG—which means that it includes all common causes of any pair of variables (Elwert, 2013)—we can algorithmically derive which variables need to be “conditioned” (see Box 1) on to identify the causal effect of interest.
-[@Deffner_et_al_2022, pp.
+[@Deffner_et_al_2022, p.
 4] -->
@@
 But they allow us to deduce which statistical models can provide valid causal inferences, assuming the DAG is true.
-[@McElreath_2020, pp.
+[@McElreath_2020, p.
 17] -->
 
-<!-- DAGs are not about adding assumptions—they are about revealing the assumptions that are otherwise made implicitly [@Sterner_et_al_2024, pp. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
-
-<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, pp. 4; @Sterner_et_al_2024, pp. 749]. -->
+<!-- DAGs are not about adding assumptions—they are about revealing the assumptions that are otherwise made implicitly [@Sterner_et_al_2024, p. 755; @Deffner_et_al_2022; @Pearl_et_al_2014] -->
+
+<!-- It has been argued that causal DAGs can be considered graphical representations of nonparametric SEMs [@Pearl_2012; @Kunicki_et_al_2023, p. 4; @Sterner_et_al_2024, p. 749]. -->
 
 <!-- DAG can be thought of as a “qualitative schematic” for a particular class of structural equation modeling (i.e., linear structural equation modeling), or a nonparametric SEM.
-Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, pp.
+Conversely, structural equation modeling may be thought of as an algebraic system for a causal DAG [@VanderWeele_et_al_2021; Kunicki_et_al_2023, p.
 12].
@@
 
-<!-- Consistency: An estimator is said to be consistent if it converges to its estimand as the sample size tends to infinity. [@Everitt_et_al_2010, pp. 100] -->
+<!-- Consistency: An estimator is said to be consistent if it converges to its estimand as the sample size tends to infinity. [@Everitt_et_al_2010, p. 100] -->
 
 <!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
-So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, p.
 5].
@@
 <!-- we not only underline the value of causal graphs as an intuitive and powerful tool for survey researchers to communicate their assumptions, be it for inferences on descriptive or causal parameters, but also highlight their value in uncovering nontrivial insights for improving more complicated statistical practices.
-[@Schuessler_et_al_2023, pp.
+[@Schuessler_et_al_2023, p.
 3] -->
@@
 
-# Conclussion {#sec-conclussion}
+# Conclusion {#sec-conclusion}
 
@@
 Researchers then proceed to identify the estimands.
-*Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp.
+*Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, p.
 4].
@@
 <!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests.
-Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp.
+Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, p.
 4].
@@
 
-![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}
+![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [p. 32] ](images/png/IEflow.png){#fig-IEflow fig-align="center" width=35%}
 
@@
 Without identification, researchers cannot achieve consistency, even with "infinite" and error-free data.
-As a result, deriving meaningful insights about an estimand from finite data becomes impossible [@Schuessler_et_al_2023, pp.
+As a result, deriving meaningful insights about an estimand from finite data becomes impossible [@Schuessler_et_al_2023, p.
 5].
@@
 <!-- The main reason to focus on identification is that if these parameters cannot be identified using "infinite" and error free data, we certainly cannot learn anything about it using finite data.
-So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, pp.
+So, for the time being, we assume large samples and error-free measurements [@Schuessler_et_al_2023, p.
 5].
@@
 When such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \mid X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].
-Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp.
+Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, p.
 32] (see @eq-CACE).
@@
 <!-- Conditioning on T=t just means that we are restricting our focus to the subset of the population to those who received treatment t.
-In contrast, an intervention would be to take the whole population and give everyone treatment t [@Neal_2020, pp.
+In contrast, an intervention would be to take the whole population and give everyone treatment t [@Neal_2020, p.
 32].
@@
 Naturally, the validity of claims about the causal effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set.
-However, as @Kohler_et_al_2019 [pp.
+However, as @Kohler_et_al_2019 [p.
 150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions.
@@
 
-Notably, the approach extends the ACE identification for a continuous variable $T$ as in @eq-mACE_cont, ensuring broad applicability across different causal scenarios [@Neal_2020, pp. 45]
+Notably, the approach extends the ACE identification for a continuous variable $T$ as in @eq-mACE_cont, ensuring broad applicability across different causal scenarios [@Neal_2020, p. 45]
 
@@
 The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-Essentially, these tools serve as *conceptual (theoretical) models* on which identification analysis rests [@Schuessler_et_al_2023, pp.
+Essentially, these tools serve as *conceptual (theoretical) models* on which identification analysis rests [@Schuessler_et_al_2023, p.
 4].
@@
 It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool.
-[@Hernan_et_al_2025, pp.
+[@Hernan_et_al_2025, p.
 75] -->
@@
 Second, they depict causal relationships in a non-parametric and fully interactive way.
-This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp.
+This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, p.
 35].
@@
 Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020].
-In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp.
+In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, p.
 27,68].
@@
 Notably, this encoding relies on two fundamental assumptions: the causal Markov and the faithfulness assumption.
-The *causal Markov assumption* links the graph structure to the joint probability distribution of variables by asserting that any variable in a graph is independent of all its non-descendants conditional on its parents [@Neal_2020, pp.
-20; @Hernan_et_al_2025, pp.
+The *causal Markov assumption* links the graph structure to the joint probability distribution of variables by asserting that any variable in a graph is independent of all its non-descendants conditional on its parents [@Neal_2020, p.
+20; @Hernan_et_al_2025, p.
 80].
-The *faithfulness assumption*, in turn, states that the observed statistical (in)dependencies in the data reflect the underlying causal structure among variables--except in rare degenerate cases [@Neal_2020, pp.
-100; @Hernan_et_al_2025, pp.
+The *faithfulness assumption*, in turn, states that the observed statistical (in)dependencies in the data reflect the underlying causal structure among variables--except in rare degenerate cases [@Neal_2020, p.
+100; @Hernan_et_al_2025, p.
 81].
@@
 That is, conditional on its parents, Vj is independent of its non-descendants; hence, a causal DAG is Markov with respect to the DAG G.
-[@Hernan_et_al_2025, pp.
+[@Hernan_et_al_2025, p.
 74] -->
@@
 See Fine Point 6.1 for a systematic summary of the graphical rules, and Fine Point 6.2 for an introduction to the concept of faithfulness.
-[@Hernan_et_al_2025, pp.
+[@Hernan_et_al_2025, p.
 79] -->
@@
 The assumption that the converse holds, i.e., that A is d-separated from B conditional on C if A is statistically independent of B given C, is a separate assumption—the faithfulness assumption described in Fine Point 6.2.
-[@Hernan_et_al_2025, pp.
+[@Hernan_et_al_2025, p.
 80] -->
@@
 Matching leads to unfaithfulness.
-[@Hernan_et_al_2025, pp.
+[@Hernan_et_al_2025, p.
 81] -->
 
-<!-- ![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%} -->
+<!-- ![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [p. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%} -->
 
 <!-- 2. What are the general implications of this? -->
-A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can represent the joint distribution of variables more efficiently [@Pearl_et_al_2016, pp.
+A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can represent the joint distribution of variables more efficiently [@Pearl_et_al_2016, p.
 29].
 This expression takes the form of a product of conditional probability distributions (CPDs) of the type $P(child \mid parents)$.
-This property is formally known as the *Bayesian Network factorization* (BNF, @eq-net_factor) [@Pearl_et_al_2016, pp.
-29; @Neal_2020, pp.
+This property is formally known as the *Bayesian Network factorization* (BNF, @eq-net_factor) [@Pearl_et_al_2016, p.
+29; @Neal_2020, p.
 21].
@@
 This encoding enables researchers with conceptual (theoretical) knowledge in the form of an SCM and DAG to predict patterns of (in)dependencies in the data.
-As highlighted by @Pearl_et_al_2016 [pp.
+As highlighted by @Pearl_et_al_2016 [p.
 35], these predictions depend solely on the structure of these conceptual models without requiring the quantitative details of the equations or the distributions of the errors.
@@
 
-<!-- The net result is that a researcher who has scientific knowledge in the form of a structural equation model is able to predict patterns of independencies in the data, based solely on the structure of the model’s graph, without relying on any quantitative information carried by the equations or by the distributions of the errors [@Pearl_et_al_2016, pp.
+<!-- The net result is that a researcher who has scientific knowledge in the form of a structural equation model is able to predict patterns of independencies in the data, based solely on the structure of the model’s graph, without relying on any quantitative information carried by the equations or by the distributions of the errors [@Pearl_et_al_2016, p.
 35].
@@
 The five fundamental building blocks described in @sec-appendixB31 clearly illustrate which (in)dependencies can SMCs and DAGs predict.
-For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] enables researchers to express the joint probability distribution of the observed variables as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, supporting the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$) [@Neal_2020, pp.
+For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] enables researchers to express the joint probability distribution of the observed variables as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, supporting the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$) [@Neal_2020, p.
 24].
@@
 Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of(in)dependencies from chains, forks, and colliders.
-These (in)dependencies apply to any data set generated by a causal model with those structures, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, pp.
+These (in)dependencies apply to any data set generated by a causal model with those structures, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, p.
 36].
@@
 This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent $(X_{1} \dsep X_{3})$, but conditionally dependent when controlling for $X_{2}$ $(X_{1} \ndsep X_{3} \mid X_{2})$.
-The authors @Pearl_et_al_2016 [pp.
-37, 40, 41] and @Neal_2020 [pp.
+The authors @Pearl_et_al_2016 [p.
+37, 40, 41] and @Neal_2020 [p.
 25–26] provide the mathematical proofs for these conclusions.
 
-<!-- These (in)dependencies will be true of every data set generated by a causal model with that graphical structure, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, pp.
+<!-- These (in)dependencies will be true of every data set generated by a causal model with that graphical structure, regardless of the specific functions attached to the SCM [@Pearl_et_al_2016, p.
 36].
@@
 The structural approach to causal inference translates probabilistic insights into actionable strategies seeking to identify the ACE from associational quantities.
-The approach achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp.
+The approach achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, p.
 34].
@@
 The modularity assumption underpins the concepts of manipulated graphs and Truncated Factorization, which are essential for representing interventions $P(Y_{i} \mid do(T_{i}=t))$ within SCMs and DAGs.
-*Manipulated graphs* simulate physical interventions by removing specific edges from a DAG, while preserving the remaining structure unchanged [@Neal_2020, pp.
+*Manipulated graphs* simulate physical interventions by removing specific edges from a DAG, while preserving the remaining structure unchanged [@Neal_2020, p.
 34].
@@
 The sufficient adjustment set (potentially empty) must block all non-causal paths between $X_{a}$ and $X_{b}$ without introducing new paths.
-If such a set exists, then $X_{a}$ and $X_{b}$ are *d-separated* by $X_{p}$ ($X_{a} \dsep X_{b} \mid X_{p}$) [@Pearl_2009], and $X_{p}$ satisfies the *backdoor criterion* [@Neal_2020, pp.
+If such a set exists, then $X_{a}$ and $X_{b}$ are *d-separated* by $X_{p}$ ($X_{a} \dsep X_{b} \mid X_{p}$) [@Pearl_2009], and $X_{p}$ satisfies the *backdoor criterion* [@Neal_2020, p.
 37].

### 2025-07-03 _tex/paper2.tex
Built site for gh-pages

@@
 \begin{abstract}
-This study revisits Thurstone's law of comparative judgment (CJ),
-focusing on two prominent issues of traditional approaches. First, it
-critiques the heavy reliance on Thurstone's Case V assumptions and, by
-extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
-Specifically, the study raises concerns about the assumptions of equal
-discriminal dispersions and zero correlation between the stimuli. While
-these assumptions simplify the trait measurement model, they may fail to
-capture the complexity of CJ data, potentially leading to unreliable and
-inaccurate trait estimates. Second, the study highlights the apparent
-disconnect between CJ's trait measurement and hypothesis testing
-processes. Although separating these processes simplifies the analysis
-of CJ data, it may also undermine the reliability of various statistical
-results derived from these processes.
-
-To address these issues, the study extends Thurstone's general form
-using a systematic and integrated approach based on Causal and Bayesian
-inference methods. This extension integrates core theoretical principles
-alongside key CJ assessment design features, such as the selection of
-judges, stimuli, and comparisons. It then translates these elements into
-a probabilistic statistical model for analyzing dichotomous CJ data,
-overcoming the rigid assumptions of Case V and the BTL model.
+This study revisits Thurstone's law of comparative judgment (CJ) by
+addressing two central issues in the CJ literature. First, it critiques
+the widespread reliance on Thurstone's Case V assumptions and, by
+extension, the Bradley-Terry-Luce (BTL) model in the analysis of CJ
+data. Specifically, this study argues that while the assumptions of
+equal discriminal dispersions and zero correlation between stimuli
+simplify the trait measurement model, they may fail to capture the
+complexity of some traits or account for heterogeneous stimuli,
+potentially leading to unreliable and inaccurate trait estimates.
+Second, the study highlights the apparent disconnect between trait
+measurement and hypothesis testing procedures in CJ applications. It
+contends that while separating these procedures simplifies the analysis
+of CJ data, this practice can also undermine the resulting statistical
+inferences.
+
+To address these issues, this study aims to extend Thurstone's general
+form using a systematic and integrated approach based on causal and
+Bayesian inference methods. This extension integrates Thurstone's core
+theoretical principles alongside key CJ assessment design features. It
+then translates these elements into a probabilistic statistical model
+for analyzing dichotomous CJ data, overcoming the rigid assumptions of
+Case V and the BTL model.
 
@@
 produces a dichotomous outcome, indicating which stimulus is perceived
-to have a higher trait level. For example, when assessing writing
-quality, judges compare pairs of written texts (the stimuli) to
-determine the relative writing quality each text exhibit (the trait)
+to have a higher trait level. For example, judges can compare pairs of
+written texts (the stimuli) to determine the relative writing quality
+each text exhibit (the trait)
 \citep{Pollitt_2012b, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
@@
 traits and competencies over the past decade. These studies have
-highlighted three aspects of the method's effectiveness: its
-reliability, validity, and practical applicability. Research on
-reliability suggests that CJ requires a relatively modest number of
-pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
-to generate trait scores that are as precise and consistent as those
-generated by other assessment methods
+highlighted three key strengths of the method: its reliability, its
+validity, and practical applicability. Research on reliability suggests
+that CJ requires a relatively modest number of pairwise comparisons
+\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to generate trait
+scores that are as precise and consistent as those generated by other
+assessment methods
 \citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
@@
 Lastly, research on its practical applicability highlights CJ's
-versatility across both educational and non-educational contexts
+versatility across educational and non-educational contexts
 \citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020b, Boonen_et_al_2020}.
@@
 critical issues unresolved. This study identifies and discusses two
-prominent issues of traditional approaches that can undermine the
-reliability and validity of CJ's trait estimates. First, it critiques
-the heavy reliance on Thurstone's Case V assumptions
-\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
-model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
-Specifically, the study raises concerns about the assumptions of equal
-discriminal dispersions and zero correlation between the stimuli. While
-these assumptions simplify the trait measurement model, they may fail to
-capture the complexity of some traits or account for heterogeneous
-stimuli, potentially leading to unreliable and inaccurate trait
-estimates. Second, the study highlights the disconnect between CJ's
-trait measurement and hypothesis testing processes. Although separating
-these processes simplifies the analysis of CJ data, it may also
-undermine the reliability of various statistical inferences derived from
-these processes
+prominent issues in the CJ literature that can undermine the reliability
+and validity of CJ's trait estimates. First, it critiques the widespread
+reliance on Thurstone's Case V assumptions \citep{Thurstone_1927b} and,
+by extension, the Bradley-Terry-Luce (BTL) model
+\citep{Bradley_et_al_1952, Luce_1959} in the analysis of CJ data.
+Specifically, this study argues that while the assumptions of equal
+discriminal dispersions and zero correlation between stimuli simplify
+the trait measurement model, they may fail to capture the complexity of
+some traits or account for heterogeneous stimuli, potentially leading to
+unreliable and inaccurate trait estimates. Second, the study highlights
+the disconnect between trait measurement and hypothesis testing
+procedures in CJ applications. In particular, it contends that while
+separating these procedures simplifies the analysis of CJ data, this
+practice can also undermine the resulting statistical inferences
 \citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.
 
-To address these issues, this study extends Thurstone's general form
-through a systematic and integrated approach that combines causal and
-Bayesian inference methods. In addition to potentially enhancing
-measurement reliability and validity, and improving statistical accuracy
-in hypothesis testing, this approach offers two key advantages. First,
-it clarifies the interactions among all actors and processes involved in
-CJ assessments. Second, it shifts the current comparative data analysis
-paradigm from passively accepting Case V and the BTL model assumptions
-to actively testing whether those assumptions fit the data under
-analysis.
-
-The study divides its content into six main sections.
+To address these issues, this study aims to extend Thurstone's general
+form through a systematic and integrated approach that combines causal
+and Bayesian inference methods. This extension integrates Thurstone's
+core theoretical principles alongside key CJ assessment design features,
+such as the selection of judges, stimuli, and comparisons. In addition
+to potentially enhancing measurement reliability and validity, and
+improving statistical accuracy in hypothesis testing, this approach
+offers two key advantages. First, it clarifies the interactions among
+all actors and processes involved in CJ assessments. Second, it shifts
+the current comparative data analysis paradigm from passively accepting
+Case V and the BTL model assumptions to actively testing whether those
+assumptions fit the data under analysis.
+
+The remainder of this study is organized into seven sections.
 Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
-theory. Section~\ref{sec-theory-issues} discusses the identified issues
-in detail. Section~\ref{sec-theoretical} extends Thurstone's general
-form to address these challenges. The extension integrates core
-theoretical principles alongside key CJ assessment design features, such
-as the selection of judges, stimuli, and comparisons.
-Section~\ref{sec-statistical} translates these theoretical and practical
-elements into a probabilistic statistical model to analyze dichotomous
-pairwise comparison data. Finally, Section~\ref{sec-discussion}
-discusses the findings, explores avenues for future research, and detail
-the challenges for future researchers.
+theory. Section~\ref{sec-theory-issues} examines the central issues
+identified in the CJ literature. Section~\ref{sec-theoretical} extends
+Thurstone's general form to address these issues. The extension
+integrates core theoretical principles alongside key CJ assessment
+design features, such as the selection of judges, stimuli, and
+comparisons. Section~\ref{sec-statistical} translates these theoretical
+and design elements into a probabilistic statistical model to analyze
+dichotomous pairwise comparison data. Section~\ref{sec-discussion}
+reviews the findings, outline directions for future research, discusses
+the study's limitations and details the challenges that applied
+researchers may encounter. Finally, Section~\ref{sec-conclusion}
+provides the concluding remarks.
 
@@
 
-In its most general form, Thurstone's theory addresses pairwise
-comparisons wherein a single judge evaluates multiple stimuli
-\citep{Thurstone_1927b}. The theory posits that two key factors
-determine the dichotomous outcome of these comparisons: the discriminal
-process of each stimulus and their discriminal difference. The
-\emph{discriminal process} captures the psychological impact each
+In its \emph{general form}, Thurstone's theory addresses pairwise
+comparisons of a single judge who evaluates multiple stimuli
+\citep{Thurstone_1927a, Thurstone_1927b}. The theory posits that two key
+factors determine the dichotomous outcome of these comparisons: the
+discriminal process of each stimulus and their discriminal difference.
+The \emph{discriminal process} captures the psychological impact each
 stimulus exerts on the judge or, more simply, his perception of the
@@
 
+These ideas become clearer with an example.
 Figure~\ref{fig-discriminal_process} illustrates the hypothetical
-discriminal processes along a quality trait continuum for two written
-texts. The figure indicates that the modal discriminal process for Text
-B is positioned further along the continuum than that of Text A
-\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality.
-Additionally, the figure highlights that Text B has a broader
-distribution compared to Text A, which arises from its larger
+discriminal processes for two written texts along a \emph{quality} trait
+continuum. The figure shows that the modal discriminal process for Text
+B lies further along the continuum than that of Text A
+\((T_{B} > T_{A})\), suggesting that Text B exhibits higher quality. The
+figure also shows that Text B has a broader distribution than Text A,
+indicating a greater variability in perception due to its larger
 discriminal dispersion \((\sigma_{B} > \sigma_{A})\).
@@
 
-\subcaption{\label{fig-discriminal_process}Discriminal processes}
+\subcaption{\label{fig-discriminal_process}Hypothetical discriminal
+processes for two written texts}
 
@@
 
-\subcaption{\label{fig-discriminal_difference}Discriminal difference}
+\subcaption{\label{fig-discriminal_difference}Hypothetical discriminal
+difference of two written texts}
 
@@
 However, since the individual discriminal processes of the stimuli are
-not directly observable, the theory introduces the \emph{law of
-comparative judgment}. This law posits that in pairwise comparisons, a
-judge perceives the stimulus with a discriminal process positioned
-further along the trait continuum as possessing more of the trait
-\citep{Bramley_2008}. This suggests that pairwise comparison outcomes
-depend on the relative distance between stimuli, not their absolute
-positions on the continuum. Indeed, the theory assumes that the
+not directly observable \citep{Thurstone_1927a}, the theory introduces
+the \emph{law of comparative judgment}. This law posits that in pairwise
+comparisons, a judge perceives the stimulus with a discriminal process
+positioned further along the trait continuum as possessing more of the
+trait \citep{Bramley_2008}. This suggests that pairwise comparison
+outcomes depend on the relative distance between stimuli, not their
+absolute positions on the continuum. Indeed, the theory assumes that the
 difference between the underlying discriminal processes of the stimuli,
@@
 Normal distribution \citep{Andrich_1978}. In this distribution, the mode
-(mean) represents the average relative separation between the stimuli,
-and its dispersion indicates the variability of that separation.
+(mean) represents the average relative separation between the stimuli
+\((T_{B} - T_{A})\), and its dispersion \(\sigma_{BA}\) indicates the
+variability of that separation.
 
@@
 modal discriminal processes, five discriminal dispersions, and \(10\)
-correlations--one per comparison (see Table~\ref{tbl-thurstone_cases}).
-However, a single judge could only provide \({5 \choose 2} = 10\) unique
-comparisons, an insufficient data set to estimate the required
-parameters.
-
-\begin{table}
-
-\caption{\label{tbl-thurstone_cases}Thurstones cases and their
-asumptions}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/thurstone_cases.png}
-
-}
-
-\end{table}%
+correlations--one per comparison. However, a single judge could only
+provide \({5 \choose 2} = 10\) unique comparisons, an ``insufficient''
+data set to estimate the required parameters.
 
@@
 assumptions of the general form and the five cases. For a detailed
-discussion of these cases and their progression, refer to
+discussion of these cases and their growing simplification, refer to
 \citet{Thurstone_1927b} and \citet{Bramley_2008}.
 
-However, Thurstone developed Case V prioritizing statistical simplicity
-over precise trait measurement and offering no guidance on how to use
-its trait estimates for statistical inference or hypothesis testing.
-Specifically, Thurstone cautioned that its use ``should not be made
-without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as it
-imposes the most extensive set of simplifying assumptions
-\citep{Bramley_2008, Kelly_et_al_2022} (see
+\begin{table}
+
+\caption{\label{tbl-thurstone_cases}Thurstones cases and their
+asumptions}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/thurstone_cases.png}
+
+}
+
+\end{table}%
+
+As the table suggest, Thurstone developed Case V with an emphasis on
+statistical simplicity, but this simplicity comes at the expense of
+accurate and precise trait measurement and practical guidance for
+inference. Specifically, Thurstone cautioned that Case V use ``should
+not be made without (an) experimental test''
+\citep[p.~270]{Thurstone_1927b}, as it imposes the most extensive set of
+simplifying assumptions \citep{Bramley_2008, Kelly_et_al_2022} (see
 Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
@@
 the compared stimuli on this continuum''
-\citep[pp.~269]{Thurstone_1927b}, his theory did not support formal
-statistical inference. Despite these limitations, CJ research has
-predominantly relied on Case V to measure different traits, which raises
-significant concerns about the reliability and validity of such
-measurements in contexts where the case's assumptions may not hold
+\citep[p.~269]{Thurstone_1927b}, his theory did not support formal
+statistical inference. Nevertheless, despite Case V's limitations, CJ
+research has predominantly relied on it to measure various traits,
+raising significant concerns about the reliability and validity of such
+measurements in contexts where its assumptions may not hold
 \citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
-tradition has attempted to address the gap in hypothesis testing by
+tradition has attempted to address the gap in statistical inference by
 relying on the point estimates of traits--or their transformations--the
 statistical literature cautions against using these estimates as the
-sole basis for statistical inference, as such practices introduce bias
-in the analysis and reduce the precision of hypothesis tests
-\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Next, both
-issues are discussed in more depth.
+sole basis for inference, as such practices introduce bias and reduce
+the precision of hypothesis tests
+\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. The next
+sections discuss both issues in greater depth.
 
@@
 
-As previously discussed, Case V remains the most widely used model in CJ
-literature. This preference largely stems from the widespread adoption
-of the BTL model, which provides a simplified statistical representation
-of the case. The BTL model mirrors most of Case V's assumptions, with
-one notable distinction. While Case V assumes a Normal distribution for
-the stimuli' discriminal processes, the BTL model uses the more
-mathematically tractable Logistic distribution
-\citep{Andrich_1978, Bramley_2008} (see
+Case V is the most widely used model in the CJ literature. This
+preference largely stems from the widespread adoption of the BTL model,
+which provides a simplified statistical representation of the case. The
+BTL model incorporates most of Case V's assumptions, with one notable
+exception. While Case V assumes a Normal distribution for the stimuli'
+discriminal processes, the BTL model uses the more mathematically
+tractable Logistic distribution \citep{Andrich_1978, Bramley_2008} (see
 Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
@@
 non-monotonic transformation
-\citep{vanderLinden_et_al_2017_I, McElreath_2021}. That is, as long as
-the substitution (transformation) preserves the data rank order, the
-choice of distribution for the discriminal processes is inconsequential.
-This condition is satisfied in this case, as the Normal and Logistic
-distributions exhibit analogous statistical properties, differing only
-by a scaling factor of approximately \(1.7\)
+\citep{vanderLinden_et_al_2017_I, McElreath_2021}. In other words, as
+long as the substitution (transformation) preserves the rank order of
+the data, the choice of distribution for the discriminal processes is
+inconsequential. The BTL model satisfies this condition because the
+Normal and Logistic distributions exhibit analogous statistical
+properties, differing only by a scaling factor of approximately \(1.7\)
 \citep{vanderLinden_et_al_2017_I}.
 
-However, Thurstone acknowledged that some assumptions of Case V could be
+However, Thurstone acknowledged that some assumptions of Case V are
 problematic when researchers assess complex traits or heterogeneous
-stimuli \citep{Thurstone_1927a}. Thus, given that modern CJ applications
-often involve such traits and stimuli, two key assumptions of Case V,
-and by extension, the BTL model, may not always hold in theory or
-practice. These assumptions are the equal dispersion and zero
+stimuli \citep{Thurstone_1927a}. Thus, given that contemporary CJ
+applications often involve such traits and stimuli, two key assumptions
+of Case V, and by extension, the BTL model, may not always hold in
+theory or practice. These assumptions are the equal dispersion and zero
 correlation between stimuli.
@@
 influencing the outcome of pairwise comparisons. A thought experiment
-can help illustrate this idea. In it, researchers observe the
+can help illustrate this idea. Suppose researchers observe the
 discriminal processes for two texts, A and B, assuming that the
@@
 their respective discriminal processes \citep{Thurstone_1927a}.
-Consequently, the outcome's ability to reflect \emph{true} differences
-between stimuli largely depends on the validity of the model's
-assumptions \citep{Kohler_et_al_2019}, in this case, the assumption of
-equal dispersions. When the assumption accurately captures the
-complexity of the data, the BTL model estimates a discriminal difference
-distribution that accurately represents the \emph{true} discriminal
-difference between the texts. This scenario is illustrated in
-Figure~\ref{fig-dispersion}, when the model's discriminal difference
+Consequently, the effectiveness of the outcome to reflect \emph{true}
+differences between stimuli largely depends on the validity of the
+model's assumptions \citep{Kohler_et_al_2019}, in this case, the
+assumption of equal dispersions. When the assumption accurately captures
+the complexity of the data, the BTL model estimates a discriminal
+difference distribution that accurately represents the \emph{true}
+discriminal difference between the texts. This scenario is illustrated
+in Figure~\ref{fig-dispersion}, when the model's discriminal difference
 distribution aligns with the \emph{true} discriminal difference
@@
 measurement model, evidence from the CJ literature suggests that this
-assumption may fail to accommodate for heterogeneous stimuli, such as
-handwritten texts or English compositions
+assumption may not hold for heterogeneous stimuli, such as handwritten
+texts or English compositions
 \citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
-The presence of the so-called \emph{misfit texts}--texts that elicit
-more judgment discrepancies than others
-\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}--may
-signal this limitation, as such discrepancies can arise from larger
-discriminal dispersions due to stimulus heterogeneity or because the
-texts are genuine outliers--that is, texts with distinctive
-characteristics that deviate markedly from the rest of the sample
+The presence of the so-called misfit texts signals this limitation.
+\emph{Misfit texts} are those that elicit more judgment discrepancies
+than others
+\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018},
+and these discrepancies may arise from larger discriminal dispersions
+caused by the stimulus' heterogeneity or because the texts are genuine
+outliers--i.e., texts with distinctive characteristics that deviate
+markedly from the rest of the sample in which they occur
 \citep{Grubbs_1969}. In either case, the BTL model's assumptions prevent
 it from adequately accounting for or addressing these anomalies, leaving
-exclusion of the ``problematic'' texts as the primary remedy
+exclusion of such ``problematic'' texts as the primary remedy
 \citep{Pollitt_2012a, Pollitt_2012b}.
@@
 assessment designs often adopt the assumption of no correlation between
-stimuli based on an old theoretical justification. Specifically,
-\citet{Thurstone_1927b} argued that stimuli could be treated as
+stimuli based on Thurstone's early theoretical justification
+\citep{Thurstone_1927b}. He argued that stimuli could be treated as
 uncorrelated because judges' biases--arising from two opposing and
@@
 CJ literature indicates that the assumption of zero correlation does not
-hold in practice in at least two cases: when intricate aspects of
+hold in practice in at least two cases: (1) when intricate aspects of
 multidimensional, complex traits or heterogeneous stimuli influence
-judges' perceptions or when additional hierarchical structures are
+judges' perceptions, (2) when additional hierarchical structures are
 relevant to the stimuli.
 
-Research on text quality assessments suggests that when judges evaluate
-complex, multidimensional traits or heterogeneous stimuli, they often
-rely on a variety of intricate stimulus characteristics to inform their
-judgments
+Regarding the first case, research on text quality assessments suggests
+that when judges evaluate complex, multidimensional traits or
+heterogeneous stimuli, they often rely on a variety of intricate
+stimulus characteristics to inform their judgments
 \citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
-Regardless of their relevance, these characteristics may not receive
-equal weight or consistently oppose one another across comparisons. As a
-result, they may exert a disproportionate influence on judges'
-perceptions, generating biases that persist rather than cancel out. For
-example, this could occur when a judge assessing the argumentative
-quality of a text may place disproportionate emphasis on handwriting
-clarity, thereby favoring neatly written texts despite their weaker
-arguments. Moreover, because the discriminal process of stimuli becomes
-an observable outcome only through the judges' perceptions, these biases
+Regardless of their relevance, these characteristics may not be equally
+weighted or consistently opposed across comparisons. As a result, they
+may exert a disproportionate influence on judges' perceptions,
+generating biases that persist rather than cancel out. For example, this
+might occur when a judge assessing the argumentative quality of a text
+is disproportionately influenced by the clarity of the handwriting,
+thereby favoring neatly written texts even if their arguments are
+weaker. Moreover, because the discriminal process of stimuli becomes an
+observable outcome only through the judges' perceptions, these biases
 could introduce dependencies between the stimuli
 \citep{vanderLinden_et_al_2017_II}. While direct evidence for this exact
-scenario is limited, existing studies document the presence of judge
-bias in CJ
+scenario is limited, existing studies document the presence of judge's
+biases in CJ contexts
 \citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
@@
 produces multiple texts, those texts often share several features such
-as writing style or overall quality. Although the CJ literature
-acknowledges the existence of such hierarchical structures
-\citep[e.g.,][]{Boonen_et_al_2020}, the statistical approaches to
-account for this additional source of dependence have been insufficient.
-For instance, when CJ data incorporates multiple samples of stimuli from
-the same individuals, researchers frequently rely on (averaged) point
-estimates of the BTL scores to conduct subsequent analyses and tests at
-the individual level
-\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
+as writing style or overall quality. Although some CJ studies
+acknowledge the presence of such hierarchical structures and account for
+them \citep[e.g.,][]{Boonen_et_al_2020}, the treatment of this
+additional source of dependence in other research has often been
+insufficient. For instance, when CJ data include multiple samples of
+stimuli from the same individuals, researchers frequently rely on
+(averaged) point estimates of the BTL scores to conduct subsequent
+analyses and tests at the individual level
+\citep{Bramley_et_al_2019, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
 
@@
 One such spurious conclusion could be the incorrect classification of
-stimuli or judges as \emph{misfits}. Figure~\ref{fig-correlation}
+stimuli (or judges) as \emph{misfits}. Figure~\ref{fig-correlation}
 illustrates how assuming zero correlation can undermine trait
@@
 
-Finally, removing \emph{misfit} judges--that is, judges whose
-assessments deviate markedly from the shared consensus
+Finally, similar to misfit stimuli, removing \emph{misfit judges} risks
+discarding valuable information and introducing bias into trait
+estimates \citep{Miller_2023}. The direction and magnitude of these
+biases remain unpredictable, as they depend on which judges researchers
+exclude from the analysis
+\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. \emph{Misfit}
+judges are those whose assessments deviate markedly from the shared
+consensus
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
-and may appear as outliers under the BTL model
-\citep{Wu_et_al_2022}--risks discarding valuable information and
-introducing bias into trait estimates \citep{Miller_2023}. The direction
-and magnitude of these biases remain unpredictable, as they depend on
-which judges researchers exclude from the analysis
-\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
+often appearing as outliers under the BTL model \citep{Wu_et_al_2022}.
 
@@
 Nevertheless, while separating the trait measurement and hypothesis
-testing processes simplifies the analysis of CJ data, the statistical
+testing procedures simplifies the analysis of CJ data, the statistical
 literature cautions against relying solely on the point estimates of BTL
 scores to conduct statistical inference or hypothesis tests, as this
-practice can undermine the resulting statistical conclusions. A key
+practice can undermine the resulting statistical inferences. A key
 consideration is that BTL scores are parameter estimates that inherently
@@
 the disconnect between the trait measurement and hypothesis testing
-processes. The structural approach to causal inference can address these
-issues by offering a systematic and integrated framework to extend
+procedures. The structural approach to causal inference can address
+these issues by offering a systematic and integrated framework to extend
 Thurstone's general form. This approach can also strengthen measurement
@@
 using the structural approach to causal inference. Specifically, it
-combines the core theoretical principles outlined in
+combines Thurstone's core theoretical principles outlined in
 Section~\ref{sec-thurstone_theory} with key CJ assessment design
@@
 times, i.e., a \emph{repeated measures designs}
-\citep[pp.~366-376]{Lawson_2015}. Thus, the indexing system supports
+\citep[p.~366-376]{Lawson_2015}. Thus, the indexing system supports
 comparisons between different texts written by the same student
@@
 estimation of the proportion \(P(B>A)\) for each stimulus pair
-\citep[pp.~267]{Thurstone_1927b}.
+\citep[p.~267]{Thurstone_1927b}.
 
@@
   distribution--represent a maximum entropy distribution for binary
-  events \citep[pp.~34]{McElreath_2020}. This means that the Bernoulli
+  events \citep[p.~34]{McElreath_2020}. This means that the Bernoulli
   distribution is the most consistent alternative when only two
   un-ordered outcomes are possible and their expected frequencies are
-  assumed to be constant \citep[pp.~310]{McElreath_2020}. For a detailed
+  assumed to be constant \citep[p.~310]{McElreath_2020}. For a detailed
   discussion of the binomial as a maximum entropy distribution, see
@@
 scaling'' of traits and allocate stimuli along this continuum
-\citep[pp.~269]{Thurstone_1927a}, his theory offered no guidance on how
+\citep[p.~269]{Thurstone_1927a}, his theory offered no guidance on how
 to use trait estimates for statistical inference. The CJ tradition has
@@
 comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
-pair selection process suggests that ACJ's comparison mechanism is
+pair selection procedure suggests that ACJ's comparison mechanism is
 outcome-dependent, potentially classifying the method as a generator of
@@
 
-\section{Conclussion}\label{sec-conclussion}
+\section{Conclusion}\label{sec-conclusion}
 
@@
 compute the estimand based solely on its assumptions, regardless of
-random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
+random variability \citep[p.~4]{Schuessler_et_al_2023}. An
 \emph{estimator} refers to a method or function that transforms data
@@
 \caption{\label{fig-IEflow}Identification-Estimation flowchart.
-Extracted and slightly modified from \citet[pp.~32]{Neal_2020}}
+Extracted and slightly modified from \citet[p.~32]{Neal_2020}}
 
@@
 As a result, deriving meaningful insights about an estimand from finite
-data becomes impossible \citep[pp.~5]{Schuessler_et_al_2023}. Therefore,
+data becomes impossible \citep[p.~5]{Schuessler_et_al_2023}. Therefore,
 to ensure accurate and reliable estimates, researchers prioritize
@@
 the subset of the population defined by the conditioning variable
-\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).
+\citep[p.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).
 
@@
 that \(X\) serves as a sufficient adjustment set. However, as
-\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
+\citet[p.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
 real world from observed data inevitably requires assumptions. This
@@
 variable \(T\) as in Equation~\ref{eq-mACE_cont}, ensuring broad
-applicability across different causal scenarios
-\citep[pp.~45]{Neal_2020}
+applicability across different causal scenarios \citep[p.~45]{Neal_2020}
 
@@
 on which identification analysis rests
-\citep[pp.~4]{Schuessler_et_al_2023}. Thus, using these tools,
+\citep[p.~4]{Schuessler_et_al_2023}. Thus, using these tools,
 researchers can determine which statistical models can identify (ACE,
@@
 defining the variables' data types, the functional form between them, or
-their parameters \citep[pp.~35]{Pearl_et_al_2016}.
+their parameters \citep[p.~35]{Pearl_et_al_2016}.
 
@@
 factors that the investigator chooses not to model explicitly
-\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
+\citep[p.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
 \emph{structural equations}, express the endogenous variables as
@@
 non-descendants conditional on its parents
-\citetext{\citealp[pp.~20]{Neal_2020}; \citealp[pp.~80]{Hernan_et_al_2025}}.
+\citetext{\citealp[p.~20]{Neal_2020}; \citealp[p.~80]{Hernan_et_al_2025}}.
 The \emph{faithfulness assumption}, in turn, states that the observed
@@
 structure among variables--except in rare degenerate cases
-\citetext{\citealp[pp.~100]{Neal_2020}; \citealp[pp.~81]{Hernan_et_al_2025}}.
+\citetext{\citealp[p.~100]{Neal_2020}; \citealp[p.~81]{Hernan_et_al_2025}}.
 
@@
 represent the joint distribution of variables more efficiently
-\citep[pp.~29]{Pearl_et_al_2016}. This expression takes the form of a
+\citep[p.~29]{Pearl_et_al_2016}. This expression takes the form of a
 product of conditional probability distributions (CPDs) of the type
@@
 Equation~\ref{eq-net_factor})
-\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
+\citetext{\citealp[p.~29]{Pearl_et_al_2016}; \citealp[p.~21]{Neal_2020}}.
 In this expression, \(pa(X_{i})\) denotes the set of variables that are
@@
 (in)dependencies in the data. As highlighted by
-\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
+\citet[p.~35]{Pearl_et_al_2016}, these predictions depend solely on the
 structure of these conceptual models without requiring the quantitative
@@
 \(X_{1}\) and \(X_{3}\) are unconditionally independent
-(\(X_{1} \:\bot\:X_{3}\)) \citep[pp.~24]{Neal_2020}. Conversely, when
+(\(X_{1} \:\bot\:X_{3}\)) \citep[p.~24]{Neal_2020}. Conversely, when
 \(X_{3}\) is unconditionally dependent on \(X_{1}\)
@@
 causal model with those structures, regardless of the specific functions
-attached to the SCM \citep[pp.~36]{Pearl_et_al_2016}. For instance,
+attached to the SCM \citep[p.~36]{Pearl_et_al_2016}. For instance,
 applying the BNF to the chain structure depicted in the SCM
@@
 dependent when controlling for \(X_{2}\)
-\((X_{1} \:\not\bot\:X_{3} \mid X_{2})\). The authors \citet[pp.~37, 40,
-41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
+\((X_{1} \:\not\bot\:X_{3} \mid X_{2})\). The authors \citet[p.~37, 40,
+41]{Pearl_et_al_2016} and \citet[p.~25--26]{Neal_2020} provide the
 mathematical proofs for these conclusions.
@@
 alters only the causal mechanism of that node, leaving others unchanged
-\citep[pp.~34]{Neal_2020}.
+\citep[p.~34]{Neal_2020}.
 
@@
 specific edges from a DAG, while preserving the remaining structure
-unchanged \citep[pp.~34]{Neal_2020}. In parallel, \emph{Truncated
+unchanged \citep[p.~34]{Neal_2020}. In parallel, \emph{Truncated
 Factorization} (TF) achieves a similar simulation by removing specific
@@
 (\(X_{a} \:\bot\:X_{b} \mid X_{p}\)) \citep{Pearl_2009}, and \(X_{p}\)
-satisfies the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020}.
+satisfies the \emph{backdoor criterion} \citep[p.~37]{Neal_2020}.
 

### 2025-07-03 paper2.qmd
final review and editing

@@
 <!-- 2. But what are limitations of Case V (and the BTL model)? -->
-However, Thurstone acknowledged that some assumptions of Case V are problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a].
+However, Thurstone acknowledged that some assumptions of Case V are problematic when assessing complex traits or heterogeneous stimuli [@Thurstone_1927a].
 Thus, given that contemporary CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice.
@@
 @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.
-Furthermore, if researchers acknowledge that *misfit statistics* may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics may unintentionally discard valuable information [@Miller_2023], and introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
-The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.
+Furthermore, if it is acknowledged that *misfit statistics* may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics may unintentionally discard valuable information [@Miller_2023], and introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020].
+The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli are excluded from the analysis.
 
@@
 Finally, similar to misfit stimuli, removing *misfit judges* risks discarding valuable information and introducing bias into trait estimates [@Miller_2023].
-The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
+The direction and magnitude of these biases remain unpredictable, as they depend on which judges are excluded from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
 *Misfit* judges are those whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], often appearing as outliers under the BTL model [@Wu_et_al_2022].
@@
 <!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
-Researchers in CJ studies typically use the BTL model to measure traits and position the compared stimuli along a latent continuum [@Thurstone_1927b].
-The CJ literature shows that research frequently relies on point estimates of these traits--typically the BTL scores or its transformations--to conduct statistical inference or hypothesis testing.
+In CJ studies, the BTL model is typically used to measure traits and position the compared stimuli along a latent continuum [@Thurstone_1927b].
+The CJ literature shows that studies frequently relies on point estimates of these traits--typically the BTL scores or its transformations--to conduct statistical inference or hypothesis testing.
 For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].
@@
 <!-- 1. The structural approach to causal inference -->
-The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data.
-The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ assessments.
-Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests.
-*Identification analysis* determines whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023].
+The *structural approach* to causal inference [@Pearl_2009; @Pearl_et_al_2016] offers a formal framework for identifying causes and estimating their effects using data.
+The approach relies on structural causal models (SCMs) and directed acyclic graphs (DAGs) to formally and graphically represent the assumed causal structure of a system [@Morgan_et_al_2014; @Gross_et_al_2018; @Neal_2020], such as the one found in CJ assessments.
+In essence, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023].
+*Identification analysis* determines whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability.
 Here, *estimands* represent the specific quantities researchers aim to determine (i.e., a parameter) [@Everitt_et_al_2010].
@@
 <!-- 3. Example of a CJ estimand and estimator, and its limitations -->
-A motivating example that will appear throughout this study clarifies these concepts.
-In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ assessment by randomly assigning students (individuals) to two groups, each receiving a different teaching method.
+As an illustration, consider a researcher seeking to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, the researcher designs a CJ assessment by randomly assigning students (individuals) to two groups, each exposed to a different teaching method.
 Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait).
-Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".
-
-Following standard CJ practices, researchers would typically use estimates from the BTL model--or its transformations--to approximate this estimand.
-However, as discussed in @sec-theory-issues, Thurstone's Case V and the BTL model exhibit several statistical and measurement limitations.
-These limitations hinder the model's ability to identify various estimands relevant to CJ inquiries, including the one described in the example.
+Based on this setup, researchers can reformulate the research question as the estimand: *On average, is there a difference in the ability to produce high-quality written texts between the two groups of students?* Following standard CJ practices, the researcher would then use point estimates from the BTL model, or its transformations, to approximate this estimand.
+However, as discussed in @sec-theory-issue1, Thurstone's Case V and the BTL model exhibit several statistical and measurement limitations.
+These limitations hinder the model's ability to identify and accurately estimate a range of estimands relevant to CJ inquiries, including the one described in this illustration.
 
@@
 First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2024].
-This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis.
+This feature allows the decomposition of complex structures into manageable components, facilitating their analysis.
 Second, they depict causal relationships in a non-parametric way.
@@
 <!-- 6. What the section will do? -->
-Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to causal inference.
-Specifically, it combines Thurstone's core theoretical principles outlined in @sec-thurstone_theory with key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
-@sec-theory-theoretical_P introduces the *conceptual-population model*, which incorporates these theoretical principles and assumes an idealized setting where researchers observe a *conceptual population* of comparative judgment data--that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
-Conversely, @sec-theory-theoretical_SC presents the *sample-comparison model*, which integrates the assessment design features and reflects a more realistic setting where researchers access only a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
+Thus, using SCMs and DAGs, this study extends Thurstone's general form to address the issues identified in @sec-theory-issues.
+This extension combines Thurstone's core theoretical principles (see @sec-thurstone_theory) with key CJ assessment design features, such as the selection of judges, stimuli, and comparisons.
+Specifically, @sec-theory-theoretical_P introduces the *conceptual-population model* (henceforth CPM), which incorporates these theoretical principles and assumes an idealized setting where researchers observe a *conceptual population* of comparative judgment data--that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
+Conversely, @sec-theory-theoretical_SC presents the *sample-comparison model* (hereafter SCM), which integrates the assessment design features and reflects a more realistic setting where researchers access only a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
 
@@
 
-## The conceptual-population model {#sec-theory-theoretical_P}
+## The conceptual-population model (CPM) {#sec-theory-theoretical_P}
 
@@
 
-<!-- 1. What is the assumption on the conceptual-population model -->
-In the conceptual-population model, the idealized scenario of a *conceptual population* of comparative data enables the integration of Thurstone's theoretical principles and provides a foundation for proposing innovations aimed at addressing some of the issues discussed in @sec-theory-issues.
+<!-- 1. What is the assumption on the CPM -->
+In the CPM, the idealized scenario of a *conceptual population* of comparative data enables the integration of Thurstone's theoretical principles and provides a foundation for proposing innovations aimed at addressing some of the issues discussed in @sec-theory-issues.
 
@@
 <!-- 2. the preliminary SCM for CJ -->
-[SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the conceptual-population model, which evaluates the impact of different teaching methods on students' writing ability.
+[SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the CPM evaluating the impact of different teaching methods on students' writing ability.
 This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{iahbjk})$ and several related variables.
@@
 The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli).
-The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for assessment conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015, p.
-366-376].
+The index $j$ indicates the judge conducting the comparison, while the index $k$ accounts for assessment conditions where a judge compares the same pair of stimuli multiple times, i.e., a *repeated measures designs* [@Lawson_2015].
 Thus, the indexing system supports comparisons between different texts written by the same student $(i = h;$ $a \neq b)$ and between texts written by distinct students $(i \neq h;$ where $a = b$ is permitted$)$, each compared once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K;$ where $n_{J}>1$ and $n_{K}\geq1)$.
@@
 Here, $n_{J}$ indicates the total number of judges, and $n_{K}$ denotes the number of repeated judgments each judge performs.
+
+<!-- repeated measures designs [@Lawson_2015, p. 366-376] -->
 
@@
 
-Conceptual-population model, scalar form.
+Conceptual-population model (CPM), scalar form.
 :::
@@
 
-Conceptual-population model, initial vectorized form.
+Conceptual-population model (CPM), initial vectorized form.
 :::
@@
 <!-- 1. What are we going to do? -->
-Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023] and Item Response Theory (IRT) [@Fox_2010; @vanderLinden_et_al_2017_I], the conceptual-population model integrates two *hierarchical structural components* to examine how different *relevant* [^2] variables--whether observed or latent--affect the primary latent variable of interest [@Everitt_et_al_2010].
-This hierarchical design enables researchers to formulate and test hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for a discussion of these considerations).
+Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023] and Item Response Theory (IRT) [@Fox_2010; @vanderLinden_et_al_2017_I], the CPM integrates two *hierarchical structural components* to examine how different *relevant* [^2] variables--whether observed or latent--affect the primary latent variable of interest [@Everitt_et_al_2010].
+This hierarchical design supports the formulation and testing of hypotheses that account for both the nested structure of stimuli and the uncertainties inherent in trait estimation (see @sec-theory-issue1b and @sec-theory-issue2 for a discussion of these considerations).
 
@@
 
-Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the conceptual-population model.
+Notably, all variables and functions shown in [SCM -@fig-cj09_scm] and [DAG -@fig-cj09_dag] are part of the set of endogenous variables $V$, structural equations $F$, and exogenous variables $E$ for the CPM.
 Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$ and the absence of connecting arrows.
@@
 <!-- 3. segway for next section -->
-Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
+Overall, the CPM extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
 These enhancements include accounting for judges' biases and integrating hierarchical structural components.
 Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
-Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario.
+Since this assumption is rarely met in practice, the study must consider a more realistic scenario.
 
@@
 
-Conceptual-population model, final vectorized form.
-:::
-
-## The sample-comparison model {#sec-theory-theoretical_SC}
+Conceptual-population model (CPM), final vectorized form.
+:::
+
+## The sample-comparison model (SCM) {#sec-theory-theoretical_SC}
 
@@
 <!-- 1. What is the more realistic scenario: the sample mechanism -->
-The sample-comparison model presents a more realistic scenario than the conceptual-population model.
-First, it explicitly assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population (@sec-theory-theoretical_SC1).
+The SCM presents a more realistic scenario than the CPM.
+First, it explicitly assumes a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population (@sec-theory-theoretical_SC1).
 Second, the model assumes that judges do not perform *all repeated judgments* within the data sample (@sec-theory-theoretical_SC2).
@@
 Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled repeated judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.
-Notice that in this example, because the design systematically excludes every third repeated judgment, researchers can also express $S_{JK}$ using $n_{K} = n^{s}_{K} = 2$.
+Notice that in this example, because the design systematically excludes every third repeated judgment, $S_{JK}$ can also be expressed using $n_{K} = n^{s}_{K} = 2$.
 
@@
 <!-- 2. Integrate the sample mechanism -->
-With the definition of $S$, we incorporate the sample mechanism into the conceptual-population model.
-Following the convention of @McElreath_2020 and @Deffner_et_al_2022, [DAG -@fig-cj14_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.
+With the definition of $S$, we incorporate the sample mechanism into the CPM.
+Following the convention of @McElreath_2020 and @Deffner_et_al_2022, [DAG -@fig-cj14_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that this outcome cannot be directly accessed due to the sampling mechanism.
 The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$.
 A square encloses $S$, indicating that it is a conditioned variable.
-In this context, *conditioning* means that researchers restrict their focus to the elements of $O^{cp}_{R}$ that satisfy $s_{(i,a,h,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
+In this context, *conditioning* means that the analysis is restricted to the elements of $O^{cp}_{R}$ that satisfy $s_{(i,a,h,b,j,k)}=1$ [@Neal_2020; @McElreath_2020].
 In essence, $S$ is a vector that selects *all repeated judgments made by a subset of judges for a subset of stimuli produced by the sampled individuals*.
@@
 <!-- 2. What is the effect of the comparison mechanism -->
-The [DAG -@fig-cj14_dag] also incorporates the *comparison mechanism* $C$ into the conceptual-population model.
-It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, emphasizing that researchers cannot directly access this variable because of the comparison mechanism.
+The [DAG -@fig-cj14_dag] also incorporates the *comparison mechanism* $C$ into the CPM.
+It shows the sample-comparison outcome $O^{sc}_{R}$ as unobserved, emphasizing that this variable cannot be directly accessed because of the comparison mechanism.
 The DAG further shows $C$ as a conditioned variable (enclosed in a square) that causally influences the observed outcome $O_{R}$.
@@
 In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model.
-These extensions account for judge biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and even clarify the role of the sample and comparison mechanisms in CJ assessments (see @sec-theory-theoretical_SC).
+These extensions account for judges' biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and even clarify the role of the sample and comparison mechanisms in CJ assessments (see @sec-theory-theoretical_SC).
 However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a.
@@
 <!-- 1. How to derive an stat model from an SCM? -->
-Using the structural causal model (SCM) [-@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption (see @sec-theory-issue1a).
+Using the [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption (see @sec-theory-issue1a).
 This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016].
@@
 However, Thurstone identified one key challenge in this general formulation: the measurement model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
-To address this issue and facilitate the theory's practical applicability, he formulated five cases, each progressively incorporating several simplifying assumptions.
+To address this issue and to facilitate the theory's practical applicability, he formulated five cases, each progressively incorporating several simplifying assumptions.
 
@@
 Among these, Case V remains the most widely used model in empirical CJ research, mainly due to the widespread adoption of the BTL model.
-The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008].
+The BTL model incorporates the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008].
 Although this substitution has minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], the simplifying assumptions of the BTL model--and by extension, of Case V--may fail to capture the complexity of some traits or account for heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022], potentially leading to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].
@@
 <!-- 4. General summary of what we have done. -->
-To address the limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods.
-The approach begins with the development of a conceptual model, represented as a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
-This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms, to disentangle the causal processes underlying the CJ system.
-
-The approach then translates the SCM into a bespoke statistical model that allows researchers to analyze CJ data when violations to the assumptions of equal dispersion and zero correlation occur, and when statistical inference is necessary.
-In particular, this model accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error into the hypothesis testing process, and accommodates heterogeneity in discriminal dispersions.
-By addressing all these issues, these methodological innovations have the potential to enhance the reliability and validity of trait measurement in CJ [@Perron_et_al_2015], while also improving the accuracy of statistical inferences.
+To address the limitations of Thurstone's Case V and the BTL model, this study extended Thurstone's general form using a systematic, integrated approach that combined causal and Bayesian inference methods.
+The approach began with the development of a conceptual model, formalized as a Structural Causal Model (SCM) and represented graphically by a Directed Acyclic Graph (DAG) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020].
+This model integrated Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ assessment design features, including judges' bias, sampling procedures, and comparison mechanisms.
+Together, these components allowed the causal processes underlying the CJ system to be disentangled.
+
+The approach then translated the SCM into a bespoke statistical model that enabled the analysis of CJ data in cases where the assumptions of equal dispersion and zero correlation were violated, and where statistical inference was required.
+In particular, the model accounted for judge biases, captured the hierarchical structure of stimuli, incorporated measurement error into the hypothesis testing process, and accommodated heterogeneity in discriminal dispersions.
+By addressing these issues, the methodological innovations aimed to enhance the reliability and validity of trait measurement in CJ, while also improving the accuracy of statistical inferences.
 
@@
 <!-- 5. Are there any additional advantages? -->
-Beyond these potential benefits, the approach offers two additional advantages.
-First, it clarifies the roles and interactions of all actors and processes involved in CJ assessments.
-Second, it shifts the analytic paradigm from passively accepting the assumptions of Case V and the BTL model to actively testing their fit with observed data.
-Together, these advantages establish a principled framework for evaluating best practices in designing CJ assessments, one that better aligns with the demands of contemporary CJ contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.
+Beyond these potential benefits, the approach offered two additional advantages.
+First, it clarified the roles and interactions of all actors and processes involved in CJ assessments.
+Second, it shifted the analytic paradigm from passively accepting the assumptions of Case V and the BTL model to actively testing their fit with observed data.
+Together, these advantages established a principled framework for evaluating best practices in designing CJ assessments--one that better aligns with the demands of contemporary CJ contexts [@Kelly_et_al_2022]--providing new insights into existing research and opening promising avenues for future inquiry.
 
@@
 While these studies offer valuable insights, they also overlook the broader role that these mechanisms play within the CJ system.
-As this oversight likely stems from a more fundamental lack of conceptual clarity about how these mechanisms function within the system, this study integrates these mechanisms into the conceptual model of CJ.
+As this oversight likely stems from a more fundamental lack of conceptual clarity about how these mechanisms function within the system, the present study integrated these mechanisms into the conceptual model of CJ.
 
@@
 The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process.
-Specifically, it clarifies their role as sources of missing data in CJ's data-generating process, that is, as mechanisms that determine which observations are missing from the final data sample.
-This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], allowing a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex assessments setups.
+Specifically, it clarified their role as sources of missing data in CJ's data-generating process--that is, as mechanisms that determine which observations are missing from the final data sample.
+This new perspective invites the application of Little and Rubin's principled missing data framework [@Little_et_al_2020], allowing a more rigorous evaluation of existing claims about these missing data mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex assessments setups.
 
 <!-- 3. Leveraging MCAR assumptions to simplify CJ analysis in this study -->
-This study circumvents the need to apply this missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome.
-In other words, these mechanisms are designed to produce data that are *missing completely at random* (MCAR) [@Little_et_al_2020].
-This design offers one key advantage: it generates simple random samples that satisfy the condition of *ignorability*, allowing researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019; @Neal_2020].
+This study circumvented the need to apply the missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome.
+In other words, these mechanisms were intentionally designed to yield data that are *missing completely at random* (MCAR) [@Little_et_al_2020].
+This design choice offered one key advantage: it generated simple random samples that satisfied the condition of *ignorability*, thereby allowing researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019; @Neal_2020].
 
@@
 However, many modern CJ applications rely on more complex assessment designs, in which the sampling and comparison mechanisms introduce more intricate forms of missingness such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020].
-A prominent example is the previously discussed ACJ design, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
+One prominent example is ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015].
 This pair selection procedure suggests that ACJ's comparison mechanism is outcome-dependent, potentially classifying the method as a generator of MNAR data.
-If this classification holds, the mixed findings on ACJ's capabilities become more comprehensible: some studies find that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others contend that it artificially inflates these gains [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
+If this classification holds, ACJ might violate the condition of ignorability, making it an unsuitable pair selection procedure for reliable and valid trait estimation and inference.
+Moreover, under this interpretation, the mixed findings on ACJ's effectiveness become more comprehensible: while some studies find that the method improves trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], others argue that that these gains may be artificially inflated [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].
 
@@
 <!-- 5. The importance of clarifying assumptions behind CJ assessment designs -->
-Regardless of the underlying missingness mechanisms, any CJ assessment design would benefit from explicitly defining its assumptions--a practice supported by our approach.
-This clarity enables researchers to evaluate how the sampling and comparison mechanisms affect trait estimation and statistical inference within each design.
+Regardless of the underlying missingness mechanisms, any CJ assessment design would benefit from explicitly defining its assumptions--a practice supported by this study's approach.
+This clarity enables a thorough evaluation of how the sampling and comparison mechanisms affect trait estimation and statistical inference in each design.
 Such assessments are particularly relevant given the common misconception in the CJ literature that Thurstone's model can naturally handle even non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].
@@
 
-### The effects of judges' bias on the reliability of traits {#sec-discussion_RA2}
+### The effects of judges' bias on the reliability of trait estimates {#sec-discussion_RA2}
 <!-- recording: Sven and Tine 25.05.21; time: 00:24:00 - 00:31:45 -->
@@
 <!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
-Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
-This gap likely persists not due to a lack of interest or research but because researchers often rely on ad-hoc detection methods, such as 'misfit' statistics, that may not be well-suited for the task [@Kelly_et_al_2022].
-To address this limitation, this study treats judges' biases as an integral component of the CJ system from the outset.
-This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.
+Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, empirical evidence of judges’ biases remains scarce in the CJ literature [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
+This gap likely persists not from a lack of interest or research but because the identification of such biases often depends on ad-hoc detection methods, like 'misfit' statistics, that may not be well-suited for the task [@Kelly_et_al_2022].
+To overcome this limitation, the present study treated judges' biases as an integral component of the CJ system from the outset.
+This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons, one that acknowledges that the discriminal processes of stimuli becomes an observable outcome only through judges' perceptions, which may exhibit bias.
 
@@
 The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions.
-One key question is whether researchers can validly analyze CJ data under the assumption of "sample-free" trait calibration, specifically under the hypothesis that judges exhibit no systematic bias.
-This question is particularly relevant because many researchers still regard "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases.
+One key question is whether CJ data can be validly analyzed under the assumption of "sample-free" trait calibration, specifically under the hypothesis that judges exhibit no systematic bias.
+This question is particularly relevant because "sample-freeness" is still regarded as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases.
 Another critical question is whether training or expertise can help judges avoid focusing on irrelevant stimulus features, such as handwriting, over more central criteria like argumentative quality in writing assessments [@Kelly_et_al_2022].
@@
 <!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
-Moreover, since judges rely on these stimulus-related factors when evaluating complex, multidimensional traits [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these factors account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that assessments also vary according to judge-specific attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022].
-Prior studies support this view [@Bartholomew_et_al_2020a; @McMahon_et_al_2015].
-Thus, building on the discussion in @sec-discussion_RA1, researchers could further explore how judges' selection influences the formation of a "shared consensus" and whether these attributes introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022].
-Furthermore, if such attributes indeed undermine the assumption of "sample-freeness," it becomes essential to explore strategies for mitigating their influence and to determine how many judges (and how many judgments per judge) are needed to produce reliable trait estimates under these conditions.
-In addition, it is worth considering whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
-As anticipated, the approach presented in this study provides the necessary structure to investigate rigorously these questions.
+Moreover, since judges rely on these stimulus-related factors when evaluating complex, multidimensional traits (e.g., structure, style) [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022], and these factors account for variation in judgment accuracy [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020; @Gijsen_et_al_2021], it is reasonable to expect that assessments also vary according to judge-specific attributes such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022].
+Prior studies support this view [e.g., @Bartholomew_et_al_2020a; @McMahon_et_al_2015].
+Thus, building on the discussion in @sec-discussion_RA1, further exploration is warranted into how judge selection influences the formation of a "shared consensus" and whether certain judge attributes introduce systematic biases or distortions in the stimuli trait distribution [@Deffner_et_al_2022].
+Furthermore, if such attributes do in fact compromise the assumption of "sample-freeness," it becomes essential to consider strategies for mitigating their effects and to determine how many judges (and how many judgments per judge) are required to produce reliable trait estimates under these conditions.
+Additionally, it is worth examining whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.
+As anticipated, the approach presented in this study provides a structured foundation to rigorously investigate  these questions.
 
@@
 Although the CJ literature clearly defines *misfit* judges and stimuli, CJ researchers have rarely examined how these observations relate to Thurstonian theory.
-In particular, they have not identified which elements of Thurstone's theory might account for the occurrence of misfits.
-This disconnect likely stems from the fact that CJ researchers derive misfit statistics from residual analysis and outlier detection methods rather than from Thurstonian principles.
+In particular, they have not identified which elements of Thurstone's theory account for the occurrence of misfits.
+This disconnect likely stems from the fact that misfit statistics are derived from residual analysis and outlier detection methods rather than from Thurstonian principles.
 Specifically, *misfit judges* are typically defined as those whose assessments diverge significantly from the "shared consensus" [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018; @Wu_et_al_2022], while *misfit stimuli* are those that elicit more judgment discrepancies than others [@Pollitt_2004; @Pollitt_2012a; @Pollitt_2012b; @Goossens_et_al_2018].
 Both definitions closely mirror the statistical concept of outliers, that is, observations that deviate markedly from the rest of the sample in which they occur [@Grubbs_1969].
-But this resemblance extends beyond the definitions themselves, as CJ researchers often identify misfits using conventional outlier detection procedures, such as transforming BTL model residuals into diagnostic statistics and comparing them against predefined thresholds [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022].
+But this resemblance extends beyond the definitions themselves, as misfits are often identified using conventional outlier detection procedures, such as transforming BTL model residuals into diagnostic statistics and comparing them against predefined thresholds [@Pollitt_2012a; @Pollitt_2012b; @Wu_et_al_2022].
 
@@
 
-Nevertheless, it is not the classification of misfits as outliers that raises concerns for trait measurement and inference.
-Rather, the concern lies in the prevailing CJ practice of classify them through ad-hoc detection procedures and then excluding them from analysis [@Pollitt_2012a; @Pollitt_2012b], often without providing empirical evidence for the various hypotheses proposed to explain their occurrence.
-In this regard, it is essential to recognize that outliers can only be defined relative to a specific model [@McElreath_2020].
-As such, detection procedures based on models like the BTL, which relies on strong assumptions, should be approached with caution, as they may not be well-suited for the task [@Kelly_et_al_2022].
-In particular, because the BTL model may not always accurately reflect the underlying data-generating process of the CJ system, misfit classification based on this model risks misidentifying observations.
-Furthermore, the exclusion of these observations carries additional risks.
+Nevertheless, is not the classification of misfits as outliers that raises concerns for trait measurement and inference.
+Rather, the concern lies in the prevailing CJ practice of identifying these observations through ad-hoc procedures and excluding them from analysis [@Pollitt_2012a; @Pollitt_2012b], often without empirical support for the various hypotheses suggested to explain their occurrence.
+It is essential to recognize that outliers are defined relative to a specific model [@McElreath_2020].
+Thus, detection procedures based on models like the BTL--which rests on strong assumptions--should be approached with caution, as they may not be appropriate for this purpose [@Kelly_et_al_2022].
+If the BTL model does not accurately represent the actual data-generating process of the CJ system, the method may miss classify valid observations as misfits.
+On top this, excluding these observations carries additional risks.
 The statistical literature cautions that removing outliers can discard valuable information [@Miller_2023] and introduce bias into trait estimates.
-The direction and magnitude of this bias are often unpredictable, as they depend on which observations researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
-Finally, precisely because of its rigid assumptions, the BTL model lacks the flexibility to adequately test many of the hypotheses proposed to explain the presence of misfits.
+The direction and magnitude of this bias are often unpredictable, as they depend on which observations are excluded from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].
+Finally, even when the model aligns well with the data, its rigid assumptions limit the model's ability to adequately test many of the hypotheses proposed to explain misfit behavior.
 
@@
 
-In contrast, the approach presented in this study provides a rigorous framework that enables the examination of several relevant hypotheses.
-For instance, researchers could investigate whether misfit judges are those who exhibit (an outlying degree of) systematic bias or greater variability in their judgments compared to their peers.
-Similarly, researchers might explore whether misfit stimuli exhibit more variable discriminal processes relative to other stimuli or if they are genuinely outlying cases.
-Moreover, since outliers are not necessarily "bad data" [@McElreath_2020], our approach also offers a principled alternative that retains misfits in the analysis without compromising trait estimation or inference.
-This alternative involves adapting the proposed model into robust measurement models [@McElreath_2020], a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].
-This alternative also relates to a broader discussion in social science, which is: that when researchers face low predictive capacity from their models, they should not only search for "new" variables or alternative procedures but also consider employing more sophisticated measurement models [@Wainer_et_al_1978].
+In contrast, the approach presented in this study provides a rigorous framework for testing several relevant hypotheses.
+For instance, it allows to investigate whether misfit judges are those who exhibit (an outlying degree of) systematic bias or greater variability in their judgments compared to their peers.
+Similarly, the approach makes possible to explore whether misfit stimuli exhibit more variable discriminal processes relative to other stimuli or if they are genuinely outlying cases.
+Moreover, since outliers are not inherently "bad data" [@McElreath_2020], the present approach offers a principled alternative to exclusion, one that retains misfits in the analysis without compromising trait estimation or inference.
+This is achieved by adapting the proposed model into robust measurement models [@McElreath_2020], a broad class of procedures designed to reduce the sensitivity of parameter estimates to mild or moderate departures from model assumptions [@Everitt_et_al_2010].
+This strategy also speaks to a broader concern in the social sciences: when predictive power is low, the solution may not lie solely in seeking new variables or procedures, but also in adopting more sophisticated measurement models [@Wainer_et_al_1978].
 
@@
 
-The process of deriving conclusions from observed data always requires assumptions, whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022].
+Drawing conclusions from observed data always requires assumptions, whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022].
 The proposed approach is not an exception to this fundamental principle.
-As with all approaches grounded on causal inference, it relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
-As such, the approach does not seek to yield automatic answers when applied to a given CJ dataset.
-Instead, it aims to encourage researchers to formulate precise questions and to make their assumptions explicit, fostering a generalizable understanding of the CJ system under study [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024].
-This clarity is critical because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019].
-Although this alignment remains empirically untested for the proposed models, the theory-driven nature of our approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], which are well supported by both theory and existing evidence.
+Like all approaches grounded on causal inference, it relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025].
+However, its purpose is not to deliver automatic answers when applied to a given CJ dataset.
+Instead, it encourages the formulation of precise questions and the explicit articulation of assumptions, fostering a generalizable understanding of the CJ system under study [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024].
+This clarity is crucial because the accuracy of estimates and validity of inferences depend on how well the data and inferential goals align with a model's assumptions [@Kohler_et_al_2019].
+Although this alignment remains to be empirically tested for the proposed models, the theory-driven nature of this approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], grounded in both established theory and existing evidence.
 
@@
 <!-- 1. Challenges for the approach: background knowledge and research context -->
-Moreover, this theoretical commitment to causal inference also introduces several practical challenges that applied CJ researchers must navigate.
+The theoretical commitment to causal inference also introduces several practical challenges that applied CJ researchers must navigate.
 These fall into two main categories: first, acquiring the foundational knowledge necessary to apply the approach effectively; and second, dedicating greater attention to both conceptual and statistical modeling.
@@
 
-To apply the approach effectively, CJ researchers require foundational knowledge in two areas.
-First, they need a solid understanding of causal inference principles.
-Second, they must learn to translate the functional and probabilistic content of conceptual models into bespoke statistical models.
-One example that illustrates the importance of a sound knowledge of causal inference is the recurrent assumption that predictor variables are "relevant" to the research context--interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
-However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference.
-As a result, CJ researchers must deeply engage with these and other complex ideas, including how SCMs encode functional and probabilistic information.
-To assist this effort, this study includes key references throughout the study to guide applied CJ researchers toward a deeper understanding of these concepts [^10].
-
-[^10]: refer to footnote 1, 2, and the detailed online document referenced in the Declarations section
+Applying the approach effectively requires foundational knowledge in two areas.
+First, a solid understanding of causal inference principles.
+Second, the ability to translate the functional and probabilistic aspects of conceptual models into bespoke statistical models.
+A clear example illustrating the importance of causal inference is the recurrent assumption that predictor variables are "relevant" to the research context--interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
+However, this study does not explore the full implications of that assumption for model specification, estimation, and inference.
+Nevertheless, to assist with the effort of engaging with this and other complex causal inference concepts, key references are provided throughout the study to guide applied CJ researchers toward a deeper understanding of these ideas [^10].
+
+[^10]: Also refer to the detailed online document referenced in the Declarations section
 
 Developing the skills to translate conceptual models into bespoke statistical models also presents challenges.
-Although Bayesian inference methods offer a more accessible path for many applied CJ researchers to develop these skills--by reducing the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--they still demand a working knowledge of other technical concepts such as probabilistic programming languages (PPLs), probability distributions, and convergence.
-These requirements can be non-trivial to master.
-To support CJ researchers in developing these skills, this study provides a link to the statistical model code and alternative model specifications in the **Declarations** section of this document [^11].
+Although Bayesian inference methods offer a more accessible path for many applied CJ researchers to develop these skills--by reducing the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--they still demand familiarity with technical concepts such as probabilistic programming languages (PPLs), probability distributions, and convergence diagnostics.
+Thus, to support this skill development, this study provides a link to the statistical model code and alternative model specifications in the **Declarations** section of this document [^11].
 
@@
 
-Even after acquiring the necessary foundational knowledge, CJ researchers still face two additional challenges when applying our approach to their specific research context.
-First, they must verify whether the conceptual model provides a *faithful* [^12] representation of the CJ system under study.
-Second, they need to assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data.
-To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets.
-While these models may prove adequate in some contexts, researchers cannot assume this apriori.
-Instead, they need to adapt these models to their specific contexts, paying close attention to assessment design features and causal assumptions.
-As discussed in @sec-discussion_challenges, our approach facilitates this process by offering a transparent framework for articulating new assumptions and guiding the design of CJ assessments.
+Even after acquiring the necessary foundational knowledge, applying this approach to a specific CJ context involves two additional challenges.
+First, it is essential to verify whether the conceptual model offers a *faithful* [^12] representation of the CJ system under study.
+Second, the statistical translation of the model must be assessed to determine if it can accurately estimate the intended estimands (i.e., parameters) from empirical data.
+To ensure fidelity, the conceptual and statistical models presented here should be treated as starting points, not universal solutions for all CJ designs or datasets.
+While these models may be adequate in some contexts, assuming so without evaluation is unwise.
+Instead, adaptation to the specific context is necessary, with careful attention to assessment design features and causal assumptions.
+As discussed in @sec-discussion_challenges, this approach supports such adaptation by providing a transparent framework for articulating new assumptions and guiding CJ assessment design.
 
@@
 
-Conversely, to evaluate the estimation capabilities of the statistical model, researchers must tackle the challenge of performing identification analysis.
+Conversely, evaluating the estimation capabilities of the statistical model requires addressing the challenge of identification analysis.
 As outlined in @sec-theoretical, *identification analysis* determines whether a statistical model can accurately compute a given estimand (e.g., a parameter) based solely on its (causal) assumptions, independent of random variability [@Schuessler_et_al_2023].
 Identification is crucial because it is a necessary condition for consistency.
-*Consistency* is the property of an estimator (e.g., a statistical model) whose estimates converge to the "true" value of an estimand as data size approaches infinity [@Everitt_et_al_2010].
-Without identification, consistency is impossible--even with infinite, error-free data--and thus, meaningful inference from finite samples cannot be achieved [@Schuessler_et_al_2023].
-Although performing identification analysis through formal derivations may seem like a natural next step, the complexity of the CJ system renders such an approach impractical at the outset.
-Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs.
-Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research.
-Notably, our approach supports both strategies by providing the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
+*Consistency* is the property of an estimator (e.g., a statistical model) whose estimates converge to the "true" value of an estimand as the data size approaches infinity [@Everitt_et_al_2010].
+Without identification, consistency is impossible--even with infinite, error-free data--and meaningful inference from finite samples cannot be achieved [@Schuessler_et_al_2023].
+While formal derivations of identification may seem a natural next step, the complexity of the CJ system makes this approach impractical at the outset.
+Instead, simulation-based methods like power analysis offer a more practical and flexible alternative, enabling examination of estimate consistency without relying on complex mathematical proofs.
+Notably, the approach presented here supports both strategies by providing the probabilistic foundation for formal derivations and the statistical structure needed for simulation-based methods.
 
@@
 
-This paper underscores the need to extend Thurstone's theory to meet the demands of contemporary empirical CJ research.
-It specifically advocates for the development of bespoke CJ models aligned with the underlying data-generating processes of specific CJ assessment designs, as these models have the potential to strengthen the robustness of trait estimates and the clarity of inferences.
-The work also outlines a clear trajectory for advancing CJ research in both theoretical and applied domains.
-Finally, the study also lays the groundwork for broader applications, encouraging researchers across the social sciences to adopt more robust and interpretable methodologies.
+The present study highlights the need to extend Thurstone's theory to address the demands of contemporary empirical CJ research.
+It advocates for developing bespoke CJ models tailored to the specific data-generating processes of different CJ assessment designs.
+These models aim to enhance the robustness of trait estimates and clarity of inferences.
+Moreover, this work outlines a clear path for advancing both theoretical and applied CJ research and lays the foundation for broader adoption of more robust and interpretable methodologies across the social sciences.
 

### 2025-07-03 _tex/paper2.tex
Built site for gh-pages

@@
 However, Thurstone acknowledged that some assumptions of Case V are
-problematic when researchers assess complex traits or heterogeneous
-stimuli \citep{Thurstone_1927a}. Thus, given that contemporary CJ
-applications often involve such traits and stimuli, two key assumptions
-of Case V, and by extension, the BTL model, may not always hold in
-theory or practice. These assumptions are the equal dispersion and zero
+problematic when assessing complex traits or heterogeneous stimuli
+\citep{Thurstone_1927a}. Thus, given that contemporary CJ applications
+often involve such traits and stimuli, two key assumptions of Case V,
+and by extension, the BTL model, may not always hold in theory or
+practice. These assumptions are the equal dispersion and zero
 correlation between stimuli.
@@
 discriminal difference follows any discontinuous line where
-\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
-acknowledge that \emph{misfit statistics} may represent texts with
-different dispersions or outlying observations, the common CJ practice
-of excluding stimuli based on these statistics may unintentionally
-discard valuable information \citep{Miller_2023}, and introduce bias
-into the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
-direction and magnitude of these biases remain unpredictable, as they
-depend on which stimuli researchers exclude from the analysis.
+\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if it is acknowledged
+that \emph{misfit statistics} may represent texts with different
+dispersions or outlying observations, the common CJ practice of
+excluding stimuli based on these statistics may unintentionally discard
+valuable information \citep{Miller_2023}, and introduce bias into the
+trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
+and magnitude of these biases remain unpredictable, as they depend on
+which stimuli are excluded from the analysis.
 
@@
 estimates \citep{Miller_2023}. The direction and magnitude of these
-biases remain unpredictable, as they depend on which judges researchers
-exclude from the analysis
-\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. \emph{Misfit}
-judges are those whose assessments deviate markedly from the shared
-consensus
+biases remain unpredictable, as they depend on which judges are excluded
+from the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
+\emph{Misfit} judges are those whose assessments deviate markedly from
+the shared consensus
 \citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
@@
 
-Researchers in CJ studies typically use the BTL model to measure traits
-and position the compared stimuli along a latent continuum
-\citep{Thurstone_1927b}. The CJ literature shows that research
-frequently relies on point estimates of these traits--typically the BTL
-scores or its transformations--to conduct statistical inference or
-hypothesis testing. For example, researchers have used these scores to
-identify `misfit' judges and stimuli
+In CJ studies, the BTL model is typically used to measure traits and
+position the compared stimuli along a latent continuum
+\citep{Thurstone_1927b}. The CJ literature shows that studies frequently
+relies on point estimates of these traits--typically the BTL scores or
+its transformations--to conduct statistical inference or hypothesis
+testing. For example, researchers have used these scores to identify
+`misfit' judges and stimuli
 \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
@@
 
-The \emph{structural approach} to causal inference provides a formal
-framework for identifying causes and estimating their effects using
-data. The approach uses structural causal models (SCMs) and directed
-acyclic graphs (DAGs)
-\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
-formally and graphically represent the assumed causal structure of a
-system, such as the one found in CJ assessments. Essentially, SCMs and
-DAGs function as \emph{conceptual models} on which identification
-analysis rests. \emph{Identification analysis} determines whether an
-estimator can accurately compute an estimand based solely on its
-(causal) assumptions, regardless of random variability
-\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
-specific quantities researchers aim to determine (i.e., a parameter)
-\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
-functions that transform data into an estimate (e.g., a statistical
-model), while \emph{estimates} are the numerical values approximating
-the estimand \citep{Neal_2020, Everitt_et_al_2010}.
-
-A motivating example that will appear throughout this study clarifies
-these concepts. In this example, researchers aim to answer the question:
-``To what extent do different teaching methods influence students'
-ability to produce high-quality written texts?'' To investigate this, a
-researcher designs a CJ assessment by randomly assigning students
-(individuals) to two groups, each receiving a different teaching method.
-Judges then compare pairs of students' written texts (stimuli) to
-produce a dichotomous outcome reflecting the relative quality of each
-text (trait). Based on this setup, researchers can reformulate the
-research question as the estimand: ``\emph{On average}, is there a
-difference in the ability to produce high-quality written texts between
-the two groups of students?''.
-
-Following standard CJ practices, researchers would typically use
-estimates from the BTL model--or its transformations--to approximate
-this estimand. However, as discussed in Section~\ref{sec-theory-issues},
-Thurstone's Case V and the BTL model exhibit several statistical and
-measurement limitations. These limitations hinder the model's ability to
-identify various estimands relevant to CJ inquiries, including the one
-described in the example.
+The \emph{structural approach} to causal inference
+\citep{Pearl_2009, Pearl_et_al_2016} offers a formal framework for
+identifying causes and estimating their effects using data. The approach
+relies on structural causal models (SCMs) and directed acyclic graphs
+(DAGs) to formally and graphically represent the assumed causal
+structure of a system
+\citep{Morgan_et_al_2014, Gross_et_al_2018, Neal_2020}, such as the one
+found in CJ assessments. In essence, SCMs and DAGs function as
+\emph{conceptual models} on which identification analysis rests
+\citep{Schuessler_et_al_2023}. \emph{Identification analysis} determines
+whether an estimator can accurately compute an estimand based solely on
+its (causal) assumptions, regardless of random variability. Here,
+\emph{estimands} represent the specific quantities researchers aim to
+determine (i.e., a parameter) \citep{Everitt_et_al_2010}.
+\emph{Estimators} denote the methods or functions that transform data
+into an estimate (e.g., a statistical model), while \emph{estimates} are
+the numerical values approximating the estimand
+\citep{Neal_2020, Everitt_et_al_2010}.
+
+As an illustration, consider a researcher seeking to answer the
+question: ``To what extent do different teaching methods influence
+students' ability to produce high-quality written texts?'' To
+investigate this, the researcher designs a CJ assessment by randomly
+assigning students (individuals) to two groups, each exposed to a
+different teaching method. Judges then compare pairs of students'
+written texts (stimuli) to produce a dichotomous outcome reflecting the
+relative quality of each text (trait). Based on this setup, researchers
+can reformulate the research question as the estimand: \emph{On average,
+is there a difference in the ability to produce high-quality written
+texts between the two groups of students?} Following standard CJ
+practices, the researcher would then use point estimates from the BTL
+model, or its transformations, to approximate this estimand. However, as
+discussed in Section~\ref{sec-theory-issue1}, Thurstone's Case V and the
+BTL model exhibit several statistical and measurement limitations. These
+limitations hinder the model's ability to identify and accurately
+estimate a range of estimands relevant to CJ inquiries, including the
+one described in this illustration.
 
@@
 fundamental building blocks \citep{Neal_2020, McElreath_2024}. This
-feature allows researchers to decompose complex structures into
-manageable components, facilitating their analysis. Second, they depict
-causal relationships in a non-parametric way. This flexibility enables
-feasible identification strategies without requiring specification of
-the types of variables, the functional forms relating them, or the
-parameters of those functional forms \citep{Pearl_et_al_2016}.
-
-Thus, this section addresses the issues identified in
-Section~\ref{sec-theory-issues} by extending Thurstone's general form
-using the structural approach to causal inference. Specifically, it
-combines Thurstone's core theoretical principles outlined in
-Section~\ref{sec-thurstone_theory} with key CJ assessment design
+feature allows the decomposition of complex structures into manageable
+components, facilitating their analysis. Second, they depict causal
+relationships in a non-parametric way. This flexibility enables feasible
+identification strategies without requiring specification of the types
+of variables, the functional forms relating them, or the parameters of
+those functional forms \citep{Pearl_et_al_2016}.
+
+Thus, using SCMs and DAGs, this study extends Thurstone's general form
+to address the issues identified in Section~\ref{sec-theory-issues}.
+This extension combines Thurstone's core theoretical principles (see
+Section~\ref{sec-thurstone_theory}) with key CJ assessment design
 features, such as the selection of judges, stimuli, and comparisons.
-Section~\ref{sec-theory-theoretical_P} introduces the
-\emph{conceptual-population model}, which incorporates these theoretical
-principles and assumes an idealized setting where researchers observe a
-\emph{conceptual population} of comparative judgment data--that is, data
-representing all repeated judgments made by every available judge for
-each pair of stimuli produced by each pair of individuals in the
-population. Conversely, Section~\ref{sec-theory-theoretical_SC} presents
-the \emph{sample-comparison model}, which integrates the assessment
-design features and reflects a more realistic setting where researchers
-access only a sample of judges, individuals, stimuli, and comparisons
-from the conceptual population.
-
-\subsection{The conceptual-population
-model}\label{sec-theory-theoretical_P}
-
-In the conceptual-population model, the idealized scenario of a
-\emph{conceptual population} of comparative data enables the integration
-of Thurstone's theoretical principles and provides a foundation for
-proposing innovations aimed at addressing some of the issues discussed
-in Section~\ref{sec-theory-issues}.
+Specifically, Section~\ref{sec-theory-theoretical_P} introduces the
+\emph{conceptual-population model} (henceforth CPM), which incorporates
+these theoretical principles and assumes an idealized setting where
+researchers observe a \emph{conceptual population} of comparative
+judgment data--that is, data representing all repeated judgments made by
+every available judge for each pair of stimuli produced by each pair of
+individuals in the population. Conversely,
+Section~\ref{sec-theory-theoretical_SC} presents the
+\emph{sample-comparison model} (hereafter SCM), which integrates the
+assessment design features and reflects a more realistic setting where
+researchers access only a sample of judges, individuals, stimuli, and
+comparisons from the conceptual population.
+
+\subsection{The conceptual-population model
+(CPM)}\label{sec-theory-theoretical_P}
+
+In the CPM, the idealized scenario of a \emph{conceptual population} of
+comparative data enables the integration of Thurstone's theoretical
+principles and provides a foundation for proposing innovations aimed at
+addressing some of the issues discussed in
+Section~\ref{sec-theory-issues}.
 
@@
 SCM~\ref{fig-cj03_scm} presents the first theoretical principles
-embedded in the conceptual-population model, which evaluates the impact
-of different teaching methods on students' writing ability. This SCM
-outlines the relationship between the conceptual-population outcome
-\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
-\(i\) and \(h\) identify the students who authored the texts (i.e., the
-individuals). The indices \(a\) and \(b\) represent the texts under
-comparison (i.e., the stimuli). The index \(j\) indicates the judge
-conducting the comparison, while the index \(k\) accounts for assessment
-conditions where a judge compares the same pair of stimuli multiple
-times, i.e., a \emph{repeated measures designs}
-\citep[p.~366-376]{Lawson_2015}. Thus, the indexing system supports
+embedded in the CPM evaluating the impact of different teaching methods
+on students' writing ability. This SCM outlines the relationship between
+the conceptual-population outcome \((O^{cp}_{iahbjk})\) and several
+related variables. The subscripts \(i\) and \(h\) identify the students
+who authored the texts (i.e., the individuals). The indices \(a\) and
+\(b\) represent the texts under comparison (i.e., the stimuli). The
+index \(j\) indicates the judge conducting the comparison, while the
+index \(k\) accounts for assessment conditions where a judge compares
+the same pair of stimuli multiple times, i.e., a \emph{repeated measures
+designs} \citep{Lawson_2015}. Thus, the indexing system supports
 comparisons between different texts written by the same student
@@
 
-\caption{\label{fig-cj03}Conceptual-population model, scalar form.}
+\caption{\label{fig-cj03}Conceptual-population model (CPM), scalar
+form.}
 
@@
 
-\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
-form.}
+\caption{\label{fig-cj04}Conceptual-population model (CPM), initial
+vectorized form.}
 
@@
 \citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
-\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
-model integrates two \emph{hierarchical structural components} to
-examine how different \emph{relevant} \footnote{\emph{Relevant
-  variables} are those that satisfy the \emph{backdoor criterion}
-  \citep[pp 37]{Neal_2020}, that is, they belong to a \emph{sufficient
-  adjustment set}
+\citep{Fox_2010, vanderLinden_et_al_2017_I}, the CPM integrates two
+\emph{hierarchical structural components} to examine how different
+\emph{relevant} \footnote{\emph{Relevant variables} are those that
+  satisfy the \emph{backdoor criterion} \citep[pp 37]{Neal_2020}, that
+  is, they belong to a \emph{sufficient adjustment set}
   \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
@@
 observed or latent--affect the primary latent variable of interest
-\citep{Everitt_et_al_2010}. This hierarchical design enables researchers
-to formulate and test hypotheses that account for both the nested
+\citep{Everitt_et_al_2010}. This hierarchical design supports the
+formulation and testing of hypotheses that account for both the nested
 structure of stimuli and the uncertainties inherent in trait estimation
@@
 \(V\), structural equations \(F\), and exogenous variables \(E\) for the
-conceptual-population model. Additionally, the figures demonstrate that
-all exogenous variables are independent of one another, as indicated by
-the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
+CPM. Additionally, the figures demonstrate that all exogenous variables
+are independent of one another, as indicated by the relationships
+\(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
 \(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\) and
@@
 
-Overall, the conceptual-population model extends Thurstone's general
-form by introducing key innovations to address the limitations discussed
-in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
+Overall, the CPM extends Thurstone's general form by introducing key
+innovations to address the limitations discussed in
+Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
 These enhancements include accounting for judges' biases and integrating
@@
 the unrealistic assumption that researchers have access to data from the
-\emph{conceptual population}. Since researchers rarely meet this
-assumption in practice, they must consider a more realistic scenario.
+\emph{conceptual population}. Since this assumption is rarely met in
+practice, the study must consider a more realistic scenario.
 
@@
 
-\caption{\label{fig-cj09}Conceptual-population model, final vectorized
-form.}
+\caption{\label{fig-cj09}Conceptual-population model (CPM), final
+vectorized form.}
 
@@
 
-\subsection{The sample-comparison
-model}\label{sec-theory-theoretical_SC}
-
-The sample-comparison model presents a more realistic scenario than the
-conceptual-population model. First, it explicitly assumes researchers
-work with a data sample consisting of a limited number of repeated
-judgments \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a
-specific number of texts \((n^{s}_{A})\) from a sample of students
-\((n^{s}_{I})\), all drawn from the conceptual population
-(Section~\ref{sec-theory-theoretical_SC1}). Second, the model assumes
-that judges do not perform \emph{all repeated judgments} within the data
-sample (Section~\ref{sec-theory-theoretical_SC2}). Instead, they conduct
-a sufficient number of stimuli comparisons, \(n_{C}\), to ensure an
-accurate estimation of the proportion \(P(B>A)\), as proposed by
-\citet{Thurstone_1927b}.
+\subsection{The sample-comparison model
+(SCM)}\label{sec-theory-theoretical_SC}
+
+The SCM presents a more realistic scenario than the CPM. First, it
+explicitly assumes a data sample consisting of a limited number of
+repeated judgments \((n^{s}_{K})\) from a sample of judges
+\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
+sample of students \((n^{s}_{I})\), all drawn from the conceptual
+population (Section~\ref{sec-theory-theoretical_SC1}). Second, the model
+assumes that judges do not perform \emph{all repeated judgments} within
+the data sample (Section~\ref{sec-theory-theoretical_SC2}). Instead,
+they conduct a sufficient number of stimuli comparisons, \(n_{C}\), to
+ensure an accurate estimation of the proportion \(P(B>A)\), as proposed
+by \citet{Thurstone_1927b}.
 
@@
 example, because the design systematically excludes every third repeated
-judgment, researchers can also express \(S_{JK}\) using
+judgment, \(S_{JK}\) can also be expressed using
 \(n_{K} = n^{s}_{K} = 2\).
@@
 With the definition of \(S\), we incorporate the sample mechanism into
-the conceptual-population model. Following the convention of
-\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
-DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
-\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
-directly access this outcome due to the sampling mechanism. The DAG also
-depicts the \emph{sample design} vector \(S\) as a causal factor
-influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
-encloses \(S\), indicating that it is a conditioned variable. In this
-context, \emph{conditioning} means that researchers restrict their focus
-to the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
-\citep{Neal_2020, McElreath_2020}. In essence, \(S\) is a vector that
-selects \emph{all repeated judgments made by a subset of judges for a
-subset of stimuli produced by the sampled individuals}.
+the CPM. Following the convention of \citet{McElreath_2020} and
+\citet{Deffner_et_al_2022}, DAG~\ref{fig-cj14_dag} represents the
+conceptual-population outcome \(O^{cp}_{R}\) as unobserved, emphasizing
+that this outcome cannot be directly accessed due to the sampling
+mechanism. The DAG also depicts the \emph{sample design} vector \(S\) as
+a causal factor influencing the sample-comparison outcome
+\(O^{sc}_{R}\). A square encloses \(S\), indicating that it is a
+conditioned variable. In this context, \emph{conditioning} means that
+the analysis is restricted to the elements of \(O^{cp}_{R}\) that
+satisfy \(s_{(i,a,h,b,j,k)}=1\) \citep{Neal_2020, McElreath_2020}. In
+essence, \(S\) is a vector that selects \emph{all repeated judgments
+made by a subset of judges for a subset of stimuli produced by the
+sampled individuals}.
 
@@
 The DAG~\ref{fig-cj14_dag} also incorporates the \emph{comparison
-mechanism} \(C\) into the conceptual-population model. It shows the
-sample-comparison outcome \(O^{sc}_{R}\) as unobserved, emphasizing that
-researchers cannot directly access this variable because of the
-comparison mechanism. The DAG further shows \(C\) as a conditioned
-variable (enclosed in a square) that causally influences the observed
-outcome \(O_{R}\). This structure implies that \(C\) determines
-\emph{which repeated judgments judges make for the stimuli produced by
-the individuals}. In essence, \(C\) reflects the assumption that judges
-\emph{do not} perform all possible repeated judgments but instead
-complete a sufficient number, \(n_{C}\), to enable the accurate
-estimation of the proportion \(P(B>A)\) for each stimulus pair
-\citep[p.~267]{Thurstone_1927b}.
+mechanism} \(C\) into the CPM. It shows the sample-comparison outcome
+\(O^{sc}_{R}\) as unobserved, emphasizing that this variable cannot be
+directly accessed because of the comparison mechanism. The DAG further
+shows \(C\) as a conditioned variable (enclosed in a square) that
+causally influences the observed outcome \(O_{R}\). This structure
+implies that \(C\) determines \emph{which repeated judgments judges make
+for the stimuli produced by the individuals}. In essence, \(C\) reflects
+the assumption that judges \emph{do not} perform all possible repeated
+judgments but instead complete a sufficient number, \(n_{C}\), to enable
+the accurate estimation of the proportion \(P(B>A)\) for each stimulus
+pair \citep[p.~267]{Thurstone_1927b}.
 
@@
 Thurstone's general form to address several limitations of the BTL
-model. These extensions account for judge biases (see
+model. These extensions account for judges' biases (see
 Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
@@
 
-Using the structural causal model (SCM) \ref{fig-cj15_scm}, we can
-derive a statistical model that addresses violations of the equal
-dispersion assumption (see Section~\ref{sec-theory-issue1a}). This
-derivation is possible because a fully specified SCM encodes functional
-and probabilistic information, which we can replace with suitable
-functions and probabilistic assumptions \citep{Pearl_et_al_2016}.
-Specifically, SCM~\ref{fig-cj15_scm} allows us to express the joint
-distribution of our complex CJ system as a product of simpler
-conditional probability distributions (CPDs)\footnote{This re-expression
-  is possible because the \emph{chain rule} of probability and the
-  \emph{Bayesian Network Factorization (BNF)} property. For further
-  details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as
-shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
-such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
-as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
-CPD of \(Y\) given \(X\).
-\begin{equation}\phantomsection\label{eq-mat51}{
+Using the SCM~\ref{fig-cj15_scm}, we can derive a statistical model that
+addresses violations of the equal dispersion assumption (see
+Section~\ref{sec-theory-issue1a}). This derivation is possible because a
+fully specified SCM encodes functional and probabilistic information,
+which we can replace with suitable functions and probabilistic
+assumptions \citep{Pearl_et_al_2016}. Specifically,
+SCM~\ref{fig-cj15_scm} allows us to express the joint distribution of
+our complex CJ system as a product of simpler conditional probability
+distributions (CPDs)\footnote{This re-expression is possible because the
+  \emph{chain rule} of probability and the \emph{Bayesian Network
+  Factorization (BNF)} property. For further details, see
+  \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as shown in Equation
+(\ref{eq-mat51}). For clarity, we treat expressions such as
+\(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\) as
+equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the CPD
+of \(Y\) given \(X\). \begin{equation}\phantomsection\label{eq-mat51}{
 \begin{aligned}
@@
 parameters than the number of available pairwise comparisons
-\citep{Thurstone_1927b}. To address this issue and facilitate the
+\citep{Thurstone_1927b}. To address this issue and to facilitate the
 theory's practical applicability, he formulated five cases, each
@@
 research, mainly due to the widespread adoption of the BTL model. The
-BTL model mirrors the core assumptions of Case V--namely, equal
+BTL model incorporates the core assumptions of Case V--namely, equal
 discriminal dispersions and zero correlation among stimuli' discriminal
@@
 To address the limitations of Thurstone's Case V and the BTL model, this
-study extends Thurstone's general form using a systematic, integrated
-approach that combines causal and Bayesian inference methods. The
-approach begins with the development of a conceptual model, represented
-as a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG)
+study extended Thurstone's general form using a systematic, integrated
+approach that combined causal and Bayesian inference methods. The
+approach began with the development of a conceptual model, formalized as
+a Structural Causal Model (SCM) and represented graphically by a
+Directed Acyclic Graph (DAG)
 \citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. This
-model integrates Thurstone's core theoretical principles, such as the
+model integrated Thurstone's core theoretical principles, such as the
 discriminal processes of stimuli, alongside key CJ assessment design
 features, including judges' bias, sampling procedures, and comparison
-mechanisms, to disentangle the causal processes underlying the CJ
-system.
-
-The approach then translates the SCM into a bespoke statistical model
-that allows researchers to analyze CJ data when violations to the
-assumptions of equal dispersion and zero correlation occur, and when
-statistical inference is necessary. In particular, this model accounts
-for judge biases, captures the hierarchical structure of stimuli,
-incorporates measurement error into the hypothesis testing process, and
-accommodates heterogeneity in discriminal dispersions. By addressing all
-these issues, these methodological innovations have the potential to
-enhance the reliability and validity of trait measurement in CJ
-\citep{Perron_et_al_2015}, while also improving the accuracy of
-statistical inferences.
-
-Beyond these potential benefits, the approach offers two additional
-advantages. First, it clarifies the roles and interactions of all actors
-and processes involved in CJ assessments. Second, it shifts the analytic
-paradigm from passively accepting the assumptions of Case V and the BTL
-model to actively testing their fit with observed data. Together, these
-advantages establish a principled framework for evaluating best
-practices in designing CJ assessments, one that better aligns with the
-demands of contemporary CJ contexts \citep{Kelly_et_al_2022}, offering
-new insights into existing research and opening promising avenues for
-future inquiry.
+mechanisms. Together, these components allowed the causal processes
+underlying the CJ system to be disentangled.
+
+The approach then translated the SCM into a bespoke statistical model
+that enabled the analysis of CJ data in cases where the assumptions of
+equal dispersion and zero correlation were violated, and where
+statistical inference was required. In particular, the model accounted
+for judge biases, captured the hierarchical structure of stimuli,
+incorporated measurement error into the hypothesis testing process, and
+accommodated heterogeneity in discriminal dispersions. By addressing
+these issues, the methodological innovations aimed to enhance the
+reliability and validity of trait measurement in CJ, while also
+improving the accuracy of statistical inferences.
+
+Beyond these potential benefits, the approach offered two additional
+advantages. First, it clarified the roles and interactions of all actors
+and processes involved in CJ assessments. Second, it shifted the
+analytic paradigm from passively accepting the assumptions of Case V and
+the BTL model to actively testing their fit with observed data.
+Together, these advantages established a principled framework for
+evaluating best practices in designing CJ assessments--one that better
+aligns with the demands of contemporary CJ contexts
+\citep{Kelly_et_al_2022}--providing new insights into existing research
+and opening promising avenues for future inquiry.
 
@@
 likely stems from a more fundamental lack of conceptual clarity about
-how these mechanisms function within the system, this study integrates
-these mechanisms into the conceptual model of CJ.
+how these mechanisms function within the system, the present study
+integrated these mechanisms into the conceptual model of CJ.
 
@@
 offers a new perspective on how these mechanisms shape the CJ process.
-Specifically, it clarifies their role as sources of missing data in CJ's
-data-generating process, that is, as mechanisms that determine which
+Specifically, it clarified their role as sources of missing data in CJ's
+data-generating process--that is, as mechanisms that determine which
 observations are missing from the final data sample. This new
-perspective encourages the application of Little and Rubin's principled
-missing data framework \citeyearpar{Little_et_al_2020}, allowing a more
+perspective invites the application of Little and Rubin's principled
+missing data framework \citep{Little_et_al_2020}, allowing a more
 rigorous evaluation of existing claims about these missing data
@@
 
-This study circumvents the need to apply this missing data framework by
+This study circumvented the need to apply the missing data framework by
 deliberately structuring the sampling and comparison mechanisms to be
 independent of any observed or unobserved variables, including the
-outcome. In other words, these mechanisms are designed to produce data
-that are \emph{missing completely at random} (MCAR)
-\citep{Little_et_al_2020}. This design offers one key advantage: it
-generates simple random samples that satisfy the condition of
-\emph{ignorability}, allowing researchers to legitimately \emph{ignore}
-missing data during analysis without introducing bias
+outcome. In other words, these mechanisms were intentionally designed to
+yield data that are \emph{missing completely at random} (MCAR)
+\citep{Little_et_al_2020}. This design choice offered one key advantage:
+it generated simple random samples that satisfied the condition of
+\emph{ignorability}, thereby allowing researchers to legitimately
+\emph{ignore} missing data during analysis without introducing bias
 \citep{Everitt_et_al_2010, Kohler_et_al_2019, Neal_2020}.
@@
 intricate forms of missingness such as \emph{missing at random} (MAR) or
-\emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
-prominent example is the previously discussed ACJ design, where prior
-judgment outcomes inform the selection of stimulus pairs for subsequent
-comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
-pair selection procedure suggests that ACJ's comparison mechanism is
-outcome-dependent, potentially classifying the method as a generator of
-MNAR data. If this classification holds, the mixed findings on ACJ's
-capabilities become more comprehensible: some studies find that the
-method improves trait reliability
-\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
-contend that it artificially inflates these gains
+\emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. One
+prominent example is ACJ designs, where prior judgment outcomes inform
+the selection of stimulus pairs for subsequent comparisons
+\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This pair selection
+procedure suggests that ACJ's comparison mechanism is outcome-dependent,
+potentially classifying the method as a generator of MNAR data. If this
+classification holds, ACJ might violate the condition of ignorability,
+making it an unsuitable pair selection procedure for reliable and valid
+trait estimation and inference. Moreover, under this interpretation, the
+mixed findings on ACJ's effectiveness become more comprehensible: while
+some studies find that the method improves trait reliability
+\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
+that that these gains may be artificially inflated
 \citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
@@
 design would benefit from explicitly defining its assumptions--a
-practice supported by our approach. This clarity enables researchers to
-evaluate how the sampling and comparison mechanisms affect trait
-estimation and statistical inference within each design. Such
+practice supported by this study's approach. This clarity enables a
+thorough evaluation of how the sampling and comparison mechanisms affect
+trait estimation and statistical inference in each design. Such
 assessments are particularly relevant given the common misconception in
@@
 
-\subsubsection{The effects of judges' bias on the reliability of
-traits}\label{sec-discussion_RA2}
+\subsubsection{The effects of judges' bias on the reliability of trait
+estimates}\label{sec-discussion_RA2}
 
@@
 \citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}
-and that these influences may not always cancel each other out, few
-studies in the CJ literature provide empirical evidence for judges'
-biases
+and that these influences may not always cancel each other out,
+empirical evidence of judges' biases remains scarce in the CJ literature
 \citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
-This gap likely persists not due to a lack of interest or research but
-because researchers often rely on ad-hoc detection methods, such as
-`misfit' statistics, that may not be well-suited for the task
-\citep{Kelly_et_al_2022}. To address this limitation, this study treats
-judges' biases as an integral component of the CJ system from the
-outset. This approach offers one key advantage: it provides a more
-accurate representation of the data-generating process behind pairwise
-comparisons, one that acknowledges that the discriminal processes of
-stimuli become an observable outcome only through judges' perceptions,
-which may exhibit bias.
+This gap likely persists not from a lack of interest or research but
+because the identification of such biases often depends on ad-hoc
+detection methods, like `misfit' statistics, that may not be well-suited
+for the task \citep{Kelly_et_al_2022}. To overcome this limitation, the
+present study treated judges' biases as an integral component of the CJ
+system from the outset. This approach offers one key advantage: it
+provides a more accurate representation of the data-generating process
+behind pairwise comparisons, one that acknowledges that the discriminal
+processes of stimuli becomes an observable outcome only through judges'
+perceptions, which may exhibit bias.
 
@@
 paves the way for investigating several relevant research questions. One
-key question is whether researchers can validly analyze CJ data under
-the assumption of ``sample-free'' trait calibration, specifically under
-the hypothesis that judges exhibit no systematic bias. This question is
-particularly relevant because many researchers still regard
-``sample-freeness'' as an inherent property of the BTL model
-\citep{Bramley_2008, Andrich_1978} despite growing evidence of
-persistent biases. Another critical question is whether training or
-expertise can help judges avoid focusing on irrelevant stimulus
-features, such as handwriting, over more central criteria like
-argumentative quality in writing assessments \citep{Kelly_et_al_2022}.
-Exploring these questions may also provide insights into what it truly
-means to be an ``expert'' within the CJ context
-\citep{Kelly_et_al_2022}.
+key question is whether CJ data can be validly analyzed under the
+assumption of ``sample-free'' trait calibration, specifically under the
+hypothesis that judges exhibit no systematic bias. This question is
+particularly relevant because ``sample-freeness'' is still regarded as
+an inherent property of the BTL model \citep{Bramley_2008, Andrich_1978}
+despite growing evidence of persistent biases. Another critical question
+is whether training or expertise can help judges avoid focusing on
+irrelevant stimulus features, such as handwriting, over more central
+criteria like argumentative quality in writing assessments
+\citep{Kelly_et_al_2022}. Exploring these questions may also provide
+insights into what it truly means to be an ``expert'' within the CJ
+context \citep{Kelly_et_al_2022}.
 
 Moreover, since judges rely on these stimulus-related factors when
-evaluating complex, multidimensional traits
+evaluating complex, multidimensional traits (e.g., structure, style)
 \citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022},
@@
 studies support this view
-\citep{Bartholomew_et_al_2020a, McMahon_et_al_2015}. Thus, building on
-the discussion in Section~\ref{sec-discussion_RA1}, researchers could
-further explore how judges' selection influences the formation of a
-``shared consensus'' and whether these attributes introduce systematic
-biases or distortions in the observed trait distribution
-\citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
-undermine the assumption of ``sample-freeness,'' it becomes essential to
-explore strategies for mitigating their influence and to determine how
-many judges (and how many judgments per judge) are needed to produce
-reliable trait estimates under these conditions. In addition, it is
-worth considering whether \emph{repeated measures designs}, in which
-judges evaluate the same stimulus pairs multiple times
-\citep{Lawson_2015}, can improve judgment consistency and accuracy. As
-anticipated, the approach presented in this study provides the necessary
-structure to investigate rigorously these questions.
+\citep[e.g.,][]{Bartholomew_et_al_2020a, McMahon_et_al_2015}. Thus,
+building on the discussion in Section~\ref{sec-discussion_RA1}, further
+exploration is warranted into how judge selection influences the
+formation of a ``shared consensus'' and whether certain judge attributes
+introduce systematic biases or distortions in the stimuli trait
+distribution \citep{Deffner_et_al_2022}. Furthermore, if such attributes
+do in fact compromise the assumption of ``sample-freeness,'' it becomes
+essential to consider strategies for mitigating their effects and to
+determine how many judges (and how many judgments per judge) are
+required to produce reliable trait estimates under these conditions.
+Additionally, it is worth examining whether \emph{repeated measures
+designs}, in which judges evaluate the same stimulus pairs multiple
+times \citep{Lawson_2015}, can improve judgment consistency and
+accuracy. As anticipated, the approach presented in this study provides
+a structured foundation to rigorously investigate these questions.
 
@@
 relate to Thurstonian theory. In particular, they have not identified
-which elements of Thurstone's theory might account for the occurrence of
-misfits. This disconnect likely stems from the fact that CJ researchers
-derive misfit statistics from residual analysis and outlier detection
+which elements of Thurstone's theory account for the occurrence of
+misfits. This disconnect likely stems from the fact that misfit
+statistics are derived from residual analysis and outlier detection
 methods rather than from Thurstonian principles. Specifically,
@@
 in which they occur \citep{Grubbs_1969}. But this resemblance extends
-beyond the definitions themselves, as CJ researchers often identify
-misfits using conventional outlier detection procedures, such as
-transforming BTL model residuals into diagnostic statistics and
-comparing them against predefined thresholds
+beyond the definitions themselves, as misfits are often identified using
+conventional outlier detection procedures, such as transforming BTL
+model residuals into diagnostic statistics and comparing them against
+predefined thresholds
 \citep{Pollitt_2012a, Pollitt_2012b, Wu_et_al_2022}.
 
-Nevertheless, it is not the classification of misfits as outliers that
+Nevertheless, is not the classification of misfits as outliers that
 raises concerns for trait measurement and inference. Rather, the concern
-lies in the prevailing CJ practice of classify them through ad-hoc
-detection procedures and then excluding them from analysis
-\citep{Pollitt_2012a, Pollitt_2012b}, often without providing empirical
-evidence for the various hypotheses proposed to explain their
-occurrence. In this regard, it is essential to recognize that outliers
-can only be defined relative to a specific model \citep{McElreath_2020}.
-As such, detection procedures based on models like the BTL, which relies
-on strong assumptions, should be approached with caution, as they may
-not be well-suited for the task \citep{Kelly_et_al_2022}. In particular,
-because the BTL model may not always accurately reflect the underlying
-data-generating process of the CJ system, misfit classification based on
-this model risks misidentifying observations. Furthermore, the exclusion
-of these observations carries additional risks. The statistical
-literature cautions that removing outliers can discard valuable
-information \citep{Miller_2023} and introduce bias into trait estimates.
-The direction and magnitude of this bias are often unpredictable, as
-they depend on which observations researchers exclude from the analysis
-\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. Finally, precisely
-because of its rigid assumptions, the BTL model lacks the flexibility to
-adequately test many of the hypotheses proposed to explain the presence
-of misfits.
+lies in the prevailing CJ practice of identifying these observations
+through ad-hoc procedures and excluding them from analysis
+\citep{Pollitt_2012a, Pollitt_2012b}, often without empirical support
+for the various hypotheses suggested to explain their occurrence. It is
+essential to recognize that outliers are defined relative to a specific
+model \citep{McElreath_2020}. Thus, detection procedures based on models
+like the BTL--which rests on strong assumptions--should be approached
+with caution, as they may not be appropriate for this purpose
+\citep{Kelly_et_al_2022}. If the BTL model does not accurately represent
+the actual data-generating process of the CJ system, the method may miss
+classify valid observations as misfits. On top this, excluding these
+observations carries additional risks. The statistical literature
+cautions that removing outliers can discard valuable information
+\citep{Miller_2023} and introduce bias into trait estimates. The
+direction and magnitude of this bias are often unpredictable, as they
+depend on which observations are excluded from the analysis
+\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. Finally, even when
+the model aligns well with the data, its rigid assumptions limit the
+model's ability to adequately test many of the hypotheses proposed to
+explain misfit behavior.
 
 In contrast, the approach presented in this study provides a rigorous
-framework that enables the examination of several relevant hypotheses.
-For instance, researchers could investigate whether misfit judges are
-those who exhibit (an outlying degree of) systematic bias or greater
-variability in their judgments compared to their peers. Similarly,
-researchers might explore whether misfit stimuli exhibit more variable
+framework for testing several relevant hypotheses. For instance, it
+allows to investigate whether misfit judges are those who exhibit (an
+outlying degree of) systematic bias or greater variability in their
+judgments compared to their peers. Similarly, the approach makes
+possible to explore whether misfit stimuli exhibit more variable
 discriminal processes relative to other stimuli or if they are genuinely
-outlying cases. Moreover, since outliers are not necessarily ``bad
-data'' \citep{McElreath_2020}, our approach also offers a principled
-alternative that retains misfits in the analysis without compromising
-trait estimation or inference. This alternative involves adapting the
-proposed model into robust measurement models \citep{McElreath_2020}, a
-broad class of procedures designed to reduce the sensitivity of
-parameter estimates to mild or moderate departures from model
-assumptions \citep{Everitt_et_al_2010}. This alternative also relates to
-a broader discussion in social science, which is: that when researchers
-face low predictive capacity from their models, they should not only
-search for ``new'' variables or alternative procedures but also consider
-employing more sophisticated measurement models
-\citep{Wainer_et_al_1978}.
+outlying cases. Moreover, since outliers are not inherently ``bad data''
+\citep{McElreath_2020}, the present approach offers a principled
+alternative to exclusion, one that retains misfits in the analysis
+without compromising trait estimation or inference. This is achieved by
+adapting the proposed model into robust measurement models
+\citep{McElreath_2020}, a broad class of procedures designed to reduce
+the sensitivity of parameter estimates to mild or moderate departures
+from model assumptions \citep{Everitt_et_al_2010}. This strategy also
+speaks to a broader concern in the social sciences: when predictive
+power is low, the solution may not lie solely in seeking new variables
+or procedures, but also in adopting more sophisticated measurement
+models \citep{Wainer_et_al_1978}.
 
@@
 
-The process of deriving conclusions from observed data always requires
-assumptions, whether the data are observational or experimental
+Drawing conclusions from observed data always requires assumptions,
+whether the data are observational or experimental
 \citep{Kohler_et_al_2019, Deffner_et_al_2022}. The proposed approach is
-not an exception to this fundamental principle. As with all approaches
+not an exception to this fundamental principle. Like all approaches
 grounded on causal inference, it relies on expert knowledge and
 assumptions about the variables' causal structure that are often
-untestable at the outset \citep{Hernan_et_al_2025}. As such, the
-approach does not seek to yield automatic answers when applied to a
-given CJ dataset. Instead, it aims to encourage researchers to formulate
-precise questions and to make their assumptions explicit, fostering a
-generalizable understanding of the CJ system under study
+untestable at the outset \citep{Hernan_et_al_2025}. However, its purpose
+is not to deliver automatic answers when applied to a given CJ dataset.
+Instead, it encourages the formulation of precise questions and the
+explicit articulation of assumptions, fostering a generalizable
+understanding of the CJ system under study
 \citep{Rohrer_et_al_2022, Deffner_et_al_2022, Sterner_et_al_2024}. This
-clarity is critical because a model's ability to produce accurate
-estimates and valid inferences depends heavily on how well the data and
-inferential goals align with its underlying assumptions
-\citep{Kohler_et_al_2019}. Although this alignment remains empirically
-untested for the proposed models, the theory-driven nature of our
-approach provides a solid foundation for future empirical evaluation of
-its causal assumptions \citep{Deffner_et_al_2022}, which are well
-supported by both theory and existing evidence.
-
-Moreover, this theoretical commitment to causal inference also
-introduces several practical challenges that applied CJ researchers must
-navigate. These fall into two main categories: first, acquiring the
-foundational knowledge necessary to apply the approach effectively; and
-second, dedicating greater attention to both conceptual and statistical
+clarity is crucial because the accuracy of estimates and validity of
+inferences depend on how well the data and inferential goals align with
+a model's assumptions \citep{Kohler_et_al_2019}. Although this alignment
+remains to be empirically tested for the proposed models, the
+theory-driven nature of this approach provides a solid foundation for
+future empirical evaluation of its causal assumptions
+\citep{Deffner_et_al_2022}, grounded in both established theory and
+existing evidence.
+
+The theoretical commitment to causal inference also introduces several
+practical challenges that applied CJ researchers must navigate. These
+fall into two main categories: first, acquiring the foundational
+knowledge necessary to apply the approach effectively; and second,
+dedicating greater attention to both conceptual and statistical
 modeling.
@@
 
-To apply the approach effectively, CJ researchers require foundational
-knowledge in two areas. First, they need a solid understanding of causal
-inference principles. Second, they must learn to translate the
-functional and probabilistic content of conceptual models into bespoke
-statistical models. One example that illustrates the importance of a
-sound knowledge of causal inference is the recurrent assumption that
-predictor variables are ``relevant'' to the research
+Applying the approach effectively requires foundational knowledge in two
+areas. First, a solid understanding of causal inference principles.
+Second, the ability to translate the functional and probabilistic
+aspects of conceptual models into bespoke statistical models. A clear
+example illustrating the importance of causal inference is the recurrent
+assumption that predictor variables are ``relevant'' to the research
 context--interpreting this relevance as their inclusion in a
 \emph{sufficient adjustment set}
-\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
-not fully explore what this assumption entails or its implications for
-model specification, estimation, and inference. As a result, CJ
-researchers must deeply engage with these and other complex ideas,
-including how SCMs encode functional and probabilistic information. To
-assist this effort, this study includes key references throughout the
-study to guide applied CJ researchers toward a deeper understanding of
-these concepts \footnote{refer to footnote 1, 2, and the detailed online
-  document referenced in the Declarations section}.
+\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, this
+study does not explore the full implications of that assumption for
+model specification, estimation, and inference. Nevertheless, to assist
+with the effort of engaging with this and other complex causal inference
+concepts, key references are provided throughout the study to guide
+applied CJ researchers toward a deeper understanding of these ideas
+\footnote{Also refer to the detailed online document referenced in the
+  Declarations section}.
 
@@
 areas like optimization theory, which frequentist methods often
-require--they still demand a working knowledge of other technical
-concepts such as probabilistic programming languages (PPLs), probability
-distributions, and convergence. These requirements can be non-trivial to
-master. To support CJ researchers in developing these skills, this study
-provides a link to the statistical model code and alternative model
-specifications in the \textbf{Declarations} section of this document
-\footnote{Seminal texts on Bayesian inference methods, such as
+require--they still demand familiarity with technical concepts such as
+probabilistic programming languages (PPLs), probability distributions,
+and convergence diagnostics. Thus, to support this skill development,
+this study provides a link to the statistical model code and alternative
+model specifications in the \textbf{Declarations} section of this
+document \footnote{Seminal texts on Bayesian inference methods, such as
   \citet{Gelman_et_al_2014} and \citet{McElreath_2020}, offer valuable
@@
 
-Even after acquiring the necessary foundational knowledge, CJ
-researchers still face two additional challenges when applying our
-approach to their specific research context. First, they must verify
-whether the conceptual model provides a \emph{faithful} \footnote{Avoid
-  confusing this term with the \emph{faithfulness assumption}
-  \citep{Neal_2020, Hernan_et_al_2025} described in the causal inference
-  literature. For further details, see footnote 1.} representation of
-the CJ system under study. Second, they need to assess whether the
-statistical translation of the model can accurately estimate the
-intended estimands (i.e., parameters) from empirical data. To ensure a
-faithful representation of the CJ system, we encourage researchers to
-treat our conceptual and statistical models as starting points rather
-than universal solutions for all CJ designs or datasets. While these
-models may prove adequate in some contexts, researchers cannot assume
-this apriori. Instead, they need to adapt these models to their specific
-contexts, paying close attention to assessment design features and
-causal assumptions. As discussed in
-Section~\ref{sec-discussion_challenges}, our approach facilitates this
-process by offering a transparent framework for articulating new
-assumptions and guiding the design of CJ assessments.
-
-Conversely, to evaluate the estimation capabilities of the statistical
-model, researchers must tackle the challenge of performing
-identification analysis. As outlined in Section~\ref{sec-theoretical},
-\emph{identification analysis} determines whether a statistical model
-can accurately compute a given estimand (e.g., a parameter) based solely
-on its (causal) assumptions, independent of random variability
+Even after acquiring the necessary foundational knowledge, applying this
+approach to a specific CJ context involves two additional challenges.
+First, it is essential to verify whether the conceptual model offers a
+\emph{faithful} \footnote{Avoid confusing this term with the
+  \emph{faithfulness assumption} \citep{Neal_2020, Hernan_et_al_2025}
+  described in the causal inference literature. For further details, see
+  footnote 1.} representation of the CJ system under study. Second, the
+statistical translation of the model must be assessed to determine if it
+can accurately estimate the intended estimands (i.e., parameters) from
+empirical data. To ensure fidelity, the conceptual and statistical
+models presented here should be treated as starting points, not
+universal solutions for all CJ designs or datasets. While these models
+may be adequate in some contexts, assuming so without evaluation is
+unwise. Instead, adaptation to the specific context is necessary, with
+careful attention to assessment design features and causal assumptions.
+As discussed in Section~\ref{sec-discussion_challenges}, this approach
+supports such adaptation by providing a transparent framework for
+articulating new assumptions and guiding CJ assessment design.
+
+Conversely, evaluating the estimation capabilities of the statistical
+model requires addressing the challenge of identification analysis. As
+outlined in Section~\ref{sec-theoretical}, \emph{identification
+analysis} determines whether a statistical model can accurately compute
+a given estimand (e.g., a parameter) based solely on its (causal)
+assumptions, independent of random variability
 \citep{Schuessler_et_al_2023}. Identification is crucial because it is a
@@
 of an estimator (e.g., a statistical model) whose estimates converge to
-the ``true'' value of an estimand as data size approaches infinity
+the ``true'' value of an estimand as the data size approaches infinity
 \citep{Everitt_et_al_2010}. Without identification, consistency is
-impossible--even with infinite, error-free data--and thus, meaningful
+impossible--even with infinite, error-free data--and meaningful
 inference from finite samples cannot be achieved
-\citep{Schuessler_et_al_2023}. Although performing identification
-analysis through formal derivations may seem like a natural next step,
-the complexity of the CJ system renders such an approach impractical at
-the outset. Instead, simulation-based methods, such as power analysis,
-provide a more practical and flexible alternative, enabling researchers
-to examine the consistency of estimates without relying on cumbersome
-mathematical proofs. Nonetheless, the development of formal
-identification proofs remains a significant goal for future CJ research.
-Notably, our approach supports both strategies by providing the
-probabilistic foundation for formal derivations and the necessary
-statistical structure for simulation-based methods.
+\citep{Schuessler_et_al_2023}. While formal derivations of
+identification may seem a natural next step, the complexity of the CJ
+system makes this approach impractical at the outset. Instead,
+simulation-based methods like power analysis offer a more practical and
+flexible alternative, enabling examination of estimate consistency
+without relying on complex mathematical proofs. Notably, the approach
+presented here supports both strategies by providing the probabilistic
+foundation for formal derivations and the statistical structure needed
+for simulation-based methods.
 
@@
 
-This paper underscores the need to extend Thurstone's theory to meet the
-demands of contemporary empirical CJ research. It specifically advocates
-for the development of bespoke CJ models aligned with the underlying
-data-generating processes of specific CJ assessment designs, as these
-models have the potential to strengthen the robustness of trait
-estimates and the clarity of inferences. The work also outlines a clear
-trajectory for advancing CJ research in both theoretical and applied
-domains. Finally, the study also lays the groundwork for broader
-applications, encouraging researchers across the social sciences to
-adopt more robust and interpretable methodologies.
+The present study highlights the need to extend Thurstone's theory to
+address the demands of contemporary empirical CJ research. It advocates
+for developing bespoke CJ models tailored to the specific
+data-generating processes of different CJ assessment designs. These
+models aim to enhance the robustness of trait estimates and clarity of
+inferences. Moreover, this work outlines a clear path for advancing both
+theoretical and applied CJ research and lays the foundation for broader
+adoption of more robust and interpretable methodologies across the
+social sciences.
 

### 2025-07-05 paper2.qmd
final review and editing

@@
 abstract: |
-  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two central issues in the CJ literature.
-First, it critiques the widespread reliance on Thurstone's Case V assumptions and, by extension, the Bradley-Terry-Luce (BTL) model in the analysis of CJ data.
-Specifically, this study argues that while the assumptions of equal discriminal dispersions and zero correlation between stimuli simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates.
-Second, the study highlights the apparent disconnect between trait measurement and hypothesis testing procedures in CJ applications.
-It contends that while separating these procedures simplifies the analysis of CJ data, this practice can also undermine the resulting statistical inferences.
-
-  To address these issues, this study aims to extend Thurstone's general form using a systematic and integrated approach based on causal and Bayesian inference methods.
-This extension integrates Thurstone's core theoretical principles alongside key CJ assessment design features.
-It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.
-
-  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, particularly stressing the need for bespoke CJ models tailored to the experiments and data assumptions.
-It also lays the foundation for broader applications, encouraging researchers across the social sciences to adopt more robust and interpretable methodologies.
+  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two key issues in the CJ literature.
+First, it critiques the reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce (BTL) model in the analysis of CJ data.
+Specifically, it argues that while the assumptions of equal discriminal dispersions and zero correlation simplify the trait measurement model, they may fail to capture the complexity of some traits or stimuli, potentially leading to unreliable and inaccurate trait estimates.
+Second, the study highlights the disconnect between trait measurement and statistical inference in CJ applications.
+It contends that while separating these processes simplifies the analysis, this practice can also undermine the resulting statistical inferences.
+  To address these issues, this study extends Thurstone's general form using a systematic, integrated approach based on causal and Bayesian inference methods.
+This extension integrates CJ's core theoretical principles alongside key assessment design features.
+It then translates these elements into a statistical model for analyzing dichotomous CJ data.
+  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, stressing the need for CJ models tailored to the experiments and data assumptions.
+It also encourages researchers across the social sciences to adopt more robust and interpretable methodologies.
 date: last-modified
@@
 
-![Hypothetical discriminal processes for two written texts](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}
-
-![Hypothetical discriminal difference of two written texts](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}
+![Discriminal processes](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}
+
+![Discriminal difference](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}
 
@@
 Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b].
-For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison.
+For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases).
 However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an "insufficient" data set to estimate the required parameters.
@@
 <!-- 3. But are our specific problems? -->
-As the table suggest, Thurstone developed Case V with an emphasis on statistical simplicity, but this simplicity comes at the expense of accurate and precise trait measurement and practical guidance for inference.
+As the table suggests, Thurstone developed Case V with an emphasis on statistical simplicity, but this simplicity comes at the expense of accurate and precise trait measurement and practical guidance for inference.
 Specifically, Thurstone cautioned that Case V use "should not be made without (an) experimental test" [@Thurstone_1927b, p.
@@
 Specifically, @sec-theory-theoretical_P introduces the *conceptual-population model* (henceforth CPM), which incorporates these theoretical principles and assumes an idealized setting where researchers observe a *conceptual population* of comparative judgment data--that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population.
-Conversely, @sec-theory-theoretical_SC presents the *sample-comparison model* (hereafter SCM), which integrates the assessment design features and reflects a more realistic setting where researchers access only a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
+Conversely, @sec-theory-theoretical_SC presents the *sample-comparison model* (hereafter CSM), which integrates the assessment design features and reflects a more realistic setting where researchers access only a sample of judges, individuals, stimuli, and comparisons from the conceptual population.
 
@@
 Thus, we define the matrix as in Equation ([-@eq-mat11]).
-
 $$
@@
 The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$.
-Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns
+Here, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ independent columns (variables), while $e_{JK}$ and $B_{JK}$ are latent latent matrices with $n_{J}$ rows and $n_{K}$ columns.
 
@@
 Additionally, the figures demonstrate that all exogenous variables are independent of one another, as indicated by the relationships $e_{IA} \dsep \{ e_{I}, e_{JK}, e_{J} \}$, $e_{I} \dsep \{ e_{JK}, e_{J} \}$ and $e_{JK} \dsep e_{J}$ and the absence of connecting arrows.
-
-<!-- 3. segway for next section -->
-Overall, the CPM extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
-These enhancements include accounting for judges' biases and integrating hierarchical structural components.
-Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
-Since this assumption is rarely met in practice, the study must consider a more realistic scenario.
 
@@
 
-## The sample-comparison model (SCM) {#sec-theory-theoretical_SC}
+<!-- 3. segway for next section -->
+Overall, the CPM extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2.
+These enhancements include accounting for judges' biases and integrating hierarchical structural components.
+Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to data from the *conceptual population*.
+Since this assumption is rarely met in practice, the study must consider a more realistic scenario.
+
+## The sample-comparison model (CSM) {#sec-theory-theoretical_SC}
 
@@
 <!-- 1. What is the more realistic scenario: the sample mechanism -->
-The SCM presents a more realistic scenario than the CPM.
+The CSM presents a more realistic scenario than the CPM.
 First, it explicitly assumes a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population (@sec-theory-theoretical_SC1).
@@
 <!-- 1. start by defining the sampling vector and matrices  -->
-To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the *data sampling process* using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:
+To incorporate the sampling mechanism and facilitate the interpretation of the CSM, we first define the *data sampling process* using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:
 $$
@@
 
-Sample-comparison model, final vectorized form
+Sample-comparison model (CSM), final vectorized form.
 :::
@@
 <!-- 1. start by defining the comparison vector  -->
-As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model.
+As in the previous section, we begin defining the *comparison mechanism* using the binary vector variable $C$ to facilitate the interpretation of the CSM.
 Equation ([-@eq-mat27]) shows that $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element $c_{(i,a,h,b,j,k)}$ being a binary value indicating the presence or absence of data rows in $R$, a definition similar to that of $s_{(i,a,h,b,j,k)}$ in Equation ([-@eq-mat26]).
@@
 Notably, [DAG -@fig-cj14_dag] also shows that $C$ is independent of all other variables in the model.
-This independence implies that the conceptual model represented by the DAG applies exclusively to Random Allocation Comparative Designs [@Bramley_2015], or Incomplete Block Designs [@Lawson_2015], where every repeated judgment has an equal probability of being included in the sample.
+This independence implies that the conceptual model represented by the DAG applies exclusively to *Random Allocation Comparative Designs* [@Bramley_2015], or *Incomplete Block Designs* [@Lawson_2015], where every repeated judgment has an equal probability of being included in the sample.
 
@@
 <!-- 3. preparing to transform the conceptual model into an statistical one -->
-Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the sample-comparison model in @fig-cj14 into the equivalent form shown in @fig-cj15.
+Finally, since it is standard to assume that the distribution of the conceptual-population outcome $O^{cp}_{R}$ also holds for $O^{sc}_{R}$ and $O_{R}$, we can reformulate the SCM in @fig-cj14 into the equivalent form shown in @fig-cj15.
 This reformulation produces a model that applies directly to a sample of comparative data.
@@
 For example, if I have the distribution of the population outcome, and it is a Bernoulli, and even if I do not have it because of S and C, the distributional assumption still holds -->
+
+<!-- 4. segway for next section -->
+In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model.
+These extensions account for judges' biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and even clarify the role of the sample and comparison mechanisms in CJ assessments (see @sec-theory-theoretical_SC).
+However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a.
+Since this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.
 
@@
 
-Comparative judgment model
-:::
-
-<!-- 4. segway for next section -->
-In summary, the [SCM -@fig-cj15_scm] and [DAG -@fig-cj15_dag] extend Thurstone's general form to address several limitations of the BTL model.
-These extensions account for judges' biases (see @sec-theory-theoretical_P1), reflect the hierarchical structure of stimuli and incorporate measurement error in trait estimation and hypothesis testing (see @sec-theory-theoretical_P3), and even clarify the role of the sample and comparison mechanisms in CJ assessments (see @sec-theory-theoretical_SC).
-However, they do not resolve concerns about the assumption of equal dispersions among stimuli discussed in @sec-theory-issue1a.
-Since this concern relates to the statistical assumption underlying the distribution of the discriminal process, we develop a formal statistical model to address it in the next section.
+Comparative judgment model.
+:::
 
@@
 
-Comparative judgment model, SCM, probabilistic and statistical model assuming different discriminal dispersions for the student's traits
+Comparative judgment model. SCM, probabilistic and statistical model assuming different discriminal dispersions for the student's traits
 :::
@@
 Put simply, this mathematical description captures the idea that the stimuli' discriminal processes become an observable outcome only through the lens of judges' perceptions (i.e., their biases).
-For clarity, the square brackets in $D_{R}$ indicate the relevant indices for each trait vector; they do not imply any subsetting of the data.
+For clarity, the square brackets in $D_{R}$ indicate the relevant indices for each trait vector.
 
@@
 The **Declarations** section of this document provides a link to the model code, along with an alternative specification that assumes equal discriminal dispersions.
-We tested both versions of the model with success using `Stan` [@Stan_2020, version 2.26.1].
+Both versions of the model have been tested with success using `Stan` [@Stan_2020, version 2.26.1].
 
@@
 
-[^10]: Also refer to the detailed online document referenced in the Declarations section
+[^10]: Also refer to the detailed online document referenced in the **Declarations** section
 
@@
 Even after acquiring the necessary foundational knowledge, applying this approach to a specific CJ context involves two additional challenges.
-First, it is essential to verify whether the conceptual model offers a *faithful* [^12] representation of the CJ system under study.
+First, it is essential to verify whether the conceptual model achieves sufficient *fidelity* in representing the CJ system under study.
 Second, the statistical translation of the model must be assessed to determine if it can accurately estimate the intended estimands (i.e., parameters) from empirical data.
@@
 While these models may be adequate in some contexts, assuming so without evaluation is unwise.
-Instead, adaptation to the specific context is necessary, with careful attention to assessment design features and causal assumptions.
+Instead, adaptation to the specific context is necessary, with careful attention to assessment design features and their implied causal assumptions.
 As discussed in @sec-discussion_challenges, this approach supports such adaptation by providing a transparent framework for articulating new assumptions and guiding CJ assessment design.
-
-[^12]: Avoid confusing this term with the *faithfulness assumption* [@Neal_2020; @Hernan_et_al_2025] described in the causal inference literature. For further details, see footnote 1.
 

### 2025-07-05 paper2.tex
final review and editing

--- 2025-07-03 _tex/paper2.tex
@@
-% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   1p]{elsarticle}
-\usepackage{xcolor}
+
 \usepackage{amsmath,amssymb}
-\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
+  \usepackage{unicode-math}
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
-\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-\makeatother
-
-\usepackage{longtable,booktabs,array}
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 % Set default figure placement to htbp
+\makeatletter
 \def\fps@figure{htbp}
@@
 
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\usepackage[]{natbib}
-\bibliographystyle{elsarticle-harv}
-
-\usepackage{float}
-\floatplacement{table}{H}
 \makeatletter
@@
 \journal{Psychometrika}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-harv}
 \usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
@@
 This study revisits Thurstone's law of comparative judgment (CJ) by
-addressing two central issues in the CJ literature. First, it critiques
-the widespread reliance on Thurstone's Case V assumptions and, by
-extension, the Bradley-Terry-Luce (BTL) model in the analysis of CJ
-data. Specifically, this study argues that while the assumptions of
-equal discriminal dispersions and zero correlation between stimuli
-simplify the trait measurement model, they may fail to capture the
-complexity of some traits or account for heterogeneous stimuli,
-potentially leading to unreliable and inaccurate trait estimates.
-Second, the study highlights the apparent disconnect between trait
-measurement and hypothesis testing procedures in CJ applications. It
-contends that while separating these procedures simplifies the analysis
-of CJ data, this practice can also undermine the resulting statistical
-inferences.
-
-To address these issues, this study aims to extend Thurstone's general
-form using a systematic and integrated approach based on causal and
-Bayesian inference methods. This extension integrates Thurstone's core
-theoretical principles alongside key CJ assessment design features. It
-then translates these elements into a probabilistic statistical model
-for analyzing dichotomous CJ data, overcoming the rigid assumptions of
-Case V and the BTL model.
-
-Finally, the study emphasizes the relevance of this extension for
-contemporary empirical CJ research, particularly stressing the need for
-bespoke CJ models tailored to the experiments and data assumptions. It
-also lays the foundation for broader applications, encouraging
-researchers across the social sciences to adopt more robust and
-interpretable methodologies.
+addressing two key issues in the CJ literature. First, it critiques the
+reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce
+(BTL) model in the analysis of CJ data. Specifically, it argues that
+while the assumptions of equal discriminal dispersions and zero
+correlation simplify the trait measurement model, they may fail to
+capture the complexity of some traits or stimuli, potentially leading to
+unreliable and inaccurate trait estimates. Second, the study highlights
+the disconnect between trait measurement and statistical inference in CJ
+applications. It contends that while separating these processes
+simplifies the analysis, this practice can also undermine the resulting
+statistical inferences. To address these issues, this study extends
+Thurstone's general form using a systematic, integrated approach based
+on causal and Bayesian inference methods. This extension integrates CJ's
+core theoretical principles alongside key assessment design features. It
+then translates these elements into a statistical model for analyzing
+dichotomous CJ data. Finally, the study emphasizes the relevance of this
+extension for contemporary empirical CJ research, stressing the need for
+CJ models tailored to the experiments and data assumptions. It also
+encourages researchers across the social sciences to adopt more robust
+and interpretable methodologies.
 \end{abstract}
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_process.png}
-
-}
-
-\subcaption{\label{fig-discriminal_process}Hypothetical discriminal
-processes for two written texts}
+\includegraphics[width=1\textwidth,height=\textheight]{./images/png/discriminal_process.png}
+
+}
+
+\subcaption{\label{fig-discriminal_process}Discriminal processes}
 
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_difference.png}
-
-}
-
-\subcaption{\label{fig-discriminal_difference}Hypothetical discriminal
-difference of two written texts}
+\includegraphics[width=1\textwidth,height=\textheight]{./images/png/discriminal_difference.png}
+
+}
+
+\subcaption{\label{fig-discriminal_difference}Discriminal difference}
 
@@
 modal discriminal processes, five discriminal dispersions, and \(10\)
-correlations--one per comparison. However, a single judge could only
-provide \({5 \choose 2} = 10\) unique comparisons, an ``insufficient''
-data set to estimate the required parameters.
+correlations--one per comparison (see Table~\ref{tbl-thurstone_cases}).
+However, a single judge could only provide \({5 \choose 2} = 10\) unique
+comparisons, an ``insufficient'' data set to estimate the required
+parameters.
 
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/thurstone_cases.png}
+\includegraphics[width=1\textwidth,height=\textheight]{./images/png/thurstone_cases.png}
 
@@
 
-As the table suggest, Thurstone developed Case V with an emphasis on
+As the table suggests, Thurstone developed Case V with an emphasis on
 statistical simplicity, but this simplicity comes at the expense of
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/dispersion.png}
+\includegraphics[width=1\textwidth,height=\textheight]{./images/png/dispersion.png}
 
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/correlation.png}
+\includegraphics[width=1\textwidth,height=\textheight]{./images/png/correlation.png}
 
@@
 Section~\ref{sec-theory-theoretical_SC} presents the
-\emph{sample-comparison model} (hereafter SCM), which integrates the
+\emph{sample-comparison model} (hereafter CSM), which integrates the
 assessment design features and reflects a more realistic setting where
@@
 
-\includegraphics[width=0.77\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}
+\includegraphics[width=0.77\textwidth,height=\textheight]{./images/png/CJ_TM_03.png}
 
@@
 individuals in the population. Thus, we define the matrix as in Equation
-(\ref{eq-mat11}).
-
-\begin{equation}\phantomsection\label{eq-mat11}{
+(\ref{eq-mat11}). \begin{equation}\phantomsection\label{eq-mat11}{
 I = \begin{bmatrix}
@@
 
-\includegraphics[width=0.87\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}
+\includegraphics[width=0.87\textwidth,height=\textheight]{./images/png/CJ_TM_04.png}
 
@@
 (variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent matrices
-with \(n_{J}\) rows and \(n_{K}\) columns
+with \(n_{J}\) rows and \(n_{K}\) columns.
 
@@
 the absence of connecting arrows.
-
-Overall, the CPM extends Thurstone's general form by introducing key
-innovations to address the limitations discussed in
-Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
-These enhancements include accounting for judges' biases and integrating
-hierarchical structural components. Nevertheless, despite its promise of
-enhancing measurement accuracy and precision, the model still depends on
-the unrealistic assumption that researchers have access to data from the
-\emph{conceptual population}. Since this assumption is rarely met in
-practice, the study must consider a more realistic scenario.
 
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}
+\includegraphics[width=0.7\textwidth,height=\textheight]{./images/png/CJ_TM_09.png}
 
@@
 
+Overall, the CPM extends Thurstone's general form by introducing key
+innovations to address the limitations discussed in
+Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
+These enhancements include accounting for judges' biases and integrating
+hierarchical structural components. Nevertheless, despite its promise of
+enhancing measurement accuracy and precision, the model still depends on
+the unrealistic assumption that researchers have access to data from the
+\emph{conceptual population}. Since this assumption is rarely met in
+practice, the study must consider a more realistic scenario.
+
 \subsection{The sample-comparison model
-(SCM)}\label{sec-theory-theoretical_SC}
-
-The SCM presents a more realistic scenario than the CPM. First, it
+(CSM)}\label{sec-theory-theoretical_SC}
+
+The CSM presents a more realistic scenario than the CPM. First, it
 explicitly assumes a data sample consisting of a limited number of
@@
 To incorporate the sampling mechanism and facilitate the interpretation
-of the sample-comparison model, we first define the \emph{data sampling
-process} using the binary vector variables \(S_{I}\), \(S_{J}\),
-\(S_{IA}\), and \(S_{JK}\) as follows:
-\begin{equation}\phantomsection\label{eq-mat21}{
+of the CSM, we first define the \emph{data sampling process} using the
+binary vector variables \(S_{I}\), \(S_{J}\), \(S_{IA}\), and \(S_{JK}\)
+as follows: \begin{equation}\phantomsection\label{eq-mat21}{
 S_{I} = \begin{bmatrix}
@@
 
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}
+\includegraphics[width=1\textwidth,height=\textheight]{./images/png/CJ_TM_14.png}
 
@@
 
-\caption{\label{fig-cj14}Sample-comparison model, final vectorized form}
+\caption{\label{fig-cj14}Sample-comparison model (CSM), final vectorized
+form.}
 
@@
 mechanism} using the binary vector variable \(C\) to facilitate the
-interpretation of the sample-comparison model. Equation (\ref{eq-mat27})
-shows that \(C\) contains \(n\) elements corresponding to the number of
-rows in the \(R\) matrix, with each element \(c_{(i,a,h,b,j,k)}\) being
-a binary value indicating the presence or absence of data rows in \(R\),
-a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in Equation
-(\ref{eq-mat26}). \begin{equation}\phantomsection\label{eq-mat27}{
+interpretation of the CSM. Equation (\ref{eq-mat27}) shows that \(C\)
+contains \(n\) elements corresponding to the number of rows in the \(R\)
+matrix, with each element \(c_{(i,a,h,b,j,k)}\) being a binary value
+indicating the presence or absence of data rows in \(R\), a definition
+similar to that of \(s_{(i,a,h,b,j,k)}\) in Equation (\ref{eq-mat26}).
+\begin{equation}\phantomsection\label{eq-mat27}{
 C = \begin{bmatrix}
@@
 all other variables in the model. This independence implies that the
-conceptual model represented by the DAG applies exclusively to Random
-Allocation Comparative Designs \citep{Bramley_2015}, or Incomplete Block
-Designs \citep{Lawson_2015}, where every repeated judgment has an equal
-probability of being included in the sample.
+conceptual model represented by the DAG applies exclusively to
+\emph{Random Allocation Comparative Designs} \citep{Bramley_2015}, or
+\emph{Incomplete Block Designs} \citep{Lawson_2015}, where every
+repeated judgment has an equal probability of being included in the
+sample.
 
@@
 conceptual-population outcome \(O^{cp}_{R}\) also holds for
-\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the sample-comparison
-model in Figure~\ref{fig-cj14} into the equivalent form shown in
+\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the SCM in
+Figure~\ref{fig-cj14} into the equivalent form shown in
 Figure~\ref{fig-cj15}. This reformulation produces a model that applies
@@
 and the sample and comparison mechanisms, \(S\) and \(C\).
-
-\begin{figure}[H]
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\[
-\begin{aligned}
-  O_{R} & := f_{O}(D_{R}, S, C) \\
-  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
-  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
-  T_{I} & := f_{T}(X_{I}, e_{I}) \\
-  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
-  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
-  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
-  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
-  e_{IA} & \:\bot\:e_{JK} 
-\end{aligned}
-\]
-
-}
-
-\subcaption{\label{fig-cj15_scm}SCM}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}
-
-}
-
-\subcaption{\label{fig-cj15_dag}DAG}
-
-\end{minipage}%
-
-\caption{\label{fig-cj15}Comparative judgment model}
-
-\end{figure}%
 
@@
 in the next section.
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{R} & := f_{O}(D_{R}, S, C) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj15_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.72\textwidth,height=\textheight]{./images/png/CJ_TM_15.png}
+
+}
+
+\subcaption{\label{fig-cj15_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj15}Comparative judgment model.}
+
+\end{figure}%
 
@@
 
-\caption{\label{fig-cj16}Comparative judgment model, SCM, probabilistic
+\caption{\label{fig-cj16}Comparative judgment model. SCM, probabilistic
 and statistical model assuming different discriminal dispersions for the
@@
 perceptions (i.e., their biases). For clarity, the square brackets in
-\(D_{R}\) indicate the relevant indices for each trait vector; they do
-not imply any subsetting of the data.
+\(D_{R}\) indicate the relevant indices for each trait vector.
 
@@
 the model code, along with an alternative specification that assumes
-equal discriminal dispersions. We tested both versions of the model with
-success using \texttt{Stan} \citep[version 2.26.1]{Stan_2020}.
+equal discriminal dispersions. Both versions of the model have been
+tested with success using \texttt{Stan} \citep[version
+2.26.1]{Stan_2020}.
 
@@
 \footnote{Also refer to the detailed online document referenced in the
-  Declarations section}.
+  \textbf{Declarations} section}.
 
@@
 approach to a specific CJ context involves two additional challenges.
-First, it is essential to verify whether the conceptual model offers a
-\emph{faithful} \footnote{Avoid confusing this term with the
-  \emph{faithfulness assumption} \citep{Neal_2020, Hernan_et_al_2025}
-  described in the causal inference literature. For further details, see
-  footnote 1.} representation of the CJ system under study. Second, the
-statistical translation of the model must be assessed to determine if it
-can accurately estimate the intended estimands (i.e., parameters) from
-empirical data. To ensure fidelity, the conceptual and statistical
-models presented here should be treated as starting points, not
-universal solutions for all CJ designs or datasets. While these models
-may be adequate in some contexts, assuming so without evaluation is
-unwise. Instead, adaptation to the specific context is necessary, with
-careful attention to assessment design features and causal assumptions.
-As discussed in Section~\ref{sec-discussion_challenges}, this approach
-supports such adaptation by providing a transparent framework for
-articulating new assumptions and guiding CJ assessment design.
+First, it is essential to verify whether the conceptual model achieves
+sufficient \emph{fidelity} in representing the CJ system under study.
+Second, the statistical translation of the model must be assessed to
+determine if it can accurately estimate the intended estimands (i.e.,
+parameters) from empirical data. To ensure fidelity, the conceptual and
+statistical models presented here should be treated as starting points,
+not universal solutions for all CJ designs or datasets. While these
+models may be adequate in some contexts, assuming so without evaluation
+is unwise. Instead, adaptation to the specific context is necessary,
+with careful attention to assessment design features and their implied
+causal assumptions. As discussed in
+Section~\ref{sec-discussion_challenges}, this approach supports such
+adaptation by providing a transparent framework for articulating new
+assumptions and guiding CJ assessment design.
 
@@
 
-\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{images/png/IEflow.png}
+\includegraphics[width=0.35\textwidth,height=\textheight]{images/png/IEflow.png}
 
@@
 
-\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}
+\includegraphics[width=0.54\textwidth,height=\textheight]{images/png/mdag_bb1.png}
 
@@
 
-\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}
+\includegraphics[width=0.54\textwidth,height=\textheight]{images/png/mdag_bb2.png}
 
@@
 
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}
+\includegraphics[width=0.65\textwidth,height=\textheight]{images/png/mdag_bb3.png}
 
@@
 
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}
+\includegraphics[width=0.65\textwidth,height=\textheight]{images/png/mdag_bb4.png}
 
@@
 
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}
+\includegraphics[width=0.65\textwidth,height=\textheight]{images/png/mdag_bb5.png}
 
@@
 
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}
+\includegraphics[width=0.65\textwidth,height=\textheight]{images/png/mdag_example1.png}
 
@@
 
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1a.png}
+\includegraphics[width=0.65\textwidth,height=\textheight]{images/png/mdag_example1a.png}
 
@@
 
-\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1b.png}
+\includegraphics[width=0.65\textwidth,height=\textheight]{images/png/mdag_example1b.png}
 

### 2025-07-05 _tex/paper2.tex
Built site for gh-pages

@@
 
-\usepackage{float}
-\floatplacement{table}{H}
 \makeatletter
@@
 
-\subcaption{\label{fig-discriminal_process}Hypothetical discriminal
-processes for two written texts}
+\subcaption{\label{fig-discriminal_process}Discriminal processes}
 
@@
 
-\subcaption{\label{fig-discriminal_difference}Hypothetical discriminal
-difference of two written texts}
+\subcaption{\label{fig-discriminal_difference}Discriminal difference}
 
@@
 in the next section.
+
+\begin{figure}[H]
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\[
+\begin{aligned}
+  O_{R} & := f_{O}(D_{R}, S, C) \\
+  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
+  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
+  T_{I} & := f_{T}(X_{I}, e_{I}) \\
+  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
+  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
+  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
+  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
+  e_{IA} & \:\bot\:e_{JK} 
+\end{aligned}
+\]
+
+}
+
+\subcaption{\label{fig-cj15_scm}SCM}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.72\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}
+
+}
+
+\subcaption{\label{fig-cj15_dag}DAG}
+
+\end{minipage}%
+
+\caption{\label{fig-cj15}Comparative judgment model.}
+
+\end{figure}%
 

### 2025-08-13 paper2.qmd
psychological methods, submission

@@
       given: Tine
-      family: van Daal
+      family: Daal
       non-dropping-particle: van
@@
       given: Sven
-      family: De Maeyer
+      family: Maeyer
       non-dropping-particle: De
@@
 This clarity is crucial because the accuracy of estimates and validity of inferences depend on how well the data and inferential goals align with a model's assumptions [@Kohler_et_al_2019].
-Although this alignment remains to be empirically tested for the proposed models, the theory-driven nature of this approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], grounded in both established theory and existing evidence.
+Although this alignment remains to be empirically tested for the proposed statistical model, the theory-driven nature of this approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], grounded in both established theory and existing evidence.
 
@@
 
-**Materials and code availability:** The `CODE LINK` section at the top of the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/) provides access to all materials and code.
+**Materials and code availability:** A previous version of this manuscript, along with the associated materials and code (see the section titled `CODE LINK`), has been made publicly available at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).
 
@@
 **CRediT authorship contribution statement:** *Conceptualization:* J.M.R.E, T.vD., S.DM., and S.G.; *Methodology:* J.M.R.E, T.vD., and S.DM.; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* T.vD.
-and S.DM.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review and editing:* T.vD., S.DM., and S.G.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
+and S.DM.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review and editing:* J.M.R.E., T.vD., S.DM., and S.G.; *Visualization:* J.M.R.E.; *Supervision:* S.G.
 and S.DM.; *Project administration:* S.G.

### 2025-08-13 _tex/paper2.tex
Built site for gh-pages

@@
 \makeatother
-\journal{Psychometrika}
+\journal{Psychological Methods}
 \usepackage{bookmark}
@@
   pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
-  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
+  pdfauthor={Jose Manuel Rivera Espejo; Tine van Daal; Sven De Maeyer; Steven Gillis},
   pdfkeywords={causal inference, directed acyclic graphs, structural
@@
  \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
-\author[1]{Tine van Daal%
+\author[1]{Tine Daal%
 %
@@
  \ead{tine.vandaal@uantwerpen.be} 
-\author[1]{Sven De Maeyer%
+\author[1]{Sven Maeyer%
 %
@@
 a model's assumptions \citep{Kohler_et_al_2019}. Although this alignment
-remains to be empirically tested for the proposed models, the
+remains to be empirically tested for the proposed statistical model, the
 theory-driven nature of this approach provides a solid foundation for
@@
 
-\textbf{Materials and code availability:} The \texttt{CODE\ LINK}
-section at the top of the digital document located at:
-\url{https://jriveraespejo.github.io/paper2_manuscript/} provides access
-to all materials and code.
+\textbf{Materials and code availability:} A previous version of this
+manuscript, along with the associated materials and code (see the
+section titled \texttt{CODE\ LINK}), has been made publicly available
+at: \url{https://jriveraespejo.github.io/paper2_manuscript/}.
 
@@
 \emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
-J.M.R.E.; \emph{Writing - review and editing:} T.vD., S.DM., and S.G.;
-\emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;
-\emph{Project administration:} S.G. and S.DM.; \emph{Funding
+J.M.R.E.; \emph{Writing - review and editing:} J.M.R.E., T.vD., S.DM.,
+and S.G.; \emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and
+S.DM.; \emph{Project administration:} S.G. and S.DM.; \emph{Funding
 acquisition:} S.G. and S.DM.

### 2025-08-13 paper2.qmd
psychological methods, submission

@@
       given: Tine
-      family: Daal
-      non-dropping-particle: van
+      family: van Daal
     orcid: 0000-0001-9398-9775
@@
       given: Sven
-      family: Maeyer
-      non-dropping-particle: De
+      family: De Maeyer
     orcid: 0000-0003-2888-1631

### 2025-08-15 paper2.qmd
Psychological methods, submission

@@
   - structural causal models
-  - bayesian statistical methods
+  - bayesian inference
   - thurstonian model
@@
 It contends that while separating these processes simplifies the analysis, this practice can also undermine the resulting statistical inferences.
+
   To address these issues, this study extends Thurstone's general form using a systematic, integrated approach based on causal and Bayesian inference methods.
@@
 It then translates these elements into a statistical model for analyzing dichotomous CJ data.
+
   Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, stressing the need for CJ models tailored to the experiments and data assumptions.
@@
 <!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
-Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, empirical evidence of judges’ biases remains scarce in the CJ literature [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
+Despite the growing notion that various stimulus-related factors influence judges' perceptions [@vanDaal_et_al_2016; @Lesterhuis_et_al_2018; @Chambers_et_al_2022] and that these influences may not always cancel each other out, empirical evidence of judges' biases remains scarce in the CJ literature [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a].
 This gap likely persists not from a lack of interest or research but because the identification of such biases often depends on ad-hoc detection methods, like 'misfit' statistics, that may not be well-suited for the task [@Kelly_et_al_2022].

### 2025-08-15 _tex/paper2.tex
Built site for gh-pages

@@
   pdfkeywords={causal inference, directed acyclic graphs, structural
-causal models, bayesian statistical methods, thurstonian
-model, comparative judgement, probability, statistical modeling},
+causal models, bayesian inference, thurstonian model, comparative
+judgement, probability, statistical modeling},
   colorlinks=true,
@@
  \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
-\author[1]{Tine Daal%
+\author[1]{Tine van Daal%
 %
@@
  \ead{tine.vandaal@uantwerpen.be} 
-\author[1]{Sven Maeyer%
+\author[1]{Sven De Maeyer%
 %
@@
 simplifies the analysis, this practice can also undermine the resulting
-statistical inferences. To address these issues, this study extends
-Thurstone's general form using a systematic, integrated approach based
-on causal and Bayesian inference methods. This extension integrates CJ's
-core theoretical principles alongside key assessment design features. It
-then translates these elements into a statistical model for analyzing
-dichotomous CJ data. Finally, the study emphasizes the relevance of this
-extension for contemporary empirical CJ research, stressing the need for
-CJ models tailored to the experiments and data assumptions. It also
-encourages researchers across the social sciences to adopt more robust
-and interpretable methodologies.
+statistical inferences.
+
+To address these issues, this study extends Thurstone's general form
+using a systematic, integrated approach based on causal and Bayesian
+inference methods. This extension integrates CJ's core theoretical
+principles alongside key assessment design features. It then translates
+these elements into a statistical model for analyzing dichotomous CJ
+data.
+
+Finally, the study emphasizes the relevance of this extension for
+contemporary empirical CJ research, stressing the need for CJ models
+tailored to the experiments and data assumptions. It also encourages
+researchers across the social sciences to adopt more robust and
+interpretable methodologies.
 \end{abstract}
@@
     causal inference \sep directed acyclic graphs \sep structural causal
-models \sep bayesian statistical methods \sep thurstonian
-model \sep comparative judgement \sep probability \sep 
+models \sep bayesian inference \sep thurstonian model \sep comparative
+judgement \sep probability \sep 
     statistical modeling
